# JavaScript

```{r}
#| include: false
source("../R/setup.R")
source("../R/javascript.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means what it means whoever wrote it, and
everything in this book up to here is as true in JavaScript as it is in R.

The *sentence*, though, is spelled differently here, and this is the one chapter
in the book where that is so. Every other binding writes the same words joined by
the same four operators. JavaScript cannot join them, because `+`, `*`, `|` and
`/` cannot be given a meaning on objects: whatever you put on either side, `a + b`
is a number and `a | b` is a bitwise integer. There is no version of the language
in which `data(gm) + point` builds a plot.

So JavaScript spells the four operators as four words. That is all that changes,
and it is worth reading the two lines side by side before anything else:

```r
data(gm) + bar * bin + x(life) | facet(era)
```

```js
plot(data(gm), layer(bar, bin), x(col.life), across(col.era))
```

Every JavaScript block below was executed to draw the plot beside it, through the
same engine every other chapter uses.

## The comma is `+`

A gog sentence is a list of words, and in JavaScript it is a list of arguments.
`plot()` is what gives the comma its meaning:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log' }), y(col.life), color(col.continent))")
```

That is [A first plot](../first-plot.qmd) exactly: the same words, in the same
order, reading the same way. Position still decides scope ([Encoding
scope](../encoding-scope.qmd)), so a channel written after a mark belongs to that
mark and one written before belongs to the plot. The commas change nothing about
that; they are the `+` signs.

## `layer()` is `*`

In R, `*` binds tighter than `+`, which is how `bar * bin + x(life)` reads as a
binned bar rather than a bar plus a bin. JavaScript has no precedence to lean on,
so the tighter binding becomes nesting:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), layer(bar, mean), x(col.continent), y(col.life))")
```

Nesting is arguably the better of the two. A reader who does not already know
that `*` binds tighter than `+` has to be told; a reader looking at
`layer(bar, mean)` can see it.

Most sentences need no `layer()` at all. Of the 500 sentences in this book, 264
are a flat list of words, and the ones that do use it usually hold two:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), layer(bar, count), x(col.continent), style({ color: 'steelblue' }))")
```

## `across()` and `down()` are `|` and `/`

Faceting in R is `plot | facet(era)`, panels side by side, or `plot / facet(era)`,
panels stacked. The `|` and the `facet()` were always one phrase (neither means
anything without the other), so JavaScript writes the phrase as one word:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log' }), y(col.life), across(col.continent))")
```

`down()` stacks them instead, and writing both crosses the two into a grid, which
is what `plot | facet(a) / facet(b)` does in R.

Because the pair is one word here, there is no separate `facet()` to write. Typing
one from memory says where it went:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x(col.gdp), y(col.life), facet(col.continent))")
```

## Composing two plots: `beside()` and `below()`

R's `|` means two things, and what is on its right decides which: `plot | facet(g)`
splits one plot into panels, `plot_a | plot_b` arranges two plots on a page. With
no operator to overload there is nothing to read, so the second reading is
its own word: `beside()`, and `below()` for `/`:

```{r}
#| echo: false
js_plot("below(plot(data(gapminder_2007), layer(bar, bin), x(col.gdp), theme({ height: 130 })), plot(data(gapminder_2007), point, x(col.gdp), y(col.life)))")
```

The two words are not new coinages: Henderson's *Functional Geometry*
[@henderson1982] built its algebra of pictures out of `beside` and `above`, and
that is the picture language SICP teaches. Both take pages as well as plots, so
`below(top, beside(main, right))` is the marginal plot, exactly as
`top / (main | right)` is in R. What the composed
plots share is decided by the engine and not by the spelling: the same column on
the same axis in two of them is one axis. See [Composition](../composition.qmd).

## A column is written `col.gdp`

R writes a column as a bare name, Julia as a symbol, Python and JavaScript through
a small accessor. The two languages with no bare names give the same answer, and
they give it for the same reason: in this grammar a plain string is how you spell
a *value*, as in `style({ color: "tomato" })`, `title("…")`, `palette("okabe")`.
Without the accessor a column and a value would look identical, and the grammar's
oldest rule (a channel takes a column, never a value) would be invisible in the
one place it most needs saying.

So a string where a column belongs is refused, and told what to write:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x('gdp'), y(col.life))")
```

The message offers both spellings because both are real: one *maps* a column and
earns a legend, the other *sets* one value and earns none ([Setting is not
mapping](../style.qmd)). The mistake in the other direction is refused the same
way:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x(col.gdp), y(col.life), style({ color: col.continent }))")
```

A column whose name is not a JavaScript identifier takes the bracket form
instead, `col["life exp"]`, which is also how you reach a name with a space or a
dot in it. Names in other scripts need no such thing: `col.지역` (*jiyeok*,
region) is a perfectly ordinary identifier.

## A named argument is one trailing object

R writes `x(gdp, scale = "log")`. JavaScript has no syntax for a named argument,
so the named ones gather into a single object at the end while the positional
ones stay where they are:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log', base: 2 }), y(col.life), title('Doublings of income'))")
```

`style()` is the case where every argument is named, so it is written as one
object throughout the book: `style({ color: "tomato", size: 3 })`.

## The table, and what is lost

A table is an object of columns, so a first plot needs nothing installed:

```{r}
#| echo: false
js_plot(
  "plot(data(heights), bar, x(col.person), y(col.cm))",
  setup = "const heights = { person: ['Ada', 'Alan', 'Grace'], cm: [160, 172, 181] };"
)
```

R reads the table's *name* off the expression you wrote, and Python off the
frame you called from. JavaScript can do neither, and the name matters: a layer
resolves its columns against the nearest table by that name ([Bind
once](../data.qmd)). So an unnamed table is given a unique one, which is enough
for two tables in one sentence to stay apart, and `data(df, { name: "notes" })`
is there for when a message should say `notes` rather than `data2`.

## A sub-expression is a value

One thing this spelling gains over R's. A piece of a sentence is an ordinary
JavaScript value, so it can be named, reused, and passed around, and it means the
same thing everywhere it lands:

```{r}
#| echo: false
js_show(
  "const piled = layer(point, bin, stack);\nconsole.log(String(plot(data(gapminder_2007), piled, x(col.life))));"
)
```

That is the sixth of the [nine laws](../design-laws.qmd) holding in a language
that had to be talked into the other eight.

## Reading from a database

Node has no database standard, so `query()` accepts a connection by the shape of
its methods rather than by its type. A connection that can prepare a statement
and return all its rows works:

```javascript
import { DatabaseSync } from "node:sqlite";
import { plot, query, bar, x, y, col, render_svg } from "grammar-of-graphics";

const con = new DatabaseSync("sales.db");
render_svg(plot(query(con, "SELECT status, revenue FROM orders"),
  bar, x(col.status), y(col.revenue)));
```

`node:sqlite` is included with Node from version 22, so that example needs nothing
installed. `better-sqlite3` has the same shape and works the same way.

### Asynchronous drivers are refused

`render_svg()` is synchronous. It returns the SVG rather than a Promise, which
is what lets a plot be written inline in an expression. A driver whose query
returns a Promise cannot be awaited inside it.

So `pg` and `mysql2` do not work with `query()`, and gog says so rather than
letting a Promise reach the engine. The direction it gives is the fix: fetch the
rows yourself, then bind them as an ordinary table.

```javascript
const { rows } = await con.query("SELECT status, revenue FROM orders");
render_svg(plot(data(rows), bar, x(col.status), y(col.revenue)));
```

That works with every driver, asynchronous ones included. `data()` takes the
rows an async driver returns, and the sentence after it is unchanged. This is
the only place in the four bindings where a language's own shape reaches into
the grammar, and it costs one `await`.

## Getting it

```bash
npm install grammar-of-graphics
```

That command works today. The package is on
[npm](https://www.npmjs.com/package/grammar-of-graphics), and the name is spelled
out because npm already had a package called `gog`. The registry also forbids
capital letters in a new name, so this is Julia's `GrammarOfGraphics` written the
way npm writes names.

The engine arrives with it. There are five more packages on npm, one per
platform, each holding the engine compiled for that platform, and the main
package lists all five as optional dependencies. Installing takes the one that
matches your computer and skips the other four. So there is nothing to put on
your `PATH`, nothing to build, and no Rust toolchain to install.

The binding itself is a thin layer over that engine, and it depends on nothing
else:

```js
import { plot, data, point, x, y, col, render_svg } from "grammar-of-graphics";

const gm = { gdp: [1000, 2000, 3000], life: [60, 70, 80] };
console.log(render_svg(plot(data(gm), point, x(col.gdp), y(col.life))));
```

`render_svg()` returns the SVG as a string, which in a browser goes straight into
the DOM and on a server goes straight into a response. There is no viewer and no
raster step, which is the one place JavaScript fits the engine better than any
other language does.
