# JavaScript

```{r}
#| include: false
source("../R/setup.R")
source("../R/javascript.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means the same thing no matter who wrote it.
Everything in this book up to here is as true in JavaScript as it is in R.

The *sentence* is spelled differently here, and the operators are the reason.
Every other binding joins the same words with the same four operators.
JavaScript cannot join words that way, because `+`, `*`, `|` and `/` cannot be
given a meaning on objects. For two objects `a` and `b`, `a + b` joins them into
one piece of text, `a * b` is `NaN`, and `a | b` is the number `0`. There is no
version of the language in which `data(gm) + point` builds a plot.

So JavaScript spells the four operators as six words: `plot()`, `layer()`,
`across()`, `down()`, `beside()` and `below()`. That is the change this chapter
starts from. Read the two lines together, the R first and the JavaScript
under it:

```r
data(gm) + bar * bin + x(life) | facet(era)
```

```js
plot(data(gm), layer(bar, bin), x(col.life), across(col.era))
```

Every JavaScript block with a plot, a message or a printed line under it was run
when the page was built, through the same engine every other chapter uses. The
database and install commands are shown, not run. In your own code, import the
words you use, such as
`import { plot, data, point, x, y, col, render_svg } from "grammar-of-graphics"`.
The blocks on this page leave that line out, and also the line that reads the
table: `const gapminder_2007 = await gog_table("gapminder_2007");`.

JavaScript brings in only the names you list, and none of gog's names is already
a name in the language. So, unlike the R, Python and Julia chapters, this one has
no section on names the language already uses. If a gog word matches a name your
own script uses, such as `path` beside Node.js's `node:path`, import it under
another name with `as`. The sections below take the four operators first. Then
come how a column, a named argument and a table are written, a plot for a web
page, reading from a database, and installing the package.

## The comma is `+`

A gog sentence is a list of words, and in JavaScript it is a list of arguments.
A comma means nothing on its own. Inside `plot()`, the commas join the words,
exactly as `+` joins them in R. A named argument is written differently: it goes
into an object at the end of the call.
[The section on named arguments](#a-named-argument-is-one-trailing-object),
below, sets that out in full.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log' }), y(col.life), color(col.continent))")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

Those are the words of [A first plot](../first-plot.qmd) without its title, in
the same order, reading the same way. Where you write a word still decides its
scope ([Encoding scope](../encoding-scope.qmd)). A channel written after a mark
belongs to that mark, and one written before any mark belongs to the plot. The
commas change nothing about that; they are the `+` signs.

## `layer()` is `*`

The commas replaced `+`. `*` cannot become a comma, because R groups `*` before
`+`. That order is called **precedence**, and it is how `bar * bin + x(life)`
reads as a binned bar rather than a bar plus a bin. JavaScript writes function
calls rather than operators here, so there is no precedence to use. Nesting does
the same work:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), layer(bar, mean), x(col.continent), y(col.life))")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

Nesting is clearer than precedence. A reader who does not know that `*` groups
before `+` has to be told. A reader looking at `layer(bar, mean)` can see it.

Just over half the sentences in this book need no `layer()` at all, because
they are a flat list of words. Where one is needed, it usually holds two words:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), layer(bar, count), x(col.continent), style({ color: 'steelblue' }))")
```

*"Given gapminder 2007: bars derived by count, x is continent."*

The two words are a mark and the transform that derives it. The color here is a
setting, written as an object inside `style()`.

## `across()` and `down()` are `|` and `/`

Two operators are left, `|` and `/`. With `facet()` on their right, both split
one plot into panels. Faceting in R is `plot | facet(era)`, panels side by side,
or `plot / facet(era)`, panels stacked. In R, `facet()` means nothing on its
own: it always follows a `|` or a `/`. JavaScript writes that whole phrase as
one word:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log' }), y(col.life), across(col.continent))")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split into
panel columns by continent."*

`down()` stacks the panels instead. Writing both gives a grid: one panel column
per value of the column in `across()`, and one panel row per value of the column
in `down()`. That is what `plot | facet(a) / facet(b)` does in R.

Because the operator and `facet()` are one word here, there is no separate
`facet()` to write. Writing `facet()`, as the other three languages do, is
refused, and the message names the two words to use instead:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x(col.gdp), y(col.life), facet(col.continent))")
```

## Composing two plots: `beside()` and `below()`

The `|` you have just replaced had a second job in R. What is on its right
decides which job it does: `plot | facet(g)` splits one plot into panels, and
`plot_a | plot_b` arranges two plots on a page. JavaScript cannot overload `|`,
so the two meanings never share a symbol. The second one gets its own word,
`beside()`, and `below()` for `/`:

```{r}
#| echo: false
js_plot("below(plot(data(gapminder_2007), layer(bar, bin), x(col.gdp), theme({ height: 130 })), plot(data(gapminder_2007), point, x(col.gdp), y(col.life)))")
```

*"Given gapminder 2007: bars derived by bin, x is gdp, above points, x is gdp, y
is life."*

The two words come from an algebra of pictures published in 1982. Henderson's
*Functional Geometry* [@henderson1982] combines pictures with `beside` and
`above`. His `above(a, b)` and gog's `below(a, b)` draw the same picture, with
`a` on top and `b` under it. [Composition](../composition.qmd) gives the
history. Both words also take a
page, the result of an earlier `beside()` or `below()`, so one page can hold
another. `below(top, beside(main, right))` is the marginal plot, exactly as
`top / (main | right)` is in R: a histogram above a scatter, and another to its
right. The spelling does not decide what two composed plots share; the engine
does. Plots one above the other align their `x` axes, and plots side by side
align their `y` axes. When both plots put the same column on that aligned axis,
it becomes one axis: one scale, both panels the same length along it, drawn
once.

That marginal plot, `below(top, beside(main, right))`, needs no parentheses
beyond the calls themselves. In R, Python and Julia, this book puts each plot on
a page inside its own parentheses. Without them, precedence can group parts of
two plots together ([Parentheses on a
page](../operators.qmd#parentheses-on-a-page)). JavaScript needs none of those
parentheses. Each plot is already one `plot()` call, and `beside()` and
`below()` take whole plots and pages as their arguments, so the nesting shows
the grouping. This book writes a `theme()` for the whole page as the last
argument of `beside()` or `below()`:

```{r}
#| echo: false
js_plot("beside(plot(data(gapminder_2007), point, x(col.gdp), y(col.life)), plot(data(gapminder_2007), layer(bar, count), x(col.continent)), theme({ height: 260 }))")
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

In R, Python and Julia, a `theme()` for the whole page follows parentheses
around the page, as in `((…) | (…)) + theme(height = 260)`. In JavaScript, the
`theme()` goes inside the call that makes the page, so no extra parentheses are
needed.

## A column is written `col.gdp`

The operators are the largest change JavaScript makes to a sentence. Three
smaller ones are left, and the first is how a column is named. R writes a
column as a bare name and Julia as a symbol. Python and JavaScript write it
through `col`, a small object called an **accessor**: `col.gdp` names the column
`gdp`. Both languages make that choice for the same reason. In this grammar a
plain string is how you spell a *value*, as in `style({ color: "tomato" })`,
`title("…")`, `palette("okabe")`. Without the accessor, a column and a value
would both be plain strings, and gog could not tell which one you meant.

So a string where a column belongs is refused, and the message says what to
write:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x('gdp'), y(col.life))")
```

The message names the column spelling, `x(col.gdp)`. A channel takes a
*column*, and a string is a *value* ([Setting vs mapping](../style.qmd)). The
opposite mistake, a column where a value belongs, is refused the same way:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x(col.gdp), y(col.life), style({ color: col.continent }))")
```

A column name with a space or a dot in it is not a JavaScript identifier, the
kind of name that can follow a dot. Write it in the bracket form instead:
`col["life exp"]`. A name in another writing system needs no brackets.
JavaScript accepts letters from any writing system in an identifier, so
`col.지역` (*region*) works.

## A named argument is one trailing object

That was the first of the smaller differences. The second is how a named
argument is written. R writes `x(gdp, scale = "log")`. JavaScript has no syntax
for a named argument. You put the named ones into a single object at the end,
and the positional ones stay where they are:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log', base: 2 }), y(col.life), title('Doublings of income'))")
```

*"Given gapminder 2007: points, x is gdp on a log scale in base 2, y is life."*

`style()` is the case where every argument is named, so it is written as one
object throughout the book: `style({ color: "tomato", size: 3 })`.

## The table, and its name

The third difference is how a table gets its name. A table is an object of
columns, or an array of rows, one object per row. Neither needs a package
besides gog, and this one is written in place:

```js
const heights = { person: ["Ada", "Alan", "Grace"], cm: [160, 172, 181] };
```

```{r}
#| echo: false
js_plot(
  "plot(data(heights), bar, x(col.person), y(col.cm))",
  setup = "const heights = { person: ['Ada', 'Alan', 'Grace'], cm: [160, 172, 181] };"
)
```

R reads the table's *name* from the expression you wrote, and Python from the
variable you passed. JavaScript can do neither. The name matters, because a
layer looks up its columns in the nearest table, and it finds that table by name
([Data](../data.qmd)). So an unnamed table gets a generated name instead, such
as `data` or `data2`, which is enough to tell two tables in one sentence apart.
Some messages name the table they are about, such as the warning about missing
values. Write `data(df, { name: "notes" })` when you want such a message to say
`notes`.

A column of `Date` values is drawn on a calendar axis. gog reads each `Date` in
local time, the time shown by the clock of the computer that runs your code. So
`new Date(2024, 0, 1, 12)`, noon on January 1, is drawn at 12:00. JavaScript's
`Date` type holds both a date and a date-time, so gog distinguishes the two by
their values. If every value in a column is midnight in local time, as
`new Date(2024, 0, 1)` is, gog reads the column as dates and never gives it clock
ticks. JavaScript reads the text in `new Date("2024-01-01")` as midnight in UTC,
the time every time zone is counted from. On a computer in New York, gog draws
that value at 19:00 on December 31, 2023. On a computer in Seoul it is 09:00 on
January 1: the right day, but not midnight, so the column is read as date-times.
To get midnight in every time zone, write a time after the date:
`new Date("2024-01-01T00:00")` is midnight in local time.

Three more things have a JavaScript spelling. A missing value is `null`,
`undefined` or `NaN`: a row with one in a column the plot maps is left out, and
gog says how many rows it left out. `ordered(values, levels)` declares the order
of a column's categories, which the [Data](../data.qmd) chapter describes. And
the book's own tables come from `gog_table()`, which fetches them over the
network. It returns a Promise, JavaScript's object for a value that arrives
later, so you write `await` before it, as [The book's data](../book-data.qmd)
shows.

## A sub-expression is a value

Writing a sentence as plain function calls makes one law easy to see. A piece
of a sentence is an ordinary JavaScript value, so you can give it a name, use it
again, and pass it to a function. It means the same thing in every sentence that
holds it:

```{r}
#| echo: false
js_show(
  "const piled = layer(point, bin, stack);\nconsole.log(String(plot(data(gapminder_2007), piled, x(col.life))));"
)
```

Printing a plot gives one line, naming its mark and its table, here the
generated name `data`. It does not give the picture. `piled` could go into a
second sentence unchanged, and those three words would mean the same thing
there. That law is Compositional Invariance, the sixth of the
[nine laws](../design-laws.qmd).

## A plot for a web page: `html_block()`

Everything so far has been about writing a sentence. The rest of this chapter
is about using one: on a web page, with a table from a database, and from an
installed package.

`render_svg()` returns a plot as an SVG string: the drawing and nothing else.
For a web page, `html_block()` returns the same drawing with
[the controls under every plot](../first-plot.qmd#the-boxes-under-every-plot).
Some plots also get a copy of the engine compiled for the browser, which redraws
the plot as you turn or brush it. Four kinds of plot need it: a plot in the
cube, a plot on the globe, a network given a viewing angle, and a plot with a
brush. Those are the plots of [Space](../space.qmd),
[Globe](../globe.qmd), [Network](../network.qmd) and
[Selection](../selection.qmd).

```{r}
#| echo: false
js_show(paste0(
  "const cube = plot(data(gapminder_2007), point,\n",
  "                  x(col.gdp), y(col.life), z(col.population));\n",
  "console.log(html_block(cube).split(\"\\n\")[0]);"
))
```

The function returns HTML as a string. Put it in a page, a notebook cell, or a
dashboard panel. The drawing comes first in that string, so a reader sees the
plot before the copy of the engine has finished loading. That copy is compiled
to WebAssembly, and a page whose security policy forbids WebAssembly still shows
the plot. What stops working is everything the engine does. The cube, the globe
and a network given a viewing angle no longer turn, the brush no longer moves,
and those plots show no controls. A flat plot carries no engine, so it keeps its
controls.

R, Python and Julia never call this function. A notebook in those languages asks
the plot object how it would be displayed, and each binding answers: R with
`repr_html`, Python with `_repr_html_`, Julia with a show method. JavaScript has
no such request to answer, so here it is a function you call.

A script run outside a notebook has nowhere to put that HTML, so `show(plot)`
saves it to a file in your computer's folder for temporary files. It returns the
file's path. Unlike Python's `plot.show()`, it opens no browser.

Three more functions are for building a tool with gog, such as a web server,
and never for writing a sentence. `svg_block(svg)` puts an SVG string in centered
HTML that shrinks to fit a web page, without the controls. `find_gog_cli()`
returns the path of the engine the package found
([Installing gog](#installing-gog)). `to_wire(table, name)` returns the table as
gog sends it to the engine, before it becomes JSON; `name` labels its messages.
Julia exports these three as well; R and Python do not.

## Reading from a database

Your table may be stored in a database rather than held in your session. R,
Python and Julia each have a database standard, but Node.js, the program that
runs JavaScript outside a browser, has none. So `query()` accepts a connection
by the methods it has rather than by its type. Any connection will work if it
can prepare a statement and return all its rows:

```javascript
import { DatabaseSync } from "node:sqlite";
import { plot, data, query, bar, x, y, col, render_svg } from "grammar-of-graphics";

const con = new DatabaseSync("sales.db");
render_svg(plot(query(con, "SELECT status, revenue FROM orders"),
  bar, x(col.status), y(col.revenue)));
```

`node:sqlite` is built into Node.js. It works without a flag from Node.js 22.13,
and from 23.4 in the version 23 series, so on those versions that example needs
nothing installed. Node.js prints a warning that the module is experimental, and
the plot is drawn anyway. `better-sqlite3` has the same shape and works the same
way.

### Asynchronous drivers

Not every driver, the package that connects to one kind of database, returns
its rows that way. `render_svg()` is synchronous: it returns
the SVG rather than a Promise. That is what lets a plot be written inline in an
expression. A driver whose query returns a Promise cannot be awaited inside a
synchronous function.

So `pg` and `mysql2` do not work with `query()`, and gog says so rather than
letting a Promise reach the engine. Fetch the rows yourself instead. Both
drivers return an array of rows, one object per row, and `data()` takes that
array directly:

```javascript
const { rows } = await con.query("SELECT status, revenue FROM orders");
render_svg(plot(data(rows), bar, x(col.status), y(col.revenue)));
```

The first line is written for `pg`. With `mysql2` it begins `const [rows] =`
instead. The line after it is the sentence you would write for any table. The
driver also decides each column's type: `pg` returns a count or a `numeric`
column as text, so cast it in the SQL, as in `COUNT(*)::int` or
`revenue::float8`. Everywhere else, a rule of JavaScript changes only how a
sentence is spelled. Here it changes what a sentence can *do*: the query runs
before the sentence, with one `await`.

## Installing gog

Every plot is drawn by the engine, a separate program written in Rust, so the
npm package has to be able to find it.

```bash
npm install grammar-of-graphics
```

That command works. The package is on
[npm](https://www.npmjs.com/package/grammar-of-graphics), and the name is
written in full because npm already had a package called `gog`. The registry also
forbids capital letters in a new name, so this is Julia's `GrammarOfGraphics`
written the way npm writes names.

The engine is installed with it. There are five more packages on npm, one per
platform, each holding the engine compiled for that platform. The main package
lists all five as optional dependencies. The install takes the one that
matches your computer and skips the other four. So there is nothing to put on
your `PATH`, nothing to build, and no Rust toolchain to install. The five cover
macOS on Intel and ARM processors, Linux on Intel, AMD and ARM processors, and
Windows on Intel and AMD processors. On any other computer the install still
succeeds, and the first plot stops with an error that says no engine is
published for that computer.

The binding itself only builds the specification and hands it to that engine,
and it depends on nothing else:

```js
import { plot, data, point, x, y, col, render_svg } from "grammar-of-graphics";

const gm = { gdp: [1000, 2000, 3000], life: [60, 70, 80] };
console.log(render_svg(plot(data(gm), point, x(col.gdp), y(col.life))));
```

The package runs in Node.js and not in a browser, because it starts the engine
as a separate program. On a server, you send the SVG string as the response, and
the page puts it straight into its DOM. Nothing has to
open a viewer window, and nothing has to be turned into pixels first. What the
engine returns is already the finished picture.

To work from a checkout instead, a copy of gog's source repository, build the
engine once. Then import the package by its path, from the repository's top
folder:

```bash
cargo build --release -p gog-cli
```

```js
import { plot, data, point, x, y, col, render_svg } from "./js-pkg/gog/src/index.js";
```

The engine is looked for in four places, in this order: the `GOG_CLI_PATH`
environment variable, the platform package installed beside the main one,
`gog-cli` on your `PATH`, and a `target/release` build in a surrounding
checkout. The search stops at the first place that holds the engine, so an
installed copy normally stops at the platform package.
`GOG_CLI_PATH` is how this page was rendered.

## What comes next

These four chapters are about spelling, and none of them changes what a
sentence means. [The afterword](../afterword.qmd) closes the book with the
ten-day challenge, and the one goal that remains.
