# JavaScript

```{r}
#| include: false
source("../R/setup.R")
source("../R/javascript.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means the same thing no matter who wrote it.
Everything in this book up to here is as true in JavaScript as it is in R.

The *sentence* is spelled differently here, and the operators are the reason.
Every other binding joins the same words with the same four operators.
JavaScript cannot join words that way, because `+`, `*`, `|` and `/` cannot be
given a meaning on objects. `a + b` joins two objects into one piece of text,
`a * b` is `NaN`, and `a | b` is a bitwise integer. There is no version of the
language in which `data(gm) + point` builds a plot.

So JavaScript spells the four operators as four words. That is the change this
chapter starts from. Read the two lines side by side first:

```r
data(gm) + bar * bin + x(life) | facet(era)
```

```js
plot(data(gm), layer(bar, bin), x(col.life), across(col.era))
```

Every JavaScript block with a plot under it was executed, through the same
engine every other chapter uses.

## The comma is `+`

A gog sentence is a list of words, and in JavaScript it is a list of arguments.
The commas mean nothing on their own. Inside `plot()` they join the words,
exactly as `+` joins them in R. A named argument is written differently: it goes
into an object at the end of the call, which [A named argument is one trailing
object](#a-named-argument-is-one-trailing-object) sets out in full.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log' }), y(col.life), color(col.continent))")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

Those are the words of [A first plot](../first-plot.qmd), in the same order,
reading the same way. Position still decides scope ([Encoding
scope](../encoding-scope.qmd)), so a channel written after a mark belongs to
that mark and one written before belongs to the plot. The commas change nothing
about that; they are the `+` signs.

## `layer()` is `*`

In R, `*` binds tighter than `+`, which is how `bar * bin + x(life)` reads as a
binned bar rather than a bar plus a bin. JavaScript has no precedence to use
here, so the tighter binding becomes nesting:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), layer(bar, mean), x(col.continent), y(col.life))")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

Nesting is the clearer of the two. A reader who does not already know
that `*` binds tighter than `+` has to be told; a reader looking at
`layer(bar, mean)` can see it.

Just over half the sentences in this book need no `layer()` at all, because
they are a flat list of words. Where one is needed, it almost always holds two
words:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), layer(bar, count), x(col.continent), style({ color: 'steelblue' }))")
```

*"Given gapminder 2007: bars derived by count, x is continent."*

## `across()` and `down()` are `|` and `/`

Faceting in R is `plot | facet(era)`, panels side by side, or `plot /
facet(era)`, panels stacked. In R, `facet()` means nothing on its own: it
always follows a `|` or a `/`. JavaScript writes that whole phrase as one word:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log' }), y(col.life), across(col.continent))")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split into
panel columns by continent."*

`down()` stacks the panels instead. Writing both gives a grid: one panel column
per value of the first column, and one panel row per value of the second. That
is what `plot | facet(a) / facet(b)` does in R.

Because the pair is one word here, there is no separate `facet()` to write.
Writing `facet()` from memory is refused, and the message names the word to
use instead:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x(col.gdp), y(col.life), facet(col.continent))")
```

## Composing two plots: `beside()` and `below()`

The `|` you have just replaced had a second job in R. What is on its right
decides which job it does: `plot | facet(g)` splits one plot into panels, and
`plot_a | plot_b` arranges two plots on a page. JavaScript has no `|` to
overload, so nothing has to decide between those two meanings. The second one
gets its own word, `beside()`, and `below()` for `/`:

```{r}
#| echo: false
js_plot("below(plot(data(gapminder_2007), layer(bar, bin), x(col.gdp), theme({ height: 130 })), plot(data(gapminder_2007), point, x(col.gdp), y(col.life)))")
```

*"Given gapminder 2007: bars derived by bin, x is gdp, above points, x is gdp, y
is life."*

The two words come from an older algebra. Henderson's *Functional Geometry*
[@henderson1982] combines pictures with `beside` and `above`, and `below()` is
this book's name for `above`. [Composition](../composition.qmd) tells that story
in full. Both words take pages as well as plots, so `below(top, beside(main,
right))` is the marginal plot, exactly as `top / (main | right)` is in R. The
spelling does not decide what two composed plots share; the engine does. The
same column on the aligned axis of both is one axis.

## A column is written `col.gdp`

The operators are the whole of what JavaScript changes about a sentence. Two
smaller differences are left, and the first is how a column is named. R writes a
column as a bare name, Julia as a symbol, Python and JavaScript through a small
accessor. The two languages with no bare names give the same answer, and for the
same reason. In this grammar a plain string is how you spell a *value*, as in
`style({ color: "tomato" })`, `title("…")`, `palette("okabe")`. Without the
accessor a column and a value would look identical. The grammar's oldest rule,
that a channel takes a column and never a value, would then be invisible in the
two languages that most need it stated.

So a string where a column belongs is refused, and the message says what to
write:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x('gdp'), y(col.life))")
```

The message names the spelling to write, because a string here is a *value* and
a channel takes a *column* ([Setting vs mapping](../style.qmd)). The mistake in
the other direction is refused the same way:

```{r}
#| echo: false
js_error("plot(data(gapminder_2007), point, x(col.gdp), y(col.life), style({ color: col.continent }))")
```

A column name with a space or a dot in it is not a JavaScript identifier. Write
it in the bracket form instead: `col["life exp"]`. A name in another writing
system needs no such thing: `col.지역` (*jiyeok*, region) is an ordinary
identifier.

## A named argument is one trailing object

R writes `x(gdp, scale = "log")`. JavaScript has no syntax for a named argument.
The named ones gather into a single object at the end, and the positional ones
stay where they are:

```{r}
#| echo: false
js_plot("plot(data(gapminder_2007), point, x(col.gdp, { scale: 'log', base: 2 }), y(col.life), title('Doublings of income'))")
```

*"Given gapminder 2007: points, x is gdp on a log scale in base 2, y is life."*

`style()` is the case where every argument is named, so it is written as one
object throughout the book: `style({ color: "tomato", size: 3 })`.

## The table, and its name

A table is an object of columns, `{ person: ['Ada', 'Alan', 'Grace'], cm: [160,
172, 181] }`, so a first plot needs nothing installed:

```{r}
#| echo: false
js_plot(
  "plot(data(heights), bar, x(col.person), y(col.cm))",
  setup = "const heights = { person: ['Ada', 'Alan', 'Grace'], cm: [160, 172, 181] };"
)
```

R reads the table's *name* from the expression you wrote, and Python from the
variable you passed. JavaScript can do neither, and the name matters: a layer
resolves its columns against the nearest table by that name
([Data](../data.qmd)). So an unnamed table is given a name of its own, which is
enough to tell two tables in one sentence apart. Write `data(df, { name:
"notes" })` when a message should say `notes` rather than `data2`.

## A sub-expression is a value

Writing a sentence as plain function calls makes one law easy to see. A piece
of a sentence is an ordinary JavaScript value, so you can give it a name, use it
again, and hand it to a function. It means the same thing in every sentence that
holds it:

```{r}
#| echo: false
js_show(
  "const piled = layer(point, bin, stack);\nconsole.log(String(plot(data(gapminder_2007), piled, x(col.life))));"
)
```

Printing a sentence gives one line naming its mark and its table, not the
picture. `piled` could go into a second sentence unchanged, and it would mean
the same three words there. That law is Compositional Invariance, the sixth of
the [nine laws](../design-laws.qmd), and here the language provides it rather
than the binding.

## Reading from a database

Node.js, where JavaScript runs outside a browser, has no database standard, so
`query()` accepts a connection by the methods it has rather than by its type.
A connection that can prepare a statement and return all its rows works:

```javascript
import { DatabaseSync } from "node:sqlite";
import { plot, data, query, bar, x, y, col, render_svg } from "grammar-of-graphics";

const con = new DatabaseSync("sales.db");
render_svg(plot(query(con, "SELECT status, revenue FROM orders"),
  bar, x(col.status), y(col.revenue)));
```

`node:sqlite` is built into Node.js. Version 22 needs a flag to turn it on,
and from version 23.4 it is on already, so that example needs nothing
installed. `better-sqlite3` has the same shape and works the same way.

### Asynchronous drivers are refused

Not every driver answers that way. `render_svg()` is synchronous: it returns
the SVG rather than a Promise, which is what lets a plot be written inline in an
expression. A driver whose query returns a Promise cannot be awaited inside a
synchronous function.

So `pg` and `mysql2` do not work with `query()`, and gog says so rather than
letting a Promise reach the engine. The refusal gives the direction: fetch the
rows yourself, then bind them as an ordinary table.

```javascript
const { rows } = await con.query("SELECT status, revenue FROM orders");
render_svg(plot(data(rows), bar, x(col.status), y(col.revenue)));
```

That works with every driver, asynchronous ones included. `data()` takes the
rows an asynchronous driver returns, and the sentence after it is unchanged.
Everywhere else a language rule changes only how a sentence is spelled. This is
the one place where it changes what a sentence can *do*, and it costs one
`await`.

## Getting it

A plot is drawn by the engine, which is a compiled Rust binary, so the npm
package has to be able to find one.

```bash
npm install grammar-of-graphics
```

That command works today. The package is on
[npm](https://www.npmjs.com/package/grammar-of-graphics), and the name is
spelled out because npm already had a package called `gog`. The registry also
forbids capital letters in a new name, so this is Julia's `GrammarOfGraphics`
written the way npm writes names.

The engine arrives with it. There are five more packages on npm, one per
platform, each holding the engine compiled for that platform, and the main
package lists all five as optional dependencies. The install takes the one that
matches your computer and skips the other four. So there is nothing to put on
your `PATH`, nothing to build, and no Rust toolchain to install.

The binding itself only builds the specification and hands it to that engine,
and it depends on nothing else:

```js
import { plot, data, point, x, y, col, render_svg } from "grammar-of-graphics";

const gm = { gdp: [1000, 2000, 3000], life: [60, 70, 80] };
console.log(render_svg(plot(data(gm), point, x(col.gdp), y(col.life))));
```

`render_svg()` returns the SVG as a string. In a browser that string goes
straight into the DOM, and on a server it is sent as the response. Nothing has
to open a viewer window, and nothing has to be turned into pixels first. What
the engine returns is already the finished thing.

To work from a checkout instead, build the engine once and import the package
from where it sits:

```bash
cargo build --release -p gog-cli
```

```js
import { plot, data, point, x, y, col, render_svg } from "./js-pkg/gog/src/index.js";
```

The engine is looked for in four places, in this order: the `GOG_CLI_PATH`
environment variable, the platform package installed beside this one, `gog-cli`
on your `PATH`, and a `target/release` build in a surrounding checkout. The
second and third are how an installed copy works; `GOG_CLI_PATH` is how this
page was rendered.

## A plot for a web page: `html_block()`

The string `render_svg()` returns is the drawing and nothing else. For a web
page, `html_block()` returns the same drawing with two things added: the line of
controls under the picture, and, for some plots, a copy of the engine compiled
for the browser. That copy redraws the plot as you turn or brush it. Four kinds
of plot need it: a plot in the cube, a plot on the globe, a turned network, and
a plot with a brush. Those are the plots of [Space](../space.qmd),
[Globe](../globe.qmd), [Network](../network.qmd) and
[Selection](../selection.qmd).

```{r}
#| echo: false
js_show(paste0(
  "const cube = plot(data(gapminder_2007), point,\n",
  "                  x(col.gdp), y(col.life), z(col.population));\n",
  "console.log(html_block(cube).split(\"\\n\")[0]);"
))
```

The function returns HTML as a string. Put it in a page, a notebook cell, or a
dashboard panel. The drawing comes first in that string, so a reader sees the
plot before the copy of the engine has finished loading. That copy is compiled
to WebAssembly, and a page whose security policy forbids WebAssembly still shows
the plot. What stops working is the dragging: the cube and the globe no longer
turn, and the brush no longer moves.

R, Python and Julia never call this function. A notebook in those languages asks
the plot object how it would be displayed, and each binding answers: R with
`repr_html`, Python with `_repr_html_`, Julia with a show method. JavaScript has
no such request to answer, so here it is a function you call.

## What comes next

These four chapters are about spelling, and none of them changes what a
sentence means. [The afterword](../afterword.qmd) closes the book with the
ten-day challenge, and the one goal the project has not built yet.
