# Julia

```{r}
#| include: false
source("../R/setup.R")
source("../R/julia.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means the same thing no matter who wrote it.
Everything in this book up to here is as true in Julia as it is in R.

Of the three languages that are not R, Julia asks for the fewest changes to the
sentence you already know. Like R and Python, it lets a package give `+`, `*`,
`|` and `/` new meanings. Like R, it also keeps a column apart from a value in
the syntax itself. So the sentence below differs from the R version only by a
colon in front of each column:

```r
data(gm) + bar * bin + x(life) | facet(era)
```

```julia
data(gm) + bar * bin + x(:life) | facet(:era)
```

The sections below cover how a column is written, how to break a sentence
across lines, and where Julia's precedence differs from R's. Then come the
tables Julia accepts, showing a plot, the nine names gog shares with `Base`,
reading from a database, and installing the package.

Every Julia block on this page that shows a plot, a refusal or printed output
ran when the page was built, through the same engine every other chapter uses.
The other blocks are shown but not run. Each block ran after `gapminder_2007`
was fetched as [The table](#the-table) shows, and after these two lines, which
[Nine names Julia already uses](#nine-names-julia-already-uses) explains:

```julia
using GrammarOfGraphics
using GrammarOfGraphics: count, sum, min, max, range, size, step, stack, map
```

## A column is a symbol

The colon is the main change to the sentence, so this section shows it in a
plot. R writes a column as a bare name, Python as `col.gdp`, and Julia as a
symbol, a name written after a colon, such as `:gdp`:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) + color(:continent)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

That is the sentence from [A first plot](../first-plot.qmd), without its title
and with a colon in front of each column. In this grammar a plain string spells
a *value*, as in `style(color = "tomato")` or `palette("okabe")`. A symbol is
visibly not a string, so the colon does the same work as R's bare name and
Python's `col.` accessor. A channel takes a column and never a value. Because a
symbol and a string look different, both you and the package can see when a
value is written where a column belongs.

So a string where a column belongs is refused, and the message says what to
write:

```{r}
#| echo: false
jl_error("data(gapminder_2007) + point + x(\"gdp\") + y(:life)")
```

The opposite mistake, a column where a value belongs, is refused the same way
and for the same reason ([Setting vs mapping](../style.qmd)):

```{r}
#| echo: false
jl_error("data(gapminder_2007) + point + x(:gdp) + y(:life) + style(color = :continent)")
```

A column name with a space or a dot in it is written `Symbol("life exp")`. A
name in another writing system needs no `Symbol()`: `:지역` (*region*) is an
ordinary symbol. Julia accepts letters from any writing system in a name.

## A sentence over several lines

A long sentence can be broken across lines, and Julia asks for nothing extra.
Like R, it continues a line that ends in `+`:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) +\n    color(:continent) +\n    size(:population)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent, size by population."*

What does not work is the opposite arrangement, where a line *starts* with the
operator. Julia ends the statement at the line break and reads the next line on
its own, so it reports an error about `+` instead of drawing. To write a
sentence that way, put the whole sentence inside parentheses. The Python chapter
writes every multi-line sentence this way ([Python](python.qmd)), and inside
parentheses a line may begin with `+` in Julia too.

## The operators, and their precedence

A sentence can use up to four operators, so it matters which one Julia groups
first. Julia's precedence table is not R's. R puts `|` below `+`; Julia gives
`|` the same precedence as `+`. That difference could change how a sentence
groups, so the table compares four expressions in both languages:

| Written | Julia builds | R builds |
|---|---|---|
| `a + b + c \| d` | `(a + b + c) \| d` | the same |
| `a \| b / c` | `a \| (b / c)` | the same |
| `a * b + c` | `(a * b) + c` | the same |
| `a \| b + c` | `(a \| b) + c` | `a \| (b + c)` |

Only the last row differs: it is the only one with a `+` after the `|`. After a
facet, the difference does not change the plot. R joins the atom to the facet
first, and Julia joins it to the plot directly. Both then draw the same plot as
the sentence that writes the atom before the facet. A composed page can
have a `+` after the `|`, because the plot on the right starts with its own
table and adds atoms to it. So every composed page in this book puts each plot
in its own parentheses, and each plot is complete before `|` joins them. The
first row is safe for a different reason: Julia groups operators of equal
precedence from left to right, so everything left of the `|` is joined first.

Faceting therefore reads exactly as it does in R:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) | facet(:continent)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split into
panel columns by continent."*

A transform groups as it does in R, because `*` groups before `+` in both
languages:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + bar * mean + x(:continent) + y(:life)")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

## The table

Every sentence starts from a table. The simplest table in Julia is a named tuple
of columns, which the language provides, so a first plot needs nothing
installed:

```{r}
#| echo: false
jl_plot("data((person = [\"Ada\", \"Alan\", \"Grace\"], cm = [160.0, 172.0, 181.0]),\n     name = \"heights\") + bar + x(:person) + y(:cm)")
```

The table holds three people and their heights, written directly inside the
call. A `Dict` of columns works too. So does any object that supports
`names(table)` and `table[!, col]`, as a DataFrame does. The package never
imports DataFrames, so if you do not use one, you install nothing extra.

The other plots use `gapminder_2007`, one of the book's shared tables, and
`gapminder_2007 = gog_table("gapminder_2007")` fetches it. `gog_table()` fetches
any of the shared tables by name, as it does in the other three languages
([The book's data](../book-data.qmd)).

Julia has both a `Date` and a `DateTime`. R and Python each have the same pair
of types under their own names. A `Date` column never gets clock ticks, and a
`DateTime` column can, because it stores a time of day
([Scales](../scales.qmd)). The package tells the two apart by type, so nothing
is guessed.

In R, `data()` reads a table's name from the expression you wrote. In Julia,
`data()` receives only the table itself, so an unnamed table gets a generated
name instead: `data`, then `data2`. Some messages name the table they are about,
such as the warning about missing values, so write `data(df, name = "notes")`
when you want that message to say `notes` ([Data](../data.qmd)).

## Showing the plot

The plots on this page appear under their code. In your own session, where a
plot appears depends on where you run it. In a notebook, such as Jupyter or
Pluto, the notebook asks the package for a picture, and the plot is drawn under
the cell. At the REPL, Julia's prompt in a terminal, a plot prints as one line
of text, such as `<gog plot: bar on heights>` for the heights plot in [The
table](#the-table). To see the picture, assign the plot to a variable such as
`p`, write it to a file with `save_svg(p, "plot.svg")`, and open that file in a web
browser. [`render_svg(p)`](../utilities.qmd#render_svg) returns the same drawing
as text.

Julia also exports three functions that R and Python do not. You need them
only to build a tool with gog, such as a web server, and never to write a
sentence. `svg_block(svg)` puts the text from `render_svg()` in centered HTML for
a web page. Given the plot too, `svg_block(svg, p)` adds its controls.
`find_gog_cli()` returns the path of the engine the package found
([Installing gog](#installing-gog)). `to_wire(table, name)` returns the table as
gog sends it to the engine, before it becomes JSON; `name` labels its messages.
JavaScript exports the same three.

## Nine names Julia already uses

Every plot above ran after two lines that its block leaves out, and this section
explains them. Nine of the kernel's words are also names in `Base`: `count`,
`sum`, `min`, `max`, `range`, `size`, `step`, `stack` and `map`. Julia will not
choose between two modules that export the same name, and the rule is the
language's own. Writing one of these nine without an import stops with an
`UndefVarError` and a hint, rather than quietly calling the wrong function. The
fix is one more import line:

```julia
using GrammarOfGraphics
using GrammarOfGraphics: count, sum, min, max, range, size, step, stack, map
```

With them, a bare `sum` is gog's transform, and `Base.sum` still adds numbers:

```{r}
#| echo: false
jl_show("println(sum)\nprintln(Base.sum([1, 2, 3]))")
```

`bin` is not in the import line, because `Base.bin` is not exported, so there is
no second `bin` to choose between.

The import line treats the nine names alike, but your existing code does not
use them equally. `map` and `size` appear in ordinary Julia code constantly.
After the import, a call such as `map(f, xs)` or `size(A)` in that code reaches
gog's word instead, and the error appears on a line you did not change. Write
`Base.map` or `Base.size` there, and the same prefix works for any of the
nine.

R, Python and Julia all have this problem, and each one solves it in its own
way. R's `range`, `sum` and `min` belong to base R, Python's `bin`,
`sum` and `max` are builtins, and Julia's belong to `Base`. A grammar keeps its
own vocabulary; what changes is how each language asks you to say which one you
meant.

## Reading from a database

Your table may be stored in a database rather than in your session. `query()`
runs a SQL query on that database, and it takes a **DBInterface.jl** connection.
DBInterface.jl is Julia's database standard, and SQLite.jl, LibPQ.jl, MySQL.jl
and DuckDB.jl all implement it.

```julia
using DBInterface, SQLite, GrammarOfGraphics

con = SQLite.DB("sales.db")
query(con, "SELECT status, revenue FROM orders") + bar + x(:status) + y(:revenue)
```

`GrammarOfGraphics` does not depend on DBInterface. Every dependency the package
declares is a Julia standard library, and many readers never write SQL, so none
of them should have to install a database package. Instead, the package uses the
DBInterface that your driver loads. If no driver is loaded, drawing the plot
stops with a message that names DBInterface and the line that loads it.

## Installing gog

The Julia package is named `GrammarOfGraphics`. The name is longer than the one
R and Python use, and the registry is the reason. Julia's registry checks a new
name automatically. The name must be at least five letters long and must start
with a capital. A lowercase `gog` is too short and has no capital.

Every plot is drawn by the engine, a separate program written in Rust, so the
Julia package has to be able to find it. Python and JavaScript each publish the
engine built for every common platform, and the install picks the one for your
computer. R builds it during the install, or installs one already built. Julia
fetches it as an **artifact**, which is how a Julia package distributes a
program it needs. `Pkg` downloads the copy of the engine built for your computer
along with the package, so one call to `Pkg.add` installs everything:

```julia
using Pkg
Pkg.add("GrammarOfGraphics")
```

To work from a checkout, a copy of gog's source repository on your own
computer, first build the engine once from the checkout's top folder (this
needs Rust):

```bash
cargo build --release -p gog-cli
```

Then run `Pkg.develop` from the same folder:

```julia
using Pkg
Pkg.develop(path = "jl-pkg/GrammarOfGraphics")

using GrammarOfGraphics
```

The Julia package looks for the engine in six places. In order: the
`GOG_CLI_PATH` environment variable, a build in the checkout the package sits
in, the artifact the install fetched, a copy you placed inside the package
yourself, `gog-cli` on your `PATH`, and last a build found by searching upward
from the working directory. The search stops at the first place that holds the
engine. An installed copy is not inside a checkout, so it uses the artifact.
Inside a checkout, your own build is searched before the artifact on purpose.
The artifact is the engine of the last release, and using it would draw with an
old engine while you are changing the engine's source.

## What comes next

R, Python and Julia write one sentence three ways, and the main difference is
how a column is named. [JavaScript](javascript.qmd) reads differently, because
it cannot give `+ * | /` new meanings, so it spells them as words.
