# Julia

```{r}
#| include: false
source("../R/setup.R")
source("../R/julia.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means the same thing no matter who wrote it.
Everything in this book up to here is as true in Julia as it is in R.

Of the four languages, Julia asks for the fewest changes to the sentence you
already know. Like R and Python, it gives `+`, `*`, `|` and `/` a meaning of
their own. Like R, it also keeps a column apart from a value in the syntax
itself. So a sentence differs from the R version only by a colon in front of
each column:

```r
data(gm) + bar * bin + x(life) | facet(era)
```

```julia
data(gm) + bar * bin + x(:life) | facet(:era)
```

Every Julia block with a plot under it was executed, through the same engine
every other chapter uses.

## A column is a symbol

R writes a column as a bare name, Python through a small accessor, Julia as a
symbol:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) + color(:continent)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

That is the sentence from [A first plot](../first-plot.qmd), with a colon in
front of each column. In this grammar a plain string spells a *value*, as in
`style(color = "tomato")` or `palette("okabe")`. A symbol is visibly not a
string, so the colon does the same work as R's bare name and Python's `col.`
accessor. The grammar's oldest rule is that a channel takes a column and never a value. In
Julia the syntax itself enforces it, so no check has to.

So a string where a column belongs is refused, and told what to write:

```{r}
#| echo: false
jl_error("data(gapminder_2007) + point + x(\"gdp\") + y(:life)")
```

The mistake in the other direction is refused the same way, and for the same
reason ([Setting vs mapping](../style.qmd)):

```{r}
#| echo: false
jl_error("data(gapminder_2007) + point + x(:gdp) + y(:life) + style(color = :continent)")
```

A column name Julia cannot read as an identifier, because it holds a space or a
dot, is written `Symbol("life exp")`. A
name in another writing system needs no such thing: `:지역` (*jiyeok*, region) is
an ordinary symbol, because Julia identifiers are Unicode.

## A sentence over several lines

A long sentence can be broken across lines, and Julia asks for nothing extra.
Like R, it continues a line that ends in `+`:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) +\n    color(:continent) +\n    size(:population)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent, size by population."*

What does not work is the opposite arrangement, where a line *starts* with the
operator. Julia ends the statement at the line break and reads the next line on
its own, so it reports an error about `+` instead of drawing. To write a sentence
that way, wrap the whole of it in parentheses. That is what Python requires for
every multi-line sentence ([Python](python.qmd)), and inside parentheses a line
may begin with `+` in Julia too.

## The operators, and their precedence

Julia's precedence table is not R's. R puts `|` below `+`; Julia puts it in the
*addition* tier, alongside `+`. That could have broken every faceted sentence in
this book, so it was measured rather than assumed:

| Written | Julia builds | R builds |
|---|---|---|
| `a + b + c \| d` | `(a + b + c) \| d` | the same |
| `a \| b / c` | `a \| (b / c)` | the same |
| `a * b + c` | `(a * b) + c` | the same |
| `a \| b + c` | `(a \| b) + c` | `a \| (b + c)` |

Only the last row differs, and it is the row where an atom is written *after* a
`|`. A facet sentence never does that, and R refuses it anyway. A composed page
comes closer, because the plot on the right names its own table and then adds
atoms to it. So every composed page in this book is written in parentheses, and
they close the right-hand plot before the `|` is read. The first row is safe for
a different reason: Julia reads operators of one tier from left to right, so
everything left of the `|` is joined first.

Faceting therefore reads exactly as it does in R:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) | facet(:continent)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split into
panel columns by continent."*

A transform reads the same way, because `*` binds tighter than `+`:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + bar * mean + x(:continent) + y(:life)")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

## Nine names Julia already uses

Nine of the kernel's words are also names in `Base`: `count`, `sum`, `min`,
`max`, `range`, `size`, `step`, `stack` and `map`. Julia will not choose between
two modules that export the same name, and the rule is the language's own.
Writing one of these nine without an import gives an `UndefVarError` with a
hint, rather than the wrong function. That is the safe outcome, and it means one
more import line:

```julia
using GrammarOfGraphics
using GrammarOfGraphics: count, sum, min, max, range, size, step, stack, map
```

Those two lines open every block on this page that drew a plot. With them in
place, a gog sentence builds and `Base.sum` still answers:

```{r}
#| echo: false
jl_show("println(bar * bin)\nprintln(Base.sum([1, 2, 3]))")
```

`bin` is not in the import line, because `Base.bin` is not exported, so there is
no second `bin` to choose between.

The import line treats the nine names alike, but their effect on code you
already wrote is not equal, and `map` is where that shows. `Base.map` appears in
ordinary Julia code constantly, while `Base.step` and `Base.stack` appear far
less often. Code you already wrote that calls `map` now reaches gog's `map`
instead, and the error names a line you did not change. Write `Base.map` there,
and the same prefix works for any of the nine.

Every language meets this problem, and each one asks you to solve it
differently. R's `range`, `sum` and `min` belong to base R, Python's `bin`,
`sum` and `max` are builtins, and Julia's belong to `Base`. A grammar keeps its
own vocabulary; what changes is how each language asks you to say which one you
meant.

## The table

A table is a named tuple of columns, so a first plot needs nothing installed:

```{r}
#| echo: false
jl_plot("data((person = [\"Ada\", \"Alan\", \"Grace\"], cm = [160.0, 172.0, 181.0]),\n     name = \"heights\") + bar + x(:person) + y(:cm)")
```

A `Dict` works too, and so does anything that implements `names` and `[!, col]`,
which is what a DataFrame does. The package never imports DataFrames, so a
reader who does not use one installs nothing extra.
`gog_table("gapminder_2007")` fetches any of the book's shared tables, as it
does in the other three languages ([The book's data](../book-data.qmd)).

Julia has both a `Date` and a `DateTime`, as R and Python do. A `Date` column
gets day ticks, and a `DateTime` column keeps its hours and minutes. The engine
reads the unit off the type, so nothing is guessed.

R reads a table's name off the expression you wrote, and Julia cannot. An
unnamed table gets a generated name instead, such as `data2`. A warning or a
refusal names the table it is about, so write `data(df, name = "notes")` when
you want that message to say `notes` ([Data](../data.qmd)).

## Reading from a database

`query()` takes a **DBInterface.jl** connection. That is Julia's database
standard, and it covers SQLite.jl, LibPQ.jl, MySQL.jl and DuckDB.jl.

```julia
using DBInterface, SQLite, GrammarOfGraphics

con = SQLite.DB("sales.db")
query(con, "SELECT status, revenue FROM orders") + bar + x(:status) + y(:revenue)
```

One detail is specific to this binding. `GrammarOfGraphics` does not depend on
DBInterface. Every dependency the package declares is a Julia standard library,
and a database package would be the first exception. Every reader would install
it, and many never write SQL. So the package finds DBInterface in your session
instead.

In practice this means one extra line: you must write `using DBInterface`
yourself. Without it, `query()` names the missing package and the line that
fixes it, rather than failing later with a message about something else.

## Getting it

The Julia package is named `GrammarOfGraphics`. The name is longer than the one
R and Python use, and the registry is the reason. Julia's registry checks a new
name automatically, and that check asks for at least five letters in capitalized
words. A lowercase `gog` fails on both counts.

Installing takes one command. A plot is drawn by the engine, which is a compiled
Rust binary, and every one of the four packages carries one. Python and
JavaScript ship it, and R builds or bundles it. Julia fetches it as an
**artifact**, which is how a Julia package distributes a binary. `Pkg` downloads
the build for your computer along with the package, so there is nothing else to
install.

```julia
using Pkg
Pkg.add("GrammarOfGraphics")
```

To work from a checkout instead, add the package from the path where it sits:

```julia
using Pkg
Pkg.develop(path = "jl-pkg/GrammarOfGraphics")

using GrammarOfGraphics
```

The engine is looked for in five places, in this order: the `GOG_CLI_PATH`
environment variable, the artifact the install fetched, a copy you placed inside
the package yourself, `gog-cli` on your `PATH`, and a local `target/release` build.
The search stops at the first place that holds the engine. An installed copy
finds the artifact and never looks further. A checkout has no artifact, so the
search runs to the end of the list and uses the build you compiled yourself.

`render_svg(plot)` returns the SVG as a string, and a plot displays itself in a
notebook, because the binding hands the notebook a picture when it asks for
one.

## What comes next

The first three languages write one sentence three ways, differing only in how a
column is named. [JavaScript](javascript.qmd) is the one that reads differently,
because it cannot overload `+ * | /` and spells them as four words.
