# Julia

```{r}
#| include: false
source("../R/setup.R")
source("../R/julia.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means the same thing no matter who wrote it, and
everything in this book up to here is as true in Julia as it is in R.

Of the four languages, Julia is the one that changes least about the sentence
you already know. Like R and Python, it gives `+`, `*`, `|` and `/` a meaning of
their own, and like R it keeps a column apart from a value in the syntax: a
symbol is not a string. So a sentence differs from the R above it by one colon:

```r
data(gm) + bar * bin + x(life) | facet(era)
```

```julia
data(gm) + bar * bin + x(:life) | facet(:era)
```

Every Julia block with a plot under it was executed, through the same engine
every other chapter uses.

## A column is a symbol

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) + color(:continent)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

That is the sentence from [A first plot](../first-plot.qmd), with a colon in
front of each column. In this grammar a plain string spells a *value*, as in
`style(color = "tomato")` or `palette("okabe")`. A symbol is visibly not a
string, so the colon does the same work as R's bare name and Python's `col.`
accessor. The grammar's oldest rule, that a channel takes a column and never a
value, is therefore in the syntax rather than in a check.

So a string where a column belongs is refused, and told what to write:

```{r}
#| echo: false
jl_error("data(gapminder_2007) + point + x(\"gdp\") + y(:life)")
```

The mistake in the other direction is refused the same way, and for the same
reason ([Setting vs mapping](../style.qmd)):

```{r}
#| echo: false
jl_error("data(gapminder_2007) + point + x(:gdp) + y(:life) + style(color = :continent)")
```

A column whose name is not a Julia identifier is written `Symbol("life exp")`. A
name in another writing system needs no such thing: `:지역` (*jiyeok*, region) is
an ordinary symbol, because Julia identifiers are Unicode.

## The operators, and their precedence

Julia's precedence table is not R's. R puts `|` below `+`; Julia puts it in the
*addition* tier, alongside `+`. That could have broken every faceted sentence in
this book, so it was measured rather than assumed:

| Written | Julia builds | R builds |
|---|---|---|
| `a + b + c \| d` | `(a + b + c) \| d` | the same |
| `a \| b / c` | `a \| (b / c)` | the same |
| `a * b + c` | `(a * b) + c` | the same |
| `a \| b + c` | `(a \| b) + c` | `a \| (b + c)` |

Only the last row differs, and it is the shape where something is written
*after* a facet. No sentence in this book has it, and R refuses that shape
anyway, so every sentence this book teaches parses identically in both
languages. The first row is safe because Julia reads operators of one tier from
left to right: everything left of the `|` is joined before the `|` is reached.

Faceting therefore reads exactly as it does in R:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) | facet(:continent)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split into
panel columns by continent."*

And so does a transform, with `*` binding tighter than `+`:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + bar * mean + x(:continent) + y(:life)")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

## Nine names Julia already uses

Nine of the kernel's words are also names in `Base`: `count`, `sum`, `min`,
`max`, `range`, `size`, `step`, `stack` and `map`. Julia will not silently
choose between two modules that export the same name, and every Julia
programmer has met that rule before. Writing one of
them without the import gives an `UndefVarError` with a hint, rather than the
wrong function, which is the good outcome, but it does mean a second import
line:

```{r}
#| echo: false
jl_show("println(bar * bin)\nprintln(Base.sum([1, 2, 3]))")
```

Those two lines open every block on this page:

```julia
using GrammarOfGraphics
using GrammarOfGraphics: count, sum, min, max, range, size, step, stack, map
```

`Base.sum` is still there, as the second line of output shows. `bin` is not in
the import line above, because `Base.bin` is not exported and so there is no
second `bin` to choose between.

The import line treats the nine alike, and one of them costs more than the
others. `map` is the one most likely to break code you already wrote. `Base.map`
is everyday Julia. `Base.step` and `Base.stack` are not. A script that called
`map` before loading this package will stop working, and the error will name a
line that did not change.

Every language meets this problem, and each one asks you to solve it
differently: R's `range`, `sum` and `min` belong to base R, Python's `bin`,
`sum` and `max` are builtins, and Julia's belong to `Base`. A grammar keeps its
own vocabulary; what changes is how each language asks you to say which one you
meant.

## The table

A table is a named tuple of columns, so a first plot needs nothing installed:

```{r}
#| echo: false
jl_plot("data((person = [\"Ada\", \"Alan\", \"Grace\"], cm = [160.0, 172.0, 181.0]),\n     name = \"heights\") + bar + x(:person) + y(:cm)")
```

A `Dict` works too, and so does anything that answers `names` and `[!, col]`,
which is what a DataFrame does. DataFrames is never imported, so a reader who
does not use one installs nothing for it. `gog_table("gapminder_2007")` fetches
any of the book's shared tables, as it does in the other three languages.

Julia has both a `Date` and a `DateTime`, as R and Python do. A `Date` column
gets day ticks, and a `DateTime` column keeps its hours and minutes. The engine
reads the unit off the type, so nothing is guessed.

R reads a table's name off the expression you wrote, and Julia cannot. An
unnamed table gets a generated name instead, such as `data2`. Write `data(df,
name = "notes")` when a message should say `notes` ([Data](../data.qmd)).

## Reading from a database

`query()` takes a **DBInterface.jl** connection. That is Julia's database
standard, and it covers SQLite.jl, LibPQ.jl, MySQL.jl and DuckDB.jl.

```julia
using DBInterface, SQLite, GrammarOfGraphics

con = SQLite.DB("sales.db")
query(con, "SELECT status, revenue FROM orders") + bar + x(:status) + y(:revenue)
```

One detail is specific to this binding. `GrammarOfGraphics` does not depend on
DBInterface. Every dependency the package declares is a Julia standard library,
and a database package would be the first that is not, installed for every
reader although many never write SQL. So the package looks DBInterface up in
your session instead.

The practical consequence is one line: you must write `using DBInterface`
yourself. Without it, `query()` names the missing package and the line that
fixes it, rather than failing later with a message about something else.

## Getting it

The Julia package is named `GrammarOfGraphics`. The name is longer than the one
R and Python use, and the registry is the reason. Julia requires a package name
of at least five letters, in capitalized words, so a lowercase `gog` cannot be
registered.

Installing takes two commands here, and the engine comes first. A plot is drawn
by the engine, which is a compiled Rust binary. The Python and JavaScript
packages carry an engine built for your computer, and the R package carries one
on macOS and Windows. The Julia package carries none, so this is the binding
that asks you to install it:

```bash
cargo install --git https://github.com/psychometrician/gog gog-cli
```

That command needs Rust, which [rustup](https://rustup.rs) installs in one step.
`cargo install` then puts `gog-cli` on your `PATH`, which is one of the places
the package looks. If you started Julia from an editor or a notebook, that
session may not see your `PATH`, so set `ENV["GOG_CLI_PATH"]` to the binary
instead. Then the package:

```julia
using Pkg
Pkg.add("GrammarOfGraphics")
```

The order is advice rather than a rule. The package loads without an engine and
looks for one only when it draws, so the other order works as well. Installing
the engine first means you never meet the message that says the engine is
missing.

To work from a checkout instead, add the package from the path where it sits:

```julia
using Pkg
Pkg.develop(path = "jl-pkg/GrammarOfGraphics")

using GrammarOfGraphics
```

The engine is looked for in four places, in this order: the `GOG_CLI_PATH`
environment variable, the copy inside the installed package, `gog-cli` on your
`PATH`, and a local `target/release` build. The Julia package carries no engine
of its own, so the third and the fourth places are the ones that find it.

`render_svg(plot)` returns the SVG as a string, and a plot shown in a notebook
draws itself, because the binding hands the notebook a picture when it asks for
one.
