# Julia

```{r}
#| include: false
source("../R/setup.R")
source("../R/julia.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means what it means whoever wrote it, and everything
in this book up to here is as true in Julia as it is in R.

Of the four languages, Julia is the one with the least to say for itself, and
that is the interesting fact about it. It is the only binding besides R that can
spell all four assembly operators, and its symbols keep a column apart from a
value the way R's bare names do. So a sentence differs from the R above it by one
colon:

```r
data(gm) + bar * bin + x(life) | facet(era)
```

```julia
data(gm) + bar * bin + x(:life) | facet(:era)
```

Every Julia block below was executed to draw the plot beside it, through the same
engine every other chapter uses.

## A column is a symbol

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) + color(:continent)")
```

That is [A first plot](../first-plot.qmd) with a colon in front of each column.
The colon is doing the work R's bare name does and Python's `col.` accessor has to
be built for: in this grammar a plain string is how you spell a *value*, as in
`style(color = "tomato")` or `palette("okabe")`, and a symbol is visibly not a
string. The grammar's oldest rule, that a channel takes a column and never a
value, is therefore in the syntax rather than in a check.

So a string where a column belongs is refused, and told what to write:

```{r}
#| echo: false
jl_error("data(gapminder_2007) + point + x(\"gdp\") + y(:life)")
```

The mistake in the other direction is refused the same way, and for the same
reason ([Setting is not mapping](../style.qmd)):

```{r}
#| echo: false
jl_error("data(gapminder_2007) + point + x(:gdp) + y(:life) + style(color = :continent)")
```

A column whose name is not a Julia identifier is written `Symbol("life exp")`.
Names in other scripts need no such thing: `:지역` (*jiyeok*, region) is an
ordinary symbol, because Julia identifiers are Unicode.

## The operators, and their precedence

Julia's precedence table is not R's. R puts `|` below `+`; Julia puts it in the
*addition* tier, alongside `+`. That could have broken every faceted sentence in
this book, so it was measured rather than assumed:

| Written | Julia builds | R builds |
|---|---|---|
| `a + b + c \| d` | `(a + b + c) \| d` | the same |
| `a \| b / c` | `a \| (b / c)` | the same |
| `a * b + c` | `(a * b) + c` | the same |
| `a \| b + c` | `(a \| b) + c` | `a \| (b + c)` |

Only the last row differs, and it is the shape where something is written *after*
a facet. None of the 500 sentences in this book has it, and R refuses that shape
anyway, so every sentence the manual teaches parses identically in both
languages. Left associativity is what saves the first row: `|` sitting beside `+`
rather than below it does not matter when everything to its left was written
first.

Faceting therefore reads exactly as it does in R:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + point + x(:gdp, scale = \"log\") + y(:life) | facet(:continent)")
```

And so does a transform, with `*` binding tighter than `+`:

```{r}
#| echo: false
jl_plot("data(gapminder_2007) + bar * mean + x(:continent) + y(:life)")
```

## Ten words need naming twice

This is Julia's one real wrinkle, and every Julia programmer has met it before.
Ten of the fifty-one kernel words are also names in `Base` (`bin`, `count`, `sum`,
`min`, `max`, `range`, `size`, `step`, `stack`, `map`), and Julia will not silently
pick a winner between two modules that export the same name. Using one bare gives an
`UndefVarError` with a hint rather than the wrong function, which is the good
outcome, but it does mean a second import line:

```{r}
#| echo: false
jl_show("println(bar * bin)\nprintln(Base.sum([1, 2, 3]))")
```

The preamble that made that work is the one every snippet on this page runs, and
it is worth seeing in full:

```julia
using GrammarOfGraphics
using GrammarOfGraphics: bin, count, sum, min, max, range, size, step, stack, map
```

`Base.sum` is still there, as the second line of output shows.

The ten are listed as equals, and one is not. `map` is the one most likely to
reach code you already wrote. `Base.map` is everyday Julia. `Base.step` and
`Base.stack` are not. A script that called `map` before loading this package
will stop working, and the error will name a line that did not change.

This is the third language to hit the same wall from a different side: R's
`range`, `sum` and `min` belong to base R, Python's `bin`, `sum` and `max` are
builtins, and Julia's are `Base`. A grammar keeps its own vocabulary; what
changes is how each language asks you to say which one you meant.

## The table

A table is a named tuple of columns, so a first plot needs nothing installed:

```{r}
#| echo: false
jl_plot(
  "data(heights) + bar + x(:person) + y(:cm)",
  setup = "heights = (person = [\"Ada\", \"Alan\", \"Grace\"], cm = [160.0, 172.0, 181.0])"
)
```

A `Dict` works too, and so does anything that answers `names` and `[!, col]`,
which is what a DataFrame does. That last one is duck-typed rather than imported,
so the binding depends on nothing but `Dates`.

Julia and R are also the only two languages here that draw a *date* apart from a
*timestamp*, which is the distinction the engine's time axis wants: a `Date`
column gets day ticks and a `DateTime` column keeps its clock, with no inference
in between.

R reads a table's name off the expression you wrote, and Julia cannot, so an
unnamed table is given a unique one and `data(df, name = "notes")` is there when
a message should say `notes` rather than `data2` ([Bind once](../data.qmd)).

## Reading from a database

`query()` takes a **DBInterface.jl** connection. That is Julia's database
standard, and it covers SQLite.jl, LibPQ.jl, MySQL.jl and DuckDB.jl.

```julia
using DBInterface, SQLite, GrammarOfGraphics

con = SQLite.DB("sales.db")
query(con, "SELECT status, revenue FROM orders") + bar + x(:status) + y(:revenue)
```

One detail is specific to this binding. `GrammarOfGraphics` **does not depend on
DBInterface**. The package declares one dependency, `Dates`, and adding a
database stack for a feature many readers never use would be a poor trade. So
the package looks DBInterface up in your session instead.

The practical consequence is one line: you must write `using DBInterface`
yourself. Without it, `query()` says so and names the fix rather than failing
somewhere further in.

## Getting it

The Julia package is named `GrammarOfGraphics`. The name is longer than the one
R and Python use, and the registry is the reason: Julia asks a package name for
at least five letters and capitalized words, so a lowercase `gog` is not legal
there.

Installing takes two commands here, and the engine comes first. A plot is drawn
by the engine, which is a compiled Rust binary. The R, Python and JavaScript
packages each ship one built for your computer, and the Julia package does not
carry one yet, so this is the binding that asks you to install it:

```bash
cargo install --git https://github.com/psychometrician/gog gog-cli
```

That command needs Rust, which [rustup](https://rustup.rs) installs in one step.
It puts `gog-cli` on your `PATH`, which is where the package looks. If you
started Julia from an editor or a notebook, that session may not see your
`PATH`, so set `ENV["GOG_CLI_PATH"]` to the binary instead. Then the package:

```julia
using Pkg
Pkg.add("GrammarOfGraphics")
```

The order is advice rather than a rule. The package loads without an engine and
looks for one only when it draws, so the other order works as well. Installing
the engine first means you never meet the error.

The proper answer is an artifact, which is how Julia normally distributes a
compiled binary, and it is owed rather than optional. Until it lands, the
examples in this chapter assume you installed the engine yourself.

To work from a checkout instead, add the package from the path where it sits:

```julia
using Pkg
Pkg.develop(path = "jl-pkg/GrammarOfGraphics")

using GrammarOfGraphics
```

The engine is looked for in four places, in this order: the `GOG_CLI_PATH`
environment variable, the copy inside the installed package, `gog-cli` on your
`PATH`, and a local `target/release` build. The second place is the empty one
today, which is why the third and the fourth matter more in Julia than they do in
the other three languages.

`render_svg(plot)` returns the SVG as a string, and a plot shown in a notebook
draws itself, because the binding answers the `image/svg+xml` MIME type.
