# Python

```{r}
#| include: false
source("../R/setup.R")
source("../R/python.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means what it means whoever wrote it, and
everything in this book up to here is as true in Python as it is in R. What
differs is small enough to state on one page, and this chapter is that page.

Every Python block below was executed to draw the plot beside it, through the
same engine every other chapter uses.

## A column is written `col.gdp`

R writes a column as a bare name, Julia as a symbol, Python through a small
accessor:

```{r}
#| echo: false
py_plot("data(gapminder_2007) + point + x(col.gdp, scale='log') + y(col.life) + color(col.continent)")
```

The R form of that sentence is the one from [A first plot](../first-plot.qmd),
with `col.` in front of each column. Those four characters buy something worth
paying for. In R a bare `gdp` cannot be anything except a name; Python has no
bare names, and in this grammar a plain string is how you spell a *value*:
`style(color="tomato")`, `title("…")`, `palette("okabe")`. Without the accessor
a column and a value would look identical, and the grammar's oldest rule (a
channel takes a column, never a value) would be invisible in the one language
that cannot see it.

So a string where a column belongs is refused, and told what to write:

```{r}
#| echo: false
py_error("data(gapminder_2007) + point + x('gdp') + y(col.life)")
```

The message offers both spellings because both are real: one *maps* a column and
earns a legend, the other *sets* one value and earns none ([Setting is not
mapping](../style.qmd)). The mistake in the other direction is refused the same
way:

```{r}
#| echo: false
py_error("data(gapminder_2007) + point + x(col.gdp) + y(col.life) + style(color=col.continent)")
```

A column whose name Python cannot spell as an attribute takes the bracket form
instead, `col["life exp"]`, which is also how you reach a name with a space or a
dot in it.

## A sentence over several lines takes parentheses

R continues a line that ends in `+`. Python ends the statement there, so a
sentence spanning several lines is wrapped in parentheses:

```{r}
#| echo: false
py_plot("(data(gapminder_2007) + point + x(col.gdp, scale='log') + y(col.life)\n + color(col.continent) + size(col.population)\n + title('Wealth and life expectancy, 2007'))")
```

That is Python's rule about lines rather than anything of GOG's, and together
with the accessor it accounts for almost every difference you will type.

## The table

A table is a dict of columns, so a first plot needs nothing installed at all:

```{r}
#| echo: false
py_plot("data({'level': ['Low', 'Medium', 'High'], 'count': [30.0, 20.0, 10.0]},\n     name='small') + bar + x(col.level) + y(col.count)")
```

Anything carrying `.columns` and `df[name]` is also a table, which covers pandas
and polars. They are duck-typed rather than imported, so neither is a dependency
of the binding, and the ordinary way in is the one you already use:

```{r}
#| echo: false
py_show("import pandas as pd\n\ngapminder_2007 = pd.read_csv(GAPMINDER_CSV)\nprint(gapminder_2007.shape, list(gapminder_2007.columns))")
```

Every plot on this page reads that frame. Types cross as you would expect:
numbers become positions, text becomes categories, `datetime64` and Python's own
`date` and `datetime` become a calendar axis, and `NaN` or `None` is a missing
value the engine drops from the rows it cannot place, saying how many. A pandas
`Categorical` carries its declared order across, which is the counterpart of R's
`factor(x, levels = …)`; a plain dict of strings has nowhere to put that, so
there the categories fall back to the order the rows came in, which
[Data](../data.qmd) describes.

## Computation is pandas' job, not the grammar's

A channel takes a column name and never an expression. There is no
`x(log(gdp))`, and GOG grows no small language for deriving values, because
Python already is one: compute the column, then bind it by name.

```{r}
#| include: false
# One definition, used by the block the reader sees and again by the plot below
# it: each snippet on this page is its own process, so the frame is rebuilt
# rather than carried, and rebuilding it by re-running the shown code is what
# keeps the two from drifting.
europe <- paste0(
  "europe = gapminder_2007[gapminder_2007.continent == 'Europe'].copy()\n",
  "europe['gdp_thousands'] = europe.gdp / 1000\n",
  "europe = europe.sort_values('life', ascending=False)\n"
)
```

```{r}
#| echo: false
py_show(paste0(europe, "print(europe[['country', 'gdp_thousands', 'life']].head(3).to_string(index=False))"))
```

That frame is a table like any other, and the sentence over it is the sentence
you already know:

```{r}
#| echo: false
py_plot("(data(europe) + point + x(col.gdp_thousands) + y(col.life)\n + size(col.population) + style(color='steelblue')\n + x_label('GDP per capita (thousands)')\n + title('Europe, 2007'))", setup = europe)
```

The division stayed in pandas, where the reader can see it, and the plot names a
column that exists. That split is deliberate rather than a missing feature: a
plot is a mapping from a table ([Data](../data.qmd)), so a channel names one of
its columns, and anything you would rather compute is computed before the
sentence starts.

## Showing the plot

In a Jupyter notebook a plot displays itself, as it does in the R console. A
script displays nothing on its own, so say where the plot goes:

```{r}
#| echo: false
py_show("plot = data(gapminder_2007) + point + x(col.gdp) + y(col.life)\nprint(plot)\nprint(plot.save('/tmp/life.svg'))")
```

`save()` writes the SVG, `show()` opens it in a browser, and `render()` returns
it as a string. R's `print()` opens the viewer without being asked; Python waits
to be told, which is the convention each language already has.

## Five names Python already uses

`bin`, `sum`, `min`, `max` and `range` are transforms here and builtins there,
so `from gog import *` puts GOG's over Python's for the rest of the file. The R
package does the same thing to `base::range` and `base::sum`: a grammar keeps
its own vocabulary. Calling one says so rather than failing obscurely:

```{r}
#| echo: false
py_show("try:\n    range(10)\nexcept GogError as error:\n    print(error)")
```

A module that needs both can `import gog` and write `gog.point`, or take the
builtin back with `from builtins import sum`.

## What does not change

The four operators, including how tightly they bind. `*` groups before `+`, `+`
before `|`, and `/` sits beside `*`, which are Python's own precedences as well
as R's, so a derived layer and a crossed facet group exactly as they do in
[Operators](../operators.qmd):

```{r}
#| echo: false
py_plot("(data(gapminder_2007) + bar * mean + x(col.continent) + y(col.life)\n + order(col.life, desc=True)\n + title('Mean life expectancy, highest first'))")
```

The vocabulary does not change either: every mark, transform, channel, setting
and plot-level atom in [the kernel](../kernel-card.qmd) exists in Python under
the same name. Neither do the refusals, because they are the engine's and not
the binding's, word for word:

```{r}
#| echo: false
py_error("data(gapminder_2007) + bar + x(col.continent)")
```

How far apart are the two spellings, then? That is measured rather than argued.
Every sentence in this book, 481 of them, was translated by rule and run through
the Python binding: 379 drew a **byte-identical** plot, 88 earned a
word-identical refusal, and seven differed only in the example inside the
message, where each binding teaches its own syntax. Seven sentences do not carry
over at all, and every one of them is in [R](r.qmd), about R's pipes. Two
rewrites account for nearly all the rest: the accessor, 471 times, and the
parentheses, 326.

## Reading from a database

Python has a database standard, and `query()` speaks it. It also speaks Spark,
because that is where a large share of Python's data actually lives.

The standard is DB-API, also called PEP 249. Anything with a `.cursor()`
method works: the built-in `sqlite3`, DuckDB, `psycopg`, `pyodbc`, Snowflake's
connector, and the Databricks SQL connector.

```python
import sqlite3
from gog import query, bar, x, y, col, render_svg

con = sqlite3.connect("sales.db")
query(con, "SELECT status, revenue FROM orders") + bar + x(col.status) + y(col.revenue)
```

A Spark session works too. Hand `query()` the session itself, and the table
name is whatever your catalog calls it. This is the route to a Databricks table:

```python
query(spark, "SELECT status, revenue FROM main.sales.orders") + bar + x(col.status) + y(col.revenue)
```

In a Databricks notebook the session is already there, under the name `spark`.
`gog` installs from PyPI with its engine included, so `%pip install gog` is the
only setup step.

Neither route adds a dependency to `gog`. You bring the connection; `gog` reads
rows from it.

## Getting it running

A plot is drawn by the engine, which is a compiled Rust binary, so the Python
package has to be able to find one. The released wheel **carries it**: one wheel
per platform, each holding the engine built for that platform, so installing
needs no Rust toolchain.

```bash
pip install gog
```

That command works today. `gog` is on [PyPI](https://pypi.org/project/gog/).
Five wheels cover macOS, Linux and Windows, on both common processor families.
Python was the first of the four bindings to be published, and it is still the
shortest install line of the four. The other three now have registry commands of
their own, listed in the [preface](../index.qmd).

To work from a checkout instead, build the engine once and import the package
from where it sits:

```bash
cargo build --release -p gog-cli
```

```python
import sys
sys.path.insert(0, "py-pkg/gog")

from gog import *
```

The engine is looked for in four places, in this order: the `GOG_CLI_PATH`
environment variable, the copy inside the installed package, `gog-cli` on your
`PATH`, and a local `target/release` build. The first three are how an installed
copy works; the last is how this page was rendered.
