# Python

```{r}
#| include: false
source("../R/setup.R")
source("../R/python.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means the same thing no matter who wrote it.
Everything in this book up to here is as true in Python as it is in R. What
differs is short enough to state in one chapter. The sections below cover how a
column is named, how a sentence is written over several lines, how values are
written, and what counts as a table. Then come computed columns, showing a
plot, and the six names Python already uses. After them comes the operators
section. Python groups the operators in the same order as R, and a page needs
two kinds of parentheses. The vocabulary and the refusals follow, and they do
not change. Reading from a database and installing the
package come last.

Every Python block on this page was run when the page was built, through the
same engine every other chapter uses. Each block runs after `from gog import *`,
and after `gapminder_2007` is read as the section [The table](#the-table)
shows. The blocks leave those lines out. The database and install commands at
the end are shown, not run.

## A column is written `col.gdp`

In Python, gog marks each column with four characters, `col.`, so that a column
is never mistaken for a value. R writes a column as a bare name and Julia as a
symbol. Python writes it through `col`, a small object called an **accessor**:
`col.gdp` names the column `gdp`:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
py_plot("data(gapminder_2007) + point + x(col.gdp, scale='log') + y(col.life) + color(col.continent)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

This is the sentence from [A first plot](../first-plot.qmd), without its title
and with `col.` in front of each column. The four characters are needed because
R and Python pass an argument differently. In R, the call `x(gdp)` receives the
word `gdp` itself, not a value. Python always passes values, so a bare `gdp`
would have to be a variable you defined earlier. In this grammar, a plain string
is how you spell a *value*: `style(color="tomato")`, `title("…")`,
`palette("okabe")`. Without the accessor, a column and a value would both be
plain strings, and gog could not tell which one you meant.

So a string where a column belongs is refused, and the message says what to
write:

```{r}
#| echo: false
py_error("data(gapminder_2007) + point + x('gdp') + y(col.life)")
```

The message names the column spelling, `x(col.gdp)`. Where one name is both a
channel and a setting, such as `color`, the message gives both spellings,
because the two do different things. A channel *maps* a column and the plot
gets a legend for it; a setting *sets* one value and gets none
([Setting vs mapping](../style.qmd)). The opposite mistake, a column where a
value belongs, is refused the same way:

```{r}
#| echo: false
py_error("data(gapminder_2007) + point + x(col.gdp) + y(col.life) + style(color=col.continent)")
```

### Names that need brackets

After `col.`, Python accepts only an **identifier**: a name made of letters,
digits and underscores that does not start with a digit. Any other name is
written in square brackets instead: `col["life exp"]`, `col["gdp.2007"]`,
`col["2007"]`. Inside the brackets the name is a string, but `col` still marks
it as a column. Be careful with a hyphen: Python reads `col.life-exp` as
`col.life` minus a variable named `exp`, so write `col["life-exp"]`.

Python also keeps some words for its own syntax, such as `class`, `from` and
`global`. These **reserved words** need brackets too: `col["class"]`. If you
write `col.class`, Python stops with `SyntaxError: invalid syntax` before any
gog code runs. Unlike gog's messages above, this message comes from Python, so
it cannot tell you to write `col["class"]`. `help("keywords")` prints the full
list.

A name in another writing system needs no brackets: `col.국가` (*country*)
works with the dot. Python accepts letters from any writing system in a name.
The bracket form works for any name.

## A sentence over several lines

The second thing you type differently is about lines rather than names. R
continues a line that ends in `+`. Python ends the statement there, so a
sentence spanning several lines is wrapped in parentheses:

```{r}
#| echo: false
py_plot("(data(gapminder_2007) + point + x(col.gdp, scale='log') + y(col.life)\n + color(col.continent) + size(col.population)\n + title('Wealth and life expectancy, 2007'))")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent, size by population."*

That is Python's rule about lines, not a rule of gog's. Those parentheses and
the accessor are nearly the only things you will type differently. The rest is
as in R, except that values take Python's spelling.

## How values are written

In the other chapters, a code block that is a sentence has a tab for each
language. Code written inside the text has none, so it is written in R, as in
`x(life, limits = c(37.5, 87.5))`. Most of it reads the same in Python once
`col.` is written before each column. The values are what change. R writes
several values inside `c()`, and Python writes them in a list, inside square
brackets. The table gives the Python spelling of each kind of value the book
uses:

| Kind of value | R | Python |
|---|---|---|
| true or false | `TRUE`, `FALSE` | `True`, `False` |
| several numbers | `c(37.5, 87.5)` | `[37.5, 87.5]` |
| several strings, such as colors | `c("tomato", "steelblue")` | `["tomato", "steelblue"]` |
| a color for each category | `c(Asia = "tomato", Europe = "steelblue")` | `{"Asia": "tomato", "Europe": "steelblue"}` |
| a missing value | `NA` | `None` |
| the number *e* | `exp(1)` | `math.e`, after `import math` |
| a declared order | `factor(values, levels = c("Low", "High"))` | `ordered(values, ["Low", "High"])` |

So `x(life, limits = c(37.5, 87.5))` in the text is
`x(col.life, limits=[37.5, 87.5])` in Python. A tuple works too:
`limits=(37.5, 87.5)`. A named argument is written as in R, because Python has
named arguments of its own: `legend = FALSE` is `legend=False`. `ordered()`
declares the order of a column's categories, as R's `factor()` does. The next
section says more about it, and about a pandas column that already carries an
order.

## The table

The simplest table is a dict of columns, so a first plot needs no package
besides gog:

```{r}
#| echo: false
py_plot("data({'level': ['Low', 'Medium', 'High'], 'count': [30.0, 20.0, 10.0]},\n     name='small') + bar + x(col.level) + y(col.count)")
```

*"Given the small table: bars, x is level, y is count."*

Any object with a `.columns` attribute that you can index by column name is also
a table, which covers pandas and polars. The binding imports neither one, so
neither is a dependency. You read a file the way you already do.
`GAPMINDER_CSV` here is the path to the book's `gapminder_2007.csv`, and [The
book's data](../book-data.qmd) says where to get that file:

```{r}
#| echo: false
py_show("import pandas as pd\n\ngapminder_2007 = pd.read_csv(GAPMINDER_CSV)\nprint(gapminder_2007.shape, list(gapminder_2007.columns))")
```

Apart from the small bar plot above, every plot on this page draws from this
table or from a table computed from it.

The column kinds follow the Python types. A column of numbers is read as
continuous, and a column of text as categorical. A column of `datetime64`,
`date` or `datetime` values is drawn on a calendar axis. `NaN` and `None` are
missing values. A row with one in a column the plot maps is left out, and the
engine says how many rows it left out. A pandas `Categorical` keeps its declared
order when the engine reads it. It is the pandas counterpart of R's
`factor(x, levels = …)`. A plain dict of lists declares the same order with
`ordered(values, levels)`, which the [Data](../data.qmd) chapter describes. A
column with a declared order is categorical even when its values are numbers,
such as years.

## Computing a column first

A channel takes a column name and never an expression. That rule is the same in
all four languages. There is no `x(log(col.gdp))`, and gog has no language of
its own for deriving values, because Python already has one. So compute the
column first, then bind it by name.

Dividing a column inside the channel is a natural thing to try. gog refuses
it, and says where the computation belongs:

```{r}
#| echo: false
py_error("data(gapminder_2007) + point + x(col.gdp / 1000) + y(col.life)")
```

```{r}
#| include: false
# One definition, used by the block the reader sees and again by the plot below
# it: each snippet on this page is its own process, so the frame is rebuilt
# rather than carried, and rebuilding it by re-running the shown code is what
# keeps the two from drifting.
europe <- paste0(
  "europe = gapminder_2007[gapminder_2007.continent == 'Europe'].copy()\n",
  "europe['gdp_thousands'] = europe.gdp / 1000\n",
  "europe = europe.sort_values('life', ascending=False)\n"
)
```

```{r}
#| echo: false
py_show(paste0(europe, "print(europe[['country', 'gdp_thousands', 'life']].head(3).to_string(index=False))"))
```

That result is a table like any other, and the sentence that draws from it is
one you already know:

```{r}
#| echo: false
py_plot("(data(europe) + point + x(col.gdp_thousands) + y(col.life)\n + size(col.population) + style(color='steelblue')\n + x_label('GDP per person (thousands)')\n + title('Europe, 2007'))", setup = europe)
```

You did the division in pandas, in a line of your own code that you can check.
The plot then names a column that exists. Keeping the computing and the plotting
apart is deliberate rather than a missing feature. A plot is a mapping from a
table ([Data](../data.qmd)), so a channel names one of its columns. Anything you
want to compute is computed before you write the sentence.

## Showing the plot

The plots above appear on this page because the book drew each one when it was
built. In your own session, where a plot appears depends on how you are running
Python. In a Jupyter notebook, a plot on the last line of a cell is drawn below
that cell, as the R console draws a plot you type. A script displays nothing on
its own, so you ask for the picture yourself:

```{r}
#| echo: false
py_show("plot = data(gapminder_2007) + point + x(col.gdp) + y(col.life)\nprint(plot)\nprint(save_svg(plot, '/tmp/life.svg'))")
```

Printing a plot gives one line, naming its mark and its table. It does not give
the picture. `save_svg()` writes the SVG to a file and returns its path, which
is the second line printed above. `plot.show()` opens the SVG in a browser, and
`render_svg(plot)` returns it as a string. R's `print()` draws the picture, and
Python's does not. Each language is following the convention it already has.

## Six names Python already uses

Six of gog's names are also Python **builtins**, the functions Python provides
without any import: `bin`, `map`, `max`, `min`, `range` and `sum`. After
`from gog import *`, those six names mean gog's atoms, not Python's builtins,
for the rest of the file or notebook. gog's
[R package shares names with R](r.qmd#twenty-names-r-already-uses) as well, and
`max`, `min`, `range` and `sum` are four of them. gog does not rename an atom
to avoid a name the language already uses. If you call `sum`, `min`, `max` or
`map` as you would call Python's builtin, gog refuses the call. So does `range`
with whole numbers outside 0 to 1. The message says the name is gog's, and how
to reach Python's builtin:

```{r}
#| echo: false
py_show("try:\n    range(10)\nexcept GogError as error:\n    print(error)")
```

That is `range`'s message. Its word for this is *shadows*: after the import, the
name reaches gog's atom instead of Python's builtin. `sum`, `min`, `max` and
`map` share a different message, which lists all six names and says how to
reach Python's builtin.

One name, `bin`, gives no message at all. `bin(5)` is also how you ask gog for
five bins, so it returns a transform instead of Python's string `'0b101'`.

If a file needs Python's builtins as well as gog, write `import gog` and then
`gog.point`, or restore one builtin with `from builtins import sum`.

## The operators, and their precedence

The four operators do not change from R to Python, and neither does their
**precedence**, the order in which they group. `*` and `/` group first, then
`+`, then `|`. That order is Python's own, and R's is the same. So `bar * mean`
below becomes one compound mark before any `+` joins it, exactly as in
[Operators](../operators.qmd#what-groups-first):

```{r}
#| echo: false
py_plot("(data(gapminder_2007) + bar * mean + x(col.continent) + y(col.life)\n + order(col.life, desc=True)\n + title('Mean life expectancy, highest first'))")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life, ordered
by life, largest first."*

The same order decides where a page needs parentheses. On a page, Python has
two kinds of parentheses, and they do different jobs. The outer pair, around the
whole sentence, lets it continue on the next line ([A sentence over several
lines](#a-sentence-over-several-lines)). It groups nothing inside the sentence.
So a page inside that pair alone is still refused when `/` joins its plots. `/`
groups before `+`, so it joins the first plot's last atom to the second plot's
`data()`:

```{r}
#| echo: false
py_error("(data(gapminder_2007) + point + x(col.gdp) + y(col.life) /\n  data(gapminder_2007) + bar * count + x(col.continent))")
```

The inner pairs, one around each plot, are the ones that group. Each plot is
then complete before `/` joins it. This book writes both kinds whenever it
composes plots:

```{r}
#| echo: false
py_plot("((data(gapminder_2007) + point + x(col.gdp) + y(col.life)) /\n  (data(gapminder_2007) + bar * count + x(col.continent)))")
```

*"Given gapminder 2007: points, x is gdp, y is life, above bars derived by
count, x is continent."*

With `|` between the plots, gog draws the page without the inner pairs,
because `|` groups after `+`. Write them anyway. The page then means the same
thing with either operator, and the same as in Julia, which groups `|` together
with `+`. A page inside a page needs a pair of its own too, and so does a whole
page before a `theme()` for it. Inside a page, `/` groups before `|`, so
`a | b / c` is read as `a | (b / c)`. Without the pair around a whole page, `+`
groups before `|`, and the `theme()` goes to the second plot ([Parentheses on a
page](../operators.qmd#parentheses-on-a-page)).

## The same vocabulary and the same refusals

Like the operators, the vocabulary does not change: every mark, transform,
channel, setting and plot-level atom in [the kernel card](../kernel-card.qmd)
keeps its name in Python. An engine refusal does not change either. The engine
does not know which binding asked, so it sends you the same words it sends an R
user:

```{r}
#| echo: false
py_error("data(gapminder_2007) + bar + x(col.continent)")
```

The refusals that differ are the binding's own, like the two about the accessor.
Each of those has to teach the spelling of the language you are writing in.

This chapter's examples are only a small sample. On every change to the project,
its automated tests translate the book's sentences into Python and draw them
again. How far the two spellings differ was measured that way, sentence by
sentence. Of the book's 926 sentences, eleven have no Python spelling at all.
Nine are in the [R](r.qmd) chapter and use R's own syntax, such as the pipe. Two
are the GIF exports in the [Play](../play.qmd) chapter. Every one of the rest was
translated by rule and run through the Python binding. Of those, 708 drew the
same plot, byte for byte, and 187 gave the same refusal, word for word. The
other 20 were refused in both languages, in a message worded for each one. Two
rewrites do nearly all of the translating. The accessor is needed in 909
sentences, and the parentheses in 635.

## Reading from a database

Your table may be stored in a database rather than held in your session.
Python's database standard is DB-API, also called PEP 249, and `query()` reads
any connection that follows it. It also reads a Spark session, because many
Python users reach their data through Spark.

Any connection with a `.cursor()` method works: `sqlite3` from Python's
standard library, DuckDB, `psycopg`, `pyodbc`, Snowflake's connector, and the
Databricks SQL connector.

```python
import sqlite3
from gog import query, bar, x, y, col, save_svg

con = sqlite3.connect("sales.db")
plot = query(con, "SELECT status, revenue FROM orders") + bar + x(col.status) + y(col.revenue)
save_svg(plot, "orders.svg")
```

Pass the Spark session itself to `query()`. In the SQL, write the table's full
name as Spark lists it, such as `main.sales.orders`. This is how you read a
Databricks table:

```python
plot = query(spark, "SELECT status, revenue FROM main.sales.orders") + bar + x(col.status) + y(col.revenue)
```

In a Databricks notebook the session is already there, under the name `spark`,
and `%pip install gog` is the only setup step.

Neither way adds a dependency to gog. You open the connection, and gog only
reads rows from it.

## Installing gog

A plot is drawn by the engine, which is a compiled Rust binary, so the Python
package has to be able to find one. The released package carries it. A Python
package is released as **wheels**, ready-built files for one platform each, and
each wheel holds the engine built for its platform. So installing needs no Rust
toolchain.

```bash
pip install gog
```

That command works. gog is on [PyPI](https://pypi.org/project/gog/) as five
wheels. They cover macOS on Intel and ARM processors, Linux on Intel, AMD and
ARM processors, and Windows on Intel and AMD processors. On any other platform,
`pip install gog` stops with an error, because there is no wheel for it. The
other three bindings install from their own registries, and the
[preface](../index.qmd#installing-gog) lists all four commands.

To work from a checkout instead, a copy of gog's source repository, build the
engine once. Then start Python in the repository's top folder and import the
package from there:

```bash
cargo build --release -p gog-cli
```

```python
import sys
sys.path.insert(0, "py-pkg/gog")

from gog import *
```

The engine is looked for in four places, in this order: the `GOG_CLI_PATH`
environment variable, the copy inside the installed package, `gog-cli` on your
`PATH`, and a local `target/release` build. Only a checkout has the fourth,
which also accepts a `target/debug` build. This page used the first: the book
sets `GOG_CLI_PATH` to a local build before any Python runs. If `GOG_CLI_PATH` names no file, gog refuses to draw and says which path it
was given. It does not move on to the next place, because a mistyped path would
then draw with an engine you did not choose.

## What comes next

Python's differences are spelling rather than grammar. [Julia](julia.qmd) has
four differences of its own. A column is a symbol, `|` has a different
precedence, nine names need one extra import line, and the package has a
different name.
