# Python

```{r}
#| include: false
source("../R/setup.R")
source("../R/python.R")
```

The grammar does not change when the language does. One engine reads one
specification, so a sentence means the same thing no matter who wrote it.
Everything in this book up to here is as true in Python as it is in R. What
differs is short enough to state in one chapter.

Every Python block on this page was run when the page was built, through the
same engine every other chapter uses. The database and install commands at the
end are shown, not run.

## A column is written `col.gdp`

R writes a column as a bare name, Julia as a symbol, Python through a small
accessor:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
py_plot("data(gapminder_2007) + point + x(col.gdp, scale='log') + y(col.life) + color(col.continent)")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

The R form of that sentence is the one from [A first plot](../first-plot.qmd),
with `col.` in front of each column. Those four characters are not decoration.
In R a bare `gdp` cannot be anything except a name. Python has no bare names,
and in this grammar a plain string is how you spell a *value*:
`style(color="tomato")`, `title("…")`,
`palette("okabe")`. Without the accessor a column and a value would look
identical, and the rule that a channel takes a column, never a value, would be
invisible in Python.

So a string where a column belongs is refused, and the message says what to
write:

```{r}
#| echo: false
py_error("data(gapminder_2007) + point + x('gdp') + y(col.life)")
```

The message names the column spelling, `x(col.gdp)`. Where one name is both a
channel and a setting, the message gives both spellings, because the two do
different things: a channel *maps* a column and earns a legend, and a setting
*sets* one value and earns none ([Setting vs mapping](../style.qmd)). The mistake in the other direction is
refused the same way:

```{r}
#| echo: false
py_error("data(gapminder_2007) + point + x(col.gdp) + y(col.life) + style(color=col.continent)")
```

A name Python cannot spell after a dot takes the bracket form instead:
`col["life exp"]` for a space, `col["gdp.2007"]` for a dot. A non-Latin column
name needs nothing special either: `col.국가` works, and the bracket form covers
the rest.

## A sentence over several lines takes parentheses

R continues a line that ends in `+`. Python ends the statement there, so a
sentence spanning several lines is wrapped in parentheses:

```{r}
#| echo: false
py_plot("(data(gapminder_2007) + point + x(col.gdp, scale='log') + y(col.life)\n + color(col.continent) + size(col.population)\n + title('Wealth and life expectancy, 2007'))")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent, size by population."*

That is Python's rule about lines, not a rule of gog's. Those parentheses and
the accessor are almost everything you will type differently from the R in this
book.

## The table

A table is a dict of columns, so a first plot needs nothing installed:

```{r}
#| echo: false
py_plot("data({'level': ['Low', 'Medium', 'High'], 'count': [30.0, 20.0, 10.0]},\n     name='small') + bar + x(col.level) + y(col.count)")
```

*"Given the small table: bars, x is level, y is count."*

Any object with a `.columns` attribute that you can index by column name is also
a table, which covers pandas and polars. The binding imports neither one, so
neither is a dependency. You read a file the way you already do.
`GAPMINDER_CSV` here is the path to the book's `gapminder_2007.csv`, and [The
book's data](../book-data.qmd) says where to get that file:

```{r}
#| echo: false
py_show("import pandas as pd\n\ngapminder_2007 = pd.read_csv(GAPMINDER_CSV)\nprint(gapminder_2007.shape, list(gapminder_2007.columns))")
```

Every plot on this page reads that table, apart from the small table written in
place above.

The column kinds follow the Python types. A column of numbers is read as
continuous, and a column of text as categorical. A column of `datetime64`,
`date` or `datetime` values is drawn on a calendar axis. `NaN` and `None` are
missing values: the engine leaves out the whole row and says how many it left
out. A pandas `Categorical` keeps its declared order when the engine reads it,
which is the counterpart of R's `factor(x, levels = …)`. A plain dict of lists
declares the same order with `ordered(values, levels)`, which
[Data](../data.qmd) describes.

## Computing a column first

A channel takes a column name and never an expression. That rule is the same in
all four languages. There is no `x(log(col.gdp))`, and gog has no language of
its own for deriving values, because Python already has one. So compute the
column first, then bind it by name.

Dividing a column inside the channel is the obvious thing to try. gog refuses
it, and says where the computation belongs:

```{r}
#| echo: false
py_error("data(gapminder_2007) + point + x(col.gdp / 1000) + y(col.life)")
```

```{r}
#| include: false
# One definition, used by the block the reader sees and again by the plot below
# it: each snippet on this page is its own process, so the frame is rebuilt
# rather than carried, and rebuilding it by re-running the shown code is what
# keeps the two from drifting.
europe <- paste0(
  "europe = gapminder_2007[gapminder_2007.continent == 'Europe'].copy()\n",
  "europe['gdp_thousands'] = europe.gdp / 1000\n",
  "europe = europe.sort_values('life', ascending=False)\n"
)
```

```{r}
#| echo: false
py_show(paste0(europe, "print(europe[['country', 'gdp_thousands', 'life']].head(3).to_string(index=False))"))
```

That result is a table like any other, and the sentence that draws from it is
one you already know:

```{r}
#| echo: false
py_plot("(data(europe) + point + x(col.gdp_thousands) + y(col.life)\n + size(col.population) + style(color='steelblue')\n + x_label('GDP per capita (thousands)')\n + title('Europe, 2007'))", setup = europe)
```

You did the division in pandas, where you can see it. The plot then names a
column that exists. Keeping the two apart is deliberate rather than a missing
feature. A plot is a mapping from a table ([Data](../data.qmd)), so a channel
names one of its columns. Anything you would rather compute is computed first.

## Showing the plot

In your own session, where a plot appears depends on how you are running Python.
In a Jupyter notebook a plot displays itself, as it does in the R console. A
script displays nothing on its own, so you print the plot or save it:

```{r}
#| echo: false
py_show("plot = data(gapminder_2007) + point + x(col.gdp) + y(col.life)\nprint(plot)\nprint(plot.save('/tmp/life.svg'))")
```

Printing a plot gives one line, naming its mark and its table. It does not give
the picture. `save()` writes the SVG, `show()` opens it in a browser, and
`render()` returns it as a string. R's `print()` opens the viewer on its own,
and Python draws only when you ask it to. Each language is following the
convention it already has.

## Six names Python already uses

`bin`, `map`, `max`, `min`, `range` and `sum` are words in gog's grammar and
also builtins in Python. After `from gog import *`, those six names carry gog's
meaning for the rest of the file. The R package does the same to base R, with
`max`, `min`, `range` and `sum` among them: a grammar keeps its own vocabulary.
Call four of the six and gog says the name is its own, and how to reach
Python's:

```{r}
#| echo: false
py_show("try:\n    range(10)\nexcept GogError as error:\n    print(error)")
```

That is `range`'s message, and `sum`, `min` and `max` read the same way.

The other two say nothing useful. `bin(5)` returns a transform where Python
would have returned a string, so nothing is reported at all. `map` is the harder
one. Python's `map` takes a function and a sequence, and gog's takes at most one
argument. So Python reports the wrong number of arguments, and names nothing you
can act on.

A module that needs both can `import gog` and write `gog.point`, or import the
builtin again with `from builtins import sum`.

## What does not change

The four operators do not change, and neither does how tightly they bind. `*`
groups before `+`, `+` before `|`, and `/` has the same precedence as `*`. Those
are Python's own precedences, and R's are the same, so a derived layer and a
crossed facet are grouped exactly as they are in [Operators](../operators.qmd):

```{r}
#| echo: false
py_plot("(data(gapminder_2007) + bar * mean + x(col.continent) + y(col.life)\n + order(col.life, desc=True)\n + title('Mean life expectancy, highest first'))")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life, ordered
by life, largest first."*

The vocabulary does not change either: every mark, transform, channel, setting
and plot-level atom in [the kernel card](../kernel-card.qmd) exists in Python
under the same name. Neither does an engine refusal. The engine does not know
which binding asked, so its message reaches you in the same words R gets:

```{r}
#| echo: false
py_error("data(gapminder_2007) + bar + x(col.continent)")
```

The refusals that differ are the binding's own, like the two about the accessor,
because each of those teaches its caller's spelling.

How far the two spellings differ was measured, sentence by sentence. Of the
book's 849 sentences, eleven have no Python spelling at all: nine are in
[R](r.qmd), about R-only spellings, and two are the GIF exports in
[Play](../play.qmd). Every one of the rest was translated by rule and run
through the Python binding. Of those, 651 drew a *byte-identical* plot, 167
earned a word-identical refusal, and 20 differed only in the example inside the
message, where each binding teaches its own syntax. Two rewrites do almost all
the translating: the accessor, 832 times, and the parentheses, 573.

## Reading from a database

Python's database standard is DB-API, also called PEP 249, and `query()` reads
any connection that follows it. It reads a Spark session too, because that is
where much of the data a Python user works with is stored.

Anything with a `.cursor()` method works: the built-in `sqlite3`, DuckDB,
`psycopg`, `pyodbc`, Snowflake's connector, and the Databricks SQL connector.

```python
import sqlite3
from gog import query, bar, x, y, col

con = sqlite3.connect("sales.db")
plot = query(con, "SELECT status, revenue FROM orders") + bar + x(col.status) + y(col.revenue)
plot.save("orders.svg")
```

Hand `query()` the Spark session itself, and the table name is whatever your
catalog calls it. This is the route to a Databricks table:

```python
plot = query(spark, "SELECT status, revenue FROM main.sales.orders") + bar + x(col.status) + y(col.revenue)
```

In a Databricks notebook the session is already there, under the name `spark`,
and `%pip install gog` is the only setup step.

Neither route adds a dependency to gog. You open the connection, and gog only
reads rows from it.

## Getting it

A plot is drawn by the engine, which is a compiled Rust binary, so the Python
package has to be able to find one. The released wheel carries it: one wheel per
platform, each holding the engine built for that platform, so installing needs
no Rust toolchain.

```bash
pip install gog
```

That command works today. gog is on [PyPI](https://pypi.org/project/gog/).
There are five wheels. They cover macOS and Linux on both Intel and ARM
processors, and Windows on Intel. The other three bindings install from their own registries, and the
[preface](../index.qmd) lists all four commands.

To work from a checkout instead, build the engine once and import the package
from where it sits:

```bash
cargo build --release -p gog-cli
```

```python
import sys
sys.path.insert(0, "py-pkg/gog")

from gog import *
```

The engine is looked for in four places, in this order: the `GOG_CLI_PATH`
environment variable, the copy inside the installed package, `gog-cli` on your
`PATH`, and a local `target/release` build. The first three are the places an
installed copy looks, and the fourth is for a checkout. This page was rendered
through the first: the book points
`GOG_CLI_PATH` at a local build before any Python runs.

## What comes next

Python's differences are spelling rather than grammar. [Julia](julia.qmd) has
three differences of its own: a symbol for a column, keyword arguments the
language already had, and a package under a different name.
