# R

```{r}
#| include: false
source("../R/setup.R")
```

Every sentence in this book so far was written in R, so this chapter adds
nothing to the grammar. What it adds is the four questions only R raises: how
the two pipes reach `data()`, which of gog's words base R already uses, how a
database connection arrives, and how the package installs. What a table is, and
how one is passed, is in [Data](../data.qmd). A table is an ordinary R data
frame, and needs no package of its own.

Every block marked `{r}` below was run when this page was built, through the
same engine every other chapter uses. The install commands at the end are
shown, not run.

## Pipes

R has two pipes in common use, and R users reasonably want to start a plot
with a table they already have. The native pipe works, and draws exactly the
plot the unpiped form draws.

```{r}
gapminder_2007 |> data() + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

That is not a second way of saying it, and gog does not implement it. By the
time the engine sees anything, the two forms are the *same expression*. You can
watch R rewrite one into the other, because `quote()` shows what the parser
built rather than what you typed:

```{r}
quote(gapminder_2007 |> data() + point + x(gdp) + y(life))
```

Two facts about R combine to make that work, and both have to hold.

The first is precedence. `|>` binds more tightly than `+`, so the pipe finishes
before the sentence begins: it takes the table into `data()` and hands the
finished result to `+`. Derivation with `*` sits between the two, which is
where it needs to be, so a transform still binds to its mark first:

```{r}
gm_all |> data() + bar * count + x(continent)
```

*"Given all the gapminder years: bars derived by count, x is continent."*

The second is *when* it happens. The native pipe is a parser transformation:
writing `df |> data()` produces the code `data(df)` before any of it runs.
That timing decides everything, because `data()` reads the table's name off
the expression you wrote. The pipe leaves it a plain name, so nothing is lost.
Naming the table is the whole job of `data()` (see [Data](../data.qmd)), and it
is what a later layer resolves its bare columns against.

Arguments still land where you would expect, since the piped value fills the
first one:

```{r}
gapminder_2007 |> data(name = "gapminder") + point + x(gdp) + y(life)
```

## What `%>%` does to the table's name

The magrittr pipe, `%>%`, is the one most R code still uses, and here it
behaves differently. It is an ordinary function rather than a parser rule, so
it *evaluates* the left side and passes the value along under its own
placeholder, `.`. The name never reaches `data()`:

```{r}
#| warning: true
library(magrittr)
gapminder_2007 %>% data() + point + x(gdp) + y(life)
```

The plot is right, and for a single table it stays right. What is gone is the
name, so gog says so and tells you how to put it back. Either use the native
pipe, or name the table yourself:

```{r}
gapminder_2007 %>% data(name = "gapminder_2007") + point + x(gdp) + y(life)
```

Only the placeholder raises that warning, so it is worth saying what does not
raise it. Filtering before `data()` is the native pipe's most useful form, and
it raises no warning. The expression is not a plain name, but it is still
distinct, so gog uses it as the table's name.

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data() +
  point + x(gdp) + y(life)
```

The table there is called `subset(gm_all, year == 2007 & continent == Europe)`,
which is ugly and still works. `name =` gives the table a shorter name, and is
worth using once a second table is involved:

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data(name = "europe") +
  point + x(gdp) + y(life)
```

## Why the name is worth a warning

An unnamed table is fine until there are two of them. A second `data()` gives a
layer its own table, and "nearest table wins" is decided *by name*, so two
tables under one name leave the rule nothing to choose between:

```{r}
#| echo: false
peek(actuals)
peek(forecast)
```

```{r}
#| error: true
data(actuals, name = "series") + x(year) + y(sales) + line +
  data(forecast, name = "series") + point
```

Without that refusal the plot would draw the first table twice and report the
second table's columns as misspellings, which blames the reader for something
the binding lost. Two `%>%` pipes in one expression reach that state, because
both tables arrive as `.`.

## What the pipe does not change

The pipe feeds `data()`. It does not replace `+`, and there is no piped form of
the rest of the sentence: marks, channels, and transforms compose with `+` and
`*` as they do everywhere else in this book. gog has no pipe of its own, because
that would be a second spelling for something the language already spells.

This section is about R specifically. Julia's `|>` is a function-application
operator with different behavior, and Python and JavaScript have no pipe of
their own, so the question does not arise in the other three chapters.

## Twenty names base R already uses

Twenty of the package's exports are also names in the packages R attaches for
you, and `library(gog)` warns you about eleven of them. Here are all twenty:

| from | names |
|---|---|
| `stats` | `density` `line` `median` `quantile` `smooth` `step` |
| `graphics` | `box` `layout` `text` `title` |
| `grDevices` | `palette` |
| `utils` | `data` `stack` |
| `base` | `jitter` `max` `mean` `min` `order` `range` `sum` |

Nine of the twenty cannot affect you, which is why `library(gog)` leaves them
out, and the reason is a rule in R itself. A mark that takes no argument is an
*object* rather than a function, and so is a transform that takes no parameter.
When R evaluates a call, it looks for a function of that name and skips any
binding that is not one. So base R still answers, even though `library(gog)` put
its own `mean` in front:

```{r}
is.function(mean)   # gog's `mean` is an object, so a call cannot reach it
mean(1:10)          # and base R's function answers anyway
```

That holds for the functions that take another function as an argument, because
they resolve the name the same way:

```{r}
sapply(gapminder_2007["life"], mean)
```

The other eleven are functions, and a call does reach gog's: `box`, `data`,
`density`, `jitter`, `layout`, `order`, `palette`, `quantile`, `range`,
`stack` and `title`. Qualify the one you want, and both remain available:

| you write | you get | for base R's, write |
|---|---|---|
| `order(population)` | the plot's sort order | `base::order(df$population)` |
| `data(gapminder_2007)` | the plot's table | `utils::data(diamonds)` |
| `density(2)` | the smoothing transform | `stats::density(x)` |
| `jitter(0.5)` | the offset transform | `base::jitter(x)` |
| `quantile(0.9)` | the 90th percentile transform | `stats::quantile(df$gdp)` |
| `range(0.25, 0.75)` | the band transform | `base::range(df$gdp)` |
| `stack(share = TRUE)` | the piling transform | `utils::stack(df)` |
| `palette("soft")` | the plot's colors | `grDevices::palette()` |
| `title("Life expectancy")` | the plot's title | `graphics::title(main = "…")` |
| `box()` | the box-and-whisker mark | `graphics::box()` |

`range` is a function because it takes a band, so the skip rule above does not
save you. That is what writing `range(0.25, 0.75)` costs you. The refusal reads
the argument it was given: a vector where a probability belongs means you wanted
base R's, and the message says so.

```{r}
#| error: true
range(c(3, 1, 4, 1, 5))
```

`order` is the one that breaks the most ordinary code. Sorting a table beside a
plot is ordinary work. gog's `order` returns an atom, not the row positions that
`[` expects. Left alone, the failure arrives far from the line that caused it
and names neither `order` nor gog. Whenever the argument is not a bare name, the
refusal says which function you reached and how to reach the other one:

```{r}
#| error: true
gapminder_2007[order(gapminder_2007$population), ]
```

Qualified, it sorts:

```{r}
ranked <- gapminder_2007[base::order(-gapminder_2007$population), ]
head(ranked[c("country", "population")], 3)
```

This book does that in several other chapters, for exactly this reason.

Three of the eleven refuse nothing, because nothing separates the two readings.
`data(mtcars)` is well formed in both readings. It returns a plot's table and no
complaint, so write `utils::data()` when you mean to load a dataset.
`title("...")` is the same case: a string is what both want. So is `box()`,
which both readings accept with no argument at all. The other eight report the
mistake, because the argument you passed is one gog would refuse. `order` is
the partial case: it catches `order(df$population)` and not `order(pop)`, where
`pop` is a bare name holding a vector.

The names stay as they are. A grammar keeps its own vocabulary, which is Law 3,
and the question is settled. Renaming `order` to `sort_by` would replace a word
every reader knows with one only this package uses. Where another package
defines a name gog also defines, attach gog after that package, so R finds gog's
name first, or write `gog::interval`. Every language meets this problem, and
each one asks you to solve it differently; [the grammar
chapter](../grammar.qmd#names-the-host-language-already-uses) compares them.

## Five spellings that still need a prefix

The nine objects above are safe in ordinary code, and safe inside `sapply()`,
`lapply()`, `tapply()`, `mapply()`, `Map()`, `Reduce()` and `apply()`. Each of
those asks R for a function of the name it was handed, which finds base R's.

Five spellings still fail, for two different reasons. The first is a name that
is passed on twice before anyone looks it up:

```{r}
#| error: true
aggregate(life ~ continent, gapminder_2007, mean)
```

By the time the lookup happens, the name being searched for is `FUN` and not
`mean`, so the message names neither `mean` nor gog. `by()` and `ave(..., FUN =
mean)` fail the same way. Writing `base::mean` fixes all three. The data frame
form of `aggregate()` looks the name up one step earlier, so
`aggregate(gapminder_2007["life"], gapminder_2007["continent"], mean)` works
unchanged.

The second reason is a name used as a value rather than called. `do.call(sum,
list(x))` wants a function and receives an object, so write `do.call("sum",
list(x))` or `do.call(base::sum, list(x))`. And `f <- mean` copies the object,
so `f(x)` reports that it could not find a function called `f`.

That is the whole cost, and it is worth stating as one rule. If you hand one of
those nine words to another function *by name*, and that function hands it on
again, write `base::` in front of it.

## Reading from a database

`query()` takes a **DBI** connection. DBI is R's database standard, so the same
call reaches SQLite, Postgres, MariaDB, BigQuery, and anything with an ODBC
driver.

```r
con <- DBI::dbConnect(RSQLite::SQLite(), "sales.db")

query(con, "SELECT status, revenue FROM orders") + bar + x(status) + y(revenue)
```

DBI and its drivers are **Suggests**, not hard dependencies. A reader who never
writes SQL installs neither. If you do write SQL, install DBI and the one driver
your database needs:

```r
install.packages(c("DBI", "RSQLite"))    # or RPostgres, RMariaDB, odbc, bigrquery
```

gog never opens a connection and never stores a password. You connect the way
you already do, and hand the open connection over.

## Getting it

A plot is drawn by the engine, which is a compiled Rust binary, so the R package
has to be able to find one. An installed copy carries its own. The package's
`configure` script bundles the engine at install time, and refuses to install at
all rather than leave you a package that cannot draw.

```r
install.packages("gog", repos = c("https://psychometrician.r-universe.dev",
                                  "https://cloud.r-project.org"))
```

That command works today. gog is published on
[r-universe](https://psychometrician.r-universe.dev/gog). On macOS and Windows
it builds a binary with the engine inside it, so installing needs no Rust
toolchain.

On Linux the same command installs from source, and the source build compiles
the engine. That needs Rust and a network connection. If the machine does not
have Rust, install it first and then run the command above:

```bash
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh -s -- -y
source "$HOME/.cargo/env"
```

The build runs once, while the package installs.

The `repos` argument names two addresses because r-universe is not CRAN. The
first address is where gog lives. The second is an ordinary CRAN mirror, and it
is there so your other packages still install as usual. gog is not on CRAN, so
both addresses are needed.

There is another route, and it installs straight from the source repository:

```r
remotes::install_github("psychometrician/gog", subdir = "r-pkg/gog")
```

A tarball from GitHub carries the engine's Rust sources, and `configure`
compiles them during the install. So this route needs a Rust toolchain, which
the macOS and Windows builds above do not. It also needs the `remotes` package,
which `install.packages("remotes")` adds if you do not have it. Take it when you
want one particular branch or commit rather than the latest build.

You can also work from a checkout. On that route nothing installs other packages
for you, so `jsonlite` and `pkgload` have to be installed first. Outside base R,
gog needs one package, `jsonlite`. It turns the sentence you write into the JSON
that the engine reads. `install.packages()` and `remotes::install_github()` both
install it along with gog. Loading from a checkout also needs `pkgload`, which
loads a package from a folder instead of from a library. Install both if you do
not have them:

```r
install.packages(c("jsonlite", "pkgload"))
```

If one is missing, the step that needs it stops immediately, and the message
names the missing package.

Then build the engine once and load the package from where it sits:

```bash
cargo build --release -p gog-cli
```

```r
pkgload::load_all("r-pkg/gog")
```

Installing from that same checkout is the better test, because it runs
`configure` and proves the bundling works:

```bash
R CMD INSTALL r-pkg/gog
```

The engine is looked for in five places, in this order: the `GOG_CLI_PATH`
environment variable, a `target/` build in a surrounding checkout, the copy
inside the installed package, `gog-cli` on your `PATH`, and one last search
upward from the working directory. A checkout is searched before the bundled
copy on purpose. Inside a checkout the bundled engine is a leftover build, and
using it would give a stale engine to the person editing the engine.

## What comes next

R's sentence is the one this book has been writing all along.
[Python](python.qmd) writes the same sentence with `col.gdp` where R writes a
bare name, and draws the same picture, byte for byte.

