# R

```{r}
#| include: false
source("../R/setup.R")
```

Every sentence in this book so far was written in R, so this chapter adds
nothing to the grammar. It answers the four questions only R raises: how the
two pipes reach `data()`, which of gog's words base R already uses, how a
database table reaches a plot, and how the package installs.
[Data](../data.qmd) covers what a table is and how you pass one. A table is an
ordinary R data frame, and needs no package of its own.

Every block marked `{r}` below was run when this page was built, through the
same engine every other chapter uses. The database and install commands are
shown, not run.

## Pipes

R has two pipes in common use, and you will want to start a plot with a table
you already have. Which one you write decides whether the table keeps its
name. The native pipe works, and draws exactly the plot the unpiped form
draws.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
gapminder_2007 |> data() + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

That is not a second way of writing the sentence, and gog does nothing to
support it. By the time the engine sees anything, the two forms are the *same
expression*. `quote()` shows what the parser built rather than what you typed,
so you can see R rewrite one form into the other:

```{r}
quote(gapminder_2007 |> data() + point + x(gdp) + y(life))
```

Two facts about R make that work, and both are needed.

The first is precedence. `|>` binds more tightly than `+`, so
`gapminder_2007 |> data()` groups first and `+` sees the finished `data()`
call. Derivation with `*` binds more tightly than `+` and less tightly than
`|>`, so a transform still binds to its mark first:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
gm_all |> data() + bar * count + x(continent)
```

*"Given all the gapminder years: bars derived by count, x is continent."*

The second is *when* it happens. The native pipe is a parser transformation:
writing `df |> data()` produces the code `data(df)` before any of it runs.
That timing is what matters, because `data()` reads the table's name from the
expression you wrote, and the pipe leaves that expression a plain name. Naming
the table is the whole job of `data()` (see [Data](../data.qmd)). A later
layer resolves its bare columns against that name.

The piped table fills the first argument, so any argument you write yourself
follows it:

```{r}
gapminder_2007 |> data(name = "gapminder") + point + x(gdp) + y(life)
```

## What `%>%` does to the table's name

The magrittr pipe, `%>%`, is the one most R code still uses, and here it
behaves differently. It is an ordinary function rather than a parser rule, so
it *evaluates* the left side and passes the value along under its own
placeholder, `.`. The name never reaches `data()`:

```{r}
#| warning: true
library(magrittr)
gapminder_2007 %>% data() + point + x(gdp) + y(life)
```

The plot is right, and with one table it stays right. Only the name is
missing, so gog warns you and says how to supply it. Either use the native
pipe, or name the table yourself:

```{r}
gapminder_2007 %>% data(name = "gapminder_2007") + point + x(gdp) + y(life)
```

Only the placeholder raises that warning. The native pipe raises none, even
when what you pipe in is not a plain name. Filtering before `data()` is its
most useful form. The expression there is not a plain name, but it is still
distinct, so gog uses the whole expression as the table's name.

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data() +
  point + x(gdp) + y(life)
```

The table there is called
`subset(gm_all, year == 2007 & continent == Europe)`, which is long and still
works. `name =` gives the table a shorter name, and is worth using once a
second table is involved:

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data(name = "europe") +
  point + x(gdp) + y(life)
```

## Why the name is worth a warning

An unnamed table is fine until there are two of them. A second `data()` gives
a layer its own table. The "nearest table wins" rule decides *by name*, so two
tables under one name make it impossible to apply:

```{r}
#| echo: false
peek(actuals)
peek(forecast)
```

```{r}
#| error: true
data(actuals, name = "series") + x(year) + y(sales) + line +
  data(forecast, name = "series") + point
```

Without that refusal the plot would draw the first table twice and report the
second table's columns as misspellings. Your spelling would be correct, and
the table it was checked against would be the wrong one. Two `%>%` pipes in
one expression produce exactly this, because both tables arrive as `.`.

## What the pipe does not change

The pipe supplies `data()`. It does not replace `+`, and there is no piped
form of the rest of the sentence. Marks, channels, and transforms compose with
`+` and `*` as they do everywhere else in this book. gog has no pipe of its
own, because that would be a second spelling for something the language
already spells.

The pipe question is R's alone. Julia's `|>` is a function-application
operator and behaves differently, and neither Python nor JavaScript has a
pipe. The other three chapters do not raise it.

## Twenty names base R already uses

That was the first of the four questions. The next is that some of gog's words
are already R's. Twenty of the package's exports are also names in the
packages R attaches for you, and `library(gog)` tells you about eleven of
them. Here are all twenty:

| from | names |
|---|---|
| `stats` | `density` `line` `median` `quantile` `smooth` `step` |
| `graphics` | `box` `layout` `text` `title` |
| `grDevices` | `palette` |
| `utils` | `data` `stack` |
| `base` | `jitter` `max` `mean` `min` `order` `range` `sum` |

Nine of the twenty cannot affect you, and `library(gog)` leaves those nine out
of its message. A rule in R itself is the reason. A mark that takes no
argument is an *object* rather than a function, and so is a transform that
takes no parameter. When R evaluates a call, it looks for a function of that
name and skips any binding that is not one. So a call still reaches base R's
`mean`, even though `library(gog)` attached its own:

```{r}
is.function(mean)   # gog's `mean` is an object, so a call cannot reach it
mean(1:10)          # and base R's function answers anyway
```

The same is true of a function you hand another function to, because it looks
the name up the same way:

```{r}
sapply(gapminder_2007["life"], mean)
```

The other eleven are functions, and a call does reach gog's: `box`, `data`,
`density`, `jitter`, `layout`, `order`, `palette`, `quantile`, `range`,
`stack` and `title`. Qualify the one you want, and both remain available:

| you write | you get | for base R's, write |
|---|---|---|
| `order(population)` | the plot's sort order | `base::order(df$population)` |
| `data(gapminder_2007)` | the plot's table | `utils::data(diamonds)` |
| `density(2)` | the smoothing transform | `stats::density(x)` |
| `jitter(0.5)` | the offset transform | `base::jitter(x)` |
| `layout(exporter, importer)` | the network transform | `graphics::layout(mat)` |
| `quantile(0.9)` | the 90th percentile transform | `stats::quantile(df$gdp)` |
| `range(0.25, 0.75)` | the band transform | `base::range(df$gdp)` |
| `stack(share = TRUE)` | the piling transform | `utils::stack(df)` |
| `palette("soft")` | the plot's colors | `grDevices::palette()` |
| `title("Life expectancy")` | the plot's title | `graphics::title(main = "…")` |
| `box()` | the box-and-whisker mark | `graphics::box()` |

`range` is a function because it takes a band, so R does not skip past it.
Writing `range(0.25, 0.75)` therefore reaches gog's `range` and never base
R's. The refusal reads the argument you passed. A vector where a probability
belongs means you wanted base R's `range`, and the message says so.

```{r}
#| error: true
range(c(3, 1, 4, 1, 5))
```

Of the eleven, `order` is the one that breaks the most code, because sorting a
table beside a plot is everyday work. gog's `order` returns an atom, not the
row positions that `[` expects. Without a refusal, the error would appear far
from the line that caused it, and it would name neither `order` nor gog. So
whenever the argument is not a bare name, the refusal says which function you
reached and how to reach the other one:

```{r}
#| error: true
gapminder_2007[order(gapminder_2007$population), ]
```

Qualified, it sorts:

```{r}
ranked <- gapminder_2007[base::order(-gapminder_2007$population), ]
head(ranked[c("country", "population")], 3)
```

This book does that in several other chapters, for exactly this reason.

Three of the eleven refuse nothing, because the argument does not separate the
two readings. `data(mtcars)` is well formed for both. It returns a plot's
table and raises no complaint. Write `utils::data()` when you mean to load a
dataset. `title("...")` is the same case: a string is what both want. So is
`box()`, which both readings accept with no argument at all. The other eight
report the mistake, because the argument you passed is one gog would refuse.
`order` catches only part of the mistake: it reports `order(df$population)`
and not `order(pop)`, where `pop` is a bare name holding a vector.

The names stay as they are. A grammar keeps its own vocabulary, which is Law
3. Renaming `order` to `sort_by` would replace a word every reader knows with
one only this package uses. Another package may define a name gog also
defines. Attach gog after that package, so R finds gog's name first, or write
the name with `gog::` in front. Every language meets this problem, and each
one asks you to solve it differently; [the grammar
chapter](../grammar.qmd#names-the-host-language-already-uses) compares them.

## Five calls that still need a prefix

Those nine objects are safe in ordinary code, and `sapply()` was only one
example: `lapply()`, `tapply()`, `mapply()`, `Map()`, `Reduce()` and `apply()`
are all safe for the same reason.

Five ways of writing a call still fail, for two different reasons. The first
is a name passed on twice before R looks it up:

```{r}
#| error: true
aggregate(life ~ continent, gapminder_2007, mean)
```

By the time the lookup happens, the name being searched for is `FUN` and not
`mean`, so the message names neither `mean` nor gog. `by()` and `ave(..., FUN =
mean)` fail the same way. Writing `base::mean` fixes all three. The data frame
form of `aggregate()` looks the name up one step earlier, so
`aggregate(gapminder_2007["life"], gapminder_2007["continent"], mean)` works
unchanged.

The second reason is a name used as a value rather than called. `do.call(sum,
list(x))` wants a function and receives an object, so write `do.call("sum",
list(x))` or `do.call(base::sum, list(x))`. And `f <- mean` copies the object,
so `f(x)` reports that it could not find a function called `f`.

That is the whole cost, and it is worth stating as one rule. If you hand one
of those nine words to another function *by name*, and that function hands it
on again, qualify it with the package it belongs to. Only four of the nine are
base R's own: `max`, `mean`, `min` and `sum`. `line`, `median`, `smooth` and
`step` belong to `stats`, and `text` belongs to `graphics`, so `base::median`
is itself an error.

## Reading from a database

Your table may live in a database rather than in your session. `query()` reads
one, and it takes a **DBI** connection. DBI is R's database standard, so the
same call reaches SQLite, Postgres, MariaDB, BigQuery, and anything with an
ODBC driver.

```r
con <- DBI::dbConnect(RSQLite::SQLite(), "sales.db")

query(con, "SELECT status, revenue FROM orders") + bar + x(status) + y(revenue)
```

DBI and its drivers are **Suggests**, not hard dependencies. A reader who never
writes SQL installs neither. If you do write SQL, install DBI and the one driver
your database needs:

```r
install.packages(c("DBI", "RSQLite"))    # or RPostgres, RMariaDB, odbc, bigrquery
```

gog never opens a connection and never stores a password. You connect the way
you already do, and hand the open connection over.

## Getting it

A plot is drawn by the engine, which is a compiled Rust binary, so the R
package has to be able to find one. An installed copy carries its own engine.
The package's `configure` script bundles the engine at install time, and
refuses to install at all rather than leave you a package that cannot draw.

```r
install.packages("gog", repos = c("https://psychometrician.r-universe.dev",
                                  "https://cloud.r-project.org"))
```

That command works today. gog is published on
[r-universe](https://psychometrician.r-universe.dev/gog). On macOS and Windows
r-universe builds a binary with the engine inside it, so installing needs no
Rust toolchain.

On Linux the same command installs from source, and the source build compiles
the engine. That needs Rust and a network connection. If the machine does not
have Rust, install it first:

```bash
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh -s -- -y
source "$HOME/.cargo/env"
```

Then install gog with the `install.packages()` call at the start of this
section. The build runs once, while the package installs.

The `repos` argument names two addresses because gog is not on CRAN. The first
address is the r-universe repository that holds gog. The second is an ordinary
CRAN mirror, and it is there so your other packages still install as usual.

There is another route, for when you want the source repository itself rather
than the last build r-universe published. It installs gog directly from
GitHub. It needs the `remotes` package first, so install that if you do not
have it:

```r
install.packages("remotes")
```

Then install gog:

```r
remotes::install_github("psychometrician/gog", subdir = "r-pkg/gog",
                        build = FALSE)
```

A download from GitHub carries the whole repository, so the engine's Rust
source code comes with the package. `configure` compiles it during the
install. The `build = FALSE` argument is what keeps that source code in the
download. Without it, `remotes` first copies the package folder alone into a
new archive. The Rust sources are left out, and the install stops with a
message that names the argument. So this route needs a Rust toolchain on
every system, including the macOS and Windows machines where the r-universe
install needs none. To install one particular branch, tag or commit instead
of the newest, add `ref = "name"` to the call.

You can also work from a checkout. Outside base R, gog needs one package,
`jsonlite`, which turns the sentence you write into the JSON that the engine
reads. `install.packages()` and `remotes::install_github()` both install it
along with gog. Loading from a checkout needs `pkgload` as well, which loads a
package from a folder instead of from a library. A checkout installs neither
for you, so install both if you do not have them:

```r
install.packages(c("jsonlite", "pkgload"))
```

If one is missing, the step that needs it stops immediately, and the message
names the missing package.

Then build the engine once and load the package from where it sits:

```bash
cargo build --release -p gog-cli
```

```r
pkgload::load_all("r-pkg/gog")
```

Installing from that same checkout is a stronger test than loading it, because
it runs `configure` and proves the bundling works:

```bash
R CMD INSTALL r-pkg/gog
```

The engine is looked for in five places, in this order: the `GOG_CLI_PATH`
environment variable, a `target/` build in a surrounding checkout, the copy
inside the installed package, `gog-cli` on your `PATH`, and one last search
upward from the working directory. A checkout is searched before the bundled
copy on purpose. Inside a checkout the bundled engine is a copy an earlier
install left behind. Using it would draw with an old engine while you are
changing the engine's source.

## What comes next

R's sentence is the one every chapter before this has written.
[Python](python.qmd) writes the same sentence with `col.gdp` where R writes a
bare name, and draws the same picture, byte for byte.

