# R

```{r}
#| include: false
source("../R/setup.R")
```

## Pipes

R has two pipes in common use, and R users reasonably want to start a plot
with the table they already have in hand. The short answer: the native pipe
works, and gives you exactly the plot you would have written by hand.

```{r}
gapminder_2007 |> data() + point + x(gdp) + y(life)
```

That is not a second way of saying it, and GOG does not implement it. By the
time the engine sees anything, the two forms are the *same expression*. You can
watch R rewrite one into the other, because `quote()` shows what the parser
built rather than what you typed:

```{r}
quote(gapminder_2007 |> data() + point + x(gdp) + y(life))
```

Two facts about R combine to make that work, and both have to hold.

The first is precedence. `|>` binds more tightly than `+`, so the pipe closes
before the chain begins: it takes the table into `data()` and hands the
finished result to `+`. Derivation with `*` sits between the two, which is
where it needs to be, so a transform still binds to its mark first:

```{r}
gm_all |> data() + bar * count + x(continent)
```

The second is *when* it happens. The native pipe is a parser transformation:
writing `df |> data()` produces the code `data(df)` before any of it runs.
That matters more than it sounds, because `data()` reads the table's name off
the expression you wrote. The pipe leaves it a plain name, so nothing is lost.
Naming the table is the whole job of `data()` (see [Data](../data.qmd)), and it
is what a later layer resolves its bare columns against.

Arguments still land where you would expect, since the piped value fills the
first one:

```{r}
gapminder_2007 |> data(name = "gapminder") + point + x(gdp) + y(life)
```

## What `%>%` does to the table's name

The magrittr pipe, `%>%`, is the one most R code still uses, and here it
behaves differently. It is an ordinary function rather than a parser rule, so
it *evaluates* the left side and passes the value along under its own
placeholder, `.`. The name never reaches `data()`:

```{r}
#| warning: true
library(magrittr)
gapminder_2007 %>% data() + point + x(gdp) + y(life)
```

The plot is right, and for a single table it stays right. What is gone is the
name, so GOG says so and tells you how to put it back. Either use the native
pipe, or name the table yourself:

```{r}
gapminder_2007 %>% data(name = "gapminder_2007") + point + x(gdp) + y(life)
```

Only the placeholder draws that warning, and it is worth being clear about what
does *not*. Filtering on the way in is the pipe's most useful shape, and it stays
silent: the expression you piped is not a plain name, but it is still yours and
still distinct, so GOG keeps it as the table's name rather than lecturing you
about it.

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data() +
  point + x(gdp) + y(life)
```

The table there is called `subset(gm_all, year == 2007 & continent == "Europe")`,
which is ugly and works. `name =` is available when you would rather read
something else, and becomes worth using once a second table is involved:

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data(name = "europe") +
  point + x(gdp) + y(life)
```

## Why the name is worth a warning

An unnamed table is fine until there are two of them. A second `data()` gives
a layer its own table, and "nearest table wins" is decided *by name*, so two
tables that both lost theirs leave the rule nothing to choose between:

```{r}
#| error: true
data(actuals, name = "series") + x(year) + y(sales) + line +
  data(forecast, name = "series") + point
```

Before the warning existed, that plot did not refuse. It drew the first table
twice and reported the second table's columns as misspellings, which blames
the reader for something the binding lost. Two `%>%` pipes in one expression
reached exactly that state, both tables having arrived as `.`.

## What the pipe does not change

The pipe feeds `data()`. It does not replace `+`, and there is no piped form
of the rest of the sentence: marks, channels, and transforms compose with `+`
and `*` as they do everywhere else in this book. GOG has no pipe of its own
and will not grow one, because that would be a second spelling for something
the language already spells.

This section is about R specifically. Julia's `|>` is a function-application
operator with different behavior, and Python has no pipe at all, so their
chapters will answer the same question their own way.

## Eighteen names base R already uses

`library(gog)` prints a long list of masked names, and the list is alarming out
of proportion to the problem. Eighteen of the package's exports are also names in
the packages R attaches for you:

| from | names |
|---|---|
| `stats` | `density` `line` `median` `smooth` `step` |
| `graphics` | `box` `text` `title` |
| `grDevices` | `palette` |
| `utils` | `data` `stack` |
| `base` | `jitter` `max` `mean` `min` `order` `range` `sum` |

**Ten of the eighteen cannot hurt you, and the reason is a rule in R itself.**
Every mark and every value transform is an *object*, not a function. When R
evaluates a call, it looks for a function of that name and skips any binding that
is not one. So base R still answers, even with GOG's `mean` sitting closer:

```{r}
is.function(mean)   # GOG's `mean` is an object, so a call cannot reach it
mean(1:10)          # and base R's function answers anyway
```

That holds for the functions that take another function as an argument, because
they resolve the name the same way:

```{r}
sapply(gapminder_2007["life"], mean)
```

**The other eight are functions, and those do take over**: `box`, `data`,
`density`, `jitter`, `order`, `palette`, `stack` and `title`. Qualify the one you
want, and both remain available:

| you write | you get | for base R's, write |
|---|---|---|
| `order(population)` | the plot's sort order | `base::order(df$population)` |
| `data(gapminder_2007)` | the plot's table | `utils::data(diamonds)` |
| `density(2)` | the smoothing transform | `stats::density(x)` |
| `jitter(0.5)` | the offset transform | `base::jitter(x)` |
| `stack(share = TRUE)` | the piling transform | `utils::stack(df)` |
| `palette("soft")` | the plot's colors | `grDevices::palette()` |
| `title("Life expectancy")` | the plot's title | `graphics::title(main = "…")` |
| `box()` | the box-and-whisker mark | `graphics::box()` |

`order` is the one that used to hurt. Sorting a data frame beside a plot is
ordinary work. GOG's `order` returns an atom, not the row positions that `[`
expects. So the failure used to arrive far from the line that caused it, and it
named neither `order` nor GOG. It now says which function you reached, and how to
reach the other one:

```{r}
#| error: true
gapminder_2007[order(gapminder_2007$population), ]
```

Qualified, it sorts:

```{r}
ranked <- gapminder_2007[base::order(-gapminder_2007$population), ]
head(ranked[c("country", "population")], 3)
```

This book does that in three chapters, for exactly this reason.

Two of the eight refuse nothing, because nothing tells the two intentions apart.
`data(mtcars)` is well formed in both readings. It returns a plot's table and no
complaint, so write `utils::data()` when you mean to load a dataset.
`title("...")` is the same case: a string is what both want.

The names stay as they are. A grammar keeps its own vocabulary, which is Law 3 and
a decided question (spec §18). Renaming `order` to `sort_by` would trade a word
every reader knows for one only this package uses. Where another *package* owns a
name you want, attach `gog` after it so its names win, or qualify it as
`gog::interval`.

All four languages meet this wall, and each one hits it differently. Julia will
not choose between two modules that export one name. It raises, and you import
the one you meant. Python has no rule about skipping non-functions, so
`from gog import *` really does replace five builtins. Calling one of them raises
rather than misbehaving. R is the quiet one of the four, which is why the eight
names above refuse out loud. JavaScript has no collision at all, because nothing
reaches your scope unless you import it by name.

## Getting it

A plot is drawn by the engine, which is a compiled Rust binary, so the R package
has to be able to find one. An installed copy **carries its own**. The package's
`configure` script bundles the engine at install time, and refuses to install at
all rather than leave you a package that cannot draw.

```r
install.packages("gog")
```

That is the shape of the answer rather than a command you can run today. The
package is not on CRAN yet. `remotes::install_github()` will work before CRAN
does, because a source tarball carries the engine's crates and `configure` builds
them, but that route needs a public repository and there is not one yet.

Until then, work from a checkout. Build the engine once, then load the package
from where it sits:

```bash
cargo build --release -p gog-cli
```

```r
pkgload::load_all("r-pkg/gog")
```

Installing from that same checkout is the better test, because it runs
`configure` and proves the bundling works:

```bash
R CMD INSTALL r-pkg/gog
```

The engine is looked for in five places, in this order: the `GOG_CLI_PATH`
environment variable, a `target/` build in a surrounding checkout, the copy
inside the installed package, `gog-cli` on your `PATH`, and one last search
upward from the working directory. A checkout outranks the bundled copy on
purpose. Inside a checkout the bundled engine is a leftover build artifact, and
preferring it is how a stale engine gets served to the person editing the engine.
