# R

```{r}
#| include: false
source("../R/setup.R")
```

Every sentence in this book so far was written in R, so this chapter adds
nothing to the grammar. What it adds is the four questions only R raises: how
the two pipes reach `data()`, which of gog's words base R already uses, how a
database connection arrives, and how the package installs. What a table is, and
how one is passed, is in [Data](../data.qmd); it needs nothing installed beyond
R.

Every R block below was executed to draw the plot beside it, through the same
engine every other chapter uses.

## Pipes

R has two pipes in common use, and R users reasonably want to start a plot
with a table they already have. The native pipe works, and draws exactly the
plot the unpiped form draws.

```{r}
gapminder_2007 |> data() + point + x(gdp) + y(life)
```

That is not a second way of saying it, and gog does not implement it. By the
time the engine sees anything, the two forms are the *same expression*. You can
watch R rewrite one into the other, because `quote()` shows what the parser
built rather than what you typed:

```{r}
quote(gapminder_2007 |> data() + point + x(gdp) + y(life))
```

Two facts about R combine to make that work, and both have to hold.

The first is precedence. `|>` binds more tightly than `+`, so the pipe finishes
before the sentence begins: it takes the table into `data()` and hands the
finished result to `+`. Derivation with `*` sits between the two, which is
where it needs to be, so a transform still binds to its mark first:

```{r}
gm_all |> data() + bar * count + x(continent)
```

The second is *when* it happens. The native pipe is a parser transformation:
writing `df |> data()` produces the code `data(df)` before any of it runs.
That timing decides everything, because `data()` reads the table's name off
the expression you wrote. The pipe leaves it a plain name, so nothing is lost.
Naming the table is the whole job of `data()` (see [Data](../data.qmd)), and it
is what a later layer resolves its bare columns against.

Arguments still land where you would expect, since the piped value fills the
first one:

```{r}
gapminder_2007 |> data(name = "gapminder") + point + x(gdp) + y(life)
```

## What `%>%` does to the table's name

The magrittr pipe, `%>%`, is the one most R code still uses, and here it
behaves differently. It is an ordinary function rather than a parser rule, so
it *evaluates* the left side and passes the value along under its own
placeholder, `.`. The name never reaches `data()`:

```{r}
#| warning: true
library(magrittr)
gapminder_2007 %>% data() + point + x(gdp) + y(life)
```

The plot is right, and for a single table it stays right. What is gone is the
name, so gog says so and tells you how to put it back. Either use the native
pipe, or name the table yourself:

```{r}
gapminder_2007 %>% data(name = "gapminder_2007") + point + x(gdp) + y(life)
```

Only the placeholder raises that warning, and it is worth saying what does not.
Filtering before `data()` is the pipe's most useful form, and it raises no
warning. The expression is not a plain name, but it is still distinct, so gog
uses it as the table's name.

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data() +
  point + x(gdp) + y(life)
```

The table there is called `subset(gm_all, year == 2007 & continent == "Europe")`,
which is ugly and works. `name =` is available when you would rather read
something else, and becomes worth using once a second table is involved:

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data(name = "europe") +
  point + x(gdp) + y(life)
```

## Why the name is worth a warning

An unnamed table is fine until there are two of them. A second `data()` gives
a layer its own table, and "nearest table wins" is decided *by name*, so two
tables that both lost theirs leave the rule nothing to choose between:

```{r}
#| error: true
data(actuals, name = "series") + x(year) + y(sales) + line +
  data(forecast, name = "series") + point
```

Before the warning existed, that plot did not refuse. It drew the first table
twice and reported the second table's columns as misspellings, which blames
the reader for something the binding lost. Two `%>%` pipes in one expression
reached exactly that state, both tables having arrived as `.`.

## What the pipe does not change

The pipe feeds `data()`. It does not replace `+`, and there is no piped form
of the rest of the sentence: marks, channels, and transforms compose with `+`
and `*` as they do everywhere else in this book. gog has no pipe of its own
and will not grow one, because that would be a second spelling for something
the language already spells.

This section is about R specifically. Julia's `|>` is a function-application
operator with different behavior, and Python and JavaScript have no pipe of
their own, so the question does not arise in the other three chapters.

## Nineteen names base R already uses

`library(gog)` prints a long list of masked names, and the list is alarming out
of proportion to the problem. Nineteen of the package's exports are also names in
the packages R attaches for you:

| from | names |
|---|---|
| `stats` | `density` `line` `median` `quantile` `smooth` `step` |
| `graphics` | `box` `text` `title` |
| `grDevices` | `palette` |
| `utils` | `data` `stack` |
| `base` | `jitter` `max` `mean` `min` `order` `range` `sum` |

Nine of the nineteen cannot affect you, and the reason is a rule in R itself.
A mark is an *object* rather than a function, and so is a transform that takes no
parameter. When R evaluates a call, it looks for a function of that name and
skips any binding that is not one. So base R still answers, even with gog's
`mean` sitting closer:

```{r}
is.function(mean)   # gog's `mean` is an object, so a call cannot reach it
mean(1:10)          # and base R's function answers anyway
```

That holds for the functions that take another function as an argument, because
they resolve the name the same way:

```{r}
sapply(gapminder_2007["life"], mean)
```

The other ten are functions, and a call does reach gog's: `box`, `data`,
`density`, `jitter`, `order`, `palette`, `quantile`, `range`, `stack` and
`title`. Qualify the one you want, and both remain available:

| you write | you get | for base R's, write |
|---|---|---|
| `order(population)` | the plot's sort order | `base::order(df$population)` |
| `data(gapminder_2007)` | the plot's table | `utils::data(diamonds)` |
| `density(2)` | the smoothing transform | `stats::density(x)` |
| `jitter(0.5)` | the offset transform | `base::jitter(x)` |
| `quantile(0.9)` | the 90th percentile transform | `stats::quantile(df$gdp)` |
| `range(0.25, 0.75)` | the band transform | `base::range(df$gdp)` |
| `stack(share = TRUE)` | the piling transform | `utils::stack(df)` |
| `palette("soft")` | the plot's colors | `grDevices::palette()` |
| `title("Life expectancy")` | the plot's title | `graphics::title(main = "…")` |
| `box()` | the box-and-whisker mark | `graphics::box()` |

`range` is a function because it takes a band, and R finds a function first. That
is what writing `range(0.25, 0.75)` costs you. The refusal reads the argument it
was given: a vector where a probability belongs means you wanted base R's, and
the message says so.

```{r}
#| error: true
range(c(3, 1, 4, 1, 5))
```

`order` is the one that used to break. Sorting a data frame beside a plot is
ordinary work. gog's `order` returns an atom, not the row positions that `[`
expects. So the failure used to arrive far from the line that caused it, and it
named neither `order` nor gog. It now says which function you reached, and how to
reach the other one:

```{r}
#| error: true
gapminder_2007[order(gapminder_2007$population), ]
```

Qualified, it sorts:

```{r}
ranked <- gapminder_2007[base::order(-gapminder_2007$population), ]
head(ranked[c("country", "population")], 3)
```

This book does that in three chapters, for exactly this reason.

Three of the ten refuse nothing, because nothing separates the two intentions.
`data(mtcars)` is well formed in both readings. It returns a plot's table and no
complaint, so write `utils::data()` when you mean to load a dataset.
`title("...")` is the same case: a string is what both want. So is `box()`,
which takes no argument in either language. The other seven all report the
mistake, because in each of them the argument you passed is one gog would
refuse.

The names stay as they are. A grammar keeps its own vocabulary, which is Law 3 and
a decided question (spec §18). Renaming `order` to `sort_by` would replace a word
every reader knows with one only this package uses. Where another *package* owns a
name you want, attach `gog` after it, so R finds gog's names first, or qualify it as
`gog::interval`. Every language meets this problem, and each one asks you to
solve it differently;
[the grammar chapter](../grammar.qmd#names-the-host-language-already-uses)
compares them.

## Five spellings that still need a prefix

The nine objects above are safe in ordinary code, and safe inside `sapply()`,
`lapply()`, `tapply()`, `mapply()`, `Map()`, `Reduce()` and `apply()`. Each of
those asks R for a function of the name it was handed, which finds base R's.

Five spellings still fail, for two different reasons. The first is a name that
travels through two layers before anyone looks it up:

```{r}
#| error: true
aggregate(life ~ continent, gapminder_2007, mean)
```

By the time the lookup happens, the name being searched for is `FUN` and not
`mean`, so the message names neither `mean` nor gog. `by()` and
`ave(..., FUN = mean)` fail the same way. Writing `base::mean` fixes all three.
The data frame form of `aggregate()` looks the name up one layer earlier, so
`aggregate(gapminder_2007["life"], gapminder_2007["continent"], mean)` works
unchanged.

The second reason is a name used as a value rather than called. `do.call(sum,
list(x))` wants a function and receives an object, so write
`do.call("sum", list(x))` or `do.call(base::sum, list(x))`. And `f <- mean`
copies the object, so `f(x)` reports that `f` is not a function.

That is the whole cost, and it is worth stating as one rule. If you hand one of
those nine words to another function *by name*, and that function hands it on
again, write `base::` in front of it.

## Reading from a database

`query()` takes a **DBI** connection. DBI is R's database standard, so the same
call reaches SQLite, Postgres, MariaDB, BigQuery, and anything with an ODBC
driver.

```r
con <- DBI::dbConnect(RSQLite::SQLite(), "sales.db")

query(con, "SELECT status, revenue FROM orders") + bar + x(status) + y(revenue)
```

DBI and its drivers are **Suggests**, not hard dependencies. A reader who never
writes SQL installs neither. If you do write SQL, install DBI and the one driver
your database needs:

```r
install.packages(c("DBI", "RSQLite"))    # or RPostgres, RMariaDB, odbc, bigrquery
```

`gog` never opens a connection and never stores a password. You connect the way
you already do, and hand the open connection over.

## Getting it

A plot is drawn by the engine, which is a compiled Rust binary, so the R package
has to be able to find one. An installed copy **carries its own**. The package's
`configure` script bundles the engine at install time, and refuses to install at
all rather than leave you a package that cannot draw.

```r
install.packages("gog", repos = c("https://psychometrician.r-universe.dev",
                                  "https://cloud.r-project.org"))
```

That command works today. `gog` is published on
[r-universe](https://psychometrician.r-universe.dev/gog), which builds a binary
for macOS, Windows and Linux, and every one of them has the engine inside it. So
installing needs no Rust toolchain.

The `repos` argument names two addresses because r-universe is not CRAN. The
first address is where `gog` lives. The second is an ordinary CRAN mirror, and it
is there so your other packages still install as usual. `gog` is not on CRAN yet;
when it is, the plain `install.packages("gog")` will work and this line can lose
its second half.

There is another route, and it installs straight from the source repository:

```r
remotes::install_github("psychometrician/gog", subdir = "r-pkg/gog")
```

A tarball from GitHub carries the engine's Rust sources, and `configure` compiles
them during the install. So this route does need a Rust toolchain, and the
r-universe binaries do not. Take it when you want one particular branch or commit
rather than the latest build.

You can also work from a checkout. Build the engine once, then load the package
from where it sits:

```bash
cargo build --release -p gog-cli
```

```r
pkgload::load_all("r-pkg/gog")
```

Installing from that same checkout is the better test, because it runs
`configure` and proves the bundling works:

```bash
R CMD INSTALL r-pkg/gog
```

The engine is looked for in five places, in this order: the `GOG_CLI_PATH`
environment variable, a `target/` build in a surrounding checkout, the copy
inside the installed package, `gog-cli` on your `PATH`, and one last search
upward from the working directory. A checkout is searched before the bundled copy
on purpose. Inside a checkout the bundled engine is a leftover build, and using
it would give a stale engine to the person editing the engine.
