# R

```{r}
#| include: false
source("../R/setup.R")
```

Every sentence in this book so far was written first in R, and each tab for
another language shows a translation of it. So this chapter adds nothing to the
grammar. It answers the five questions R raises: how the two
pipes reach `data()`, where a plot appears, which of gog's words R's own
packages already use, how a database table reaches a plot, and how the package
installs. [Data](../data.qmd) covers what a table is and how you pass one. A
table is an ordinary R data frame, and needs no package of its own.

Every R block on this page ran when the page was built, except the database and
install commands, which are shown but not run. Every plot came from the same
engine every other chapter uses.

## Pipes

Much R code prepares a table in a chain of steps joined by a pipe, and you will
often want the last step to be a plot. R has two pipes in common use, and which
one you write decides whether gog learns the table's name ([Data](../data.qmd)
explains why the name matters). Write the native pipe, `|>`: it keeps the name,
and draws exactly the plot that
`data(gapminder_2007) + point + x(gdp) + y(life)` draws.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
gapminder_2007 |> data() + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

gog has no code for the pipe. Before anything runs, R rewrites the piped form as
the unpiped one, so the two are the *same expression*. `quote()` shows the
expression R built rather than what you typed:

```{r}
quote(gapminder_2007 |> data() + point + x(gdp) + y(life))
```

Two facts about R make that work, and both are needed.

The first is precedence, which decides what groups first
([Operators](../operators.qmd#what-groups-first)). `|>` groups before `+`, so
`gapminder_2007 |> data()` is complete before `+` joins anything to it. `*`
groups after `|>` and before `+`, so a transform still joins its mark first:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
gm_all |> data() + bar * count + x(continent)
```

*"Given all the gapminder years: bars derived by count, x is continent."*

The second is *when* it happens. The native pipe is a rule of R's parser, the
part of R that reads your code before running it: writing `df |> data()`
produces the code `data(df)` before any of it runs. That timing matters because
`data()` reads the table's name from the expression you wrote, and the native
pipe leaves that expression a bare name. `data()` keeps the name as well as the
rows ([Data](../data.qmd)), and a later layer looks up its columns in the table
with that name.

The piped table fills the first argument, so any argument you write yourself
follows it:

```{r}
gapminder_2007 |> data(name = "gapminder") + point + x(gdp) + y(life)
```

## If your code uses `%>%`

The magrittr pipe, `%>%`, is the one most R code still uses, and it does not
keep the table's name. It is an ordinary function rather than a rule of the
parser, so it *evaluates* the left side and gives `data()` the placeholder `.`
instead of the table's name:

```{r}
#| warning: true
library(magrittr)
gapminder_2007 %>% data() + point + x(gdp) + y(life)
```

The plot is correct. With one table, the missing name changes nothing in the
plot, but gog warns you anyway and says how to supply it. Either use the native
pipe, or name the table yourself:

```{r}
gapminder_2007 %>% data(name = "gapminder_2007") + point + x(gdp) + y(life)
```

Only the placeholder `.` raises that warning. The native pipe raises none, even
when what you pipe in is an expression rather than a bare name. Its most useful
case is filtering the table before `data()`. Unlike `.`, that expression is
different for every table, so gog uses the whole expression as the table's
name.

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data() +
  point + x(gdp) + y(life)
```

gog drops the quotation marks, so the table there is called
`subset(gm_all, year == 2007 & continent == Europe)`. That name is long, and it
still works. `name =` gives the table a shorter name:

```{r}
gm_all |> subset(year == 2007 & continent == "Europe") |> data(name = "europe") +
  point + x(gdp) + y(life)
```

## Two tables with one name

The name matters once a plot has two tables. A second `data()` gives a layer its
own table, and the "nearest table wins" rule finds each table *by name*. So gog
refuses two different tables that you gave the same name:

```{r}
#| echo: false
peek(actuals)
peek(forecast)
```

```{r}
#| error: true
data(actuals, name = "series") + x(year) + y(sales) + line +
  data(forecast, name = "series") + point
```

Without that refusal the plot would draw the first table twice and report the
second table's columns as misspellings. You would have spelled the columns
correctly, and gog would have checked them against the wrong table. Two `%>%`
tables do not cause this refusal. gog calls them `data` and `data2`, warns once
for each, and draws the plot. Naming the tables yourself removes both
warnings.

## What the pipe does not change

The pipe supplies the table that `data()` names. It does not replace `+`, and
there is no piped form of the rest of the sentence. Marks, channels, and
transforms compose with `+` and `*` as they do everywhere else in this book. gog
has no pipe of its own, because R already has two.

The pipe question is R's alone. Julia has a `|>`, but Julia's `data()` never
reads a table's name from your code, so its pipe changes nothing. Neither Python
nor JavaScript has a pipe operator. The other three chapters do not raise it.

## Showing the plot

That was the first question. The second is where a plot appears once you have
written its sentence. You do not call anything to show it: R shows a plot as
soon as you run its sentence, and where it appears depends on where you work:

- At the console, it opens in the viewer pane of RStudio or Positron. Without a
  viewer pane, it opens in your web browser.
- In a Quarto or R Markdown document, it is drawn under the chunk, as every plot
  in this book is.
- In a Jupyter notebook with the R kernel, it appears under the cell. `print()`
  shows it too, so a loop can show several plots one after another.

To keep the drawing itself, [`render_svg()`](../utilities.qmd#render_svg)
returns it as text.

## Twenty names R already uses

The third question is which of gog's words are already R's. Twenty of the names
gog provides are also names in the packages R loads when it starts, and
`library(gog)` lists eleven of them. Here are all twenty:

| from | names |
|---|---|
| `stats` | `density` `line` `median` `quantile` `smooth` `step` |
| `graphics` | `box` `layout` `text` `title` |
| `grDevices` | `palette` |
| `utils` | `data` `stack` |
| `base` | `jitter` `max` `mean` `min` `order` `range` `sum` |

Nine of the twenty cannot affect an ordinary call, and `library(gog)` leaves
those nine out of its message. A rule in R itself is the reason. A mark that
takes no argument is an *object* rather than a function, and so is a transform
that takes no argument. When R evaluates a call, it looks for a function of that
name and skips any object of that name that is not a function. So a call still
reaches base R's `mean`, even though `library(gog)` attached its own:

```{r}
is.function(mean)   # gog's `mean` is an object, so a call cannot reach it
mean(1:10)          # and base R's function answers anyway
```

The same is usually true when you pass `mean` to another function, because most
functions look the name up the same way; the exceptions come [later in this
chapter](#five-calls-that-still-need-a-prefix):

```{r}
sapply(gapminder_2007["life"], mean)
```

The other eleven are functions, and a call does reach gog's: `box`, `data`,
`density`, `jitter`, `layout`, `order`, `palette`, `quantile`, `range`,
`stack` and `title`. To choose one, give it a prefix: the package's name and
`::`, as in `base::order`. Both remain available:

| you write | you get | for R's own, write |
|---|---|---|
| `order(population)` | the plot's sort order | `base::order(df$population)` |
| `data(gapminder_2007)` | the plot's table | `utils::data(mtcars)` |
| `density(2)` | the density transform | `stats::density(x)` |
| `jitter(0.5)` | the offset transform | `base::jitter(x)` |
| `layout(exporter, importer)` | the layout transform | `graphics::layout(mat)` |
| `quantile(0.9)` | the 90th percentile transform | `stats::quantile(df$gdp)` |
| `range(0.25, 0.75)` | the band transform | `base::range(df$gdp)` |
| `stack(share = TRUE)` | the piling transform | `utils::stack(df)` |
| `palette("soft")` | the plot's colors | `grDevices::palette()` |
| `title("Life expectancy")` | the plot's title | `graphics::title(main = "…")` |
| `box()` | the box-and-whisker mark | `graphics::box()` |

gog's `range` takes two numbers, so it is a function rather than an object, and
R does not skip past it. Writing `range(0.25, 0.75)` therefore reaches gog's
`range` and never base R's. gog then reads the argument you passed. If you pass
a vector where gog expects two numbers between 0 and 1, you meant base R's
`range`, and gog's refusal says so:

```{r}
#| error: true
range(c(3, 1, 4, 1, 5))
```

Of the eleven, `order` causes the most errors, because you often sort a table
while you are making plots. gog's `order` returns an atom, not the row positions
that `[` expects. Without a refusal, R would stop with
`invalid subscript type 'list'`, which names neither `order` nor gog. So
whenever the argument is not a bare name, the refusal says which function you
reached and how to reach the other one:

```{r}
#| error: true
gapminder_2007[order(gapminder_2007$population), ]
```

With the prefix, it sorts:

```{r}
ranked <- gapminder_2007[base::order(-gapminder_2007$population), ]
head(ranked[c("country", "population")], 3)
```

Several other chapters write `base::order` for exactly this reason.

Three of the eleven refuse nothing, because gog cannot tell from the argument
which function you meant. `data(mtcars)` is well formed for both. It starts a
plot and raises no complaint, so write `utils::data()` when you mean to load a
dataset. `title("...")` is the same case: a string is what both want. So is
`box()`, which both accept with no argument at all. Six of the others report the
mistake, because what R's function usually takes, a vector, a matrix or a table,
is an argument gog refuses. The last two catch only part of it. `order` reports
`order(df$population)` but not `order(pop)`, where `pop` is a bare name holding
a vector. `palette` reports `palette()` but not `palette("R4")`, because a
palette name is what both want.

The names stay as they are, because renaming `order` to `sort_by` would replace
a word every reader knows with one that only gog uses. [The grammar
chapter](../grammar.qmd#names-the-host-language-already-uses) gives the full
reason and compares how each language handles a name it already uses. R's own
packages are not the only source of a collision. Another package may define a
name gog also defines. Load gog with `library(gog)` after that package, so R
finds gog's name first, or write the name with the prefix `gog::`.

## Five calls that still need a prefix

The nine words that are objects in gog, such as `mean`, are safe in ordinary
code, and `sapply()` was only one example: `lapply()`, `tapply()`, `mapply()`,
`Map()`, `Reduce()` and `apply()` are all safe for the same reason.

Five ways of writing a call still fail, for two different reasons. The first is
a name that one function passes to a second function before R looks it up:

```{r}
#| error: true
aggregate(life ~ continent, gapminder_2007, mean)
```

When R finally looks the name up, it is looking for `FUN`, the name
`aggregate()` gave the function you passed, and not for `mean`. So the message
names neither `mean` nor gog. `by()` and `ave(..., FUN = mean)` fail the same
way. Writing `base::mean` fixes all three. The data frame
form of `aggregate()` looks the name up one step earlier, so
`aggregate(gapminder_2007["life"], gapminder_2007["continent"], mean)` works
unchanged.

The second reason is a name used as a value rather than called.
`do.call(sum, list(x))` wants a function and receives an object, so write
`do.call("sum", list(x))` or `do.call(base::sum, list(x))`. And `f <- mean`
copies the object, so `f(x)` reports that it could not find a function called
`f`.

That is the whole cost, and one habit avoids all five failures: when you pass
one of those nine words to another function, or store it in a variable, write it
with its prefix, as in `base::mean`. The functions listed above work without the
prefix, but the prefixed name works everywhere. Only four of the nine are base
R's own: `max`, `mean`, `min` and `sum`. `line`, `median`, `smooth` and `step`
belong to `stats`, and `text` belongs to `graphics`, so write `stats::median`;
`base::median` is itself an error.

## Reading from a database

The fourth question is how a table in a database reaches a plot. `query()` runs
a query on a database, and it takes a **DBI** connection. DBI is R's database
standard, so the same call reaches SQLite, Postgres, MariaDB, BigQuery, and
anything with an ODBC driver.

```r
con <- DBI::dbConnect(RSQLite::SQLite(), "sales.db")

query(con, "SELECT status, revenue FROM orders") + bar + x(status) + y(revenue)
```

DBI and the drivers are optional: gog's DESCRIPTION file lists DBI and RSQLite
under **Suggests**, R's field for optional packages, so installing gog does not
install them. If you never write SQL, you need neither. If you do, install DBI
and the one driver your database needs:

```r
install.packages(c("DBI", "RSQLite"))    # or RPostgres, RMariaDB, odbc, bigrquery
```

You open the connection with DBI yourself, as the example above does, and pass
it to `query()`. gog never opens one and never stores a password.
[Data](../data.qmd#a-table-in-a-database) covers what `query()` sends and how
much data comes back.

## Installing gog

The last question is how you install gog. Every plot is drawn by the engine, a
separate program written in Rust, so the R package has to be able to find it.
An installed copy of the package carries its own engine. The package's
`configure` script, which runs during the install, puts the engine inside the
package, and it stops the install rather than leave you a package that cannot
draw.

```r
install.packages("gog", repos = c("https://psychometrician.r-universe.dev",
                                  "https://cloud.r-project.org"))
```

gog is published on [r-universe](https://psychometrician.r-universe.dev/gog).
For the current version of R and the one before it, r-universe builds a
ready-to-install copy of the package for macOS and Windows, with the engine
already inside it, so installing needs no Rust toolchain. With an older version
of R, the command installs from source, as it does on Linux.

On Linux the same command installs from source, and the source build compiles
the engine. That needs Rust and a network connection. If the machine does not
have Rust, install it first:

```bash
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh -s -- -y
source "$HOME/.cargo/env"
```

Then install gog with the `install.packages()` call at the start of this
section. The build runs once, while the package installs.

The `repos` argument names two addresses because gog is not on CRAN. The first
address is the r-universe repository that holds gog. The second is an ordinary
CRAN mirror. It is there for `jsonlite`, the one package gog needs that
r-universe does not hold.

There is another way to install gog: straight from its source code on GitHub,
for when you want a particular version rather than the build r-universe
published. It needs the `remotes` package first, so install that if you do not
have it:

```r
install.packages("remotes")
```

Then install gog:

```r
remotes::install_github("psychometrician/gog", subdir = "r-pkg/gog",
                        build = FALSE)
```

A download from GitHub holds the whole repository, including the engine's Rust
source code, and `configure` compiles the engine during the install. So
installing this way needs a Rust toolchain on every system, including macOS and
Windows, where the r-universe install needs none. The `build = FALSE` argument
keeps the Rust source in the download. Without it, `remotes` first copies the
package folder alone into a new archive, the Rust source is left out, and the
install stops with a message that names the argument. To install one particular
branch, tag or commit instead of the newest, add `ref = "name"` to the call.

You can also work from a checkout, a copy of gog's source repository on your own
computer; that is how you change gog itself. Besides the packages that come with
R, gog needs one package, `jsonlite`, which turns the sentence you write into
the JSON that the engine reads. `install.packages()` and
`remotes::install_github()` both install it along with gog. Loading from a
checkout needs `pkgload` as well, which loads a package from a folder instead of
from a library. Working from a checkout installs neither, so install both if you
do not have them:

```r
install.packages(c("jsonlite", "pkgload"))
```

If one is missing, the step that needs it stops immediately, and the message
names the missing package.

Then, from the checkout's top folder, build the engine once (this needs Rust)
and load the package from its folder:

```bash
cargo build --release -p gog-cli
```

```r
pkgload::load_all("r-pkg/gog")
```

If you change gog's source, installing from the checkout checks more than
loading it does, because the install runs `configure`, which puts the engine
inside the installed package:

```bash
R CMD INSTALL r-pkg/gog
```

The R package looks for the engine in five places. In order: the `GOG_CLI_PATH`
environment variable, a `target/` build in a surrounding checkout, the copy
inside the installed package, `gog-cli` on your `PATH`, and last a search upward
from the working directory. A checkout is searched before the bundled copy on
purpose. Inside a checkout the bundled engine is a copy an earlier install left
behind. Using it would draw with an old engine while you are changing the
engine's source.

## What comes next

R's sentence is the one every chapter before this has written.
[Python](python.qmd) writes the same sentence with `col.gdp` where R writes a
bare name, and draws the same picture, byte for byte.

