# Comparing categories

```{r}
#| include: false
source("../R/setup.R")
```

How do many categories compare, and how did their order change over time?
[Many groups](groups.qmd) splits one plot by a few groups, with color or with
panels. This chapter compares one value for each of many categories. A ranked
bar chart answers the first question, and the chapter starts from that chart
and changes a few atoms at a time. Each change draws a chart that has a name of
its own: the Cleveland dot plot, the lollipop chart, the dumbbell chart, the
slope chart and the bump chart. All of them show twelve countries of East,
Southeast and South Asia, and how long their people lived, from the gapminder
data.

## Which country lives longest?

The question is which of the twelve countries had the longest life expectancy
in 2007. The ranked bar chart is the usual answer, and [Which group is
biggest?](groups.qmd#which-group-is-biggest) shows how `order()` sorts the
bars. `gm_twelve_2007` holds the twelve countries' life expectancy in 2007, one
row per country:

```{r}
#| echo: false
peek(gm_twelve_2007)
```

```{r}
data(gm_twelve_2007) + bar + y(country) + x(life) + order(life, desc = TRUE) +
  title("Life expectancy in 2007")
```

*"Given the gapminder twelve in 2007: bars, y is country, x is life, ordered by
life, largest first."*

Japan is first, at 82.6 years, and India is last, at 64.7. The country names
are long, so they go on `y`, where each one is written in full on one line.
Every bar starts at zero, because a bar shows its value by its length. So all
twelve bars end in the right part of the panel, between 64 and 83 years.

## What if the differences are small?

In the bar chart, most of the twelve values are between 70 and 80 years. When
every bar is more than 60 years long, a difference of one or two years is hard
to see. A **Cleveland dot plot** puts one point at each value instead. It is
named after the statistician William Cleveland, who recommended it
[@cleveland1984]. Write `point` in place of `bar` ([Point](../marks/point.qmd)):

```{r}
data(gm_twelve_2007) + point + y(country) + x(life) + order(life, desc = TRUE) +
  title("The same values as dots")
```

*"Given the gapminder twelve in 2007: points, y is country, x is life, ordered
by life, largest first."*

A point shows its value by its position, not by a length, so the axis does not
need to start at zero. It now covers only the values, from about 65 to 83
years, and the twelve points spread across the whole panel. China and Malaysia
differ by 1.3 years, and the gap between their points is now easy to see.

## What if the length matters too?

A bar's length says how large a value is, not only where it ranks. The
Cleveland dot plot shows each value by a position only, with no length. A
**lollipop chart** shows both: a thin stem from zero to the value, and a dot at
the value. Add one layer to the Cleveland dot plot's sentence:
`interval * bounds(zero, life)` draws each stem as an interval between two
columns ([Interval](../marks/interval.qmd)). The table has no column of zeros,
so add one before you draw:

```{r}
#| echo: false
spellings(
  r = r"(gm_twelve_2007$zero <- 0

data(gm_twelve_2007) + y(country) + x(life) + order(life, desc = TRUE) +
  interval * bounds(zero, life) + style(caps = FALSE) + point +
  title("A stem from zero, and a dot"))",
  python = r"(gm_twelve_2007["zero"] = [0] * len(gm_twelve_2007["life"])

(data(gm_twelve_2007) + y(col.country) + x(col.life) + order(col.life, desc = True) +
  interval * bounds(col.zero, col.life) + style(caps = False) + point +
  title("A stem from zero, and a dot")))",
  julia = r"(gm_twelve_2007["zero"] = zeros(length(gm_twelve_2007["life"]))

data(gm_twelve_2007) + y(:country) + x(:life) + order(:life, desc = true) +
  interval * bounds(:zero, :life) + style(caps = false) + point +
  title("A stem from zero, and a dot"))",
  js = r"(gm_twelve_2007.zero = gm_twelve_2007.life.map(() => 0);

plot(data(gm_twelve_2007), y(col.country), x(col.life),
  order(col.life, { desc: true }),
  layer(interval, bounds(col.zero, col.life)), style({ caps: false }),
  point, title("A stem from zero, and a dot"));)"
)
```

```{r}
#| echo: false
# The plot the four tabs above describe.
gm_twelve_2007$zero <- 0

data(gm_twelve_2007) + y(country) + x(life) + order(life, desc = TRUE) +
  interval * bounds(zero, life) + style(caps = FALSE) + point +
  title("A stem from zero, and a dot")
```

*"Given the gapminder twelve in 2007: y is country, x is life, ordered by life,
largest first, intervals from zero to life, and also points."*

The stems start at zero, as the bars did, and the dots mark the values. An
interval draws a short crossbar, its cap, at each end, and
`style(caps = FALSE)` leaves both out, so each stem is a plain line. The axis
starts at zero again, so the 1.3 years between China and Malaysia are as hard to
see as in the bar chart. Use the lollipop chart when the size of each value
matters more than the small gaps between values.

## How did each country change?

A ranking for one year does not show how each country's value changed over
time. `gm_twelve_ends` holds the same twelve countries in 1952 and in 2007, two
rows each. Its `year` names the two ends of the period and measures nothing, so
the table keeps it as text ([Labels that look like
numbers](../book-data.qmd#labels-that-look-like-numbers)):

```{r}
#| echo: false
peek(gm_twelve_ends)
```

A **dumbbell chart** joins each country's two values with a line. It keeps the
lollipop chart's two layers and changes the stem's transform.
`interval * range` draws one line for each country, from its smaller value to
its larger one. `point + color(year)` marks the two ends:

```{r}
data(gm_twelve_ends) + y(country) + x(life) + order(life, desc = TRUE) +
  interval * range + style(caps = FALSE) +
  point + color(year) +
  title("1952 and 2007 on one line")
```

*"Given the gapminder twelve in 1952 and 2007: y is country, x is life, ordered
by life, largest first, intervals derived by range, and also points colored by
year."*

Every country lived longer in 2007 than in 1952, so every orange point is to the
right of its blue one. The longest lines belong to the countries that gained
the most: Vietnam gained 33.8 years, Indonesia 33.2 and Korea 31.2. Japan was
already first in 1952 and gained 19.6. `order(life, desc = TRUE)` sorts the
countries by their two values added together, which is why Korea is 5th here
but 4th in 2007 alone.

`color(year)` comes after `point`, so it colors only the points ([Encoding
scope](../encoding-scope.qmd#channels-plot-scoped-or-layer-scoped)). Written
before the marks, it would also split each country's range by year. A range
over a single value has no length, so the lines would disappear.

## What if the change matters most?

The dumbbell chart shows each change as a length along one axis. A **slope
chart** puts the two years on an axis of their own. Each change then becomes
the slope of a line, and two lines that cross belong to two countries whose
order changed. The same three columns move: `year` from `color` to `x`, `life`
from `x` to `y`, and `country` from `y` to `color` ([Line](../marks/line.qmd)):

```{r}
data(gm_twelve_ends) + x(year) + y(life) + color(country) + line + point +
  title("The two years as an axis")
```

*"Given the gapminder twelve in 1952 and 2007: x is year, y is life, color by
country, lines, and also points."*

`year` is text, so the axis has two slots, 1952 and 2007. `color(country)`
gives each country its own line, which joins that country's two values.
Korea's line crosses four others. Korea was 8th in 1952, below Taiwan,
Thailand, Malaysia and the Philippines, and 4th in 2007, above all four.

## How did the order change over the years?

A slope chart has two years. A **bump chart** follows the order through every
year: it draws each country's rank instead of its value. `gm_twelve` holds the
twelve years from 1952 to 2007, five years apart, with each country's rank in
its year, from `1st` to `12th`:

```{r}
#| echo: false
peek(gm_twelve)
```

The ranks were computed from `life` before the table was written. gog draws a
rank, but it does not compute one, so you compute it in your host language.
Here `rank` is text with a declared order, so its values sit on a categorical
axis with `1st` at the top. The sentence is the slope chart's, with `gm_twelve`
as the table, `rank` in place of `life` and `path` in place of `line`. A `path`
joins each country's points in the order of the rows, and the table lists the
years in order ([Path](../marks/path.qmd#what-line-draws-instead)):

```{r}
data(gm_twelve) + x(year) + y(rank) + color(country) + path + point +
  title("Rank in every year")
```

*"Given the gapminder twelve: x is year, y is rank, color by country, paths, and
also points."*

Japan is 1st and Hong Kong is 2nd in every year. Korea rises from 8th in 1952
to 4th in 2002 and stays there. China jumps from 10th in 1962 to 5th in 1972,
and then drops to 8th by 1997. The countries take the same colors as in the
slope chart, because both tables list them in the same order.

## What this section refuses

A bump chart looks like a line chart, so `line` is the mark you might try
first. A `line` needs a continuous `y`, and a rank stored as text is
categorical, so the same sentence with `line` is refused:

```{r}
#| error: true
data(gm_twelve) + x(year) + y(rank) + color(country) + line
```

The message says that `y` on a `line` needs a continuous (numeric) column. It
suggests the [profile
plot](../marks/line.qmd#a-category-on-the-domain-the-profile), which puts the
category on `x`, but a bump chart needs the rank on `y`. A `path` has no such
rule: it joins its points in the order of the rows, whatever kind of column `y`
is.
