# Many groups

```{r}
#| include: false
source("../R/setup.R")
```

How do several groups compare? There are two ways to split the picture. You can
split the marks *inside* one panel, with `color` or `group`. You can split the
plot *into* panels, with `|` or `/`. The recipes below use both, and the last
two also put the groups in an order you choose.

## How do the groups' distributions compare?

The first route is to split the marks. The iris flowers table holds three
species, and every row carries a petal length. The question is whether those
lengths overlap or sit apart. An overlaid histogram answers it. `bin` cuts one
set of intervals from all the rows, so every species is counted into the same
intervals. Each species is drawn as a translucent fill under its own solid
outline:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."* Setosa sits apart, sharing no bin with the other two. Versicolor and
virginica overlap near 5, where their fills blend into a darker band. The solid
outline is what keeps each species readable there.
Owning chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd),
[Channels](../channels.qmd).

## What if the overlap gets worse?

Three translucent fills are still readable. Add more groups, or move them
closer together, and the overlaps blend into one dark band where no single fill
can be followed. Drop the fills entirely: `step * bin` draws each group's
outline as an unfilled staircase, and an outline covers only the thin line it
draws:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species."*

Only the mark changed, from `bar` to `step`. Each group is now one line rather
than a filled area, so far less of the plot is covered.
[Step](../marks/step.qmd) owns the staircase.

## Grouping without a legend

Both recipes above named their groups with `color`, and a legend decoded the
names. Sometimes the grouping matters and the names do not: you want the *shape*
of many series, not a legend naming each. `group` divides the rows into one line
per country. It binds no color, so there is nothing for a legend to decode:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country)
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country."*

Five countries, five lines, no legend. What you read is the shape the five lines
make together, not any one country's line. Bind [`color`](../channels.qmd)
instead when the reader needs to know which is which.

## What if each group needs its own axes?

gapminder 2007 holds five continents, and overlaying them on one histogram would
draw five fills on one axis. The first two recipes had only three groups to
separate. When each group needs its own pair of axes,
split the plot into panels instead of splitting the marks. The bins and the
scale stay shared, and each panel counts only its own rows, so the panels can be
compared:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

*"Given gapminder 2007: bars derived by bin, x is life, split into panel columns
by continent."*

Each panel holds one continent's distribution of life expectancy. Compare the
panels one at a time. Africa's bars
sit at the low end of the axis. Europe's bars sit at the high end. Bins and the
count axis are shared, so equal heights mean equal counts.
[Faceting](../faceting.qmd) owns the rules, including why the cut is shared when
the count is not.

## What if two columns group the rows?

Continent is one grouping and year is another. You want each continent drawn
in 1957 and again in 2007, to see what changed between the two years. `|`
splits into panel columns, `/` into panel rows; written together they cross
into a grid. Every combination gets a panel. The two years are taken from
`gm_all`, the whole gapminder table. Subsetting is the host language's job, so
it happens in R, and the other three bindings do it their own way:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
gm_years <- gm_all[gm_all$year %in% c(1957, 2007), ]
gm_years$period <- as.character(gm_years$year)
```

What follows is the sentence, and it has four spellings like every other:

```{r}
data(gm_years) + point + x(gdp, scale = "log") + y(life) |
  facet(continent) / facet(period)
```

*"Given the gm years table: points, x is gdp on a log scale, y is life, split
into panel columns by continent and panel rows by period."*

Each column holds one continent, and its two panels are fifty years apart. Read
a column downward to see what changed. `year` is a number,
and a facet variable has to name its panels, so `period` holds the same years as
a category. That is what the [Over time](over-time.qmd) refusal is about.

## Which group is biggest?

The `medals` table lists five countries, and the question is which of them won
the most gold. A comparison is easiest to read when it is sorted, and a column
of plain text carries no order of its own. `order()` is how you give it one:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold) + order(gold, desc = TRUE)
```

*"Given the medals: bars, x is country, y is gold, ordered by gold, largest
first."*

**One atom away.** `desc = FALSE`, the default, sorts smallest first. `order()`
names `gold` here, not the category column `country`. You name the column that
holds the values, and the categories are arranged to follow it. It overrides a
[declared order](../data.qmd) as well.

## Where does a population pyramid come from?

Two cities each have a population count in every age band, and the question is
how the two age structures compare. The chart drawn for that question is the
**population pyramid**. It is a named chart type, and no atom in the kernel
carries that name. You never ask for a pyramid; you describe one.

The `census` table has three columns: an `age` band, a `city` (Busan or Seoul),
and a `population`. Each city has 18 bands, so the table holds 36 rows. The
numbers are invented; the columns are the ones a census would hold.

```{r}
#| echo: false
peek(census)
```

The mirrored shape comes from one minus sign. Busan's populations are negated,
so Busan's bars run to the left of zero. That is arithmetic, so it happens in
the table, before the sentence:

```{r}
mirrored <- census
mirrored$population <- ifelse(mirrored$city == "Busan",
                              -mirrored$population, mirrored$population)
```

```{r}
#| echo: false
peek(mirrored)
```

The arithmetic is written in R, because arithmetic is the host language's job,
and the other three bindings negate a column the same way. What follows is the
sentence, and it has four spellings like every other:

```{r}
data(mirrored) + bar + y(age) + x(population) + color(city) +
  order(age, desc = TRUE)
```

*"Given the mirrored table: bars, y is age, x is population, color by city,
ordered by age, largest first."*

No new atom, and none was needed. The pyramid comes from three rules the grammar
already states. First, orientation is read off the bindings, so a category on
`y` makes the bars horizontal. Second, a bar runs from the zero baseline to its
value, in whichever direction that value lies, so a negative bar is not a
special case. It is the same rectangle, pointing the other way. Third, `stack`
is something you ask for, never something the engine does on its own. Nothing
here asks for it, so two bars in one slot sit on opposite sides of zero.
There is no `coord_flip()` as in ggplot2, and no `barh()` as in matplotlib,
because either would be a second way to say one thing.

`order(age, desc = TRUE)` is the atom from the section above, in its other form.
There it named `gold`, a different column from the one on the axis. Here it
names the category column itself, so it asks for that column's own order and
reverses it. gog puts the first category at the top when the category is on `y`,
so age 0 would sit at the top. A pyramid needs its youngest band at the bottom,
and `desc = TRUE` reverses the order.

Now look at the left half of the population axis. It reads -40K, not 40K.

Many versions of this chart replace each tick's text with its absolute value.
The minus sign that placed one city's bars then disappears. gog leaves every
tick reading the number it drew at, and this plot shows why. Relabeled, the axis
carries two different ticks both reading 40K, and only a reader who knows the
convention can tell which is which. That reader never needed the absolute
values. The reader who did is left with bars whose positions no longer match the
numbers on the axis. gog gives you no way to write your own tick text, and
[Coverage](../coverage.qmd) lists that among the capabilities with no decision
recorded. The paragraph above is the reason gog has not added one.

## What this section refuses

`shape` separates categories on a point, one glyph per category, and you might
expect the same on a bar. A bar is a rectangle; it has no glyph to vary, so
`shape` on a `bar` is refused outright. One table says what every mark accepts,
and it has no per-mark exceptions:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + shape(country)
```
