# Many groups

```{r}
#| include: false
source("../R/setup.R")
```

How do several groups compare? The comparison questions: several groups, one
framework. Two families of answer: split the marks *inside* one panel (`color`,
`group`), or split the plot *into* panels (`|`, `/`). The recipes run from one
to the other.

## How do the groups' distributions compare?

The overlaid histogram: `bin` cuts one set of intervals for everyone, so
the groups sit on shared edges, each a translucent fill under its own
solid outline:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Bars derived by bin: x is petal length, color by species."* The
outline is what keeps each species' silhouette readable where the fills
overlap.

## And when the overlap gets worse?

Drop the fills entirely: `step * bin` traces each group's silhouette as
an unfilled staircase, so nothing can hide anything:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species)
```

One mark swapped, nothing else. [Step](../marks/step.qmd) owns the
staircase.

## Split, but don't decorate

Sometimes the grouping matters and the identities do not: you want the
*shape* of many series, not a legend naming each. `group` splits
without coloring:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country)
```

Five countries, five lines, no legend: the shape of the bundle is the
answer. Bind [`color`](../channels.qmd) instead when the reader needs to
know which is which.

## One panel per group

When each group deserves its own pair of axes, multiply the plot
instead of the marks. Statistics run *within* each panel; the scale stays
shared so panels stay comparable:

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

Five distributions on one common axis, cut into one common set of bins and
counted continent by continent. Compare the shapes column by column: Africa's
bars sit left, Europe's right, and because every bar covers the same span of
years the heights mean the same thing in each panel.
[Faceting](../faceting.qmd) owns the rules, including why the cut is shared
when the count is not.

## Two grouping variables at once

`|` splits into panel columns, `/` into panel rows; written together they
cross into a grid. Every combination gets a panel:

```{r}
gm_years <- gm_all[gm_all$year %in% c(1957, 2007), ]
gm_years$period <- as.character(gm_years$year)

data(gm_years) + point + x(gdp, scale = "log") + y(life) |
  facet(continent) / facet(period)
```

Fifty years of change, read down each continent's column. The `period`
column exists because a facet variable names panels, which is also what
the [Over time](over-time.qmd) refusal is about.

## Which group is biggest?

A comparison is easiest to read when it is sorted, and no bare text column can
say "order me by value"; that is what `order()` is for:

```{r}
data(medals) + bar + x(country) + y(gold) + order(gold, desc = TRUE)
```

**One atom away.** `desc = FALSE` (the default) for ascending; naming a column
other than the axis's own, as here, also overrides a
[factor's declared levels](../data.qmd).

## Two groups, back to back

A **population pyramid** is a named chart type, and nothing in the kernel knows
its name. It follows from rules that were already there, for other reasons.

The `census` table holds 36 rows: an `age` band, a `city` (Busan or Seoul), and
a `population` for each of 18 bands per city. It is invented data, shaped like a
census.

The mirror comes from a minus sign: one city's populations are negated so its
bars grow the other way. That is arithmetic, so it happens in the table, before
the sentence:

```{r}
mirrored <- census
mirrored$population <- ifelse(mirrored$city == "Busan",
                              -mirrored$population, mirrored$population)
```

The arithmetic is written in R, because it is R's job, and the other three
bindings negate a column the same way. What follows is the sentence, and it has
four spellings like every other:

```{r}
data(mirrored) + bar + y(age) + x(population) + color(city) +
  order(age, desc = TRUE)
```

No new atom, and none was needed. Three rules the grammar already stated met
here on their own. Orientation is read off the bindings, so `y(age)` being a
category is what lays the bars down, and there is no `flip` atom because that
would be a second way to say one thing. A bar runs from the zero baseline to its
value in whichever direction that value sits, so a negative bar is not a special
case; it is the same rectangle pointing the other way. And `stack` is something
you ask for rather than something that happens, so two bars sharing one slot sit
back to back instead of piling up.

`order(age, desc = TRUE)` is the atom from the section above, doing exactly what
it did there. gog puts the first category at the top of a horizontal chart. A
pyramid needs its youngest band at the bottom instead.

Now look at the left half of that axis. It reads -40K, not 40K.

This chart is often drawn with the tick text replaced by its absolute value,
which hides the minus sign the bars were placed with. gog keeps the coordinates
it drew, and this is the clearest case for why. Relabeled, the axis carries two
different ticks both reading 40K, and only a reader who knows the convention
can see which is which. That reader did not need the help; the one who did has been
handed a plot that contradicts its own coordinates. Choosing your own tick text
is a capability nothing here has ruled on in either direction, and
[Coverage](../coverage.qmd) keeps it on the open list. This is the argument
against it.

## What this section refuses

A bar is a rectangle; it has no glyph to vary, so `shape` on a `bar` is
refused outright. The table of what each mark accepts is one table, with
no per-mark exceptions:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + shape(country)
```
