# Many groups

```{r}
#| include: false
source("../R/setup.R")
```

How do several groups compare? There are two ways to separate the groups. You
can split the marks *inside* one panel, with `color` or `group`. You can split
the plot *into* panels, with `|` or `/`. The recipes below use both. Two of them
also put the groups in an order you choose.

## How do the groups' distributions compare?

The first way is to split the marks. The `iris_flowers` table holds three
species, and every row carries a petal length. The question is whether those
lengths overlap or sit apart. An overlaid histogram answers it: one histogram
per species, drawn in one panel. `bin` cuts the axis into intervals, called
bins, and it cuts one set of bins from all the rows, so every species is counted
into the same bins. Nothing in the sentence below asks for transparency. The
engine draws each species as a translucent fill under its own solid outline, so
a fill that lies under another is still visible:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."*

Setosa sits apart, sharing no bin with the other two. Versicolor and virginica
overlap near 5 cm, where one fill lies over the other and the color looks
darker. The solid outline is what keeps each species readable there. Owning
chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd),
[Channels](../channels.qmd).

## What if the overlap gets worse?

Three translucent fills are still readable. With more groups, or with groups
closer together, the overlaps blend into one dark band, and you can no longer
see where one fill ends. Remove the fills entirely: `step * bin` draws each
group's outline as an unfilled staircase. A thin outline hides almost nothing
behind it:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species."*

Only the mark changed, from `bar` to `step`. Each group is now one outline
rather than a filled area, so far less of the plot is covered. Near 5 cm,
versicolor's outline steps down where virginica's steps up, and both lines stay
visible. [Step](../marks/step.qmd) owns the staircase.

## What if the names do not matter? {#grouping-without-a-legend}

Both recipes above gave each group its own `color`, and a legend matched each
color to a name. Sometimes the grouping matters and the names do not: you want
the *shape* of many series, not a legend naming each. `group` divides the rows
by country, and `line` draws one stroke for each. `group` binds no color, so
there is nothing for a legend to decode:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country)
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country."*

Five countries, five lines, no legend. All five lines end higher than they
start, and only one ever falls, between 1957 and 1962. Nothing on the plot says
which country that is: what you read is the shape the five lines make together.
Bind [`color`](../channels.qmd) instead when the reader needs each line's
country.

## What if each group needs its own panel?

The first two recipes had only three groups to separate. The `gapminder_2007`
table holds five continents, and on one histogram up to four of their fills
would lie in the same bin. Split the plot instead of the marks, so that each
continent gets a panel of its own. Each panel counts only its own rows:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

*"Given gapminder 2007: bars derived by bin, x is life, split into panel columns
by continent."*

Each panel holds one continent's distribution of life expectancy. Africa's 52
countries run from about 40 to 76 years, and 40 of them are below 60. All 30 of
Europe's countries are above 70. Bins and the count axis are shared, so equal
heights mean equal counts. [Faceting](../faceting.qmd) owns the rules,
including why the bins are shared while each panel counts only its own rows.

## What if two columns group the rows?

The recipe above split the plot by one column, `continent`. This recipe splits
the plot a second time. You want each continent drawn in 1957 and again in 2007,
to see what changed between the two years. `|` splits into panel
columns, `/` into panel rows; written together they cross into a grid. Every
combination gets a panel. The `gm_eras` table holds the rows for those two
years. It records each year twice, as the number `year` and as the text `era`:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + point + x(gdp, scale = "log") + y(life) |
  facet(continent) / facet(era)
```

*"Given the gapminder eras: points, x is gdp on a log scale, y is life, split
into panel columns by continent and panel rows by era."*

Each panel column holds one continent, and its two panels show 1957 and 2007,
fifty years apart. Read a panel column downward to see what changed. The plot
is split by `era`, not by `year`. A facet column must be categorical, because it
names its panels, and `year` holds numbers.
[Over time](over-time.qmd#what-this-section-refuses) shows the refusal you get
for `facet(year)`.

## Which group is biggest?

The `medals` table lists five countries, and the question is which of them won
the most gold. A comparison is easiest to read when it is sorted. gog draws the
categories of a column of text in the order their rows appear in the table, and
`order()` lets you choose another order:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold) + order(gold)
```

*"Given the medals: bars, x is country, y is gold, ordered by gold, smallest
first."*

The bars rise from left to right, from Germany's 17 gold medals to the USA's
46, so the country with the most gold comes last. `order()` names `gold` here,
not the categorical column `country`. You name the column that holds the
values, and the categories are sorted by those values.

**One atom away.** `desc = TRUE` puts the largest first. The table already lists
the countries from most gold to least, so with `desc = TRUE` the bars keep the
table's own order. A sentence with no `order()` draws the same plot. Writing the
order anyway keeps the plot sorted if a later version of the table lists the
countries in a different order.

`order()` overrides a
[declared order](../data.qmd#declaring-the-order-of-categories) as well. The
next recipe uses it another way: it sorts by a column's declared order, not by
size.

## Where does a population pyramid come from?

Two cities each have a population count in every age band, and the question is
how the two age structures compare. The chart drawn for that question is the
**population pyramid**. It is a named chart type, and no atom in the kernel
carries that name. You never ask for a pyramid; you describe one.

The `census` table has three columns: an `age` band, a `city` (Busan or Seoul),
and a `population`. `age` holds the band names as text, from 0 to 85, in a
declared order from youngest to oldest. Each city has 18 bands, so the table
holds 36 rows. The numbers are invented; the columns are the ones a census would
hold.

```{r}
#| echo: false
peek(census)
```

The mirrored shape comes from one minus sign. Busan's populations are negated,
so Busan's bars run to the left of zero. That is arithmetic, so it happens in
the table, before the sentence:

```{r}
mirrored <- census
mirrored$population <- ifelse(mirrored$city == "Busan",
                              -mirrored$population, mirrored$population)
```

```{r}
#| echo: false
peek(mirrored)
```

The negation above is shown in R only; Python, Julia and JavaScript each negate
a column in their own way. What follows is the sentence:

```{r}
data(mirrored) + bar + y(age) + x(population) + color(city) +
  order(age, desc = TRUE)
```

*"Given the mirrored table: bars, y is age, x is population, color by city,
ordered by age, largest first."*

No new atom, and none was needed. The pyramid comes from three rules the grammar
already states. First, the engine takes the orientation from the bindings, so a
category on `y` makes the bars horizontal. Second, a bar runs from the zero
baseline to its value, on whichever side of zero that value is, so a negative
bar is not a special case. It is the same rectangle, extending the other way.
Third, `stack` is something you ask for, never something the engine does on its
own. This sentence does not ask for it, so the two bars in one slot sit on
opposite sides of zero. If you know ggplot2's `coord_flip()` or matplotlib's
`barh()`, gog has no such word, because it would be a second way to say one
thing.

`order(age, desc = TRUE)` is the atom from the medals recipe, used in a second
way. There it named `gold`, a different column from the one on the axis. Here it
names the categorical column itself, so it asks for that column's own order: the
declared order, youngest first. `desc = TRUE` reverses it. gog puts the first
category at the top when the category is on `y`, so without `desc = TRUE` age 0
would sit at the top. A pyramid needs its youngest band at the bottom.

Now look at the left half of the population axis. It reads -40K, not 40K.

Many versions of this chart replace each tick's text with its absolute value.
The minus sign that placed one city's bars then disappears. gog keeps every
tick showing the number at its own position, and this plot shows why. If the
ticks were relabeled, the axis would carry two ticks reading 40K. Only a reader
who knows the convention would know that the left 40K marks the position -40K.
That reader could read -40K just as well, so the relabeling does not help. Any
other reader gets bars whose positions no longer match the numbers on the axis.
gog gives you no way to write your own tick text, and
[Coverage](../coverage.qmd#not-decided-either-way) lists that among the
capabilities with no decision recorded. These reasons argue against adding one.

## What this section refuses

Most recipes here separated groups with `color`, `group` or a panel. `shape`
separates categories too, with one glyph per category on a point, and you might
expect the same on a bar. A bar is a rectangle; it has no glyph to vary, so
`shape` on a `bar` is refused outright:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + shape(country)
```

The message says a bar has no shape feature, and tells you to remove `shape` or
to use a mark that has one. `point` is the only mark that does. One rule table
decides every pairing of a mark and a channel, and it makes no exception for any
mark. The [companion grid](../combinations.qmd#marks-and-channels) prints all of
it.
