# Many groups

```{r}
#| include: false
source("../R/setup.R")
```

How do several groups compare? There are two ways to answer. You can split the
marks *inside* one panel, with `color` or `group`. You can split the plot *into*
panels, with `|` or `/`. The recipes below use both.

## How do the groups' distributions compare?

Three species of iris share one petal length axis. The question is whether
their petal lengths overlap or sit apart. The overlaid histogram: `bin` cuts
one set of intervals for all three species, so the groups sit on shared
edges, each a translucent fill under its own solid outline:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."* The outline is what keeps each species' silhouette readable where the
fills overlap. Owning chapters: [Bar](../marks/bar.qmd),
[Transforms](../transforms.qmd), [Channels](../channels.qmd).

## What if the overlap gets worse?

Three translucent fills are still readable. More groups, or groups that sit
closer together, hide more of each fill. Drop the fills entirely:
`step * bin` traces each group's silhouette as an unfilled staircase, so
nothing can hide anything:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species."*

One mark swapped, nothing else. [Step](../marks/step.qmd) owns the
staircase.

## Grouping without a legend

Sometimes the grouping matters and the identities do not: you want the
*shape* of many series, not a legend naming each. `group` splits
without coloring:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country)
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country."*

Five countries, five lines, no legend. The shape the five lines make together is
the answer. Bind [`color`](../channels.qmd) instead when the reader needs to
know which is which.

## One panel per group

Five continents on one histogram would be five overlaid fills; the recipes
above had only three to separate. When each group needs its own pair of axes,
split the plot into panels instead of splitting the marks. Statistics run
*within* each panel; the scale stays shared so panels stay comparable:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

*"Given gapminder 2007: bars derived by bin, x is life, split into panel columns
by continent."*

Five distributions on one common axis, cut into one common set of bins and
counted continent by continent. Compare the panels one at a time. Africa's bars
sit at the low end of the axis. Europe's bars sit at the high end. Bins and the
count axis are shared, so equal heights mean equal counts.
[Faceting](../faceting.qmd) owns the rules, including why the cut is shared when
the count is not.

## Two grouping variables at once

Continent is one grouping and year is another. You want each continent drawn
in 1957 and again in 2007, to see what changed between the two years. `|`
splits into panel columns, `/` into panel rows; written together they cross
into a grid. Every combination gets a panel. The two years are selected in R,
because subsetting a table is the host language's job, and the other three
bindings do it their own way:

```{r}
gm_years <- gm_all[gm_all$year %in% c(1957, 2007), ]
gm_years$period <- as.character(gm_years$year)

data(gm_years) + point + x(gdp, scale = "log") + y(life) |
  facet(continent) / facet(period)
```

*"Given the gm years table: points, x is gdp on a log scale, y is life, split
into panel columns by continent and panel rows by period."*

Fifty years of change, read down each continent's column. `year` is a number,
and a facet variable has to name its panels, so `period` is `year` written as
text. That is what the [Over time](over-time.qmd) refusal is about.

## Which group is biggest?

The `medals` table lists five countries, and the question is which of them won
the most gold. A comparison is easiest to read when it is sorted, and a column
of plain text carries no order of its own. `order()` is how you give it one:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold) + order(gold, desc = TRUE)
```

*"Given the medals: bars, x is country, y is gold, ordered by gold, largest
first."*

**One atom away.** `desc = FALSE`, the default, sorts smallest first. `order()`
names `gold` here, not the category column `country`: that is how a category is
sorted by a value. It overrides a [declared order](../data.qmd) as well.

## Two groups, mirrored

Two cities each have a population count in every age band, and the question is
how the two age structures compare. The chart drawn for that question is the
**population pyramid**. It is a named chart type, and nothing in the kernel
knows its name. It follows from rules that were already there, for other
reasons.

The `census` table holds 36 rows: an `age` band, a `city` (Busan or Seoul), and
a `population` for each of 18 bands per city. It is invented data, shaped like a
census.

```{r}
#| echo: false
peek(census)
```

The mirror comes from a minus sign: one city's populations are negated so its
bars point the other way. That is arithmetic, so it happens in the table, before
the sentence:

```{r}
mirrored <- census
mirrored$population <- ifelse(mirrored$city == "Busan",
                              -mirrored$population, mirrored$population)
```

```{r}
#| echo: false
peek(mirrored)
```

The arithmetic is written in R, because it is R's job, and the other three
bindings negate a column the same way. What follows is the sentence, and it has
four spellings like every other:

```{r}
data(mirrored) + bar + y(age) + x(population) + color(city) +
  order(age, desc = TRUE)
```

*"Given the mirrored table: bars, y is age, x is population, color by city,
ordered by age, largest first."*

No new atom, and none was needed. The pyramid comes from three rules the grammar
already states. Orientation is read off the bindings, so a category on `y` lays
the bars down. There is no `coord_flip()` as in ggplot2, and no `barh()` as in
matplotlib, because either would be a second way to say one thing. A bar runs
from the zero baseline to its value, in whichever direction that value lies.
A negative bar is not a special case. It is the same
rectangle, pointing the other way. And `stack` is something you ask for rather
than something that happens, so two bars sharing one slot sit on opposite
sides of zero instead of stacking.

`order(age, desc = TRUE)` is the atom from the section above, doing exactly what
it did there. gog puts the first category at the top of a horizontal chart, so
age 0 would sit at the top. A pyramid needs its youngest band at the bottom, and
`desc = TRUE` reverses the order.

Now look at the left half of the population axis. It reads -40K, not 40K.

This chart is often drawn with the tick text replaced by its absolute value,
which hides the minus sign the bars were placed with. gog keeps the coordinates
it drew, and this is the clearest case for why. Relabeled, the axis carries two
different ticks both reading 40K, and only a reader who knows the convention can
see which is which. That reader did not need the help. The reader who did need
it is now given a plot that disagrees with its own axis. gog has no way to set
your own tick text, and [Coverage](../coverage.qmd) lists it among the
capabilities with no decision recorded. The paragraph above is the case against
adding one.

## What this section refuses

`shape` separates categories on a point, one glyph per category, and you might
expect the same on a bar. A bar is a rectangle; it has no glyph to vary, so
`shape` on a `bar` is refused outright. The table of what each mark accepts is
one table, with no per-mark exceptions:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + shape(country)
```
