# One variable

```{r}
#| include: false
source("../R/setup.R")
```

You have one column in front of you. What can it tell you? The questions it can
answer, and the sentence for each.

## How many of each?

A text column's first question. Counting rows is a derivation, not a
different kind of chart, so it is `bar` with `* count`:

```{r}
data(gapminder_2007) + bar * count + x(continent)
```

*"Bars derived by count: x is continent."* No `y()` is bound because the
count is invented, and the axis labels itself.

**One atom away.** Long category names? Put the categories on `y` and the
bars lie down, same sentence, positions swapped:

```{r}
data(gapminder_2007) + bar * count + y(continent)
```

The count still measures along the invented axis, which is now x.
Owning chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd).

## What share of the whole is each?

Swap `count` for `proportion` and the tallies become fractions that sum
to one:

```{r}
data(iris_flowers) + bar * proportion + x(species)
```

**One atom away.** Back to raw tallies: `* count`. Sorted by size:
add [`order()`](../data.qmd).

## How is a measurement distributed?

The continuous column's first question. `bin` cuts the axis into
intervals and counts rows into each; `bar` draws the counts touching,
because the intervals are adjacent:

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

**One atom away.** `bin` takes a count or a width when the default
guesses wrong. `bin(30)` for thirty bins, or intervals in the data's own
units:

```{r}
data(iris_flowers) + bar * bin(width = 0.5) + x(petal_length)
```

A `line * bin` draws the same counts as a frequency polygon. Owning
chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd).

## What is its smooth shape?

The binned view depends on where the bin edges fall; the density estimate
smooths that decision away:

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"A line derived by density: x is life."* Same question as the
histogram, one derivation over. Use both when the bumps matter. One atom tunes
the smoothing: `density(0.5)` sharpens the bumps, `density(2)` irons them out,
and `density(bandwidth = 1)` fixes the kernel width in years.

## Where does the distribution step?

The third rendering of `bin`: an unfilled staircase silhouette, tracing
bin tops without any fill. It is most useful when several distributions
share an axis, see [Many groups](groups.qmd), but reads cleanly alone
too:

```{r}
data(gapminder_2007) + step * bin + x(life)
```

Three marks (`bar`, `line`, `step`), one `bin`. The transform never
changed; only the geometry drawing its output did. That is the
[No Exceptions law](../design-laws.qmd) doing its quiet work.

## And when there are only thirty rows?

Every answer above summarizes: `bin` needs a bin width, `density` needs a
bandwidth, and both choices show more when there are fewer rows to
average over. With thirty numbers you can skip the choice and draw all
thirty. Add `stack` and a `point` piles one dot per observation:

```{r}
data(gm_europe) + point * bin * stack + x(life) +
  x_label("Life expectancy (years)") + title("Thirty countries, thirty dots")
```

*"Points derived by bin and stack: x is life."* The pile's height is a count, so
the tallest column can be read off the axis or simply counted, and no
observation has been averaged into anything. Compare the same thirty as
bars, where a single bin edge moving would redraw the shape:

```{r}
data(gm_europe) + bar * bin + x(life) +
  x_label("Life expectancy (years)") + title("The same thirty, summarized")
```

**One atom away.** A categorical column instead: `point * count * stack`
piles one dot per row in each category. Owning chapter:
[Transforms](../transforms.qmd).

## What this section refuses

A distribution question needs a continuous column. Ask `density` about a
category and the refusal names the type the estimate needs. The
categorical version of this question is the share recipe above:

```{r}
#| error: true
data(gapminder_2007) + line * density + x(continent)
```
