# One variable

```{r}
#| include: false
source("../R/setup.R")
```

You have one column. What can it tell you? The answer depends on the column's
type. A categorical column asks how many of each; a continuous column asks how
the values spread.

## How many of each?

Your column is categorical, one continent per country. The first question is
how many countries each continent holds. Counting rows is a derivation, not a
different kind of chart, so the sentence is `bar` with `* count`:

```{r}
data(gapminder_2007) + bar * count + x(continent)
```

*"Given gapminder 2007: bars derived by count, x is continent."* No `y()` is
bound because the
count is invented, and the engine labels the axis for you.

**One atom away.** Long category names? Move the categories to `y` and the bars
lie down. The sentence is otherwise the same.

```{r}
data(gapminder_2007) + bar * count + y(continent)
```

*"Given gapminder 2007: bars derived by count, y is continent."*

The count still measures along the invented axis, which is now x.
Owning chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd).

## What share of the whole is each?

The count above shows 52 countries in Africa. That number means little until
you know it is 52 out of 142. When the question is each continent's share of
the whole, swap `count` for `proportion`. The counts become fractions that sum
to one:

```{r}
data(gapminder_2007) + bar * proportion + x(continent)
```

*"Given gapminder 2007: bars derived by proportion, x is continent."*

**One atom away.** Back to raw counts: `* count`. [`order()`](../data.qmd)
reorders the categories by a column the table holds, never by the invented
share.

## How is a measurement distributed?

Your column is continuous, one petal length per flower. The first question is
where the values cluster and how far they spread. `bin` cuts the axis into
equal-width bins and counts the rows in each; `bar` draws them touching, so a
gap means a bin with no rows. The result is the **histogram**:

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

*"Given the iris flowers: bars derived by bin, x is petal length."*

**One atom away.** `bin` takes a count or a width when the default does not suit
the data. `bin(30)` asks for thirty bins; `bin(width = 0.5)` asks for bins half
a unit wide:

```{r}
data(iris_flowers) + bar * bin(width = 0.5) + x(petal_length)
```

*"Given the iris flowers: bars derived by bin at width 0.5, x is petal length."*

A `line * bin` is the frequency polygon: the same counts, joined through the bin
centers. Owning chapters: [Bar](../marks/bar.qmd),
[Transforms](../transforms.qmd).

## What is its smooth shape?

A histogram's shape depends on where its bin edges fall. A reader can mistake
that choice for a feature of the data. When the shape of the distribution
itself is the answer, the density estimate removes that decision:

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"Given gapminder 2007: a line derived by density, x is life."* The same
question the histogram answered, with one derivation changed. Draw both when the
number of peaks is the thing you want to show. One atom tunes the smoothing:
`density(0.5)` sharpens the peaks, `density(2)` flattens them, and
`density(bandwidth = 1)` fixes the kernel width in years. Owning chapters:
[Line](../marks/line.qmd), [Transforms](../transforms.qmd).

## How does the distribution look in steps?

A third mark on the same `bin`: a staircase outline, tracing the bin tops
with no fill. It is most useful when several distributions share an axis, as in
[Many groups](groups.qmd), and it reads well alone:

```{r}
data(gapminder_2007) + step * bin + x(life)
```

*"Given gapminder 2007: a step outline derived by bin, x is life."*

Two marks (`bar`, `step`) and one `bin`, with `line * bin` above making three.
The transform never changed. Only the geometry that draws its output changed.
That is the [No Exceptions law](../design-laws.qmd) doing its job. Owning
chapters: [Step](../marks/step.qmd), [Transforms](../transforms.qmd).

## When there are only thirty rows

Every answer above summarizes: `bin` needs a bin width, `density` needs a
bandwidth, and both choices matter more when there are fewer rows to
average over. With thirty numbers the bins can stay and every row still
shows. Add `stack` and a `point` piles one dot per observation:

```{r}
data(gm_europe) + point * bin * stack + x(life) +
  x_label("Life expectancy (years)") + title("Thirty countries, thirty dots")
```

*"Given gapminder Europe: points derived by bin and stack, x is life."* The
pile's height is a count, so the tallest pile can be read off the axis or
counted dot by dot. No row has been averaged into anything. Compare the same
thirty as bars, where each pile becomes one length and the rows disappear into
it:

```{r}
data(gm_europe) + bar * bin + x(life) +
  x_label("Life expectancy (years)") + title("The same thirty, summarized")
```

*"Given gapminder Europe: bars derived by bin, x is life."*

**One atom away.** A categorical column instead: `point * count * stack`
piles one dot per row in each category. Owning chapter:
[Transforms](../transforms.qmd).

## What this section refuses

A distribution question needs a continuous column. Give `density` a category and
the refusal names the column type it needs. The categorical version of this
question is the share recipe above:

```{r}
#| error: true
data(gapminder_2007) + line * density + x(continent)
```
