# One variable

```{r}
#| include: false
source("../R/setup.R")
```

You have one column. What can it tell you? The column's type decides the
question. A categorical column asks how many of each; a continuous column asks
how the values spread.

## How many of each?

In `gapminder_2007` the `continent` column is categorical, one continent per
country. The first question is how many countries each continent holds.
Counting rows is a derivation, not a different kind of chart. The sentence is
`bar` with `* count`:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * count + x(continent)
```

*"Given gapminder 2007: bars derived by count, x is continent."* No `y()` is
bound, because `count` invents the y column itself. The engine labels that axis
for you.

**One atom away.** When the category names are long, put them on `y` instead.
The bars are then horizontal, and the sentence is otherwise the same.

```{r}
data(gapminder_2007) + bar * count + y(continent)
```

*"Given gapminder 2007: bars derived by count, y is continent."*

The invented count axis has moved to x, so the bars are read from left to right.
Owning chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd).

## What share of the whole is each?

The count above shows 52 countries in Africa. A count alone cannot say whether
that is many or few, and the table holds 142 countries in all. When the question
is each continent's share of the whole, swap `count` for `proportion`. The
counts become fractions that sum to one:

```{r}
data(gapminder_2007) + bar * proportion + x(continent)
```

*"Given gapminder 2007: bars derived by proportion, x is continent."*

**One atom away.** Swap `proportion` back to `count` for the counts themselves.
To change which order the categories appear in, name the category column in
[`order()`](../data.qmd): here `order(continent)` sorts them alphabetically. The
share is invented by the transform, so `order()` has no column to read it from.

## How is a measurement distributed?

The two recipes above counted categories. A continuous column asks a different
question, so the table changes: in the iris flowers, `petal_length` is
continuous, one measurement per flower. The first question is where the values
cluster and how far they spread. `bin` cuts the axis into equal-width bins and
counts the rows in each. `bar` draws those counts as touching rectangles, so a
gap is a bin with no rows. The result is the **histogram**:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

*"Given the iris flowers: bars derived by bin, x is petal length."* The gap near
2.5 cm is one of those empty bins. No flower in the table has a petal length
between 1.9 and 3.0 cm, because one species is separated from the other two.

**One atom away.** `bin` takes a number of bins or a bin width when the
automatic choice hides the shape. `bin(30)` asks for thirty bins;
`bin(width = 0.5)` asks for bins 0.5 cm wide:

```{r}
data(iris_flowers) + bar * bin(width = 0.5) + x(petal_length)
```

*"Given the iris flowers: bars derived by bin at width 0.5, x is petal length."*

A `line * bin` is the frequency polygon: the same counts, joined through the bin
centers. Owning chapters: [Bar](../marks/bar.qmd),
[Transforms](../transforms.qmd).

## What is its smooth shape?

A histogram's shape depends on where its bin edges fall. A reader can mistake
that choice for a feature of the data. When the shape of the distribution itself
is the answer, `density` draws a curve with no bin edges at all. Back to the
gapminder countries, and their life expectancies:

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"Given gapminder 2007: a line derived by density, x is life."* This answers the
question the histogram answered, with one derivation changed. Draw both when the
number of peaks matters, because a histogram and a density curve can disagree
about it.

One argument controls the smoothing. `density(0.5)` is half as smooth as the
default and `density(2)` is twice as smooth. `density(bandwidth = 1)` sets the
bandwidth to one year instead, in the column's own units. Owning chapters:
[Line](../marks/line.qmd), [Transforms](../transforms.qmd).

## How does the distribution look in steps?

A third mark can draw the same `bin`. `step` traces the bin tops as a staircase
outline, with no fill. It is most useful when several distributions share one
axis, as in [Many groups](groups.qmd), because unfilled outlines cannot hide
each other:

```{r}
data(gapminder_2007) + step * bin + x(life)
```

*"Given gapminder 2007: a step outline derived by bin, x is life."*

`bar` and `step` have now drawn the same `bin`, and the `line * bin` named above
would make three. The transform never changed. Only the geometry that draws its
output changed. That is the [No Exceptions law](../design-laws.qmd): a transform
behaves the same on every mark. Owning chapters: [Step](../marks/step.qmd),
[Transforms](../transforms.qmd).

## When there are only thirty rows

Every answer above replaces the rows with a summary. `bin` needs a bin width,
and `density` needs a bandwidth. With few rows, either choice can suggest
structure the data does not have. The `gm_europe` table holds the thirty
European countries of 2007, and thirty rows are few enough to draw one by one.
`bin` still cuts the axis, and adding `stack` piles one `point` per row. The
result is the **dot plot**:

```{r}
#| echo: false
peek(gm_europe)
```

```{r}
data(gm_europe) + point * bin * stack + x(life) +
  x_label("Life expectancy (years)") + title("Thirty countries, thirty dots")
```

*"Given gapminder Europe: points derived by bin and stack, x is life."* The
pile's height is a count. You can read the tallest pile off the axis, or count
its dots one by one. No row has been replaced by a summary. Compare the same
thirty countries as bars, where each pile becomes one bar and the individual
rows are no longer drawn:

```{r}
data(gm_europe) + bar * bin + x(life) +
  x_label("Life expectancy (years)") + title("The same thirty, summarized")
```

*"Given gapminder Europe: bars derived by bin, x is life."*

**One atom away.** A categorical column instead: `point * count * stack`
piles one dot per row in each category. Owning chapter:
[Transforms](../transforms.qmd).

## What this section refuses

A distribution question needs a continuous column. Give `density` a category and
it refuses. The message names the kind of column `density` needs, and the
categorical version of this question is the share recipe above:

```{r}
#| error: true
data(gapminder_2007) + line * density + x(continent)
```
