# One variable

```{r}
#| include: false
source("../R/setup.R")
```

You have one column. What can it tell you? Which question you can ask depends on
the column's type. With a categorical column you ask how many rows fall in each
category. With a continuous column you ask how the values spread. When the table
is small, there is a third question: whether to summarize the rows at all.

## How many of each?

In `gapminder_2007` the `continent` column is categorical, one continent per
country. The first question is how many countries each continent holds.
Counting rows is a derivation, not a different kind of chart. The sentence is
`bar` with `* count`:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * count + x(continent)
```

*"Given gapminder 2007: bars derived by count, x is continent."* No `y()` is
bound, because `count` invents the y column itself. The engine labels that axis
for you.

**One atom away.** When the category names are long, put them on `y` instead.
The bars are then horizontal, and the sentence is otherwise the same.

```{r}
data(gapminder_2007) + bar * count + y(continent)
```

*"Given gapminder 2007: bars derived by count, y is continent."*

The invented count axis has moved to x, so each bar's length is read from left
to right. Owning chapters: [Bar](../marks/bar.qmd),
[Transforms](../transforms.qmd).

## What share of the whole is each?

In both count plots above, Africa has the longest bar: 52 countries. A count
alone cannot say whether 52 is many or few. The table holds 142 countries in
all, so Africa holds 52 of the 142, a share of about 0.37. When the question is
each continent's share of the whole, swap `count` for `proportion`. The counts
become fractions that sum to one:

```{r}
data(gapminder_2007) + bar * proportion + x(continent)
```

*"Given gapminder 2007: bars derived by proportion, x is continent."*

**One atom away.** Swap `proportion` back to `count` for the counts themselves.
The bars above follow the order the continents first appear in the table. To
draw them in another order, use
[`order()`](../data.qmd#declaring-the-order-of-categories), which sorts them by
the values in a column of the table: `order(continent)` sorts them
alphabetically, by their own names. The transform invents the share,
so the table has no column of shares, and gog refuses `order(proportion)`.
Owning chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd).

## How is a measurement distributed?

The two recipes above counted categories. With a continuous column you ask a
different question. In the `iris_flowers` table, `petal_length` is continuous,
one measurement per flower. The first question is where the values
cluster and how far they spread. `bin` cuts the axis into equal-width bins and
counts the rows in each. `bar` draws those counts as touching rectangles, so a
gap is a bin with no rows. The result is the **histogram**:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

*"Given the iris flowers: bars derived by bin, x is petal length."* The gap near
2.5 cm is an empty bin. No flower in the table has a petal length between 1.9
and 3.0 cm. One species has much shorter petals than the other two, and the
empty bin is part of the space between them.

**One atom away.** When the automatic choice makes the shape hard to read, give
`bin` a number of bins or a bin width. `bin(30)` asks for thirty bins;
`bin(width = 0.5)` asks for bins 0.5 cm wide:

```{r}
data(iris_flowers) + bar * bin(width = 0.5) + x(petal_length)
```

*"Given the iris flowers: bars derived by bin at width 0.5, x is petal length."*

The width 0.5 suits these lengths: they are recorded to one decimal place, so
every bin holds exactly five possible values.
[Choosing a width](../transforms.qmd#choosing-a-width) explains how to pick a
width for your own column.

A `line * bin` is the **frequency polygon**: a line through the same counts,
placed at the center of each bin. Owning chapters: [Bar](../marks/bar.qmd),
[Transforms](../transforms.qmd).

## What is its smooth shape?

A histogram's shape depends on the choice of bin edges. A reader can mistake
that choice for a feature of the data. When the shape of the distribution itself
is the answer, `density` draws a curve with no bin edges at all. Here is the
histogram's own column again, the petal lengths:

```{r}
data(iris_flowers) + line * density + x(petal_length)
```

*"Given the iris flowers: a line derived by density, x is petal length."* This
answers the question the histogram answered, for the same column, with
`density` in place of `bin` and a `line` to draw it. The two groups of petals
are now two peaks, and the valley between them lies over the empty bin. Draw
both when the number of peaks matters, because a histogram and a density curve
can show a different number of peaks.

The curve starts below zero, although no petal is shorter than 1 cm. A density
spreads each row into a small smooth bump, and the bumps of the shortest petals
reach past zero.

The width of each bump is the **bandwidth**, and it sets how smooth the curve
is. A number written alone multiplies the default bandwidth: `density(0.5)` is
half as smooth as the default and `density(2)` is twice as smooth.
`density(bandwidth = 0.5)` sets the bandwidth itself instead, in the column's
own unit, so here it is half a centimeter. Write one of the two forms, never
both. Owning chapters:
[Line](../marks/line.qmd), [Transforms](../transforms.qmd).

## How does the distribution look as a staircase?

The histogram drew `bin` with `bar`. `step` can draw it too, tracing the bin
tops as a staircase outline with no fill. It is most useful when several
distributions share one axis, as in [Many groups](groups.qmd), because unfilled
outlines do not cover each other. This plot draws the histogram's distribution
again, so you can compare the two marks directly:

```{r}
data(iris_flowers) + step * bin + x(petal_length)
```

*"Given the iris flowers: a step outline derived by bin, x is petal length."*

This is the histogram's sentence with one atom changed, `step` in place of
`bar`. The bins are the same, the empty one included, and only the mark that
draws them changed. With the `line * bin` named earlier, that is three marks on
one transform. That is the [No Exceptions
law](../design-laws.qmd#law-2-no-exceptions): a transform behaves the same on
every mark. Owning chapters: [Step](../marks/step.qmd),
[Transforms](../transforms.qmd).

## When there are only thirty rows

Every answer above replaces the rows with a summary. `bin` needs a bin width,
and `density` needs a bandwidth. With few rows, either choice can suggest
structure the data does not have. The `gm_europe` table holds the thirty
European countries of 2007, and thirty rows are few enough to draw one by one.
`bin` still cuts the axis, and adding `stack` piles one `point` per row. The
result is the **dot plot**:

```{r}
#| echo: false
peek(gm_europe)
```

```{r}
data(gm_europe) + point * bin * stack + x(life) +
  x_label("Life expectancy (years)") + title("Thirty countries, thirty dots")
```

*"Given gapminder Europe: points derived by bin and stack, x is life."* Each
pile's height is a count. You can read the tallest pile's height from the axis,
or count its dots one by one. No row has been replaced by a summary. Compare the
same thirty countries as bars, where each pile becomes one bar and the individual
rows are no longer drawn:

```{r}
data(gm_europe) + bar * bin + x(life) + x_label("Life expectancy (years)") +
  title("The same thirty countries, summarized")
```

*"Given gapminder Europe: bars derived by bin, x is life."*

**One atom away.** For a categorical column, `point * count * stack` piles one
dot per row in each category. Owning chapter:
[Transforms](../transforms.qmd).

## What this section refuses

Every recipe above that used `density` or `bin` gave it a continuous column.
Suppose you give `density` the `continent` column instead. `density` estimates
a continuous distribution, so gog refuses the sentence. For a categorical
column, use the share recipe in [What share of the whole is
each?](#what-share-of-the-whole-is-each). The message says which kind of column
`density` needs, and it names the same fix, `bar * proportion`:

```{r}
#| error: true
data(gapminder_2007) + line * density + x(continent)
```
