# Over time

```{r}
#| include: false
source("../R/setup.R")
```

How did it change over time? Time is not a chart type in this grammar; it is a
column, and you choose which channel carries it. On `x` it makes the familiar
time series, and every recipe below puts it there.

## How did it change?

The `actuals` table holds one sales figure per year. The question is whether
sales rose or fell, and a `line` with `year` on `x` answers it:

```{r}
#| echo: false
peek(actuals)
```

```{r}
data(actuals) + line + x(year) + y(sales)
```

*"Given the actuals: a line, x is year, y is sales."* A `line` connects
the rows in x order; with years on x, that order is time.

**One atom away.** If the amount matters more than the rise and fall, swap
`line` for [`area`](../marks/area.qmd): it fills to zero, so the filled area is
the amount.

## How did several things change?

The `gapminder_asia` table holds five countries across the same years. One
`line` over all of them would join the five into a single stroke. A categorical
`color` splits the rows into one line per category and colors them at the same
time:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

**One atom away.** Too many categories for a legend to stay readable?
`group` splits without coloring; see [Many groups](groups.qmd).

## What about a date column?

The `six_weeks` table records orders by day, and its `day` column holds
calendar dates. The `year` above is a plain number. A column of calendar dates
*is* temporal, so you get a calendar axis without asking for one. Ticks land on
calendar boundaries, and weekly steps land on Mondays:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + line + x(day) + y(orders)
```

*"Given the six weeks table: a line, x is day, y is orders."*

You never have to write `scale = "time"`: the column's type already says it.
[Scales](../scales.qmd) owns the calendar's rules: three-month steps that read
as quarters, and a midnight tick labeled with the day's name.

## What happened, and what comes next?

The `actuals` table ends in 2023, and the `forecast` table holds the three
years after it. The two belong on one plot, with the actuals and the forecast
drawn as different marks. The first `data()` is the plot's table; a later one
applies to the mark written after it.

```{r}
#| echo: false
peek(forecast)
```

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points."* The line ends at 2023, the dots start at 2024, and each layer reads
its own rows. [Data](../data.qmd) owns multi-table plots.

## What this section refuses

You may want a scatter of wealth against life expectancy for each year, to
compare the years side by side. One panel per year is faceting, but a facet
variable *names* its panels, and `year` here is a number. The refusal says what
to do with a year you mean as a label:

```{r}
#| error: true
data(gapminder_asia) + point + x(gdp) + y(life) | facet(year)
```
