# Over time

```{r}
#| include: false
source("../R/setup.R")
```

How did it change over time? Time is not a chart type in this grammar; it is a
column, and you choose which channel carries it. On `x` it makes the familiar
time series; the recipes here work through what that unlocks.

## How did it change?

```{r}
data(actuals) + line + x(year) + y(sales)
```

*"Given the actuals: a line, x is year, y is sales."* A `line` connects
the rows in x order; with years on x, that order is time.

**One atom away.** If the quantity (not the path) is the story, swap in
[`area`](../marks/area.qmd): it fills to zero, so the ink is the amount.

## How did several things change?

A categorical `color` splits the rows into one line per category, then
colors them, the same one-two it performs everywhere:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

**One atom away.** Too many categories for a legend to earn its space?
`group` splits without coloring; see [Many groups](groups.qmd).

## What about real dates?

A `Date` column *is* temporal, so the axis reads as a calendar without
your saying anything. Ticks land on calendar boundaries, and weekly
steps land on Mondays:

```{r}
data(six_weeks) + line + x(day) + y(orders)
```

There is no `scale = "time"` to remember: the column's type already said
it. [Scales](../scales.qmd) owns the calendar's rules: months that read
as quarters, midnight borrowing the day's name.

## What happened, and what do we expect?

History in one table, expectation in another. Bind each layer to its own
table: each `data()` applies to the mark written directly after it.

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point
```

*"Given the actuals: x is year, y is sales, a line; then given the
forecast: points."* The forecast dots pick up exactly where the actual
line ends, and each layer read its own rows.
[Data](../data.qmd) owns multi-table plots.

## What this section refuses

One panel per year sounds like faceting, but a facet variable *names*
panels, and a number names nothing. The refusal says what to do with a
year you mean as a label:

```{r}
#| error: true
data(gapminder_asia) + point + x(gdp) + y(life) | facet(year)
```
