# Over time

```{r}
#| include: false
source("../R/setup.R")
```

How did a measurement change over time? Time is not a chart type in this
grammar. It is a column like any other, and you decide where it goes. Most
recipes here put it on `x`, which gives the familiar time series. One gives it
to `play` instead, to show where earthquakes happened, one week at a time. The
last section asks for one panel per year, which gog refuses.

## How did it change?

The `actuals` table holds one sales figure per year. The question is whether
sales rose or fell. A `line` with `year` on `x` answers it:

```{r}
#| echo: false
peek(actuals)
```

```{r}
data(actuals) + line + x(year) + y(sales)
```

*"Given the actuals: a line, x is year, y is sales."*

Sales rose over the five years, from 120 to 168, and fell only once, in 2021. A
`line` sorts the rows by `x` and joins each row's point to the next. With years
on `x`, the stroke runs from the earliest year to the latest. Owning chapter:
[Line](../marks/line.qmd).

**One atom away.** If the question is how large sales were in each year, not
whether they rose, swap `line` for [`area`](../marks/area.qmd). An `area` fills
from zero up to each value, so the height of the fill is the amount itself.
[Two variables](two-variables.qmd#how-much-was-there-at-each-time) draws that
plot on this table.

## How did several things change?

The recipe above needed only one line, because `actuals` holds one row per year.
The `gapminder_asia` table holds five countries, each with one row for the same
twelve years, 1952 to 2007. One `line` reads every row, so without `color` the
five countries join into a single stroke. A categorical `color` does two jobs:
it *splits* the rows into one line per category, and it gives each line its own
color:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

Five countries, five lines, and a legend that names them. Life expectancy rose
in all five countries from 1952 to 2007. Only China's line ever falls, from
50.5 years in 1957 to 44.5 in 1962. Owning chapters: [Line](../marks/line.qmd),
[Channels](../channels.qmd).

**One atom away.** When the lines must be kept apart but their names do not
matter, swap `color` for `group`. It splits the rows in the same way, draws
every line in one color, and adds no legend.
[Many groups](groups.qmd#grouping-without-a-legend) draws that plot.

## How did many things change?

The recipe above drew five countries, and each of the five lines can be
followed on its own. Thirty-three lines cannot: the lines cross each other, and no
reader can match 33 legend colors to 33 lines. The `gm_asia` table holds the 33
countries of Asia in the gapminder data, each with one row for the same twelve
years. In the gapminder data, Asia includes Western Asia, so Iraq, Israel and
Saudi Arabia are among the 33. A **heatmap** gives each country a row and each
year a column, and colors each cell by life expectancy:

```{r}
#| echo: false
peek(gm_asia)
```

```{r}
data(gm_asia) + zone * bin(width = 5) * mean + x(year) + y(country) +
  color(life) + order(life, desc = TRUE) + palette("cividis") +
  style(border_color = "white") + theme(legend = "top") +
  title("Life expectancy in Asia, 1952 to 2007")
```

*"Given the gm Asia table: zones derived by bin at width 5 and mean, x is year,
y is country, color by life, ordered by life, largest first, with the cividis
palette."*

A zone needs sides, and a year is a number, which has no width.
`bin(width = 5)` gives the years their sides. It cuts the axis into cells five
years wide, from 1950 to 2010, and each cell holds one of the table's twelve
years. `mean` then colors each cell by `life`. A cell holds one row, so its mean
is that row's own value. Leave out either atom and gog refuses the sentence, as
[Zone](../marks/zone.qmd#what-it-refuses) shows.

`order(life, desc = TRUE)` sorts the countries by their mean life expectancy
over all twelve years, with the longest lives at the top. `palette("cividis")`
runs from dark blue for short lives to yellow for long ones.
`style(border_color = "white")` draws a thin line around each cell, and
`theme(legend = "top")` puts the legend above the panel.

Read across a row for one country's change, and down a column to compare the
countries in one year.
Sorted this way, the colors run from dark at the bottom left (Afghanistan in
1952) to light at the top right (Japan in 2007). Life expectancy was higher in
2007 than in 1952 in every country, and Oman gained the most, from 38 years to
76. Life expectancy fell by more than five years from one column to the next in
only three countries. Each fall is a cell much darker than the cell to its left.
In China, it fell from 50.5 years in 1957 to 44.5 in 1962. In Cambodia, it fell
twice: from 45.4 years in 1967 to 40.3 in 1972, then to 31.2 in 1977. In Iraq,
it fell from 65.0 years in 1987 to 59.5 in 1992, and it was still below 65 in
2007. Owning chapters:
[Zone](../marks/zone.qmd#cells-cut-then-summarized-the-summary-heatmap),
[Channels](../channels.qmd).

**One atom away.** Leave out `order()`, and the countries keep the table's
order, which is alphabetical. Each cell keeps its color, but dark rows and light
rows are now mixed together. The pattern from dark at the bottom left to light
at the top right is gone.

## What about a date column?

The recipes above put a plain number on `x`. The `six_weeks` table holds one row
for each of 42 days, with the date, the number of orders, the weekday and the
week. Its `day` column holds calendar dates rather than plain numbers, so you
get a calendar axis without asking for one. Each tick falls at the start of a
day, a week, a month or a year, and a week starts on a Monday:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + line + x(day) + y(orders)
```

Orders climb from 20 to 33 in the first week, fall to 13 on March 22 and 23,
and end at 34 on April 11. The table starts on a Friday, March 1, so the first
tick is on the first Monday, March 4. After that there is a tick every Monday.

You never have to write `scale = "time"`, because the column's type already
says that it holds dates. A longer span of dates gets a longer step between
ticks: days, then weeks, then months, then years. A column of date-times also
stores the time of day, so over a span of hours its ticks are clock times, such
as 06:00. The midnight tick shows its date instead of `00:00`.
[Scales](../scales.qmd#time-a-date-column-gets-a-calendar-axis) owns the
calendar's rules.

## What happened, and what comes next?

The sales line in the first recipe stops at 2023, and the `forecast` table
holds the three years after it. Both tables are drawn on one plot, and a
different mark for each shows which values were measured and which are
forecast. The first `data()` names the table that every mark reads, unless a
mark is given another. A later `data()` gives its table to the one mark written
directly after it:

```{r}
#| echo: false
peek(forecast)
```

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points."*

The line stops at 2023 and the points start at 2024, because the line reads
only `actuals` and the points read only `forecast`. `x` and `y` are written
once, before both marks, so both marks use them. That works because `forecast`
has the same two columns, `year` and `sales`.
[Data](../data.qmd#the-plots-table-and-a-layers-table) owns plots that read
more than one table.

## Where did it happen, week by week?

The recipes above follow one measurement along `x`. Some questions are about
places instead: where did something happen, and when? The `quakes_2011` table
holds every earthquake of magnitude 5 or more that the US Geological Survey
recorded in 2011. That was the year of the magnitude 9.1 earthquake off the
coast of Tohoku, in Japan, on March 11.

The table is shaped for a sequence of weekly frames. `week` holds the first day
of each seven-day week, counted from January 1, so every week here starts on a
Saturday. A frame draws only its own rows, so an earthquake with a single row
would appear for one week and then vanish. Each earthquake therefore has four
rows. The first is in the week it struck, with `age` "this week". The other
three repeat it in each of the next three weeks, with `age` "1 week ago",
"2 weeks ago" and "3 weeks ago". The land comes from `world_borders`, drawn as
in the [Globe](../globe.qmd) chapter:

```{r}
#| echo: false
peek(quakes_2011)
peek(world_borders)
```

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  style(color = "#1c2636") +
  data(quakes_2011) + point + x(lon) + y(lat) + size(magnitude) + color(age) +
  palette(c("gold", "darkorange", "orangered", "darkred")) + play(week, speed = 3) +
  globe(turn = 150, tilt = 15) + theme(background = "#050a14", grid = "none")
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country; then
given quakes 2011: points, x is lon, y is lat, sized by magnitude, colored by age,
with a palette of gold, darkorange, orangered and darkred, played by week at
speed 3, on the globe."*

Fifty-three frames: 52 full weeks, and a last one that holds only December 31.
A typical week adds about 40 new earthquakes, most of them along the edges of
the Pacific. The frame for the week of March 5 adds 335, and 303 of them lie off
the coast of Japan. Twenty-six of them came in the two days before the magnitude
9.1 earthquake. The rest are the magnitude 9.1 earthquake itself and 276
**aftershocks**, the smaller earthquakes that follow a large one, in the 18
hours that remained in the week. Over the next three frames that cluster turns
from gold to dark red. Meanwhile new aftershocks keep arriving off Japan: 201 in
the following week, then 52, then 23.

The palette runs from gold to dark red in the declared order of `age`, so an
earthquake starts bright and darkens as it ages. The four rows were in the table
before gog read it, because `play` cannot keep a row visible for several frames
by itself. The same preparation works for any event that should stay visible for
several frames. Repeat its row once for each of those frames, and add a column
that says how old each copy is. Owning chapters: [Play](../play.qmd),
[Globe](../globe.qmd).

**One atom away.** Swap `globe(turn = 150, tilt = 15)` for `map()`, and the
whole earth shows at once, flattened. The earthquakes on the far side of this
globe, in the Americas and along the middle of the Atlantic, then come into
view. [Map](../map.qmd) owns the flat version.

## What this section refuses

The recipes so far put time on `x` or gave it to `play`. Suppose instead you
want one scatter of GDP per person and life expectancy for each year, with the
years side by side. One panel per year is what [faceting](../faceting.qmd)
does. A facet column *names* its panels, so it must be categorical, but `year`
holds numbers. gog refuses the sentence, and the message names two fixes:

```{r}
#| error: true
data(gapminder_asia) + point + x(gdp) + y(life) | facet(year)
```

The first fix turns `year` into text. The `gm_eras` table already holds each
year as text, in the column `era`.
[Many groups](groups.qmd#what-if-two-columns-group-the-rows) draws a plot split
by `era`, with one row of panels for each of the two years.
