# Over time

```{r}
#| include: false
source("../R/setup.R")
```

How did a measurement change over time? Time is not a chart type in this
grammar. It is a column, and you choose which channel carries it. Time on `x`
gives the familiar time series, and every recipe below puts it there.

## How did it change?

The `actuals` table holds one sales figure per year. The question is whether
sales rose or fell. A `line` with `year` on `x` answers it:

```{r}
#| echo: false
peek(actuals)
```

```{r}
data(actuals) + line + x(year) + y(sales)
```

*"Given the actuals: a line, x is year, y is sales."* A `line` connects the
rows in x order. With years on `x`, the stroke runs from the earliest year to
the latest.

**One atom away.** If the quantity matters more than the rise and fall, swap
`line` for [`area`](../marks/area.qmd). The fill is the quantity, so it starts
from zero. [Two variables](two-variables.qmd) draws that plot on this table.

## How did several things change?

The recipe above draws one line, because `actuals` holds one row per year. The
`gapminder_asia` table holds five countries, and each has a row for every year.
One `line` reads every row, so all five countries would join into a single
stroke. A categorical `color` does two things, and you write it once. It first
*splits* the rows into one line per category, then colors them:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

**One atom away.** When too many categories make the legend unreadable, `group`
splits the rows without coloring them; see [Many groups](groups.qmd).

## What about a date column?

The recipes above put a plain number on `x`. The `six_weeks` table holds one
order count per day, and its `day` column holds calendar dates instead. A
column of dates *is* temporal, so you get a calendar axis without asking for
one. Ticks land on calendar boundaries, and weekly steps land on Mondays:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + line + x(day) + y(orders)
```

*"Given the six weeks table: a line, x is day, y is orders."*

You never have to write `scale = "time"`. The column holds dates, and gog reads
the calendar from its type. [Scales](../scales.qmd) owns the calendar's rules.
Over a longer span a three-month step marks each quarter. Over a span of hours
the tick at midnight is labeled with its date, not `00:00`.

## What happened, and what comes next?

The sales line in the first recipe stops at 2023, and the `forecast` table
holds the three years after it. Both go on one plot, and a different mark for
each keeps them apart. The first `data()` names the table for the whole plot. A
later `data()` names the table for the mark written after it.

```{r}
#| echo: false
peek(forecast)
```

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points."* The line stops one year before the points start. Each layer reads
only the table named for it. [Data](../data.qmd) owns multi-table plots.

## What this section refuses

Every recipe so far carried time on `x`. Suppose instead you want one scatter
of wealth against life expectancy per year, so the years sit side by side. One
panel per year is faceting. A facet variable *names* its panels, and `year`
here is a number rather than a name. The refusal names the fix, and [Many
groups](groups.qmd) draws it:

```{r}
#| error: true
data(gapminder_asia) + point + x(gdp) + y(life) | facet(year)
```
