# Over time

```{r}
#| include: false
source("../R/setup.R")
```

How did a measurement change over time? Time is not a chart type in this
grammar. It is a column like any other, and you decide where it goes. Every
recipe here puts it on `x`, which gives the familiar time series. The last
section asks for one panel per year instead, which gog refuses.

## How did it change?

The `actuals` table holds one sales figure per year. The question is whether
sales rose or fell. A `line` with `year` on `x` answers it:

```{r}
#| echo: false
peek(actuals)
```

```{r}
data(actuals) + line + x(year) + y(sales)
```

*"Given the actuals: a line, x is year, y is sales."*

Sales rose over the five years, from 120 to 168, and fell only once, in 2021. A
`line` sorts the rows by `x` and joins each row's point to the next. With years
on `x`, the stroke runs from the earliest year to the latest. Owning chapter:
[Line](../marks/line.qmd).

**One atom away.** If the question is how large sales were in each year, not
whether they rose, swap `line` for [`area`](../marks/area.qmd). An `area` fills
from zero up to each value, so the height of the fill is the amount itself.
[Two variables](two-variables.qmd#how-much-was-there-at-each-time) draws that
plot on this table.

## How did several things change?

The recipe above needed only one line, because `actuals` holds one row per year.
The `gapminder_asia` table holds five countries, each with one row for the same
twelve years, 1952 to 2007. One `line` reads every row, so without `color` the
five countries join into a single stroke. A categorical `color` does two jobs:
it *splits* the rows into one line per category, and it gives each line its own
color:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

Five countries, five lines, and a legend that names them. Life expectancy rose
in all five countries from 1952 to 2007. Only China's line ever falls, from
50.5 years in 1957 to 44.5 in 1962. Owning chapters: [Line](../marks/line.qmd),
[Channels](../channels.qmd).

**One atom away.** When the lines must be kept apart but their names do not
matter, swap `color` for `group`. It splits the rows in the same way, draws
every line in one color, and adds no legend.
[Many groups](groups.qmd#grouping-without-a-legend) draws that plot.

## What about a date column?

The recipes above put a plain number on `x`. The `six_weeks` table holds one row
for each of 42 days, with the date, the number of orders, the weekday and the
week. Its `day` column holds calendar dates rather than plain numbers, so you
get a calendar axis without asking for one. Each tick falls at the start of a
day, a week, a month or a year, and a week starts on a Monday:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + line + x(day) + y(orders)
```

Orders climb from 20 to 33 in the first week, fall to 13 on March 22 and 23,
and end at 34 on April 11. The table starts on a Friday, March 1, so the first
tick is on the first Monday, March 4. After that there is a tick every Monday.

You never have to write `scale = "time"`, because the column's type already
says that it holds dates. A longer span of dates gets a longer step between
ticks: days, then weeks, then months, then years. A column of date-times also
stores the time of day, so over a span of hours its ticks are clock times, such
as 06:00. The midnight tick shows its date instead of `00:00`.
[Scales](../scales.qmd#time-a-date-column-gets-a-calendar-axis) owns the
calendar's rules.

## What happened, and what comes next?

The sales line in the first recipe stops at 2023, and the `forecast` table
holds the three years after it. Both tables are drawn on one plot, and a
different mark for each shows which values were measured and which are
forecast. The first `data()` names the table that every mark reads, unless a
mark is given another. A later `data()` gives its table to the one mark written
directly after it:

```{r}
#| echo: false
peek(forecast)
```

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points."*

The line stops at 2023 and the points start at 2024, because the line reads
only `actuals` and the points read only `forecast`. `x` and `y` are written
once, before both marks, so both marks use them. That works because `forecast`
has the same two columns, `year` and `sales`.
[Data](../data.qmd#the-plots-table-and-a-layers-table) owns plots that read
more than one table.

## What this section refuses

Every recipe so far put time on `x`. Suppose instead you want one scatter of GDP
per capita and life expectancy for each year, with the years side by side. One
panel per year is what [faceting](../faceting.qmd) does. A facet column *names*
its panels, so it must be categorical, but `year` holds numbers. gog refuses the
sentence, and the message names two fixes:

```{r}
#| error: true
data(gapminder_asia) + point + x(gdp) + y(life) | facet(year)
```

The first fix turns `year` into text. The `gm_eras` table already holds each
year as text, in the column `era`.
[Many groups](groups.qmd#what-if-two-columns-group-the-rows) draws a plot split
by `era`, with one row of panels for each of the two years.
