# Parts of a whole

```{r}
#| include: false
source("../R/setup.R")
```

What share of a whole does each part take, and how do the parts compare across
several wholes? The book draws many charts that answer these questions, each in
the chapter that teaches its atoms. This chapter draws them from two tables and
changes a few atoms at a time. The charts are the pie, the share-of-total
column, the donut, the treemap, the stacked and 100% stacked bars, the mosaic, a
pie for each whole and the sunburst. Most of them show `commutes`, 3,790 people
in four cities, counted by their **mode** of travel to work: car, bus or tram,
bicycle or on foot. The two sunbursts at the end show a household's month of
spending, whose parts split more than once.

## What share does each mode of travel take?

The first question is how the 3,790 people divide among the four modes. The pie
is the usual answer, and [The pie: where you put the
category](../polar.qmd#the-pie-where-you-put-the-category) explains that a pie
is a stacked bar in a circle. `commutes` holds one row for each city and mode:

```{r}
#| echo: false
peek(commutes)
```

```{r}
data(commutes) + bar * sum * stack + y(people, axis = FALSE) + color(mode) +
  polar() + title("Each mode's share of all trips")
```

*"Given the commutes: bars derived by sum and stack, y is people without an
axis, color by mode, in polar."*

`sum` adds the four cities together for each mode, and `stack` puts the four
totals end to end around the circle. The car takes almost half the circle:
1,780 of the 3,790 people drive.

## What if the shares are close?

The bus or tram carries 900 people and the bicycle 690, 24% and 18% of the
whole. Angles this close can be hard to judge. Remove `polar()` and
`axis = FALSE`, and the same sentence draws a **share-of-total column**, where
the parts are lengths along one counted axis ([What a circle makes harder to
read](../polar.qmd#what-a-circle-makes-harder-to-read)):

```{r}
data(commutes) + bar * sum * stack + y(people) + color(mode) +
  title("The same shares as one column")
```

Each part now starts where the one below it ends, and the axis counts people,
so the bus or tram part is visibly longer than the bicycle part.

## What if the names go in the ring?

The pie needs its legend: you read a color, then look beside the plot for its
name. A **donut** writes each name in its ring and leaves its middle empty. The
pie's sentence cannot leave a hole, because it binds nothing to the radius. So
in gog a donut is a hierarchy one level deep. `zone * partition(mode)` draws one
part for each mode, and `x(people)` sets each part's share of the circle.
`partition` also writes two columns of its own, `depth`, the ring a part sits
on, and `name`, the part's name. The ring sits between 1 and 2 on `depth`, so
`y(depth, limits = c(0, 2))` leaves the inner half of the radius empty. A `text`
layer with the same partition writes each `name` in its part ([The donut: one
level is still a tree](../polar.qmd#the-donut-one-level-is-still-a-tree)):

```{r}
data(commutes) + x(people, axis = FALSE) + y(depth, limits = c(0, 2)) +
  zone * partition(mode) + color(mode) +
  text * partition(mode) + label(name) + style(color = "white") +
  polar() + title("The names in the ring")
```

*"Given the commutes: x is people without an axis, y is depth from 0 to 2, zones
derived by partition through mode, colored by mode, and also text derived by
partition through mode, label by name, in polar."*

The shares are the same as in the pie, and each name sits in its own part.

## What if there are many parts?

So far every chart has added the four cities together. Kept apart, the four
cities and four modes make sixteen parts, and sixteen wedges in one circle would
be hard to compare. A **treemap** gives each part an area instead. Its sentence
uses `nest()` where the pie used `polar()`, and `bar` alone, because the rows
already hold one count for each part. `x(city)` divides the panel into one
region for each city, and each region is packed with its city's modes.
`color(city)` colors the regions, and a `text` layer writes each mode in its
part ([Two levels, from a position](../nest.qmd#two-levels-from-a-position)):

```{r}
data(commutes) + x(city) + y(people) +
  bar + color(city) +
  text + label(mode) + style(color = "white", size = 11) +
  nest() + title("Sixteen parts in four cities")
```

*"Given the commutes: x is city, y is people, bars colored by city, and also
text, label by mode, in the nest."*

Ashford's region is the largest, because 1,540 of the people counted are in
Ashford. Inside it, the car takes the largest area. In Millbrook the largest
area is the bicycle's.

## How do the parts compare across several wholes?

Each city is a whole of its own. Go back to the share-of-total column and add
`x(city)`. Stacked bars then draw one column for each city and split it by
mode, so you can compare the cities' sizes and their parts at once. The rows
already hold one count for each city and mode, so `stack` needs no `sum`
([Stacked bars: piled up](../marks/bar.qmd#stacked-bars-piled-up)):

```{r}
data(commutes) + bar * stack + x(city) + y(people) + color(mode) +
  title("Each city a whole")
```

*"Given the commutes: bars derived by stack, x is city, y is people, color by
mode."*

Ashford's column is three times as tall as Millbrook's, 1,540 people against
500. Its car part is also the longest, so you cannot tell whether a larger share
of Ashford's people drive, or Ashford only has more people.

## What if only the shares matter?

The question is whether a larger share of people drive in Ashford than in
Millbrook. `stack(share = TRUE)` divides each column by its own total, so every
column reaches 1 and the parts become shares ([The 100% stacked
bar](../marks/bar.qmd#the-100-stacked-bar)):

```{r}
data(commutes) + bar * stack(share = TRUE) + x(city) + y(people) + color(mode) +
  title("Shares only")
```

*"Given the commutes: bars derived by stack as shares, x is city, y is people,
color by mode."*

Now you can compare the shares directly. In Ashford 64% of the people drive,
and in Millbrook only 24%; in Millbrook 42% ride a bicycle. The columns no
longer show how many people each city has, because every column has the same
height.

## What if each whole's size matters too?

A **mosaic plot**, also called a **Marimekko chart**, shows each city's size and
its shares at once. Each city's column is as wide as its number of people.
Inside a column, each part's height is its share of that city, so each cell's
area is the number of people in it ([The mosaic](../transforms.qmd#the-mosaic)):

```{r}
data(commutes) + x(people) +
  zone * partition(city, mode, cross = TRUE) + color(mode) +
  style(border_color = "white", border_size = 2) +
  text * partition(city, cross = TRUE) + label(name) +
  title("Width is the city's size")
```

*"Given the commutes: x is people, zones derived by partition through city and
mode, crossed, colored by mode, and also text derived by partition through
city, crossed, label by name."*

Ashford's column takes about two fifths of the width, and its car cell is the
largest area on the plot. Millbrook's column is narrow, so its large bicycle
share is only 210 people.

## What about a pie for each whole?

Many reports draw one pie for each city. That is the first pie's sentence with
two changes, `stack(share = TRUE)` and `| facet(city)`:

```{r}
data(commutes) + bar * sum * stack(share = TRUE) + y(people, axis = FALSE) +
  color(mode) + polar() + title("A pie for each city") | facet(city)
```

*"Given the commutes: bars derived by sum and stack as shares, y is people
without an axis, color by mode, in polar, split into panel columns by city."*

`stack(share = TRUE)` is what makes each pie a whole circle. With plain `stack`,
the panels share one count axis, so only Ashford's pie would be a whole circle.
Each pie shows its city's parts, but comparing a share between cities means
comparing angles in separate circles. When the comparison between cities is
the point, use the 100% stacked bars above: they put the same shares side by
side on one axis, which is usually easier to read.

## What if the whole splits more than once?

Some wholes split more than once. `commutes` splits only into cities and modes,
so this section uses `spending`, one month of a household's spending. It splits
into groups, then items, and for two items into details:

```{r}
#| echo: false
peek(spending, n = 12)
```

A **sunburst** draws one ring for each level, from the center outward. Only
Utilities and Car split into details, so only their parts reach the outer ring.
The whole itself goes at the center, as a first level that every row shares.
The table has no such column, so add one before you draw:

```{r}
#| echo: false
spellings(
  r = r"(spending$whole <- "All spending"

data(spending) + x(amount, axis = FALSE) +
  zone * partition(whole, group, item, detail) + color(group) +
  style(border_color = "white", border_size = 2) +
  text * partition(whole, group, item, detail) + label(name) +
  style(color = "white", size = 10) +
  polar() + theme(grid = "none", frame = "none", background = "white") +
  title("A month of spending, split three times"))",
  python = r"(spending["whole"] = ["All spending"] * len(spending["group"])

(data(spending) + x(col.amount, axis = False) +
  zone * partition(col.whole, col.group, col.item, col.detail) + color(col.group) +
  style(border_color = "white", border_size = 2) +
  text * partition(col.whole, col.group, col.item, col.detail) + label(col.name) +
  style(color = "white", size = 10) +
  polar() + theme(grid = "none", frame = "none", background = "white") +
  title("A month of spending, split three times")))",
  julia = r"(spending["whole"] = fill("All spending", length(spending["group"]))

data(spending) + x(:amount, axis = false) +
  zone * partition(:whole, :group, :item, :detail) + color(:group) +
  style(border_color = "white", border_size = 2) +
  text * partition(:whole, :group, :item, :detail) + label(:name) +
  style(color = "white", size = 10) + polar() +
  theme(grid = "none", frame = "none", background = "white") +
  title("A month of spending, split three times"))",
  js = r"(spending.whole = spending.group.map(() => "All spending");

plot(data(spending), x(col.amount, { axis: false }),
  layer(zone, partition(col.whole, col.group, col.item, col.detail)),
  color(col.group), style({ border_color: "white", border_size: 2 }),
  layer(text, partition(col.whole, col.group, col.item, col.detail)),
  label(col.name), style({ color: "white", size: 10 }), polar(),
  theme({ grid: "none", frame: "none", background: "white" }),
  title("A month of spending, split three times"));)"
)
```

```{r}
#| echo: false
# The plot the four tabs above describe.
spending$whole <- "All spending"

data(spending) + x(amount, axis = FALSE) +
  zone * partition(whole, group, item, detail) + color(group) +
  style(border_color = "white", border_size = 2) +
  text * partition(whole, group, item, detail) + label(name) +
  style(color = "white", size = 10) +
  polar() + theme(grid = "none", frame = "none", background = "white") +
  title("A month of spending, split three times")
```

*"Given the spending table: x is amount without an axis, zones derived by
partition through whole, group, item and detail, colored by group, and also
text derived by partition through whole, group, item and detail, label by name,
in polar."*

Housing takes more than half of the 2,490 spent, and Rent takes most of the
Housing part. The center holds every group, so it has no group of its own to
color, and it is drawn in gray. The white borders separate the parts, and
`theme()` removes the grid, the frame and the gray background, so nothing is
drawn around the uneven rim. A name too wide for its part turns, so the names
of thin parts, such as Hobbies, each run along its own radius. Subscriptions and
Season ticket still do not fit, so they are drawn smaller ([Naming the
nodes](../polar.qmd#naming-the-nodes)).

## What if the center should stay empty?

In the first sunburst, the center says only All spending, and it takes the
middle of the plot. To leave the center empty, as the donut does, leave out the
`whole` level. Then `y(depth, limits = c(0, 4))` leaves the inner quarter of
the radius empty ([The sunburst: depth is the
radius](../polar.qmd#the-sunburst-depth-is-the-radius)):

```{r}
data(spending) + x(amount, axis = FALSE) + y(depth, limits = c(0, 4)) +
  zone * partition(group, item, detail) + color(group) +
  style(border_color = "white", border_size = 2) +
  text * partition(group, item, detail) + label(name) +
  style(color = "white", size = 10) +
  polar() + theme(grid = "none", frame = "none", background = "white") +
  title("The same month, with an empty center")
```

*"Given the spending table: x is amount without an axis, y is depth from 0 to 4,
zones derived by partition through group, item and detail, colored by group,
and also text derived by partition through group, item and detail, label by
name, in polar."*

The rings are the same three levels, and the groups are now the innermost ring.

## What this section refuses

A treemap of the four modes alone is the pie's sentence with `nest()` in place
of `polar()`, but one more atom must be removed. Keep `stack`, and the sentence
is refused:

```{r}
#| error: true
data(commutes) + bar * sum * stack + y(people) + color(mode) + nest()
```

The message says that `stack` decides what happens when two marks land in the
same place. In a packing that cannot happen, because every part has a region of
its own. The message says to drop `stack`, and that the shares are the same
without it.
