# Above and below a reference

```{r}
#| include: false
source("../R/setup.R")
```

How far is each value from a reference, and is it above or below?
[Comparing categories](comparing.qmd) ranks twelve countries by how long their
people lived. This chapter measures each one against a reference instead: the
world median. The chapter starts from the Cleveland dot plot and changes a few
atoms at a time. Each change adds a reference line or draws a chart that has a
name of its own: the diverging lollipop chart, the diverging bar chart and the
back-to-back bar chart. Every section but one uses the same twelve countries of
East, Southeast and South Asia, from the gapminder data. The back-to-back bar
chart uses the Titanic.

## Which countries live longer than the world median?

In 2007, half of the 142 countries in the gapminder data had a life expectancy
above 71.94 years, and half had one below it. That number is the **world
median**. `gm_twelve_2007` holds the twelve countries' life expectancy in 2007,
one row per country:

```{r}
#| echo: false
peek(gm_twelve_2007)
```

A reference is a single number, so give it a table with a single row. A `rule`
draws one line for each row of its own table, on the axis whose column that
table holds ([A line at a threshold](../marks/rule.qmd#a-line-at-a-threshold)).
Here the table holds `life`, the plot's `x`, so the line runs up the panel:

```{r}
reference <- data.frame(life = 71.94)

data(gm_twelve_2007) + y(country) + x(life) + order(life, desc = TRUE) +
  point + data(reference) + rule + style(pattern = "dashed") +
  title("The world median as a line")
```

*"Given the gapminder twelve in 2007: y is country, x is life, ordered by life,
largest first, points, and also rules from the reference table."*

This is the Cleveland dot plot with one layer added, a **reference line**. The
eight countries from Japan to China are to the right of the line, so their
people lived longer than the median. The Philippines, Indonesia, Thailand and
India are to its left. The line is dashed so that it does not read as a
country's value.

## How far is each country from the median?

The reference line shows the side of each country, but the reader has to judge
each distance from the line. The [lollipop
chart](comparing.qmd#what-if-the-length-matters-too) draws a stem from zero to
each value. Start each stem at the median instead, and its length is the
distance from the reference. That is the **diverging lollipop chart**.
`interval * bounds(world, life)` draws each stem between two columns, so the
table needs a column `world` that holds the median in every row. Add it, and a
column `side` that says on which side of the median each country falls:

```{r}
#| echo: false
spellings(
  r = r"(gm_twelve_2007$world <- 71.94
gm_twelve_2007$side <- ifelse(gm_twelve_2007$life >= 71.94, "Above", "Below")

data(gm_twelve_2007) + y(country) + x(life) + order(life, desc = TRUE) +
  interval * bounds(world, life) + style(caps = FALSE) +
  point + color(side) +
  title("A stem from the median to each value"))",
  python = r"(gm_twelve_2007["world"] = [71.94] * len(gm_twelve_2007["life"])
gm_twelve_2007["side"] = ["Above" if v >= 71.94 else "Below"
                          for v in gm_twelve_2007["life"]]

(data(gm_twelve_2007) + y(col.country) + x(col.life) + order(col.life, desc = True) +
  interval * bounds(col.world, col.life) + style(caps = False) +
  point + color(col.side) +
  title("A stem from the median to each value")))",
  julia = r"(gm_twelve_2007["world"] = fill(71.94, length(gm_twelve_2007["life"]))
gm_twelve_2007["side"] = [v >= 71.94 ? "Above" : "Below" for v in gm_twelve_2007["life"]]

data(gm_twelve_2007) + y(:country) + x(:life) + order(:life, desc = true) +
  interval * bounds(:world, :life) + style(caps = false) +
  point + color(:side) + title("A stem from the median to each value"))",
  js = r"(gm_twelve_2007.world = gm_twelve_2007.life.map(() => 71.94);
gm_twelve_2007.side = gm_twelve_2007.life.map((v) => (v >= 71.94 ? "Above" : "Below"));

plot(data(gm_twelve_2007), y(col.country), x(col.life),
  order(col.life, { desc: true }),
  layer(interval, bounds(col.world, col.life)), style({ caps: false }),
  point, color(col.side), title("A stem from the median to each value"));)"
)
```

```{r}
#| echo: false
# The plot the four tabs above describe.
gm_twelve_2007$world <- 71.94
gm_twelve_2007$side <- ifelse(gm_twelve_2007$life >= 71.94, "Above", "Below")

data(gm_twelve_2007) + y(country) + x(life) + order(life, desc = TRUE) +
  interval * bounds(world, life) + style(caps = FALSE) +
  point + color(side) +
  title("A stem from the median to each value")
```

*"Given the gapminder twelve in 2007: y is country, x is life, ordered by life,
largest first, intervals from world to life, and also points colored by side."*

Every stem starts at 71.94, and each dot is at the country's own value. Japan's
stem is the longest above the median, and India's is the longest below it. The
Philippines is only 0.25 years below the median, so its stem is very short.
`color(side)` comes after `point`, so it colors only the points, and the stems
stay gray ([Encoding
scope](../encoding-scope.qmd#channels-plot-scoped-or-layer-scoped)). As in the
lollipop chart, `style(caps = FALSE)` leaves out the short crossbar that an
interval draws at each end.

## What if only the distance matters?

The axis of the diverging lollipop chart still shows life expectancy, from about
65 to 83 years. When the distance itself is the answer, subtract the median from
every value. The median then becomes zero. A bar runs from zero to its value, on
whichever side of zero the value is. So a country above the median gets a bar to
the right of zero, and a country below it gets a bar to the left. That is the
**diverging bar chart**. Add the difference as a column, `gap`:

```{r}
#| echo: false
spellings(
  r = r"(gm_twelve_2007$gap <- gm_twelve_2007$life - 71.94

data(gm_twelve_2007) + bar + y(country) + x(gap) + order(gap, desc = TRUE) +
  color(side) + title("Years above or below the world median"))",
  python = r"(gm_twelve_2007["gap"] = [v - 71.94 for v in gm_twelve_2007["life"]]

(data(gm_twelve_2007) + bar + y(col.country) + x(col.gap) + order(col.gap, desc = True) +
  color(col.side) + title("Years above or below the world median")))",
  julia = r"(gm_twelve_2007["gap"] = gm_twelve_2007["life"] .- 71.94

data(gm_twelve_2007) + bar + y(:country) + x(:gap) + order(:gap, desc = true) +
  color(:side) + title("Years above or below the world median"))",
  js = r"(gm_twelve_2007.gap = gm_twelve_2007.life.map((v) => v - 71.94);

plot(data(gm_twelve_2007), bar, y(col.country), x(col.gap),
  order(col.gap, { desc: true }), color(col.side),
  title("Years above or below the world median"));)"
)
```

```{r}
#| echo: false
# The plot the four tabs above describe.
gm_twelve_2007$gap <- gm_twelve_2007$life - 71.94

data(gm_twelve_2007) + bar + y(country) + x(gap) + order(gap, desc = TRUE) +
  color(side) + title("Years above or below the world median")
```

*"Given the gapminder twelve in 2007: bars, y is country, x is gap, ordered by
gap, largest first, color by side."*

The axis now measures years from the median. Japan is 10.66 years above it and
India 7.24 years below. The bars of Vietnam and Malaysia are nearly the same
length, 2.31 and 2.30 years, and so are the bars of Indonesia and Thailand, 1.29
and 1.32. Bars on both sides of zero need no new atom, only negative values.

## How did the distance change over the years?

The world median rose in every five-year step, from 45.14 years in 1952 to
71.94 in 2007. So each year needs a reference of its own. `gm_twelve_gap` holds
the twelve countries in every year, with that year's median in `world`, the
difference in `gap`, and `side`. Each year's median comes from all 142
countries, not from these twelve, so the table's file already holds `world` and
`gap`:

```{r}
#| echo: false
peek(gm_twelve_gap)
```

The diverging bar chart for one year becomes one panel for each country, with
the years on `x` and the gap on `y`. `| facet(country, wrap = 4)` draws the
panels in rows of four ([Faceting](../faceting.qmd)):

```{r}
data(gm_twelve_gap) + bar + x(year) + y(gap) + color(side) +
  title("Each year's gap from the world median") | facet(country, wrap = 4)
```

*"Given the gapminder twelve's gaps: bars, x is year, y is gap, color by side,
split into panel columns by country, wrapped at 4."*

The panels follow the order of 2007, because the table lists the countries in
that order. Japan, Hong Kong and Singapore stayed far above the median, but
their gaps became smaller, because the median rose faster than their life
expectancy. Japan's fell from 17.89 years in 1952 to 10.66 in 2007. Three
countries moved to the other side of the median and stayed there. Vietnam was
below the median until 1992 and above it from 1997. The Philippines moved below
it in 1982, and Thailand in 1992. China was below the median in 1952 and 1962
and above it in every other year. India and Indonesia were below it in every
year.

## What if zero separates two groups?

So far the reference has been a median, computed from the data. It does not
have to be. In a **back-to-back bar chart**, zero separates two groups: the bars
of one group run to the right, and the bars of the other run to the left.
`titanic` holds the passengers and crew of the Titanic, the ship that sank in
1912, as 32 groups, one for each combination of `class`, `sex`, `age` and
`survived`. `n` is the number of people in each group:

```{r}
#| echo: false
peek(titanic)
```

Make the counts of the people who died negative, and their bars run left of
zero, while the bars of the survivors run right. Add the counts, with the
deaths now negative, as a column, `people`:

```{r}
#| echo: false
spellings(
  r = r"(titanic$people <- ifelse(titanic$survived == "No", -titanic$n, titanic$n)

data(titanic) + bar * sum + y(class) + x(people) + color(survived) +
  title("Who survived, and who died, in each class"))",
  python = r"(titanic["people"] = [-n if s == "No" else n
                     for n, s in zip(titanic["n"], titanic["survived"])]

(data(titanic) + bar * sum + y(col["class"]) + x(col.people) + color(col.survived) +
  title("Who survived, and who died, in each class")))",
  julia = r"(titanic["people"] = [s == "No" ? -n : n for (n, s) in zip(titanic["n"], titanic["survived"])]

data(titanic) + bar * sum + y(:class) + x(:people) + color(:survived) +
  title("Who survived, and who died, in each class"))",
  js = r"(titanic.people = titanic.n.map((n, i) => (titanic.survived[i] === "No" ? -n : n));

plot(data(titanic), layer(bar, sum), y(col.class), x(col.people),
  color(col.survived), title("Who survived, and who died, in each class"));)"
)
```

```{r}
#| echo: false
# The plot the four tabs above describe.
titanic$people <- ifelse(titanic$survived == "No", -titanic$n, titanic$n)

data(titanic) + bar * sum + y(class) + x(people) + color(survived) +
  title("Who survived, and who died, in each class")
```

*"Given titanic: bars derived by sum, y is class, x is people, color by
survived."*

`sum` adds up the groups that share a class and a value of `survived`, so each
class has one bar on each side. The crew had the most deaths, 673. First class
is the only class in which more people survived than died: 203 survived and 122
died. The left half of the axis reads -600, -400 and -200, because those are
counts of the dead, drawn as negative numbers. The population pyramid in [Many
groups](groups.qmd#where-does-a-population-pyramid-come-from) is a back-to-back
bar chart too, with age bands on `y` and the populations of two cities. It also
says why gog keeps the minus sign on the axis.

## What this section refuses

A [diverging ramp](../channels.qmd#diverging-ramps) colors each value by its
distance from a center, so it might seem the natural way to color the
diverging bar chart. On a `bar`, `color` takes only a categorical column, so the
sentence with `color(gap)` is refused:

```{r}
#| error: true
data(gm_twelve_2007) + bar + y(country) + x(gap) + color(gap) +
  palette("blue_red")
```

The message says that `color` on a `bar` needs a categorical column, and that
`opacity` can show a continuous one. That is why the diverging bar chart is
colored by `side`, a categorical column, and not by `gap`.
