# A third variable, and more

```{r}
#| include: false
source("../R/setup.R")
```

You have more columns than the flat page has positions. Where do the rest go?
`x` and `y` are already taken. Every further variable is carried by a refining
channel (`color`, `size`, `shape`, `pattern`, `opacity`) or by the third
position, `z`. Each channel answers one kind of question, and a refining channel
earns the legend that decodes it.

## Which group does each mark belong to?

The scatter of wealth against life expectancy draws every country as an
identical dot. Nothing in it says which continent a country belongs to, and
that third column is categorical. A category is carried by `color`, and the
palette gives each category one hue:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

**One atom away.** The same split without color, when you want the grouping but
not the hues, is the `group` channel; see [Channels](../channels.qmd).

## Will it read in gray?

Hue alone separates groups on a screen but not in grayscale print, in poor
light, or for a colorblind reader. When the plot must survive all three, bind
`color` *and* `shape` to the same column. Each group is then said twice:

```{r}
data(iris_flowers) + point + x(petal_length) + y(sepal_length) +
  color(species) + shape(species) +
  palette("okabe")
```

*"Given the iris flowers: points, x is petal length, y is sepal length, color by
species, shape by species, with the okabe palette."*

`palette()` chooses which colors the `color` channel uses: here the Okabe and
Ito colorblind-safe set. Saying one column twice is redundant encoding. Owning
chapters: [Channels](../channels.qmd), [Setting vs mapping](../style.qmd).

## How big is each observation?

Population is a third measurement of each country, and both positions are
already taken by wealth and health. A magnitude is carried by `size`; the
channel's question is *how much*, so it only accepts a continuous column:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  size(population)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, size by
population."* A legend decodes the radius. Owning chapter:
[Channels](../channels.qmd).

## What if four variables matter?

The recipes above bound `color` to the continent and `size` to the population,
one column at a time. You may need both at once, to see which continent a
country is in and how many people live there. Channels compose: every one of
them works beside every other. That rule is
[Orthogonality](../design-laws.qmd). Wealth, health, population, and continent
in one sentence:

```{r}
data(gapminder_2007) + point +
  x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent, size by population."*

Every channel keeps its own rule and its own legend; none of them changed
meaning by being combined. Four variables is usually the most a single plot can
carry. Past that, use [panels](../faceting.qmd) instead of more channels.

## What if the third variable is a number?

The third column can be continuous rather than categorical. Population is one,
carried here by hue rather than by radius, so every dot keeps one size.
`color` accepts either kind of column. A continuous column gets a
sequential ramp instead of a palette, and a skewed one reads better on a
log ramp, exactly like a skewed axis:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
population on a log scale."*

The legend becomes a labeled gradient strip: same channel, same rule, different
column type. [Scales](../scales.qmd) owns the log scale.

## What if three measurements matter equally?

Each iris here has three measurements, and none of them is less important than
the others. `color` or `size` would carry the third one as a refinement of a
plot about the other two. When three *measurements* matter equally, the third
one is a position, not a decoration. One more vowel and the plot stands up:

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is
petal length, color by species."*

**One atom away.** The viewing angle is a property of the space, not of
the data; `turn` rotates the scene sideways, `tilt` raises the viewing height:

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species) +
  space(turn = -50, tilt = 15)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is
petal length, color by species, turned -50, tilted 15."*

[Space](../space.qmd) owns the third dimension.

## What if the rows form a route?

The scatter above is a cloud: each flower is its own dot and the table's order
means nothing. Sometimes the order is what matters. Here are the two gliders
from [Path](../marks/path.qmd) again. Each row records where that glider was a
moment later, so
the marks belong joined in the order they were recorded. The mark that joins
rows in table order is `path`:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(glider) +
  title("Two gliders, one thermal")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider."*

Each route rises because the gliders do. Nothing in the sentence draws a spiral:
`path` was told to connect the rows in order, the rows happen to go around while
they go up, and the coil is the result. Watch where the two routes cross, too.
Whichever route is nearer covers the other, and the nearer one changes as they
climb, so neither glider is simply drawn on top.

Turn the view and the same route answers a different question, with no binding
changed:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(glider) +
  space(turn = 20, tilt = 80) +
  title("The same two routes, seen from overhead")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider, turned 20, tilted 80."*

From nearly overhead the climb collapses and you are looking down at two
circles. The plot has become a picture of how wide each circle is, and it shows
what the first view hid: Bravo is flying a far tighter circle than Alpha. The
side view above showed Bravo climbing faster; this view shows the tighter turn
that goes with it.

Of `line`, `step`, `area`, `ribbon` and `path`, only `path` takes `z`, and the
reason is the one that separates it from `line` in the plane. A path joins rows
in the table's order, and an order belongs to no axis, so it survives the third
dimension untouched. A `line` sorts by `x`, and the cube has no left to right:
at some angles `x` runs straight into the page and becomes depth rather than
position, which would leave a line sorted by an axis you cannot see. So a rising
route is `path`, and `line` refuses to pretend otherwise:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude) + color(glider)
```

## What if the shape needs three dimensions?

`quakes_fiji` is the thousand earthquakes from [Map](../map.qmd), each with a
place on the globe and a depth below it. The reason it is here is that `east`,
`north` and `elevation` show together what no pair of them shows alone:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  color(magnitude) + palette("viridis") + space(turn = 120, tilt = 20) +
  title("A thousand earthquakes, and the sheet they lie on")
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, color by
magnitude."*

The column is named `elevation`, so a measurement below sea level is a negative
number: the deepest quake in the table sits at −680 km. The quakes are not
scattered through the cube. They trace a sheet of ocean floor sinking into the
earth. It starts shallow in the east and runs down past 600 km to the west. That
is a shape, and a shape needs three positions to be a shape.

Now cut the quakes into slices of depth and run the slices one after another.
`play(slab)` steps through `slab`, a column that holds the depth in 90 km bands:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  space(turn = 120, tilt = 20) + play(slab)
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, played by
slab."*

The band of points moves down and west as the sequence runs, which is the slope
of that sheet read one slice at a time. The frames are not time here. `play`
advances through the levels of an ordered column, and whether that column
measures years or kilometers is the column's job, not the channel's. Depth
happens to be the one variable a cross-section is usually cut on, so this is the
ordinary cross-section view, spelled with the atom that also animates a year.

The same slices laid out across the page instead of shown one after another in
time:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  space(turn = 120, tilt = 20) | facet(slab, wrap = 4)
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, split into
panel columns by slab, wrapped at 4."*

Eight cubes, on one shared scale, so the reader compares them side by side
instead of remembering them. Choosing between them is a question about the
reader: panels let them look back, frames let them see the movement.
[Play](../play.qmd) owns that choice.

## Which rows and columns are alike?

The `nutrients` table gives the amount of five nutrients in each of eight foods.
The question is which foods resemble each other, and which nutrients are found
in the same foods. Two categorical axes and one value make a tile plot, and
nothing in the data chose its slot order. Build a [cluster](../cluster.qmd)
tree on each axis, then compose the three plots: the shared axes take the
trees' leaf order, and tiles of similar color sit next to each other.

```{r}
(data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
   theme(height = 150)) /
  ((data(nutrients) + zone + x(food) + y(nutrient) + color(amount)) |
     (data(nutrients) + path * cluster(amount, over = food) + y(nutrient) +
        theme(width = 150)))
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, x is food;
above zones, x is food, y is nutrient, color by amount, beside paths derived by
cluster on amount, over food, y is nutrient."* Owning chapters:
[Cluster](../cluster.qmd), [Zone](../marks/zone.qmd).

## What this section refuses

Population was bound to `size` earlier, and `shape` may look like another
channel that could carry it. Five glyph shapes have no order from smallest to
largest, so `shape` refuses a continuous column and names the channels that do
carry magnitude:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + shape(population)
```
