# A third variable, and more

```{r}
#| include: false
source("../R/setup.R")
```

You have more columns than the flat page has positions. Where do the rest go?
`x` and `y` are already taken. A further variable can go in one of four places.
The first is a refining channel: `color`, `size`, `shape`, `pattern` or
`opacity`. The second is the third position, `z`. The third is `play`, which
gives the variable one frame per value, and the fourth is `facet`, which gives
it one panel per value. This chapter uses all four. Each channel answers one
kind of question, and a refining channel gets the legend that decodes it.

## Which group does each row belong to?

The first scatter in [Two variables](two-variables.qmd), GDP per capita against
life expectancy, draws every country as an identical dot. Nothing in it says
which continent a country belongs to. The `continent` column is categorical. A
category is carried by `color`, and the palette gives each category one hue:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

**One atom away.** For the same split without hues, swap `color` for `shape`,
and each continent gets its own glyph instead. The `group` channel cannot do
this on a scatter. It splits a mark that joins rows, such as `line`, and a point
joins nothing. See [Channels](../channels.qmd).

## Will it read in gray?

That plot separates the continents by hue alone. Hue works on a screen, but not
in grayscale print, in poor light, or for a colorblind reader. When a plot has
to work in all three of those conditions, bind `color` *and* `shape` to the same
column, so each group is said twice. The `iris_flowers` table has three species
of flower, and here each one gets both its own color and its own glyph:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + point + x(petal_length) + y(sepal_length) +
  color(species) + shape(species) +
  palette("okabe")
```

*"Given the iris flowers: points, x is petal length, y is sepal length, color by
species, shape by species, with the okabe palette."*

Printed in gray, *setosa* and *versicolor* become nearly the same shade. Their
glyphs, circles for *setosa* and squares for *versicolor*, still show which
group is which. Saying one column twice is called **redundant encoding**.
`palette()` chooses which colors the `color` channel uses: here the Okabe and
Ito colorblind-safe set. Owning chapters: [Channels](../channels.qmd), [Setting
vs mapping](../style.qmd).

## How big is each observation?

The last two recipes put a category on a channel. This one puts a measurement
there, and it returns to the countries. Population is a third measurement of
each country, and both positions are already taken by GDP per capita and life
expectancy. A magnitude is carried by `size`. That channel answers the question
*how much*, so it accepts a continuous column and nothing else:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  size(population)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, size by
population."*

The two largest circles are China and India. A legend decodes the radius.
Owning chapter: [Channels](../channels.qmd).

## What if four variables matter?

The recipes so far bound `color` to the continent and `size` to the population,
one column at a time. You may need both at once, to see which continent a
country is in and how many people live there. Channels compose: any channel a
mark accepts works beside any other channel it accepts. That rule is
[Orthogonality](../design-laws.qmd#law-1-orthogonality). GDP per capita, life expectancy,
population and continent in one sentence:

```{r}
data(gapminder_2007) + point +
  x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent, size by population."*

Every channel keeps its own rule and its own legend; none of them changed
meaning by being combined. Four variables are usually the most a single plot
can carry. Beyond four, use [panels](../faceting.qmd) instead of more channels.

## What if the third variable is continuous?

Now return to three columns. A continuous third column does not have to go on
`size`. Here population goes on `color` instead, so every dot keeps the same
size and the color carries the number. `color` accepts either kind of column. A
continuous column gets a sequential ramp, from light to dark, instead of a
categorical palette. A skewed column reads better on a log ramp. On a linear
ramp, China and India would sit alone at the dark end, and almost every other
country would get the same light shade:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
population on a log scale."*

The legend becomes a labeled ramp. The channel and its rule did not
change; only the kind of column did. [Scales](../scales.qmd) owns the log scale.

## What if three measurements matter equally?

The iris flowers from the grayscale recipe above carry three measurements each,
and none of them is less important than the others. `color` or `size` would make
the third measurement a refinement, and the plot would still be about the other
two. When three *measurements* matter equally, the third one is a position, not
a refinement. Put one measurement on each of `x`, `y` and `z`, and the plot
becomes a cube:

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is
petal length, color by species."*

*Setosa* sits apart, near the bottom of the cube, because its petals are short.
*Versicolor* and *virginica* sit higher, and from this side they overlap.

**One atom away.** The viewing angle is a property of the space, not of the
data. `turn` says which side of the cube you look at, and `tilt` says how high
your eye is:

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species) +
  space(turn = -50, tilt = 15)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is
petal length, color by species, turned -50, tilted 15."*

From this side *versicolor* and *virginica* overlap less.
[Space](../space.qmd) owns the third dimension.

## What if the rows form a route?

Each flower in the iris cube is its own dot, and the order of the rows means
nothing. Sometimes the order is what matters. Here are the two gliders from
[Path](../marks/path.qmd) again. Each row records where one glider was at one
moment, and that glider's next row is two seconds later. So the rows have to be
joined in the order they were recorded. The mark that joins rows in table order
is `path`:

```{r}
#| echo: false
peek(thermals)
```

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(glider) +
  title("Two gliders, one thermal")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider."*

Each route rises because each glider gains altitude. Both gliders start at 900
meters and were recorded for the same 338 seconds. Bravo's first point is drawn
higher than Alpha's, but only because it sits at the back of the cube. In this
view, the back is drawn higher on the page. By the end, Bravo has reached about
2,500 meters and Alpha about 2,100, so Bravo climbs faster.

Nothing in the sentence draws a spiral. The sentence asks `path` to connect the
rows in order. Each glider moves in a circle while it rises, so the connected
rows form a spiral. Look at the crossings as well. At each one the route nearer
to the eye is drawn over the route behind it, and which route is nearer changes
as the two gliders climb. Neither route is drawn entirely in front of the other:
the pieces of both routes are sorted by depth together, as
[Path](../marks/path.qmd#a-path-in-the-cube) explains.

Turn the view, and the same two routes answer a different question. No binding
changes:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(glider) +
  space(turn = 20, tilt = 80) +
  title("The same two routes, seen from overhead")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider, turned 20, tilted 80."*

Seen from nearly overhead the climb is no longer visible, and you are looking
down at two circles, one inside the other. Bravo's circle is half as wide as
Alpha's. Together, the two views show that Bravo climbs faster and turns in a
tighter circle.

These two plots raise one more question. `line`, `step`, `area` and `ribbon`
also join their rows, so why is `path` the only one of the five that takes `z`?
The answer is the difference that separates `path` from `line` on the flat
page: the order in which each one joins its rows. A path joins rows in the
table's order. That order does not depend on any axis, so a third axis does not
change it. A `line` sorts its rows by `x`. In the cube, `x` is one of three
equal positions. At some angles it points straight into the page, and then you
cannot see the order the line was sorted in. That is why a rising route is drawn
with `path`, and why `line` refuses `z` rather than draw a stroke you cannot
read:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude) + color(glider)
```

## What if the structure needs three dimensions?

A route is one kind of data that needs three positions, and a physical structure
is another. `quakes_fiji` is the thousand earthquakes from [Map](../map.qmd),
each with a place on the globe and a depth below it. Those three columns,
`east`, `north` and `elevation`, show together what no pair of them shows alone:

```{r}
#| echo: false
peek(quakes_fiji)
```

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  color(magnitude) + palette("viridis") + space(turn = 120, tilt = 20) +
  title("A thousand earthquakes, and the sheet they lie on")
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, color by
magnitude."*

Color carries a fourth column here, the magnitude of each quake. The vertical
axis is `elevation`, the height above sea level, so a quake 680 km down has the
value −680. That is the deepest quake in the table.

The quakes are not scattered through the cube. They trace a sheet of ocean floor
sinking into the earth. At this angle, east is on the left. The sheet begins
near the surface at the eastern edge and dips west. About a third of the way
from that edge, it is more than 600 km down. The shallower quakes farther west
do not lie on this sheet. They form a separate group. That sheet lies in three
dimensions, so drawing it takes three positions. Any two of them show it
flattened onto the page.

Now cut the quakes into slices of depth and show the slices one after another.
`play(slab)` steps through `slab`, a column that holds the depth in 90 km bands:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  space(turn = 120, tilt = 20) + play(slab)
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, played by
slab."*

In the web edition the frames play one after another. As they run, the points
move down, and their eastern end moves west. That movement is the slope of the
sheet, read one slice at a time. The frames are not time here. `play` advances
through the levels of an ordered column, and whether that column measures years
or kilometers is a fact about the column, not about `play`. The same atom that
steps through years steps through depths here.

In the PDF, only the first frame is drawn, the slice from 0 to 90 km, and the
panels below replace the motion. They lay the same slices out across the page
instead of showing them one after another in time:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  space(turn = 120, tilt = 20) | facet(slab, wrap = 4)
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, split into
panel columns by slab, wrapped at 4."*

The eight panels share one set of scales. You compare two depths by looking from
one panel to the next, not by remembering the frame that just played. Whether
you choose panels or frames depends on your readers. Panels let them compare any
two slices, and frames let them see the movement. [Play](../play.qmd) owns that
choice.

## Which categories are alike?

The `nutrients` table gives the amount of five nutrients in each of eight foods.
The question is which foods resemble each other, and which nutrients are found
in the same foods. Two categorical axes and one value make a tile plot, with the
amount on `color`. The slot order on each axis is the order the rows arrived in,
and that order carries no meaning. So here the amount also sets the order of
both axes. Build a [cluster](../cluster.qmd) tree on each axis, then compose the
three plots: the tree above, the tiles below it, and the second tree on the
right. Each shared axis takes its leaf order from the tree along that side.
Tiles of similar color then sit next to each other.

```{r}
#| echo: false
peek(nutrients)
```

```{r}
(data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
   theme(height = 150)) /
  ((data(nutrients) + zone + x(food) + y(nutrient) + color(amount)) |
     (data(nutrients) + path * cluster(amount, over = food) + y(nutrient) +
        theme(width = 150)))
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, x is food,
above zones, x is food, y is nutrient, color by amount, beside paths derived by
cluster on amount, over food, y is nutrient."*

Chicken and salmon sit side by side, both dark on protein, and oats, rice,
beans and lentils sit together, dark on carbs. Almonds join the top tree last,
and their fat tile is the darkest on the page. In the tree on the right, iron
and fiber join first: their amounts are the most alike across the eight foods.
Owning chapters: [Cluster](../cluster.qmd), [Zone](../marks/zone.qmd).

## What this section refuses

An earlier recipe put population on `size`, and `shape` may look like a channel
that could carry it too. The seven glyphs have no order from smallest to
largest, so `shape` refuses a continuous column and names channels that do carry
magnitude:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + shape(population)
```
