# A third variable, and more

```{r}
#| include: false
source("../R/setup.R")
```

You have more columns than the flat page has positions. Where do the rest go?
`x` and `y` are already taken. A further variable has three routes: a refining
channel (`color`, `size`, `shape`, `pattern`, `opacity`), the third position
`z`, or `play`, which gives the variable one frame per value. This chapter uses
all three. Each channel answers one kind of question, and a refining channel
earns the legend that decodes it.

## Which group does each mark belong to?

The scatter of wealth against life expectancy draws every country as an
identical dot. Nothing in it says which continent a country belongs to, and
that third column is categorical. A category is carried by `color`, and the
palette gives each category one hue:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

**One atom away.** The same split without color, when you want the grouping but
not the hues, is the `group` channel; see [Channels](../channels.qmd).

## Will it read in gray?

That plot separates the continents by hue alone. Hue works on a screen, but not
in grayscale print, in poor light, or for a colorblind reader. When a plot has
to work in all three of those conditions, bind `color` *and* `shape` to the same
column. Each group is then said twice:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + point + x(petal_length) + y(sepal_length) +
  color(species) + shape(species) +
  palette("okabe")
```

*"Given the iris flowers: points, x is petal length, y is sepal length, color by
species, shape by species, with the okabe palette."*

Saying one column twice is redundant encoding. `palette()` chooses which colors
the `color` channel uses: here the Okabe and Ito colorblind-safe set. Owning
chapters: [Channels](../channels.qmd), [Setting vs mapping](../style.qmd).

## How big is each observation?

Back to the countries. Population is a third measurement of each one, and both
positions are already taken by wealth and health. A magnitude is carried by
`size`. That channel asks *how much*, so it accepts a continuous column and
nothing else:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  size(population)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, size by
population."* A legend decodes the radius. Owning chapter:
[Channels](../channels.qmd).

## What if four variables matter?

The recipes so far bound `color` to the continent and `size` to the population,
one column at a time. You may need both at once, to see which continent a
country is in and how many people live there. Channels compose: every one of
them works beside every other. That rule is
[Orthogonality](../design-laws.qmd). Wealth, health, population, and continent
in one sentence:

```{r}
data(gapminder_2007) + point +
  x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent, size by population."*

Every channel keeps its own rule and its own legend; none of them changed
meaning by being combined. Four variables is usually the most a single plot can
carry. Past that, use [panels](../faceting.qmd) instead of more channels.

## What if the third variable is continuous?

The third column can be continuous rather than categorical. Population is one.
Here it goes on `color` instead of `size`, so every dot keeps the same size and
the hue carries the number. `color` accepts either kind of column. A continuous
column gets a sequential ramp instead of a palette. A skewed column reads better
on a log ramp, for the same reason a skewed axis does:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
population on a log scale."*

The legend becomes a labeled gradient strip. The channel and its rule did not
change; only the kind of column did. [Scales](../scales.qmd) owns the log scale.

## What if three measurements matter equally?

The iris flowers from the gray-print recipe carry three measurements each, and
none of them is less important than the others. `color` or `size` would make the
third measurement a refinement, and the plot would still be about the other two.
When three *measurements* matter equally, the third one is a position, not a
decoration. One more vowel, and the plot lifts into space:

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is
petal length, color by species."*

**One atom away.** The viewing angle is a property of the space, not of the
data. `turn` rotates the scene sideways, and `tilt` raises the viewing height:

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species) +
  space(turn = -50, tilt = 15)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is
petal length, color by species, turned -50, tilted 15."*

[Space](../space.qmd) owns the third dimension.

## What if the rows form a route?

That scatter is a cloud: each flower is its own dot and the table's order means
nothing. Sometimes the order is what matters. Here are the two gliders from
[Path](../marks/path.qmd) again. Each row records where that glider was a moment
later, so the marks have to be joined in the order the rows were recorded. The
mark that joins rows in table order is `path`:

```{r}
#| echo: false
peek(thermals)
```

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(glider) +
  title("Two gliders, one thermal")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider."*

Each route rises because the gliders do. Nothing in the sentence draws a spiral:
the sentence asks `path` to connect the rows in order, the rows happen to go
around while they go up, and the coil is the result. Look at the crossings as
well. At each one the route nearer to the eye is drawn over the route behind it,
and which route is nearer changes as the two gliders climb. Neither glider is
simply drawn on top of the other.

Turn the view, and the same route answers a different question. No binding
changes:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(glider) +
  space(turn = 20, tilt = 80) +
  title("The same two routes, seen from overhead")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider, turned 20, tilted 80."*

Seen from nearly overhead the climb is no longer visible, and you are looking
down at two circles. The plot now shows how wide each circle is. Bravo's circle
is far tighter than Alpha's, which the side view did not show. That side view
showed Bravo climbing faster, and this one shows that Bravo also turns more
tightly.

One question is left over from that plot. `line`, `step`, `area` and `ribbon`
also join rows into one stroke, so why is `path` the only one of the five that
takes `z`? The reason is the one that separates `path` from `line` in the
plane. A path joins rows in the table's
order, and an order is not a property of any axis, so a third axis does not
change it. A `line` sorts by `x`, and the cube has no left to right. There `x`
is one of three equal positions. At some viewing angles it runs straight into
the page and becomes depth, which would leave a line sorted by an axis you
cannot see. So a rising route is `path`, and `line` refuses `z` rather than
drawing part of it:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude) + color(glider)
```

## What if the shape needs three dimensions?

A route is one kind of three-dimensional fact, and a physical structure is
another. `quakes_fiji` is the thousand earthquakes from [Map](../map.qmd), each
with a place on the globe and a depth below it. Those three columns, `east`,
`north` and `elevation`, show together what no pair of them shows alone:

```{r}
#| echo: false
peek(quakes_fiji)
```

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  color(magnitude) + palette("viridis") + space(turn = 120, tilt = 20) +
  title("A thousand earthquakes, and the sheet they lie on")
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, color by
magnitude."*

The column is named `elevation`, so a measurement below sea level is a negative
number: the deepest quake in the table sits at −680 km.

The quakes are not scattered through the cube. They trace a sheet of ocean floor
sinking into the earth. It begins near the surface at the eastern edge and dips
west, reaching past 600 km about a third of the way across. The shallow quakes
further west belong to a second zone. That sheet lies in three dimensions, so three positions
are what draw it, and any two of them show a flat slice.

Now cut the quakes into slices of depth and run the slices one after another.
`play(slab)` steps through `slab`, a column that holds the depth in 90 km bands:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  space(turn = 120, tilt = 20) + play(slab)
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, played by
slab."*

The band of points moves down and west as the sequence runs. That movement is
the slope of the sheet, read one slice at a time. The frames are not time here.
`play` advances through the levels of an ordered column, and whether that column
measures years or kilometers is a fact about the column, not about `play`. A
cross-section is usually cut on depth, so this is the ordinary cross-section
view. The only new thing is that `play` draws it, and `play` also animates a
year.

The same slices laid out across the page instead of shown one after another in
time:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  space(turn = 120, tilt = 20) | facet(slab, wrap = 4)
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, split into
panel columns by slab, wrapped at 4."*

The eight panels share one set of scales. You compare two depths by looking from
one panel to the next, not by remembering the frame that just played. Choosing
between them is a question about the reader: panels let them look back, frames let them see the movement.
[Play](../play.qmd) owns that choice.

## Which rows and columns are alike?

The recipes so far put a third variable on a channel or on the third position. A
third way is to let the plot's own layout carry it. The `nutrients` table gives
the amount of five nutrients in each of eight foods. The question is which foods
resemble each other, and which nutrients are found in the same foods. Two
categorical axes and one value make a tile plot, and the slot order on each axis
comes from the order the rows arrived in, which means nothing. Build a
[cluster](../cluster.qmd) tree on each axis, then compose the three plots. Each
shared axis takes its leaf order from the tree beside it. Tiles of similar color
then sit next to each other.

```{r}
#| echo: false
peek(nutrients)
```

```{r}
(data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
   theme(height = 150)) /
  ((data(nutrients) + zone + x(food) + y(nutrient) + color(amount)) |
     (data(nutrients) + path * cluster(amount, over = food) + y(nutrient) +
        theme(width = 150)))
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, x is food;
above zones, x is food, y is nutrient, color by amount, beside paths derived by
cluster on amount, over food, y is nutrient."* Owning chapters:
[Cluster](../cluster.qmd), [Zone](../marks/zone.qmd).

## What this section refuses

Population was bound to `size` earlier, and `shape` may look like another
channel that could carry it. The seven glyphs have no order from smallest to
largest, so `shape` refuses a continuous column and names the channels that do
carry magnitude:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + shape(population)
```
