# A third variable, and more

```{r}
#| include: false
source("../R/setup.R")
```

You have more columns than positions. Where do the rest go? Two positions are
spoken for. Every further variable rides a refining channel (color, shape,
size, opacity) or the third position, `z`. Each channel answers one kind of
question, and each mapping earns the legend that decodes it.

## Which group is each mark?

A category rides `color`, and the palette gives each category one hue:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by continent."*

**One atom away.** The same split without color, when you want the
grouping but not the rainbow, is the `group` channel; see
[Many groups](groups.qmd).

## Will it survive a photocopier?

Bind `color` *and* `shape` to the same column and each group is said
twice, readable for color-blind readers, in grayscale print, and in bad
light:

```{r}
data(iris_flowers) + point + x(petal_length) + y(sepal_length) +
  color(species) + shape(species) +
  palette("okabe")
```

`palette()` chooses which colors the `color` channel hands out: here
the Okabe–Ito color-blind-safe set. Owning chapters:
[Channels](../channels.qmd), [Setting vs mapping](../style.qmd).

## How big is each observation?

A magnitude rides `size`; the channel's question is *how much*, so it
only accepts a numeric column:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  size(population)
```

## All of it at once

Channels compose: that is what being orthogonal gives you.
Wealth, health, population, and continent in one sentence:

```{r}
data(gapminder_2007) + point +
  x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) +
  title("The health and wealth of nations, 2007")
```

Every channel keeps its own rule and its own legend; none of them changed
meaning by being combined. Four variables is a sensible ceiling in
practice; beyond it, consider [panels](groups.qmd) instead of more
channels.

## What if the third variable is a number?

`color` accepts either kind of column. A numeric column gets a
sequential ramp instead of a palette, and a skewed one reads better on a
log ramp, exactly like a skewed axis:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log")
```

The legend becomes a labeled gradient strip: same channel, same rule,
different column type. [Scales](../scales.qmd) owns the log story.

## The third position

When three *measurements* deserve equal footing, the third one is a
position, not a decoration. One more vowel and the plot stands up:

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

**One atom away.** The viewing angle is a property of the space, not of
the data; `turn` swings the scene, `tilt` lifts your eye:

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species) +
  space(turn = -50, tilt = 15)
```

[Space](../space.qmd) owns the third dimension.

## What if the rows form a route?

The scatter above is a cloud: each flower is its own dot and the table's
order means nothing. Sometimes the order is the whole story. Two gliders
circling one thermal are measured every few seconds, and each row is
*where that glider was next*, so the marks should be joined, in the
order they were recorded, and the mark that joins rows in table order is
`path`:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(glider) +
  title("Two gliders, one thermal")
```

Each thread rises because the gliders do. Nothing in the sentence draws a
spiral: `path` was told to connect the rows in order, the rows happen to
go round while they go up, and the coil is what that *is*. Watch where the
two threads cross, too. Whichever is nearer covers the other, and which
one that is changes as they climb, so neither glider is simply drawn on
top.

Swing the view and the same route answers a different question, without a
single binding changing:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(glider) +
  space(turn = 20, tilt = 80) +
  title("The same two routes, seen from overhead")
```

From nearly overhead the climb collapses and you are looking down at two
circles. The plot has become a picture of the turn radius, and the reading
it hands you is one the first view buried: Bravo is flying a far tighter
circle than Alpha, which is why Bravo is also the one climbing faster.

**One atom away.** `path` is the only stroke that takes `z`, and the
reason is the one that separates it from `line` in the plane. A path
joins rows in the table's order, and an order belongs to no axis, so it
survives the third dimension untouched. A `line` sorts by `x`, and the
cube has no left to right: at some angles `x` runs straight into the page
and becomes depth rather than position, which would leave a line sorted
by an axis you cannot see. So a rising route is `path`; `line` says so:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude) + color(glider)
```

## What if the shape needs three dimensions?

`quakes_fiji` is a thousand earthquakes recorded near Fiji, each with a place on
the globe and a depth below it. The reason it is here is that no pair of its
columns shows what all three do:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  color(magnitude) + palette("viridis") + space(turn = 120, tilt = 20) +
  title("A thousand earthquakes, and the plate they sit on")
```

They are not scattered through the volume. They lie on a sheet, tilted, running
from shallow near the trench down past 600 km to the west, which is a slab of
ocean floor sinking under the plate beside it. That is a shape, and a shape needs
three positions to be a shape.

Now cut it into 90 km slices of depth and run them:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  space(turn = 120, tilt = 20) + play(slab)
```

The cluster walks down and west as the sequence runs, which is the tilt of that
sheet read one slice at a time. The frames are not time here. `play` advances
through the levels of an ordered column, and whether that column measures years
or kilometers is the column's business, not the channel's. Depth happens to be
the one variable a cross-section is usually cut on, so this is the ordinary
tomographic slice, spelled with the atom that was designed for animation.

The same slices laid out across the page instead of along the clock, which is the
other half of the same split:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  space(turn = 120, tilt = 20) | facet(slab, wrap = 4)
```

Eight cubes, on one shared scale, so the march is a comparison rather than a
memory. Which of the two you want is a question about the reader: panels let them
look back, frames let them see the movement. [Play](../play.qmd) is where that
choice is argued.

## What this section refuses

Glyphs cannot order themselves, so `shape` refuses a numeric column, and
names the channels that do carry magnitude:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + shape(population)
```
