# Two variables

```{r}
#| include: false
source("../R/setup.R")
```

You have two columns. Do they move together, or do they differ by group? The
questions split by type: two measurements ask about
*relationship*; a category and a measurement ask about *comparison*.

## Are two measurements related?

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Points, x is gdp, y is life."* The scatter is the plainest sentence in
the grammar: a mark and its two positions, nothing else.

**One atom away.** When one variable spans orders of magnitude, read its
axis logarithmically. A property of the binding, not a new plot:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life)
```

Owning chapters: [Point](../marks/point.qmd), [Scales](../scales.qmd).

## What is the trend through the cloud?

Layer a derived line over the points; `+` stacks layers in one panel,
and `* smooth` fits where the cloud is heading:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

The positions are written before any mark, so both layers share them;
`color` is written after `point`, so only the points split.
[Encoding scope](../encoding-scope.qmd) owns that rule.

## How much was there, over the range?

A line says *where the value went*; an area says *how much there was*.
It fills to zero so the quantity is the ink:

```{r}
data(actuals) + area + x(year) + y(sales)
```

**One atom away.** If zero is not a meaningful floor for your variable,
you wanted [`line`](../marks/line.qmd), not `area`.

## What is the typical value per group?

A category and a measurement, summarized. The aggregation family
(`mean`, `median`, `sum`, `max`, `min`) all read a bound `y` and collapse
each group to one number:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

**One atom away.** Robust to outliers? Swap the word:

```{r}
data(gapminder_2007) + bar * median + x(continent) + y(life)
```

## What does *every* value per group look like?

Summaries hide spread. The strip plot is the same sentence with the
derivation removed. One dot per row, stacked over each category's slot:

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

When the dots pile onto one line, `* jitter` spreads them sideways, along
the category axis only, so every dot keeps its exact height:

```{r}
data(gapminder_2007) + point * jitter + x(continent) + y(life) + style(opacity = 0.6)
```

**One atom away.** Swap the positions and the strips lie down, useful
when the category names are long. See [Point](../marks/point.qmd) and
[`jitter`](../transforms.qmd).

## The numbers are already the answer: just draw them

When the table already holds one value per category, no transform is
involved; `bar` draws data as it arrives:

```{r}
data(medals) + bar + x(country) + y(gold)
```

**One atom away.** Ranked, because comparison reads best sorted:

```{r}
data(medals) + bar + x(country) + y(gold) + order(gold, desc = TRUE)
```

There is no hidden counting here to surprise you: if you want a
computation, you write its name. That is why this recipe and
[How many of each?](one-variable.qmd) are different sentences.

## What this section refuses

A bar needs exactly one of its axes to be a number. Two categories give
it nothing to measure, and the refusal points at the counting derivation:

```{r}
#| error: true
data(gapminder_2007) + bar + x(country) + y(continent)
```
