# Two variables

```{r}
#| include: false
source("../R/setup.R")
```

You have two columns. What can they tell you together? Their types decide the
question. Two measurements ask about *relationship*: do high values of one go
with high values of the other? A category and a measurement ask about
*comparison*: is the measurement larger in some groups than in others?

## Are two measurements related?

Each country in the table has two measurements, its wealth and its life
expectancy. The question is whether the richer countries also live longer. Two
measurements, one on each position, and nothing else:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."* The scatter is the
plainest sentence in the grammar: a mark and its two positions.

**One atom away.** When one column is spread across orders of magnitude, put its
axis on a log scale. The scale is written on the binding, not as an atom of its
own:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life."*

Owning chapters: [Point](../marks/point.qmd), [Scales](../scales.qmd).

## What is the trend through the cloud?

The scatter above is a cloud of 142 countries. The direction of that cloud is
hard to judge from the dots alone. When the trend is the answer and every
country still has to be seen, layer a derived line over the points. `+` puts
both marks in one panel, and `* smooth` fits a curve through the middle of the
cloud:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

The positions are written before any mark, so both layers share them; `color` is
written after `point`, so only the points split. [Transforms](../transforms.qmd)
owns `smooth`; [Encoding scope](../encoding-scope.qmd) owns that rule.

## How much was there at each time?

The `actuals` table holds one sales figure per year. Sometimes the question is
not whether sales rose but how large they were in each year. A line says *where
the value went*; an area says *how much there was*. It fills to zero, so the
filled area is the amount:

```{r}
data(actuals) + area + x(year) + y(sales)
```

*"Given the actuals: an area, x is year, y is sales."*

**One atom away.** If your column has no meaningful zero, you want
[`line`](../marks/line.qmd), not `area`. Owning chapters:
[Line](../marks/line.qmd), [Area](../marks/area.qmd).

## What is the typical value per group?

Each continent holds many countries, and each country has its own life
expectancy. To compare the continents you need one number for each, so you
summarize the measurement within each category. The aggregation family is
`sum`, `mean`, `median`, `max`, `min` and `quantile`. Each one reads a bound
`y` and reduces every group to a single number. `quantile` also needs a
probability, written `quantile(0.9)`:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

**One atom away.** The median is robust to outliers, so swap `mean` for
`median`:

```{r}
data(gapminder_2007) + bar * median + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by median, x is continent, y is life."*
Owning chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd).

## What does *every* value per group look like?

The bars above gave each continent one number, and that number hides how far
its countries spread. Sometimes you want to see every country, not only the
mean. The strip plot drops the derivation and changes the mark: `point` where
`bar * mean` stood. Each row is one dot, at its own height in its category's
slot:

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Given gapminder 2007: points, x is continent, y is life."*

When the dots pile onto one line, `* jitter` spreads them sideways, along
the category axis only, so every dot keeps its exact height:

```{r}
data(gapminder_2007) + point * jitter + x(continent) + y(life) + style(opacity = 0.6)
```

*"Given gapminder 2007: points derived by jitter, x is continent, y is life."*

**One atom away.** Swap the positions and the strips lie down, useful
when the category names are long. See [Point](../marks/point.qmd) and
[`jitter`](../transforms.qmd).

## What if the table already has the answer?

The `medals` table holds one row per country, and the question is how many
bronze medals each one won. That count is already in the table, so no transform
is involved; `bar` draws the numbers as they arrive:

```{r}
data(medals) + bar + x(country) + y(bronze)
```

*"Given the medals: bars, x is country, y is bronze."*

**One atom away.** Ranked, because a comparison is easiest to read when it
is sorted:

```{r}
data(medals) + bar + x(country) + y(bronze) + order(bronze, desc = TRUE)
```

*"Given the medals: bars, x is country, y is bronze, ordered by bronze, largest
first."*

There is no hidden counting here to surprise you: if you want a
computation, you write its name. That is why this recipe and
[How many of each?](one-variable.qmd#how-many-of-each) are different sentences.

## What this section refuses

Suppose both columns are categorical, a country and its continent, and you
write `bar` with one on each axis. A bar needs at least one of its axes to be a
number: that number is the length of the bar. Two categories leave it nothing
to measure, so the refusal names `count`:

```{r}
#| error: true
data(gapminder_2007) + bar + x(country) + y(continent)
```
