# Two variables

```{r}
#| include: false
source("../R/setup.R")
```

You have two columns. What can they tell you together? The two kinds of column,
continuous and categorical, decide which question you can ask. Two measurements
ask about *relationship*: when one is high, is the other high too? A category
and a measurement ask about *comparison*: is the measurement larger in some
groups than in others?

## Are two measurements related?

The `gapminder_2007` table measures each country's wealth and its life
expectancy. The question is whether the richer countries also live longer. Two
measurements, one on each position, and nothing else:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."* The scatter is the
plainest sentence in the grammar: a mark and its two positions.

**One atom away.** When one column is spread across orders of magnitude, put its
axis on a log scale. The scale is an argument inside the binding, written beside
the column name:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life."*

Owning chapters: [Point](../marks/point.qmd), [Scales](../scales.qmd).

## What is the trend through the points?

The first scatter in this chapter draws 142 countries as a cloud. Its direction
is hard to judge from the dots alone. You want the trend, and you still want
every country on the page. Layering a derived line over the points gives you
both. `+` puts both marks in one panel, and `* smooth` fits one trend curve
through all the points. `color` is written on the points alone, so the
continents show without splitting the curve:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

The positions are written before any mark, so both layers share them. `color` is
written after `point`, so only the points are split into one group per
continent. Owning chapters: [Transforms](../transforms.qmd) for `smooth`,
[Encoding scope](../encoding-scope.qmd) for that rule.

## How much was there at each time?

Two columns can also be a time and an amount. The `actuals` table holds one
sales figure per year. Sometimes the question is not whether sales rose but how
large they were in each year. A `line` says *where* the value was; an `area`
says *how much* there was. The fill is the quantity, so it starts from zero:

```{r}
#| echo: false
peek(actuals)
```

```{r}
data(actuals) + area + x(year) + y(sales)
```

*"Given the actuals: an area, x is year, y is sales."*

**One atom away.** If your column has no meaningful zero, you want
[`line`](../marks/line.qmd), not `area`. Owning chapters:
[Line](../marks/line.qmd), [Area](../marks/area.qmd).

## What is the typical value per group?

The second question this chapter opened with was comparison, a category against
a measurement. Each continent holds many countries, and each country has its own
life expectancy. To compare the continents you need one number for each, so you
summarize the measurement within each category. The aggregation family is
`sum`, `mean`, `median`, `max`, `min` and `quantile`. Each one groups the rows
by `x` and reduces `y` to a single value per group. `quantile` also needs a
probability, written `quantile(0.9)`:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

**One atom away.** A few unusual countries pull a mean toward themselves. A
median barely moves. When that matters, swap `mean` for `median`:

```{r}
data(gapminder_2007) + bar * median + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by median, x is continent, y is life."*
Owning chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd).

## What does *every* value per group look like?

A `bar * mean` gives each continent one number. That number hides how far the
continent's countries spread, and sometimes you want to see every country
rather than the mean. The plot that shows every row is the **strip plot**. It
makes two changes: it removes the derivation, and it writes `point` where
`bar * mean` was. Each row is one dot, at its own height in its category's
slot:

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Given gapminder 2007: points, x is continent, y is life."*

When the dots pile onto one vertical line, `* jitter` spreads them sideways. It
moves them along the category axis only, so every dot keeps its exact height.
The lowered `opacity` lets you see where the dots overlap:

```{r}
data(gapminder_2007) + point * jitter + x(continent) + y(life) + style(opacity = 0.6)
```

*"Given gapminder 2007: points derived by jitter, x is continent, y is life."*

**One atom away.** Swap the positions and the strips run horizontally, which
helps when the category names are long. Owning chapters:
[Point](../marks/point.qmd), [Transforms](../transforms.qmd).

## What if the table already has the answer?

The `medals` table holds one row per country, and the question is how many
bronze medals each one won. That count is already in the table, so no transform
is needed. `bar` draws the numbers exactly as they arrive in the table:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(bronze)
```

*"Given the medals: bars, x is country, y is bronze."*

**One atom away.** `order()` sorts the categories by a column you name, and
`desc = TRUE` puts the largest first. Only Germany and Russia move here, because
the table already lists the countries by their gold medals:

```{r}
data(medals) + bar + x(country) + y(bronze) + order(bronze, desc = TRUE)
```

*"Given the medals: bars, x is country, y is bronze, ordered by bronze, largest
first."*

The engine counts nothing on its own. If you want a computation, you write its
name, which is why this recipe and
[How many of each?](one-variable.qmd#how-many-of-each) are different sentences.
Owning chapters: [Bar](../marks/bar.qmd), [Data](../data.qmd).

## What this section refuses

Every recipe in this chapter has had at least one continuous column. Suppose
both of your columns are categorical, a country and its continent. You write
`bar`, with one column on each axis. A bar needs at least one continuous axis,
because the value on that axis is what sets the length of the bar. Two
categories leave it nothing to measure, so the refusal names `count`:

```{r}
#| error: true
data(gapminder_2007) + bar + x(country) + y(continent)
```
