# Two variables

```{r}
#| include: false
source("../R/setup.R")
```

You have two columns. What can they tell you together? The two kinds of column,
continuous and categorical, decide which question you can ask. Two measurements
ask about *relationship*: when one is high, is the other high too? When one
measurement is a time, the question can be *quantity*: how large was the value
in each year? A category and a measurement ask about *comparison*: is the
measurement larger in some groups than in others?

## Are two measurements related?

The `gapminder_2007` table records each country's GDP per person, in dollars,
and its life expectancy. The question is whether the richer countries also live
longer. The sentence needs only the two measurements, one on each position:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."* The dots rise from left
to right: in general, the richer countries live longer. The sentence for this
plot holds nothing but a mark and its two positions.

**One atom away.** When one column spans several orders of magnitude, as `gdp`
does from about $300 to about $49,000, put its axis on a log scale. The scale is
an argument written inside `x()`, beside the column name:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life."*

The poorer countries are no longer crowded at the left edge, and the dots now
rise in a band that is close to straight. Owning chapters:
[Point](../marks/point.qmd), [Scales](../scales.qmd).

## What is the trend through the points?

The first scatter in this chapter draws one dot for each of 142 countries. The
dots show that life expectancy rises with GDP per person. Below $5,000, though,
they spread from 40 to 74 years, so the typical value is hard to see. You want
the trend, and you still want every country on the page. Layering a derived
line over the points gives you both. `+` puts both marks in one panel, and
`* smooth` fits one trend curve through all the points:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

The curve climbs steeply on the left and much more slowly on the right. Most of
its rise happens below $10,000 of GDP per person. At the far right the curve
runs above the dots of the five richest countries. A smooth curve is least
reliable at the edge of the data, where there are few points.

The positions are written before any mark, so both layers share them. `color` is
written after `point`, so only the points are split into one group per
continent. The curve is fitted to all 142 countries, not to any one continent.
Owning chapters: [Transforms](../transforms.qmd) for `smooth`,
[Encoding scope](../encoding-scope.qmd) for that rule.

## How much was there at each time?

The second question in the opening was quantity, which pairs a time with an
amount. The `actuals` table holds one sales figure per year. Sometimes the
question is not whether sales rose but how large they were in each year. A
`line` says *where* the value was; an `area` says *how much* there was. The fill
is the quantity, so it starts from zero:

```{r}
#| echo: false
peek(actuals)
```

```{r}
data(actuals) + area + x(year) + y(sales)
```

*"Given the actuals: an area, x is year, y is sales."*

Sales stay between 120 and 168 in every year. Measured from zero, the fall in
2021, from 135 to 128, is small.

**One atom away.** Some quantities have no true zero, where zero means none at
all, such as a temperature in degrees Celsius. For those, write
[`line`](../marks/line.qmd) instead of `area`. Owning chapters:
[Line](../marks/line.qmd), [Area](../marks/area.qmd).

## What is the typical value per group?

The last question in the opening was comparison, which pairs a category with a
measurement. Each continent holds many countries, and each country has its own
life expectancy. To compare the continents you need one number for each, so you
summarize the measurement within each category. The transforms that summarize
this way are the aggregation family: `sum`, `mean`, `median`, `max`, `min` and
`quantile`. Each one groups the rows by the category and reduces the
measurement to one value per group. Only `quantile` needs a number of its own,
the probability: `quantile(0.9)` gives the 90th percentile of each group. Here
is `mean`, the average life expectancy in each continent:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

Oceania's countries average 81 years and Europe's 78, while Africa's average 55.

**One atom away.** A few very high or very low values can change a mean
noticeably. They change a median much less. When that matters, swap `mean` for
`median`:

```{r}
data(gapminder_2007) + bar * median + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by median, x is continent, y is life."*

Africa's bar falls by about two years and Asia's rises by about the same. The
other three change by less than a year. Where the two bars differ, a few very
high or very low values have moved the mean. In Africa, seven countries live 71
to 76 years, well above the other 45. Owning chapters: [Bar](../marks/bar.qmd),
[Transforms](../transforms.qmd).

## What does *every* value per group look like?

A `bar * mean` gives each continent one number. That number hides how much life
expectancy differs among the continent's countries. Sometimes you want to see
every country, not the mean. The plot that shows every row is the **strip
plot**. Its sentence differs from the `bar * mean` sentence in two places:
`* mean` is gone, and `point` is written where `bar` was. Each row is one dot, at
its own height in its category's slot:

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Given gapminder 2007: points, x is continent, y is life."*

On each vertical line, dots with nearly the same value cover each other, and
`* jitter` spreads them sideways. It moves them along the category axis only, so
every dot keeps its exact height. `style()` sets a value instead of mapping a
column. Here it sets `opacity` to 0.6, below the default, so you can see where
the dots overlap:

```{r}
data(gapminder_2007) + point * jitter + x(continent) + y(life) + style(opacity = 0.6)
```

*"Given gapminder 2007: points derived by jitter, x is continent, y is life."*

**One atom away.** Swap the positions and the strips run horizontally, which
helps when the category names are long. Owning chapters:
[Point](../marks/point.qmd), [Transforms](../transforms.qmd),
[Setting vs mapping](../style.qmd).

## What if the table already holds the count?

The `mean` and `median` recipes derived a number that was not in the table. The
`medals` table holds one row per country, and the question is how many bronze
medals each one won. That count is already a column, `bronze`, so no transform
is needed. `bar` draws the numbers exactly as the table gives them:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(bronze)
```

*"Given the medals: bars, x is country, y is bronze."*

**One atom away.** `order()` sorts the categories by a column you name, and
`desc = TRUE` puts the largest first. The table already lists the countries by
gold medals, and their bronze counts are in nearly the same order. Only Germany
and Russia change places:

```{r}
data(medals) + bar + x(country) + y(bronze) + order(bronze, desc = TRUE)
```

*"Given the medals: bars, x is country, y is bronze, ordered by bronze, largest
first."*

The engine counts nothing on its own. If you want a computation, you write its
name. That is why [How many of each?](one-variable.qmd#how-many-of-each) writes
`* count` and this recipe does not. Owning chapters: [Bar](../marks/bar.qmd),
[Data](../data.qmd).

## What this section refuses

Every recipe in this chapter used at least one continuous column. Suppose both
of your columns are categorical, a country and its continent. You write `bar`,
with one column on each axis. A bar needs at least one continuous axis, because
the value on that axis is what sets the length of the bar. Two categories give
the bar nothing to measure. The one number you can still make from them is a
count of rows, so the refusal names `count`:

```{r}
#| error: true
data(gapminder_2007) + bar + x(country) + y(continent)
```

Adding `* count` to this sentence still refuses. `bar * count` draws its count
along the axis that has no column, so one axis must stay empty. To count the
countries in each continent, remove `country` and write
`bar * count + x(continent)`, the sentence of
[How many of each?](one-variable.qmd#how-many-of-each).
