# Two variables

```{r}
#| include: false
source("../R/setup.R")
```

You have two columns. What can they tell you together? Their types decide the
question. Two measurements ask about *relationship*: do high values of one go
with high values of the other? A category and a measurement ask about
*comparison*: is the measurement larger in some groups than in others?

## Are two measurements related?

Two measurements, one on each position, and nothing else:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."* The scatter is the
plainest sentence in the grammar: a mark and its two positions.

**One atom away.** When one column is spread across orders of magnitude, put its
axis on a log scale. The scale is written on the binding, not as an atom of its
own:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life."*

Owning chapters: [Point](../marks/point.qmd), [Scales](../scales.qmd).

## What is the trend through the cloud?

Layer a derived line over the points. `+` puts both marks in one panel, and `*
smooth` fits a curve through the middle of the cloud:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

The positions are written before any mark, so both layers share them; `color` is
written after `point`, so only the points split. [Transforms](../transforms.qmd)
owns `smooth`; [Encoding scope](../encoding-scope.qmd) owns that rule.

## How much was there at each time?

A line says *where the value went*; an area says *how much there was*.
It fills to zero, so the filled area is the amount:

```{r}
data(actuals) + area + x(year) + y(sales)
```

*"Given the actuals: an area, x is year, y is sales."*

**One atom away.** If your column has no meaningful zero, you want
[`line`](../marks/line.qmd), not `area`. Owning chapters:
[Line](../marks/line.qmd), [Area](../marks/area.qmd).

## What is the typical value per group?

A category and a measurement, summarized. The aggregation family is `sum`,
`mean`, `median`, `max`, `min` and `quantile`. Each one reads a bound `y` and
reduces every group to a single number. `quantile` also needs a probability,
written `quantile(0.9)`:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

**One atom away.** The median is robust to outliers, so swap `mean` for
`median`:

```{r}
data(gapminder_2007) + bar * median + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by median, x is continent, y is life."*
Owning chapters: [Bar](../marks/bar.qmd), [Transforms](../transforms.qmd).

## What does *every* value per group look like?

Summaries hide spread. The strip plot drops the derivation and changes the
mark: `point` where `bar * mean` stood. Each row is one dot, at its own height
in its category's slot:

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

When the dots pile onto one line, `* jitter` spreads them sideways, along
the category axis only, so every dot keeps its exact height:

```{r}
data(gapminder_2007) + point * jitter + x(continent) + y(life) + style(opacity = 0.6)
```

*"Given gapminder 2007: points derived by jitter, x is continent, y is life."*

**One atom away.** Swap the positions and the strips lie down, useful
when the category names are long. See [Point](../marks/point.qmd) and
[`jitter`](../transforms.qmd).

## What if the table already has the answer?

When the table already holds one value per category, no transform is
involved; `bar` draws data as it arrives:

```{r}
data(medals) + bar + x(country) + y(bronze)
```

**One atom away.** Ranked, because a comparison is easiest to read when it
is sorted:

```{r}
data(medals) + bar + x(country) + y(bronze) + order(bronze, desc = TRUE)
```

*"Given the medals: bars, x is country, y is bronze, ordered by bronze, largest
first."*

There is no hidden counting here to surprise you: if you want a
computation, you write its name. That is why this recipe and
[How many of each?](one-variable.qmd#how-many-of-each) are different sentences.

## What this section refuses

A bar needs at least one of its axes to be a number: that number is the length
of the bar. Two categories leave it nothing to measure, so the refusal names
`count`:

```{r}
#| error: true
data(gapminder_2007) + bar + x(country) + y(continent)
```
