# Area

```{r}
#| include: false
source("../R/setup.R")
```

How large was the quantity at each date? `area` fills the
region between the data and a baseline at zero. Its top edge is the boundary a
`line` would draw. The fill below that edge is what makes the plot read as a
quantity rather than as a path.

## Basic usage

How did sales change over five years? Everything in this chapter is added to
one sentence: a source, the mark, and its two positions.

```{r}
#| echo: false
peek(actuals)
```

```{r}
data(actuals) + area + x(year) + y(sales) +
  y_label("Sales") + title("Sales, 2019 to 2023")
```

*"Given the actuals: an area, x is year, y is sales."*

Notice the fill reaches the left and right edges of the panel. With `point`,
the axis ends a little past the data. A dot drawn at the last value would
otherwise be cut in half by the panel edge. An area has no dots. Its
edge is the data itself, so nothing can be cut off, and the axis ends exactly
at the first and last year. [Scales](../scales.qmd#the-axis-fits-your-data)
states the rule in full.

An area's two axes are not interchangeable, and the rest of this chapter
follows from that. `x` is the **domain**, the axis the area is read along. `y`
is the **measure**, the quantity it fills to. A `line` has the same two axes,
and [the profile](line.qmd#a-category-on-the-domain-the-profile) section of
the Line chapter states the rule. A bar draws horizontally when you swap its
bindings. A line does not, because its measure is always `y`. An area follows
the same rule, so it has no horizontal form either. The one exception is
`density`, whose curve spreads along whichever axis is continuous;
[With transforms](#with-transforms) shows both of its forms.

## The baseline has to be zero

A `line` says *where* the value was; an `area` says *how much* there was. The
fill is the quantity, so it has to start from zero. An area drawn from any
other baseline would show a height that measures nothing. gog does not stop
you from starting the axis somewhere else. `y(sales, limits = c(100, NA))`
starts it at 100, the fill starts there too, and no warning is printed. That
decision is yours to make: the fill then measures from 100 rather than from
zero, and the reader of your plot has to know it.
[Scales](../scales.qmd#limits-when-your-data-is-not-the-authority) explains
`limits`.

So the choice of mark is about the data, not the appearance. Sales have a true
zero, so the first plot is correct: the fill is the sales. Life expectancy in
years has a true zero too, and several plots below fill to it. A quantity whose
zero is only a convention has no amount for a fill to show. A temperature in
degrees Celsius is one, and so is a calendar year. Draw those with `line`.

Because life expectancy has a true zero, either mark suits it, and the choice
depends on what the reader should compare. An area and a line draw the same
top edge. Drawn with `line`, split by `color` as the [Line](line.qmd) chapter
showed, that edge says where each country's value was, and nothing about how
much:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  title("Life expectancy as paths, not quantities")
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

## Multiple series

The table holds five countries, and the sentence has to name the split.
Without a split, the boundary runs through all sixty rows as one region.
`color` splits the mark into one region per category and names each one in a
legend. It is the same split that gives `line` one stroke per category:

```{r}
data(gapminder_asia) + area + x(year) + y(life) + color(country) +
  style(opacity = 0.45) +
  title("One region per country")
```

*"Given gapminder Asia: areas, x is year, y is life, color by country."*

Five lines can cross without hiding each other, because a line is a thin
stroke. Five areas cannot: a fill covers whatever lies behind it, and at the
default opacity only a little shows through. Write the same sentence with no
opacity, and gog prints a warning before it draws:

```{r}
#| message: true
data(gapminder_asia) + area + x(year) + y(life) + color(country) +
  title("Five regions, no opacity set: the last one drawn covers the rest")
```

Compare the legend with the plot. Every country is named there, but you cannot
see five regions. Each one covers the ones drawn before it, and only the part
of a region that rises above every later one is left to read. The warning is
advice rather than a refusal: the sentence is legal, so gog draws it.

The other way to draw it is to stack the regions instead of overlaying them.
That is written `area * stack`. `stack` is a collision modifier: it computes
nothing, and it moves marks that would otherwise be drawn in the same place.
Its sibling `dodge` subdivides a *width*, and an area has no width. So of the
two, only `stack` applies to an area. Each region starts at the top of the ones
below it, so the five form one band and none is covered:

```{r}
data(gapminder_asia) + area * stack + x(year) + y(population) + color(country) +
  y_label("Population") +
  title("Five Asian countries, population stacked")
```

*"Given gapminder Asia: areas derived by stack, x is year, y is population,
color by country."*

Each country's region is as thick as its population, and the band's top edge is
the five-country total. This plot measures `population` instead of `life` on
purpose. You stack quantities that *add*: populations add up to a total that
means something, and life expectancies do not. gog does not stop you from
stacking them. `area * stack + y(life)` draws five bands to a top edge near
370 years, with no warning, and that number means nothing. So use `stack`
with care: you always get a plot, and only you know whether the parts add up.

Which country's share of the five grew? The stacked plot cannot say, because
every region's baseline moves with the ones below it. `stack(share = TRUE)`
divides each year's pile by that year's total. The band's top edge is then flat
at 1, and each region's thickness is that country's share:

```{r}
data(gapminder_asia) + area * stack(share = TRUE) + x(year) + y(population) +
  color(country) +
  y_label("Share of population") +
  title("Five Asian countries, population as shares")
```

*"Given gapminder Asia: areas derived by stack as shares, x is year, y is
population, color by country."*

India's share rose from 0.33 to 0.39, and Japan's fell from 0.08 to 0.05. The
total is gone from this plot: every year reaches 1, so it answers the share
question and not the size question. Draw both when you need both answers. The
[Transforms](../transforms.qmd#stacked-areas-the-case-stack-was-built-for)
chapter covers `stack` in full, with a section on stacked areas.

## Giving an area an edge

A low opacity, like the one on the split regions above, leaves a faint top
edge, and the top edge is where the data is. A stroke along it makes the
boundary easy to read. An area takes no
border of its own. The edge on top of a filled region is a `line`, so you layer
one over the other:

```{r}
data(actuals) + x(year) + y(sales) +
  area + style(color = "steelblue", opacity = 0.25) +
  line + style(color = "steelblue", size = 2.5) +
  title("area + line: the fill and its edge, composed")
```

*"Given the actuals: x is year, y is sales, an area and also a line."*

There is no border setting on `area` to use instead. Layering a `line` gives
the edge its own color, width and opacity, which one border setting could not.

## One region, one fill

Can you map a column to an area's opacity, the way the plot above set one? No,
and the reason is the shape. An area is a single filled region, so `opacity` is
a *setting* here and not a channel.
Each row of the table is one corner of the top edge, not a piece of the fill.
No part of the region belongs to one row, so a column cannot make one row
lighter or darker than another. `opacity` is a setting on `line` for the same
reason.

`size` is refused both ways: `area` neither maps it nor sets it. A stroke has a
width, so `line` lets you set one. An area's extent comes from its x, its y and
its baseline, so there is no size to choose. Both marks take `style(color = )`
and `style(opacity = )`, and [What it refuses](#what-it-refuses) shows the
refusals for `opacity` and `size`, after the two that come from an axis.

## With transforms

Apart from the stack, every plot so far filled to values already in the table.
A transform computes new values first, and the area fills to what it computed.
A distribution, for example, can be drawn as a filled shape. The question is
the one a histogram answers: how many rows fall in each interval? `bin`,
`density` and the aggregation family mean the same thing on an area as on a
bar. `bin` counts the rows in each interval.
The boundary then runs through the bin centers, and the region below it is
filled. That shape is the frequency polygon `line * bin` draws, with its inside
filled:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + area * bin + x(life) +
  x_label("Life expectancy (years)") +
  title("area * bin: a filled frequency polygon")
```

*"Given gapminder 2007: an area derived by bin, x is life."*

`bin(30)` and `bin(width = 5)` set the intervals here exactly as they do for a
bar; [Choosing the bin count](../transforms.qmd#choosing-the-bin-count) shows
both.

`density` replaces the counts with a smooth curve, and the area fills the
region under it:

```{r}
data(gapminder_2007) + area * density + x(life) +
  x_label("Life expectancy (years)") +
  title("area * density")
```

*"Given gapminder 2007: an area derived by density, x is life."*

`density` is not the area's own transform. Five other marks estimate the same
curve, and only the shape changes:

| Mark derived by `density` | Shape |
|---|---|
| `line * density` | the curve as a stroke |
| `step * density` | the curve as a staircase |
| `point * density` | one dot at each point the curve was estimated |
| `bar * density` | one thin bar at each of those points |
| `ribbon * density` | the curve mirrored, which is the violin |
| `area * density` | the curve filled to the baseline, the plot above |

Put `line`, `step`, `point` or `bar` in place of `area` in the sentence above,
and each one draws. `ribbon` refuses it, because a ribbon spans between two
boundaries and one curve gives only one. With a category,
`ribbon * density + x(<category>) + y(<number>)` draws the violin.

Give that same sentence a category on `x` and a number on `y`, and the curve
turns. Life now runs along `y`, and the estimate is drawn across each
category's slot. The domain takes a category as well as a number, and [A
category on the domain](#a-category-on-the-domain) uses that in full.
[`ribbon * density`](ribbon.qmd#the-violin-a-distribution-per-category) draws
the mirrored form, and that section says what the width of a violin can mean.
This plot is the half violin:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  color(continent) +
  y_label("Life expectancy") +
  title("area * density + x(continent): the half violin")
```

*"Given gapminder 2007: areas derived by density, x is continent, y is life,
color by continent."*

The two sentences differ in three places: `x` moved from `life` to
`continent`, `y(life)` was added, and so was `color(continent)`. With only
`x(life)`, life runs along `x` and `y` is free, so `y` carries the density
estimate, as it carried the count for `bin`. With a category on `x` and a
number on `y`, no axis is free, so each estimate is drawn across its
category's slot, one distribution per group. `color` does what it did in
[Multiple series](#multiple-series): a categorical column gives each region
its own fill and a legend. On a `line` the same channel colors the stroke; on
an area it colors the fill. The slots already keep the continents apart, so
here `color` changes only the fill of each one.

Can the two bindings be swapped? For `density`, yes. Put life on `x` and the
category on `y`, and each curve runs along life, rising from its own
category's place on the `y` axis. That is the **ridgeline plot**, and it is the
half violin's sentence with `x` and `y` exchanged:

```{r}
data(gapminder_2007) + area * density + x(life) + y(continent) +
  color(continent) +
  x_label("Life expectancy") + title("area * density + x(life): the ridgeline")
```

*"Given gapminder 2007: areas derived by density, x is life, y is continent,
color by continent."*

[Ribbon](ribbon.qmd#sideways-and-halved) draws all four forms together: one
side or two, and the category on either axis.

After `bin` and `density`, the third kind is the aggregation family. An
aggregation runs before the fill: it leaves one value per year, and the area
fills to those values, exactly as a bar would:

```{r}
data(gapminder_asia) + area * mean + x(year) + y(life) +
  y_label("Mean life expectancy") +
  title("area * mean: five countries averaged per year")
```

*"Given gapminder Asia: an area derived by mean, x is year, y is life."*

## A category on the domain

That mean ran over years, a continuous domain. Put a category there, and one
summary per continent is usually drawn as bars. An area draws the five
continent means as one connected shape instead. Use it when the reader should
compare the shape rather than five heights. Each category gets one slot, the
boundary runs from slot to slot in axis order, and the region still fills down
to zero:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("area * mean + x(continent): the filled profile")
```

gog draws that sentence. The boundary runs from Asia to Oceania in axis order,
and the fill reaches from each mean down to zero. Whether the plot claims more
than was measured depends on the segments between the categories. Nothing was
measured between Africa and the Americas, and a
straight segment is drawn there anyway. A line across that gap claims a change.
A fill claims a quantity as well, at every point between the two categories. If
you read only the vertices, the segments help you compare the five heights. If
you read the segments themselves, they claim what was never measured. To
compare the five means and nothing else, swap `area` for `bar`: one atom away,
and no segments to read.

Keep the fill when the shape itself is what the reader should compare. A shape
can also be drawn on a circular axis. Add
[`polar()`](../polar.qmd#the-radar-a-categorical-angle) and the same sentence
draws the filled radar:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) + polar() +
    title("area * mean + polar(): the filled radar")
```

*"Given gapminder 2007: an area derived by mean, x is continent, y is life, in
polar."*

## What you can set

A setting changes how the region looks without reading a column, and each mark
takes its own. These are an area's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("area")
```

That is the shortest settings list in the book, shared with `ribbon`. `line`
has one more, `size`, and `bar` has two, `border_color` and `border_size`.
Neither gap is an omission: [One region, one fill](#one-region-one-fill)
explains `size`, and [Giving an area an edge](#giving-an-area-an-edge) explains
the border. The one setting this chapter has not shown yet is `pattern`, a
texture on the fill. The edge is still a `line`:

```{r}
data(actuals) + x(year) + y(sales) +
  area + style(color = "seagreen", pattern = "hatch", opacity = 0.5) +
  line + style(color = "seagreen", size = 2) +
  y_label("Sales") + title("A textured fill, and a line for the edge")
```

*"Given the actuals: x is year, y is sales, an area colored seagreen, with
pattern hatch and opacity 0.5, and also a line colored seagreen, with size 2."*

That layered line can be dashed, colored differently, or smoothed, and it traces
only the data boundary. A border would also draw a line along the two sides and
the baseline.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The first two refusals come from an axis. The first is the measure. A
region fills to a baseline at zero, and a category has none. So a categorical
`y` is refused, and the message asks for the category on `x` instead:

```{r}
#| error: true
data(gapminder_2007) + area * mean + x(life) + y(continent)
```

`density` is the exception, because its curve spreads along the continuous
axis instead of filling to a baseline. With `y(continent)` it draws the
ridgeline in [With transforms](#with-transforms).

The second comes from the domain. `bin` cuts a continuous axis into intervals,
and a category is one slot with no width to cut. `density` needs a continuous
axis too, and either `x` or `y` can be that axis, so it refuses only when
neither is continuous. The message names the transform that works on
categories: `count` in place of `bin`:

```{r}
#| error: true
data(gapminder_2007) + area * bin + x(continent)
```

The message for a `density` on a category names `proportion` in the same way.

The last two come from the fill. A mapped `opacity` is the third.
Each row is one corner of the top edge, so no part of the region belongs to a
single row, as [One region, one fill](#one-region-one-fill) explained.

```{r}
#| error: true
data(gapminder_asia) + area + x(year) + y(life) + opacity(gdp)
```

A `size` is the fourth, and it is refused both ways. An area's extent comes
from its x, its y and its baseline, so there is no size to map and none to
set:

```{r}
#| error: true
data(actuals) + area + x(year) + y(sales) + style(size = 4)
```

A `line` accepts a size, because a stroke has a width to choose.

## In short

An area measures on `y`, and never draws horizontally the way a bar does.
`density` is the one exception, and its curve follows whichever axis is
continuous. Its interior belongs to no single row, so `opacity` can only be
set. Its extent comes from its x, its y and its baseline, so `size` is refused
both ways.

## What comes next

An area slopes from one measurement to the next, and that slope claims values
nobody recorded. [Step](step.qmd) holds each value until the next measurement,
and jumps where it changes.
