# Area

```{r}
#| include: false
source("../R/setup.R")
```

How large was the quantity at each date? `area` fills the
region between the data and a baseline at zero. Its top edge is the boundary a
`line` would draw. The fill below that edge is what makes the plot read as a
quantity rather than as a stroke.

## Basic usage

How large were sales in each of five years? Everything in this chapter is added
to one sentence, the smallest that draws an area: a source, the mark, and its
two positions.

```{r}
#| echo: false
peek(actuals)
```

```{r}
data(actuals) + area + x(year) + y(sales) +
  y_label("Sales") + title("Sales, 2019 to 2023")
```

*"Given the actuals: an area, x is year, y is sales."*

Notice that the fill reaches the left and right edges of the panel. With `point`,
the axis ends a little past the data. A dot drawn at the last value would
otherwise be cut in half by the panel edge. An area has no dots. Its
edge is the data itself, so nothing can be cut off, and the axis ends exactly
at the first and last year. [Scales](../scales.qmd#the-axis-fits-your-data)
states the rule in full.

An area's two axes are not interchangeable, and the rest of this chapter
follows from that. `x` is the **domain**, the axis the area is read along. `y`
is the **measure**, the quantity it fills to. A `line` has the same two axes. A
bar draws horizontally when you swap its bindings. A line does not, because its
measure is always `y`, as [the
profile](line.qmd#a-category-on-the-domain-the-profile) section of the Line
chapter explains. An area follows the same rule, so it has no horizontal form
either. The one exception is `density`, whose curve runs along whichever axis
is continuous; [With transforms](#with-transforms) shows both of its forms.

## The baseline has to be zero

A `line` says *where* the value was; an `area` says *how much* there was. The
fill is the quantity, so it has to start from zero. An area drawn from any
other baseline would show a height that measures nothing. gog does not stop
you from starting the axis somewhere else. `y(sales, limits = c(100, NA))`
starts it at 100, the fill starts there too, and no warning is printed. That
decision is yours to make: the fill then shows sales minus 100 rather than the
sales, and the reader of your plot has to know it.
[Scales](../scales.qmd#limits-stating-the-range-yourself) explains
`limits`.

So the choice of mark is about the data, not the appearance. Sales have a
**true zero**, where zero means none at all, so the first plot is correct: the
fill is the sales. Life expectancy in years has a true zero too, and several
plots below fill to it. A quantity whose zero is only a convention has no
amount for a fill to show. A temperature in degrees Celsius is one, and so is a
calendar year. Draw those with `line`.

Life expectancy has a true zero, so the data allows either mark, and the
question you are asking decides between them. An area and a line draw the same
top edge. Here is that edge drawn with `line` and split by `color`, as in the
[Line](line.qmd) chapter. It shows where each country's value was, and nothing
about how much:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  title("Life expectancy as paths, not quantities")
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

## Multiple series

The table holds five countries, so the sentence has to name the column that
splits them. Without a split, the boundary runs through all sixty rows as one
region. `color` splits the mark into one region per category and names each
one in a legend. It is the same split that gives `line` one stroke per
category. The sentence also sets a low opacity, so the regions drawn first
show through the ones drawn over them:

```{r}
data(gapminder_asia) + area + x(year) + y(life) + color(country) +
  style(opacity = 0.45) +
  title("One region per country")
```

*"Given gapminder Asia: areas, x is year, y is life, color by country."*

Five lines can cross without hiding each other, because a line is a thin
stroke. Five areas cannot: a fill covers whatever lies behind it, and at the
default opacity only a little shows through. Write the same sentence with no
opacity, and gog prints a warning before it draws:

```{r}
#| message: true
data(gapminder_asia) + area + x(year) + y(life) + color(country) +
  title("Five regions, no opacity set: each one covers the ones before it")
```

Compare the legend with the plot. Every country is named there, but five
regions cannot be distinguished. Each region covers the ones drawn before it,
so you can read a region only where it rises above every region drawn after
it. The warning is advice rather than a refusal: the sentence is legal, so gog
draws it.

The other way to draw it is to stack the regions instead of overlaying them.
That is written `area * stack`. `stack` is a collision modifier: it moves marks
that would otherwise be drawn in the same place, rather than summarizing them.
Its sibling `dodge` subdivides a *width*, and an area has no width. So of the
two, only `stack` applies to an area. Each region starts at the top of the ones
below it, so the five form one band and none is covered:

```{r}
data(gapminder_asia) + area * stack + x(year) + y(population) + color(country) +
  y_label("Population") +
  title("Five Asian countries, population stacked")
```

*"Given gapminder Asia: areas derived by stack, x is year, y is population,
color by country."*

Each country's region is as thick as its population, and the band's top edge is
the five-country total. This plot measures `population` instead of `life` on
purpose. You stack quantities that *add*: populations add up to a total that
means something, and life expectancies do not. gog does not stop you from
stacking them. Put `y(life)` in place of `y(population)` in the sentence above,
and gog draws five bands to a top edge near 370 years, with no warning. That
number means nothing. So use `stack` with care: you always get a plot, and only
you know whether the sum of the parts means something.

Which country's share of the five grew? The stacked plot cannot say, because
every region's baseline moves with the ones below it. `stack(share = TRUE)`
divides each year's pile by that year's total. The band's top edge is then flat
at 1, and each region's thickness is that country's share:

```{r}
data(gapminder_asia) + area * stack(share = TRUE) + x(year) + y(population) +
  color(country) +
  y_label("Share of population") +
  title("Five Asian countries, population as shares")
```

*"Given gapminder Asia: areas derived by stack as shares, x is year, y is
population, color by country."*

Between 1952 and 2007, India's share rose from 0.33 to 0.39, and Japan's fell
from 0.08 to 0.05. The total is gone from this plot: every year reaches 1, so it
shows which shares grew and no longer how large the total was. Draw both when
you need both answers. The
[Transforms](../transforms.qmd#stacked-areas-the-case-stack-was-built-for)
chapter covers `stack` in full, with a section on stacked areas.

## Giving an area an edge

A low opacity, like the 0.45 in [Multiple series](#multiple-series), leaves a
faint top edge, and the top edge is where the data is. A stroke along it makes
the boundary easy to read. An area takes no border of its own. The edge on top
of a filled region is a `line`, so you layer one over the other:

```{r}
data(actuals) + x(year) + y(sales) +
  area + style(color = "steelblue", opacity = 0.25) +
  line + style(color = "steelblue", size = 2.5) +
  title("area + line: the fill and its edge, composed")
```

*"Given the actuals: x is year, y is sales, an area and also a line."*

There is no border setting on `area` to use instead. A layered `line` gives the
edge its own color, width, opacity and dash. A bar's border settings,
`border_color` and `border_size`, give only a color and a width.

## One region, one fill

The plot above set one opacity for the whole area. Can you map a column to an
area's opacity instead? No, and the reason is the shape. An area is a single
filled region, so `opacity` is a *setting* here and not a channel. Each row of
the table is one vertex, a corner of the top edge, not a piece of the fill.
No part of the region belongs to one row, so a column cannot make one row
lighter or darker than another. `opacity` is a setting on `line` for the same
reason.

`size` is refused both ways: `area` neither maps it nor sets it. A stroke has a
width, so `line` lets you set one. An area's extent comes from its x, its y and
its baseline, so there is no size to choose. Both marks take `style(color = )`
and `style(opacity = )`. [What it refuses](#what-it-refuses) shows the refusals
for `opacity` and `size`.

## With transforms

Apart from the stack, every plot so far filled to values already in the table.
A transform computes new values first, and the area fills to what it computed.
A distribution, for example, can be drawn as a filled shape. The question is
the one a histogram answers: how many rows fall in each interval? `bin`,
`density` and the
[aggregation family](../transforms.qmd#aggregation-family-sum-mean-median-max-min-quantile)
mean the same thing on an area as on a bar. `bin` counts the rows in each
interval. The boundary then runs through the bin centers, and the region below
it is filled. That shape is the frequency polygon `line * bin` draws, with its
inside filled:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + area * bin + x(life) +
  x_label("Life expectancy (years)") +
  title("area * bin: a filled frequency polygon")
```

*"Given gapminder 2007: an area derived by bin, x is life."*

`bin(30)` and `bin(width = 5)` set the intervals here exactly as they do for a
bar; [Choosing the bin count](../transforms.qmd#choosing-the-bin-count) shows
both.

`density` replaces the counts with a smooth curve, and the area fills the
region under it:

```{r}
data(gapminder_2007) + area * density + x(life) +
  x_label("Life expectancy (years)") +
  title("area * density")
```

*"Given gapminder 2007: an area derived by density, x is life."*

`density` is not the area's own transform. Five other marks estimate the same
curve, and only the shape changes:

| Mark derived by `density` | Shape |
|---|---|
| `line * density` | the curve as a stroke |
| `step * density` | the curve as a staircase |
| `point * density` | one dot at each point where the curve was estimated |
| `bar * density` | one thin bar at each of those points |
| `ribbon * density` | with a category, the curve mirrored: the violin |
| `area * density` | the curve filled to the baseline, the plot above |

Put `line`, `step`, `point` or `bar` in place of `area` in the sentence above,
and each one draws. `ribbon` refuses it, because a ribbon spans between two
boundaries and one curve gives only one. With a category,
`ribbon * density + x(<category>) + y(<number>)` draws the violin.

Give the `area * density` sentence a category on `x` and a number on `y`, and
the curve runs along `y` instead of `x`. One estimate is drawn across each
category's slot. This plot is the **half violin**:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  color(continent) +
  y_label("Life expectancy") +
  title("area * density + x(continent): the half violin")
```

*"Given gapminder 2007: areas derived by density, x is continent, y is life,
color by continent."*

The two sentences differ in three places: `x` moved from `life` to
`continent`, `y(life)` was added, and so was `color(continent)`. With only
`x(life)`, life runs along `x` and `y` holds no column, so `y` carries the
density estimate, as it carried the count for `bin`. With a category on `x` and
a number on `y`, both axes hold a column, so each estimate is drawn across its
category's slot, one distribution per continent. `color` does what it did in
[Multiple series](#multiple-series): a categorical column gives each region
its own fill and a legend. On a `line` the same channel colors the stroke; on
an area it colors the fill. The slots already keep the continents apart, so
here `color` splits nothing; it gives each slot its own fill and a legend
entry.

Can the two bindings be swapped? For `density`, yes. Put life on `x` and the
category on `y`, and each curve runs along life, rising from its own
category's place on the `y` axis. That is the **ridgeline plot**, and it is the
half violin's sentence with `x` and `y` exchanged:

```{r}
data(gapminder_2007) + area * density + x(life) + y(continent) +
  color(continent) +
  x_label("Life expectancy") + title("area * density + x(life): the ridgeline")
```

*"Given gapminder 2007: areas derived by density, x is life, y is continent,
color by continent."*

[Ribbon](ribbon.qmd#sideways-and-halved) shows all four forms in one section:
one side or two, and the category on either axis. Its
[violin section](ribbon.qmd#the-violin-a-distribution-per-category) says what
the width of a violin can mean.

After `bin` and `density`, the third kind of transform is the aggregation
family: `sum`, `mean`, `median`, `max`, `min` and `quantile`. Each leaves one
value per year, and the area fills to those values, exactly as a bar would:

```{r}
data(gapminder_asia) + area * mean + x(year) + y(life) +
  y_label("Mean life expectancy") +
  title("area * mean: five countries averaged per year")
```

*"Given gapminder Asia: an area derived by mean, x is year, y is life."*

## A category on the domain

That mean was computed for each year, and year is a continuous domain. Put a
category there, and one summary per continent is usually drawn as bars. An
area draws the five continent means as one connected shape instead. Use it when
the reader should see one overall shape rather than compare five heights. Each
category gets one slot, the boundary runs from slot to slot in axis order, and
the region still fills down to zero:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("area * mean + x(continent): the filled profile")
```

gog draws that sentence. The boundary runs from Asia to Oceania in axis order,
and the fill reaches from each mean down to zero. The segments between the
categories are where the plot can claim more than was measured. Nothing was
measured between Africa and the Americas, and a straight segment is drawn there
anyway. A line across that gap claims a change.
A fill claims a quantity as well, at every point between the two categories. If
you read only the vertices, the segments help you compare the five heights. If
you read the segments themselves, they claim what was never measured. To
compare the five means and nothing else, swap `area` for `bar`: one atom away,
and no segments to read.

Keep the fill when the overall shape is what the reader should see. Continents
have no order, but a straight axis still puts one of them first and another
last. A circle has no first or last place. Add
[`polar()`](../polar.qmd#the-radar-a-categorical-angle), and the same sentence
draws the filled **radar chart**:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) + polar() +
    title("area * mean + polar(): the filled radar")
```

*"Given gapminder 2007: an area derived by mean, x is continent, y is life, in
polar."*

## What you can set

A setting changes how the region looks without reading a column, and each mark
takes its own. These are an area's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("area")
```

That is the shortest settings list in the book, shared with `ribbon`. `line`
has one more, `size`, and `bar` has two, `border_color` and `border_size`.
Both are left out on purpose: [One region, one fill](#one-region-one-fill)
explains `size`, and [Giving an area an edge](#giving-an-area-an-edge) explains
the border. The one setting this chapter has not shown yet is `pattern`, a
texture on the fill. The plot below hatches the fill and draws the edge with a
layered `line`, as before:

```{r}
data(actuals) + x(year) + y(sales) +
  area + style(color = "seagreen", pattern = "hatch", opacity = 0.5) +
  line + style(color = "seagreen", size = 2) +
  y_label("Sales") + title("A textured fill, and a line for the edge")
```

*"Given the actuals: x is year, y is sales, an area colored seagreen, with
pattern hatch and opacity 0.5, and also a line colored seagreen, with size 2."*

That layered line traces only the top boundary. A border would also draw a line
along the two sides and the baseline.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The first two refusals come from an axis. The first comes from the measure. A
region fills to a baseline at zero, and a category has no zero. So a
categorical `y` is refused, and the message asks for the category on `x`
instead:

```{r}
#| error: true
data(gapminder_2007) + area * mean + x(life) + y(continent)
```

`density` is the exception. Its curve runs along the continuous axis, and each
category's place on `y` is the baseline its curve fills to. With `y(continent)`
it draws the ridgeline in [With transforms](#with-transforms).

The second comes from the domain. `bin` cuts a continuous axis into intervals,
and a category is one slot with no width to cut. The message names `count`, the
transform that tallies the rows in each category:

```{r}
#| error: true
data(gapminder_2007) + area * bin + x(continent)
```

`density` needs a continuous axis too, but either `x` or `y` can be that axis,
so it refuses only when neither is continuous. Its message names `proportion`
in the same way.

The last two come from the fill. A mapped `opacity` is the third. Each row is
one vertex of the top edge, so no part of the region belongs to a single row,
as [One region, one fill](#one-region-one-fill) explained.

```{r}
#| error: true
data(gapminder_asia) + area + x(year) + y(life) + opacity(gdp)
```

A `size` is the fourth, and it is refused both ways. An area's extent comes
from its x, its y and its baseline, so there is no size to map and none to
set:

```{r}
#| error: true
data(actuals) + area + x(year) + y(sales) + style(size = 4)
```

A `line` accepts a size, because a stroke has a width to choose.

## In short

An area measures on `y`, so it does not draw horizontally the way a bar does.
The one exception is `density`, whose curve runs along whichever axis is
continuous. An area's interior belongs to no single row, so `opacity` can only
be set. Its extent comes from its x, its y and its baseline, so `size` is
refused both ways.

## What comes next

An area slopes from one measurement to the next, and that slope claims the value
changed steadily in between. [Step](step.qmd) claims instead that each value
held until the next measurement, and then jumped.
