# Area

```{r}
#| include: false
source("../R/setup.R")
```

How much was there at each point in time, for each group? `area` fills the
region between the data and a baseline at zero. It is `line` plus everything
between it and the baseline: the same boundary, drawn as a quantity rather than
as a path.

## Basic usage

How did life expectancy move in five Asian countries, drawn as a quantity
rather than as a path? Everything in this chapter is added to one sentence: a
source, the mark, and its two positions. The table holds five countries, so
`group` keeps their regions apart, and the opacity lets each region show
through the others.

```{r}
data(gapminder_asia) + area + x(year) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("Life expectancy, five Asian countries")
```

*"Given gapminder Asia: areas, x is year, y is life, grouped by country."*

An area's two axes are different kinds of thing, and the rest of this chapter
follows from that difference. `x` is the **domain**, the axis the boundary is
read along; `y` is the **measure**, the quantity the region fills to. That is
why an area has no horizontal form the way a `bar` does. A bar takes its measure
axis from the columns you bind. An area's measure is always `y`.

Notice the fill reaches the left and right edges of the panel. A filled region
has no glyph to keep clear of the panel edge. So the domain axis stops at the
data, instead of leaving the margin a scatter gets at each end.
[Scales](../scales.qmd#the-axis-fits-your-data) explains why.

## The baseline has to be zero

A `line` says *where*; an `area` says *how much*. The fill is the quantity, so
it has to start from zero. An area drawn from any other baseline would show a
height that measures nothing.

That makes the mark a real choice rather than decoration. Life expectancy has no
meaningful zero, and the plot above fills every year from zero up to each curve.
Sales do have a meaningful zero, and the same mark reads correctly:

```{r}
data(actuals) + area + x(year) + y(sales) +
  y_label("Sales") + title("Sales: zero means zero")
```

When the quantity you are drawing has no zero, `line` is the mark to use:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  title("The same data, as paths rather than quantities")
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

## Multiple series

The opening plot drew five regions in one color, so nothing on it says which
country is which. `color` names them, with a legend. A categorical column
*splits* a mark, one region per category, exactly as
`color` splits `line` into one stroke per category:

```{r}
data(gapminder_asia) + area + x(year) + y(life) + color(country) +
  style(opacity = 0.45) +
  title("One region per country")
```

Regions are filled, so unlike strokes they hide one another. Write that plot
without an opacity and gog warns you, rather than letting you read a plot with
a region hidden underneath:

```{r}
#| message: true
data(gapminder_asia) + area + x(year) + y(life) + color(country) +
  title("Opaque regions: the last one drawn covers the rest")
```

Compare the legend with the plot: every country is named in the legend, and the
lowest regions are painted over by the ones drawn after them. The note above the
plot is advice rather than a refusal. The grammar has no objection to this
sentence, so gog guides your taste and still draws it. Stacking the regions
instead of overlaying them is the other answer, and it is written `area *
stack`, an area's collision modifier. Its sibling `dodge` subdivides a *width*,
and an area has no width. An area is offset by *accumulating*, and that is
`stack`. Each region sits on the cumulative height of the ones below, and they
meet edge to edge in one band instead of hiding each other:

```{r}
data(gapminder_asia) + area * stack + x(year) + y(population) + color(country) +
  y_label("Population") +
  title("Five Asian countries, population stacked")
```

*"Given gapminder Asia: areas derived by stack, x is year, y is population,
color by country."*

The thickness of each color is that country's population, and the band's top
edge is the five-country total. The measure moved from `life` to `population` on
purpose: you stack quantities that *add* (populations sum to a real total),
never averages (a stack of life expectancies means nothing). The
[Transforms](../transforms.qmd) chapter covers `stack`, and stacked bars, in
full.

## Giving an area an edge

A pale fill has a soft top edge, and the top edge is where the data is. A
stroke along it makes the boundary easy to read. An area takes no border of
its own. The edge on top of a filled region is a
`line`, and layering one over the other is how you ask for it:

```{r}
data(actuals) + x(year) + y(sales) +
  area + style(color = "steelblue", opacity = 0.25) +
  line + style(color = "steelblue", size = 2.5) +
  title("area + line: the fill and its edge, composed")
```

*"Given the actuals: x is year, y is sales, an area and also a line."*

There is deliberately no border setting on `area` to use instead. Two marks
that already exist, combined with `+`, are better than a new setting on one
of them. The composed version is also more capable: the edge takes its own
color, width and opacity as a `line`.

## One region, one fill

An area is a single filled region, so `opacity` is a *setting* here and not a
channel. A row is a vertex of the boundary, not a region of its own, so a
per-row opacity has nothing to vary along. The same reasoning makes `opacity`
a setting on `line`, one dimension down:

```{r}
#| error: true
data(gapminder_asia) + area + x(year) + y(life) + opacity(gdp)
```

`size` is refused entirely: on `area` it cannot even be set. A stroke has a
width to choose, so `line` lets you set it. An area's extent comes from its
x, its y and its baseline, so there is no size to choose:

```{r}
#| error: true
data(actuals) + area + x(year) + y(sales) + style(size = 4)
```

Two marks, two different answers, both derived from what the mark *is* rather
than listed in a table of options. `style(color = )` and `style(opacity = )`
work on both.

## With transforms

A distribution can be drawn as a filled shape. The question is the one a
histogram answers: how many rows fall in each interval? `bin`, `density` and the
aggregations mean the same thing on an area as on a bar. Binned, it is a filled
histogram:

```{r}
data(gapminder_2007) + area * bin + x(life) +
  x_label("Life expectancy (years)") +
  title("area * bin: a filled histogram")
```

*"Given gapminder 2007: an area derived by bin, x is life."*

Smoothed into a density, it is the distribution as a shape:

```{r}
data(gapminder_2007) + area * density + x(life) +
  x_label("Life expectancy (years)") +
  title("area * density")
```

*"Given gapminder 2007: an area derived by density, x is life."*

Give that same sentence a category on `x` and a number on `y`, and the estimate
changes axis. It is drawn *across each category's slot* instead of along the
whole axis, one distribution per group. It is the half violin. [`ribbon *
density`](ribbon.qmd#the-violin-a-distribution-per-category) draws the mirrored
form, and explains what the width of a violin can mean:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("area * density + x(continent): the half violin")
```

The two sentences differ by one binding, and that difference is exactly what
the extra binding changes. With only `x(life)`, the `y` axis is unused, so
the estimate is measured along it. With a category there is no unused axis,
so it is measured across the slot instead.

And an aggregation summarizes before filling, exactly as it does for `bar`:

```{r}
data(gapminder_asia) + area * mean + x(year) + y(life) +
  y_label("Mean life expectancy") +
  title("area * mean: five countries averaged per year")
```

*"Given gapminder Asia: an area derived by mean, x is year, y is life."*

## A category on the domain

One summary per continent is usually a bar chart. Draw it as a filled profile
when the five values should read as one shape, which is what a radar chart asks
for. The domain takes a category as readily as a number. Each category gets a
slot, the boundary runs across them in axis order, and the region fills to the
same baseline it always fills to:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("area * mean + x(continent): the filled profile")
```

Whether that is the right picture is your decision, not the grammar's. The
straight segments pass through space where there is no data, and that is a real
objection: no continent sits between Africa and the Americas. If you read only
the vertices, each segment only joins one value to the next. If you read the
segments themselves, they claim a change that was never measured. When the
comparison is what matters, the bar chart of the same sentence is the safer
reading, and it is one atom away.

What the fill gives you is a *shape*, and a shape can also be read in a circle.
Add [`polar()`](../polar.qmd) and the same sentence draws the filled radar:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) + polar() +
    title("area * mean + polar(): the filled radar")
```

*"Given gapminder 2007: an area derived by mean, x is continent, y is life, in
polar."*

## What you can set

A setting changes how the region looks without reading a column, and each mark
takes its own. These are an area's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("area")
```

That is the shortest settings list in the book, shared with `ribbon`. `line` has
one setting more and `bar` has two, and each gap has a reason. There is no
`size`, because the region's extent comes from the data and the baseline. And
there is no border, because the edge you would want is a *data curve*, which is
already a `line`:

```{r}
data(actuals) + x(year) + y(sales) +
  area + style(color = "seagreen", pattern = "hatch", opacity = 0.5) +
  line + style(color = "seagreen", size = 2) +
  y_label("Sales") + title("A textured fill, and a line for the edge")
```

*"Given the actuals: x is year, y is sales, an area colored seagreen, with
pattern hatch and opacity 0.5, and also a line colored seagreen, with size 2."*

That layered line can be dashed, colored differently, or smoothed, and it traces
only the data boundary. A border would also stroke the two sides and the
baseline.

## What it refuses

The first refusal comes from the measure axis. A region fills to a baseline at
zero, and a category has none. So a categorical `y` is refused, and the message
asks for the category on `x` instead:

```{r}
#| error: true
data(gapminder_2007) + area * mean + x(life) + y(continent)
```

The second refusal comes from the domain. `bin` cuts a continuous axis into
intervals, and a category is one slot with no width to cut. `density` also
needs a continuous axis, so it refuses only when the sentence leaves it no
numeric axis, which is why the half violin above draws. Each message names the
transform for categories: `count` after `bin`, `proportion` after `density`.

```{r}
#| error: true
data(gapminder_2007) + area * bin + x(continent)
```
