# Area

```{r}
#| include: false
source("../R/setup.R")
```

How much was there across a span, for each group? `area` fills the region
between the data and a baseline at zero. It is `line` plus the ground beneath
it: the same boundary, drawn as a quantity rather than as a path.

## Basic usage

```{r}
data(gapminder_asia) + area + x(year) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("Life expectancy, five Asian countries")
```

*"Given gapminder Asia: areas, x is year, y is life, grouped by country."*

An area's two axes are different kinds of thing, and it is worth naming the
difference early because most of the mark follows from it. `x` is the **domain**,
the axis the boundary is read along; `y` is the **measure**, the quantity the
region fills to. That is why an area has no horizontal form the way a `bar` does:
a bar reads which of its axes measures off the bindings, and an area does not
need to, because the answer is always `y`.

Notice the fill reaches the left and right edges of the panel. A filled region
has no glyph to clip against the frame, so its position axis sits flush to the
data rather than leaving the breathing margin a scatter gets; the reason is in
[Scales](../scales.qmd#the-axis-fits-your-data).

## The baseline has to be zero

A `line` says *where*; an `area` says *how much*. The fill is the quantity, so
it has to start from zero: an area drawn from an arbitrary floor would put a
number on the page that nothing measured.

That makes the mark a real choice rather than decoration. Life expectancy has
no meaningful zero, and the plot above spends most of its ink on the fifty
years below every observation. Sales do have a meaningful zero, and the same
mark reads correctly:

```{r}
data(actuals) + area + x(year) + y(sales) +
  y_label("Sales") + title("Sales: zero means zero")
```

When the quantity you are drawing has no zero, `line` is the mark that says
so:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  title("The same data, as paths rather than quantities")
```

## Multiple series

A categorical column *splits* a mark, one region per category, exactly as
`color` splits `line` into one polyline each:

```{r}
data(gapminder_asia) + area + x(year) + y(life) + color(country) +
  style(opacity = 0.45) +
  title("One region per country")
```

Regions are filled, so unlike polylines they hide one another. Write that plot
without an opacity and gog says so rather than letting you read a chart with a
series missing underneath:

```{r}
#| message: true
data(gapminder_asia) + area + x(year) + y(life) + color(country) +
  title("Opaque regions: the last one drawn wins")
```

Compare the legend with the picture: every country is named in the key, and the
ones with the lowest values are painted over by the ones drawn after them. The note
above the plot is advice rather than a refusal: the plot still renders,
because the grammar has no objection to it, and gog guides taste without
blocking it. Stacking the regions instead of overlaying them is the real answer,
and it ships as `area * stack`, an area's collision modifier. Its sibling `dodge`
subdivides a *width*, which an area has none of; an area is offset by
*accumulating*, so `stack` is its tool. Each region sits on the cumulative height
of the ones below, and they abut into one band instead of hiding each other:

```{r}
data(gapminder_asia) + area * stack + x(year) + y(population) + color(country) +
  y_label("Population") +
  title("Five Asian countries, population stacked")
```

Read one color's *thickness* for that country; the band's top edge is the
five-country total. The measure moved from `life` to `population` on purpose:
you stack quantities that *add* (populations sum to a real total), not indices
that do not (a stack of life expectancies would be meaningless). The
[Transforms](../transforms.qmd) chapter covers `stack`, and stacked bars, in
full.

## Giving an area an edge

An area takes no outline of its own. The edge on top of a filled region is a
`line`, and layering one over the other is how you ask for it:

```{r}
data(actuals) + x(year) + y(sales) +
  area + style(color = "steelblue", opacity = 0.25) +
  line + style(color = "steelblue", size = 2.5) +
  title("area + line: the fill and its edge, composed")
```

There is deliberately no outline option on `area` to reach for instead. Two
marks that already exist, combined with `+`, beat a new knob on one of them,
and the composed version is more capable anyway, since the edge can take its
own color, width and opacity as a `line`.

## One region, one fill

An area is a single filled region, so `opacity` is a *setting* here and not a
channel: there is nothing for a per-row opacity to vary along, since a row is
a vertex of the boundary rather than a region of its own. That is the same
reasoning that makes `opacity` a setting on `line`, one dimension up:

```{r}
#| error: true
render_svg(data(gapminder_asia) + area + x(year) + y(life) + opacity(gdp))
```

`size` goes further: on `area` it cannot even be set. A stroke's width is free,
so `line` lets you set it, but an area's extent is pinned by its x, its y and
its baseline, and there is no size left over to choose:

```{r}
#| error: true
render_svg(data(actuals) + area + x(year) + y(sales) + style(size = 4))
```

Two marks, two different answers, both derived from what the mark *is* rather
than listed in a table of options. `style(color = )` and `style(opacity = )`
work on both.

## With transforms

Transforms behave identically on every mark, so an area takes the same ones a
bar or a line does. Binned, it is a filled histogram:

```{r}
data(gapminder_2007) + area * bin + x(life) +
  x_label("Life expectancy (years)") +
  title("area * bin: a filled histogram")
```

Smoothed into a density, it is the distribution as a shape:

```{r}
data(gapminder_2007) + area * density + x(life) +
  x_label("Life expectancy (years)") +
  title("area * density")
```

Give that same sentence a category to stand in and a number to spread along,
and the estimate is drawn *across each category's slot* instead of along the
whole axis: one distribution per group, closed on the slot's center line. It
is the half violin, and the mirrored form is [`ribbon *
density`](ribbon.qmd#the-violin-a-distribution-per-category), where the choice
between the two readings of its width is also explained:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("area * density + x(continent): the half violin")
```

The two sentences differ by one binding, and that difference is exactly what
the extra binding buys. With only `x(life)` there is a free axis for the
estimate to be measured along. With a category there is not, so it is measured
across the slot instead.

And an aggregation summarizes before filling, exactly as it does for `bar`:

```{r}
data(gapminder_asia) + area * mean + x(year) + y(life) +
  y_label("Mean life expectancy") +
  title("area * mean: five countries averaged per year")
```

## A category on the domain

The domain takes a category as readily as a number. Each category gets a slot,
the boundary runs across them in axis order, and the region fills to the same
baseline it always fills to:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("area * mean + x(continent): the filled profile")
```

Whether that is the right picture is your call, not the grammar's. The straight
segments between two categories pass through space where there is no data, which
is a real objection: nothing lives between Africa and the Americas. Read the
vertices and the segments are a guide for the eye; read the segments and they
say something the data does not. When the comparison is what matters, the bar
chart of the same sentence is the safer reading, and it is one atom away.

What the fill does earn you is a *shape*, and shapes survive being bent. This is
the same expression in [polar coordinates](../polar.qmd), where it is the plot
everyone calls a radar:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) + polar() +
  title("the same sentence, read in a circle")
```

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("area")
```

An area takes the shortest list in the book, and both absences are arguments.
There is no `size`, because the region's extent comes from the data and the
baseline. And there is no border, because the edge you would want is a *data
curve*, which is already a `line`:

```{r}
data(actuals) + x(year) + y(sales) +
  area + style(color = "seagreen", pattern = "hatch", opacity = 0.5) +
  line + style(color = "seagreen", size = 2) +
  y_label("Sales") + title("A textured fill, and a line for the edge")
```

Layering the line is strictly more capable than a border setting would be: it can
be dashed, colored differently, or smoothed, and it traces only the data
boundary rather than the closure down to the baseline that a rim would also
stroke.

## What it refuses

The measure is where the mark draws its line. A region fills to a baseline at
zero, and a category has no zero to fill to, so a categorical `y` is refused
with the direction back to the mark that lays categories along the measured
axis:

```{r}
#| error: true
render_svg(data(gapminder_2007) + area * mean + x(life) + y(continent))
```

The transforms that describe how values spread along an axis refuse a category
too, on either axis, and for the same reason: they need a number line. `area *
bin` and `area * density` want a continuous domain, and each refusal names the
atom that asks their question of categories.

```{r}
#| error: true
render_svg(data(gapminder_2007) + area * bin + x(continent))
```
