# Bar

```{r}
#| include: false
source("../R/setup.R")
```

How large is each category? `bar` draws a rectangle from the baseline (zero) to
the value for each row. Bars stand up by default and lie down when the
categories are on `y`; see [Horizontal bars](#horizontal-bars).

The bar chart exists because William Playfair ran out of data. His *Commercial
and Political Atlas* of 1786 drew England's imports and exports as two curves
across thirty years, and for Scotland he had a single year, which no curve can
show [@playfair1786]. So he drew a bar for each of Scotland's trading partners
instead, and wrote that the result was the weaker picture for having no time in
it. The form he apologized for outlasted the ones he was proud of.

Two names for it are still in circulation, and they disagree. Much of the
business world, Excel included, calls this a **column chart** when the bars stand
up and saves **bar chart** for when they lie down. GOG has one `bar`, and which
way it runs is read off the sentence rather than chosen: the categorical axis
decides, which is what [Horizontal bars](#horizontal-bars) is about. Two names
for one mark would be a silent letter, and refusing those is what the No
Exceptions law is for. A **histogram** is a third name, and it is a genuinely
different plot drawn with the same mark. Karl Pearson coined the word in the
1890s [@pearson1895] for the picture of a *distribution*, where the bars are cut from a
continuous axis and touch; that is `bar * bin`, in [Histogram](#histogram) below.

## Basic usage

```{r}
data(medals) + bar + x(country) + y(gold)
```

*"Given the medals: bars, x is country, y is gold."*

The x-axis ticks are placed at each bar position automatically.
The y-axis always includes zero so bars have a truthful baseline.

## Color

`color` fills each bar with a distinct color:

```{r}
data(medals) + bar + x(country) + y(gold) +
  color(country)
```

## Ordering bars

`order()` controls the sequence of categories along whichever axis carries
them: `x` on a vertical chart, `y` on a horizontal one. The key can be any
column in the data: the category column itself, the measured column, or a third
variable entirely. Naming the category column asks for that column's own order,
which is alphabetical here because `country` is plain text, and the declared
levels if it were a [factor](../data.qmd).

**Default, data order** (rows as they appear in the dataset):

```{r}
data(medals) + bar + x(country) + y(gold) +
  title("Gold Medals: Data Order")
```

**By value, descending**, tallest bar first, the most common pattern for
ranked comparisons:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(gold, desc = TRUE) +
  title("Gold Medals: Ranked (desc)")
```

**By value, ascending**, smallest first, useful when scanning for the lowest values:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(gold) +
  title("Gold Medals: Ascending")
```

**By category name, A → Z**, natural for names and labels:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(country) +
  title("Gold Medals: Alphabetical")
```

**By category name, Z → A**, the reverse:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(country, desc = TRUE) +
  title("Gold Medals: Reverse Alphabetical")
```

**By a third column**, order countries by their silver medal count, while plotting gold:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(silver, desc = TRUE) +
  title("Gold Medals: Ordered by Silver")
```

| Expression | Order |
|-----------|-------|
| *(none)* | Data order (rows as they appear) |
| `order(gold)` | Ascending by gold (smallest first) |
| `order(gold, desc = TRUE)` | Descending by gold (largest first) |
| `order(country)` | A → Z by country name |
| `order(country, desc = TRUE)` | Z → A by country name |
| `order(silver, desc = TRUE)` | By any other column in the dataset |

## Duplicate categories

If the same category appears in multiple rows, `bar` draws one bar per row;
the last silently overlaps the earlier ones.

Use the `sum` transform to total the values per category first:

```{r}
medals_dup <- data.frame(
  country = c("USA", "USA", "GBR", "GBR", "JPN"),
  gold    = c(10, 5, 8, 4, 6)
)
data(medals_dup) + bar * sum + x(country) + y(gold) +
  order(gold, desc = TRUE) + title("Totals: bar * sum")
```

`bar * sum` and `order()` compose freely, because they do different jobs. `sum`
totals the values; `order()` sorts the categories those totals land on.

`bar` handles negative values correctly: the bar grows away from zero in
whichever direction the value sits.

```{r}
balance <- data.frame(quarter = c("Q1", "Q2", "Q3", "Q4"),
                      profit  = c(12.0, -5.0, 8.0, -2.0))
data(balance) + bar + x(quarter) + y(profit) +
  title("Negative values grow downward from zero")
```

To color the losses and the gains differently, notice what is being asked for:
color carries a *variable*, so the variable has to exist. "Is this negative" is a
fact about the data rather than about the rectangle, so compute it into a column
and map that column, and `palette()` says which colors the mapping hands out:

```{r}
balance$result <- factor(ifelse(balance$profit < 0, "loss", "profit"),
                         levels = c("loss", "profit"))
data(balance) + bar + x(quarter) + y(profit) + color(result) +
  palette(c("firebrick", "seagreen")) +
  title("Red below zero, green above")
```

`palette()` hands its colors out in **category order**, and that is why `result`
is a factor here instead of a plain character column. Left to itself a character
column takes the order its values first appear in, and this data opens on a
profit, so the palette would arrive the wrong way round and paint the gains red.
Declaring `levels = c("loss", "profit")` pins the order, the legend reads in it
too, and the plot cannot silently invert if next quarter's numbers start with a
loss. This is the same mechanism as
[Factors decide the order of categories](../data.qmd).

Red against green is the convention for money and also the hardest pair to
separate for the commonest color blindness. Texture answers that without giving
up the colors, by giving each result its own hatch as well as its own hue: see
[Texture](#texture) above, and
[Will it survive a photocopier?](../cookbook/three-or-more.qmd) in the cookbook.

## Horizontal bars

Swap the bindings. There is no `flip` atom, because that would be a second way
to say one thing. Put the categories on `y` and the measure on `x`:

```{r}
data(medals) + bar + x(gold) + y(country) +
  title("Categories on y: the bars lie down")
```

GOG reads the orientation off the bindings: the axis carrying a number is the
one the bars measure along; the other is the one they sit on. Nothing else
changes: `order`, transforms, color and `style()` all behave identically.

This is what you want whenever the category names are long, because a vertical
axis gives each label a whole line instead of a cramped slot:

```{r}
long <- data.frame(
  department = c("Research & Development", "Sales and Marketing",
                 "Customer Operations", "Finance", "People & Culture"),
  headcount  = c(128.0, 96.0, 74.0, 31.0, 22.0)
)
data(long) + bar + x(headcount) + y(department) +
  order(headcount, desc = TRUE) +
  x_label("Headcount") +
  title("Long labels fit on a vertical axis")
```

`order` means the same thing in both orientations: first in sort order reads
first. Descending puts the largest bar leftmost on a vertical chart and
topmost on a horizontal one.

Orientation is only ambiguous if neither axis is categorical. There GOG keeps
the long-standing default and draws vertically; `x(year) + y(sales)` is a
column chart, as it always was. To lay that out horizontally, name the measure
on `x`.

A bar needs something to measure, so two categorical axes are refused:

```{r}
#| error: true
data(medals) + bar + x(country) + y(country)
```

## Texture

A bar's fill can carry a **texture** as well as a color: a hatch instead of a
flat wash, set with [`style(pattern = )`](../style.qmd). A texture reads without
hue, so a hatched bar chart stays legible in grayscale or a black-and-white print:

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(pattern = "hatch") +
  title("A hatched bar reads without color")
```

The four fill textures are `"hatch"`, `"crosshatch"`, `"grid"`, and `"dots"`; the
[style chapter](../style.qmd) lays them out beside the strokes' dash, `style(pattern
= )` realized once per geometry. One texture covers the whole layer; to give each
series in a color split its *own* hatch, map a column with the
[`pattern()` channel](../channels.qmd#pattern), `shape()`'s texture twin.

## Histogram

Combine `bar` with the `bin` transform to create a histogram.
`bin` groups continuous `x` values into equal-width buckets and counts rows:

```{r}
data(gapminder_2007) + bar * bin + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Histogram of Life Expectancy, 2007")
```

Thickness is automatic, and it is where a histogram and a bar chart part ways.
A **categorical** bar fills 80 % of the gap to its neighbor, leaving the space
that reads as "separate categories". A histogram cuts a continuous axis into
adjacent intervals, so nothing sits between one bin and the next and the bars
**touch**, with a hairline in the panel color to keep them readable. Nothing was
set either time: the transform decides it, measured across whichever axis the
bars sit on.

A histogram turns on its side by the same rule. Bind the measured variable to
`y` and the count is drawn along `x`:

```{r}
data(gapminder_2007) + bar * bin + y(life) +
  y_label("Life expectancy (years)") +
  title("The same histogram, lying down")
```

### Overlaid histograms: one distribution per group {#overlaid-histograms}

Add `color` and the histogram **splits**: one histogram per category, cut on the
same bins and drawn in the same panel. This answers the question a single
combined histogram cannot: not "how is petal length distributed?" but "how is it
distributed *for each species?*"

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  x_label("Petal length (cm)") + title("Petal length by species")
```

The three histograms **overlay**: each drawn in place, measured from the shared
baseline. Opaque, the last species drawn would bury the others; so each bar is a
translucent fill under a solid outline in its own color. The outline is what
carries the shape where the fills pile up: `setosa`, well separated on the left,
reads as cleanly as `versicolor` and `virginica`, which overlap in the middle and
blend where they meet. The bins are shared across the groups, so the bars line up
and the overlap is truthful rather than an artifact of three different binnings.

The fill is faint by default so stacked series show through. `style(opacity = )`
sets how far; the outline stays solid regardless:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(opacity = 0.6) + x_label("Petal length (cm)") +
  title("A heavier fill: style(opacity = 0.6)")
```

The outline is the series color by default, but you can recolor it:
`style(border_color = )` sets the rim, `style(border_size = )` its width. A white
border cleanly parts the overlapping fills, the way many overlaid histograms are
drawn:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(border_color = "white", border_size = 1.5) +
  x_label("Petal length (cm)") + title("White borders between the fills")
```

`color` is the fill and `border_color` the outline: two independent settings, so
a bar can be one color inside and another at its edge. See
[Setting vs mapping](../style.qmd).

Or go the other way and drop the outline entirely: `style(border_size = 0)` draws
no border at all, so the translucent fills overlap with nothing between them, the
plainest overlaid histogram, blending where the groups meet:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(border_size = 0) +
  x_label("Petal length (cm)") + title("No border: just the fills")
```

The split reads the orientation off the bindings, like every other bar: put
`petal_length` on `y` and the overlaid histograms lie down. And when the groups
overlap so heavily that even the outlines crowd, give each its own panel instead:
`… | `[`facet(species)`](../faceting.qmd) draws the same three histograms side
by side on a shared scale.

See [Transforms](../transforms.qmd) for details on `bin` and other transforms.

## Grouped bars: side by side

A color split stacks the groups at one position; to set them **side by side**,
the grouped bar chart, add [`dodge`](../transforms.qmd). Each group's bar narrows
to share the slot, and the split reads as a comparison within each category:

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  y_label("Mean life expectancy") +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

`dodge` is a collision modifier: it moves the bars rather than summarizing them,
so it composes with the statistic (`bar * mean * dodge` here, `bar * count * dodge`
for tallies). See [Transforms](../transforms.qmd) for the whole family.

## Stacked bars: piled up

The other way to place a color split: instead of side by side, pile the groups
**on top of each other** with [`stack`](../transforms.qmd). Each bar's height
becomes the group total, and the segments read as parts of a whole:

```{r}
data(gm_eras) + bar * sum * stack + x(continent) + y(population) + color(era) +
  y_label("Population") +
  title("Population by continent, 1957 and 2007 stacked")
```

`dodge` compares the groups (equal baselines, side by side); `stack` sums them
(one bar, segments piled). Reach for `stack` when the parts add to a meaningful
total (populations, counts, revenue) and for `dodge` when you are comparing the
groups against each other. Note the measure: you stack `sum population` (quantities
add), where the dodge example compared `mean life` (indices do not).

### The 100% stacked bar

The plot above answers two questions at once, and one of them drowns out the
other: Asia's two eras come to 5.4 billion against Oceania's 0.04, so the axis is
scaled for Asia and the *split* inside the small continents is unreadable. When the
composition is what you came for and the total is not, `stack(share = TRUE)`
divides each pile by its own bar's total, so every bar reaches 1 and the segments
are the only thing left to read:

```{r}
data(gm_eras) + bar * sum * stack(share = TRUE) + x(continent) + y(population) +
  color(era) + title("Each continent's 1957/2007 split, filled to one")
```

Now the comparison is between continents rather than inside one, and the two ends
of it are worth naming. **Africa** is the most lopsided: 78% of its two-era total
sits in 2007, so it more than tripled. **Europe** is the flattest at 57%, the only
continent whose halves come close to even. Neither is legible in the plot above at
any size. What you give up is the totals, which have gone entirely. That is the
trade, not a side effect, and Oceania now occupies as much width as Asia.

The axis says `Share`, because whatever the numbers were before they are fractions
of one now, and that holds even though `y(population)` names a column: the fill
rescales an axis somebody already named. It is not
[`proportion`](../transforms.qmd#proportion-relative-frequency) under another
name. The two divide by different totals. `proportion` divides by the whole plot's,
so its bars still say how big each slot is. This one divides by the bar's own.
And only this one works on a `sum` of a column, which `proportion` has no way
to say.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("bar")
```

A bar has no `size`, and the absence is the point: its extent is already pinned
by its position and its value, so a width setting would be a second, conflicting
answer to how big it is. What it does have, being a closed fill, is a rim.

```{r}
data(medals) + bar * sum + x(country) + y(gold) +
  style(color = "goldenrod", pattern = "crosshatch",
        border_color = "black", border_size = 1) +
  y_label("Gold medals") + title("A fill, a texture, and a rim")
```

The `pattern` values here are the five **fill textures**, not the three dashes a
`line` takes. One setting name, one realization per geometry, so a plot that has
to survive grayscale printing can separate its bars by texture instead of hue.

The bar's own knob is not a setting at all: `bin(bins = 20)` and
`bin(width = 5)` control the histogram `bar * bin` draws, and they belong to the
transform because binning is what has a parameter, not the rectangle.

## What it refuses

A bar measures a length from a baseline, and its refusals mark the edges of that.

Binning cuts a continuous axis into intervals, so there is nothing for it to cut
on a categorical one:

```{r}
#| error: true
data(gapminder_2007) + bar * bin + x(continent)
```

A category is already one slot per value, which is what `count` is for. The
refusal says so rather than drawing five bins of nothing.

Texture is the second edge. A bar is a fill, so it takes a fill's textures, and a
dash is a stroke's:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(pattern = "dashed")
```

The dash is available on the bar's *outline* through `border_*`, which is a
stroke and does take one. And a bar given two categorical positions has nothing
left to measure, which is the refusal shown above under horizontal bars.
