# Bar

```{r}
#| include: false
source("../R/setup.R")
```

How large is each category? `bar` draws a rectangle from the baseline (zero) to
the value for each row. Bars are vertical when the categories are on `x` and
horizontal when they are on `y`; see [Horizontal bars](#horizontal-bars).

The bar chart exists because William Playfair had too little data. His
*Commercial and Political Atlas* of 1786 drew England's imports and exports as
two curves across thirty years [@playfair1786]. For Scotland he had a single
year, and no curve can show one year. So he drew a bar for each of Scotland's
trading partners instead, and wrote that the result was the weaker picture,
because it showed no change over time. The form he apologized for became the
one everyone knows.

Two names for it are still in use, and they name opposite orientations. Much of
the business world, Excel included, calls this a **column chart** when the bars
are vertical, and uses **bar chart** only when they are horizontal. gog has one
`bar`, and its direction comes from whichever axis carries the category. A
third name, **histogram**, belongs to a different plot drawn with the same
mark. Karl Pearson introduced the word in the 1890s [@pearson1895] for the
picture of a *distribution*. Its bars cover intervals cut from a continuous
axis, and neighboring bars touch. In gog that plot is `bar * bin`, shown in
[Histogram](#histogram) below.

## Basic usage

Which country won the most gold medals? Everything in this chapter is added to
one sentence, the smallest that draws a bar: a source, the mark, and its two
positions.

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold)
```

*"Given the medals: bars, x is country, y is gold."*

The x axis places a tick at each bar position automatically. The y axis reaches
zero unless you state your own `limits`, so a bar's length is its value.

## Color

Every bar is the same color until you map a column to `color`. With a color for
each bar, a reader can find one country quickly and match it to the same
country in another plot. Bind `color` to the column that already sits on `x`.
`color` fills each bar with a distinct color:

```{r}
data(medals) + bar + x(country) + y(gold) +
  color(country)
```

*"Given the medals: bars, x is country, y is gold, color by country."*

The legend repeats the names on the x axis, because the same column sits on
both.

## Ordering bars

In what order should the bars appear? A reader looking for the largest value
wants it first. A reader looking for one name wants the names in alphabetical
order. `order()` controls the sequence of categories along whichever axis
carries them: `x` on a vertical plot, `y` on a horizontal one. You can sort on
the categorical column itself, on the measured column, or on a third column of
numbers. Naming the categorical column asks for that column's own order. Here
that is alphabetical, because `order()` sorts plain text such as `country` from
A to Z. A column with a
[declared order](../data.qmd#declaring-the-order-of-categories) uses that order
instead.

**Default, data order** (rows as they appear in the table):

```{r}
data(medals) + bar + x(country) + y(gold) +
  title("Gold medals: data order")
```

**By value, descending**, tallest bar first, the most common pattern for
ranked comparisons:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(gold, desc = TRUE) +
  title("Gold medals: largest first")
```

*"Given the medals: bars, x is country, y is gold, ordered by gold, largest
first."*

These bars are in the same order as the plot titled *Gold medals: data order*.
The medals table already lists the largest count first, so ordering by gold
changes nothing here. Writing the order anyway is still safer, because the plot
then stays sorted even if next year's rows arrive in a different order. The
plots below each put the bars in a different order.

**By value, ascending**, smallest first, useful when scanning for the lowest
values:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(gold) +
  title("Gold medals: smallest first")
```

**By category name, A → Z**, natural for names and labels:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(country) +
  title("Gold medals: A to Z")
```

**By category name, Z → A**, the reverse:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(country, desc = TRUE) +
  title("Gold medals: Z to A")
```

**By a third column**, order countries by their silver medal count, while
plotting gold:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(silver, desc = TRUE) +
  title("Gold medals: ordered by silver")
```

| Expression | Order |
|-----------|-------|
| *(none)* | Data order (rows as they appear) |
| `order(gold)` | Ascending by gold (smallest first) |
| `order(gold, desc = TRUE)` | Descending by gold (largest first) |
| `order(country)` | A → Z by country name |
| `order(country, desc = TRUE)` | Z → A by country name |
| `order(silver, desc = TRUE)` | By another column of numbers |

## Duplicate categories

A table often holds several rows for one category, one per event or one per
year. The question is still how large each category is in total. Without a
transform, `bar` draws one bar per row, and rows that share a category share a
slot. This table has two rows each for USA and GBR:

```{r}
medals_dup <- data.frame(
  country = c("USA", "USA", "GBR", "GBR", "JPN"),
  gold    = c(10, 5, 8, 4, 6)
)
data(medals_dup) + bar + x(country) + y(gold) +
  title("Duplicate rows: no transform")
```

Every bar grows from zero, so the shorter ones sit inside the tallest. USA has
rows of 10 and 5, and its bar reads 10 rather than 15. Each bar is drawn
slightly translucent, so the color is darker where two bars overlap. For USA,
that darker band ends at 5, where the shorter bar stops. That band is the only
sign that a category has extra rows.

Use the `sum` transform to total the values per category first:

```{r}
data(medals_dup) + bar * sum + x(country) + y(gold) +
  order(gold, desc = TRUE) + title("Totals: bar * sum")
```

*"Given the medals dup table: bars derived by sum, x is country, y is gold,
ordered by gold, largest first."*

`sum` gives one number for each category. It reads every row that shares a
category and totals them. USA's 10 and 5 become 15, and GBR's 8 and 4 become
12. Five rows become three bars, one for each country, and the darker band is
gone.

## Negative values

A bar can also point the other way, when a value is below zero, such as the
profit of a quarter that lost money. Every bar starts at zero. That rule made
the duplicate bars overlap, and it is also why a negative bar points down. So a
negative value needs no special treatment:

```{r}
balance <- data.frame(quarter = c("Q1", "Q2", "Q3", "Q4"),
                      profit  = c(12.0, -5.0, 8.0, -2.0))
data(balance) + bar + x(quarter) + y(profit) +
  title("Negative values grow downward from zero")
```

To color the losses and the profits differently, you need a column that says
which is which. `color` maps a column, so that column has to exist. "Is this
negative" is a fact about the data, not about the rectangle. So put it in a
column, map that column, and let
[`palette()`](../style.qmd#palette-choosing-the-colors-a-mapping-uses) say
which colors the mapping uses. The table below sets the column's values by
hand. In practice you compute it from `profit` with one condition. Above zero
is a profit, and anything else is a loss.

```{r}
balance <- data.frame(
  quarter = c("Q1", "Q2", "Q3", "Q4"),
  profit  = c(12.0, -5.0, 8.0, -2.0),
  result  = factor(c("profit", "loss", "profit", "loss"),
                   levels = c("loss", "profit"))
)
data(balance) + bar + x(quarter) + y(profit) + color(result) +
  palette(c("firebrick", "seagreen")) +
  title("Red below zero, green above")
```

*"Given the balance table: bars, x is quarter, y is profit, color by result,
with a palette of firebrick and seagreen."*

`palette()` assigns its colors in **category order**, and that is why `result`
carries a declared order here instead of being plain text. `palette()` does not
sort plain text, and `order()` does not move its colors: it assigns them in the
order the values first appear in the table. The first row here is a profit, so
the palette would give firebrick to the profits. Declaring the order as loss,
then profit, gives firebrick to the losses, and the legend lists them in that
order. The colors then stay with their results, even if next year's table
starts with a loss. R spells that with a factor's levels, and the other three
bindings with `ordered()`. This is the same mechanism as [Declaring the order
of categories](../data.qmd#declaring-the-order-of-categories).

Red for loss and green for profit is one common convention. It is also the
hardest pair to separate for red-green color blindness, the most common kind.
Texture solves that and keeps the colors, by giving each result its own texture
as well as its own color: see [Texture](#texture) below, and [Will it read in
gray?](../cookbook/three-or-more.qmd#will-it-read-in-gray) in the cookbook.

## Horizontal bars

Sometimes the bars should be horizontal. The category names may be too long for
the bottom axis. Or there may be many categories, and a tall list of horizontal
bars reads more easily than a crowded row of vertical ones. Swap the bindings,
and put the categories on `y` and the measure on `x`:

```{r}
data(medals) + bar + x(gold) + y(country) +
  title("Categories on y: horizontal bars")
```

If you know ggplot2's `coord_flip()` or matplotlib's `barh()`, gog has no such
word, because it would be a second way to say one thing. gog takes the
orientation from the bindings. The axis with the number is the one the bars
grow along, so their length is read on it. The axis with the
category is the one the bars are placed on, one slot each. Nothing else
changes: `order`, transforms such as `sum` and `bin`, `color` and `style()` all
behave identically.

Horizontal bars help most when the category names are long. On a vertical axis
each label is written across the page, so it does not have to fit inside one
slot:

```{r}
#| echo: false
peek(departments)
```

```{r}
data(departments) + bar + x(headcount) + y(department) +
  order(headcount, desc = TRUE) +
  x_label("Headcount") +
  title("Long labels fit on a vertical axis")
```

`order` means the same thing in both orientations: the first category in sort
order takes the first slot. Descending puts the largest bar leftmost on a
vertical plot and topmost on a horizontal one.

Orientation is only ambiguous if neither axis is categorical. In that case gog
draws vertically, so `x(year) + y(sales)` gives vertical bars. Swapping the two
does not make them horizontal. With two continuous columns, gog always
measures the one on `y`. So `x(sales) + y(year)` still draws vertical bars, and
each bar's length is a year, such as 2019. For horizontal bars, make the year a
category and put it on `y`: store the years as text, such as `"2019"`, and each
year gets a slot of its own.

A bar needs something to measure, so two categorical axes are refused.
[What it refuses](#what-it-refuses) shows that refusal.

## Texture

The profit and loss bars in [Negative values](#negative-values) are red and
green, the pair hardest to distinguish for a reader with red-green color
blindness. Give each result its own texture as well, and that reader can
separate the bars without seeing color at all.

A bar's fill can carry a **texture** instead of one solid color. On a fill, a
texture is a **hatch**: a repeated pattern of lines or dots, drawn in the bar's
color. You set it with [`style(pattern = )`](../style.qmd). A texture does not
need color, so a hatched bar chart stays legible in grayscale or in a
black-and-white print:

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(pattern = "hatch") +
  title("A hatched bar reads without color")
```

*"Given the medals: bars, x is country, y is gold, with pattern hatch."*

`style(pattern = )` gives every bar in the layer the same texture. To give each
category its own texture, map a column with the [`pattern()`
channel](../channels.qmd#pattern), which does for texture what `shape()` does
for glyphs. The plot below maps the `texture` column with `pattern()`. Each bar
then shows one of the six fill textures, in this order: `"solid"` (the default,
a plain fill), `"hatch"`, `"crosshatch"`, `"stripes"`, `"grid"` and `"dots"`:

```{r}
texture_order <- c("solid", "hatch", "crosshatch", "stripes", "grid", "dots")
textures <- data.frame(texture = ordered(texture_order, levels = texture_order),
                       height = c(6, 5, 4, 3, 2, 1))
data(textures) + bar + x(texture) + y(height) + pattern(texture) +
  x_label("") + y_label("") + title("The six fill textures")
```

*"Given the textures table: bars, x is texture, y is height, pattern by
texture."*

The [style chapter](../style.qmd) shows them beside the dashes a stroke takes,
because `style(pattern = )` is one name whose meaning depends on the shape: a
hatch on a fill, a dash on a stroke. For the profit and loss bars, write
`pattern(result)` beside `color(result)`. The losses then stay solid and the
profits are hatched, so each result has its own color and its own texture.

## Histogram

How are the world's life expectancies spread out, and where do most countries
fall? That is a question about one continuous column, and a bar needs a
category to sit in. The `bin` transform makes those categories: it cuts the
`x` axis into equal-width intervals, called bins, and counts the rows in each.
Combine `bar` with `bin` to create a histogram:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * bin + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Histogram of life expectancy, 2007")
```

*"Given gapminder 2007: bars derived by bin, x is life, y is count."*

Thickness is automatic, and it is the clearest way a histogram and a bar chart
differ on the page. A categorical bar fills 80% of its slot, which leaves a gap
between neighboring bars. That gap is what reads as separate categories. A
histogram cuts a continuous axis into adjacent intervals, so nothing sits
between one bin and the next and the bars *touch*. A thin line in the panel's
background color is drawn between neighboring bars, so each one can still be
seen. You set nothing in either plot. The gold medal chart in
[Basic usage](#basic-usage) and this histogram both take their thickness from
the axis the bars sit on.

`bin` is not the bar's own transform. Four other marks can draw the same bins,
and only the shape changes:

| Mark derived by `bin` | Shape |
|---|---|
| `point * bin` | one dot at the top of each bin |
| `line * bin` | the frequency polygon: straight lines through the bin centers |
| `area * bin` | the same polygon, filled to the baseline |
| `step * bin` | the staircase outline, drawn in [Step](step.qmd) |
| `bar * bin` | filled bars, the histogram above |

`zone * bin` and `surface * bin` cut two axes instead of one: the heatmap in
[Zone](zone.qmd) and the sheet in [Surface](surface.qmd).

The width of each bin decides what the histogram shows. With no argument, `bin`
picks the number of bins from the number of rows, by Sturges' rule
[@sturges1926]. The 142 countries here get 9 bins. Name a width to cut the axis
in the data's own units, here five years:

```{r}
data(gapminder_2007) + bar * bin(width = 5) + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Bins five years wide")
```

*"Given gapminder 2007: bars derived by bin at width 5, x is life, y is count."*

`bin(30)` sets the number of bins instead. Both belong to the transform, and
[Choosing the bin count](../transforms.qmd#choosing-the-bin-count) shows what
each does to the picture.

A histogram can also be horizontal, by the rule in
[Horizontal bars](#horizontal-bars). Bind the column you are binning to `y`,
and the count is drawn along `x`. The sentence does not have to name the
count: `bin` computes it and draws it on the other axis, so `y(count)` in the
first histogram was optional:

```{r}
data(gapminder_2007) + bar * bin + y(life) +
  y_label("Life expectancy (years)") +
  title("The same histogram, horizontal")
```

### Overlaid histograms: one distribution per group {#overlaid-histograms}

The iris flowers table holds measurements of 150 flowers from three species. A
single histogram of all the flowers answers "how is petal length distributed?"
It cannot answer "how is it distributed *for each species*?" Add `color` and
the histogram *splits*: one histogram per category, cut on the same bins and
drawn in the same panel.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  x_label("Petal length (cm)") + title("Petal length by species")
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."*

The three histograms overlay. None of them is moved aside, and all three
measure from the same baseline. If the fills were opaque, the species drawn
last would hide the others wherever they overlap. So each bar is a translucent
fill under a solid outline in its own color. The outline is what keeps the
shape readable where the fills overlap. `setosa` sits alone on the left and
reads clearly. `versicolor` and `virginica` overlap in the middle, and their
outlines still separate them. The groups share the same bins, so their bars
align, and an overlap in the plot is an overlap in the data.

The fill is faint by default so the groups underneath show through.
`style(opacity = )` sets how faint; the outline stays solid whatever you choose:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(opacity = 0.6) + x_label("Petal length (cm)") +
  title("A heavier fill: style(opacity = 0.6)")
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species, with opacity 0.6."*

The outline is the group's color by default, but you can recolor it:
`style(border_color = )` sets the border, `style(border_size = )` its width. A
black border separates the overlapping fills, the way many overlaid histograms
are drawn. It also stays visible against the light panel, where a white one
disappears:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(border_color = "black", border_size = 1.5) +
  x_label("Petal length (cm)") + title("Black borders between the fills")
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species, with border color black and border size 1.5."*

The fill and the outline are independent: a bar can be one color inside and
another at its edge, whether the fill was mapped (as here) or set. See
[Setting vs mapping](../style.qmd).

Or remove the outline entirely. `style(border_size = 0)` draws no border, so the
translucent fills overlap with nothing between them. This is the plainest
overlaid histogram:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(border_size = 0) +
  x_label("Petal length (cm)") + title("No border: just the fills")
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species, with border size 0."*

When the groups overlap so much that the outlines are hard to distinguish, give
each species its own panel instead. Adding
[`| facet(species)`](../faceting.qmd) to the sentence draws the same three
histograms side by side, on one shared scale.

## Grouped bars: side by side

How did each continent's life expectancy change between 1957 and 2007? That is a
comparison inside each category, so each continent needs two bars, one per era.
A color split alone draws both bars in one slot, the shorter inside the taller.
To set them side by side as a **grouped bar chart**, add
[`dodge`](../transforms.qmd). Each group's bar narrows to share the slot, and
the split reads as a comparison within each category. The `gm_eras` table holds
two years, 1957 and 2007, with one row per country in each. `mean` averages
the countries into one value for each continent and era. `era` is the column
that splits each continent into two bars.

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  y_label("Mean life expectancy") +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

*"Given the gapminder eras: bars derived by mean and dodge, x is continent, y is
life, color by era."*

`dodge` is a collision modifier: it moves the bars rather than summarizing them,
so it composes with the statistic (`bar * mean * dodge` here, `bar * count *
dodge` for tallies). See [Transforms](../transforms.qmd) for all four collision
modifiers.

`mean` is the right statistic here, and gog will not choose it for you. Write
`sum` in its place and gog still draws it:

```{r}
data(gm_eras) + bar * sum * dodge + x(continent) + y(life) + color(era) +
  y_label("Sum of life expectancies") +
  title("Sum in place of mean: legal, and meaningless")
```

*"Given the gapminder eras: bars derived by sum and dodge, x is continent, y is
life, color by era."*

Africa's fifty-two life expectancies for 2007 are added into one bar of 2,850
years. A sum of life expectancies has no meaning. You may write either
statistic. Only the data says which one is meaningful, and
[Law 8](../design-laws.qmd#law-8-pronounceable-usable) says why gog leaves
that choice to you.

## Stacked bars: piled up

Sometimes the question is the total as well as the parts: how much is there
altogether, and how much of it does each group hold? A color split can be
placed a second way. Instead of side by side, [`stack`](../transforms.qmd) piles
the groups on top of each other. Each bar's height becomes the category's total
across the groups, and the segments read as parts of a whole:

```{r}
data(gm_eras) + bar * sum * stack + x(continent) + y(population) + color(era) +
  y_label("Population") +
  title("Population by continent, 1957 and 2007 stacked")
```

*"Given the gapminder eras: bars derived by sum and stack, x is continent, y is
population, color by era."*

`dodge` sets the groups side by side, all on one baseline; `stack` piles them
into one bar, whose height is their total. Use `stack` when the parts add to a
meaningful total (populations, counts, revenue), and `dodge` when you are
comparing the groups. The two examples also use different statistics. You stack
the `sum` of population, because quantities add, and you dodge the `mean` of
life, because averages do not.

A stack accepts the wrong statistic as well. gog draws `mean` in place of
`sum`, and piles two averages into a bar that reads as their total:

```{r}
data(gm_eras) + bar * mean * stack + x(continent) + y(life) + color(era) +
  y_label("Mean life expectancy, stacked") +
  title("Mean in place of sum: legal, and meaningless")
```

*"Given the gapminder eras: bars derived by mean and stack, x is continent, y is
life, color by era."*

Oceania's pile reaches 150 years. A stack reads as parts of a total, so its
values must be quantities that add up. That is why this section sums
`population`, while the grouped bars show the `mean` of `life`.

### The 100% stacked bar

The population stack answers two questions at once: how large each total is,
and how it splits. The totals differ so much that the small continents' splits
are hard to read. Together, Asia's two eras hold 5.4 billion people, and
Oceania's hold 0.04 billion. The axis is scaled for Asia, so Oceania's bar is
too short to show its split at all. When the split is the question and the
total is not, ask for shares. `stack(share = TRUE)` divides each segment by its
own bar's total. Each segment then shows its group's share of the bar, a number
from 0 to 1. Every bar reaches 1, so all the splits sit on one scale and can be
compared across continents:

```{r}
data(gm_eras) + bar * sum * stack(share = TRUE) + x(continent) + y(population) +
  color(era) + title("Each continent's 1957/2007 split, filled to one")
```

*"Given the gapminder eras: bars derived by sum and stack as shares, x is
continent, y is population, color by era."*

With every bar at 1, the question changes. The population stack compared the
two eras inside one continent, and only for the large ones. This plot compares
the splits across continents: which population grew the most between the two
eras. Africa's population grew the most. Of its two-era total, 78% falls in
2007 and 22% in 1957, so it more than tripled. Europe's grew the least: 57% of
its total falls in 2007, the closest any continent comes to an even split.
Neither number can be read from the population stack above. This plot does not
show the totals at all, which is why Oceania's bar is now as tall as Asia's.
The two plots are complementary, each answering half of one question. `stack`
says how much each continent has, and `stack(share = TRUE)` says how it splits.
When you need both answers, draw both plots.

The axis is labeled `Share` rather than `Population`, because every value is
now a fraction of its bar's total. `Share` alone does not say a share of what,
so name it yourself with `y_label("Share of two-era population")`.
`stack(share = TRUE)` is not
[`proportion`](../transforms.qmd#proportion-relative-frequency) under another
name. The two divide by different totals. `proportion` divides by the whole
plot's total, so its bars still show how large each category is. This one
divides by the bar's own total.

## What you can set

A setting changes how every bar looks without reading a column, and each mark
takes its own. These are a bar's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("bar")
```

A bar has no `size`, and that is deliberate. Its slot and its value already
decide how large it is, so a thickness setting would be a second, conflicting
answer.
A bar is a shape with an inside and an outline, so it does have a border.

A fill, a texture and a border compose, so one layer can carry all three:

```{r}
data(medals) + bar * sum + x(country) + y(gold) +
  style(color = "goldenrod", pattern = "crosshatch",
        border_color = "black", border_size = 1) +
  y_label("Gold medals") + title("A fill, a texture, and a border")
```

*"Given the medals: bars derived by sum, x is country, y is gold, colored
goldenrod, with pattern crosshatch, border color black and border size 1."*

The `pattern` values here are the six fill textures. A `line` takes five values
of its own, and only `"solid"` is in both lists. The setting has one name, and
each mark draws it the way its shape allows. A fill takes a hatch, and a stroke
takes a dash.

The histogram's bins are not a setting. `bin(30)` and `bin(width = 5)` belong
to the transform, because they change what is counted, not how a bar is drawn.
[Histogram](#histogram) shows `bin(width = 5)`, and [Choosing the bin
count](../transforms.qmd#choosing-the-bin-count) shows both.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The three refusals below come from different facts. The first belongs to the
transform rather than the mark: `bin` needs a continuous axis to cut. The other
two belong to the mark: a bar is a fill rather than a stroke, and a bar needs a
number to measure.

Binning cuts a continuous axis into intervals, so there is nothing for it to cut
on a categorical one:

```{r}
#| error: true
data(gapminder_2007) + bar * bin + x(continent)
```

To count the rows in each category's slot, use `count`. The refusal names
`bar * count` rather than drawing empty bins.

Texture is the second refusal. A bar is a fill, so it takes a fill's textures,
and a dash is a stroke's:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(pattern = "dashed")
```

A bar's outline takes a color and a width, never a dash. `border_color` and
`border_size` are the only two settings the outline has.

Two categorical positions are the third refusal. The axis with the category
places the bars, one slot each, and the axis with the number measures how far
they reach, as [Horizontal bars](#horizontal-bars) explained. With a category
on both, there is no length left to draw:

```{r}
#| error: true
data(medals) + bar + x(country) + y(country)
```

The message names two fixes: put a number on one axis, or count the rows per
category with `bar * count`.

## In short

The axis with the category places the bars, and the axis with the number
measures them. A bar's length is its value, and that axis reaches
zero unless you state your own `limits`. Without a transform one bar is drawn
per row, and rows sharing a category overlap in one slot. Thickness is not a
setting. A categorical bar fills 80% of its slot. A histogram bar fills its
whole bin, and `bin()` decides how wide a bin is.

## What comes next

The gap on each side of a bar says that the categories are separate. When the
positions are continuous, such as the years of a sales series, [Area](area.qmd)
joins each value to the next and fills the region below them, with no gaps.
