# Bar

```{r}
#| include: false
source("../R/setup.R")
```

How large is each category? `bar` draws a rectangle from the baseline (zero) to
the value for each row. Bars are vertical when the categories are on `x` and
horizontal when they are on `y`; see [Horizontal bars](#horizontal-bars).

The bar chart exists because William Playfair had too little data. His
*Commercial and Political Atlas* of 1786 drew England's imports and exports as
two curves across thirty years [@playfair1786]. For Scotland he had a single
year, and no curve can show one year. So he drew a bar for each of Scotland's
trading partners instead, and wrote that the result was the weaker picture for
having no time in it. The form he apologized for became the one everyone knows.

Two names for it are still in use, and they name opposite orientations. Much of
the business world, Excel included, calls this a **column chart** when the bars
are vertical, and keeps **bar chart** for when they are horizontal. gog has one
`bar`, and its direction comes from the sentence rather than from a choice.
Whichever axis carries the category decides it, which is what [Horizontal
bars](#horizontal-bars) is about. A **histogram** is a third name, and it is a
genuinely different plot drawn with the same mark. Karl Pearson coined the word
in the 1890s [@pearson1895] for the picture of a *distribution*, where the bars are cut
from a continuous axis and touch; that is `bar * bin`, in
[Histogram](#histogram) below.

## Basic usage

Which country won the most gold medals? Everything in this chapter is added to
one sentence, the smallest that draws a bar: a source, the mark, and its two
positions.

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold)
```

*"Given the medals: bars, x is country, y is gold."*

The x axis places a tick at each bar position automatically. The y axis reaches
zero unless you state your own `limits`, so a bar's length is its value.

## Color

Every bar is the same color until a column separates them. A color of its own
lets a reader find one country quickly, and match it to the same country in
another plot. Bind `color` to the column that already sits on `x`. `color`
fills each bar with a distinct color:

```{r}
data(medals) + bar + x(country) + y(gold) +
  color(country)
```

*"Given the medals: bars, x is country, y is gold, color by country."*

## Ordering bars

In what order should the bars appear? A reader looking for the largest value
wants it first. A reader looking for one name wants the names in alphabetical
order. `order()` controls the sequence of categories along whichever axis
carries them: `x` on a vertical plot, `y` on a horizontal one. You can sort on
any column in the table: the categorical column itself, the measured column, or
a third column entirely. Naming the categorical column asks for that column's
own order. Here that is alphabetical, because `country` is plain text. A column
with a [declared order](../data.qmd) uses that order instead.

**Default, data order** (rows as they appear in the table):

```{r}
data(medals) + bar + x(country) + y(gold) +
  title("Gold medals: data order")
```

**By value, descending**, tallest bar first, the most common pattern for
ranked comparisons:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(gold, desc = TRUE) +
  title("Gold medals: largest first")
```

*"Given the medals: bars, x is country, y is gold, ordered by gold, largest
first."*

These bars are in the same order as the bars in the data order plot above. The
medals arrive with the leader at the top, so ordering by gold changes nothing
here. Writing the order anyway is still worth it. The plot then stays sorted
even if next year's rows arrive differently. The plots below each put the bars
in a different order.

**By value, ascending**, smallest first, useful when scanning for the lowest
values:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(gold) +
  title("Gold medals: smallest first")
```

**By category name, A → Z**, natural for names and labels:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(country) +
  title("Gold medals: A to Z")
```

**By category name, Z → A**, the reverse:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(country, desc = TRUE) +
  title("Gold medals: Z to A")
```

**By a third column**, order countries by their silver medal count, while
plotting gold:

```{r}
data(medals) + bar + x(country) + y(gold) +
  order(silver, desc = TRUE) +
  title("Gold medals: ordered by silver")
```

| Expression | Order |
|-----------|-------|
| *(none)* | Data order (rows as they appear) |
| `order(gold)` | Ascending by gold (smallest first) |
| `order(gold, desc = TRUE)` | Descending by gold (largest first) |
| `order(country)` | A → Z by country name |
| `order(country, desc = TRUE)` | Z → A by country name |
| `order(silver, desc = TRUE)` | By any other column in the table |

## Duplicate categories

A table often holds several rows for one category, one per event or one per
year. The question is still how large each category is in total. Without a
transform, `bar` draws one bar per row, and rows that share a category share a
slot. This table has two rows each for USA and GBR:

```{r}
medals_dup <- data.frame(
  country = c("USA", "USA", "GBR", "GBR", "JPN"),
  gold    = c(10, 5, 8, 4, 6)
)
data(medals_dup) + bar + x(country) + y(gold) +
  title("Duplicate rows: no transform")
```

Every bar grows from zero, so the shorter ones sit inside the tallest. USA has
rows of 10 and 5, and its bar reads 10 rather than 15. The darker band ends at
5, which is the row hidden inside. That band is the only sign that a category
has extra rows.

Use the `sum` transform to total the values per category first:

```{r}
data(medals_dup) + bar * sum + x(country) + y(gold) +
  order(gold, desc = TRUE) + title("Totals: bar * sum")
```

*"Given the medals dup table: bars derived by sum, x is country, y is gold,
ordered by gold, largest first."*

`sum` gives one number for each category. It reads every row that shares a
category and totals them. USA's 10 and 5 become 15, and GBR's 8 and 4 become
12. Five rows become three bars, one for each country, and the darker band is
gone.

A negative value needs no special treatment. The bar extends from zero in
whichever direction the value lies, so it reaches down instead of up.

```{r}
balance <- data.frame(quarter = c("Q1", "Q2", "Q3", "Q4"),
                      profit  = c(12.0, -5.0, 8.0, -2.0))
data(balance) + bar + x(quarter) + y(profit) +
  title("Negative values grow downward from zero")
```

To color the losses and the gains differently, start from what `color` requires:
it maps a column, so that column has to exist. "Is this negative" is a fact
about the data, not about the rectangle. So put it in a column, map that
column, and let [`palette()`](../style.qmd#palette-choosing-the-colors-a-mapping-uses)
say which colors the mapping uses. The table below types the column's values
in by hand. In practice you compute it from `profit` with one condition. Above
zero is a profit, and anything else is a loss.

```{r}
balance <- data.frame(
  quarter = c("Q1", "Q2", "Q3", "Q4"),
  profit  = c(12.0, -5.0, 8.0, -2.0),
  result  = factor(c("profit", "loss", "profit", "loss"),
                   levels = c("loss", "profit"))
)
data(balance) + bar + x(quarter) + y(profit) + color(result) +
  palette(c("firebrick", "seagreen")) +
  title("Red below zero, green above")
```

*"Given the balance table: bars, x is quarter, y is profit, color by result,
with a palette of firebrick and seagreen."*

`palette()` assigns its colors in **category order**, and that is why `result`
carries a declared order here instead of being plain text. A column of plain text
takes the order its values first appear in. The first row here is a profit, so
the palette would give firebrick to the gains. Declaring the order as loss,
then profit, sets it, and the legend reads in it too. The plot then cannot
silently invert if next quarter's numbers start with a loss. R spells that
with a factor's levels, and the other three with `ordered()`. This is the same
mechanism as [Declaring the order of categories](../data.qmd).

Red for loss and green for gain is one common convention. It is also the hardest
pair to separate for red-green color blindness, the most common kind. Texture
solves that and keeps the colors, by giving each result its own hatch as well as
its own color: see [Texture](#texture) below, and [Will it survive a
photocopier?](../cookbook/three-or-more.qmd) in the cookbook.

## Horizontal bars

Sometimes the bars should be horizontal. The category names may be too long for
the
bottom axis. Or there may be many categories, and a list of rows reads more
easily than a row of columns. Swap the bindings. If you know ggplot2's
`coord_flip()` or matplotlib's `barh()`, there is no such word here, because it
would be a second way to say one thing. Put the categories on `y` and the
measure on `x`:

```{r}
data(medals) + bar + x(gold) + y(country) +
  title("Categories on y: horizontal bars")
```

gog takes the orientation from the bindings. The axis with the number is the
one the bars grow along, so their length is read on it. The axis with the
category is the one the bars are placed on, one slot each. Nothing else
changes: `order`, transforms such as `sum` and `bin`, `color` and `style()` all
behave identically.

This is what you want whenever the category names are long, because a vertical
axis gives each label a whole line instead of a cramped slot:

```{r}
long <- data.frame(
  department = c("Research & Development", "Sales and Marketing",
                 "Customer Operations", "Finance", "People & Culture"),
  headcount  = c(128.0, 96.0, 74.0, 31.0, 22.0)
)
data(long) + bar + x(headcount) + y(department) +
  order(headcount, desc = TRUE) +
  x_label("Headcount") +
  title("Long labels fit on a vertical axis")
```

`order` means the same thing in both orientations: the first category in sort
order is drawn first. Descending puts the largest bar leftmost on a vertical
plot and topmost on a horizontal one.

Orientation is only ambiguous if neither axis is categorical. In that case gog
draws vertically, so `x(year) + y(sales)` is a column chart. To lay that out
horizontally, name the measure on `x`.

A bar needs something to measure, so two categorical axes are refused:

```{r}
#| error: true
data(medals) + bar + x(country) + y(country)
```

## Texture

A bar's fill can carry a **texture** as well as a color: a hatch instead of a
solid block of paint, set with [`style(pattern = )`](../style.qmd). A texture
reads without color, so a hatched bar chart stays legible in grayscale or a
black-and-white print:

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(pattern = "hatch") +
  title("A hatched bar reads without color")
```

*"Given the medals: bars, x is country, y is gold, with pattern hatch."*

The five fill textures are `"solid"` (the default), `"hatch"`, `"crosshatch"`,
`"grid"`, and `"dots"`; the [style chapter](../style.qmd) shows them beside the
dashes a stroke takes, because `style(pattern = )` is one name with one meaning
per shape. One texture covers the whole layer; to give each color category its
*own* hatch, map a column with the [`pattern()`
channel](../channels.qmd#pattern), which does for texture what `shape()` does
for glyphs.

## Histogram

How are the world's life expectancies spread out, and where do most countries
fall? That is a question about one continuous column, and a bar needs a
category to stand on. Combine `bar` with the `bin` transform to create a
histogram. `bin` groups
continuous `x` values into equal-width bins and counts the rows in each:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * bin + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Histogram of life expectancy, 2007")
```

*"Given gapminder 2007: bars derived by bin, x is life, y is count."*

Thickness is automatic, and it is the clearest way a histogram and a bar chart
differ on the page. A **categorical** bar fills 80% of the slot its category
owns, and the empty fifth is what reads as "separate categories". A histogram
cuts a continuous axis into adjacent intervals, so nothing sits between one bin
and the next and the bars **touch**, with a thin line in the panel's background
color to keep them readable. You set nothing in either plot. The bar chart at
the top of this chapter and this histogram both take their thickness from the
axis the bars sit on.

How fine the cut is decides what the histogram shows. Left alone, `bin` picks
the number of bins from the number of rows, by Sturges' rule. Name a width to
cut the axis in the data's own units, here five years:

```{r}
data(gapminder_2007) + bar * bin(width = 5) + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Bins five years wide")
```

*"Given gapminder 2007: bars derived by bin at width 5, x is life, y is count."*

`bin(30)` sets a count instead. Both belong to the transform, and [Choosing the
bin count](../transforms.qmd#choosing-the-bin-count) shows what each does to
the picture.

A histogram turns on its side by the same rule. Bind the measured column to `y`
and the count is drawn along `x`:

```{r}
data(gapminder_2007) + bar * bin + y(life) +
  y_label("Life expectancy (years)") +
  title("The same histogram, horizontal")
```

### Overlaid histograms: one distribution per group {#overlaid-histograms}

Add `color` and the histogram **splits**: one histogram per category, cut on the
same bins and drawn in the same panel. This answers the question a single
combined histogram cannot: not "how is petal length distributed?" but "how is it
distributed *for each species?*"

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  x_label("Petal length (cm)") + title("Petal length by species")
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."*

The three histograms **overlay**: each drawn in place, measured from the shared
baseline. If the fills were opaque, the last species drawn would hide the
others. So each bar is a translucent fill under a solid outline in its own
color. The outline is what keeps the shape readable where the fills overlap.
`setosa` sits alone on the left and reads clearly. `versicolor` and `virginica`
overlap in the middle, and their outlines still separate them. The bins are
shared across the groups, so the bars align and the overlap is real rather than
an artifact of three different binnings.

The fill is faint by default so the series underneath show through.
`style(opacity = )` sets how faint; the outline stays solid whatever you choose:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(opacity = 0.6) + x_label("Petal length (cm)") +
  title("A heavier fill: style(opacity = 0.6)")
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species, with opacity 0.6."*

The outline is the series color by default, but you can recolor it:
`style(border_color = )` sets the border, `style(border_size = )` its width. A
black border separates the overlapping fills, the way many overlaid histograms
are drawn. It also stays visible against the light panel, where a white one
fades:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(border_color = "black", border_size = 1.5) +
  x_label("Petal length (cm)") + title("Black borders between the fills")
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species, with border color black and border size 1.5."*

The fill and the outline are independent: a bar can be one color inside and
another at its edge, whether the fill was mapped (as here) or set. See
[Setting vs mapping](../style.qmd).

Or remove the outline entirely. `style(border_size = 0)` draws no border, so the
translucent fills overlap with nothing between them. This is the plainest
overlaid histogram:

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species) +
  style(border_size = 0) +
  x_label("Petal length (cm)") + title("No border: just the fills")
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species, with border size 0."*

When the groups overlap so much that the outlines run together, give each
species its own panel instead. Adding [`| facet(species)`](../faceting.qmd) to
the sentence draws the same three histograms side by side, on one shared scale.

See [Transforms](../transforms.qmd) for details on `bin` and other transforms.

## Grouped bars: side by side

How did each continent's life expectancy change between 1957 and 2007? That is a
comparison inside each category, so each continent needs two bars, one per era.
A color split alone draws both bars in one slot, the shorter inside the taller.
To set them **side by side** as a grouped bar chart, add
[`dodge`](../transforms.qmd). Each group's bar narrows to share the slot, and
the split reads as a comparison within each category. The `gm_eras` table holds
two years, 1957 and 2007, with one row per country in each. `mean` averages
those rows down to one value per continent, and `era` is the column that splits
them.

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  y_label("Mean life expectancy") +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

*"Given the gapminder eras: bars derived by mean and dodge, x is continent, y is
life, color by era."*

`dodge` is a collision modifier: it moves the bars rather than summarizing them,
so it composes with the statistic (`bar * mean * dodge` here, `bar * count *
dodge` for tallies). See [Transforms](../transforms.qmd) for the whole family.

`mean` is the right statistic here, and gog will not choose it for you. Write
`sum` in its place and the sentence still draws:

```{r}
data(gm_eras) + bar * sum * dodge + x(continent) + y(life) + color(era) +
  y_label("Sum of life expectancies") +
  title("Sum in place of mean: legal, and meaningless")
```

*"Given the gapminder eras: bars derived by sum and dodge, x is continent, y is
life, color by era."*

Africa's fifty-two life expectancies are added into one bar of 2,850 years.
Life expectancies do not add. The grammar writes either; the data decides which
is honest, and [Law 8](../design-laws.qmd#law-8-pronounceable-usable) says why
gog refuses to decide for you.

## Stacked bars: piled up

Sometimes the question is the total as well as the parts: how much is there
altogether, and how much of it does each group hold? The other way to place a
color split: instead of side by side, pile the groups
**on top of each other** with [`stack`](../transforms.qmd). Each bar's height
becomes the category's total across the groups, and the segments read as parts
of a whole:

```{r}
data(gm_eras) + bar * sum * stack + x(continent) + y(population) + color(era) +
  y_label("Population") +
  title("Population by continent, 1957 and 2007 stacked")
```

*"Given the gapminder eras: bars derived by sum and stack, x is continent, y is
population, color by era."*

`dodge` compares the groups (equal baselines, side by side); `stack` sums them
(one bar, segments piled). Use `stack` when the parts add to a meaningful total
(populations, counts, revenue), and `dodge` when you are comparing the groups.
Note which statistic each one takes: you stack `sum` of population, because
quantities add, while the dodge example compared `mean` of life, because averages
do not.

The grouped bars above have a mirror here. `mean` in place of `sum` draws too,
and piles two averages into a bar that reads as their total:

```{r}
data(gm_eras) + bar * mean * stack + x(continent) + y(life) + color(era) +
  y_label("Mean life expectancy, stacked") +
  title("Mean in place of sum: legal, and meaningless")
```

*"Given the gapminder eras: bars derived by mean and stack, x is continent, y is
life, color by era."*

Oceania's pile reaches 150 years. A stack reads as parts of a total, so it
wants a quantity that has one, which is why this section sums `population` and
the grouped bars averaged `life`.

### The 100% stacked bar

The plot above answers two questions at once, and the larger answer covers the
smaller one. Asia's two eras total 5.4 billion and Oceania's total 0.04 billion,
so the axis is scaled for Asia and the *split* inside the small continents
cannot be read. When the composition is the question and the total is not,
`stack(share = TRUE)` divides each pile by its own bar's total, so every bar
reaches 1 and the segments are the only thing left to read:

```{r}
data(gm_eras) + bar * sum * stack(share = TRUE) + x(continent) + y(population) +
  color(era) + title("Each continent's 1957/2007 split, filled to one")
```

*"Given the gapminder eras: bars derived by sum and stack as shares, x is
continent, y is population, color by era."*

Now the comparison is between continents rather than inside one, and two
continents show the range. **Africa** is the most uneven: 78% of its two-era
total falls in 2007, so its population more than tripled. **Europe** is the most
even at 57%, the closest any continent comes to a half-and-half split. Neither
split can be read in the stacked plot above this section, at any size. What you
lose is the totals: the plot no longer shows them at all. That is the cost of
the choice, not a fault. Oceania's bar is now as tall as Asia's.

The axis says `Share`, because the numbers are fractions of one now, whatever
they were before. That is true even though `y(population)` names a column:
`stack(share = TRUE)` rescales an axis you already named. It is not
[`proportion`](../transforms.qmd#proportion-relative-frequency) under another
name. The two divide by different totals. `proportion` divides by the whole
plot's total, so its bars still show how large each category is. This one
divides by the bar's own total. And only this one works on a `sum` of a column,
which `proportion` has no way to say.

## What you can set

A setting changes how every bar looks without reading a column, and each mark
takes its own. These are a bar's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("bar")
```

A bar has no `size`, and that is deliberate. Its position and its value already
fix how large it is, so a width setting would be a second, conflicting answer.
What it does have, being a shape with an inside and an outline, is a border.

A fill, a texture and a border compose, so one layer can carry all three:

```{r}
data(medals) + bar * sum + x(country) + y(gold) +
  style(color = "goldenrod", pattern = "crosshatch",
        border_color = "black", border_size = 1) +
  y_label("Gold medals") + title("A fill, a texture, and a border")
```

*"Given the medals: bars derived by sum, x is country, y is gold, colored
goldenrod, with pattern crosshatch, border color black and border size 1."*

The `pattern` values here are the five **fill textures**, not the three dashes a
`line` takes. One setting name, and each shape realizes it in its own way, so a
plot that has to survive grayscale printing can separate its bars by texture
instead of color.

The one thing left to adjust is the histogram's bins, and that is not a setting.
`bin(bins = 20)` and `bin(width = 5)` control the histogram `bar * bin` draws.
They belong to the transform, because binning is what has a parameter, not the
rectangle.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The two refusals below come from different facts. The first belongs to the
transform rather than the mark: `bin` needs a continuous axis to cut. The second
belongs to the mark: a bar is a fill rather than a stroke.

Binning cuts a continuous axis into intervals, so there is nothing for it to cut
on a categorical one:

```{r}
#| error: true
data(gapminder_2007) + bar * bin + x(continent)
```

A category is already one slot per value, which is what `count` is for. The
refusal says so rather than drawing empty bins.

Texture is the second refusal. A bar is a fill, so it takes a fill's textures,
and a dash is a stroke's:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(pattern = "dashed")
```

A bar's outline takes a color and a width, never a dash. `border_color` and
`border_size` are the only two settings the outline has. And a bar given two
categorical positions has nothing left to measure, which is the refusal shown
above under [Horizontal bars](#horizontal-bars).