# Box

```{r}
#| include: false
source("../R/setup.R")
```

How is each group spread out? `box` draws the **box-and-whisker** of a
distribution, one per group. In each category it summarizes the measured column.
A box spans the lower quartile to the upper, with a line at the median. Whiskers
reach to the most extreme values *within 1.5 times the interquartile range (IQR)
of the box*. Any values beyond that limit are outliers, drawn individually as
dots. `bar * mean` gives only one number per group, while `box` shows how the
whole group is spread. An [`interval`](interval.qmd) needs a transform to supply
its low and high values, while a box computes its whole summary itself. Boxes
are vertical when the categories are on `x` and horizontal when they are on
`y`; see [Horizontal boxes](#horizontal-boxes).

The construction is John Tukey's, from *Exploratory Data Analysis* [@tukey1977],
where he called it a **schematic plot**. The name readers kept was his informal
one, box-and-whisker, now usually shortened to **box plot**. Mary Eleanor
Spear's **range bar** of 1952 [@spear1952] had drawn the box and the extremes a
quarter century earlier. The range bar had no rule for the **fence**, the cutoff
that decides which values are outliers. That rule, 1.5 times the IQR, is
Tukey's: it decides where ordinary values stop and unusual ones begin.

There is one difference from his book. Tukey's box spans **hinges**. Sort the
values and cut the sorted list in half; when the count is odd, the middle value
belongs to both halves. Take the median of each half: those two medians are the
hinges. gog spans the lower and upper quartiles instead, as most software does.
The two agree when a group has an odd number of values, and can differ slightly
when the number is even. A box drawn by gog is therefore not always the box
Tukey's own rule gives.

## A summary per group

How do the three species differ in petal length, not only on average but in how
widely each varies? Put the categorical column, `species`, on `x`, and put the
measured column, `petal_length`, on `y`. The mark computes the summary itself:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Distribution of petal length per species")
```

*"Given the iris flowers: boxes, x is species, y is petal length."*

Every box is built the same way. Half the flowers of a species fall inside the
box, the middle 50% from the lower quartile to the upper. The line across the
box is the median. Each whisker runs from the box to the most extreme value
inside the fence. The separate dots are the outliers: flowers more than 1.5
times the IQR from the box. The whisker stops before them, and each one is
drawn on its own.

## Distribution across many groups

Which continent varies the most, and which countries fall far below the rest
of their own continent? Five boxes answer both, and you read each one exactly
as you read the species boxes. In this plot each outlier is one country. This
is life expectancy across the continents in 2007, five distributions in one
panel:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("Life expectancy by continent, 2007")
```

Africa's box is the tallest, so the middle half of its countries is the most
spread out. The low dot below Asia is Afghanistan, and the one below the
Americas is Haiti. Each sits far enough below its continent's box that no
whisker reaches it. That is what Tukey's construction is for: the box and
whiskers cover the ordinary values, and the dots mark the unusual ones.

One more case can occur in your own data. When every value in a group's middle
50% is the same, the box has no height and draws as a flat line. gog does not
give it a minimum height to make the spread look larger. Oceania's 1957 box in
[Coloring the boxes](#coloring-the-boxes) is almost this case: its two
countries differ by 0.07 years, so the box draws as a flat line.

## Horizontal boxes

When the category names are long, or there are many of them, the boxes should
be horizontal. Swap the two positions, and the plot is drawn with its axes
exchanged. If you know ggplot2's `coord_flip()`, there is no such word here,
because it would be a second way to say one thing. gog reads the two bindings
and decides from them which axis carries the numbers, exactly as
[`bar`](bar.qmd) does.

```{r}
data(gapminder_2007) + box + x(life) + y(continent) +
  x_label("Life expectancy (years)") + y_label("Continent") +
  title("The same five distributions, horizontal")
```

Compare it with the vertical version in [Distribution across many
groups](#distribution-across-many-groups). Every box, whisker and outlier dot
keeps its place relative to its neighbors. Only the two axes have changed
places. Afghanistan and Haiti are still the two dots beyond their continents'
whiskers. Everything later in this chapter works the same on either axis: the
whisker rule, the color split, `dodge` and `style()`.

The swap matters most when the names are long. The `tenure` table is invented:
five teams, thirty people in each, and the years each person has been at the
company. With the categories on `x`, each name would have to fit into a slot
one box wide. With the categories on `y`, as below, each name is written in
full on one line:

```{r}
#| echo: false
peek(tenure)
```

```{r}
data(tenure) + box + x(years) + y(team) +
  x_label("Years at the company") +
  title("Team names fit on a vertical axis")
```

A box needs a continuous column to summarize, so two categorical axes are
refused, exactly as they are on a bar. [What it refuses](#what-it-refuses)
shows that refusal.

## Whiskers to the extremes

Sometimes you do not want outliers drawn separately. You want the whiskers to
run to the true minimum and maximum. The box then shows the **five-number
summary**: the minimum, the lower quartile, the median, the upper quartile and
the maximum. The mark takes one **parameter** for this, a named option inside
its parentheses: `box(whiskers = "range")`:

```{r}
data(gapminder_2007) + box(whiskers = "range") + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("The same, with whiskers to the full range")
```

*"Given gapminder 2007: boxes with whiskers to the range, x is continent, y is
life."*

Now Afghanistan and Haiti are the ends of their whiskers rather than dots, and
no value is drawn as an outlier. The default is `"tukey"`: fences and outliers,
the standard box plot. `"range"` draws the whole five-number summary.
`whiskers` changes only how the summary is computed; the settings that change
how a box looks are in [What you can set](#what-you-can-set).

## Why `box` needs no transform

An [`interval`](interval.qmd) cannot draw without a transform to give it a low
and a high. A `box` never needs one. An `interval` is only a shape, a span from
a low value to a high one. *Which* low and high it spans is a real choice: a
min–max range, a confidence interval, or a pre-computed pair. A `box` is a
named statistical glyph: the summary *is* what a box plot means. A statistic
that is always drawn by the same mark belongs inside that mark, not in a
separate transform. So the whole box is `box + x + y`. Adding any transform
other than `dodge` is refused, and the message says what to write instead, as
[What it refuses](#what-it-refuses) shows.

The same principle keeps the [channel
vocabulary](../grammar.qmd#channels-a-column-made-visible) small: something that
only one mark can use belongs inside that mark, not in the shared vocabulary.
Taking ggplot2's `ymin`, `lower`, `middle`, `upper` and `ymax` as channels
would add *five* channels meaningful to a single mark. Instead the summary is
computed inside the mark, and the channel vocabulary does not grow.

The table below sets the four marks side by side. Three of them read a low and
a high value from a transform, and `box` computes its own summary:

| Mark | The spread it draws |
|---|---|
| `interval * range` | a whisker from the low value to the high |
| `ribbon * range` | the band between the two values, filled |
| `line * range` | the two values as two separate strokes |
| `box` | quartiles, median and whiskers, with no transform to write |

`box` is also one of several marks that draw a distribution, and each answers a
different question. [`bar * bin`](bar.qmd) draws one variable's histogram, and
[`interval`](interval.qmd) draws a range or an error bar. `box` gives the
summary per group. The violin
([`ribbon * density`](ribbon.qmd#the-violin-a-distribution-per-category))
draws an estimate of the whole distribution that summary is computed from. See
[Transforms](../transforms.qmd) for how `mean`, `median` and `range` differ from
the box's own summary.

## Coloring the boxes

The three species boxes in [A summary per group](#a-summary-per-group) are all
one color, so only the axis labels identify each box. A color per species lets
a reader match each box to the same species in another plot. A color also
splits each slot by a second category, such as an era. `color` gives each box
its own hue, and adds one legend that names the hues. It is the same categorical
split that `bar` and `interval` make. Here `color` is mapped to the same column
as `x`, so it colors one box per group:

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) + color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same distributions, colored by species")
```

*"Given the iris flowers: boxes, x is species, y is petal length, color by
species."*

Mapping `color` to a *second* category splits each category on x into one box
per color. Without `dodge` those boxes sit at the same x and overlap, and their
fills turn translucent so each stays readable.
[`dodge`](../transforms.qmd#dodge-side-by-side-instead-of-on-top) sets them
side by side, each continent's two eras next to each other:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + box * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy") +
  title("Life expectancy per continent, 1957 vs 2007")
```

*"Given the gapminder eras: boxes derived by dodge, x is continent, y is life,
color by era."*

In every continent, the 2007 box sits entirely above the 1957 box.

On a box, `size` and `opacity` are settings rather than channels. `color` can
split one group into several boxes, but `size` and `opacity` cannot. A single
box is one shape, so it cannot show two sizes or two opacities inside itself.
`style(size = )` sets the width of every line a box draws, for the whole layer.
That covers the outline, the whiskers and the caps, the short crossbars at the
whisker ends; the median is one pixel thicker. A box always draws its caps:
`caps` is a setting of `interval` alone.

## What you can set

A setting changes how every box looks without reading a column, and each mark
takes its own. These are a box's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("box")
```

`border_color` and `border_size` are useful on a box in particular, because a
box draws lines *and* a fill. The lines are the outline, the whiskers, the caps
and the median. `border_color` gives them a color of their own; without it,
they take the fill's color, so the median barely shows against the fill. A pale
fill with a dark border keeps the median easy to see. `border_size` sets the
same width as `size`, and when both are given, the box uses `border_size`.

A box is also the one mark in the kernel that takes a parameter. The other
parameters you have met belong to a transform, such as `bin(width = )`, or to a
position, such as `x(gdp, scale = "log")`. A box computes its own summary, so
`whiskers`, which changes how that summary is computed, belongs to the mark.
The plot below shows both, whiskers to the range and a pale fill with a navy
border. It reads `gm_continents`: the Americas, Asia and Europe, measured every
five years from 1952 to 2007. That is twelve years in each box, so a box here
is drawn from twelve times as many values as one drawn from gapminder 2007:

```{r}
#| echo: false
peek(gm_continents)
```

```{r}
data(gm_continents) + box(whiskers = "range") + x(continent) + y(life) +
  style(color = "lightsteelblue", border_color = "navy", border_size = 1.2) +
  y_label("Life expectancy") + title("Whiskers to the true extremes, and a border")
```

*"Given the gapminder continents: boxes with whiskers to the range, x is
continent, y is life, colored lightsteelblue, with border color navy and border
size 1.2."*

Both values of `whiskers` still draw a box. `whiskers` changes the rule the
summary follows, not the shape the summary is drawn as.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

Each of the four refusals below has a different reason. The first two concern
how a box is drawn, and the last two concern what it summarizes. A dashed
`pattern` is the first. A box's `pattern` textures its fill while its lines
stay solid, so it takes a fill's texture and not a stroke's dash. `"hatch"`,
`"crosshatch"`, `"stripes"`, `"grid"` and `"dots"` are accepted, and
`"dashed"` is refused.

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + style(pattern = "dashed")
```

A mapped `size` is the second. No column can vary a box's size. The box's width
is fixed by the slot it occupies. The width of the lines it draws is a setting
for the whole layer:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + size(life)
```

Two categorical axes are the third. gog reads a box's two bindings to find the
axis that carries the numbers, so one axis has to be continuous. Two categories
leave it nothing to summarize, exactly as they do on a bar:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(continent)
```

A statistic is the fourth. A box already summarizes each group into quartiles,
a median and whiskers, so the only transform it takes is `dodge`, which
computes nothing. [Why `box` needs no transform](#why-box-needs-no-transform)
gave the reason:

```{r}
#| error: true
data(iris_flowers) + box * mean + x(species) + y(petal_length)
```

The message names the fix: drop the transform, and the box draws its own
summary.

## In short

The summary is inside the mark: a `box` refuses every statistic, and takes only
`dodge`, which moves boxes and computes nothing. `whiskers` is its one
parameter. A box's width is fixed by its slot, so no column can vary it. gog's
box spans the quartiles, and Tukey's spans hinges, so on a group with an even
number of values the two can differ slightly.

## What comes next

A box occupies one position on its axis, and its width carries no value from
the table. [Ribbon](ribbon.qmd) fills the band between a low value and a high
one at every position along the axis.
