# Box

```{r}
#| include: false
source("../R/setup.R")
```

How is each group spread out? `box` draws the **box-and-whisker of a
distribution**, one per group. At each category it summarizes the measured
column. A box spans the lower quartile to the upper, with a line at the median.
Whiskers reach to the most extreme values *within 1.5×IQR of the box*. Any
points beyond that are outliers, drawn individually as dots. It is the answer
to *how is this spread out, within each group?* Whereas `bar * mean` gives one
number per group, `box` gives the shape of all of them, and flags the few that
sit apart. Boxes stand up by default and lie down when the categories are on
`y`; see [Boxes on their side](#boxes-on-their-side).

The construction is John Tukey's, from *Exploratory Data Analysis*
[@tukey1977], where he called it a **schematic plot**. The name that stuck was
his informal one, box-and-whisker, now usually shortened to **box plot**. Mary
Eleanor Spear's **range bar** of 1952 [@spear1952] had drawn the box and the extremes a
quarter century earlier, without the fence that makes this version worth
drawing: the 1.5×IQR rule is Tukey's, and so is the decision it encodes.

One difference is worth knowing if you arrived from his book. Tukey's box spans
**hinges**, the medians of the sorted batch's two halves; GOG spans the lower and
upper **quartiles**, as most software does. They agree on most samples and can
differ by a little on small ones, so a box drawn here is not always the box *EDA*
would have drawn.

## A summary per group

Name the category that splits the data on x and the measured column on y. The box
does the rest:

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Distribution of petal length per species")
```

*"Given the iris flowers: boxes, x is species, y is petal length."*

Every box reads the same way. Half the flowers of a species fall inside the box
(the middle 50%, from the lower quartile to the upper); the line across it is the
median; the whiskers run to the rest of the bulk. The stray dots are the
outliers, petals far enough from the box to stand out, which the whisker rule
separates out rather than stretching a whisker to reach.

## Distribution across many groups

The comparison scales, and the outliers start to carry information. Here is life expectancy
across the continents in 2007. Five distributions read at a glance, and two
countries that sit apart from their neighbors:

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("Life expectancy by continent, 2007")
```

The low dot under Asia is Afghanistan; the one under the Americas is Haiti, each
far enough below its continent's box to be flagged rather than swept into a
whisker. That is the whole point of the Tukey construction: the whiskers describe
the bulk, and the dots say *these are unusual*. A zero-height box, a group whose
middle 50% share almost one value, collapses to a flat line honestly, rather than
being nudged open.

## Boxes on their side

Swap the two positions and the same sentence draws the same plot, turned. There
is no `flip` atom, for the reason there is none on [`bar`](bar.qmd): GOG reads
which axis measures off the bindings, and a second way to say one thing is a
second thing to learn.

```{r}
data(gapminder_2007) + box + x(life) + y(continent) +
  x_label("Life expectancy (years)") + y_label("Continent") +
  title("The same five distributions, lying down")
```

Set it beside the upright version above. Every box, whisker and outlier dot sits
in the same place relative to its neighbors, rotated a quarter turn, and
Afghanistan and Haiti are still the two dots hanging off their continents.
Nothing else in the chapter changes with the orientation either: `color`,
`dodge`, the `whiskers` knob and `style()` all behave identically whichever axis
holds the categories.

Use it when the category names are long. A vertical axis gives each label
a whole line, where an upright chart has to fit it into a slot one box wide:

```{r}
set.seed(4)
.teams <- c("Research & Development", "Sales and Marketing",
            "Customer Operations", "Finance", "People & Culture")
tenure <- data.frame(
  team  = rep(.teams, each = 30),
  years = round(pmax(rnorm(150, rep(c(6.2, 3.1, 2.4, 7.8, 4.5), each = 30),
                           rep(c(2.4, 1.6, 1.2, 3.0, 2.0), each = 30)), 0.2), 1)
)
data(tenure) + box + x(years) + y(team) +
  x_label("Years at the company") +
  title("Long labels fit on a vertical axis")
```

A box needs a number to summarize, so two categorical axes are refused, exactly
as they are on a bar:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(continent)
```

## Whiskers to the extremes

Sometimes you do not want outliers split off: you want the whiskers to run to the
true minimum and maximum, the plain five-number summary. That is the one knob the
box mark carries, `box(whiskers = "range")`:

```{r}
data(gapminder_2007) + box(whiskers = "range") + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("The same, with whiskers to the full range")
```

Now Afghanistan and Haiti are the ends of their whiskers rather than dots. No
point is called an outlier. The default is `"tukey"` (fences and outliers, the
standard box plot); `"range"` is the whole five-number summary with nothing set
apart. It is a parameter of the mark, the way `bin(30)` and `confidence(0.99)`
parameterize their statistics, the only choice a box offers.

## Why `box` needs no transform

An [`interval`](interval.qmd) cannot draw without a `range` transform to hand it a
low and a high; a `box` never asks for one. The reason is a small but real
distinction. `interval` is pure geometry ("span from here to there") and *which*
low and high it spans is a genuine choice (a min–max range, a confidence
interval, a precomputed pair). A `box` is a named statistical glyph: the summary
*is* what a box plot means. A statistic that draws as exactly one mark belongs *to*
that mark, not beside it as a transform you compose. So the whole box is
`box + x + y`, and asking for a transform on top is refused with direction:

```{r}
#| error: true
data(iris_flowers) + box * mean + x(species) + y(petal_length)
```

This is the same principle that keeps the [channel vocabulary](../grammar.qmd)
small, seen from the other side. Borrowing `ymin`/`ymax`/`lower`/`middle`/`upper`
as channels would add *five* channels meaningful to a single mark; instead the
summary is invented inside the mark, and the channel set never grows.

## Coloring the boxes

`color` gives each box its own hue and a legend, the same discrete split `bar`
and `interval` make. Mapped to the same column that sits on x, it colors one box
per group:

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) + color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same distributions, colored by species")
```

Mapping `color` to a *second* category splits each x-group into one box per
color. Left alone those boxes share an x and overlap (their fills turn
translucent to stay readable) and [`dodge`](../transforms.qmd) sets them side by
side, each continent's two eras next to each other:

```{r}
data(gm_eras) + box * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy") +
  title("Life expectancy per continent, 1957 vs 2007")
```

Width and opacity stay *settings*: a box is one shape, so there is nothing for a
per-row `size` to vary, and `style()` sets them for the whole layer.

`box` completes the distribution tools: [`bar * bin`](bar.qmd) for one variable's
histogram, [`interval`](interval.qmd) for a range or an error bar, and `box` for
the summary per group. The same low/high pair drawn as a continuous fill is the
[`ribbon`](ribbon.qmd) band. See [Transforms](../transforms.qmd) for `dodge` and
how the summarizing statistics relate.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("box")
```

A box is the one mark in the kernel with a **parameter of its own**. Every other
knob in the grammar belongs to a transform (`bin(bins = )`, `confidence(0.99)`)
or to `style()`, but a box carries its summary intrinsically, so the one choice
that summary offers has nowhere else to live:

```{r}
data(gm_continents) + box(whiskers = "range") + x(continent) + y(life) +
  style(color = "lightsteelblue", border_color = "navy", border_size = 1.2) +
  y_label("Life expectancy") + title("Whiskers to the true extremes, and a rim")
```

`box(whiskers = "range")` runs the whiskers to the true minimum and maximum with
no outliers, against the default `"tukey"`, which stops at 1.5 times the
interquartile range and draws the points beyond as dots. Both still draw as a
box, which is why it is a parameter and not a transform: it changes the rule the
summary follows, not the geometry it becomes.

`border_color` and `border_size` are worth using on a box specifically,
because a box is line-work *and* fill. The setting takes the outline, the
whiskers and the median together, so a pale fill with a dark edge reads at
small sizes where a solid fill does not.

## What it refuses

A box is line-work *and* fill, which is exactly what its texture refusal turns
on: a fill takes a texture, and a dash belongs to a stroke.

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + style(pattern = "dashed")
```

Width is the other one. A box has no width of its own to vary, because the width
it draws is the slot it sits in:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + size(life)
```

The statistic refusal a box makes is above, under why it needs no transform.
