# Box

```{r}
#| include: false
source("../R/setup.R")
```

How is each group spread out? `box` draws the **box-and-whisker** of a
distribution, one per group. In each category it summarizes the measured column.
A box spans the lower quartile to the upper, with a line at the median. Whiskers
reach to the most extreme values *within 1.5 times the interquartile range (IQR)
of the box*. Any points beyond that are outliers, drawn individually as dots.
`bar * mean` gives one number per group. `box` shows how the whole group is
spread, and draws the few values far from the rest as separate dots. Boxes are
vertical when the categories are on `x` and horizontal when they are on `y`;
see [Horizontal boxes](#horizontal-boxes).

The construction is John Tukey's, from *Exploratory Data Analysis* [@tukey1977],
where he called it a **schematic plot**. The name readers kept was his informal
one, box-and-whisker, now usually shortened to **box plot**. Mary Eleanor
Spear's **range bar** of 1952 [@spear1952] had drawn the box and the extremes a
quarter century earlier. The range bar had no rule for the **fence**, the cutoff
that decides which values are outliers. That rule, 1.5 times the IQR, is
Tukey's, and so is the choice of where ordinary values stop and unusual ones
begin.

There is one difference from his book. Tukey's box spans **hinges**. Sort the
values, cut the sorted list in half, and take the median of each half. gog spans
the lower and upper **quartiles** instead, as most software does. The two agree
when a group has an odd number of values and differ slightly when the number is
even. A box drawn by gog is therefore not always the box Tukey's own rule gives.

## A summary per group

How do the three species differ in petal length, not only on average but in how
widely each varies? Put the category that splits the data on x, and put the
measured column on y. The mark computes the summary itself:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Distribution of petal length per species")
```

*"Given the iris flowers: boxes, x is species, y is petal length."*

Every box is built the same way. Half the flowers of a species fall inside the
box, the middle 50% from the lower quartile to the upper. The line across the
box is the median. The whiskers run to the rest of the values inside the fence.
The separate dots are the outliers: flowers more than 1.5 IQR from the box. The
whisker stops before them, and each one is drawn on its own.

## Distribution across many groups

Which continent varies the most, and which countries fall far below the rest
of their own continent? Five boxes show both at once. The comparison still
works with more groups. In this plot each outlier is one country. This is life
expectancy across the continents in 2007, five distributions in one panel, with
two countries far below the rest of their continent:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("Life expectancy by continent, 2007")
```

The low dot below Asia is Afghanistan, and the one below the Americas is Haiti.
Each sits far enough below its continent's box that no whisker reaches it. That
is what the Tukey construction is for: the whiskers cover the values near the
middle, and the dots are the values far from it. When every value in a group's
middle 50% is the same, the box has no height and draws as a flat line. gog does
not give it a minimum height to make the spread look larger.

## Horizontal boxes

When the category names are long, or there are many of them, the boxes should
be horizontal. Swap the two positions, and the same sentence draws the same
plot with the two axes exchanged. If you know ggplot2's `coord_flip()`, there
is no such word here, for the same reason [`bar`](bar.qmd) has none. gog reads
the two bindings and decides from them which axis carries the numbers. A second
way to say one thing is a second thing to learn.

```{r}
data(gapminder_2007) + box + x(life) + y(continent) +
  x_label("Life expectancy (years)") + y_label("Continent") +
  title("The same five distributions, horizontal")
```

Compare it with the vertical version in [Distribution across many
groups](#distribution-across-many-groups). Every box, whisker and outlier dot
keeps its place relative to its neighbors. Only the two axes have changed
places. Afghanistan and Haiti are still the two dots beyond their continents'
whiskers. Everything the rest of this chapter adds behaves the same whichever
axis holds the categories: the color split, boxes set side by side, the whisker
rule and `style()`.

Put the categories on `y` when the names are long. The `tenure` table is
invented: five teams, thirty people in each, and the years each person has been
at the company. With the categories on `y`, each name gets a whole line of its
own. With the categories on `x`, each name has to fit into a slot one box wide:

```{r}
#| echo: false
peek(tenure)
```

```{r}
data(tenure) + box + x(years) + y(team) +
  x_label("Years at the company") +
  title("Team names fit on a vertical axis")
```

A box needs a number to summarize, so two categorical axes are refused, exactly
as they are on a bar. [What it refuses](#what-it-refuses) shows that refusal.

## Whiskers to the extremes

Sometimes you do not want outliers drawn separately. You want the whiskers to
run to the true minimum and maximum, the plain five-number summary. That is the
one parameter the box mark carries, `box(whiskers = "range")`:

```{r}
data(gapminder_2007) + box(whiskers = "range") + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("The same, with whiskers to the full range")
```

*"Given gapminder 2007: boxes with whiskers to the range, x is continent, y is
life."*

Now Afghanistan and Haiti are the ends of their whiskers rather than dots. No
point is called an outlier. The default is `"tukey"`: fences and outliers, the
standard box plot. `"range"` is the whole five-number summary, with no value
drawn as an outlier. `whiskers` is the only choice a box offers.

## Why `box` needs no transform

An [`interval`](interval.qmd) cannot draw without a transform to give it a low
and a high. A `box` never needs one. An `interval` is only a shape, a span from
a low value to a high one. *Which* low and high it spans is a real choice: a
min–max range, a confidence interval, or a precomputed pair. A `box` is a named
statistical glyph: the summary *is* what a box plot means. A statistic that
draws as exactly one mark belongs *to* that mark, not beside it as a transform
you compose. So the whole box is `box + x + y`, and adding a transform to it is
refused with direction, as [What it refuses](#what-it-refuses) shows.

This is the principle that keeps the [channel
vocabulary](../grammar.qmd#channels-a-column-made-visible) small, applied to a
mark instead of a channel. Borrowing `ymin`/`ymax`/`lower`/`middle`/`upper` as
channels would add *five* channels meaningful to a single mark. Instead the
summary is invented inside the mark, and the channel set never grows.

The difference is one row in a table. Three marks read a low and a high value
from a transform, and the fourth carries its own:

| Mark | The spread it draws |
|---|---|
| `interval * range` | a whisker from the low value to the high |
| `ribbon * range` | the band between them, filled |
| `line * range` | the two values as two separate strokes |
| `box` | the five-number summary, with no transform to write |

## Coloring the boxes

Three boxes in one color look like one series. A color per species lets a
reader match each box to the same species in another plot. A color is also how
a second category, such as an era, shares each slot. `color` gives each box its
own hue and a legend, the same categorical split `bar` and `interval` make.
Here `color` is mapped to the same column as `x`, so it colors one box per
group:

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) + color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same distributions, colored by species")
```

*"Given the iris flowers: boxes, x is species, y is petal length, color by
species."*

Mapping `color` to a *second* category splits each x-group into one box per
color. Without `dodge` those boxes share one x and overlap, and their fills turn
translucent so each stays readable.
[`dodge`](../transforms.qmd#dodge-side-by-side-instead-of-on-top) sets them
side by side, each continent's two eras next to each other:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + box * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy") +
  title("Life expectancy per continent, 1957 vs 2007")
```

*"Given the gapminder eras: boxes derived by dodge, x is continent, y is life,
color by era."*

Stroke width and opacity stay *settings*. `color` can split one group into
several boxes. `size` and `opacity` cannot. A single box is one shape, so it
cannot show two sizes or two opacities inside itself. `style(size = )` sets
the width of every line a box draws, for the whole layer: the outline, the
whiskers, the median and the caps.

`box` is one of the distribution marks, and each of them answers a different
question. [`bar * bin`](bar.qmd) draws one variable's histogram, and
[`interval`](interval.qmd) draws a range or an error bar. `box` gives the
summary per group. The violin
([`ribbon * density`](ribbon.qmd#the-violin-a-distribution-per-category))
draws the whole distribution that summary is computed from. See
[Transforms](../transforms.qmd) for `dodge`, and for how `mean`, `median` and
`range` differ from the box's own summary.

## What you can set

A setting changes how every box looks without reading a column, and each mark
takes its own. These are a box's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("box")
```

A box is the one mark in the kernel that takes a **parameter**. The other
parameters you have met belong to a transform (`bin(bins = )`), to a position
(`x(gdp, scale = "log")`), or to `style()`. A box carries its summary inside
itself, so the one choice that summary offers belongs to the mark. The plot
below reads `gm_continents`: three continents, every year from 1952 to 2007.
Each box therefore summarizes every country in twelve years rather than in one,
so it is drawn from twelve times as many values as a box drawn from gapminder
2007:

```{r}
#| echo: false
peek(gm_continents)
```

```{r}
data(gm_continents) + box(whiskers = "range") + x(continent) + y(life) +
  style(color = "lightsteelblue", border_color = "navy", border_size = 1.2) +
  y_label("Life expectancy") + title("Whiskers to the true extremes, and a border")
```

Both values of `whiskers` still draw a box, which is why it is a parameter and
not a transform. It changes the rule the summary follows, not the shape the
summary is drawn as.

Two more settings apply whichever value `whiskers` takes. `border_color` and
`border_size` are useful on a box in particular, because a box draws lines
*and* a fill. Both settings apply to every line the box draws: the outline, the
whiskers, the median and the caps. A pale fill with a dark edge stays legible
at small sizes, and a solid fill does not.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A box draws lines and a fill, so its `pattern` takes a fill's texture and not a
stroke's dash. `"hatch"`, `"crosshatch"`, `"stripes"`, `"grid"` and `"dots"`
are accepted, and `"dashed"` is refused.

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + style(pattern = "dashed")
```

A mapped `size` is the second. No row can vary a box's size. The box's own
width is fixed by the slot it occupies, and the width of its lines is a setting
for the whole layer:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + size(life)
```

Two categorical axes are the third. gog decides which axis carries the numbers
from a box's two bindings, so one axis has to be continuous. Two categories
leave it nothing to summarize, exactly as they do on a bar:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(continent)
```

A transform is the fourth. A box already summarizes each group into quartiles,
a median and whiskers, so it takes none. [Why `box` needs no
transform](#why-box-needs-no-transform) gave the reason:

```{r}
#| error: true
data(iris_flowers) + box * mean + x(species) + y(petal_length)
```

The message names the fix: drop the transform, and the box draws its own
summary.

## In short

gog's box spans the quartiles, and Tukey's spans hinges, so on a group with an
even number of values the two differ slightly. The summary is inside the mark:
a `box` refuses every statistic, and takes only `dodge`, which moves boxes and
computes nothing. `whiskers` is its one parameter. A box's width is fixed by
its slot, so no column can vary it.

## What comes next

A box occupies one position on its axis, and its width does not depend on the
axis. [Ribbon](ribbon.qmd) carries a spread along the whole axis, filling the
band between a low boundary and a high one.
