# Box

```{r}
#| include: false
source("../R/setup.R")
```

How is each group spread out? `box` draws the **box-and-whisker of a
distribution**, one per group. At each category it summarizes the measured
column. A box spans the lower quartile to the upper, with a line at the median.
Whiskers reach to the most extreme values *within 1.5 times the interquartile
range (IQR) of the box*. Any points beyond that are outliers, drawn
individually as dots. Whereas `bar * mean` gives one number per group, `box`
gives the shape of all of them, and marks the few values far from the rest. Boxes stand up by default and lie down when the categories are on
`y`; see [Boxes on their side](#boxes-on-their-side).

The construction is John Tukey's, from *Exploratory Data Analysis*
[@tukey1977], where he called it a **schematic plot**. The name readers kept
was his informal one, box-and-whisker, now usually shortened to **box plot**.
Mary Eleanor Spear's **range bar** of 1952 [@spear1952] had drawn the box and
the extremes a quarter century earlier. What it had no rule for is the
**fence**, the cutoff that decides which values count as outliers. That rule,
1.5 times the IQR, is Tukey's, and so is the decision it encodes.

One difference is worth knowing if you know his book. Tukey's box spans
**hinges**, the medians of the sorted batch's two halves; gog spans the lower and
upper **quartiles**, as most software does. They agree on most samples and can
differ by a little on small ones, so a box drawn here is not always the box *EDA*
would have drawn.

## A summary per group

Name the category that splits the data on x and the measured column on y. The
mark computes the summary itself:

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Distribution of petal length per species")
```

*"Given the iris flowers: boxes, x is species, y is petal length."*

Every box reads the same way. Half the flowers of a species fall inside the box
(the middle 50%, from the lower quartile to the upper); the line across it is the
median; the whiskers run to the rest of the bulk. The separate dots are the
outliers: petals more than 1.5 IQR from the box, so the whisker stops short
and each is drawn on its own.

## Distribution across many groups

The comparison still works with more groups, and now the outliers carry
information. Here is life expectancy across the continents in 2007: five
distributions in one panel, and two countries far below their neighbors:

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("Life expectancy by continent, 2007")
```

The low dot under Asia is Afghanistan; the one under the Americas is Haiti,
each far enough below its continent's box to be drawn as its own dot rather
than reached by a whisker. That is what the Tukey construction is for: the
whiskers describe the bulk, and the dots say *these are unusual*. When a
group's middle 50% share almost one value, the box has zero height and draws
as a flat line; gog does not widen it to make it look like more.

## Boxes on their side

Swap the two positions and the same sentence draws the same plot, turned. There
is no `flip` atom, for the reason there is none on [`bar`](bar.qmd): gog
decides which axis measures by reading the two bindings, and a second way to
say one thing is a second thing to learn.

```{r}
data(gapminder_2007) + box + x(life) + y(continent) +
  x_label("Life expectancy (years)") + y_label("Continent") +
  title("The same five distributions, lying down")
```

Set it beside the upright version above. Every box, whisker and outlier dot sits
in the same place relative to its neighbors, rotated a quarter turn, and
Afghanistan and Haiti are still the two dots below their continents'
whiskers. Nothing else in the chapter changes with the orientation either:
`color`, `dodge`, the `whiskers` parameter and `style()` all behave
identically whichever axis holds the categories.

Use it when the category names are long. A vertical axis gives each label
a whole line, while an upright plot has to fit it into a slot one box wide:

```{r}
set.seed(4)
.teams <- c("Research & Development", "Sales & Marketing",
            "Customer Operations", "Finance", "People & Culture")
tenure <- data.frame(
  team  = rep(.teams, each = 30),
  years = round(pmax(rnorm(150, rep(c(6.2, 3.1, 2.4, 7.8, 4.5), each = 30),
                           rep(c(2.4, 1.6, 1.2, 3.0, 2.0), each = 30)), 0.2), 1)
)
data(tenure) + box + x(years) + y(team) +
  x_label("Years at the company") +
  title("Long labels fit on a vertical axis")
```

A box needs a number to summarize, so two categorical axes are refused, exactly
as they are on a bar:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(continent)
```

## Whiskers to the extremes

Sometimes you do not want outliers split off: you want the whiskers to run to the
true minimum and maximum, the plain five-number summary. That is the one
parameter the box mark carries, `box(whiskers = "range")`:

```{r}
data(gapminder_2007) + box(whiskers = "range") + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("The same, with whiskers to the full range")
```

Now Afghanistan and Haiti are the ends of their whiskers rather than dots. No
point is called an outlier. The default is `"tukey"` (fences and outliers, the
standard box plot); `"range"` is the whole five-number summary with nothing set
apart. It is a parameter of the mark, the way `bin(bins = 30)` and
`confidence(0.99)` parameterize their statistics, and it is the only choice a
box offers.

## Why `box` needs no transform

An [`interval`](interval.qmd) cannot draw without a `range` transform to give
it a low and a high; a `box` never needs one. The reason is a small but real
distinction. `interval` is pure geometry ("span from here to there") and *which*
low and high it spans is a genuine choice (a min–max range, a confidence
interval, a precomputed pair). A `box` is a named statistical glyph: the summary
*is* what a box plot means. A statistic that draws as exactly one mark belongs *to*
that mark, not beside it as a transform you compose. So the whole box is
`box + x + y`, and asking for a transform on top is refused with direction:

```{r}
#| error: true
data(iris_flowers) + box * mean + x(species) + y(petal_length)
```

This is the principle that keeps the [channel vocabulary](../grammar.qmd)
small, applied to a mark instead of a channel. Borrowing `ymin`/`ymax`/`lower`/`middle`/`upper`
as channels would add *five* channels meaningful to a single mark; instead the
summary is invented inside the mark, and the channel set never grows.

## Coloring the boxes

`color` gives each box its own hue and a legend, the same discrete split `bar`
and `interval` make. Mapped to the same column that sits on x, it colors one box
per group:

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) + color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same distributions, colored by species")
```

Mapping `color` to a *second* category splits each x-group into one box per
color. Without `dodge` those boxes share one x and overlap, and their fills
turn translucent so both stay readable. [`dodge`](../transforms.qmd) sets
them side by side, each continent's two eras next to each other:

```{r}
data(gm_eras) + box * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy") +
  title("Life expectancy per continent, 1957 vs 2007")
```

Stroke width and opacity stay *settings*: a box is one shape, so there is no
row for `size` or `opacity` to vary over. `style(size = )` is the width of
the line-work, set for the whole layer.

`box` joins the other distribution marks: [`bar * bin`](bar.qmd) for one
variable's histogram, [`interval`](interval.qmd) for a range or an error bar,
`box` for the summary per group, and the violin
([`ribbon * density`](ribbon.qmd#the-violin-a-distribution-per-category)) for
the whole shape behind that summary. See [Transforms](../transforms.qmd) for
`dodge` and how the summarizing statistics relate.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("box")
```

A box is the one mark in the kernel with a **parameter of its own**. Every
other parameter in the grammar belongs to a transform (`bin(bins = )`,
`confidence(0.99)`) or to `style()`, but a box carries its summary inside
itself, so the one choice that summary offers belongs to the mark:

```{r}
data(gm_continents) + box(whiskers = "range") + x(continent) + y(life) +
  style(color = "lightsteelblue", border_color = "navy", border_size = 1.2) +
  y_label("Life expectancy") + title("Whiskers to the true extremes, and a border")
```

Both values of `whiskers` still draw as a box, which is why it is a parameter
and not a transform: it changes the rule the summary follows, not the
geometry it becomes.

`border_color` and `border_size` are worth using on a box specifically,
because a box is line-work *and* fill. The setting takes the outline, the
whiskers and the median together, so a pale fill with a dark edge stays legible at small sizes, while a solid fill does not.

## What it refuses

A box is line-work *and* fill, so its `pattern` takes a fill's texture and
not a stroke's dash: `"hatch"` is accepted, and `"dashed"` is refused.

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + style(pattern = "dashed")
```

A mapped `size` is the other one. No row can vary a box's size: the box's
width is the slot it sits in, and the line-work's width is a layer setting:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + size(life)
```

The statistic refusal a box makes is above, under why it needs no transform.
