# Box

```{r}
#| include: false
source("../R/setup.R")
```

How is each group spread out? `box` draws the **box-and-whisker** of a
distribution, one per group. In each category it summarizes the measured column.
A box spans the lower quartile to the upper, with a line at the median. Whiskers
reach to the most extreme values *within 1.5 times the interquartile range (IQR)
of the box*. Any points beyond that are outliers, drawn individually as dots.
`bar * mean` gives one number per group. `box` shows how the whole group is
spread, and draws the few values far from the rest as separate dots. Boxes stand
up by default and lie down when the categories are on `y`; see [Boxes on their
side](#boxes-on-their-side).

The construction is John Tukey's, from *Exploratory Data Analysis* [@tukey1977],
where he called it a **schematic plot**. The name readers kept was his informal
one, box-and-whisker, now usually shortened to **box plot**. Mary Eleanor
Spear's **range bar** of 1952 [@spear1952] had drawn the box and the extremes a
quarter century earlier. The range bar had no rule for the **fence**, the cutoff
that decides which values are outliers. That rule, 1.5 times the IQR, is
Tukey's, and so is the choice of where ordinary values stop and unusual ones
begin.

One difference is worth knowing if you know his book. Tukey's box spans
**hinges**. Sort the values, cut the sorted list in half, and take the median of
each half. gog spans the lower and upper **quartiles** instead, as most software
does. The two agree on most samples and differ slightly on small ones. A box
drawn by gog is therefore not always the box Tukey's own rule gives.

## A summary per group

Put the category that splits the data on x, and put the measured column on y.
The mark computes the summary itself:

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Distribution of petal length per species")
```

*"Given the iris flowers: boxes, x is species, y is petal length."*

Every box reads the same way. Half the flowers of a species fall inside the box
(the middle 50%, from the lower quartile to the upper); the line across it is
the median; the whiskers run to the rest of the values inside the fence. The
separate dots are the outliers: flowers more than 1.5 IQR from the box. The
whisker stops before them, and each one is drawn on its own.

## Distribution across many groups

The comparison still works with more groups, and here the outliers name
particular countries. This is life expectancy across the continents in 2007,
five distributions in one panel, with two countries far below the rest of their
continent:

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("Life expectancy by continent, 2007")
```

The low dot below Asia is Afghanistan, and the one below the Americas is Haiti.
Each sits far enough below its continent's box that no whisker reaches it. That
is what the Tukey construction is for: the whiskers cover the values near the
middle, and the dots are the values far from it. When every value in a group's
middle 50% is the same, the box has no height and draws as a flat line. gog does
not give it a minimum height to make the spread look larger.

## Boxes on their side

Swap the two positions and the same sentence draws the same plot, turned. There
is no `flip` atom, for the reason there is none on [`bar`](bar.qmd). gog reads
the two bindings and decides from them which axis carries the numbers. A second
way to say one thing is a second thing to learn.

```{r}
data(gapminder_2007) + box + x(life) + y(continent) +
  x_label("Life expectancy (years)") + y_label("Continent") +
  title("The same five distributions, lying down")
```

Compare it with the upright version above. Every box, whisker and outlier dot
keeps its place relative to its neighbors, rotated a quarter turn. Afghanistan
and Haiti are still the two dots beyond their continents' whiskers. Everything
the rest of this chapter adds behaves the same whichever axis holds the
categories: the color split, boxes set side by side, the whisker rule and
`style()`.

Use it when the category names are long. The table below is invented: five
teams, thirty people in each, and the years each person has been at the company.
A vertical axis gives each name a whole line, while an upright plot has to fit
it into a slot one box wide:

```{r}
set.seed(4)
.teams <- c("Research & Development", "Sales & Marketing",
            "Customer Operations", "Finance", "People & Culture")
tenure <- data.frame(
  team  = rep(.teams, each = 30),
  years = round(pmax(rnorm(150, rep(c(6.2, 3.1, 2.4, 7.8, 4.5), each = 30),
                           rep(c(2.4, 1.6, 1.2, 3.0, 2.0), each = 30)), 0.2), 1)
)
data(tenure) + box + x(years) + y(team) +
  x_label("Years at the company") +
  title("Team names fit on a vertical axis")
```

A box needs a number to summarize, so two categorical axes are refused, exactly
as they are on a bar:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(continent)
```

## Whiskers to the extremes

Sometimes you do not want outliers split off: you want the whiskers to run to
the true minimum and maximum, the plain five-number summary. That is the one
parameter the box mark carries, `box(whiskers = "range")`:

```{r}
data(gapminder_2007) + box(whiskers = "range") + x(continent) + y(life) +
  x_label("Continent") + y_label("Life expectancy (years)") +
  title("The same, with whiskers to the full range")
```

*"Given gapminder 2007: boxes with whiskers to the range, x is continent, y is
life."*

Now Afghanistan and Haiti are the ends of their whiskers rather than dots. No
point is called an outlier. The default is `"tukey"` (fences and outliers, the
standard box plot); `"range"` is the whole five-number summary with nothing set
apart. `whiskers` is the only choice a box offers.

## Why `box` needs no transform

An [`interval`](interval.qmd) cannot draw without a `range` transform to give it
a low and a high; a `box` never needs one. `interval` is pure geometry ("span
from here to there") and *which* low and high it spans is a genuine choice (a
min–max range, a confidence interval, a precomputed pair). A `box` is a named
statistical glyph: the summary *is* what a box plot means. A statistic that
draws as exactly one mark belongs *to* that mark, not beside it as a transform
you compose. So the whole box is `box + x + y`, and adding a transform to it is
refused with direction:

```{r}
#| error: true
data(iris_flowers) + box * mean + x(species) + y(petal_length)
```

This is the principle that keeps the [channel vocabulary](../grammar.qmd) small,
applied to a mark instead of a channel. Borrowing
`ymin`/`ymax`/`lower`/`middle`/`upper` as channels would add *five* channels
meaningful to a single mark; instead the summary is invented inside the mark,
and the channel set never grows.

## Coloring the boxes

`color` gives each box its own hue and a legend, the same discrete split `bar`
and `interval` make. Mapped to the same column that sits on x, it colors one box
per group:

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) + color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same distributions, colored by species")
```

*"Given the iris flowers: boxes, x is species, y is petal length, color by
species."*

Mapping `color` to a *second* category splits each x-group into one box per
color. Without `dodge` those boxes share one x and overlap, and their fills turn
translucent so each stays readable. [`dodge`](../transforms.qmd) sets them side
by side, each continent's two eras next to each other:

```{r}
data(gm_eras) + box * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy") +
  title("Life expectancy per continent, 1957 vs 2007")
```

*"Given the gapminder eras: boxes derived by dodge, x is continent, y is life,
color by era."*

Stroke width and opacity stay *settings*. `color` and `pattern` can split one
group into several boxes. `size` and `opacity` cannot: they would have to change
inside a single box, and a box is one shape. `style(size = )` sets the width of
every line a box draws, for the whole layer: the outline, the whiskers, the
median and the end caps.

`box` joins the other distribution marks. [`bar * bin`](bar.qmd) draws one
variable's histogram, and [`interval`](interval.qmd) draws a range or an
error bar. `box` gives the summary per group, and the violin
([`ribbon * density`](ribbon.qmd#the-violin-a-distribution-per-category))
gives the whole shape behind that summary. See
[Transforms](../transforms.qmd) for `dodge`, and for how `mean`, `median` and
`range` differ from the box's own summary.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("box")
```

A box is the one mark in the kernel that takes a **parameter**. The other
parameters you have met belong to a transform (`bin(bins = )`), to a position
(`x(gdp, scale = "log")`), or to `style()`. A box carries its summary inside
itself, so the one choice that summary offers belongs to the mark. The plot
below reads `gm_continents`: three continents, every year from 1952 to 2007.
Each box therefore covers many more countries than the 2007 plots above:

```{r}
data(gm_continents) + box(whiskers = "range") + x(continent) + y(life) +
  style(color = "lightsteelblue", border_color = "navy", border_size = 1.2) +
  y_label("Life expectancy") + title("Whiskers to the true extremes, and a border")
```

Both values of `whiskers` still draw a box, which is why it is a parameter and
not a transform. It changes the rule the summary follows, not the shape the
summary is drawn as.

`border_color` and `border_size` are worth using on a box specifically,
because a box draws lines *and* a fill. Both settings apply to every line the
box draws: the outline, the whiskers, the median and the caps. A pale fill
with a dark edge stays legible at small sizes, and a solid fill does not.

## What it refuses

A box draws lines and a fill, so its `pattern` takes a fill's texture and not a
stroke's dash. `"hatch"`, `"crosshatch"`, `"grid"` and `"dots"` are accepted;
`"dashed"` is refused.

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + style(pattern = "dashed")
```

A mapped `size` is the second. No row can vary a box's size: the box's width is
the slot it sits in, and the line-work's width is a layer setting:

```{r}
#| error: true
data(gapminder_2007) + box + x(continent) + y(life) + size(life)
```

The third refusal is above, under Why `box` needs no transform: a box already
summarizes, so it takes none.