# edge

```{r}
#| include: false
source("../R/setup.R")
```

What draws a relation? Every other stroke in the kernel connects positions
the data placed: a `line` runs through its domain, a `path` follows the
table's order, an `interval` spans between two values. An `edge` connects two
*named things*, and one row of the table is one such connection.

That difference decides everything about the mark. Its two endpoints are not
bound to channels; they come from a [`layout`](../network.qmd), which reads
the two endpoint columns and computes a position for every name. So the
mark's smallest sentence carries three words, the mark, its layout, and the
space the result is read in:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in the network."*

An `edge` does not draw outside `network()`, the way `surface` does not draw
outside the cube. A stroke whose ends mean nothing on data axes has nowhere
to stand in a panel that promises its axes mean something.

## What varies by edge

An edge is one row, so its paint can follow any column of the table. `color`
maps a category per edge, and `opacity` maps a number, which is the reading a
weighted network wants:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) +
  network() +
  title("Each relation, faded by its value")
```

`opacity` maps on this stroke and sets on `line`, and the difference is the
mark's own: a `line` is one stroke with one opacity, while an edge layer is
many strokes, each its own row with its own value.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("edge")
```

`style(size = )` sets the stroke width for the layer, `style(color = )` its
one color when nothing maps, and `style(pattern = )` dashes it.

## What it refuses

An `edge` without its `layout` has no endpoints to draw, and a position bound
to it would say nothing the layout has not already said:

```{r}
#| error: true
data(trade_partners) + edge + x(billions) + y(exporter) + network()
```

And its layout draws only in its space:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

The [Network](../network.qmd) chapter has the whole family: the nodes, the
names, the cube, and the rest of the refusals.
