# Edge

```{r}
#| include: false
source("../R/setup.R")
```

Which things in your table are connected to which? `edge` draws one stroke per
row of the table, between the two things named in that row. Every other mark
that draws a stroke is placed by the axes. A `line` and a `path` join points at
their `x` and `y` values, and an `interval` spans two values on one axis. An
`edge` is different: each row gives it only two names, and no column says where
a name sits on the page. The picture an `edge` layer makes is the **network
diagram**.

So an edge takes no `x` and no `y`, and no other mark refuses both. An edge's
endpoints come from a [`layout`](../network.qmd), a transform that reads the two
columns of names and places each distinct name once. Write the *from* column
first and the *to* column second, so the sentence states the direction of each
relation. If you swap the two columns, the picture does not change. It changes
only when you add arrowheads: `style(arrow = "end")` points each edge from the
first column to the second ([What you can set](#what-you-can-set)).

## The network diagram

A network diagram needs a table with one row per relation. In
`trade_partners`, each row names an exporter and an importer, and gives the
value of the goods in billions of dollars, rounded:

```{r}
#| echo: false
peek(trade_partners)
```

Name the table, then write three things: the mark, the `layout` that derives
it, and the space `network()`:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in
the network."*

The 15 countries become 15 positions on the panel, and the 22 rows become 22
strokes. Five pairs of countries trade in both directions, and each pair's two
strokes lie on one line. So you can count only 17 lines, and those five look
slightly darker than the rest. No country is labeled: the layout placed the
countries, but no layer draws their names.

An `edge` draws only inside `network()`. A layout computes its positions from
the **graph**. Here a graph is not a chart: it is a set of things and the
relations between them. Each thing is a **node**, and each row is one relation
between two nodes. A node is one name, however many rows it appears in: Germany
is one node in seven rows. A layout chooses positions that keep the picture
readable, so the distance between two nodes is not a quantity. An axis would put
numbers on that distance, so gog refuses every other space.

A `layout` is not for edges alone. It gives every node one position, and it adds
a column, `name`, that holds each node's name. Two other marks read a layout and
draw at those same positions:

| Mark derived by `layout` | Shape |
|---|---|
| `point * layout` | one dot per node |
| `text * layout + label(name)` | each node's name |
| `edge * layout` | one stroke per relation, the plot above |

The next section gives each stroke a value from its row, and adds the first of
those three: a dot at each node. The [Network](../network.qmd) chapter draws all
three in one plot.

## Color, opacity and dash, one per row

In the first plot every stroke is drawn the same way, so a relation worth 500
billion dollars looks like one worth 50 billion. The table holds that value, and
a stroke can carry it as opacity. Each edge is drawn from one row, so it can
take its color, its opacity and its dash from that row's values. `opacity` takes a
continuous column such as `billions`. A faint stroke's ends are hard to find. So
a second layer, `point * layout(exporter, importer)`, reads the same layout and
draws a dot at each node:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) +
  network() +
  title("More trade, darker stroke")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, in the network."*

The three darkest lines meet at one node. The table shows which: China, Mexico
and Canada each export more than 400 billion dollars of goods to the United
States. On `line`, `opacity` is only a setting, because a line joins many rows
into one stroke. An edge draws one stroke per row, so each row gives its stroke
an opacity of its own.

Which strokes share an exporter? `color` needs a categorical column such as
`exporter`, and each stroke takes its exporter's color. So the five strokes with
Germany as exporter share one color:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + color(exporter) +
  point * layout(exporter, importer) +
  network() +
  title("Each stroke, colored by its exporter")
```

*"Given trade partners: edges derived by layout from exporter to importer, color
by exporter, and also points derived by layout from exporter to importer, in the
network."*

Mexico, Canada and the Netherlands are in the legend, but no stroke shows their
color. Each trades both ways with one partner, and the partner's stroke is drawn
on top of theirs.

The dash can come from a column too. `pattern(exporter)` gives each exporter a
dash, the way `color(exporter)` gave each a color above. There are only five
dashes: solid, dashed, dotted, dotdash and longdash. When a column has more than
five categories, some of them share a dash, and gog prints a message that says
how many:

```{r}
#| message: true
data(trade_partners) +
  edge * layout(exporter, importer) + pattern(exporter) +
  point * layout(exporter, importer) +
  network() +
  title("Each stroke, dashed by its exporter")
```

*"Given trade partners: edges derived by layout from exporter to importer,
pattern by exporter, and also points derived by layout from exporter to
importer, in the network."*

The legend lists the exporters in the order the table first names them, and the
dashes start again after the fifth. So Germany, sixth, has China's solid stroke,
and so does India, eleventh. Where two countries trade both ways, their two
strokes lie on one line, and the lower dash shows through the gaps in the upper
one. `color(exporter)` gave each of the twelve its own color.

## What you can set

Some choices come from no column: how thick every stroke is, or one dash for
all of them. You set those in the sentence, with `style()`, and each mark takes
its own settings. These are an edge's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("edge")
```

`style(size = )` sets the stroke width for the whole layer. `style(color = )`,
`style(opacity = )` and `style(pattern = )` give every stroke one value, when no
column is mapped to that channel. The settings table above lists every dash that
`style(pattern = )` accepts.

`style(arrow = )` draws an arrowhead on each stroke. With `"end"`, it is at the
node in the second column of `layout()`, the *to* column. With `"start"`, it is
at the node in the first column, and with `"both"`, at both nodes. Each
arrowhead ends outside the dot that a `point * layout` layer draws at the node,
so the dot does not cover it:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + style(arrow = "end") +
  point * layout(exporter, importer) +
  network() +
  title("Each stroke points at its importer")
```

*"Given trade partners: edges derived by layout from exporter to importer, with
arrow end, and also points derived by layout from exporter to importer, in the
network."*

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The other marks take their positions from `x` and `y`, so binding them to an
edge is the first thing most readers try. An `edge` with no `layout` has no
endpoints to draw, and `x` and `y` cannot supply them. gog refuses the `x`
binding, then the `y` binding, and then the `network()`, because no layer gives
it a `layout` to draw:

```{r}
#| error: true
data(trade_partners) + edge + x(billions) + y(exporter) + network()
```

A `layout` is read only inside `network()`, so the sentence has to name that
space:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

The [Network](../network.qmd) chapter adds a label beside each dot and sizes
each dot by its number of relations. It also draws the same graph in a cube, and
lists the other refusals.

## What comes next

An edge never has an axis, because a layout places both of its endpoints. Most
marks are placed by an axis instead, and [Scales](../scales.qmd) is where a
value becomes a position on an axis.
