# Edge

```{r}
#| include: false
source("../R/setup.R")
```

Which things in your table are connected to which? `edge` draws one stroke per
row of the table, between the two things that row names. The other marks that
draw a stroke take both ends from columns the axes measure. A `line` runs along
its domain, a `path` follows the table's order, an `interval` spans two values.
An `edge` is different. Its two endpoints are names rather than numbers, and an
axis can only place a number. The picture an `edge` layer makes is the
**network diagram**.

So an edge takes no `x` and no `y`, and it is the only mark in the kernel that
takes neither. Its two endpoints come from a
[`layout`](../network.qmd), a transform that reads the two name columns and
computes a position for every name. Write the source column first and the
destination second.

## The network diagram

`trade_partners` holds one row per trade relation: who exports, who imports,
and the value of the goods, in billions of dollars, rounded:

```{r}
#| echo: false
peek(trade_partners)
```

Name the table, then write three things: the mark, the `layout` it is derived
by, and the space `network()`:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in
the network."*

An `edge` draws only inside `network()`. A layout computes its positions from
the **graph**, the set of things and the relations between them, rather than
from any column. Ticks under those positions would put numbers on a distance
that measures nothing, so gog refuses every other space.

`layout` is not the edge's own transform. It gives every name in the table one
position. Two other marks read a layout, and both draw at those same positions:

| Mark derived by `layout` | Shape |
|---|---|
| `point * layout` | one dot per node |
| `text * layout` | each node's name |
| `edge * layout` | one stroke per relation, the shape this chapter opens with |

[Network](../network.qmd) layers all three into one plot.

## Color and opacity, one per row

Every stroke in the first plot is drawn alike, so a trade of 500 billion cannot
be told from a trade of 50. The table holds that value, and a stroke can carry
it as opacity.
Each edge is drawn from one row, so its color and its opacity can be mapped from
a column of that row. `color` takes a categorical column, and `opacity` a
continuous one such as `billions`. A second layer,
`point * layout(exporter, importer)`, reads the same layout and draws a dot at
each name:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) +
  network() +
  title("Heavier trade, darker stroke")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, in the network."*

On `line`, `opacity` is only a setting. A line is drawn as a single stroke, and
a single stroke can hold only one opacity. An edge layer is many strokes, and
each row gives its stroke an opacity of its own.

Which strokes start at the same country?
`color` reads a column the same way `opacity` does, one value from each row.
Each stroke takes the color of
its exporter, so Germany's five strokes share one color:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + color(exporter) +
  point * layout(exporter, importer) +
  network() +
  title("Each stroke, colored by its exporter")
```

*"Given trade partners: edges derived by layout from exporter to importer, color
by exporter, and also points derived by layout from exporter to importer, in the
network."*

## What you can set

Whether the strokes are thin or thick, solid or dashed, is not in any column of
the table. The value comes from the sentence rather than from the table, and
each mark takes its own settings. These are an edge's, generated from the
engine's own legality table, so this page cannot differ from what `style()`
accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("edge")
```

`style(size = )` sets the stroke width for the whole layer. `style(color = )`
and `style(opacity = )` give every stroke one value, for when no column is
mapped to them. `style(pattern = )` chooses the dash, and the table above lists
every name it accepts.

Mapping a column to `pattern` is legal grammar that this engine does not draw.
gog refuses such a binding rather than ignoring it. The dash has to come from
`style(pattern = )`.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The other marks take their positions from `x` and `y`, so writing them on an
edge is the first thing most readers try.
An `edge` with no `layout` has no endpoints to draw. `x` and `y` cannot supply
those endpoints. gog refuses each binding in turn, and then refuses the
`network()` that has nothing left to place:

```{r}
#| error: true
data(trade_partners) + edge + x(billions) + y(exporter) + network()
```

A `layout` is read only inside `network()`, so the space has to be written:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

The [Network](../network.qmd) chapter adds the layers this one leaves out. It
draws a dot for every name, a label beside each dot, the same graph in a cube,
and the other refusals.

## What comes next

An edge is placed by a layout rather than by an axis, which is why it draws only
inside `network()`. Most marks are placed by an axis instead, and
[Scales](../scales.qmd) is where a value becomes a place on one.
