# Edge

```{r}
#| include: false
source("../R/setup.R")
```

What draws a relation? `edge` does: one stroke per row of the table, between
the two things that row relates. Every other stroke joins positions bound
from columns. A `line` runs along its domain, a `path` follows the table's
order, an `interval` spans two values. An `edge` is different, because the
ends of an edge are names, not numbers, so no axis can measure them.

That difference is what makes this mark unusual. Its two endpoints are not
bound to channels; they come from a [`layout`](../network.qmd), which reads
the two endpoint columns and computes a position for every name.
`trade_partners` holds one row per trade relation: who exports, who imports,
and the value of the goods, in billions of dollars, rounded:

```{r}
#| echo: false
peek(trade_partners)
```

The smallest sentence needs three parts: the mark, its layout, and the space:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in the network."*

An `edge` draws only inside `network()`. Outside it, every axis promises a
measurement, and an edge's ends have none to offer, so the sentence is
refused rather than drawn.

## What varies by edge

An edge is one row, so its color and its opacity can follow any column of
the table. `color` maps a category, and `opacity` maps a number such as
`billions`:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) +
  network() +
  title("Heavier trade, darker stroke")
```

On `line`, `opacity` is only a setting, because a line is one stroke with one
value. An edge layer is many strokes, and each row carries its own.

The same is true of `color`. Each stroke takes the color of its exporter, so
Germany's five strokes share one color:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + color(exporter) +
  point * layout(exporter, importer) +
  network() +
  title("Each stroke, colored by its exporter")
```

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("edge")
```

`style(size = )` sets the stroke width for the layer, `style(color = )` sets
one color when nothing maps, `style(opacity = )` sets one opacity the same
way, and `style(pattern = )` dashes the strokes.

## What it refuses

An `edge` without its `layout` has no endpoints to draw, and a position bound
to it would say nothing the layout has not already said:

```{r}
#| error: true
data(trade_partners) + edge + x(billions) + y(exporter) + network()
```

And a `layout` is read only in its space:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

The [Network](../network.qmd) chapter has the whole family: the nodes, the
names, the cube, and the rest of the refusals.
