# Edge

```{r}
#| include: false
source("../R/setup.R")
```

Which things in your table are connected to which? `edge` draws one stroke per
row of the table, between the two things that row names. The other strokes draw
between positions taken from columns. A `line` runs along its domain, a `path`
follows the table's order, an `interval` spans two values. An `edge` is
different, because the endpoints of an edge are names, not numbers, so no axis
can measure them. The picture an `edge` layer makes is the **network diagram**.

So an edge takes no `x` and no `y`. Its two endpoints come from a
[`layout`](../network.qmd), a transform that reads the two name columns and
computes a position for every name. Write the source column first and the
destination second.

`trade_partners` holds one row per trade relation: who exports, who imports,
and the value of the goods, in billions of dollars, rounded:

```{r}
#| echo: false
peek(trade_partners)
```

Name the table, then write three things: the mark, its layout, and the space
`network()`:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in
the network."*

An `edge` draws only inside `network()`. Every other space labels its axes with
numbers, and an edge's two endpoints are names. gog refuses the sentence rather
than drawing it.

## Color and opacity, one per row

Every stroke in the network above looks the same, so a trade of 500 billion
looks like a trade of 50. The table holds the value, and a stroke can show it
by how dark it is.
An edge is one row, so its color and its opacity can follow a column of the
table. `color` maps a category, and `opacity` maps a number such as
`billions`. A second layer, `point * layout(exporter, importer)`, reads the
same layout and draws a dot at each name:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) +
  network() +
  title("Heavier trade, darker stroke")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, in the network."*

On `line`, `opacity` is only a setting, because a line is one stroke with one
value. An edge layer is many strokes, and each row gives its stroke an opacity
of its own.

Which strokes start at the same country?
`color` works the same way, one value per row. Each stroke takes the color of
its exporter, so Germany's five strokes share one color:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + color(exporter) +
  point * layout(exporter, importer) +
  network() +
  title("Each stroke, colored by its exporter")
```

*"Given trade partners: edges derived by layout from exporter to importer, color
by exporter, and also points derived by layout from exporter to importer, in the
network."*

## What you can set

Whether the strokes are thin or thick, solid or dashed, is not in any column of
the table. A setting decides that without reading one, and each mark takes its
own. These are an edge's, generated from the engine's own legality table, so
this page cannot differ from what `style()` accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("edge")
```

`style(size = )` sets the stroke width for the whole layer. `style(color = )`
and `style(opacity = )` give every stroke one value, for the case where no
column maps. `style(pattern = )` chooses the dash: `solid`, `dashed` or
`dotted`.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The other marks take their positions from `x` and `y`, so writing them on an
edge is the natural first attempt.
An `edge` with no `layout` has no endpoints to draw. `x` and `y` cannot supply
them, so gog refuses each binding, then refuses the empty `network()`:

```{r}
#| error: true
data(trade_partners) + edge + x(billions) + y(exporter) + network()
```

A `layout` is read only inside `network()`, so the space cannot be left out:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

The [Network](../network.qmd) chapter draws the rest: a dot for every name, a
label beside each dot, the same graph in a cube, and the other refusals.