# Choosing a mark

```{r}
#| include: false
source("../R/setup.R")
```

Which visible form should your data take? A **mark** is
that visible form: a dot, a bar, a line. Choosing the mark is the first
decision in every plot. The answer depends on the *types of your variables*
and the *question you are asking*.

There are fourteen marks, and they are not a list to learn before you draw.
This chapter is a guide to the choice: it shows which mark answers which
question, draws the common cases, and stops. Each mark then has a chapter of
its own, with examples. The tables in [Quick reference](#quick-reference) are
for later, when you have data in front of you and want the row that fits.

If you have used ggplot2, a mark is similar to a geom, but the two differ in
one way. Many geoms compute a statistic before they draw.
`geom_histogram()` bins your data and then draws bars. `geom_smooth()` fits a
curve and then draws a line. A mark never does that. A mark is only the shape,
and the statistic is a separate word: a histogram is `bar * bin`, a trend line
is `line * smooth`. A geom tends to name a chart. A mark names a shape.

Each column of your data is one variable, and gog reads its type from the data.
There are two types:

- **Continuous.** A column that holds numbers and has no declared order (`gdp`,
  `temperature`, `score`). A date or date-time column is continuous too, and it
  gets a calendar axis (see [Scales](../scales.qmd#time-a-date-column-gets-a-calendar-axis)).
- **Categorical.** A column that holds text (`continent`, `country`, `species`), or a
  column with a declared order.

---

## One variable

With one variable there is one question: how are its values spread out? The
answer depends on whether the column is continuous or categorical.

### Continuous variable

Are the life expectancies of 142 countries close to one value, or spread
across a wide range? You can cut a continuous column into bands, or estimate a
smooth curve over it. `bin` cuts the column into bands and counts
the rows in each. `bar * bin` draws one bar per band, which is the **histogram**:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
# How is life expectancy distributed? → histogram
data(gapminder_2007) + bar * bin + x(life) +
  x_label("Life expectancy (years)") + title("Distribution: bar * bin")
```

*"Given gapminder 2007: bars derived by bin, x is life."*

When sharp changes in height from one bar to the next hide the shape, a curve
shows the same spread without them. `density` estimates a smooth curve over the same column,
and `line * density` draws that curve:

```{r}
# Smooth shape of the distribution → density curve
data(gapminder_2007) + line * density + x(life) +
  x_label("Life expectancy (years)") + title("Density: line * density")
```

*"Given gapminder 2007: a line derived by density, x is life."*

### Categorical variable

A categorical column cannot be cut into bands, so the same question becomes:
how often does each category occur? `count` tallies the rows in each category, and
`bar * count` draws one bar per category:

```{r}
# How many countries per continent? → frequency bar
data(gapminder_2007) + bar * count + x(continent) +
  title("Frequency: bar * count")
```

*"Given gapminder 2007: bars derived by count, x is continent."*

A count does not say what part of the whole each continent is. Africa holds 52
of the 142 countries, which is more than a third of them. `proportion` is the
same tally as a share of the whole, so the bars sum to 1:

```{r}
# What share of countries per continent? → proportion bar
data(gapminder_2007) + bar * proportion + x(continent) +
  title("Relative frequency: bar * proportion")
```

*"Given gapminder 2007: bars derived by proportion, x is continent."*

---

## Two variables

With two variables, the pair of column types decides the mark.

### Continuous × Continuous

Both variables are continuous, so the questions are about relationships,
trends and patterns. Do people in richer countries live longer? `point` places one glyph
per row, and the shape those glyphs make is the first answer:

```{r}
# Is there a relationship? → scatter
data(gapminder_2007) + point + x(gdp) + y(life) + x_label("GDP per capita") +
  y_label("Life expectancy") + title("Relationship: point")
```

*"Given gapminder 2007: points, x is gdp, y is life."*

The direction of the cloud can matter more than any one country in it.
`smooth` fits a trend line through the points, and `line * smooth` draws it
over the scatter. `color(continent)` is written right after `point`, so it
colors only the points, and one line is fitted to all of them:

```{r}
# Smooth the relationship → fitted trend line
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Trend: point + line * smooth")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

A question about two continuous columns is not always about a relationship.
When one column is time and the other an amount, the question can be how much
there was. The `actuals` table records yearly sales:

```{r}
#| echo: false
peek(actuals)
```

```{r}
# How much, not just where → filled area
data(actuals) + area + x(year) + y(sales) + y_label("Sales") +
  title("Quantity: area")
```

*"Given the actuals: an area, x is year, y is sales."*

`line` and `area` draw the same boundary and give it different meanings. A
line says *where* the value was; an area says *how much* there was, so it
fills to zero and the filled region is the quantity. Choose `area` when zero
is a true baseline for your variable, and `line` when it is not.

Sometimes each x has a *spread* to show, not one value. [`ribbon`](ribbon.qmd)
fills between a low boundary and a high one, and a `range` transform supplies
that pair: the minimum and maximum at each x. Name two quantiles and `range`
uses those instead. Add a `line * mean` over the ribbon, and the line of means
is drawn inside the band.

When the *order of the rows* is what you want to show, [`path`](path.qmd)
connects them in that order. `line` sorts its points by x before it joins them.
`path` does not, so the stroke can move left as well as right, cross itself, or
return to where it started. That is the **connected scatterplot**: the two axes
carry measurements, and the joining line shows the order in time, here for
five Asian countries.

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
# The reading order is the table's → path
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Route: path (the arrow points to 2007)")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
end."*

Layered *over* any of the marks above, [`rule`](rule.qmd) draws a straight
stroke at one value on one axis, across the whole of the other: a threshold
line, or a rug. A rug is a short tick at the edge of the panel for each
observation, so it shows where the observations are. `style(reach = "edge")`
shortens a rule into a rug.

How many countries have a life expectancy below 60, 70 or 80 years, and where
are they on the income axis? `life_bands` holds three threshold values of life
expectancy, each with a name:

```{r}
#| echo: false
peek(life_bands)
```

```{r}
# One position, the panel supplies the other → rule
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  rule + x(gdp) + style(reach = "edge") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Threshold and rug: rule")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules, x is gdp,
with reach edge, and also rules from the life bands, color by band."*

The rule names `x(gdp)` even though the points already named it. That second
`x(gdp)` does not add an axis. Every layer shares the plot's one x and one y.
A rule sits on one axis and spans the other, so it has to say which. This rule
reads the scatter's table, which holds both `gdp` and `life`, so either column
could place it. Without `x(gdp)`, gog refuses rather than guessing; the
[Rule](rule.qmd#what-it-refuses) chapter shows the refusal. The three rules
from `life_bands` need no position, because that table holds `life` and no
`gdp`, so only one axis can answer.

[`text`](text.qmd) is also layered over other marks: it writes a label from a
column at each position where `point` would draw a dot. Layered *under* the
other marks, [`zone`](zone.qmd) shades a rectangle. On an axis where you give
it a pair of columns, the rectangle starts at the value in one column and ends
at the value in the other. On an axis where you give it no pair, the rectangle
spans the whole panel. Write `zone * bin` instead, and the same mark covers the
panel with a grid of cells and counts the rows in each: the
[heatmap](zone.qmd#cells-cut-from-the-data-the-heatmap).

### Categorical × Continuous

One axis groups, the other measures. The natural question is *how the
measurement differs across groups*. `mean` reduces each group to one value, its
average. `bar * mean` draws one bar per group, and the bar's height is that
value:

```{r}
# Mean value per group → bar
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("Group summary: bar * mean")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

A mean hides how far apart the countries inside one continent are. Sometimes
that spread is what you want to see. Drop the transform, and `point` draws every
country in its continent's slot:

```{r}
# All individual points per group → strip plot
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("Strip plot: point (one dot per country)")
```

*"Given gapminder 2007: points, x is continent, y is life."*

Nothing is summarized: each dot is one country, placed by its continent and its
life expectancy. Dots for countries with the same life expectancy overlap,
so a crowded slot hides two things: how many dots it holds, and how they are
spread.

There are three ways to draw a crowded group more clearly, and each shows
something different. For the *shape* of each group rather than every point,
write `box + x(continent) + y(life)`. [`box`](box.qmd) draws a box between the
quartiles, a line at the median, whiskers, and outliers as separate dots.
[`ribbon * density`](ribbon.qmd#the-violin-a-distribution-per-category) draws
the whole estimated distribution that the box summarizes: the **violin plot**.
When every point matters but the points overlap,
[`point * jitter`](../transforms.qmd#jitter-spread-sideways-along-the-category)
spreads them sideways: the **jittered strip plot**.

The question is not always which group is highest. It can be how the value
changes from one group to the next. `line`, `area`, `step` and `ribbon` accept a
category on `x` too. They join one summary per group instead of drawing a bar on
each, and the result is a **profile**:

```{r}
# One value per group, joined → profile
data(gapminder_2007) + x(continent) + y(life) +
  line * mean + point * mean +
  y_label("Mean life expectancy") +
  title("Profile: line * mean")
```

*"Given gapminder 2007: x is continent, y is life, a line derived by mean and
also points derived by mean."*

A bar compares one group with another; a profile traces a change across the
groups. A profile needs more care than a bar, because its segments cross parts
of the panel where nothing was measured. Use it when the categories have an
order, such as a rating scale or size groups from smallest to largest.
Continents have no order, so that advice says not to draw this plot. gog draws
it anyway. The grammar refuses a sentence that is ill-formed, never one that is
only unwise; whether a profile suits your categories is your judgment
([Law 8](../design-laws.qmd)).

`line` is not the only mark that draws a profile. [`area`](area.qmd) fills
under the same boundary, [`step`](step.qmd) holds it flat across each slot, and
[`ribbon * range`](ribbon.qmd) fills the spread as a band. Drawn around a
circle, the same `line * mean` profile is the **radar chart**, in
[Polar](../polar.qmd).

---

## Quick reference

The table below holds the common cases, one question per row. Start from the
types of the columns you put on x and y, and find the rows that match. Each row
names the mark and transform that answer its question. This chapter drew plots
for the most common rows, and each mark's own chapter covers the rest. A `—`
means that no column of yours goes on that position.

| x | y | Mark + transform | Question |
|---|---|---|---|
| Continuous | — | [`bar * bin`](bar.qmd#histogram) | Distribution? |
| Continuous | — | [`line * density`](../transforms.qmd#density-kernel-density-estimate) | Smooth distribution? |
| Continuous | — | [`point * bin * stack`](../transforms.qmd#the-dot-plot) | Distribution, every observation shown? |
| Categorical | — | [`bar * count`](../transforms.qmd#count-frequency-aggregation) | Frequency? |
| Categorical | — | [`bar * proportion`](../transforms.qmd#proportion-relative-frequency) | Share? |
| Continuous | Continuous | [`point`](point.qmd#basic-usage) | Relationship? |
| Continuous | Continuous | [`line`](line.qmd#basic-usage) | Trend / time series? |
| Continuous | Continuous | [`line * smooth`](line.qmd#smoothed-trend-line) | Smoothed trend? |
| Continuous | Continuous | [`area`](area.qmd#basic-usage) | How much, over a range? |
| Continuous | Continuous | [`ribbon * range`](ribbon.qmd#a-band-from-a-range) | Spread / band over a range? |
| Continuous | Continuous | [`path`](path.qmd#the-connected-scatterplot) | The order of the rows? |
| Categorical | Continuous | [`bar`](bar.qmd#basic-usage) | One value per group, already in the table? |
| Categorical | Continuous | [`bar * mean/sum/…`](../transforms.qmd#aggregation-family-sum-mean-median-max-min-quantile) | Value per group? |
| Categorical | Continuous | [`point`](point.qmd#a-categorical-x-the-strip-plot) | All values per group? |
| Categorical | Continuous | [`box`](box.qmd#a-summary-per-group) | Distribution per group, summarized? |
| Categorical | Continuous | [`ribbon * density`](ribbon.qmd#the-violin-a-distribution-per-category) | Distribution per group, in full? |
| Categorical | Continuous | [`interval * range`](interval.qmd#a-range-per-group) | Spread per group? |
| Categorical | Continuous | [`line * mean`](line.qmd#a-category-on-the-domain-the-profile) | Value per group, as a profile? |
| Categorical | Continuous | [`area * mean`](area.qmd#with-transforms) | Value per group, filled? |
| Categorical | Continuous | [`line * mean` + `polar()`](../polar.qmd#the-radar-a-categorical-angle) | Value per group, as a radar? |

The table has no row for `edge`, because its question has no x and no y: what
is connected to what? [`edge`](edge.qmd) draws one stroke for each connection
in a table of connections, such as who trades with whom. Nothing in the data
says where the two ends of that stroke should sit, so a `layout` transform
inside the `network()` space places them. The [Network](../network.qmd) chapter builds the
whole diagram.

A third variable takes the third position, `z`, rather than a new mark. The
last row below is different. It has only two columns, both categorical, so
`count` supplies the third number: how many rows have each pair. `space()` gives
that count an axis of its own.

| x | y | z | Mark | Question |
|---|---|---|---|---|
| Continuous | Continuous | Continuous | [`point`](../space.qmd#z-one-more-vowel) | Relationship, in three dimensions? |
| Continuous | Continuous | Continuous | [`path`](../space.qmd#path-the-route-through-the-cube) | A route through three dimensions? |
| Continuous | Continuous | Continuous | [`surface`](surface.qmd) | A height over a plane? (needs a **grid**: a z value at every x and y) |
| Categorical | Categorical | — | [`bar * count` + `space()`](../space.qmd#transforms-in-the-cube) | How many, per pair? |

### Swapping `x` and `y`

Long category names, such as country names, are easier to read on the vertical
axis than along the bottom. In the first table, each Categorical / Continuous
row whose mark sits in a slot can be drawn that way, with the two axes swapped.
`bar`, `box`, `interval` and `point` sit in a slot on one axis and measure along
the other, and so does the violin, `ribbon * density`. The column types decide
which axis holds the slots: the one with the categorical column. Swapping the
two positions therefore draws the same plot horizontally.

`box + x(life) + y(continent)` is the horizontal box plot, and
`bar + x(gold) + y(country)`, on the medals table, is the horizontal bar.
Nothing else is needed. If you know ggplot2's `coord_flip()` or matplotlib's
`barh()`, this is the same result with no extra word. gog reads the
orientation from which axis carries the categories, as ggplot2 does.

Only the marks with a slot can be drawn horizontally this way. For `line`,
`step` and `area`, `x` is always the **domain**, the axis the mark is read
along, and `y` is the measure. The same is true of a `ribbon` drawn as a band
rather than as a violin. Put the category on `y` for any of them, and gog
refuses the sentence. Their two columns are often both numbers, so no column
type could say which one is the domain, and the mark decides it instead. The
[`line`](line.qmd#a-category-on-the-domain-the-profile) chapter shows what a
swap of two continuous columns draws, and what to write when the domain must be
the vertical axis.

---

## Mark families

The mark list looks like a set to memorize, one entry at a time. It is not.
The fourteen consonants of Hangeul (한글) are not fourteen unrelated shapes
either. The letters are *featural*: most of them are another letter with a
single feature changed. ㅋ (*k*) is ㄱ (*g*) with one stroke added, so learning
ㄱ teaches you most of ㅋ.

The marks are built the same way. Some pairs have the same rules for every
channel: `x`, `y`, `color`, `size` and the rest. If one mark requires, accepts
or refuses a channel, the other mark does the same. They differ in the geometry
they draw:

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules`, the same dump the grids in Combinations read.
# The *membership* of each family is computed, never typed: if a mark's channel
# row changes it drops out of its family here on the next build, rather than this
# page going quietly on claiming a parity the engine no longer has.
rules <- .gog_rules()
marks <- rules$marks$name[rules$marks$drawable]
cells <- rules$cells

# `base::order` on purpose: gog exports an `order` atom, which masks the base
# function everywhere in the book. Calling the bare name here returns a spec
# object and the chunk dies.
row_signature <- function(m) {
  r <- cells[cells$mark == m, , drop = FALSE]
  r <- r[base::order(r$channel), , drop = FALSE]
  paste(r$channel, r$obligation, r$accepts, collapse = " | ")
}

families <- split(marks, vapply(marks, row_signature, character(1)))
families <- families[vapply(families, length, integer(1)) > 1]

# The geometric difference is prose about what each mark draws, so it is written
# here rather than derived. It is keyed on the *generated* membership, so a mark
# joining a family it has no sentence for prints a visible gap instead of being
# silently absorbed.
differs <- c(
  "line, step"    = "`line` **slopes** from one value to the next; `step` **holds** each value until it changes",
  "area, ribbon"  = "`area` closes its region on the **baseline**; `ribbon` closes it on a **second data boundary**",
  "box, interval" = "`box` computes its **own summary** of each group; `interval` spans **two ends a transform supplies**"
)

cat("| Family | Same rules for every channel | They differ in |\n")
cat("|---|---|---|\n")
for (f in families) {
  key <- paste(sort(f), collapse = ", ")
  cat(sprintf("| `%s` | ✓ | %s |\n",
              paste(sort(f), collapse = "` and `"),
              if (key %in% names(differs)) differs[[key]]
              else "**(no description yet for this family)**"))
}
```

The second member of a pair is quick to learn. It follows the same rules for
every channel as its sibling, so you already know those. The grid of marks and
channels in [Combinations](../combinations.qmd) shows this: the two marks of a
family have identical rows there. What is left is what the second mark draws.
`ribbon` and `interval` also need a transform, such as `range`, to supply their
two ends.

When does gog add a mark? Only for a shape that no existing mark can draw.
`step` was added because "hold the value until it changes" is such a shape: no
channel, transform or setting can say it. A thick line was not added as a mark,
because `style(size = )` already says it. A new chart is never the reason. A
histogram, a radar and a pyramid are arrangements of marks that already exist.
A new word is spelled with the letters you have, and a new chart is written with
the marks you have. The families make the marks quick to learn, but gog never
adds a mark only to complete a pair.

---

## Unpronounceable combinations

Just as some letter combinations cannot be pronounced, some pairs of a transform
and a column type have no meaning. gog refuses each one and draws nothing.

`bin`, `density` and `smooth` all work along a continuous axis, so the column on
that axis must be continuous. A category on that axis is refused, and each
refusal names the atom that does the same job for categories:

| Combination | Problem | Use instead |
|---|---|---|
| `bar * bin + x(category)` | `bin` cuts a *continuous* axis into bands | `bar * count + x(category)` |
| `bar * density + x(category)` | `density` estimates a *continuous* distribution | `bar * proportion + x(category)` |
| `point * smooth + x(category)` | `smooth` fits a curve along a *continuous* axis | `bar * mean + x(category) + y(value)` |

All three appear as live refusals in [What `bin`, `density` and `smooth`
refuse](../transforms.qmd#what-bin-density-and-smooth-refuse); the table
above is this page's summary of them.

## What comes next

Choosing the mark is the first decision in a plot, not the last one.
[Point](point.qmd) opens the fourteen chapters that follow, each describing one
mark: its channels, its transforms, its settings and its refusals.
