# Choosing a mark

```{r}
#| include: false
source("../R/setup.R")
```

Which visible form should your data take? A **mark** is
that visible form: a dot, a bar, a line. Choosing the mark is the first
decision in every plot. The answer depends on the *types of your variables*
and the *question you are asking*.

If you have used ggplot2, a mark is close to a geom, and one difference is
worth knowing early. Many geoms compute a statistic before they draw.
`geom_histogram()` bins your data and then draws bars. `geom_smooth()` fits a
curve and then draws a line. A mark never does that. A mark is only the shape,
and the statistic is a separate word: a histogram is `bar * bin`, a trend line
is `line * smooth`. A geom tends to name a chart. A mark names a shape, and
there are fourteen of them.

gog reads the type of each column from your data:

- **Continuous.** Any numeric column (`gdp`, `temperature`, `score`).
- **Categorical.** Any text column (`continent`, `country`, `species`), or a
  column with a declared order.
- **Date/time.** Any date or date-time column; the axis reads as a calendar
  (see [Scales](../scales.qmd)).

---

## One variable

With one variable there is one question: how are its values spread out? The
answer depends on whether the column is continuous or categorical.

### Continuous variable

A continuous column has a number line, so you can cut it into intervals or
estimate a smooth curve over it.

```{r}
# How is life expectancy distributed? → histogram
data(gapminder_2007) + bar * bin + x(life) +
  x_label("Life expectancy (years)") + title("Distribution: bar * bin")
```

*"Given gapminder 2007: bars derived by bin, x is life."*

```{r}
# Smooth shape of the distribution → density curve
data(gapminder_2007) + line * density + x(life) +
  x_label("Life expectancy (years)") + title("Density: line * density")
```

*"Given gapminder 2007: a line derived by density, x is life."*

### Categorical variable

A categorical column has no number line, so the question is how often each
category occurs.

```{r}
# How many countries per continent? → frequency bar
data(gapminder_2007) + bar * count + x(continent) +
  title("Frequency: bar * count")
```

*"Given gapminder 2007: bars derived by count, x is continent."*

```{r}
# What share of countries per continent? → proportion bar
data(gapminder_2007) + bar * proportion + x(continent) +
  title("Relative frequency: bar * proportion")
```

*"Given gapminder 2007: bars derived by proportion, x is continent."*

---

## Two variables

With two variables, the pair of column types decides the mark.

### Continuous × Continuous

Both variables are numeric. The natural questions are about relationships,
trends, and patterns.

```{r}
# Is there a relationship? → scatter
data(gapminder_2007) + point + x(gdp) + y(life) + x_label("GDP per capita") +
  y_label("Life expectancy") + title("Relationship: point")
```

*"Given gapminder 2007: points, x is gdp, y is life."*

```{r}
# Smooth the relationship → fitted trend line
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Trend: point + line * smooth")
```

*"Given gapminder 2007: x is gdp, y is life, points, color by continent, and
also a line derived by smooth."*

```{r}
# How much, not just where → filled area
data(actuals) + area + x(year) + y(sales) + y_label("Sales") +
  title("Quantity: area")
```

*"Given the actuals: an area, x is year, y is sales."*

`line` and `area` draw the same boundary and give it different meanings. A
line says *where* the value was; an area says *how much* there was, so it
fills to zero and the filled region is the quantity. Choose `area` when zero
is a true baseline for your variable, and `line` when it is not.

[`ribbon`](ribbon.qmd) fills between a low boundary and a high one, and a
`range` transform supplies the pair. Use it when each x has a *spread* to
show, such as a lowest-to-highest range or a confidence band. Add a
`line * mean` on top and you get a trend line inside its band.

When the *order of the rows* is what you want to show, [`path`](path.qmd)
connects them in that order. It does not sort by x, so the stroke can double
back or return to its starting point. That is the **connected scatterplot**:
the two axes carry measurements, and the joining line carries time.

```{r}
# The reading order is the table's → path
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Route: path (the arrow points to 2007)")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
end."*

Layered *over* any of these, [`rule`](rule.qmd) marks a value on one axis and
spans the other: a threshold line, or a rug. A rug reaches only a little way
in from the edge and shows where the observations are. Layered *under* them,
[`zone`](zone.qmd) shades a rectangle the same way: bounded where you give it a
pair of columns, spanning the panel where you do not. Give `zone` a `bin`
instead of the pair of columns, and the same mark tiles the panel with
counted cells: the heatmap.

`gdp_rug` holds one column, one row per country, and `life_bands` holds three
named bands of life expectancy:

```{r}
#| echo: false
peek(gdp_rug)
```

```{r}
#| echo: false
peek(life_bands)
```

```{r}
# One position, the panel supplies the other → rule
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  data(gdp_rug) + rule + style(reach = "edge") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Threshold and rug: rule")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the gdp
rug, with reach edge, and also rules from the life bands, color by band."*

### Categorical × Continuous

One axis groups, the other measures. The natural question is
*how the measurement differs across groups*.

```{r}
# Mean value per group → bar
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("Group summary: bar * mean")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

```{r}
# All individual points per group → strip plot
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("Strip plot: point (one dot per country)")
```

For the *shape* of each group rather than every point, [`box`](box.qmd) draws
the five-number summary as `box + x(continent) + y(life)`: median, quartiles,
and whiskers, with far values drawn as separate dots. [`ribbon *
density`](ribbon.qmd#the-violin-a-distribution-per-category) draws the whole
estimated distribution that those five numbers summarize: the violin plot.
When every point matters but the points overlap,
[`point * jitter`](../transforms.qmd#jitter-spread-sideways-along-the-category)
spreads them sideways: the jittered strip plot.

`line`, `area`, `step` and `ribbon` take a category here too, joining one
summary per group instead of drawing a bar on each:

```{r}
# One value per group, joined → profile
data(gapminder_2007) + x(continent) + y(life) +
  line * mean + point * mean +
  y_label("Mean life expectancy") +
  title("Profile: line * mean")
```

A bar compares one group with another; a profile traces a change across the
groups. The profile is the weaker claim of the two, since its segments cross
parts of the panel where nothing was measured. Use it when the categories have
an order, such as a rating scale or size groups from smallest to largest. The
same categories drawn around a circle give the radar chart,
in [Polar](../polar.qmd). [`area`](area.qmd) fills the same boundary,
[`step`](step.qmd) holds it flat across each slot, and [`ribbon *
range`](ribbon.qmd) fills the spread as a band.

---

## Quick reference

A `—` means the question needs no variable in that column.

| x | y | Mark + transform | Question |
|---|---|---|---|
| Continuous | — | `bar * bin` | Distribution? |
| Continuous | — | `line * density` | Smooth distribution? |
| Continuous | — | `point * bin * stack` | Distribution, every observation shown? |
| Categorical | — | `bar * count` | Frequency? |
| Categorical | — | `bar * proportion` | Share? |
| Continuous | Continuous | `point` | Relationship? |
| Continuous | Continuous | `line` | Trend / time series? |
| Continuous | Continuous | `line * smooth` | Smoothed trend? |
| Continuous | Continuous | `area` | How much, over a range? |
| Continuous | Continuous | `ribbon * range` | Spread / band over a range? |
| Categorical | Continuous | `bar * mean/sum/…` | Value per group? |
| Categorical | Continuous | `point` | All values per group? |
| Categorical | Continuous | `box` | Distribution per group, summarized? |
| Categorical | Continuous | `ribbon * density` | Distribution per group, in full? |
| Categorical | Continuous | `interval * range` | Spread per group? |
| Categorical | Continuous | `line * mean` | Value per group, as a profile? |
| Categorical | Continuous | `area * mean` | Value per group, filled? |
| Categorical | Continuous | `line * mean` + `polar()` | Value per group, as a radar? |

One mark answers a question that x and y cannot answer: what is connected to
what? [`edge`](edge.qmd) draws one relation per row. Its two endpoints are
placed by `layout` inside `network()`, and the [Network](../network.qmd)
chapter builds the whole diagram.

A third variable takes the third position, `z`, rather than a new mark. The
last row below is different: it uses two categories, and `space()` gives the
counts an axis of their own.

| x | y | z | Mark | Question |
|---|---|---|---|---|
| Continuous | Continuous | Continuous | `point` | Relationship, in three? |
| Continuous | Continuous | Continuous | `path` | A route through three? |
| Continuous | Continuous | Continuous | [`surface`](surface.qmd) | A height over a plane? (needs a **grid**: a z value at every x and y) |
| Categorical | Categorical | — | `bar * count` + `space()` | How many, per pair? |

Every Categorical / Continuous row above can also be read with the two axes
swapped. `bar`, `box` and `interval` sit in a slot on one axis and measure
along the other. The column types decide which is which: the categorical
column gets the slots. Swapping the two positions therefore draws the same
plot horizontally.
`box + x(life) + y(continent)` is the horizontal box plot, and
`bar + x(gold) + y(country)` the horizontal bar. Nothing else is needed,
because there is no `flip` atom to learn.

---

## Mark families

The mark list looks like a set to memorize, one entry at a time. It is not.
The fourteen consonants of Hangeul (한글) are not fourteen unrelated shapes
either. The letters are *featural*: most of them are another letter with a
single feature changed. ㅋ (*k*) is ㄱ (*g*) with one stroke added, so learning
ㄱ teaches you most of ㅋ.

The marks are built the same way. Some pairs share an entire *channel row*:
every channel one requires, accepts or refuses, the other does too. They
differ only in the geometry they draw:

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules`, the same dump the grids in Combinations read.
# The *membership* of each family is computed, never typed: if a mark's channel
# row changes it drops out of its family here on the next build, rather than this
# page going quietly on claiming a parity the engine no longer has.
rules <- .gog_rules()
marks <- rules$marks$name[rules$marks$drawable]
cells <- rules$cells

# `base::order` on purpose: gog exports an `order` atom, which masks the base
# function everywhere in the book. Calling the bare name here returns a spec
# object and the chunk dies.
row_signature <- function(m) {
  r <- cells[cells$mark == m, , drop = FALSE]
  r <- r[base::order(r$channel), , drop = FALSE]
  paste(r$channel, r$obligation, r$accepts, collapse = " | ")
}

families <- split(marks, vapply(marks, row_signature, character(1)))
families <- families[vapply(families, length, integer(1)) > 1]

# The geometric difference is prose about what each mark draws, so it is written
# here rather than derived. It is keyed on the *generated* membership, so a mark
# joining a family it has no sentence for prints a visible gap instead of being
# silently absorbed.
differs <- c(
  "line, step"    = "a stroke that **slopes** between values, against one that **holds** each value until it changes",
  "area, ribbon"  = "a region closed on the **baseline**, against one closed on a **second data boundary**",
  "box, interval" = "a body that computes its **own five-number summary**, against a span whose **ends you supply**"
)

cat("| Family | Identical channel row | They differ only in |\n")
cat("|---|---|---|\n")
for (f in families) {
  key <- paste(sort(f), collapse = ", ")
  cat(sprintf("| `%s` | ✓ | %s |\n",
              paste(sort(f), collapse = "` and `"),
              if (key %in% names(differs)) differs[[key]]
              else "**(no description yet for this family)**"))
}
```

The second member of a pair takes almost nothing to learn. It asks exactly the
questions its sibling asks, so you already know all of them. One sentence about
what it draws is all that is left. The two marks of a family also sit near each
other on the grid in [Combinations](../combinations.qmd), which is where the
structure you can see on that page comes from.

This is a fact about *learning* the set, never a reason to grow it. A mark that
is another mark with one feature changed belongs only when that feature
is a geometry nothing else can express. `step` qualifies, because "hold the
value until it changes" is not something a channel, a transform or a setting can
say. A thick line does not, because `style(size = )` already says it.

Marks are building blocks, and this rule follows from that. You do not get a
new block because you want to build a new model. Almost every model is a new
arrangement of the blocks you already have. A new block is made only when no
existing block has the shape you need. A mark is added for a new geometry,
never for a new chart.

---

## Unpronounceable combinations

Just as some letter combinations cannot be pronounced, some mark + variable type
combinations cannot be rendered meaningfully. Each is refused outright, and
nothing is drawn: if a plot was drawn, the grammar accepted it.

`bin`, `density` and `smooth` all work along a number line, so the axis they
measure must carry numbers. A category on that axis is refused, and each
refusal names the atom that asks the same question of categories:

| Combination | Problem | Use instead |
|---|---|---|
| `bar * bin + x(category)` | `bin` cuts a **continuous** axis into intervals | `bar * count + x(category)` |
| `bar * density + x(category)` | `density` estimates a **continuous** distribution | `bar * proportion + x(category)` |
| `point * smooth + x(category)` | `smooth` fits a curve along a **number line** | `bar * mean + x(category) + y(value)` |

All three are rendered as live refusals in [What `bin`, `density` and `smooth`
refuse](../transforms.qmd#what-bin-density-and-smooth-refuse); the table
above is this page's summary of them.

The next fourteen chapters give each mark its own page, starting with
`point`.
