# Choosing a mark

```{r}
#| include: false
source("../R/setup.R")
```

Which visible form should your data take? A **mark** is
that visible form: a dot, a bar, a line. Choosing the mark is the first
decision in every plot. The answer depends on the *types of your variables*
and the *question you are asking*.

There are fourteen marks, and they are not a list to learn before you draw.
This chapter is a guide to the choice: it shows which mark answers which
question, draws the common cases, and stops. Each mark then has a chapter of
its own, with examples. The table at the end is for later, when you have data
in front of you and want the row that fits.

If you have used ggplot2, a mark is close to a geom, and one difference is
worth knowing early. Many geoms compute a statistic before they draw.
`geom_histogram()` bins your data and then draws bars. `geom_smooth()` fits a
curve and then draws a line. A mark never does that. A mark is only the shape,
and the statistic is a separate word: a histogram is `bar * bin`, a trend line
is `line * smooth`. A geom tends to name a chart. A mark names a shape.

gog reads the type of each column from your data:

- **Continuous.** A column that holds numbers (`gdp`, `temperature`, `score`).
- **Categorical.** A column that holds text (`continent`, `country`, `species`), or a
  column with a declared order.
- **Date/time.** Any date or date-time column; the axis reads as a calendar
  (see [Scales](../scales.qmd)).

---

## One variable

With one variable there is one question: how are its values spread out? The
answer depends on whether the column is continuous or categorical.

### Continuous variable

A continuous column can be cut into bands, or a smooth curve can be estimated
over it. `bin` cuts the column into bands and counts the rows in each. `bar *
bin` draws one bar per band, which is the histogram:

```{r}
# How is life expectancy distributed? → histogram
data(gapminder_2007) + bar * bin + x(life) +
  x_label("Life expectancy (years)") + title("Distribution: bar * bin")
```

*"Given gapminder 2007: bars derived by bin, x is life."*

`density` estimates a smooth curve over the same column, and `line * density`
draws that curve:

```{r}
# Smooth shape of the distribution → density curve
data(gapminder_2007) + line * density + x(life) +
  x_label("Life expectancy (years)") + title("Density: line * density")
```

*"Given gapminder 2007: a line derived by density, x is life."*

### Categorical variable

A categorical column cannot be cut into bands, so the same question reads as
how often each category occurs. `count` tallies the rows in each category, and
`bar * count` draws one bar per category:

```{r}
# How many countries per continent? → frequency bar
data(gapminder_2007) + bar * count + x(continent) +
  title("Frequency: bar * count")
```

*"Given gapminder 2007: bars derived by count, x is continent."*

`proportion` is the same tally as a share of the whole, so the bars sum to 1:

```{r}
# What share of countries per continent? → proportion bar
data(gapminder_2007) + bar * proportion + x(continent) +
  title("Relative frequency: bar * proportion")
```

*"Given gapminder 2007: bars derived by proportion, x is continent."*

---

## Two variables

With two variables, the pair of column types decides the mark.

### Continuous × Continuous

Both variables are numeric. The natural questions are about relationships,
trends, and patterns.

```{r}
# Is there a relationship? → scatter
data(gapminder_2007) + point + x(gdp) + y(life) + x_label("GDP per capita") +
  y_label("Life expectancy") + title("Relationship: point")
```

*"Given gapminder 2007: points, x is gdp, y is life."*

`smooth` fits a trend line through the points, and `line * smooth` draws it
over the scatter. The points are colored by continent here, and the one line
runs through all of them: 

```{r}
# Smooth the relationship → fitted trend line
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Trend: point + line * smooth")
```

*"Given gapminder 2007: x is gdp, y is life, points, color by continent, and
also a line derived by smooth."*

```{r}
# How much, not just where → filled area
data(actuals) + area + x(year) + y(sales) + y_label("Sales") +
  title("Quantity: area")
```

*"Given the actuals: an area, x is year, y is sales."*

`line` and `area` draw the same boundary and give it different meanings. A
line says *where* the value was; an area says *how much* there was, so it
fills to zero and the filled region is the quantity. Choose `area` when zero
is a true baseline for your variable, and `line` when it is not.

[`ribbon`](ribbon.qmd) fills between a low boundary and a high one. A `range`
transform supplies that pair: the minimum and maximum at each x, unless you
name two quantiles. Use `ribbon` when each x has a *spread* to show. Add a
`line * mean` on top and you get a trend line inside its band.

When the *order of the rows* is what you want to show, [`path`](path.qmd)
connects them in that order. `line` sorts its points by x before it joins them.
`path` does not, so the stroke can move left as well as right, cross itself, or
return to where it started. That is the **connected scatterplot**: the two axes
carry measurements, and the joining line carries time.

```{r}
# The reading order is the table's → path
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Route: path (the arrow points to 2007)")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
end."*

Layered *over* any of these, [`rule`](rule.qmd) marks a value on one axis and
spans the other: a threshold line, or a rug. A rug reaches only a little way in
from the edge, which is `style(reach = "edge")`, and shows where the
observations are. Layered *under* them, [`zone`](zone.qmd) shades a rectangle
the same way: bounded where you give it a pair of columns, spanning the panel
where you do not. Give `zone` a `bin` instead of the pair of columns, and the
same mark tiles the panel with counted cells: the heatmap.

`life_bands` holds three named bands of life expectancy:

```{r}
#| echo: false
peek(life_bands)
```

```{r}
# One position, the panel supplies the other → rule
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  rule + x(gdp) + style(reach = "edge") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Threshold and rug: rule")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules, x is gdp,
with reach edge, and also rules from the life bands, color by band."*

The rule names `x(gdp)` even though the points already named it. That second
`x(gdp)` does not add an axis. Every layer shares the plot's one x and one y.
This rule reads the scatter's table, which holds both `gdp` and `life`, so both
axes answer. A rule sits on one axis and spans the other, so it has to say
which. Leave that out and gog refuses rather than guessing; the
[Rule](rule.qmd#what-it-refuses) chapter shows the refusal. The three rules
from `life_bands` need no position, because that table holds `life` and no
`gdp`, so only one axis can answer.

### Categorical × Continuous

One axis groups, the other measures. The natural question is *how the
measurement differs across groups*. `mean` reduces each group to one value, its
average. `bar * mean` draws one bar per group, and the bar's height is that
value:

```{r}
# Mean value per group → bar
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("Group summary: bar * mean")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

Drop the transform, and `point` draws every country in its continent's slot:

```{r}
# All individual points per group → strip plot
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("Strip plot: point (one dot per country)")
```

*"Given gapminder 2007: points, x is continent, y is life."*

Nothing is summarized: each dot is one country, placed by its continent and its
life expectancy. Countries with the same life expectancy land on the same spot,
so a crowded column hides two things: how many dots it holds, and how they are
spread. The next paragraph names three marks that show both.

For the *shape* of each group rather than every point, [`box`](box.qmd) draws
the five-number summary as `box + x(continent) + y(life)`: median, quartiles,
and whiskers, with far values drawn as separate dots. [`ribbon *
density`](ribbon.qmd#the-violin-a-distribution-per-category) draws the whole
estimated distribution that those five numbers summarize: the violin plot.
When every point matters but the points overlap,
[`point * jitter`](../transforms.qmd#jitter-spread-sideways-along-the-category)
spreads them sideways: the jittered strip plot.

`line`, `area`, `step` and `ribbon` take a category here too, joining one
summary per group instead of drawing a bar on each:

```{r}
# One value per group, joined → profile
data(gapminder_2007) + x(continent) + y(life) +
  line * mean + point * mean +
  y_label("Mean life expectancy") +
  title("Profile: line * mean")
```

*"Given gapminder 2007: x is continent, y is life, a line derived by mean and
also points derived by mean."*

A bar compares one group with another; a profile traces a change across the
groups. The profile is the weaker claim of the two, since its segments cross
parts of the panel where nothing was measured. Use it when the categories have
an order, such as a rating scale or size groups from smallest to largest.
Continents have no order, so this plot does what that advice says not to do,
and gog draws it anyway. The grammar refuses a sentence that is ill-formed,
never one that is only unwise; whether a profile suits your categories is your
call ([Law 8](../design-laws.qmd)). The same categories drawn around a circle
give the radar chart, in [Polar](../polar.qmd). [`area`](area.qmd) fills the same boundary,
[`step`](step.qmd) holds it flat across each slot, and [`ribbon *
range`](ribbon.qmd) fills the spread as a band.

---

## Quick reference

Everything above, in one table. Start from the types of your two columns, x and
y, find the rows that match, and read the mark and transform that answer each
question. The chapter drew the common rows; every mark has a chapter of its own
for the rest. A `—` means the question needs no variable in that column.

| x | y | Mark + transform | Question |
|---|---|---|---|
| Continuous | — | [`bar * bin`](bar.qmd#histogram) | Distribution? |
| Continuous | — | [`line * density`](../transforms.qmd#density-kernel-density-estimate) | Smooth distribution? |
| Continuous | — | [`point * bin * stack`](../transforms.qmd#the-dot-plot) | Distribution, every observation shown? |
| Categorical | — | [`bar * count`](../transforms.qmd#count-frequency-aggregation) | Frequency? |
| Categorical | — | [`bar * proportion`](../transforms.qmd#proportion-relative-frequency) | Share? |
| Continuous | Continuous | [`point`](point.qmd#basic-usage) | Relationship? |
| Continuous | Continuous | [`line`](line.qmd#basic-usage) | Trend / time series? |
| Continuous | Continuous | [`line * smooth`](line.qmd#smoothed-trend-line) | Smoothed trend? |
| Continuous | Continuous | [`area`](area.qmd#basic-usage) | How much, over a range? |
| Continuous | Continuous | [`ribbon * range`](ribbon.qmd#a-band-from-a-range) | Spread / band over a range? |
| Categorical | Continuous | [`bar`](bar.qmd#basic-usage) | One value per group, already in the table? |
| Categorical | Continuous | [`bar * mean/sum/…`](../transforms.qmd#aggregation-family-sum-mean-median-max-min-quantile) | Value per group? |
| Categorical | Continuous | [`point`](point.qmd#a-categorical-x-the-strip-plot) | All values per group? |
| Categorical | Continuous | [`box`](box.qmd#a-summary-per-group) | Distribution per group, summarized? |
| Categorical | Continuous | [`ribbon * density`](ribbon.qmd#the-violin-a-distribution-per-category) | Distribution per group, in full? |
| Categorical | Continuous | [`interval * range`](interval.qmd#a-range-per-group) | Spread per group? |
| Categorical | Continuous | [`line * mean`](line.qmd#a-category-on-the-domain-the-profile) | Value per group, as a profile? |
| Categorical | Continuous | [`area * mean`](area.qmd#with-transforms) | Value per group, filled? |
| Categorical | Continuous | [`line * mean` + `polar()`](../polar.qmd#the-radar-a-categorical-angle) | Value per group, as a radar? |

One mark is not in the table, because its question has no x and no y: what is
connected to what? [`edge`](edge.qmd) draws one line for each connection in a
table of connections, such as who trades with whom. Nothing in the data says
where the two ends of that line should sit, so a `layout` transform inside the
`network()` space places them. The [Network](../network.qmd) chapter builds the
whole diagram.

A third variable takes the third position, `z`, rather than a new mark. The
last row below is different: it uses two categories, and `space()` gives the
counts an axis of their own.

| x | y | z | Mark | Question |
|---|---|---|---|---|
| Continuous | Continuous | Continuous | [`point`](../space.qmd#z-one-more-vowel) | Relationship, in three? |
| Continuous | Continuous | Continuous | [`path`](../space.qmd#path-the-route-through-the-cube) | A route through three? |
| Continuous | Continuous | Continuous | [`surface`](surface.qmd) | A height over a plane? (needs a **grid**: a z value at every x and y) |
| Categorical | Categorical | — | [`bar * count` + `space()`](../space.qmd#transforms-in-the-cube) | How many, per pair? |

Every Categorical / Continuous row above can also be read with the two axes
swapped. `bar`, `box` and `interval` sit in a slot on one axis and measure
along the other. The column types decide which is which: the categorical
column gets the slots. Swapping the two positions therefore draws the same
plot horizontally.
`box + x(life) + y(continent)` is the horizontal box plot, and `bar + x(gold) +
y(country)` the horizontal bar. Nothing else is needed. If you know ggplot2's
`coord_flip()` or matplotlib's `barh()`, this is the same result with no extra
word. gog reads the orientation from which axis carries the categories, as
ggplot2 itself now does.

---

## Mark families

The mark list looks like a set to memorize, one entry at a time. It is not.
The fourteen consonants of Hangeul (한글) are not fourteen unrelated shapes
either. The letters are *featural*: most of them are another letter with a
single feature changed. ㅋ (*k*) is ㄱ (*g*) with one stroke added, so learning
ㄱ teaches you most of ㅋ.

The marks are built the same way. Some pairs have the same rules for every
channel: whatever one mark requires, accepts or refuses, for `x`, `y`, `color`,
`size` and the rest, the other does too. They differ only in the geometry they
draw:

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules`, the same dump the grids in Combinations read.
# The *membership* of each family is computed, never typed: if a mark's channel
# row changes it drops out of its family here on the next build, rather than this
# page going quietly on claiming a parity the engine no longer has.
rules <- .gog_rules()
marks <- rules$marks$name[rules$marks$drawable]
cells <- rules$cells

# `base::order` on purpose: gog exports an `order` atom, which masks the base
# function everywhere in the book. Calling the bare name here returns a spec
# object and the chunk dies.
row_signature <- function(m) {
  r <- cells[cells$mark == m, , drop = FALSE]
  r <- r[base::order(r$channel), , drop = FALSE]
  paste(r$channel, r$obligation, r$accepts, collapse = " | ")
}

families <- split(marks, vapply(marks, row_signature, character(1)))
families <- families[vapply(families, length, integer(1)) > 1]

# The geometric difference is prose about what each mark draws, so it is written
# here rather than derived. It is keyed on the *generated* membership, so a mark
# joining a family it has no sentence for prints a visible gap instead of being
# silently absorbed.
differs <- c(
  "line, step"    = "a stroke that **slopes** between values, against one that **holds** each value until it changes",
  "area, ribbon"  = "a region closed on the **baseline**, against one closed on a **second data boundary**",
  "box, interval" = "a body that computes its **own five-number summary**, against a span whose **ends you supply**"
)

cat("| Family | Same rules for every channel | They differ only in |\n")
cat("|---|---|---|\n")
for (f in families) {
  key <- paste(sort(f), collapse = ", ")
  cat(sprintf("| `%s` | ✓ | %s |\n",
              paste(sort(f), collapse = "` and `"),
              if (key %in% names(differs)) differs[[key]]
              else "**(no description yet for this family)**"))
}
```

The second member of a pair takes almost nothing to learn. It asks exactly the
questions its sibling asks, so you already know all of them. One sentence about
what it draws is all that is left. The two marks of a family also sit near each
other on the grid in [Combinations](../combinations.qmd), which is where the
structure you can see on that page comes from.

When does gog add a mark? Only for a shape that no existing mark can draw.
`step` was added because "hold the value until it changes" is such a shape: no
channel, transform or setting can say it. A thick line was not, because
`style(size = )` already says it. A new chart is never the reason. A histogram,
a radar and a pyramid are arrangements of marks that already exist. A new word
is spelled with the letters you have, and a new chart is written with the marks
you have. The families make the marks quick to learn. They are not a reason to
add more marks.

---

## Unpronounceable combinations

Just as some letter combinations cannot be pronounced, some mark + variable type
combinations cannot be rendered meaningfully. Each is refused outright, and
nothing is drawn: if a plot was drawn, the grammar accepted it.

`bin`, `density` and `smooth` all work along a continuous axis, so the column
they measure must be numbers. A category on that axis is refused, and each
refusal names the atom that asks the same question of categories:

| Combination | Problem | Use instead |
|---|---|---|
| `bar * bin + x(category)` | `bin` cuts a **continuous** axis into intervals | `bar * count + x(category)` |
| `bar * density + x(category)` | `density` estimates a **continuous** distribution | `bar * proportion + x(category)` |
| `point * smooth + x(category)` | `smooth` fits a curve along a **continuous axis** | `bar * mean + x(category) + y(value)` |

All three are rendered as live refusals in [What `bin`, `density` and `smooth`
refuse](../transforms.qmd#what-bin-density-and-smooth-refuse); the table
above is this page's summary of them.

The next fourteen chapters give each mark its own page, starting with
`point`.
