# Choosing a mark

```{r}
#| include: false
source("../R/setup.R")
```

Which visible form should your data take? A **mark** is the visible form your
data takes: a dot, a bar, a line. Choosing the right mark is the first question
in every plot. The answer depends on the **types of your variables** and the
**question you are asking**.

If you come from ggplot2, a mark is close to a geom, and one difference is
worth knowing early. Many geoms carry a statistic inside them.
`geom_histogram()` bins your data and then draws bars. `geom_smooth()` fits a
curve and then draws a line. A mark never does that. A mark is only the shape,
and the statistic is a separate word: a histogram is `bar * bin`, a trend line
is `line * smooth`. A geom tends to name a chart. A mark names a shape, and
there are thirteen of them.

GOG infers variable type automatically from your data:

- **Continuous**, any numeric column (`gdp`, `temperature`, `score`)
- **Categorical**, any string column (`continent`, `country`, `species`)
- **Date/time**, any `Date` or `POSIXct` column; the axis reads as a calendar
  (see [Scales](../scales.qmd))

---

## One variable

### Continuous variable

```{r}
# How is life expectancy distributed? → histogram
data(gapminder_2007) + bar * bin + x(life) +
  x_label("Life expectancy (years)") + title("Distribution — bar * bin")
```

*"Given gapminder 2007: bars derived by bin, x is life."*

```{r}
# Smooth shape of the distribution → density curve
data(gapminder_2007) + line * density + x(life) +
  x_label("Life expectancy (years)") + title("Density — line * density")
```

### Categorical variable

```{r}
# How many countries per continent? → frequency bar
data(gapminder_2007) + bar * count + x(continent) +
  title("Frequency — bar * count")
```

```{r}
# What share of countries per continent? → proportion bar
data(gapminder_2007) + bar * proportion + x(continent) +
  title("Relative frequency — bar * proportion")
```

---

## Two variables

### Continuous × Continuous

Both variables are numeric. The natural questions are about **relationships**,
**trends**, and **patterns**.

```{r}
# Is there a relationship? → scatter
data(gapminder_2007) + point + x(gdp) + y(life) + x_label("GDP per capita") +
  y_label("Life expectancy") + title("Relationship — point")
```

```{r}
# Smooth the relationship → LOESS trend line
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Trend — point + line * smooth")
```

```{r}
# How much, not just where → filled area
data(actuals) + area + x(year) + y(sales) + y_label("Sales") +
  title("Quantity — area")
```

`line` and `area` draw the same boundary and mean different things by it. A
line says *where* the value went; an area says *how much* there was, so it
fills to zero and the quantity is the ink. Choose `area` when zero is a real
number for your variable, and `line` when it is not.

When there is a *spread* to show at each x, such as a min–max range or a
confidence band, [`ribbon`](ribbon.qmd) fills between a low and a high
boundary that a `range` transform supplies. Layer a `line * mean` through it
and you have the familiar trend with a band around it.

And when the *order of the rows* is itself the story, [`path`](path.qmd) connects
them in the table's order rather than sorting by x, so the stroke may double back
or return where it began. That is the connected scatterplot: two measurements on
the axes and time in the joining.

```{r}
# The reading order is the table's → path
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Route — path (the head is 2007)")
```

Layered *over* any of these, [`rule`](rule.qmd) marks a value on one axis and
spans the other — a threshold line, or, reaching only a little way in from the
edge, a rug showing where the observations actually sit. Layered *under* them,
[`zone`](zone.qmd) shades a rectangle the same way: bounded where you give it a
pair of columns, spanning the panel where you do not. Let `bin` cut the pairs
instead of naming them and the same mark tiles the panel with counted cells,
which is a heatmap.

```{r}
# One position, the panel supplies the other → rule
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  data(gdp_rug) + rule + style(reach = "edge") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Threshold and rug — rule")
```

### Categorical × Continuous

One axis groups, the other measures. The natural question is
**how the measurement differs across groups**.

```{r}
# Mean value per group → bar
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("Group summary — bar * mean")
```

```{r}
# All individual points per group → strip plot
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("Strip plot — point (one dot per country)")
```

For the *shape* of each group rather than every point, [`box`](box.qmd) draws
the five-number summary as `box + x(continent) + y(life)`: median, quartiles,
and whiskers to the extremes. [`ribbon *
density`](ribbon.qmd#the-violin-a-distribution-per-category) draws the whole
estimated distribution that summary is a summary of, which is the violin. When the
points *are* the picture but overlap hides them,
[`point * jitter`](../transforms.qmd) spreads the pile sideways, the jittered
strip plot.

The path and region marks take a category here too, joining one summary per
group instead of standing a bar on each:

```{r}
# One value per group, joined → profile
data(gapminder_2007) + x(continent) + y(life) +
  line * mean + point * mean +
  y_label("Mean life expectancy") +
  title("Profile — line * mean")
```

Bars compare, profiles trace. A profile is the weaker claim of the two, since
the segments run through space where nothing was measured. It is worth drawing
when the categories have a direction, such as an ordered scale or a size band,
or no direction at all. A closed set of categories bent into a circle is the
radar, in [Polar](../polar.qmd). [`area`](area.qmd) fills the same boundary,
[`step`](step.qmd) holds it flat across each slot, and [`ribbon *
range`](ribbon.qmd) bands the spread.

---

## Quick reference

| x | y | Mark + transform | Question |
|---|---|---|---|
| Continuous | — | `bar * bin` | Distribution? |
| Continuous | — | `line * density` | Smooth distribution? |
| Continuous | — | `point * bin * stack` | Distribution, every observation shown? |
| Categorical | — | `bar * count` | Frequency? |
| Categorical | — | `bar * proportion` | Share? |
| Continuous | Continuous | `point` | Relationship? |
| Continuous | Continuous | `line` | Trend / time series? |
| Continuous | Continuous | `line * smooth` | Smoothed trend? |
| Continuous | Continuous | `area` | How much, over a range? |
| Continuous | Continuous | `ribbon * range` | Spread / band over a range? |
| Categorical | Continuous | `bar * mean/sum/…` | Value per group? |
| Categorical | Continuous | `point` | All values per group? |
| Categorical | Continuous | `box` | Distribution per group, summarized? |
| Categorical | Continuous | `ribbon * density` | Distribution per group, in full? |
| Categorical | Continuous | `interval * range` | Spread per group? |
| Categorical | Continuous | `line * mean` | Value per group, as a profile? |
| Categorical | Continuous | `area * mean` | The same, filled? |
| Categorical | Continuous | `line * mean` + `polar()` | The same, as a radar? |

Three variables put a third position in the sentence rather than a new mark family:

| x | y | z | Mark | Question |
|---|---|---|---|---|
| Continuous | Continuous | Continuous | `point` | Relationship, in three? |
| Continuous | Continuous | Continuous | `path` | A route through three? |
| Continuous | Continuous | Continuous | [`surface`](surface.qmd) | A height over a plane? (needs a **grid**) |
| Categorical | Categorical | — | `bar * bin` + `space()` | How many, per pair? |

Every **Categorical / Continuous** row above can be read the other way round.
`bar`, `box` and `interval` sit in a slot on one axis and measure along the
other, and they read which is which off the bindings. Swapping the two
positions therefore lays the chart on its side. `box + x(life) + y(continent)`
is the horizontal box plot, and `bar + x(gold) + y(country)` the horizontal
bar. That is
the whole of it, because there is no `flip` atom to learn.

---

## Marks come in families

The mark list looks like a set to memorize, one entry at a time. It is not, for
the same reason the consonants of Hangeul (한글) are not fourteen shapes: the
letters are *featural*, built so that most of them are another letter with a
single feature changed. ㅋ (*k*) is ㄱ (*g*) with a stroke added, and once you
know ㄱ you have most of ㅋ already.

The marks are built the same way. Some pairs share their **entire channel row**,
meaning everything they require, accept and refuse of your columns is identical,
and they part only on the geometry they draw:

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules`, the same dump the grids in Combinations read.
# The *membership* of each family is computed, never typed: if a mark's channel
# row changes it drops out of its family here on the next build, rather than this
# page going quietly on claiming a parity the engine no longer has.
rules <- .gog_rules()
marks <- rules$marks$name[rules$marks$drawable]
cells <- rules$cells

# `base::order` on purpose: gog exports an `order` atom, which masks the base
# function everywhere in the book. Calling the bare name here returns a spec
# object and the chunk dies.
row_signature <- function(m) {
  r <- cells[cells$mark == m, , drop = FALSE]
  r <- r[base::order(r$channel), , drop = FALSE]
  paste(r$channel, r$obligation, r$accepts, collapse = " | ")
}

families <- split(marks, vapply(marks, row_signature, character(1)))
families <- families[vapply(families, length, integer(1)) > 1]

# The geometric difference is prose about what each mark draws, so it is written
# here rather than derived. It is keyed on the *generated* membership, so a mark
# joining a family it has no sentence for prints a visible gap instead of being
# silently absorbed.
differs <- c(
  "line, step"    = "a stroke that **slopes** between values, against one that **holds** each value until it changes",
  "area, ribbon"  = "a region closed on the **baseline**, against one closed on a **second data boundary**",
  "box, interval" = "a body that computes its **own five-number summary**, against a span whose **ends you supply**"
)

cat("| Family | Identical channel row | They part only on |\n")
cat("|---|---|---|\n")
for (f in families) {
  key <- paste(sort(f), collapse = ", ")
  cat(sprintf("| `%s` | ✓ | %s |\n",
              paste(sort(f), collapse = "` and `"),
              if (key %in% names(differs)) differs[[key]]
              else "**(no description yet for this family)**"))
}
```

The second member of a pair is nearly free to learn. You already know every
question it will ask of your data, because it asks exactly the ones its sibling
asks, so what is left is a single sentence about what it draws. The families sit
close to one another too, often a single cell apart, which is where the
structure visible in [Combinations](../combinations.qmd) comes from.

This is a fact about *learning* the set, never a license to grow it. A mark that
is another mark with one feature changed earns its place only when that feature
is a geometry nothing else can express. `step` qualifies, because "hold the
value until it changes" is not something a channel, a transform or a setting can
say. A thick line does not, because `style(size = )` already says it.

Marks are Lego bricks, and this rule is what that comparison is good for. You
do not get a new brick because you want to build a new model. Almost every
model is a new arrangement of bricks you already have. A new brick is cut only
when the shape you need is one that no existing brick has. That is the test a
mark has to pass here: a new geometry, never a new chart.

---

## Unpronounceable combinations

Just as some letter combinations cannot be pronounced, some mark + variable type
combinations cannot be rendered meaningfully. Each is refused outright, and
nothing is drawn: a plot you can see is a plot the grammar agreed to.

The three distributional transforms need a number line to spread along, so a
category is refused on the axis they read, and each refusal names the atom that
asks the same question of categories:

| Combination | Problem | Use instead |
|---|---|---|
| `bar * bin + x(category)` | `bin` cuts a **continuous** axis into intervals | `bar * count + x(category)` |
| `line * density + x(category)` | `density` estimates a **continuous** distribution | `bar * proportion + x(category)` |
| `point * smooth + x(category)` | `smooth` fits a curve along a **number line** | `bar * mean + x(category) + y(value)` |

All three are rendered as live refusals in [What `bin`, `density` and `smooth`
refuse](../transforms.qmd), so the messages on this page cannot drift from the
ones the engine actually gives.
