# Interval

```{r}
#| include: false
source("../R/setup.R")
```

How far does each group reach, from its lowest value to its highest? `interval`
draws a span from a low value to a high one at each position: a whisker, capped
at both ends. It is the mark behind error bars and min–max ranges. A `bar`
measures from a baseline up to a value. An `interval` spans a low value and a
high value instead: it has no baseline, and it needs both before it can draw.
Whiskers stand up by default and lie down when the categories are on `y`; see
[Whiskers on their side](#whiskers-on-their-side).

Almost everyone calls this an **error bar**, and the name hides a choice no
drawing can show. Three different quantities get drawn as the same picture: a
standard deviation, a standard error, and a confidence interval. On the same
data, one can be several times longer than another. Nothing in the drawing says
which one it is, so a reader without the caption is guessing at a number's
meaning.

gog makes the sentence name the statistic instead. `interval * range` spans the
observed extremes, `interval * confidence(0.95)` computes the interval and names
its level, and `interval * bounds(lower, upper)` takes two columns you computed
yourself. The mark is the same whisker in all three; what differs is the
transform, and the transform is written down. That is the law of [Explicit Over
Implicit](../design-laws.qmd), shown in one mark: the name *error bar* can hide
the choice, and a written sentence cannot.

## A range per group

How far do the petal lengths of each species reach, from the shortest to the
longest? One whisker per species answers it, and a whisker needs a low value and
a high value to span. The low value and the high value come from a transform,
not from two extra channels. `range` reduces `y` to its minimum and maximum
within each `x` group, and `interval` spans between them:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Range of petal length per species")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal
length."*

Read it exactly like `bar * mean`. Here the transform gives a range, while
`mean` gives one number. You name the one variable whose spread you are asking
about; the transform invents the two extents and the mark spans them.

## The middle half of each group

A minimum and a maximum each come from a single row, so one unusual flower can
set the whole whisker. To show the middle of a group instead of its extremes,
give `range` two **quantiles**:

```{r}
data(iris_flowers) + interval * range(0.25, 0.75) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Middle half of petal length per species")
```

*"Given the iris flowers: intervals derived by range from 0.25 to 0.75, x is
species, y is petal length."*

This spans the first quartile to the third, which is the **interquartile
range**. `range(0.1, 0.9)` keeps the middle 80%, and any other pair works the
same way. Bare `range` is `range(0, 1)`, so the plain sentence above is
unchanged. [Transforms](../transforms.qmd#a-band-between-two-quantiles) gives
the full account of the two arguments.

## Three layers instead of one mark

A [`box`](box.qmd) draws five numbers in one mark, and you take all five or
none. Draw the band and the full range as two layers, and the median as a third.
You can restyle any of the three layers, drop one, or replace one:

```{r}
data(iris_flowers) +
  interval * range + style(color = "lightgray") +
  interval * range(0.25, 0.75) + style(color = "steelblue", size = 8) +
  point * median + style(color = "white", size = 3) +
  x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Petal length per species, drawn in three layers")
```

*"Given the iris flowers: intervals derived by range, and also intervals derived
by range from 0.25 to 0.75, and also points derived by median, x is species, y
is petal length."*

The gray whisker is the full range, the thick blue band is the middle half, and
the white dot is the median. The three layers name the three statistics. A box
plot draws the same three and names none of them.

This is not a `box` redrawn, and the difference is at the ends. A `box` stops
its whiskers at a cutoff computed from the quartiles, and draws every point
beyond that cutoff as an outlier. Drawn as a box, setosa shows four outlier dots
that are absent here, because these whiskers run to the true minimum and
maximum. What the layers give you instead is a plot you can change. Widen the
band to `range(0.1, 0.9)` and the plot reports a different statistic, with the
same marks. Drop the gray layer and the y axis covers the band rather than the
extremes. Add a third layer with another pair of quantiles, and you have the
[fan chart](ribbon.qmd#a-fan-chart-of-nested-bands) drawn over categories
instead of along a line. A `box` is still the shorter sentence when you want the
five numbers and the outlier rule together.

## A confidence interval, with a center

How well is the average petal length of each species known? That is a different
question from how far the flowers spread, and it needs a different statistic.
`range` shows the full spread. For the **uncertainty of the mean**, use
`confidence`. It computes a t-interval for each group's mean (a t-interval
widens when the group is small) and keeps that mean as the center. The interval
then draws a dot at that center: a pointrange.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length."*

The level is a parameter, so `confidence(0.99)` draws wider than
`confidence(0.95)`.

## The spread of the data

How far is a typical flower from the average of its species? That is a question
about the spread of the data, not about how well the average is known. The
opening of this chapter named a standard deviation, a standard error and a
confidence interval. `confidence` draws the third. The first is `deviation`:

```{r}
data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Mean petal length, with one standard deviation")
```

*"Given the iris flowers: intervals derived by deviation, x is species, y is
petal length."*

`deviation(2)` reaches two standard deviations on each side of the mean, so that
band is four standard deviations wide. The default is one each side. Like
`confidence` it carries a center, so it draws a pointrange rather than a bare
whisker.

The standard error has no transform of its own, and that is deliberate.
`confidence` computes it internally and reports the interval, which is what a
reader needs from it.

`deviation` and `confidence` look alike and answer different questions, so here
they are on the same fifty flowers per species, sharing one y scale:

```{r}
(data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
   y_label("Petal length (cm)") + title("Spread of the data")) |
  (data(iris_flowers) + interval * confidence + x(species) + y(petal_length) +
     y_label("") + title("Uncertainty of the mean"))
```

*"Given the iris flowers: intervals derived by deviation, x is species, y is
petal length, beside intervals derived by confidence, x is species, y is petal
length."*

The bands on the right are less than a third as long, on identical data, and the
ratio is the same in all three species. The two numbers do not contradict each
other. `deviation` describes how far the flowers are from their average, and
measuring more of them does not make that band shorter. `confidence` describes
how precisely the average itself is known. More flowers make it shorter: four
times as many flowers make it half as long. Fifty flowers per species is what
sets that ratio here.

So choose by the sentence you would write under the plot. *"A versicolor petal
is usually within half a centimeter of its species average"* describes the left
panel. *"The average versicolor petal is 4.26 cm, and the interval reaches 0.13
cm on either side"* describes the right one. Reporting one and meaning the other
is the mistake the name *error bar* allows, and the reason gog asks you to name
the statistic.

Which band is wider is not fixed, either. `confidence` narrows as the group gets
larger. On fifty flowers it is under a third of the spread; on three flowers it
is the wider of the two.

Whether a center exists is decided by the statistic, not by a setting.
`confidence` and `deviation` have a mean and `range` does not, so the statistic
chooses the geometry: a bare error bar or a pointrange. A setting can *hide* an
existing center ([below](#hiding-the-center-dot)), but no setting adds one to
`range`. See [Transforms](../transforms.qmd) for the whole range family.

## Why `interval` needs a transform

`bar` needs no transform, `bar + x + y`, because a bar measures from a baseline
the grammar already supplies: zero. An `interval` has no such default. A low
*and* a high are two numbers, and a plain `y()` supplies one. So the shortest
interval that draws, its minimum syllable, includes a transform that produces a
range. Ask for an interval without one and the engine says so, rather than
drawing an empty panel:

```{r}
#| error: true
data(iris_flowers) + interval + x(species) + y(petal_length)
```

This is the design choice that keeps the [channel vocabulary](../grammar.qmd)
small. The obvious alternative would add `ymin` and `ymax` as two more channels
meaningful to a single mark, and a box plot would need five channels. Instead a
transform invents the extents, exactly as `bin` invents a count, so the channel
set never grows. The same low and high rows are what [`box`](box.qmd) and
[`ribbon`](ribbon.qmd) draw from. When you already have the extents, from a
fitted model or a published standard error, [`bounds(lower,
upper)`](../transforms.qmd) supplies them without computing anything: `interval
* bounds(lower, upper)` draws one whisker per row.

## Whiskers on their side

Whiskers lie down for the reasons bars do: the names are long, or there are
many of them. A list of rows is then easier to scan than a row of columns. An
interval reads its orientation off the bindings, the way a [`bar`](bar.qmd)
and a [`box`](box.qmd) do. Put the categories on `y` and the measured column on
`x`, and the whiskers lie down:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(petal_length) + y(species) +
  x_label("Petal length (cm)") + y_label("Species") +
  title("The same 95% CI, lying down")
```

The caps turn with the whisker, since a cap always crosses the span at its end,
and the center dot stays at the center. Everything else in this chapter is
unchanged: `pattern`, `color` and `dodge` behave the same whichever axis carries
the categories, and so do the two settings the later sections introduce.

This is the usual shape for a **coefficient plot**, where each row is a named
estimate and the reader scans down a column of intervals. Estimates like these
normally arrive already computed, from a model rather than from raw
observations. That is what [`bounds`](../transforms.qmd) is for: it reads two
columns you already have as the low and the high, and computes nothing. With the
terms on `y`, the measured axis is the one `bounds` invents, so there is no
`x()` to write:

```{r}
coefs <- data.frame(
  term = c("Age", "Education", "Experience", "Region: North", "Region: South"),
  lo   = c(0.02, 0.31, 0.11, -0.24, -0.05),
  hi   = c(0.18, 0.55, 0.29,  0.06,  0.21)
)
data(coefs) + interval * bounds(lo, hi) + y(term) +
  x_label("Effect on log wage") +
  title("A coefficient plot, from bounds you already have")
```

*"Given the coefs table: intervals from lo to hi, y is term."*

The two intervals that cross zero cover positive and negative values alike, so
the data does not settle whether those two effects are positive or negative.

An interval needs a number to span, so two categorical axes are refused:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(continent)
```

## Splitting by color

A color says which whisker is which without reading the axis. It is also the
only way to put a second category on the same axis. `color` splits an
interval into one whisker per group, the same categorical
split `line` and `bar` make. The statistic is computed inside each group, so
each gets its own interval, its own hue, and a legend:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length, by species")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species."*

When the split is a *second* category, so several whiskers share one x, they
overlap, and [`dodge`](../transforms.qmd) sets them side by side, the same
collision modifier that separates grouped bars and boxes:

```{r}
data(gm_eras) + interval * range * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy range") +
  title("Life expectancy range per continent, 1957 vs 2007")
```

*"Given the gapminder eras: intervals derived by range and dodge, x is
continent, y is life, color by era."*

Width and opacity stay *settings*. One whisker is a single stroke, so there
is no row for `size` or `opacity` to vary over, and `style()` sets them for
the layer:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  style(color = "tomato", size = 2.5) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A heavier, one-color range")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal
length, colored tomato, with size 2.5."*

## Caps, or a bare linerange

Whether a whisker ends in a crossbar is a matter of style, and a crowded plot
often reads more cleanly without them. The end caps, the short crossbars, are
decoration, not data; they only
emphasize where a whisker stops. So whether to draw them is a *setting*:
`style(caps = FALSE)` drops them for a bare linerange, and `caps = TRUE` (the
default) keeps the capped error bar.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(caps = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, drawn as a linerange")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with caps off."*

Because caps are geometry only `interval` has, `style(caps = )` on any other
mark is refused with direction, the same way a bar's `border_size` is refused on
a `line`. `caps` combines with `range`, `confidence` and `deviation`, and with
the `color` split. So error bar, linerange, pointrange and their colored
versions are one mark, one transform and one setting.

## Hiding the center dot

A pointrange's dot is the mean; sometimes you want the confidence whiskers
without it, a plain error bar. `style(center = FALSE)` drops the dot, and
`center = TRUE` (the default) keeps it. Like `caps`, it only *hides*: it cannot
add a center to `range`, which has none to draw.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(center = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, without its center dots")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with center off."*

`center` and `caps` are the interval's two settings, one for its middle and one
for its ends, so a confidence interval draws four ways. A capped pointrange is
the default. A capped bare error bar is `center = FALSE`. A linerange with a dot
is `caps = FALSE`, and a plain linerange is both `FALSE`. Like `style(caps = )`,
`style(center = )` belongs to `interval` alone; on any other mark it is refused
with direction.

## A dashed whisker

A dash separates one layer from another without a second color, and it
survives a print in black and white. A whisker is a **stroke**, so it takes
`style(pattern = )` the same way a
[`line`](line.qmd) does. This is the [settable rule](../design-laws.qmd): a dash
patterns every path-stroke mark (`line`, `step`, `interval`, `path`, `rule`).
The dash is drawn on the whisker span only. The caps and the center dot stay
solid, since a dashed crossbar is hard to read as a cap:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(pattern = "dashed") +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A dashed 95% CI")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with pattern dashed."*

The five-number summary has its own mark, [`box`](box.qmd), and the same
low and high pair drawn as a continuous fill is the [`ribbon`](ribbon.qmd)
band. See
[Transforms](../transforms.qmd) for where `range`, `confidence` and `deviation`
sit among the others.

## What you can set

A setting changes how every whisker looks without reading a column, and each
mark takes its own. These are an interval's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("interval")
```

`caps` and `center` are the interval's own settings, and both only ever *hide*.
`caps` turns the end crossbars off, leaving a bare linerange; `center` hides the
dot that `confidence` supplies. Neither can add geometry the statistic did not
produce, so `style(center = TRUE)` on `interval * range` changes nothing: a
range has no center. The plot below reads `gm_continents`, three continents
across every year from 1952 to 2007, so each interval pools twelve years of
countries.

```{r}
#| echo: false
peek(gm_continents)
```

```{r}
data(gm_continents) + interval * confidence(0.99) + x(continent) + y(life) +
  style(caps = FALSE, size = 3, color = "darkslateblue", pattern = "dotted") +
  y_label("Life expectancy") + title("A bare, thick, dotted 99% interval")
```

One number is missing from that table, and the transform carries it:
`confidence(0.99)` widens the interval from the default 0.95. That belongs to
the transform rather than to `style()`, because it changes *what is computed*,
not how it is drawn. The test is whether the numbers change: they do here, so it
is not a setting.

## What it refuses

An interval is a stroke, however much it looks like a bar, so its pattern is a
dash and a hatch is refused:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  style(pattern = "hatch")
```

That refusal is worth explaining, because a capped range looks like a shape with
an inside. The engine draws it as a stroke, which is why the whisker takes a
dash and not a hatch. Two more refusals appear earlier in this chapter. A bare
`interval` has no extents to span, under [Why `interval` needs a
transform](#why-interval-needs-a-transform), and an interval with two
categorical axes has nothing to measure, under [Whiskers on their
side](#whiskers-on-their-side).