# Interval

```{r}
#| include: false
source("../R/setup.R")
```

How far does each group reach, from its lowest value to its highest? `interval`
draws a span from a low value to a high one at each position: a whisker,
capped at both ends. It is the mark behind error bars and min–max ranges. Where
`bar` measures *from a baseline* up to a value, an `interval` **floats between
two extents**: it has no baseline, and it needs both a low and a high before it
can draw anything at all. Whiskers stand up by default and lie down when the
categories are on `y`; see [Whiskers on their side](#whiskers-on-their-side).

Almost everyone calls this an **error bar**, and that name carries a problem the
mark cannot fix. Three different quantities get drawn as the same picture, a
standard deviation, a standard error, and a confidence interval, and on the same
data they differ by large factors. Nothing in the drawing says which one it is, so
a reader without the caption is guessing at a number's meaning.

gog makes the sentence say it instead. `interval * range` spans the observed
extremes, `interval * confidence(0.95)` computes the interval and names its
level, and `interval * bounds(low, high)` takes two columns you worked out
yourself. The mark is the same whisker in all three; what differs is the
transform, and the transform is written down. That is Explicit Over Implicit in
one mark: the name *error bar* can hide the choice, and a sentence cannot.

## A range per group

Those two extents come from a transform, not from two extra channels. `range`
reduces `y` to its minimum and maximum within each `x` group, and `interval`
spans between them:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Range of petal length per species")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal length."*

Read it exactly like `bar * mean`: one range where the other draws a single
summarized value. You name the one variable whose spread you are asking about;
the transform invents the two extents and the mark spans them.

## The middle half of each group

A minimum and a maximum each come from a single row, so one unusual flower can
set the whole whisker. To show the middle of a group instead of its extremes,
give `range` two **quantiles**:

```{r}
data(iris_flowers) + interval * range(0.25, 0.75) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Middle half of petal length per species")
```

This spans the first quartile to the third, which is the **interquartile range**.
`range(0.1, 0.9)` keeps the middle 80%, and any other pair works the same
way. Bare `range` is `range(0, 1)`, so the plain sentence above it is unchanged.
[Transforms](../transforms.qmd#a-band-between-two-quantiles) gives the full
account of the two arguments.

## Three layers instead of one mark

A [`box`](box.qmd) draws five numbers in one mark, and you take all five or none.
Draw the band and the full range as two layers, and the median as a third. Style,
drop, or replace any of them:

```{r}
data(iris_flowers) +
  interval * range + style(color = "lightgray") +
  interval * range(0.25, 0.75) + style(color = "steelblue", size = 8) +
  point * median + style(color = "white", size = 3) +
  x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Petal length per species, drawn in three layers")
```

The gray whisker is the full range, the thick blue band is the middle half, and
the white dot is the median. The three layers name the three statistics. A box
plot draws the same three and names none of them.

This is not a `box` redrawn, and the difference is at the ends. A `box` stops its
whiskers at Tukey's fences and draws the points beyond as outliers, so setosa
gains four dots that are not here, while these whiskers run to the true minimum
and maximum. What the layers give you instead is a plot you can change. Widen the
band to `range(0.1, 0.9)` and the picture changes meaning without changing marks.
Drop the gray layer and the y axis fits the band rather than the extremes. Add a
third layer with another pair, and you have the [fan
chart](ribbon.qmd#a-fan-chart-of-nested-bands) at categories rather than along a
line. A `box` is still the shorter sentence when you want the five numbers and
the outlier rule together.

## A confidence interval, with a center

`range` shows the full spread. For the **uncertainty of the mean**, an error bar,
use `confidence`, which computes the mean's t-interval per group and carries a
center (the mean) with it. The interval then draws a dot at that center: a
pointrange.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length")
```

The level is a parameter (`confidence(0.99)` is wider) and it uses the
t-distribution, so a small sample's interval is honestly wider than a fixed 1.96
would draw.

## The spread of the data

The opening of this chapter named a standard deviation, a standard error and a
confidence interval. `confidence` draws the third. The first is `deviation`:

```{r}
data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Mean petal length, with one standard deviation")
```

`deviation(2)` reaches two standard deviations on each side of the mean, so that
band is four standard deviations wide. The default is one each side. Like
`confidence` it carries a center, so it draws a pointrange rather than a bare
whisker.

The standard error has no transform of its own, and that is deliberate: it is
`confidence`'s ingredient rather than a band anyone draws. What a reader wants
from it is the interval, and `confidence` gives that with its level named.

`deviation` and `confidence` look alike and answer different questions, so here
they are over the same fifty flowers per species, sharing one y scale:

```{r}
(data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
   y_label("Petal length (cm)") + title("Spread of the data")) |
  (data(iris_flowers) + interval * confidence + x(species) + y(petal_length) +
     y_label("") + title("Uncertainty of the mean"))
```

The bands on the right are less than a third as long, on identical data, and the
factor is the same in all three species. The two numbers do not contradict each
other. `deviation` describes how far the flowers are from their average, and
measuring more of them does not shrink it. `confidence` describes how precisely
the average itself is known. More flowers shrink it: four times as many halves
it. Fifty flowers per species is what sets the factor here.

So choose by the sentence you would write under the plot. *"A versicolor petal is
usually within half a centimeter of its species average"* describes the left
panel. *"The average versicolor petal is 4.26 cm, give or take 0.13"* describes
the right one. Reporting one and meaning the other is the mistake the name
*error bar* allows, and the reason gog asks you to name the statistic.

Which band is wider is not fixed, either. `confidence` narrows as the group gets
larger. On fifty flowers it is under a third of the spread; on three flowers it
is the wider of the two.

Whether a center exists is decided by the statistic, not by a setting.
`confidence` and `deviation` have a mean and `range` does not, so the statistic
chooses the geometry: a bare error bar or a pointrange. A setting can *hide* an
existing center ([below](#hiding-the-center-dot)), but no setting adds one to
`range`. See [Transforms](../transforms.qmd) for the whole range family.

## Why `interval` needs a transform

`bar` can stand alone, `bar + x + y`, because a bar measures from a baseline the
grammar already knows: zero. An `interval` has no such default. A low *and* a high
are two numbers, and a plain `y()` supplies one, so the mark's minimum syllable
includes a range-producing transform. Ask for an interval without one and the
engine says so, rather than drawing an empty panel:

```{r}
#| error: true
data(iris_flowers) + interval + x(species) + y(petal_length)
```

This is the design choice that keeps the [channel vocabulary](../grammar.qmd)
small. The obvious alternative, `ymin`/`ymax` aesthetics, would add two channels
meaningful to a single mark, and a box plot would need five. Instead the extents
are **invented by a transform**, exactly as `bin` invents a count: the
channel set never grows, and the same
low/high-rows mechanism feeds the [`box`](box.qmd) mark and the
[`ribbon`](ribbon.qmd) band as well. When you have the extents *already*, a
pre-computed error bar from a model or an SEM, [`bounds(lower,
upper)`](../transforms.qmd) supplies them without computing anything:
`interval * bounds(lower, upper)` draws one whisker per row.

## Whiskers on their side

An interval reads its orientation off the bindings, the way a [`bar`](bar.qmd)
and a [`box`](box.qmd) do. Put the categories on `y` and the measured column on
`x`, and the whiskers lie down:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(petal_length) + y(species) +
  x_label("Petal length (cm)") + y_label("Species") +
  title("The same 95% CI, lying down")
```

The caps turn with the whisker, since a cap always crosses the span it ends, and
the center dot rides along. Everything else in this chapter is unchanged:
`caps`, `center`, `pattern`, `color` and `dodge` behave the same whichever axis
carries the categories.

This is the usual shape for a **coefficient plot**, where each row is a named
estimate and the reader scans down a column of intervals. Estimates like these
normally arrive already computed, from a model rather than from raw
observations, which is what [`bounds`](../transforms.qmd) is for: it reshapes two
columns you already have into the low/high pair, computing nothing. With the
terms on `y`, the measured axis is the one `bounds` invents, so there is no `x()`
to write:

```{r}
coefs <- data.frame(
  term = c("Age", "Education", "Experience", "Region: North", "Region: South"),
  lo   = c(0.02, 0.31, 0.11, -0.24, -0.05),
  hi   = c(0.18, 0.55, 0.29,  0.06,  0.21)
)
data(coefs) + interval * bounds(lo, hi) + y(term) +
  x_label("Effect on log wage") +
  title("A coefficient plot, from bounds you already have")
```

Read the two intervals that cross zero as the two effects the data cannot sign.

An interval needs a number to span, so two categorical axes are refused:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(continent)
```

## Splitting by color

`color` splits an interval into one whisker per group, the same discrete split
`line` and `bar` make, and the statistic runs within each group, so every group
gets its own interval, its own hue, and a legend to decode them:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length, by species")
```

When the split is a *second* category, so several whiskers share one x, they
overlap, and [`dodge`](../transforms.qmd) sets them side by side, the same
collision modifier that separates grouped bars and boxes:

```{r}
data(gm_eras) + interval * range * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy range") +
  title("Life expectancy range per continent, 1957 vs 2007")
```

Width and opacity stay *settings*, though; one whisker is a single stroke, so
there is nothing for a per-row `size` or `opacity` to vary along, and `style()`
sets them for the whole layer:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  style(color = "tomato", size = 2.5) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A heavier, one-color range")
```

## Caps, or a bare linerange

The end caps, the short crossbars, are decoration, not data; they only
emphasize where a whisker stops. So whether to draw them is a *setting*:
`style(caps = FALSE)` drops them for a bare linerange, and `caps = TRUE` (the
default) keeps the capped error bar.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(caps = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, drawn as a linerange")
```

Because caps are geometry only `interval` has, `style(caps = )` on any other mark
is refused with direction, the same way a bar's `border_size` is refused on a
`line`. It pairs with the two range statistics and the `color` split, so
error bar, linerange, pointrange, and colored versions of each are all
`interval` plus a transform and a setting, never a new mark to learn.

## Hiding the center dot

A pointrange's dot is the mean; sometimes you want the confidence whiskers
without it, a plain error bar. `style(center = FALSE)` drops the dot, and
`center = TRUE` (the default) keeps it. Like `caps`, it only *hides*: it cannot
add a center to `range`, which has none to draw.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(center = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, without its center dots")
```

`center` and `caps` are the interval's two display toggles, its middle and its
ends, so a confidence interval draws four ways. A capped pointrange is the
default. A capped bare error bar is `center = FALSE`. A linerange with a dot
is `caps = FALSE`, and a plain linerange is both `FALSE`. Like `style(caps = )`,
`style(center = )` belongs to `interval` alone; on any other mark it is refused
with direction.

## A dashed whisker

A whisker is a **stroke**, so it takes `style(pattern = )` the same way a
[`line`](line.qmd) does, the [settable rule](../design-laws.qmd) at work: a dash
patterns any of the path-stroke marks (`line`, `step`, `interval`). The pattern
falls on the whisker span; the caps and center dot stay solid, since a dashed
crossbar reads as noise:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(pattern = "dashed") +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A dashed 95% CI")
```

The five-number summary lives in its own mark, [`box`](box.qmd), and the same
low/high pair drawn as a continuous fill is the [`ribbon`](ribbon.qmd) band. See
[Transforms](../transforms.qmd) for where `range` and `confidence` sit among the
others.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("interval")
```

The last two are the interval's own, and both only ever *hide*. `caps` turns the
end crossbars off, leaving a bare linerange; `center` hides the dot that
`confidence` supplies. Neither can conjure geometry the statistic did not
produce, which is why `style(center = TRUE)` on `interval * range` adds nothing:
a range has no center to draw.

```{r}
data(gm_continents) + interval * confidence(0.99) + x(continent) + y(life) +
  style(caps = FALSE, size = 3, color = "darkslateblue", pattern = "dotted") +
  y_label("Life expectancy") + title("A bare, thick, dotted 99% interval")
```

The statistic carries the other parameter worth knowing: `confidence(0.99)`
widens the interval from the default 0.95. That belongs to the transform rather
than to `style()`, because it changes *what is computed*, not how it is painted.
The test is whether the numbers move: they do here, so it is not a setting.

## What it refuses

An interval is a stroke, however much it looks like a bar, so its pattern is a
dash and a hatch is refused:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  style(pattern = "hatch")
```

That is worth meeting, because the eye reads a capped range as a shape with an
inside. The engine reads it as line-work, and the dashed whisker above is the
same fact arrived at from the other side. Its second refusal, a bare `interval`
with no transform to give it extents, opens the chapter.
