# Interval

```{r}
#| include: false
source("../R/setup.R")
```

How far does each group reach, from its lowest value to its highest? `interval`
draws a span from a low value to a high one at each position: a whisker, capped
at both ends. It is the mark that draws error bars and min–max ranges. A `bar`
measures from a baseline up to a value. An `interval` has no baseline. It spans
from a low value to a high value, and it needs both of them before it can draw.
Whiskers are vertical when the categories are on `x` and horizontal when they
are on `y`; see [Horizontal whiskers](#horizontal-whiskers).

Almost everyone calls this an **error bar**, and that name does not say what the
whisker measures. Three different quantities are drawn as the same picture: a
standard deviation, a standard error, and a confidence interval. On the same
data, one can be several times longer than another. Nothing in the drawing says
which one it is, and a reader with no caption cannot tell.

gog makes the sentence name the statistic instead. `interval * range` spans the
observed extremes. `interval * confidence(0.95)` computes a confidence interval
and states its level in the sentence. `interval * bounds(lower, upper)` reads
two columns you computed yourself. The mark is the same whisker in all three;
what differs is the transform, and the transform is written down. That is the
law of [Explicit Over
Implicit](../design-laws.qmd#law-5-explicit-over-implicit), shown in one mark:
the name *error bar* does not say which statistic it drew, and a written
sentence must.

## A range per group

How far do the petal lengths of each species reach, from the shortest to the
longest? One whisker per species answers it, and a whisker needs a low value and
a high value to span. The low value and the high value come from a transform,
not from two extra channels. `range` reduces `y` to its minimum and maximum
within each `x` group, and `interval` spans between them:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Range of petal length per species")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal
length."*

This sentence has the same shape as `bar * mean`. The difference is what the
transform returns: `mean` returns one number per group, and `range` returns two.
You name only the one variable whose spread you are asking about. The transform
invents the two extents, and the mark spans them.

`range` is not the interval's own transform. Three other marks read the same
low and high value, and only the shape changes:

| Mark derived by `range` | Shape |
|---|---|
| `ribbon * range` | the band between the two values, filled |
| `line * range` | the two values as two separate strokes |
| `step * range` | the same two boundaries, as staircases |
| `interval * range` | one whisker per group |

`range` is one of four **pair transforms**. A pair transform supplies a low
value and a high value, where most transforms supply a single number. The other
three are `confidence`, `deviation` and `bounds`, and each one works the same
way in every row of that table. A mark that spans a pair takes all four.

## The middle half of each group

A minimum and a maximum each come from a single row, so one unusual flower can
set the whole whisker. To show the middle of a group instead of its extremes,
give `range` two **quantiles**:

```{r}
data(iris_flowers) + interval * range(0.25, 0.75) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Middle half of petal length per species")
```

*"Given the iris flowers: intervals derived by range from 0.25 to 0.75, x is
species, y is petal length."*

This spans the first quartile to the third, which is the **interquartile
range**. `range(0.1, 0.9)` keeps the middle 80%, and any other pair works the
same way. `range` written with no arguments means `range(0, 1)`, so it still
spans from the minimum to the maximum.
[Transforms](../transforms.qmd#a-band-between-two-quantiles) gives the full
account of the two arguments.

## Three layers instead of one mark

A [`box`](box.qmd) draws five numbers in one mark, and you take all five or
none. Draw the band and the full range as two layers, and the median as a third.
You can give any of the three layers its own style, drop one, or replace one:

```{r}
data(iris_flowers) +
  interval * range + style(color = "darkgray") +
  interval * range(0.25, 0.75) + style(color = "steelblue", size = 8) +
  point * median + style(color = "gold", size = 5, opacity = 1) +
  x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Petal length per species, drawn in three layers")
```

*"Given the iris flowers: intervals derived by range, and also intervals derived
by range from 0.25 to 0.75, and also points derived by median, x is species, y
is petal length."*

The gray whisker is the full range, the thick blue band is the middle half, and
the gold dot is the median. Order decides what you can see. A later layer is
painted over an earlier one, so the median is written last. Move it to the
front of the sentence and the band draws over it, leaving almost nothing to
see. If you have used ggplot2, this is the rule you already know: a `geom_`
added later draws on top. The dot is set opaque as well, since every mark is
slightly translucent by default.

The three layers name the three statistics, and a box plot draws the same three
and names none of them. This is not a `box` redrawn, though, and the difference
is at the ends. A `box` stops its whiskers at a cutoff computed from the
quartiles, and draws every point beyond that cutoff as an outlier. A `box` on
this same data draws outlier dots for setosa and versicolor. They are absent
here, because these whiskers run to the true minimum and maximum.

Because each statistic is its own layer, you can change one without touching the
others. Widen the band to `range(0.1, 0.9)` and the plot reports a different
statistic, with the same marks. Drop the gray layer and the y axis is scaled to
the band rather than to the extremes. Add a third layer with another pair of
quantiles, and you have the [fan chart](ribbon.qmd#a-fan-chart-of-nested-bands)
drawn over categories instead of along a line. A `box` is still the shorter
sentence when you want the five numbers and the outlier rule together.

## A confidence interval, with a center

How well is the average petal length of each species known? That is a different
question from how far the flowers spread, and it needs a different statistic.
`range` shows the full spread. For the **uncertainty of the mean**, use
`confidence`. It computes a **t-interval** for each group's mean, which is the
confidence interval the opening named, and that interval widens when the group
is small. It keeps the mean as the center of the span, and the mark draws a dot
there. A whisker with a dot at its center is a **pointrange**.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length."*

The level is the only parameter, so `confidence(0.99)` draws wider than
`confidence(0.95)`. `confidence` computes a t-interval and takes no other. For
a z-interval, a bootstrap, or the standard errors a model already gave you,
compute the two numbers yourself and pass them to
[`bounds`](../transforms.qmd#bounds-a-pre-computed-lowhigh-pair).

## The spread of the data

How far is a typical flower from the average of its species? That is a question
about the spread of the data, not about how well the average is known. The
opening of this chapter named three quantities: a standard deviation, a standard
error and a confidence interval. `confidence` draws the confidence interval. The
standard deviation has a transform of its own, `deviation`:

```{r}
data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Mean petal length, with one standard deviation")
```

*"Given the iris flowers: intervals derived by deviation, x is species, y is
petal length."*

`deviation` alone reaches one standard deviation on each side of the mean.
`deviation(2)` reaches two, so that band is four standard deviations wide. Like
`confidence` it carries a center, so it draws a pointrange rather than a bare
whisker.

The standard error has no transform of its own, and it does not need one. It is
the standard deviation divided by the square root of the group size.
`confidence` computes it internally and reports the interval built from it,
which is what a reader of the plot needs.

`deviation` and `confidence` look alike and answer different questions. `|` puts
two plots side by side on one page, and because both name `petal_length` on `y`
they share that scale, drawn once down the left. The same operator splits one
plot into panels when `facet()` is on its right instead, and
[Composition](../composition.qmd) and [Faceting](../faceting.qmd) have the rules
for each. Here are the two statistics on the same fifty flowers per species:

```{r}
(data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
   y_label("Petal length (cm)") + title("Spread of the data")) |
  (data(iris_flowers) + interval * confidence + x(species) + y(petal_length) +
     y_label("") + title("Uncertainty of the mean"))
```

*"Given the iris flowers: intervals derived by deviation, x is species, y is
petal length, beside intervals derived by confidence, x is species, y is petal
length."*

The bands on the right are less than a third as long, and the data is identical.
The ratio is the same in all three species. The two statistics do not contradict
each other. `deviation` describes how far the flowers are from their average,
and measuring more of them does not make that band shorter. `confidence`
describes how precisely the average itself is known. More flowers make it
shorter: four times as many flowers make it half as long. That ratio comes from
the group size, fifty flowers per species.

So choose by the caption you would write under the plot. *"A versicolor petal
is usually within half a centimeter of its species average"* describes the left
panel. *"The average versicolor petal is 4.26 cm, and the interval reaches 0.13
cm on either side"* describes the right one. Reporting one and meaning the other
is the mistake the name *error bar* allows, and the reason gog asks you to name
the statistic.

Which band is wider is not fixed, either. `confidence` narrows as the group gets
larger. On fifty flowers it is under a third of the spread; on three flowers it
is the wider of the two.

Whether a center exists is decided by the statistic, not by a setting.
`confidence` and `deviation` compute a mean and `range` does not. So the shape
follows from the statistic: a bare error bar for `range`, a pointrange for the
other two. A setting can
*hide* an existing center, as [Hiding the center dot](#hiding-the-center-dot)
shows, but no setting adds one to `range`. See
[Transforms](../transforms.qmd) for the whole range family.

## Why `interval` needs a transform

Every plot so far has named a transform, and the rule behind that is worth
stating on its own. A bar needs no transform, because the grammar already
supplies its baseline: zero. `bar + x + y` draws. An `interval` has no such
default. A low *and* a high are two numbers, and a plain `y()` supplies one. So
the shortest interval that will draw is a mark, two positions, and a transform
that produces a range. That is its minimum syllable. Ask for an interval without
a transform and the engine refuses the sentence rather than drawing an empty
panel. [What it refuses](#what-it-refuses) shows that refusal.

This is the design choice that keeps the [channel
vocabulary](../grammar.qmd#channels-a-column-made-visible) small. The obvious
alternative would add `ymin` and `ymax` as two more channels meaningful to a
single mark, and a box plot would need five channels. Instead a transform
invents the extents, exactly as `bin` invents a count, so the channel set never
grows. [`box`](box.qmd) and [`ribbon`](ribbon.qmd) read the same low and high
values. The pair transform
[`bounds(lower, upper)`](../transforms.qmd#bounds-a-pre-computed-lowhigh-pair)
takes extents you already hold, and
[Horizontal whiskers](#horizontal-whiskers) draws a coefficient plot with it.

## Horizontal whiskers

Whiskers turn horizontal for the reasons bars do: the names are long, or there
are many of them. A list of rows is then easier to scan than a row of columns.
An interval takes its orientation from the bindings, the way a [`bar`](bar.qmd)
and a [`box`](box.qmd) do. Put the categories on `y` and the measured column on
`x`:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(petal_length) + y(species) +
  x_label("Petal length (cm)") + y_label("Species") +
  title("The same 95% CI, horizontal")
```

The caps turn with the whisker, since a cap always crosses the span at its end.
The center dot stays at the center. Everything else in this chapter is
unchanged: `pattern`, `color` and `dodge` behave the same whichever axis carries
the categories, and so do `caps` and `center`.

This is the usual shape for a **coefficient plot**, where each row is a named
estimate and the reader reads down the list of intervals. Estimates like these
normally arrive already computed, from a model rather than from raw
observations. That is what
[`bounds`](../transforms.qmd#bounds-a-pre-computed-lowhigh-pair) is for: it
reads two columns you already have as the low and the high, and computes
nothing. The terms go on `y`. `bounds` supplies the values for the other axis,
so there is no `x()` to write:

```{r}
#| echo: false
peek(coefs)
```

```{r}
data(coefs) + interval * bounds(lo, hi) + y(term) +
  x_label("Effect on log wage") +
  title("A coefficient plot, from bounds you already have")
```

*"Given the coefs table: intervals from lo to hi, y is term."*

Two intervals, Region: North and Region: South, run from a negative low to a
positive high. Each one includes zero, so this data cannot say whether either
effect is positive or negative.

An interval needs a number to span, so two categorical axes are refused, as
they are on a bar and a box. [What it refuses](#what-it-refuses) shows that
refusal.

## Splitting by color

A color says which whisker is which without reading the axis. The x axis
already carries one category, so a channel is what separates a second one.
`color` splits an interval
into one whisker per group, the same categorical split `line` and `bar` make.
The statistic is computed inside each group, so every group gets its own
interval and its own hue. One legend names all of them:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length, by species")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species."*

A split by a *second* category puts several whiskers at one x, and they overlap.
[`dodge`](../transforms.qmd#dodge-side-by-side-instead-of-on-top) sets them side
by side instead. It is the same collision modifier that separates grouped bars
and boxes:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + interval * range * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy range") +
  title("Life expectancy range per continent, 1957 vs 2007")
```

*"Given the gapminder eras: intervals derived by range and dodge, x is
continent, y is life, color by era."*

Width and opacity stay *settings*. A channel needs one value per row, and one
whisker is a single stroke drawn from a whole group, so `size` and `opacity`
have no rows to follow. `style()` sets them once for the layer:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  style(color = "tomato", size = 2.5) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A heavier, one-color range")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal
length, colored tomato, with size 2.5."*

## Caps, or a bare linerange

The end caps are the short crossbars at each end of a whisker. They are
decoration rather than data, because they only emphasize where a whisker stops,
and a crowded plot often reads more cleanly without them. So whether to draw
them is a *setting*. `style(caps = FALSE)` drops them for a bare **linerange**,
a whisker with no crossbars at its ends. `caps = TRUE`, the default, keeps the
capped error bar.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(caps = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, drawn as a linerange")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with caps off."*

No other mark has caps. On any other mark `style(caps = )` is refused with
direction, the same way a bar's `border_size` is refused on a `line`. `caps`
combines with `range`, `confidence` and `deviation`, and with the `color` split.
So the error bar, the linerange, the pointrange and their colored versions are
all the same mark. Only the transform and one setting change.

## Hiding the center dot

A pointrange's dot is the mean. Sometimes you want the whiskers without it,
which is a plain error bar. `style(center = FALSE)` drops the dot, and
`center = TRUE` (the default) keeps it. Like `caps`, it only *hides*: it cannot
add a center to `range`, which has none to draw.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(center = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, without its center dots")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with center off."*

`center` and `caps` are the interval's two settings. One is for its middle and
one for its ends. Each takes two values, so a confidence interval draws four
ways. A capped pointrange is
the default. A capped bare error bar is `center = FALSE`. A linerange with a dot
is `caps = FALSE`, and a plain linerange is both `FALSE`. Like `style(caps = )`,
`style(center = )` belongs to `interval` alone; on any other mark it is refused
with direction.

## A dashed whisker

A dash separates one layer from another without a second color, and it still
separates them when the plot is printed in black and white. A whisker is a
**stroke**, so it takes `style(pattern = )` the same way a [`line`](line.qmd)
does. This is the [settable rule](../design-laws.qmd#the-settable-rule): a dash
patterns every path-stroke mark (`line`, `step`, `interval`, `path`, `rule`).
The dash is drawn on the whisker span only. The caps and the center dot stay
solid, since a dashed crossbar is hard to read as a cap:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(pattern = "dashed") +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A dashed 95% CI")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with pattern dashed."*

## A bar with its error bar

Two marks can answer two questions about the same groups. A `bar` shows where
each group's mean sits. An `interval` shows how well that mean is known. Layer
them and you have a plot that appears throughout the social sciences, usually
with a bare error bar rather than a pointrange:

```{r}
data(iris_flowers) + bar * mean + style(color = "lightgray") +
  interval * confidence(0.95) + style(center = FALSE, color = "black") +
  x(species) + y(petal_length) +
  x_label("Species") + y_label("Mean petal length (cm)") +
  title("Mean petal length, with a 95% confidence interval")
```

*"Given the iris flowers: bars derived by mean, colored lightgray, and also
intervals derived by confidence at 95%, with center off, colored black, x is
species, y is petal length."*

Each layer names its own statistic. `bar * mean` reduces each group to one
number, and `interval * confidence(0.95)` computes the interval around that
number. Both read the same column and report different things, and the sentence
says which is which. The dot is turned off because the top of the bar already
marks the mean, and the engine draws the two at exactly the same height.

The opening of this chapter named three quantities that all draw as the same
whisker. A published figure of this shape has to say in its caption which one it
used, because the drawing cannot show it. Here the sentence carries it:
`confidence(0.95)` sits beside the plot, and `deviation` would be a different
sentence and a different picture.

A second category goes side by side, with the same `dodge` that separates
grouped bars. Bind `color` to the bars and `group` to the whiskers. Both layers
then split into the same slots, and the plot keeps one legend:

```{r}
data(gm_eras) +
  bar * mean * dodge + color(era) +
  interval * confidence(0.95) * dodge + group(era) +
  style(center = FALSE, color = "black") +
  x(continent) + y(life) +
  y_label("Mean life expectancy (years)") +
  title("Mean life expectancy by continent, 1957 and 2007")
```

*"Given the gapminder eras: bars derived by mean and dodge, color by era, and
also intervals derived by confidence at 95% and dodge, grouped by era, with
center off, colored black, x is continent, y is life."*

What this plot does not show is the data. Two groups can have the same mean and
the same interval and still hold very different values, and a bar reports only
the one number. [Three layers instead of one
mark](#three-layers-instead-of-one-mark) draws the spread itself, and
[`box`](box.qmd) draws five numbers at once.

## What you can set

A setting changes how every whisker looks without reading a column, and each
mark takes its own. These are an interval's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("interval")
```

`caps` and `center` are the interval's own settings, and neither can add
geometry the statistic did not produce. A range has no center, so `style(center
= TRUE)` on `interval * range` changes nothing. It draws the same picture as the
sentence without the setting. Several settings can also be given at once. This
plot reads `gm_continents`, three continents across every year from 1952 to
2007, so each interval is computed from every country in that continent, across
all twelve years.

```{r}
#| echo: false
peek(gm_continents)
```

```{r}
data(gm_continents) + interval * confidence(0.99) + x(continent) + y(life) +
  style(caps = FALSE, size = 3, color = "darkslateblue", pattern = "dotted") +
  y_label("Life expectancy") + title("A bare, thick, dotted 99% interval")
```

The confidence level is not in that table, because it is not a setting. It is
written into the transform instead. `confidence(0.99)` changes *what is
computed*, not how it is drawn. The test is whether the numbers change: they do
here, so it is not a setting.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

An interval is a stroke, however much it looks like a bar, so its pattern is a
dash and a hatch is refused:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  style(pattern = "hatch")
```

That refusal has a reason. A capped range looks like a shape with an inside. The
engine draws it as a stroke, which is why the whisker takes a dash and not a
hatch.

A bare `interval` is the second refusal. A plain `y()` supplies one number, and
a span needs two, as [Why `interval` needs a
transform](#why-interval-needs-a-transform) explained. The engine says so
rather than drawing an empty panel:

```{r}
#| error: true
data(iris_flowers) + interval + x(species) + y(petal_length)
```

The message names two of those transforms, `range` and `bounds`, and writes out
a sentence the engine would accept. Two categorical axes are the third refusal.
An interval takes its orientation from the bindings, so one axis has to be
continuous:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(continent)
```

That continuous column is where the low and the high come from. Two categories
leave the mark nothing to span.

## In short

An interval is geometry and nothing else. The transform says which two numbers
it spans, and whether there is a center to draw. Caps, the center dot and the
dash are settings, so none of the three reads a column.

## What comes next

Which two numbers an interval spans is the transform's answer, never the
mark's. [Box](box.qmd) answers it inside the mark, with the five-number summary
that gives the box plot its name.
