# Interval

```{r}
#| include: false
source("../R/setup.R")
```

How far does each group reach, from its lowest value to its highest? `interval`
draws a span from a low value to a high one at each position: a whisker,
capped at both ends. It is the mark behind error bars and min–max ranges. Where
`bar` measures *from a baseline* up to a value, an `interval` **floats between
two extents**: it has no baseline, and it needs both a low and a high before it
can draw anything at all. Whiskers stand up by default and lie down when the
categories are on `y`; see [Whiskers on their side](#whiskers-on-their-side).

Almost everyone calls this an **error bar**, and that name carries a problem the
mark cannot fix. Three different quantities get drawn as the same picture, a
standard deviation, a standard error, and a confidence interval, and on the same
data they differ by large factors. Nothing in the drawing says which one it is, so
a reader without the caption is guessing at a number's meaning.

gog makes the sentence say it instead. `interval * range` spans the observed
extremes, `interval * confidence(0.95)` computes the interval and names its
level, and `interval * bounds(low, high)` takes two columns you worked out
yourself. The mark is the same whisker in all three; what differs is the
transform, and the transform is written down. That is Explicit Over Implicit in
one mark: the name *error bar* can hide the choice, and a sentence cannot.

## A range per group

Those two extents come from a transform, not from two extra channels. `range`
reduces `y` to its minimum and maximum within each `x` group, and `interval`
spans between them:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Range of petal length per species")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal length."*

Read it exactly like `bar * mean`: one range where the other draws a single
summarized value. You name the one variable whose spread you are asking about;
the transform invents the two extents and the mark spans them.

## A confidence interval, with a center

`range` shows the full spread. For the **uncertainty of the mean**, an error bar,
use `confidence`, which computes the mean's t-interval per group and carries a
center (the mean) with it. The interval then draws a dot at that center: a
pointrange.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length")
```

The level is a parameter (`confidence(0.99)` is wider) and it uses the
t-distribution, so a small sample's interval is honestly wider than a fixed 1.96
would draw. Whether a center *exists* is the statistic's call, not a toggle:
`confidence` has a mean and `range` does not, so choosing the statistic chooses
the geometry, a bare error bar or a pointrange, with no second mark to learn. A
setting can *hide* an existing center ([below](#hiding-the-center-dot)), but none
can conjure one onto `range`. See [Transforms](../transforms.qmd) for the whole
range family.

## Why `interval` needs a transform

`bar` can stand alone, `bar + x + y`, because a bar measures from a baseline the
grammar already knows: zero. An `interval` has no such default. A low *and* a high
are two numbers, and a plain `y()` supplies one, so the mark's minimum syllable
includes a range-producing transform. Ask for an interval without one and the
engine says so, rather than drawing an empty panel:

```{r}
#| error: true
data(iris_flowers) + interval + x(species) + y(petal_length)
```

This is the design choice that keeps the [channel vocabulary](../grammar.qmd)
small. The obvious alternative, `ymin`/`ymax` aesthetics, would add two channels
meaningful to a single mark, and a box plot would need five. Instead the extents
are **invented by a transform**, exactly as `bin` invents a count: the
channel set never grows, and the same
low/high-rows mechanism feeds the [`box`](box.qmd) mark and the
[`ribbon`](ribbon.qmd) band as well. When you have the extents *already*, a
pre-computed error bar from a model or an SEM, [`bounds(lower,
upper)`](../transforms.qmd) supplies them without computing anything:
`interval * bounds(lower, upper)` draws one whisker per row.

## Whiskers on their side

An interval reads its orientation off the bindings, the way a [`bar`](bar.qmd)
and a [`box`](box.qmd) do. Put the categories on `y` and the measured column on
`x`, and the whiskers lie down:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(petal_length) + y(species) +
  x_label("Petal length (cm)") + y_label("Species") +
  title("The same 95% CI, lying down")
```

The caps turn with the whisker, since a cap always crosses the span it ends, and
the center dot rides along. Everything else in this chapter is unchanged:
`caps`, `center`, `pattern`, `color` and `dodge` behave the same whichever axis
carries the categories.

This is the usual shape for a **coefficient plot**, where each row is a named
estimate and the reader scans down a column of intervals. Estimates like these
normally arrive already computed, from a model rather than from raw
observations, which is what [`bounds`](../transforms.qmd) is for: it reshapes two
columns you already have into the low/high pair, computing nothing. With the
terms on `y`, the measured axis is the one `bounds` invents, so there is no `x()`
to write:

```{r}
coefs <- data.frame(
  term = c("Age", "Education", "Experience", "Region: North", "Region: South"),
  lo   = c(0.02, 0.31, 0.11, -0.24, -0.05),
  hi   = c(0.18, 0.55, 0.29,  0.06,  0.21)
)
data(coefs) + interval * bounds(lo, hi) + y(term) +
  x_label("Effect on log wage") +
  title("A coefficient plot, from bounds you already have")
```

Read the two intervals that cross zero as the two effects the data cannot sign.

An interval needs a number to span, so two categorical axes are refused:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(continent)
```

## Splitting by color

`color` splits an interval into one whisker per group, the same discrete split
`line` and `bar` make, and the statistic runs within each group, so every group
gets its own interval, its own hue, and a legend to decode them:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length, by species")
```

When the split is a *second* category, so several whiskers share one x, they
overlap, and [`dodge`](../transforms.qmd) sets them side by side, the same
collision modifier that separates grouped bars and boxes:

```{r}
data(gm_eras) + interval * range * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy range") +
  title("Life expectancy range per continent, 1957 vs 2007")
```

Width and opacity stay *settings*, though; one whisker is a single stroke, so
there is nothing for a per-row `size` or `opacity` to vary along, and `style()`
sets them for the whole layer:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  style(color = "tomato", size = 2.5) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A heavier, one-color range")
```

## Caps, or a bare linerange

The end caps, the short crossbars, are decoration, not data; they only
emphasize where a whisker stops. So whether to draw them is a *setting*:
`style(caps = FALSE)` drops them for a bare linerange, and `caps = TRUE` (the
default) keeps the capped error bar.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(caps = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, drawn as a linerange")
```

Because caps are geometry only `interval` has, `style(caps = )` on any other mark
is refused with direction, the same way a bar's `border_size` is refused on a
`line`. It pairs with the two range statistics and the `color` split, so
error bar, linerange, pointrange, and colored versions of each are all
`interval` plus a transform and a setting, never a new mark to learn.

## Hiding the center dot

A pointrange's dot is the mean; sometimes you want the confidence whiskers
without it, a plain error bar. `style(center = FALSE)` drops the dot, and
`center = TRUE` (the default) keeps it. Like `caps`, it only *hides*: it cannot
add a center to `range`, which has none to draw.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(center = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, without its center dots")
```

`center` and `caps` are the interval's two display toggles, its middle and its
ends, so a confidence interval draws four ways. A capped pointrange is the
default. A capped bare error bar is `center = FALSE`. A linerange with a dot
is `caps = FALSE`, and a plain linerange is both `FALSE`. Like `style(caps = )`,
`style(center = )` belongs to `interval` alone; on any other mark it is refused
with direction.

## A dashed whisker

A whisker is a **stroke**, so it takes `style(pattern = )` the same way a
[`line`](line.qmd) does, the [settable rule](../design-laws.qmd) at work: a dash
patterns any of the path-stroke marks (`line`, `step`, `interval`). The pattern
falls on the whisker span; the caps and center dot stay solid, since a dashed
crossbar reads as noise:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(pattern = "dashed") +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A dashed 95% CI")
```

The five-number summary lives in its own mark, [`box`](box.qmd), and the same
low/high pair drawn as a continuous fill is the [`ribbon`](ribbon.qmd) band. See
[Transforms](../transforms.qmd) for where `range` and `confidence` sit among the
others.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("interval")
```

The last two are the interval's own, and both only ever *hide*. `caps` turns the
end crossbars off, leaving a bare linerange; `center` hides the dot that
`confidence` supplies. Neither can conjure geometry the statistic did not
produce, which is why `style(center = TRUE)` on `interval * range` adds nothing:
a range has no center to draw.

```{r}
data(gm_continents) + interval * confidence(0.99) + x(continent) + y(life) +
  style(caps = FALSE, size = 3, color = "darkslateblue", pattern = "dotted") +
  y_label("Life expectancy") + title("A bare, thick, dotted 99% interval")
```

The statistic carries the other parameter worth knowing: `confidence(0.99)`
widens the interval from the default 0.95. That belongs to the transform rather
than to `style()`, because it changes *what is computed*, not how it is painted.
The test is whether the numbers move: they do here, so it is not a setting.

## What it refuses

An interval is a stroke, however much it looks like a bar, so its pattern is a
dash and a hatch is refused:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  style(pattern = "hatch")
```

That is worth meeting, because the eye reads a capped range as a shape with an
inside. The engine reads it as line-work, and the dashed whisker above is the
same fact arrived at from the other side. Its second refusal, a bare `interval`
with no transform to give it extents, opens the chapter.
