# Interval

```{r}
#| include: false
source("../R/setup.R")
```

How far does each group reach, from its lowest value to its highest? `interval`
draws a **whisker** for each group: a span from a low value to a high one, with
a short line across each end called a **cap**. When the whisker shows a
statistic with a center, such as a mean, a dot marks that center. `interval` is
the mark that draws error bars and min–max ranges. A `bar` measures from a
baseline up to a value. An `interval` has no baseline, so it needs both a low
value and a high value before it can draw. Whiskers are vertical when the
categories are on `x` and horizontal when they are on `y`; see [Horizontal
whiskers](#horizontal-whiskers).

Almost everyone calls this an **error bar**, and that name does not say what the
whisker measures. Three different quantities are drawn as the same picture: a
standard deviation, a standard error, and a confidence interval (CI). On the same
data, one can be several times longer than another. Nothing in the drawing says
which one it is, and a reader without a caption cannot know.

gog makes the sentence name the statistic instead. `interval * range` spans the
observed extremes. `interval * confidence(0.95)` computes a confidence interval
and states its level in the sentence. `interval * bounds(lower, upper)` reads
two columns you computed yourself. The mark is the same whisker in all three;
what differs is the transform, and the transform is written down. That is the
law [Explicit Over Implicit](../design-laws.qmd#law-5-explicit-over-implicit),
applied to one mark. The name *error bar* never says which statistic was drawn.
A written sentence has to say it.

## A range per group

How far do the petal lengths of each species reach, from the shortest to the
longest? One whisker per species answers it, and a whisker needs a low value and
a high value to span. The low value and the high value come from a transform,
not from two extra channels. `range` reduces `y` to its minimum and maximum
within each `x` group, and `interval` spans between them:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Range of petal length per species")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal
length."*

`interval * range` has the same shape as `bar * mean`. The difference is what
the transform returns: `mean` returns one number per group, and `range` returns
two. You name only the column whose spread you are asking about. The transform
computes its two **extents**, the low value and the high value, and the mark
spans them.

`range` does not belong to `interval` alone. Three other marks read the same low
and high values, and only the shape changes:

| Mark derived by `range` | Shape |
|---|---|
| `ribbon * range` | the band between the two values, filled |
| `line * range` | the two values as two separate strokes |
| `step * range` | the same two boundaries, as staircases |
| `interval * range` | one whisker per group |

`range` is one of five **pair transforms**. A pair transform supplies a low
value and a high value, where most transforms supply a single number. The other
four are `confidence`, `deviation`, `smooth_band` and `bounds`, and every mark in
that table takes all five.

## The middle half of each group

A minimum and a maximum each come from a single row, so one unusual flower can
decide where a whisker ends. To show the middle of a group instead of its
extremes,
give `range` two **quantiles**:

```{r}
data(iris_flowers) + interval * range(0.25, 0.75) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Middle half of petal length per species")
```

*"Given the iris flowers: intervals derived by range from 0.25 to 0.75, x is
species, y is petal length."*

This spans the first quartile to the third, which is the **interquartile
range**. `range(0.1, 0.9)` keeps the middle 80%, and any other pair works the
same way. `range` written with no arguments means `range(0, 1)`, so it still
spans from the minimum to the maximum.
[Transforms](../transforms.qmd#a-band-between-two-quantiles) explains the two
arguments in full.

## Three layers instead of one mark

You now have two ranges of the same groups, the extremes and the middle half. A
[`box`](box.qmd) draws a similar summary, five numbers in one mark, and you
cannot remove one of them. Draw the middle half and the full range as two
layers, and the median as a third. You can give any of the three layers its
own style, drop one, or replace one:

```{r}
data(iris_flowers) +
  interval * range + style(color = "darkgray") +
  interval * range(0.25, 0.75) + style(color = "steelblue", size = 8) +
  point * median + style(color = "gold", size = 5, opacity = 1) +
  x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Petal length per species, drawn in three layers")
```

*"Given the iris flowers: intervals derived by range, and also intervals derived
by range from 0.25 to 0.75, and also points derived by median, x is species, y
is petal length."*

The gray whisker is the full range, the thick blue band is the middle half, and
the gold dot is the median. The order of the layers decides what stays
visible. A later layer is drawn over an earlier one, so the median is written
last. Move it to the front of the sentence and the band draws over it, and
almost none of the dot is left. If you have used ggplot2, this is the rule you
already know: a geom added later draws on top. The dot is set opaque as well,
since a point is slightly translucent by default.

The three layers name the three statistics they draw. A box plot draws similar
information and names none of it. It differs from these layers at the ends of
the whiskers. A `box` stops its whiskers at a cutoff computed
from the quartiles, and draws every point beyond that cutoff as an outlier. A
`box` on this same data draws outlier dots for setosa and versicolor. They are
absent here, because these whiskers run to the true minimum and maximum.

Because each statistic is its own layer, you can change one and leave the
others as they are. Widen the band to `range(0.1, 0.9)` and the plot reports a
different statistic, with the same marks. Drop the gray layer and the y axis is
scaled to the band rather than to the extremes. Add another layer with a
different pair of quantiles, and you have a
[fan chart](ribbon.qmd#a-fan-chart-of-nested-bands): nested spans for each
category, instead of nested bands along a continuous axis. A `box` is still
the shorter sentence when you want the five numbers and the outlier rule
together.

## A confidence interval, with a center

How well is the average petal length of each species known? That is a different
question from how far the flowers spread, and it needs a different statistic.
`range` shows the full spread. For the **uncertainty of the mean**, use
`confidence`. It computes a **t-interval** for each group's mean. That is the
confidence interval named at the start of this chapter, and it widens when the
group is small. `confidence` also keeps the mean as the center of the span, and
the mark draws a dot there. A whisker with a dot at its center is a
**pointrange**.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length."*

The level is `confidence`'s only argument, and it is 0.95 when you leave it out.
A higher level draws a longer whisker, so `confidence(0.99)` is longer than
`confidence(0.95)`. The interval is always a t-interval. For a z-interval, a
bootstrap, or the standard errors a model already gave you, compute the two
numbers yourself and pass them to
[`bounds`](../transforms.qmd#bounds-a-pre-computed-lowhigh-pair).

## The spread of the data

How far is a typical flower from the average of its species? That is a question
about the spread of the data, not about how well the average is known. The
opening of this chapter named three quantities: a standard deviation, a standard
error and a confidence interval. `confidence` draws the confidence interval. The
standard deviation has a transform of its own, `deviation`:

```{r}
data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Mean petal length, with one standard deviation")
```

*"Given the iris flowers: intervals derived by deviation, x is species, y is
petal length."*

`deviation` with no argument reaches one standard deviation on each side of the
mean. `deviation(2)` reaches two, so that whisker is four standard deviations
long. Like `confidence`, it computes a mean as its center, so it draws a
pointrange rather than a bare error bar.

The standard error has no transform of its own, and it does not need one. It is
the standard deviation divided by the square root of the group size.
`confidence` computes it internally and reports the interval built from it,
which is what a reader of the plot needs. To draw one standard error on each
side of the mean, compute the two ends yourself and pass them to `bounds`.

`deviation` and `confidence` look alike and answer different questions. `|` puts
two plots side by side, and because both name `petal_length` on `y` they share
that scale, drawn once down the left. [Composition](../composition.qmd) explains
when two plots share a scale. Here are the two statistics on the same fifty
flowers per species:

```{r}
(data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
   y_label("Petal length (cm)") + title("Spread of the data")) |
  (data(iris_flowers) + interval * confidence + x(species) + y(petal_length) +
     y_label("") + title("Uncertainty of the mean"))
```

*"Given the iris flowers: intervals derived by deviation, x is species, y is
petal length, beside intervals derived by confidence, x is species, y is petal
length."*

The whiskers on the right are less than a third as long, and the data is
identical. The ratio is the same in all three species. The two statistics do
not contradict each other. `deviation` describes how far the flowers are from
their average, and measuring more of them does not make that whisker shorter.
`confidence` describes how precisely the average itself is known. More flowers
make its whisker shorter: four times as many flowers make it half as long. So
the ratio between the two whiskers comes from the group size. With only a few
flowers per species, the confidence whisker can be the longer one, because a
t-interval widens when the group is small.

So choose the statistic by asking which caption you would write under the plot.
*"A versicolor petal is usually within half a centimeter of its species
average"* describes the left panel. *"The average versicolor petal is 4.26 cm,
and the interval reaches 0.13 cm on either side"* describes the right one.
Reporting one and meaning the other is the mistake the name *error bar* allows,
and the reason gog requires the statistic in the sentence.

## Why `interval` needs a transform

Every plot so far has named a transform. A mark needs one when it draws a
number that nothing else supplies. A bar needs no transform, because the
grammar already supplies its baseline: zero. `bar + x + y` draws. An `interval`
has no such default. A low *and* a high are two numbers, and a plain `y()`
supplies one. So the shortest interval that will draw is a mark, its positions,
and a pair transform. That is its
[minimum syllable](../design-laws.qmd#law-7-minimum-syllable). Ask for an
interval without a transform and the engine refuses the sentence rather than
drawing an empty panel. [What it refuses](#what-it-refuses) shows that refusal.

This is the design choice that keeps the [channel
vocabulary](../grammar.qmd#channels-a-column-made-visible) small. The obvious
alternative would add `ymin` and `ymax` as two more channels, useful to only a
few marks, and a box plot would then need five. Instead a transform computes
the extents, the way `mean` computes one value per group, so the channel set
never grows. [`ribbon`](ribbon.qmd) reads the same low and high values, and
[`box`](box.qmd) computes its five numbers inside the mark. The pair transform
[`bounds(lower, upper)`](../transforms.qmd#bounds-a-pre-computed-lowhigh-pair)
takes a low and a high you already have as two columns of the table, and
[Horizontal whiskers](#horizontal-whiskers) draws a coefficient plot with it.

## Horizontal whiskers

Whiskers turn horizontal for the reasons bars do: the names are long, or there
are many of them. A tall list of horizontal whiskers is then easier to read
than a crowded row of vertical ones. An interval takes its orientation from the
bindings, the way a [`bar`](bar.qmd) and a [`box`](box.qmd) do. Put the
categories on `y` and the measured column on `x`:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(petal_length) + y(species) +
  x_label("Petal length (cm)") + y_label("Species") +
  title("The same 95% CI, horizontal")
```

The caps turn too, since a cap always crosses the span at its end, and the
center dot stays at the center. The sections after this one work the same way
on horizontal whiskers. `color`, `dodge`, `caps`, `center` and `pattern`, all
introduced below, behave the same whichever axis carries the categories.

This is the usual shape for a **coefficient plot**, where each row is a named
estimate and the reader reads the intervals from top to bottom. Estimates like
these normally arrive already computed, from a model rather than from raw
observations. That is what
[`bounds`](../transforms.qmd#bounds-a-pre-computed-lowhigh-pair) is for: it
reads two columns you already have as the low and the high, and computes
nothing. The `term` column goes on `y`. In `coefs`, `lo` and `hi` are the ends
of a 95% confidence interval for each term. `bounds` supplies the values for
the other axis, so there is no `x()` to write:

```{r}
#| echo: false
peek(coefs)
```

```{r}
data(coefs) + interval * bounds(lo, hi) + y(term) +
  x_label("Effect on log wage") +
  title("A coefficient plot, from bounds you already have")
```

*"Given the coefs table: intervals from lo to hi, y is term."*

The two Region intervals, North and South, run from a negative low to a
positive high. Each one includes zero, so this data cannot say whether either
effect is positive or negative.

An interval needs a number to span, so two categorical axes are refused, as
they are on a bar and a box. [What it refuses](#what-it-refuses) shows that
refusal.

## Splitting by color

With color, a reader can see which whisker is which without reading the axis.
`color` splits an interval into one whisker per group, the same categorical
split `line` and `bar` make. The statistic is computed inside each group, so
every group gets its own interval. Here `color` repeats the species already on
`x`, so each species gets its own hue, and one legend names them all:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length, by species")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species."*

A split by a *second* category puts several whiskers on the same category of
`x`, and they overlap.
[`dodge`](../transforms.qmd#dodge-side-by-side-instead-of-on-top) sets them side
by side instead. It is the same collision modifier that separates grouped bars
and boxes:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + interval * range * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy range") +
  title("Life expectancy range per continent, 1957 vs 2007")
```

*"Given the gapminder eras: intervals derived by range and dodge, x is
continent, y is life, color by era."*

Oceania's 1957 whisker looks like a single cap. The table has only two
countries there, Australia and New Zealand. In 1957 their life expectancies
were within a tenth of a year of each other. That span is shorter than a pixel,
so the two caps draw on top of each other. In 2007 the two countries were a
year apart, and the caps separate.

Color splits an interval into groups, so each whisker has one color. `size` and
`opacity` cannot be mapped on an interval, and they stay *settings*. A whisker
is one **stroke** drawn from a whole group, so no single row can give it a
width or an opacity. `style()` sets them once for the layer:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  style(color = "tomato", size = 2.5) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A heavier, one-color range")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal
length, colored tomato, with size 2.5."*

Every whisker is drawn in tomato and 2.5 pixels wide, whatever its species.

## Caps, or a linerange

The caps are the short lines across each end of a whisker. They are decoration
rather than data, because they only mark where a whisker stops. A crowded plot
is often easier to read without them. So whether to draw them is a *setting*.
`style(caps = FALSE)` removes them, and a whisker with no caps is a
**linerange**. `caps = TRUE`, the default, keeps the capped error bar. Here is
the confidence interval from [Splitting by color](#splitting-by-color) with its
caps off:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(caps = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, drawn as a linerange")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with caps off."*

`caps` belongs to the two marks that draw whiskers, `interval` and
[`box`](box.qmd). On any other mark, `style(caps = )` is refused, and the
message says what to use instead. A bar's `border_size` is
refused on a `line` the same way. `caps` combines with all five pair
transforms. So the error bar, the linerange and the pointrange are all the same
mark. Only the transform and one setting change.

## Hiding the center dot

Whether a center exists is decided by the statistic, not by a setting.
`confidence` and `deviation` compute a mean as their center, and `smooth_band`
computes the fitted value. `range` and `bounds` compute none. So the first three
draw a pointrange, and `range` and `bounds` draw a bare error bar. A second
layer can add a center where the statistic has none. `point * mean` over
`interval * range` draws a pointrange from two layers, as [Three layers instead
of one mark](#three-layers-instead-of-one-mark) shows with the median.

A setting can only hide a center that exists. A pointrange's dot is the center
the statistic computed.
Sometimes you want the whisker without it. `style(center = FALSE)` drops the
dot and leaves a bare error bar. `center = TRUE`, the default, keeps it:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(center = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, without its center dots")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with center off."*

`center` and `caps` each show or hide one part of an interval: `center` its
middle dot, `caps` its ends. Each takes two values, so a confidence interval can be
drawn in four ways. A capped pointrange is the default. A capped bare error bar
is `center = FALSE`. A linerange with a dot is `caps = FALSE`. Setting both to
`FALSE` gives a bare linerange. `style(center = )` belongs to `interval` alone,
because a box has no center dot: its middle is the median line. On any other mark
it is refused, and the message says what to use instead.

## A dashed whisker

A dash separates one layer from another without a second color, and it still
separates them when the plot is printed in black and white. A whisker is a
stroke, so it takes `style(pattern = )` the same way a [`line`](line.qmd) does.
This is the [settable rule](../design-laws.qmd#the-settable-rule): a dash is
the pattern on every path-stroke mark (`line`, `step`, `interval`, `path`,
`rule`, `edge`). The dash is drawn on the whisker span only. The caps and the
center dot stay solid, since a dashed cap would not look like a cap and a dot
has no line to dash:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(pattern = "dashed") +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A dashed 95% CI")
```

*"Given the iris flowers: intervals derived by confidence at 95%, x is species,
y is petal length, color by species, with pattern dashed."*

`pattern` can also be mapped to a column. Like `color`, it splits an interval
into one whisker per group, and each group's whisker gets its own dash. Mapped
to a second category, `pattern` puts several whiskers on one category of `x`,
where they overlap. `dodge` sets them side by side, as it does for [a split by
color](#splitting-by-color).

## A bar with its error bar

Two marks can answer two questions about the same groups. A `bar` shows where
each group's mean sits. An `interval` shows how well that mean is known. Layer
them and you have a plot that appears throughout the social sciences, usually
with a bare error bar rather than a pointrange:

```{r}
data(iris_flowers) + bar * mean + style(color = "lightgray") +
  interval * confidence(0.95) + style(center = FALSE, color = "black") +
  x(species) + y(petal_length) +
  x_label("Species") + y_label("Mean petal length (cm)") +
  title("Mean petal length, with a 95% confidence interval")
```

*"Given the iris flowers: bars derived by mean, colored lightgray, and also
intervals derived by confidence at 95%, with center off, colored black, x is
species, y is petal length."*

Each layer names its own statistic. `bar * mean` reduces each group to one
number, and `interval * confidence(0.95)` computes the interval around that
number. Both read the same column and report different things, and the sentence
says which is which. The dot is turned off because it would sit exactly on the
top of the bar, which already marks the mean.

A bar with an error bar is where the three quantities named at the start of
this chapter are most often confused, and nothing in the picture separates
them. In gog the sentence names the statistic: this one says `confidence(0.95)`,
and `deviation` in its place would give a different sentence and a much longer
whisker.

The groups of a second category go side by side, with the same `dodge` that
separates grouped bars. Bind `color` to the bars and `group` to the whiskers.
`group` splits the whiskers into the same slots as the bars but draws no legend
of its own, so the plot keeps one:

```{r}
data(gm_eras) +
  bar * mean * dodge + color(era) +
  interval * confidence(0.95) * dodge + group(era) +
  style(center = FALSE, color = "black") +
  x(continent) + y(life) +
  y_label("Mean life expectancy (years)") +
  title("Mean life expectancy by continent, 1957 and 2007")
```

*"Given the gapminder eras: bars derived by mean and dodge, color by era, and
also intervals derived by confidence at 95% and dodge, grouped by era, with
center off, colored black, x is continent, y is life."*

Oceania has both the shortest whisker and the longest, from the same two
countries. A confidence interval from two values is wide compared with the
distance between them, because two values say little about the spread. In 2007
Australia and New Zealand were a year apart, and the interval is thirteen years
wide. In 1957 the two countries were 0.07 years apart, and the confidence
interval is still nine tenths of a year wide. That is more than twelve times
the 1957 range, which showed as a single cap in
[Splitting by color](#splitting-by-color).

What this plot does not show is the data. Two groups can have the same mean and
the same interval and still hold very different values, and a bar reports only
the one number. [Three layers instead of one
mark](#three-layers-instead-of-one-mark) draws the spread itself, and
[`box`](box.qmd) draws five numbers at once.

## What you can set

A setting changes how every whisker looks without reading a column, and each
mark takes its own. These are an interval's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("interval")
```

An interval takes the four settings of a [`line`](line.qmd), because a whisker
is a stroke. It also takes two more, `caps` and `center`. The plot below gives
several settings at once. It reads `gm_continents`: the Americas, Asia and
Europe, measured every five years from 1952 to 2007, so each interval is
computed from every country in that continent, across all twelve years.

```{r}
#| echo: false
peek(gm_continents)
```

```{r}
data(gm_continents) + interval * confidence(0.99) + x(continent) + y(life) +
  style(caps = FALSE, size = 3, color = "darkslateblue", pattern = "dotted") +
  y_label("Life expectancy") + title("A thick, dotted 99% linerange")
```

The confidence level is not in the settings table, because it is not a
setting. It is written into the transform instead. `confidence(0.99)` changes
*what is computed*, not how it is drawn. To decide whether something is a
setting, ask whether it changes the computed numbers. The level does, so it
belongs to the transform.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

An interval is a stroke, not a filled shape, so its pattern is a dash and a
hatch is refused:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  style(pattern = "hatch")
```

You might expect a hatch to work, because a capped range looks like a shape
with an inside. The engine draws it as strokes, and a stroke takes a dash.

An `interval` with no transform is the second refusal. A plain `y()` supplies
one number, and
a span needs two, as [Why `interval` needs a
transform](#why-interval-needs-a-transform) explained. The engine says so
rather than drawing an empty panel:

```{r}
#| error: true
data(iris_flowers) + interval + x(species) + y(petal_length)
```

The message calls them range transforms. It names two of the four, `range` and
`bounds`, and writes out a sentence the engine would accept.

Two categorical axes are the third refusal. An interval needs one continuous
axis, because that column is where the low and the high come from:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(continent)
```

Two categories leave the mark nothing to span.

## In short

An interval computes nothing. Its transform decides which two numbers it spans,
and whether there is a center to draw. `caps` and `center` are settings, so
neither reads a column. A dash can be set for the whole layer, or mapped to a
column, which splits the whiskers the way `color` does.

## What comes next

Which two numbers an interval spans is the transform's answer, never the
mark's. [Box](box.qmd) answers it inside the mark: it computes the quartiles
and the median itself, and draws them as a box and two whiskers.
