# Interval

```{r}
#| include: false
source("../R/setup.R")
```

How far does each group reach, from its lowest value to its highest? `interval`
draws a span from a low value to a high one at each position: a whisker,
capped at both ends. It is the mark behind error bars and min–max ranges. Where
`bar` measures *from a baseline* up to a value, an `interval` **spans a low
and a high**: it has no baseline, and it needs both before it can draw
anything at all. Whiskers stand up by default and lie down when the
categories are on `y`; see [Whiskers on their side](#whiskers-on-their-side).

Almost everyone calls this an **error bar**, and the name hides a choice no
drawing can show. Three different quantities get drawn as the same picture, a
standard deviation, a standard error, and a confidence interval, and on the same
data they differ by large factors. Nothing in the drawing says which one it is, so
a reader without the caption is guessing at a number's meaning.

gog makes the sentence say it instead. `interval * range` spans the observed
extremes, `interval * confidence(0.95)` computes the interval and names its
level, and `interval * bounds(lower, upper)` takes two columns you computed
yourself. The mark is the same whisker in all three; what differs is the
transform, and the transform is written down. That is Explicit Over Implicit,
shown in one mark: the name *error bar* can hide the choice, and a written
sentence cannot hide it.

## A range per group

Those two extents come from a transform, not from two extra channels. `range`
reduces `y` to its minimum and maximum within each `x` group, and `interval`
spans between them:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Range of petal length per species")
```

*"Given the iris flowers: intervals derived by range, x is species, y is petal length."*

Read it exactly like `bar * mean`. Here the transform gives a range, while
`mean` gives one number. You name the one variable whose spread you are asking about;
the transform invents the two extents and the mark spans them.

## The middle half of each group

A minimum and a maximum each come from a single row, so one unusual flower can
set the whole whisker. To show the middle of a group instead of its extremes,
give `range` two **quantiles**:

```{r}
data(iris_flowers) + interval * range(0.25, 0.75) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Middle half of petal length per species")
```

This spans the first quartile to the third, which is the **interquartile range**.
`range(0.1, 0.9)` keeps the middle 80%, and any other pair works the same
way. Bare `range` is `range(0, 1)`, so the plain sentence above is unchanged.
[Transforms](../transforms.qmd#a-band-between-two-quantiles) gives the full
account of the two arguments.

## Three layers instead of one mark

A [`box`](box.qmd) draws five numbers in one mark, and you take all five or none.
Draw the band and the full range as two layers, and the median as a third. Style,
drop, or replace any of them:

```{r}
data(iris_flowers) +
  interval * range + style(color = "lightgray") +
  interval * range(0.25, 0.75) + style(color = "steelblue", size = 8) +
  point * median + style(color = "white", size = 3) +
  x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Petal length per species, drawn in three layers")
```

The gray whisker is the full range, the thick blue band is the middle half, and
the white dot is the median. The three layers name the three statistics. A box
plot draws the same three and names none of them.

This is not a `box` redrawn, and the difference is at the ends. A `box` stops
its whiskers at Tukey's fences and draws the points beyond as outliers.
Setosa then shows four dots that are absent here, because these whiskers run
to the true minimum and maximum. What the layers give you instead is a plot you can change. Widen the
band to `range(0.1, 0.9)` and the picture changes meaning without changing marks.
Drop the gray layer and the y axis fits the band rather than the extremes. Add a
third layer with another pair, and you have the [fan
chart](ribbon.qmd#a-fan-chart-of-nested-bands) at categories rather than along a
line. A `box` is still the shorter sentence when you want the five numbers and
the outlier rule together.

## A confidence interval, with a center

`range` shows the full spread. For the **uncertainty of the mean**, use
`confidence`. It computes a t-interval for each group's mean (a t-interval
widens when the group is small) and keeps that mean as the center. The interval then draws a dot at that center: a
pointrange.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length")
```

The level is a parameter, so `confidence(0.99)` draws wider. It uses the
t-distribution, which widens the interval further when the group is small.

## The spread of the data

The opening of this chapter named a standard deviation, a standard error and a
confidence interval. `confidence` draws the third. The first is `deviation`:

```{r}
data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Mean petal length, with one standard deviation")
```

`deviation(2)` reaches two standard deviations on each side of the mean, so that
band is four standard deviations wide. The default is one each side. Like
`confidence` it carries a center, so it draws a pointrange rather than a bare
whisker.

The standard error has no transform of its own, and that is deliberate.
`confidence` computes it internally and reports the interval, which is what a
reader needs from it.

`deviation` and `confidence` look alike and answer different questions, so here
they are over the same fifty flowers per species, sharing one y scale:

```{r}
(data(iris_flowers) + interval * deviation + x(species) + y(petal_length) +
   y_label("Petal length (cm)") + title("Spread of the data")) |
  (data(iris_flowers) + interval * confidence + x(species) + y(petal_length) +
     y_label("") + title("Uncertainty of the mean"))
```

The bands on the right are less than a third as long, on identical data, and the
ratio is the same in all three species. The two numbers do not contradict each
other. `deviation` describes how far the flowers are from their average, and
measuring more of them does not shrink it. `confidence` describes how precisely
the average itself is known. More flowers shrink it: four times as many halves
it. Fifty flowers per species is what sets that ratio here.

So choose by the sentence you would write under the plot. *"A versicolor petal is
usually within half a centimeter of its species average"* describes the left
panel. *"The average versicolor petal is 4.26 cm, and the interval reaches 0.13
either side"* describes the right one. Reporting one and meaning the other is the mistake the name
*error bar* allows, and the reason gog asks you to name the statistic.

Which band is wider is not fixed, either. `confidence` narrows as the group gets
larger. On fifty flowers it is under a third of the spread; on three flowers it
is the wider of the two.

Whether a center exists is decided by the statistic, not by a setting.
`confidence` and `deviation` have a mean and `range` does not, so the statistic
chooses the geometry: a bare error bar or a pointrange. A setting can *hide* an
existing center ([below](#hiding-the-center-dot)), but no setting adds one to
`range`. See [Transforms](../transforms.qmd) for the whole range family.

## Why `interval` needs a transform

`bar` needs no transform, `bar + x + y`, because a bar measures from a
baseline the grammar already supplies: zero. An `interval` has no such default. A low *and* a high
are two numbers, and a plain `y()` supplies one, so the mark's minimum syllable
includes a range-producing transform. Ask for an interval without one and the
engine says so, rather than drawing an empty panel:

```{r}
#| error: true
data(iris_flowers) + interval + x(species) + y(petal_length)
```

This is the design choice that keeps the [channel vocabulary](../grammar.qmd)
small. The obvious alternative would add `ymin` and `ymax` as two more
channels meaningful to a single mark, and a box plot would need five. Instead
the extents are **invented by a transform**, exactly as `bin` invents a
count, so the channel set never grows; the same low and high rows are what
[`box`](box.qmd) and [`ribbon`](ribbon.qmd) draw from. When you already have
the extents, from a fitted model or a published standard error,
[`bounds(lower, upper)`](../transforms.qmd) supplies them without computing
anything:
`interval * bounds(lower, upper)` draws one whisker per row.

## Whiskers on their side

An interval reads its orientation off the bindings, the way a [`bar`](bar.qmd)
and a [`box`](box.qmd) do. Put the categories on `y` and the measured column on
`x`, and the whiskers lie down:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(petal_length) + y(species) +
  x_label("Petal length (cm)") + y_label("Species") +
  title("The same 95% CI, lying down")
```

The caps turn with the whisker, since a cap always crosses the span it ends,
and the center dot stays at the center. Everything else in this chapter is unchanged:
`caps`, `center`, `pattern`, `color` and `dodge` behave the same whichever axis
carries the categories.

This is the usual shape for a **coefficient plot**, where each row is a named
estimate and the reader scans down a column of intervals. Estimates like these
normally arrive already computed, from a model rather than from raw
observations, which is what [`bounds`](../transforms.qmd) is for: it reshapes two
columns you already have into the low/high pair, computing nothing. With the
terms on `y`, the measured axis is the one `bounds` invents, so there is no `x()`
to write:

```{r}
coefs <- data.frame(
  term = c("Age", "Education", "Experience", "Region: North", "Region: South"),
  lo   = c(0.02, 0.31, 0.11, -0.24, -0.05),
  hi   = c(0.18, 0.55, 0.29,  0.06,  0.21)
)
data(coefs) + interval * bounds(lo, hi) + y(term) +
  x_label("Effect on log wage") +
  title("A coefficient plot, from bounds you already have")
```

The two intervals that cross zero cover positive and negative values alike,
so those two effects could go either way.

An interval needs a number to span, so two categorical axes are refused:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(continent)
```

## Splitting by color

`color` splits an interval into one whisker per group, the same discrete
split `line` and `bar` make. The statistic is computed inside each group, so
each gets its own interval, its own hue, and a legend:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length, by species")
```

When the split is a *second* category, so several whiskers share one x, they
overlap, and [`dodge`](../transforms.qmd) sets them side by side, the same
collision modifier that separates grouped bars and boxes:

```{r}
data(gm_eras) + interval * range * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy range") +
  title("Life expectancy range per continent, 1957 vs 2007")
```

Width and opacity stay *settings*. One whisker is a single stroke, so there
is no row for `size` or `opacity` to vary over, and `style()` sets them for
the layer:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  style(color = "tomato", size = 2.5) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A heavier, one-color range")
```

## Caps, or a bare linerange

The end caps, the short crossbars, are decoration, not data; they only
emphasize where a whisker stops. So whether to draw them is a *setting*:
`style(caps = FALSE)` drops them for a bare linerange, and `caps = TRUE` (the
default) keeps the capped error bar.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(caps = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, drawn as a linerange")
```

Because caps are geometry only `interval` has, `style(caps = )` on any other mark
is refused with direction, the same way a bar's `border_size` is refused on a
`line`. `caps` combines with the two range statistics and the `color` split,
so error bar, linerange, pointrange, and colored versions of each are
`interval` plus a transform and a setting, never a new mark to learn.

## Hiding the center dot

A pointrange's dot is the mean; sometimes you want the confidence whiskers
without it, a plain error bar. `style(center = FALSE)` drops the dot, and
`center = TRUE` (the default) keeps it. Like `caps`, it only *hides*: it cannot
add a center to `range`, which has none to draw.

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(center = FALSE) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("The same CI, without its center dots")
```

`center` and `caps` are the interval's two switches, its middle and its
ends, so a confidence interval draws four ways. A capped pointrange is the
default. A capped bare error bar is `center = FALSE`. A linerange with a dot
is `caps = FALSE`, and a plain linerange is both `FALSE`. Like `style(caps = )`,
`style(center = )` belongs to `interval` alone; on any other mark it is refused
with direction.

## A dashed whisker

A whisker is a **stroke**, so it takes `style(pattern = )` the same way a
[`line`](line.qmd) does, the [settable rule](../design-laws.qmd) in action: a
dash patterns every path-stroke mark (`line`, `step`, `interval`, `path`,
`rule`). The pattern falls on the whisker span; the caps and center dot stay
solid, since a dashed crossbar is hard to read as a cap:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  color(species) + style(pattern = "dashed") +
  x_label("Species") + y_label("Petal length (cm)") +
  title("A dashed 95% CI")
```

The five-number summary has its own mark, [`box`](box.qmd), and the same
low and high pair drawn as a continuous fill is the [`ribbon`](ribbon.qmd)
band. See
[Transforms](../transforms.qmd) for where `range`, `confidence` and `deviation`
sit among the others.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("interval")
```

The last two are the interval's own, and both only ever *hide*. `caps` turns the
end crossbars off, leaving a bare linerange; `center` hides the dot that
`confidence` supplies. Neither can add geometry the statistic did not
produce, so `style(center = TRUE)` on `interval * range` changes nothing: a
range has no center.

```{r}
data(gm_continents) + interval * confidence(0.99) + x(continent) + y(life) +
  style(caps = FALSE, size = 3, color = "darkslateblue", pattern = "dotted") +
  y_label("Life expectancy") + title("A bare, thick, dotted 99% interval")
```

The statistic carries the other parameter: `confidence(0.99)` widens the
interval from the default 0.95. That belongs to the transform rather than to
`style()`, because it changes *what is computed*, not how it is drawn. The
test is whether the numbers change: they do here, so it is not a setting.

## What it refuses

An interval is a stroke, however much it looks like a bar, so its pattern is a
dash and a hatch is refused:

```{r}
#| error: true
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  style(pattern = "hatch")
```

That refusal deserves a look, because the eye reads a capped range as a shape
with an inside. The engine draws it as a stroke, which is why the whisker
takes a dash and not a hatch. Its second refusal, a bare `interval` with no
transform to give it extents, is shown under
[Why `interval` needs a transform](#why-interval-needs-a-transform).
