# Line

```{r}
#| include: false
source("../R/setup.R")
```

How did one value change over time? `line` connects the rows in **x order**,
drawing a polyline through each `(x, y)` point.

## Basic usage

```{r}
data(gapminder_asia) + line + x(year) + y(life)
```

*"Given gapminder Asia: a line, x is year, y is life."*

## You do not have to sort first

`line` sorts its own vertices by `x`, so the rows may arrive in any order. The
frame below is deliberately shuffled, and the line still runs left to right:

```{r}
scrambled <- data.frame(year = c(1992, 1962, 2002, 1972, 1982),
                        life  = c(  68,   52,   71,   58,   63))

data(scrambled) + line + x(year) + y(life)
```

That is a property of the mark rather than a convenience. A line says *this value
changed along this axis*, and an axis has one direction, so there is only one order
the vertices could be joined in.

[`path`](path.qmd) is the mark that does the opposite. It joins the rows in the
order your table holds them, which is what draws a route or a loop. So choosing
between the two is choosing whether the order of your rows carries meaning: with
`line` it cannot, and with `path` it is the whole point.

## Multiple series

Use `color` to split into one line per group:

```{r}
data(gapminder_asia) + line + x(year) + y(life) +
  color(country)
```

Each unique value in `country` becomes a separate line with its own color.
A color legend is generated automatically.

## Layering line over other marks

`line` and `point` are often combined: the point layer adds
visible markers at each data position.
Write `color` *before* the marks and it belongs to the plot, so both layers
take it:

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  line + point
```

Both the lines and the points share the same color per country.

### Giving each layer its own appearance

`style()` attaches to the nearest preceding mark and **stops there**, so each
layer can be styled independently:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "steelblue") +
  point + style(color = "firebrick", size = 3) +
  title("Blue line, red points: styled per layer")
```

Channels scope the same way. Written after a mark, one binds to that mark alone,
so each layer can carry a different encoding:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population) +
  title("Color on the lines, size on the points")
```

That works because channels bind **forward only**. A channel that also reached
backwards would put `size` on the `line`, which has no size, and the plot would
be refused with no way to say what you meant. See
[Encoding scope](../encoding-scope.qmd).

## A category on the domain: the profile

A line's two axes are not the same kind of thing. `x` is the **domain**, the axis
the path is read along; `y` is the **measure**, the quantity it traces. The domain
takes a category as readily as a number, and connecting one summary per category
is the profile plot:

```{r}
data(gapminder_2007) + line * mean + point * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("line * mean + x(continent): one value per category, joined")
```

The transform is doing real work there. Without one, a `line` would connect every
row in the frame, and 142 countries sharing five slots is a scribble rather than a
profile; the engine says so on stderr when it sees a line with many ungrouped
rows. `mean` leaves exactly one point per category, which is what makes the
connected path mean anything.

Whether the connection is honest is a question about your data, and it is
worth asking. The segment from Africa to the Americas crosses ground where
nothing was measured. Read the vertices, and the segments guide the eye down a
ranking; read the segments, and they assert a change over a distance that does
not exist. The grammar draws what you said and leaves the judgment with you.
Two cases make the judgment easy. An ordered category, such as a grade, a size
band or a survey scale, genuinely has a direction for the path to run along. A
**circular** one has no ends at all, which is the plot the next section is
really about.

Because it is the same sentence, the same profile bends into a circle. That is
the radar, and it needs no new atom:

```{r}
data(gapminder_2007) + line * mean + x(continent) + y(life) + polar() +
  title("line * mean + polar(): the radar")
```

Look at the segment from Oceania back to Asia. On the flat profile the path has
two ends, because a straight axis has two ends; in the circle the categories fill
the turn between them and the last is next to the first, so the path closes.
Nothing was added to say so: a categorical angular axis wraps, and the mark reads
that off the space it is drawn in. Full treatment in
[Polar](../polar.qmd).

A categorical **measure** is a different matter and is refused, because a line
traces a quantity and category names are not one:

```{r}
#| error: true
render_svg(data(gapminder_2007) + line * mean + x(life) + y(continent))
```

## Smoothed trend line

Combine `line` with the `smooth` transform to draw a fitted trend
instead of raw data:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

## Two boundaries from a pair

Most transforms give a `line` one value per x. A **pair** transform
([`range`](../transforms.qmd), `confidence`, or `bounds`) gives it two (a low and a
high), and a `line` then draws **both boundaries** as separate curves rather than
connecting them in a zigzag. It is the *unfilled* counterpart to a
[`ribbon`](ribbon.qmd) filling the same pair. Here the min–max envelope of five
countries, as two lines:

```{r}
data(gapminder_asia) + line * range + x(year) + y(life) +
  y_label("Life expectancy") + title("A min–max envelope, as two boundary lines")
```

Add `style(pattern = "dashed")` for the dashed ± band publications use; see
[Ribbon](ribbon.qmd) for the filled and unfilled versions side by side.

## Stroke width and opacity

A polyline is drawn with a *single* stroke, so `size` and `opacity` are not
channels here: there is no per-row anything for them to vary along. That one
stroke still has a width and an opacity, and `style()` sets them:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(size = 4, opacity = 0.45) +
  title("One width, one opacity, for the whole layer")
```

This is the clearest case of the split described in
[Setting vs mapping](../style.qmd): *"can a column be mapped here?"* and *"does
this mark have this property?"* are different questions, and `line` answers them
differently.

A set color is the usual way to push a layer into the background:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "lightgrey") +
  line * smooth + style(color = "firebrick", size = 3) +
  title("Context in gray, message in red")
```

`style()` scopes like a channel: it attaches to the mark it is written after.
Unlike a channel it has no plot-scoped form: writing it before any mark is an
error rather than a shorthand, because a plot has no appearance of its own to
style.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("line")
```

A line is a stroke, so `size` is its width and `pattern` its dash. Neither can
vary along the line, because a width and a dash are properties of the whole
stroke: they are settings here even though `size` maps per row on `point`.
(`color` is the one that *can* vary along it, and the reason is exactly that a
color is a property of a **place** on the stroke rather than of the stroke:
see [What it refuses](#what-it-refuses).)

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  style(size = 3, pattern = "dashed") +
  y_label("Life expectancy") + title("One width and one dash, for every series")
```

`color` is mapped there, so it is not also set; a layer that both maps and sets
the same feature is refused, because one of the two would have to be silently
dropped.

The transforms a line takes carry their own parameters, which is where the rest
of the control lives: `line * density(adjust = 1.5)` widens the kernel,
`line * bounds(lo, hi)` traces a pre-computed band's two edges.

## What it refuses

A line is a stroke, and its texture vocabulary says so: a line's pattern is a
dash, while a hatch fills an area, and a line has no area to fill.

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + style(pattern = "hatch")
```

The refusal lists the three dashes a stroke does take. Its other two refusals
are above rather than here, each in the section that earns it. One is a
statistic across a category, which `line` cannot lay down. The other is a
mapped width or opacity, which one stroke cannot vary row by row even though
it can be set. That second pair is the worked example in [Setting vs
mapping](../style.qmd).

`color` is deliberately *not* in that pair, and the line between them is worth
drawing precisely, because it is the reason none of this is arbitrary. A width
and an opacity are properties of the stroke as a whole (there is no answer to
"how wide is the line *here*"). A color has an answer everywhere: point at a
place on the line and it has one. So a measure can ride `color` and vary along
the stroke, and it earns a gradient key rather than one swatch per series:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Each country's line, colored by its income")
```

Read one thread from left to right and you are watching two variables at once:
how long people live (the height) and how rich the country was when they did
(the color). `group` still splits the mark into one line per country; `color`
answers the other question. A *categorical* `color` splits and colors in one
go, exactly as it always has, which is the sentence at the top of this chapter.
