# Line

```{r}
#| include: false
source("../R/setup.R")
```

How did one value change over time? `line` sorts the rows by `x`, from
smallest to largest, and draws one stroke through their `(x, y)` points in that
order, the **x order**. The stroke is straight pieces laid end to end, each
joining one row's point to the next, with nothing smoothed between them.

## Basic usage

How did life expectancy change in five Asian countries? Most examples in this
chapter start from the smallest sentence that draws a line: a source, the mark,
and its two positions.

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
#| message: true
data(gapminder_asia) + line + x(year) + y(life)
```

*"Given gapminder Asia: a line, x is year, y is life."*

The stroke looks wrong, but gog drew exactly what the sentence asks for. All
five countries have a
row for each year, so the one stroke passes through five values at every year.
The vertical segments are where it joins the five countries inside one year,
before it continues to the next year. That is exactly what a line does when nothing
splits the rows: it draws one **series**, a single stroke in x order. The
message printed above the plot is gog's warning, and it names what to add: a
`color` or a `group`. [Multiple series](#multiple-series), two sections below,
draws the same table as five lines, one per country.

## Automatic sorting by x

Rows rarely arrive in the order a line needs: a file is sorted by name, or by
nothing at all. `line` sorts its own vertices, the points it joins, by `x`, so
the rows may arrive in any order. The table below is deliberately shuffled, and the line
still runs from left to right:

```{r}
#| echo: false
peek(scrambled)
```

```{r}
data(scrambled) + line + x(year) + y(life)
```

That sorting is part of what the mark means. It is not an extra step that the
engine performs for you. A line says *this value changed along this axis*, and
an axis has one direction, so there is only one order the vertices could be
joined in.

[`path`](path.qmd) is the mark that does the opposite. It joins the rows in the
order your table holds them, which is what draws a route or a loop. So choosing
between the two is choosing whether the order of your rows carries meaning. With
`line` the row order cannot carry meaning; with `path` the row order sets the
whole shape.

## Multiple series

The chapter's first plot drew one stroke through all five countries, and gog
asked for a `color` or a `group`. Here is the same table with `color(country)`
added, and the single stroke splits into a line per country. `color` also tells
the reader which line belongs to which country.

```{r}
data(gapminder_asia) + line + x(year) + y(life) +
  color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

Each unique value in `country` becomes a separate line with its own color, and
a color legend is generated automatically. `group` makes the same split without
a color or a legend. Use it when the lines matter as a whole and their names do
not, as in the [Many groups](../cookbook/groups.qmd#grouping-without-a-legend)
recipe.

## Layering a line over other marks

A line shows the trend and hides where the measurements were. A dot at each
row shows them again, which is why `line` and `point` are so often combined.
Write `color` *before* the marks and it belongs to the plot, so both layers
use it:

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  line + point
```

*"Given gapminder Asia: x is year, y is life, color by country, lines and also
points."*

Both the lines and the points share the same color per country.

### Giving each layer its own appearance

You do not have to map a column to `color` at all. Sometimes you want one color
for the lines and another for the points, chosen by you rather than read from
the data. That is a *setting*, written `style(color = )`, and no legend appears.
`group(country)` still keeps one line per country: it splits without coloring.
`style()` attaches to the nearest preceding mark and *stops there*, so each
layer can be styled independently:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "steelblue") +
  point + style(color = "firebrick", size = 3) +
  title("Blue line, red points: styled per layer")
```

*"Given gapminder Asia: x is year, y is life, lines grouped by country, colored
steelblue, and also points, colored firebrick, with size 3."*

A channel can belong to the whole plot or to one layer. Written before any mark,
it belongs to the plot and reaches every layer, as `color(country)` did above.
Written after a mark, it belongs to that mark alone, like `style()`, so each
layer can map a different column. The whole rule, for every atom, is one table
in [Encoding scope](../encoding-scope.qmd#the-rule):

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population) +
  title("Color on the lines, size on the points")
```

*"Given gapminder Asia: x is year, y is life, lines colored by country, and also
points sized by population."*

That works because a channel written after a mark belongs only to that mark. If
it also applied to the earlier `line`, the line would get a `size` for each
row, which it cannot take, and gog would refuse the plot. See
[Encoding scope](../encoding-scope.qmd).

## A category on the domain: the profile

In [Point](point.qmd#either-axis-a-horizontal-strip), swapping the two bindings
turned a strip plot on its side. A line answers the same swap differently. A
line's two axes are not interchangeable. `x` is the **domain**, the axis the
line is read along. `y` is the **measure**, the quantity it traces.

A bar draws horizontally when you swap its bindings, because its categorical
column says which axis holds the slots. A line usually has no such column,
because its two columns are both numbers, such as year and life. So gog does
not read the direction from the columns: the domain is always `x`. The
`quarterly` table, despite its name, holds one sales figure for each year. With
the years on `x`, the line is the plot you expect:

```{r}
#| echo: false
peek(quarterly)
```

```{r}
data(quarterly) + line + x(year) + y(sales) +
  title("Years on x: sales read along the years")
```

*"Given quarterly: a line, x is year, y is sales."*

Sales climbed for most of the years. They fell in 2008 and 2009, and again in
2020, and rose again after each fall. Swap the two bindings and the line still
sorts along `x`, which is now sales. The stroke starts at 2009, the year with the
lowest sales, and puts 2020 between 2012 and 2013, so the years are scrambled:

```{r}
data(quarterly) + line + x(sales) + y(year) +
  title("The bindings swapped: the line still sorts along x")
```

*"Given quarterly: a line, x is sales, y is year."*

The zigzag seems to show a value that rose and fell many times. It does not.
Each segment joins a year to the year with the next higher sales, not to the
next year, so the rises and falls come from the sorting. Only the overall slope
is real, and only because sales were mostly higher in later years.

Put a category on `y` instead and gog refuses it, as [What it
refuses](#what-it-refuses) shows. So the swap has two outcomes. When both
columns are numbers, gog accepts it, and the line sorts along the new `x` and
draws a different picture. When one column is a category, gog refuses the swap,
because that column must stay on `x`.

Sometimes the domain has to run up the page, as depth does in a plot of ocean
temperature, or as the years do here. Then write [`path`](path.qmd) instead. A
path joins the rows in the order the table holds them, so a table sorted along
the domain draws the vertical series the line could not:

```{r}
data(quarterly) + path + x(sales) + y(year) +
  title("The same rows, joined in table order")
```

*"Given quarterly: a path, x is sales, y is year."*

The domain takes a category as readily as a number, and connecting one summary
per category is the **profile plot**:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + line * mean + point * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("line * mean + x(continent): one value per category, joined")
```

*"Given gapminder 2007: a line derived by mean and also points derived by mean,
x is continent, y is life."*

The `mean` transform is what makes the profile readable. Without it, a `line`
connects all 142 rows across five category slots:

```{r}
data(gapminder_2007) + line + x(continent) + y(life) +
  title("Without the transform: 142 rows across five slots")
```

*"Given gapminder 2007: a line, x is continent, y is life."*

No reader can follow that. Inside each slot the stroke runs up and down through
every country, and between slots it joins the last country of one slot to the
first country of the next. gog prints a warning in your session when a line has
more than five rows and nothing splits or summarizes them. A single series,
such as the `quarterly` line, gets the same message; there one stroke is what
you want. `mean` leaves exactly one point per category, so the stroke joins
five summaries instead of 142 rows.

Does the connection mean anything? That depends on your data. Nothing was
measured between Africa and the Americas, and the segment is still drawn there.
If you read only the vertices, the segments help you compare the five heights.
If you read the segments themselves, they claim a change that never happened.
The grammar draws what you asked for. You decide whether your data supports the
connection. When it does not, keep the `mean` and change the mark to
[`bar`](bar.qmd). A bar stands in its own slot and claims nothing between one
slot and the next. An ordered category, such as a grade, a clothing size or a
survey scale, is the case that suits a line, because it has a direction for the
line to follow. Two years stored as text are such a category. One line for each
country, from its value in the first year to its value in the second, is the
**slope chart** ([Comparing
categories](../cookbook/comparing.qmd#what-if-the-change-matters-most)).

Continents have no order, but a straight axis must still put one of them first
and another last. A circle has no first or last place. Drop the point layer and
add `polar()`, and the same profile bends into a circle. That is the **radar
chart**, and it needs no new atom:

```{r}
data(gapminder_2007) + line * mean + x(continent) + y(life) + polar() +
  title("line * mean + polar(): the radar")
```

*"Given gapminder 2007: a line derived by mean, x is continent, y is life, in
polar."*

Look at the segment from Oceania back to Asia. On the flat profile the line has
two ends, because a straight axis has two ends. In the circle the five
categories are spread around the whole ring, so the last one sits beside the
first and the line closes. No atom was added to close the ring. A categorical
angular axis wraps, and the mark follows the space it is drawn in. The
[Polar](../polar.qmd) chapter covers this in full.

## Smoothed trend line

People in richer countries tend to live longer. The relationship between `gdp`
and `life` is positive: as one rises, the other tends to rise as well. Sometimes
the direction of a relationship is all a reader needs to know. Combine `line`
with the `smooth` transform to draw a trend line fitted through all the rows
instead of the rows themselves:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

The point layer keeps every country visible, and `smooth` reduces the same rows
to one fitted curve. Only the `line` layer carries the transform, so the points
are untouched.

`smooth` is not the line's own transform. Four other marks take it too, and each
draws the same fitted curve in its own way:

| Mark derived by `smooth` | What it draws |
|---|---|
| `point * smooth` | one dot at each fitted value |
| `area * smooth` | the curve filled to the baseline |
| `bar * smooth` | one thin bar at each fitted value |
| `step * smooth` | the curve as a staircase |
| `line * smooth` | the fitted curve above |

## Two boundaries from a pair

Sometimes the answer at each `x` is two values rather than one: a lowest and a
highest, or the spread around a mean. Most transforms give a
`line` one value per `x`. A **pair** transform gives it two, a low and a high. A
`line` then draws *both boundaries* as two separate lines, not one zigzag
joining them. A [`ribbon`](ribbon.qmd) takes the same pair and fills the band
between the two boundaries; a line draws only the boundaries. There are five
pair transforms: [`range`](../transforms.qmd#range-a-lowhigh-pair),
[`confidence`](../transforms.qmd#confidence-the-interval-of-the-mean),
[`deviation`](../transforms.qmd#deviation-the-spread-of-the-data),
[`smooth_band`](../transforms.qmd#smooth_band-the-band-around-the-curve) and
[`bounds`](../transforms.qmd#bounds-a-pre-computed-lowhigh-pair). Here are the
lowest and the highest life expectancy among the five countries in each year,
drawn as two lines:

```{r}
data(gapminder_asia) + line * range + x(year) + y(life) +
  y_label("Life expectancy") +
  title("The lowest and highest life expectancy, as two lines")
```

*"Given gapminder Asia: a line derived by range, x is year, y is life."*

Add `style(pattern = "dashed")` for the dashed band often used in published
figures. [Ribbon](ribbon.qmd#filled-or-two-dashed-edges) shows the same pair
filled and then unfilled.

## Stroke width and opacity

You may want thicker lines for a slide, or fainter ones where many lines cross.
Each line is drawn as a *single* stroke from its first point to its last. So
`size` and `opacity` are not channels here: one stroke has one width and one
opacity, and no row can change them. That one stroke still has a width and an
opacity, and `style()` sets them:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(size = 4, opacity = 0.45) +
  title("One width, one opacity, for the whole layer")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, with
size 4 and opacity 0.45."*

Two questions look alike and are not the same: *"Does this mark have this
property?"* and *"Can a column be mapped here?"* For a line's width, the first
answer is yes and the second is no. [Setting vs
mapping](../style.qmd#two-questions-not-one) uses a line as its clearest case.

Setting one pale color on a layer is the usual way to make another layer easier
to see:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "lightgray") +
  line * smooth + style(color = "firebrick", size = 3) +
  title("Every country in gray, the trend in red")
```

Each layer here carries its own `style()`. Unlike a channel, `style()` cannot
belong to the whole plot: written before any mark, it is an error, because a
plot has no appearance of its own to style.

## Each country and its continent's mean or median

The plot above draws five countries and one trend. The `gm_all` table holds 142
countries, measured every five years from 1952 to 2007. In one panel, 142 gray
lines are too many to read. One trend through all of them would also hide how
the continents differ. So draw one panel for each continent, with its countries
in gray and their mean as one blue line.

The sentence below has two `line` layers. The first draws one gray line for each
country, as above. The second is `line * mean`. In each year, `mean` averages
the values of the countries, so it leaves one value per year, and the line joins
those values. `| facet(continent)` splits the whole plot into one panel per
continent. Every layer is split the same way, so the blue line in each panel is
the mean of that continent's countries. [Faceting](../faceting.qmd) explains the
operator in full.

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + x(year) + y(life) +
  line + group(country) + style(color = "lightgray") +
  line * mean + style(color = "steelblue", size = 3) +
  y_label("Life expectancy") +
  title("Every country, and the continent's mean") | facet(continent)
```

*"Given all the gapminder years: x is year, y is life, lines grouped by country,
and also a line derived by mean, split into panel columns by continent."*

The blue line is the continent's mean, and the gray lines show how far each
country was from it. Every continent's mean rose, but Africa's stopped rising
for fifteen years. It rose from 39 years in 1952 to 53 in 1987, and in 2002 it
was still 53. In 22 of Africa's 52 countries, life expectancy was lower in 2002
than in 1987. In Zimbabwe it fell from 62 years to 40. The table has only two
countries in Oceania, Australia and New Zealand, so the blue line almost covers
their two gray lines.

Income is a second way to compare the continents. The same two layers can draw
GDP per person, with two changes to the sentence. The first change is the scale.
GDP per person runs from a few hundred dollars to more than a hundred thousand,
so the next plot draws it on a log scale. On this scale each labeled tick is ten
times the one below it, so the poorer countries are not crowded together at the
bottom.

The second change is the summary. gog computes `mean` in dollars, and only then
places the result on the log axis ([The measured value keeps its own
units](../scales.qmd#the-measured-value-keeps-its-own-units)). In dollars, one
very rich country can raise the mean far above most of the others. In 1952,
Kuwait's GDP per person was 108,382 dollars, and Asia's mean was 5,195. Sorted
from poorest to richest, the middle one of Asia's 33 countries had 1,207, less
than a quarter of the mean. In the Asia panel, a mean line would therefore run
above most of the gray lines. The
[`median`](../transforms.qmd#aggregation-family-sum-mean-median-max-min-quantile)
is that middle value, and sorting the countries gives the same order in dollars
as in logarithms. So `line * median` has half of the gray lines above it and
half below, on a log axis too:

```{r}
data(gm_all) + x(year) + y(gdp, scale = "log") +
  line + group(country) + style(color = "lightgray") +
  line * median + style(color = "steelblue", size = 3) +
  y_label("GDP per person") +
  title("GDP per person on five continents") | facet(continent)
```

*"Given all the gapminder years: x is year, y is gdp on a log scale, lines
grouped by country, and also a line derived by median, split into panel columns
by continent."*

The medians grew at very different rates. Europe's rose from 5,142 dollars in
1952 to 28,054 in 2007, more than five times its 1952 value. Africa's rose from
987 dollars to 1,452, less than one and a half times its 1952 value. In 2002 it
was lower than in 1972.

## What you can set

A setting changes how the stroke looks without reading a column, and each
mark takes its own. These are a line's settings, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("line")
```

A line is a stroke, so `size` is its width and `pattern` its dash. A setting
gives the whole layer one value. A channel reads a column instead, so what is
drawn follows the data.

Two of the four above are also channels. `color(country)` draws one color per
country. A mapped `pattern` dashes the series instead. `pattern` gives a stroke
five values, solid among them, shown here in the order gog assigns them:

```{r}
dash_order <- c("solid", "dashed", "dotted", "dotdash", "longdash")
dashes <- data.frame(
  dash = ordered(rep(dash_order, each = 2), levels = dash_order),
  at = rep(c(1, 2), times = 5),
  level = rep(c(5, 4, 3, 2, 1), each = 2))
data(dashes) + line + x(at) + y(level) + pattern(dash) +
  style(size = 2) + x_label("") + y_label("") + title("The five dashes")
```

*"Given the dashes table: lines, x is at, y is level, pattern by dash, with size
2."*

A sixth series gets the first of the five again, and gog says so: see
[`pattern`](../channels.qmd#pattern). `opacity` and `size` can only be set,
because a width and an opacity belong to the whole stroke, so no row can change
them. `color` is different. It is the only one that can
vary *along* a single stroke, because a color belongs to a place on the stroke
rather than to the whole stroke. [What it refuses](#what-it-refuses) ends with
that case, a line colored by a continuous column.

Set `size` and `pattern` together, and every series takes the same width and the
same dash:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  style(size = 3, pattern = "dashed") +
  y_label("Life expectancy") + title("One width and one dash, for every series")
```

*"Given gapminder Asia: lines, x is year, y is life, color by country, with size
3 and pattern dashed."*

Each property takes its value from one source: a column, or you. Here `color`
reads a column, and `size` and `pattern` are set, so no property is given a
value twice. Ask for both on one property and gog refuses, rather than dropping
one in silence.
[Setting vs mapping](../style.qmd#mapping-and-setting-the-same-property)
shows the refusal.

A transform's own arguments are neither a mapping nor a setting. They change
what is computed, not how it is drawn. `line * density(adjust = 1.5)` widens the
bandwidth, so the curve is smoother than the automatic one.
[Transforms](../transforms.qmd) lists what each one takes.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A dash is the only texture a stroke can carry. A hatch fills an area, and a line
has no area to fill.

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + style(pattern = "hatch")
```

The refusal lists the five values `pattern` does take on a stroke. Two more
refusals follow from earlier sections. The first is a categorical column on `y`, refused
because `y` carries the quantity a line traces, and category names are not a
quantity. [A category on the
domain](#a-category-on-the-domain-the-profile) showed where that rule comes
from:

```{r}
#| error: true
data(gapminder_2007) + line * mean + x(life) + y(continent)
```

The second refusal was described under [Stroke width and
opacity](#stroke-width-and-opacity). A mapped `size` or `opacity` is refused,
because one stroke has one width and one opacity, and no row can vary them.

`color` is the exception, because a color belongs to a place on the stroke
rather than to the whole stroke. So a continuous column mapped to `color`
changes the color along one line, and it gets a ramp legend rather than
one color per series:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Each country's line, colored by its income")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, color
by gdp on a log scale, with the viridis palette."*

`palette()` chooses the colors a mapping uses. `"viridis"` is a
[ramp](../channels.qmd#ramps), and the named palettes and what each is for are
in [Palettes](../channels.qmd#palettes).

Read one country's line from left to right and you see two variables. The height
is how long people live, and the color is how rich the country was then. `group`
still splits the mark into one line per country; `color` carries the second
variable, income. A *categorical* `color` splits and colors at the same time,
which is what Multiple series showed above.

## In short

A categorical `color`, a `group` or a mapped `pattern` splits the rows into one
line per category. Without them, one stroke joins every row. A line always sorts
along `x`, so row order means nothing to it. Swap the two bindings and the line
re-sorts: you get a different plot, not the same one on its side. Width and
opacity belong to the whole stroke, and only `color` varies along it.

## What comes next

You read a line by comparing its values with each other, so its axis need not
reach zero. [Bar](bar.qmd) measures from zero instead, so one bar's length has
a meaning by itself.
