# Line

```{r}
#| include: false
source("../R/setup.R")
```

How did one value change over time? `line` connects the rows in **x order**,
drawing one unbroken stroke (a polyline) through all the `(x, y)` points.

## Basic usage

```{r}
data(gapminder_asia) + line + x(year) + y(life)
```

*"Given gapminder Asia: a line, x is year, y is life."*

## Automatic sorting by x

`line` sorts its own vertices, the points it joins, by `x`, so the rows may
arrive in any order. The table below is deliberately shuffled, and the line
still runs from left to right:

```{r}
scrambled <- data.frame(year = c(1992, 1962, 2002, 1972, 1982),
                        life  = c(  68,   52,   71,   58,   63))

data(scrambled) + line + x(year) + y(life)
```

That sorting is part of what the mark means. It is not an extra step that the
engine performs for you. A line says *this value changed along this axis*, and
an axis has one direction, so there is only one order the vertices could be
joined in.

[`path`](path.qmd) is the mark that does the opposite. It joins the rows in the
order your table holds them, which is what draws a route or a loop. So choosing
between the two is choosing whether the order of your rows carries meaning: With
`line` the row order cannot carry meaning; with `path` it decides the whole
shape.

## Multiple series

Use `color` to split into one line per group:

```{r}
data(gapminder_asia) + line + x(year) + y(life) +
  color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

Each unique value in `country` becomes a separate line with its own color.
A color legend is generated automatically.

## Layering line over other marks

`line` and `point` are often combined: the point layer adds a dot at each
data position.
Write `color` *before* the marks and it belongs to the plot, so both layers
use it:

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  line + point
```

*"Given gapminder Asia: x is year, y is life, color by country, lines and also
points."*

Both the lines and the points share the same color per country.

### Giving each layer its own appearance

`style()` attaches to the nearest preceding mark and *stops there*, so each
layer can be styled independently:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "steelblue") +
  point + style(color = "firebrick", size = 3) +
  title("Blue line, red points: styled per layer")
```

*"Given gapminder Asia: x is year, y is life, lines grouped by country, colored
steelblue, and also points, colored firebrick, with size 3."*

Channels scope the same way. A channel written after a mark binds to that mark
alone, so each layer can carry a different encoding:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population) +
  title("Color on the lines, size on the points")
```

*"Given gapminder Asia: x is year, y is life, lines, color by country, and also
points, size by population."*

That works because channels bind *forward only*. A channel that also bound
backwards would put `size` on the `line`, which takes no per-row size, and the
plot would be refused. See
[Encoding scope](../encoding-scope.qmd).

## A category on the domain: the profile

A line's two axes are not the same kind of thing. `x` is the **domain**, the
axis the line is read along; `y` is the **measure**, the quantity it traces. The
domain takes a category as readily as a number, and connecting one summary per
category is the profile plot:

```{r}
data(gapminder_2007) + line * mean + point * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("line * mean + x(continent): one value per category, joined")
```

*"Given gapminder 2007: a line derived by mean and also points derived by mean,
x is continent, y is life."*

The transform is what makes that plot readable. Without one, a `line` would
connect all 142 rows across five category slots, and no reader can follow that.
The engine prints a warning on your console when a line has many ungrouped rows.
`mean` leaves exactly one point per category, which is what makes the connected
line mean anything.

Does the connection mean anything? That depends on your data. Nothing was
measured between Africa and the Americas, and the segment is still drawn there.
If you read only the vertices, the segments help you compare the five heights.
If you read the segments themselves, they claim a change that never happened.
The grammar draws what you asked for. You decide whether your data supports the
connection. Two cases make the judgment easy. An ordered category, such as a
grade, a size band or a survey scale, genuinely has a direction for the line to
follow. A *circular* category has no ends at all, and that is the next plot.

Drop the point layer, add `polar()`, and the same profile bends into a
circle. That is the **radar chart**, and it needs no new atom:

```{r}
data(gapminder_2007) + line * mean + x(continent) + y(life) + polar() +
  title("line * mean + polar(): the radar")
```

*"Given gapminder 2007: a line derived by mean, x is continent, y is life, in
polar."*

Look at the segment from Oceania back to Asia. On the flat profile the line has
two ends, because a straight axis has two ends. In the circle the five
categories are spread around the whole ring, so the last one sits beside the
first and the line closes. Nothing was added to say so: a categorical angular
axis wraps, and the mark takes that from the space it is drawn in. Full
treatment in [Polar](../polar.qmd).

A category on the measure axis is refused, because a line traces a quantity and
category names are not a quantity:

```{r}
#| error: true
render_svg(data(gapminder_2007) + line * mean + x(life) + y(continent))
```

## Smoothed trend line

Combine `line` with the `smooth` transform to draw a fitted trend
instead of raw data:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

*"Given gapminder 2007: x is gdp, y is life, points, color by continent, and
also a line derived by smooth."*

The point layer keeps every country visible, and `smooth` reduces the same rows
to one fitted curve. Only the `line` layer carries the transform, so the points
are untouched.

## Two boundaries from a pair

Most transforms give a `line` one value per x. A **pair** transform
([`range`](../transforms.qmd), `confidence`, `deviation`, or `bounds`) gives it
two (a low and a high), and a `line` then draws *both boundaries* as two
separate lines rather than connecting them into one zigzag. It is the *unfilled*
counterpart to a [`ribbon`](ribbon.qmd) filling the same pair. Here are the
lowest and highest life expectancies among five countries, drawn as two lines:

```{r}
data(gapminder_asia) + line * range + x(year) + y(life) +
  y_label("Life expectancy") +
  title("The lowest and highest life expectancy, as two lines")
```

*"Given gapminder Asia: a line derived by range, x is year, y is life."*

Add `style(pattern = "dashed")` for the dashed band that publications often
draw. [Ribbon](ribbon.qmd) shows the filled and unfilled versions next to each
other.

## Stroke width and opacity

A polyline is drawn with a *single* stroke, so `size` and `opacity` are not
channels here: there is nothing per row for them to vary along. That one
stroke still has a width and an opacity, and `style()` sets them:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(size = 4, opacity = 0.45) +
  title("One width, one opacity, for the whole layer")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, with
size 4 and opacity 0.45."*

Two questions look alike and are not. *"Can a column be mapped here?"* and
*"Does this mark have this property?"* get different answers on a `line`,
which is why [Setting vs mapping](../style.qmd) uses it as its clearest
case.

Setting one pale color is the usual way to keep a layer behind the one you want
read:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "lightgray") +
  line * smooth + style(color = "firebrick", size = 3) +
  title("Every country in gray, the trend in red")
```

`style()` scopes like a channel: it attaches to the mark it is written after.
Unlike a channel it has no plot-scoped form. Writing it before any mark is an
error, because a plot has no appearance of its own to style.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("line")
```

A line is a stroke, so `size` is its width and `pattern` its dash. Neither
varies *along* one line, because a width and a dash belong to the whole stroke.
A mapped `pattern` still gives each series its own dash, and `size` is a setting
here even though it maps per row on `point`. `color` is the one that *can* vary
along the stroke, because a color belongs to a place on the stroke rather than
to the whole stroke. The last section of this chapter shows that case.

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  style(size = 3, pattern = "dashed") +
  y_label("Life expectancy") + title("One width and one dash, for every series")
```

*"Given gapminder Asia: lines, x is year, y is life, color by country, with size
3 and pattern dashed."*

`color` is mapped there, so it is not also set. A layer that both maps and sets
the same feature is refused, because one of the two would have to be dropped in
silence.

The transforms a line takes carry their own parameters, and those give you the
rest of the control. `line * density(adjust = 1.5)` makes the density curve
smoother, and `line * bounds(lower, upper)` traces a band's two edges.

## What it refuses

A dash is the only texture a stroke can carry. A hatch fills an area, and a line
has no area to fill.

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + style(pattern = "hatch")
```

The refusal lists the three values a stroke's pattern does take. The two other
refusals this chapter shows appeared in earlier sections. One is shown above: a
category on the measure axis, which a line cannot trace. The other is stated
under Stroke width and opacity: a mapped width or opacity, which one stroke
cannot vary row by row even though it can be set. That second pair is the worked
example in [Setting vs mapping](../style.qmd).

`color` is deliberately *not* in that pair, and the difference is worth
stating exactly, because it is the reason none of this is arbitrary. A width
and an opacity are properties of the stroke as a whole (there is no answer to
"how wide is the line *here*"). A color has an answer everywhere: point at a
place on the line and it has one. So a measure can be mapped to `color` and
vary along the stroke, and it earns a gradient legend rather than one color
per series:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Each country's line, colored by its income")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, color
by gdp on a log scale, with the viridis palette."*

Read one country's line from left to right and you see two variables. The height
is how long people live, and the color is how rich the country was then. `group`
still splits the mark into one line per country; `color` answers the other
question. A *categorical* `color` splits and colors at the same time, which is
what Multiple series showed above.