# Line

```{r}
#| include: false
source("../R/setup.R")
```

How did one value change over time? `line` connects the rows in **x order**,
drawing one unbroken stroke (a polyline) through each `(x, y)` point.

## Basic usage

```{r}
data(gapminder_asia) + line + x(year) + y(life)
```

*"Given gapminder Asia: a line, x is year, y is life."*

## You do not have to sort first

`line` sorts its own vertices by `x`, so the rows may arrive in any order. The
frame below is deliberately shuffled, and the line still runs left to right:

```{r}
scrambled <- data.frame(year = c(1992, 1962, 2002, 1972, 1982),
                        life  = c(  68,   52,   71,   58,   63))

data(scrambled) + line + x(year) + y(life)
```

That sorting is part of what the mark means, not an extra step the engine
saves you. A line says *this value
changed along this axis*, and an axis has one direction, so there is only one order
the vertices could be joined in.

[`path`](path.qmd) is the mark that does the opposite. It joins the rows in the
order your table holds them, which is what draws a route or a loop. So choosing
between the two is choosing whether the order of your rows carries meaning: with
`line` it cannot, and with `path` it carries everything.

## Multiple series

Use `color` to split into one line per group:

```{r}
data(gapminder_asia) + line + x(year) + y(life) +
  color(country)
```

Each unique value in `country` becomes a separate line with its own color.
A color legend is generated automatically.

## Layering line over other marks

`line` and `point` are often combined: the point layer adds a dot at each
data position.
Write `color` *before* the marks and it belongs to the plot, so both layers
use it:

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  line + point
```

Both the lines and the points share the same color per country.

### Giving each layer its own appearance

`style()` attaches to the nearest preceding mark and **stops there**, so each
layer can be styled independently:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "steelblue") +
  point + style(color = "firebrick", size = 3) +
  title("Blue line, red points: styled per layer")
```

Channels scope the same way. Written after a mark, one binds to that mark alone,
so each layer can carry a different encoding:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population) +
  title("Color on the lines, size on the points")
```

That works because channels bind **forward only**. A channel that also bound
backwards would put `size` on the `line`, which has no size, and the plot
would be refused. See
[Encoding scope](../encoding-scope.qmd).

## A category on the domain: the profile

A line's two axes are not the same kind of thing. `x` is the **domain**, the axis
the path is read along; `y` is the **measure**, the quantity it traces. The domain
takes a category as readily as a number, and connecting one summary per category
is the profile plot:

```{r}
data(gapminder_2007) + line * mean + point * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("line * mean + x(continent): one value per category, joined")
```

The transform is what makes that plot readable. Without one, a `line` would
connect all 142 rows across five category slots, and no reader can follow
that. The engine prints a warning on your console when a line has many
ungrouped rows. `mean` leaves exactly one point per category, which is what
makes the connected path mean anything.

Does the connection mean anything? That depends on your data. Nothing was
measured between Africa and the Americas, and the segment still draws a path
there. Read the vertices, and the segments only help you compare the five
heights. Read the segments themselves, and they claim a change that never
happened. The grammar draws what you asked for; you decide whether the
connection is fair.
Two cases make the judgment easy. An ordered category, such as a grade, a size
band or a survey scale, genuinely has a direction for the path to run along. A
**circular** one has no ends at all, and that is the plot this section turns
to next.

Drop the point layer, add `polar()`, and the same profile bends into a
circle. That is the radar, and it needs no new atom:

```{r}
data(gapminder_2007) + line * mean + x(continent) + y(life) + polar() +
  title("line * mean + polar(): the radar")
```

Look at the segment from Oceania back to Asia. On the flat profile the path has
two ends, because a straight axis has two ends; in the circle the categories fill
the turn between them and the last is next to the first, so the path closes.
Nothing was added to say so: a categorical angular axis wraps, and the mark reads
that off the space it is drawn in. Full treatment in
[Polar](../polar.qmd).

A categorical **measure** is refused instead, because a line
traces a quantity and category names are not one:

```{r}
#| error: true
render_svg(data(gapminder_2007) + line * mean + x(life) + y(continent))
```

## Smoothed trend line

Combine `line` with the `smooth` transform to draw a fitted trend
instead of raw data:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

## Two boundaries from a pair

Most transforms give a `line` one value per x. A **pair** transform
([`range`](../transforms.qmd), `confidence`, `deviation`, or `bounds`) gives it
two (a low and a high), and a `line` then draws **both boundaries** as two
separate lines rather than connecting them into one zigzag. It is the *unfilled* counterpart to a
[`ribbon`](ribbon.qmd) filling the same pair. Here the min–max envelope of five
countries, as two lines:

```{r}
data(gapminder_asia) + line * range + x(year) + y(life) +
  y_label("Life expectancy") + title("A min–max envelope, as two boundary lines")
```

Add `style(pattern = "dashed")` for the dashed band publications often draw; see
[Ribbon](ribbon.qmd) for the filled and unfilled versions side by side.

## Stroke width and opacity

A polyline is drawn with a *single* stroke, so `size` and `opacity` are not
channels here: there is no per-row anything for them to vary along. That one
stroke still has a width and an opacity, and `style()` sets them:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(size = 4, opacity = 0.45) +
  title("One width, one opacity, for the whole layer")
```

This is the clearest case of the split described in
[Setting vs mapping](../style.qmd): *"can a column be mapped here?"* and *"does
this mark have this property?"* are different questions, and `line` answers them
differently.

A set color is the usual way to turn a layer into background:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "lightgray") +
  line * smooth + style(color = "firebrick", size = 3) +
  title("Every country in gray, the trend in red")
```

`style()` scopes like a channel: it attaches to the mark it is written after.
Unlike a channel it has no plot-scoped form: writing it before any mark is an
error rather than a shorthand, because a plot has no appearance of its own to
style.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("line")
```

A line is a stroke, so `size` is its width and `pattern` its dash. Neither can
vary along the line, because a width and a dash are properties of the whole
stroke: they are settings here even though `size` maps per row on `point`.
(`color` is the one that *can* vary along it, because a color is a property
of a **place** on the stroke rather than of the stroke; the last section of
this chapter ends on that case.)

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  style(size = 3, pattern = "dashed") +
  y_label("Life expectancy") + title("One width and one dash, for every series")
```

`color` is mapped there, so it is not also set; a layer that both maps and sets
the same feature is refused, because one of the two would have to be silently
dropped.

The transforms a line takes carry their own parameters, and those hold the
rest of the control: `line * density(adjust = 1.5)` widens the bandwidth, and
`line * bounds(lo, hi)` traces a band's two edges.

## What it refuses

A line is a stroke, so its texture is a dash. A hatch fills an area, and a
line has no area to fill.

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + style(pattern = "hatch")
```

The refusal lists the three values a stroke's pattern does take. Its other
two refusals belong to earlier sections. One is shown above: a category on
the measure axis, which a line cannot trace. The other is stated under Stroke
width and opacity: a mapped width or opacity, which one stroke cannot vary
row by row even though it can be set. That second pair is the worked example
in [Setting vs mapping](../style.qmd).

`color` is deliberately *not* in that pair, and the difference is worth
stating exactly, because it is the reason none of this is arbitrary. A width
and an opacity are properties of the stroke as a whole (there is no answer to
"how wide is the line *here*"). A color has an answer everywhere: point at a
place on the line and it has one. So a measure can be mapped to `color` and
vary along the stroke, and it earns a gradient legend rather than one color
per series:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Each country's line, colored by its income")
```

Read one country's line from left to right and you watch two variables at
once: how long people live (the height) and how rich the country was then
(the color). `group` still splits the mark into one line per country; `color`
answers the other question. A *categorical* `color` splits and colors at the
same time, which is the sentence under Multiple series above.
