# Line

```{r}
#| include: false
source("../R/setup.R")
```

How did one value change over time? `line` sorts the rows by `x`, from
smallest to largest, and draws one stroke through their `(x, y)` points in that
order, the **x order**. The stroke is straight pieces laid end to end, each
joining one row's point to the next, with nothing smoothed between them.

## Basic usage

Everything in this chapter is added to one sentence, the smallest that draws a
line: a source, the mark, and its two positions.

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life)
```

*"Given gapminder Asia: a line, x is year, y is life."*

The stroke looks wrong, and the plot is correct. All five countries have a
row for each year, so the one stroke passes through five values at every year.
The vertical segments are where it joins the five countries inside one year,
before it continues to the next year. That is exactly what a line does when nothing
splits the rows: it draws one **series**, a single stroke in x order. gog warns
about this sentence and says what to add, a `color` or a `group`. [Multiple
series](#multiple-series) below draws the same table as five lines, one per
country.

## Automatic sorting by x

Rows rarely arrive in the order a line needs: a file is sorted by name, or by
nothing at all. `line` sorts its own vertices, the points it joins, by `x`, so
the rows may arrive in any order. The table below is deliberately shuffled, and the line
still runs from left to right:

```{r}
#| echo: false
peek(scrambled)
```

```{r}
data(scrambled) + line + x(year) + y(life)
```

That sorting is part of what the mark means. It is not an extra step that the
engine performs for you. A line says *this value changed along this axis*, and
an axis has one direction, so there is only one order the vertices could be
joined in.

[`path`](path.qmd) is the mark that does the opposite. It joins the rows in the
order your table holds them, which is what draws a route or a loop. So choosing
between the two is choosing whether the order of your rows carries meaning. With
`line` the row order cannot carry meaning; with `path` the row order sets the
whole shape.

## Multiple series

This is the plot the warning above asked for. The table is the same one the
chapter opened with, and one word splits the single stroke into a line per
country: `color`, which also tells the reader which line is which.

```{r}
data(gapminder_asia) + line + x(year) + y(life) +
  color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*

Each unique value in `country` becomes a separate line with its own color, and
a color legend is generated automatically. `group` makes the same split without
a color or a legend, which the [Groups](../cookbook/groups.qmd#grouping-without-a-legend)
recipe shows, for when the shape of the lines together is the answer and their
names are not.

## Layering line over other marks

A line shows the trend and hides where the measurements were. A dot at each
row puts them back, which is why `line` and `point` are so often combined.
Write `color` *before* the marks and it belongs to the plot, so both layers
use it:

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  line + point
```

*"Given gapminder Asia: x is year, y is life, color by country, lines and also
points."*

Both the lines and the points share the same color per country.

### Giving each layer its own appearance

You do not have to map a column to `color` at all. If you want one color for
the lines and another for the points, chosen by you rather than read from the
data, that is a *setting*, `style(color = )`, and no legend appears.
`group(country)` still keeps one line per country: it splits without coloring.
`style()` attaches to the nearest preceding mark and *stops there*, so each
layer can be styled independently:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "steelblue") +
  point + style(color = "firebrick", size = 3) +
  title("Blue line, red points: styled per layer")
```

*"Given gapminder Asia: x is year, y is life, lines grouped by country, colored
steelblue, and also points, colored firebrick, with size 3."*

Channels scope the same way. Written before any mark, a channel belongs to the
plot and reaches every layer, as `color(country)` did above. Written after a
mark, it binds to that mark alone, and not to any mark written after it, so
each layer can carry a different encoding. The whole rule, for every atom, is
one table in [Encoding scope](../encoding-scope.qmd#the-rule):

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population) +
  title("Color on the lines, size on the points")
```

*"Given gapminder Asia: x is year, y is life, lines, color by country, and also
points, size by population."*

That works because channels bind *forward only*. A channel that also bound
backwards would put `size` on the `line`, which takes no per-row size, and the
plot would be refused. See
[Encoding scope](../encoding-scope.qmd).

## A category on the domain: the profile

A line's two axes are not the same kind of thing. `x` is the **domain**, the
axis the line is read along. `y` is the **measure**, the quantity it traces.

A bar turns on its side when you swap its bindings, because its categorical
column says which axis holds the slots. A line has no such column. Its two columns are
usually both numbers, such as year and life, so the rule fixes the domain: it
is always `x`. The `quarterly` table holds one sales figure for each year.
With the years on `x`, the line is the plot you expect:

```{r}
#| echo: false
peek(quarterly)
```

```{r}
data(quarterly) + line + x(year) + y(sales) +
  title("Years on x: sales read along the years")
```

*"Given quarterly: a line, x is year, y is sales."*

Sales climbed for most of the years. They fell in 2008 and 2009, and again in
2020, and climbed back after each fall. Swap the two bindings and the line still
sorts along `x`, which is now sales. The stroke starts at 2009, the year with the
lowest sales, and puts 2020 between 2012 and 2013, so the years are scrambled:

```{r}
data(quarterly) + line + x(sales) + y(year) +
  title("The bindings swapped: the line still sorts along x")
```

*"Given quarterly: a line, x is sales, y is year."*

Read as a history of sales, this seems to say that sales fluctuated while
climbing. It does not. Each segment joins the two years with the nearest sales,
not two consecutive years, so the rises and falls come from the sorting, not
from the sales. Only the overall slope survives, and only because sales were
mostly higher in later years.

Put a category on `y` instead and gog refuses it, as [What it
refuses](#what-it-refuses) shows. So there are two cases. When both columns are numbers, the swap
is accepted, but you do not get the same plot on its side: the line sorts
along `x` and draws a different picture. When one column is a category, it must be on `x`.

Sometimes the domain has to run up the page, as depth does in an ocean profile,
or as the years do here. Then write [`path`](path.qmd) instead. A path joins
the rows in the order the table holds them, so a table sorted along the domain
draws the vertical series the line could not:

```{r}
data(quarterly) + path + x(sales) + y(year) +
  title("The same rows, joined in table order")
```

*"Given quarterly: a path, x is sales, y is year."*

The domain takes a category as readily as a number, and connecting one summary
per category is the profile plot:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + line * mean + point * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("line * mean + x(continent): one value per category, joined")
```

*"Given gapminder 2007: a line derived by mean and also points derived by mean,
x is continent, y is life."*

The transform is what makes that plot readable. Without one, a `line`
connects all 142 rows across five category slots:

```{r}
data(gapminder_2007) + line + x(continent) + y(life) +
  title("Without the transform: 142 rows across five slots")
```

*"Given gapminder 2007: a line, x is continent, y is life."*

No reader can follow that. Inside each slot the stroke runs up and down through
every country, and between slots it joins whichever two countries the sort put
last and first. The engine prints a warning on your console when a line has
many ungrouped rows. `mean` leaves exactly one point per category, which is
what makes the connected line mean anything.

Does the connection mean anything? That depends on your data. Nothing was
measured between Africa and the Americas, and the segment is still drawn there.
If you read only the vertices, the segments help you compare the five heights.
If you read the segments themselves, they claim a change that never happened.
The grammar draws what you asked for. You decide whether your data supports the
connection. Two cases make the judgment easy. An ordered category, such as a
grade, a size band or a survey scale, genuinely has a direction for the line to
follow. A *circular* category has no ends at all, and that is the next plot.

Drop the point layer, add `polar()`, and the same profile bends into a
circle. That is the **radar chart**, and it needs no new atom:

```{r}
data(gapminder_2007) + line * mean + x(continent) + y(life) + polar() +
  title("line * mean + polar(): the radar")
```

*"Given gapminder 2007: a line derived by mean, x is continent, y is life, in
polar."*

Look at the segment from Oceania back to Asia. On the flat profile the line has
two ends, because a straight axis has two ends. In the circle the five
categories are spread around the whole ring, so the last one sits beside the
first and the line closes. Nothing was added to say so: a categorical angular
axis wraps, and the mark follows the space it is drawn in. The
[Polar](../polar.qmd) chapter covers this in full.

## Smoothed trend line

Richer countries tend to live longer. The relationship between gdp and life is
positive: as one rises, the other tends to rise as well. Sometimes the direction
of a relationship is all a reader needs to know. Combine `line` with the
`smooth` transform to draw a trend line fitted through all the rows instead of
the rows themselves:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth
```

*"Given gapminder 2007: x is gdp, y is life, points, color by continent, and
also a line derived by smooth."*

The point layer keeps every country visible, and `smooth` reduces the same rows
to one fitted curve. Only the `line` layer carries the transform, so the points
are untouched.

## Two boundaries from a pair

Most transforms give a `line` one value per x. A **pair** transform gives it
two, a low and a high, and a `line` then draws *both boundaries* as two separate
lines rather than connecting them into one zigzag. That is the *unfilled*
counterpart to a [`ribbon`](ribbon.qmd) filling the same pair. There are four
pair transforms: [`range`](../transforms.qmd#range-a-lowhigh-pair),
[`confidence`](../transforms.qmd#confidence-the-interval-of-the-mean),
[`deviation`](../transforms.qmd#deviation-the-spread-of-the-data) and
[`bounds`](../transforms.qmd#bounds-a-pre-computed-lowhigh-pair). Here are the
lowest and highest life expectancies among five countries, drawn as two lines:

```{r}
data(gapminder_asia) + line * range + x(year) + y(life) +
  y_label("Life expectancy") +
  title("The lowest and highest life expectancy, as two lines")
```

*"Given gapminder Asia: a line derived by range, x is year, y is life."*

Add `style(pattern = "dashed")` for the dashed band often used in published
figures. [Ribbon](ribbon.qmd) shows the filled and unfilled versions next to each
other.

## Stroke width and opacity

A line is drawn with a *single* stroke from its first point to its last, so
`size` and `opacity` are not channels here: there is nothing per row for them
to vary along. That one
stroke still has a width and an opacity, and `style()` sets them:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(size = 4, opacity = 0.45) +
  title("One width, one opacity, for the whole layer")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, with
size 4 and opacity 0.45."*

Two questions look alike but are not the same. *"Can a column be mapped here?"* and
*"Does this mark have this property?"* get different answers on a `line`,
which is why [Setting vs mapping](../style.qmd) uses it as its clearest
case.

Setting one pale color is the usual way to keep a layer behind the one you want
the reader to read:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "lightgray") +
  line * smooth + style(color = "firebrick", size = 3) +
  title("Every country in gray, the trend in red")
```

`style()` scopes like a channel: it attaches to the mark it is written after.
Unlike a channel it has no plot-scoped form. Writing it before any mark is an
error, because a plot has no appearance of its own to style.

## What you can set

A setting changes how the stroke looks without reading a column, and each
mark takes its own. These are a line's settings, with the values each accepts: 

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("line")
```

A line is a stroke, so `size` is its width and `pattern` its dash. A setting
gives the whole layer one value. A channel reads a column instead, so what is
drawn follows the data.

Two of the four above are also channels. `color(country)` draws one color per
country. A mapped `pattern` dashes the series instead, though a stroke has only
three dashes, so they repeat past the third series: see
[`pattern`](../channels.qmd#pattern). `opacity` and `size` can only be set,
because a width and an opacity belong to the whole stroke, and there is nothing
per row for a column to give. `color` goes one step further. It is the only one
that can vary *along* a single stroke, because a color belongs to a place on the
stroke rather than to the whole stroke. The gradient at the end of this chapter
shows that case.

Set both at once, and every series takes the same width and the same dash:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  style(size = 3, pattern = "dashed") +
  y_label("Life expectancy") + title("One width and one dash, for every series")
```

*"Given gapminder Asia: lines, x is year, y is life, color by country, with size
3 and pattern dashed."*

Each property takes its value from one place: a column, or you. Here `color`
reads a column, and `size` and `pattern` are set, so nothing collides. Ask for
both on one property and gog refuses, rather than dropping one in silence.
[Setting vs mapping](../style.qmd#you-cannot-map-and-set-the-same-property)
shows the refusal.

A third kind of number belongs to neither the column nor you. A transform
carries its own, and they change what is computed rather than how it is drawn.
`line * density(adjust = 1.5)` widens the bandwidth, so the curve comes out
smoother than the automatic one. [Transforms](../transforms.qmd) lists what each
one takes.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A dash is the only texture a stroke can carry. A hatch fills an area, and a line
has no area to fill.

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + style(pattern = "hatch")
```

The refusal lists the three values a stroke's pattern does take. Two more
refusals belong to a line. The first is a categorical column on `y`, refused
because `y` carries the quantity a line traces, and category names are not a
quantity. [A category on the
domain](#a-category-on-the-domain-the-profile) showed where that rule comes
from:

```{r}
#| error: true
data(gapminder_2007) + line * mean + x(life) + y(continent)
```

The second is only described, under [Stroke width and
opacity](#stroke-width-and-opacity): a mapped width or opacity, refused because
one stroke has a single width and a single opacity, so no row is left for a
column to answer.

`color` is the exception, for the reason the section above gave: a color belongs
to a place on the stroke rather than to the whole stroke. A continuous column
mapped to `color` therefore varies along one line, and it earns a gradient
legend rather than one color per series:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Each country's line, colored by its income")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, color
by gdp on a log scale, with the viridis palette."*

`palette()` names the colors a mapping uses. `"viridis"` is a ramp, and the
named palettes and what each is for are in [Palettes](../channels.qmd#palettes).

Read one country's line from left to right and you see two variables. The height
is how long people live, and the color is how rich the country was then. `group`
still splits the mark into one line per country; `color` answers the other
question. A *categorical* `color` splits and colors at the same time, which is
what Multiple series showed above.
