# Path

```{r}
#| include: false
source("../R/setup.R")
```

Where did each country travel, in the order it happened? `path` strokes the
rows **in the order the table gives them**. That is all of it, and it is
the one thing `line` will not do. A line sorts its vertices by `x` first,
because a line draws a *function*: one `y` for each `x`, read left to
right along a domain. A path draws a *route*. It visits row 1, then row 2,
and it may turn back, cross itself, or return to where it started.

Everything the two marks share, they share exactly: same stroke, same `color` and
`group` splits, same `style(size =)`, same dash. One question separates them, and
this chapter shows what that question gives a path: an arrowhead, a spiral in
polar, and the third dimension. A `line` can have none of the three. All three are the same
answer read in different places, so none of them is a special case.

## The connected scatterplot

Here are five Asian countries, each measured twelve times between 1952 and 2007.
Put income on one axis and life expectancy on the other, and connect each
country's measurements in the order they were taken:

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Fifty-five years, five countries, one plane")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country."*

Each thread is one country's history. Time is nowhere on the plot, and yet the
whole chart is about time: it is the order the points are joined in. Japan
climbs steeply, then turns right and flattens: past that corner, income keeps
rising and life expectancy barely moves. Korea follows the same corner two
decades later.
China goes almost straight up at very low income, which is the shape of a country
that got healthier long before it got rich, and Indonesia's path bends the
same way. This is a **connected scatterplot**, and no other arrangement of these atoms
draws it.

## Why `line` cannot draw it

Change one word and the same sentence collapses:

```{r}
data(gapminder_asia) + line + x(gdp) + y(life) + color(country) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("the same data, sorted by x")
```

Look at China's thread. Its income did not rise steadily: it fell and
recovered, so the year order and the income order are different sequences,
and `line` draws the income one. Every point is in the right place, and every
route between them is wrong. That is not a bug in `line`; it is what a line
*is*. A function has one value per `x`, so sorting changes nothing, and for a
time series read along a date axis the sort helps.

The two marks answer different questions, so a spec that names the wrong one gets
a plausible picture rather than an error. The rule is this: if the reader
should follow the marks in a particular order, that order has to come from
the table, and only `path` takes it from there.

## Which end is the last row?

A route has a direction, and the plot above does not show it. Japan's thread
could be read from either end. So `path` takes a setting no other mark can:

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("style(arrow = \"end\"): the head is 2007")
```

Now each route points at its last row. `arrow` takes `"end"`, `"start"`, or
`"both"`: a head at the last row, at the first, or at each.

It belongs to `path` alone, and every other mark refuses it, which is worth
seeing because the refusal is really the argument for the mark:

```{r}
#| error: true
render_svg(data(gapminder_asia) + line + x(year) + y(life) + style(arrow = "end"))
```

A line's last vertex is wherever the domain happens to end, so a head there
marks the sort rather than anything the data did. Only a path has an end that
comes from the data.

## An arrow is a layer

That makes `path` the other half of the annotation the [Text](text.qmd)
chapter began. A note is a one-row table; the arrow that points from the note
to the thing is a **two-row** table, one row for each end:

```{r}
seg  <- data.frame(gdp = c(20000, 13100), life = c(45.5, 50.2))
note <- data.frame(gdp = 20500, life = 44.8, what = "Botswana")

data(gapminder_2007) + point + x(gdp) + y(life) +
  data(seg)  + path + style(arrow = "end", color = "#c0392b") +
  data(note) + text + label(what) +
  title("a callout, in the grammar you already have")
```

There is no `annotate()` function in this grammar and there will not be one. An
annotation is a layer over a small table, and the marks it uses are the marks
everything else uses: `text` for the words, `path` for the arrow, `point` for a
marker. What the table holds is data, so it is scaled like data, which is why
the arrow's ends are written in the data's own units rather than in pixels.

## One transform, and the rest refused

The other stroke and band marks compose with any transform. `path` composes
with one, and the refusals say why:

```{r}
#| error: true
render_svg(data(gapminder_asia) + path * mean + x(country) + y(life))
```

A summary replaces the rows. `mean` reduces each country to one number, and
what is left to connect is one point per country, in whatever order the
summary produced them. That is a `line`, reached by a detour. The order a path *is*
does not survive being summarized, and that one sentence empties almost the whole
column of the [Combinations](../combinations.qmd) grid for this mark.

There is one exception: `density`, and it is an exception to the *sentence*
rather
than to the rule: a summary reduces each group to one value, but a density estimated
over the plane does not reduce anything to keys at all. What it produces is
vertices **in the order they were traced**, which is the one thing a path is for.

## The contour: `path * density`

Bind both positions and compose `density`, and gog estimates how thick the cloud
is everywhere on the plane, then traces the rings where it is equally thick:

```{r}
data(iris_flowers) + path * density +
  x(sepal_length) + y(petal_length) +
  title("where the flowers cluster")
```

Nothing new was added to the grammar to draw that. The rule that made a
[heatmap](zone.qmd) out of `zone * bin` is the same one at work here: how many
axes a transform cuts is read off the mark, never asked for. A `bar` leaves an
axis free to measure the density along, so `bar * density` spreads it along one
axis and draws a curve. A `path` is a route through two data columns and privileges
neither, so it has no axis to spare, and `density` cuts both. There is no
`contour` atom for the same reason there is no `pie()`
([Polar](../polar.qmd) makes that argument): the sentence already exists.

Read the outermost ring, then the ones inside it. The outermost is a *single*
ring around both clusters, and the next level inward splits into two separate
rings, which is
why a level is not enough to tell gog where one stroke ends. Each ring is its own
stroke, and the level only decides its color, so the two inner rings never get
joined by a line drawn across the empty ground between them.

The color is the level, and nothing had to bind it, exactly as `bar * bin`
binds `y` for you. Say it out loud if you prefer, and change how many
rings are traced:

```{r}
data(iris_flowers) + path * density(levels = 10) +
  x(sepal_length) + y(petal_length) + color(level) +
  title("ten levels, named out loud")
```

`density` is one transform read three ways, so each of its parameters belongs
to one reading and is refused in the other two. `levels` cuts the **field**
(the estimated thickness everywhere on the plane) into that many steps, so it
needs a field to cut: the one-dimensional reading is a single curve, with
nothing to cut.

```{r}
#| error: true
render_svg(data(iris_flowers) + line * density(levels = 8) + x(sepal_length))
```

`bandwidth` is refused for the opposite reason. It is a width in the data's own units, and a
contour spreads over two columns that measure different things. Here they
happen to agree, sepal length and petal length both in centimeters, but there
is no reason two axes ever do. One number cannot be a width in both:

```{r}
#| error: true
render_svg(data(iris_flowers) + path * density(bandwidth = 0.5) +
  x(sepal_length) + y(petal_length))
```

`adjust` is the parameter that means the same thing on either axis, because it is a
multiplier rather than a length, so it is the one that reaches all three readings:

```{r}
data(iris_flowers) + path * density(adjust = 1.6) +
  x(sepal_length) + y(petal_length) +
  title("the same cloud, smoothed harder")
```

To estimate one field per category, `group` splits it. The level already
holds `color`, which is exactly why the split is `group` and not `color`:

```{r}
petals <- iris_flowers[iris_flowers$species != "setosa", ]

data(petals) + path * density(levels = 4) +
  x(sepal_length) + y(petal_length) + group(species) +
  title("two species, estimated separately")
```

A `zone` has no measure axis either, so it reads `density` in two dimensions too,
and it *fills* what this mark strokes. Give both marks the same `levels` and you get
the same curves twice, once as lines and once as the edges of filled bands, which is
the plot usually meant by a *filled* density. It belongs to the mark that draws it:
[`zone * density`](zone.qmd), and the zone chapter covers its ramp. One
field, two marks, [the same split](../grammar.qmd) as `bar * bin` and
`line * bin`: the mark chooses the geometry, the transform stays constant.

## In polar, a route is a spiral

`path` bends with everything else. Give it an angle and a radius that both grow,
and the route winds outward:

```{r}
spiral <- data.frame(a = seq(0, 6.2, length.out = 90))
spiral$r <- seq(0.1, 4, length.out = 90)

data(spiral) + path + x(a) + y(r) + polar() +
  title("a path in polar")
```

One thing `path` does not copy from `line` is the radar's closing segment
([Polar](../polar.qmd) draws it). A
categorical `line` in polar joins its last vertex back to its first, because
the categories fill the whole turn, and a gap there would leave one sector
blank. A path is not indexed by the angular axis at all, so its last vertex is
simply where the data stopped, and closing it would draw a segment nothing asked
for. If a route does return to its start, say so by repeating the first row.

## In space, the route is unchanged

Bind `z` and the route runs through the cube instead of across the page.
`path` is the *only* stroke that takes
the third position, and the reason is the one this whole chapter is about. Two
gliders circling the same thermal (a rising column of warm air), sampled as
they climb:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  title("Two gliders, one thermal")
```

An order belongs to no axis, so it survives the third dimension untouched: the
route is the same route however the scene is turned. A `line` sorts by `x`, and
the cube has no left to right. There `x` is one of three equal positions, and at
some viewing angles it runs straight into the page and becomes depth rather than
position, which would leave a line sorted by an axis the reader cannot see. So
the sort that makes a `line` a line is the very thing that will not follow it
into space, and `line`, `step` and `area` say so rather than half-drawing it:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude)
```

Watch the two threads where they cross: whichever is nearer covers the other,
and which one that is changes as they climb. A glyph sits at one distance from
the eye and can be sorted whole. A stroke running through the cube has a far
end and a near end, so the unit that gets sorted is the **segment**. The
segments of both routes are sorted together. If each stroke were sorted as a
whole, one glider would sit entirely in front of the other: two coils that
never meet, rather than two that thread through each other.

Cutting the route into segments is also what lets `color` carry a **measure**
along it. Bind the altitude to the color as well as to the height, and the
climb is said twice: once by where the thread is, once by what color it is:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(altitude) + group(glider) + palette("viridis") +
  space(turn = -25, tilt = 20) +
  title("Altitude, said twice")
```

`palette("viridis")` picks the ramp; `group(glider)` keeps the two routes
apart. Drop that last atom and the plot is wrong. A *category* on `color` splits the
mark into one stroke per group, but a *measure* has no categories to split on.
Without `group` the two gliders would be joined into a single route, and a
straight line would run from where one glider's rows stop to where the
other's begin. Color
answers what the route was carrying; `group` answers which route it is. Two
questions, two atoms.

A dash survives the segmentation too. Each segment is drawn as its own piece,
and a dash normally restarts at the start of a piece:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  style(pattern = "dashed") +
  space(turn = -25, tilt = 20)
```

The dashes run along the route rather than resetting at every sample, because the
dash is a property of the route and not of the segmentation the depth sort
happens to need. An arrowhead behaves the same way: it is sorted at the depth
of its own tip, so a head that ends deep in the cube is covered by whatever passes
in front of it.

[Space](../space.qmd) owns the third dimension and the viewing angle.

## What a path is for

The connected scatterplot is the best known use, but the mark is more general
than that. Anything whose reading order is the table's order is a path: a
trajectory through any two measurements over time, a route through space, a
cycle that returns to where it began, a hysteresis loop (a cycle whose
forward and return paths differ), an arrow. The test is always the
same one question. Does the order of the rows mean something? If it does, the
mark is `path`; if the reader should read left to right, it is `line`.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("path")
```

`arrow` is the path's own, and it is a one-mark setting for a reason about
geometry rather than taste. A head marks a direction, and a path is the only mark
that has one: `line`, `step` and `area` sort their vertices by `x`, so their last
point is wherever the domain ends rather than where the data stopped.

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "both", size = 2.5, pattern = "dotted") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both ends marked, on a dotted route")
```

It takes a *value* rather than `TRUE`/`FALSE`, because `"both"` is an ordinary
want and a flag would have needed a second setting to express it. Note that the
heads draw solid on a dotted path: the dash is the route's texture, and a head
chopped into dashes stops reading as an arrow.

The one transform a path takes parameters from is `density`, whose `levels`
and `adjust` are above. Everything else that varies is either mapped or set.

## What it refuses

A path is one stroke, and a stroke has one width (there is no answer to "how
thick is the route *here*"), so a width cannot vary along it:

```{r}
#| error: true
data(gapminder_asia) + path + x(gdp) + y(life) + size(gdp)
```

Set it instead and every segment takes the same width, which is the [Setting
vs mapping](../style.qmd) distinction again. The refusals a path makes that
no other mark does are above rather than here. One is the summaries it cannot
take, because a summary replaces the very rows whose order *is* the path. One
is the two `density` parameters, each refused in the reading it cannot mean.
The last is an arrow asked of `line`, which has no end to point from.