# Path

```{r}
#| include: false
source("../R/setup.R")
```

How did income and life expectancy change together in each country, year by
year? `path` draws one stroke through the rows' points, *in the order the table
gives them*. That is the whole definition of the mark, and it is the one thing
`line` does not do. A
line sorts its vertices by `x` first, because a line draws a *function*: one `y`
for each `x`, read left to right along a domain. A path draws a *route*. It
visits row 1, then row 2, and it may turn back, cross itself, or return to where
it started.

The two marks draw the same kind of stroke, and they take the same `color` and
`group` splits and the same `style(size = )` and `style(pattern = )` settings.
One question separates them: does the order of the rows mean something? A path
keeps that order, and three things follow from it. It can carry an arrowhead,
because its last vertex is the table's last row. In polar it can go around the
circle more than once, because two rows with the same angle stay apart, each on
its own turn. And it can run through the cube, because the row order does not
depend on any axis. None of the three is a special case.

## The connected scatterplot

Two measures that change over time can be drawn as two separate time series, one
above the other. The reader then has to find the same year in both plots before
comparing them. One plot can show both measures and time together, if the rows
are already in time order. Here are five Asian countries, each measured twelve
times from 1952 to 2007. Put income on one axis and life expectancy on the
other, and connect each country's measurements in the order they were taken:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Fifty-five years, five countries, one plane")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country."*

Each route is one country's history. The plot has no time axis, but it still
shows time: time is the order in which the points are joined. China's
route goes almost straight up at very low income. That shape belongs to a
country whose life expectancy rose long before its income did, and India's and
Indonesia's routes rise the same way. Korea's route starts upward and then turns
a sharp corner to the right. Past that corner its income climbs fast and its
life expectancy rises more slowly. Japan's route is the flattest of the five. It
begins where Korea's route turns: Japan had that income and life expectancy in
1952, twenty years before Korea. This is a **connected scatterplot**, and `path`
is the only mark that draws it.

## What `line` draws instead

`line` is the mark you would probably write first, because it is the familiar
way to join rows with a stroke. So compare what `line` draws from the same
table. One word changes in the specification, and three of the five routes
change shape:

```{r}
data(gapminder_asia) + line + x(gdp) + y(life) + color(country) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("the same data, sorted by x")
```

*"Given gapminder Asia: lines, x is gdp, y is life, color by country."*

Look at China's route. Its income did not rise steadily: it fell once and then
rose again, so the year order and the income order differ, and `line` joins the
points in income order. Every point is still in the right place, but the stroke
between them no longer follows the years. Indonesia's and Japan's routes change
the same way, Japan's only slightly. India's and Korea's income rose at every
measurement, so for them both marks draw the same picture. This is not a defect
in `line`. A line draws a function, and a function has one value for each `x`.
Sorting the rows by `x` cannot change which value belongs to which `x`. For a
time series read along a date axis, the sort simply puts the rows in date order.

The two marks answer different questions. If a specification names the wrong
one, the engine still draws a plot, and the plot still looks reasonable. Nothing
reports the mistake. The rule is this. If the reader should follow the points in
any order other than left to right, that order has to come from the table, and
only `path` takes its order from the table.

## Which end is the last row?

A route has a direction, and a plain stroke does not show it. Japan's route
could be read from either end. So `path` takes a setting no other mark can:

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("style(arrow = \"end\"): the head sits at 2007")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
end."*

Now each route points at its last row. `arrow` takes `"end"`, `"start"`, or
`"both"`: an arrowhead (a *head*, for short) at the last row, at the first row,
or at both ends.

Two marks take it: `path`, and [`edge`](edge.qmd), whose direction comes from
the two columns of its layout. Every other mark refuses it. A line's last vertex
is the row with the largest `x`, so a head there would point at the end of the
sort, not at the last row. A path's end comes from the data. [What it
refuses](#what-it-refuses) shows a line refusing an arrowhead.

## An arrow is a layer

`arrow` has a second use, and it needs no new atom. A note placed in an empty
part of the plot has to say which point it is about. An arrow from the words to
the point says it. The [Text](text.qmd#a-note-is-a-layer) chapter drew a note by
itself; here `path` adds the arrow beside one. A note is a one-row table; the
arrow that points from the note to the point it is about is a *two-row* table,
one row for each end:

```{r}
#| echo: false
peek(gapminder_2007)
peek(botswana_arrow)
peek(botswana_label)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(botswana_arrow)  + path + style(arrow = "end", color = "#c0392b") +
  data(botswana_label) + text + label(what) +
  title("an annotation, in the grammar you already have")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also a path from the
Botswana arrow, and text from the Botswana label, label by what."*

This grammar has no `annotate()` function, and it needs none. An annotation, a
note with its arrow, is made of layers, each over a small table. Its marks are
the marks everything else uses: `text` for the words, `path` for the arrow, and
`point` for a dot at a place you choose. The arrow's
table holds data, so gog scales it like any other table. That is why the
arrow's ends are written in the data's own units rather than in pixels.

## Two transforms, and the rest refused

Every path so far has been drawn directly from the rows. The next question is
what happens when a transform changes the rows before the path draws them.
`line`, `step`, `area` and `ribbon` compose with many transforms. `path`
composes with two, `density` and [`cluster`](../cluster.qmd), and refuses every
other transform. The summaries show the reason most clearly. [What it
refuses](#what-it-refuses) shows `path * mean` being refused.

A summary replaces the rows. `mean` reduces each country to one number, and what
is left to connect is one point per country, in whatever order the summary
produced them. One point per country, joined left to right, is what `line`
already draws. A path's meaning is the row order, and a summary discards that
order. That is why this mark refuses almost every transform, and why its row on
the [Combinations](../combinations.qmd#marks-and-transforms) grid is nearly
empty.

`density` is the transform this chapter shows, and [Cluster](../cluster.qmd)
shows the other. Like a summary, `density` replaces the rows. Unlike a summary,
its new rows have an order that means something: they are the points around
each ring, in the order the ring is traced.

## The contour: `path * density`

Where are the flowers most crowded, and how tightly? A scatterplot of these 150
flowers would show two groups. Where several flowers share a position their dots
are drawn on top of each other, and you cannot tell how many are there. A
contour plot draws the shape of the cloud as closed rings, without drawing the
points. Bind both
positions and compose `density`. gog estimates the density of points at every
place on the plane, then traces the rings on which that density is the same:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + path * density +
  x(sepal_length) + y(petal_length) +
  title("where the flowers cluster")
```

*"Given the iris flowers: paths derived by density, x is sepal length, y is
petal length."*

Over two positions a density is a **field**: one estimated value at every place
on the plane. Three marks read that field, and each one draws it
differently:

| Mark derived by `density`, over two positions | Shape |
|---|---|
| `zone * density` | the field as filled cells |
| `surface * density` | the field as a sheet in the cube |
| `path * density` | the field as contour rings, the shape this section draws |

[Surface](surface.qmd) draws all three from one table, one after another.

That contour needed nothing new in the grammar. One rule produced it, and the
same rule makes `zone * bin` draw a [heatmap](zone.qmd). The rule is that the
mark decides how many axes a transform cuts. You never state the number
yourself. In `bar * density` you bind only one position, so one axis is still
free. The density is estimated along the bound axis and drawn on the free one,
as one thin bar at each point where it was estimated. A `path` is a route
through two columns and treats them equally, so it has no axis left over.
`density` therefore cuts both. There is no `contour` atom for the same reason
there is no `pie()` ([Polar](../polar.qmd) makes that argument): the atoms you
already have spell it.

Read the outermost ring, then the ones inside it. The outermost is a *single*
ring around both groups, and the next level inward splits into two separate
rings. One level can therefore be several separate rings. So a level is not one
stroke, the way one `color` value is one line. Each ring is drawn as its own
stroke, and the level decides only its color. The two inner rings are never
joined across the gap between them.

### The parameters

The rings are colored by their level, and you did not bind `color` to get that.
The rows `density` produces carry a new column, `level`, and gog bound `color`
to it for you, the same way `bar * bin` binds `y` for you. Bind it yourself if
you prefer, and change how many rings are traced:

```{r}
data(iris_flowers) + path * density(levels = 10) +
  x(sepal_length) + y(petal_length) + color(level) +
  title("ten levels, named out loud")
```

*"Given the iris flowers: paths derived by density at 10 levels, x is sepal
length, y is petal length, color by level."*

The rings now come in ten levels instead of the default six, so they sit closer
together.

`density` is one transform, and the mark decides what it computes. With `line`
it computes a curve over one column. With `zone` or `path` it computes a field
over two columns: a zone fills the field, and a path draws its rings. The
parameters follow from this. A width suits the curve, and a number of levels
suits the field.

`levels` cuts the field into that many steps. Of the three marks that read the
field, `zone` and `path` accept it, and `surface` refuses it, for a reason
[Surface](surface.qmd) gives. A one-dimensional density is a single curve, and a
curve has no field to cut, so a `line` refuses it.

`bandwidth` is how far each row is spread when the density is estimated, in the
column's own units, so it belongs to the curve over one column. A contour
spreads over two columns, and two columns are usually in different units, such
as dollars and years. One number cannot be a width in both. Sepal length and
petal length are both in centimeters, but gog refuses `bandwidth` here too,
because the rule is about two columns, whatever their units. [What it
refuses](#what-it-refuses) shows both refusals.

`adjust` multiplies the bandwidth gog chooses by itself. It is a plain number
rather than a length, so it means the same thing on either axis. It is the one
parameter `line`, `zone`, `surface` and `path` all accept:

```{r}
data(iris_flowers) + path * density(adjust = 1.6) +
  x(sepal_length) + y(petal_length) +
  title("the same cloud, smoothed harder")
```

*"Given the iris flowers: paths derived by density adjusted by 1.6, x is sepal
length, y is petal length."*

Each flower now spreads further when the density is estimated, so the rings are
smoother.

To estimate one field per category, split the rows with `group`. The level is
already on `color`, which is why the split has to be `group`. Versicolor and
virginica overlap on this plot, so the chunk below first keeps only their rows,
in a table named `petals`:

```{r}
petals <- iris_flowers[iris_flowers$species != "setosa", ]

data(petals) + path * density(levels = 4) +
  x(sepal_length) + y(petal_length) + group(species) +
  title("two species, estimated separately")
```

*"Given the petals: paths derived by density at 4 levels, x is sepal length, y
is petal length, grouped by species."*

Each species has its own set of rings. Both sets take their colors from the
levels, so the lower set, with the shorter petals, is versicolor.

### One field, two marks

A `path` has no axis left over, and a `zone` has none either, so a zone also
reads `density` over both positions. The difference is in the geometry. A path
draws the rings as strokes. A zone given `levels` fills the bands between them,
and without `levels` it fills small cells. Give both marks the same `levels` and
you get the same curves twice: once as strokes, once as the edges of filled
bands. The filled version is the *filled contour*, and [`zone *
density`](zone.qmd) draws it; the zone chapter covers the color ramp it uses.
One field, two marks. It is [the same division of
work](../grammar.qmd#the-operator-combining) as in `bar * bin` and `line * bin`:
the mark chooses the geometry, and the transform stays the same.

## A path in polar

The arrowhead was the first of the three things the row order gives a path. The
second appears in polar, when the same angle appears in more than one row. Some
routes turn: a bearing that moves through all 360 degrees, or a value that
cycles through the hours while another grows. The `spiral` table holds two days
of readings, one every 15 minutes. `a` is the hour of the day, and `r` grows the
whole time:

```{r}
#| echo: false
peek(spiral)
```

On flat axes the route has to jump back at midnight:

```{r}
data(spiral) + path + x(a) + y(r) +
  title("a path on flat axes")
```

*"Given the spiral: a path, x is a, y is r."*

The long stroke across the middle joins 23:45 on the first day to midnight on
the second, and no reading lies along it. On a straight axis the end of one day
is drawn far from the start of the next. On a circle the two are at the same
angle, so the route continues instead of jumping. `path` works in polar like
every other mark:

```{r}
data(spiral) + path + x(a) + y(r) + polar() +
  title("a path in polar")
```

*"Given the spiral: a path, x is a, y is r, in polar."*

The route goes around twice, one day after the other, and turns outward in a
spiral. A `line` cannot draw this. It sorts the rows by `a`, so each moment of
the first day is joined to the same moment of the second, and the stroke
zigzags between them inside one turn:

```{r}
data(spiral) + line + x(a) + y(r) + polar() +
  title("a line in polar")
```

*"Given the spiral: a line, x is a, y is r, in polar."*

There is one thing `path` does not copy from `line`: the segment that closes a
radar chart, like the one in
[Line](line.qmd#a-category-on-the-domain-the-profile). A categorical `line` in
polar joins its last vertex back to its first, because the categories fill the
whole turn, and a gap there would leave one sector blank. A path puts each
category on the same spoke as a line does, but it does not join its last vertex
back to its first. Its last vertex is where the data stopped, so closing the
route would add a segment that is not in the table. If a route does return to
its start, say so by repeating the first row.

## A path in the cube

In polar the path still joined the same rows in the same order. A third axis
does not change that either, because the row order does not depend on any axis.
A glider's route has three positions: east, north and altitude. A flat page
holds only two of them. Drawn flat, two circling gliders are two rings, one
inside the other, and nothing on the page shows that they are rising. The
`thermals` table holds one row for each recorded position of two gliders. Both
circle the same thermal, a rising column of warm air, as they climb:

```{r}
#| echo: false
peek(thermals)
```

Bind `z` and the route runs through the cube instead of across the page. Of
the marks that join their rows into one stroke, `path` is the only one that
takes the third position. The reason is the row order this whole chapter is
about:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  title("Two gliders, one thermal")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider."*

An order is not a property of any axis, so adding a third axis does not change
it. The route joins the same rows in the same sequence, whatever angle the scene
is turned to. A `line` sorts by `x`, and the cube has no left to right. In the
cube `x` is one of three equal positions. At some viewing angles `x` points
straight into the page, so a line would be sorted by an axis the reader cannot
see. So a line's sort by `x` has no meaning in the cube, and `line`, `step`,
`area` and `ribbon` refuse `z` rather than draw a stroke the reader cannot read,
as [What it refuses](#what-it-refuses) shows.

Look again at the two routes in the cube. They cross several times. At each
crossing the route nearer to the eye is drawn over the one behind it, and which
route is nearer changes as the two gliders climb. To draw the cube on a flat
page, gog draws the pieces farthest from the eye first and the nearest last, so
that near pieces cover far ones. This drawing order is the **depth sort**. A
glyph sits at one distance from the eye, so it takes one place in that order. A
stroke through the cube has a far end and a near end, so each segment takes its
own place. The segments of both routes share one depth sort. If each stroke took
one place in the depth sort, one glider's route would be drawn entirely in front
of the other's, and the two coils could not pass through each other as they do
here. An arrowhead takes its place the same way, at the depth of its own tip. A
head far from the eye is covered by whatever is drawn in front of it.

As on a flat line, `color` can vary along the route, because a color belongs to
one place on the stroke rather than to the whole stroke. Bind altitude to
`color` as well as to `z`.
The climb is then said twice: once by where the route is, once by what color it
is:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(altitude) + group(glider) + palette("viridis") +
  space(turn = -25, tilt = 20) +
  title("Altitude, said twice")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
altitude, grouped by glider, with the viridis palette."*

`palette("viridis")` chooses the color ramp that shows the altitudes.
`space(turn = -25, tilt = 20)` sets the viewing angle, which
[Space](../space.qmd#the-viewing-angle) explains; a `z` binding alone would draw
the cube from its default angle. `group(glider)` tells gog that the rows belong
to two routes rather than one. Remove `group(glider)` and the plot is wrong. A
*category* on `color` splits the mark into one stroke per group, but a *measure*
has no categories to split on. Without `group` the two gliders would be joined
into a single route, with one straight segment from the last row of one glider
to the first row of the other. `color` shows a value along the route; `group`
says which route each row belongs to. They are two different questions, so they
need two atoms.

The depth sort draws the route as separate segments. Left alone, the dash
pattern would start again at every segment, so gog continues it from one
segment to the next:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  style(pattern = "dashed") +
  space(turn = -25, tilt = 20)
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider, with pattern dashed."*

The dashes run along the route rather than resetting at every row. The dash
belongs to the route, not to the segments the depth sort needs.

[Space](../space.qmd) owns the third dimension and the viewing angle.

## What a path is for

In the three spaces this chapter used (the flat plane, polar and the cube), the
path did one thing: it joined the rows in the table's order. The connected
scatterplot is the best known use, but the mark is more general than that. A
path is the right mark whenever the reader should follow the rows in the order
the table gives them. A route through space is one, and so is a cycle that
returns to where it began. So is a hysteresis loop, which is a cycle whose
outward and return routes differ, and so is an arrow. The
test is always the same question. Does the order of the rows mean something? If
it does, the mark is `path`; if the reader should read left to right, it is
`line`.

## What you can set

A plain stroke does not show which way the route runs, and it looks like every
other stroke on the panel. The table has no column for the direction or for the
look of the stroke, so both are set rather than mapped, and each mark takes its
own settings. These are a path's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("path")
```

`arrow` belongs to `path` and [`edge`](edge.qmd) only, for the reason [Which
end is the last row?](#which-end-is-the-last-row) gave.

A route can carry the heads, a width and a dash together:

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "both", size = 2.5, pattern = "dotted") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both ends marked, on a dotted route")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
both, size 2.5 and pattern dotted."*

`arrow` takes a *value* rather than a true-or-false flag. `"both"` is an
ordinary thing to ask for, and a flag would have needed a second setting to say
it. The heads are drawn solid on a dotted path. The dash is the route's texture,
and a head cut into dashes no longer reads as an arrow.

This chapter covers the parameters of one transform, `density`: `levels`,
`bandwidth` and `adjust`, in [The parameters](#the-parameters).
[Cluster](../cluster.qmd) covers those of `cluster`. Every other property of a
path is either mapped to a column or set with `style()`.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

You might want the route to thicken as income rises, so that the stroke itself
carries a third measure.
`size` reads one value from every row. A path joins all of its rows into a
single stroke, and a stroke is drawn at one width from end to end. There is
nowhere to put a second width:

```{r}
#| error: true
data(gapminder_asia) + path + x(gdp) + y(life) + size(gdp)
```

Set it instead and every segment takes the same width, which is the [Setting vs
mapping](../style.qmd) distinction again.

A summary is the second refusal. A path draws the rows in their order, and a
summary replaces those rows, as [Two transforms, and the rest
refused](#two-transforms-and-the-rest-refused) explained:

```{r}
#| error: true
data(gapminder_asia) + path * mean + x(country) + y(life)
```

`density(bandwidth = )` is the third. It is a width in one column's own units,
and the two columns of a contour are usually in different units, so one number
cannot be a width in both:

```{r}
#| error: true
data(iris_flowers) + path * density(bandwidth = 0.5) +
  x(sepal_length) + y(petal_length)
```

Three more refusals follow. In each one, `line` refuses something that `path`
accepts, and two of them come from the sort. An arrowhead is the first of them.
A line sorts its rows by `x`, so its last vertex is simply the largest `x` in
the table. A head there would point at the result of the sort, not at the last
measurement taken:

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + style(arrow = "end")
```

Here `x` is the year, so the sort and the table order agree, and the head would
land on 2007. gog refuses it anyway, because on a line the end of the stroke
always comes from the sort.

`density(levels = )` is the second. It cuts the estimated field into steps, and
a one-dimensional density is a single curve with no field to cut:

```{r}
#| error: true
data(iris_flowers) + line * density(levels = 8) + x(sepal_length)
```

`z` is the third. A line sorts by `x`, and the cube has no left to right, so
the sort has no meaning in the cube, as [A path in the
cube](#a-path-in-the-cube) explained:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude)
```

Two of the three have one cause: a line's shape comes from the sort, and a
path's comes from the rows. The third has another. A curve over one column has
no field to cut into levels.

## In short

A path joins the rows in the order the table holds them, so sorting the table
changes the picture. That order is the mark's meaning. The arrowhead belongs to
`path` and `edge` only, and the two transforms a path takes are `density` and
`cluster`.

## What comes next

Every part of a path comes from the data. [Rule](rule.qmd) draws a different
kind of stroke, one that crosses the whole panel at a value you give it.
