# Path

```{r}
#| include: false
source("../R/setup.R")
```

How did income and life expectancy change together in each country, year by
year? `path` strokes the rows *in the order the table gives them*. That is all
of it, and it is the one thing `line` does not do. A line sorts its vertices by
`x` first, because a line draws a *function*: one `y` for each `x`, read left to
right along a domain. A path draws a *route*. It visits row 1, then row 2, and
it may turn back, cross itself, or return to where it started.

Everything the two marks share, they share exactly: same stroke, same `color`
and `group` splits, same `style(size = )`, same `style(pattern = )` dash. One
question separates them: does the order of the rows mean something? Because a
path keeps that order, it can carry three things a `line` cannot in general: an
arrowhead, a spiral, and the third dimension. All three follow from the row
order, so none of them is a special case.

## The connected scatterplot

Two measures that change over time can be drawn as two time series, one above
the other. The reader is then left to match the years by eye. One plane can
hold both, if the order of the rows carries the time.
Here are five Asian countries, each measured twelve times between 1952 and 2007.
Put income on one axis and life expectancy on the other, and connect each
country's measurements in the order they were taken:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Fifty-five years, five countries, one plane")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country."*

Each route is one country's history. Time is nowhere on the plot, and yet the
whole plot is about time: it is the order the points are joined in. Japan climbs
steeply, then turns right and flattens: past that corner, income keeps rising
and life expectancy barely moves. Korea's route bends in the same place two
decades later. China's route goes almost straight up at very low income. That
shape belongs to a country that got healthier long before it got rich, and
Indonesia's route bends the same way. This is a **connected scatterplot**, and
no other arrangement of these atoms draws it.

## Why `line` cannot draw it

`line` is the mark most readers would write first, because it is the familiar
way to join rows with a stroke. It is worth seeing what `line` does with this
table.
Change one word and the same sentence draws the wrong picture:

```{r}
data(gapminder_asia) + line + x(gdp) + y(life) + color(country) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("the same data, sorted by x")
```

*"Given gapminder Asia: lines, x is gdp, y is life, color by country."*

Look at China's route. Its income did not rise steadily: it fell and recovered,
so the year order and the income order are different sequences, and `line` draws
the income one. Every point is in the right place, and every route between them
is wrong. That is not a bug in `line`; it is what a line *is*. A function has
one value per `x`, so sorting changes nothing. For a time series read along a
date axis, the sort puts the rows in date order.

The two marks answer different questions, so a specification that names the
wrong one gets a plausible picture rather than an error. The rule is this. If
the reader should follow the marks in a particular order, that order has to
come from the table. Only `path` reads it there.

## Which end is the last row?

A route has a direction, and the plot above does not show it. Japan's route
could be read from either end. So `path` takes a setting no other mark can:

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("style(arrow = \"end\"): the head sits at 2007")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
end."*

Now each route points at its last row. `arrow` takes `"end"`, `"start"`, or
`"both"`: a head at the last row, at the first row, or at both ends.

It belongs to `path` alone, and every other mark refuses it. The refusal is
worth reading, because it states why `path` is a separate mark:

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + style(arrow = "end")
```

A line's last vertex is wherever the domain happens to end, so a head there
marks the sort rather than anything the data did. Only a path has an end that
comes from the data.

## An arrow is a layer

A note placed in empty space has to say which point it is about. An arrow from
the words to the point says it.
The [Text](text.qmd) chapter drew the note; `path` draws the arrow beside it. A
note is a one-row table; the arrow that points from the note to the thing is a
*two-row* table, one row for each end:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
seg  <- data.frame(gdp = c(20000, 13100), life = c(45.5, 50.2))
note <- data.frame(gdp = 20500, life = 44.8, what = "Botswana")

data(gapminder_2007) + point + x(gdp) + y(life) +
  data(seg)  + path + style(arrow = "end", color = "#c0392b") +
  data(note) + text + label(what) +
  title("an annotation, in the grammar you already have")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also a path from the
seg table and text from the note table, label by what."*

This grammar has no `annotate()` function, and it needs none. An annotation is a
layer over a small table, and the marks it uses are the marks everything else
uses: `text` for the words, `path` for the arrow, `point` for a marker. What the
table holds is data, so it is scaled like data, which is why the arrow's ends
are written in the data's own units rather than in pixels.

## Two transforms, and the rest refused

A summary such as `mean` composes with `line`. Writing the same sentence with
`path` is a natural next step.
The other stroke and band marks compose with many transforms. `path` composes
with two, `density` and [`cluster`](../cluster.qmd), and the refusals say
why:

```{r}
#| error: true
data(gapminder_asia) + path * mean + x(country) + y(life)
```

A summary replaces the rows. `mean` reduces each country to one number, and what
is left to connect is one point per country, in whatever order the summary
produced them. That is a `line`, written the long way around. The order a path
*is* does not survive being summarized, and that one fact empties almost the
whole row this mark has on the [Combinations](../combinations.qmd) grid.

`density` is the one exception, and it does not break the rule above. A
summary reduces each group to one value. A density estimated over the plane
groups nothing and reduces nothing. It produces vertices *in the order they
were traced*, which is the one thing a path is for.

## The contour: `path * density`

Where do the flowers cluster, and how tightly? A scatter of 150 flowers shows
two groups, and points that share a position hide their count. A contour draws
the shape of the cloud without drawing the points.
Bind both positions and compose `density`. gog estimates how thick the cloud of
points is everywhere on the plane, then traces the rings where it is equally
thick:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + path * density +
  x(sepal_length) + y(petal_length) +
  title("where the flowers cluster")
```

*"Given the iris flowers: paths derived by density, x is sepal length, y is
petal length."*

Nothing new was added to the grammar to draw that. One rule decides this, and
the same rule turns `zone * bin` into a [heatmap](zone.qmd). The mark decides
how many axes a transform cuts, and you never state it. A `bar` leaves an axis
free to measure the density along, so `bar * density` spreads it along one axis
and draws a curve. A `path` is a route through two columns and treats them
equally, so it has no axis left over. `density` therefore cuts both. There is no
`contour` atom for the same reason there is no `pie()` ([Polar](../polar.qmd)
makes that argument): the sentence already exists.

Read the outermost ring, then the ones inside it. The outermost is a *single*
ring around both clusters, and the next level inward splits into two separate
rings. So a level alone cannot tell gog where one stroke ends. Each ring is its
own stroke, and the level only decides its color, so the two inner rings are
never joined by a stroke across the gap between them.

The color is the level, and nothing had to bind it, exactly as `bar * bin`
binds `y` for you. Say it out loud if you prefer, and change how many
rings are traced:

```{r}
data(iris_flowers) + path * density(levels = 10) +
  x(sepal_length) + y(petal_length) + color(level) +
  title("ten levels, named out loud")
```

*"Given the iris flowers: paths derived by density at 10 levels, x is sepal
length, y is petal length, color by level."*

`density` is one transform read three ways. A `line` draws a curve, a `zone`
paints the whole field, and a `path` strokes the level sets that `levels` cuts
the field into. Each parameter belongs to one reading and is refused in the
other two. `levels` cuts the **field**, the estimated thickness everywhere on
the plane, into that many steps. A one-dimensional density is a single curve,
and a curve has no field to cut.

```{r}
#| error: true
data(iris_flowers) + line * density(levels = 8) + x(sepal_length)
```

`bandwidth` is refused for the mirror reason: it needs one axis, and a contour
has two. It is a width in the data's own units, and a contour spreads over two
columns that measure different things. Here the two agree, since sepal length
and petal length are both in centimeters, but two axes usually do not. One
number cannot be a width in both:

```{r}
#| error: true
data(iris_flowers) + path * density(bandwidth = 0.5) +
  x(sepal_length) + y(petal_length)
```

`adjust` is a multiplier rather than a length, so it means the same thing on
either axis. It is the one parameter all three readings accept:

```{r}
data(iris_flowers) + path * density(adjust = 1.6) +
  x(sepal_length) + y(petal_length) +
  title("the same cloud, smoothed harder")
```

*"Given the iris flowers: paths derived by density adjusted by 1.6, x is sepal
length, y is petal length."*

To estimate one field per category, split the rows with `group`. The level is
already on `color`, which is why the split has to be `group`:

```{r}
petals <- iris_flowers[iris_flowers$species != "setosa", ]

data(petals) + path * density(levels = 4) +
  x(sepal_length) + y(petal_length) + group(species) +
  title("two species, estimated separately")
```

*"Given the petals: paths derived by density at 4 levels, x is sepal length, y
is petal length, grouped by species."*

A `zone` has no measure axis either, so it reads `density` in two dimensions
too, and it *fills* what this mark strokes. Give both marks the same `levels`
and you get the same curves twice: once as strokes, once as the edges of filled
bands. That pair is the plot usually meant by a *filled* density. That plot
belongs to the mark that fills it, [`zone * density`](zone.qmd), and the zone
chapter covers the color ramp it uses. One field, two marks. It is [the same
split](../grammar.qmd) as `bar * bin` and `line * bin`: the mark chooses the
geometry, and the transform stays the same.

## In polar, a route is a spiral

Some routes turn: a bearing that swings around the compass, or a value that
cycles through the hours while another grows. On a straight axis the end of one
turn is far from the start of the next, and on a circle they meet.
`path` bends into polar like every other mark. Give it an angle and a radius
that both grow, and the route winds outward:

```{r}
spiral <- data.frame(a = seq(0, 6.2, length.out = 90))
spiral$r <- seq(0.1, 4, length.out = 90)

data(spiral) + path + x(a) + y(r) + polar() +
  title("a path in polar")
```

*"Given the spiral: a path, x is a, y is r, in polar."*

There is one thing `path` does not copy from `line`: the segment that closes a
radar chart ([Polar](../polar.qmd) draws one). A categorical `line` in polar
joins its last vertex back to its first, because the categories fill the whole
turn, and a gap there would leave one sector blank. A path does not read the
angular axis as a set of categories. Its last vertex is where the data stopped,
so closing the route would add a segment nobody asked for. If a route does
return to its start, say so by repeating the first row.

## In space, the route is unchanged

A glider's route has three positions, and a plane holds two of them. Drawn
flat, two circling gliders are two rings, one inside the other, and the climb
is lost.
The `thermals` table holds two gliders circling the same thermal, a rising
column of warm air, sampled as they climb:

```{r}
#| echo: false
peek(thermals)
```

Bind `z` and the route runs through the cube instead of across the page. Of
the marks that join their rows into one stroke, `path` is the only one that
takes the third position. The reason is the row order this whole chapter is
about:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  title("Two gliders, one thermal")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider."*

An order belongs to no axis, so it survives the third dimension untouched: the
route is the same route however the scene is turned. A `line` sorts by `x`, and
the cube has no left to right. There `x` is one of three equal positions. At
some viewing angles it runs straight into the page and becomes depth, which
would leave a line sorted by an axis the reader cannot see. So the sort that
makes a `line` a line cannot follow it into the cube. `line`, `step`, `area` and
`ribbon` refuse `z` rather than half-drawing it:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude)
```

Watch the two routes where they cross: whichever is nearer to the eye covers the
other, and which one that is changes as they climb. A glyph sits at one distance
from the eye and can be sorted whole. A stroke running through the cube has a
far end and a near end, so the unit that gets sorted is the **segment**. The
segments of both routes are sorted together. If each stroke were sorted as a
whole, one glider would sit entirely in front of the other: two coils that never
meet, rather than two that thread through each other.

Cutting the route into segments is also what lets `color` carry a **measure**
along it. Bind the altitude to the color as well as to the height, and the climb
is said twice: once by where the route is, once by what color it is:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(altitude) + group(glider) + palette("viridis") +
  space(turn = -25, tilt = 20) +
  title("Altitude, said twice")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
altitude, grouped by glider, with the viridis palette."*

`palette("viridis")` picks the ramp; `group(glider)` keeps the two routes apart.
Drop that last atom and the plot is wrong. A *category* on `color` splits the
mark into one stroke per group, but a *measure* has no categories to split on.
Without `group` the two gliders would be joined into a single route, and a
straight line would run from where one glider's rows stop to where the other's
begin. Color answers what the route was carrying; `group` answers which route it
is. Two questions, two atoms.

A dash survives the segmentation too. Each segment is drawn as its own piece,
and a dash would normally restart at the start of each piece:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  style(pattern = "dashed") +
  space(turn = -25, tilt = 20)
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider, with pattern dashed."*

The dashes run along the route rather than resetting at every sample. The dash
belongs to the route, not to the segments the depth sort needs. An arrowhead is
sorted the same way, at the depth of its own tip. A head that sits deep in the
cube is covered by whatever passes in front of it.

[Space](../space.qmd) owns the third dimension and the viewing angle.

## What a path is for

The connected scatterplot is the best known use, but the mark is more general
than that. Any picture whose reading order is the table's order is a path. A
trajectory through two measurements over time is one. So is a route through
space, and a cycle that returns to where it began. So is a hysteresis loop,
which is a cycle whose forward and return paths differ, and so is an arrow. The
test is always the same question. Does the order of the rows mean something? If
it does, the mark is `path`; if the reader should read left to right, it is
`line`.

## What you can set

A route drawn as a plain stroke leaves the reader two questions. Which way does
it run, and is it solid or dashed? Neither answer is in a column, so both are
set rather than mapped, and each mark takes its own settings. These are a
path's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("path")
```

`arrow` belongs to `path` alone, and geometry decides that rather than taste. A
head marks a direction, and a path is the only mark that has one. `line`, `step`
and `area` sort their vertices by `x`, so their last point is wherever the
domain ends rather than where the data stopped.

A route can carry the heads, a width and a dash together:

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "both", size = 2.5, pattern = "dotted") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both ends marked, on a dotted route")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
both, size 2.5 and pattern dotted."*

It takes a *value* rather than a true-or-false flag, because `"both"` is an
ordinary thing to ask for, and a flag would have needed a second setting to say
it. The heads draw solid on a dotted path. The dash is the route's texture, and
a head cut into dashes no longer reads as an arrow.

The transform whose parameters this chapter covers is `density`, and `levels`
and `adjust` are above. `cluster` takes its own, in
[Cluster](../cluster.qmd). Everything else that varies is either mapped or
set.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

You might want the route to thicken as income rises, so that the stroke itself
carries a third measure.
A path is one stroke, and a stroke has one width. There is no answer to how
thick the route is at a single point, so width cannot vary along it:

```{r}
#| error: true
data(gapminder_asia) + path + x(gdp) + y(life) + size(gdp)
```

Set it instead and every segment takes the same width, which is the [Setting vs
mapping](../style.qmd) distinction again. The refusals that belong to `path`
alone are stated above rather than here. One is the summaries it cannot take,
because a summary replaces the very rows whose order *is* the path. One is
`density(bandwidth = )`, which is a width in one column's units and cannot be a
width in two. The last is an arrow asked of `line`, which has no end for a head
to sit on.