# Path

```{r}
#| include: false
source("../R/setup.R")
```

How did income and life expectancy change together in each country, year by
year? `path` strokes the rows *in the order the table gives them*. That is all
of it, and it is the one thing `line` does not do. A line sorts its vertices by
`x` first, because a line draws a *function*: one `y` for each `x`, read left to
right along a domain. A path draws a *route*. It visits row 1, then row 2, and
it may turn back, cross itself, or return to where it started.

Everything the two marks share, they share exactly: same stroke, same `color`
and `group` splits, same `style(size = )`, same `style(pattern = )` dash. One
question separates them: does the order of the rows mean something? A path keeps
that order, and three things follow from it. It can carry an arrowhead, because
it has a last row. It can wind into a spiral, because it may pass beyond where
it started. And it can run through the cube, because an order needs no axis.
None of the three is a special case.

## The connected scatterplot

Two measures that change over time can be drawn as two separate time series,
one above the other. The reader then has to find the same year in both plots
before comparing them. One plot can carry both measures, if the order of the
rows supplies the time.
Here are five Asian countries, each measured twelve times from 1952 to 2007.
Put income on one axis and life expectancy on the other, and connect each
country's measurements in the order they were taken:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Fifty-five years, five countries, one plane")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country."*

Each route is one country's history. Time is nowhere on the plot, and yet the
whole plot is about time: it is the order the points are joined in. China's
route goes almost straight up at very low income. That shape belongs to a
country whose life expectancy rose long before its income did, and India's and
Indonesia's routes rise the same way. Korea's route starts upward and then turns a sharp
corner to the right. Past that corner its income climbs fast and its life
expectancy rises more slowly. Japan's route is the flattest of the five, and it
begins at the point where Korea turns, two decades earlier. This is a
**connected scatterplot**, and no other arrangement of these atoms draws it.

## Why `line` cannot draw it

`line` is the mark most readers would write first, because it is the familiar
way to join rows with a stroke. So it is worth seeing what `line` draws from the
same table. One word changes in the specification, and the plot changes
completely:

```{r}
data(gapminder_asia) + line + x(gdp) + y(life) + color(country) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("the same data, sorted by x")
```

*"Given gapminder Asia: lines, x is gdp, y is life, color by country."*

Look at China's route. Its income did not rise steadily: it fell and recovered,
so the year order and the income order are different sequences, and `line` draws
the income one. Every point is still in the right place, and the route between
them is not. Three of the five routes change this way. India's and Korea's
income rose every year, so for them the two orders agree and the picture is the
same either way. This is not a defect in `line`. A line draws a function, and a
function has one value for each `x`. Sorting the rows by `x` cannot change which
value belongs to which `x`. For a time series read along a date axis, the sort
simply puts the rows in date order.

The two marks answer different questions. If a specification names the wrong
one, the engine still draws a plot, and the plot still looks reasonable. Nothing
reports the mistake. The rule is this. If the reader should follow the marks in
a particular order, that order has to come from the table. Only `path` reads it
there.

## Which end is the last row?

A route has a direction, and a plain stroke does not show it. Japan's route
could be read from either end. So `path` takes a setting no other mark can:

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("style(arrow = \"end\"): the head sits at 2007")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
end."*

Now each route points at its last row. `arrow` takes `"end"`, `"start"`, or
`"both"`: a head at the last row, at the first row, or at both ends.

It belongs to `path` alone, and every other mark refuses it. A line's last
vertex is wherever the domain happens to end, so a head there marks the sort
rather than anything the data did. Only a path has an end that comes from the
data. [What it refuses](#what-it-refuses) shows a line refusing an arrowhead.

## An arrow is a layer

A note placed in empty space has to say which point it is about. An arrow from
the words to the point says it.
The [Text](text.qmd) chapter drew the note; `path` draws the arrow beside it. A
note is a one-row table; the arrow that points from the note to the thing is a
*two-row* table, one row for each end:

```{r}
#| echo: false
peek(gapminder_2007)
peek(botswana_arrow)
peek(botswana_label)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(botswana_arrow)  + path + style(arrow = "end", color = "#c0392b") +
  data(botswana_label) + text + label(what) +
  title("an annotation, in the grammar you already have")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also a path from the
Botswana arrow, and text from the Botswana label, label by what."*

This grammar has no `annotate()` function, and it needs none. An annotation is a
layer over a small table. The marks it uses are the marks everything else uses:
`text` for the words, `path` for the arrow, `point` for a marker. What the
table holds is data, so it is scaled like data, which is why the arrow's ends
are written in the data's own units rather than in pixels.

## Two transforms, and the rest refused

Everything so far has been a path drawn straight from the rows. The next
question is what happens when a transform stands between the table and the
stroke. `line`, `step`, `area` and `ribbon` compose with many transforms. `path`
composes with two, `density` and [`cluster`](../cluster.qmd), and every summary
is refused. [What it refuses](#what-it-refuses) shows `path * mean` being
refused.

A summary replaces the rows. `mean` reduces each country to one number, and what
is left to connect is one point per country, in whatever order the summary
produced them. That is a `line`, and `line` already draws it. A path's meaning
is the row order, and a summary throws that order away. That is why this mark
refuses almost every transform, and why its row on the
[Combinations](../combinations.qmd) grid is nearly empty.

`density` is the one exception, and it does not break that rule. A
summary reduces each group to one value. A density estimated over the plane
groups nothing and reduces nothing. It produces vertices *in the order they
were traced*, which is the one thing a path is for.

## The contour: `path * density`

Where do the flowers cluster, and how tightly? A scatter of 150 flowers shows
two groups. Where several flowers share a position their dots are drawn on top
of each other, and you cannot tell how many are there. A contour draws the shape
of the cloud without drawing the points.
Bind both positions and compose `density`. gog estimates the density of points
at every place on the plane, then traces the rings where that density is equal:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + path * density +
  x(sepal_length) + y(petal_length) +
  title("where the flowers cluster")
```

*"Given the iris flowers: paths derived by density, x is sepal length, y is
petal length."*

Over two positions a density is a **field**: an estimated thickness at every
place on the plane. Three marks read that field, and each one draws it
differently:

| Mark derived by `density`, over two positions | Shape |
|---|---|
| `zone * density` | the field as filled cells |
| `surface * density` | the field as a sheet in the cube |
| `path * density` | the field as contour rings, the shape this section draws |

[Surface](surface.qmd) draws all three from one table, side by side.

Nothing new was added to the grammar to draw that. One rule produced that
contour, and the same rule turns `zone * bin` into a [heatmap](zone.qmd). The
rule is that the mark decides how many axes a transform cuts. You never state
the number yourself. A `bar`
binds only one position, so one axis is still free. `bar * density` reads the
density over the bound axis and writes the result to the free one, which draws a
curve. A `path` is a route through two columns and treats them equally, so it
has no axis left over. `density` therefore cuts both. There is no `contour` atom
for the same reason there is no `pie()` ([Polar](../polar.qmd) makes that
argument): the atoms you already have spell it.

Read the outermost ring, then the ones inside it. The outermost is a *single*
ring around both clusters, and the next level inward splits into two separate
rings. One level can therefore be several separate rings, and the level alone
cannot say where a stroke ends. Each ring is drawn as its own stroke, and the
level decides only its color. The two inner rings are never joined across the
gap between them.

### The parameters

The rings are colored by their level, and you did not bind `color` to get that.
gog bound it for you, the same way `bar * bin` binds `y` for you. Bind it
yourself if you prefer, and change how many rings are traced:

```{r}
data(iris_flowers) + path * density(levels = 10) +
  x(sepal_length) + y(petal_length) + color(level) +
  title("ten levels, named out loud")
```

*"Given the iris flowers: paths derived by density at 10 levels, x is sepal
length, y is petal length, color by level."*

`density` is one transform, and the mark decides how it is read. A `line` draws
a curve over one column. A `zone` and a `path` both read a field over two
columns, one filling it and the other stroking its level sets. Which parameters
a sentence accepts follows from that, and the dividing line is one column
against two.

`levels` cuts the field into that many steps. Both readings of the field take
it, so `zone * density`
and `path * density` each accept `levels`. A one-dimensional density is a single
curve, and a curve has no field to cut, so a `line` refuses it.

`bandwidth` divides the other way. It is a width in one column's own units, so
only the one-column reading can take it. A contour spreads over two columns that
measure different things. Here the two agree, since sepal length and petal
length are both in centimeters, but two axes usually do not. One number cannot
be a width in both. [What it refuses](#what-it-refuses) shows both refusals.

`adjust` is a multiplier rather than a length, so it means the same thing on
either axis. It is the one parameter every reading accepts:

```{r}
data(iris_flowers) + path * density(adjust = 1.6) +
  x(sepal_length) + y(petal_length) +
  title("the same cloud, smoothed harder")
```

*"Given the iris flowers: paths derived by density adjusted by 1.6, x is sepal
length, y is petal length."*

To estimate one field per category, split the rows with `group`. The level is
already on `color`, which is why the split has to be `group`:

```{r}
petals <- iris_flowers[iris_flowers$species != "setosa", ]

data(petals) + path * density(levels = 4) +
  x(sepal_length) + y(petal_length) + group(species) +
  title("two species, estimated separately")
```

*"Given the petals: paths derived by density at 4 levels, x is sepal length, y
is petal length, grouped by species."*

### One field, two marks

A `zone` has no measure axis either, so it reads `density` over both positions,
just as `path` does. The difference is in the geometry: a path strokes the
rings, and a zone fills the bands between them. Give both marks the same
`levels`
and you get the same curves twice: once as strokes, once as the edges of filled
bands. That pair is the plot usually meant by a *filled* density. That plot
belongs to the mark that fills it, [`zone * density`](zone.qmd), and the zone
chapter covers the color ramp it uses. One field, two marks. It is [the same
split](../grammar.qmd) as `bar * bin` and `line * bin`: the mark chooses the
geometry, and the transform stays the same.

## In polar, a route is a spiral

A route does not have to run across a flat panel. Some routes turn: a bearing
that moves through all 360 degrees, or a value that cycles through the hours
while another grows. On a straight axis the end of one turn is drawn far from
the start of the next. On a circle the two land at the same angle, so the route
continues instead of jumping.
`path` bends into polar like every other mark. Give it an angle and a radius
that both grow, and the route winds outward:

```{r}
#| echo: false
peek(spiral)
```

```{r}
data(spiral) + path + x(a) + y(r) + polar() +
  title("a path in polar")
```

*"Given the spiral: a path, x is a, y is r, in polar."*

There is one thing `path` does not copy from `line`: the segment that closes a
radar chart ([Polar](../polar.qmd) draws one). A categorical `line` in polar
joins its last vertex back to its first, because the categories fill the whole
turn, and a gap there would leave one sector blank. A path does not read the
angular axis as a set of categories. Its last vertex is where the data stopped,
so closing the route would add a segment that is not in the table. If a route
does return to its start, say so by repeating the first row.

## In space, the route is unchanged

Bending the plane into a circle left the route itself untouched. Adding a third
axis does the same, and for the same reason. A glider's route has three
positions: east, north and altitude. A flat page holds only two of them. Drawn
flat, two circling gliders are two rings, one inside the other, and nothing on
the page shows that they are rising.
The `thermals` table holds two gliders circling the same thermal, a rising
column of warm air, sampled as they climb:

```{r}
#| echo: false
peek(thermals)
```

Bind `z` and the route runs through the cube instead of across the page. Of
the marks that join their rows into one stroke, `path` is the only one that
takes the third position. The reason is the row order this whole chapter is
about:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  title("Two gliders, one thermal")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider."*

An order is not a property of any axis, so adding a third axis does not change
it. The route joins the same rows in the same sequence, whatever angle the scene
is turned to. A `line` sorts by `x`, and
the cube has no left to right. There `x` is one of three equal positions. At
some viewing angles it runs straight into the page and becomes depth, which
would leave a line sorted by an axis the reader cannot see. So the sort that
makes a `line` a line cannot follow it into the cube. `line`, `step`, `area` and
`ribbon` refuse `z` rather than drawing part of it, as [What it
refuses](#what-it-refuses) shows.

Look again at the two routes in the cube. They cross several times. At each
crossing the route nearer to the eye is drawn over the one behind it, and which
route is nearer changes as the two gliders climb. A glyph occupies one distance
from the eye, so it can be sorted as a whole. A stroke running through the cube
has a far end and a near end, so the unit that gets sorted is the **segment**.
The segments of both routes are sorted together. If each stroke were sorted as a
whole, one glider would sit entirely in front of the other. The two coils would
never meet, instead of winding through each other. An arrowhead is sorted the
same way, at the depth of its own tip. A head far from the eye is covered by
whatever is drawn in front of it.

Each segment is a separate element in the drawing, so each can take a color of
its own. That is what lets `color` read a **measure** along the route instead of
giving the whole stroke one color. Bind the altitude to the color as well as to
the height. The climb is
then said twice: once by where the route is, once by what color it is:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) +
  color(altitude) + group(glider) + palette("viridis") +
  space(turn = -25, tilt = 20) +
  title("Altitude, said twice")
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
altitude, grouped by glider, with the viridis palette."*

`palette("viridis")` chooses the color ramp the altitudes are read on.
`group(glider)` tells gog that the rows belong to two routes rather than one.
Remove `group(glider)` and the plot is wrong. A *category* on `color` splits the
mark into one stroke per group, but a *measure* has no categories to split on.
Without `group` the two gliders would be joined into a single route, and a
straight line would run from where one glider's rows stop to where the other's
begin. Color answers what the route was carrying; `group` answers which route it
is. They are two different questions, so they need two atoms.

A dash survives the segmentation too. Each segment is drawn as its own piece,
and a dash would normally restart at the start of each piece:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  style(pattern = "dashed") +
  space(turn = -25, tilt = 20)
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider, with pattern dashed."*

The dashes run along the route rather than resetting at every sample. The dash
belongs to the route, not to the segments the depth sort needs.

[Space](../space.qmd) owns the third dimension and the viewing angle.

## What a path is for

The connected scatterplot is the best known use, but the mark is more general
than that. A path is the right mark whenever the reader should follow the marks
in the order the table gives them. A
trajectory through two measurements over time is one. So is a route through
space, and a cycle that returns to where it began. So is a hysteresis loop,
which is a cycle whose forward and return paths differ, and so is an arrow. The
test is always the same question. Does the order of the rows mean something? If
it does, the mark is `path`; if the reader should read left to right, it is
`line`.

## What you can set

A route drawn as a plain stroke leaves the reader two questions. Which way does
it run, and is it solid or dashed? Neither answer is in a column, so both are
set rather than mapped, and each mark takes its own settings. These are a
path's, with the values each accepts. The two this chapter explains are `arrow`
and `pattern`:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("path")
```

`arrow` belongs to `path` alone, and geometry decides that rather than taste. A
head marks a direction, and a path is the only mark that has one. `line`,
`step`, `area` and `ribbon` sort their vertices by `x`, so their last point is
wherever the domain ends rather than where the data stopped.

A route can carry the heads, a width and a dash together:

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "both", size = 2.5, pattern = "dotted") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both ends marked, on a dotted route")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
both, size 2.5 and pattern dotted."*

It takes a *value* rather than a true-or-false flag. `"both"` is an ordinary
thing to ask for, and a flag would have needed a second setting to say it. The
heads draw solid on a dotted path. The dash is the route's texture, and
a head cut into dashes no longer reads as an arrow.

The transform whose parameters this chapter covers is `density`, and [The
contour](#the-contour-path-density) covers `levels`, `bandwidth` and `adjust`.
`cluster` takes its own, in [Cluster](../cluster.qmd). Everything else that
varies is either mapped or set.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

You might want the route to thicken as income rises, so that the stroke itself
carries a third measure.
`size` reads one value from every row. A path joins all of its rows into a
single stroke, and a stroke is drawn at one width from end to end. There is
nowhere to put a second width:

```{r}
#| error: true
data(gapminder_asia) + path + x(gdp) + y(life) + size(gdp)
```

Set it instead and every segment takes the same width, which is the [Setting vs
mapping](../style.qmd) distinction again.

A summary is the second refusal. It replaces the rows whose order is the
path, as [Two transforms, and the rest
refused](#two-transforms-and-the-rest-refused) explained:

```{r}
#| error: true
data(gapminder_asia) + path * mean + x(country) + y(life)
```

`density(bandwidth = )` is the third. It is a width in one column's own units,
and a contour spreads over two columns that measure different quantities, so
one number cannot be a width in both:

```{r}
#| error: true
data(iris_flowers) + path * density(bandwidth = 0.5) +
  x(sepal_length) + y(petal_length)
```

Three more refusals follow, and all three belong to `line` rather than to
`path`. Each one says why the two are separate marks. An arrowhead is the first
of them. A line sorts its rows by `x`, so its last vertex is simply the largest
`x` in the table. A head there would point at the result of the sort, not at the
last measurement taken:

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + style(arrow = "end")
```

`density(levels = )` is the second. It cuts the estimated field into steps, and
a one-dimensional density is a single curve with no field to cut:

```{r}
#| error: true
data(iris_flowers) + line * density(levels = 8) + x(sepal_length)
```

`z` is the third. A line sorts by `x`, and the cube has no left to right, so
that sort cannot follow it in, as [In space, the route is
unchanged](#in-space-the-route-is-unchanged) explained:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude)
```

All three have the same cause. A line's shape comes from the sort, and a path's
comes from the rows.

## In short

A path joins the rows in the order the table holds them, so sorting the table
changes the picture. That order is the mark's meaning. The arrowhead belongs to
`path` alone, and the two transforms it takes are `density` and `cluster`.

## What comes next

A path follows the rows in the order the table gives them, so every part of it
comes from data. [Rule](rule.qmd) draws a different kind of stroke, one that
crosses the whole panel at a value you give it.
