# Point

```{r}
#| include: false
source("../R/setup.R")
```

How does one measurement change as another changes? `point` draws one glyph
per row of data at the `(x, y)` position.

The plot it makes is the **scatterplot**, the last of the plain chart types
to arrive. Playfair had the line and the bar in 1786, and the pie by 1801;
the scatterplot came three decades later. A picture that plots one measured
column against another is usually dated to John Herschel. In 1833 he plotted
the position of twin stars against the year, then drew a smooth curve through
the dots so the trend could be seen [@friendly2005]. Other names for it are the same picture:
scatter diagram, scattergram, and, once `size` carries a third column, the
bubble chart.

None of those is an atom here, and the reason is the design itself. `point` names
the **mark**, not the chart. `point + x(gdp) + y(life)` is a scatterplot and
`point + x(continent) + y(life)` is a strip plot, and nothing about the mark
changed between them: the chart name is a property of the sentence, not a word in
the vocabulary. That is why the names live in an appendix, [Chart
names](../chart-names.qmd), instead of in the kernel.

## Basic usage

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

## The dot is a default, not the mark

Write `point` and you get a dot, and treating it as *the dot mark* is right
almost every time. But the dot is a **default**, not the mark's whole
identity. A point is *a glyph at a position*, and the glyph is yours to change:
`shape` swaps the dot for one of five.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + shape(continent) +
  title("The glyph is not always a dot")
```

So "a dot" really means "the first of five shapes, drawn when you name none."
That is what makes `point` and [`text`](text.qmd) **siblings**: both draw one
glyph per row at `(x, y)`, differing only in *where the glyph comes from*. A
point picks its glyph from the closed set of five (`shape`); a text mark takes it
from a column: the string itself (`label`). Put both on one plot and each row is
located by a dot and named by its label:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("point and text, siblings on one plot")
```

## A categorical x: the strip plot

`x` accepts a text column as readily as a numeric one. Each category gets a
position on the axis and every row is drawn at it, showing the whole
distribution rather than a summary of it:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") + title("One dot per country")
```

This is the same rule `bar` uses for a categorical axis, no per-mark special
case. Compare it with `bar * mean`, which collapses each group to one number:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("The same data, summarized")
```

The strip plot shows Africa's spread and its overlap with Asia; the bar chart
shows neither. The strip plot's one weakness: points at the same category sit
on one line, so heavy overlap can hide density. `opacity` softens that, and
[`point * jitter`](../transforms.qmd#jitter-spread-sideways-along-the-category)
spreads the stacked points sideways so the density shows, nudging each point
along the category axis only, never off its measured value.

A second answer is to stop hiding the coincident points and
*count* them: [`point * bin * stack`](../transforms.qmd) piles one dot per
observation up the measure axis, which is the dot plot. One mark answers
overlap in two ways: a nudge along a category, or a pile along a measure.

### Either axis: a horizontal strip

Because the categorical-axis rule is `bar`'s rule, the category can sit on
`y` instead of `x`. The
strip then lies down, the same countries and the same distributions, read left
to right along each row:

```{r}
data(gapminder_2007) + point + x(life) + y(continent) +
  style(opacity = 0.6) +
  x_label("Life expectancy") +
  title("The same distributions, lying down")
```

This is the orientation freedom `bar` already has, and gog reads it the same
way: whichever axis carries the category decides the direction, and there is
no `flip` atom to set. Use the horizontal form when the category names are
too long to fit across the bottom, or when there are so many categories that
rows read more easily than columns.

## Color

Bind `color` to a categorical column to color points by group.
gog assigns a distinct color from a 20-color palette automatically.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent)
```

A color legend is generated automatically whenever a column is mapped to
`color`.

## Shape

`shape` maps a categorical column to one of five glyphs:
circle, square, triangle, diamond, or cross. Assignment is automatic and
consistent: the first category gets the circle, the second the square, and
so on.

Combine `color` and `shape` on the same column to make groups
distinguishable for colorblind readers and in black-and-white print:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + shape(continent)
```

## Size

`size` maps a numeric column to the point radius (3–12 pixels by default).

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population)
```

A size legend showing the minimum, midpoint, and maximum of the column is
generated automatically.

## Opacity

`opacity` maps a numeric column to transparency.
Useful for dense scatterplots where overplotting hides structure.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + opacity(population) +
  title("Opacity: population")
```

See [Channels and legends](../channels.qmd) for the opacity range and why `line`
does not accept this channel.

That example maps opacity to a column. To reduce plain overplotting you want
the same opacity on every point, which is a *setting* rather than a mapping.
`style(opacity = )`, and no legend; see [Setting vs
mapping](../style.qmd):

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(opacity = 0.2, size = 3) +
  title("Every point at one opacity")
```

## All channels at once

The optional channels can all be combined freely:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  shape(continent) +
  size(population) +
  title("GDP, Life Expectancy, and Population")
```

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("point")
```

A point is a closed glyph, and its settings row is wide because of it:
it is the only mark that picks a `shape`, and one of five that carry a border.
`size` is the radius in pixels, and `border_*` draws that border distinct from
the fill.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(shape = "diamond", size = 7, color = "steelblue",
        border_color = "white", border_size = 1.5) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A shape, a size, and a border")
```

The white border is what separates the overlapping points, which is what a
border is for in a dense cloud. A `cross` has no fill to
outline, so a border on that shape has nothing to draw on.

There is no `pattern`. A texture needs an area to fill or a stroke to dash, and
a small glyph is neither; a point's form is `shape`. The other way to answer
overplotting is the collision modifier, which takes a parameter of its own:
`point * jitter(0.4)` spreads coincident points along a categorical axis.

## What it refuses

A point is a glyph, and both of its refusals follow from that.

`size` measures, so it needs something to measure:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

Continents are names, not amounts, and a larger dot would claim an order the data
does not have. The refusal names the three channels that separate categories;
on a point, `color` and `shape` are the two that apply, and the third is
refused next.

Texture is the other. A pattern is a stroke's dash or a fill's hatch, and a
small closed glyph is neither:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + style(pattern = "dashed")
```

Both refusals end at `shape`, which is a point's own form.
