# Point

```{r}
#| include: false
source("../R/setup.R")
```

How does one measurement change as another changes? `point` draws one glyph,
a small symbol, at the `(x, y)` position of each row. The glyph is a dot unless
you choose another.

The plot it makes is the **scatterplot**, and it was invented later than the
other basic charts. Playfair published the line chart and the bar chart in 1786
[@playfair1786], and the pie chart in 1801 [@playfair1801]. The scatterplot came
about thirty years after the pie chart. A picture that plots one measured column
against another is usually dated to John Herschel. In 1833 he plotted the
position of twin stars against the year, then drew a smooth curve through the
dots so the trend could be seen [@friendly2005]. Other names for the same
picture are scatter diagram and scattergram. When `size` maps a third column,
the same plot is called the bubble chart.

None of those chart names is a word in gog, and that is deliberate. `point`
is a **mark**, not a chart. `point + x(gdp) + y(life)` is a scatterplot,
and `point + x(continent) + y(life)` is a strip plot. The mark is the same in
both. The chart name belongs to the whole sentence, not to any word in the
vocabulary. That is why the names live in an appendix, [Chart
names](../chart-names.qmd), instead of in the kernel.

## Basic usage

Everything in this chapter is added to one sentence, the smallest that draws a
point: a source, the mark, and its two positions.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

Each dot is one row, one country, placed by its `gdp` and its `life`. Nothing
else about the row is shown yet. Both columns hold numbers, so
each axis is a linear scale. Most of the dots sit on the left, because most
countries have far lower incomes than the richest few. The
[Scales](../scales.qmd) chapter shows the log scale that spreads them out.

## The dot as a default

Write `point` and you get a dot, and treating it as *the dot mark* is right
almost every time. But the dot is only a *default*, and you can change it. A
point is *a glyph at a position*, and `shape` replaces the dot with another of
the five glyphs gog draws, one per category. [Shape](#shape) below names the
five, says which category gets which, and what to do with more than five.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + shape(continent) +
  title("The glyph is not always a dot")
```

*"Given gapminder 2007: points, x is gdp, y is life, shape by continent."*

So "a dot" means "the first of the five shapes, drawn when you name no shape."
That is what makes `point` and [`text`](text.qmd) **siblings**: both draw one
glyph per row at `(x, y)`, differing only in *where the glyph comes from*. A
point picks its glyph from the fixed set of five, with `shape`. A text mark
takes its glyph from a column, with `label`, so what you see is the value in
that column. Put both on one plot and each row is located by a dot and named by
its label:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("point and text, siblings on one plot")
```

*"Given the medals: points and also text, x is gold, y is silver, label by
country."*

## A categorical x: the strip plot

`x` accepts a categorical column as readily as a continuous one. Each category gets a
position on the axis, and every row is drawn there. You see the whole
distribution rather than a summary of it:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") + title("One dot per country")
```

This is the same rule `bar` uses for a categorical axis, and no mark gets a
special case. Compare it with `bar * mean`, which collapses each group to one
number:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("The same data, summarized")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

The strip plot shows Africa's spread and its overlap with Asia; the bar chart
shows neither. The strip plot has one weakness: every point in a category sits
on the same vertical line, so heavy overlap can hide how many there are.
`opacity` makes each point partly transparent, so overlap reads as a darker
patch. [`point *
jitter`](../transforms.qmd#jitter-spread-sideways-along-the-category) spreads
overlapping points along the category axis, never off their measured value.

A second answer is to count the overlapping points instead of separating them.
[`point * bin * stack`](../transforms.qmd#the-dot-plot) stacks one dot per row
along the measure axis, which is the dot plot. The cookbook draws it with
[thirty rows](../cookbook/one-variable.qmd#when-there-are-only-thirty-rows),
where every row still shows. Which to write is your choice, by what you want
the reader to see. `jitter` keeps the spread: every value sits where it was
measured. `bin * stack` gives the count: how many rows fall in each bin, which
is close to a bar chart, with the rows still visible instead of summarized into
one length.

### Either axis: a horizontal strip

gog reads the direction of a plot from its bindings: whichever axis carries the
category decides it, for a point as for a [bar](bar.qmd#horizontal-bars). So
the category can sit on `y` instead of `x`, and the strips lie down. The
countries and the distributions are the same, and each row now reads left to
right:

```{r}
data(gapminder_2007) + point + x(life) + y(continent) +
  style(opacity = 0.6) +
  x_label("Life expectancy") +
  title("The same distributions, lying down")
```

If you know ggplot2's `coord_flip()`, there is no such word here: swapping the
two bindings was the whole change, as swapping the two arguments of
matplotlib's `scatter()` is. Use the horizontal form when the category names
are too long for the bottom axis. It also helps when there are many categories,
since a list of rows is easier to read than a list of columns.

## Color

Which continent is which? Every dot looks the same until a column tells them
apart. Bind `color` to a categorical column to color points by group. gog
assigns a distinct color from a 20-color palette automatically.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."*

A color legend is generated automatically whenever a column is mapped to
`color`.

## Shape

`shape` maps a categorical column to one of five glyphs. In order, they are
the circle, the square, the triangle, the diamond and the cross, one per
category. Categories are taken as they first appear in the table, or in the
[order you declare](../data.qmd#declaring-the-order-of-categories), so the
first category gets the circle, the second the square, and so on. Asia comes
first in the gapminder table, so Asia is the circle, and the legend below reads
the same way. Five is the whole set, and about as many as a
reader can separate on one plot. Do not map `shape` to a column with more than
five categories: the sixth starts over at the circle, two categories then share
a glyph and look the same, and gog does not warn. The book shows four ways to
draw more groups than that:

- `color`, whose default palette has twenty colors, as [Color](#color) above
  does with the continents.
- One panel per category with `facet`, folded into rows so the panels stay
  wide: [Folding a long line of
  panels](../faceting.qmd#folding-a-long-line-of-panels).
- The name on the point instead of a glyph: a `text` layer with `label`, the
  [labeled scatter](text.qmd#both-at-once-the-labeled-scatter), and
  [`repel`](text.qmd#repel-moving-labels-off-each-other) when the names
  overlap. This is the one route that reaches a hundred categories, and in the
  web edition a [brush](../selection.qmd#which-one-is-that) leaves only the
  selected names readable.
- Every category in one pale color, with the few that matter drawn over it in
  a second layer: [What settings are for](../style.qmd#what-settings-are-for).

A column with fewer categories often carries the encoding: `continent` groups
the countries and fits the five glyphs, so `country` goes on the label or the
facet.

Combine `color` and `shape` on the same column so that colorblind readers, and
black-and-white printing, can still separate the groups:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + shape(continent)
```

## Size

A third measurement has nowhere to go once both positions are taken. `size`
gives it one: it maps a continuous column to the point radius, from 3 pixels
at the smallest value to 12 at the largest.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, size by
population."*

A size legend showing the minimum, midpoint, and maximum of the column is
generated automatically.

## Opacity

`opacity` maps a continuous column to transparency. It helps in a dense
scatterplot, where marks cover one another and hide the shape of the data.
Population is skewed, so on a linear scale two countries are dark and the rest
are faint. The log scale spreads them across the range:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  opacity(population, scale = "log") +
  title("Opacity: population, on a log scale")
```

*"Given gapminder 2007: points, x is gdp, y is life, opacity by population on a
log scale."*

See [Channels and legends](../channels.qmd) for the opacity range and why `line`
does not accept this channel.

That example maps opacity to a column. Often there is no column to show: the
only problem is that many points overlap. Then you want one opacity on every
point instead. That is a *setting* rather than a mapping, so you write
`style(opacity = )` and no legend appears; see [Setting vs
mapping](../style.qmd). The table below, `gm_all`, holds every gapminder country
in every year, so the points overlap heavily:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(opacity = 0.2, size = 3) +
  title("Every point at one opacity")
```

*"Given all the gapminder years: points, x is gdp, y is life, with opacity
0.2."*

## All channels at once

A table often has more to show than two positions can hold. The refining
channels carry the rest, and they can all be on one plot at once. Here
`continent` goes to `color` and `shape` together, so the groups survive
black-and-white printing, and `population` goes to `size`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  shape(continent) +
  size(population) +
  title("GDP, life expectancy, and population")
```

## What you can set

A setting changes how every point looks without reading a column, and each
mark takes its own. These are a point's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("point")
```

A point is a closed glyph, a shape with an inside to fill, so it takes more
settings than most marks. It is the only mark that picks a `shape`, and one of
the five marks that draw a border.
`size` is the radius in pixels, and `border_*` draws that border distinct from
the fill. 

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(shape = "diamond", size = 7, color = "steelblue",
        border_color = "navy", border_size = 1.5) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A shape, a size, and a border")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored steelblue, with
shape diamond, size 7, border color navy and border size 1.5."*

The navy border is what separates the overlapping points, which is what a
border is for in a dense cloud. A `cross` has no fill to outline, so
`border_color` and `border_size` change nothing on that shape.

`pattern` is not on the list. A texture needs an area to fill or a stroke to
dash, and a small glyph has neither, so a point's only form is its `shape`;
[What it refuses](#what-it-refuses) shows the refusal. When points overlap, a
border keeps them apart, and so does `jitter`, one of the four collision
modifiers: `point * jitter(0.4)` moves each point a little along the
categorical axis, by that amount, as the [strip
plot](#a-categorical-x-the-strip-plot) showed.

## What it refuses

A point is a glyph, and the two refusals below both follow from that.

`size` shows an amount, so it needs a column of amounts:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

Continents are names, not amounts, and a larger dot would claim an order the
data does not have. The refusal names three channels that separate categories:
`color`, `shape` and `pattern`. A point takes the first two, and the next
section shows why it refuses `pattern`.

The second refusal is texture. A pattern is a stroke's dash or a fill's hatch,
and a small closed glyph is neither:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + style(pattern = "dashed")
```

Both messages send you to `shape`, which is the one form a point can change.
