# Point

```{r}
#| include: false
source("../R/setup.R")
```

How does one measurement change as another changes? `point` draws one glyph,
a small symbol, at the `(x, y)` position of each row. The glyph is a dot unless
you choose another.

The plot it makes is the **scatterplot**, and it was invented later than the
other basic charts. William Playfair published the line chart and the bar chart
in 1786 [@playfair1786], and the pie chart in 1801 [@playfair1801]. The first
scatterplot was drawn about thirty years after the pie chart. Historians
usually name John Herschel as the first person to plot one measured column
against another. In 1833 he plotted the position of a double star against the
year, then drew a smooth curve through the dots so the trend could be seen
[@friendly2005]. Other names for the same
picture are scatter diagram and scattergram. When `size` maps a third column,
the same plot is called the bubble chart.

None of those chart names is a word in gog, and that is deliberate. `point`
is a **mark**, not a chart. `point + x(gdp) + y(life)` is a scatterplot,
and `point + x(continent) + y(life)` is a **strip plot**. The mark is the same in
both. The chart name belongs to the whole sentence, not to any word in the
vocabulary. That is why the names are listed in an appendix, [Chart
names](../chart-names.qmd), instead of in the kernel.

## Basic usage

Do people in richer countries live longer? Everything in this chapter is added
to one sentence, the smallest that draws a point: a source, the mark, and its
two positions.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

Each dot is one row, one country, placed by its `gdp` and its `life`. Nothing
else about the row is shown yet. The answer to the opening question is yes:
life expectancy rises quickly with gdp among the poorer countries, then slowly
among the richer ones. Both columns hold numbers, so
each axis is a linear scale. Most of the dots sit on the left, because most
countries have far lower incomes than the richest few. The
[Scales](../scales.qmd) chapter shows the log scale that spreads them out.

## The dot as a default

Write `point` and you get a dot. Calling it *the dot mark* is accurate almost
every time. But the dot is only a *default*, and you can change it. A
point is *a glyph at a position*, and `shape` replaces the dot with one of the
seven glyphs gog draws, a different one for each category of a column. [Shape](#shape) below names the
seven, says which category gets which, and what to do with more than seven.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + shape(continent) +
  title("The glyph is not always a dot")
```

*"Given gapminder 2007: points, x is gdp, y is life, shape by continent."*

So "a dot" means the circle, the first of the seven glyphs, drawn when you name
no `shape`.
That is what makes `point` and [`text`](text.qmd) **siblings**: both draw one
glyph per row at `(x, y)`, differing only in *where the glyph comes from*. A
point picks its glyph from the fixed set of seven, with `shape`. A text mark
takes its glyph from a column, with `label`, so what you see is the value in
that column. Put both on one plot and each row is located by a dot and named by
its label:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("point and text, siblings on one plot")
```

*"Given the medals: points and also text, x is gold, y is silver, label by
country."*

The dot shows where each country is, and the label names it.
`style(nudge = "up")` moves each label above its dot, so the two do not cover
each other.

## A categorical x: the strip plot

Every plot so far has put a measurement on `x`, but `x` also accepts a
categorical column. Each category gets its own slot on the axis, and every row
is drawn in its category's slot. You see the whole distribution rather than a
summary of it:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") + title("One dot per country")
```

Every mark places a categorical column this way, `bar` included. Compare the
strip plot with `bar * mean`, which reduces each group to one number:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("The same data, summarized")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

`mean` is not the bar's own transform. Other marks take it too, and each draws
the same means in its own way:

| Mark derived by `mean` | What it draws |
|---|---|
| `point * mean` | one dot at each mean |
| `line * mean` | the means joined into a profile |
| `area * mean` | the same profile, filled to the baseline |
| `step * mean` | each mean held flat across its slot |
| `bar * mean` | one bar per category, the plot above |

`median`, `sum`, `min` and `max` can replace `mean` in every row of that table.
So can `quantile`, which needs a probability: `quantile(0.9)` is the value nine
tenths of each group falls below.
The statistic decides the number, and the mark decides how it is drawn.

The strip plot shows Africa's spread and its overlap with Asia; the bar chart
shows neither. The strip plot has one weakness: every point in a category sits
on the same vertical line. Where many countries have close values, their dots
cover each other, and you cannot see how many there are. `opacity` makes each
point partly transparent, so overlap reads as a darker patch. [`point *
jitter`](../transforms.qmd#jitter-spread-sideways-along-the-category) moves
overlapping points sideways along the categorical axis, and never changes their
measured value:

```{r}
data(gapminder_2007) + point * jitter + x(continent) + y(life) +
  y_label("Life expectancy") + title("The same dots, spread sideways")
```

*"Given gapminder 2007: points derived by jitter, x is continent, y is life."*

Every dot keeps its life expectancy, so its height has not changed. Only its
place inside its continent's slot has changed, and that place means nothing:
the dot still reads as its continent. `jitter` moves a dot's category, never its
value. In a scatterplot both axes are measurements, so a moved dot would show a
value the data does not hold. gog therefore refuses `point * jitter` when `x`
and `y` are both continuous, as [What it refuses](#what-it-refuses) shows.

A third answer is to count the overlapping points instead of separating them.
[`point * bin * stack`](../transforms.qmd#the-dot-plot) cuts the measured
column into bins and piles one dot per row above each bin, which is the
**dot plot**. It shows one distribution, with the measure on `x`;
`| facet(continent)` draws one for each continent. The cookbook draws it with [thirty
rows](../cookbook/one-variable.qmd#when-there-are-only-thirty-rows), where every
row is still visible. Choose between `jitter` and `bin * stack` by what you want
the reader to see. `jitter` keeps the spread: every value sits where it was
measured. `bin * stack` gives the count: how many rows fall in each bin. The
result looks like a bar chart, but each row is still its own dot.

### Either axis: a horizontal strip

gog reads the direction of a plot from its bindings. Whichever axis carries the
category decides the direction, for a point exactly as for a
[bar](bar.qmd#horizontal-bars). So the category can sit on `y` instead of `x`,
and the strips run horizontally. The countries and the distributions are the
same, and each continent's dots now run from left to right:

```{r}
data(gapminder_2007) + point + x(life) + y(continent) +
  style(opacity = 0.6) +
  x_label("Life expectancy") +
  title("The same distributions, lying down")
```

If you know ggplot2's `coord_flip()`, there is no such word here. Swapping the
two bindings was the whole change, the same way you swap the two arguments of
matplotlib's `scatter()`. Use the horizontal form when the category names are
too long for the bottom axis. It also helps when there are many categories,
since names listed down the left side are easier to read than names crowded
along the bottom.

## Color

In the scatterplot of `gdp` against `life`, every dot looks the same, so you
cannot see which continent each country is in. Bind `color` to a categorical
column to color points by group. gog assigns a distinct color from a 20-color
palette automatically.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."*

Most African countries sit at the lower left, and most European countries at
the upper right. gog adds a color legend whenever you map a column to `color`.

## Shape

Which glyph does each category get, and what happens when there are more
categories than glyphs? `shape` maps a categorical column to one of seven
glyphs. In order, they are the circle, the square, the triangle, the diamond,
the cross, the star and the wye, which looks like the letter Y. The categories
are ordered by where each first appears in the table, or by the
[order you declare](../data.qmd#declaring-the-order-of-categories). The first
category gets the circle, the second the square, and so on. Here is the whole
set, in the order gog assigns them:

```{r}
glyph_order <- c("circle", "square", "triangle", "diamond", "cross", "star", "wye")
glyphs <- data.frame(glyph = ordered(glyph_order, levels = glyph_order),
                     row = c(1, 1, 1, 1, 1, 1, 1))
data(glyphs) + point + x(glyph) + y(row) + shape(glyph) +
  style(size = 8) + x_label("") + y_label("") +
  title("The seven glyphs, in the order they are assigned")
```

*"Given the glyphs table: points, x is glyph, y is row, shape by glyph, with
size 8."*

Asia comes first in the gapminder table, so Asia is the circle, and every
continent legend in this chapter lists Asia first. Seven is the whole set gog
draws. It is also about as many glyphs as a reader can separate on one plot. Map
`shape` to a column with more than seven categories and the eighth category
gets the circle again, so two categories share a glyph and look the same. gog says so when
that happens, naming the count and the limit, and the plot still draws. The
`nutrients` table has eight foods, one more than there are glyphs:

```{r}
#| echo: false
peek(nutrients)
```

```{r}
#| message: true
data(nutrients) + point + x(nutrient) + y(amount) + shape(food) +
  y_label("Amount per 100g") + title("Eight foods, seven glyphs")
```

*"Given the nutrients: points, x is nutrient, y is amount, shape by food."*

Oats is the eighth food in the table, so its points are circles, the glyph
salmon already has. Nothing in the picture shows which circles are salmon and
which are oats, so the message counts the foods and the glyphs for you. The book
has four ways to draw more than seven groups:

- `color`, whose default palette has 20 colors, as [Color](#color) above
  does with the continents.
- One panel per category with `facet`, folded into rows so the panels stay
  wide: [Folding a long line of
  panels](../faceting.qmd#folding-a-long-line-of-panels).
- The name on the point instead of a glyph: a `text` layer with `label`, the
  [labeled scatter](text.qmd#both-at-once-the-labeled-scatter), and
  [`repel`](text.qmd#repel-moving-labels-off-each-other) when the names
  overlap. Of the four, only this one can name each of a hundred categories.
  In the web edition, a
  [brush](../selection.qmd#which-one-is-that) leaves only the selected names
  readable, as long as the `text` layer does not use `repel`.
- Every category in one pale color, and the few that matter drawn again on
  top, in a second layer: [What settings are
  for](../style.qmd#what-settings-are-for).

When a table has two categorical columns, map `shape` to the one with fewer
categories. `continent` has five, and seven glyphs are enough for five.
`country` has 142, so it goes on a label or a facet instead.

Some readers cannot see the difference between colors, and a black-and-white
printout removes it. Map `color` and `shape` to the same column, and the glyphs
still separate the groups:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + shape(continent)
```

Each continent now differs in two ways, so either one alone still separates the
groups.

## Size

Once `x` and `y` are both bound, no position is free for a third measurement.
The `size` channel can show it instead. `size` maps a continuous column to the
point's radius, from 3 pixels at the smallest value to 12 at the largest.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, size by
population."*

The two largest dots are China and India, and both sit at the left, among the
poorer countries. gog adds a size legend showing the column's minimum,
midpoint, and maximum.

## Opacity

The horizontal strip plot above made every point partly transparent with one
setting, `style(opacity = 0.6)`. `opacity` is also a channel: it maps a
continuous column to transparency, and a larger value draws a more solid point.
Population is skewed, so a linear scale would leave China and India dark and
the rest faint. The log scale spreads the values across the range:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  opacity(population, scale = "log") +
  title("Opacity: population, on a log scale")
```

*"Given gapminder 2007: points, x is gdp, y is life, opacity by population on a
log scale."*

See [Channels and legends](../channels.qmd) for the opacity range and why `line`
does not accept this channel.

That example maps opacity to a column. Often there is no column to show: the
only problem is that many points overlap. Then you want one opacity on every
point instead. That is a *setting* rather than a mapping, so you write
`style(opacity = )` and no legend appears; see [Setting vs
mapping](../style.qmd). The `gm_all` table holds every gapminder country in
every year, so the points overlap heavily:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(opacity = 0.2, size = 3) +
  title("Every point at one opacity")
```

*"Given all the gapminder years: points, x is gdp, y is life, with opacity
0.2."*

Where many points overlap, the color is darker, and the darkest area is at the
left, where most countries were in most years. A hollow point is the other way
to show the same crowding, and [What you can set](#what-you-can-set) draws this
table with it.

## All channels at once

A table often has more to show than two positions can hold. The refining
channels carry the rest, and they can all be on one plot at once. Here
`continent` is mapped to both `color` and `shape`, so the groups stay separate
in black-and-white printing, and `population` is mapped to `size`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  shape(continent) +
  size(population) +
  title("GDP, life expectancy, and population")
```

Four columns now show in one picture: `gdp` and `life` by position, `continent`
by color and glyph, and `population` by size.

## What you can set

A setting changes how every point looks without reading a column, and each
mark takes its own. These are a point's settings, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("point")
```

A point takes more settings than most marks. It is the only mark that can set
`shape`. Six of its seven glyphs are filled with color, so a point is also one
of the five marks that draw a border. `size` is the radius in pixels, and
`border_color` and `border_size` give that border its own color and width.

A point layer can carry several settings at once:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(shape = "diamond", size = 7, color = "steelblue",
        border_color = "navy", border_size = 1.5) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A shape, a size, and a border")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored steelblue, with
shape diamond, size 7, border color navy and border size 1.5."*

Where two diamonds overlap, the navy border shows where one ends and the next
begins. That is what a border is for in a dense cloud of points. A `cross` has
no fill to outline, so `border_color` and `border_size` change nothing on that
glyph.

A border can also be the whole point. Set the fill color to `"none"` and give
the border a color, and each point is drawn hollow: an outline with no fill. A
hollow point can be partly transparent as well. `opacity` fades a point's fill,
and a hollow point has none, so the transparency goes in the border's color
instead. In `"#4682b44d"`, the first six hex digits are steelblue, and the last
two, `4d`, make the outline 30% opaque. Here are all 1,704 rows of `gm_all`
again, the table the [Opacity](#opacity) section drew at one opacity:

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(color = "none", border_color = "#4682b44d", size = 3) +
  title("Every point hollow, its outline partly transparent")
```

Where hollow points overlap, every outline stays visible, and where many pile
up, the outlines add up to a darker area. The plot shows each point and the
crowding at once. Neither this nor opacity moves a point off its measured value,
which is why they, and not `jitter`, are the answers for a crowded scatterplot. On a categorical axis,
`jitter`, one of the four collision modifiers, moves points apart instead:
`point * jitter(0.4)` spreads them 0.4 times as wide as a plain `jitter`, which
the [strip plot](#a-categorical-x-the-strip-plot) showed.

`pattern` is not on the list. A hatch needs a large area and a dash needs a
long stroke, and a glyph is too small for either. To separate groups by glyph,
a point uses `shape` instead, and [What it refuses](#what-it-refuses) shows the
refusal.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A point refuses three things: a category on `size`, any pattern, and `jitter` on
a scatterplot.

`size` shows an amount, so it needs a column of amounts:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

Continents are names, not amounts, and a larger dot would claim an order the
data does not have. The refusal names three channels that separate categories:
`color`, `shape` and `pattern`. A point takes the first two, and the refusal
below, of the `pattern` setting, shows why it takes no pattern at all.

The second refusal is about texture. A pattern is a dash along a stroke or a
hatch across an area, and a glyph is too small for either:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + style(pattern = "dashed")
```

Both messages name `shape`, which chooses a point's glyph.

The third refusal is `jitter` on a scatterplot. In a strip plot, `jitter` moves
a dot sideways inside its continent's slot. A place inside a slot has no scale,
so the move changes nothing a reader could measure. In a scatterplot both axes
are measurements, and every place on them is a value. A jittered dot would sit
at a gdp and a life expectancy that are not in the table. Two countries with the
same values would no longer look the same, and a reader who reads a dot against
the axes would read a number that was never measured. So gog refuses:

```{r}
#| error: true
data(gapminder_2007) + point * jitter + x(gdp) + y(life)
```

This is a design rule, not a missing feature: `jitter` moves a dot's category,
never its value. ggplot2's `geom_jitter()` moves points along both axes by a
small random amount, and gog does not. To show overlapping points in a
scatterplot, keep every point at its value and change how it is drawn.
`style(opacity = )`, which the message suggests, makes overlap read as a darker
area, as in the plot of every gapminder year under [Opacity](#opacity). A
hollow point, an outline with no fill, keeps every point visible where points
overlap, as the same table shows under [What you can set](#what-you-can-set).

## In short

A point is a glyph at a position. The `point` mark draws one per row, unless a
transform groups the rows first. It is the only mark that maps `shape`, because
it is the only mark that picks its glyph from a fixed set. There are seven
glyphs, and an eighth category gets the circle again.

## What comes next

The `point` mark draws each row as a separate glyph, and nothing joins one row
to the next. When one column puts the rows in order, such as years, the change
from one value to the next often matters most. [Line](line.qmd) joins the rows into one
stroke that shows whether the value rose or fell.
