# Point

```{r}
#| include: false
source("../R/setup.R")
```

How do two measurements move together? `point` draws one glyph per row of data
at the `(x, y)` position.

The plot it makes is the **scatterplot**, which is the youngest of the plain
chart types by half a century. Playfair had the line and the bar in 1786, and
the pie by 1801. A picture that sets two measured columns against each other
is usually dated to John Herschel. In 1833 he plotted the position of twin
stars against the year, then drew a smooth curve through the dots to read a
trend off them [@friendly2005]. Other names for it are the same picture:
scatter diagram, scattergram, and, once `size` carries a third column, the
bubble chart.

None of those is an atom here, and the reason is the whole design. `point` names
the **mark**, not the chart. `point + x(gdp) + y(life)` is a scatterplot and
`point + x(continent) + y(life)` is a strip plot, and nothing about the mark
changed between them: the chart name is a property of the sentence, not a word in
the vocabulary. That is why the names live in an appendix, [Chart
names](../chart-names.qmd), instead of in the kernel.

## Basic usage

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

## A point is a dot: the dot is just its default

Reach for `point` and you get a dot, and treating it as *the dot mark* will serve
you well almost every time. But the dot is a **default**, not the mark's whole
identity. A point is *a glyph at a position*, and the glyph is yours to change:
`shape` swaps the dot for one of five.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + shape(continent) +
  title("The glyph is not always a dot")
```

So "a dot" really means "the first of five shapes, drawn when you name none."
That is what makes `point` and [`text`](text.qmd) **siblings**: both draw one
glyph per row at `(x, y)`, differing only in *where the glyph comes from*. A
point picks its glyph from the closed set of five (`shape`); a text mark takes it
from a column: the string itself (`label`). Put both on one plot and each row is
located by a dot and named by its label:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("point and text, siblings on one plot")
```

## A categorical x: the strip plot

`x` accepts a text column as readily as a numeric one. Each category gets a
position on the axis and every row is drawn at it, showing the whole
distribution rather than a summary of it:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") + title("One dot per country")
```

This is the same rule `bar` uses for a categorical axis, no per-mark special
case. Compare it with `bar * mean`, which collapses each group to one number:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("The same data, summarized")
```

The strip plot shows Africa's spread and its overlap with Asia; the bar chart
shows neither. Its one weakness: points at the same category sit on one line,
so heavy overlap can hide density. `opacity` helps, and
[`point * jitter`](../transforms.qmd) spreads the pile sideways to reveal it
outright, nudging each point along the category axis only, never off its
measured value.

The other way to resolve that overlap is to stop hiding the coincident points and
*count* them: [`point * bin * stack`](../transforms.qmd) piles one dot per
observation up the measure axis, which is the dot plot. One mark, two overlap
answers: a nudge along a category, a pile along a measure.

### Either axis: a horizontal strip

Because it *is* the same rule, the category can sit on `y` instead of `x`. The
strip then lies down, the same countries and the same distributions, read left
to right along each row:

```{r}
data(gapminder_2007) + point + x(life) + y(continent) +
  style(opacity = 0.6) +
  x_label("Life expectancy") +
  title("The same distributions, lying down")
```

This is the orientation freedom `bar` already has, and GOG reads it the same
way: whichever axis carries the category decides which way the plot faces;
there is no `flip` atom to set. Reach for the horizontal form when the category
names are long enough to collide across the bottom, or when there are enough of
them that rows stacked down the page scan more easily than columns across it.

## Color

Bind `color` to a categorical column to color points by group.
GOG assigns a distinct color from a 10-color palette automatically.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent)
```

A color legend is generated automatically whenever `color` is active.

## Shape

`shape` maps a categorical column to one of five glyphs:
circle, square, triangle, diamond, or cross.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  shape(continent)
```

Combine `color` and `shape` on the same column to make groups
distinguishable for colorblind readers and in black-and-white print:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + shape(continent)
```

## Size

`size` maps a numeric column to the point radius (3 – 12 px by default).

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population)
```

A size legend showing the min, midpoint, and max values is generated automatically.

## Opacity

`opacity` maps a numeric column (0 – 1) to transparency.
Useful for dense scatterplots where overplotting hides structure.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + opacity(population) +
  title("Opacity — population")
```

See [Channels and legends](../channels.qmd) for the opacity range and why `line`
does not accept this channel.

That example maps opacity to a column. For plain overplotting relief you want
the same opacity for every point, which is a *setting* rather than a mapping.
`style(opacity = )`, and no legend:

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(opacity = 0.2, size = 3) +
  title("Every point at one opacity — see Setting vs mapping")
```

## All channels at once

All decorative channels can be combined freely:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  shape(continent) +
  size(population) +
  title("GDP, Life Expectancy, and Population")
```

## Available shapes

| Index | Shape |
|-------|-------|
| 0 | Circle (default) |
| 1 | Square |
| 2 | Triangle |
| 3 | Diamond |
| 4 | Cross |

Shape assignment is automatic and consistent: the first unique category
always gets circle, the second gets square, and so on.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("point")
```

A point is a closed glyph, and its row is the widest in the book because of it:
it is the only mark that picks a `shape`, and one of five that carry a rim.
`size` is the radius in pixels, and `border_*` draws that rim distinct from the
fill.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(shape = "diamond", size = 7, color = "steelblue",
        border_color = "white", border_size = 1.5) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A shape, a size, and a rim")
```

The white rim is doing real work there: it separates points that overlap, which
is what a rim is for on a glyph dense enough to pile up. A `cross` has no fill to
outline, so a border on that shape has nothing to draw on.

There is no `pattern`. A texture needs an area to fill or a stroke to dash, and a
5-pixel glyph is neither; a point's form is `shape`. The other way to answer
overplotting is the collision modifier, which takes a parameter of its own:
`point * jitter(0.4)` spreads coincident points along a categorical axis.

## What it refuses

A point is a glyph, and both refusals worth meeting follow from that.

`size` measures, so it needs something to measure:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

Continents are names, not amounts, and a larger dot would claim an order the data
does not have. The refusal names the three channels that do separate categories.

Texture is the other. A pattern is a stroke's dash or a fill's hatch, and a glyph
five pixels across is neither:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + style(pattern = "dashed")
```

Both refusals point the same way, at `shape`, which is a point's own form.
