# Point

```{r}
#| include: false
source("../R/setup.R")
```

How does one measurement change as another changes? `point` draws one glyph
per row of data at the `(x, y)` position.

The plot it makes is the **scatterplot**, and it was invented later than the
other basic charts. Playfair published the line chart and the bar chart in 1786,
and the pie chart by 1801. The scatterplot came about thirty years after the pie
chart. A picture that plots one measured column against another is usually dated
to John Herschel. In 1833 he plotted the position of twin stars against the
year, then drew a smooth curve through the dots so the trend could be seen
[@friendly2005]. Other names for the same picture are scatter diagram and
scattergram. When `size` maps a third column, the same plot is called the bubble
chart.

None of those is an atom here, and the reason is the design itself. `point`
names the **mark**, not the chart. `point + x(gdp) + y(life)` is a scatterplot,
and `point + x(continent) + y(life)` is a strip plot. The mark is the same in
both. The chart name belongs to the whole sentence, not to any word in the
vocabulary. That is why the names live in an appendix, [Chart
names](../chart-names.qmd), instead of in the kernel.

## Basic usage

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

## The dot as a default

Write `point` and you get a dot, and treating it as *the dot mark* is right
almost every time. But the dot is only a *default*, and you can change it. A
point is *a glyph at a position*, and `shape` replaces the dot with one of five
glyphs.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + shape(continent) +
  title("The glyph is not always a dot")
```

*"Given gapminder 2007: points, x is gdp, y is life, shape by continent."*

So "a dot" means "the first of the five shapes, drawn when you name no shape."
That is what makes `point` and [`text`](text.qmd) **siblings**: both draw one
glyph per row at `(x, y)`, differing only in *where the glyph comes from*. A
point picks its glyph from the fixed set of five, with `shape`. A text mark
takes its glyph from a column, with `label`, so what you see is the value in
that column. Put both on one plot and each row is located by a dot and named by
its label:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("point and text, siblings on one plot")
```

*"Given the medals: points and also text, x is gold, y is silver, label by
country."*

## A categorical x: the strip plot

`x` accepts a text column as readily as a numeric one. Each category gets a
position on the axis, and every row is drawn there. You see the whole
distribution rather than a summary of it:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) +
  y_label("Life expectancy") + title("One dot per country")
```

This is the same rule `bar` uses for a categorical axis, and no mark gets a
special case. Compare it with `bar * mean`, which collapses each group to one
number:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") + title("The same data, summarized")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*

The strip plot shows Africa's spread and its overlap with Asia; the bar chart
shows neither. The strip plot has one weakness: every point in a category sits
on the same vertical line, so heavy overlap can hide how many there are.
`opacity` makes each point partly transparent, so overlap reads as a darker
patch. [`point *
jitter`](../transforms.qmd#jitter-spread-sideways-along-the-category) spreads
overlapping points along the category axis, never off their measured value.

A second answer is to count the overlapping points instead of separating them.
[`point * bin * stack`](../transforms.qmd) stacks one dot per row along the
measure axis, which is the dot plot. One mark handles overlap in two ways:
spread along a category, or count along a measure.

### Either axis: a horizontal strip

Because the categorical-axis rule is `bar`'s rule, the category can sit on `y`
instead of `x`. The strips then lie down. The countries and the distributions
are the same, and each row now reads left to right:

```{r}
data(gapminder_2007) + point + x(life) + y(continent) +
  style(opacity = 0.6) +
  x_label("Life expectancy") +
  title("The same distributions, lying down")
```

`bar` already works this way, and gog reads a point the same way. Whichever axis
carries the category decides the direction, and there is no `flip` atom to set.
Use the horizontal form when the category names are too long for the bottom
axis. It also helps when there are many categories, since a list of rows is
easier to read than a list of columns.

## Color

Bind `color` to a categorical column to color points by group.
gog assigns a distinct color from a 20-color palette automatically.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."*

A color legend is generated automatically whenever a column is mapped to
`color`.

## Shape

`shape` maps a categorical column to one of five glyphs:
circle, square, triangle, diamond, or cross. Assignment is automatic and
consistent: the first category gets the circle, the second the square, and
so on.

Combine `color` and `shape` on the same column so that colorblind readers, and
black-and-white printing, can still separate the groups:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + shape(continent)
```

## Size

`size` maps a numeric column to the point radius, from 3 pixels at the smallest
value to 12 at the largest.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, size by
population."*

A size legend showing the minimum, midpoint, and maximum of the column is
generated automatically.

## Opacity

`opacity` maps a numeric column to transparency. It helps in a dense
scatterplot, where marks cover one another and hide the shape of the data.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + opacity(population) +
  title("Opacity: population")
```

*"Given gapminder 2007: points, x is gdp, y is life, opacity by population."*

See [Channels and legends](../channels.qmd) for the opacity range and why `line`
does not accept this channel.

That example maps opacity to a column. When points simply cover one another, you
want one opacity on every point instead. That is a *setting* rather than a
mapping, so you write `style(opacity = )` and no legend appears; see [Setting vs
mapping](../style.qmd). The table below, `gm_all`, holds every gapminder country
in every year, so the points overlap heavily:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(opacity = 0.2, size = 3) +
  title("Every point at one opacity")
```

*"Given all the gapminder years: points, x is gdp, y is life, with opacity
0.2."*

## All channels at once

The refining channels can all be combined on one plot:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  shape(continent) +
  size(population) +
  title("GDP, life expectancy, and population")
```

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("point")
```

A point is a closed glyph, a shape with an inside to fill, so it takes more
settings than most marks. It is the only mark that picks a `shape`, and one of
the five marks that draw a border.
`size` is the radius in pixels, and `border_*` draws that border distinct from
the fill.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(shape = "diamond", size = 7, color = "steelblue",
        border_color = "white", border_size = 1.5) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A shape, a size, and a border")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored steelblue, with
shape diamond, size 7, border color white and border size 1.5."*

The white border is what separates the overlapping points, which is what a
border is for in a dense cloud. A `cross` has no fill to outline, so
`border_color` and `border_size` change nothing on that shape.

There is no `pattern`. A texture needs an area to fill or a stroke to dash, and
a small glyph is neither; a point's form is `shape`. The other way to handle
points that cover one another is `jitter`, one of the four collision modifiers.
It takes an amount: `point * jitter(0.4)` spreads overlapping points along a
categorical axis.

## What it refuses

A point is a glyph, and the two refusals below both follow from that.

`size` shows an amount, so it needs a column of amounts:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

Continents are names, not amounts, and a larger dot would claim an order the
data does not have. The refusal names three channels that separate categories:
`color`, `shape` and `pattern`. A point takes the first two, and the next
section shows why it refuses `pattern`.

The second refusal is texture. A pattern is a stroke's dash or a fill's hatch,
and a small closed glyph is neither:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + style(pattern = "dashed")
```

Both messages send you to `shape`, which is the one form a point can change.
