# Ribbon

```{r}
#| include: false
source("../R/setup.R")
```

How wide is the band of values at each point along x? `ribbon` draws a filled
band from a low boundary to a high one across x: a confidence band, a spread
band, the shaded region around a trend. It fills one region, the way an
[`area`](area.qmd) does. It reads a low value and a high value at each
position, the way an [`interval`](interval.qmd) does. An area fills from the
data down to a baseline at zero. A ribbon has no baseline, and both of its
edges are values the table supplies.

## A band from a range

Did the five countries move closer together over the years, or further apart?
The gap between the lowest and the highest in each year answers that. A band
drawn between those two values shows every year at once, as one shape. The two
boundaries come from a transform, exactly as an interval's two extents do.
`range` reduces `y` to its minimum and maximum within each `x`, and the ribbon
fills between them. Here, the spread of life expectancy across five Asian
countries, year by year:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + ribbon * range + x(year) + y(life) +
  y_label("Life expectancy") +
  title("The min–max spread across five Asian countries, per year")
```

*"Given gapminder Asia: a ribbon derived by range, x is year, y is life."*

In each year the band's lower edge is the country with the lowest life
expectancy, and its upper edge the country with the highest. The band is the
region that holds all five. Read it as `interval * range` drawn as a
*continuous* fill rather than as a whisker at each x. The low and high pair is
the same. Only the shape drawn from it changes.

## The mean and its band

On its own a band shows spread but not center. A band is usually drawn with a
line through the middle. You compose the two by layering, not by adding an
option to either mark. A `ribbon * range` for the spread, a `line * mean` for
the trend, sharing one `x` and `y`:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  ribbon * range + style(color = "steelblue", opacity = 0.25) +
  line * mean + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Mean life expectancy, with its min–max band")
```

*"Given gapminder Asia: x is year, y is life, a ribbon derived by range and also
a line derived by mean."*

Each layer runs its own transform on the same table. `mean` reduces each year to
one point for the line, and `range` reduces the same year to a low and high pair
for the band. This is the most common ribbon plot: a trend with its uncertainty
or spread drawn around it. It is two marks the grammar already has, added with
`+`.

## A fan chart of nested bands

`gm_all` is the whole gapminder table: every country, in every year, 142
countries in all. A min–max band over it spans the full range of life
expectancy in each year, but only the two extreme countries set its edges.
Naming two quantiles instead gives a band that contains most of the countries.
Layer two bands of different widths to draw a **fan chart**:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + x(year) + y(life) +
  ribbon * range(0.1, 0.9) + style(color = "steelblue", opacity = 0.2) +
  ribbon * range(0.25, 0.75) + style(color = "steelblue", opacity = 0.35) +
  line * median + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Life expectancy across 142 countries, 1952 to 2007")
```

*"Given all the gapminder years: x is year, y is life, a ribbon derived by range
from 0.1 to 0.9, and also a ribbon derived by range from 0.25 to 0.75, and also
a line derived by median."*

The pale band contains the middle 80% of countries. The darker one contains the
middle half, and the line is the median. The plot then says something the
min–max version cannot. The median rose from 45 years to 72, and the 90th
percentile from 67 to 80, so the distance between them fell from 22 years to 8.

The lower edges of the two bands do not rise steadily. Both lower edges, the
10th percentile and the 25th, decline between 1992 and 2002 while the median
rises throughout. A min–max band cannot show this, because its lower edge is one
country.

The same layering goes further. Each band is one more layer, so a third pair of
quantiles adds a third band. Write the widest band first, because a later layer
is painted over an earlier one.

## Why a ribbon needs a transform

A low *and* a high are two numbers, and a plain `y()` supplies one. An `area`
closes on zero, which the grammar supplies. A ribbon has no such second number,
so like `interval` its minimum syllable includes a range-producing transform.
Write a ribbon without such a transform and the engine refuses it, rather than
drawing an empty panel. [What it refuses](#what-it-refuses) shows the message.

This is the same design choice that keeps the [channel
vocabulary](../grammar.qmd) small. The alternative would add `ymin` and `ymax`
as two more channels, and only one mark would ever use them. Instead the
boundaries are invented by a transform, exactly as `bin` invents a count. The
same low and high rows then serve the whisker, the box and the band. See
[Transforms](../transforms.qmd) for the `range`, `confidence` and `deviation`
statistics.

Read the requirement as *where do the two boundaries come from?* rather than
as a list of transforms. One answer is unlike the rest. `density` over a
category gives a ribbon its boundaries by **reflection**, and that is
[the violin](#the-violin-a-distribution-per-category), at the end of this
chapter.

## A confidence band

How well is the mean life expectancy known in each year? That is a different
question from how far the five countries spread. `range` shows the full
spread. For the **uncertainty of the mean**, use
`confidence`. It computes a t-interval for the mean at each `x`, and the ribbon
fills between its two bounds. The interval widens when the group is small:

```{r}
data(gapminder_asia) + ribbon * confidence(0.95) + x(year) + y(life) +
  y_label("Life expectancy") +
  title("95% confidence band for mean life expectancy")
```

*"Given gapminder Asia: a ribbon derived by confidence at 95%, x is year, y is
life."*

The band is wide because each mean comes from only five countries. Those five
differ by more than fifteen years, so the interval around their mean is wide. A
higher level draws a wider band still. What the band means is decided by the
statistic you name: `range` draws the observed spread, and `confidence` draws
the uncertainty of the mean. Neither one is a new mark to learn, which is the
rule the [`interval`](interval.qmd) family follows too.

## A pre-computed band

`range`, `confidence` and `deviation` compute the band *from raw values in the
plot*. Most confidence bands are not computed in the plot at all. A model's
standard error, a standard error of measurement, a bootstrap interval: each of
these is computed **before the plot**. Its two bounds then become two columns in
the table. gog draws what you computed, and it does not re-fit the model. The
`score_band` table holds one row per raw test score. Each row also holds the
score a model predicts for that raw score, and the two edges of the band around
the prediction:

```{r}
#| echo: false
peek(score_band)
```

The ribbon reads those edge columns directly, through `bounds(lower, upper)`:

```{r}
data(score_band) + ribbon * bounds(lower, upper) + x(score) +
  y_label("Expected true score") + x_label("Raw score") +
  title("A pre-computed measurement band")
```

*"Given the score band: a ribbon from lower to upper, x is score."*

`bounds` is `range`'s counterpart, and it computes nothing. It reshapes the two
columns you give it into the same low and high pair `range` produces. The band
needs no `y()`, because the two columns *are* the boundaries. This is the one
case `range`, `confidence` and `deviation` cannot serve: a band that was
computed outside the plot, before gog ever read the table.

A band like this is usually drawn together with the values it surrounds. Here
those values are the **expected true score**, the score the model predicts for
each raw score. A dashed [`line`](line.qmd) draws them, with a point at every
raw score. The band, the line and the points are three layers composed with
`+`:

```{r}
data(score_band) + x(score) + y(expected) +
  ribbon * bounds(lower, upper) + style(color = "steelblue", opacity = 0.18) +
  line + style(color = "steelblue", size = 1.4, pattern = "dashed") +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("Expected score with its measurement error band")
```

*"Given the score band: x is score, y is expected, a ribbon from lower to upper,
and also a line and points."*

The dashed line is [`style(pattern = )`](../style.qmd). On a stroke the texture
is a **dash**, so here it applies to the `line` and not to the band. The band
takes the same setting, with different values. On a fill the
texture is a **hatch**: `"hatch"`, `"crosshatch"`, `"stripes"`, `"grid"` or
`"dots"`, never `"dashed"`.

## Filled, or two dashed edges

A band is a low/high pair, and the *mark* chooses how to draw it. This is the
[No Exceptions](../design-laws.qmd#law-2-no-exceptions) rule again: one `bin`
draws as a bar chart, as a line and as a step histogram, and one pair draws in
as many ways. `ribbon` **fills** the pair. A [`line`](line.qmd) traces its
**two boundaries** instead, which is the unfilled band many journals prefer.
With `style(pattern = )` it becomes the dashed pair of curves those journals
often print:

```{r}
data(score_band) + x(score) + y(expected) +
  line * bounds(lower, upper) + style(color = "gray", pattern = "dashed", size = 1.2) +
  line + style(color = "steelblue", size = 1.5) +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("The same band, unfilled: two dashed edges")
```

*"Given the score band: x is score, y is expected, lines from lower to upper,
and also a line and points."*

`line * bounds` is one layer that draws *both* edges, splitting the pair into a
low curve and a high one. It therefore needs no second `y`. Two plain `line`
layers can draw two columns you already have. They cannot split a pair that
`range` computes, because that pair never becomes a column you can name.
`line * bounds` reads any pair the same way. `line * range` traces a min–max
envelope, and `line * confidence` traces the two edges of a confidence interval.
[`step`](step.qmd) `* bounds` draws the same two edges as staircases. Filled or
unfilled, solid or dashed, it is one pair and a choice of mark, never a new
option added to the band.

## One band per group

How does the spread of one continent compare with another's, year by year? One
band per continent puts the three ranges in one panel. A categorical column
*splits* a ribbon into one band per group. It is the same split that gives
[`area`](area.qmd) one region per group and `line` one stroke per group. Here
each continent gets its own band, the range of its countries' life expectancy
over time:

```{r}
#| echo: false
peek(gm_continents)
```

```{r}
data(gm_continents) + ribbon * range + x(year) + y(life) + color(continent) +
  y_label("Life expectancy") +
  title("Life expectancy spread, three continents")
```

*"Given the gapminder continents: ribbons derived by range, x is year, y is
life, color by continent."*

The bands overlap where two continents cover the same range of life expectancy.
A split `area` fills every region at the same opacity it uses for one, so only
a little of the region behind shows through. The `stack` transform is an area's
answer to that: it starts each region at the top of the ones below, so no two
regions overlap at all. A ribbon instead draws its split **translucent** by
default, so a band behind another is still readable.
The difference follows from what each mark measures. An area measures a height
from the baseline, so its groups can be piled into a total. A ribbon already
spans a low to a high and measures no such height, so there is nothing to stack.
Overlapping bands are read through transparency and never stacked, which is why
`ribbon * stack` is refused and the refusal names `style(opacity = )`.

## One region, one fill

A ribbon is a single filled region, so `opacity` is a *setting* here, not a
channel, exactly as on `area`. Each row of the table is one corner of a
boundary, not a piece of the fill. No part of the band belongs to one row, so a
column cannot make one row lighter or darker than another. [What it
refuses](#what-it-refuses) shows the refusal.

`size` cannot be set either. A band's extent comes from its x and its two
boundaries, so there is no width to choose. Its edge, if you want one, is a
`line` layered along a boundary, never an outline option on the mark. That is
the `ribbon + line` composition drawn in [A pre-computed
band](#a-pre-computed-band).

## A category on the domain

What happens when `x` holds categories instead of years? The spread of each
continent and its center can be drawn as one band across the categories. The
reader then reads the whole panel as one shape, the way they read an area's
filled profile. A ribbon takes a category on `x`, the same way an
[`area`](area.qmd) does, and fills the spread across the categories in axis
order:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + ribbon * range + line * mean + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * range + line * mean: spread and center, per continent")
```

For most such data an [`interval`](interval.qmd) is the better sentence. It
draws a separate whisker at each category and claims nothing about the space
between them. A ribbon instead slopes from one category to the next, across
positions where no rows exist. The grammar does not decide that for you: both
sentences are well formed, and which one fits *your* data is a question about
the data.

```{r}
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("interval * range: the same numbers, claiming less")
```

*"Given gapminder 2007: intervals derived by range, x is continent, y is life."*

## The violin: a distribution per category

A `range` on a category gives two numbers per group, the lowest and the highest.
`density` gives a whole *shape* per group, and a ribbon has room for that shape.
The estimate spreads along the measure. It is drawn across the category's slot
and mirrored, so the two halves are the same curve facing opposite ways. That is
the violin plot [@hintze1998], and it is this chapter's sentence with one word
changed:

```{r}
data(gapminder_2007) + ribbon * density + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * density: one distribution per continent")
```

*"Given gapminder 2007: ribbons derived by density, x is continent, y is life."*

`density` is not the ribbon's own transform. Five other marks estimate the same
curve, and only the shape changes:

| Mark derived by `density` | Shape |
|---|---|
| `line * density` | the curve as a stroke |
| `step * density` | the curve as a staircase |
| `area * density` | the curve filled to the baseline |
| `point * density` | one dot at each point the curve was estimated |
| `bar * density` | one thin bar at each of those points |
| `ribbon * density` | the curve mirrored, the violin above |

`line`, `step` and `area` also read a category on one axis, exactly as the
violin does. `point` and `bar` need a continuous axis instead: one dot or one
thin bar per estimated point has no reading inside a category's slot, so gog
refuses both there.

Nothing new was added to the grammar to draw it. A ribbon closes on a second
boundary the data supplies, and a violin closes on the mirror image of its own
curve. `density` is the transform that estimates a distribution, and this plot
is that estimate read per group. There is no `violin` word to learn. Such a word
would have to say four things at once: the mark, the statistic, the mirror and
the slot. Every one of those is already said by something you know.

Read it against the [`box`](box.qmd) of the same two columns. A box gives you
five numbers you can name. The violin gives you the shape those five numbers
summarize. Africa's violin is widest near 52 years, and it has a second, smaller
bulge near 73. Its median is 53, and a single number cannot report that second
group at all.

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  y_label("Life expectancy") + title("box + x + y: the same distributions, summarized")
```

*"Given gapminder 2007: boxes, x is continent, y is life."*

### What the width means

Every group's density integrates to 1 on its own. If the estimate stopped there,
the widths would say nothing about how many rows each group has. A continent of
two countries would be drawn as wide as a continent of fifty-two. So by default
the estimate is weighted by the group's row count, and width means the same
thing across the whole panel. Oceania holds two countries, so it is the
narrowest violin in the plot above.

The weighting is on the shape's **area**, not on its width, which is why Europe
is drawn wider there than Africa. Europe's thirty countries sit within ten years
of each other, so Europe's area is packed into a short stretch of the axis and
the shape has to be wide. Africa's fifty-two countries are spread over
thirty-seven years, so a larger area is spread thin. What the width reports at
any height is countries per year of life expectancy, and that reading is the
same in every slot.

`compare = "shape"` removes that weighting and draws every violin to the same
area. It is the right reading when the question is about the shapes alone.
It is the wrong one when a group whose values are packed together takes the
width away from every other violin:

```{r}
data(gapminder_2007) + ribbon * density(compare = "shape") + x(continent) + y(life) +
  y_label("Life expectancy") +
  title('density(compare = "shape"): every violin the same area')
```

*"Given gapminder 2007: ribbons derived by density compared by shape, x is
continent, y is life."*

Two countries now fill more space than fifty-two. Oceania's two values are one
year apart, so its equal share of area has to fit inside one year of the axis.
The only way to fit that area there is to be very wide. The widest shape in a
panel fixes the scale, so every other violin is drawn thinner against it. Each
plot answers a different question, which is why the choice is a word you write
rather than a default nobody stated.

### Sideways, and halved

The category can go on either axis, and the violins turn to follow it. The
orientation is read off the bindings, exactly as a [`bar`](bar.qmd)'s or a
[`box`](box.qmd)'s is. A category on `y` is the form with room for long category
names:

```{r}
data(gapminder_2007) + ribbon * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("The same violins, turned sideways")
```

An [`area`](area.qmd) draws the same estimate closed on the slot's center line
instead of on its own reflection. So the area draws one side of the shape and
the ribbon draws two, which is what separates these two marks everywhere else in
the book:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  y_label("Life expectancy") + title("area * density: the half violin")
```

*"Given gapminder 2007: areas derived by density, x is continent, y is life."*

There are two choices here, and they are made independently: which mark closes
the region, and which axis carries the category. That is four plots, and the
fourth one has a name of its own. Halve the violin *and* turn it sideways. Each
category then gets a filled curve on its own line above a shared measure axis,
which is the **ridgeline plot**:

```{r}
data(gapminder_2007) + area * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("area * density + x(life): the ridgeline")
```

Nothing was added for it either. It is the half violin's sentence with `x` and
`y` exchanged. The plot exists because the two choices really are independent:
nothing in the grammar ties *where a region closes* to *which axis holds the
category*. All four combinations are sentences, and the named chart is simply
one of them.

Read the four plots together and you can see what
[orthogonality](../design-laws.qmd#law-1-orthogonality) gives you here. A word
for the violin and a word for the ridgeline would be two names for one idea, and
the other two combinations would have no word at all.

### Letting the ridges overlap

The widest shape in a panel reaches four tenths of a slot by default, and every
other shape is drawn to scale against it. Two of them in neighboring slots
would reach four tenths each, filling four fifths of the space between their
categories and leaving the last fifth empty. That is a [`bar`](bar.qmd)'s rule,
and for a bar's reason: the gap is what marks the categories as separate rather
than a divided continuum.

The ridgeline breaks that rule on purpose. As it is usually drawn (the form some
call a joyplot) it does not stay inside its slot. Its ridges are tall enough to
reach into the row above, and that overlap is part of the design.
`density(reach = )` says how far, **in slots**:

```{r}
data(gapminder_2007) + area * density(reach = 2.5) + x(life) + y(continent) +
  x_label("Life expectancy") + title("reach = 2.5: each ridge two and a half slots tall")
```

*"Given gapminder 2007: areas derived by density reaching 2.5 slots, x is life,
y is continent."*

The number is measured from the line a category sits on to the shape's furthest
point. That is why the number means the same thing to both marks. A `ribbon`
reaches that far on each side, an `area` on one side only. The half violin
therefore stays exactly half of the violin, at any reach. Past `0.5` the shapes
leave their own slots. The grammar does not guard against that. It is the plot
being asked for, and the later category simply draws over the earlier one.

The category axis grows to hold the ridges. Half a slot of margin is right for a
[`bar`](bar.qmd) or a [`box`](box.qmd), which stand *in* their slots. A ridge
2.5 slots tall does not stand in its slot, so the axis adds the room it needs.
Without that room the panel edge would cut the top ridge off, and you would see
the cut.

The edge is a layer here, as it is for every filled band in this chapter. A
`ribbon` takes no border setting, and neither does the violin. A
[`line`](line.qmd) given the same transform traces the outline of the fill. That
is the [Filled, or two dashed edges](#filled-or-two-dashed-edges) rule again,
this time inside a category's slot:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.5) + style(color = "steelblue", opacity = 0.9) +
  line * density(reach = 2.5) + style(color = "black", size = 1) +
  x_label("Life expectancy") + title("The same ridges, with their edges drawn")
```

Both layers read the same `density(reach = 2.5)`, so the stroke lands exactly on
the edge of the fill it belongs to. That is the reason an edge is a layer rather
than a setting. Color each ridge by its own category and the plot is complete:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.2) + color(continent) +
  line * density(reach = 2.2) + style(color = "black", size = 0.9) +
  x_label("Life expectancy") + title("One hue per ridge")
```

The `color(continent)` is written *after* the area, so it belongs to that layer
alone. Written before both marks it would color the `line` too, and a layer
cannot both map color and set it. Where a channel sits in the sentence is what
decides its scope, and [Encoding
scope](../encoding-scope.qmd#channels-plot-scoped-or-layer-scoped-by-position)
explains that in full.

`step * density` traces the same estimate as a staircase, for the same reason
`step * bin` is the outline histogram. [No
Exceptions](../design-laws.qmd#law-2-no-exceptions) means that a mark draws
whatever rows a transform produces, in the mark's own way.

## What you can set

A setting changes how the band looks without reading a column, and each mark
takes its own. These are a ribbon's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("ribbon")
```

A ribbon takes `area`'s list, and for `area`'s reasons. There is no `size`,
because the band's extent comes from the two boundaries its transform supplies.
There is no border, because the edge is a `line` traced along the same pair.

Here is the textured band with a `line` along both of its edges:

```{r}
data(score_band) + x(score) +
  ribbon * bounds(lower, upper) +
    style(color = "slateblue", pattern = "grid", opacity = 0.45) +
  line * bounds(lower, upper) + style(color = "slateblue", pattern = "dashed") +
  y_label("Expected true score") + title("A textured band, and its two edges")
```

*"Given the score band: x is score, a ribbon from lower to upper, with pattern
grid, and also lines from lower to upper, with pattern dashed."*

Both marks read the same `bounds(lower, upper)` there, exactly as the two ridge
layers did, so the edges stay on the band. Note the two `pattern` values in one
sentence. `"grid"` is a fill texture, and the ribbon is a fill. `"dashed"` is a
stroke dash, and the line is a stroke. The setting has one name, and each
geometry has its own set of values for it.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The measure stays numeric. A band from `range` spans a low value to a high one,
and both are quantities, so a category on `y` is refused. The violins above put
a category on `y`, but there `density` makes it a slot rather than the measure:

```{r}
#| error: true
data(gapminder_2007) + ribbon * range + x(life) + y(continent)
```

A bare `ribbon` is the second refusal. A plain `y()` supplies one number, and a
band needs two boundaries, as [Why a ribbon needs a
transform](#why-a-ribbon-needs-a-transform) explained. The engine says so
rather than drawing an empty panel:

```{r}
#| error: true
data(gapminder_asia) + ribbon + x(year) + y(life)
```

A mapped `opacity` is the third. Each row is one corner of a boundary, so no
part of the band belongs to a single row, as [One region, one
fill](#one-region-one-fill) explained. Opacity is set with `style(opacity = )`
and never mapped:

```{r}
#| error: true
data(gapminder_asia) + ribbon * range + x(year) + y(life) + opacity(gdp)
```

`ribbon` completes the low/high family: [`interval`](interval.qmd) draws the
pair as a whisker at each x, [`line`](line.qmd) and [`step`](step.qmd) trace its
two boundaries, and `ribbon` fills between them. One `range`, four marks, which
is [No Exceptions](../design-laws.qmd#law-2-no-exceptions) made visible. See
[Transforms](../transforms.qmd) for the `range`, `confidence` and `deviation`
statistics all four share.

## In short

A ribbon fills a pair, and refuses to draw until a pair transform gives it one.
The pair you choose is what decides the band's meaning: the extremes from
`range`, the middle from two quantiles, the mean's uncertainty from
`confidence`. Both edges are data, so the band's shape is never a setting.

## What comes next

A ribbon shows a region and names nothing inside it. [Text](text.qmd) draws the
words: one string per row, read from a column, wherever the positions put it.
