# Ribbon

```{r}
#| include: false
source("../R/setup.R")
```

How wide is the band the values fall in, across a span? `ribbon` draws a
filled band from a low boundary to a high one across x: a confidence band,
a spread band, the shaded region around a trend. It is [`area`](area.qmd) fed
by [`interval`](interval.qmd)'s machinery: `area`'s single filled region, but
closed on a *second data boundary* instead of a baseline at zero. Where an area
fills from the data down to the ground, a ribbon fills between two lines the
data draws.

## A band from a range

The two boundaries come from a transform, exactly as an interval's two extents
do. `range` reduces `y` to its minimum and maximum within each `x`, and the
ribbon fills between them. Here, the spread of life expectancy across five Asian
countries, year by year:

```{r}
data(gapminder_asia) + ribbon * range + x(year) + y(life) +
  y_label("Life expectancy") +
  title("The min–max spread across five Asian countries, per year")
```

*"Given gapminder Asia: a ribbon derived by range, x is year, y is life."*

Every year, the band's lower edge is the shortest-lived country and its upper
edge the longest-lived; the band is the region that contains all five. Read it as
`interval * range` drawn as a *continuous* fill rather than a whisker at each x:
the same low/high pair, one geometry apart.

## The mean and its band

On its own a band shows spread but not center. The band's natural partner is a
line through the middle, and superposition, not a new option on either mark,
is how you compose them. A `ribbon * range` for the spread, a `line * mean` for
the trend, sharing one `x` and `y`:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  ribbon * range + style(color = "steelblue", opacity = 0.25) +
  line * mean + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Mean life expectancy, with its min–max band")
```

Each layer runs its own transform on the shared data: `mean` collapses each year
to one point for the line, `range` to a low/high pair for the band. This is the
archetypal ribbon plot, a trend with the uncertainty or spread drawn around it,
and it is two marks the grammar already has, added with `+`.

## A fan chart of nested bands

A min–max band shows what every country did. With 142 countries, only the two
extreme countries set its edges. Naming two quantiles instead gives a band that
contains most of the countries. `gm_all` is the whole gapminder table, every
country in every year, so layer two bands of different widths over it and you
have a **fan chart**:

```{r}
data(gm_all) + x(year) + y(life) +
  ribbon * range(0.1, 0.9) + style(color = "steelblue", opacity = 0.2) +
  ribbon * range(0.25, 0.75) + style(color = "steelblue", opacity = 0.35) +
  line * median + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Life expectancy across 142 countries, 1952 to 2007")
```

The pale band holds the middle 80% of countries. The darker one holds the
middle half, and the line is the median. The plot then says something the min–max
version cannot. The median rose from 45 years to 72, and the 90th percentile from
67 to 80, so the distance between them fell from 22 years to 8.

The bottom of the plot moves the other way. Both lower edges fall between 1992
and 2002, the 10th percentile and the 25th, while the median rises throughout.
A min–max band cannot show this, because its lower edge is one country.

Each band is one more layer, so a third pair adds a third. The bands must be
drawn widest first, because a later layer is painted over an earlier one.

## Why a ribbon needs a transform

A low *and* a high are two numbers, and a plain `y()` supplies one. So (like
`interval`, and unlike `area`, which closes on the zero the grammar already
knows) a ribbon's minimum syllable includes a range-producing transform. Ask for
one without it and the engine says so rather than drawing an empty panel:

```{r}
#| error: true
data(gapminder_asia) + ribbon + x(year) + y(life)
```

This is the same design choice that keeps the [channel
vocabulary](../grammar.qmd) small. The alternative, `ymin`/`ymax` aesthetics,
would add two channels meaningful to a single mark. Instead the boundaries are
**invented by a transform**, exactly as `bin` invents a count, so the same
low/high-rows mechanism feeds the whisker, the box, and the band alike; see
[Transforms](../transforms.qmd) for the `range` and `confidence` statistics.

Read the requirement as *where are my two boundaries?* rather than as a list of
three transforms. There is a fourth answer, and it looks nothing like the other
three: `density` over a category gives a ribbon its boundaries by
**reflection**. That is [the violin](#the-violin-a-distribution-per-category),
at the end of this chapter.

## A confidence band

`range` shows the full spread. For the **uncertainty of the mean**, the shaded
band of a regression or a group mean, use `confidence`, which computes the
mean's t-interval per `x` and fills between its bounds:

```{r}
data(gapminder_asia) + ribbon * confidence(0.95) + x(year) + y(life) +
  y_label("Life expectancy") +
  title("95% confidence band for mean life expectancy")
```

The band is narrow because the five countries agree closely on the mean in each
year; a wider level draws a wider band. Which statistic you ask for chooses the
band's meaning, the observed spread (`range`) or the uncertainty of the mean
(`confidence`), with no new mark to learn, the same rule the
[`interval`](interval.qmd) family follows.

## A pre-computed band

`range` and `confidence` compute the band *from raw values in the plot*. But most
confidence bands are not computed at the plot at all: a model's standard error, a
psychometric standard error of measurement, a bootstrap interval are all worked out
**upstream**, and you arrive with the bounds already in two columns. gog draws what
you computed; it does not re-fit the model. So the ribbon reads those columns
directly, through `bounds(lower, upper)`:

```{r}
data(score_band) + ribbon * bounds(lower, upper) + x(score) +
  y_label("Expected true score") + x_label("Raw score") +
  title("A pre-computed measurement band")
```

`bounds` is `range`'s counterpart that **computes nothing**: it reshapes the two
columns you hand it into the same low/high pair `range` produces, so the band draws
with no new machinery and no `y()` (the two columns *are* the extents). It is the
common case the summary transforms cannot serve: a band whose bounds came from
somewhere gog will never be.

Its natural companion is the curve the band surrounds. Here the scoring model's
**expected true score** runs through the middle, drawn as a dashed
[`line`](line.qmd) with a marker at each integer raw score (the band, the trend,
and the score points, composed with `+`):

```{r}
data(score_band) + x(score) + y(expected) +
  ribbon * bounds(lower, upper) + style(color = "steelblue", opacity = 0.18) +
  line + style(color = "steelblue", size = 1.4, pattern = "dashed") +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("Expected score with its 95% CSEM band")
```

The dashed line is [`style(pattern = )`](../style.qmd): on a stroke the texture is a
**dash**, so it rides the `line` here, not the band. The band can take the same
setting too: there it is a **hatch** (a fill texture), not a dash.

## Filled, or two dashed edges

A band is a low/high pair, and the *mark* chooses how to draw it: the same
[No Exceptions](../design-laws.qmd) rule that gives one `bin` a bar, a line, and a
step histogram. `ribbon` **fills** the pair; a [`line`](line.qmd) traces its **two
boundaries**, the unfilled band publications often prefer, and with
`style(pattern = )`, the classic dashed ± curves:

```{r}
data(score_band) + x(score) + y(expected) +
  line * bounds(lower, upper) + style(color = "gray", pattern = "dashed", size = 1.2) +
  line + style(color = "steelblue", size = 1.5) +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("The same band, unfilled: two dashed edges")
```

`line * bounds` is one layer that draws *both* edges, splitting the pair into a
low locus and a high one. It therefore needs no second `y`. That is why the two
boundaries come out as separate curves, where two hand-written `line` layers
could not manage it. It reads any pair the same way: `line * range` traces a min–max envelope,
`line * confidence` the CI edges, and [`step`](step.qmd) `* bounds` draws them as
staircases (stepped control limits). Filled or unfilled, solid or dashed, it is one
pair and a choice of mark, never a new option bolted onto the band.

## One band per group

A categorical column *splits* a ribbon into one band per group, the same split
[`area`](area.qmd) and `line` make. Here each continent gets its own band, the
range of its countries' life expectancy over time:

```{r}
data(gm_continents) + ribbon * range + x(year) + y(life) + color(continent) +
  y_label("Life expectancy") +
  title("Life expectancy spread, three continents")
```

The bands overlap where the continents do, and unlike a split `area`, which
stays opaque and points you at `stack`, a ribbon draws its split **translucent**
by default, so a band behind another still shows through. That difference is
principled: an area measures a height from the baseline, so its groups can be
piled into a total; a ribbon already spans a low to a high and measures no such
height, so there is nothing to stack. Transparency, not accumulation, is a band's
honest answer to overlap, which is why `ribbon * stack` is refused with
direction toward `style(opacity = )`.

## One region, one fill

A ribbon is a single filled region, so, exactly as on `area`, `opacity` is a
*setting*, not a channel: a row is a vertex of a boundary, not a region of its
own, so there is nothing for a per-row opacity to vary along.

```{r}
#| error: true
render_svg(data(gapminder_asia) + ribbon * range + x(year) + y(life) + opacity(gdp))
```

And `size` cannot even be set: a band's extent is pinned by its x and its two
boundaries, with no width left to choose. Its edge, if you want one, is a `line`
layered along a boundary (`ribbon + line`, the composition the previous section
already drew), never an outline option on the mark.

## A category on the domain

A ribbon takes a category on `x`, the same way an [`area`](area.qmd) does, and
fills the spread across the categories in axis order:

```{r}
data(gapminder_2007) + ribbon * range + line * mean + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * range + line * mean: spread and center, per continent")
```

For most such data an [`interval`](interval.qmd) is the better sentence. It draws
a separate whisker at each category and claims nothing about the space between
them, where the band's edges slope across ground that holds no observations. The
grammar does not decide that for you: both expressions are well formed, and which
one is honest about *your* data is a question about the data.

```{r}
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("interval * range: the same numbers, claiming less")
```

## The violin: a distribution per category

A `range` on a category gives two numbers per group, the lowest and the highest.
`density` gives a whole *shape* per group, and a ribbon has somewhere to put it:
the estimate spreads along the measure and is drawn across the category's slot,
mirrored. That is the violin plot [@hintze1998], and it is this chapter's
sentence with one word changed:

```{r}
data(gapminder_2007) + ribbon * density + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * density: one distribution per continent")
```

Nothing new was added to the grammar to draw it. A ribbon is the mark that closes
on a second data boundary, and a violin closes on its own reflection, which is
one; `density` is the transform that estimates a distribution, and this is that
estimate read per group. There is no `violin` word to learn, because a word for
it would be four things at once (the mark, the statistic, the mirror, and the
slot), and each of those is already said by something you know.

Read it against the [`box`](box.qmd) of the same two columns. A box gives you
five numbers you can name; the violin gives you the shape those five numbers are
a summary of, and shows the two humps a median hides.

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  y_label("Life expectancy") + title("box + x + y: the same distributions, summarized")
```

### What the width means

Every group's density integrates to 1 on its own, so left alone the widths would
say nothing about how much data each group has: a continent of two countries
would be drawn as boldly as a continent of fifty-two. So by default the estimate
is weighted by the group's row count, and width means the same thing across the
whole panel. Oceania above is two countries, and it is drawn as two countries.

`compare = "shape"` takes that weighting back off and draws every violin to the
same area. It is the right reading when the question is about the shapes and the
groups are of unequal size on purpose, and the wrong one when a small group's
spike can take the panel:

```{r}
data(gapminder_2007) + ribbon * density(compare = "shape") + x(continent) + y(life) +
  y_label("Life expectancy") +
  title('density(compare = "shape"): every violin the same area')
```

Two countries now occupy as much ink as fifty-two, and because two nearby values
make a tall narrow estimate, Oceania takes the width and everything else is
squeezed. Both plots are honest about something different, which is why the
choice is a word you write rather than a default nobody stated.

### Sideways, and halved

The category can go on either axis, and the violins turn to follow it. Read off
the bindings, exactly as a [`bar`](bar.qmd)'s or a [`box`](box.qmd)'s orientation
is. It is the form with room for long category names:

```{r}
data(gapminder_2007) + ribbon * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("the same violins, lying down")
```

And an [`area`](area.qmd) draws the same estimate closed on the slot's center
line instead of on its own reflection: one side rather than two, since that is
the difference between the two marks everywhere else in the book:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  y_label("Life expectancy") + title("area * density: the half violin")
```

Two choices, made independently: which mark closes the region, and which axis
carries the category. That is four plots, and the fourth one has a name of its
own. Halve the violin *and* lay it down, and each category gets a filled curve
on its own line above a shared measure axis, which is the **ridgeline plot**:

```{r}
data(gapminder_2007) + area * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("area * density + x(life): the ridgeline")
```

Nothing was added for it either. It is the previous chunk's sentence with `x`
and `y` exchanged. It arrives because the two choices are genuinely independent:
nothing in the grammar ties *where a region closes* to *which axis holds the
category*. All four combinations are sentences, and the one the world gave a
name to is simply one of them.

Read the four together and the point of
[orthogonality](../design-laws.qmd) is easier to see here than anywhere else in
this chapter. A word for the violin and a word for the ridgeline would be two
names for one idea, and the other two combinations would have no word at all.

### Letting the ridges overlap

Each shape reaches four tenths of a slot by default, so two violins face to face
fill four fifths of the space between their categories and the remaining fifth is
air. That is a [`bar`](bar.qmd)'s own rule, for a bar's reason: the gap is what
says the categories are separate rather than a divided continuum.

The joyplot as it is usually drawn is not like that. Its ridges are tall enough
to run into the row above, and that overlap is the look. `density(reach = )` says
how far, **in slots**:

```{r}
data(gapminder_2007) + area * density(reach = 2.5) + x(life) + y(continent) +
  x_label("Life expectancy") + title("reach = 2.5: each ridge two and a half slots tall")
```

The number is measured from the line a category sits on to the shape's furthest
point. That is why it means the same thing to both marks: a `ribbon` reaches it
each way and an `area` one way, so the half violin stays exactly half of the
violin at any value. Past `0.5` the shapes leave their own slots. The grammar
does not guard against that. It is the plot being asked for, and the later
category simply draws over the earlier one.

The category axis grows to hold them. Half a slot of margin is right for a
[`bar`](bar.qmd) or a [`box`](box.qmd), which stand *in* their slots. A ridge
2.5 slots tall does not, and without the extra room the frame would cut the top
one off. That would be a mark drawn outside its own panel: a plot that is wrong
in a way nothing but a reader would notice.

The edge is a layer, as it is for every filled band in this chapter: a `ribbon`
takes no border setting, and neither does the violin, because
[`line`](line.qmd) reads the same slot reading and **traces** what the fill
fills. That is the "filled, or two edges" rule again, now against a slot:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.5) + style(color = "steelblue", opacity = 0.9) +
  line * density(reach = 2.5) + style(color = "black", size = 1) +
  x_label("Life expectancy") + title("The same ridges, with their edges drawn")
```

Both layers read the same `density(reach = 2.5)`, so the stroke cannot drift off
the fill it belongs to, the reason an edge is a layer rather than a setting.
Color each ridge by its own category and the plot is complete:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.2) + color(continent) +
  line * density(reach = 2.2) + style(color = "black", size = 0.9) +
  x_label("Life expectancy") + title("One hue per ridge")
```

The `color(continent)` is written *after* the area, so it belongs to that layer
alone: a plot-scoped one would reach the `line` too, and a layer cannot both map
color and set it. [Scope is position](../encoding-scope.qmd), and this is the
sentence where it matters.

`step * density` traces the same estimate as a staircase, for the same reason
`step * bin` is the outline histogram: [No Exceptions](../design-laws.qmd) means
a mark draws whatever a transform hands it, in the mark's own way.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("ribbon")
```

A ribbon takes `area`'s list, and for `area`'s reasons: no `size`, because the
band's extent is pinned by the two boundaries a range transform supplies, and no
border, because the edge is a `line` traced along the same pair.

```{r}
data(score_band) + x(score) +
  ribbon * bounds(lower, upper) +
    style(color = "slateblue", pattern = "grid", opacity = 0.45) +
  line * bounds(lower, upper) + style(color = "slateblue", pattern = "dashed") +
  y_label("Expected true score") + title("A textured band, and its two edges")
```

Both marks read the same `bounds(lower, upper)` there, so the dashed edges cannot
drift from the band they belong to. Note the two `pattern` values in one
sentence: `"grid"` is a fill texture on the ribbon and `"dashed"` a stroke dash on
the line, the same setting realized once per geometry.

## What it refuses

The measure stays numeric. A band spans from a low value to a high one, and the
two extents are quantities, so a categorical `y` is refused:

```{r}
#| error: true
render_svg(data(gapminder_2007) + ribbon * range + x(life) + y(continent))
```

`ribbon` completes the low/high family: [`interval`](interval.qmd) draws the pair
as a whisker at each x, [`box`](box.qmd) as a five-number summary, and `ribbon` as
a filled band across x. One `range`, three marks: [No
Exceptions](../design-laws.qmd) made visible. See
[Transforms](../transforms.qmd) for the `range` and `confidence` statistics all
three share.
