# Ribbon

```{r}
#| include: false
source("../R/setup.R")
```

How wide is the band of values at each point along x? `ribbon` draws a filled
band from a low boundary to a high one across x: a confidence band, a spread
band, the shaded region around a trend. It takes [`area`](area.qmd)'s single
filled region and gives it [`interval`](interval.qmd)'s low and high pair. The
region then closes on a *second data boundary* instead of on a baseline at zero.
Where an area fills from the data down to zero, a ribbon fills between two
boundaries the data supplies.

## A band from a range

Did the five countries move closer together over the years, or further apart?
The gap between the lowest and the highest in each year answers that. A band
drawn between the two shows it as one shape. The two boundaries come from a
transform, exactly as an interval's two extents do. `range` reduces `y` to its
minimum and maximum within each `x`, and the ribbon fills between them. Here,
the spread of life expectancy across five Asian countries, year by year:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + ribbon * range + x(year) + y(life) +
  y_label("Life expectancy") +
  title("The min–max spread across five Asian countries, per year")
```

*"Given gapminder Asia: a ribbon derived by range, x is year, y is life."*

Every year, the band's lower edge is the shortest-lived country and its upper
edge the longest-lived; the band is the region that contains all five. Read it
as `interval * range` drawn as a *continuous* fill rather than a whisker at each
x. The pair is the same; only the geometry differs.

## The mean and its band

On its own a band shows spread but not center. A band is usually drawn with a
line through the middle. You compose the two by layering, not by adding an
option to either mark. A `ribbon * range` for the spread, a `line * mean` for
the trend, sharing one `x` and `y`:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  ribbon * range + style(color = "steelblue", opacity = 0.25) +
  line * mean + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Mean life expectancy, with its min–max band")
```

*"Given gapminder Asia: x is year, y is life, a ribbon derived by range and also
a line derived by mean."*

Each layer runs its own transform on the shared data: `mean` collapses each year
to one point for the line, `range` to a low/high pair for the band. This is the
most common ribbon plot: a trend with its uncertainty or spread drawn around it.
It is two marks the grammar already has, added with `+`.

## A fan chart of nested bands

`gm_all` is the whole gapminder table: every country, in every year, 142
countries in all. A min–max band over it shows what every country did, but only
the two extreme countries set its edges. Naming two quantiles instead gives a
band that contains most of the countries. Layer two bands of different widths
and you have a **fan chart**:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + x(year) + y(life) +
  ribbon * range(0.1, 0.9) + style(color = "steelblue", opacity = 0.2) +
  ribbon * range(0.25, 0.75) + style(color = "steelblue", opacity = 0.35) +
  line * median + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Life expectancy across 142 countries, 1952 to 2007")
```

*"Given all the gapminder years: x is year, y is life, a ribbon derived by range
from 0.1 to 0.9, and also a ribbon derived by range from 0.25 to 0.75, and also
a line derived by median."*

The pale band contains the middle 80% of countries. The darker one contains the
middle half, and the line is the median. The plot then says something the
min–max version cannot. The median rose from 45 years to 72, and the 90th
percentile from 67 to 80, so the distance between them fell from 22 years to 8.

The lower edges of the two bands do not rise steadily. Both lower edges, the
10th percentile and the 25th, decline between 1992 and 2002 while the median
rises throughout. A min–max band cannot show this, because its lower edge is one
country.

Each band is one more layer, so a third pair adds a third. Write the widest band
first, because a later layer is painted over an earlier one.

## Why a ribbon needs a transform

A low *and* a high are two numbers, and a plain `y()` supplies one. An `area`
closes on zero, which the grammar supplies. A ribbon has no such second
number, so like `interval` its minimum syllable includes a range-producing
transform. Ask for
one without it and the engine says so rather than drawing an empty panel, as
[What it refuses](#what-it-refuses) shows.

This is the same design choice that keeps the [channel
vocabulary](../grammar.qmd) small. The alternative would add `ymin` and `ymax`
as two more channels meaningful to a single mark. Instead the boundaries are
invented by a transform, exactly as `bin` invents a count. The same low and high
rows then serve the whisker, the box and the band alike. See
[Transforms](../transforms.qmd) for the `range`, `confidence` and `deviation`
statistics.

Read the requirement as *where do the two boundaries come from?* rather than
as a list of transforms. One answer looks nothing like the others: `density`
over a category gives a ribbon its boundaries by
**reflection**. That is [the violin](#the-violin-a-distribution-per-category),
at the end of this chapter.

## A confidence band

How well is the mean life expectancy known in each year? That is a different
question from how far the five countries spread. `range` shows the full
spread. For the **uncertainty of the mean**, use
`confidence`. It computes a t-interval for the mean at each `x`, and the ribbon
fills between its two bounds. The interval widens when the group is small:

```{r}
data(gapminder_asia) + ribbon * confidence(0.95) + x(year) + y(life) +
  y_label("Life expectancy") +
  title("95% confidence band for mean life expectancy")
```

*"Given gapminder Asia: a ribbon derived by confidence at 95%, x is year, y is
life."*

The band is wide because each mean comes from only five countries. Those five
differ by more than fifteen years, so the interval around their mean is wide. A
higher level draws a wider band still. The statistic you ask for chooses the
band's meaning: the observed spread with `range`, the uncertainty of the mean
with `confidence`. There is no new mark to learn, which is the rule the
[`interval`](interval.qmd) family follows.

## A pre-computed band

`range`, `confidence` and `deviation` compute the band *from raw values in the
plot*. But most confidence bands are not computed in the plot at all. A model's
standard error, a standard error of measurement, a bootstrap interval: each is
computed **before the plot**, and its bounds arrive as two columns. gog draws
what you computed; it does not re-fit the model. The `score_band` table holds
one row per raw test score, with the score a model predicts for it and the two
edges of the band around that prediction:

```{r}
#| echo: false
peek(score_band)
```

The ribbon reads those edge columns directly, through `bounds(lower, upper)`:

```{r}
data(score_band) + ribbon * bounds(lower, upper) + x(score) +
  y_label("Expected true score") + x_label("Raw score") +
  title("A pre-computed measurement band")
```

*"Given the score band: a ribbon from lower to upper, x is score."*

`bounds` is `range`'s counterpart, and it computes nothing. It reshapes the two
columns you give it into the same low and high pair `range` produces. The band
needs no `y()`, because the two columns *are* the boundaries. It serves the case
`range`, `confidence` and `deviation` cannot: a band computed before the data
reached the plot.

It is usually drawn with the curve the band surrounds. Here that curve is the
**expected true score**, the score the model predicts for each raw score. A
dashed [`line`](line.qmd) draws it, with a point at every raw score. The
band, the line and the points are three layers composed with `+`:

```{r}
data(score_band) + x(score) + y(expected) +
  ribbon * bounds(lower, upper) + style(color = "steelblue", opacity = 0.18) +
  line + style(color = "steelblue", size = 1.4, pattern = "dashed") +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("Expected score with its measurement error band")
```

*"Given the score band: x is score, y is expected, a ribbon from lower to upper,
and also a line and points."*

The dashed line is [`style(pattern = )`](../style.qmd): on a stroke the texture
is a **dash**, so it applies to the `line` here, not to the band. The band takes
the same setting, with different values. On a fill the texture is a **hatch**:
`"hatch"`, `"crosshatch"`, `"grid"` or `"dots"`, never `"dashed"`.

## Filled, or two dashed edges

A band is a low/high pair, and the *mark* chooses how to draw it: the same [No
Exceptions](../design-laws.qmd) rule that gives one `bin` a bar, a line, and a
step histogram. `ribbon` **fills** the pair; a [`line`](line.qmd) traces its
**two boundaries**. That is the unfilled band many journals prefer, and with
`style(pattern = )` it becomes the dashed pair of curves they often draw:

```{r}
data(score_band) + x(score) + y(expected) +
  line * bounds(lower, upper) + style(color = "gray", pattern = "dashed", size = 1.2) +
  line + style(color = "steelblue", size = 1.5) +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("The same band, unfilled: two dashed edges")
```

*"Given the score band: x is score, y is expected, lines from lower to upper,
and also a line and points."*

`line * bounds` is one layer that draws *both* edges, splitting the pair into a
low curve and a high one. It therefore needs no second `y`. Two plain `line`
layers can draw two columns you already have. They cannot split a pair that
`range` computes, because that pair never becomes a column you can name. `line *
bounds` reads any pair the same way: `line * range` traces a min–max envelope,
`line * confidence` the two edges of a confidence interval, and
[`step`](step.qmd) `* bounds` draws them as staircases. Filled or unfilled,
solid or dashed, it is one pair and a choice of mark, never a new option added
to the band.

## One band per group

How does the spread of one continent compare with another's, year by year? One
band per continent puts the three ranges in one panel. A categorical column
*splits* a ribbon into one band per group, the same split
[`area`](area.qmd) and `line` make. Here each continent gets its own band, the
range of its countries' life expectancy over time:

```{r}
#| echo: false
peek(gm_continents)
```

```{r}
data(gm_continents) + ribbon * range + x(year) + y(life) + color(continent) +
  y_label("Life expectancy") +
  title("Life expectancy spread, three continents")
```

*"Given the gapminder continents: ribbons derived by range, x is year, y is
life, color by continent."*

The bands overlap where two continents cover the same range of life expectancy.
A split `area` stays opaque, and `stack` is its answer. A ribbon instead draws
its split **translucent** by default, so a band behind another still shows
through. That difference is principled: an area measures a height from the
baseline, so its groups can be piled into a total; a ribbon already spans a low
to a high and measures no such height, so there is nothing to stack. Overlapping
bands are read through transparency, never stacked, which is why `ribbon *
stack` is refused and the refusal names `style(opacity = )`.

## One region, one fill

A ribbon is a single filled region, so `opacity` is a *setting* here, not a
channel, exactly as on `area`. Each row of the table is one corner of a
boundary, not a piece of the fill. No part of the band belongs to one row, so a
column cannot make one row lighter or darker than another. [What it
refuses](#what-it-refuses) shows the refusal.

And `size` cannot even be set: a band's extent comes from its x and its two
boundaries, so there is no width to choose. Its edge, if you want one, is a
`line` layered along a boundary (`ribbon + line`, the composition the
pre-computed band drew above), never an outline option on the mark.

## A category on the domain

The spread of each continent and its center can be drawn as one band across
the categories. It is then read as a shape, the way an area's filled profile
is. A ribbon takes a category on `x`, the same way an [`area`](area.qmd) does, and
fills the spread across the categories in axis order:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + ribbon * range + line * mean + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * range + line * mean: spread and center, per continent")
```

For most such data an [`interval`](interval.qmd) is the better sentence. It
draws a separate whisker at each category and claims nothing about the space
between them. A ribbon instead slopes from one category to the next, across
positions where no rows exist. The grammar does not decide that for you: both
sentences are well formed, and which one fits *your* data is a question about
the data.

```{r}
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("interval * range: the same numbers, claiming less")
```

*"Given gapminder 2007: intervals derived by range, x is continent, y is life."*

## The violin: a distribution per category

A `range` on a category gives two numbers per group, the lowest and the highest.
`density` gives a whole *shape* per group, and a ribbon has room for that shape:
the estimate spreads along the measure and is drawn across the category's slot,
mirrored. That is the violin plot [@hintze1998], and it is this chapter's
sentence with one word changed:

```{r}
data(gapminder_2007) + ribbon * density + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * density: one distribution per continent")
```

*"Given gapminder 2007: ribbons derived by density, x is continent, y is life."*

Nothing new was added to the grammar to draw it. A ribbon closes on a second
boundary the data supplies, and a violin closes on the mirror image of its own
curve. `density` is the transform that estimates a distribution, and this plot
is that estimate read per group. There is no `violin` word to learn, because a
word for it would be four things at once (the mark, the statistic, the mirror,
and the slot), and each of those is already said by something you know.

Read it against the [`box`](box.qmd) of the same two columns. A box gives you
five numbers you can name; the violin gives you the shape those five numbers
summarize. Africa's violin has two peaks, which its median hides.

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  y_label("Life expectancy") + title("box + x + y: the same distributions, summarized")
```

*"Given gapminder 2007: boxes, x is continent, y is life."*

### What the width means

Every group's density integrates to 1 on its own. Left at that, the widths would
say nothing about how many rows each group has. A continent of two countries
would be drawn as wide as a continent of fifty-two. So by default the estimate
is weighted by the group's row count, and width means the same thing across the
whole panel. Oceania above holds two countries, so its violin is drawn narrow.

`compare = "shape"` removes that weighting and draws every violin to the same
area. It is the right reading when the question is about the shapes alone.
It is the wrong one when a small group's narrow, tall estimate sets the width
for every other violin:

```{r}
data(gapminder_2007) + ribbon * density(compare = "shape") + x(continent) + y(life) +
  y_label("Life expectancy") +
  title('density(compare = "shape"): every violin the same area')
```

*"Given gapminder 2007: ribbons derived by density compared by shape, x is
continent, y is life."*

Two countries now fill as much space as fifty-two. Two nearby values make a
tall narrow estimate, so Oceania sets the width and every other violin is
drawn thinner. Both plots are honest about something different, which is why the
choice is a word you write rather than a default nobody stated.

### Sideways, and halved

The category can go on either axis, and the violins turn to follow it. The
orientation is read off the bindings, exactly as a [`bar`](bar.qmd)'s or a
[`box`](box.qmd)'s is. A category on `y` is the form with room for long category
names:

```{r}
data(gapminder_2007) + ribbon * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("the same violins, lying down")
```

An [`area`](area.qmd) draws the same estimate closed on the slot's center line
instead of on its own reflection. One side rather than two is the difference
between the two marks everywhere else in the book:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  y_label("Life expectancy") + title("area * density: the half violin")
```

*"Given gapminder 2007: areas derived by density, x is continent, y is life."*

Two choices, made independently: which mark closes the region, and which axis
carries the category. That is four plots, and the fourth one has a name of its
own. Halve the violin *and* lay it down, and each category gets a filled curve
on its own line above a shared measure axis, which is the **ridgeline plot**:

```{r}
data(gapminder_2007) + area * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("area * density + x(life): the ridgeline")
```

Nothing was added for it either. It is the half violin's sentence with `x` and
`y` exchanged. It arrives because the two choices are genuinely independent:
nothing in the grammar ties *where a region closes* to *which axis holds the
category*. All four combinations are sentences, and the named chart is simply
one of them.

Read the four plots together and you can see what
[orthogonality](../design-laws.qmd) buys. A word for the violin and a word for
the ridgeline would be two names for one idea, and the other two combinations
would have no word at all.

### Letting the ridges overlap

Each shape reaches four tenths of a slot by default, so two violins face to face
fill four fifths of the space between their categories, and the remaining fifth
stays empty. That is a [`bar`](bar.qmd)'s rule, and for a bar's reason: the gap
is what says the categories are separate rather than a divided continuum.

The ridgeline as it is usually drawn (the form some call a joyplot) does not
stay inside its slot. Its ridges are tall enough to reach into the row above,
and that overlap is part of the design. `density(reach = )` says how far, **in
slots**:

```{r}
data(gapminder_2007) + area * density(reach = 2.5) + x(life) + y(continent) +
  x_label("Life expectancy") + title("reach = 2.5: each ridge two and a half slots tall")
```

*"Given gapminder 2007: areas derived by density reaching 2.5 slots, x is life,
y is continent."*

The number is measured from the line a category sits on to the shape's furthest
point. That is why the number means the same thing to both marks. A `ribbon`
reaches that far on each side, an `area` on one side only. The half violin
therefore stays exactly half of the violin, at any reach. Past `0.5` the shapes
leave their own slots. The grammar does not guard against that. It is the plot
being asked for, and the later category simply draws over the earlier one.

The category axis grows to hold the ridges. Half a slot of margin is right for a
[`bar`](bar.qmd) or a [`box`](box.qmd), which stand *in* their slots. A ridge
2.5 slots tall does not stand in its slot, so the axis adds the room it needs.
Without that room the panel edge would cut the top ridge off, and you would see
the cut.

The edge is a layer here, as it is for every filled band in this chapter. A
`ribbon` takes no border setting, and neither does the violin. A
[`line`](line.qmd) given the same transform traces the outline of the fill.
That is the *filled, or two dashed edges* rule again, read against a slot:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.5) + style(color = "steelblue", opacity = 0.9) +
  line * density(reach = 2.5) + style(color = "black", size = 1) +
  x_label("Life expectancy") + title("The same ridges, with their edges drawn")
```

Both layers read the same `density(reach = 2.5)`, so the stroke cannot drift off
the fill it belongs to, the reason an edge is a layer rather than a setting.
Color each ridge by its own category and the plot is complete:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.2) + color(continent) +
  line * density(reach = 2.2) + style(color = "black", size = 0.9) +
  x_label("Life expectancy") + title("One hue per ridge")
```

The `color(continent)` is written *after* the area, so it belongs to that layer
alone: a plot-scoped one would color the `line` too, and a layer cannot both map
color and set it. [Scope is position](../encoding-scope.qmd), and this is the
sentence where it matters.

`step * density` traces the same estimate as a staircase, for the same reason
`step * bin` is the outline histogram: [No Exceptions](../design-laws.qmd) means
a mark draws whatever a transform hands it, in the mark's own way.

## What you can set

A setting changes how the band looks without reading a column, and each mark
takes its own. These are a ribbon's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("ribbon")
```

A ribbon takes `area`'s list, and for `area`'s reasons. There is no `size`,
because the band's extent comes from the two boundaries its transform supplies.
There is no border, because the edge is a `line` traced along the same pair.

Here is the textured band with a `line` along both of its edges:

```{r}
data(score_band) + x(score) +
  ribbon * bounds(lower, upper) +
    style(color = "slateblue", pattern = "grid", opacity = 0.45) +
  line * bounds(lower, upper) + style(color = "slateblue", pattern = "dashed") +
  y_label("Expected true score") + title("A textured band, and its two edges")
```

*"Given the score band: x is score, a ribbon from lower to upper, with pattern
grid, and also lines from lower to upper, with pattern dashed."*

Both marks read the same `bounds(lower, upper)` there, exactly as in the ridge
section, so the edges stay on the band. Note the two `pattern` values in one
sentence: `"grid"` is a fill texture on the ribbon and `"dashed"` a stroke dash
on the line, the same setting realized once per geometry.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The measure stays numeric. A band from `range` spans a low value to a high one,
and both are quantities, so a category on `y` is refused. The violins above put
a category on `y`, but there `density` makes it a slot rather than the measure:

```{r}
#| error: true
data(gapminder_2007) + ribbon * range + x(life) + y(continent)
```

A bare `ribbon` is the second refusal. A plain `y()` supplies one number, and a
band needs two boundaries, as [Why a ribbon needs a
transform](#why-a-ribbon-needs-a-transform) explained. The engine says so
rather than drawing an empty panel:

```{r}
#| error: true
data(gapminder_asia) + ribbon + x(year) + y(life)
```

A mapped `opacity` is the third. Each row is one corner of a boundary, so no
part of the band belongs to a single row, as [One region, one
fill](#one-region-one-fill) explained. Opacity is set with `style(opacity = )`
and never mapped:

```{r}
#| error: true
data(gapminder_asia) + ribbon * range + x(year) + y(life) + opacity(gdp)
```

`ribbon` completes the low/high family: [`interval`](interval.qmd) draws the
pair as a whisker at each x, [`line`](line.qmd) and [`step`](step.qmd) trace its
two boundaries, and `ribbon` fills between them. One `range`, four marks: [No
Exceptions](../design-laws.qmd) made visible. See
[Transforms](../transforms.qmd) for the `range`, `confidence` and `deviation`
statistics all four share.