# Ribbon

```{r}
#| include: false
source("../R/setup.R")
```

How far apart were the lowest and highest values at each date? `ribbon` draws a
filled band from a low boundary to a high one across x: a confidence band, a
spread band, the shaded region around a trend. It fills one region, the way an
[`area`](area.qmd) does. It reads a low value and a high value at each
position, the way an [`interval`](interval.qmd) does. An area fills from the
data down to a baseline at zero. A ribbon has no baseline, and both of its
edges come from the data, usually through a transform. With a category on one
axis and `density`, the same mark draws a violin plot.

## A band from a range

The `gapminder_asia` table holds five Asian countries, measured every five
years. Did those five move closer together over the years, or further apart?
The gap between the lowest and the highest in each year answers that. A band
drawn between those two values shows every year at once, as one shape. The two
boundaries come from a transform, exactly as an interval's two extents do.
`range` reduces `y` to its minimum and maximum within each `x`, and the ribbon
fills between them. Here, the spread of life expectancy across five Asian
countries, year by year:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + ribbon * range + x(year) + y(life) +
  y_label("Life expectancy") +
  title("The min–max spread across five Asian countries, per year")
```

*"Given gapminder Asia: a ribbon derived by range, x is year, y is life."*

In each year the band's lower edge sits at the lowest life expectancy of the
five, and its upper edge sits at the highest. The band is the region that holds
all five. It narrows from about 26 years wide in 1952 to about 18 in 2007, so
the five countries moved closer together.

The low and high pair is the one `interval * range` draws, filled continuously
instead of drawn as a whisker at each x. Only the shape drawn from it changes.

## A band with a mean line

On its own, a band shows the spread but not the center. A band is usually drawn
with a line through the middle. You compose the two by layering, not by adding
an option to either mark. A `ribbon * range` for the spread, a `line * mean` for
the trend, sharing one `x` and `y`:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  ribbon * range + style(color = "steelblue", opacity = 0.25) +
  line * mean + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Mean life expectancy, with its min–max band")
```

*"Given gapminder Asia: x is year, y is life, a ribbon derived by range and also
a line derived by mean."*

Each layer runs its own transform on the same table. `mean` reduces each year to
one point for the line, and `range` reduces the same year to a low and high pair
for the band. This is the most common ribbon plot: a trend with its uncertainty
or spread drawn around it. It is two marks the grammar already has, added with
`+`.

## A fan chart of nested bands

Which values of life expectancy do most countries have in each year? `gm_all`
is the whole gapminder table: every country, in every year, 142 countries in
all. A min–max band over it spans the full range of life expectancy in each
year, but only the two extreme countries set its edges. Naming two quantiles
instead gives a band that contains most of the countries. Layer two bands of
different widths to draw a **fan chart**:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + x(year) + y(life) +
  ribbon * range(0.1, 0.9) + style(color = "steelblue", opacity = 0.2) +
  ribbon * range(0.25, 0.75) + style(color = "steelblue", opacity = 0.35) +
  line * median + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Life expectancy across 142 countries, 1952 to 2007")
```

*"Given all the gapminder years: x is year, y is life, a ribbon derived by range
from 0.1 to 0.9, and also a ribbon derived by range from 0.25 to 0.75, and also
a line derived by median."*

The pale band contains the middle 80% of countries. The darker one contains the
middle half, and the line is the median. This plot shows something the min–max
version cannot. The median rose from 45 years to 72, and the 90th
percentile from 67 to 80, so the distance between them fell from 22 years to 8.

The lower edges of the two bands do not rise steadily. Both lower edges, the
10th percentile and the 25th, decline between 1992 and 2002 while the median
rises throughout. A min–max band cannot show this, because its lower edge is one
country.

Each band is one more layer, so a third pair of quantiles adds a third band.
Write the widest band first, because a later layer is drawn over an earlier
one.

## Why a ribbon needs a transform

A low *and* a high are two numbers, and a plain `y()` supplies one. An `area`
closes on zero, which the grammar supplies. A ribbon has no such second number.
Like `interval`, a ribbon needs a transform that supplies both boundaries, and
that transform is part of its minimum syllable. Write a ribbon without one and
the engine refuses it, rather than drawing an empty panel. [What it
refuses](#what-it-refuses) shows the message.

Computing the boundaries in a transform also keeps the [channel
vocabulary](../grammar.qmd#channels-a-column-made-visible) small. The
alternative would add `ymin` and `ymax` as two more channels, useful to only a
few marks. A box plot would then need five such channels, one for each number
it summarizes. With a transform instead, the same low and high values can be
drawn as a whisker, as two lines, or as a filled band. See
[Transforms](../transforms.qmd) for the `range`, `confidence`, `deviation` and
`smooth_band` statistics.

Usually that transform is one of the five pair transforms: `range`,
`confidence`, `deviation`, `smooth_band` and `bounds`. `density` over a category
also works. It gives one curve per category, and the ribbon draws that curve and
its **mirror image** as the two boundaries. That is [the
violin](#the-violin-a-distribution-per-category), later in this chapter.

## A confidence band

How well is the mean life expectancy known in each year? That is a different
question from how far the five countries spread. `range` shows the full
spread. For the **uncertainty of the mean**, use
`confidence`. It computes a t-interval for the mean at each `x`, and the ribbon
fills between its two boundaries. Here is the 95% band for the same five
countries:

```{r}
data(gapminder_asia) + ribbon * confidence(0.95) + x(year) + y(life) +
  y_label("Life expectancy") +
  title("95% confidence band for mean life expectancy")
```

*"Given gapminder Asia: a ribbon derived by confidence at 95%, x is year, y is
life."*

The band is wide because each mean comes from only five countries. Those five
differ by more than fifteen years, so the interval around their mean is wide. A
confidence level above 0.95 draws a wider band still.

Compare it with the first plot in this chapter: the two bands are almost the
same width. That comes from having only five values, not from the two
statistics agreeing. Over the 142 countries of `gm_all`, the 95% band in 2007 is
about 4 years wide, and their range is 43 years. What the band means is decided
by the statistic you name: `range` draws the observed spread, and `confidence`
draws the uncertainty of the mean. The mark stays the same in both plots,
exactly as it does for [`interval`](interval.qmd).

## A pre-computed band

`range`, `confidence`, `deviation` and `smooth_band` compute the band from the
raw values in the table, at the moment gog draws the plot. Many bands are
computed earlier, by the analysis itself. A model's standard error, a standard
error of measurement, a bootstrap interval: each of these is computed *before
the plot*. Each gives a low value and a high one, and those two become two
columns in the table. gog draws what you computed, and it does not re-fit the
model. The `score_band` table holds one row per raw score on a test scored from
0 to 20. Each row also holds the **expected true score** for that raw score, the
score a model predicts, and the two ends of a 95% band around it. In this table
the expected true score equals the raw score, so a line through it is a
diagonal:

```{r}
#| echo: false
peek(score_band)
```

The ribbon reads those edge columns directly, through `bounds(lower, upper)`:

```{r}
data(score_band) + ribbon * bounds(lower, upper) + x(score) +
  y_label("Expected true score") + x_label("Raw score") +
  title("A pre-computed measurement band")
```

*"Given the score band: a ribbon from lower to upper, x is score."*

`bounds` is the pair transform that computes nothing. It reshapes the two
columns you give it into the same low and high pair `range` produces. The band
needs no `y()`, because the two columns *are* the boundaries. This is the one
case `range`, `confidence`, `deviation` and `smooth_band` cannot produce: a band
that was computed outside the plot, before gog ever read the table.

A band like this is usually drawn together with the values it surrounds. Here
those values are the expected true scores. A dashed [`line`](line.qmd) draws
them, with a point at every raw score. The band, the line and the points are
three layers composed with `+`:

```{r}
data(score_band) + x(score) + y(expected) +
  ribbon * bounds(lower, upper) + style(color = "steelblue", opacity = 0.18) +
  line + style(color = "steelblue", size = 1.4, pattern = "dashed") +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("Expected score with its measurement error band")
```

*"Given the score band: x is score, y is expected, a ribbon from lower to upper,
and also a line and points."*

The dash on that line comes from [`style(pattern = )`](../style.qmd). That
`style()` is written after `line`, so it belongs to that layer, and the band is
not dashed. On a stroke the texture is a **dash**. The band takes the same
setting, with different values. On a fill the
texture is a **hatch**: `"hatch"`, `"crosshatch"`, `"stripes"`, `"grid"` or
`"dots"`, never `"dashed"`.

## Filled, or two dashed edges

Some plots need the band without a fill. A band starts as a low and high pair,
and the *mark* decides how that pair is drawn. This follows the [No
Exceptions](../design-laws.qmd#law-2-no-exceptions) law. One `bin`
draws as a bar chart, as a line and as a step histogram, and one pair draws in
just as many ways: filled, as two strokes, or as two staircases. `ribbon`
**fills** the pair. A [`line`](line.qmd) traces its **two boundaries** instead,
which is the unfilled band many journals prefer.
With `style(pattern = )` it becomes the dashed pair of curves those journals
often print:

```{r}
data(score_band) + x(score) + y(expected) +
  line * bounds(lower, upper) + style(color = "gray", pattern = "dashed", size = 1.2) +
  line + style(color = "steelblue", size = 1.5) +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("The same band, unfilled: two dashed edges")
```

*"Given the score band: x is score, y is expected, lines from lower to upper,
and also a line and points."*

`line * bounds` is one layer that draws *both* edges, splitting the pair into a
low curve and a high one. So you do not write one `y()` for the low edge and
another for the high edge. Two plain `line` layers can draw two columns you
already have. They cannot split a pair that `range` computes, because that pair
never becomes a column you can name. A `line` draws every pair transform the
same way. `line * range` traces the lowest and the highest values, and `line *
confidence` traces the two ends of a confidence interval.
[`step`](step.qmd) `* bounds` draws the same two edges as staircases. Filled or
unfilled, solid or dashed, it is one pair and a choice of mark, never a new
option added to the band.

Four marks draw a low and high pair: [`interval`](interval.qmd) draws it as a
whisker at each x, `line` and `step` trace its two boundaries, and `ribbon`
fills between them. All four marks draw `range`, `confidence`, `deviation`,
`smooth_band` and `bounds`.

## One band per group

How does the spread of one continent compare with another's, year by year? The
`gm_continents` table holds three continents. One band per continent puts all
three spreads in one panel. A categorical column *splits* a ribbon into one band
per group. It is the same split that gives [`area`](area.qmd) one region per
group and `line` one stroke per group. Here each continent gets its own band,
the range of its countries' life expectancy over time:

```{r}
#| echo: false
peek(gm_continents)
```

```{r}
data(gm_continents) + ribbon * range + x(year) + y(life) + color(continent) +
  y_label("Life expectancy") +
  title("Life expectancy spread, three continents")
```

*"Given the gapminder continents: ribbons derived by range, x is year, y is
life, color by continent."*

Europe's band narrows from about 29 years wide in 1952 to about 10 in 2007, and
the Americas' from 31 to 20. Asia's stays about 37 to 40 years wide, except for
a dip in its lower edge in 1977.

The bands overlap where two continents cover the same range of life expectancy.
A split `area` draws each region nearly opaque, as it draws a single one, so
little of the region behind shows through. For an area, `stack` solves this: it
starts each region at the top of the ones below, so no two regions overlap. A
ribbon split into groups draws each band translucent by default instead, so a
band behind another stays readable. The difference follows from what each mark
measures. An area measures a height from the baseline, so its groups can be
piled into a total. A ribbon already spans a low to a high and measures no such
height, so there is nothing to stack. Overlapping bands are read through
transparency and never stacked, which is why `ribbon * stack` is refused and the
refusal names `style(opacity = )`.

## One region, one fill

A ribbon is a single filled region, so `opacity` is a *setting* here, not a
channel, exactly as on `area`. Its two boundaries pass through the low and high
values at each x, and its inside belongs to no single row. So a column cannot
make one part of the band lighter or darker than another. [What it
refuses](#what-it-refuses) shows the refusal.

`size` is refused both ways: a ribbon neither maps it nor sets it. A band's
shape comes entirely from its x values and its two boundaries, so there is no
width to choose. To outline the band, layer a `line` along its boundaries rather
than setting an outline on the mark. The plot in [What you can
set](#what-you-can-set) does this with `line * bounds`.

## A category on the domain

What happens when `x` holds categories instead of years? The spread of each
continent and its center can be drawn as one band across the categories. You
then see one shape across the panel, as with an area's filled profile. A ribbon
takes a category on `x`, the same way an
[`area`](area.qmd) does, and fills the spread across the categories in axis
order:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + ribbon * range + line * mean + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * range + line * mean: spread and center, per continent")
```

The band runs from each continent's lowest life expectancy to its highest, and
the line marks each continent's mean.

For most such data an [`interval`](interval.qmd) is the better sentence. It
draws a separate whisker at each category and claims nothing about the space
between them. A ribbon instead slopes from one category to the next, across
positions where no rows exist. The grammar does not decide that for you: both
sentences are well formed, and which one fits *your* data is a question about
the data.

```{r}
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("interval * range: the same numbers, claiming less")
```

*"Given gapminder 2007: intervals derived by range, x is continent, y is life."*

Each whisker spans the same lowest and highest values as the band, and nothing
is drawn between the continents.

## The violin: a distribution per category

A `range` on a category gives two numbers per group, the lowest and the highest.
`density` gives a whole *shape* per group, and a ribbon can draw that shape.
The curve runs along `y`, the measure. Its height at each life expectancy is
drawn as a width across the continent's slot, on both sides, so the two halves
are the same curve facing opposite ways. That is the violin plot
[@hintze1998], and it is the band across the continents drawn above, with
`density` in place of `range` and without the line:

```{r}
data(gapminder_2007) + ribbon * density + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * density: one distribution per continent")
```

*"Given gapminder 2007: ribbons derived by density, x is continent, y is life."*

`density` is not the ribbon's own transform. Five other marks draw the same
estimated curve, and only the shape changes. That is the [No
Exceptions](../design-laws.qmd#law-2-no-exceptions) law: a mark draws whatever
rows a transform produces, in the mark's own way.

| Mark derived by `density` | Shape |
|---|---|
| `line * density` | the curve as a stroke |
| `step * density` | the curve as a staircase |
| `area * density` | the curve filled to the baseline |
| `point * density` | one dot at each point the curve was estimated |
| `bar * density` | one thin bar at each of those points |
| `ribbon * density` | the curve mirrored, the violin above |

`line`, `step` and `area` also read a category on one axis, exactly as the
violin does. `point` and `bar` need a continuous axis instead: one dot or one
thin bar per estimated point cannot be read inside a category's slot, so gog
refuses both there.

Nothing new was added to the grammar to draw the violin. A ribbon closes on a
second boundary the data supplies, and a violin closes on the mirror image of
its own curve. `density` estimates a distribution, and this plot computes one
estimate for each group. There is no `violin` word to learn. Such a word would
have to say four things at once: the mark, the statistic, the mirror and the
slot. The sentence already says all four: `ribbon` is the mark and supplies the
mirror, `density` is the statistic, and the category on `x` gives the slot.

Compare it with the [`box`](box.qmd) drawn from the same two columns:

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  y_label("Life expectancy") + title("box + x + y: the same distributions, summarized")
```

*"Given gapminder 2007: boxes, x is continent, y is life."*

A box gives you five numbers you can name. The violin gives you the shape those
five numbers summarize. Africa's violin is widest near 52 years, and it has a
second, smaller bulge near 73. Africa's median is 53, and none of the box's five
numbers shows that second cluster.

### What the width means

Every group's density integrates to 1 on its own. If that were the whole
estimate, the widths would say nothing about how many rows each group has. A
continent of two countries would be drawn as large as a continent of fifty-two.
So by default the estimate is weighted by the group's row count, and width means
the same thing across the whole panel. Oceania holds two countries, so it is the
narrowest of the five violins.

The weighting is on the shape's **area**, the space it covers, not on its width,
which is why Europe's violin is wider than Africa's. Europe's thirty countries
differ in life expectancy by no more than ten years, so Europe's area is packed
into a short stretch of the axis and the shape has to be wide. Africa's
fifty-two countries cover thirty-seven years of life expectancy, so its larger
area is spread across a much longer stretch of the axis. What the width reports
at any height is countries per year of life expectancy, and that reading is the
same in every slot.

`compare = "shape"` removes that weighting and draws every violin to the same
area. It is the right choice when your question is about the shapes alone.
It is the wrong one when a group whose values are packed together is drawn so
wide that every other violin is drawn thin:

```{r}
data(gapminder_2007) + ribbon * density(compare = "shape") + x(continent) + y(life) +
  y_label("Life expectancy") +
  title('density(compare = "shape"): every violin the same area')
```

*"Given gapminder 2007: ribbons derived by density compared by shape, x is
continent, y is life."*

Oceania is now the widest violin and Africa the thinnest, so two countries are
drawn wider than fifty-two. Oceania's two values are one year apart, so its
equal share of area has to fit into a very short stretch of the axis. To hold
that area in so little height, the shape has to be very wide. The widest shape
in a panel sets the scale, so every other violin is drawn in proportion to it.
Each plot answers a different question, so gog draws the weighted width unless
you write `compare = "shape"`.

### Sideways, and halved

A band cannot turn, because its `x` is always the domain. A violin can:
`density` makes the category a slot, and gog reads the orientation from the
bindings, as it does for a [`bar`](bar.qmd) or a [`box`](box.qmd). A category on
`y` is the form with room for long category names:

```{r}
data(gapminder_2007) + ribbon * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("The same violins, turned sideways")
```

An [`area`](area.qmd) draws the same estimate closed on the slot's center line
instead of on its mirror image. So `area` draws one half of the shape, and
`ribbon` draws both halves:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  y_label("Life expectancy") + title("area * density: the half violin")
```

*"Given gapminder 2007: areas derived by density, x is continent, y is life."*

There are two choices here, and they are made independently: which mark closes
the region, and which axis carries the category. That gives four plots. The one
this chapter has not drawn yet, an `area` with the category on `y`, has a name
of its own. Halve the violin *and* turn it sideways. Each category then gets a
filled curve rising from its slot's center line, above one shared measure axis.
That is the **ridgeline plot**:

```{r}
data(gapminder_2007) + area * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("area * density + x(life): the ridgeline")
```

Nothing was added for it either. It is the half violin's sentence with `x` and
`y` exchanged.

Read the four plots together and you can see what
[orthogonality](../design-laws.qmd#law-1-orthogonality) gives you here. One
idea draws all four plots: `density` estimated per category, filled on one side
of the slot's center line or on both, with the category on either axis. A word
for the violin and a word for the ridgeline would name two of the four. The
other two would have no word, and neither would the idea behind all four.

### Letting the ridges overlap

The ridgeline above keeps every ridge inside its own slot. That comes from how
far a shape reaches by default. The widest shape in a panel reaches `0.4` of a
slot from the slot's center line, and every other shape is drawn in proportion
to it. A slot ends `0.5` from that line, so each shape stays inside its own
slot, with empty space before the next category. [Bars](bar.qmd) follow the same
rule, for the same reason: the gap shows that the categories are separate
groups, not pieces of one continuous scale.

A ridgeline, as it is usually drawn, breaks that rule on purpose: its ridges are
tall enough to reach into the slot above, and that overlap is part of the
design. `density(reach = )` says how far, *in slots*:

```{r}
data(gapminder_2007) + area * density(reach = 2.5) + x(life) + y(continent) +
  x_label("Life expectancy") + title("reach = 2.5: each ridge two and a half slots tall")
```

*"Given gapminder 2007: areas derived by density reaching 2.5 slots, x is life,
y is continent."*

The number is measured from the slot's center line to the shape's furthest
point. That is why the number means the same thing to both marks. A `ribbon`
reaches that far on each side, an `area` on one side only. The half violin
therefore stays exactly half of the violin, at any reach. Past `0.5` the shapes
leave their own slots. gog does not refuse this, because a reach above `0.5` is
what your sentence asks for. Where two ridges overlap, the lower one is drawn
later and covers the one above it.

The category axis grows to hold the ridges. Normally the axis leaves half a slot
of space beyond the first and last categories, which is enough for a
[`bar`](bar.qmd) or a [`box`](box.qmd), since each stays inside its slot. A
ridge 2.5 slots tall does not stay in its slot, so the axis adds the room it
needs. Without that room, the panel edge would hide the top of the highest
ridge.

The edge is a layer here, as it is for every filled band in this chapter.
Neither `area` nor `ribbon` takes a border setting. A [`line`](line.qmd) given
the same transform traces the outline of the fill. That
is the [Filled, or two dashed edges](#filled-or-two-dashed-edges) rule again,
this time inside a category's slot:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.5) + style(color = "steelblue", opacity = 0.9) +
  line * density(reach = 2.5) + style(color = "black", size = 1) +
  x_label("Life expectancy") + title("The same ridges, with their edges drawn")
```

Both layers read the same `density(reach = 2.5)`, so the stroke lies exactly on
the outline of the fill. That is why no border setting is needed: a layer draws
the outline, with its own color, width and dash. Color each ridge by its own
category and the plot is complete:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.2) + color(continent) +
  line * density(reach = 2.2) + style(color = "black", size = 0.9) +
  x_label("Life expectancy") + title("One hue per ridge")
```

The `color(continent)` is written *after* the area, so it belongs to that layer
alone. Written before both marks it would color the `line` too. A layer cannot
both map `color` and set it, so the sentence would be refused. Where a channel
sits in the sentence is what decides its scope, and [Encoding
scope](../encoding-scope.qmd#channels-plot-scoped-or-layer-scoped-by-position)
explains that in full.

## What you can set

A setting changes how the band looks without reading a column, and each mark
takes its own. These are a ribbon's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("ribbon")
```

A ribbon takes `area`'s list, and for `area`'s reasons. There is no `size`,
because the band's shape comes from the two boundaries its transform supplies.
There is no border, because the edge is a `line` traced along the same pair.

One setting this chapter has not shown on a band is a fill `pattern`. Here is a
band hatched with a grid, with a dashed `line` along both of its boundaries:

```{r}
data(score_band) + x(score) +
  ribbon * bounds(lower, upper) +
    style(color = "slateblue", pattern = "grid", opacity = 0.45) +
  line * bounds(lower, upper) + style(color = "slateblue", pattern = "dashed") +
  y_label("Expected true score") + title("A textured band, and its two edges")
```

*"Given the score band: x is score, a ribbon from lower to upper, with pattern
grid, and also lines from lower to upper, with pattern dashed."*

Both marks read the same `bounds(lower, upper)` there, exactly as the two ridge
layers did, so the edges stay on the band. The two `pattern` values follow the
rule from [A pre-computed band](#a-pre-computed-band): a hatch on the fill, a
dash on the stroke.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The first refusal concerns the measure, which must be continuous. A band from
`range` spans a low value to a high one, and both are quantities, so a category
on `y` is refused. The violins above
put a category on `y`, but there `density` makes it a slot rather than the
measure:

```{r}
#| error: true
data(gapminder_2007) + ribbon * range + x(life) + y(continent)
```

A bare `ribbon` is the second refusal. A plain `y()` supplies one number, and a
band needs two boundaries, as [Why a ribbon needs a
transform](#why-a-ribbon-needs-a-transform) explained. The engine says so
rather than drawing an empty panel:

```{r}
#| error: true
data(gapminder_asia) + ribbon + x(year) + y(life)
```

The message calls them range transforms. It names two of the four, `range` and
`bounds`, and writes out a sentence the engine would accept.

A mapped `opacity` is the third. The inside of the band belongs to no single
row, as [One region, one fill](#one-region-one-fill) explained. Opacity is set
with `style(opacity = )` and never mapped:

```{r}
#| error: true
data(gapminder_asia) + ribbon * range + x(year) + y(life) + opacity(gdp)
```

## In short

A ribbon fills the space between two boundaries, and refuses to draw until a
transform supplies them. The transform decides what the band means: the
extremes from `range`, the middle from two quantiles, the mean's uncertainty
from `confidence`, and a band computed elsewhere from `bounds`. Over a category,
`density` makes the ribbon a violin. Both edges come from the data, so the
band's shape is never a setting.

## What comes next

A ribbon shows a region and names nothing inside it. [Text](text.qmd) draws a
string at each position, read from a column.
