# Ribbon

```{r}
#| include: false
source("../R/setup.R")
```

How wide is the band the values fall in, across a span? `ribbon` draws a
filled band from a low boundary to a high one across x: a confidence band,
a spread band, the shaded region around a trend. It is [`area`](area.qmd)'s
single filled region reading [`interval`](interval.qmd)'s low and high pair:
closed on a *second data boundary* instead of a baseline at zero. Where an
area fills from the data down to zero, a ribbon fills between two boundaries
the data supplies.

## A band from a range

The two boundaries come from a transform, exactly as an interval's two extents
do. `range` reduces `y` to its minimum and maximum within each `x`, and the
ribbon fills between them. Here, the spread of life expectancy across five Asian
countries, year by year:

```{r}
data(gapminder_asia) + ribbon * range + x(year) + y(life) +
  y_label("Life expectancy") +
  title("The min–max spread across five Asian countries, per year")
```

*"Given gapminder Asia: a ribbon derived by range, x is year, y is life."*

Every year, the band's lower edge is the shortest-lived country and its upper
edge the longest-lived; the band is the region that contains all five. Read it as
`interval * range` drawn as a *continuous* fill rather than a whisker at each
x. The pair is the same; only the geometry differs.

## The mean and its band

On its own a band shows spread but not center. A band is usually drawn with a
line through the middle, and layering, not a new option on either mark,
is how you compose them. A `ribbon * range` for the spread, a `line * mean` for
the trend, sharing one `x` and `y`:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  ribbon * range + style(color = "steelblue", opacity = 0.25) +
  line * mean + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Mean life expectancy, with its min–max band")
```

Each layer runs its own transform on the shared data: `mean` collapses each year
to one point for the line, `range` to a low/high pair for the band. This is the
most common ribbon plot, a trend with the uncertainty or spread drawn around it,
and it is two marks the grammar already has, added with `+`.

## A fan chart of nested bands

A min–max band shows what every country did. With 142 countries, only the two
extreme countries set its edges. Naming two quantiles instead gives a band that
contains most of the countries. `gm_all` is the whole gapminder table, every
country in every year, so layer two bands of different widths over it and you
have a **fan chart**:

```{r}
data(gm_all) + x(year) + y(life) +
  ribbon * range(0.1, 0.9) + style(color = "steelblue", opacity = 0.2) +
  ribbon * range(0.25, 0.75) + style(color = "steelblue", opacity = 0.35) +
  line * median + style(color = "steelblue", size = 2) +
  y_label("Life expectancy") +
  title("Life expectancy across 142 countries, 1952 to 2007")
```

The pale band contains the middle 80% of countries. The darker one contains
the middle half, and the line is the median. The plot then says something the min–max
version cannot. The median rose from 45 years to 72, and the 90th percentile from
67 to 80, so the distance between them fell from 22 years to 8.

The bottom of the plot moves the other way. Both lower edges, the 10th
percentile and the 25th, decline between 1992 and 2002 while the median rises
throughout.
A min–max band cannot show this, because its lower edge is one country.

Each band is one more layer, so a third pair adds a third. The bands must be
drawn widest first, because a later layer is painted over an earlier one.

## Why a ribbon needs a transform

A low *and* a high are two numbers, and a plain `y()` supplies one. An `area`
closes on zero, which the grammar supplies. A ribbon has no such second
number, so like `interval` its minimum syllable includes a range-producing
transform. Ask for
one without it and the engine says so rather than drawing an empty panel:

```{r}
#| error: true
data(gapminder_asia) + ribbon + x(year) + y(life)
```

This is the same design choice that keeps the [channel
vocabulary](../grammar.qmd) small. The alternative would add `ymin` and
`ymax` as two more channels meaningful to a single mark. Instead the boundaries are
**invented by a transform**, exactly as `bin` invents a count, so the same
low/high-rows mechanism feeds the whisker, the box, and the band alike; see
[Transforms](../transforms.qmd) for the `range`, `confidence` and `deviation`
statistics.

Read the requirement as *where do the two boundaries come from?* rather than
as a list of transforms. One answer looks nothing like the others: `density`
over a category gives a ribbon its boundaries by
**reflection**. That is [the violin](#the-violin-a-distribution-per-category),
at the end of this chapter.

## A confidence band

`range` shows the full spread. For the **uncertainty of the mean**, the shaded
band of a regression or a group mean, use `confidence`, which computes the
mean's t-interval per `x` (wider when the group is small) and fills between
its bounds:

```{r}
data(gapminder_asia) + ribbon * confidence(0.95) + x(year) + y(life) +
  y_label("Life expectancy") +
  title("95% confidence band for mean life expectancy")
```

The band is narrow because the five countries agree closely on the mean in each
year; a higher level draws a wider band. Which statistic you ask for chooses the
band's meaning, the observed spread (`range`) or the uncertainty of the mean
(`confidence`), with no new mark to learn, the same rule the
[`interval`](interval.qmd) family follows.

## A pre-computed band

`range`, `confidence` and `deviation` compute the band *from raw values in the
plot*. But most confidence bands are not computed in the plot at all. A
model's standard error, a standard error of measurement, a bootstrap
interval: each is computed **before the plot**, and its bounds arrive as two
columns. gog draws what
you computed; it does not re-fit the model. So the ribbon reads those columns
directly, through `bounds(lower, upper)`:

```{r}
data(score_band) + ribbon * bounds(lower, upper) + x(score) +
  y_label("Expected true score") + x_label("Raw score") +
  title("A pre-computed measurement band")
```

`bounds` is `range`'s counterpart that **computes nothing**: it reshapes the two
columns you hand it into the same low/high pair `range` produces, so the band
draws with nothing new and no `y()` (the two columns *are* the boundaries).
It is the common case the computing transforms cannot serve: a band whose
bounds were computed before the data reached the plot.

It is usually drawn with the curve the band surrounds. Here the scoring model's
**expected true score** runs through the middle, drawn as a dashed
[`line`](line.qmd) with a marker at each integer raw score (the band, the trend,
and the score points, composed with `+`):

```{r}
data(score_band) + x(score) + y(expected) +
  ribbon * bounds(lower, upper) + style(color = "steelblue", opacity = 0.18) +
  line + style(color = "steelblue", size = 1.4, pattern = "dashed") +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("Expected score with its measurement error band")
```

The dashed line is [`style(pattern = )`](../style.qmd): on a stroke the texture is a
**dash**, so it applies to the `line` here, not to the band. The band can take the same
setting too: there it is a **hatch** (a fill texture), not a dash.

## Filled, or two dashed edges

A band is a low/high pair, and the *mark* chooses how to draw it: the same
[No Exceptions](../design-laws.qmd) rule that gives one `bin` a bar, a line, and a
step histogram. `ribbon` **fills** the pair; a [`line`](line.qmd) traces its **two
boundaries**, the unfilled band publications often prefer, and with
`style(pattern = )`, the dashed pair of curves publications often draw:

```{r}
data(score_band) + x(score) + y(expected) +
  line * bounds(lower, upper) + style(color = "gray", pattern = "dashed", size = 1.2) +
  line + style(color = "steelblue", size = 1.5) +
  point + style(color = "steelblue", size = 2.4) +
  y_label("Expected true score") + x_label("Raw score") +
  title("The same band, unfilled: two dashed edges")
```

`line * bounds` is one layer that draws *both* edges, splitting the pair into
a low curve and a high one. It therefore needs no second `y`, and that is why
the two boundaries are drawn as separate curves: two hand-written `line`
layers cannot do it, because each takes only one `y`. `line * bounds` reads
any pair the same way: `line * range` traces a min–max envelope,
`line * confidence` the CI edges, and [`step`](step.qmd) `* bounds` draws them as
staircases (stepped control limits). Filled or unfilled, solid or dashed, it is one
pair and a choice of mark, never a new option added to the band.

## One band per group

A categorical column *splits* a ribbon into one band per group, the same split
[`area`](area.qmd) and `line` make. Here each continent gets its own band, the
range of its countries' life expectancy over time:

```{r}
data(gm_continents) + ribbon * range + x(year) + y(life) + color(continent) +
  y_label("Life expectancy") +
  title("Life expectancy spread, three continents")
```

The bands overlap where the continents do, and unlike a split `area`, which
stays opaque and leaves `stack` as the answer, a ribbon draws its split
**translucent** by default, so a band behind another still shows through. That difference is
principled: an area measures a height from the baseline, so its groups can be
piled into a total; a ribbon already spans a low to a high and measures no such
height, so there is nothing to stack. Overlapping bands are read through
transparency, never stacked, which is why `ribbon * stack` is refused and the
refusal names `style(opacity = )`.

## One region, one fill

A ribbon is a single filled region, so, exactly as on `area`, `opacity` is a
*setting*, not a channel: a row is a vertex of a boundary, not a region of its
own, so there is nothing for a per-row opacity to vary along.

```{r}
#| error: true
render_svg(data(gapminder_asia) + ribbon * range + x(year) + y(life) + opacity(gdp))
```

And `size` cannot even be set: a band's extent comes from its x and its two
boundaries, so there is no width to choose. Its edge, if you want one, is a `line`
layered along a boundary (`ribbon + line`, the composition the previous section
already drew), never an outline option on the mark.

## A category on the domain

A ribbon takes a category on `x`, the same way an [`area`](area.qmd) does, and
fills the spread across the categories in axis order:

```{r}
data(gapminder_2007) + ribbon * range + line * mean + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * range + line * mean: spread and center, per continent")
```

For most such data an [`interval`](interval.qmd) is the better sentence. It draws
a separate whisker at each category and claims nothing about the space
between them, where the band's edges slope across x values that have no rows.
The grammar does not decide that for you: both sentences are well formed, and
which one fits *your* data is a question about the data.

```{r}
data(gapminder_2007) + interval * range + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("interval * range: the same numbers, claiming less")
```

## The violin: a distribution per category

A `range` on a category gives two numbers per group, the lowest and the highest.
`density` gives a whole *shape* per group, and a ribbon has somewhere to put it:
the estimate spreads along the measure and is drawn across the category's slot,
mirrored. That is the violin plot [@hintze1998], and it is this chapter's
sentence with one word changed:

```{r}
data(gapminder_2007) + ribbon * density + x(continent) + y(life) +
  y_label("Life expectancy") +
  title("ribbon * density: one distribution per continent")
```

Nothing new was added to the grammar to draw it. A ribbon is the mark that closes
on a second data boundary, and a violin closes on its own reflection, which is
one; `density` is the transform that estimates a distribution, and this is that
estimate read per group. There is no `violin` word to learn, because a word for
it would be four things at once (the mark, the statistic, the mirror, and the
slot), and each of those is already said by something you know.

Read it against the [`box`](box.qmd) of the same two columns. A box gives you
five numbers you can name; the violin gives you the shape those five numbers
summarize. Africa's violin has two peaks, which its median hides.

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  y_label("Life expectancy") + title("box + x + y: the same distributions, summarized")
```

### What the width means

Every group's density integrates to 1 on its own, so without the weighting
the widths would say nothing about how much data each group has: a continent of two countries
would be drawn as boldly as a continent of fifty-two. So by default the estimate
is weighted by the group's row count, and width means the same thing across the
whole panel. Oceania above is two countries, and it is drawn as two countries.

`compare = "shape"` takes that weighting back off and draws every violin to the
same area. It is the right reading when the question is about the shapes and the
groups are of unequal size on purpose, and the wrong one when a small group's
spike sets the width for the whole panel:

```{r}
data(gapminder_2007) + ribbon * density(compare = "shape") + x(continent) + y(life) +
  y_label("Life expectancy") +
  title('density(compare = "shape"): every violin the same area')
```

Two countries now fill as much space as fifty-two. Two nearby values make a
tall narrow estimate, so Oceania sets the width and every other violin is
drawn thinner. Both plots are honest about something different, which is why the
choice is a word you write rather than a default nobody stated.

### Sideways, and halved

The category can go on either axis, and the violins turn to follow it. The
orientation is read off the bindings, exactly as a [`bar`](bar.qmd)'s or a
[`box`](box.qmd)'s is. It is the form with room for long category names:

```{r}
data(gapminder_2007) + ribbon * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("the same violins, lying down")
```

And an [`area`](area.qmd) draws the same estimate closed on the slot's center
line instead of on its own reflection: one side rather than two, since that is
the difference between the two marks everywhere else in the book:

```{r}
data(gapminder_2007) + area * density + x(continent) + y(life) +
  y_label("Life expectancy") + title("area * density: the half violin")
```

Two choices, made independently: which mark closes the region, and which axis
carries the category. That is four plots, and the fourth one has a name of its
own. Halve the violin *and* lay it down, and each category gets a filled curve
on its own line above a shared measure axis, which is the **ridgeline plot**:

```{r}
data(gapminder_2007) + area * density + x(life) + y(continent) +
  x_label("Life expectancy") + title("area * density + x(life): the ridgeline")
```

Nothing was added for it either. It is the previous chunk's sentence with `x`
and `y` exchanged. It arrives because the two choices are genuinely independent:
nothing in the grammar ties *where a region closes* to *which axis holds the
category*. All four combinations are sentences, and the one the world gave a
name to is simply one of them.

Read the four together and the value of
[orthogonality](../design-laws.qmd) is easier to see here than anywhere else in
this chapter. A word for the violin and a word for the ridgeline would be two
names for one idea, and the other two combinations would have no word at all.

### Letting the ridges overlap

Each shape reaches four tenths of a slot by default, so two violins face to
face fill four fifths of the space between their categories, and the
remaining fifth stays empty. That is a [`bar`](bar.qmd)'s own rule, for a bar's reason: the gap is what
says the categories are separate rather than a divided continuum.

The ridgeline as it is usually drawn (the form some call a joyplot) is not
like that. Its ridges are tall enough to reach into the row above, and that
overlap is part of the design. `density(reach = )` says
how far, **in slots**:

```{r}
data(gapminder_2007) + area * density(reach = 2.5) + x(life) + y(continent) +
  x_label("Life expectancy") + title("reach = 2.5: each ridge two and a half slots tall")
```

The number is measured from the line a category sits on to the shape's furthest
point. That is why it means the same thing to both marks: a `ribbon` reaches it
each way and an `area` one way, so the half violin stays exactly half of the
violin at any value. Past `0.5` the shapes leave their own slots. The grammar
does not guard against that. It is the plot being asked for, and the later
category simply draws over the earlier one.

The category axis grows to hold them. Half a slot of margin is right for a
[`bar`](bar.qmd) or a [`box`](box.qmd), which stand *in* their slots. A ridge
2.5 slots tall does not, and without the extra room the frame would cut the top
one off. That would be a mark drawn outside its own panel. No check would
catch it, and a reader would see the cut ridge.

The edge is a layer, as it is for every filled band in this chapter: a `ribbon`
takes no border setting, and neither does the violin: a [`line`](line.qmd)
given the same transform **traces** the boundary the fill stops at. That is
the "filled, or two dashed edges" rule again, now against a slot:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.5) + style(color = "steelblue", opacity = 0.9) +
  line * density(reach = 2.5) + style(color = "black", size = 1) +
  x_label("Life expectancy") + title("The same ridges, with their edges drawn")
```

Both layers read the same `density(reach = 2.5)`, so the stroke cannot drift off
the fill it belongs to, the reason an edge is a layer rather than a setting.
Color each ridge by its own category and the plot is complete:

```{r}
data(gapminder_2007) + x(life) + y(continent) +
  area * density(reach = 2.2) + color(continent) +
  line * density(reach = 2.2) + style(color = "black", size = 0.9) +
  x_label("Life expectancy") + title("One hue per ridge")
```

The `color(continent)` is written *after* the area, so it belongs to that layer
alone: a plot-scoped one would reach the `line` too, and a layer cannot both map
color and set it. [Scope is position](../encoding-scope.qmd), and this is the
sentence where it matters.

`step * density` traces the same estimate as a staircase, for the same reason
`step * bin` is the outline histogram: [No Exceptions](../design-laws.qmd) means
a mark draws whatever a transform hands it, in the mark's own way.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("ribbon")
```

A ribbon takes `area`'s list, and for `area`'s reasons: no `size`, because
the band's extent comes from the two boundaries its transform supplies, and
no border, because the edge is a `line` traced along the same pair.

```{r}
data(score_band) + x(score) +
  ribbon * bounds(lower, upper) +
    style(color = "slateblue", pattern = "grid", opacity = 0.45) +
  line * bounds(lower, upper) + style(color = "slateblue", pattern = "dashed") +
  y_label("Expected true score") + title("A textured band, and its two edges")
```

Both marks read the same `bounds(lower, upper)` there, exactly as in the
ridge section, so the edges stay on the band. Note the two `pattern` values in one
sentence: `"grid"` is a fill texture on the ribbon and `"dashed"` a stroke dash on
the line, the same setting realized once per geometry.

## What it refuses

The measure stays numeric. A band spans from a low value to a high one, and the
two extents are quantities, so a categorical `y` is refused:

```{r}
#| error: true
render_svg(data(gapminder_2007) + ribbon * range + x(life) + y(continent))
```

`ribbon` completes the low/high family: [`interval`](interval.qmd) draws the pair
as a whisker at each x, [`box`](box.qmd) as a five-number summary, and `ribbon` as
a filled band across x. One `range`, three marks: [No
Exceptions](../design-laws.qmd) made visible. See
[Transforms](../transforms.qmd) for the `range`, `confidence` and `deviation`
statistics all three share.
