# Rule

```{r}
#| include: false
source("../R/setup.R")
```

Where is the line you want to compare every value against? `rule` draws a line
at one position. Along the other axis the line crosses the whole panel. Most
marks read two positions. A rule reads one. That gives two familiar plots: a
reference line at a threshold, and a rug of ticks showing where the
observations sit.

In most libraries that is three marks: a vertical reference line, a horizontal
one, and a rug. gog has one. A vertical and a horizontal reference line differ
only in which axis carries the position. The grammar reads that off the
bindings, exactly as it does for a bar.

## A line at a threshold

Which countries have a life expectancy above 70? On a scatter the reader has to
find 70 on the axis and judge each point against it. A line drawn at 70 does
that work once, for every reader.
Here is the 2007 gapminder scatter, income against life expectancy, with
lines drawn at three life-expectancy thresholds:

```{r}
#| echo: false
peek(gapminder_2007)
peek(life_bands)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Three thresholds, one table")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands."*

The sentence names `rule` once, and the plot draws three lines. That is the
mark's most useful property: its position is a **column**, not a number.
`life_bands` holds three values, so it draws three lines. Add a fourth row and a
fourth line appears with no change to the plot's sentence at all.

Every atom that takes a column works this way (`color(band)` names a
column, not a color), and that is why there is no `rule(70)`. A number written
into the plot would be a value the grammar could not scale, facet, or bend into
another coordinate space. A column is data, so it scales, facets and bends with
everything else.

A rule takes no transform, as [What it refuses](#what-it-refuses) shows, so it
cannot compute a reference for you. Which reference you need decides which mark
you write:

| Mark | The reference it draws |
|---|---|
| `rule` | a line at each value in a column you supply |
| `line * mean` | a line at a value gog computes from the data |
| `ribbon * bounds` | a band between a low column and a high one |
| `zone * bounds` | a block across the panel, between the same two |

## Which axis it lands on

Nothing in the read-aloud sentence under that plot said "horizontal". The mark
read it from the columns. `life_bands` has a `life` column and no `gdp` column,
and the plot's `y` is `life`. One axis answers and the other does not. The rule
sits on the axis that answers, and spans the axis that does not.

A threshold on income needs a vertical line, and there is no second mark for it.
Change which column the table holds and the same word draws the line
vertically:

```{r}
#| echo: false
peek(gdp_threshold)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(gdp_threshold) + rule + style(color = "firebrick", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("The same mark, the other axis")
```

`gdp_threshold` holds `gdp`, the plot's `x`, so the line is vertical. The mark
did not change and the sentence did not change. gog has one `rule`, and the
column the table holds chooses the axis. ggplot2 needs two marks for this,
`geom_vline()` and `geom_hline()`, and matplotlib needs `axvline()` and
`axhline()`.

## A rug is the same mark

A scatter shows where each country sits on two axes at once. A second question
asks only how the countries spread along the income axis.
A rug tick has no second position either: it sits at one value and extends a
short fixed distance in from the edge. The geometry is the same, and only the
distance the line crosses is shorter. A shorter distance is a setting rather
than a second mark:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  rule + x(gdp) + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Where the countries actually sit on the income axis")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules, x is gdp,
with reach edge."*

The rule names `x(gdp)` even though the points already named it. This layer has
no table of its own, so it reads the scatter's, which holds both `gdp` and
`life`. Both axes answer, so the rule has to say which one it sits on. The
`life_bands` table holds only `life`, which is why those rules named no
position.

Look along the bottom. Each tick is one country, drawn at its own income.
Together they answer a question the points themselves cannot. How many countries
sit at each income, counting the ones hidden under other points? Points overlap,
so the 66 countries below $5,000 look like far fewer than 66 dots. Each tick has
one coordinate rather than two. Two countries hide each other only when their
incomes are almost equal, so the crowding at the left stays visible.

In general a rug puts the one-dimensional distribution of one column in the
margin of a plot that already shows something else. You read the spread of
`gdp` and still see the scatter, with no second plot whose axis you would have
to match.

Everything else about the mark is unchanged. The only new setting is
`style(reach = "edge")`, and there is no distance to set. A tick's length
comes from the panel, the way a label's nudge distance comes from the font size.
A number there would be a pixel count that meant something different in every
facet and at every plot size.

### The other axis

The rug so far has been along the bottom, on income. Life expectancy has a
distribution of its own, and the overlapping points hide it the same way.
A rug is not only for the horizontal axis. The position you name on the layer
decides which axis it lands on. Write `y(life)` instead of `x(gdp)` and the
ticks run up the left edge. Write both, as two layers, and the scatter gets a
rug along the bottom and another up the left at once:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  rule + x(gdp)  + style(reach = "edge") +
  rule + y(life) + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both margins, from two layers")
```

That takes two layers rather than one. Each rug names one axis, and both read
the scatter's table.

Read the left edge and the distribution looks different. The ticks are dense
above 70 and sparse between 60 and 70, and they spread more widely from 60 down
to 40. The sparse middle is what to notice: of the 142 countries, 83 sit above
70 and 43 below 60, with only 16 between. This is a distribution with two
groups and a gap, not one spread. The bottom rug cannot show it, because it
reads the income axis, and the scatter spreads the same countries over two
axes, which makes the gap harder to see.

## Coloring the lines

Leave the rug and go back to the three threshold lines. All three are drawn
alike, so a reader has to match each line to its value on the axis. Give
each line its own color instead, and the legend prints the band's name
beside that color.
A rule's table draws one line per row, so `color` gives each row its own color.
The `band` column in `life_bands`, which names each threshold, is for this:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgray") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("One line per band, keyed by the legend")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands, color by band."*

A thin line gives a reader no area to compare shades in, so `color` here takes
categories, not a continuous ramp. A rule's color stands for a name, and the
legend prints that name beside it.

## Thickening a rule

A thin line marks a threshold without drawing attention to it. Sometimes one
value is the reason the plot exists. Then the reader should find that line
first, before reading the data around it.
Width is the rule's second visible setting, after color. `style(size = )` sets
it, and a wide, faint rule reads as a highlighted band:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.55) +
  data(life_bands) + rule + style(color = "seagreen", size = 26, opacity = 0.20) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A thick rule: emphasis, measured in pixels")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands, colored seagreen, with size 26 and opacity 0.20."*

That is a good way to mark a value. Remember what the width is measured in:
pixels, not data. The band in that plot covers a little over two years of life
expectancy. It covers that much only because this axis runs from 40 to 80.
Facet it, resize it, or change the data's range and the same `size = 26`
will cover a different number of years while looking identical. Nothing is wrong
with that as long as the width is decoration. It is wrong as soon as the width
is supposed to *mean* something.

### Thickness as data

Sometimes the thickness is the point: "the healthy range is 65 to 75", not
"look here". Then the two edges are data, and you need a mark whose edges are
positions. That is [`ribbon`](ribbon.qmd), given a low column and a high column
you computed. The table has two rows, one at the lowest income and one at the
highest, so the band reaches across the data:

```{r}
#| echo: false
peek(healthy_band)
```

```{r}
data(healthy_band) + ribbon * bounds(lo, hi) + style(color = "seagreen", opacity = 0.18) +
  data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(color = "seagreen", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A band whose edges are 65 and 75, and rules on top")
```

*"Given the healthy band table: a ribbon from lo to hi; then given gapminder 2007:
points, x is gdp, y is life, and also rules from the life bands."*

Now 65 and 75 are positions. They rescale with the axis and keep the same values
in every facet. On a log axis they land where the data puts them, not at a fixed
number of pixels. The highlight is written *first*, and layers are drawn in the
order they are written, so the points land on top of it.

Give that same table two different `gdp` values instead of the data's range,
and the band becomes a box bounded on both axes. That is how you shade a
corner of a scatter, or one span of a time series.

Neither the band nor the box reaches the edge of the panel. A ribbon is bounded
by its data, so it stops exactly at the numbers you gave it. To make it wider
you have to widen the scale. For a rectangle that takes one side from the data
and the other from the panel, use [`zone`](zone.qmd). A zone takes a pair of
values on one axis and spans the other, as a rule takes one value and spans the
other.

## In polar: the ring and the spoke

A wind plot is drawn on a circle, and a target speed on it is a threshold like
any other. In polar the axis the line crosses is bent into a circle. The
question is what shape the line takes then.
The mark does one thing: it spans the whole of the axis it does not name. In
polar that is unchanged, and it draws two shapes rather than one, because a
circle has two axes to span. [Polar](../polar.qmd) builds the circle itself.

`winds` is the table that [The book's data](../book-data.qmd) introduced.
`speed_target` holds one target speed:

```{r}
#| echo: false
peek(winds)
peek(speed_target)
```

The angular axis runs all the way around the circle, and that whole sweep is the
**turn**. A rule that spans the whole turn closes into a ring. Here it marks a
target wind speed on a circular wind plot, and `speed` is the radial axis:

```{r}
data(winds) + line * mean + x(direction) + y(speed) + polar() +
  data(speed_target) + rule + style(color = "firebrick", pattern = "dashed") +
  title("A target speed, ringed")
```

*"Given the winds: a line derived by mean, x is direction, y is speed, in polar,
and also rules from the speed target."*

A rule that spans the whole *radius* draws a spoke, a line from the center to
the rim. Give the rule a table holding `direction` instead of `speed`, and the
line sits at that compass point:

```{r}
data(winds) + line * mean + x(direction) + y(speed) + polar() +
  data(data.frame(direction = "E")) + rule +
    style(color = "firebrick", pattern = "dashed") +
  title("One compass point, center to rim")
```

*"Given the winds: a line derived by mean, x is direction, y is speed, in polar;
then given one row with direction E: a rule colored firebrick, with pattern
dashed."*

Neither shape is a special case in the mark. In a circle the radial axis runs
from the center outward and the angular axis runs all the way around, so one
rule draws two shapes.

`rule` is still one mark in polar because nothing about it changed. It is stated
once, its axis is read off the bindings, and it means the same thing in every
space. A rule that needed a special case in polar would not be one mark. It
would be three: a flat line, a ring, and a spoke.

## What you can set

A threshold line has to look different from the data it crosses, and no column
carries that difference. A setting supplies it instead, and each mark accepts
its own list. This is a rule's list, generated from the engine's own legality
table, so this page cannot differ from what `style()` accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("rule")
```

Four of those are a stroke's ordinary controls, and one is the mark's own.
`style(reach = )` sets how far the rule crosses the axis it does not name. The
value `edge` shortens it to a tick, which is why the rug is this mark and not a
second one. The other four behave here as on `line` or `path`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgray") +
  data(data.frame(life = 50.0)) + rule + style(color = "firebrick", size = 4) +
  data(data.frame(life = 60.0)) + rule + style(color = "seagreen", pattern = "dashed") +
  data(data.frame(life = 70.0)) + rule +
    style(color = "steelblue", pattern = "dotted", size = 2.5) +
  data(data.frame(life = 80.0)) + rule +
    style(color = "black", size = 6, opacity = 0.25) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Bottom to top: size, dashed, dotted, a faded thick rule")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored lightgray; then
given one row with life 50: a rule colored firebrick, with size 4; then given
one row with life 60: a rule colored seagreen, with pattern dashed; then given
one row with life 70: a rule colored steelblue, with pattern dotted and size
2.5; then given one row with life 80: a rule colored black, with size 6 and
opacity 0.25."*

The `pattern` values are the five **dashes**, because a rule is a stroke.
`style(pattern = "hatch")` is refused, and the message lists the dashes
instead. A hatch is a *fill*'s texture, and it is what [`zone`](zone.qmd) takes.
The setting has one name, and each mark draws it the way its shape allows. A
stroke takes a dash, and a fill takes a hatch.

What a rule refuses follows from the same geometry. There is no `shape`,
because a stroke has no glyph to choose, and there is no border. [What it
refuses](#what-it-refuses) shows the border refused.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list.
The [companion grid](../combinations.qmd#marks-and-channels) has a row per mark,
listing what it *maps* rather than sets.

## What it refuses

The first sentence a reader tries is a bare `rule` on the plot's own table. That
table holds both position columns, so both `x` and `y` find a value in it.
Nothing in the sentence says which axis is meant, so gog refuses rather than
choosing one:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + rule
```

That refusal is why the two rugs in this chapter name their axis,
`rule + x(gdp)` and `rule + y(life)`. A threshold with a table of its own needs
no position. That table holds only one of the two columns, so the axis is read
off it. A bar with categories on both axes is refused for a related reason.
Neither axis holds a number, so the bar has nothing to measure and no length to
draw, and gog will not guess.

A rule takes no transform either. One of its two axes is the whole panel, so
there is no measured axis for a statistic to compute on and nowhere to put the
answer:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule * mean
```

The message restates what a rule is. A line at the mean is a mean you computed
yourself, supplied as a column. A rule's position always comes from a column.

A rule has no border either. A border belongs to a filled shape, and a rule is
the outline already:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(border_color = "black")
```

So `style(color = )` is a rule's color and `style(size = )` its width. That is
the same answer `line`, `step` and `path` give, because it is the same
geometry.

## In short

A rule's position is a column and never a number, so three thresholds are three
rows. Bind `x` and the line is vertical. Bind `y` and it is horizontal. Over the
plot's own table both axes answer, and that sentence is refused. It takes no
transform: a line at a computed value is `line * mean`, or a rule over a value
you computed yourself.

## What comes next

A rule marks one value, and its width is pixels rather than data.
[Zone](zone.qmd) is the region between two values, a block of the panel rather
than a line across it.
