# Rule

```{r}
#| include: false
source("../R/setup.R")
```

Where is the threshold your values should be compared against? `rule` draws a
line at one position and spans the whole of the other axis. Every other mark
reads two positions; a rule reads one and takes the rest from the panel. That
gives two familiar plots: a reference line at a threshold, and a rug of ticks
showing where the observations sit.

Those are often three separate marks: a vertical reference line, a horizontal
one, and a rug. gog has one, because a vertical
and a horizontal reference line differ only in **which axis
carries the position**, and the grammar already reads that off the bindings,
the same way it decides whether a bar stands up or lies down.

## A line at a threshold

Here is the 2007 gapminder scatter, income against life expectancy, with
lines drawn at three life-expectancy thresholds:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Three thresholds, one table")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the life bands."*

Three lines, and only one `rule` in the sentence. That is the mark's most
useful property: **its position is a column**, not a number. `life_bands` holds three
values, so it draws three lines. Add a fourth row and a fourth line appears with
no change to the plot's sentence at all.

Every atom that takes a column works this way (`color(species)` names a
column, not a color), and that is why there is no `rule(70)`. A number
written into the plot would be a value the grammar could not scale, facet, or
bend into another coordinate space; a column is data, so it scales, facets
and bends with everything else.

## Which axis it lands on

Nothing in that sentence said "horizontal". The mark decided it, and this is
how: `life_bands` has a `life` column and no `gdp` column, and
the plot's `y` is `life`. One axis answers, the other does not, so the rule sits
on the one that does and spans the one that does not.

Change which column the table holds and the same word draws the line
vertically:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(gdp_threshold) + rule + style(color = "firebrick", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("The same mark, the other axis")
```

`gdp_threshold` holds `gdp`, the plot's `x`, so the line stands up. The mark did
not change, the sentence did not change, and there is no `flip` anywhere: the
data said which axis it was about.

## A rug is the same mark

A rug tick has no second position either: it sits at one value and reaches a
little way in from the edge. That is not a different geometry, only a different
distance, so it is a setting rather than a second mark:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  data(gdp_rug) + rule + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Where the countries actually sit on the income axis")
```

Look along the bottom. Each tick is one country, drawn at its own income, and the
row of them answers a question the cloud above cannot: how many countries are
there at each income, including the ones hidden underneath each other? Points
overlap, and the 66 countries below $5,000 look much like half that many. The
ticks share only one coordinate instead of two, so they overlap less, and the
crowding at the left stays visible.

That is what a rug is for in general. It puts the one-dimensional
distribution of a variable in the margin of a plot that is already showing
something else. You can read the spread of `gdp` without losing the scatter,
and without a second plot whose axis you would have to match.

Everything else about the mark is unchanged. `style(reach = "edge")` is the only
difference between this and the threshold lines above, and there is no distance to
set: a tick's length comes from the panel, the way a label's `nudge` comes from
the font size. A number there would be a pixel count that meant something
different in every facet and at every plot size.

### The other axis

A rug is not tied to the horizontal. Which axis it lands on is decided the way it
was for the threshold lines: by the column its own table holds. Hand it `life`
instead of `gdp` and the ticks run up the left edge. Hand it both, as two layers,
and you get the margins of the scatter on both sides at once:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  data(gdp_rug)  + rule + style(reach = "edge") +
  data(life_rug) + rule + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both margins, from two tables")
```

Two layers rather than one, because each rug reads its own single-column
table. A table holding `gdp` *and* `life` at once is the ambiguous case the
grammar refuses, and the refusal offers two ways out: a single-column table,
or naming the axis on the layer, `rule + x(gdp)`.

Read the left edge and the distribution looks different. The ticks crowd
above 70, thin out through the sixties, then spread out again from 60 all the
way down to 40. That thin band in the middle is what to notice: of the 142 countries,
83 sit above 70 and 43 below 60, with only 16 in the decade between. It is a
distribution with two groups and a gap, not one spread, and neither the scatter
nor the bottom rug shows it.

## Coloring the lines

Each row of a rule's table is its own segment, so `color` maps per row. The
second column in `life_bands` is for this:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgray") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("One line per band, keyed by the legend")
```

A rule has no area for a reader to decode a ramp from, so `color` takes
categories rather than a continuous ramp, the same answer `line` and `path`
give.

## Thickening a rule

`style(size = )` makes a rule as thick as you want, and a thick translucent
one looks like a highlighted band:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.55) +
  data(life_bands) + rule + style(color = "seagreen", size = 26, opacity = 0.20) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A thick rule: emphasis, measured in pixels")
```

That is a good way to draw attention to a value, and be clear about what it
is: a stroke width is pixels, not data. The band above
covers about four years of life expectancy, but only because of where this
plot's vertical axis happens to sit. Facet it, resize it, or change the data's range
and the same `size = 26` will cover a different number of years while looking
identical. Nothing is wrong with that as long as the width is decoration; it is
wrong the moment the width is supposed to *mean* something.

When it should mean something ("the healthy range is 65 to 75", not "look
here"), the thickness belongs to the data, and then you want a region with real
edges. That is [`ribbon`](ribbon.qmd) given a low and a high column you
computed, with a two-row table placing the left and right edges:

```{r}
healthy <- data.frame(gdp = base::range(gapminder_2007$gdp), lo = 65.0, hi = 75.0)

data(healthy) + ribbon * bounds(lo, hi) + style(color = "seagreen", opacity = 0.18) +
  data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(color = "seagreen", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A band whose edges are 65 and 75, and rules on top")
```

Now 65 and 75 are positions. They rescale with the axis, they hold across
facets, and on a log axis they land where the data puts them rather than
where the pixels did. The highlight is written *first* so the points draw
over it.

Give that same table two different `gdp` values instead of the data's range,
and the band becomes a box bounded on both axes. That is how you shade a
corner of a scatter, or one span of a time series.

Neither the band nor the box reaches the **panel**. A ribbon is bounded
by its data, so it stops exactly at the numbers you gave it, and you cannot pad
them outward without widening the scale itself. A rectangle that takes one
extent from the data and the other from the panel is `rule`'s sibling one
dimension up, and that is exactly what [`zone`](zone.qmd) is.

## Bent: the ring and the spoke

The mark does one thing: it spans the whole of the axis it does not name. In
polar that is unchanged, and it draws two shapes rather than one, because a
circle has two ways of going all the way around.

Span the **turn** whole and you get a ring. Here it marks a target wind speed
on a circular wind plot, and `speed` is the radial axis:

```{r}
data(winds) + line * mean + x(direction) + y(speed) + polar() +
  data(speed_target) + rule + style(color = "firebrick", pattern = "dashed") +
  title("A target speed, ringed")
```

Span the **radius** whole and you get a spoke, a line from the center to the
rim. Nothing was added to the mark for either one. Spanning the axis a rule
does not name is what it always did, and in a circle the radial axis runs
from the center outward while the angular axis runs all the way around.

That is what a mark must show before polar counts as the same mark: stated
once, read off the bindings, meaning the same thing in every space. A rule
that needed a special case in polar would not have been one mark; it would
have been three marks under one name.

## What you can set

Everything a rule can be told to look like, generated from the engine's own
legality table, so this page cannot differ from what `style()` accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("rule")
```

Four of those are a stroke's ordinary controls, and one is the mark's own.
`style(reach = )` decides how far the rule crosses the axis it does not name, and
it is what makes the rug the same mark rather than a second one. The other four
behave here as on `line` or `path`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgray") +
  data(data.frame(life = 50.0)) + rule + style(color = "firebrick", size = 4) +
  data(data.frame(life = 60.0)) + rule + style(color = "seagreen", pattern = "dashed") +
  data(data.frame(life = 70.0)) + rule +
    style(color = "steelblue", pattern = "dotted", size = 2.5) +
  data(data.frame(life = 80.0)) + rule +
    style(color = "black", size = 6, opacity = 0.25) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Bottom to top: size, dashed, dotted, a faded thick rule")
```

The `pattern` values are the three **dashes**, because a rule is a stroke.
`style(pattern = "hatch")` is refused, pointing at the dashes: a hatch is a
*fill*'s texture, and it is what [`zone`](zone.qmd) takes instead. One setting
name, one realization per geometry.

What a rule refuses shows the same thing. There is no `shape`,
because a stroke has no glyph to choose. And there is no border:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(border_color = "black")
```

A border belongs to a filled shape, and a rule is the outline already, so
`style(color = )` is its color and `style(size = )` its width. That is the same
answer `line`, `step` and `path` give, because it is the same geometry.

## What it refuses

Give a rule the plot's own table and both position columns answer at once. There
is then nothing to say which axis is meant, so the grammar refuses rather
than picking one:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + rule
```

That refusal is why the tables above each hold a single column, and its
message names the other way out: `rule + x(gdp)` says the axis on the layer,
and then a both-column table works. It is the same shape as a bar with a
category on both axes: the grammar reads an orientation only when the
bindings distinguish the two axes, and guessing would teach the reader a
pattern the grammar does not follow.

A rule takes no transform either. It has handed one whole axis to the panel, so
there is no measure for a statistic to compute and nowhere to put the answer:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule * mean
```

The direction it gives is the one the mark is built on. A line at the mean is a
mean you computed, handed over as a column, which is what a rule's position
always is.
