# Rule

```{r}
#| include: false
source("../R/setup.R")
```

Where is the threshold these values should be compared against? Every other
mark reads two positions and draws something between them. `rule` reads *one*,
and asks the panel for the rest. That is the whole of it, and two familiar
things fall out of it: a reference line at a threshold, and a rug of ticks
showing where the observations actually sit.

Those are often four separate marks: a vertical reference line, a horizontal one,
a line from a slope and an intercept, and a rug. gog has one, because a vertical
and a horizontal reference line differ only in **which axis
carries the position**, and the grammar already reads that off the bindings,
the same way it decides whether a bar stands up or lies down.

## A line at a threshold

Here is the 2007 gapminder cloud, income against life expectancy, with a line
drawn where life expectancy reaches 70:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Three thresholds, one table")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the life bands."*

Three lines, and only one `rule` in the sentence. That is the mark's most useful
habit: **its position is a column**, not a number. `life_bands` holds three
values, so it draws three lines. Add a fourth row and a fourth line appears with
no change to the plot's sentence at all.

This is the same rule every column-taking atom keeps (`color(species)` names a
column, not a color), and it is why there is no `rule(70)`. A number written
into the plot would be a value the grammar could not scale, facet, or bend into
another coordinate space; a column is data, and data goes everywhere data goes.

## Which axis it lands on

Nothing in that sentence said "horizontal". The rule worked it out, and here is
the whole of how: `life_bands` has a `life` column and no `gdp` column, and
the plot's `y` is `life`. One axis answers, the other does not, so the rule sits
on the one that does and spans the one that does not.

Change which column the table holds and the same word stands the line up:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(gdp_threshold) + rule + style(color = "firebrick", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("The same mark, the other axis")
```

`gdp_threshold` holds `gdp`, the plot's `x`, so the line stands up. The mark did
not change, the sentence did not change, and there is no `flip` anywhere: the
data said which axis it was about.

## A rug is the same mark

A rug tick has no second position either: it sits at one value and reaches a
little way in from the edge. That is not a different geometry, only a different
distance, so it is a setting rather than a second mark:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  data(gdp_rug) + rule + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Where the countries actually sit on the income axis")
```

Look along the bottom. Each tick is one country, drawn at its own income, and the
row of them answers a question the cloud above cannot: how many countries are
there at each income, including the ones hidden underneath each other? Points
overlap, and a hundred countries crowded below $5,000 look much like forty. The
ticks do not overlap in the same way, because they only have to agree on one
coordinate instead of two, so the crowding at the left is visible as crowding.

That is what a rug is for in general. It puts the one-dimensional distribution
of a variable in the margin of a plot that is busy showing something else. You
can read the spread of `gdp` without giving up the scatter, and without a
second plot that would have its own axis to reconcile.

Everything else about the mark is unchanged. `style(reach = "edge")` is the only
difference between this and the threshold lines above, and there is no distance to
set: a tick's length comes from the panel, the way a label's `nudge` comes from
the font size. A number there would be a pixel count that meant something
different in every facet and at every plot size.

### The other axis

A rug is not tied to the horizontal. Which axis it lands on is decided the way it
was for the threshold lines: by the column its own table holds. Hand it `life`
instead of `gdp` and the ticks run up the left edge. Hand it both, as two layers,
and you get the margins of the scatter on both sides at once:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  data(gdp_rug)  + rule + style(reach = "edge") +
  data(life_rug) + rule + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both margins, from two tables")
```

Two layers rather than one, and that is the constraint the chapter started
with rather than a limitation of rugs. A rule reads its axis off its table, so
a single table holding `gdp` *and* `life` would be the ambiguous case the
grammar refuses. One table per axis is how you say which you meant.

Read the left edge and the picture changes character. The ticks crowd above 70,
go noticeably sparse through the sixties, then spread out again from 60 all the
way down to 40. That thin band in the middle is the point: of the 142 countries,
83 sit above 70 and 43 below 60, with only 16 in the decade between. It is a
distribution with two groups and a gap, not one spread, and neither the scatter
nor the bottom rug shows it.

## Coloring the lines

Each row of a rule's table is its own segment, so `color` maps per row. The
second column in `life_bands` was waiting for this:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgrey") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("One line per band, keyed by the legend")
```

A rule is paint on a hairline, so `color` takes categories rather than a
continuous ramp, the same answer `line` and `path` give.

## Thickening a rule

`style(size = )` will make a rule as fat as you like, and a fat translucent one
reads as a highlighted band:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.55) +
  data(life_bands) + rule + style(color = "seagreen", size = 26, opacity = 0.20) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A thick rule: emphasis, measured in pixels")
```

That is a perfectly good way to draw attention to a value, and it is worth being
clear about what it is: a stroke width is pixels, not data. The band above
covers about four years of life expectancy, but only because of where this
plot's vertical axis happens to sit. Facet it, resize it, or change the data's range
and the same `size = 26` will cover a different number of years while looking
identical. Nothing is wrong with that as long as the width is decoration; it is
wrong the moment the width is supposed to *mean* something.

When it should mean something ("the healthy range is 65 to 75", not "look
here"), the thickness belongs to the data, and then you want a region with real
edges. That is [`ribbon`](ribbon.qmd) fed a pre-computed pair, with a two-row
table putting the left and right edges where you want them:

```{r}
healthy <- data.frame(gdp = base::range(gapminder_2007$gdp), lo = 65.0, hi = 75.0)

data(healthy) + ribbon * bounds(lo, hi) + style(color = "seagreen", opacity = 0.18) +
  data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(color = "seagreen", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A band whose edges are 65 and 75, and rules on top")
```

Now 65 and 75 are positions. They rescale with the axis, they survive a facet,
and on a log axis they land where the data says rather than where the pixels
fell. The highlight is written *first* so the points draw over it.

Give that same table two different `gdp` values instead of the data's range and
it is a box rather than a band, bounded on both axes, which is how you shade a
corner of a scatter, or one span of a time series.

The one thing neither spelling does is reach the **panel**. A ribbon is bounded
by its data, so it stops exactly at the numbers you gave it, and you cannot pad
them outward without widening the scale itself. A rectangle that takes one
extent from the data and the other from the panel is `rule`'s trick one
dimension up, and that is exactly what [`zone`](zone.qmd) is.

## Bent: the ring and the spoke

The mark's one sentence is *a rule spans the axis it does not name, whole*. In
polar that sentence is unchanged, and it draws two shapes rather than one,
because a circle has two ways of going all the way round.

Span the **turn** whole and you get a ring. Here it marks a target wind speed on
the rose, and `speed` is the radial axis:

```{r}
data(winds) + line * mean + x(direction) + y(speed) + polar() +
  data(speed_target) + rule + style(color = "firebrick", pattern = "dashed") +
  title("A target speed, ringed")
```

Span the **radius** whole and you get a spoke. Nothing was added to the mark to
get either one: spanning the axis a rule does not name is what it always did,
and in a circle the radial axis runs from the center outward while the angular
axis runs all the way round.

That is the test a mark has to pass here before it is given a shortcut. The pie
earned one the same way: stated once, read off the bindings, and meaning the
same thing in every coordinate space. A rule that needed a special case in polar
would not have been one rule; it would have been three marks wearing one name.

## What you can set

Everything a rule can be told to look like, generated from the engine's own rule
table so this page cannot drift from what `style()` actually accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("rule")
```

Four of those are a stroke's ordinary controls, and one is the mark's own.
`style(reach = )` decides how far the rule crosses the axis it does not name, and
it is what makes the rug the same mark rather than a second one. The other four
behave here as on `line` or `path`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgrey") +
  data(data.frame(life = 50.0)) + rule + style(color = "firebrick", size = 4) +
  data(data.frame(life = 60.0)) + rule + style(color = "seagreen", pattern = "dashed") +
  data(data.frame(life = 70.0)) + rule +
    style(color = "steelblue", pattern = "dotted", size = 2.5) +
  data(data.frame(life = 80.0)) + rule +
    style(color = "black", size = 6, opacity = 0.25) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Bottom to top: size, dashed, dotted, a faded thick rule")
```

The `pattern` values are the three **dashes**, because a rule is a stroke.
`style(pattern = "hatch")` is refused, pointing at the dashes: a hatch is a
*fill*'s texture, and it is what [`zone`](zone.qmd) takes instead. One setting
name, one realization per geometry.

What a rule refuses says the same thing from the other side. There is no `shape`,
because a stroke has no glyph to choose. And there is no border:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(border_color = "black")
```

A border belongs to a filled shape, and a rule is the outline already, so
`style(color = )` is its color and `style(size = )` its width. That is the same
answer `line`, `step` and `path` give, because it is the same geometry.

## What it refuses

Give a rule the plot's own table and both position columns answer at once. There
is then nothing to say which axis is meant, so the grammar says so rather than
picking one:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + rule
```

That refusal is the reason the tables above each hold a single column. It is the
same shape as a bar with a category on both axes: the grammar can only read an
orientation when the bindings actually distinguish the two axes, and guessing
would teach a rule that is not real.

A rule takes no transform either. It has handed one whole axis to the panel, so
there is no measure for a statistic to compute and nowhere to put the answer:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule * mean
```

The direction it gives is the one the mark is built on. A line at the mean is a
mean you computed, handed over as a column, which is what a rule's position
always is.
