# Rule

```{r}
#| include: false
source("../R/setup.R")
```

Where is the line you want to compare every value against? `rule` draws a line
at one position and lets the other axis decide how far it reaches. Every other
mark reads two positions; a rule reads one and takes the rest from the panel.
That gives two familiar plots: a reference line at a threshold, and a rug of
ticks showing where the observations sit.

Those are often three separate marks: a vertical reference line, a horizontal
one, and a rug. gog has one. A vertical and a horizontal reference line differ
only in which axis carries the position. The grammar reads that off the
bindings, exactly as it does for a bar.

## A line at a threshold

Which countries have a life expectancy above 70? A scatter leaves the reader
to find 70 on the axis and hold a line there by eye. A line drawn at the
threshold does that once, for every reader.
Here is the 2007 gapminder scatter, income against life expectancy, with
lines drawn at three life-expectancy thresholds:

```{r}
#| echo: false
peek(gapminder_2007)
peek(life_bands)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Three thresholds, one table")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands."*

Three lines, and only one `rule` in the sentence. That is the mark's most useful
property: its position is a **column**, not a number. `life_bands` holds three
values, so it draws three lines. Add a fourth row and a fourth line appears with
no change to the plot's sentence at all.

Every atom that takes a column works this way (`color(species)` names a
column, not a color), and that is why there is no `rule(70)`. A number
written into the plot would be a value the grammar could not scale, facet, or
bend into another coordinate space; a column is data, so it scales, facets
and bends with everything else.

## Which axis it lands on

Nothing in that sentence said "horizontal". The mark decided it, and this is
how: `life_bands` has a `life` column and no `gdp` column, and
the plot's `y` is `life`. One axis answers, the other does not, so the rule sits
on the one that does and spans the one that does not.

A threshold on income wants a line standing up rather than lying across, and
there is no second mark for it.
Change which column the table holds and the same word draws the line
vertically:

```{r}
#| echo: false
peek(gdp_threshold)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(gdp_threshold) + rule + style(color = "firebrick", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("The same mark, the other axis")
```

`gdp_threshold` holds `gdp`, the plot's `x`, so the line stands up. The mark did
not change and the sentence did not change. Where ggplot2 has `geom_vline()`
and `geom_hline()`, and matplotlib `axvline()` and `axhline()`, gog has one
`rule`: the column the table holds is what chose the axis.

## A rug is the same mark

A scatter says where each country sits. How the countries are spread along the
income axis alone is a second question.
A rug tick has no second position either: it sits at one value and reaches a
little way in from the edge. That is not a different geometry, only a different
distance, so it is a setting rather than a second mark:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  rule + x(gdp) + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Where the countries actually sit on the income axis")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules, x is gdp,
with reach edge."*

The rule names `x(gdp)` even though the points already named it. This layer has
no table of its own, so it reads the scatter's, which holds both `gdp` and
`life`. Both axes answer, so the rule has to say which one it sits on.
`life_bands` above held only one, which is why those rules named nothing.

Look along the bottom. Each tick is one country, drawn at its own income.
Together they answer a question the points above cannot. How many countries sit
at each income, counting the ones hidden under other points? Points overlap, so
the 66 countries below $5,000 look like far fewer dots than that. The ticks
share only one coordinate instead of two, so they overlap less, and the crowding
at the left stays visible.

That is what a rug is for in general. It puts the one-dimensional
distribution of a variable in the margin of a plot that is already showing
something else. You can read the spread of `gdp` without losing the scatter,
and without a second plot whose axis you would have to match.

Everything else about the mark is unchanged. The only new setting is
`style(reach = "edge")`, and there is no distance to set. A tick's length
comes from the panel, the way a label's nudge distance comes from the font size.
A number there would be a pixel count that meant something different in every
facet and at every plot size.

### The other axis

Income is not the only axis worth a rug. Life expectancy has a distribution of
its own, and the scatter hides it the same way.
A rug is not tied to the horizontal. Which axis it lands on is decided by the
position you name on the layer. Write `y(life)` instead of `x(gdp)` and the
ticks run up the left edge. Write both, as two layers, and the scatter gets a
rug along the bottom and another up the left at once:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  rule + x(gdp)  + style(reach = "edge") +
  rule + y(life) + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both margins, from two layers")
```

Two layers rather than one, because each rug names one axis, and both read the
scatter's table.

Read the left edge and the distribution looks different. The ticks crowd
together above 70, thin out between 60 and 70, then run down to 40 in a looser
second band. That thin band in the middle is what to notice: of the 142
countries, 83 sit above 70 and 43 below 60, with only 16 between. It is a
distribution with two groups and a gap, not one spread, and neither the scatter
nor the bottom rug shows it.

## Coloring the lines

Three lines drawn alike leave the reader to work out which threshold is which
from the axis. A name on each, carried by its color and decoded by the legend,
removes that step.
Each row of a rule's table is its own segment, so `color` maps per row. The
`band` column in `life_bands`, which names each threshold, is for this:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgray") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("One line per band, keyed by the legend")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands, color by band."*

A thin line gives a reader no area to compare shades in, so `color` here takes
categories, not a continuous ramp. `line` and `path` answer the same way.

## Thickening a rule

A thin line marks a threshold without making it prominent. When one value is
the point of the plot, it should be seen before the data around it.
Width is the rule's other visible setting. `style(size = )` sets it, and a wide,
faint rule reads as a highlighted band:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.55) +
  data(life_bands) + rule + style(color = "seagreen", size = 26, opacity = 0.20) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A thick rule: emphasis, measured in pixels")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands, colored seagreen, with size 26 and opacity 0.20."*

That is a good way to mark a value, as long as you remember what the width is:
pixels, not data. The band above covers about two and a half years of life
expectancy. It covers that much only because this panel holds the range from 40
to 80. Facet it, resize it, or change the data's range and the same `size = 26`
will cover a different number of years while looking identical. Nothing is wrong
with that as long as the width is decoration; it is wrong the moment the width
is supposed to *mean* something.

When it should mean something ("the healthy range is 65 to 75", not "look
here"), the thickness belongs to the data, and then you want a region with real
edges. That is [`ribbon`](ribbon.qmd), given a low column and a high column you
computed. The table has two rows, one at each end of the income axis, so the
band reaches from side to side:

```{r}
#| echo: false
peek(healthy_band)
```

```{r}
data(healthy_band) + ribbon * bounds(lo, hi) + style(color = "seagreen", opacity = 0.18) +
  data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(color = "seagreen", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A band whose edges are 65 and 75, and rules on top")
```

*"Given the healthy band table: a ribbon from lo to hi; then given gapminder 2007:
points, x is gdp, y is life, and also rules from the life bands."*

Now 65 and 75 are positions. They rescale with the axis and keep the same values
in every facet. On a log axis they land where the data puts them, not at a fixed
number of pixels. The highlight is written *first* so the points draw over it.

Give that same table two different `gdp` values instead of the data's range,
and the band becomes a box bounded on both axes. That is how you shade a
corner of a scatter, or one span of a time series.

Neither the band nor the box stretches to the edge of the panel. A ribbon is
bounded by its data, so it stops exactly at the numbers you gave it, and you
cannot pad them outward without widening the scale itself. For a rectangle that
takes one side from the data and the other from the panel, use
[`zone`](zone.qmd). It is `rule` with one more dimension.

## In polar: the ring and the spoke

A wind plot is drawn on a circle, and a target speed on it is a threshold like
any other. The question is what a line at one value becomes when the axis it
spans is bent.
The mark does one thing: it spans the whole of the axis it does not name. In
polar that is unchanged, and it draws two shapes rather than one, because a
circle has two axes to span. [Polar](../polar.qmd) builds the circle itself;
what matters here is that a rule spans whichever axis it does not name.

`winds` is the table [The book's data](../book-data.qmd) introduced, and
`speed_target` holds the one target speed:

```{r}
#| echo: false
peek(winds)
peek(speed_target)
```

The angular axis runs all the way around the circle, and that whole sweep is the
**turn**. Span the turn whole and you get a ring. Here it marks a target wind
speed on a circular wind plot, and `speed` is the radial axis:

```{r}
data(winds) + line * mean + x(direction) + y(speed) + polar() +
  data(speed_target) + rule + style(color = "firebrick", pattern = "dashed") +
  title("A target speed, ringed")
```

*"Given the winds: a line derived by mean, x is direction, y is speed, in polar,
and also rules from the speed target."*

Span the *radius* whole and you get a spoke, a line from the center to the rim.
Give the rule a table holding `direction` instead of `speed`, and the line
stands at that compass point. Neither shape is a special case in the mark.
Spanning the axis a rule does not name is what it always did, and in a circle
the radial axis runs from the center outward while the angular axis runs all the
way around.

That is what a mark must show before polar counts as the same mark: stated once,
read off the bindings, meaning the same thing in every space. A rule that needed
a special case in polar would not be one mark. It would be three: a flat line, a
ring, and a spoke.

## What you can set

A threshold line has to look different from the data it crosses, and no column
says how. A setting decides that without reading one, and each mark takes its
own. These are a rule's, generated from the engine's own legality table, so
this page cannot differ from what `style()` accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("rule")
```

Four of those are a stroke's ordinary controls, and one is the mark's own.
`style(reach = )` decides how far the rule crosses the axis it does not name,
and it is what makes the rug the same mark rather than a second one. The other
four behave here as on `line` or `path`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgray") +
  data(data.frame(life = 50.0)) + rule + style(color = "firebrick", size = 4) +
  data(data.frame(life = 60.0)) + rule + style(color = "seagreen", pattern = "dashed") +
  data(data.frame(life = 70.0)) + rule +
    style(color = "steelblue", pattern = "dotted", size = 2.5) +
  data(data.frame(life = 80.0)) + rule +
    style(color = "black", size = 6, opacity = 0.25) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Bottom to top: size, dashed, dotted, a faded thick rule")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored lightgray; then
given one row with life 50: a rule colored firebrick, with size 4; then given
one row with life 60: a rule colored seagreen, with pattern dashed; then given
one row with life 70: a rule colored steelblue, with pattern dotted and size
2.5; then given one row with life 80: a rule colored black, with size 6 and
opacity 0.25."*

The `pattern` values are the three **dashes**, because a rule is a stroke.
`style(pattern = "hatch")` is refused, pointing at the dashes: a hatch is a
*fill*'s texture, and it is what [`zone`](zone.qmd) takes instead. One setting
name, one realization per geometry.

What a rule refuses follows from the same geometry. There is no `shape`,
because a stroke has no glyph to choose, and there is no border. [What it
refuses](#what-it-refuses) shows the border refused.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The shortest sentence to try is a bare `rule` over the plot's own table,
expecting a default axis.
Give a rule the plot's own table and both position columns answer at once.
There is then nothing to say which axis is meant, so gog refuses rather than
picking one:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + rule
```

That refusal is why the rugs above name their axis, `rule + x(gdp)` and `rule +
y(life)`. A threshold with a table of its own needs no position. That table
holds only one of the two columns, so the axis is read off it. A bar with
categories on both axes is refused for a related reason. Neither axis measures
anything, so there is nothing to read the orientation from, and gog will not
guess.

A rule takes no transform either. It has handed one whole axis to the panel, so
there is no measure for a statistic to compute and nowhere to put the answer:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule * mean
```

The message points back at what a rule is. A line at the mean is a mean you
computed yourself, passed in as a column. A rule's position always comes from a
column.

A rule has no border either. A border belongs to a filled shape, and a rule is
the outline already:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(border_color = "black")
```

So `style(color = )` is a rule's color and `style(size = )` its width. That is
the same answer `line`, `step` and `path` give, because it is the same
geometry.
