# Rule

```{r}
#| include: false
source("../R/setup.R")
```

Where is the line you want to compare every value against? Most marks read two
positions. `rule` reads one, and draws a line there. Along the other axis, the
line crosses the whole panel. That gives two familiar plots: a **reference
line** at a threshold, and a **rug** of ticks showing where the observations
sit.

Most plotting libraries draw these two plots with three marks: a vertical
reference line, a horizontal one, and a rug. gog has one. A vertical and a
horizontal reference line differ only in which axis holds the position. gog
works that out from the columns, as it does for a bar.

## A line at a threshold

How many countries have a life expectancy above 70? On a scatter the reader has
to find 70 on the axis and judge each point against it. With a line at 70, each
point is compared with the line instead of with the axis.
Here is the 2007 gapminder scatter, income against life expectancy, with
lines drawn at three life-expectancy thresholds:

```{r}
#| echo: false
peek(gapminder_2007)
peek(life_bands)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Three thresholds, one table")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands."*

The sentence names `rule` once, and the plot draws three lines. That is the
mark's most useful property: its position is a *column*, not a number.
`life_bands` holds three values, one per line. Add a fourth row and a fourth
line appears with no change to the plot's sentence at all.

Every channel works this way: `color(band)` names a column, not a color. That is
why there is no `rule(70)`. A number typed into the plot would be one value for
the whole plot. A column is data, so it can hold a different threshold for each
panel of a faceted plot, and it follows the scale and the space like every other
column.

A rule takes no transform, so it cannot compute a reference for you. [What it
refuses](#what-it-refuses) shows that refusal. Which reference you need decides
which mark you write:

| Mark | The reference it draws |
|---|---|
| `rule` | a line at each value in a column you supply |
| `line * mean` | the means joined into a profile |
| `ribbon * bounds` | a band between a low column and a high one |
| `zone * bounds` | a block across the panel, between the same two |

For a line at the overall mean, compute the mean yourself and give it to `rule`
as a column.

## Horizontal or vertical

Nothing in the read-aloud sentence under that plot said "horizontal". The mark
took the direction from the columns instead. `life_bands` has a `life` column
and no `gdp` column, and the plot's `y` is `life`. One axis answers and the
other does not. The rule sits on the axis that answers, and spans the axis that
does not.

A threshold on income needs a vertical line, and there is no second mark for it.
Give the rule a table that holds `gdp` instead, and the same word draws the line
vertically:

```{r}
#| echo: false
peek(gdp_threshold)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(gdp_threshold) + rule + style(color = "firebrick", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("The same mark, the other axis")
```

`gdp_threshold` holds `gdp`, the plot's `x`, so the line is vertical. gog draws
both directions with one `rule`. ggplot2 needs two marks for them,
`geom_vline()` and `geom_hline()`, and matplotlib needs `axvline()` and
`axhline()`.

## A rug of ticks

A scatter shows where each country sits on two axes at once. A second question
is only about how the countries spread along the income axis.
A rug tick reads one position too. It sits at one value and extends a short
distance into the panel from its edge. That is a threshold line, only shorter.
One setting makes the line short, so a rug needs no second mark:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  rule + x(gdp) + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Where the countries sit on the income axis")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules, x is gdp,
with reach edge."*

The rule names `x(gdp)` even though the points already named it. This layer has
no table of its own, so it reads the scatter's, which holds both `gdp` and
`life`. Both axes answer, so the rule has to say which one it sits on. The
`life_bands` table holds only `life`, which is why those rules named no
position.

Look along the bottom. Each tick is one country, drawn at its own income. Below
$5,000 the ticks merge into one solid block: 66 of the 142 countries sit in the
first tenth of the income axis. The scatter holds the same countries but spreads
them up the other axis, so the crowding is harder to see.

In general a rug puts the one-dimensional distribution of one column in the
margin of a plot that already shows something else. You read the spread of
`gdp` and still see the scatter, with no second plot whose axis you would have
to match.

Everything else about the mark is unchanged. `style(reach = )` takes one of two
words: `panel`, the default, and `edge`. You do not give it a length. A tick's
length comes from the panel: a small share of it, held between 4 and 14 pixels
so that it stays visible in a small facet. A number in place of `edge` would be
a pixel count, and it would mean something different in every facet and at
every plot size.

### The other axis

The rug so far has been along the bottom, on income. Life expectancy has a
distribution of its own, and the overlapping points hide it the same way.
A rug is not only for the horizontal axis. The position you name on the layer
decides which axis it sits on. Write `y(life)` instead of `x(gdp)` and the
ticks run up the left edge. Write both, as two layers, and the scatter has a
rug along the bottom and another up the left at once:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.45) +
  rule + x(gdp)  + style(reach = "edge") +
  rule + y(life) + style(reach = "edge") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Both margins, from two layers")
```

Each rug names one axis, and both read the scatter's table. A single `rule`
layer that names both is refused, because a rule draws one line per row, placed
by one column. The message suggests the two layers above, as [What it
refuses](#what-it-refuses) shows.

The left edge shows a different distribution from the bottom edge. The ticks
are dense above 70, thinnest between 60 and 70, and a little denser again from
60 down to 40. Of the 142 countries, 83 sit above 70 and 43 below 60, with only
16 between. This is a distribution with two groups and a thin stretch between
them, not a single group. The bottom rug cannot show this, because its ticks are
placed by income. In the scatter it is harder to see, because the same
countries are spread over two axes.

## Coloring the lines

The three threshold lines in the first plot are all drawn alike, and nothing
says what each one stands for. Give each line its own color instead, and the
legend prints the band's name beside that color.
A rule draws one line per row of its table, so `color` gives each row its own
color. `life_bands` has a `band` column naming each threshold, and that is the
column to use:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgray") +
  data(life_bands) + rule + color(band) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("One line per band, named in the legend")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands, color by band."*

A line colored by a number changes shade along its length. A rule is one
segment in one shade, and a single shade on a thin line is too hard to read a
number from. So `color` on a rule takes categories, not a continuous ramp.

## Thickening a rule

A thin line marks a threshold, but it is not the first thing a reader sees.
Sometimes the thresholds are the reason the plot exists. Then the reader should
find those lines first, before reading the data around them.
`style(size = )` sets a rule's width, and a wide, faint rule reads as a
highlighted band:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(opacity = 0.55) +
  data(life_bands) + rule + style(color = "seagreen", size = 26, opacity = 0.20) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A thick rule: emphasis, measured in pixels")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also rules from the
life bands, colored seagreen, with size 26 and opacity 0.20."*

That is a good way to mark a value. The width is measured in pixels, not in
data. Each wide rule in that plot covers a little over two years of life
expectancy, only because this axis runs from 40 to 80. Facet the plot, resize
it, or change the range of the data, and the same `size = 26` covers a
different number of years while looking identical. Nothing is wrong
with that as long as the width is decoration. It is wrong as soon as the width
is supposed to *mean* something.

### Thickness as data

Sometimes the thickness is itself a measurement: "the healthy range is 65 to
75", not "look here". Then the two edges are data, and you need a mark whose
edges are positions. That mark is [`ribbon`](ribbon.qmd). You give it a low
column and a high column. `healthy_band` has two rows, one at the lowest income
and one at the highest. The band therefore runs from the poorest
country to the richest:

```{r}
#| echo: false
peek(healthy_band)
```

```{r}
data(healthy_band) + ribbon * bounds(lo, hi) + style(color = "seagreen", opacity = 0.18) +
  data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(color = "seagreen", pattern = "dashed") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("A band whose edges are 65 and 75, and rules on top")
```

*"Given the healthy band table: a ribbon from lo to hi; then given gapminder 2007:
points, x is gdp, y is life, and also rules from the life bands."*

Now 65 and 75 are positions. They rescale with the axis and keep the same values
in every facet. On a log axis they are still drawn at 65 and 75, not a fixed
number of pixels apart. The band is written *first*, and layers are drawn in the
order they are written, so the points are drawn over it.

Give that same table two different `gdp` values instead of the data's range,
and the band becomes a box bounded on both axes. That is how you shade a
corner of a scatter, or one period of a time series.

Neither the band nor the box extends to the edge of the panel. A ribbon is
bounded by its data, so it stops exactly at the numbers you gave it. Give it
wider values and the scale grows to fit them, so a ribbon never touches the
edge. For a rectangle whose sides come from the data on one axis and from the
panel on the other, use [`zone`](zone.qmd). A zone takes a pair of
values on one axis and spans the other, as a rule takes one value and spans the
other.

## In polar: the ring and the spoke

Wind comes from a compass direction, so its speed can be drawn on a circle, as
a radar chart. A target speed on that chart is a threshold like any other. In
polar one of the two axes is bent into a circle. The question is what shape the
line takes then.
The mark does one thing: it draws a line at one position and spans the whole of
the other axis. In polar that is unchanged. A polar plot has two axes, the angle and the radius. A
rule that spans the angle is a ring, and one that spans the radius is a spoke:
a straight line from the center to the rim, the outer edge of the circle. The
[Polar](../polar.qmd) chapter explains how the circle itself is built.

`winds` is the table that [The book's data](../book-data.qmd) introduced.
`speed_target` holds one target speed:

```{r}
#| echo: false
peek(winds)
peek(speed_target)
```

The angular axis runs all the way around the circle, and one full circle of it
is called the **turn**. A rule that spans the whole turn closes into a ring.
Here it marks a target speed on the radar chart of mean wind speed by
direction, and `speed` is on the radial axis, the one that runs from the center
outward:

```{r}
data(winds) + line * mean + x(direction) + y(speed) + polar() +
  data(speed_target) + rule + style(color = "firebrick", pattern = "dashed") +
  title("A target speed, drawn as a ring")
```

*"Given the winds: a line derived by mean, x is direction, y is speed, in polar,
and also rules from the speed target."*

A rule that spans the whole *radius* draws a spoke. Give the rule a table
holding `direction` instead of `speed`, and the line sits at that compass point:

```{r}
data(winds) + line * mean + x(direction) + y(speed) + polar() +
  data(data.frame(direction = "E")) + rule +
    style(color = "firebrick", pattern = "dashed") +
  title("One compass point, center to rim")
```

*"Given the winds: a line derived by mean, x is direction, y is speed, in polar;
then given one row with direction E: a rule."*

Neither shape is a special case in the sentence you write. `rule` is still one
mark in polar because nothing about it changed. It is written the same way, it
takes its axis from its column, and it means the same thing in every space. A
rule that needed a special case in polar would not be one mark. It would be
three: a flat line, a ring, and a spoke.

## What you can set

A threshold line has to look different from the data it crosses, and no column
in the data says how it should look. You say it with a setting, and each mark
accepts its own list. These are a rule's settings, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("rule")
```

Four of those are a stroke's ordinary controls, and one is the mark's own.
`style(reach = )` sets how far the rule reaches across the panel. The value
`edge` shortens it to a tick. The other four behave here as on `line` or
`path`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + style(color = "lightgray") +
  data(data.frame(life = 50.0)) + rule + style(color = "firebrick", size = 4) +
  data(data.frame(life = 60.0)) + rule + style(color = "seagreen", pattern = "dashed") +
  data(data.frame(life = 70.0)) + rule +
    style(color = "steelblue", pattern = "dotted", size = 2.5) +
  data(data.frame(life = 80.0)) + rule +
    style(color = "black", size = 6, opacity = 0.25) +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Bottom to top: thick, dashed, dotted, faded")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored lightgray; then
given one row with life 50: a rule colored firebrick, with size 4; then given
one row with life 60: a rule colored seagreen, with pattern dashed; then given
one row with life 70: a rule colored steelblue, with pattern dotted and size
2.5; then given one row with life 80: a rule colored black, with size 6 and
opacity 0.25."*

The `pattern` values are the five **dashes**, `solid` among them, because a rule
is a stroke. `style(pattern = "hatch")` is refused, and the message lists the
dashes instead. A hatch is a *fill*'s texture. `bar`, `area`, `box` and
`ribbon` take it, and so does [`zone`](zone.qmd), the next chapter.
The setting has one name, and each mark draws it the way its shape allows. A
stroke takes a dash, and a fill takes a hatch.

What a rule refuses follows from the same geometry. There is no `shape`,
because a stroke has no glyph to choose, and there is no border. [What it
refuses](#what-it-refuses) shows the border refused.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list.
The [companion grid](../combinations.qmd#marks-and-channels) has a row per mark,
listing what it *maps* rather than sets.

## What it refuses

The first sentence you are likely to try is a bare `rule` on the plot's own
table. That table holds both position columns, so both axes answer. Nothing in
the sentence says which axis is meant, so gog refuses rather than choosing one:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + rule
```

That refusal is why every rug in this chapter names its axis, `rule + x(gdp)`
or `rule + y(life)`. A threshold with a table of its own needs no position. That
table holds only one of the two columns, so the axis is taken from it. A bar
also takes its direction from its columns, and gog refuses a bar with
categories on both axes. The rule had a column for both axes. The bar has the
reverse problem: neither axis holds a number, so the bar has nothing to
measure, and gog will not guess.

The opposite, one layer that names both positions, is refused too. A rule draws
one line per row, placed by one column, so only one of the two columns could be
drawn:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  rule + x(gdp) + y(life) + style(reach = "edge")
```

The message suggests two layers, one per axis, as in the plot under [The other
axis](#the-other-axis).

A rule takes no transform either. It is placed by one column and spans the
whole panel on the other axis. A statistic such as `mean` summarizes one axis
for each value of the other, and a rule reads only one position:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule * mean
```

The message restates what a rule is. To draw a line at the mean, compute the
mean yourself and supply it as a column. A rule's position always comes from a
column.

A rule has no border either. A border is the outline of a filled shape, and a
rule is a line with no inside to outline:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(life_bands) + rule + style(border_color = "black")
```

So `style(color = )` is a rule's color and `style(size = )` its width. That is
the same answer `line`, `step` and `path` give, because it is the same
geometry.

## In short

A rule's position is a column and never a number, so three thresholds are three
rows. With its column on `x` the line is vertical, and on `y` horizontal.
`style(reach = "edge")` makes it a rug tick, and in polar it is a ring or a
spoke. A rule on the plot's own table must name exactly one axis, or gog
refuses it. It takes no transform: compute the value yourself and give the rule
that column.

## What comes next

A rule marks one value. Its width is measured in pixels rather than in data.
[Zone](zone.qmd) is the region between two values, a rectangle across the panel
rather than a line.
