# Step

```{r}
#| include: false
source("../R/setup.R")
```

How do you draw a value that holds until it jumps? `step` draws it as a
staircase. Where `line` slopes straight from one point to the next, `step`
**holds** each value until it changes, then jumps. That is the right shape for
anything that does not change between observations: a cumulative count, a
survival probability, an interest rate. Each holds one value until the day it
changes. It takes the same channels as `line`; only the path between the points
differs.

Several fields have their own name for this shape, and one mark draws them all.
Statisticians draw the **empirical cumulative distribution function** (the
empirical CDF) with it. Medicine draws the **survival curve**, the Kaplan–Meier
estimate [@kaplan1958], which steps down at each observed death. Reliability
engineering draws the same curve and calls it a hazard or failure plot. Finance
draws the **step chart** of a rate that holds until someone changes it. A
cumulative count climbs and a survival probability falls; both are this one
staircase.

## Histogram outline

`step * bin` draws a histogram as an **outline** rather than as bars: one
stepped line tracing the top of each bin. Split it by `color` for the
overlaid outline histogram, one staircase per group, each drawn without a
fill:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species) +
  x_label("Petal length (cm)") + title("Petal length by species: step outline")
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species."*

Only the mark changes when you draw a binned distribution. These three are the
ones you will use most:

| Mark derived by `bin` | Shape |
|---|---|
| `bar * bin` | filled bars |
| `line * bin` | frequency polygon: straight lines through the bin centers |
| `step * bin` | the staircase outline above |

All three species are cut on the same bins, so their outlines share the same
edges. Each drops to the axis where its species has no data. Because a step is a
single unfilled stroke, each group's shape stays readable through the others,
and no opacity has to be set.

## A step function

Give `step` an `x` and a `y` directly and it draws a step function. An empirical
CDF is the best-known example: with *n* observations it climbs by 1/*n* at each
observation, and stays flat between them:

```{r}
v <- sort(gapminder_2007$life)
ecdf_df <- data.frame(life = v, share = seq_along(v) / length(v))
data(ecdf_df) + step + x(life) + y(share) +
  x_label("Life expectancy (years)") + y_label("Cumulative share of countries") +
  title("Empirical CDF of life expectancy, 2007")
```

*"Given the ecdf table: a step outline, x is life, y is share."*

A `line` on this same table would slope between points, implying cumulative
shares at life expectancies no country has. The step is exact: the share stayed
at that value until the next observation.

## A value that holds until it changes

The same exactness matters for anything that jumps rather than changing
gradually: a policy rate, a price, an inventory level. Each holds its value,
then jumps on a single day; a straight line would draw a gradual change that
never happened:

```{r}
rates <- data.frame(
  year = c(2018, 2019, 2020, 2021, 2022, 2023),
  rate = c(1.50, 2.25, 0.25, 0.25, 1.75, 4.50)
)
data(rates) + step + x(year) + y(rate) +
  x_label("Year") + y_label("Policy rate (%)") +
  title("A rate holds, then jumps")
```

## Multiple series

`color` and `group` split a `step` exactly as they split a `line`: one staircase
per category. Here two series hold and jump independently:

```{r}
inventory <- data.frame(
  week  = rep(1:8, 2),
  units = c(40, 40, 25, 25, 25, 10, 55, 55,
            30, 18, 18, 18, 42, 42, 30, 30),
  site  = rep(c("North", "South"), each = 8)
)
data(inventory) + step + x(week) + y(units) + color(site) +
  x_label("Week") + y_label("Units in stock") +
  title("Inventory by site: each site holds until it changes")
```

## A category on the domain

`step` and `line` share one channel row: whatever one requires, accepts or
refuses, the other does too. They differ in how the path is drawn, never in what
you may bind. So a category on the domain works here as it does in
[Line](line.qmd), and the staircase holds each category's value across its own
slot:

```{r}
data(gapminder_2007) + step * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("step * mean + x(continent): one value per category, held")
```

*"Given gapminder 2007: a step outline derived by mean, x is continent, y is
life."*

Of `line` and `step`, the step claims less about what was measured. A `line`
slopes between two categories, so it draws a change through values nothing
recorded. A step holds flat and then jumps, which fits separate groups better.
The [value that holds](#a-value-that-holds-until-it-changes) section made this
argument about time. It applies to categories in the same way.

## Stroke width and opacity

Like `line`, a `step` is a single stroke per series, so `size` and `opacity` are
not channels; they are settings on the whole stroke:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species) +
  style(size = 2.5, opacity = 0.7) +
  x_label("Petal length (cm)") + title("A thicker, more transparent outline")
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species, with size 2.5 and opacity 0.7."*

`size` sets the stroke width; `opacity` fades the whole staircase. Both follow
the same *setting vs mapping* rule as `line`; see [Setting vs
mapping](../style.qmd).

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("step")
```

A step is a stroke, so it takes exactly what `line` takes: one color, one width,
one opacity and one dash for the whole staircase. The six weeks table holds 42
consecutive days of `orders`, with `day` a real date:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + step + x(day) + y(orders) +
  style(color = "purple", size = 2.5, pattern = "dotted") +
  y_label("Orders") + title("A dotted staircase")
```

*"Given the six weeks table: a step outline, x is day, y is orders, colored
purple, with size 2.5 and pattern dotted."*

The staircase and the dash change different things. The staircase is *geometry*:
it decides where the path goes, and puts a corner where the value changes. The
dash is *paint*: it changes how the path looks, never where it goes. That is why
`step` is a mark and `pattern` is a setting.

## What it refuses

A step is one connected stroke from its first point to its last, and that
explains the two refusals below. The third is `line`'s: a category on `y` is refused, because `y`
carries a quantity.

One stroke has one width (there is no answer to "how wide is the staircase
*here*"), so a width cannot vary along it:

```{r}
#| error: true
data(gapminder_asia) + step + x(year) + y(life) + size(gdp)
```

The stroke still *has* a width, which is why the setting above works and the
mapping does not. That distinction is the subject of
[Setting vs mapping](../style.qmd).

The second refusal is about which patterns a stroke accepts. A hatch is a fill's
texture, and a staircase is line-work with nothing inside it:

```{r}
#| error: true
data(gapminder_asia) + step + x(year) + y(life) + style(pattern = "hatch")
```

A color is different. A **tread** is a flat section held at one value, and a
**riser** is the jump between two values. Point anywhere on the staircase and
that place has a color, so a measure can vary along it. Each tread takes its
row's color, and each riser blends the colors of the two rows it joins:

```{r}
data(gapminder_asia) + step + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Each country's staircase, colored by its income")
```

*"Given gapminder Asia: step outlines, x is year, y is life, grouped by country,
color by gdp on a log scale."*
