# Step

```{r}
#| include: false
source("../R/setup.R")
```

How do you draw a value that holds until it jumps? `step` draws it as a
staircase. A `line` slopes straight from one point to the next. A `step`
**holds** each value until it changes, then jumps. The **flat part** runs from
one observation to the next, and the **jump** is where the value changes. That
is the right shape for anything that does not change between observations: a
cumulative count, a survival probability, an interest rate. `step` takes the
same channels as `line`, and only the path drawn between two points differs.

Several professions draw this shape, each under its own name, and `step` draws
all of these plots. Statisticians draw the **empirical cumulative distribution
function** (the empirical CDF) with it. Medical researchers draw the **survival
curve**, the Kaplan–Meier estimate [@kaplan1958], which drops at each observed
death. Reliability engineers draw the same staircase for parts rather than
patients, and call it a failure plot. Analysts in finance draw the **step
chart** of a rate that holds until someone changes it. A cumulative count rises
and a survival probability falls, and `step` draws both as this one staircase.

## A step function

What share of European countries have a life expectancy at or below 75 years,
or 80, or any value you choose? The empirical CDF answers all of those questions
at once. It is a **step function**: flat between observations, with a jump at
each one. The `gm_europe_cdf` table holds one row per European country in 2007:
its life expectancy, and the share of countries at or below it. gog has no
transform that computes these shares. Sort a table by the value, and each row's
share is its position divided by the number of rows. The table already holds
the shares, so `step` needs no transform: give it an `x` and a `y`, and it
draws the curve. With *n* observations, each observation adds 1/*n* to the
share:

```{r}
#| echo: false
peek(gm_europe_cdf)
```

```{r}
data(gm_europe_cdf) + step + x(life) + y(share) +
  x_label("Life expectancy (years)") + y_label("Cumulative share of countries") +
  title("Empirical CDF of life expectancy in Europe, 2007")
```

*"Given the gm europe cdf table: a step outline, x is life, y is share."*

A `line` on this same table would slope between points, so it would show
cumulative shares at life expectancies no country has. The step is exact:
between two observations no country is added, so the share does not change.

Whether you can see the flat parts depends on how many observations there are.
Thirty countries make each jump a thirtieth of the axis, and most of the flat
parts between them are easy to see. Draw the same plot over all 142 countries
in the `gapminder_2007` table, and each jump is about a fifth as tall as here.
The staircase then reads as a slightly rough line.

## A value that holds until it changes

The same exactness matters for anything that jumps rather than changing
gradually: a policy rate (the interest rate a central bank sets), a price, an
inventory level. Each holds its value, then jumps on a single day. A straight
line would draw a gradual change that never happened. The `policy_rates` table
holds one interest rate for each year from 2018 to 2023, and a step draws it:

```{r}
#| echo: false
peek(policy_rates)
```

```{r}
data(policy_rates) + step + x(year) + y(rate) +
  x_label("Year") + y_label("Policy rate (%)") +
  title("A rate rises, falls, holds two years, then rises twice")
```

The rate fell to 0.25 in 2020 and held there for two full years, then rose
twice. A line would have sloped into and out of that flat part, showing rates
that were never set.

## Multiple series

The policy rate was a single series, and often you need to compare two. Each of
two sites keeps a stock of units, and only a delivery or a sale changes it. To
compare them week by week, you want both on one plot. `color` and `group` split
a `step` exactly as they split a `line`: one staircase per category. Here two
series hold and jump independently:

```{r}
#| echo: false
peek(inventory)
```

```{r}
data(inventory) + step + x(week) + y(units) + color(site) +
  x_label("Week") + y_label("Units in stock") +
  title("Inventory by site: each site holds until it changes")
```

*"Given the inventory: step outlines, x is week, y is units, color by site."*

North holds at 40 for two weeks, then falls twice before jumping to 55. South
starts lower, rises above North in week 5, and falls below it again in week 7.
Neither series slopes between its changes.

## A category on the domain

Every `x` so far has been a number. How do the continents compare on one
summary value? A step will draw that plot, but a bar is the better mark for it,
as the step drawn below shows. A step reads `x` as its **domain**, the axis it
runs along, and `y` as its **measure**, the quantity it traces. A category on
the domain gets one slot per value, as in
[Line](line.qmd#a-category-on-the-domain-the-profile). The staircase holds each
continent's mean from the center of its slot to the center of the next one:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + step * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("step * mean + x(continent): the five means as a staircase")
```

*"Given gapminder 2007: a step outline derived by mean, x is continent, y is
life."*

Africa sits lowest at 54.8 years and Oceania highest at 80.7. Oceania, the last
slot, has no next slot, so its mean is only the top of the last jump. The five
slots run in the order the continents first appear in the table. That order
means nothing, so the rise and fall across the slots is not a trend.

A step draws only the values that were measured. A `line` slopes from one
category to the next, and nothing was measured between them. A step holds flat
and then jumps, so every height it draws is one of the five means.

That does not make it the right mark for this plot. A flat part claims that the
value held across the part of the axis under it. A policy rate does hold
between two dates, as [A value that holds until it
changes](#a-value-that-holds-until-it-changes) showed. Between two continents
there is no value to hold, so the flat part there says nothing about the data.
One summary value per category is what a [bar](bar.qmd) draws: keep the `mean`
and change the mark. Each bar stands in its own slot, joined to nothing.

Other marks take the same `mean`, and only the shape changes:

| Mark derived by `mean` | Shape |
|---|---|
| `bar * mean` | one bar per category, each in its own slot |
| `point * mean` | one dot at each mean |
| `line * mean` | the means joined into a profile |
| `area * mean` | the same profile, filled to the baseline |
| `step * mean` | each mean held flat until the next category, the plot above |

Legal is not the same as useful. gog draws this step and prints no warning.
[Law 8](../design-laws.qmd#law-8-pronounceable-usable) says the grammar
guarantees a legal plot, never a useful one, so the choice of mark is yours.

## Histogram outline

`mean` is not the only transform a step takes. With `bin`, the staircase
answers a question that filled bars make harder to read: how do three
distributions compare in one panel? Filled histograms overlap wherever two
groups share a bin. gog fades those fills so the shapes underneath still show,
but the faded colors mix where they overlap. A step has no fill at all, so
nothing needs fading. `step * bin` draws a histogram as an **outline** rather
than as bars: one staircase along the tops of the bins. Add `color` and the
outline splits, one staircase per group, cut on the same bins:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species) +
  x_label("Petal length (cm)") + title("Petal length by species: step outline")
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species."*

`bin` cuts all three species into the same bins, so their outlines share the
same edges. An outline goes to zero in any bin where its species has no rows.

The same `bin` works with other marks, and only the mark changes:

| Mark derived by `bin` | Shape |
|---|---|
| `bar * bin` | filled bars |
| `line * bin` | frequency polygon: one stroke of straight pieces through the bin centers |
| `area * bin` | the same polygon, filled to the baseline |
| `step * bin` | the staircase outline above |

## Stroke width and opacity

The three outlines above are thin, and where two of them run along the same bin
edge, the one drawn last covers the other. A thicker stroke is easier to see,
and a little transparency lets the covered stroke show through. Like `line`, a
`step` is a
single stroke per series, so `size` and `opacity` are not channels; they are
settings on the whole stroke:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species) +
  style(size = 2.5, opacity = 0.7) +
  x_label("Petal length (cm)") + title("A thicker, more transparent outline")
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species, with size 2.5 and opacity 0.7."*

`size` sets the stroke width. `opacity` makes the whole staircase more
transparent. Both are set, never mapped, exactly as on a `line`;
[Setting vs mapping](../style.qmd) explains the difference.

## What you can set

A setting changes how the staircase looks without reading a column, and each
mark takes its own. These are a step's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("step")
```

A step is a stroke, so it takes exactly what `line` takes: one color, one width,
one opacity and one dash for the whole staircase. Mapped to a column instead of
set, `color` alone can change *along* the staircase, as the last plot in
[What it refuses](#what-it-refuses) shows.

Set three of them at once and the staircase changes color, width and dash
together. The six weeks table holds 42 consecutive days of `orders`, and `day`
is a date column, so the axis shows calendar days:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + step + x(day) + y(orders) +
  style(color = "purple", size = 2.5, pattern = "dotted") +
  y_label("Orders") + title("A dotted staircase")
```

*"Given the six weeks table: a step outline, x is day, y is orders, colored
purple, with size 2.5 and pattern dotted."*

The staircase and the dash change different things. The staircase is *geometry*:
it sets where the path goes, and puts a corner where the value changes. The dash
is *paint*: it changes how the path looks, never where it goes. That is why
`step` is a mark, while a dash only changes how a mark is drawn.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A step is one connected stroke with no filled interior, and the first two
refusals follow from that. The third comes from the axes.

One stroke has one width, the same at every place along it. So no column can
make the staircase wider in one place than in another:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
#| error: true
data(gapminder_asia) + step + x(year) + y(life) + size(gdp)
```

The stroke still *has* a width, and `style(size = )` sets it. `size(column)`
asks for a width per row, which one stroke cannot give. That distinction is the
subject of [Setting vs mapping](../style.qmd).

The second refusal is about which patterns a stroke accepts. A hatch is a
texture that fills an area, and a staircase is a stroke with no area to fill:

```{r}
#| error: true
data(gapminder_asia) + step + x(year) + y(life) + style(pattern = "hatch")
```

The third refusal is one that `line`, `area` and `ribbon` give too. A step reads
`y` as its measure, and a category is not a quantity, so gog refuses it there:

```{r}
#| error: true
data(gapminder_2007) + step * mean + x(life) + y(continent)
```

`color` is not refused: it is the one property that can change along a single
staircase. This is the plot promised in [What you can set](#what-you-can-set).
A width and a dash belong to the whole stroke. A dash can still change from one
series to the next, as it does for a `line`, but never along one series. A
color belongs to each place on the stroke instead, so a column can change it
from one row to the next. Each flat part takes its row's color, and each jump
blends the colors of the two rows it joins:

```{r}
data(gapminder_asia) + step + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Each country's staircase, colored by its income")
```

*"Given gapminder Asia: step outlines, x is year, y is life, grouped by country,
color by gdp on a log scale, with the viridis palette."*

Read one country's staircase from left to right and you see two variables. The
height is how long people live, and the color is how rich the country was then.
The color stays the same along each flat part and changes only at the jumps,
where the value changes too.

## In short

A step holds each value until it changes, so it invents no value between two
observations. It takes the same channels and settings as a `line`: width and
opacity belong to the whole stroke, and only `color` varies along it. With
`color` or `group` it draws one staircase per category, and with `bin` it draws
a histogram as an outline instead of as bars.

## What comes next

Each staircase follows one value at each `x`. [Interval](interval.qmd) spans
two values at every position instead, from a low one to a high one.
