# Step

```{r}
#| include: false
source("../R/setup.R")
```

How do you draw a value that holds until it jumps? `step` draws it as a
staircase. A `line` slopes straight from one point to the next. A `step`
**holds** each value until it changes, then jumps. The **flat part** runs from
one measurement to the next, and the **jump** is where the value changes. That
is the right shape for anything that does not change between observations: a
cumulative count, a survival probability, an interest rate. `step` takes the
same channels as `line`, and only the path drawn between two points differs.

Several subjects have their own name for this shape, and one mark draws them
all. Statisticians draw the **empirical cumulative distribution function** (the
empirical CDF) with it. Doctors draw the **survival curve**, the Kaplan–Meier
estimate [@kaplan1958], which steps down at each observed death. Reliability
engineers draw the same staircase for parts rather than patients, and call it a
failure plot. Analysts in finance draw the **step chart** of a rate that holds
until someone changes it. A cumulative count rises and a survival probability
falls, and `step` draws both as this one staircase.

## A step function

What share of European countries have a life expectancy at or below 75 years,
or 80, or any value you choose? The empirical CDF answers all of those at once,
and it is the plainest step function there is. The `gm_europe_cdf` table holds
one row per European country in 2007: its life expectancy, and the share of
countries at or below it. Give `step` an `x` and a `y` with no transform, and it
draws one. With *n* observations the curve rises by 1/*n* at each observation,
and stays flat between them:

```{r}
#| echo: false
peek(gm_europe_cdf)
```

```{r}
data(gm_europe_cdf) + step + x(life) + y(share) +
  x_label("Life expectancy (years)") + y_label("Cumulative share of countries") +
  title("Empirical CDF of life expectancy in Europe, 2007")
```

*"Given the gm europe cdf table: a step outline, x is life, y is share."*

A `line` on this same table would slope between points, so it would show
cumulative shares at life expectancies no country has. The step is exact: the
share stayed at that value until the next observation.

The size of *n* decides whether you can see that. Thirty countries make each
jump a thirtieth of the axis, and the flat parts between them are plain. The
same plot over all 142 countries in the world gives each jump a fifth of that
height, and the staircase reads as a slightly rough line.

## A value that holds until it changes

The same exactness matters for anything that jumps rather than changing
gradually: a policy rate, a price, an inventory level. Each holds its value,
then jumps on a single day. A straight line would draw a gradual change that
never happened:

```{r}
#| echo: false
peek(policy_rates)
```

```{r}
data(policy_rates) + step + x(year) + y(rate) +
  x_label("Year") + y_label("Policy rate (%)") +
  title("A rate holds, falls, then rises")
```

The rate fell to 0.25 in 2020 and held there for two full years, then rose
twice. A line would have sloped into and out of that flat run, showing rates
that were never set.

## Multiple series

One rate held and jumped on its own. Each of two sites keeps a stock of units,
and only a delivery or a sale changes it. To compare them week by week, you
want both on one plot. `color` and `group` split a `step` exactly as they split
a `line`: one staircase per category. Here two series hold and jump
independently:

```{r}
#| echo: false
peek(inventory)
```

```{r}
data(inventory) + step + x(week) + y(units) + color(site) +
  x_label("Week") + y_label("Units in stock") +
  title("Inventory by site: each site holds until it changes")
```

*"Given the inventory: step outlines, x is week, y is units, color by site."*

North holds at 40 for two weeks, then falls twice before jumping to 55. South
starts lower, passes North in week 5, and ends below it again. Neither series
slopes between its changes.

## A category on the domain

How do the continents compare on one summary value, and should a staircase be
the mark that shows it? A category works on the domain here exactly as it does
in [Line](line.qmd), and the staircase holds each category's value across its
own slot:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + step * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("step * mean + x(continent): one value per category, held")
```

*"Given gapminder 2007: a step outline derived by mean, x is continent, y is
life."*

Africa sits lowest at 54.8 years and Oceania highest at 80.7. The five slots
run in alphabetical order, so the rise and fall across them is not a trend.

A step draws only the values that were measured. A `line` slopes from one
category to the next, and nothing was measured between them. A step holds flat
and jumps at the edge of each slot, so it claims no change between two
continents.

That does not make it the right mark for this plot. A flat part claims the
value held across that stretch of the axis. A policy rate does hold between
two dates, as [A value that holds until it
changes](#a-value-that-holds-until-it-changes) showed. A continent is a slot
rather than a stretch, so the flat part says nothing about the data. One
summary value per category is what a [bar](bar.qmd) draws: keep the `mean` and
change the mark. Each bar stands in its own slot, joined to nothing.

Every mark that reads one measure takes the same `mean`, and only the shape
changes:

| Mark derived by `mean` | Shape |
|---|---|
| `bar * mean` | one bar per category, each in its own slot |
| `point * mean` | one dot at each mean |
| `line * mean` | the means joined into a profile |
| `area * mean` | the same profile, filled to the baseline |
| `step * mean` | each mean held flat across its slot, the plot above |

Legal is not the same as useful. gog draws this step and prints no warning.
[Law 8](../design-laws.qmd) guarantees a legal plot, never a useful one, so the
choice of mark is yours.

## Histogram outline

How do three distributions compare in one panel? Three filled histograms
overlap in the middle of the panel, and gog fades those fills so the shapes
underneath still show. A step has no fill at all, so nothing needs fading.
`step * bin` draws a histogram as an **outline** rather than as bars: one
stepped line tracing the top of each bin. Add `color` and the outline splits,
one staircase per group, cut on the same bins:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species) +
  x_label("Petal length (cm)") + title("Petal length by species: step outline")
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species."*

`bin` cuts all three species into the same bins, so their outlines share the
same edges. An outline goes to zero in any bin where its species has no rows. A
step is a single stroke with no fill, so where two outlines cross neither hides
the other.

The same `bin` works under other marks, and only the mark changes:

| Mark derived by `bin` | Shape |
|---|---|
| `bar * bin` | filled bars |
| `line * bin` | frequency polygon: straight lines through the bin centers |
| `area * bin` | the same polygon, filled to the baseline |
| `step * bin` | the staircase outline above |

## Stroke width and opacity

The three outlines above are thin, and thin strokes are hard to follow where
they cross. A thicker stroke is easier to see, and a little transparency keeps
each one visible where it passes over another. Like `line`, a `step` is a
single stroke per series, so `size` and `opacity` are not channels; they are
settings on the whole stroke:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species) +
  style(size = 2.5, opacity = 0.7) +
  x_label("Petal length (cm)") + title("A thicker, more transparent outline")
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color
by species, with size 2.5 and opacity 0.7."*

`size` sets the stroke width. `opacity` makes the whole staircase more
transparent. Both follow the same *setting vs mapping* rule as `line`; see
[Setting vs mapping](../style.qmd).

## What you can set

A setting changes how the staircase looks without reading a column, and each
mark takes its own. These are a step's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("step")
```

A step is a stroke, so it takes exactly what `line` takes: one color, one width,
one opacity and one dash for the whole staircase. Of the four, `color` is the
one that can also vary *along* the staircase, and the plot at the end of this
chapter shows that.

Set three of them at once and the staircase changes color, weight and dash
together. The six weeks table holds 42 consecutive days of `orders`, with `day`
a real date:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + step + x(day) + y(orders) +
  style(color = "purple", size = 2.5, pattern = "dotted") +
  y_label("Orders") + title("A dotted staircase")
```

*"Given the six weeks table: a step outline, x is day, y is orders, colored
purple, with size 2.5 and pattern dotted."*

The staircase and the dash change different things. The staircase is *geometry*:
it sets where the path goes, and puts a corner where the value changes. The dash
is *paint*: it changes how the path looks, never where it goes. That is why
`step` is a mark and `pattern` is a setting.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A step is one connected stroke with no filled interior, and the first two
refusals follow from that. The third comes from the axes.

One stroke has one width. Point anywhere on the staircase and there is no width
for that place alone. So a width cannot vary along it:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
#| error: true
data(gapminder_asia) + step + x(year) + y(life) + size(gdp)
```

The stroke still *has* a width, and `style(size = )` sets it. `size(column)`
asks for a width per row, which one stroke cannot give. That distinction is the
subject of [Setting vs mapping](../style.qmd).

The second refusal is about which patterns a stroke accepts. A hatch fills an
area, and a staircase is a stroke with no area to fill:

```{r}
#| error: true
data(gapminder_asia) + step + x(year) + y(life) + style(pattern = "hatch")
```

The third is the one `line`, `area` and `ribbon` also give. A step reads `x` as
its domain and `y` as its measure, so a category on `y` has nothing to measure:

```{r}
#| error: true
data(gapminder_2007) + step * mean + x(life) + y(continent)
```

`color` is the exception, and it is not a refusal at all. This is the plot
[What you can set](#what-you-can-set) promised. A stroke has one width the whole
way along it, so asking what its width is at one point means nothing, and a dash
is the same. A color is not like that. Every point along the staircase is drawn
in some color, so the color can change from one end to the other, and a column
can decide where. Each flat part takes its row's color, and each jump blends the
colors of the two rows it joins:

```{r}
data(gapminder_asia) + step + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Each country's staircase, colored by its income")
```

*"Given gapminder Asia: step outlines, x is year, y is life, grouped by country,
color by gdp on a log scale, with the viridis palette."*

Read one country's staircase from left to right and you see two variables. The
height is how long people live, and the color is how rich the country was then.
The flat part and the jump are the whole mark: a value held, then the change to
the next one.

## In short

A step holds each value until it changes, so nothing between two measurements is
claimed. It takes the same channels as a `line`: width and opacity are settings,
and only `color` varies along the stroke. Each jump is 1/*n* of the axis, so a
large *n* reads as a rough line.

## What comes next

A step shows one value at each `x`, never a pair. [Interval](interval.qmd)
draws the pair instead, a low value and a high one at every position.
