# Step

```{r}
#| include: false
source("../R/setup.R")
```

What if a value holds steady until it jumps? `step` is the line family's
staircase. Where `line` slants straight from one point to the next, `step`
**holds each value until it changes, then jumps**. That is the right shape for
anything that stays put between observations: a cumulative count, a survival
probability, an interest rate that was 2.5 % until the day it wasn't. It takes
the same channels as `line`; only the interpolation differs.

The shape has a name in every field that needs it, and they are all this one mark.
Statisticians draw the **empirical distribution function** with it. Medicine draws
the **survival curve**, the Kaplan–Meier estimate [@kaplan1958], whose steps fall at the
observed deaths. Reliability engineering draws the same curve and calls it a
hazard or failure plot, and finance draws the **step chart** of a rate that holds
until someone changes it. A cumulative count and a survival probability are one
staircase read in opposite directions.

## Histogram outline

`step * bin` draws a histogram as a **silhouette** rather than as bars: one
stepped line tracing the top of each bin. Split it by `color` and you get the
overlaid outline histogram, one staircase per group, with nothing painted over
anything:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species) +
  x_label("Petal length (cm)") + title("Petal length by species: step outline")
```

*"Given the iris flowers: a step outline derived by bin, x is petal length, color by species."*

This is the third way to draw a binned distribution, and the mark is the only
thing that changes:

| Mark over `bin` | Shape |
|---|---|
| `bar * bin` | filled bars |
| `line * bin` | frequency polygon: straight lines through the bin centers |
| `step * bin` | the staircase silhouette above |

The bins are shared across the groups, so the three silhouettes line up on the
same edges, and each drops to the axis where its species has no data. Because a
step is a single unfilled stroke, overlapping groups never obscure one another:
no opacity to set, no "which was drawn last" to worry about.

## A step function

Give `step` an `x` and a `y` directly and it draws a step function. An empirical
cumulative distribution is the classic example; it jumps by 1/n at each
observation and is flat between:

```{r}
v <- sort(gapminder_2007$life)
ecdf_df <- data.frame(life = v, share = seq_along(v) / length(v))
data(ecdf_df) + step + x(life) + y(share) +
  x_label("Life expectancy (years)") + y_label("Cumulative share of countries") +
  title("Empirical CDF of life expectancy, 2007")
```

Drawn with `line`, this same data would slant between points and imply
cumulative shares at life expectancies that no country has. The step tells the
truth: the share *was* that value until the next country's.

## A value that holds until it changes

The same honesty matters for anything that steps rather than glides: a policy
rate, a price, an inventory level. It held its value, then jumped on a single
day; a straight line would draw a gradual change that never happened:

```{r}
rates <- data.frame(
  year = c(2018, 2019, 2020, 2021, 2022, 2023),
  rate = c(1.50, 2.25, 0.25, 0.25, 1.75, 4.50)
)
data(rates) + step + x(year) + y(rate) +
  x_label("Year") + y_label("Policy rate (%)") +
  title("A rate holds, then jumps")
```

## Multiple series

`color` and `group` split a `step` exactly as they split a `line`: one
staircase per category. Here two series step independently:

```{r}
inventory <- data.frame(
  week  = rep(1:8, 2),
  units = c(40, 40, 25, 25, 25, 10, 55, 55,
            30, 18, 18, 18, 42, 42, 30, 30),
  site  = rep(c("North", "South"), each = 8)
)
data(inventory) + step + x(week) + y(units) + color(site) +
  x_label("Week") + y_label("Units in stock") +
  title("Inventory by site: each holds until it moves")
```

## A category on the domain

`step` shares `line`'s row in the legality table, which is the No Exceptions law
made literal: the two marks differ in how the path is drawn, never in what it may
be bound to. So a category on the domain works here exactly as it does there, and
the staircase holds each category's value across its own slot:

```{r}
data(gapminder_2007) + step * mean + x(continent) + y(life) +
  y_label("Mean life expectancy") +
  title("step * mean + x(continent): held across each slot")
```

That reading is arguably the more honest of the two. A `line` slopes between two
categories and so draws a change passing through values nothing recorded; a step
holds flat and jumps, which is closer to what a set of separate groups actually
says. The same argument the [step function](#a-step-function) section makes about
time applies to categories.

## Stroke width and opacity

Like `line`, a `step` is a single stroke per series, so `size` and `opacity` are
not channels; they are settings on the whole stroke:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species) +
  style(size = 2.5, opacity = 0.7) +
  x_label("Petal length (cm)") + title("A heavier, softer outline")
```

`size` sets the stroke width; `opacity` fades the whole line. Both follow the
same *setting vs mapping* rule as `line`; see [Setting vs mapping](../style.qmd).

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("step")
```

A step is a stroke, so it takes exactly what `line` takes: one width, one
opacity, one dash for the whole staircase.

```{r}
data(six_weeks) + step + x(day) + y(orders) +
  style(color = "purple", size = 2.5, pattern = "dotted") +
  y_label("Orders") + title("A dotted staircase")
```

A dash on a step is worth knowing about, because the two say different things: the
staircase is *geometry* (it moves the path, inserting corners and holding values),
while the dash is *paint*. That is the reason `step` is a mark and `pattern` is a
setting rather than both being one or the other.

## What it refuses

A step is one polyline drawn with one stroke, and that settles both refusals.

One stroke has one width (there is no answer to "how wide is the staircase
*here*"), so a width cannot vary along it:

```{r}
#| error: true
data(gapminder_asia) + step + x(year) + y(life) + size(gdp)
```

The stroke still *has* a width, which is why the setting above works and the
mapping does not. That distinction is the subject of
[Setting vs mapping](../style.qmd).

The second is vocabulary again. A hatch is a fill's texture, and a staircase is
line-work with nothing inside it:

```{r}
#| error: true
data(gapminder_asia) + step + x(year) + y(life) + style(pattern = "hatch")
```

A color is not like a width, though, and `step` takes `line`'s answer here as
it takes `line`'s answer everywhere: you *can* point at one tread and ask what
color it is, so a measure varies along the staircase. Each tread shows the
value it holds and each riser blends between the two it joins, which is what a
tread and a riser respectively mean:

```{r}
data(gapminder_asia) + step + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("Held values, colored by income")
```
