# Surface

```{r}
#| include: false
source("../R/setup.R")
```

What shape does a landscape have, sampled on a grid? `surface` draws a sheet
through your samples. It is a sheet through *three* positions, x, y and z.
Without the third position there is no sheet at all. That makes it the only mark
that *must* be drawn in the cube. Every other mark draws in the plane, and
several of them also draw in the cube once you bind a `z`.

It is not [`area`](area.qmd) with a `z`. An area fills the space between a curve
and a baseline. A surface has no baseline. Both are fills, and that is all they
share.

## A sheet through the samples

A grid of elevations is a table of numbers, 1364 of them here. Nobody can see a
hill in a column of numbers. Draw the same numbers as a sheet and the hill
appears. Here is Maunga Whau, a volcano in Auckland. The book's copy keeps every
second row and every second column of the original grid, so these samples are
20 m apart. Ross Ihaka digitized the elevations from a topographic map, and R's
`datasets` package is where the book reads them [@rcore2026]:

```{r}
#| echo: false
peek(maunga_whau)
```

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("Maunga Whau, at 20 m")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation."*

Three positions and a mark. There is no `space()` in that sentence and none is
needed: binding `z` is what puts a plot in the cube, exactly as it is for a [3-D
scatter](../space.qmd). `space(turn =, tilt =)` still sets the angle when you
want a different one. `z_label()` names the third axis, the way `x_label()`
names the first, and [Space](../space.qmd#z_label-naming-the-third-axis)
explains it.

## One row per crossing

What shape must a table have before a surface can draw it? A scatter's table
does not have that shape, and knowing why tells you which of your tables can
become a sheet. Given two position columns and no transform, a surface reads
each row as a **crossing**: the point where one `x` value meets one `y` value.
The sheet is the faces between the crossings. The grid those faces make is the
sheet's **mesh**. You never declare which crossings are neighbors. gog recovers
that from your two position columns: the distinct `east` values and the distinct
`north` values become the **mesh lines**, one set in each direction. A face is
drawn wherever all four corners of a cell are present.

So this reading needs a table with one row per (x, y) crossing: every `x` value
paired with every `y` value.

Maunga Whau has that shape. It holds 1364 rows, 31 distinct `east` values and 44
distinct `north` values: 31 × 44 = 1364, so the grid is full, and 30 × 43 = 1290
faces get drawn.

The same count shows why a scatter cannot be a surface. The gapminder 2007 table
below holds 142 countries, and no two of them share a `gdp` value or a `life`
value. That makes a grid 142 wide and 142 deep holding only 142 crossings, so no
cell has all four corners. The sheet would be empty:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| error: true
data(gapminder_2007) + surface + x(gdp) + y(life) + z(population)
```

That refusal names the two sentences that *do* draw: a `point` with a `z` for
the cloud itself, and `surface * density` for a **field** estimated from it. A
field is one value at every place on the plane. [The third geometry of one
field](#the-third-geometry-of-one-field) draws `surface * density` for the iris
flowers.

Because a face needs all four of its corners, a gap in the grid is an
**opening** in the sheet rather than a face drawn across it. Drop the crossings
around the summit and the sheet opens there rather than covering the gap:

```{r}
#| message: true
peak <- maunga_whau[!(maunga_whau$east > 300 & maunga_whau$east < 440 &
                        maunga_whau$north > 400 & maunga_whau$north < 560), ]

data(peak) + surface + x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("A gap in the grid, left open")
```

The note above the plot counts the crossings gog did find: 1322 of 1364, so 42
crossings are missing. Each missing crossing removes every face that needed it,
so more than 42 faces are missing: 1234 are drawn here, where the full grid
draws 1290. The note is advice rather than a refusal, because a grid with gaps
is a legitimate thing to have and a sheet with openings is a true picture of
one. What would not be legitimate is quietly spanning a gap, which claims a
height nobody measured. The note also says what to check if you expected a whole
sheet: each position value must repeat exactly. If one row holds 1.2 and another
1.2000001, gog makes two mesh lines where you meant one, and the sheet gets
openings you did not intend.

## A continuous floor

A continent in one column and a year in another looks like a grid too, and a
sheet over it is a reasonable thing to ask for. The two position columns make
the sheet's **floor**, the plane under it, and both have to be continuous. A face
stretches from one sample to the next and so claims every height *between* them,
and between one category and the next there is nothing to claim. So a category
on either position is refused.

Numbers cut into cells are different, and a surface takes those, as [When each
cell holds one value](#when-each-cell-holds-one-value) shows. Cut cells meet
edge to edge. A category instead owns a **slot**: a place on the axis with empty
space on either side. So a surface refuses the slot, not the cut. The flat marks
show the same difference: bars over categories stand apart, and the bars of a
histogram touch.

The refusal, shown in [What it refuses](#what-it-refuses), names the mark that
does take a category on both positions: the [3-D bar](bar.qmd). A 3-D bar
sits *in* its slot and claims nothing about the space between slots. So
`bar * count + x(a) + y(b) + space()` draws one bar over each pair of
categories. A surface cannot draw that grid, because tiles with empty space
between them are not a sheet.

## The third geometry of one field

Which of the two iris clusters is denser, and by how much? A heatmap and a
contour plot, the first two plots below, show where each cluster is. To find how
dense it is, you have to read a legend or a contour level. A surface takes two
transforms that tile a floor, and `density` is the one that makes it the third
way gog draws a field. The same estimate over the same two columns, three times:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + zone * density +
  x(petal_length) + y(sepal_length) + title("As cells: zone")
```

*"Given the iris flowers: zones derived by density, x is petal length, y is
sepal length."*

```{r}
data(iris_flowers) + path * density +
  x(petal_length) + y(sepal_length) + title("As contours: path")
```

*"Given the iris flowers: paths derived by density, x is petal length, y is
sepal length."*

```{r}
data(iris_flowers) + surface * density +
  x(petal_length) + y(sepal_length) + space(turn = 120) +
  title("As a sheet: surface")
```

*"Given the iris flowers: a surface derived by density, x is petal length, y is
sepal length, in the cube, turned 120."*

The sheet is turned to 120 degrees because of where the two peaks are. They sit
on a diagonal of the floor. At the default angle the near peak stands directly
in front of the far one and hides it, so the sheet would show one peak where the
data has two. [One sheet, four angles](#one-sheet-four-angles) returns to that,
because every 3-D view has this problem, not only this one.

All three plots say the same thing about these two clusters: they are nearly
equally dense at their centers. The cluster with long petals is the denser one,
by about four percent. What each plot asks of the reader is different. In the
cells you compare two shades against a legend. In the contours you count rings.
On the sheet you compare two heights, and that needs no legend at all. Which of
the three you want depends on what you are asking, not on which mark is better.

One transform, three marks, and the estimate itself is the same in all three.
What differs is where each mark puts the measurement. A `zone` puts it in color.
A `path` puts it in the level each contour was cut at. A surface puts it in
*height*. That is why the third sentence never binds `z()` and still gets a
third axis, labeled Density. The transform supplies the measurement, and the
mark puts it on the height. A flat `bar * bin` does the same for `y`.

| Mark derived by `density`, over two positions | Where the measurement goes |
|---|---|
| `zone * density` | into the color of each cell |
| `path * density` | into the level each contour was cut at |
| `surface * density` | into the height of the sheet, the third plot above |

Here the transform stays the same, and each mark supplies its own geometry. That
is the rule that already made the contour and the estimated heatmap one feature
instead of two.

The third sentence above writes `space()`, and the chapter's first sentence, for
the volcano, did not. The reason matters. A bound `z` says *there are three
columns here*, which nothing else can mean. An invented height says only *this
layer has a measure of its own*. `density` invents one in the plane too, so
`space()` is what tells gog to put the answer on a third axis.

## When each cell holds one value

The volcano's sheet joins its samples with sloping faces. Does the ground rise
smoothly between your samples, or does each cell hold one value that says
nothing about its neighbors? `bin` is the other transform that tiles a floor. It
cuts the floor into cells, and `mean` gives each cell the average elevation of
the rows inside it. The surface is then flat across each cell, with a step at
each boundary. The density sheet above is drawn the same way, from much smaller
cells. Here is the same volcano, read as a coarse grid rather than a fine one:

```{r}
data(maunga_whau) + surface * bin(12) * mean +
  x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("Maunga Whau, one elevation per cell")
```

*"Given Maunga Whau: a surface derived by bin into 12 and mean, x is east, y is
north, z is elevation."*

Nothing about the volcano changed. What changed is the **claim**. The sheet at
the top of this chapter says the ground rises smoothly between its samples. This
one gives each cell one elevation, the mean of its samples, and says nothing
about how the ground varies inside the cell. Each claim is right for a different
kind of data, and which one you want is a question about how yours was measured.

That question has a clear answer in an experiment or a simulation, where the
conditions form a grid with exactly one number in each cell. Read as crossings,
such a table gives fewer faces than it has values. Nine values make a 3 × 3 grid
of crossings. A face needs four crossings at its corners, so the sheet has
2 × 2 = 4 faces, and no face stands for any one value:

```{r}
cells <- data.frame(
  a = c(-2, 0, 2, -2, 0, 2, -2, 0, 2),
  b = c(-2, -2, -2, 0, 0, 0, 2, 2, 2),
  v = c(0.039, 0.029, 0.039, 0.029, 0.019, 0.029, 0.039, 0.029, 0.039)
)

data(cells) + surface + x(a) + y(b) + z(v) + title("Nine values, four faces")
```

Nine numbers became four faces, and each face slopes between the four values at
its corners. Cut the floor into cells instead, and each cell gets a flat top of
its own, a **plateau**:

```{r}
data(cells) + surface * bin(3) * mean + x(a) + y(b) + z(v) +
  color(v) + palette("plasma") + title("Nine values, nine plateaus")
```

*"Given the cells: a surface derived by bin into 3 and mean, x is a, y is b, z
is v, color by v, with the plasma palette."*

Nine plateaus, and three colors. The nine values make a symmetric bowl: the four
corner cells share the highest value, the four edge cells share the middle one,
and the center cell holds the lowest. The ramp makes that symmetry visible. In
the plain sheet above, you had to read the symmetry from the axis.

A 3-D bar uses the same sentence: `bin` cuts the `x` and `y` axes into cells, and
`mean` averages the column named on `z` inside each one. The transforms mean the
same thing on both marks, and only the geometry laid on the cell differs. Here
`mean` does no averaging, because this table holds one row per cell. It is still
needed: without it, `bin` counts the rows in each cell and draws that count as
the height.

The step between two plateaus is drawn, as a **vertical face** standing on the
line the two cells share. That line has no width, so the vertical face claims
nothing about any area: it is simply what a jump looks like. Without it the
mark would draw disconnected tiles floating at their own heights, and a reader
could not tell a step from a missing cell.

A vertical face spans the *difference* between two neighbors, and a [3-D
bar](bar.qmd) on every cell does not. A bar always reaches down to the baseline,
so a grid of bars hides its own back rows as soon as you turn it. A vertical
face is only as tall as its own step. The next plot turns both to 60 degrees, so
you can see which one is still legible. Both are set to one color, so the only
thing that differs between them is the shape:

```{r}
((data(cells) + bar * bin(3) * mean + x(a) + y(b) + z(v) +
    space(turn = 60) + style(color = "#76b7b2") + title("Bars, turned")) |
   (data(cells) + surface * bin(3) * mean + x(a) + y(b) + z(v) +
      space(turn = 60) + style(color = "#76b7b2") + title("Plateaus, turned"))) +
  theme(height = 430)
```

*"Given the cells: bars derived by bin into 3 and mean, x is a, y is b, z is v,
beside a surface derived by bin into 3 and mean, x is a, y is b, z is v."*

`|` composes them ([Composition](../composition.qmd)), so the two are one
figure and turn together. That is what makes this a fair comparison, rather
than two pictures the reader has to match angle by angle.

`theme(height = )` sets the whole figure's height in pixels. It is set here
because two cubes side by side each get half the page's width but the full
height, so each panel is tall and narrow. A cube keeps its proportions on the
page, so it never stretches to fill a tall panel. Without the setting, blank
space would be left above and below each cube.

Empty cells need no special rule and get none. `bin` leaves them out, because a
cell with no rows is not a measurement of zero. A plateau claims only its own
cell, so a cell with nothing in it is simply a cell with no top. It is the same
opening a missing crossing makes in the mesh, for a different reason.

What a surface refuses is a grid of slots, the cells two categories make. [What
it refuses](#what-it-refuses) shows that refusal.

## The mesh, and what the shading means

A smooth sheet hides where its samples were, the way a line hides its rows.
Drawing the mesh lines shows the grid again, so a reader can see how fine the
sampling was. `border_color` and `border_size` draw them. A face is a closed
fill like a bar's or a box's, so the [settable
rule](../design-laws.qmd#the-settable-rule) gives it a border. On a surface,
those borders are the mesh lines:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  style(border_color = "white", border_size = 0.4) +
  z_label("Elevation (m)") + title("The mesh, drawn")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, with
border color white and border size 0.4."*

The mesh lines add no shapes to the picture, because every face is already
stroked in its own shade. That thin stroke covers the pale seam that would
otherwise show between neighboring faces. `border_color` replaces the stroke's
color, and `border_size` its width. Leave both out for a seamless sheet. A
`border_size` of 0 removes the stroke, and the pale seams appear.

Look at the shading on the sheet above, and on the turned one below it. The faces
are shaded by their **slope**: a steep one is darker, a flat one keeps its full
color. That is a fact about the terrain, not about where a light is, so turning
the scene rearranges the faces on the page and repaints none of them:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  space(turn = 130, tilt = 35) +
  z_label("Elevation (m)") + title("Turned, with the same shades")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, turned
130, tilted 35."*

Compare the crater walls here with the first plot. Same faces, same shades, seen
from the other side. A light that stayed in one place on the screen would have
changed the shade of every face. A color that changed while the data did not
would mislead a reader, with no way to notice.

## Color by height, per face

Slope shading says how steep a face is, not how high it stands. How high is the
summit, and how deep is the crater? `color` maps on a surface, and a
*continuous* column colors it face by face along a ramp.
[`palette()`](../style.qmd#palette-choosing-the-colors-a-mapping-uses) chooses
which ramp. `"viridis"` suits a topographic sheet, because it gets steadily
lighter from its lowest color to its highest:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) + color(elevation) +
  palette("viridis") +
  z_label("Elevation (m)") + title("Height twice: as a position and as a ramp")
```

Compare that with the first plot in this chapter. The summit and the crater rim
are in that plot too. With slope shading alone you have to infer them from which
faces are dark. The ramp states the heights, and the legend decodes them.

Mapping the height to color when `z` already shows it is redundant on purpose.
It is what a reader of a topographic map expects. It also shows why a surface
can take a continuous color when an [`area`](area.qmd) cannot. An area has one
interior, and coloring it by a measure would make a gradient fill, which is a
different and much larger job. A surface has *faces*, and a face is small enough
to hold one value.

A face has four corners and one color. Its color is the mean of its four corner
values, which stands for the field at its center. So the ramp describes the mesh
you have, not only the field you sampled from. Here is one bowl-shaped field,
sampled coarsely and then finely:

```{r}
bowl <- function(n) {
  g <- expand.grid(a = seq(-2, 2, length.out = n), b = seq(-2, 2, length.out = n))
  g$v <- 0.019 + 0.0025 * (g$a^2 + g$b^2)
  g
}
coarse <- bowl(3)
fine   <- bowl(7)

((data(coarse) + surface + x(a) + y(b) + z(v) +
    color(v) + palette("plasma") + title("Sampled 3 by 3: four faces")) |
   (data(fine) + surface + x(a) + y(b) + z(v) +
      color(v) + palette("plasma") + title("Sampled 7 by 7: thirty-six faces"))) +
  theme(height = 320)
```

*"Given coarse: a surface, x is a, y is b, z is v, color by v, with the plasma
palette, beside the same given fine."*

Two tables share this page, and `data()` keeps each table's name, `coarse` or
`fine`. gog finds each layer's table by that name, so it refuses two different
tables under one name ([Data](../data.qmd)).

The left sheet is one uniform color, and that is correct rather than broken.
Every face has one high corner, one low corner and two middle ones, so all four
faces get the same average. The mesh is too coarse to hold the shape, and the
ramp says so instead of inventing a gradient across it. The right sheet has the
same field under it and enough faces to resolve it, so the bowl appears.

This is the reading that makes the legend trustworthy. A color is the height at
one stated place, the center of a face, so two faces of one color really are at
one height there, and a sheet of one color has all its face centers at one
height.

The slope shading is still underneath the ramp, and the two combine rather than
cancel: a face takes its color from the ramp and is then darkened by how steep
it is. Viridis stays readable because only the *low* ground takes its dark
colors, and low ground is usually flat. A flat face is barely darkened by its
slope, so few faces are darkened by both the ramp and the slope.

An estimated height takes a ramp the same way, and you name it the same way.
`density` writes its estimate into a column called `density`, so `density` is
the name you give `color`, exactly as on a [`zone`](zone.qmd) or a
[`path`](path.qmd):

```{r}
data(iris_flowers) + surface * density +
  x(petal_length) + y(sepal_length) + space() + color(density) +
  palette("viridis") + title("The estimate, as a height and as a ramp")
```

Nothing there is a special case. The transform makes a column, the column has a
name, and a name can be bound. Whether the height came from your table or from
an estimate is only a difference in source. It does not change what you may say
about it.

Whether you *want* the color as well as the height is a separate question. The
plain sheet in [A sheet through the samples](#a-sheet-through-the-samples)
already shows the height by position alone. Use the ramp when the low ground
matters. Slope shading darkens a steep face but not a flat one, so a broad, low
mound and the flat ground around it are drawn almost the same color. The ramp is
what separates them.

A *categorical* color splits the mark instead, one sheet per group. gog puts
every face of both sheets into one depth order, the back-to-front order in which
faces are drawn, rather than drawing one sheet and then the other:

```{r}
band <- maunga_whau
band$side <- ifelse(band$north > 440, "north slope", "south slope")

data(band) + surface + x(east) + y(north) + z(elevation) + color(side) +
  z_label("Elevation (m)") + title("Two sheets, sorted together")
```

The two sheets meet along the line where `side` changes. Depth order runs across
that line rather than stopping at it. Each face is placed by its own distance
from you, whichever sheet it belongs to, so the two sheets are correct wherever
they overlap. Drawing one whole sheet and then the other would put every face of
the second in front, and the line where they meet would be drawn wrong.

## Opacity by a measure, per face

A color decides what a face says. An opacity decides whether you can see
through it. Part of a sheet is often not the subject. The flat ground around a
hill only hides what is behind it, and fading those faces makes the hill the
solid part of the picture. `opacity` maps on a surface, and it maps for the
reason `color` does. A face is small enough to hold one value, so it can hold an
opacity too. Every face takes its own opacity, and a face fades where its
measure is small:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) + opacity(elevation) +
  z_label("Elevation (m)") + title("Faint where the ground is low")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, opacity
by elevation."*

Over the low ground the fill fades to almost nothing and each face keeps its
outline, so that part of the sheet shows as mesh lines alone. A ramp cannot do
that; an opacity can. The legend beside the plot reads the same way `color`'s
does, with the smallest, middle and largest values of the column.

No face disappears completely. A mapped `opacity` runs from 0.15 to 0.95, so the
faintest face here is still drawn, which is the rule every mapped `opacity`
follows ([Channels](../channels.qmd)).

A face reads its opacity the way it reads its color, from the mean of its four
corners. A plateau reads its own cell's row. A vertical face stands between two
cells and reads the taller one's row, the way a bar's side takes the bar's
color. Either way one face has one reading, whichever channel asks for it.

The sentence behind the first plot in this section binds `elevation` twice: once
to `z` and once to `opacity`. A second column is the more useful case: the shape
stays in `z`, and another measure decides which faces fade.

```{r}
relief <- maunga_whau
h <- matrix(relief$elevation, nrow = length(unique(relief$north)))
up   <- rbind(h[-1, ], h[nrow(h), ])
side <- cbind(h[, -1], h[, ncol(h)])
relief$steepness <- as.vector(sqrt((up - h)^2 + (side - h)^2))

data(relief) + surface + x(east) + y(north) + z(elevation) + opacity(steepness) +
  z_label("Elevation (m)") + title("Solid where the ground is steep")
```

`steepness` combines the two elevation changes at each crossing, one toward the
next east value and one toward the next north value. The steep sides of the hill
stay solid, and the flat ground around them fades away. The sheet is the same
shape as the one above it, and a different part of it is solid.

An [`area`](area.qmd) refuses a mapped opacity for the same reason it refuses a
continuous color. A region has one interior and one fill, so a row there is a
corner of the outline rather than a part that can carry a value. A surface has
faces, and each face can carry both.

One value for the whole sheet is still [`style(opacity = )`](../style.qmd). Set
it when you want to see one sheet through another, and map it when you want each
face to fade by its own value.

## One sheet, four angles

A still 3-D plot hides something at every angle, and which thing it hides
depends on the angle you pick. So turn the same sheet rather than trust one view
of it. Each of these changes exactly *one* of `turn`/`tilt` from the first:

```{r}
((data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
    color(elevation) + palette("viridis") + space(turn = 30, tilt = 25) +
    z_label("Elevation (m)") + title("The reference three-quarter view")) |
 (data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
    color(elevation) + palette("viridis") + space(turn = 140, tilt = 25) +
    z_label("Elevation (m)") + title("Turned: the same sheet from the west"))) /
((data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
    color(elevation) + palette("viridis") + space(turn = 30, tilt = 70) +
    z_label("Elevation (m)") + title("Tilted up: the crater appears")) |
 (data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
    color(elevation) + palette("viridis") + space(turn = 30, tilt = 5) +
    z_label("Elevation (m)") + title("Tilted down: the view from the side")))
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, color by
elevation, with the viridis palette, turned 30, tilted 25, beside the same
turned 140; above the same tilted 70, beside the same tilted 5."*

Read them as a set and you can see what each angle hides. The crater is the most
interesting thing about this hill, and the first two plots do not show it. The
rim nearest you stands in front of the crater, so turning alone does not show it
while the tilt is low. Tilt up and you can see into the crater, because you are
now looking down from above. The same tilt flattens the heights, until the sheet
looks like a flat colored map. Tilt down instead and you get the opposite: the
heights read clearly, the view from above is gone, and the sheet shows as an
outline. One or the other, never both.

These four plots show why you would want to turn a plot. Here you need both the
view from above and the view from the side, and no single still view shows this
hill whole. So turn the four with the pointer. Raise the tilt until the crater
opens, and watch the heights flatten as it does. The four still views show the
problem; dragging them lets you check it yourself.

All four turn together, because `|` and `/` composed them into one figure. Each
panel keeps the angle its own sentence asked for, and the drag adds the same
change to every one. *Reset* returns each to where it began. To turn one alone,
write it alone.

That is composition rather than faceting, and the two share the same pair of
operators. A facet names a *column* ([Faceting](../faceting.qmd)), and a viewing
angle is not a column: it is a property of the coordinate space, the way a
`polar()` start angle is. gog does build small multiples of cubes, one cube for
each value of a column, but these four panels are not small multiples.
[Space](../space.qmd#faceting-one-cube-per-panel) has that example.

One limit shows in these four plots, and it belongs to the cube rather than to
the mark. Notice the Elevation axis in the tilted-up panel. You are looking down
along it, so it is drawn at about a third of its full length. gog labels three
heights there where the reference view labels five, because it drops numbers
rather than let them overlap. [Space](../space.qmd) says which numbers a crowded
axis drops, and why the scale itself never changes.

## What a surface is for

A surface is for anything that is a *height over a plane*. Terrain, and the sea
floor. A response surface from a designed experiment. A model's fit measured
over two of its parameters. A model's predictions over a grid of two predictors.
And, through `density`, the shape of a two-dimensional distribution.

What it is not for is data that merely *has* three numbers per row. That is a
[3-D scatter](../space.qmd), and the refusal in [One row per
crossing](#one-row-per-crossing) names it for you. The question to ask is
whether the space between two of your samples means anything. On a grid it
does; in a cloud it does not.

## What you can set

No column says whether the mesh lines show, or how far you can see through a
sheet. A setting decides both without reading any column, and each mark takes
its own settings. These are a surface's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("surface")
```

A surface is a closed fill, so it takes what a [`bar`](bar.qmd) and a
[`box`](box.qmd) take: a color, an opacity, and a border. `pattern` is the one
fill setting it refuses, and like a `bar` it has no `size`. The next section
says why for both.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

Every flat mark is written with only `x` and `y` as positions, so it is easy to
write a surface that way by mistake. Leaving out the height and leaving the plot
in the plane are one mistake, not two, so there is one refusal. It names both
ways into the cube, and it names the two marks that draw a field in the plane
instead:

```{r}
#| error: true
data(maunga_whau) + surface + x(east) + y(north)
```

A category on either position is the second refusal, for the reason [A
continuous floor](#a-continuous-floor) gave. A face stretches from one sample to
the next, and between two categories there is nothing for it to cross:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
#| error: true
data(gm_eras) + surface + x(continent) + y(year) + z(life)
```

The message names the mark that draws one bar in each category's slot instead,
the [3-D bar](bar.qmd). The transforms that tally into slots are refused by the
same rule. `count` and `proportion` count rows into the cells two categories
make:

```{r}
#| error: true
data(gm_eras) + surface * count + x(continent) + y(era) + space()
```

That refusal lists the two transforms that do tile a floor, `bin` and
`density`.

`density(levels = )` cuts a field into level *sets*, and a level set is a region
in the plane. A surface has already given the third axis to the measurement, so
a band drawn on it could only be a color. [`zone`](zone.qmd) already draws
those bands in the plane, as colored regions, so the sentence is refused:

```{r}
#| error: true
data(iris_flowers) + surface * density(levels = 5) +
  x(petal_length) + y(sepal_length) + space()
```

A stepped sheet can be cut on either of two axes, and only one of the two draws.
Cutting the *floor* into cells is `surface * bin`, which [When each cell holds
one value](#when-each-cell-holds-one-value) draws. Cutting the *height* into
bands is what `levels` asks for, and that one is refused. It would cut the
measurement rather than the grid, so the plateaus would be contours of equal
height rather than one per cell.

`bandwidth` is refused for the reason it is refused on every two-dimensional
field. It is a length in one column's own units, and a field spreads over two
columns that measure different quantities. One number cannot be a width in both.
Use `density(adjust = )`, which is dimensionless and scales both.

There is no `size` on a surface. Its extent is the grid it is drawn on, so a
size would be a second and contradictory answer to how big it is.
[`area`](area.qmd) refuses a size for the same reason. And a surface is the one
fill mark that refuses `pattern`:

```{r}
#| error: true
data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  style(pattern = "hatch")
```

A hatch tile is a texture in *screen* space, and every face of a projected mesh
is foreshortened differently, so one tile size would read as a different line
spacing on every face. The texture would then follow the viewing angle rather
than the data, which is the one thing this mark's shading is built to avoid.

## In short

A surface stretches between neighboring samples, so it needs a floor whose cells
tile. A scatter's table does not give it one. A face needs all four of its
corners, so a missing crossing leaves an opening rather than a guessed height.

## What comes next

A surface needs two continuous positions under it. [Edge](edge.qmd) is the mark
for tables with no positions at all: rows naming two things and saying they are
connected.
