# Surface

```{r}
#| include: false
source("../R/setup.R")
```

What shape does a landscape have, sampled on a grid? `surface` draws a sheet
through your samples. It is the last mark in the kernel to gain a renderer, and
the only one that draws in the cube alone. Every other mark was built in the
plane, and some have since learned to stand up. A surface is a sheet through
*three* positions, and there is no such thing without the third.

It is not [`area`](area.qmd) with a `z`. An area fills between a curve and a
baseline, and a surface has no baseline to close on; the family resemblance is
that both are fills, and it stops there.

## A sheet through the samples

Here is Maunga Whau, the volcano in Auckland that ships with R, on a 20 m grid.
Ross Ihaka digitized the elevations from a topographic map, and they reach this
page through R's `datasets` package [@rcore2026]:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("Maunga Whau, at 20 m")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation."*

Three positions and a mark. There is no `space()` in that sentence and none is
needed: binding `z` is what puts a plot in the cube, exactly as it is for a
[3-D scatter](../space.qmd). `space(turn =, tilt =)` still sets the angle when you
want a different one.

## One row per crossing

Handed two bare position columns, a surface reads its rows as **nodes**, and the
sheet is the faces between them. You never declare which nodes are neighbors.
GOG recovers that from your two position columns: the distinct `east` values
become the mesh's columns and the distinct `north` values its rows. A face is
drawn wherever all four corners of a cell are present.

So the table this reading wants is one row per (x, y) crossing, which is the shape
`expand.grid()` makes:

```{r}
head(maunga_whau, 4)
nrow(maunga_whau)
```

1364 rows, 31 distinct eastings and 44 distinct northings: 31 × 44 = 1364, so the
grid is full, and 30 × 43 = 1290 faces get drawn. That arithmetic is the whole
contract, and it is why a **scatter** cannot be a surface. Sixty points in general
position describe a 60 × 60 lattice with sixty nodes in it and not one complete
block of four, so the sheet would come out empty:

```{r}
#| error: true
data(gapminder_2007) + surface + x(gdp) + y(life) + z(population)
```

That refusal names the two sentences that *do* draw, and the second of them is the
interesting one — read on.

Because a face needs all four of its corners, a gap in the grid is an **opening**
in the sheet rather than a face drawn across it. Drop the crossings around the
summit and the mesh tears rather than inventing a lid:

```{r}
#| message: true
peak <- maunga_whau[!(maunga_whau$east > 300 & maunga_whau$east < 440 &
                        maunga_whau$north > 400 & maunga_whau$north < 560), ]

data(peak) + surface + x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("A hole is a hole")
```

The note above the plot is GOG counting the crossings it did not find: 1322 of
1364, so 42 are open. It is advice rather than a refusal, because a grid with gaps
is a legitimate thing to have and a torn sheet is a true picture of one. What would
not be legitimate is quietly spanning a gap, which asserts a height nobody
measured. The note is also the thing that tells you when the gaps were an accident
— a lattice recovered from coordinates that were rounded rather than repeated has
holes everywhere, and looks exactly like a mesh with a designed opening.

## The floor is continuous, and a bar's is not

Both position columns have to be numbers. A face spans the gap between two
samples and so asserts every height *inside* that gap, and between one category
and the next there is no inside to assert anything about:

```{r}
#| error: true
data(gm_eras) + surface + x(continent) + y(year) + z(life)
```

The direction there points at the mark whose floor *may* be slotted. A
[3-D bar](bar.qmd) stands *in* its cell and claims nothing about the space between
cells, so `bar * count + x(a) + y(b) + space()` is a mesh over categories and a
surface is not. A category owns a slot with air on either side of it, and tiles
that float apart are not a sheet.

Numbers cut into cells are a different matter, and a surface takes those: the
section on one value per cell, below. So the two marks part on the **slot** rather
than on the cut, which is the same cut-versus-slot distinction the flat marks make,
read one dimension up.

## The third geometry of one field

A surface takes two transforms, and `density` is the one that makes it the third
way GOG draws a field. The same estimate over the same two columns, three times:

```{r}
data(iris_flowers) + zone * density +
  x(petal_length) + y(sepal_length) + title("As cells: zone")
```

```{r}
data(iris_flowers) + path * density +
  x(petal_length) + y(sepal_length) + title("As contours: path")
```

```{r}
data(iris_flowers) + surface * density +
  x(petal_length) + y(sepal_length) + space() +
  title("As a sheet: surface")
```

Look at what the third one shows that the other two cannot: *how much* higher the
tight cluster is than the broad one. Cells and contours both spend their
information on *where* the field is, and a height spends it on *how much* — the
two setosa-versus-the-rest blobs are equally visible in the first two plots and
wildly unequal in the third. Which of the three you want is a question about your
question, not about which is better.

One transform, three marks, and the estimate itself is the same in all three.
What differs is where the measurement goes: each mark puts it where its own
geometry measures. A `zone` measures by color. A `path` measures by the level the
contour was cut at. A surface measures by **height**. That is why the third
sentence never binds `z()` and still gets an elevation axis: the transform
supplies the measurement, and the mark knows where it belongs. A flat `bar * bin`
does the same for `y`.

Here the transform stays the same and the mark chooses the geometry. That is the
rule that already made the contour and the estimated heatmap one feature instead
of two.

`space()` is in that sentence and not in the volcano's, for a reason worth
knowing. A bound `z` says *there are three columns here*, which nothing else can
mean. A synthesized one says only *this layer invents a measure*. Since
`density` invents one in the plane too, it needs the coordinate to say where the
answer goes.

## When each cell holds one value

`bin` is the other transform, and it draws a different sheet: flat across each
cell, stepping at the boundary. Here is the same volcano, read as a coarse grid
rather than a fine one:

```{r}
data(maunga_whau) + surface * bin(12) * mean +
  x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("Maunga Whau, one elevation per cell")
```

Nothing about the mountain changed. What changed is the **claim**. The sheet at the
top of this chapter says the ground rises smoothly between its samples; this one
says each cell has a single elevation and nothing is asserted about the ground
inside it. Both are honest about different data, and which you want is a question
about how yours was measured.

That question has a sharp answer whenever a design produces exactly one value per
cell, which is the ordinary case in an experiment or a simulation: a grid of
conditions, one number in each. Read as nodes, such a table loses most of itself.
Nine values on a 3 × 3 grid make 2 × 2 blocks of four corners, so the sheet is four
faces, and it spans only from the first sample to the last instead of covering the
grid:

```{r}
cells <- expand.grid(a = c(-2, 0, 2), b = c(-2, 0, 2))
cells$v <- 0.019 + 0.0025 * (cells$a^2 + cells$b^2)

data(cells) + surface + x(a) + y(b) + z(v) + title("Nine values, four faces")
```

Five of those nine numbers survive only as corners. Cut the floor instead, and each
one gets a plateau of its own:

```{r}
data(cells) + surface * bin(3) * mean + x(a) + y(b) + z(v) +
  title("Nine values, nine plateaus")
```

`bin` cuts the two floor axes into cells, and `mean` reduces the column named on
`z` inside each one. That is the sentence a 3-D bar already used. The transforms
mean the same thing on both marks, and only the geometry laid on the cell differs.
Where a table holds one row per cell, as this one does, the mean of a single value
is that value, and the transform is doing nothing but naming the column.

The step between two plateaus is drawn, as a **riser** standing on the line the two
cells share. That line has no width, so the riser asserts nothing about any area:
it is simply what a jump looks like. Without it the mark would draw disconnected
tiles floating at their own heights, and a reader could not tell a step from a
missing cell.

A riser spans the *difference* between two neighbors, and that is what separates
this from standing a [3-D bar](bar.qmd) on every cell. A bar's wall runs the whole
way from the baseline, so a floor of them walls itself off as soon as you turn it,
while risers are only as tall as the local step. Turn both and watch which survives:

```{r}
#| layout-ncol: 2
data(cells) + bar * bin(3) * mean + x(a) + y(b) + z(v) +
  space(turn = 60) + title("Columns, turned")

data(cells) + surface * bin(3) * mean + x(a) + y(b) + z(v) +
  space(turn = 60) + title("Plateaus, turned")
```

Empty cells need no special rule and get none. `bin` leaves them out, because a
count of zero and no data at all are different things, and a plateau claims only
its own cell, so a cell with nothing in it is simply a cell with no lid. That is
the same opening a missing crossing makes in a node mesh, arrived at from the
other direction.

What a surface still refuses is a floor of **slots**. `count` and `proportion`
tally into the cells two categories make, and categories leave air:

```{r}
#| error: true
data(gm_eras) + surface * count + x(continent) + y(era) + space()
```

## The mesh, and what the shading means

`border_color` and `border_size` are the **mesh lines**, and they are the reading
that made a rim worth having on this mark. A face is a closed fill like a bar's or
a box's, so the settable rule hands it a border; on a mesh, that border is the
wireframe:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  style(border_color = "white", border_size = 0.4) +
  z_label("Elevation (m)") + title("The mesh, drawn")
```

It costs nothing to draw, because every face was already stroked in its own
shade — the hairline that closes the pale seam antialiasing leaves between two
abutting polygons. `border_color` just replaces that color. Set
`border_size = 0` for a seamless sheet.

Notice what the shading is doing in both plots, and what it is *not*. The faces
are shaded by their **slope**: a steep one is darker, a level one keeps the color
whole. That is a fact about the terrain, not about where a light is, so turning the
scene rearranges the faces on the page and repaints none of them:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  space(turn = 130, tilt = 35) +
  z_label("Elevation (m)") + title("Turned, not relit")
```

Compare the crater walls here with the first plot. Same faces, same shades, seen
from the other side. A lamp fixed to the page would have repainted every one of
them, and a hue that moves while the data sits still is the kind of wrongness a
reader cannot see.

## Color by height, per face

`color` maps on a surface, and a **measured** column ramps it face by face.
[`palette()`](../style.qmd) chooses which ramp, and `"viridis"` is the one a
topographic sheet wants — it climbs in lightness the whole way, so the height
reads off the hue even where a face is turned away from you:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) + color(elevation) +
  palette("viridis") +
  z_label("Elevation (m)") + title("Height twice: as a position and as a ramp")
```

Compare that with the first plot in this chapter. The summit and the crater rim
were there all along, and slope shading alone left them to be inferred from which
faces were dark; the ramp says the heights out loud, and the key decodes them.

Mapping the height to color when it is already the height is redundant on
purpose, and it is what a reader of a topographic map expects, so the ramp earns
its place. It also marks a real boundary. An [`area`](area.qmd) refuses a measured
color, because a region has one interior and coloring an interior by a measure is
a gradient fill, which is a different and much larger job. A mesh has *faces*, and
a face is already small enough to hold one value. So the rule was never "regions
refuse it, strokes take it" — it is whether a mark has parts small enough to each
hold one value, and a stroke has segments where a surface has faces.

The slope shading is still underneath the ramp, and the two compose rather than
fight: a face takes its color from the ramp and is then darkened by how steep it
is. Viridis survives that because only the *low* ground reaches its dark end, and
on a terrain the low ground is also the flattest. The two darkenings therefore do
not pile up, and the darkest face on this sheet is lighter than a fully-shaded
viridis floor would have been.

A **categorical** color splits the mark instead, one sheet per group, and the
depth sort runs over every face of every series at once so two sheets thread
through each other rather than one landing wholly in front:

```{r}
band <- maunga_whau
band$side <- ifelse(band$north > 440, "north slope", "south slope")

data(band) + surface + x(east) + y(north) + z(elevation) + color(side) +
  z_label("Elevation (m)") + title("Two sheets, sorted together")
```

## One sheet, four angles

A still 3-D plot hides something at every angle, and which thing it hides is the
angle's choice. So it is worth turning the same sheet rather than trusting one view
of it. Each of these changes exactly *one* of `turn`/`tilt` from the first:

```{r}
#| layout-ncol: 2
data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  color(elevation) + palette("viridis") + space(turn = 30, tilt = 25) +
  z_label("Elevation (m)") + title("The reference three-quarter view")

data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  color(elevation) + palette("viridis") + space(turn = 140, tilt = 25) +
  z_label("Elevation (m)") + title("Turned: the same sheet from the west")

data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  color(elevation) + palette("viridis") + space(turn = 30, tilt = 70) +
  z_label("Elevation (m)") + title("Tilted up: the crater appears")

data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  color(elevation) + palette("viridis") + space(turn = 30, tilt = 5) +
  z_label("Elevation (m)") + title("Tilted down: the skyline")
```

Read them as a set and the trade is plain. **The crater is not visible at all in
the first two**, and it is the most interesting thing about this hill: the summit
stands in front of the hollow behind it, so no amount of turning at a low eye level
reveals it. Tilt the eye up and the hollow opens, because you are now reading the
*plan* — and by the same move the heights flatten, until the sheet is a colored map
of itself. Tilt down instead and you get the opposite bargain: the profile is exact,
the plan is gone, and the sheet reads as a skyline. Heights or plan, one at a time.

That is the argument for a live canvas made rather than asserted. A plot you can
drag is not a convenience here; it is the only way to have both readings of a shape
at once, which is why the interactive orbit is the next real piece of this
milestone and not a polish item.

Two honest edges in this tour, and neither is a surface's fault. **The tiling is the
page's, not the grammar's:** those are four sentences that Quarto laid out two
across, not one faceted plot. `|` and `/` are faceting, faceting names a
**column** ([Faceting](../faceting.qmd)), and a viewing angle is not a column: it
is a property of the coordinate space, the way a `polar()` start angle is. A small
multiple of cubes *is* built, and these four panels still are not one
([Space](../space.qmd) shows the real thing), because a facet over view angles
would need a column to name and there is none. And notice the vertical axis in
the tilted-up panel. Its numbers sit tight against each other, because that axis
is projecting at about a third of its length and they are still all being drawn.
They do clear each other, and tilting further would make the frame leave some out
rather than overlap them. Tight is simply what a foreshortened axis looks like.
[Space](../space.qmd) says what the frame does about it, and what it deliberately
does not do to the scale.

## What a surface is for

Anything that is a *height over a plane*. Terrain and bathymetry. A response
surface from a designed experiment. A likelihood or loss over two parameters. A
fitted model's predictions over a grid of two predictors. And, through
`density`, the shape of a two-dimensional distribution.

What it is not for is data that merely *has* three numbers per row. That is a
[3-D scatter](../space.qmd), and the refusal above will send you there. The
question to ask is whether the space between two of your samples means anything.
On a grid it does; in a cloud it does not.

## What you can set

Everything a surface can be told to look like, generated from the engine's own
rule table so this page cannot drift from what `style()` actually accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("surface")
```

A surface is a fill, so it takes a fill's settings — with two absences worth
noting, both of which the refusals below explain: no `size`, and no `pattern`.

## What it refuses

A surface with no height is the same mistake as a surface in the plane, so it is
one refusal rather than two, and it names both routes into the cube along with the
mark that draws a field in the plane:

```{r}
#| error: true
data(maunga_whau) + surface + x(east) + y(north)
```

`density(levels = )` cuts a field into level *sets*, and a level set is a region in
the plane. A surface has already spent the third axis on the measurement, so a
band drawn on it could only be a color — which is [`zone`](zone.qmd)'s reading of
that same request:

```{r}
#| error: true
data(iris_flowers) + surface * density(levels = 5) +
  x(petal_length) + y(sepal_length) + space()
```

A stepped sheet is a real thing, and there are two of them, cut on different axes.
Cutting the **floor** into cells is `surface * bin`, the section on one value per
cell above, and it draws. Cutting the **height** into bands is what `levels` asks
for, and that one is not drawn: it would quantize the measurement rather than the
grid, so the plateaus would be contours of equal height rather than one per cell.

`bandwidth` is refused for the reason it is refused on every two-dimensional
field: it is a length in one column's own units, and a field spreads over two
columns measuring different quantities, so one number cannot be a width in both.
Use `density(adjust = )`, which is dimensionless and scales both.

There is no `size` on a surface — its extent *is* its lattice, so a size would be
a second, contradictory answer to how big it is, which is
[`area`](area.qmd)'s refusal for `area`'s reason. And `pattern` is refused, which
is the one place a fill does not take a texture:

```{r}
#| error: true
data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  style(pattern = "hatch")
```

A hatch tile is a texture in *screen* space, and every face of a projected mesh is
foreshortened differently, so one tile size would read as a different density on
every face. That is a texture varying with the viewing angle instead of with the
data, which is exactly the light this mark's shading is defined not to be.
