# Surface

```{r}
#| include: false
source("../R/setup.R")
```

What shape does a landscape have, sampled on a grid? `surface` draws a sheet
through your samples. It is the only mark that can be drawn in the cube by
itself. Every other mark draws in the plane, and several of them also draw in
space. A surface is a sheet through *three* positions, and there is no such
thing without the third.

It is not [`area`](area.qmd) with a `z`. An area fills the space between a curve
and a baseline. A surface has no baseline. Both are fills, and that is all they
share.

## A sheet through the samples

A grid of elevations is a table of numbers, 1364 of them here. No reader can
see a hill in a column of numbers, and drawn as a sheet the hill appears.
Here is Maunga Whau, a volcano in Auckland. The book's copy keeps every second
row and column, a 20 m grid. Ross Ihaka digitized the elevations from a
topographic map, and they reach this page through R's `datasets` package
[@rcore2026]:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("Maunga Whau, at 20 m")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation."*

Three positions and a mark. There is no `space()` in that sentence and none is
needed: binding `z` is what puts a plot in the cube, exactly as it is for a [3-D
scatter](../space.qmd). `space(turn =, tilt =)` still sets the angle when you
want a different one.

## One row per crossing

What shape must a table have before a surface can draw it? A scatter's table
does not have it, and the reason decides which of your tables can be a sheet.
Given two position columns and no transform, a surface reads its rows as
**nodes**, and the sheet is the faces between them. You never declare which
nodes are neighbors. gog recovers that from your two position columns: the
distinct `east` values become the mesh's vertical grid lines and the distinct
`north` values its horizontal ones. A face is drawn wherever all four corners of
a cell are present.

So the table this reading wants is one row per (x, y) crossing: every `x` value
paired with every `y` value.

```{r}
head(maunga_whau, 4)
nrow(maunga_whau)
```

1364 rows, 31 distinct `east` values and 44 distinct `north` values: 31 × 44 =
1364, so the grid is full, and 30 × 43 = 1290 faces get drawn. That arithmetic
is the whole rule, and it is why a scatter cannot be a surface. Each of the 142
countries has its own `gdp` and its own `life`. That makes a grid 142 wide and
142 deep holding only 142 nodes, so no cell has all four corners. The sheet
would be empty:

```{r}
#| error: true
data(gapminder_2007) + surface + x(gdp) + y(life) + z(population)
```

That refusal names the two sentences that *do* draw. The section on the third
geometry of one field draws the second one.

Because a face needs all four of its corners, a gap in the grid is an
**opening** in the sheet rather than a face drawn across it. Drop the crossings
around the summit and the sheet opens there rather than covering the gap:

```{r}
#| message: true
peak <- maunga_whau[!(maunga_whau$east > 300 & maunga_whau$east < 440 &
                        maunga_whau$north > 400 & maunga_whau$north < 560), ]

data(peak) + surface + x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("A gap in the grid, left open")
```

The note above the plot counts the crossings gog did find: 1322 of 1364, so 42
are open. It is advice rather than a refusal, because a grid with gaps is a
legitimate thing to have and a torn sheet is a true picture of one. What would
not be legitimate is quietly spanning a gap, which asserts a height nobody
measured. The note also tells you when the gaps were an accident. If the two
position columns hold rounded values instead of exact repeats, the lattice fills
with holes nobody designed.

## A continuous floor

A continent in one column and a year in another looks like a grid too. A sheet
over it is a natural thing to ask for.
Both position columns have to be numbers. A face spans the gap between two
samples and so asserts every height *inside* that gap, and between one category
and the next there is no inside to assert anything about:

```{r}
#| error: true
data(gm_eras) + surface + x(continent) + y(year) + z(life)
```

The direction there points at the other mark. A category owns a slot with air on
either side of it, and a [3-D bar](bar.qmd) stands *in* its slot, claiming
nothing about the space between cells. So `bar * count + x(a) + y(b) + space()`
stands one bar over each pair of categories, and a surface cannot: tiles that
float apart are not a sheet.

Numbers cut into cells are a different matter, and a surface takes those: the
section on one value per cell, below. So the two marks differ over the **slot**,
not over the cut. That is the same distinction the flat marks make between a
category and a cut number, one dimension higher.

## The third geometry of one field

Which of the two iris clusters is denser, and by how much? The heatmap and the
contour show where each cluster is, and leave its height to a legend or a
level.
A surface takes two transforms, and `density` is the one that makes it the third
way gog draws a field. The same estimate over the same two columns, three times:

```{r}
data(iris_flowers) + zone * density +
  x(petal_length) + y(sepal_length) + title("As cells: zone")
```

*"Given the iris flowers: zones derived by density, x is petal length, y is
sepal length."*

```{r}
data(iris_flowers) + path * density +
  x(petal_length) + y(sepal_length) + title("As contours: path")
```

*"Given the iris flowers: paths derived by density, x is petal length, y is
sepal length."*

```{r}
data(iris_flowers) + surface * density +
  x(petal_length) + y(sepal_length) + space() +
  title("As a sheet: surface")
```

*"Given the iris flowers: a surface derived by density, x is petal length, y is
sepal length, in the cube."*

Look at what the third one shows that the other two cannot: *how much* higher
the tight cluster is than the broad one. Cells and contours show *where* the
field is, and a height shows *how much*. The two clusters look equally large in
the first two plots and very unequal in the third. Which of the three you want
depends on what you are asking, not on which mark is better.

One transform, three marks, and the estimate itself is the same in all three.
What differs is where the measurement goes: each mark puts it where its own
geometry measures. A `zone` measures by color. A `path` measures by the level
the contour was cut at. A surface measures by **height**. That is why the third
sentence never binds `z()` and still gets a third axis, labeled Density: the
transform supplies the measurement, and the mark puts it on the height. A flat
`bar * bin` does the same for `y`.

Here the transform stays the same and the mark chooses the geometry. That is the
rule that already made the contour and the estimated heatmap one feature instead
of two.

`space()` is in that sentence and not in the volcano's, and the reason matters.
A bound `z` says *there are three columns here*, which nothing else can mean. An
invented height says only *this layer has a measure of its own*. `density`
invents one in the plane too, so `space()` is what tells gog to put the answer
on a third axis.

## When each cell holds one value

Does the ground rise smoothly between your samples, or does each cell hold one
value that says nothing about its neighbors?
`bin` is the other transform, and it draws a different sheet: flat across each
cell, stepping at the boundary. Here is the same volcano, read as a coarse grid
rather than a fine one:

```{r}
data(maunga_whau) + surface * bin(12) * mean +
  x(east) + y(north) + z(elevation) +
  z_label("Elevation (m)") + title("Maunga Whau, one elevation per cell")
```

*"Given Maunga Whau: a surface derived by bin into 12 and mean, x is east, y is
north, z is elevation."*

Nothing about the mountain changed. What changed is the **claim**. The sheet at
the top of this chapter says the ground rises smoothly between its samples; this
one says each cell has a single elevation and nothing is asserted about the
ground inside it. Both are honest about different data, and which you want is a
question about how yours was measured.

That question has a clear answer whenever a design produces exactly one value
per cell, which is the ordinary case in an experiment or a simulation: a grid of
conditions, one number in each. Read as nodes, such a table draws a sheet
smaller than the grid it describes. Nine values on a 3 × 3 grid make 2 × 2
blocks of four corners, so the sheet is four faces, and it stops at the outer
samples instead of covering their cells:

```{r}
cells <- expand.grid(a = c(-2, 0, 2), b = c(-2, 0, 2))
cells$v <- 0.019 + 0.0025 * (cells$a^2 + cells$b^2)

data(cells) + surface + x(a) + y(b) + z(v) + title("Nine values, four faces")
```

Nine numbers became four faces, and each face is a sheet stretched over the four
corners around it. Cut the x-y plane into cells instead, and each cell gets a
flat top of its own, a **plateau**:

```{r}
data(cells) + surface * bin(3) * mean + x(a) + y(b) + z(v) +
  color(v) + palette("plasma") + title("Nine values, nine plateaus")
```

*"Given the cells: a surface derived by bin into 3 and mean, x is a, y is b, z
is v, color by v, with the plasma palette."*

Nine plateaus, and three colors. That formula makes a symmetric bowl, so the
four corner cells share one height and the four edge cells share another. The
ramp is what makes that symmetry visible, where the sheet above left it to be
read off the axis.

`bin` cuts the `x` and `y` axes into cells, and `mean` averages the column named
on `z` inside each one. That is the same sentence a 3-D bar uses. The transforms
mean the same thing on both marks, and only the geometry laid on the cell
differs. Where a table holds one row per cell, as this one does, the mean of a
single value is that value, and the transform is doing nothing but naming the
column.

The step between two plateaus is drawn, as a **riser** standing on the line the
two cells share. That line has no width, so the riser asserts nothing about any
area: it is simply what a jump looks like. Without it the mark would draw
disconnected tiles floating at their own heights, and a reader could not tell a
step from a missing cell.

A riser spans the *difference* between two neighbors, and that is what separates
this from standing a [3-D bar](bar.qmd) on every cell. A bar runs the whole way
down to the baseline, so a field of bars hides its own back rows as soon as you
turn it. A riser is only as tall as its own step. Both are turned by 60 degrees
below, so you can see which one is still legible. Both are set to one color, so
the only thing that differs between them is the shape:

```{r}
((data(cells) + bar * bin(3) * mean + x(a) + y(b) + z(v) +
    space(turn = 60) + style(color = "#76b7b2") + title("Bars, turned")) |
   (data(cells) + surface * bin(3) * mean + x(a) + y(b) + z(v) +
      space(turn = 60) + style(color = "#76b7b2") + title("Plateaus, turned"))) +
  theme(height = 430)
```

*"Given the cells: bars derived by bin into 3 and mean, x is a, y is b, z is v,
beside a surface derived by bin into 3 and mean, x is a, y is b, z is v."*

`|` composes them ([Composition](../composition.qmd)), so the two are one
figure and turn together. That is what makes this a fair comparison, rather
than two pictures the reader has to match angle by angle.

The `theme(height = )` above sets the whole figure's height in pixels, and two
cubes side by side are why it is set. They share the page's width, and each one
keeps the full height. A cube is fitted with one scale on both axes, so it never
stretches to fill a tall panel. Without the setting, blank space would be left
above and below each cube.

Empty cells need no special rule and get none. `bin` leaves them out, because a
cell with no rows is not a measurement of zero. A plateau claims only its own
cell, so a cell with nothing in it is simply a cell with no top. It is the same
opening a missing crossing makes in a node mesh, for a different reason.

What a surface refuses is a grid of **slots**. `count` and `proportion` count
rows into the cells two categories make, and categories have gaps between them:

```{r}
#| error: true
data(gm_eras) + surface * count + x(continent) + y(era) + space()
```

## The mesh, and what the shading means

A smooth sheet hides where its samples were, the way a line hides its rows.
Drawing the mesh puts the grid back, so a reader can see how fine the sampling
was.
`border_color` and `border_size` draw the **mesh lines**. A face is a closed
fill like a bar's or a box's, so the [settable rule](../design-laws.qmd) gives
it a border. On a mesh those borders draw the grid itself:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  style(border_color = "white", border_size = 0.4) +
  z_label("Elevation (m)") + title("The mesh, drawn")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, with
border color white and border size 0.4."*

The mesh adds no shapes to the picture, because every face is already stroked in
its own shade. That thin stroke covers the pale seam that would otherwise show
between neighboring faces. `border_color` replaces the stroke's color. Set
`border_size = 0` for a seamless sheet.

Notice what the shading is doing in both plots, and what it is *not*. The faces
are shaded by their **slope**: a steep one is darker, a level one keeps its full
color. That is a fact about the terrain, not about where a light is, so turning
the scene rearranges the faces on the page and repaints none of them:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) +
  space(turn = 130, tilt = 35) +
  z_label("Elevation (m)") + title("Turned, not relit")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, turned
130, tilted 35."*

Compare the crater walls here with the first plot. Same faces, same shades, seen
from the other side. A lamp fixed to the page would have repainted every one
of them, and a color that changed while the data did not would mislead a
reader with no way to notice.

## Color by height, per face

How high is the summit, and how deep is the crater?
`color` maps on a surface, and a **measured** column colors it face by face
along a ramp. [`palette()`](../style.qmd) chooses which ramp. `"viridis"` suits
a topographic sheet, because it gets lighter the whole way, so you can judge the
height from the color even where a face is turned away:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) + color(elevation) +
  palette("viridis") +
  z_label("Elevation (m)") + title("Height twice: as a position and as a ramp")
```

Compare that with the first plot in this chapter. The summit and the crater rim
are in that plot too. With slope shading alone you have to work them out from
which faces are dark. The ramp states the heights, and the legend decodes them.

Mapping the height to color when it is already the height is redundant on
purpose, and it is what a reader of a topographic map expects, so the ramp is
worth drawing. It also marks a real boundary. An [`area`](area.qmd) refuses a
measured color, because a region has one interior and coloring an interior by a
measure is a gradient fill, which is a different and much larger job. A mesh has
*faces*, and a face is already small enough to hold one value. So the rule is
not about regions and strokes. It is whether a mark has parts small enough to
each hold one value, and a stroke has segments where a surface has faces.

A face has four corners and one color. What it reads is the field at its center,
which is the mean of those four, so the ramp describes the mesh you have and not
only the field you sampled from. Sample the same bowl twice and see where the
shape appears:

```{r}
bowl <- function(n) {
  g <- expand.grid(a = seq(-2, 2, length.out = n), b = seq(-2, 2, length.out = n))
  g$v <- 0.019 + 0.0025 * (g$a^2 + g$b^2)
  g
}
coarse <- bowl(3)
fine   <- bowl(7)

((data(coarse, name = "coarse") + surface + x(a) + y(b) + z(v) +
    color(v) + palette("plasma") + title("Sampled 3 by 3: four faces")) |
   (data(fine, name = "fine") + surface + x(a) + y(b) + z(v) +
      color(v) + palette("plasma") + title("Sampled 7 by 7: thirty-six faces"))) +
  theme(height = 320)
```

*"Given coarse: a surface, x is a, y is b, z is v, color by v, with the plasma
palette, beside the same given fine."*

Two tables on one page, so each is named. A layer finds its columns in the
nearest table by that name, and two tables with the same name would leave
the second one unreachable ([Data](../data.qmd)).

The left sheet is one flat color, and that is honest rather than broken. Every
face has one high corner, one low corner and two middle ones, so all four
average to the same number. The mesh is too coarse to hold the shape, and the
ramp says so instead of inventing a gradient across it. The right sheet has the
same field under it and enough faces to resolve it, so the bowl appears.

This is the reading that makes the legend trustworthy. A color is the height at
one named place on the sheet, so two faces of one color really are at one
height, and a sheet of one color really is flat on average.

The slope shading is still underneath the ramp, and the two combine rather than
cancel: a face takes its color from the ramp and is then darkened by how steep
it is. Viridis stays readable because only the *low* ground reaches its dark
colors, and on a terrain the low ground is also the flattest. The two darkenings
therefore do not add together, and the darkest face on this sheet is lighter
than a fully-shaded viridis floor would have been.

An **estimated** height takes a ramp the same way, and you name it the same way.
`density` writes its estimate into a column called `density`, so that is the
word `color` takes, exactly as it does on a [`zone`](zone.qmd) or a
[`path`](path.qmd):

```{r}
data(iris_flowers) + surface * density +
  x(petal_length) + y(sepal_length) + space() + color(density) +
  palette("viridis") + title("The estimate, as a height and as a ramp")
```

Nothing there is a special case. The transform makes a column, the column has a
name, and a name can be bound. Whether the height came from your table or from
an estimate changes where the number came from. It does not change what you may
say about it.

Whether you *want* the color as well as the height is a separate question, and
the plain sheet earlier in this chapter is the one that makes the argument about
height. Use the ramp when the low ground matters: slope shading darkens a steep
face and leaves a level one alone, so a broad, shallow mound and the flat floor
around it are the same color, and the ramp is what separates them.

A **categorical** color splits the mark instead, one sheet per group. gog sorts
every face of both sheets by depth together, so the two sheets cross rather than
one sitting wholly in front:

```{r}
band <- maunga_whau
band$side <- ifelse(band$north > 440, "north slope", "south slope")

data(band) + surface + x(east) + y(north) + z(elevation) + color(side) +
  z_label("Elevation (m)") + title("Two sheets, sorted together")
```

## Opacity by a measure, per face

Part of a sheet is often not the subject. The flat ground around a hill only
hides what is behind it, and fading those faces leaves the hill.
`opacity` maps on a surface, and it maps for the reason `color` does. A face is
small enough to hold one value, so it can hold an opacity too. Every face takes
its own opacity, and a face fades where its measure is small:

```{r}
data(maunga_whau) + surface +
  x(east) + y(north) + z(elevation) + opacity(elevation) +
  z_label("Elevation (m)") + title("Faint where the ground is low")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, opacity
by elevation."*

Over the low ground the fill fades to almost nothing and each face keeps its
outline, so that part of the sheet shows as mesh lines alone. A ramp cannot do
that; an opacity can. A color changes what a face says. An opacity changes
whether the face hides what is behind it. The legend beside the plot reads the
same way `color`'s does, with the smallest, middle and largest values of the
column.

No face disappears completely. A mapped `opacity` runs from 0.15 to 0.95, so the
faintest face here is still drawn, which is the rule every mapped `opacity`
follows ([Channels](../channels.qmd)).

A face reads its opacity where it reads its color, at its own center, which is
the mean of its four corners. A plateau and a riser each own one cell, so each
reads its own row. Either way one face has one reading, whichever channel asks
for it.

In the plot above, the height is said twice: once as a position and once as an
opacity. A second column is the more useful case: the shape stays in `z`, and
another measure decides which faces fade.

```{r}
relief <- maunga_whau
h <- matrix(relief$elevation, nrow = length(unique(relief$north)))
up   <- rbind(h[-1, ], h[nrow(h), ])
side <- cbind(h[, -1], h[, ncol(h)])
relief$steepness <- as.vector(sqrt((up - h)^2 + (side - h)^2))

data(relief) + surface + x(east) + y(north) + z(elevation) + opacity(steepness) +
  z_label("Elevation (m)") + title("Solid where the ground is steep")
```

`steepness` combines the two elevation changes at each crossing, one toward the
next east value and one toward the next north value. The steep sides of the hill
stay solid, and the flat ground around them fades away. The sheet is the same
shape as the one above it, and a different part of it is solid.

An [`area`](area.qmd) refuses a mapped opacity for the reason it refuses a
measured color. A region has one interior and one fill, so a row there is a
corner of the outline rather than a part that can carry a value. A mesh has
faces, so it has parts. Both channels ask the same question of a mark, so a mark
that refuses one refuses the other.

One value for the whole sheet is still [`style(opacity = )`](../style.qmd). Set
it when you want to see one sheet through another, and map it when you want each
face to fade by its own value.

## One sheet, four angles

A still 3-D plot hides something at every angle, and which thing it hides is the
angle's choice. So it is worth turning the same sheet rather than trusting one
view of it. Each of these changes exactly *one* of `turn`/`tilt` from the first:

```{r}
((data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
    color(elevation) + palette("viridis") + space(turn = 30, tilt = 25) +
    z_label("Elevation (m)") + title("The reference three-quarter view")) |
 (data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
    color(elevation) + palette("viridis") + space(turn = 140, tilt = 25) +
    z_label("Elevation (m)") + title("Turned: the same sheet from the west"))) /
((data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
    color(elevation) + palette("viridis") + space(turn = 30, tilt = 70) +
    z_label("Elevation (m)") + title("Tilted up: the crater appears")) |
 (data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
    color(elevation) + palette("viridis") + space(turn = 30, tilt = 5) +
    z_label("Elevation (m)") + title("Tilted down: the skyline")))
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, color by
elevation, with the viridis palette, turned 30, tilted 25, beside the same
turned 140; above the same tilted 70, beside the same tilted 5."*

Read them as a set and you can see what each angle hides. The crater is the most
interesting thing about this hill, and the first two plots do not show it. The
summit stands in front of the hollow, so no amount of turning at a low tilt
reveals it. Tilt up and the hollow opens, because you are now looking down on
the hill from above. The same tilt flattens the heights, until the sheet is a
colored map of itself. Tilt down instead and you get the opposite: the side view
is exact, the view from above is gone, and the sheet shows as an outline against
the sky. One or the other, never both.

That is the argument for a plot you can turn, made rather than asserted. Here
you need both the view from above and the view from the side, and no single
still view shows this hill whole. So turn the four with the mouse. Lift your eye
until the crater opens, and watch the heights flatten as it does. The four still
views make the argument; dragging them is how you check it.

All four turn together, because `|` and `/` composed them into one figure,
and a composition turns as one. Each panel keeps the angle its own sentence
asked for, and the drag adds the same change to every one. They turn as a
set, and *reset* returns each to where it began. To turn one alone, write it
alone.

That is composition rather than faceting, and the two share the same pair of
operators. A facet names a **column** ([Faceting](../faceting.qmd)), and a
viewing angle is not a column: it is a property of the coordinate space, the way
a `polar()` start angle is. gog does build small multiples of cubes, and these
four panels are not one ([Space](../space.qmd) has that example).

One limit shows in these four plots, and it belongs to the cube rather than to
the mark. Notice the vertical axis in the tilted-up panel. Its numbers sit tight
against each other, because that axis is projecting at about a third of its
length and they are still all being drawn. They do clear each other, and tilting
further makes gog drop some numbers rather than overlap them. Numbers packed
close together are simply what a foreshortened axis looks like.
[Space](../space.qmd) says which numbers a crowded axis drops, and why the scale
itself never changes.

## What a surface is for

Anything that is a *height over a plane*. Terrain, and the sea floor. A response
surface from a designed experiment. A model's fit measured over two of its
parameters. A model's predictions over a grid of two predictors. And, through
`density`, the shape of a two-dimensional distribution.

What it is not for is data that merely *has* three numbers per row. That is a
[3-D scatter](../space.qmd), and the refusal above names it for you. The
question to ask is whether the space between two of your samples means anything.
On a grid it does; in a cloud it does not.

## What you can set

No column says whether the mesh lines show, or how far you can see through a
sheet. A setting decides that without reading one, and each mark takes its own.
These are a surface's, generated from the engine's own legality table, so this
page cannot differ from what `style()` accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("surface")
```

A surface is a fill, so it takes a fill's settings, with two absences the next
section explains: no `size`, and no `pattern`.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

The flat marks take no `z`, so a surface with only `x` and `y` is the first
sentence to try.
A surface with no height is the same mistake as a surface in the plane, so it is
one refusal rather than two, and it names both routes into the cube along with
the mark that draws a field in the plane:

```{r}
#| error: true
data(maunga_whau) + surface + x(east) + y(north)
```

`density(levels = )` cuts a field into level *sets*, and a level set is a region
in the plane. A surface has already given the third axis to the measurement, so
a band drawn on it could only be a color, which is [`zone`](zone.qmd)'s reading
of that same request:

```{r}
#| error: true
data(iris_flowers) + surface * density(levels = 5) +
  x(petal_length) + y(sepal_length) + space()
```

A stepped sheet can be cut on either of two axes, and only one of the two draws.
Cutting the **floor** into cells is `surface * bin`, the section on one value
per cell above, and it draws. Cutting the **height** into bands is what `levels`
asks for, and that one is refused. It would cut the measurement rather than the
grid, so the plateaus would be contours of equal height rather than one per
cell.

`bandwidth` is refused for the reason it is refused on every two-dimensional
field: it is a length in one column's own units, and a field spreads over two
columns measuring different quantities, so one number cannot be a width in both.
Use `density(adjust = )`, which is dimensionless and scales both.

There is no `size` on a surface. Its extent is the grid it is drawn on, so a
size would be a second and contradictory answer to how big it is.
[`area`](area.qmd) refuses a size for the same reason. And `pattern` is refused,
which is the one place a fill does not take a texture:

```{r}
#| error: true
data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  style(pattern = "hatch")
```

A hatch tile is a texture in *screen* space, and every face of a projected mesh
is foreshortened differently, so one tile size would read as a different density
on every face. The texture would then follow the viewing angle rather than the
data, which is the one thing this mark's shading is built to avoid.