# Text

```{r}
#| include: false
source("../R/setup.R")
```

Which point is which? `text` draws a string at each `(x, y)` position. It is
`point`'s sibling: both place one glyph per row. A point's glyph is a dot. A
text mark's glyph is a **string**, and that string comes from a column, through
the `label` channel.

## The mark and its content

A point tells you *where* a row sits; it cannot tell you *which* row. Five medal
counts drawn as dots are five points with nothing to say which is which:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + point + x(gold) + y(silver) +
  x_label("Gold") + y_label("Silver") +
  title("Five countries, and no names")
```

*"Given the medals: points, x is gold, y is silver."*

Bind `label` and draw the names themselves at those positions, and each mark
now shows which row it is:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  x_label("Gold") + y_label("Silver") +
  title("The same five, named")
```

*"Given the medals: text, x is gold, y is silver, label by country."*

That is the whole difference between the two marks. `point` draws the same dot
at every row's position. `text` draws one of the row's own values there instead.

## Both at once: the labeled scatter

A dot marks one spot on the panel. A word is much wider than a dot, so a name
on its own only shows roughly where its row sits.
The two marks compose. Layer a `point` and a `text` over one `(x, y)`.
The dot gives the exact position and the string gives the name. That is a
labeled scatter, built from two marks rather than a third mark of its own.
Each `style()` applies to the mark written before it, so
`style(size = 5, color = "#9e9e9e")` colors the dots and not the words. A label
would otherwise sit on top of its dot, so `style(nudge = "up")` moves it off:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("Located by the dot, named just above")
```

*"Given the medals: points and also text, x is gold, y is silver, label by
country, with nudge up."*

There is no "labeled point" mark to learn: a labeled scatter is just
`point + text`, layered with the `+` [operator](../operators.qmd).

Five names fit here. Fifty would not, and choosing which few to keep is the
hardest part of a labeled scatter. A plot that carries `brush()` lets you try a
few in the web edition before you write any down. Clicking a mark there prints
its values on the plot, which [Selection](../selection.qmd#stamping-a-point)
calls a stamp. A stamp disappears when the page reloads. A `text` label is part
of the specification, so it is drawn every time.

### Nudging the label off its point

Up is only one of four directions, and it is not always the side with room. A
dot near the top of the panel, or one with a neighbor just above it, has no
space for its name there. Those labels need a different direction.
`style(nudge = )` shifts a text label off its `(x, y)`. It takes one direction:
`"up"`, `"down"`, `"left"`, or `"right"`:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "right") +
  x_label("Gold") + y_label("Silver") +
  title("The same, nudged right")
```

You do not give a distance. The engine takes it from the font size, far enough
to clear a default dot, so `nudge` asks only for a direction. A nudge is one
*fixed* shift, the same for every label, and nothing in the data changes it.
`jitter` and `dodge` work the other way: they compute their offsets from the
data, so every mark moves by its own amount. Only `text` accepts a nudge. A
`point` or a `bar` has no label to move, so the engine refuses the setting there
and names the mark that does.

### `repel`: moving labels off each other

A nudge shifts every label the same way, so it moves each label off its own dot
and does nothing else. When the labels collide with *each other*, one fixed
shift moves them all together and the words still overlap. Thirty European
countries are enough to show it:

```{r}
#| echo: false
peek(gm_europe)
```

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Thirty countries, named")
```

`repel` solves that, and it is a *collision modifier* rather than a setting. It
measures where the labels and the dots actually are, then moves each label off
the words around it:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same thirty, moved apart")
```

*"Given gapminder Europe: points and also text derived by repel, x is gdp, y is
life, label by country."*

Three things in that picture are worth naming. Every label comes to rest just
outside its dot, so no word covers the point it names. A label in a crowd slides
to whichever side has room. A label with space around it stays directly above
its dot. All three follow from one rule: `repel` pulls a label toward its own
dot and pushes it away from every word and every other dot.

A label sometimes has to travel much further than any of these did. Once it ends
up more than one line of text from its point, gog draws a thin line from the
point to the word. That line is what tells you which point the name belongs to.
None of the thirty labels moves that far, so `repel` draws no line here.

The placement is **deterministic**. It uses no random-number generator and no
system time, so the same specification draws the same picture every time.

When no arrangement fits, `repel` still draws every label and reports how
many still overlap. It never drops a label without saying so. A plot missing
the names it could not fit would look like a plot of fewer countries, so gog
reports the count instead. Thirty countries all fit, so no report appears here.
Three things give crowded labels more room: fewer rows, a smaller
`style(size = )`, and a larger plot. `theme(width =, height =)` sets the plot's
size in pixels, and the
[Composition](../composition.qmd#how-thin-is-the-marginal-histogram-themewidth-height)
chapter covers it.

The setting and the collision modifier compose, and each does one thing.
`style(nudge = )` says which side a label prefers, and `repel` resolves the
overlaps that remain:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) + style(nudge = "right") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Nudged right first, then moved apart")
```

*"Given gapminder Europe: points and also text derived by repel, x is gdp, y is
life, label by country, with nudge right."*

## Numeric labels: the value on each mark

A label does not have to be a name, and it does not have to sit on a scatter. A
bar shows which country won more. To read by how much, a reader has to measure
each bar against the axis, and that is an estimate. Sometimes the number itself
is the answer. Bind a column that holds numbers and each row's value is drawn. A
whole number is printed without a decimal point. Layer `text` over `bar` to
print each bar's height at its top:

```{r}
data(medals) + bar + style(color = "steelblue") +
  text + x(country) + y(gold) + label(gold) +
  y_label("Gold medals") +
  title("Each bar labeled with its height")
```

*"Given the medals: bars and also text, x is country, y is gold, label by
gold."*

## Color

By default every label is printed in one color. You see a hue faster than you
can read a word. A color split therefore shows the grouping before the name.
Map a column to `color` and the labels split by that column. The mapping earns
a legend to name the groups, and it is the same categorical color scale `bar`
uses. Here each country's name takes its own hue:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) + color(country) +
  x_label("Gold") + y_label("Silver") +
  title("One hue per country")
```

*"Given the medals: text, x is gold, y is silver, label by country, color by
country."*

## Styling: one color, one size

A `color` channel takes its hues from a column. A color you choose yourself is
a different thing: a setting rather than a mapping. When the labels are the plot
rather than notes beside it, the default font is often too small to read
comfortably. A larger size is a setting too. Neither comes from a column,
and that is what makes them settings. `style(color = )` sets the color of the
words. `style(size = )` sets the font size in pixels. Each applies to every
label in the layer:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(color = "seagreen", size = 17) +
  x_label("Gold") + y_label("Silver") +
  title("Bigger, greener labels")
```

*"Given the medals: text, x is gold, y is silver, label by country, colored
seagreen, with size 17."*

`style(size = )` gives every label in the layer one font size. A word cloud is
the plot this rules out. It sizes each word by its frequency, so it needs a size
per row. [What you can set](#what-you-can-set) shows what the engine answers
when you ask for one.

## Why `label` is a channel

A setting is a value you type into the sentence. You might expect a label to
work the same way, the way a title is typed. A label is not a title: it is one
value per row, and that makes it data.
`text` *requires* a `label`: it is the mark's minimum syllable, the way `x` and
`y` are. Ask for a `text` mark with no `label` and the engine refuses, and says
what to write instead, as [What it refuses](#what-it-refuses) shows.

The grammar's own rule decides this. A label's string is *data*, and data
reaches a mark only through a channel. That makes `label` a channel, so no mark
takes the words themselves as an argument. A fixed note is therefore a table
with one row, handed to `data()`, and never a value handed to `style()`. Only
`text` accepts `label`; every other mark refuses it, which is why a labeled
scatter is `point + text` and never a `label` on `point`. `text` is the mark,
and `label` is the channel that carries the column it draws.

## A note is a layer

A plot often needs one remark written on it, at a place you choose. It may say
what a region of the scatter means, or why a point sits where it does. If you
have used ggplot2, a note there is a separate function, `annotate()`. In gog it
is a table with one row, and that table is worth seeing drawn. A second `data()`
opens a second layer over its own table (the [Data](../data.qmd) chapter's
rule), and a table with one row in it puts one string on the plot:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
peek(income_note)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(income_note) + text + label(what) +
  title("A note is a layer, not a function")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also text from the
notes, label by what."*

Nothing was added to the grammar to reach that. `data()` already binds the
nearest table, `text` already draws one string per row, and a note is simply a
table with one row in it. The note is data like any other, so it is scaled like
any other. Give it an `x` far from the points and the axes stretch to cover it,
which is worth checking before you settle on a position.

One rule applies while writing a note. `x` and `y` belong to the whole plot,
so the note's table needs a column for each. The `income_note` table reuses the
plot's own column names, `gdp` and `life`. A note with its own names writes its
positions after `text`: `+ text + x(at) + y(value) + label(what)`. Anything
else `annotate()` places by data values can be written this way too. A dot on
the plot is a `point` over the same one-row table, and several notes are several
rows.

## Text that a transform places

A note names a place you chose. A transform can choose the place
instead, and then the text names whatever that transform built. Each of these
sentences has a sibling that draws the shape the text is labeling:

| Text derived by | What it labels |
|---|---|
| `text * layout` | the nodes of a network, beside `point * layout` and `edge * layout` |
| `text * flow` | the blocks of a flow, beside `ribbon * flow` and `zone * flow` |
| `text * partition` | the boxes of a packing, beside `zone * partition` |
| `text * repel` | the scatter labels of [`repel`](#repel-moving-labels-off-each-other), moved apart until they fit |

The chapters that draw them are [Network](../network.qmd), [Flow](../flow.qmd)
and [Transforms](../transforms.qmd#partition-a-whole-among-nested-parts). In each one the text is a layer over the mark that
drew the shape, and both layers read the same transform.

## What you can set

The [Styling](#styling-one-color-one-size) section set a color and a size.
Those are two entries in a longer list, and the rest of it is here: which size,
which color, and which side of the dot. None of that is in a column, so it is set rather than
mapped, and each mark takes its own settings. These are a text mark's, with the
values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("text")
```

`nudge` is the one setting only `text` has. It exists so a layered `point +
text` does not draw the label on top of the dot:

```{r}
data(medals) + x(country) + y(gold) +
  point + style(color = "#9e9e9e", size = 6) +
  text + label(gold) + style(nudge = "up", size = 11, color = "black") +
  y_label("Gold medals") + title("Labels nudged off their points")
```

The four directions and the fixed offset are under [Nudging the label off its
point](#nudging-the-label-off-its-point), with `repel` for labels that collide
with each other.

A mapped `size` on `text` is legal grammar that this engine does not draw.
`size(population)` is therefore refused as **Unsupported**, rather than accepted
and quietly ignored. Set the font size with `style(size = )` instead.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. For the channels a mark *maps* rather than
sets, read its row on the
[companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A `text` with no `label` is refused first. The label is the mark's minimum
syllable, the way `x` and `y` are, for the reason [Why `label` is a
channel](#why-label-is-a-channel) gave. The engine refuses and names the
channel to add:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver)
```

A reader who knows `point` will try its channels on `text`, and one of them
cannot do anything to a string. A point's glyph can change shape because the dot
itself says nothing. A text mark's glyph is the word, so there is nothing for
`shape` to change:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) + shape(country)
```

A text mark's form comes from `label`, for the reason [Why `label` is a
channel](#why-label-is-a-channel) gave. A string takes no texture either.
`pattern` dashes a stroke or hatches a filled region, and a word is neither:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(pattern = "hatch")
```

## In short

A `text` will not draw without a `label`, and no other mark accepts one. Its
words come from a column, so a fixed note is a one-row table. `repel` reports
how many labels it could not place, because a label removed without a word would
leave a picture the reader cannot check.

## What comes next

Row order means nothing to a text mark. Each word is drawn at the position its
own row gives it, whatever order the rows are in. Row order decides everything
for a [path](path.qmd), the mark that joins the rows in the order the table
holds them.
