# Text

```{r}
#| include: false
source("../R/setup.R")
```

Which point is which? `text` draws a string at each `(x, y)` position. It is
`point`'s sibling: both place one glyph per row. A point's glyph is a dot. A
text mark's glyph is a **string**, and that string comes from a column, through
the `label` channel.

## The mark and its content

A point tells you *where* a row sits; it cannot tell you *which* row. Five medal
counts drawn as dots are five anonymous points:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + point + x(gold) + y(silver) +
  x_label("Gold") + y_label("Silver") +
  title("Five countries, and no names")
```

*"Given the medals: points, x is gold, y is silver."*

Bind `label` and draw the names themselves at those positions, and each mark
now shows which row it is:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  x_label("Gold") + y_label("Silver") +
  title("The same five, named")
```

*"Given the medals: text, x is gold, y is silver, label by country."*

That is the whole difference between the two marks: `point` draws a glyph *at*
the row, `text` draws the row's own value *as* the glyph.

## Both at once: the labeled scatter

A word is wide and a position is a point, so a name alone marks its row only
roughly.
The two marks compose. Layer a `point` and a `text` over one `(x, y)`.
The dot gives the exact position and the string gives the name. That is a
labeled scatter, built from two marks rather than a third mark of its own.
Each `style()` applies to the mark written before it, so the gray below
belongs to the dots alone. A label would otherwise sit on top of its dot, so
`style(nudge = "up")` moves it off:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("Located by the dot, named just above")
```

*"Given the medals: points and also text, x is gold, y is silver, label by
country, with nudge up."*

There is no "labeled point" mark to learn: a labeled scatter is just
`point + text`, layered with the `+` [operator](../operators.qmd).

Five names fit here. Fifty would not, and choosing which few to keep is the
hardest part of a labeled scatter. In the web edition you can try them first.
Clicking a point prints its values on the plot, which
[Selection](../selection.qmd#stamping-a-point) calls a stamp. That shows how a
few labels look before you write any. A stamp disappears when the page reloads;
a `text` label is part of the specification, so it is drawn every time.

### Nudging the label off its point

Up is one of four directions, and it is not always the empty side. A dot near
the top of the panel, or one with a neighbor just above it, wants its name
somewhere else.
`style(nudge = )` shifts a text label off its `(x, y)`. It takes one direction:
`"up"`, `"down"`, `"left"`, or `"right"`:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "right") +
  x_label("Gold") + y_label("Silver") +
  title("The same, nudged right")
```

The distance comes from the font size, just enough to sit outside a default dot,
so there is nothing to specify; `nudge` takes only the direction. It is a
*setting*, not a collision modifier, and the two differ in where the offset
comes from. `jitter` and `dodge` compute their offsets from the data, so every
mark moves by its own amount. A nudge is one *fixed* shift, the same for every
label, derived from nothing in the data. It is `text`-only. Nudging a `point` or
`bar` would move the data itself, so the engine refuses it there and says why.

### `repel`: moving labels off each other

A nudge shifts every label the same way, so it moves each label off its own dot
and does nothing else. When the labels collide with *each other*, one fixed
shift moves them all together and the words still overlap. Thirty European
countries are enough to show it:

```{r}
#| echo: false
peek(gm_europe)
```

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Thirty countries, named")
```

`repel` solves that, and it is a *collision modifier* rather than a setting. It
measures where the labels and the dots actually are, then moves each label off
the words around it:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same thirty, moved apart")
```

*"Given gapminder Europe: points and also text derived by repel, x is gdp, y is
life, label by country."*

The modifier does three things in that picture. Almost every label comes to rest
just outside its dot, so the word does not cover the point it names. A label
that moved far keeps a thin line back to its point, so you can still see which
point it names. And each label rests beside its dot, on whichever side has room.
`repel` pulls a label toward its dot and pushes it away from every word and
every other dot. A label with no neighbors comes to rest just above its dot.

The placement is **deterministic**. It uses no random-number generator and no
system time, so the same specification draws the same picture every time.

When no arrangement fits, `repel` still draws every label and reports how
many still overlap. It never drops a label in silence. A plot missing the
names it could not fit would look like a plot of fewer countries, so gog
reports the count instead. A larger plot, a smaller `style(size = )`, or
fewer rows separate the labels.

The setting and the collision modifier compose, and each does one thing.
`style(nudge = )` says which side a label prefers, and `repel` resolves the
overlaps that remain:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) + style(nudge = "right") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Nudged right first, then moved apart")
```

## Numeric labels: the value on each mark

A bar shows which country won more, and the axis only lets a reader estimate by
how much. Sometimes the number itself is the answer.
A `label` need not be a name. Bind a column that holds numbers and each row's value is
drawn. A whole number is printed without a decimal point. Layer `text` over
`bar` to print each bar's height at its top:

```{r}
data(medals) + bar + style(color = "steelblue") +
  text + x(country) + y(gold) + label(gold) +
  y_label("Gold medals") +
  title("Each bar labeled with its height")
```

*"Given the medals: bars and also text, x is country, y is gold, label by
gold."*

## Color

Every label is printed in one color until a column separates them. A reader
sees a hue before reading a word, so the group arrives before the name.
`color` splits the labels by group and earns a legend, the same categorical split
`bar` makes. Here each country's name takes its own hue:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) + color(country) +
  x_label("Gold") + y_label("Silver") +
  title("One hue per country")
```

*"Given the medals: text, x is gold, y is silver, label by country, color by
country."*

## Styling: one color, one size

When the labels are the plot rather than notes beside it, the default font can
be small for the job. A larger size for all of them, and a color of your own,
are settings, because no column decides them.
`style(color = )` sets the color of the words. `style(size = )` sets the font
size in pixels. Both cover the whole layer:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(color = "seagreen", size = 17) +
  x_label("Gold") + y_label("Silver") +
  title("Bigger, greener labels")
```

*"Given the medals: text, x is gold, y is silver, label by country, colored
seagreen, with size 17."*

On `text`, `size` is a layer setting, not a mapped channel, so one font size
covers every label. A word cloud sizes each word by its frequency, which needs a
size per row.

## Why `label` is a channel

You might expect to type the words into the sentence, the way a title is typed.
A label is not a title: it is one value per row, and that makes it data.
`text` *requires* a `label`: it is the mark's minimum syllable, the way `x` and
`y` are. Ask for a `text` mark with no `label` and the engine refuses, and says
what to write instead, as [What it refuses](#what-it-refuses) shows.

The grammar's own rule decides this. A label's string is *data*, and data
reaches a mark only through a channel. So `label` is a channel, and a mark takes
no `text("USA")` argument. So a fixed note comes from a one-row table (`data`),
not a setting. Only `text` accepts `label`; every other mark refuses it, which
is why a labeled scatter is `point + text` and never a `label` on `point`.
`text` is the mark; `label` is the column it draws.

## A note is a layer

A plot often needs one remark written on it, at a place you choose. It may say
what a region of the scatter means, or why a point sits where it does.
That one-row table is worth seeing drawn. In ggplot2 a note is a separate
function, `annotate()`. In gog it is a table with one row. A second `data()`
opens a second layer over its own table (the [Data](../data.qmd) chapter's
rule), and a table with one row in it puts one string on the plot:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
peek(income_note)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(income_note) + text + label(what) +
  title("A note is a layer, not a function")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also text from the
notes, label by what."*

Nothing was added to the grammar to reach that. `data()` already binds the
nearest table, `text` already draws one string per row, and a note is a table
that happens to have one row. The note is data like any other, so it is scaled
like any other. Place it far from the points and the axes widen to include it.

One thing to know while writing a note. `x` and `y` belong to the whole plot,
so the note's table needs a column for each. The note above reuses the base
names, `gdp` and `life`. A note with its own names writes its positions after
`text`: `+ text + x(at) + y(value) + label(what)`. Everything else
`annotate()` does in the data's own units is already sayable. A dot on the
plot is a `point` over the same one-row table, and several notes are several
rows.

## Text that a transform places

The notes above name a place you chose. A transform can choose the place
instead, and then the text names whatever that transform built. Each of these
sentences has a sibling that draws the shape the text is labeling:

| Text derived by | What it labels |
|---|---|
| `text * layout` | the nodes of a network, beside `point * layout` and `edge * layout` |
| `text * flow` | the blocks of a flow, beside `ribbon * flow` and `zone * flow` |
| `text * partition` | the boxes of a packing, beside `zone * partition` |
| `text * repel` | the same scatter labels as above, moved apart until they fit |

The chapters that draw them are [Network](../network.qmd), [Flow](../flow.qmd)
and [Transforms](../transforms.qmd#partition-a-whole-among-nested-parts). In each one the text is a layer over the mark that
drew the shape, and both layers read the same transform.

## What you can set

Once the words are on the plot, the next question is how they look. Which size,
which color, and which side of the dot? None of that is in a column, so it is
set rather than mapped, and each mark takes its own settings. These are a text
mark's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("text")
```

`nudge` is the one setting only `text` has. It exists so a layered `point +
text` does not draw the label on top of the dot:

```{r}
data(medals) + x(country) + y(gold) +
  point + style(color = "gray", size = 6) +
  text + label(gold) + style(nudge = "up", size = 11, color = "black") +
  y_label("Gold medals") + title("Labels nudged off their points")
```

The four directions and the fixed offset are under [Nudging the label off its
point](#nudging-the-label-off-its-point), with `repel` for labels that collide
with each other.

A mapped `size` on `text` is valid grammar this engine does not draw, so
`size(population)` is refused as unsupported rather than ignored. Set the font
size with `style(size = )` instead.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A `text` with no `label` is refused first. The label is the mark's minimum
syllable, the way `x` and `y` are, for the reason [Why `label` is a
channel](#why-label-is-a-channel) gave. The engine refuses and names the
channel to add:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver)
```

A reader who knows `point` will try its channels on `text`, and one of them has
no work to do on a string.
A text mark's glyph *is* its content, so it has nothing left for `shape` to
vary:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) + shape(country)
```

A text mark's form comes from `label`, for the reason that section gave. For
the same reason a string takes no texture, being neither a stroke to dash nor a
region to hatch:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(pattern = "hatch")
```

## In short

A `text` will not draw without a `label`, and no other mark accepts one. Its
words come from a column, so a fixed note is a one-row table. `repel` reports
how many labels it could not place, because a label dropped in silence is a plot
that lies.

## What comes next

Row order means nothing to a text mark, which puts each word where its
positions say. It is everything to a [path](path.qmd), the mark that joins the
rows in the order the table holds them.
