# Text

```{r}
#| include: false
source("../R/setup.R")
```

Which point is which? `text` draws a **string**, a piece of text such as a name
or a number, at each `(x, y)` position. It is `point`'s sibling: both place one
glyph per row. A point's glyph is a dot by default. A text mark's glyph is a
string, and that string comes from a column, through the `label` channel. This
chapter uses it three ways: to name the points of a scatter, to print a value on
a bar, and to write a note on a plot.

## The mark and its content

A point tells you *where* a row sits; it cannot tell you *which* row. Five
countries drawn as dots are five points with nothing to say which country is
which:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + point + x(gold) + y(silver) +
  x_label("Gold") + y_label("Silver") +
  title("Five countries, and no names")
```

*"Given the medals: points, x is gold, y is silver."*

Replace `point` with `text` and bind `label` to `country`, and each row's name
is drawn at its position:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  x_label("Gold") + y_label("Silver") +
  title("The same five, named")
```

*"Given the medals: text, x is gold, y is silver, label by country."*

That is the whole difference between the two marks. `point` draws the same dot
at every row's position. `text` draws one of the row's own values there instead.

## Both at once: the labeled scatter

A dot marks one position on the panel. A word is much wider than a dot, so a
name on its own only shows roughly where its row sits. The two marks can be
drawn together: layer a `point` mark and a `text` mark on the same `x` and `y`,
with the `+` [operator](../operators.qmd). The dot gives the exact position and
the string gives the name. That is a labeled scatter, built from two marks
rather than a third mark of its own.

Each `style()` applies to the mark written before it, so
`style(size = 5, color = "#9e9e9e")` colors the dots and not the words. Without
a nudge, each label is drawn centered on its dot, so `style(nudge = "up")` moves
every label above its dot:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("Located by the dot, named just above")
```

*"Given the medals: points and also text, x is gold, y is silver, label by
country, with nudge up."*

### Nudging the label off its point

Up is only one of four directions, and it is not always the side with room. A
dot near the top of the panel, or one with a neighbor just above it, has no
space for its name there. Those labels need a different direction.
`style(nudge = )` shifts a text label off its `(x, y)`. It takes one direction:
`"up"`, `"down"`, `"left"`, or `"right"`:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "right") +
  x_label("Gold") + y_label("Silver") +
  title("The same, nudged right")
```

You do not give a distance. The engine sets it from the font size, far enough
that the word does not touch a dot of the default size. That is why `nudge` asks
only for a direction. A nudge is one *fixed* shift, the same for every label,
and nothing in the data changes it. `jitter` and `dodge` work differently: they
compute each mark's offset from the data, so marks can move by different
amounts. Only `text`
accepts a nudge. A `point` or a `bar` has no label to move, so the engine
refuses the setting there and names the mark that does.

### `repel`: moving labels off each other

A nudge shifts every label the same way, so it moves each label off its own dot
and does nothing else. When the labels collide with *each other*, one fixed
shift moves them all together and the words still overlap. Thirty European
countries are enough to show it:

```{r}
#| echo: false
peek(gm_europe)
```

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Thirty countries, named")
```

Every name is drawn on its own dot, and where the countries crowd together the
words overlap. A nudge would move all thirty words the same distance in the same
direction, so the same words would still overlap.

`repel` solves that. It is one of the four collision modifiers, with `jitter`,
`dodge` and `stack`, so it is written after the mark with `*`, not inside
`style()`. It measures where the labels and the dots actually are, then moves
each label off the words around it:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same thirty, moved apart")
```

*"Given gapminder Europe: points and also text derived by repel, x is gdp, y is
life, label by country."*

Notice three things in this plot. Every label is drawn just outside its dot, so
no word covers the dot it names. A label with other words near it moves to
whichever side has room. A label with space around it stays where it started,
directly above its dot. The first two follow from one rule: `repel` pulls each
label toward its own dot, and pushes it away from every other word and every
dot, its own included.

In a more crowded plot, `repel` can move a label much further from its dot. Once
the gap is larger than the height of one line of text, gog draws a thin line
from the dot to the word, so you can still see which dot the name belongs to. No
label in this plot moved that far, so none has a line.

The placement is **deterministic**. It uses no random-number generator and does
not read the computer's clock, so the same specification draws the same picture
every time. It does depend on the order of the rows, though. `repel` settles the
overlaps pair by pair, in the order of the rows, so the same rows in a different
order can end with some labels in different places.

Sometimes no arrangement keeps every label apart. `repel` then still draws every
label, and gog prints a message saying how many still overlap. Which names to
leave out is your decision, so `repel` never removes one. In this plot all
thirty fit, so gog prints no message.

Three things give crowded labels more room: fewer rows, a smaller
`style(size = )`, and a larger plot. `theme(width =, height =)` sets the plot's
size in pixels, and the
[Composition](../composition.qmd#how-thin-is-the-marginal-histogram-themewidth-height)
chapter explains it.

Fewer rows is often the right choice. On a large table you usually name only the
few rows that matter: give the `text` layer its own table that holds just those
rows, as [A note is a layer](#a-note-is-a-layer) does with one row. In the web
edition, clicking a dot on a plot with `brush()` shows its values beside it,
which [Selection](../selection.qmd#stamping-a-point) calls a stamp. Stamps help
you choose those rows. A stamp is gone when the page reloads, while a `text`
label is part of the specification and is drawn every time.

In the next plot, the dots come from `gapminder_2007`, which holds all 142
countries. The names come from `gm_populous`, the ten countries with the most
people in 2007. `gm_populous` has the same columns, so the `text` layer reads
`gdp` and `life` from it:

```{r}
#| echo: false
peek(gapminder_2007)
peek(gm_populous)
```

```{r}
data(gapminder_2007) + point + style(color = "#9e9e9e") +
  x(gdp, scale = "log") + y(life) +
  data(gm_populous) + text * repel + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The ten most populous of 142 countries, named")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, and also
text from the ten most populous countries, derived by repel, label by
country."*

The `gdp` axis uses a log scale ([Scales](../scales.qmd)), which gives the many
poorer countries more room. `repel` moves the names off the 132 unnamed dots
too, though they come from another table. It moves a name off those dots only a
little, never far enough for gog to draw a thin line. So a name can still cover
some of them: United States covers several at the top right, and Bangladesh
covers two. A [halo](#a-halo-around-each-label) keeps such a name readable.

A nudge and `repel` can be used together, and each does one job.
`style(nudge = )` chooses the side each label is pulled toward, and `repel` then
moves the labels that still overlap:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) + style(nudge = "right") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Nudged right first, then moved apart")
```

*"Given gapminder Europe: points and also text derived by repel, x is gdp, y is
life, label by country, with nudge right."*

With every label pulled to the right of its dot, four labels run into their
neighbors there, and `repel` moves those four far enough to need a thin line.
Norway's dot is at the right edge of the panel, so its label has no room on its
right and is drawn over its own dot.

## Number labels: the value on each mark

A label does not have to be a name, and it does not have to sit on a scatter. A
bar shows which country won more. To read how much more, a reader has to measure
each bar against the axis, and that measurement is only an estimate. Sometimes
the number itself is the answer. Bind a column that holds numbers and each row's
value is drawn. A whole number is printed without a decimal point. Layer `text`
over `bar` to print each bar's height at its top:

```{r}
data(medals) + bar + style(color = "steelblue") +
  text + x(country) + y(gold) + label(gold) +
  y_label("Gold medals") +
  title("Each bar labeled with its height")
```

*"Given the medals: bars and also text, x is country, y is gold, label by
gold."*

Each bar now carries its count, so the gap between the USA's 46 and China's 38
can be read without the axis.

## Color

You see a hue faster than you can read a word, so a color split shows you the
groups before you read a single name. Every label so far has been printed in one
color, which is the default. Map a categorical column to `color` and the labels
split by that column; a column of numbers is refused on a text mark. The mapping
earns a legend to name the groups, and the colors come from the same palette
points and bars use. In the plot below each country is its own group, so the hue
repeats the name:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) + color(country) +
  x_label("Gold") + y_label("Silver") +
  title("One hue per country")
```

*"Given the medals: text, x is gold, y is silver, label by country, color by
country."*

The legend names the five hues. With a column that several rows share, such as
a continent, one hue would mark every name in a group before you read any of
them.

## Styling: one color, one size

The `color` channel takes its hues from a column. When the words are the whole
plot, with no dots or bars beside them, you may want one color of your own and a
larger font, because the default is often too small to read comfortably.
Neither comes from a column, so both are settings rather than mappings.
`style(color = )` sets the color of the words, and `style(size = )` sets the
font size in pixels. Each applies to every label in the layer:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(color = "seagreen", size = 17) +
  x_label("Gold") + y_label("Silver") +
  title("Bigger, greener labels")
```

*"Given the medals: text, x is gold, y is silver, label by country, colored
seagreen, with size 17."*

gog refuses a `size` mapped to a column on a text mark, so every label in a
layer has one font size. A word cloud sizes each word by how often it occurs,
so it needs a size per row, and that is one reason gog draws no word cloud.

## A halo around each label

A label drawn over a dot or a line can be hard to read, because the mark shows
around its letters. `style(border_size = )` draws a **halo**: a band of the
panel's background color around each letter, drawn under the text. It hides the
mark next to each letter and does not change the letters:

```{r}
data(medals) + point + style(color = "#9e9e9e", size = 9) +
  text + x(gold) + y(silver) + label(country) +
  style(size = 12, border_size = 3) +
  x_label("Gold") + y_label("Silver") +
  title("Names with a halo")
```

*"Given the medals: points and also text, x is gold, y is silver, label by
country, with size 12 and border size 3."*

A thin band of the panel's color now separates each name from the gray dot
behind it. `border_size` sets the width of the stroke in pixels. Half of it lies
under the letters, so a width of 3 shows as a band 1.5 pixels wide.
`border_color` sets its color, if you want one other than the background. Either
setting alone draws a halo: an unset width is 3 pixels, and an unset color is the
panel's background color. With neither, a label has no halo.

A halo hides the marks it covers, so use it where the words matter more than
those marks. It never hides another label's letters, because every halo in a
layer is drawn before any letter. It does not move overlapping labels apart.
That is the job of [`repel`](#repel-moving-labels-off-each-other).

## Why `label` is a channel

A setting is a value you type into the sentence. You might expect to type a
label into the sentence in the same way, like a title. A label is not a title:
it is one value per row, and that makes it data. One rule of the grammar decides
whether a label is a setting or a channel: data reaches a mark only through a
channel. That makes `label` a channel, so no mark takes the words themselves as
an argument. A fixed note is therefore a table with one row, given to `data()`,
and never a value given to `style()`.

Only `text` accepts `label`; every other mark refuses it, which is why a labeled
scatter is `point + text` and never a `label` on `point`. `text` *requires* a
`label`: the label is part of the mark's minimum syllable, as `x` and `y` are.
Ask for a text mark with no `label` and the engine refuses, and says what to
write instead, as [What it refuses](#what-it-refuses) shows. `text` is the mark,
and `label` is the channel that carries the column it draws.

## A note is a layer

A plot often needs one note written on it, at a place you choose. It may say
what a region of the scatter means, or why a point sits where it does. If you
have used ggplot2, a note there is a separate function, `annotate()`. In gog it
is a table with one row, and the plot below draws it. A second `data()` gives
the next layer its own table (the [Data](../data.qmd) chapter's rule), and a
table with one row in it puts one string on the plot:

```{r}
#| echo: false
peek(income_note)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(income_note) + text + label(what) +
  title("A note is a layer, not a function")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also text from the
income note, label by what."*

The note sits in an empty corner of the scatter: no country in 2007 has an
income that high with lives that short.

That plot needed nothing new in the grammar. A mark written directly after a
`data()` already reads that table, and `text` already draws one string per row.
The note is data like any other, so it is scaled like any other. Give the note
an `x` far from the points, and the axes stretch to cover it, so check the axes
each time you place a note.

One rule applies while writing a note. The `text` layer names no position of its
own, so it reads the columns the axes use, `gdp` and `life`, from its own table.
That is why `income_note` has a column with each of those names. A note with its
own names writes its positions after `text`: `+ text + x(at) + y(value) +
label(what)`. Any other mark at a place you choose is written the same way: a
small table, and a layer over it. A dot on the plot is a `point` over the same
one-row table, and several notes are several rows.

## Text that a transform places

You choose where a note goes. A transform can choose the place instead, and then
the text names whatever that transform built. In the first three rows, a second
layer with the same transform draws the shape the text labels:

| Text derived by | What it labels |
|---|---|
| `text * layout` | the nodes of a network, beside `point * layout` and `edge * layout` |
| `text * flow` | each stage's slots in a flow, beside `ribbon * flow` and `zone * flow` |
| `text * partition` | the regions of a hierarchy, beside `zone * partition` |
| `text * repel` | the scatter labels of [`repel`](#repel-moving-labels-off-each-other), moved off each other |

The chapters that draw them are [Network](../network.qmd), [Flow](../flow.qmd)
and [Transforms](../transforms.qmd#partition-a-whole-among-nested-parts). In each one the text is a layer over the mark that
drew the shape, and both layers read the same transform.

## What you can set

The [Styling](#styling-one-color-one-size) section set a color and a size. Those
are two entries in a longer list, and the whole of it is here: which color, how
transparent, which size, which way to shift the label, and the color and width
of its halo. None of that is in a
column, so it is set rather than mapped, and each mark takes its own settings.
These are a text mark's, with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("text")
```

`nudge` is the one setting only `text` has. It exists so a layered `point +
text` does not draw the label on top of the dot:

```{r}
data(medals) + x(country) + y(gold) +
  point + style(color = "#9e9e9e", size = 6) +
  text + label(gold) + style(nudge = "up", size = 11, color = "black") +
  y_label("Gold medals") + title("Labels nudged off their points")
```

Each count sits just above its dot, and neither covers the other. The four
directions and the fixed offset are under [Nudging the label off its
point](#nudging-the-label-off-its-point), with `repel` for labels that collide
with each other.

A mapped `size` on `text` is legal grammar that this engine does not draw.
`size(gold)` is therefore refused as **Unsupported**, rather than accepted
and quietly ignored. Set the font size with `style(size = )` instead.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. For the channels a mark *maps* rather than
sets, read its row on the
[companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A text mark refuses three things: a missing `label`, the `shape` channel, and
any `pattern`. The label is part of the mark's minimum syllable, as `x` and `y`
are: with no string, a text mark has nothing to draw. The engine refuses and
names the channel to add:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver)
```

A reader who knows `point` will try its channels on `text`. `color` draws with a
categorical column, and `size` and `opacity` are legal but not drawn. `shape` is
refused outright, because it cannot change a string. A point's glyph can change
shape because the dot itself says nothing. A text mark's glyph is the string
itself, so there is nothing for `shape` to change:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) + shape(country)
```

A string takes no texture either. `pattern` dashes a stroke or hatches a filled
region, and a word is neither:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(pattern = "hatch")
```

## In short

A text mark will not draw without a `label`, and no other mark accepts one. Its
words come from a column, so a fixed note is a table with one row. `repel`
draws every label and reports how many still overlap, because which names to
leave out is your decision.

## What comes next

Without `repel`, row order never moves a word: each one is drawn at the position
its own row gives it. For a [path](path.qmd), row order decides which rows are
joined: the mark joins each row to the next, in the order the table holds them.
