# Text

```{r}
#| include: false
source("../R/setup.R")
```

Which point is which? `text` draws a string at each `(x, y)` position. It is
`point`'s sibling: both place one glyph per row. A point's glyph is a dot. A
text mark's glyph is a **string**, and that string comes from a column, through
the `label` channel.

## The mark and its content

A point tells you *where* a row sits; it cannot tell you *which* row. Five medal
counts drawn as dots are five anonymous points:

```{r}
data(medals) + point + x(gold) + y(silver) +
  x_label("Gold") + y_label("Silver") +
  title("Five countries, and no names")
```

*"Given the medals: points, x is gold, y is silver."*

Bind `label` and draw the names themselves at those positions, and each mark
now shows which row it is:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  x_label("Gold") + y_label("Silver") +
  title("The same five, named")
```

*"Given the medals: text, x is gold, y is silver, label by country."*

That is the whole difference between the two marks: `point` draws a glyph *at*
the row, `text` draws the row's own value *as* the glyph.

## Both at once: the labeled scatter

They compose. Layer a `point` and a `text` over one `(x, y)`. The dot gives the
exact position and the string gives the name. That is a labeled scatter, built
from two marks rather than a third mark of its own. Each `style()` applies to
the mark written before it, so the gray below belongs to the dots alone. A label
would otherwise sit on top of its dot, so `style(nudge = "up")` moves it off:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("Located by the dot, named just above")
```

*"Given the medals: points and also text, x is gold, y is silver, label by
country, with nudge up."*

There is no "labeled point" mark to learn: a labeled scatter is just
`point + text`, layered with the `+` [operator](../operators.qmd).

Five names fit here. Fifty would not, and choosing which few to keep is the
hardest part of a labeled scatter. In the web edition you can try them first.
Clicking a point prints its values on the plot, which
[Selection](../selection.qmd#stamping-a-point) calls a stamp. That shows how a
few labels look before you write any. A stamp disappears when the page reloads;
a `text` label is part of the specification, so it is drawn every time.

### Nudging the label off its point

`style(nudge = )` shifts a text label off its `(x, y)`. It takes one direction:
`"up"`, `"down"`, `"left"`, or `"right"`:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "right") +
  x_label("Gold") + y_label("Silver") +
  title("The same, nudged right")
```

The distance comes from the font size, just enough to sit outside a default dot,
so there is nothing to specify; `nudge` takes only the direction. It is a
*setting*, not a collision modifier, and the two differ in where the offset
comes from. `jitter` and `dodge` compute their offsets from the data, so every
mark moves by its own amount. A nudge is one *fixed* shift, the same for every
label, derived from nothing in the data. It is `text`-only. Nudging a `point` or
`bar` would move the data itself, so the engine refuses it there and says why.

### `repel`: moving labels off each other

A nudge shifts every label the same way, so it moves each label off its own dot
and does nothing else. When the labels collide with *each other*, one fixed
shift moves them all together and the words still overlap. Thirty European
countries are enough to show it:

```{r}
#| echo: false
peek(gm_europe)
```

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Thirty countries, named")
```

`repel` solves that, and it is a *collision modifier* rather than a setting. It
measures where the labels and the dots actually are, then moves each label off
the words around it:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same thirty, moved apart")
```

*"Given gapminder Europe: points and also text derived by repel, x is gdp, y is
life, label by country."*

The modifier does three things in that picture. Almost every label comes to rest
just outside its dot, so the word does not cover the point it names. A label
that moved far keeps a thin line back to its point, so you can still see which
point it names. And each label rests beside its dot, on whichever side has room.
`repel` pulls a label toward its dot and pushes it away from every word and
every other dot. A label with no neighbors comes to rest just above its dot.

The placement is **deterministic**. It uses no random-number generator and no
system time, so the same specification draws the same picture every time.

When no arrangement fits, `repel` still draws every label and reports how
many still overlap. It never drops a label in silence. A plot missing the
names it could not fit would look like a plot of fewer countries, so gog
reports the count instead. A larger plot, a smaller `style(size = )`, or
fewer rows separate the labels.

The setting and the collision modifier compose, and each does one thing.
`style(nudge = )` says which side a label prefers, and `repel` resolves the
overlaps that remain:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) + style(nudge = "right") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Nudged right first, then moved apart")
```

## Numeric labels: the value on each mark

A `label` need not be a name. Bind a column that holds numbers and each row's value is
drawn. A whole number is printed without a decimal point. Layer `text` over
`bar` to print each bar's height at its top:

```{r}
data(medals) + bar + style(color = "steelblue") +
  text + x(country) + y(gold) + label(gold) +
  y_label("Gold medals") +
  title("Each bar labeled with its height")
```

*"Given the medals: bars and also text, x is country, y is gold, label by
gold."*

## Color

`color` splits the labels by group and earns a legend, the same categorical split
`bar` makes. Here each country's name takes its own hue:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) + color(country) +
  x_label("Gold") + y_label("Silver") +
  title("One hue per country")
```

*"Given the medals: text, x is gold, y is silver, label by country, color by
country."*

## Styling: one color, one size

`style(color = )` sets the color of the words. `style(size = )` sets the font
size in pixels. Both cover the whole layer:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(color = "seagreen", size = 17) +
  x_label("Gold") + y_label("Silver") +
  title("Bigger, greener labels")
```

*"Given the medals: text, x is gold, y is silver, label by country, colored
seagreen, with size 17."*

On `text`, `size` is a layer setting, not a mapped channel, so one font size
covers every label. A word cloud sizes each word by its frequency, which needs a
size per row.

## Why `label` is a channel

`text` *requires* a `label`: it is the mark's minimum syllable, the way `x` and
`y` are. Ask for a `text` mark with no `label` and the engine refuses, and says
what to write instead:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver)
```

The grammar's own rule decides this. A label's string is *data*, and data
reaches a mark only through a channel. So `label` is a channel, and a mark takes
no `text("USA")` argument. So a fixed note comes from a one-row table (`data`),
not a setting. Only `text` accepts `label`; every other mark refuses it, which
is why a labeled scatter is `point + text` and never a `label` on `point`.
`text` is the mark; `label` is the column it draws.

## A note is a layer

That one-row table is worth seeing drawn. In ggplot2 a note is a separate
function, `annotate()`. In gog it is a table with one row. A second `data()`
opens a second layer over its own table (the [Data](../data.qmd) chapter's
rule), and a table with one row in it puts one string on the plot:

```{r}
notes <- data.frame(gdp = 35000, life = 66, what = "high income, short lives")
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(notes) + text + label(what) +
  title("A note is a layer, not a function")
```

*"Given gapminder 2007: points, x is gdp, y is life, and also text from the
notes, label by what."*

Nothing was added to the grammar to reach that. `data()` already binds the
nearest table, `text` already draws one string per row, and a note is a table
that happens to have one row. The note is data like any other, so it is scaled
like any other. Place it far from the points and the axes widen to include it.

One thing to know while writing a note. `x` and `y` belong to the whole plot,
so the note's table needs a column for each. The note above reuses the base
names, `gdp` and `life`. A note with its own names writes its positions after
`text`: `+ text + x(at) + y(value) + label(what)`. Everything else
`annotate()` does in the data's own units is already sayable. A dot on the
plot is a `point` over the same one-row table, and several notes are several
rows.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("text")
```

`nudge` is the one setting only `text` has. It exists so a layered `point +
text` does not draw the label on top of the dot:

```{r}
data(medals) + x(country) + y(gold) +
  point + style(color = "gray", size = 6) +
  text + label(gold) + style(nudge = "up", size = 11, color = "black") +
  y_label("Gold medals") + title("Labels nudged off their points")
```

The four directions and the fixed offset are under [Nudging the label off its
point](#nudging-the-label-off-its-point), with `repel` for labels that collide
with each other.

A mapped `size` on `text` is valid grammar this engine does not draw, so
`size(population)` is refused as unsupported rather than ignored. Set the font
size with `style(size = )` instead.

## What it refuses

A text mark's glyph *is* its content, so it has nothing left for `shape` to
vary:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) + shape(country)
```

A text mark's form comes from `label`, for the reason [Why `label` is a
channel](#why-label-is-a-channel) gave. For the same reason a string takes no
texture, being neither a stroke to dash nor a region to hatch:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(pattern = "hatch")
```
