# Text

```{r}
#| include: false
source("../R/setup.R")
```

Which point is which? `text` draws a string at each `(x, y)` position. It is
`point`'s sibling: both place one glyph per row, but where a point's glyph is a
dot, a text mark's glyph is a **string**, and the string comes from a column,
through the `label` channel.

## The mark and its content

A point tells you *where* a row sits; it cannot tell you *which* row. Five medal
counts drawn as dots are five anonymous points:

```{r}
data(medals) + point + x(gold) + y(silver) +
  x_label("Gold") + y_label("Silver") +
  title("Five countries, and no names")
```

*"Given the medals: points, x is gold, y is silver."*

Bind `label` and draw the names themselves at those positions, and each mark
now shows which row it is:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  x_label("Gold") + y_label("Silver") +
  title("The same five, named")
```

That is the whole difference between the two marks: `point` draws a glyph *at*
the row, `text` draws the row's own value *as* the glyph.

## Both at once: the labeled scatter

They compose. Layer a `point` and a `text` over one `(x, y)` and you get a
dot located precisely *and* its name attached: a labeled scatterplot built
from two marks, not a third mark made for the job. A label would otherwise
land right on the dot, so nudge it off the dot with `style(nudge = "up")`:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "up") +
  x_label("Gold") + y_label("Silver") +
  title("Located by the dot, named just above")
```

There is no "labeled point" mark to learn: a labeled scatter is just
`point + text`, layered with the `+` [operator](../operators.qmd).

Five names fit here. Fifty would not, and choosing which few to keep is the
hardest part of a labeled scatter. In the web edition you can try them first:
clicking a point [stamps its values](../selection.qmd#stamping-a-point) onto
the plot, so you can see how a few labels look before you write any of them.
A stamp disappears when the page reloads; a `text` label is part of the
specification, so it is drawn every time.

### Nudging the label off its point

`style(nudge = )` shifts a text label a little off its `(x, y)`, one of `"up"`,
`"down"`, `"left"`, or `"right"`:

```{r}
data(medals) + point + style(size = 5, color = "#9e9e9e") +
  text + x(gold) + y(silver) + label(country) + style(nudge = "right") +
  x_label("Gold") + y_label("Silver") +
  title("The same, nudged right")
```

The distance is **derived from the font size**, just enough to sit outside a
default dot, so there is nothing to specify; `nudge` takes only the
direction. It is a *setting*, not a collision modifier, and the difference is
exact. `jitter` and `dodge` compute their offsets from the data, so every
mark moves by its own amount. A nudge is one **fixed** shift, the same for
every label, derived from nothing in the data. It is `text`-only:
nudging a `point` or `bar` would move the data itself, so the engine refuses it
there with direction.

### Moving the labels off each other: `repel`

A nudge shifts every label the same way, so it moves each label off its own dot
and does nothing else. When the labels collide with *each other*, one fixed shift
moves them all together and the words still overlap. Thirty European countries are
enough to show it:

```{r}
#| echo: false
peek(gm_europe)
```

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Thirty countries, named")
```

`repel` solves that, and it is a *collision modifier* rather than a setting. It
measures where the labels and the dots actually are, then moves each label until no
two words overlap:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same thirty, moved apart")
```

The modifier does three things in that picture. Every label sits just
outside its dot, so no word covers the data it names. A label that moved far keeps
a thin line back to its point, so you can still see which point it names. And each
label rests beside its dot, on whichever side has room. `repel` pulls a label
toward its dot and pushes it away from every word and every dot. A label with no
neighbors sits just above its dot. One with neighbors moves to whichever side is
free.

The placement is **deterministic**. It depends only on the labels and the rows,
never on a random-number generator or the system time, so the same specification
draws the same picture every time.

Where no arrangement fits, `repel` still draws every label and reports how many
still overlap. No label is removed without a message. A plot missing the names it
could not fit would look like a plot of fewer countries, so gog reports the
count instead. A larger plot, a smaller `style(size = )`, or fewer rows will separate the
labels.

The setting and the collision modifier compose, and each does one thing.
`style(nudge = )` says which side a label prefers, and `repel` resolves the
overlaps that remain:

```{r}
data(gm_europe) + point + style(color = "#9e9e9e") +
  text * repel + x(gdp) + y(life) + label(country) + style(nudge = "right") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Nudged right first, then moved apart")
```

## Numeric labels: the value on each mark

A `label` need not be a name. Bind a numeric column and its value is drawn,
formatted so a whole number is drawn without a decimal point. Layer `text` over `bar` to
print each bar's height at its top:

```{r}
data(medals) + bar + style(color = "steelblue") +
  text + x(country) + y(gold) + label(gold) +
  y_label("Gold medals") +
  title("Each bar labeled with its height")
```

## Color

`color` splits the labels by group and earns a legend, the same discrete split
`line` and `bar` make. Here each country's name takes its own hue:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) + color(country) +
  x_label("Gold") + y_label("Silver") +
  title("One hue per country")
```

## Styling: one color, one size

`style(color = )` sets the ink and `style(size = )` the font size, in pixels,
for the whole layer:

```{r}
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(color = "seagreen", size = 17) +
  x_label("Gold") + y_label("Silver") +
  title("Bigger, greener labels")
```

A per-row font size (a word cloud, where the biggest word is the most
frequent) is not built yet; for now `size` is a layer setting, not a mapped
channel.

## Why `label` is a channel

`text` **requires** a `label`: it is the mark's minimum syllable, the way `x` and
`y` are. Ask for a `text` mark with no `label` and the engine refuses with
direction rather than drawing blank:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver)
```

The reason `label` is a channel and not, say, a `text("USA")` argument is the
grammar's own rule: a string is *data*, and data reaches a mark only through a
channel, never as an argument to the mark. So a fixed note comes from a
one-row table (`data`), not a setting. Only `text` accepts `label`; every
other mark refuses it, which is why a labeled scatter is `point + text` and
never a `label` on `point`. `text` is the shape; `label` is the column it draws.

## A note is a layer

That one-row table is worth seeing drawn, because it is the whole of what
ggplot2 spells as an `annotate()` function. A second `data()` opens a second
layer over its own table (the [Data](../data.qmd) chapter's rule), and a table
with one row in it puts one string on the plot:

```{r}
notes <- data.frame(gdp = 35000, life = 66, what = "high income, short lives")
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(notes) + text + label(what) +
  title("A note is a layer, not a function")
```

Nothing was added to the grammar to reach that. `data()` already binds the
nearest table, `text` already draws one string per row, and a note is a table
that happens to have one row. Because the note is data like any other, it is
scaled like any other: place it far from the points and the axes widen to
include it, which is the same rule every other row follows.

One thing to know while writing one. `x` and `y` are shared by the whole plot,
so the note's table has to carry the base table's position column names (`gdp`
and `life` here), and a note that names its own columns needs per-layer
positions, which gog does not have yet. Everything else ggplot2's
`annotate()` does in the data's own units is already sayable: a marker is `point` over the same table, and several notes at
once are simply several rows.

## What you can set

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("text")
```

`nudge` is the text mark's own, and it exists so a layered `point + text` does
not draw the label on top of the dot:

```{r}
data(medals) + x(country) + y(gold) +
  point + style(color = "gray", size = 6) +
  text + label(gold) + style(nudge = "up", size = 11, color = "black") +
  y_label("Gold medals") + title("Labels nudged off their points")
```

There is no distance to give it: the offset comes from the font size, so a
label sits outside its point at every plot size and in every facet. The rest
of the rule, and `repel` for labels that collide with each other, is under
[Nudging the label off its point](#nudging-the-label-off-its-point).

A mapped `size` on `text` is valid grammar the engine does not draw yet, so
`size(population)` is refused as unsupported rather than ignored. Set it instead,
and the per-row font size (a word cloud) stays on the list of things not built.

## What it refuses

A text mark's glyph *is* its content, so it has nothing left for `shape` to vary:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) + shape(country)
```

Form comes from `label` here, for the reason the section above gave. For the
same reason a string takes no texture, being neither a stroke to dash nor a
region to hatch:

```{r}
#| error: true
data(medals) + text + x(gold) + y(silver) + label(country) +
  style(pattern = "hatch")
```
