# Zone

```{r}
#| include: false
source("../R/setup.R")
```

Which region of the plot should the values stay inside? `zone` shades a
rectangle. It is [`rule`](rule.qmd)'s sibling one dimension up, and it works
the same way: a rule takes one *position* and spans the axis it was not given,
and a zone takes a *pair* and spans the axis it was not given a pair for. Give
it pairs on both axes and it is a box.

That "spans the axis it was not given" is the whole reason the mark exists,
because everything else about a rectangle can already be drawn. What could not
be drawn is a rectangle that reaches the **panel**.

## A band across the panel

Here is a run of sales with a target corridor behind it, 150 to 175, whatever
the years happen to be:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) + style(color = "seagreen") +
  data(quarterly) + line + point +
  y_label("Sales") + title("A corridor to land in")
```

*"Given the quarterly figures: x is year, y is sales, a target zone from lower to upper, and also a line with points."*

`bounds(lower, upper)` names the two columns holding the bottom and top. Nothing
names a left or a right, so the zone takes those from the panel and runs edge to
edge. The zone is written *before* the line so the data draws on top of it; a
highlight belongs behind what it highlights.

## A band up the panel

Swap which pair you give it and the rectangle stands up instead:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) +
    style(color = "indianred") +
  data(quarterly) + line + point +
  y_label("Sales") + title("Two slumps, from one table")
```

Two shaded years, and only one `zone` in the sentence. **One row is one
rectangle**, so a two-row table draws two bands, the same habit that lets one
`rule` table draw three thresholds. Add a row and a third band appears with no
change to the plot.

Look at where the red reaches: the top and bottom of the panel exactly. That is
the thing a `ribbon` cannot do. A ribbon is bounded by its data, so it stops at
whatever numbers you hand it, and any number big enough to reach the top would
*widen the axis to include it*, changing the plot in order to decorate it.

The pair is called `start`/`end` rather than `left`/`right` on purpose. Left and
right are facts about a screen; this grammar reads orientation off the bindings,
and the same words have to keep working when the domain axis is an angle.

## Both pairs: a box

Name all four and neither axis comes from the panel:

```{r}
corner <- data.frame(start = 2015.0, end = 2019.0, lower = 145.0, upper = 190.0)

data(quarterly) + x(year) + y(sales) +
  data(corner) + zone * bounds(lower, upper, start = start, end = end) +
    style(color = "goldenrod", opacity = 0.3) +
  data(quarterly) + line + point +
  y_label("Sales") + title("The stretch before the fall")
```

So one atom covers the three shapes people actually want (a horizontal band, a
vertical band, and a box), and which one you get is decided by which pairs you
name. There is no `hband`, no `vband`, no `rect`.

## Which axis the pair bounds

`lower` and `upper` bound the *measure* axis. Every plot so far has measured along
`y`, so every band so far has run across the panel. Now put a category on `y`
instead. The measure moves to `x`, and the same two columns draw the plot on its
side:

```{r}
stage <- c("Website visit", "Downloads", "Potential customers",
           "Requested price", "Invoice sent")
n     <- c(39, 27.4, 20.6, 11, 2)
funnel <- data.frame(stage = factor(stage, levels = stage),
                     n = n, lo = -n / 2, hi = n / 2, mid = 0)

data(funnel) + zone * bounds(lo, hi) + y(stage) + style(opacity = 1) +
  text + label(n) + x(mid) +
  x_label("") + y_label("Stage") + title("A funnel, and no funnel atom")
```

That is a funnel chart. No atom in the sentence is a funnel.

The blocks are the `zone`, one row each. The columns `lo` and `hi` hold `-n/2` and
`n/2`, computed in R before the plot. That is the same kind of host arithmetic the
waterfall uses for its running total. The numbers inside the blocks are a second
layer, drawn by `text`.

A funnel is a bar chart whose blocks are centered on zero. Centering is arithmetic
on two columns. It is not a chart type.

The plot works because of how the pair is named. `lower` and `upper` describe the
*measure* axis. They do not mean the bottom and the top of a screen. GOG never asks
which way the screen runs; it asks what each axis carries. Here `y` carries a
category, so `y` holds slots, and the measurement can only go on `x`. This is the
rule [`bar`](bar.qmd) already follows: one categorical position, one measured
position, and their types decide the orientation. It is why there is no `flip`.

## Coloring the zones

Each row is its own rectangle, so `color` maps per row. `recessions` carries a
name for each slump:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) + color(slump) +
  data(quarterly) + line + point +
  y_label("Sales") + title("Named, and keyed by the legend")
```

Unlike every other fill in the book, a zone's `color` also takes a **number**,
reading the sequential ramp instead of a palette. A hairline has no room to
decode a scale from, which is why `rule` and `line` refuse it; a rectangle has
nothing but room. That is the property a heatmap is made of, and it is why a
heatmap cell is this mark rather than a new one. The next section is that plot.

## Rectangles that *are* the data

Every zone so far has been a highlight: a rectangle behind a line, saying where
to look. Nothing says it has to sit behind anything. Give a zone a pair on the
measure axis and a categorical position, and each category's rectangle runs from
one number to another with no plot underneath — which is what the two charts
finance draws most are made of.

A **waterfall** is a running total, broken into the steps that made it. GOG has
no transform for that and the omission is deliberate: `stack` accumulates
*within* a position, one pile per slot, and a waterfall accumulates *across*
positions. So the running total is arithmetic, and arithmetic belongs where the
data lives:

```{r}
w <- cashflow
w$top  <- cumsum(w$delta)          # where each step ends
w$base <- c(0, head(w$top, -1))    # and where it started: the step before
w$base[w$total] <- 0               # except the subtotals, which stand on the floor
w$dir  <- ifelse(w$total, "total", ifelse(w$delta >= 0, "gain", "loss"))

slot <- seq_len(nrow(w)) - 1       # category k sits at k on a categorical axis
w$left  <- slot - 0.35             # so a side named at k ± 0.35 is inside its slot
w$right <- slot + 0.35
```

That is R because it is R's job, and it stays R in the other three: every
binding's answer to a running total is the host's own. What follows is the
sentence, and it has four spellings like every other:

```{r}
data(w) + zone * bounds(base, top, start = left, end = right) +
  x(step) + color(dir) + style(opacity = 1) +
  palette(c("steelblue", "seagreen", "indianred")) +
  y_label("Cash (£k)") + title("Where the money went")
```

Six lines of host arithmetic and one sentence. The bars that rest on the floor
are not an exception the grammar had to learn; they are rows whose `base` is 0,
decided by the accounts and handed over like every other number.

Two parts of that sentence are worth pausing on, because both are this mark's
defaults being overruled on purpose.

`style(opacity = 1)` is there because a zone whose sides you **named** is
drawn translucent — 20%, on the reasoning that a rectangle someone chose is
nearly always a highlight and something is nearly always drawn over it. Here
nothing is: the rectangle *is* the measurement. A cut zone (a
[heatmap](#cells-cut-from-the-data-the-heatmap) cell) is already opaque for
the same reason, and this is how you say the named one is data too.

The `left`/`right` pair buys the gap between bars. Left to itself a zone fills
its slot whole, where a [bar](bar.qmd) takes four fifths and the fifth of air
says the categories are separate — a zone's extent is normally *constitutive*,
the region a category owns being the thing the mark names. A waterfall bar's
width means nothing, so it wants the bar's manners rather than the zone's, and
naming the sides is how to ask. The arithmetic works because a categorical axis
puts category *k* at *k*, so the slot it owns is exactly `[k−½, k+½]` and
`k ± 0.35` sits inside it. Drop those two columns and the rectangles touch,
which reads as one connected shape rather than six steps.

A **candlestick** is the same idea with the rectangle floating. Each session has
four numbers, and GOG needs no new channels for them, because the pair a
[ribbon](ribbon.qmd) and an [interval](interval.qmd) already read is the same
pair a rectangle's sides are. Low to high is a whisker; open to close is a
rectangle; that is two layers over one table:

```{r}
s <- sessions
s$body_lo <- pmin(s$open, s$close)
s$body_hi <- pmax(s$open, s$close)
s$dir     <- ifelse(s$close >= s$open, "up", "down")

slot <- seq_len(nrow(s)) - 1       # the same slot arithmetic the waterfall used
s$left  <- slot - 0.3              # a candle body is narrower still
s$right <- slot + 0.3
```

```{r}
data(s) + x(session) +
  interval * bounds(low, high) + color(dir) +
  zone * bounds(body_lo, body_hi, start = left, end = right) +
    color(dir) + style(opacity = 1) +
  palette(c("seagreen", "indianred")) +
  y_label("Price") + title("Fourteen sessions")
```

`color(dir)` is said twice, once per layer, and that is [scope](../encoding-scope.qmd)
working rather than repetition: a channel written after a mark belongs to that
mark, so the wick takes its session's color along with the body. Both readings
of the same column earn *one* legend, because it is one column.

There is no `candlestick` atom and there will not be one. The chart is a
composition of things that already exist, which is the whole claim of the
grammar: [Chart names](../chart-names.qmd) lists it beside the pie and the
rose, which are also sentences rather than words. What the two charts share is
the shape of the answer. Four numbers a session: two of them a *span*, two of
them a *rectangle's sides*. The pair a band mark already reads turns out to be
exactly the pair a rectangle needs. That is why neither chart cost a channel.

## Cells cut from the data: the heatmap

So far every zone has been a rectangle you *chose*, its sides named by columns
you already had. There is a second way to get sides, and it needs no new mark:
let `bin` cut them.

Gapminder has 1704 country-years. Drawn as points, the crowded middle is a
single blot, because a thousand dots in one place look exactly like two hundred:

```{r}
data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Where the country-years actually are")
```

Read the ridge running up the middle: wealth and life expectancy travel
together, and the darkest cell says how many country-years sit at that exact
combination, which no scatter of overlapping dots can tell you.

Three things in that sentence are worth slowing down on.

**`bin` cut both axes, and nothing asked it to.** It is the same transform that
makes a histogram, and how many axes it cuts is read off the mark rather than
requested. A `bar` leaves one axis free to measure a height along, so `bar *
bin` cuts the other one. A zone measures nothing by height, and needs an extent
on *both* axes before it can be a rectangle at all, so `zone * bin` cuts every
axis that has a width to cut, which here is both of them. One transform, one
word, readings decided by what the mark needs. There is no `bin2d`.

The qualifier in that sentence is doing real work, and the mixed mesh below is
where it pays: an axis that arrived already cut, because its values are
categories, is left exactly as it is.

**The count became the color, and nothing bound it.** Both positions were
spoken for, so the measurement went to the one channel left that can carry a
number, and the legend named itself. This is the courtesy `bar * bin` already
does for the y axis, one channel over: a transform that invents a column also
says where it goes.

The column it invents is called `count`, and you may name it out loud. That is
not a different plot, it is the same sentence with nothing left implied:

```{r}
short <- data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life)
long  <- data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
           color(count)

identical(render_svg(short), render_svg(long))
```

So why write it? Because **a binding is where a scale goes**, and the short form
has nowhere to hang one. Counts are badly skewed here: a handful of crowded
cells run to forty-odd country-years while most sit at one or two, so on a
linear ramp nearly every cell is the same pale blue and the plot has one
readable feature. Name the channel and you can say `scale = "log"` on it, the
same way you already did on `x`:

```{r}
data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
  color(count, scale = "log") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same cells, counted on a log ramp")
```

Now the sparse tail separates and the ridge is a gradient rather than a blot.
Look at the legend's middle label: 6.32, not 20.5, because the midpoint of a log
ramp is the geometric mean, so the strip names the color it actually paints
there.

**The white gaps are not zeroes.** A cell with no country-years in it is left as
panel rather than painted the pale end of the ramp, because painting it would
claim a measurement nobody made. The ragged edge you get instead *is* the shape
of the data's support, which is information rather than an unfinished job.

`bin(24)` sets the mesh, exactly as it does on a histogram, and one number cuts
both axes into that many bins. Coarser tells you less and lies less about
precision:

```{r}
data(gm_all) + zone * bin(8) + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same data, eight bins a side")
```

Notice the log scale carried through untouched. The cells are cut in log space,
so they are even on the page rather than even in dollars, which is what you
want when the variable spans from a few hundred dollars to a hundred thousand.

A binned zone also drops the translucency, because the reason for it is gone: a
zone is normally drawn *under* your data and has to let it show through, and
here the zone *is* the data, with nothing behind it to reveal.

## Hexagons: the other mesh

The cells above are rectangles, and that is a choice rather than a fact about
binning. Say `tiling = "hex"` and the plane is cut a different way:

```{r}
data(gm_all) + zone * bin(20, tiling = "hex") + x(gdp, scale = "log") + y(life) +
  palette("viridis") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same country-years, hexagonal cells")
```

Two things changed there, and only one of them is the mesh. `palette("viridis")`
swaps the sequential ramp, and it earns its place on a count: viridis is built so
that equal steps in the number are equal steps in perceived brightness, which the
default blue ramp does not promise. On a plot whose entire message is *how many*,
that is the difference between reading the ridge and guessing at it. The palette
is a free choice on either mesh, though, so do not read it as something hexagons
require.

The mesh itself is not decoration, and the reason to prefer it is a defect in the
rectangular one rather than a preference about shapes. A square grid lines its
cell centers up in rows *and* columns, and the eye is very good at seeing rows
and columns: it reads that alignment as if it were structure in the data.
Wilkinson puts it plainly, that rectangular bins "lead the eye to align bin
centers and to see regularity where there is none" [@wilkinson2005], which is why
Carr devised hexagon binning in 1987 [@carr1987]. A hexagonal mesh staggers alternate rows, so there is
no aligned lattice left to mistake for a finding.

To see that on its own, hold the palette still and change only the mesh:

```{r}
data(gm_all) + zone * bin(20) + x(gdp, scale = "log") + y(life) +
  palette("viridis") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Rectangular cells, same palette")
```

Look along the sparse edges of the two. The rectangular cells line up into faint
rows and columns that carry on past where the data thins out; the hexagons do
not, and what is left at the edge is the shape of the data rather than the shape
of the mesh.

### Why the mesh belongs to `bin` and not to the mark

Everywhere else in this book, a transform computes and the **mark** decides how
the result is drawn: `bar * bin`, `line * bin` and `step * bin` are one histogram
painted three ways. So it would be reasonable to expect the hexagon to be a mark,
or a setting on one. It is neither, and the reason is worth having.

Those three marks all receive **the same numbers**. A hexagonal mesh staggers its
rows, which changes *which rows land in which cell*, so the counts themselves
come out different. You cannot draw a hexagonal plot from rectangularly-binned
data, however clever the mark is. The mesh is upstream of any drawing, so it
belongs to the transform that does the cutting.

That is also how Wilkinson organizes it. His `bin` is not one method but a
family of ways to partition a plane, with `rect` and `hex` among the members.
The mesh both shapes the cells *and* tags each row with the cell it fell in. One
operation, both effects.

The consequence you can see: a tiling means nothing to a bin that cuts one axis,
because there the cells are intervals and an interval has no shape.

```{r}
#| error: true
data(gm_all) + bar * bin(tiling = "hex") + x(life)
```

### One caveat, and it is honest rather than hidden

The hexagons are regular in the space the mesh is cut in, which normalizes both
axes to the same number of steps. The panel then stretches that space by whatever
its own proportions are, so on a wide panel the cells come out slightly wide. Look
closely at the plot above and you can see it.

This is what the original `hexbin` package exposes as its `shape` argument. The
real fix is for the panel to be told it wants equal proportions, which is the
same thing a `polar` plot needs for its circles to be round, and GOG does not
have it yet. Until then, a roughly square panel gives
roughly regular hexagons.

## Cells the data estimated: `zone * density`

`bin` counts what landed in each cell. There is a second thing worth measuring at
a cell, and it needs no new mark either: how *thick* the cloud is there. That is
`density`, the transform that draws a smooth curve on a [line](line.qmd), read in
two dimensions for exactly the reason `bin` was.

```{r}
data(iris_flowers) + zone * density +
  x(sepal_length) + y(petal_length) +
  title("Where the flowers cluster")
```

Every sentence from the heatmap above carries over word for word. Both axes were
cut and nothing asked for it, because a zone measures nothing by height and needs
an extent on both before it can be a rectangle at all. The measurement went to
`color` with nothing bound, and the column it invents is called `density`, so
`color(density)` names it out loud and draws the identical plot. `zone * bin` and
`zone * density` are one plot measured two ways, counted and estimated.

The difference is the white gaps, and they are gone. A count of zero says the
data did not go to that cell, so `bin` leaves it unpainted. An estimate exists
at every point of the plane. There is no cell the estimator had no opinion
about, so the mesh is painted edge to edge. What it paints is a real density
rather than a shade: the cell values times the cell areas sum to one, the same
property that makes the curve's `y` a density.

`palette()` chooses the ramp, exactly as it does for a binned heatmap, and the
argument for viridis on a count is the argument for it here:

```{r}
data(iris_flowers) + zone * density +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("Filled density, viridis")
```

Look at the dark floor away from the two clusters: it is flat. Cells drawn
side by side are each smoothed against the background independently. The pixel
where two of them meet therefore keeps a few percent of the panel showing
through. On a pale ramp that is invisible; on this one it drew a fine lattice
over the whole field. That is the artifact the section above gives as the
reason hexagons exist, the eye reading the mesh's own alignment as though it
were in the data, arriving this time from the rendering side rather than from
the mesh. Cut cells now switch that smoothing off so their edges land on whole
pixels.

The knob a painted field takes is `adjust`, the multiplier on the bandwidth the
estimator chose for itself. It is dimensionless, which is why it is the one knob
that means the same thing on both axes:

```{r}
data(iris_flowers) + zone * density(adjust = 1.6) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("The same cloud, smoothed harder")
```

The same field *traced* rather than painted is [`path * density`](path.qmd), the
contour, and the division between them is the one `bar * bin` and `line * bin`
already made: the mark chooses the geometry, the transform stays constant.

## Bands the data cut: `zone * density(levels = )`

Ask for levels and the field stops being continuous. It is cut into that many
steps, and a zone fills between them:

```{r}
data(iris_flowers) + zone * density(levels = 8) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("Filled density bands")
```

That is the **filled contour**, and it is the same parameter `path` takes. `levels`
means one thing — cut the field into this many — and then each mark does what it
does: a path traces the boundaries, a zone fills between them. So the two are not
two features but one sentence read by two marks, and a band's edge is the contour
line to the pixel:

```{r}
data(iris_flowers) + path * density(levels = 8) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("The same eight levels, traced")
```

Which makes three readings of one transform, and the mark decides every time:
`line * density` is a curve, `zone * density` a painted field, and `levels` turns
the field into level sets that `path` strokes and `zone` fills.

The bands nest, and that is why plain filled shapes are enough to draw them. A
denser region is always *inside* a less dense one, so GOG paints the outermost band
first and lets each inner one cover it; nothing has to be cut out. The exception is
a crater — a ring-shaped cluster with a hollow middle, points scattered around a
circle — where the hollow really is a hole, and painting the band over it fills the
hollow in. GOG has no mark for a shape with a hole, so that is the one density this
reading draws wrongly, and the traced form shows it correctly.

## Cells the axes already made: the tile plot

Three ways to get sides so far, and every one of them put the sides in a column:
`bounds` named them, `bin` cut them, `density(levels = )` traced them. There is a
fourth, and it needs no column at all.

A **categorical axis is already divided.** Six weeks on the x axis is six
slots. The slot "Week 3" owns is a place on the panel with a left edge and a
right edge, both fixed by the axis when it laid the categories out. So a
category bounds its own axis, and two categorical positions bound a zone
completely, with no transform in the sentence:

```{r}
data(six_weeks) + zone + x(week) + y(weekday) + color(orders) +
  title("Six weeks of orders, a day at a time")
```

That is a calendar heatmap, and the whole of it is `zone + x + y + color`. What makes
it worth drawing is that 42 orders carry two rhythms at once and a single column of
numbers shows neither, where a grid separates them by direction. Read *down* a
column for the week (four is the pale one, the trough of the run) and *across* a row
for the weekday, where Thursdays are consistently darkest and Fridays lightest. The
extremes are the corners the two rhythms agree on: the darkest cell is Thursday of
week six, the palest the Friday in week four.

The rule underneath is one sentence: **a category owns a slot, a number is a point.**
A point has no width, which is why two continuous positions still leave a zone with
nothing to bound it, and why `bin` had to be asked to cut them into cells first.
Nothing else changed, and nothing was added to the kernel: no mark, no transform, no
channel.

### `bin` cuts, `count` tallies

The calendar table had one row per cell already. When it does not, when you hold the
raw observations and want to know how many fell in each cell, the transform that
tallies is `count`, exactly as it is for a bar chart:

```{r}
data(winds) + zone * count + x(direction) + y(season) +
  title("Wind observations by direction and season")
```

That is the confusion matrix's shape, and a confusion matrix is the same sentence with
different columns: `zone * count + x(actual) + y(predicted)`.

Notice what was *not* needed there. `bin` cuts a continuous axis into cells before
anything can be counted in them; a categorical axis arrives already cut, so there is
nothing left to do but tally. That is the same division `bar * bin` and `bar * count`
have always had, now with both axes in it, and it is why each refuses toward the
other. Ask a tile plot for a continuous axis and it sends you to `bin`; ask a heatmap
for *two* categorical ones, leaving it nothing at all to cut, and it sends you to
`count`.

The second dimension changes only where the answer goes. A bar has a length to spend
it on, so `bar * count` writes its tally to the y axis; a zone measures nothing by
length, so `zone * count` writes it to `color` and the legend names itself. Both
positions were spoken for, so there was nowhere else for it to go.

`proportion` is the same tally as a share of the whole, which is worth having when the
absolute numbers are not the point:

```{r}
data(winds) + zone * proportion + x(direction) + y(season) +
  palette("viridis") + title("The same observations, as shares")
```

That share is of *every* row counted, not of a row or a column of the grid. A
confusion matrix normalized per true class asks a different question, *which margin
normalizes*, and GOG does not answer it by giving one word two meanings.

It reads a cut mesh the same way, because `proportion` divides whatever measured the
cells and does not care what did. Compose it with `bin` and the heatmap's counts come
back as fractions of the whole — the [relative-frequency
histogram](../transforms.qmd#normalizing-something-other-than-a-tally) with a second
axis:

```{r}
data(gm_all) + zone * bin(20) * proportion + x(gdp, scale = "log") + y(life) +
  palette("viridis") + title("The heatmap, as shares of all 1704 rows")
```

### `count` tallies rows, `mean` reduces a column

The wind table holds more than which cells the observations fell in. Each row also
carries a **speed**, and the question "how fast, on average, in each cell?" is not a
tally at all — it is the other kind of statistic, the kind you point at a column:

```{r}
data(winds) + zone * mean + x(direction) + y(season) + color(speed) +
  title("Mean wind speed, by direction and season")
```

Compare that with `zone * count` two plots up. The sentence is the same shape and one
word longer, and the extra word is the whole difference: `count` was handed no column,
so it counts rows and names its own answer; `mean` was handed one, so `color(speed)`
says which. All five summaries read a mesh this way — `sum`, `mean`, `median`, `max`
and `min` — and each writes its answer back into the column it reduced.

This is where the sentence a page above has to be finished. "Both positions were
spoken for, so there was nowhere else for the answer to go" is true, and color is
where it went. What follows from that, and had been missed, is that **color is
therefore the channel that names the column too** — because a summary reduces *in
place*. On a flat bar, `bar * mean + x(continent) + y(life)` names `life` with `y` and
puts the mean back on `y`; the source and the destination are one binding. A zone
measures by color, so `color` plays both parts.

Which gives the rule in one sentence, and it is the same subtraction the heatmap's
dimensions come from: **a summary groups by every position the mark does not measure
with, and reduces the column named on the one it does.** A bar measures with `y`, so
it groups by `x`. A zone measures with nothing positional, so it groups by both. In
[the cube](../space.qmd) a bar measures with `z`, so it groups by the pair as well.

Say `mean` with nothing for it to reduce and the engine asks for the column rather
than guessing:

```{r}
#| error: true
data(winds) + zone * mean + x(direction) + y(season)
```

And a cell has to be a cell. These five *measure* without cutting, so they need axes
that arrive already divided: a number is a point, and a point owns no cell to
summarize into.

```{r}
#| error: true
data(winds) + zone * mean + x(speed) + y(season) + color(bearing)
```

Read the refusal rather than only the fact of it. It does not say *put a category
there*, it says *cut the axis first*, and names the transform that cuts.

### Cells cut, then summarized: the summary heatmap

A category owns a slot, and that is one way an axis can arrive already divided.
Cutting a continuous axis is the other, and `bin` is what cuts. Compose the two and
each supplies half of what a cell needs:

```{r}
data(gm_all) + zone * bin(15) * mean + x(year) + y(gdp, scale = "log") +
  color(life) + title("Mean life expectancy, by era and income")
```

`bin(15)` cuts the plane into cells, `mean` reduces `life` inside each, and `color`
both names that column and carries the answer, exactly as it does on the tile plot
above. Read the panel bottom-left to top-right and the two stories separate: life
expectancy rises with income going up the panel, and rises again with time going
across it.

This is the same mesh a histogram of those two columns would cut. Nothing about
`bin` changes when it is composed; it stops answering a question nobody asked it,
which was how many rows fell in each cell. So the plot above and `zone * bin(15) +
x(year) + y(gdp, scale = "log")` tile the panel identically, and can be read against
each other: one says how many observations there are in a cell, the other says what
they were like.

The mixed case follows without a new rule, because the two halves are still stated
one axis at a time. Cut the income axis, leave the continents in their slots:

```{r}
data(gm_all) + zone * bin(12) * mean + x(gdp, scale = "log") + y(continent) +
  color(life) + title("Mean life expectancy, by income band and continent")
```

### One of each: the mixed mesh

The pair above was stated one axis at a time, which leaves a case in the middle: one
axis continuous, one categorical. Nothing new has to be decided for it. `bin` cuts the
axis that has a width to cut and leaves alone the one that arrived already cut, so
what comes out is a row of cells per category:

```{r}
data(gm_all) + zone * bin(20) + x(life) + y(continent) +
  palette("viridis") + title("Life expectancy, a distribution per continent")
```

That is a distribution shown as shade rather than as height, which is what makes five
of them fit in the space one histogram would take. Read a row and it is a histogram
lying on its side. Read the plot and the shape of each continent's spread is the point:
Africa's mass sits far left and Europe's far right, Asia and the Americas are wide
where Africa and Europe are concentrated, and Oceania is two countries and says so.

Two things about it are worth naming, because they are what make it a *mesh* and not
five plots side by side. Every row is cut on **the same edges** — the layout is
computed once from the whole column, never per category — so a column of cells is a
straight comparison and the boundaries line up all the way down. And an **empty cell
stays empty**: Oceania's row simply starts where its data does, rather than being
padded with the floor of the ramp, so the ragged left edge is the support of each
distribution rather than a claim that anyone lived to 25 there.

The mirror is the same sentence with the axes swapped, and it is the reading that
stacks the distributions upright:

```{r}
data(gm_all) + zone * bin(20) + y(life) + x(continent) + palette("viridis")
```

Nothing was added for either. Three cases, one rule read one axis at a time: cut both
axes and it is the heatmap, cut neither and it is the tile plot, cut one and it is
this.

### Half a mesh: bounded by one slot

The rule was stated one axis at a time, so read it that way and a zone bounded on
*one* categorical axis falls out. It takes that slot and spans the panel on the other,
which is what this mark has done from the beginning. One row is still one rectangle,
so the table you give it is the list of slots to shade:

```{r}
prevailing <- data.frame(direction = c("SW", "W"))

data(winds) + point * jitter + x(direction) + y(speed) +
  data(prevailing) + zone + x(direction) + style(color = "goldenrod") +
  data(winds) + point * jitter + x(direction) + y(speed) +
  y_label("Speed") + title("The two directions the wind actually comes from")
```

Nothing was built for that. It is the same sentence as the tile plot with one axis
left unsaid, which is the test that the fourth extent is a rule rather than a special
case for one chart.

The two-row table is doing the work [`rule`](rule.qmd)'s table does, and for the same
reason: **a zone's position is a column**, so one table shades as many slots as it has
rows. Hand it the whole `winds` frame instead and you get one rectangle per
observation, hundreds of translucent golds stacked into eight solid columns. That is
the contract behaving exactly as stated, and not the plot you wanted.

One difference worth seeing rather than being told: the column highlight is
translucent and the calendar heatmap is not. A zone that tiles the panel *is* the
data, so there is nothing behind it to show through; a zone bounded on one axis is
background, and the point of it is what you can see underneath. `style(opacity = )`
overrides either.

## What you can set

Everything a zone can be told to look like, generated from the engine's own rule
table so this page cannot drift from what `style()` actually accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("zone")
```

A zone is a filled region, so those are exactly the three a fill has, and no
more. One of them has a default worth knowing: `opacity` is **0.2** for a zone
you placed and **1** for one the data cut, whether `bin` counted its cells or
`density` estimated them. A placed zone is drawn *under* your data and has to let
it show through; a cut zone *is* the data, with nothing behind it to reveal.

All three at once, one per band:

```{r}
early <- data.frame(start = 2006.0, end = 2009.0)
middle <- data.frame(start = 2011.0, end = 2014.0)
late <- data.frame(start = 2017.0, end = 2020.0)

data(quarterly) + x(year) + y(sales) +
  data(early) + zone * bounds(start = start, end = end) +
    style(color = "seagreen") +
  data(middle) + zone * bounds(start = start, end = end) +
    style(color = "indianred", opacity = 0.55) +
  data(late) + zone * bounds(start = start, end = end) +
    style(color = "steelblue", pattern = "hatch", opacity = 0.7) +
  data(quarterly) + line + point +
  y_label("Sales") + title("Color, then opacity, then a hatch")
```

The default opacity is what makes the first band readable without your saying
anything: a highlight that hides the data is not a highlight. Raise it when the
zone is the point of the plot rather than its background, as the red band does
here.

The hatch is worth reaching for when a plot has to survive being printed in gray,
or when two overlapping zones need telling apart by more than hue. Give it more
opacity than a flat fill wants, as the blue band does: a hatch is line-work with
gaps between the lines, so fading it to 0.2 leaves very little to see. It takes
the *fill* textures, not a stroke's dashes; `style(pattern = "dashed")` on a zone
is refused, pointing at the four textures instead.

One setting you might expect is refused, and the refusal says something about the
mark: `style(size = )`, because a zone's extent *is* its bounds and a size would
be a second, contradictory answer to how big it is.

A **border** is not refused, though it was until 2026-07-27, and the reversal
is worth stating because the first answer was reasonable. A zone began as a
highlight behind a line, and a highlight that draws a frame competes with the
data it sits under. The refusal pointed at two `rule`s instead. They really
are better here: dashable, colorable per row, and real positions rather than
decoration:

```{r}
edges <- data.frame(sales = c(target_band$lower, target_band$upper))

data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) + style(color = "seagreen") +
  data(edges) + rule + style(color = "seagreen", pattern = "dashed") +
  data(quarterly) + line + point +
  y_label("Sales") + title("A zone for the region, rules for its edges")
```

The two marks read their positions differently, which is why the edge table is
derived rather than reused. A zone is told which columns hold its sides
(`bounds(lower, upper)`), while a rule reads whichever of the plot's own position
columns its table carries, so the edges have to arrive as a `sales` column.

What changed is that the mark stopped being only a highlight. It draws the
waterfall, the heatmap's cells, the icicle, and the
[mosaic](../transforms.qmd#the-mosaic), and in every one of those the zone *is*
the data rather than the wash behind it. There the composition above stops being
an answer at all: a region has four sides, a mosaic has as many regions as the
table has rows, and two rules per cell hand-placed is not a sentence anyone
writes. So a zone takes `style(border_color = , border_size = )` like every other
closed-glyph fill:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) +
    style(border_color = "black") +
  data(quarterly) + line +
  y_label("Sales") + title("A zone with a rim of its own")
```

Both readings stay available, and which one is right depends on whether the zone
is the background or the subject. Nothing changed for a zone you do not ask: with
no `border_color` and no `border_size` there is no stroke, exactly as before.
Building that column *from* `target_band` is what keeps the two from drifting
apart when the corridor moves.

## What it refuses

A zone with nothing to bound it has no sides:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) + zone + line
```

And half a pair is not a side. The refusal names the half you gave, because that
is nearly always a typo rather than a misunderstanding:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower = lower) + line
```

The domain pair belongs to a rectangle and to nothing else. A band spans the
measure axis *at each position*, so it has no extent along the domain to bound:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) +
  data(recessions) + ribbon * bounds(start = start, end = end)
```

A zone says where its cells are and what is in them, and a transform can
supply either or both. `bounds` names the sides and measures nothing. `count`
tallies into the cells your categories already are, and says nothing about
where those are. `density` estimates a field and samples it. `bin` cuts cells
out of a continuous axis and supplies both, which is why it is also the one
that can give half up. `proportion` answers neither question — it divides
whatever measurement is there by its total — so it stands after any of them.
Two transforms answering the same half is the contradiction, so a `bounds`-ed
rectangle (one per row, sides you named) has no mesh for a summary to group
into:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) * mean + line
```

To shade a band the data *computed* rather than one you chose, the mark that
spans a statistic is [`ribbon`](ribbon.qmd): `ribbon * range` fills between the
minimum and maximum at each x. The division is the same one `rule` draws: a zone
shades where you say, a ribbon shades what the data says.

Two measurements is the same contradiction on the other half. `count` invents one
and `mean` reduces the column you named, and a cell holds a single number. `count`
has no cut to fall back on, so there is nothing left for it to be:

```{r}
#| error: true
data(winds) + zone * count * mean + x(direction) + y(season) + color(speed)
```

That is what separates it from `bin * mean`
[above](#cells-cut-then-summarized-the-summary-heatmap), which is the same
shape and is not a contradiction: `bin` was supplying two things and only one
of them is what makes it a `bin`.

Asking for both extents at once is refused rather than resolved by picking one,
since the two would disagree the moment the mesh moved:

```{r}
#| error: true
data(gm_all) + x(gdp) + y(life) +
  zone * bounds(lower, upper) * bin
```

A `bin` needs *something to cut*, and two categorical axes leave it nothing: a
category is one slot, with no width. The refusal does not stop there, though: a
cell per pair of *categories* is the tile plot, which does no binning at all, so
it names the transform that does tally into cells you already have:

```{r}
#| error: true
data(six_weeks) + zone * bin + x(weekday) + y(week)
```

One categorical axis is a different matter, and it draws — that is the mixed mesh
above. `density` is where the two part company, and it is not an oversight. A field
is estimated *between* the data points, so it needs somewhere to spread on both axes;
a density per category would be normalized inside each slot on its own, which makes
its cells incomparable across slots while the color ramp claims otherwise. It is the
*which margin normalizes* question again, and the answer here is the one `proportion`
gives everywhere: a share is of the whole frame, and a per-slot denominator is a
different plot that has to be asked for. So a field asks for two continuous axes and
says what to reach for instead:

```{r}
#| error: true
data(gm_all) + zone * density + x(life) + y(continent)
```

A tiling parts company for a reason of its own, one dimension along. `hex`
interleaves two lattices, weighing a step up against a step across, so it needs a
*distance* on both axes; a category's slots are an order and no more. A mixed mesh
has two axes and still no plane, and its cells are rectangles by construction:

```{r}
#| error: true
data(gm_all) + zone * bin(tiling = "hex") + x(life) + y(continent)
```

And the mirror of the first refusal, which is the same rule read the other way. A
tally needs cells that exist; a number is a point and owns none, so `count` sends you
back to `bin`:

```{r}
#| error: true
data(gm_all) + zone * count + x(gdp) + y(life)
```

And the cells' measure is already the count, so coloring by anything else asks
for a column that binning has replaced:

```{r}
#| error: true
data(gm_all) + zone * bin + x(gdp) + y(life) + color(continent)
```

A mesh GOG does not cut is refused with the list of the ones it does, rather
than quietly falling back to rectangles:

```{r}
#| error: true
data(gm_all) + zone * bin(tiling = "octagon") + x(gdp) + y(life)
```

`density`'s `bandwidth` is refused here for a reason of its own. It is a width
in one column's own units, and a field spreads over two columns measuring
different quantities, so one number cannot be a width in both. That is why
`adjust` is the knob the section above reaches for: a multiplier means the
same thing on either axis.

```{r}
#| error: true
data(iris_flowers) + zone * density(bandwidth = 0.5) +
  x(sepal_length) + y(petal_length)
```
