# Zone

```{r}
#| include: false
source("../R/setup.R")
```

Which region of the plot should the values stay inside? `zone` shades a
rectangle. It is [`rule`](rule.qmd)'s sibling one dimension up, and it works the
same way. A rule takes one *position* and spans the axis it was not given. A
zone takes a *pair* and spans the axis it was given no pair for. Give it pairs
on both axes and it is a rectangle you placed.

That "spans the axis it was not given" is the whole reason the mark exists,
because everything else about a rectangle can already be drawn. What could not
be drawn is a rectangle that reaches the **panel**.

## A band across the panel

Were sales inside the target each year? Two numbers in a caption let a reader
check, year by year. A band shaded behind the line shows it at once.
The `quarterly` table holds one sales figure for each year, and `target_band`
holds the two edges of a target band:

```{r}
#| echo: false
peek(quarterly)
peek(target_band)
```

Here are the sales with the target band behind them, 150 to 175, whatever the
years happen to be:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) + style(color = "seagreen") +
  data(quarterly) + line + point +
  y_label("Sales") + title("A target band, 150 to 175")
```

*"Given quarterly: x is year, y is sales, a target zone from lower to upper, and
also a line and points."*

`bounds(lower, upper)` names the two columns holding the bottom and top. Nothing
names a left or a right, so the zone takes those from the panel and runs edge to
edge. The zone is written *before* the line so the data draws on top of it; a
highlight belongs behind what it highlights.

## A band up the panel

Some highlights are spans of time rather than of value. The years a recession
lasted are one, when the question is how sales moved through them.
Swap which pair you give it and the rectangle stands up instead. The
`recessions` table holds the two slumps, each with a start year and an end year:

```{r}
#| echo: false
peek(recessions)
```

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) +
    style(color = "indianred") +
  data(quarterly) + line + point +
  y_label("Sales") + title("Two slumps, from one table")
```

*"Given quarterly: x is year, y is sales, recession zones from start to end, and
also a line and points."*

Two shaded spans, and only one `zone` in the sentence. **One row is one
rectangle**, so a two-row table draws two bands, the same property that lets one
`rule` table draw three thresholds. Add a row and a third band appears with no
change to the plot.

Look at where the red reaches: the top and bottom of the panel exactly. That is
the thing a `ribbon` cannot do. A ribbon is bounded by its data, so it stops at
whatever numbers you hand it, and any number big enough to reach the top would
*widen the axis to include it*, changing the plot in order to decorate it.

The pair is called `start`/`end` rather than `left`/`right` on purpose. Left and
right are facts about a screen. This grammar reads orientation off the bindings,
so the same words keep working when the category axis becomes an angle.

## Both pairs: a rectangle

Sometimes the region to highlight is bounded on both axes: these four years,
this range of sales, and nothing outside either. That is a box rather than a
band.
Name all four and neither axis comes from the panel:

```{r}
#| echo: false
peek(sales_box)
```

```{r}
data(quarterly) + x(year) + y(sales) +
  data(sales_box) + zone * bounds(lower, upper, start = start, end = end) +
    style(color = "goldenrod", opacity = 0.3) +
  data(quarterly) + line + point +
  y_label("Sales") + title("A box over four years of sales")
```

*"Given quarterly: x is year, y is sales, a corner zone from lower to upper and
from start to end, and also a line and points."*

So one atom covers the three shapes people actually want (a horizontal band, a
vertical band, and a box), and which one you get is decided by which pairs you
name. There is no `hband`, no `vband`, no `rect`.

## Which axis the pair bounds

A funnel counts how many people are left at each stage, from website visits
down to invoices sent. Its usual picture is a stack of centered blocks, one per
stage, measured across the page rather than up it.
`lower` and `upper` bound the *measure* axis. Every plot so far has measured
along `y`, so every `lower`/`upper` band so far has run across the panel. Now
put a category on `y` instead. The measure moves to `x`, and the same two
columns draw the plot on its side:

```{r}
stage <- c("Website visit", "Downloads", "Potential customers",
           "Requested price", "Invoice sent")
n     <- c(39, 27.4, 20.6, 11, 2)
funnel <- data.frame(stage = factor(stage, levels = stage),
                     n = n, lo = -n / 2, hi = n / 2, mid = 0)

data(funnel) + zone * bounds(lo, hi) + y(stage) + style(opacity = 1) +
  text + label(n) + x(mid) +
  x_label("") + y_label("Stage") + title("A funnel, and no funnel atom")
```

*"Given the funnel: zones from lo to hi, y is stage, and also text, x is mid,
label by n."*

That is a funnel chart. No atom in the sentence is a funnel.

The blocks are the `zone`, one row each. The columns `lo` and `hi` hold `-n/2`
and `n/2`, computed in R before the plot. The waterfall below does the same,
computing its running total in the language you write the sentence in. The numbers
inside the blocks are a second layer, drawn by `text`.

A funnel is a bar chart whose blocks are centered on zero. Centering is
arithmetic on two columns. It is not a chart type.

The plot works because of how the pair is named. `lower` and `upper` describe
the *measure* axis. They do not mean the bottom and the top of a screen. gog
never asks which way the screen runs; it asks what each axis carries. Here `y`
carries a category, so `y` holds slots, and the measurement can only go on `x`.
This is the rule [`bar`](bar.qmd) already follows: one categorical position, one
measured position, and their types decide the orientation. It is why there is
nothing like ggplot2's `coord_flip()` to write.

## Coloring the zones

Two bands in one red are two unnamed slumps. Color each by its name and the
legend says which is which, with no note on the plot.
Each row is its own rectangle, so `color` maps per row. `recessions` carries a
name for each slump:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) + color(slump) +
  data(quarterly) + line + point +
  y_label("Sales") + title("Named, and keyed by the legend")
```

*"Given quarterly: x is year, y is sales, recession zones from start to end,
color by slump, and also a line and points."*

A zone's `color` also takes a **number**, reading the sequential ramp instead of
a palette, where `bar`, `area` and `ribbon` keep to categories. A rectangle is
large enough to read a ramp against. That is what a heatmap needs, so a heatmap
cell is this mark rather than a new one. The heatmap section below is that plot.

## Rectangles that *are* the data

Every zone so far has been a highlight: a rectangle behind a line, saying where
to look. Nothing says it has to sit behind anything. Give a zone a pair on the
measure axis and a categorical position. Each category's rectangle then runs
from one number to another, with no plot underneath. Two common finance charts
are made of exactly that.

A **waterfall** is a running total, broken into the steps that made it. gog has
no transform for that and the omission is deliberate: `stack` accumulates
*within* a position, one pile per slot, and a waterfall accumulates *across*
positions. So the running total is arithmetic, and arithmetic belongs with the
data, in the host language. The `cashflow` table holds one row per step of a
cash statement: a `delta` for the step, and a `total` flag marking the
subtotals:

```{r}
#| echo: false
peek(cashflow)
```

```{r}
w <- cashflow
w$top  <- cumsum(w$delta)          # where each step ends
w$base <- c(0, head(w$top, -1))    # and where it started: the step before
w$base[w$total] <- 0               # except the subtotals, which start at zero
w$dir  <- ifelse(w$total, "total", ifelse(w$delta >= 0, "gain", "loss"))

slot <- seq_len(nrow(w)) - 1       # category k sits at k on a categorical axis
w$left  <- slot - 0.35             # so a side named at k ± 0.35 is inside its slot
w$right <- slot + 0.35
```

That is R because it is R's job, and the other three answer it the same way: a
running total belongs to the host language. What follows is the sentence, and it
has four spellings like every other:

```{r}
data(w) + zone * bounds(base, top, start = left, end = right) +
  x(step) + color(dir) + style(opacity = 1) +
  palette(c("steelblue", "seagreen", "indianred")) +
  y_label("Cash (thousands)") + title("Where the money went")
```

*"Given the w table: zones from base to top and from left to right, x is step,
color by dir, with opacity 1."*

A few lines of host arithmetic and one sentence. The bars that start at zero are
not a special case. They are rows whose `base` is 0, a number the accounts
supplied like every other.

Two parts of that sentence are worth pausing on, because both are this mark's
defaults being overruled on purpose.

`style(opacity = 1)` is there because a zone whose sides you named is drawn
translucent, at 20%. A rectangle someone chose is nearly always a highlight,
with something drawn over it. Here nothing is: the rectangle *is* the
measurement. A cut zone (a [heatmap](#cells-cut-from-the-data-the-heatmap) cell)
is already opaque for the same reason, and this is how you say the named one is
data too.

The `left`/`right` pair makes the gap between bars. Given no sides, a zone fills
its slot whole, where a [bar](bar.qmd) takes four fifths of it and leaves one
fifth empty, which is what separates the categories. A zone's width normally
means something: it is the region the category owns. A waterfall bar's width
means nothing, so it should behave like a bar, and naming the sides is how to
ask. The arithmetic works because a categorical axis puts category *k* at *k*,
so the slot it owns is exactly `[k−½, k+½]` and `k ± 0.35` sits inside it. Drop
those two columns and the rectangles touch, which reads as one connected shape
rather than six steps.

A **candlestick** is the same idea with the rectangle floating. Each session has
four numbers: `open`, `close`, `high` and `low`. gog needs no new channels for
them, because a rectangle's sides read the pair a [ribbon](ribbon.qmd) and an
[interval](interval.qmd) already read. Low to high is a whisker; open to close
is a rectangle; that is two layers over one table:

```{r}
#| echo: false
peek(sessions)
```

```{r}
s <- sessions
s$body_lo <- pmin(s$open, s$close)
s$body_hi <- pmax(s$open, s$close)
s$dir     <- ifelse(s$close >= s$open, "up", "down")

slot <- seq_len(nrow(s)) - 1       # the same slot arithmetic the waterfall used
s$left  <- slot - 0.3              # a candle body is narrower still
s$right <- slot + 0.3
```

```{r}
data(s) + x(session) +
  interval * bounds(low, high) + color(dir) +
  zone * bounds(body_lo, body_hi, start = left, end = right) +
    color(dir) + style(opacity = 1) +
  palette(c("seagreen", "indianred")) +
  y_label("Price") + title("Fourteen sessions")
```

*"Given the s table: x is session, intervals from low to high, color by dir, and
also zones from body lo to body hi and from left to right, color by dir."*

`color(dir)` is said twice, once per layer, and that is
[scope](../encoding-scope.qmd) working rather than repetition: a channel written
after a mark belongs to that mark, so the whisker takes its session's color
along with the body. Both readings of the same column earn *one* legend, because
it is one column.

There is no `candlestick` atom, and the grammar does not need one. The chart is
a composition of things that already exist, which is the whole claim of the
grammar: [Chart names](../chart-names.qmd) lists it beside the pie and the rose,
which are also sentences rather than words. What the two charts share is the
shape of the answer. Four numbers a session: two of them a *span*, two of them a
*rectangle's sides*. The pair a band mark already reads turns out to be exactly
the pair a rectangle needs. That is why neither chart needed a new channel.

## Cells cut from the data: the heatmap

So far every zone has been a rectangle you *chose*, its sides named by columns
you already had. There is a second way to get sides, and it needs no new mark:
let `bin` cut them.

Gapminder has 1704 country-years. Draw them as points and the crowded middle
becomes one dark mass, because a thousand dots in one place look exactly like
two hundred:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Where the country-years actually are")
```

*"Given all the gapminder years: zones derived by bin into 24, x is gdp on a log
scale, y is life."*

`bin` is not the zone's own transform. What the mark changes is how many axes
the cut runs along:

| Mark derived by `bin` | Shape |
|---|---|
| `bar * bin` | one axis cut, drawn as the bars of a histogram |
| `zone * bin` | two axes cut, drawn as the cells above |
| `surface * bin` | the same two axes, with the count raised into the cube |

Read the band of dark cells running up the middle: wealth and life expectancy
rise together. The darkest cell says how many country-years sit at that
combination, which overlapping dots cannot.

Three things in that sentence are worth slowing down on.

`bin` cut both axes, and nothing asked it to. It is the same transform that
makes a histogram, and how many axes it cuts is read off the mark rather than
requested. A `bar` leaves one axis free to measure a height along, so `bar *
bin` cuts the other one. A zone measures nothing by height, and needs an extent
on *both* axes before it can be a rectangle at all, so `zone * bin` cuts every
axis that has a width to cut, which here is both of them. One transform, one
word, readings decided by what the mark needs. There is no `bin2d`.

Note the words *has a width to cut*. The mixed mesh below is where they
matter: an axis whose values are categories arrived already cut, and `bin`
leaves it alone.

The count became the color, and nothing bound it. Both positions were taken,
so the measurement went to the one channel left that can carry a number, and
the legend named itself. This is what `bar * bin` already does for the y
axis: a transform that invents a column also says where it goes.

The column it invents is called `count`, and you may name it out loud. That is
not a different plot, it is the same sentence with nothing left implied:

```{r}
short <- data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life)
long  <- data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
           color(count)

identical(render_svg(short), render_svg(long))
```

So why write it? Because *a binding is where a scale goes*, and the short form
has nowhere to put one. Counts are badly skewed here: the fullest cells hold
about forty country-years while most hold one or two. On a linear ramp nearly
every cell is the same pale blue. Name the channel and you can say `scale =
"log"` on it, the same way you already did on `x`:

```{r}
data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
  color(count, scale = "log") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same cells, counted on a log ramp")
```

*"Given all the gapminder years: zones derived by bin into 24, x is gdp on a log
scale, y is life, color by count on a log scale."*

Now the sparse tail separates and the ridge is a gradient rather than a blot.
Look at the legend's middle label: 6.32, not 20.5, because the midpoint of a log
ramp is the geometric mean, so the strip names the color it actually paints
there.

The white gaps are not zeroes. A cell with no country-years in it is left as
panel rather than painted the pale end of the ramp, because painting it would
claim a measurement nobody made. The ragged edge you get instead *is* the
region the data actually covers, which is information rather than an
unfinished job.

`bin(24)` sets the mesh, exactly as it does on a histogram, and one number cuts
both axes into that many bins. A coarser mesh shows less detail, and claims
less precision:

```{r}
data(gm_all) + zone * bin(8) + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same data, eight bins a side")
```

Notice the log scale carried through untouched. The cells are cut in log space,
so they are even on the page rather than even in dollars, which is what you
want when the variable spans from a few hundred dollars to a hundred thousand.

A binned zone also drops the translucency: here the zone *is* the data, with
nothing behind it to show through.

## Hexagons: the other mesh

Do the rows and columns in a heatmap belong to the data or to the mesh?
The cells above are rectangles, and that is a choice rather than a fact about
binning. Say `tiling = "hex"` and the plane is cut a different way:

```{r}
data(gm_all) + zone * bin(20, tiling = "hex") + x(gdp, scale = "log") + y(life) +
  palette("viridis") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same country-years, hexagonal cells")
```

*"Given all the gapminder years: zones derived by bin into 20 on hexagons, x is
gdp on a log scale, y is life, with the viridis palette."*

Two things changed there, and only one of them is the mesh. `palette("viridis")`
replaces the default sequential ramp, and it is the right choice for a count.
Viridis is built so equal steps in the number are equal steps in perceived
brightness. On a plot whose entire message is *how many*, that is the difference
between reading the ridge and guessing at it. The palette is a free choice on
either mesh, though, so do not read it as something hexagons require.

The mesh itself is not decoration, and the reason to prefer it is a defect in
the rectangular one rather than a preference about shapes. A square grid lines
its cell centers up in rows *and* columns, and the eye is very good at seeing
rows and columns: it reads that alignment as if it were structure in the data.
Wilkinson states it directly: rectangular bins "lead the eye to align bin
centers and to see regularity where there is none" [@wilkinson2005]. That is why
Carr devised hexagon binning in 1987 [@carr1987]. A hexagonal mesh staggers
alternate rows, so there is no aligned lattice left to mistake for a finding.

To see that on its own, hold the palette still and change only the mesh:

```{r}
data(gm_all) + zone * bin(20) + x(gdp, scale = "log") + y(life) +
  palette("viridis") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Rectangular cells, same palette")
```

Look along the sparse edges of the two. The rectangular cells line up into
faint rows and columns that continue past the edge of the data. The hexagons
do not, and what is left at the edge is the shape of the data rather than the
shape of the mesh.

### Why the mesh belongs to `bin`

Everywhere else in this book, a transform computes and the **mark** decides how
the result is drawn: `bar * bin`, `line * bin` and `step * bin` are one
histogram painted three ways. So it would be reasonable to expect the hexagon to
be a mark, or a setting on one. It is neither, and the reason matters.

Those three marks all receive **the same numbers**. A hexagonal mesh staggers
its rows, which changes *which rows land in which cell*, so the counts
themselves come out different. You cannot draw a hexagonal plot from
rectangularly binned data, whatever the mark does with it. The mesh is decided
before anything is drawn, so it belongs to the transform that does the cutting.

That is also how Wilkinson organizes it. His `bin` is not one method but a
family of ways to partition a plane, with `rect` and `hex` among the members.
The mesh both shapes the cells *and* tags each row with the cell it fell in. One
operation, both effects.

The consequence you can see: a tiling means nothing to a bin that cuts one axis,
because there the cells are intervals and an interval has no shape. So `bar *
bin(tiling = "hex")` is refused, as [What it refuses](#what-it-refuses) shows.

### One caveat

The hexagons are regular in the space the mesh is cut in, which normalizes both
axes to the same number of steps. The panel then stretches that space by
whatever its own proportions are, so on a wide panel the cells come out slightly
wide. Look closely at the plot above and you can see it.

This is what the original `hexbin` package exposes as its `shape` argument. The
fix would be a panel told to keep equal proportions, which is what a `polar`
plot needs for round circles. gog has no setting for that, so a roughly square
panel gives roughly regular hexagons.

## Cells the data estimated: `zone * density`

A count per cell is honest and ragged when the table is small. The 150 flowers
spread over a mesh would leave most cells holding one or two, or none.
`bin` counts what landed in each cell. There is a second thing worth measuring
at a cell, and it needs no new mark either: how *thick* the cloud is there. That
is `density`, the transform that draws a smooth curve on a [line](line.qmd),
read in two dimensions for exactly the reason `bin` was.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + zone * density +
  x(sepal_length) + y(petal_length) +
  title("Where the flowers cluster")
```

*"Given the iris flowers: zones derived by density, x is sepal length, y is
petal length."*

Every sentence from the heatmap above is true here word for word. Both axes were
cut and nothing asked for it, because a zone measures nothing by height and
needs an extent on both before it can be a rectangle at all. The measurement
went to `color` with nothing bound, and the column it invents is called
`density`, so `color(density)` names it out loud and draws the identical plot.
`zone * bin` and `zone * density` are one plot measured two ways, counted and
estimated.

The white gaps are the difference, and here there are none. A count of zero says
the data did not go to that cell, so `bin` leaves it unpainted. An estimate
exists at every point of the plane. The estimator returns a value for every
cell, so the mesh is painted edge to edge. What it paints is a real density
rather than an arbitrary color value. The cell values times the cell areas sum
to one, the same property that makes the curve's `y` a density.

`palette()` chooses the ramp, exactly as it does for a binned heatmap, and the
argument for viridis on a count is the argument for it here:

```{r}
data(iris_flowers) + zone * density +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("Filled density, viridis")
```

Look at the dark area away from the two clusters: it is even, and the cells meet
cleanly. Blending each cell's edge into the background would leave a few percent
of the panel showing at every seam. A pale ramp hides that; this one would draw
it as a fine lattice over the whole field, the same false structure the hexagon
section describes, arriving from the renderer rather than from the mesh. So a
cut cell's edges land on whole pixels instead.

The parameter a painted field takes is `adjust`, the multiplier on the
bandwidth the estimator chose for itself. It is dimensionless, which is why
it is the one parameter that means the same thing on both axes:

```{r}
data(iris_flowers) + zone * density(adjust = 1.6) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("The same cloud, smoothed harder")
```

*"Given the iris flowers: zones derived by density adjusted by 1.6, x is sepal
length, y is petal length, with the viridis palette."*

The same field *traced* rather than painted is [`path * density`](path.qmd), the
contour, and the division between them is the one `bar * bin` and `line * bin`
already made: the mark chooses the geometry, the transform stays constant.

## Bands the data cut: `zone * density(levels = )`

A smooth ramp is hard to read a value from: two nearby cells differ by a shade
the eye cannot name. Cut the field into a few steps and each band has a level a
reader can point to.
Ask for levels and the field stops being continuous. It is cut into that many
steps, and a zone fills between them:

```{r}
data(iris_flowers) + zone * density(levels = 8) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("Filled density bands")
```

*"Given the iris flowers: zones derived by density at 8 levels, x is sepal
length, y is petal length."*

That is the **filled contour**, and it is the same parameter `path` takes.
`levels` means one thing (cut the field into this many) and then each mark does
what it does: a path traces the boundaries, a zone fills between them. So the
two are not two features but one sentence read by two marks, and a band's edge
falls exactly on the contour line:

```{r}
data(iris_flowers) + path * density(levels = 8) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("The same eight levels, traced")
```

*"Given the iris flowers: paths derived by density at 8 levels, x is sepal
length, y is petal length."*

That makes three readings of one transform, and the mark decides every time:
`line * density` is a curve, `zone * density` a painted field, and `levels`
turns the field into level sets that `path` strokes and `zone` fills.

The bands nest, and that is why plain filled shapes are enough to draw them. A
denser region is always *inside* a less dense one, so gog paints the outermost
band first and lets each inner one cover it; nothing has to be cut out. The
exception is a crater: a ring-shaped cluster with a hollow middle, points
scattered around a circle. There the hollow really is a hole, and painting the
band over it fills it in. gog has no mark for a shape with a hole, so a crater
is the one field `zone * density` draws wrongly. `path * density` traces it
correctly.

## Cells the axes already made: the tile plot

Three sources of sides so far, and every one of them put the sides in a column:
`bounds` named them, `bin` cut them, `density(levels = )` traced them. There is
a fourth, and it needs no column at all.

Six weeks of daily orders is already a grid: weeks across, weekdays down, one
count in each cell. The grid itself is the picture you want.
A categorical axis is already divided. Six weeks on the x axis is six
slots. The slot "Week 3" owns is a place on the panel with a left edge and a
right edge, both fixed by the axis when it laid the categories out. So a
category bounds its own axis, and two categorical positions bound a zone
completely, with no transform in the sentence:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + zone + x(week) + y(weekday) + color(orders) +
  title("Six weeks of orders, a day at a time")
```

*"Given the six weeks table: zones, x is week, y is weekday, color by orders."*

That is a calendar heatmap, and the whole of it is `zone + x + y + color`. What
makes it worth drawing is that 42 daily counts carry two rhythms at once. A
column of numbers shows neither, while a grid separates them by direction. Read
*down* a column for the week (four is the pale one, the lowest week) and
*across* a row for the weekday, where Thursdays are consistently darkest and
Fridays lightest. The extremes are the cells where the two rhythms agree: the
darkest cell is Thursday of week six, the palest the Friday in week four.

The rule underneath is one sentence: a category owns a slot, a number is a
point. A point has no width, which is why two continuous positions still leave a
zone with nothing to bound it, and why `bin` had to be asked to cut them into
cells first. Nothing else changed, and nothing was added to the kernel: no mark,
no transform, no channel.

### `bin` cuts, `count` tallies

The calendar table had one row per cell already. When it does not, when you hold
the raw observations and want to know how many fell in each cell, the transform
that tallies is `count`, exactly as it is for a bar chart:

```{r}
#| echo: false
peek(winds)
```

```{r}
data(winds) + zone * count + x(direction) + y(season) +
  title("Wind observations by direction and season")
```

*"Given the winds: zones derived by count, x is direction, y is season."*

That is the shape of a confusion matrix, the table that counts predicted classes
against actual ones, and a confusion matrix is the same sentence with different
columns: `zone * count + x(actual) + y(predicted)`.

Notice what was *not* needed there. `bin` cuts a continuous axis into cells
before anything can be counted in them; a categorical axis arrives already cut,
so there is nothing left to do but tally. That is the same division `bar * bin`
and `bar * count` have always had, now with both axes in it, and it is why each
one's refusal names the other. Ask a tile plot for a continuous axis and it
names `bin`. Ask a heatmap for two categorical ones, which leave it nothing to
cut, and it names `count`.

The second dimension changes only where the answer goes. A bar has a length to
put it on, so `bar * count` writes its tally to the y axis; a zone measures
nothing by length, so `zone * count` writes it to `color` and the legend names
itself. Both positions were taken, so there was nowhere else for it to go.

`proportion` is the same tally as a share of the whole, which is worth having
when the absolute numbers are not the point:

```{r}
data(winds) + zone * proportion + x(direction) + y(season) +
  palette("viridis") + title("The same observations, as shares")
```

*"Given the winds: zones derived by proportion, x is direction, y is season."*

That share is of *every* row counted, not of a row or a column of the grid. A
confusion matrix divided by the total in each true class asks a different
question, *which margin normalizes*, and gog does not answer it by giving one
word two meanings.

It reads a cut mesh the same way: `proportion` divides whatever measurement the
cells carry, whichever transform put it there. Compose it with `bin` and the
heatmap's counts come back as fractions of the whole, the [relative-frequency
histogram](../transforms.qmd#normalizing-something-other-than-a-tally) with a
second axis:

```{r}
data(gm_all) + zone * bin(20) * proportion + x(gdp, scale = "log") + y(life) +
  palette("viridis") + title("The heatmap, as shares of all 1704 rows")
```

### `count` tallies rows, `mean` reduces a column

The wind table holds more than the cell each observation fell in. Each row also
carries a **speed**, and the question "how fast, on average, in each cell?" is
not a tally at all; it is the other kind of statistic, the kind you point at a
column:

```{r}
data(winds) + zone * mean + x(direction) + y(season) + color(speed) +
  title("Mean wind speed, by direction and season")
```

*"Given the winds: zones derived by mean, x is direction, y is season, color by
speed."*

Compare that with `zone * count` two plots up. The sentence is the same shape
with one more binding, and that binding is the whole difference: `count` was
handed no column, so it counts rows and names its own answer; `mean` was handed
one, so `color(speed)` says which. All six summaries read a mesh this way
(`sum`, `mean`, `median`, `max`, `min` and `quantile`), and each writes its
answer back into the column it reduced.

This is where the rule stated under the first heatmap gets its second half.
Color is where the answer went, and color is therefore the channel that names
the column too, because a summary reduces *in place*. On a flat bar, `bar * mean
+ x(continent) + y(life)` names `life` with `y` and puts the mean back on `y`;
the source and the destination are one binding. A zone measures by color, so
`color` plays both parts.

Which gives the rule in one sentence, and it is the same rule the heatmap's
dimensions come from: a summary groups by every position the mark does not
measure with, and reduces the column named on the position it measures with. A
bar measures with `y`, so it groups by `x`. A zone measures with nothing
positional, so it groups by both. In [the cube](../space.qmd) a bar measures
with `z`, so it groups by the pair as well.

Say `mean` with nothing for it to reduce and the engine asks for the column
rather than guessing, as [What it refuses](#what-it-refuses) shows.

And a summary still needs a cell to summarize into. These six *measure* without
cutting, so they need axes that arrive already divided: a number is a point, and
a point owns no cell to summarize into. That refusal gives both ways out, a
category on each position or `bin` on the continuous axis first.

### Cells cut, then summarized: the summary heatmap

How does life expectancy depend on income and on time together? Both are
continuous, so neither owns a slot, and a mean needs a cell to compute inside.
A category owns a slot, and that is one way an axis can arrive already divided.
Cutting a continuous axis is the other, and `bin` is what cuts. Compose the two
and each supplies half of what a cell needs:

```{r}
data(gm_all) + zone * bin(15) * mean + x(year) + y(gdp, scale = "log") +
  color(life) + title("Mean life expectancy, by era and income")
```

`bin(15)` cuts the plane into cells, `mean` reduces `life` inside each, and
`color` both names that column and carries the answer, exactly as it does on the
tile plot above. Read the panel from bottom left to top right and the two
patterns separate. Life expectancy rises with income up the panel, and rises
again with time across it.

This is the same mesh a histogram of those two columns would cut. Nothing about
`bin` changes when it is composed. It only stops reporting how many rows fell in
each cell, because `mean` now supplies the measurement. So the plot above and
`zone * bin(15) + x(year) + y(gdp, scale = "log")` tile the panel identically,
and can be read against each other: one says how many observations there are in
a cell, the other says what they were like.

The mixed case follows without a new rule, because the two halves are still
stated one axis at a time. Cut the income axis, leave the continents in their
slots:

```{r}
data(gm_all) + zone * bin(12) * mean + x(gdp, scale = "log") + y(continent) +
  color(life) + title("Mean life expectancy, by income band and continent")
```

### One of each: the mixed mesh

How does the spread of life expectancy differ by continent? Five histograms
answer it, in five panels the reader must compare one by one.
The pair above was stated one axis at a time, which leaves a case in the middle:
one axis continuous, one categorical. Nothing new has to be decided for it.
`bin` cuts the axis that has a width to cut and leaves alone the one that
arrived already cut, so what comes out is a row of cells per category:

```{r}
data(gm_all) + zone * bin(20) + x(life) + y(continent) +
  palette("viridis") + title("Life expectancy, a distribution per continent")
```

That is a distribution shown as shade rather than as height, which is what makes
five of them fit in the space one histogram would take. Read a row and it is a
histogram lying on its side. Read the plot and the shape of each continent's
spread is what stands out: Africa's mass sits far left and Europe's far right,
Asia and the Americas are wide where Africa and Europe are concentrated, and
Oceania is two countries and says so.

Two things about it are worth naming, because they are what make it a *mesh* and
not five plots side by side. Every row is cut on **the same edges**, and the
layout is computed once from the whole column, never per category, so a column
of cells is a straight comparison and the boundaries line up all the way down.
And an empty cell stays empty. Oceania's row starts where its data starts,
rather than being painted the palest end of the ramp. The ragged left edge shows
where each distribution reaches, not a claim that anyone lived to 25 there.

The mirror is the same sentence with the axes swapped, and it is the reading
that stacks the distributions upright:

```{r}
data(gm_all) + zone * bin(20) + y(life) + x(continent) + palette("viridis")
```

Nothing was added for either. Three cases, one rule read one axis at a time: cut
both axes and it is the heatmap, cut neither and it is the tile plot, cut one
and it is this.

### Half a mesh: bounded by one slot

Two of the eight directions are the ones the wind comes from most often, and a
reader of the strip plot should see those two columns marked.
The rule was stated one axis at a time, so read it that way and a zone bounded
on *one* categorical axis follows. It takes that slot and spans the panel on the
other, which is what this mark has done from the beginning. One row is still one
rectangle, so the table you give it is the list of slots to shade:

```{r}
#| echo: false
peek(prevailing_winds)
```

```{r}
data(winds) + point * jitter + x(direction) + y(speed) +
  data(prevailing_winds) + zone + x(direction) + style(color = "goldenrod") +
  data(winds) + point * jitter + x(direction) + y(speed) +
  y_label("Speed") + title("The two directions the wind actually comes from")
```

*"Given the winds: points derived by jitter, x is direction, y is speed; then
given the prevailing table: zones, x is direction; then given the winds: points
derived by jitter, x is direction, y is speed."*

Nothing was built for that. It is the same sentence as the tile plot with one
axis left unsaid, which is the test that the fourth source is a rule rather than
a special case for one chart.

The two-row table is doing the work [`rule`](rule.qmd)'s table does, and for the
same reason: **a zone's position is a column**, so one table shades as many
slots as it has rows. Hand it the whole `winds` table instead and you get one
rectangle per observation, hundreds of translucent golds stacked into eight
solid columns. That is the rule behaving exactly as stated, and not the plot you
wanted.

One difference to see rather than be told: the column highlight is translucent
and the calendar heatmap is not, for the reason the settings section states in
full. `style(opacity = )` overrides either.

## Sides the data drew: the choropleth

Every section above finds a zone's sides in the same place: **the row being
drawn**. `bounds` names them, `bin` and `density` cut them, and a category owns
a slot. There is a fifth source, and it is the only one that spans many rows.

A map with every country shaded is a plot most readers have seen, and a
country's shape is not a rectangle.
A boundary is a list of points around the edge of a region: a coastline, a
border, the outline of a sales territory. It arrives as one row per point, so a
single shape needs hundreds of rows, and the sentence needs a word for which
rows belong together. That word is `group`:

```{r}
#| echo: false
peek(world_borders)
```

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + map()
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country, color
by continent, in the map space."*

That is the **choropleth**, and it lives in the [map](../map.qmd) space, where
the two positions are longitude and latitude. The mark did not change and no
atom was added: this mark was never about rectangles. It has always been
about naming an extent.

## What you can set

A highlight has to let the data show through, and a heatmap cell has to be
solid. gog picks one of the two by default, and a setting overrides it without
reading a column. Each mark takes its own settings. These are a zone's,
generated from the engine's own legality table, so this page cannot differ from
what `style()` accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("zone")
```

A zone is a filled region, so those five are exactly what a fill takes, and no
more. One of them has a default worth knowing: `opacity` is 0.2 for a zone you
placed and 1 for one whose cells a mesh made. That covers cells `bin` cut, cells
`density` estimated, and the slots two categories own. A placed zone is drawn
*under* your data and has to let it show through; a cut zone *is* the data, with
nothing behind it to reveal.

All three at once, one per band:

```{r}
#| echo: false
peek(span_early)
peek(span_middle)
peek(span_late)
```

```{r}
data(quarterly) + x(year) + y(sales) +
  data(span_early) + zone * bounds(start = start, end = end) +
    style(color = "seagreen") +
  data(span_middle) + zone * bounds(start = start, end = end) +
    style(color = "indianred", opacity = 0.55) +
  data(span_late) + zone * bounds(start = start, end = end) +
    style(color = "steelblue", pattern = "hatch", opacity = 0.7) +
  data(quarterly) + line + point +
  y_label("Sales") + title("Color, then opacity, then a hatch")
```

*"Given quarterly: x is year, y is sales, an early zone from start to end
colored seagreen, and also a middle zone from start to end colored indianred,
with opacity 0.55, and also a late zone from start to end colored steelblue,
with pattern hatch and opacity 0.7, and also a line and points."*

The default opacity is what makes the first band readable without your saying
anything: a highlight that hides the data is not a highlight. Raise it when the
zone is the subject of the plot rather than its background, as the red band does
here.

The hatch is worth using when a plot will be printed in gray, or when two
overlapping zones need more than color to separate them. Give it more opacity
than a flat fill wants, as the blue band does: a hatch is line-work with gaps
between the lines, so fading it to 0.2 leaves very little to see. It takes the
*fill* textures, not a stroke's dashes; `style(pattern = "dashed")` on a zone is
refused, pointing at the six fill textures instead.

One setting you might expect is refused, and the refusal says something about
the mark: `style(size = )`, because a zone's extent *is* its bounds and a size
would be a second, contradictory answer to how big it is.

A **border** has two good answers, and both are available. A zone highlighting a
region behind a line reads better with no border around it. Two `rule`s draw
better edges: dashed, colored per row, and real positions rather than
decoration:

```{r}
#| echo: false
peek(target_edges)
```

```{r}
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) + style(color = "seagreen") +
  data(target_edges) + rule + style(color = "seagreen", pattern = "dashed") +
  data(quarterly) + line + point +
  y_label("Sales") + title("A zone for the region, rules for its edges")
```

*"Given quarterly: x is year, y is sales, a target zone from lower to upper, and
also rules from the edges, and also a line and points."*

The two marks read their positions differently, which is why the edge table is
derived rather than reused. A zone is told which columns hold its sides
(`bounds(lower, upper)`), while a rule reads whichever of the plot's own
position columns its table carries, so the edges have to arrive as a `sales`
column. Building that column *from* `target_band` is what keeps the two from
drifting apart when the band moves.

But the mark is not only a highlight. It draws the waterfall, the heatmap's
cells, [the icicle](../transforms.qmd) and the
[mosaic](../transforms.qmd#the-mosaic), and in every one of those the zone *is*
the data rather than a background behind it. There the composition above stops
being an answer: a region has four sides, a mosaic has as many regions as the
table has rows, and two rules per cell hand-placed is not a sentence anyone
writes. So a zone takes `style(border_color = , border_size = )` like every
other closed-glyph fill:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) +
    style(border_color = "black") +
  data(quarterly) + line +
  y_label("Sales") + title("A zone with a border of its own")
```

*"Given quarterly: x is year, y is sales, recession zones from start to end,
with border color black, and also a line."*

Both readings stay available, and which one is right depends on whether the zone
is the background or the subject. A zone you do not ask for a border draws no
stroke: with no `border_color` and no `border_size` there is none at all.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A `point` or a `line` draws from the plot's own `x` and `y`, so a bare `zone`
over the same table is the natural first attempt.
A zone with nothing to bound it has no sides:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) + zone + line
```

And half a pair is not a side. The refusal names the half you gave, because that
is nearly always a typo rather than a misunderstanding:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower = lower) +
  line
```

The `start`/`end` pair belongs to a rectangle and to nothing else. A band spans
the measure axis *at each position*, so it has no extent along the domain to
bound:

```{r}
#| echo: false
peek(slump)
```

```{r}
#| error: true
data(slump) + x(year) + y(sales) +
  ribbon * bounds(start = start, end = end)
```

A zone says where its cells are and what is in them, and a transform can supply
either or both. `bounds` names the sides and measures nothing. `count` tallies
into the cells your categories already are, and says nothing about where those
are. `density` estimates a field and samples it. `bin` cuts cells out of a
continuous axis and supplies both, which is why it is also the one that can
supply just one of them. `proportion` answers neither question: it divides
whatever measurement is there by its total, so it stands after any of them. Two
transforms answering the same half is a contradiction, so a rectangle whose
sides you named, one per row, has no mesh for a summary to group into:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) * mean +
  line
```

To shade a band the data *computed* rather than one you chose, the mark that
spans a statistic is [`ribbon`](ribbon.qmd): `ribbon * range` fills between the
minimum and maximum at each x. The division is the same one `rule` draws: a zone
shades where you say, a ribbon shades what the data says.

Two measurements is the same contradiction on the other half. `count` invents
one and `mean` reduces the column you named, and a cell holds a single number.
`count` cuts nothing, so with `mean` measuring the cells it has no work left:

```{r}
#| error: true
data(winds) + zone * count * mean + x(direction) + y(season) + color(speed)
```

`bin * mean` [above](#cells-cut-then-summarized-the-summary-heatmap) has the
same shape and is not a contradiction: `bin` supplies the cells, and `mean`
supplies the measurement inside them.

Say `mean` with no column to reduce and gog asks for one rather than guessing.
A zone measures by `color`, so `color` is the binding the refusal names:

```{r}
#| error: true
data(winds) + zone * mean + x(direction) + y(season)
```

Asking for both extents at once is refused rather than resolved by picking one,
since the two would disagree the moment the mesh moved:

```{r}
#| echo: false
peek(banded)
```

```{r}
#| error: true
data(banded) + x(gdp) + y(life) +
  zone * bounds(lower, upper) * bin
```

A `bin` needs *something to cut*, and two categorical axes leave it nothing: a
category is one slot, with no width. The refusal does not stop there, though: a
cell per pair of *categories* is the tile plot, which does no binning at all, so
it names the transform that does tally into cells you already have:

```{r}
#| error: true
data(six_weeks) + zone * bin + x(weekday) + y(week)
```

One categorical axis is allowed, and it draws. That is the mixed mesh above.
`density` is where the two differ, and it is not an oversight. A field is
estimated *between* the data points, so it needs somewhere to spread on both
axes. A density per category would be normalized inside each slot on its own,
which makes its cells incomparable across slots while the color ramp claims
otherwise. It is the *which margin normalizes* question again, and the answer
here is the one `proportion` gives everywhere: a share is of the whole table,
and a per-slot denominator is a different plot that has to be asked for. So a
field asks for two continuous axes and names what to use instead:

```{r}
#| error: true
data(gm_all) + zone * density + x(life) + y(continent)
```

A tiling is refused for a reason of its own, and that reason is distance. `hex`
interleaves two lattices, weighing a step up against a step across, so it needs
a *distance* on both axes; a category's slots are an order and no more. A mixed
mesh has two axes and still no plane, and its cells are rectangles by
construction:

```{r}
#| error: true
data(gm_all) + zone * bin(tiling = "hex") + x(life) + y(continent)
```

A mark that cuts one axis gives a tiling even less to be regular against. `bar
* bin` cuts intervals, and an interval has no shape, so the refusal sends the
tiling back to `zone`:

```{r}
#| error: true
data(gm_all) + bar * bin(tiling = "hex") + x(life)
```

And the mirror of the first refusal, which is the same rule read the other way.
A tally needs cells that exist; a number is a point and owns none, so `count`
sends you back to `bin`:

```{r}
#| error: true
data(gm_all) + zone * count + x(gdp) + y(life)
```

The six summaries follow the same rule. `mean` reduces a column inside a cell,
and a continuous position gives it no cell, so the refusal offers both ways
out, a category on each position or `bin` on the continuous axis first:

```{r}
#| error: true
data(winds) + zone * mean + x(speed) + y(season) + color(bearing)
```

And the cells' measure is already the count, so coloring by anything else asks
for a column that binning has replaced:

```{r}
#| error: true
data(gm_all) + zone * bin + x(gdp) + y(life) + color(continent)
```

A mesh gog does not cut is refused with the list of the ones it does, rather
than quietly drawing rectangles instead:

```{r}
#| error: true
data(gm_all) + zone * bin(tiling = "octagon") + x(gdp) + y(life)
```

`density`'s `bandwidth` is refused here for a reason of its own. It is a width
in one column's own units, and a field spreads over two columns measuring
different quantities, so one number cannot be a width in both. That is why
`adjust` is the parameter the section above uses: a multiplier means the
same thing on either axis.

```{r}
#| error: true
data(iris_flowers) + zone * density(bandwidth = 0.5) +
  x(sepal_length) + y(petal_length)
```

## In short

A zone is bounded by the columns you give it, and reaches the panel edge where a
`ribbon` stops at the numbers you hand it. The pair you name decides which axis
it bounds, and the column types decide which one measures. Compose `bin` instead
of naming a pair, and a zone cuts both axes at once.

## What comes next

Color is how a zone carries a measure, read back from a legend.
[Surface](surface.qmd) gives that measure a height instead, and the same field
becomes a sheet standing in the cube.
