# Zone

```{r}
#| include: false
source("../R/setup.R")
```

Which region of the plot should the values stay inside? `zone` shades a
rectangle. It is [`rule`](rule.qmd)'s sibling: a rule is a line and a zone is an
area, and both work the same way. A rule takes one *position* and spans the axis
it was not given. A zone takes a *pair* and spans the axis it was given no pair
for. Give it pairs on both axes and it is a rectangle you placed.

Spanning the axis you gave no pair for is what first sets this mark apart. No
other mark draws a rectangle that reaches the edge of the **panel**.

## A band across the panel

Were sales inside the target each year? You could print the two target numbers
in the caption. A reader would then have to check one year at a time. A band
shaded behind the line shows every year together.
The `quarterly` table holds one sales figure for each year, and `target_band`
holds the two edges of a target band:

```{r}
#| echo: false
peek(quarterly)
peek(target_band)
```

Here are the sales with the target band behind them, 150 to 175, across every
year the table holds:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) + style(color = "seagreen") +
  data(quarterly) + line + point +
  y_label("Sales") + title("A target band, 150 to 175")
```

*"Given quarterly: x is year, y is sales, a target zone from lower to upper, and
also a line and points."*

`bounds(lower, upper)` names the two columns holding the bottom and top. Nothing
names a left or a right, so the zone takes those from the panel and runs edge to
edge. The zone is written *before* the line so the data draws on top of it; a
highlight belongs behind what it highlights.

## A band up the panel

Some highlights are periods of time rather than ranges of value. The years of a
recession are one such period, when the question is how sales changed during
them. Name the other pair, `start` and `end`, and the rectangle runs up the
panel instead of across it.
The `recessions` table holds the two slumps, each with a start year and an end
year:

```{r}
#| echo: false
peek(recessions)
```

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) +
    style(color = "indianred") +
  data(quarterly) + line + point +
  y_label("Sales") + title("Two slumps, from one table")
```

*"Given quarterly: x is year, y is sales, recession zones from start to end, and
also a line and points."*

Two shaded periods, and only one `zone` in the sentence. One row is one
rectangle, so a two-row table draws two bands, the same property that lets one
`rule` table draw three thresholds. Add a row and a third band appears with no
change to the sentence.

Look at where the red reaches: the top and bottom of the panel exactly. That is
the thing a `ribbon` cannot do. A ribbon is bounded by its data, so it stops at
whatever numbers you give it. Any number big enough to reach the top would
*widen the axis to include it*, so the highlight would change the plot it was
meant to mark.

The pair is called `start`/`end` rather than `left`/`right` on purpose. Left and
right are facts about a screen. This grammar takes orientation from the
bindings, so the same words still apply when the domain is drawn as an angle, in
[polar](../polar.qmd) coordinates.

## Both pairs: a rectangle

Sometimes the region to highlight is bounded on both axes: these five years,
this range of sales, and nothing outside either. That is a box rather than a
band.
Name all four and neither axis comes from the panel:

```{r}
#| echo: false
peek(sales_box)
```

```{r}
data(quarterly) + x(year) + y(sales) +
  data(sales_box) + zone * bounds(lower, upper, start = start, end = end) +
    style(color = "goldenrod", opacity = 0.3) +
  data(quarterly) + line + point +
  y_label("Sales") + title("A box over five years of sales")
```

*"Given quarterly: x is year, y is sales, a sales box zone from lower to upper
and from start to end, and also a line and points."*

So one atom covers the three shapes a highlight needs: a band across the panel,
a band up it, and a box. The pairs you name decide which one you get, so there
is no separate atom for each.

## Which axis the pair bounds: the funnel

A funnel counts how many people are left at each stage, from website visits
down to invoices sent. Its usual picture is a stack of centered blocks, one per
stage, measured across the page rather than up it. Drawing it means moving the
measurement to `x`, and a category on `y` is what moves it.

`lower` and `upper` bound the *measure* axis, the axis that carries the measured
numbers. Every plot to this point has measured along `y`, so every
`lower`/`upper` band has run across the panel. Now put a category on `y`
instead. The measure moves to `x`, and the same two
columns bound it there:

```{r}
stage <- c("Website visit", "Downloads", "Potential customers",
           "Requested price", "Invoice sent")
n     <- c(39, 27.4, 20.6, 11, 2)
funnel <- data.frame(stage = factor(stage, levels = stage),
                     n = n, lo = -n / 2, hi = n / 2, mid = 0)
```

```{r}
data(funnel) + zone * bounds(lo, hi) + y(stage) + style(opacity = 1) +
  text + label(n) + x(mid) +
  x_label("") + y_label("Stage") + title("A funnel, and no funnel atom")
```

*"Given the funnel: zones from lo to hi, y is stage, and also text, x is mid,
label by n."*

That is a funnel chart. No atom in the sentence is a funnel.

Each block is a `zone`. The table gives one row per block. The columns `lo` and
`hi` hold `-n/2` and `n/2`, computed in R before the plot. The numbers inside
the blocks are a second layer, drawn by `text`. The waterfall later in this
chapter computes its running total the same way.

A funnel is a bar chart whose blocks are centered on zero. A `bar` grows from a
baseline and refuses `bounds`, so the centered blocks are zones with both ends
named.

The plot works because `lower` and `upper` name the measure axis and not the
bottom and the top of a screen. gog never asks which way the screen runs; it
asks what each axis carries. Here `y` carries a category, so `y` holds slots,
and the measurement can only go on `x`.
This is the rule [`bar`](bar.qmd) already follows: one categorical position, one
measured position, and their types decide the orientation. It is why there is
nothing like ggplot2's `coord_flip()` to write.

## Coloring the zones

The two recession bands in the sales plot are both the same red, which leaves
them unnamed. Color each by its name and the legend says which is which,
with no note on the plot.
Each row is its own rectangle, so `color` maps per row. `recessions` carries a
name for each slump:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) + color(slump) +
  data(quarterly) + line + point +
  y_label("Sales") + title("Named, and keyed by the legend")
```

*"Given quarterly: x is year, y is sales, recession zones from start to end
colored by slump, and also a line and points."*

A zone's `color` also takes a *number*, drawn from a continuous ramp of colors,
while `bar`, `area` and `ribbon` accept only categories. A rectangle is large
enough for the eye to compare its shade with the legend. That is what a heatmap
needs, so a heatmap cell is this mark rather than a new one. [Cells cut from the
data](#cells-cut-from-the-data-the-heatmap) draws that plot, after the sections
on rectangles you place yourself.

## Rectangles that *are* the data

The funnel's blocks were the data, but every other zone so far has been a
highlight: a rectangle behind a line, saying where to look. Nothing requires it
to sit behind anything. Give a zone a pair on the measure axis and a
categorical position. Each category's rectangle then runs from one number to
another, with no plot underneath. Two common finance charts are made of exactly
that.

### The waterfall

A **waterfall** is a running total, broken into the steps that made it. gog has
no transform for that and the omission is deliberate: `stack` accumulates
*within* a position, one pile per slot, and a waterfall accumulates *across*
positions. So the running total is arithmetic, and arithmetic belongs with the
data, in the host language. The `cashflow` table holds one row per step of a
cash statement: a `delta` for the step, and a `total` flag marking the
subtotals:

```{r}
#| echo: false
peek(cashflow)
```

```{r}
w <- cashflow
w$top  <- cumsum(w$delta)          # where each step ends
w$base <- c(0, head(w$top, -1))    # and where it started: the step before
w$base[w$total] <- 0               # except the subtotals, which start at zero
w$dir  <- ifelse(w$total, "total", ifelse(w$delta >= 0, "gain", "loss"))

slot <- seq_len(nrow(w)) - 1       # category k sits at k, counting from 0
w$left  <- slot - 0.35             # so a side named at k ± 0.35 is inside its slot
w$right <- slot + 0.35
```

That chunk is R; the other three bindings compute the same columns in their own
language. The sentence has four spellings like every other:

```{r}
data(w) + zone * bounds(base, top, start = left, end = right) +
  x(step) + color(dir) + style(opacity = 1) +
  palette(c("steelblue", "seagreen", "indianred")) +
  y_label("Cash (thousands)") + title("Where the money went")
```

*"Given the w table: zones from base to top and from left to right, x is step,
color by dir, with opacity 1."*

A few lines of host arithmetic and one sentence. The blocks that start at zero
are not a special case. They are rows whose `base` is 0, a number the table
supplies like every other.

Two parts of that sentence need explaining, because both override one of this
mark's defaults on purpose.

`style(opacity = 1)` is there because a zone whose sides you named is drawn
translucent, at 20%. A rectangle someone chose is nearly always a highlight,
with something drawn over it. Here nothing is drawn over it: the rectangle *is*
the measurement. A cut zone (a [heatmap](#cells-cut-from-the-data-the-heatmap)
cell) is already opaque for the same reason, and this is how you say the named
one is data too.

The `left`/`right` pair makes the gap between the blocks. Without `start` and
`end`, a zone fills its slot whole, while a [bar](bar.qmd) takes four fifths of
it and leaves one fifth empty, which is what separates the categories. A zone's
width normally means something: it is the region the category owns. A waterfall
block's width means nothing, so it should leave a gap the way `bar` does, and
naming the sides is how to ask for one. The arithmetic works because a
categorical axis puts its categories at 0, 1, 2 and so on. Category *k* owns
the slot from `k−½` to `k+½`, and `k ± 0.35` sits inside it. Drop those two
columns and the rectangles touch, which reads as one connected shape rather than
six steps.

### The candlestick

A **candlestick** shows how a price moved in each trading session. Each session
has four numbers: the price at the start (`open`) and the end (`close`), and the
highest (`high`) and lowest (`low`) prices in between. gog needs no new channels
for them, because a rectangle's sides read the pair a [ribbon](ribbon.qmd) and
an [interval](interval.qmd) already read. Low to high is a whisker; open to
close is a rectangle; that is two layers over one table:

```{r}
#| echo: false
peek(sessions)
```

```{r}
s <- sessions
s$body_lo <- pmin(s$open, s$close)
s$body_hi <- pmax(s$open, s$close)
s$dir     <- ifelse(s$close >= s$open, "up", "down")

slot <- seq_len(nrow(s)) - 1       # the same slot arithmetic the waterfall used
s$left  <- slot - 0.3              # a candle body is narrower than a waterfall block
s$right <- slot + 0.3
```

```{r}
data(s) + x(session) +
  interval * bounds(low, high) + color(dir) +
  zone * bounds(body_lo, body_hi, start = left, end = right) +
    color(dir) + style(opacity = 1) +
  palette(c("seagreen", "indianred")) +
  y_label("Price") + title("Fourteen sessions")
```

*"Given the s table: x is session, intervals from low to high colored by dir,
and also zones from body lo to body hi and from left to right colored by dir."*

`color(dir)` is said twice, once per layer, and both are needed because of
[scope](../encoding-scope.qmd). A channel written after a mark belongs to that
mark, so the whisker takes its session's color along with the body. Both uses of
the column share *one* legend, because it is one column.

There is no `candlestick` atom, and the grammar does not need one. The chart is
made of things that already exist, which is the claim the whole grammar makes.
[Chart names](../chart-names.qmd) lists it beside the pie and the rose, which
are also sentences rather than words. Each candlestick row carries four
numbers: two for the whisker and two for the body. `ribbon` and `interval`
already read a low and a high, and a rectangle's bottom and top are the same
pair. That is why neither finance chart needs a new channel.

## Cells cut from the data: the heatmap

So far every zone has been a rectangle you *chose*, its sides named by columns
you already had. There is a second way to get sides, and it needs no new mark:
let `bin` cut them.

Gapminder has 1704 country-years. Drawn as points, the crowded middle becomes
one dark mass, because a thousand dots in one place look exactly like two
hundred. Cut the plane into cells and count the rows in each instead:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Where the country-years actually are")
```

*"Given all the gapminder years: zones derived by bin into 24, x is gdp on a log
scale, y is life."*

`bin` is not the zone's own transform. What the mark changes is how many axes
the cut runs along:

| Mark derived by `bin` | Shape |
|---|---|
| `bar * bin` | one axis cut, drawn as the bars of a histogram |
| `zone * bin` | two axes cut, drawn as the cells above |
| `surface * bin` | the same two axes, with the count raised into the cube |

Read the dark cells that rise from lower left to upper right: wealth and life
expectancy rise together. The darkest cell says how many country-years sit at
that combination, which overlapping dots cannot show.

Three things in that sentence need explaining.

`bin` cut both axes, and nothing asked it to. It is the same transform that
makes a histogram, and how many axes it cuts is taken from the mark rather than
requested. A `bar` leaves one axis free to measure a height along, so `bar *
bin` cuts the other one. A zone's cell needs an extent on *both* axes before it
can be a rectangle at all, so `zone * bin` cuts every axis that has a width to
cut, which here is both of them. One transform, one
word, readings decided by what the mark needs. There is no `bin2d`. Note the
words *has a width to cut*: [the mixed mesh](#one-of-each-the-mixed-mesh) is
where they matter, because a categorical axis arrives already cut, and `bin`
leaves it alone.

The count became the color, although no `color()` was written. Both positions
were taken, so the measurement went to the one channel left that can carry a
number, and the legend is titled Count. This is what `bar * bin` already does
for the y axis: a transform that invents a column also says where it goes.

The column it invents is called `count`, and you may name it out loud. That is
not a different plot. It is the same sentence with nothing left implied:

```{r}
short <- data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life)
long  <- data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
           color(count)

identical(render_svg(short), render_svg(long))
```

`render_svg()` returns the drawn picture as text, so comparing the two texts
shows that the pictures are the same.

Writing it has one use: *a binding is where a scale goes*, and the short form
has nowhere to put one. Counts are strongly skewed here: the fullest cell holds
forty country-years while half of them hold four or fewer. On a linear ramp
nearly every cell is the same pale blue. Name the channel and you can say `scale =
"log"` on it, the same way you already did on `x`:

```{r}
data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
  color(count, scale = "log") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same cells, counted on a log ramp")
```

*"Given all the gapminder years: zones derived by bin into 24, x is gdp on a log
scale, y is life, color by count on a log scale."*

Now the cells with few rows get distinct shades, and the dark diagonal shows a
gradient rather than one flat patch of color. The legend's middle label is
6.32, not 20.5: on a log ramp the middle color stands for the geometric mean of
the two ends.

The white gaps are cells where no country-year fell. They are left as panel,
not painted the pale end of the ramp, because that end stands for a count of
one. The uneven edge you get instead *is* the region the data actually covers,
which is information rather than a gap the renderer failed to fill.

`bin(24)` sets the **mesh**, the grid of cells the cut makes, exactly as it does
on a histogram, and one number cuts both axes into that many bins. A coarser
mesh shows less detail, and claims less precision:

```{r}
data(gm_all) + zone * bin(8) + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same data, eight bins a side")
```

Notice that the log scale still applies. The cells are cut in log space, so
they are even on the page rather than even in dollars. That is what you want
when the variable ranges from a few hundred dollars to a hundred thousand.

These cells are also drawn solid rather than translucent, for the reason the
waterfall gave: here the zone *is* the data, with nothing behind it to show
through.

## Hexagons: the other mesh

Does the grid pattern in a heatmap come from the data or from the mesh?
The cells above are rectangles, and that is a choice rather than a fact about
binning. Say `tiling = "hex"` and the plane is cut a different way:

```{r}
data(gm_all) + zone * bin(20, tiling = "hex") + x(gdp, scale = "log") + y(life) +
  palette("viridis") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("The same country-years, hexagonal cells")
```

*"Given all the gapminder years: zones derived by bin into 20 on hexagons, x is
gdp on a log scale, y is life, with the viridis palette."*

Two things changed there, and only one of them is the mesh. `palette("viridis")`
replaces the default sequential ramp, and it is the right choice for a count.
Viridis is built so equal steps in the number are equal steps in perceived
brightness. On a plot whose entire message is *how many*, that is the difference
between reading the diagonal and guessing. The palette is a free choice on
either mesh, though, so do not read it as something hexagons require.

The hexagonal mesh is chosen for a reason, not for its looks: the rectangular
mesh has a defect. A square grid lines up its cell centers in rows *and*
columns, and the eye is very good at seeing rows and columns: it reads that
alignment as if it were structure in the data. Wilkinson states it directly:
rectangular bins "lead the eye to align bin centers and to see regularity where
there is none" [@wilkinson2005]. That is why Carr and colleagues devised hexagon
binning in 1987 [@carr1987]. A hexagonal mesh staggers alternate rows, so there
is no aligned lattice left to mistake for a finding.

To see that on its own, keep the palette the same and change only the mesh:

```{r}
data(gm_all) + zone * bin(20) + x(gdp, scale = "log") + y(life) +
  palette("viridis") +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Rectangular cells, same palette")
```

Look along the sparse edges of the two plots. The rectangular cells line up in
straight rows and columns, so the edge of the data is made of horizontal and
vertical pieces. The hexagons do not line up, so the edge follows the data
rather than the mesh.

### Why the mesh belongs to `bin`

In the rest of this book, a transform computes and the *mark* decides how the
result is drawn. `bar * bin`, `line * bin` and `step * bin` are one histogram
drawn three ways. So it would be reasonable to expect the hexagon to be a mark,
or a setting on one. It is neither, and the reason is in the numbers
themselves.

Those three marks all receive *the same* numbers. A hexagonal mesh staggers
alternate rows of cells, which changes *which cell each observation falls in*,
so the counts themselves differ. You cannot draw a hexagonal plot
from data binned into rectangles, whatever the mark does with it. The mesh is
decided before anything is drawn, so it belongs to the transform that does the
cutting.

That is also how Wilkinson organizes it. His `bin` is not one method but a
family of ways to partition a plane, with `rect` and `hex` among the members.
The mesh both shapes the cells *and* tags each row with the cell it fell in. One
operation, both effects.

One consequence follows for the other marks: a tiling means nothing to a bin
that cuts one axis, because there the cells are intervals and an interval has no
shape. So `bar * bin(tiling = "hex")` is refused, as [What it
refuses](#what-it-refuses) shows.

### Hexagons and the shape of the panel

The hexagonal cells are not quite regular, meaning their six sides are not all
the same length. They are regular in the space the mesh is cut in, because that
cut gives both axes the same number of steps. The panel then stretches that
space by whatever its own proportions are, so on a wide panel the cells are
drawn slightly wide. Look closely at the hexagonal cells and you can see it.

This is what the original `hexbin` package exposes as its `shape` argument. A
panel kept at equal proportions removes it.
[`theme(ratio = 1)`](../style.qmd#theme-settings-that-belong-to-no-mark) asks
for that panel, and the hexagons are then regular.

## Cells the data estimated: `zone * density`

Both meshes so far have counted the rows that fell in each cell. A count is
exact, and when the table is small it is also ragged: the 150 flowers spread
over a mesh would leave most cells holding one or two, or none.

`bin` counts what landed in each cell. There is a second thing worth measuring
at a cell, and it needs no new mark either: how dense the cloud of points is
there. That is `density`, the transform that gives a [line](line.qmd) its smooth
curve. A zone reads it across both axes, for the reason it reads `bin` across
both, and so paints the field that [`path`](path.qmd) traces as contours.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + zone * density +
  x(sepal_length) + y(petal_length) +
  title("Where the flowers cluster")
```

*"Given the iris flowers: zones derived by density, x is sepal length, y is
petal length."*

Most of what the heatmap showed is true here too. Both axes were cut without
being asked, the measurement went to `color`, and `color(density)` names it out
loud and draws the identical plot. `zone * bin` and `zone * density` are one
plot measured two ways, counted and estimated.

The white gaps are the one difference between the two plots. This plot has none.
A count of zero says no row fell in that cell, so `bin` leaves it unpainted. An
estimate exists at every point of the plane, so the estimator returns a value
for every cell and the mesh is painted edge to edge. What it paints is a real
density rather than an arbitrary color value. The cell values times the cell
areas sum to one, the same property that makes the curve's `y` a density.

`palette()` chooses the ramp, exactly as it does for a binned heatmap, and the
argument for viridis on a count is the argument for it here:

```{r}
data(iris_flowers) + zone * density +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("Filled density, viridis")
```

Look at the dark area away from the two clusters: it is even, and the cells meet
with no gaps. gog draws each cell's edges on whole screen pixels. Smoothed edges
would leave thin lines of panel between the cells, and on a dark ramp they would
look like a grid laid over the data.

One parameter a painted field takes is `adjust`, the multiplier on the
bandwidth the estimator computed from the data. It is a multiplier rather than
a width in either column's units, which is why it is the one parameter that
means the same thing on both axes:

```{r}
data(iris_flowers) + zone * density(adjust = 1.6) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("The same cloud, smoothed harder")
```

*"Given the iris flowers: zones derived by density adjusted by 1.6, x is sepal
length, y is petal length."*

The same field *traced* rather than painted is [`path * density`](path.qmd), the
contour. The difference between them is the one between `bar * bin` and `line *
bin`: the mark chooses the shape, and the transform stays the same.

## Bands the data cut: `zone * density(levels = )`

A smooth ramp is hard to read a value from: two nearby cells differ by a shade
the eye cannot name. Cut the field into a few steps and each band has a level a
reader can point to.
Ask for levels and the field stops being continuous. It is cut into that many
steps, and a zone fills between them:

```{r}
data(iris_flowers) + zone * density(levels = 8) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("Filled density bands")
```

*"Given the iris flowers: zones derived by density at 8 levels, x is sepal
length, y is petal length."*

That is the **filled contour**, and `levels` is the same parameter `path` takes.
`levels` means one thing, cut the field into this many steps, and each mark then
draws that cut its own way: a path traces the boundaries, a zone fills between
them. So the two are not two features but one transform read by two marks, and
a band's edge falls exactly on the contour line:

```{r}
data(iris_flowers) + path * density(levels = 8) +
  x(sepal_length) + y(petal_length) + palette("viridis") +
  title("The same eight levels, traced")
```

*"Given the iris flowers: paths derived by density at 8 levels, x is sepal
length, y is petal length."*

That makes three readings of one transform: `line * density` is a curve, `zone *
density` a painted field, and `path * density` a contour.

The bands nest, and that is why plain filled shapes are enough to draw them. A
denser region usually lies *inside* a less dense one, so gog paints the
outermost band first and lets each inner one cover it; nothing has to be cut
out. The exception is a cluster shaped like a ring: points scattered around a
circle, with an empty middle. There the empty middle really is a hole, and
painting the band over it fills it in. gog has no mark for a shape with a hole,
so a ring-shaped cluster is the one field `zone * density` draws wrongly, and
gog prints no warning. `path * density` traces it correctly.

## Cells the axes already made: the tile plot

So far the sides have come from three places: `bounds` named them, `bin` cut
them, and `density` estimated them. The funnel already used a fourth without
naming it: a category's slot, which needs no column at all.

Six weeks of daily orders is already a grid: weeks across, weekdays down, one
count in each cell. The grid itself is the picture you want, and two
categorical axes already divide the panel into its cells.

A categorical axis is already divided. Six weeks on the x axis is six slots.
"Week 3" owns one of those slots, a place on the panel with a left edge and a
right edge, and the axis fixes both when it lays out the categories. So a
category bounds its own axis, and two categorical positions bound a zone
completely, with no transform in the sentence:

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
data(six_weeks) + zone + x(week) + y(weekday) + color(orders) +
  title("Six weeks of orders, a day at a time")
```

*"Given the six weeks table: zones, x is week, y is weekday, color by orders."*

That is a **tile plot**, here in the form called a calendar heatmap, and the
whole of it is `zone + x + y + color`. What makes it worth drawing is that 42
daily counts carry two patterns at once. A list of numbers shows neither, while
a grid separates them by direction. Read
*down* a column for the week (week four is the pale one, the lowest week) and
*across* a row for the weekday, where the Thursday row is the darkest overall
and the Friday row the palest. The extremes are the cells where the two patterns
agree: the darkest cell is Thursday of week six, and the palest are the Friday
and Saturday of week four.

The rule underneath is short: a category owns a slot, a number is a point. A
point has no width, so two continuous positions leave a zone with nothing to
bound it until `bin` is written into the sentence. Nothing else changed, and
nothing was added to the kernel: no mark, no transform, no channel.

### `bin` cuts, `count` tallies

The calendar table has one row per cell already. Often a table does not: you
hold the raw observations and want to know how many fell in each cell. The
transform that tallies them is `count`, exactly as it is for a bar chart:

```{r}
#| echo: false
peek(winds)
```

```{r}
data(winds) + zone * count + x(direction) + y(season) +
  title("Wind observations by direction and season")
```

*"Given the winds: zones derived by count, x is direction, y is season."*

That is the shape of a confusion matrix, the table that counts predicted classes
against actual ones. A confusion matrix is the same sentence with different
columns: `zone * count + x(actual) + y(predicted)`.

Notice what was *not* needed there. `bin` cuts a continuous axis into cells
before anything can be counted in them; a categorical axis arrives already cut,
so there is nothing left to do but tally. `bar * bin` and `bar * count` follow
the same rule on one axis; here both axes follow it. So each refusal names the
other transform: give `zone * count` a continuous axis and it names `bin`, and
give `zone * bin` two categorical axes and it names `count`.

A tile plot has two categorical positions where a bar chart has one. That
second position changes only where the tally is written. A bar has a length to
put it on, so `bar * count` writes its tally to the y axis. A tile plot's two
positions are both taken by categories, so `zone * count` writes its tally to
`color`, and the legend is titled Count.

`proportion` is the same tally as a share of the whole, which is the plot to
draw when the absolute numbers are not what you want to read:

```{r}
data(winds) + zone * proportion + x(direction) + y(season) +
  palette("viridis") + title("The same observations, as shares")
```

*"Given the winds: zones derived by proportion, x is direction, y is season."*

That share is of *every* row counted, not of a row or a column of the grid. A
confusion matrix divided by the total of each true class needs a different
denominator. `proportion` always divides by the whole table, so compute row
shares in the host language and draw them as a tile plot.

`proportion` reads a cut mesh the same way. Compose it with `bin` and the
heatmap's counts come back as fractions of the whole, the [relative-frequency
histogram](../transforms.qmd#normalizing-something-other-than-a-tally) with a
second axis:

```{r}
data(gm_all) + zone * bin(20) * proportion + x(gdp, scale = "log") + y(life) +
  palette("viridis") + title("The heatmap, as shares of all 1704 rows")
```

`proportion` leaves the cells exactly where `bin` cut them, and the legend now
reads in shares: the fullest one holds about three percent of the 1704 rows.

### `count` tallies rows, `mean` reduces a column

The wind table records more than which cell each observation fell in. Each row
also carries a `speed`, and the question "how fast, on average, in each cell?"
is not a tally at all; it is a summary of one column:

```{r}
data(winds) + zone * mean + x(direction) + y(season) + color(speed) +
  title("Mean wind speed, by direction and season")
```

*"Given the winds: zones derived by mean, x is direction, y is season, color by
speed."*

Compare that with `zone * count` in the section above. The sentence has the same
shape, with `mean` for `count` and one more binding. `count` was handed no
column, so it counts rows and labels the result `count`. `mean`
was handed one, and `color(speed)` says which column it reduced. All six
summaries read a mesh this way (`sum`, `mean`, `median`, `max`, `min` and
`quantile`), and each writes its answer back into the column it reduced.

The first heatmap said that the measurement goes to `color` when both positions
are taken. The rule has a second half: color is also the channel that names the
column, because a summary writes its result back into the column it read. On an
ordinary two-axis bar chart, `bar * mean + x(continent) + y(life)` names `life`
with `y` and puts the mean back on `y`;
the source and the destination are one binding. A zone measures by color, so
`color` is both the source and the destination.

The whole rule is short. A summary reduces the column on the channel that
carries the measurement: `y` on a bar, `color` on a zone. It groups by every
position that does not carry it. A bar measures with `y`, so it groups by `x`.
A zone measures with nothing positional, so it groups by both. In [the
cube](../space.qmd) a bar measures with `z`, so it groups by the pair as well.

Say `mean` with nothing for it to reduce and the engine asks for the column
rather than guessing, as [What it refuses](#what-it-refuses) shows.

A summary also needs cells to summarize into. The six summaries measure without
cutting, so each axis must arrive already divided. [What it
refuses](#what-it-refuses) shows the refusal, which names both fixes.

### Cells cut, then summarized: the summary heatmap

The last two sections let the axes supply the cells. This section and the next
put `bin` back in the sentence.
How does life expectancy depend on income and on time together? Both are
continuous, so neither axis arrives divided into cells. `bin` cuts them, and
`mean` measures inside each:

```{r}
data(gm_all) + zone * bin(15) * mean + x(year) + y(gdp, scale = "log") +
  color(life) + title("Mean life expectancy, by era and income")
```

`bin(15)` cuts the plane into cells, `mean` reduces `life` inside each, and
`color` both names that column and carries the answer, exactly as it does on the
tile plot above. Read the panel from bottom left to top right and you can see
both patterns. Life expectancy rises with income up the panel, and rises
again with time across it.

`bin(15)` cuts the same cells with or without `mean`. What changes is that it
stops reporting how many rows fell in each, because `mean` now supplies the
measurement. So the summary heatmap and `zone * bin(15) + x(year) + y(gdp,
scale = "log")` tile the panel identically. Read them against each other: the
count says how many observations fell in each cell, and the mean says what those
observations were like.

The same composition works when one axis is a category. Cut the income axis, and
leave the continents in their slots:

```{r}
data(gm_all) + zone * bin(12) * mean + x(gdp, scale = "log") + y(continent) +
  color(life) + title("Mean life expectancy, by income band and continent")
```

One axis was cut and the other arrived in slots already, which is the mixed mesh
the next section states as a rule.

### One of each: the mixed mesh

How does the spread of life expectancy differ by continent? Five histograms
answer it, in five panels the reader must compare one by one.
Read the rule one axis at a time and it already covers the case in the middle:
one axis continuous, one categorical. Nothing new has to be decided for it.
`bin` cuts the axis that has a width to cut and leaves alone the one that
arrived already cut, so the result is a row of cells per category:

```{r}
data(gm_all) + zone * bin(20) + x(life) + y(continent) +
  palette("viridis") + title("Life expectancy, a distribution per continent")
```

That is a distribution shown as shade rather than as height, which is what makes
five of them fit in the space one histogram would take. Read a row and it is one
continent's histogram. Read the plot and you see the shape of each continent's
spread. Africa's mass sits far left and Europe's far right. Asia spreads the
widest and Europe the narrowest of the four large rows, with Africa and the
Americas in between. Oceania has only two countries in this table, which is why
its row is the shortest.

Two things make it a *mesh* and not five plots side by side. Every row is cut on
*the same* edges: the layout is computed once from the whole `life` column,
never per category. So each column of cells covers the same range of life
expectancy for every continent, and the boundaries line up all the way down.
And an empty cell stays empty. Oceania's row starts where its data starts,
rather than being painted the palest end of the ramp. The uneven left edge shows
where each continent's values begin. Painting the empty cells would suggest life
expectancies as low as 25 in Oceania, where the lowest is 69.

Swap the axes and each distribution stands upright:

```{r}
data(gm_all) + zone * bin(20) + y(life) + x(continent) + palette("viridis")
```

Nothing was added for either. Three cases, one rule read one axis at a time:
`bin` cuts both axes in the heatmap and one in this plot, and the tile plot
needs no `bin` at all.

### One slot: a column highlight

The wind comes from two of the eight directions most often. A strip plot of
wind speeds should shade those two directions.
Reading the rule one axis at a time leaves one more case: a category on one
axis and nothing on the other. A zone bounded on *one* categorical axis
takes that slot and spans the panel on the other, which is what this mark does
whenever a pair is missing. One row is still one rectangle, so the table you
give it is the list of slots to shade:

```{r}
#| echo: false
peek(prevailing_winds)
```

```{r}
data(winds) + point * jitter + x(direction) + y(speed) +
  data(prevailing_winds) + zone + x(direction) + style(color = "goldenrod") +
  data(winds) + point * jitter + x(direction) + y(speed) +
  y_label("Speed") + title("The two directions the wind actually comes from")
```

*"Given the winds: points derived by jitter, x is direction, y is speed; then
given the prevailing winds: zones, x is direction; then given the winds: points
derived by jitter, x is direction, y is speed."*

Nothing was built for that. It is the same sentence as the tile plot with one
axis left unsaid, which shows that a slot is a general source of sides, not a
special case for one chart.

The two-row table does what [`rule`](rule.qmd)'s table does, and for the same
reason: a zone's position is a column, so one table shades as many slots as
it has rows. Hand it the whole `winds` table instead and you get one rectangle
per observation, hundreds of them in translucent gold, stacked into eight solid
columns. That is the rule behaving exactly as stated, and not the plot you
wanted.

One difference is easy to see. The column highlight is translucent and the
calendar heatmap is not. A zone that spans the panel on one axis is a highlight,
and gog draws it at 20%. A zone whose cells come from the data on both axes is
the data itself, and gog draws it opaque. `style(opacity = )` overrides either.

## Sides the data drew: the choropleth

In every section above, each drawn shape takes its sides from one row: a row of
your table, or a row a transform made. `bounds` names the sides, `bin` cuts
them, `density` estimates them, and a category owns a slot. There is a fifth
source, where one shape takes its outline from many rows.

A map with every country shaded is a plot most readers have seen, and a
country's shape is not a rectangle.
A boundary is a list of points around the edge of a region: a coastline, a
border, the outline of a sales territory. It arrives as one row per point, so a
single shape needs many rows, and the sentence needs a word for which
rows belong together. That word is `group`:

```{r}
#| echo: false
peek(world_borders)
```

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + map()
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country, color
by continent, in the map space."*

That is the **choropleth**, and it is drawn in the [map](../map.qmd) space,
where the two positions are longitude and latitude. The mark is the same `zone`,
and the kernel needs no new atom. A zone is usually a rectangle, but its shape
comes from whatever gives its sides, and here a boundary gives them.

## What you can set

A highlight has to stay in the background, and a heatmap cell has to be solid.
gog gives each kind its own default, and a value in `style()` overrides it for
the whole layer. Each mark takes its own settings. These are a zone's,
with the values each accepts:

```{r}
#| echo: false
#| results: asis
cat("| Setting | Value |\n|---|---|\n")
mark_options("zone")
```

A zone is a filled region, so those five are exactly what a fill takes, and no
more. One of them has a default that matters. `opacity` is 0.2 for a placed
zone: one whose sides you named with `bounds`, or one that spans the panel on an
axis. It is 1 when the data gives every side: cells `bin` cut, cells `density`
estimated, the slots two categories own, and the regions a boundary drew. A
placed zone is a highlight. It stays pale so the gridlines, and any layer
written before it, show through. A zone of the second kind *is* the data, with
nothing behind it to reveal.

Three bands, each adding one setting: a color, then an opacity, then a hatch:

```{r}
#| echo: false
peek(span_early)
peek(span_middle)
peek(span_late)
```

```{r}
data(quarterly) + x(year) + y(sales) +
  data(span_early) + zone * bounds(start = start, end = end) +
    style(color = "seagreen") +
  data(span_middle) + zone * bounds(start = start, end = end) +
    style(color = "indianred", opacity = 0.55) +
  data(span_late) + zone * bounds(start = start, end = end) +
    style(color = "steelblue", pattern = "hatch", opacity = 0.7) +
  data(quarterly) + line + point +
  y_label("Sales") + title("Color, then opacity, then a hatch")
```

*"Given quarterly: x is year, y is sales, an early zone from start to end
colored seagreen, and also a middle zone from start to end colored indianred,
with opacity 0.55, and also a late zone from start to end colored steelblue,
with pattern hatch and opacity 0.7, and also a line and points."*

The default opacity keeps the first band pale with nothing set, so the line and
points stay the strongest marks on the page. Raise it when the zone is the
subject of the plot rather than its background, as the red band does here.

Use the hatch when a plot will be printed in gray, or when two overlapping zones
need more than color to separate them. Give it more opacity
than a flat fill needs, as the blue band does: a hatch is a set of lines with
gaps between them, so fading it to 0.2 leaves very little to see. It takes the
*fill* textures, not a stroke's dashes; `style(pattern = "dashed")` on a zone is
refused, pointing at the six fill textures instead.

One setting you might expect is refused: `style(size = )`, because a zone's
extent *is* its bounds and a size would be a second, contradictory answer to how
big it is.

There are two good answers for a zone's **border**, and both are available. A
highlight behind a line reads better with no border. If you want its edges
drawn, two `rule`s do it better than a border: they can be dashed, and each sits
at a real value on the axis:

```{r}
#| echo: false
peek(target_edges)
```

```{r}
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) + style(color = "seagreen") +
  data(target_edges) + rule + style(color = "seagreen", pattern = "dashed") +
  data(quarterly) + line + point +
  y_label("Sales") + title("A zone for the region, rules for its edges")
```

*"Given quarterly: x is year, y is sales, a target zone from lower to upper, and
also rules from the edges, and also a line and points."*

The two marks read their positions differently, which is why the edge table is
derived rather than reused. A zone is told which columns hold its sides, with
`bounds(lower, upper)`. A rule reads whichever of the plot's own position
columns its table carries, so the edges must arrive as a `sales` column.
`target_edges` is built from `target_band`, so the edges move when the band
moves.

But the mark is not only a highlight. It draws the waterfall, the heatmap's
cells, [the icicle](../transforms.qmd#partition-a-whole-among-nested-parts) and
the [mosaic](../transforms.qmd#the-mosaic), and in every one of those the zone
*is* the data rather than a background behind it. There the composition above
cannot help. A rule spans the whole panel, or makes a short tick at its edge, so
it cannot outline one cell of a mosaic. So a zone takes `style(border_color = ,
border_size = )`, like `bar`, `box`, `point` and `surface`:

```{r}
data(quarterly) + x(year) + y(sales) +
  data(recessions) + zone * bounds(start = start, end = end) +
    style(border_color = "black") +
  data(quarterly) + line +
  y_label("Sales") + title("A zone with a border of its own")
```

*"Given quarterly: x is year, y is sales, recession zones from start to end,
with border color black, and also a line."*

Both choices stay available, and which one is right depends on whether the zone
is the background or the subject. By default a zone has no border: with no
`border_color` and no `border_size`, it draws no outline.

The [grid of every mark and every setting](../style.qmd#what-you-can-set) shows
which other marks share this list. What a mark *maps* rather than sets is its
row on the [companion grid](../combinations.qmd#marks-and-channels).

## What it refuses

A `point` or a `line` draws from the plot's own `x` and `y`, so a bare `zone`
over the same table is the natural first attempt. Two continuous positions give
a rectangle no width and no height, so it is refused:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) + zone + line
```

One column is not a pair. The refusal names the column you gave, because a
missing half is nearly always a typo:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower = lower) +
  line
```

The `start`/`end` pair belongs to a rectangle and to nothing else. A `ribbon`
spans the measure axis at each position along the domain, so it has no extent
along the domain to bound:

```{r}
#| echo: false
peek(slump)
```

```{r}
#| error: true
data(slump) + x(year) + y(sales) +
  ribbon * bounds(start = start, end = end)
```

A zone needs two answers: where its cells are, and what is in each. `bounds`
answers only where. `count` answers only what, inside cells your categories
already made. `density` answers both. `bin` answers both, and it can also answer
only where, which is how `bin * mean` works. `proportion` gives the tally as
shares of the whole. Two transforms that give the same answer contradict each
other. A summary such as `mean` answers what, inside the cells the positions
make, but a rectangle whose sides you named is one row on its own, so there are
no cells for it to group rows into:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) +
  data(target_band) + zone * bounds(lower, upper) * mean +
  line
```

To shade a region the data *computed* rather than one you chose, use
[`ribbon`](ribbon.qmd), the mark that fills between two computed values:
`ribbon * range` fills between the minimum and maximum at each x. A zone shades
where you say; a ribbon shades what the data says.

Asking for two measurements is also a contradiction. `count` makes one
measurement and `mean` makes another, but a cell holds a single number. `count`
cuts nothing, so with `mean` measuring the cells there is
nothing left for it to do:

```{r}
#| error: true
data(winds) + zone * count * mean + x(direction) + y(season) + color(speed)
```

`bin * mean` in [the summary
heatmap](#cells-cut-then-summarized-the-summary-heatmap) has the same shape and
is not a contradiction: `bin` supplies the cells, and `mean`
supplies the measurement inside them.

Say `mean` with no column to reduce and gog asks for one rather than guessing.
A zone measures by `color`, so `color` is the binding the refusal names:

```{r}
#| error: true
data(winds) + zone * mean + x(direction) + y(season)
```

Naming the sides and cutting them in one sentence is refused, not settled by
picking one, because the named sides and the cut cells would not agree:

```{r}
#| echo: false
peek(banded)
```

```{r}
#| error: true
data(banded) + x(gdp) + y(life) +
  zone * bounds(lower, upper) * bin
```

A `bin` needs *something to cut*, and two categorical axes leave it nothing:
each category is one slot, and a slot cannot be cut smaller. The refusal goes
further: a cell per pair
of *categories* is the tile plot, which does no binning at all, so it names the
transform that tallies into cells you already have:

```{r}
#| error: true
data(six_weeks) + zone * bin + x(weekday) + y(week)
```

One categorical axis is allowed, and it draws. That is [the mixed
mesh](#one-of-each-the-mixed-mesh). `density` differs from `bin` here, and the
reason is the estimate itself. A field is estimated *between* the data points,
so it needs somewhere to spread on both axes. A density per category would be
normalized inside each slot on its own, which makes its cells incomparable
across slots while the color ramp claims otherwise. A density per category is a
different plot, so a field asks for two continuous axes and names what to use
instead:

```{r}
#| error: true
data(gm_all) + zone * density + x(life) + y(continent)
```

A tiling is refused for a reason of its own, and that reason is distance. A
hexagon is regular only when a step up and a step across can be compared, so
`hex` needs a *distance* on both axes. A category's slots have an order but no
distance, so a mixed mesh always has rectangular cells:

```{r}
#| error: true
data(gm_all) + zone * bin(tiling = "hex") + x(life) + y(continent)
```

When `bin` cuts only one axis, a tiling has nothing to shape. `bar * bin`
cuts intervals, and an interval has no shape, so the refusal names `zone` as the
mark a tiling belongs to:

```{r}
#| error: true
data(gm_all) + bar * bin(tiling = "hex") + x(life)
```

The opposite case, `count` over continuous axes, is refused by the same rule. A
tally needs cells that exist; a number is a point and owns none, so the refusal
names `bin`:

```{r}
#| error: true
data(gm_all) + zone * count + x(gdp) + y(life)
```

The six summaries follow the same rule. `mean` reduces a column inside a cell,
and a continuous position gives it no cell. The refusal names both fixes: a
category on each position, or `bin` on the continuous axis first:

```{r}
#| error: true
data(winds) + zone * mean + x(speed) + y(season) + color(bearing)
```

And the cells' measure is already the count, so coloring by anything else asks
for a column that binning has replaced:

```{r}
#| error: true
data(gm_all) + zone * bin + x(gdp) + y(life) + color(continent)
```

A tiling gog does not offer is refused with the list of the ones it does,
rather than quietly drawing rectangles instead:

```{r}
#| error: true
data(gm_all) + zone * bin(tiling = "octagon") + x(gdp) + y(life)
```

`density`'s `bandwidth` is refused here for a reason of its own. It is a width
in one column's own units, and a field spreads over two columns measuring
different quantities, so one number cannot be a width in both. That is why
`adjust` is the parameter [the density
section](#cells-the-data-estimated-zone-density) uses: a multiplier means the
same thing on either axis.

```{r}
#| error: true
data(iris_flowers) + zone * density(bandwidth = 0.5) +
  x(sepal_length) + y(petal_length)
```

## In short

A zone's sides come from five places: `bounds`, `bin`, `density`, a category's
slot, or a boundary named by `group`. Where nothing gives sides, the zone
reaches the panel edge, which a `ribbon` cannot do. The pair you name decides
which axis it bounds; the column types decide which one measures.

## What comes next

Color is how a zone carries a measure, read back from a legend.
[Surface](surface.qmd) gives that measure a height instead, and the same field
is raised into the cube as a sheet.
