# The book's data

```{r}
#| include: false
source("R/setup.R")
```

Where do `gapminder_2007`, `medals` and `winds` come from, and how do you get
them?

Every chapter of this book uses the same shared tables, listed in full at the
end of this page. They are published beside the book as CSV files, one per
table, so you can run any example in this book yourself. Each file has a fixed
address:

```
https://psychometrician.github.io/gog-book/data/<name>.csv
```

`gapminder_2007.csv` holds the 142 countries the first chapter plots.
`medals.csv` holds the medal counts the reading chapter draws, and `winds.csv`
holds the wind observations the polar chapter draws.

Reading one takes a single call, `gog_table()`, which comes with the package
in all four languages. Nothing else needs installing.

## Reading a table

`gog_table()` reads any of these tables by name:

```{r}
#| echo: false
spellings(
  r = r"(gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life))",
  python = r"(gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(col.gdp) + y(col.life))",
  julia = r"(gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(:gdp) + y(:life))",
  js = r"(const gapminder_2007 = await gog_table("gapminder_2007");
plot(data(gapminder_2007), point, x(col.gdp), y(col.life));)"
)
```

Those four calls draw the same picture, because the table is the same file and
the engine is the same program.

`gog_table()` also gives each column the right type. A CSV is text, so numbers arrive as
text and have to be converted. A column becomes numbers when every value in it
is a number, and stays text otherwise. The `text` argument forces a column to
stay text. Three columns in these tables need that, and
[Labels that look like numbers](#labels-that-look-like-numbers), below, explains
which and why.

This one function is the largest thing the four packages carry that is not
grammar. R reads a CSV in one call, and Python and Julia in a few. JavaScript
has no CSV reader in its standard library, so the same work there takes dozens
of lines. One country in `gapminder_2007.csv` is `"Congo, Dem. Rep."`, and its
name holds a comma, so splitting each line on commas gives that row seven
fields where the header has six. Nobody should have to paste dozens of lines
of parsing before drawing, so each package carries `gog_table()` instead.

R's `read.csv()`, `urllib` in Python, `Downloads` in Julia and `fetch` in
JavaScript are all standard, so `gog_table()` needs nothing beyond the package
you already installed.

Reading a different table is the same call with a different name:
`gog_table("winds")`, `gog_table("medals")`.

## When the name is wrong

A table name is easy to mistype, and `gapminder2007` is not `gapminder_2007`.
gog stops instead of returning a different table. It says which name it could
not find. When one published name is close to what you typed, gog names it:

```
gog: there is no table called "gapminder2007". Did you mean "gapminder_2007"?
```

The suggestion is offered only when the two names differ by at most two letters.
When nothing is that close, the refusal points at this page instead. A wrong
suggestion makes you try a second wrong name, which is worse than no suggestion.

A site that cannot be reached is a different problem, and its refusal names the
connection rather than the table. The two need different words. Being told a
table does not exist is misleading when the connection failed, because it sends
you to correct a name that was already right.

All four languages print the same two refusals, word for word. Each is refused
the way a plot is refused, so the code that handles one refused plot handles
these too.

## Two things a CSV cannot record

A CSV is text. It records what a value is, never what kind of thing it is. Two
kinds of information are therefore dropped when the file is written, and you
have to restore them when you read it. Both change the plot, so this page
names both.

### The order of a category

The `direction` column of the winds table holds the eight points of the
compass. Their order is N, NE, E, SE, S, SW, W, NW, the order a compass runs
in. A CSV cannot record that order. A text column read from a file has no
order of its own, so gog draws its categories in the order the rows arrive.
The winds file happens to list its rows in compass order, so a plot straight
from the file reads correctly, by luck. Sort the rows by another column, wind
speed for example, and the same plot runs S, N, NE, NW, SE, E, SW, W. It still
draws, and it is wrong in a way that is easy to miss.

Declaring the order removes the luck. `gog_table("winds")` returns `direction`
as plain text, so the declaration is yours to make after reading. The example
below reads the file directly, sorts the rows by speed so that the row order
would mislead, and then declares the order:

```{r}
winds_local <- read.csv("data/winds.csv")
winds_local <- winds_local[base::order(winds_local$speed), ]
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds_local$direction <- factor(winds_local$direction, levels = compass)

data(winds_local) + bar * count + x(direction) + polar()
```

*"Given the winds read from the file: bars derived by count, x is direction, in
polar."*

The directions follow the compass, from N clockwise to NW, whatever order the
rows came in. The wind rose, the circular count of winds the polar chapter
draws, depends on that.

Each language spells the declaration differently: R uses `factor(levels = )`,
and Python, Julia and JavaScript use `ordered()`. The
[Data](data.qmd#declaring-the-order-of-categories) chapter covers this in full,
because declared order is part of the grammar and not a detail of file formats.

### Labels that look like numbers

Three columns in these tables hold text that `gog_table()` would read as
numbers unless you name them. `census.csv` has an `age`
column of `0`, `5`, `10`, which are the names of age bands rather than
quantities. `sessions.csv` has a `session` column of `01`, `02`, `03`, and a
number cannot keep a leading zero, so `01` comes back as `1`. `gm_eras.csv`
has an `era` column of `1957` and `2007`. Those two values name the eras the
rows compare; they do not measure anything. That is the whole reason
`gog_table()` takes a `text` argument:

```{r}
#| echo: false
spellings(
  r = r"(census <- gog_table("census", text = "age")
sessions <- gog_table("sessions", text = "session")
gm_eras <- gog_table("gm_eras", text = "era"))",
  python = r"(census = gog_table("census", text=("age",))
sessions = gog_table("sessions", text=("session",))
gm_eras = gog_table("gm_eras", text=("era",)))",
  julia = r"(census = gog_table("census"; text = ["age"])
sessions = gog_table("sessions"; text = ["session"])
gm_eras = gog_table("gm_eras"; text = ["era"]))",
  js = r"(const census = await gog_table("census", ["age"]);
const sessions = await gog_table("sessions", ["session"]);
const gm_eras = await gog_table("gm_eras", ["era"]);)"
)
```

Without it, `age` becomes numbers, and an axis that should show one slot per
age band is drawn as a numeric scale from 0 to 85. The plot looks reasonable,
so nothing tells you the axis is wrong.

## The tables

```{r}
#| echo: false
files <- sort(list.files("data", pattern = "[.]csv$"))
info <- data.frame(
  Table = sub("[.]csv$", "", files),
  Rows  = vapply(files, function(f) nrow(read.csv(file.path("data", f))), integer(1)),
  Columns = vapply(files, function(f) ncol(read.csv(file.path("data", f))), integer(1))
)
knitr::kable(info, row.names = FALSE)
```

Not every table here holds real data. The tables whose names begin with
`gapminder` or `gm` hold real figures from the Gapminder Foundation, released
into the public domain. `iris_flowers`, `maunga_whau`, `quakes_fiji` and
`titanic` are reshaped from tables included with R, and carry R's own
license. `world_borders` is Natural Earth's public-domain outline of the
countries. Everything else was written for this book, each table to show one
thing; `census` is two plausible city profiles, not a real census. The terms
for each are in [`data/LICENSE.md`](data/LICENSE.md).
