# The book's data

```{r}
#| include: false
source("R/setup.R")
```

Where do `gapminder_2007`, `medals` and `winds` come from, and how do you get
them?

Every chapter of this book draws from a small set of shared tables. They are
published beside the book as CSV files, one per table, so you can run any
example in this book yourself. Each file sits at a fixed address:

```
https://psychometrician.github.io/gog-book/data/<name>.csv
```

`gapminder_2007.csv` holds the 142 countries the first chapter plots.
`winds.csv` holds the wind observations behind the polar chapter. The full list
is at the end of this page.

Nothing needs installing to read them. R, Python, Julia and JavaScript each
read a CSV with their own standard library, so fetching a table adds no
dependency to your project.

## Reading a table

Here is the first sentence of the book, run from the published file, in each of
the four languages.

::: {.panel-tabset}

## R

```r
gapminder_2007 <- read.csv(
  "https://psychometrician.github.io/gog-book/data/gapminder_2007.csv"
)

data(gapminder_2007) + point + x(gdp) + y(life)
```

## Python

```python
import csv, urllib.request

url = "https://psychometrician.github.io/gog-book/data/gapminder_2007.csv"
with urllib.request.urlopen(url) as response:
    rows = list(csv.DictReader(response.read().decode("utf-8").splitlines()))

gapminder_2007 = {name: [row[name] for row in rows] for name in rows[0]}
for name in ("gdp", "life", "population"):
    gapminder_2007[name] = [float(value) for value in gapminder_2007[name]]

data(gapminder_2007) + point + x(col.gdp) + y(col.life)
```

With pandas, the whole read is one line:
`gapminder_2007 = pd.read_csv(url)`.

## Julia

```julia
using Downloads, DelimitedFiles

url = "https://psychometrician.github.io/gog-book/data/gapminder_2007.csv"
raw, header = readdlm(Downloads.download(url), ',', String; header = true)
gapminder_2007 = Dict(strip(name, '"') => raw[:, i] for (i, name) in enumerate(header))

for name in ("gdp", "life", "population")
    gapminder_2007[name] = parse.(Float64, gapminder_2007[name])
end

data(gapminder_2007) + point + x(:gdp) + y(:life)
```

## JavaScript

```javascript
const url = "https://psychometrician.github.io/gog-book/data/gapminder_2007.csv";
const text = await (await fetch(url)).text();

const [head, ...body] = text.trim().split("\n").map((line) =>
  line.split(",").map((cell) => cell.replace(/^"|"$/g, "")));
const gapminder_2007 = Object.fromEntries(
  head.map((name, i) => [name, body.map((row) => row[i])]));

for (const name of ["gdp", "life", "population"])
  gapminder_2007[name] = gapminder_2007[name].map(Number);

plot(data(gapminder_2007), point, x(col.gdp), y(col.life));
```

:::

The plot is the same in all four, because the table is the same file and the
engine is the same program.

## Two things a CSV cannot say

A CSV is text. It records what a value is, never what kind of thing it is. Two
kinds of information are therefore lost in the file and have to be restored when
you read it. Both matter to the plot, so this book states them rather than
hiding them.

**The order of a category.** A column such as `direction` holds the eight points
of the compass. Their order is N, NE, E, SE, S, SW, W, NW, which is the order a
compass runs in. A CSV cannot record that, so every language sorts them
alphabetically instead: E, N, NE, NW, S, SE, SW, W. The plot still draws, and it
is wrong in a way that is easy to miss. Declaring the order fixes it:

```{r}
winds_local <- read.csv("data/winds.csv")
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds_local$direction <- factor(winds_local$direction, levels = compass)

data(winds_local) + bar * count + x(direction) + polar()
```

*"Given the winds read from the file: bars derived by count, x is direction."*

The compass now runs the way a compass runs. Without the third line it would
start at E and the wind rose would point the wrong way.

Each language spells the declaration differently: R uses `factor(levels = )`,
Python and JavaScript use `ordered()`, and Julia uses `ordered()` too. The
[Data](data.qmd) chapter covers this in full, because declared order is part of
the grammar and not a detail of file formats.

**Labels that look like numbers.** Three columns in these tables hold text that a
reader will turn into numbers unless told otherwise. `census.csv` has an `age`
column of `0`, `5`, `10`, which are the names of age bands rather than
quantities. `sessions.csv` has a `session` column of `01`, `02`, `03`, and a
number cannot keep a leading zero, so `01` comes back as `1`. Read those columns
as text:

```r
census <- read.csv("census.csv", colClasses = c(age = "character"))
```

In Python's `csv` module every value is already text, so nothing is needed. With
pandas, pass `dtype={"age": str}`. In Julia, read with `String`. In JavaScript,
values arrive as strings and only become numbers where you convert them.

## The tables

```{r}
#| echo: false
files <- sort(list.files("data", pattern = "[.]csv$"))
info <- data.frame(
  Table = sub("[.]csv$", "", files),
  Rows  = vapply(files, function(f) nrow(read.csv(file.path("data", f))), integer(1)),
  Columns = vapply(files, function(f) ncol(read.csv(file.path("data", f))), integer(1))
)
knitr::kable(info, row.names = FALSE)
```

The tables are not all equally serious. The gapminder cuts are real figures from
the Gapminder Foundation, released into the public domain. `iris_flowers`,
`maunga_whau` and `quakes_fiji` are reshaped from tables that ship with R, and
carry R's own license. Everything else is written for this book to make a
particular point, and `census` in particular is two plausible city profiles
rather than any real census. The terms for each are in
[`data/LICENSE.md`](data/LICENSE.md).
