# The book's data

```{r}
#| include: false
source("R/setup.R")
```

Where do `gapminder_2007`, `medals` and `winds` come from, and how do you get
them?

Every chapter of this book uses the same shared tables, listed in full at the
end of this page. They are published beside the book as CSV files, one per
table, so you can run any example in this book yourself. Each file has a fixed
address:

```
https://psychometrician.github.io/gog-book/data/<name>.csv
```

`gapminder_2007.csv` holds the 142 countries the first chapter plots.
`medals.csv` holds the medal counts the reading chapter draws, and `winds.csv`
holds the wind observations the polar chapter draws.

Reading one takes a single call, `gog_table()`, which comes with the package in
all four languages. Nothing else needs installing. You do not have to use it.
The files are ordinary CSVs at a fixed address, and any reader that
takes a web address reads them.

## Reading a table

To run the first chapter's plot yourself, you need its table in your session.
One line fetches it. `gog_table()` reads any of these tables by name:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
spellings(
  r = r"(gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life))",
  python = r"(gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(col.gdp) + y(col.life))",
  julia = r"(gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(:gdp) + y(:life))",
  js = r"(const gapminder_2007 = await gog_table("gapminder_2007");
plot(data(gapminder_2007), point, x(col.gdp), y(col.life));)"
)
```

Those four calls draw the same picture, because the table is the same file and
the engine is the same program.

`gog_table()` also gives each column the right type. A CSV is text, so numbers arrive as
text and have to be converted. A column becomes numbers when every value in it
is a number, and stays text otherwise. The `text` argument forces a column to
stay text. Three columns in these tables need that, and
[Labels that look like numbers](#labels-that-look-like-numbers), below, explains
which and why.

Why does a graphics package carry a table reader at all? R reads a CSV in one
call, and Python and Julia in a few. JavaScript has no CSV reader in its
standard library, so the same work there takes dozens of lines. One country in
`gapminder_2007.csv` is `"Congo, Dem. Rep."`, and its name holds a comma, so
splitting each line on commas gives that row seven fields where the header has
six. Nobody should have to paste dozens of lines of parsing before drawing, so
each package carries `gog_table()` instead.

R's `read.csv()`, `urllib` in Python, `Downloads` in Julia and `fetch` in
JavaScript are all standard, so `gog_table()` needs nothing beyond the package
you already installed.

Read a file with a reader of your own, and the rest of this page still applies,
because `data()` takes any table your session holds. You then do two things
yourself. Your reader decides each column's type. The three columns named below
then need whatever it offers for keeping a column as text. And a declared order
is never in the file, so you declare it after reading, whichever reader you
use. [Two things a CSV cannot record](#two-things-a-csv-cannot-record) covers
both.

Reading a different table is the same call with a different name:
`gog_table("winds")`, `gog_table("medals")`. A misspelled name is refused, and
the refusal suggests the nearest published name, or points to the list at the
end of this page.

## Two things a CSV cannot record

A CSV records what a value is, and never what kind of thing it is. Two kinds of
information are therefore lost when the file is written. You restore them when
you read the file back. Both change the plot, and this page takes each in turn.

### A declared order

The eight compass directions have a fixed order, and a plot of them is correct
only when it keeps that order. The `winds` table records one wind
observation per row:

```{r}
#| echo: false
winds <- read.csv("data/winds.csv")
peek(winds)
```

The `direction` column of `winds` holds the eight points of the compass. Their
order is N, NE, E, SE, S, SW, W, NW, clockwise from north. The
[Data](data.qmd#declaring-the-order-of-categories) chapter shows how a declared
order sets the axis, the color assignment and the legend. A CSV cannot record
that declaration. `gog_table("winds")` therefore returns `direction` as plain
text, and the declaration is yours to make after reading.

Until you declare it, only the file's own row order keeps the plot correct. The
winds file lists its rows in compass order, so a plot made without the
declaration reads correctly. Sort the rows by `speed`, and the same plot
changes. The directions now follow the sorted rows, and that order has nothing
to do with the compass. The plot still draws, and nothing on it shows that the
compass is in the wrong order.

The example below reads the table, sorts it by `speed`, and then declares the
compass order:

```{r}
#| echo: false
spellings(
  r = r"(winds <- gog_table("winds")
# `base::order`, because gog's `order()` masks it
winds <- winds[base::order(winds$speed), ]
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds$direction <- factor(winds$direction, levels = compass)

data(winds) + bar * count + x(direction))",
  python = r"(winds = gog_table("winds")
by_speed = sorted(enumerate(winds["speed"]), key=lambda pair: pair[1])
winds = {name: [column[i] for i, _ in by_speed]
         for name, column in winds.items()}
compass = ["N", "NE", "E", "SE", "S", "SW", "W", "NW"]
winds["direction"] = ordered(winds["direction"], compass)

data(winds) + bar * count + x(col.direction))",
  julia = r"(winds = gog_table("winds")
by_speed = sortperm(winds["speed"])
winds = Dict{String,Any}(name => column[by_speed] for (name, column) in winds)
compass = ["N", "NE", "E", "SE", "S", "SW", "W", "NW"]
winds["direction"] = ordered(winds["direction"], compass)

data(winds) + bar * count + x(:direction))",
  js = r"(const winds = await gog_table("winds");
const by_speed = winds.speed.map((_, i) => i)
  .sort((a, b) => winds.speed[a] - winds.speed[b]);
for (const name in winds) winds[name] = by_speed.map((i) => winds[name][i]);
const compass = ["N", "NE", "E", "SE", "S", "SW", "W", "NW"];
winds.direction = ordered(winds.direction, compass);

plot(data(winds), layer(bar, count), x(col.direction));)"
)
```

```{r}
#| echo: false
# The plot the four tabs above describe, drawn from the local copy of the same
# file, because the book does not fetch over the network while it builds.
winds <- read.csv("data/winds.csv")
winds <- winds[base::order(winds$speed), ]
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds$direction <- factor(winds$direction, levels = compass)

data(winds) + bar * count + x(direction)
```

*"Given the winds: bars derived by count, x is direction."*

The rows are in speed order, and the axis still reads N to NW. The declared
order decides the axis, not the order of the rows. The [polar](polar.qmd)
chapter draws this same count as a wind rose, and the change is one word. Add
`polar()`, and the axis bends into a circle, with N at the top and the rest
clockwise from it. The declared order goes with it:

```{r}
data(winds) + bar * count + x(direction) + polar()
```

*"Given the winds: bars derived by count, x is direction, in polar."*

The polar chapter explains that plot. Here it shows that a declared order
survives a change of space. The circle starts at the first category and runs
clockwise through the rest, so the compass reads correctly in both.

The sort exists only to undo the file's compass order. The declaration is the
part you keep. Declare it once after each `gog_table("winds")`, and every plot
drawn from `winds` in the rest of the book reads the compass in order.

### Labels that look like numbers

The second thing a CSV cannot record is which columns are text. Three columns
in these tables hold text that `gog_table()` would read as numbers unless you
name them. `census.csv` has an `age` column of `0`, `5`, `10`, which are the
names of age bands rather than quantities. `sessions.csv` has a `session`
column of `01`, `02`, `03`, and a number cannot keep a leading zero, so `01`
comes back as `1`. `gm_eras.csv` has an `era` column of `1957` and `2007`.
Those two values name the eras the rows compare; they do not measure anything.
That is the whole reason `gog_table()` takes a `text` argument:

```{r}
#| echo: false
spellings(
  r = r"(census <- gog_table("census", text = "age")
sessions <- gog_table("sessions", text = "session")
gm_eras <- gog_table("gm_eras", text = "era"))",
  python = r"(census = gog_table("census", text=("age",))
sessions = gog_table("sessions", text=("session",))
gm_eras = gog_table("gm_eras", text=("era",)))",
  julia = r"(census = gog_table("census"; text = ["age"])
sessions = gog_table("sessions"; text = ["session"])
gm_eras = gog_table("gm_eras"; text = ["era"]))",
  js = r"(const census = await gog_table("census", ["age"]);
const sessions = await gog_table("sessions", ["session"]);
const gm_eras = await gog_table("gm_eras", ["era"]);)"
)
```

Without it, `age` becomes numbers, and an axis that should show one slot per
age band is drawn as a numeric scale from 0 to 85. The plot looks reasonable,
so nothing tells you the axis is wrong.

Each call above names one column, and a `text` argument can name more than
one. `gm_eras` holds the same values in two columns: `year` as a number, `era`
as a label. The book reads only `era` as text, so `year` stays a number. To
read the year as a label too, name both columns in one `text` argument:

```{r}
#| echo: false
spellings(
  r = r"(gm_eras <- gog_table("gm_eras", text = c("year", "era")))",
  python = r"(gm_eras = gog_table("gm_eras", text=("year", "era")))",
  julia = r"(gm_eras = gog_table("gm_eras"; text = ["year", "era"]))",
  js = r"(const gm_eras = await gog_table("gm_eras", ["year", "era"]);)"
)
```

Each language writes more than one name in its own way. R puts them inside
`c()`, Python inside parentheses, and Julia and JavaScript inside square
brackets. Only R's spelling changes from the block above. The other three
already put one name inside parentheses or brackets, so a second name joins it
there.

## The tables

Those three tables need a column named in a `text` argument, and no other
shared table does. The list below names every shared table this book reads,
with its rows and columns. The [preface](index.qmd#the-cast) introduces the
eight families of table that most plots read. A family can cover more than one
name below, so the list is longer than eight. Every other table is introduced
on the page that first draws it. Each name goes in place of `<name>` in the
address at the top of this page. The same name is what you give `gog_table()`:

```{r}
#| echo: false
files <- sort(list.files("data", pattern = "[.]csv$"))
info <- data.frame(
  Table = sub("[.]csv$", "", files),
  Rows  = vapply(files, function(f) nrow(read.csv(file.path("data", f))), integer(1)),
  Columns = vapply(files, function(f) ncol(read.csv(file.path("data", f))), integer(1))
)
knitr::kable(info, row.names = FALSE)
```

Not every table here holds real data. The tables whose names begin with
`gapminder` or `gm` hold real figures from the Gapminder Foundation, released
into the public domain. `iris_flowers`, `maunga_whau`, `quakes_fiji` and
`titanic` are reshaped from tables included with R, and carry R's own
license. `world_borders` is Natural Earth's public-domain outline of the
countries. Everything else was written for this book, each table to show one
thing; `census` is two plausible city profiles, not a real census. The terms
for each are in [`data/LICENSE.md`](data/LICENSE.md).
