# The book's data

```{r}
#| include: false
source("R/setup.R")
```

Where do `gapminder_2007`, `medals` and `winds` come from, and how do you get
them?

[Data](data.qmd) was about a table of your own. This chapter is about the shared
tables that the book's examples draw from, listed in full under
[The tables](#the-tables). They are published
beside the book as CSV files, one per table, so you can run any example in this
book yourself. Each file has a fixed address:

```
https://psychometrician.github.io/gog-book/data/<name>.csv
```

`gapminder_2007.csv` holds the 142 countries drawn in the chapter
[A first plot](first-plot.qmd). `medals.csv` holds the medal counts drawn in
[Reading practice](reading.qmd), and `winds.csv` holds the wind observations
drawn in [Polar](polar.qmd).

Reading one takes a single call, `gog_table()`, which comes with the package in
all four languages. Nothing else needs to be installed.

## Reading a table

To run the first chapter's plot yourself, you need its table in your session.
One line fetches it. `gog_table()` reads any of these tables by name:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| echo: false
spellings(
  r = r"(gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life))",
  python = r"(gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(col.gdp) + y(col.life))",
  julia = r"(gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(:gdp) + y(:life))",
  js = r"(const gapminder_2007 = await gog_table("gapminder_2007");
plot(data(gapminder_2007), point, x(col.gdp), y(col.life));)"
)
```

Those four calls draw the same picture, because the table is the same file and
the engine is the same program.

`gog_table()` also decides each column's type. It is right for most columns, but
not for a label written as a number, or for a date. A CSV is text, so numbers
arrive as text and have to be converted. A column becomes numbers when every
value in it is a number, and stays text otherwise. The `text` argument forces a
column to stay text. Three columns in these tables need that, and [Labels that
look like numbers](#labels-that-look-like-numbers), below, explains which and
why.

Why does a graphics package carry a table reader at all? R reads a CSV in one
call, and Python and Julia in a few. JavaScript has no CSV reader in its
standard library, so the same work there takes dozens of lines. A CSV reader
cannot simply split each line on commas. Six countries in `gapminder_2007.csv`
have a comma in the name, `"Congo, Dem. Rep."` among them, so that split gives
those rows seven fields, while the header has six. Nobody should have to paste
dozens of lines of parsing before drawing, so each package carries `gog_table()`
instead.

Each package fetches the file with what its language already has: R's
`read.csv()`, `urllib` in Python, `Downloads` in Julia and `fetch` in
JavaScript.

Read a file with a CSV reader of your own, and the rest of this page still
applies, because `data()` takes any table your session holds. You then do two
things yourself. First, check each column's type: your CSV reader decides it,
so the three columns named below need whatever that reader offers for keeping a
column as text. Second, declare the category order, because a CSV never records
one. You do this after `gog_table()` too.
[Three things a CSV cannot record](#three-things-a-csv-cannot-record) covers
both, and dates as well.

Reading a different table is the same call with a different name:
`gog_table("winds")`, `gog_table("medals")`. A misspelled name is refused, and
the refusal suggests the nearest published name, or points to the list under
[The tables](#the-tables).

## Three things a CSV cannot record

A CSV records the values in a column, and never says what kind of column it is.
Three kinds of information are therefore lost when the file is written. You
restore them when you read the file back. Each one changes the plot, and this
page explains each one.

### A declared order

The eight compass points have a fixed order, and a plot of them is correct
only when it keeps that order. The `winds` table records one wind
observation per row:

```{r}
#| echo: false
winds <- read.csv("data/winds.csv")
peek(winds)
```

The `direction` column of `winds` holds the eight compass points. Their
order is N, NE, E, SE, S, SW, W, NW, clockwise from north. The
[Data](data.qmd#declaring-the-order-of-categories) chapter shows how a declared
order sets the axis, the color assignment and the legend. A CSV cannot record
that declaration. `gog_table("winds")` therefore returns `direction` as plain
text, and the declaration is yours to make after reading.

Until you declare it, only the file's own row order keeps the plot correct. The
winds file lists its rows in compass order, so a plot of `direction` made
without the declaration reads correctly. Sort the rows by `speed`, and the same
plot changes. The compass points now follow the sorted rows, and that order is not
the compass order. The plot still draws, and nothing on it shows that the
compass is in the wrong order.

The example below reads the table, sorts it by `speed` so that the rows are no
longer in compass order, and then declares the compass order:

```{r}
#| echo: false
spellings(
  r = r"(winds <- gog_table("winds")
# `base::order`, because gog's `order()` masks it
winds <- winds[base::order(winds$speed), ]
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds$direction <- factor(winds$direction, levels = compass)

data(winds) + bar * count + x(direction))",
  python = r"(winds = gog_table("winds")
by_speed = sorted(enumerate(winds["speed"]), key=lambda pair: pair[1])
winds = {name: [column[i] for i, _ in by_speed]
         for name, column in winds.items()}
compass = ["N", "NE", "E", "SE", "S", "SW", "W", "NW"]
winds["direction"] = ordered(winds["direction"], compass)

data(winds) + bar * count + x(col.direction))",
  julia = r"(winds = gog_table("winds")
by_speed = sortperm(winds["speed"])
winds = Dict{String,Any}(name => column[by_speed] for (name, column) in winds)
compass = ["N", "NE", "E", "SE", "S", "SW", "W", "NW"]
winds["direction"] = ordered(winds["direction"], compass)

data(winds) + bar * count + x(:direction))",
  js = r"(const winds = await gog_table("winds");
const by_speed = winds.speed.map((_, i) => i)
  .sort((a, b) => winds.speed[a] - winds.speed[b]);
for (const name in winds) winds[name] = by_speed.map((i) => winds[name][i]);
const compass = ["N", "NE", "E", "SE", "S", "SW", "W", "NW"];
winds.direction = ordered(winds.direction, compass);

plot(data(winds), layer(bar, count), x(col.direction));)"
)
```

```{r}
#| echo: false
# The plot the four tabs above describe, drawn from the local copy of the same
# file, because the book does not fetch over the network while it builds.
winds <- read.csv("data/winds.csv")
winds <- winds[base::order(winds$speed), ]
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds$direction <- factor(winds$direction, levels = compass)

data(winds) + bar * count + x(direction)
```

*"Given the winds: bars derived by count, x is direction."*

The rows are in speed order, and the axis still reads N to NW. The declared
order decides the axis, not the order of the rows. A declared order is kept when
the space changes. The [polar](polar.qmd) chapter draws this same count as a
wind rose, and the change is one word. Add `polar()`, and the axis bends into a
circle, with N at the top and the rest clockwise from it:

```{r}
data(winds) + bar * count + x(direction) + polar()
```

*"Given the winds: bars derived by count, x is direction, in polar."*

The polar chapter explains that plot. The circle starts at the first
category and runs clockwise through the rest, so the compass reads correctly in
the flat plot and in the circle.

The sort exists only to undo the file's compass order. Keep the declaration in
your own code, and drop the sort. `winds` needs a second declaration, for `season`:
the book's plots show Summer before Winter, and the file lists Winter first.
Three other tables need one too: `survived` in `titanic` (Yes, then No), `slab`
in `quakes_fiji` (shallowest band first) and `weekday` in `six_weeks` (Monday
first).

### Labels that look like numbers

The second thing a CSV cannot record is which columns are text. Three columns
in these tables hold text that `gog_table()` would read as numbers unless you
name them. `census.csv` has an `age` column of `0`, `5`, `10`, which are the
names of age bands rather than quantities. `sessions.csv` has a `session`
column of `01`, `02`, `03`, and a number cannot keep a leading zero, so `01`
comes back as `1`. `gm_eras.csv` has an `era` column of `1957` and `2007`.
Those two values name the eras the rows compare; they do not measure anything.
That is the whole reason `gog_table()` takes a `text` argument:

```{r}
#| echo: false
spellings(
  r = r"(census <- gog_table("census", text = "age")
sessions <- gog_table("sessions", text = "session")
gm_eras <- gog_table("gm_eras", text = "era"))",
  python = r"(census = gog_table("census", text=("age",))
sessions = gog_table("sessions", text=("session",))
gm_eras = gog_table("gm_eras", text=("era",)))",
  julia = r"(census = gog_table("census"; text = ["age"])
sessions = gog_table("sessions"; text = ["session"])
gm_eras = gog_table("gm_eras"; text = ["era"]))",
  js = r"(const census = await gog_table("census", ["age"]);
const sessions = await gog_table("sessions", ["session"]);
const gm_eras = await gog_table("gm_eras", ["era"]);)"
)
```

Without the `text` argument, `age` becomes numbers. A `point` or a `line` then
places the bands along a continuous axis from 0 to 85. The plot looks
reasonable, so nothing tells you the axis is wrong. The population pyramid in
[Many groups](cookbook/groups.qmd) is refused, because it sorts the bands with
`order()`, and `order()` needs a categorical axis.

Each call above names one column, and a `text` argument can name more than
one. `gm_eras` holds the same values in two columns: `year` as a number, `era`
as a label. The book reads only `era` as text, so `year` stays a number. To
read the year as a label too, name both columns in one `text` argument:

```{r}
#| echo: false
spellings(
  r = r"(gm_eras <- gog_table("gm_eras", text = c("year", "era")))",
  python = r"(gm_eras = gog_table("gm_eras", text=("year", "era")))",
  julia = r"(gm_eras = gog_table("gm_eras"; text = ["year", "era"]))",
  js = r"(const gm_eras = await gog_table("gm_eras", ["year", "era"]);)"
)
```

Each language writes more than one name in its own way. R puts them inside
`c()`, Python inside parentheses, and Julia and JavaScript inside square
brackets. Python no longer needs the comma that follows a single name. Julia
and JavaScript only add the second name inside the brackets.

Only `census`, `sessions` and `gm_eras` need a `text` argument; no other shared
table does.

### A date

The third thing a CSV cannot record is which columns are dates. `six_weeks`,
`revenue` and `monitoring` each have a column of dates, and `gog_table()`
returns it as text, so the axis gets one slot per date instead of a calendar
axis. After reading, convert the column to your language's date or timestamp
type: in R, `as.Date()` for `six_weeks` and `revenue`, and `as.POSIXct()` for
`monitoring`, which also records the hour.
[Scales](scales.qmd#time-a-date-column-gets-a-calendar-axis) shows the calendar
axis you get.

## The tables

The list below names every shared table this book reads, and how many rows and
columns each one has. The [preface](index.qmd#the-cast) introduces the eight
families of table that most plots read. A family can cover more than one name
below, so the list is longer than eight.

Tables outside those families are introduced on the page that first draws them.
Each name goes in place of `<name>` in the address at the top of this page. The
same name is what you give `gog_table()`:

```{r}
#| echo: false
files <- sort(list.files("data", pattern = "[.]csv$"))
info <- data.frame(
  Table = sub("[.]csv$", "", files),
  Rows  = vapply(files, function(f) nrow(read.csv(file.path("data", f))), integer(1)),
  Columns = vapply(files, function(f) ncol(read.csv(file.path("data", f))), integer(1))
)
knitr::kable(info, row.names = FALSE)
```

Not every table here holds real data. The tables whose names begin with
`gapminder` or `gm` hold figures from the Gapminder Foundation, which released
them into the public domain. `healthy_band`, `world_median` and
`population_spikes` are computed from those figures. `iris_flowers`,
`maunga_whau`, `quakes_fiji` and `titanic` are reshaped from tables included
with R, and carry R's own license. `world_borders` is Natural Earth's
public-domain outline of the countries. Everything else was written for this
book, each table to show one thing; `census` is two plausible city profiles, not
a real census. The terms for each are in [`data/LICENSE.md`](data/LICENSE.md).

## What comes next

A table holds the rows, and says nothing about how they look on the page.
[Choosing a mark](marks/index.qmd) shows how to pick the mark that draws them.
There are fourteen, and your question and your columns decide which one fits.
