# The book's data

```{r}
#| include: false
source("R/setup.R")
```

Where do `gapminder_2007`, `medals` and `winds` come from, and how do you get
them?

Every chapter of this book draws from a small set of shared tables. They are
published beside the book as CSV files, one per table, so you can run any
example in this book yourself. Each file sits at a fixed address:

```
https://psychometrician.github.io/gog-book/data/<name>.csv
```

`gapminder_2007.csv` holds the 142 countries the first chapter plots.
`winds.csv` holds the wind observations behind the polar chapter. The full list
is at the end of this page.

Reading one takes a single call, `gog_table()`, which comes with the package
in all four languages. Nothing else needs installing.

## Reading a table

`gog_table()` fetches any of the thirty-five by name. It is included with the
package, so there is nothing to copy:

::: {.panel-tabset}

## R

```r
gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life)
```

## Python

```python
gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(col.gdp) + y(col.life)
```

## Julia

```julia
gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(:gdp) + y(:life)
```

## JavaScript

```javascript
const gapminder_2007 = await gog_table("gapminder_2007");
plot(data(gapminder_2007), point, x(col.gdp), y(col.life));
```

:::

It also gives each column the right type. A CSV is text, so numbers arrive as
text and have to be converted. A column becomes numbers when every value in it
is a number, and stays text otherwise. The `text` argument forces a column to
stay text. Three columns in these tables need that, and the next section
explains which and why.

This one function is the whole reason the four packages carry anything that is
not grammar. R reads a CSV in a single call and Python and Julia in a handful,
but JavaScript has no CSV reader in its standard library, so the same work there
is more than thirty lines. One country in `gapminder_2007.csv` is
`"Congo, Dem. Rep."`, and its name holds a comma, so splitting each line on
commas gives that row seven fields where the header has six. Asking a reader to
paste that before drawing anything is what moved it into the packages.

Nothing else is needed. `Downloads` in Julia, `urllib` in Python and `fetch` in
JavaScript are all standard, so the helper costs you no dependency beyond the
package you already installed to draw.

The plot is the same in all four, because the table is the same file and the
engine is the same program.

Reading a different table is the same call with a different name:
`gog_table("winds")`, `gog_table("medals")`. The list of names is at the end of
this page.

## When the name is wrong

A table name is easy to mistype, and `gapminder2007` is not `gapminder_2007`.
gog stops instead of giving you something else. It says which name it could not
find. When one table is a near miss, it names that table:

```
gog: there is no table called "gapminder2007". Did you mean "gapminder_2007"?
```

The suggestion is offered only when the two names differ by at most two letters.
When nothing is that close, the refusal points at this page instead. A wrong
suggestion would cost you a second attempt, which is worse than no suggestion.

A site that cannot be reached is a different problem, so it gets different
words. That refusal names the connection rather than the table. The distinction
matters. Being told a table does not exist is misleading when the network is
simply down. It sends you to correct a name that was already right.

All four languages print the same sentences. Each one refuses the way the rest
of the grammar refuses, so whatever catches a refused plot catches these too.

## Two things a CSV cannot say

A CSV is text. It records what a value is, never what kind of thing it is. Two
kinds of information are therefore lost in the file and have to be restored when
you read it. Both matter to the plot, so this book states them rather than
hiding them.

The order of a category. A column such as `direction` holds the eight points
of the compass. Their order is N, NE, E, SE, S, SW, W, NW, which is the order a
compass runs in. A CSV cannot record that, so every language sorts them
alphabetically instead: E, N, NE, NW, S, SE, SW, W. The plot still draws, and it
is wrong in a way that is easy to miss. Declaring the order fixes it:

```{r}
winds_local <- read.csv("data/winds.csv")
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds_local$direction <- factor(winds_local$direction, levels = compass)

data(winds_local) + bar * count + x(direction) + polar()
```

*"Given the winds read from the file: bars derived by count, x is direction."*

The compass now runs the way a compass runs. Without the third line it would
start at E and the wind rose would point the wrong way.

Each language spells the declaration differently: R uses `factor(levels = )`,
Python and JavaScript use `ordered()`, and Julia uses `ordered()` too. The
[Data](data.qmd) chapter covers this in full, because declared order is part of
the grammar and not a detail of file formats.

Labels that look like numbers. Three columns in these tables hold text that a
reader will turn into numbers unless told otherwise. `census.csv` has an `age`
column of `0`, `5`, `10`, which are the names of age bands rather than
quantities. `sessions.csv` has a `session` column of `01`, `02`, `03`, and a
number cannot keep a leading zero, so `01` comes back as `1`. That is the whole
reason `gog_table()` takes a `text` argument:

::: {.panel-tabset}

## R

```r
census <- gog_table("census", text = "age")
sessions <- gog_table("sessions", text = "session")
```

## Python

```python
census = gog_table("census", text=("age",))
sessions = gog_table("sessions", text=("session",))
```

## Julia

```julia
census = gog_table("census"; text = ["age"])
sessions = gog_table("sessions"; text = ["session"])
```

## JavaScript

```javascript
const census = await gog_table("census", ["age"]);
const sessions = await gog_table("sessions", ["session"]);
```

:::

Without it, `age` becomes the numbers 0, 5 and 10, and an axis that should read
as five labels is drawn as a scale from 0 to 10. The plot looks reasonable, which
is what makes the mistake worth naming.

## The tables

```{r}
#| echo: false
files <- sort(list.files("data", pattern = "[.]csv$"))
info <- data.frame(
  Table = sub("[.]csv$", "", files),
  Rows  = vapply(files, function(f) nrow(read.csv(file.path("data", f))), integer(1)),
  Columns = vapply(files, function(f) ncol(read.csv(file.path("data", f))), integer(1))
)
knitr::kable(info, row.names = FALSE)
```

The tables are not all equally serious. The gapminder cuts are real figures from
the Gapminder Foundation, released into the public domain. `iris_flowers`,
`maunga_whau` and `quakes_fiji` are reshaped from tables included with R, and
carry R's own license. Everything else is written for this book to make a
particular point, and `census` in particular is two plausible city profiles
rather than any real census. The terms for each are in
[`data/LICENSE.md`](data/LICENSE.md).
