# The book's data

```{r}
#| include: false
source("R/setup.R")
```

Where do `gapminder_2007`, `medals` and `winds` come from, and how do you get
them?

Every chapter of this book draws from a small set of shared tables. They are
published beside the book as CSV files, one per table, so you can run any
example in this book yourself. Each file sits at a fixed address:

```
https://psychometrician.github.io/gog-book/data/<name>.csv
```

`gapminder_2007.csv` holds the 142 countries the first chapter plots.
`winds.csv` holds the wind observations behind the polar chapter. The full list
is at the end of this page.

Nothing needs installing to read them. R, Python, Julia and JavaScript each
read a CSV with their own standard library, so fetching a table adds no
dependency to your project.

## Reading a table

Every table on this page is read the same way, so what follows is one function
per language. It fetches any of the thirty-five by name. Copy it once, and you
can run every example in this book.

The function also gives each column the right type. A CSV is text, so numbers
arrive as text and have to be converted. Each version converts a column when
every value in it is a number, and leaves the column alone otherwise. The `text`
argument forces a column to stay text. Three columns in these tables need that,
and the next section explains which and why.

::: {.panel-tabset}

## R

```r
gog_table <- function(name, text = character()) {
  url <- paste0("https://psychometrician.github.io/gog-book/data/", name, ".csv")
  read.csv(url, colClasses = setNames(rep("character", length(text)), text))
}

gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life)
```

`read.csv()` chooses the column types itself, so almost all of this function is
the address. `colClasses` is how you overrule it for one column.

## Python

```python
import csv, urllib.request

def gog_table(name, text=()):
    url = f"https://psychometrician.github.io/gog-book/data/{name}.csv"
    with urllib.request.urlopen(url) as response:
        rows = list(csv.DictReader(response.read().decode("utf-8").splitlines()))
    table = {}
    for key in rows[0]:
        values = [row[key] for row in rows]
        try:
            table[key] = values if key in text else [float(value) for value in values]
        except ValueError:
            table[key] = values
    return table

gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(col.gdp) + y(col.life)
```

`csv.DictReader` reads every value as text, so the loop is where the types are
decided. A column that will not convert is a text column, which is what the
`except` branch means.

With pandas the read is one line, `pd.read_csv(url)`, and the equivalent of
`text` is `dtype={"session": str}`.

## Julia

```julia
using Downloads, DelimitedFiles

function gog_table(name; text = String[])
    url = "https://psychometrician.github.io/gog-book/data/$name.csv"
    raw, header = readdlm(Downloads.download(url), ',', String; header = true)
    table = Dict{String,Any}()
    for (i, column) in enumerate(header)
        key = strip(column, '"')
        values = raw[:, i]
        numbers = tryparse.(Float64, values)
        table[key] = key in text || any(isnothing, numbers) ? values : numbers
    end
    table
end

gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(:gdp) + y(:life)
```

`tryparse` returns `nothing` instead of failing, so one pass over a column
answers both questions at once: what the numbers are, and whether they are all
numbers.

## JavaScript

```javascript
// A comma can sit inside a quoted field, so a line cannot be split on commas.
function parse_csv(text) {
  const rows = [[]];
  let field = "", quoted = false;
  for (let i = 0; i < text.length; i++) {
    const c = text[i];
    if (quoted) {
      if (c !== '"') field += c;
      else if (text[i + 1] === '"') { field += '"'; i++; }
      else quoted = false;
    } else if (c === '"') quoted = true;
    else if (c === ",") { rows.at(-1).push(field); field = ""; }
    else if (c === "\n") { rows.at(-1).push(field); field = ""; rows.push([]); }
    else if (c !== "\r") field += c;
  }
  rows.at(-1).push(field);
  if (rows.at(-1).length === 1 && rows.at(-1)[0] === "") rows.pop();
  return rows;
}

async function gog_table(name, text = []) {
  const url = `https://psychometrician.github.io/gog-book/data/${name}.csv`;
  const [head, ...body] = parse_csv((await (await fetch(url)).text()).trim());
  return Object.fromEntries(head.map((key, i) => {
    const values = body.map((row) => row[i]);
    if (text.includes(key)) return [key, values];
    const numbers = values.map(Number);
    return [key, numbers.some((n) => Number.isNaN(n)) ? values : numbers];
  }));
}

const gapminder_2007 = await gog_table("gapminder_2007");
plot(data(gapminder_2007), point, x(col.gdp), y(col.life));
```

JavaScript has no CSV reader in its standard library, which is why this tab is
the long one. The fourteen lines are not decoration. One country in
`gapminder_2007.csv` is `"Congo, Dem. Rep."`, and its name holds a comma, so
splitting each line on commas gives that row seven fields where the header has
six. R, Python and Julia all handle quoting for you.

:::

The plot is the same in all four, because the table is the same file and the
engine is the same program.

Reading a different table is the same call with a different name:
`gog_table("winds")`, `gog_table("medals")`. The list of names is at the end of
this page.

## Two things a CSV cannot say

A CSV is text. It records what a value is, never what kind of thing it is. Two
kinds of information are therefore lost in the file and have to be restored when
you read it. Both matter to the plot, so this book states them rather than
hiding them.

**The order of a category.** A column such as `direction` holds the eight points
of the compass. Their order is N, NE, E, SE, S, SW, W, NW, which is the order a
compass runs in. A CSV cannot record that, so every language sorts them
alphabetically instead: E, N, NE, NW, S, SE, SW, W. The plot still draws, and it
is wrong in a way that is easy to miss. Declaring the order fixes it:

```{r}
winds_local <- read.csv("data/winds.csv")
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds_local$direction <- factor(winds_local$direction, levels = compass)

data(winds_local) + bar * count + x(direction) + polar()
```

*"Given the winds read from the file: bars derived by count, x is direction."*

The compass now runs the way a compass runs. Without the third line it would
start at E and the wind rose would point the wrong way.

Each language spells the declaration differently: R uses `factor(levels = )`,
Python and JavaScript use `ordered()`, and Julia uses `ordered()` too. The
[Data](data.qmd) chapter covers this in full, because declared order is part of
the grammar and not a detail of file formats.

**Labels that look like numbers.** Three columns in these tables hold text that a
reader will turn into numbers unless told otherwise. `census.csv` has an `age`
column of `0`, `5`, `10`, which are the names of age bands rather than
quantities. `sessions.csv` has a `session` column of `01`, `02`, `03`, and a
number cannot keep a leading zero, so `01` comes back as `1`. That is the whole
reason `gog_table()` takes a `text` argument:

::: {.panel-tabset}

## R

```r
census <- gog_table("census", text = "age")
sessions <- gog_table("sessions", text = "session")
```

## Python

```python
census = gog_table("census", text=("age",))
sessions = gog_table("sessions", text=("session",))
```

## Julia

```julia
census = gog_table("census"; text = ["age"])
sessions = gog_table("sessions"; text = ["session"])
```

## JavaScript

```javascript
const census = await gog_table("census", ["age"]);
const sessions = await gog_table("sessions", ["session"]);
```

:::

Without it, `age` becomes the numbers 0, 5 and 10, and an axis that should read
as five labels is drawn as a scale from 0 to 10. The plot looks reasonable, which
is what makes the mistake worth naming.

## The tables

```{r}
#| echo: false
files <- sort(list.files("data", pattern = "[.]csv$"))
info <- data.frame(
  Table = sub("[.]csv$", "", files),
  Rows  = vapply(files, function(f) nrow(read.csv(file.path("data", f))), integer(1)),
  Columns = vapply(files, function(f) ncol(read.csv(file.path("data", f))), integer(1))
)
knitr::kable(info, row.names = FALSE)
```

The tables are not all equally serious. The gapminder cuts are real figures from
the Gapminder Foundation, released into the public domain. `iris_flowers`,
`maunga_whau` and `quakes_fiji` are reshaped from tables that ship with R, and
carry R's own license. Everything else is written for this book to make a
particular point, and `census` in particular is two plausible city profiles
rather than any real census. The terms for each are in
[`data/LICENSE.md`](data/LICENSE.md).
