# The book's data

```{r}
#| include: false
source("R/setup.R")
```

Where do `gapminder_2007`, `medals` and `winds` come from, and how do you get
them?

Every chapter of this book draws from a small set of shared tables. They are
published beside the book as CSV files, one per table, so you can run any
example in this book yourself. Each file has a fixed address:

```
https://psychometrician.github.io/gog-book/data/<name>.csv
```

`gapminder_2007.csv` holds the 142 countries the first chapter plots.
`winds.csv` holds the wind observations the polar chapter draws. The full list
is at the end of this page.

Reading one takes a single call, `gog_table()`, which comes with the package
in all four languages. Nothing else needs installing.

## Reading a table

`gog_table()` reads any of them by name:

::: {.panel-tabset}

## R

```r
gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life)
```

## Python

```python
gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(col.gdp) + y(col.life)
```

## Julia

```julia
gapminder_2007 = gog_table("gapminder_2007")
data(gapminder_2007) + point + x(:gdp) + y(:life)
```

## JavaScript

```javascript
const gapminder_2007 = await gog_table("gapminder_2007");
plot(data(gapminder_2007), point, x(col.gdp), y(col.life));
```

:::

It also gives each column the right type. A CSV is text, so numbers arrive as
text and have to be converted. A column becomes numbers when every value in it
is a number, and stays text otherwise. The `text` argument forces a column to
stay text. Three columns in these tables need that, and the next section
explains which and why.

This one function is the whole reason the four packages carry anything that is
not grammar. R reads a CSV in one call, and Python and Julia in a few.
JavaScript has no CSV reader in its standard library, so the same work there
is more than thirty lines. One country in `gapminder_2007.csv` is
`"Congo, Dem. Rep."`, and its name holds a comma, so splitting each line on
commas gives that row seven fields where the header has six. Nobody should
have to paste thirty lines of parsing before drawing, so each package carries
`gog_table()` instead.

Nothing else is needed. `urllib` in Python, `Downloads` in Julia and `fetch`
in JavaScript are all standard, so `gog_table()` needs nothing beyond the
package you already installed.

The plot is the same in all four, because the table is the same file and the
engine is the same program.

Reading a different table is the same call with a different name:
`gog_table("winds")`, `gog_table("medals")`.

## When the name is wrong

A table name is easy to mistype, and `gapminder2007` is not `gapminder_2007`.
gog stops instead of giving you something else. It says which name it could not
find. When one table is a near miss, it names that table:

```
gog: there is no table called "gapminder2007". Did you mean "gapminder_2007"?
```

The suggestion is offered only when the two names differ by at most two letters.
When nothing is that close, the refusal points at this page instead. A wrong
suggestion would cost you a second attempt, which is worse than no suggestion.

A site that cannot be reached is a different problem, so it gets different
words. That refusal names the connection rather than the table. The distinction
matters. Being told a table does not exist is misleading when the network is
simply down. It sends you to correct a name that was already right.

All four languages print the same sentences. Each one refuses the way the rest
of the grammar refuses, so whatever catches a refused plot catches these too.

## Two things a CSV cannot record

A CSV is text. It records what a value is, never what kind of thing it is. Two
kinds of information are therefore lost in the file and have to be restored when
you read it. Both matter to the plot, so this book states them rather than
hiding them.

### The order of a category

A column such as `direction` holds the eight points
of the compass. Their order is N, NE, E, SE, S, SW, W, NW, which is the order a
compass runs in. A CSV cannot record that, so every language sorts them
alphabetically instead: E, N, NE, NW, S, SE, SW, W. The plot still draws, and it
is wrong in a way that is easy to miss. This example reads the raw file on
purpose: `gog_table()` already declares the order for you, and the point here
is what happens without it. Declaring the order fixes it:

```{r}
winds_local <- read.csv("data/winds.csv")
compass <- c("N", "NE", "E", "SE", "S", "SW", "W", "NW")
winds_local$direction <- factor(winds_local$direction, levels = compass)

data(winds_local) + bar * count + x(direction) + polar()
```

*"Given the winds read from the file: bars derived by count, x is direction."*

The directions now follow the compass, from N clockwise to NW. Without the
third line the circle would start at E, and the wind rose (the circular count
of winds the polar chapter draws) would be wrong in a way that is easy to
miss.

Each language spells the declaration differently: R uses `factor(levels = )`,
and Python, Julia and JavaScript use `ordered()`. The
[Data](data.qmd) chapter covers this in full, because declared order is part of
the grammar and not a detail of file formats.

### Labels that look like numbers

Three columns in these tables hold text that `gog_table()` would read as
numbers unless you name them. `census.csv` has an `age`
column of `0`, `5`, `10`, which are the names of age bands rather than
quantities. `sessions.csv` has a `session` column of `01`, `02`, `03`, and a
number cannot keep a leading zero, so `01` comes back as `1`. `gm_eras.csv`
has an `era` column of `1957` and `2007`, which name the two eras its rows
compare rather than measure anything. That is the whole
reason `gog_table()` takes a `text` argument:

::: {.panel-tabset}

## R

```r
census <- gog_table("census", text = "age")
sessions <- gog_table("sessions", text = "session")
gm_eras <- gog_table("gm_eras", text = "era")
```

## Python

```python
census = gog_table("census", text=("age",))
sessions = gog_table("sessions", text=("session",))
gm_eras = gog_table("gm_eras", text=("era",))
```

## Julia

```julia
census = gog_table("census"; text = ["age"])
sessions = gog_table("sessions"; text = ["session"])
gm_eras = gog_table("gm_eras"; text = ["era"])
```

## JavaScript

```javascript
const census = await gog_table("census", ["age"]);
const sessions = await gog_table("sessions", ["session"]);
const gm_eras = await gog_table("gm_eras", ["era"]);
```

:::

Without it, `age` becomes numbers, and an axis that should show one slot per
age band is drawn as a numeric scale from 0 to 85. The plot looks reasonable,
so nothing tells you the axis is wrong.

## The tables

```{r}
#| echo: false
files <- sort(list.files("data", pattern = "[.]csv$"))
info <- data.frame(
  Table = sub("[.]csv$", "", files),
  Rows  = vapply(files, function(f) nrow(read.csv(file.path("data", f))), integer(1)),
  Columns = vapply(files, function(f) ncol(read.csv(file.path("data", f))), integer(1))
)
knitr::kable(info, row.names = FALSE)
```

Not every table here holds real data. The tables whose names begin with
`gapminder` or `gm` hold real figures from the Gapminder Foundation, released
into the public domain. `iris_flowers`, `maunga_whau`, `quakes_fiji` and
`titanic` are reshaped from tables included with R, and carry R's own
license. `world_borders` is Natural Earth's public-domain outline of the
countries. Everything else was written for this book, each table to show one
thing; `census` is two plausible city profiles, not a real census. The terms
for each are in [`data/LICENSE.md`](data/LICENSE.md).
