# Channels and legends

```{r}
#| include: false
source("R/setup.R")
```

Which visual property should carry which column? A **channel** binds a data
column to a visual property of a mark, and the channels do not all do the same
work. `x`, `y` and `z` are the **vowels**: without one a mark does not render.
Five other channels get a legend: `color`, `size`, `shape`, `pattern` and
`opacity`. Those five are the **final consonants**, the optional ㅁ (*m*) that
turns 가 (*ga*) into 감 (*gam*): never required, and each one changes what the
plot says. Every channel atom has the same form:

```
channel(field)
```

No exceptions. `x` and `y` are channels too: the same rule applies everywhere.

Different fields use different words for this, so you may know it by another
name. Bertin wrote the vocabulary down first, in *Sémiologie graphique*
[@bertin1967]. He separated the two dimensions of the page from six **retinal
variables**: size, value (his word for lightness), texture, color, orientation,
and shape. The eye does not read the two groups the same way. Wilkinson and
ggplot2 both say **aesthetic**, which is why ggplot2 code has `aes()` and gog
code has nothing like it. The visualization literature chose *channel* instead,
paired with the mark it modifies [@munzner2014], and that is the word gog uses.
In everyday English, a channel carries a signal. Here the column is the signal,
and color or size is the channel that carries it to the reader. Choosing an
everyday word follows gog's [Plain Names](design-laws.qmd#law-3-plain-names) law. **Encoding** is
the same idea named as an act rather than a thing, and this book keeps both
words. A channel is what you write; an encoding is what it does. [Encoding
scope](encoding-scope.qmd) is about the second.

| Atom | Visual property | Variable type |
|------|----------------|---------------|
| `x(field)` | horizontal position | any |
| `y(field)` | vertical position | any |
| `z(field)` | depth: the third position, which makes a plot 3-D | continuous |
| `color(field)` | fill / stroke color | continuous *or* categorical |
| `size(field)` | point radius | continuous |
| `shape(field)` | glyph shape | categorical |
| `pattern(field)` | fill hatch / stroke dash | categorical |
| `opacity(field)` | transparency | continuous |
| `group(field)` | splits a mark into one series per category, without encoding it | categorical |
| `label(field)` | the string a `text` mark draws | any |
| `play(field)` | cuts the plot into frames and plays them in sequence | any |

A legend is generated automatically whenever one of the five refining channels
(`color`, `size`, `shape`, `pattern`, `opacity`) is bound to a column, with no
configuration required. `x`, `y` and `z` get an axis instead, and `group` gets
neither: it splits a mark without encoding anything, so there is nothing for a
legend to decode. `label` belongs to the [`text`](marks/text.qmd) mark, and
`play` has [a chapter of its own](play.qmd).

---

## `color`

Which points belong together? A plain scatter cannot say, so the grouping has to
come from a column.

`color(field)` maps a column to color, and a legend appears automatically. On
`point`, `line`, `step`, `path`, `zone` and `surface` it takes *either* kind of
column, and the kind decides what color means. On the other marks it takes only
a categorical column.

It is spelled `color`, and only `color`. gog writes American English throughout
and accepts no second spelling, so `colour`, which ggplot2 accepts, does not
exist here. Writing it is refused, and the refusal names the fix (see [Law
3](design-laws.qmd#law-3-plain-names)).

A **categorical** column answers *"which one?"* Each category gets its own color
from the palette:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."*

A **continuous** column answers *"how much?"* Its values run along a **ramp**.
A ramp is a range of colors that a number is placed along, the way an axis is a
range of positions. This one runs light to dark, and the legend is the ramp
itself, labeled at its minimum, midpoint, and maximum:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + style(size = 6)
```

Notice the legend: color gets one continuous strip. The `size` and `opacity`
legends later in this chapter show three sampled examples instead. A legend
should show the *scale*. A range of colors fits in one strip, but there is no
way to draw a continuum of circles. The rule is the same for every channel; only
the form of the legend changes. The difference matters most for a multi-hue
ramp. Three sampled colors from `viridis`, one of the ramps [below](#ramps), say
nothing about the colors between them. Three circles of growing size do, because
a reader can imagine the sizes in between and cannot imagine the colors.

The default ramp is one hue, not a rainbow, and that is deliberate. A
**sequential** ramp, one that runs from low values to high, encodes magnitude. A
reader reads a darker color as *more*, so **lightness** is what such a ramp
varies. One hue running light to dark is read quickly, stays readable in
grayscale and under color blindness, and matches the categorical palette when
both appear in one figure. A rainbow encodes magnitude as *hue*, which has no
intrinsic order, so the reader has to consult the legend for every point. gog's
default is a single blue built on the same hue as the first categorical color,
so the two scales read as one system. [Ramps](#ramps) has the multi-hue
alternatives, and the sections after it have the diverging ramps and the ramps
you build from colors you choose.

A categorical `color` also implies grouping for line marks. Each unique value
gets its own connected line:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  line + point
```

*"Given gapminder Asia: x is year, y is life, color by country, lines and also
points."*

Where you write a channel decides which layers it reaches. `color(country)` is
written *before* any mark in the plot above, so it belongs to the plot and
colors both the lines and the points. Written *after* a mark it binds to that
mark alone:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

*"Given gapminder Asia: x is year, y is life, lines colored by country, and also
points sized by population."*

After the first mark, a channel belongs to the mark just before it, never to an
earlier one. See [Encoding scope](encoding-scope.qmd).

---

## `size`

The `color` section used `size` without saying what it does. Here is the problem
it solves. In the 2007 scatter, `x` carries GDP per capita and `y` carries life
expectancy. Population is a third number with no position left for it.

`size(field)` scales point radius linearly from 3px (min) to 12px (max). A size
legend shows the minimum, the midpoint and the maximum of the column.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  size(population)
```

Combine `color` and `size` and the result is a **bubble chart**, four columns on
one set of points:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population)
```

---

## `shape`

A journal prints your figure in black and white, and the colors you chose print
as grays that look almost the same. The grouping has to come from the glyph
instead.

`shape(field)` maps a categorical column to one of seven glyphs: circle, square,
triangle, diamond, cross, star, wye (a Y shape). Seven is the whole set. It is
also about as many shapes as a reader can recognize quickly. Map `shape` to a
column with more than seven categories and the eighth category starts again at
the circle. Two categories then share a glyph, and gog prints a message to say
so. The [Shape](marks/point.qmd#shape) section of the point chapter shows what
to do when a column has more than seven categories.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  shape(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, shape by continent."*

`color` and `shape` can be bound to the same column. Binding one column to two
channels is **redundant encoding**, and it makes the plot easier to read:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + shape(continent)
```

Either channel alone would separate the continents. Using both separates them in
two ways, so a reader who cannot see the colors still has the shapes. gog draws
one legend for the two, with each glyph in its continent's color.

---

## `pattern`

`shape` answers the black-and-white problem for points. A bar or a line has no
glyph to change, so it needs a different answer, and that answer is texture.

`pattern(field)` maps a categorical column to a **texture**, the way `shape`
maps one to a glyph. The mark's geometry decides which one: a **hatch**, a
repeated pattern of lines or dots, on the filled regions (`bar`, `box`, `area`,
`ribbon`, `zone`), and a **dash** on the strokes (`line`, `step`, `interval`,
`path`, `rule`, `edge`). Textures do not depend on color, so they work in
grayscale and for a colorblind reader. The first category is always the plain
one (solid), the rest textured:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * mean + x(species) + y(petal_length) +
  pattern(species) + y_label("Mean petal length (cm)")
```

*"Given the iris flowers: bars derived by mean, x is species, y is petal length,
pattern by species."*

A second use of the channel is redundant encoding: the same column on `color`
*and* `pattern`, so a reader separates the series by hue and by texture at once.
As with `shape`, the two share one legend:

```{r}
#| echo: false
peek(channel_sales)
```

```{r}
data(channel_sales) + bar * sum * dodge + x(quarter) + y(revenue) + color(channel) +
  pattern(channel) + title("Grouped bars, separated by hue and hatch")
```

*"Given the channel sales table: bars derived by sum and dodge, x is quarter, y is
revenue, color by channel, pattern by channel."*

On a stroke the same channel dashes each series instead. The plot below carries
two teams, separated by hue and by dash:

```{r}
#| echo: false
peek(team_trend)
```

```{r}
data(team_trend) + line + x(month) + y(sales) + color(team) + pattern(team) +
  title("Two series, by hue and dash")
```

The set of textures is deliberately small: six on a fill and five on a stroke,
and the first of each is solid. A fill covers a large area, so a reader can
distinguish more textures on it. A dash needs a long enough line to be seen, so
a reader can distinguish fewer on a stroke. Because both sets are small,
`pattern` suits a few categories rather than dozens. A seventh category on a fill, or a sixth on
a stroke, repeats a texture, and gog says so. To set one texture for a whole
layer instead of mapping a column, use [`style(pattern = )`](style.qmd).

---

## `opacity`

Sometimes a number should decide how faint a mark is drawn, so that small values
fade and large ones stay opaque.

`opacity(field)` maps a continuous column to transparency: a larger value draws
a more opaque mark.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + opacity(population) +
  title("Opacity: larger populations drawn more solidly")
```

*"Given gapminder 2007: points, x is gdp, y is life, opacity by population."*

Only China and India are drawn nearly opaque, because their populations are
several times larger than any other country's. [A ramp on a skewed
column](#a-ramp-on-a-skewed-column) shows the fix, a log scale, and it works on
`opacity` too.

Mapped opacities run from 0.15 to 0.95 rather than 0 to 1: a fully transparent
mark is invisible, which is never what a reader wants. With no `opacity`
binding, a filled mark is drawn at 0.82 so that overlapping fills show through
each other, while most strokes and a `text` are drawn fully opaque. A
[`zone`](marks/zone.qmd) sets its own: a highlight is drawn pale so the plot
behind it shows, and a zone of cells, such as the heatmaps below, is drawn fully
opaque. [`style(opacity = )`](style.qmd) overrides any of them.

Like `size`, `opacity` produces a legend showing the minimum, midpoint, and
maximum of the column.

On `line`, `opacity` cannot be *mapped*: a stroke has *one* opacity, and one
value cannot carry a column. This is the same rule that refuses a per-row
`size` on a line, not a special case for `opacity`:

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + opacity(population)
```

That one stroke still *has* an opacity, so you can *set* it for the whole line:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(opacity = 0.4)
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, with
opacity 0.4."*

To split the rows into one line per country, use [`group`](#group), as above; to
give each line its own color as well, use `color`.

That refusal looks like an inconsistency: `color` maps onto a line and `opacity`
does not. It is not one. A width, an opacity and a dash are properties of the
*whole* stroke: ask where along the line the width is 3px and the question has
no answer. A color is not like that. You can point at one place on a line and
ask what color it is, so a measure can vary along it, and `color` on `line`,
`step` and `path` accepts either kind of column:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("One line per country, colored by GDP per capita")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, color
by gdp on a log scale, with the viridis palette."*

A category on `color` splits the mark into one stroke per group. A measure
varies the color *along* each stroke and gets a ramp legend, not one swatch (a
small sample of the color) per series. One channel and one mark give
two behaviors, and the column's type decides which behavior you get. `group`
still splits the mark either way, which is why the sentence above draws one line
per country and runs the ramp along each line. See [Setting vs
mapping](style.qmd).

---

## `group`

The section above used `group` twice without saying what it does. It is the one
channel that splits a mark and encodes nothing.

`group(field)` splits the mark into one series per value in the column. On a
`line`, that is one stroke per value, and every stroke gets the same color. Use
it when the lines must be kept apart but one color is what you want.

```{r}
data(gapminder_asia) + line + x(year) + y(life) +
  group(country)
```

When you want color *and* grouping, use a categorical `color(field)` alone: it
implies grouping automatically.

---

## Multiple channels

Two positions and one refining channel carry three columns. A table usually has
a fourth you want on the page.

You have already seen two refining channels on one mark several times. Channels
compose freely, so you can bind more of them at once:

```{r}
# Two positions and three refining channels
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population) + shape(continent) +
  title("GDP and life expectancy, 2007")
```

Five channels carry four columns here. The three refining channels draw only two
legends, because `shape` repeats `color` and the two share one. Each column on a
refining channel adds a legend for the reader to check. So add a channel only
when the plot needs its column, or when it makes a column easier to read, as
`shape` does here.

---

## Palettes

Most plots so far used the colors gog chose by default. To choose your own, use
`palette()`.

`palette()` chooses the colors the `color` channel assigns. There is no separate
atom for ramps, which ggplot2 calls gradients. One atom serves both, because the
column's type decides which meaning the colors get. For a categorical column,
gog reads a set of colors as *one color per category*. For a continuous column,
it reads the same set as **stops**, the fixed colors a ramp passes through.

There are three kinds of named palette, and they say three different things: a
categorical palette gives each category its own color, a sequential ramp runs
from low values to high, and a **diverging** ramp measures distance from a
center in either direction. The type of the column decides which kinds you may
name: a categorical column takes a categorical palette, and a continuous column
takes a sequential or a diverging ramp. The sections below follow that order.

gog's default palette holds 20 colors, which is more than most tables need. When
a column has more categories than that, gog builds the extra colors from the HSL
color wheel (hue, saturation and lightness), which places every hue around a
circle. No two categories share a color.

### Categorical palettes

Most of the time the default is what you want. You change it when a reader may
be colorblind, when a journal requires particular colors, or when the marks are
large enough that the default is too strong.

A categorical palette gives one color to each category, so a reader matches a
swatch in the legend to a name. gog provides three, and the last column says
which one to pick:

| Name | Colors | Best for |
|------|--------|----------|
| `"gog"` | 20 (default) | General use |
| `"okabe"` | 8 base + 8 light variants | Colorblind accessibility |
| `"soft"` | 8, lighter | Filled marks: bars, areas, zones |

`"okabe"` is Okabe and Ito's set [@okabe2008], chosen so that its eight colors
stay distinct for readers with the common forms of color blindness. gog adds
eight lighter versions of these colors, used when a column has more than eight
categories:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("okabe")
```

The default is designed for points and lines, where a color occupies a few
pixels and must stay visible at reading distance. A fill is large, so it is
already easy to see, and a color strong enough to make a small point easy to
find is more than a bar needs. In a row of such bars, the strong color takes
attention away from the heights, which are what the bars show. `soft` is a
different set of eight colors, lighter than the default. Compare these two.
Apart from the title, the palette is the only difference:

```{r}
data(gapminder_2007) + bar * count + x(continent) + color(continent) +
  title("The default, over filled bars")
```

*"Given gapminder 2007: bars derived by count, x is continent, color by
continent."*

```{r}
data(gapminder_2007) + bar * count + x(continent) + color(continent) +
  palette("soft") + title("palette(\"soft\")")
```

The default still works on bars, and on a scatter of the same data it would be
the better choice. The difference is what the color is doing: a point has to be
easy to find, while a bar is already visible and only has to differ from its
neighbors.

### Sequential ramps {#ramps}

A continuous column has no categories to give a color to, so a palette has to
mean something else. For a continuous color column, `palette()` names a ramp
instead. The ramps below run one way, from low values to high. Lightness changes
steadily along the ramp, and that is what shows the order.

| Name | What it is | Use it when |
|------|-----------|-------------|
| `"blue"` | the default: one hue, light→dark | almost always |
| `"viridis"` | many hues, with equal color steps for equal value steps | dense data where small differences must stay visible |
| `"magma"` | viridis's family: black through purple to cream | a dark plot with strong contrast |
| `"inferno"` | the same family: black through red and orange to pale yellow | as for magma, with more red and orange |
| `"plasma"` | the same family, blue through magenta to yellow | a dark plot with strong contrast, starting at dark blue rather than black |
| `"cividis"` | the family's colorblind-optimized member | the ramp itself has to stay readable under two-color vision (dichromacy) |
| `"gray"` | no hue at all: light gray to near-black | figures that will be printed in black and white |

Those seven names do not tell you what the ramps look like. Here are six of the
seven, every one except the default `blue`, on Maunga Whau, a volcano in
Auckland. The volcano is cut into cells, each colored by its mean elevation. A
dense grid of continuous values like this is what a ramp is for. `palette()` is
a property of the plot, so six ramps means six plots, and `|` and `/` put them
on one page:

```{r}
#| echo: false
peek(maunga_whau)
```

```{r}
#| fig-height: 7
((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
    color(elevation) + palette("viridis") + title("viridis")) |
   (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("magma") + title("magma"))) /
  ((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("inferno") + title("inferno")) |
     (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
        color(elevation) + palette("plasma") + title("plasma"))) /
  ((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("cividis") + title("cividis")) |
     (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
        color(elevation) + palette("gray") + title("gray")))
```

*"Given Maunga Whau: zones derived by bin into 16 and mean, x is east, y is
north, color by elevation, with the viridis palette, beside the same with magma;
above the same pair with inferno and plasma; above the same pair with cividis
and gray."*

That is the same sentence six times with one word changed. It is written out in
full because a loop or a variable is written differently in each of the book's
four languages.

Those six plots show four things, one in each paragraph.

`gray` runs from light to dark, and the other five run from dark to light. In
the `gray` plot the summit is dark, while in the other five it is bright. Both
directions are correct. gog's own ramps (`gray` and the default `blue`) run
light to dark, because with one hue a darker color reads as a larger value, and
a photocopier keeps those lightness differences. The viridis-family ramps run
dark to light because their designers made them that way, and reversing them
would give a ramp named `magma` that no longer looks like magma. In any plot,
the legend shows which end is high: its top label names the value at that end of
the ramp.

Of the five viridis-family ramps, `magma` and `inferno` end on a near-white.
`magma`'s last stop is nearly the same color as the panel behind it, so the very
top of the scale almost disappears. You can see this in the pale, flat summits
in the `magma` and `inferno` plots. The viridis family ends on very light colors
by design. `blue` and `gray` never reach white, so their pale end stays visible
against the panel, and that is why `blue` is the default rather than one of
these.

`cividis` uses the fewest hues of the five, and that is deliberate: it uses only
the two that stay distinct under the common forms of color blindness. The other
four run through many more, so they separate neighboring values by hue as well
as by lightness. `cividis` separates them mostly by lightness, and that loss is
acceptable whenever the ramp has to stay readable for a colorblind reader.

`gray` is for the figure that will be photocopied. It is a ramp only, and there
is no categorical gray palette. Grays differ only in lightness, so readers can
distinguish only a few of them. For categories in a print figure, use
[`shape`](#shape) on points and [`pattern`](#pattern) on bars and lines, which
separate them and leave the `color` channel free. gog refuses `palette("gray")` on a categorical column, and
the refusal says why:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("gray")
```

### Diverging ramps

A sequential ramp means *more*. A diverging ramp means *away from a center, in
two directions*. A change, a residual, a correlation and an anomaly all fall on
both sides of zero. gog provides two:

| Name | What it is | Use it when |
|------|-----------|-------------|
| `"blue_red"` | blue → neutral → red | the usual choice |
| `"brown_teal"` | brown → neutral → teal | the sign of the value is what matters, and readers may be colorblind |

The name reads low to high, so `blue_red` paints the smallest values blue and
the largest red. Both pass through the same gray at the center of the scale.

The palette does not decide which value gets the neutral color. That value is
the middle of the scale's domain, which `limits` sets. Without `limits`, the
middle falls on the data's own midpoint. Here `delta` runs from -32 to 120, so
the midpoint is 44. That value is painted gray, and zero is painted blue.

```{r}
#| echo: false
peek(cashflow)
```

```{r}
data(cashflow) + point + x(step) + y(delta) +
  color(delta) + palette("blue_red") + style(size = 9) +
  title("44 is painted gray, not 0")
```

State a symmetric domain and zero becomes the center of the scale, so zero is
the value painted gray. Read the "Closing" point, which is exactly zero:

```{r}
data(cashflow) + point + x(step) + y(delta) +
  color(delta, limits = c(-120, 120)) + palette("blue_red") + style(size = 9) +
  title("Symmetric limits paint zero gray")
```

*"Given the cashflow table: points, x is step, y is delta, color by delta from
-120 to 120, with the blue red palette."*

Two things follow from centering the ramp with `limits`. The ramp stays
**linear**: equal distance in color is equal distance in the data on both sides
of the center. Some scales set a midpoint and then rescale each side to the
data's own ends, and a scale built that way is no longer linear. And `limits`
[removes rows outside it](scales.qmd), so the domain has to be wide enough to
hold the data: `c(-50, 50)` here would drop the opening balance, and gog would
say so.

Symmetric limits have one drawback. If a column runs from -2 to 100, limits of
-100 to 100 leave almost all of the blue half unused, because the data reaches
only a little way below zero. Accept that when the sign is what matters, since
zero has to be painted gray. When the sign does not matter, a sequential ramp
such as `blue` uses its whole range.

### Stops of your own

Sometimes you need a ramp between two particular colors, and no named ramp runs
between them. Pass those colors to `palette()`: they become the stops, and gog
fills in the colors between them:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + palette(c("white", "navy")) + style(size = 6) +
  title("Custom stops")
```

*"Given gapminder 2007: points, x is gdp, y is life, color by life, with a
palette of white and navy."*

gog refuses a palette that does not match the column, rather than guessing what
you meant. A categorical palette cannot serve a continuous column:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + palette("okabe")
```

### A ramp on a skewed column

No ramp reads well on a skewed column, and changing the palette does not fix it.
A linear ramp spreads its colors evenly across the domain, so when most rows sit
in a narrow part of it they all receive almost the same color. When you color
by `population`, nearly every country falls at the light end:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(population) +
  x_label("GDP per capita")
```

The two dark points are China and India, whose populations are several times
larger than any other country's. The cause is the data, not the palette, so the
fix is a [log scale](scales.qmd) on the channel, and on `x` as well, since GDP
per capita is skewed the same way:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log") +
  x_label("GDP per capita")
```

On the log scale the colors spread along the whole ramp, so countries of
different sizes get different colors.

`color`, `size` and `opacity` all take `scale =`, because all three answer "how
far along?". gog refuses it on `shape`, `pattern` and `group`: they answer "which
one?".

### Custom colors

Often the colors are decided before you begin: a journal asks for them, or an
earlier figure already used them.

Pass the colors yourself: hex values, CSS color names such as `firebrick`, or a
mix of the two:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  palette(c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00"))
```

The same call takes CSS color names, which are easier to recognize when you read
the code later:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  palette(c("firebrick", "steelblue", "seagreen", "rebeccapurple", "darkorange"))
```

gog refuses a color it cannot recognize, and offers the nearest match. Otherwise
the misspelled name would go into the plot, and a browser draws an unknown color
as black:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette(c("firebrick", "stelblue"))
```

Colors are assigned in **first-appearance order**: the first unique value gets
the first palette color, the second gets the second, and so on. If the column
[declares its order](data.qmd), that order is used instead. Either way, the same
rows in the same order always get the same colors.

One mistake to know about: a single color is not a palette. `palette()` reads a
single word as a palette name, and `"red"` is not one. What `palette("red")`
attempts is one constant color, which is [`style(color = "red")`](style.qmd):

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("red")
```

### Colors by name

Colors passed without names are assigned in first-appearance order, so a
category's color depends on where it falls in that order. Sorting the rows can
change a category's color, and so can another table that lists the categories
in a different order. Two plots placed side by side in a report then disagree
about which color a continent is. To assign a color to a category, write the
category's name beside the color:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  palette(c(Africa = "seagreen", Americas = "goldenrod", Asia = "tomato",
            Europe = "steelblue", Oceania = "orchid"))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, with a
palette naming Africa seagreen, Americas goldenrod, Asia tomato, Europe
steelblue and Oceania orchid."*

Each continent now takes the color written beside its name, whatever order the
rows arrive in. The legend still lists the continents in first-appearance order,
with Asia first. Once colors are named, the names and the categories must match.
A name that is not a category would color nothing. A category without a name
would get a color nobody chose. gog refuses both. So a palette that names all
five continents is refused on a table that holds only three, unless the column's
declared order includes all five. Usually a mismatch is a misspelled name, and
the message suggests the category you probably meant:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  palette(c(Africa = "seagreen", Americas = "goldenrod", Asai = "tomato",
            Europe = "steelblue", Oceania = "orchid"))
```

---

## Legend placement

Every plot above that maps a refining channel drew a legend automatically, and
gog also decides where the legend goes: to the right of the plot area. The plot
area shrinks by itself to make room, so there is no margin to set and no
position to choose. A legend too tall for the plot is left out rather than drawn
past the edge. gog says so, and names `theme(height = )` as the way to give it
room.

### Leaving a legend out

A legend decodes a channel, and sometimes an axis already decodes the same
channel. Here `color` and `x` map the same column, so the axis already names the
continent under every bar. A legend would repeat those names. Write
`legend = FALSE` on the channel and the legend is left out:

```{r}
data(gapminder_2007) + bar * count + x(continent) +
  color(continent, legend = FALSE)
```

*"Given gapminder 2007: bars derived by count, x is continent, color by
continent, without a legend."*

The bars keep their colors, because the channel still maps the column. Only the
legend is gone, and the plot area takes the legend's space. `legend` belongs to
the five refining channels, the ones that draw a legend. `x`, `y` and `z` are
read from their axes, and `group`, `label` and `play` draw no legend. gog
refuses `legend` on all six:

```{r}
#| error: true
data(gapminder_2007) + bar * count + x(continent, legend = FALSE)
```

## In short

A channel takes a column and never a value, so gog refuses `color("red")` and
points you to `style()`. The five refining channels get a legend unless
`legend = FALSE` leaves it out; `x`, `y` and `z` get an axis. `group` and
`label` get neither, and `play` gets a strip that names the frame. Which
channels a mark accepts follows from its geometry. `palette()` chooses the
colors: a categorical palette for a categorical column, and a ramp for a
continuous one.

## What comes next

Every channel here reads a column, so every plot answers a question about the
data. [Setting vs mapping](style.qmd) covers the opposite case: a value you choose
yourself, which reads no column and gets no legend.
