# Channels and legends

```{r}
#| include: false
source("R/setup.R")
```

Which visual property should carry which column? Channels animate a mark by
binding a data column to a visual property, and they do not all play the same
part. `x`, `y` and `z` are the **vowels**: without one a mark is silent. The
five below that earn a legend are the **final consonants**, the optional ㅁ
(*m*) that turns 가 (*ga*) into 감 (*gam*): never required, and each one changes
what the plot says. Every channel atom follows the same pattern:

```
channel(field)
```

No exceptions.  `x` and `y` are channels too: the same rule applies everywhere.

The word for this is not settled outside gog, so you may know it as something
else. Bertin set the vocabulary down first, in *Sémiologie graphique*
[@bertin1967], separating the plane's two dimensions from six **retinal
variables** (size, value, texture, color, orientation, shape) on the grounds
that the eye does not read them the same way. Wilkinson and ggplot2 both say
**aesthetic**, which is why you write `aes()` there and nothing like it here.
The visualization literature settled instead on **channel**, paired with the
mark it modifies [@munzner2014], and that is the word gog takes. A channel is
something a signal travels down, which is what this is, and Plain Names asks
for the everyday word that says the job. **Encoding**
is the same idea named as an act rather than a thing, and both uses survive here:
a channel is what you write, an encoding is what it does, and [Encoding
scope](encoding-scope.qmd) is about the second.

| Atom | Visual property | Variable type |
|------|----------------|---------------|
| `x(field)` | horizontal position | any |
| `y(field)` | vertical position | any |
| `z(field)` | depth: the third position, which makes a plot 3-D | continuous |
| `color(field)` | fill / stroke color | categorical *or* continuous |
| `size(field)` | point radius | continuous |
| `shape(field)` | glyph shape | categorical |
| `pattern(field)` | fill hatch / stroke dash | categorical |
| `opacity(field)` | transparency | continuous |
| `group(field)` | splits a mark into one series per category, without encoding it | categorical |
| `label(field)` | the string a `text` mark draws | any |
| `play(field)` | cuts the plot into frames and plays them in sequence | any |

A legend is generated automatically whenever one of the five refining channels
is bound to a column, with no configuration required. `x` and `y` earn an axis
instead, and `group` earns neither: it splits a mark without encoding anything,
so there is nothing for a key to decode.

---

## `color`

`color(field)` maps a column to color, and a legend appears automatically.
It takes *either* kind of column, and the kind decides what color means.

It is spelled `color`, and only `color`. gog writes American English throughout
and accepts no second spelling, so there is no `colour` here as there is in
ggplot2; writing one earns a refusal that names the fix (see
[Law 3](design-laws.qmd)).

A **text** column asks *"which one?"* Each category gets its own hue from the
palette:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent)
```

A **numeric** column asks *"how much?"* Its values run along a **ramp**. A ramp
is a range of colors that a number is placed along, the way an axis is a range
of positions. This one runs light to dark, and the legend is the ramp itself,
labeled at its minimum, midpoint, and maximum:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + style(size = 6)
```

Notice the legend: color gets one continuous strip while `size` and `opacity`
get three sampled examples. A legend should show the *scale*, and color is
the only continuous channel whose whole range fits in a fixed space; there is
no way to draw a continuum of circles. Same rule, different shapes. The
difference matters most for a multi-hue ramp: given three swatches of
`viridis` a reader cannot guess what lies between them, whereas three circles
obviously interpolate.

The default ramp is one hue, not a rainbow, and that too is deliberate. A
sequential ramp encodes magnitude, and **lightness** is what a reader
perceives as *more*: one hue running light→dark is readable at a glance,
survives grayscale printing and color blindness, and does not compete with
the categorical palette when both appear in one figure. A rainbow encodes
magnitude as *hue*, which has no intrinsic order, so the reader has to consult
the legend for every point. gog's default is a single blue built on the same
hue as the first categorical color, so the two scales read as one system.
See [Ramps](#ramps) below for the multi-hue alternatives, the diverging ramps,
and custom stops.

`color` also implies grouping for line marks. Each unique value gets its own
connected line:

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  line + point
```

Position decides which layers a channel reaches. `color(country)` is written
*before* any mark above, so it belongs to the plot and colors both the
lines and the points. Written *after* a mark it binds to that mark alone:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

Channels bind forward, never backward. See
[Encoding scope](encoding-scope.qmd).

---

## `size`

`size(field)` scales point radius linearly from 3 px (min) to 12 px (max).
A size legend shows three representative circles.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  size(population)
```

Combine `color` and `size` for a classic bubble chart:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population)
```

---

## `shape`

`shape(field)` maps a categorical column to one of five glyphs:
circle, square, triangle, diamond, cross.
Useful for distinguishing groups in black-and-white prints.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  shape(continent)
```

`color` and `shape` can be bound to the same field. This double-encodes
the grouping, improving accessibility:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + shape(continent)
```

---

## `pattern`

`pattern(field)` is `shape`'s twin one geometry class over: it maps a categorical
column to a **texture** rather than a glyph, realized by the mark's geometry: a
**hatch** on the fills (`bar`, `box`, `area`, `ribbon`), a **dash** on the strokes
(`line`, `step`, `interval`). Textures read without hue, so a plot mapped by
`pattern` stays legible in grayscale or with color-blindness. The first category is
always the plain one (solid), the rest textured:

```{r}
data(iris_flowers) + bar * mean + x(species) + y(petal_length) +
  pattern(species) + y_label("Mean petal length (cm)")
```

The point of the channel is **redundant encoding**: the same column on `color`
*and* `pattern`, so a reader separates the series by hue *and* by texture at once (the
accessibility best practice). gog draws these as **one merged legend**, not two:

```{r}
sales <- data.frame(
  quarter = rep(c("Q1", "Q2", "Q3"), each = 2),
  channel = rep(c("Retail", "Online"), 3),
  revenue = c(4, 3, 5, 6, 4, 8)
)
data(sales) + bar * sum * dodge + x(quarter) + y(revenue) + color(channel) +
  pattern(channel) + title("Grouped bars, separated by hue and hatch")
```

On a stroke the same channel dashes each series instead: the color-free line
chart:

```{r}
trend <- data.frame(
  month = rep(1:6, 2),
  team  = rep(c("North", "South"), each = 6),
  sales = c(2, 3, 4, 5, 6, 7,  5, 4, 4, 3, 2, 2)
)
data(trend) + line + x(month) + y(sales) + color(team) + pattern(team) +
  title("Two series, by hue and dash")
```

The texture palette is deliberately small (five hatches on a fill, three dashes on
a stroke), so `pattern` suits a handful of categories, not dozens; past the palette
size the textures repeat. To fix one texture for a whole layer rather than map a
column, that is the [`style(pattern = )`](style.qmd) setting.

---

## `opacity`

`opacity(field)` maps a continuous column to transparency.  Useful when marks
overlap and you want the crowded regions to read as denser.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + opacity(population) +
  title("Opacity: larger populations drawn more solidly")
```

Mapped values run from 0.15 to 0.95 rather than 0 to 1: a fully transparent mark
is invisible, which is never what a reader wants.  Marks with no `opacity`
binding are drawn at 0.82, unless [`style(opacity = )`](style.qmd) sets one.

Like `size`, `opacity` produces a legend showing the minimum, midpoint, and
maximum of the column.

On `line`, `opacity` cannot be *mapped*: a stroke has *one* opacity, and one
value cannot carry a column. This is the same rule that rules out per-row `size`
on a line, not a special case for `opacity`:

```
gog: `opacity` cannot be bound to `line` — a line has no opacity feature.
     Remove `opacity(dist)`, or use a mark that has one.
```

That one stroke still *has* an opacity, though, so you can *set* it for the
whole line:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(opacity = 0.4)
```

To distinguish lines *by data*, use `color` or `group`.

`color` is the one that does not follow `size` and `opacity` here, and the
difference is worth being exact about, because it looks like an exception and is
not. A width, an opacity and a dash are properties of the **whole stroke**: ask
where along the line the width is 3px and the question has no answer. A color is
not like that. You can point at one place on a line and ask what color it is,
so a measure can vary along it, and `color` on a stroke mark accepts either kind
of column:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("One line per country, colored by income")
```

A category on `color` splits the mark into one stroke per group, as it always
has. A measure varies the color *along* each stroke and earns the gradient key
instead of a swatch per series. Same channel, same mark, two questions, and the
column decides which one you asked. `group` still does the splitting either way,
which is why the sentence above draws one line per country *and* ramps each.
See [Setting vs mapping](style.qmd).

---

## `group`

`group(field)` connects line segments within the same group without
adding a color distinction.  Use it when you want a single color for
all lines but still need them separated by category.

```{r}
data(gapminder_asia) + line + x(year) + y(life) +
  group(country)
```

When you want color *and* grouping, use `color(field)` alone:
it implies grouping automatically.

---

## Multiple channels

Channels compose freely.  Every combination that makes visual sense is valid:

```{r}
# Color + size + shape on the same plot
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population) + shape(continent) +
  title("GDP vs Life Expectancy, 2007")
```

---

## Palettes

`palette()` chooses the colors the `color` channel hands out. There is no
separate atom for ramps, which ggplot2 calls gradients. The same word covers
both, because whether a color vector means *one color per category* or *stops to
interpolate between* follows from the column `color` is bound to.

There are three kinds of named palette, and they say three different things: a
**categorical** palette gives each category its own color, a **sequential** ramp
runs from little to much, and a **diverging** ramp measures distance from a
center in either direction. Which kinds a name may meet is decided by the
column, so the sections below are in that order.

gog uses a 20-color default palette so you won't run out of colors for
typical datasets. When categories exceed the palette size, additional
colors are generated automatically via the HSL color wheel, guaranteed
unique for any number of categories.

### Categorical palettes

| Name | Colors | Best for |
|------|--------|----------|
| `"gog"` | 20 (default) | General use |
| `"okabe"` | 8 base + 8 light variants | Colorblind accessibility |
| `"soft"` | 8, muted | Paint that covers ground: bars, areas, zones |

`"okabe"` is Okabe and Ito's set [@okabe2008], chosen so that its eight colors
stay distinct for readers with the common forms of color blindness:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("okabe")
```

The default is tuned for points and lines, where a color occupies a few pixels
and has to carry from across the page. A fill has the opposite problem. It
already has the reader's attention by covering ground, so the saturation that
made a 4.5px dot findable is more weight than a bar needs. A row of them
competes with the numbers they are there to show. `soft` is the same set of
jobs done at lower saturation. Compare these two, which differ only in the
palette:

```{r}
data(gapminder_2007) + bar * count + x(continent) + color(continent) +
  title("The default, over filled bars")
```

```{r}
data(gapminder_2007) + bar * count + x(continent) + color(continent) +
  palette("soft") + title("palette(\"soft\")")
```

Nothing is wrong with the first one, and on a scatter of the same data it is
the better choice. The difference is what the ink is doing: a point asks to be
found, a bar is already found and only has to be distinct from its
neighbors.

### Ramps {#ramps}

For a **numeric** color column, `palette()` names a ramp instead. These run one
way, from little to much, and lightness is what carries the ordering.

| Name | What it is | Use it when |
|------|-----------|-------------|
| `"blue"` | the default: one hue, light→dark | almost always |
| `"viridis"` | perceptually uniform, multi-hue | dense data where fine levels must stay distinct |
| `"magma"` | viridis's family: black through purple to cream | a dark, high-contrast field |
| `"inferno"` | the same family, hotter | as above, warmer |
| `"plasma"` | the same family, blue through magenta to yellow | as above, cooler at the low end |
| `"cividis"` | the family's colorblind-optimized member | the ramp itself has to survive dichromacy |
| `"gray"` | one hue, no hue | figures that will be printed in black and white |

A table is a poor way to choose a ramp. Here are six of the seven on one field,
the Maungawhau crater cut into cells and colored by mean height, which is the
kind of dense continuous surface a ramp is picked for. `palette()` is a
property of the plot, so six ramps means six plots, and `|` and `/` put them on
one page:

```{r}
#| fig-height: 7
((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
    color(elevation) + palette("viridis") + title("viridis")) |
   (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("magma") + title("magma"))) /
  ((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("inferno") + title("inferno")) |
     (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
        color(elevation) + palette("plasma") + title("plasma"))) /
  ((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("cividis") + title("cividis")) |
     (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
        color(elevation) + palette("gray") + title("gray")))
```

That is the same sentence six times with one word changed, written out rather
than wrapped in a helper, because a helper would be an R function and this book
draws every sentence in four languages.

Three things are worth reading off that page.

`gray` runs the other way. Its crater is dark at the summit while the other
five are bright there. That is not a mistake in either: gog's own ramps (`gray`
and the default `blue`) run light to dark, because for a single hue more ink
reads as more, which is also how the figure survives a photocopier. The
viridis family runs dark to light, because that is what those ramps are, and
re-pitching them would leave you with something called `magma` that is not
magma. The legend is what settles it in any given plot, and it is the reason a
mapped channel always gets one: read the top label and you know which end is
which.

The four imported ramps end on a near-white. `magma`'s last stop holds
barely any contrast against the panel, so the very top of the scale nearly
disappears, which you can see in the pale flat summits above. That is the price
of perceptual uniformity rather than an oversight, and it is why `blue` is the
default rather than one of these. `blue` and `gray` are pitched to keep their
pale end on the page.

**`cividis` is the quiet one**, and deliberately: it spends its range on the
two hues that survive the common forms of color blindness, so it has less
contrast to give than `inferno` does. That is the trade, and it is the right
one whenever the ramp itself has to be readable rather than merely pretty.

`gray` is for the figure that has to survive photocopying. It is a ramp only:
there is no categorical gray palette, because grays stop being tellable apart
after about four and the grammar already has a better answer for categories in
a print figure. That answer is [`pattern`](style.qmd), which separates them
without spending color at all, and asking for `palette("gray")` on a text
column says so:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("gray")
```

### Diverging ramps

A sequential ramp means *more*. A **diverging** ramp means *away from a center,
in two directions*, which is what a change, a residual, a correlation and an
anomaly all are. Two ship:

| Name | What it is | Use it when |
|------|-----------|-------------|
| `"blue_red"` | blue → neutral → red | the usual choice |
| `"brown_teal"` | brown → neutral → teal | the sign of the value is the whole point, and readers may be colorblind |

The name reads low to high, so `blue_red` paints the smallest values blue and
the largest red. Both turn through the same gray, and that gray is the color of
"nothing here".

Which value gets the neutral is not a property of the palette. It is the
middle of the scale's domain, so **`limits` is what puts zero in the middle**,
and without it the middle falls wherever the data's own midpoint happens to
land. Here the cash flow runs from -32 to 120, so the midpoint is 44: that is
what gets painted gray, and zero comes out blue.

```{r}
data(cashflow) + point + x(step) + y(delta) +
  color(delta) + palette("blue_red") + style(size = 9) +
  title("The neutral lands at 44, not at 0")
```

State a symmetric domain and it lands where it means something. Read the
"Closing" point, which is exactly zero:

```{r}
data(cashflow) + point + x(step) + y(delta) +
  color(delta, limits = c(-120, 120)) + palette("blue_red") + style(size = 9) +
  title("limits = c(-120, 120) puts zero on the neutral")
```

Two things follow from doing it this way. The ramp stays **linear**: equal
distance in color is equal distance in the data on both arms, which is not
true of a scale that pins a midpoint and then stretches each side to reach the
data's own ends. And `limits` [removes rows outside it](scales.qmd), so the
domain has to be wide enough to hold the data: `c(-50, 50)` here would drop
the opening balance and say so. The cost is the other side of the same coin,
and it is worth knowing before you use this. A signed column that is
lopsided, -2 to 100 say, spends almost none of its blue arm, because a
symmetric domain is mostly empty on the short side.

A vector of colors becomes the **stops**, interpolated between:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + palette(c("white", "navy")) + style(size = 6) +
  title("Custom stops")
```

Asking for a palette of the wrong kind is refused rather than guessed at. A
categorical palette cannot serve a numeric column:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + palette("okabe")
```

No ramp survives a skewed column, and swapping palettes will not help. A
linear ramp spends most of its range on the values you have most of: color
`population` and nearly every country lands at the light end, because two
countries are an order of magnitude larger than the rest. That is the data,
not the palette, so the fix is a [log scale](scales.qmd) on the channel:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log") +
  x_label("GDP per capita")
```

`color`, `size` and `opacity` all take `scale =`, because all three answer "how
far along?". `shape` and `group` do not: they answer "which one?".

### Custom colors

Pass a character vector (hex values, CSS color names, or a mix):

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  palette(c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00"))
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  palette(c("firebrick", "steelblue", "seagreen", "rebeccapurple", "darkorange"))
```

A color gog cannot recognize is refused with the nearest match offered, rather
than handed to the renderer to be painted black:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette(c("firebrick", "stelblue"))
```

Colors are assigned in **first-appearance order** in the data (the first
unique value gets the first palette color, the second unique value gets the
second, and so on), so color assignment is deterministic and predictable.
(Unless the column [declares its order](data.qmd), whose levels win.)

One near-miss to know about: a single color is not a palette. `palette()`
takes a palette *name* or a *vector* of colors, and `palette("red")` is
neither; it is an attempt to set one constant color, which is
[`style(color = "red")`](style.qmd):

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("red")
```

---

## Legend placement

Legends are always placed to the right of the plot area.
The plot area shrinks automatically to make room, no manual margin setting.
Legend suppression (`show_legend = FALSE`) and manual legend position
(`legend_position`) are not built yet.
