# Channels and legends

```{r}
#| include: false
source("R/setup.R")
```

Which visual property should carry which column? A channel binds a data
column to a visual property of a mark, and the channels do not all do the
same work. `x`, `y` and `z` are the **vowels**: without one a mark does not
render. Five other channels earn a legend: `color`, `size`, `shape`, `pattern`
and `opacity`. Those five are the **final consonants**, the optional ㅁ (*m*)
that turns 가 (*ga*) into 감 (*gam*): never required, and each one changes what
the plot says. Every channel atom follows the same pattern:

```
channel(field)
```

No exceptions. `x` and `y` are channels too: the same rule applies everywhere.

The word for this is not settled outside gog, so you may know it as something
else. Bertin wrote the vocabulary down first, in *Sémiologie graphique*
[@bertin1967]. He separated the plane's two dimensions from six **retinal
variables**: size, value, texture, color, orientation, and shape. The eye does
not read the two groups the same way. Wilkinson and ggplot2 both say
**aesthetic**, which is why you write `aes()` there and nothing like it here.
The visualization literature settled instead on **channel**, paired with the
mark it modifies [@munzner2014], and that is the word gog takes. A channel is a
path a signal travels down. Here the signal is a column, and color or size is
the path it takes. gog takes the everyday word for the job, as its [Plain
Names](design-laws.qmd) law asks.
**Encoding** is the same idea named as an act rather than a thing, and this book
keeps both words. A channel is what you write; an encoding is what it does.
[Encoding scope](encoding-scope.qmd) is about the second.

| Atom | Visual property | Variable type |
|------|----------------|---------------|
| `x(field)` | horizontal position | any |
| `y(field)` | vertical position | any |
| `z(field)` | depth: the third position, which makes a plot 3-D | continuous |
| `color(field)` | fill / stroke color | continuous *or* categorical |
| `size(field)` | point radius | continuous |
| `shape(field)` | glyph shape | categorical |
| `pattern(field)` | fill hatch / stroke dash | categorical |
| `opacity(field)` | transparency | continuous |
| `group(field)` | splits a mark into one series per category, without encoding it | categorical |
| `label(field)` | the string a `text` mark draws | any |
| `play(field)` | cuts the plot into frames and plays them in sequence | any |

A legend is generated automatically whenever one of the five refining
channels (`color`, `size`, `shape`, `pattern`, `opacity`) is bound to a
column, with no configuration required. `x`, `y` and `z` earn an axis
instead, and `group` earns neither: it splits a mark without encoding anything,
so there is nothing for a key to decode. `label` belongs to the
[`text`](marks/text.qmd) mark, and `play` has [a chapter of its own](play.qmd).

---

## `color`

Which points belong together? A plain scatter cannot say, so the grouping has to
come from a column.

`color(field)` maps a column to color, and a legend appears automatically.
It takes *either* kind of column, and the kind decides what color means.

It is spelled `color`, and only `color`. gog writes American English throughout
and accepts no second spelling, so there is no `colour` here as there is in
ggplot2. Writing one is refused, and the refusal names the fix (see
[Law 3](design-laws.qmd)).

A **categorical** column answers *"which one?"* Each category gets its own hue
from the palette:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."*

A **continuous** column answers *"how much?"* Its values run along a **ramp**.
A ramp is a range of colors that a number is placed along, the way an axis is a
range of positions. This one runs light to dark, and the legend is the ramp
itself, labeled at its minimum, midpoint, and maximum:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + style(size = 6)
```

Notice the legend: color gets one continuous strip. The `size` and `opacity`
legends later in this chapter show three sampled examples instead. A legend
should show the *scale*, and color is the
only continuous channel whose whole range fits in a fixed space; there is no way
to draw a continuum of circles. The rule is the same for every channel; only the
shape of the legend changes. The difference matters most for a multi-hue ramp.
Three sampled colors from `viridis` say nothing about the colors between them,
while three circles of growing size do.

The default ramp is one hue, not a rainbow, and that is deliberate. A sequential
ramp encodes magnitude. A reader reads **lightness** as *more*, so lightness is
what such a ramp varies. One hue running light to dark is read quickly, stays
readable in grayscale and under color blindness, and stays distinct from
the categorical palette when both appear in one figure. A rainbow encodes
magnitude as *hue*, which has no intrinsic order, so the reader has to consult
the legend for every point. gog's default is a single blue built on the same
hue as the first categorical color, so the two scales read as one system.
[Ramps](#ramps) has the multi-hue alternatives, and the sections after it have
the diverging ramps and the stops you write yourself.

`color` also implies grouping for line marks. Each unique value gets its own
connected line:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  line + point
```

*"Given gapminder Asia: x is year, y is life, color by country, lines and also
points."*

Where you write a channel decides which layers it reaches. `color(country)` is
written *before* any mark in the plot above, so it belongs to the plot and
colors both the lines and the points. Written *after* a mark it binds to that
mark alone:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

*"Given gapminder Asia: x is year, y is life, lines, color by country, and also
points, size by population."*

Channels bind forward, never backward. See
[Encoding scope](encoding-scope.qmd).

---

## `size`

The plot above already used `size` without saying what it does. Go back to the
2007 scatter: `x` carries income, `y` carries life expectancy, and population is
a third number with no position left for it.

`size(field)` scales point radius linearly from 3px (min) to 12px (max). A size
legend shows the minimum, the midpoint and the maximum of the column.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  size(population)
```

Combine `color` and `size` and the result is a **bubble chart**, four columns on
one set of points:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population)
```

---

## `shape`

A journal prints your figure in black and white, and every color you chose
arrives as the same gray. The grouping has to come from the symbol instead.

`shape(field)` maps a categorical column to one of seven glyphs: circle, square,
triangle, diamond, cross, star, wye. Seven is the whole set. It is also about as
many shapes as a reader can recognize at a glance. Map `shape` to a column with
more than seven categories and the eighth category starts again at the circle.
Two categories then share a glyph, and gog says so rather than leaving it to be
noticed. The [Shape](marks/point.qmd#shape) section of the point chapter shows
what to do when a column has more than seven categories.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  shape(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, shape by continent."*

`color` and `shape` can be bound to the same column. Saying one column twice is
**redundant encoding**, and it makes the plot easier to read:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + shape(continent)
```

Either channel alone would separate the continents. Together they separate them
twice, so a reader who cannot use one still has the other.

---

## `pattern`

`shape` answers the black-and-white problem for points. A bar or a line has no
glyph to change, so it needs a different answer, and that answer is texture.

`pattern(field)` maps a categorical column to a **texture**, the way `shape`
maps one to a glyph. The mark's geometry decides which one: a **hatch** on the
fills (`bar`, `box`, `area`, `ribbon`, `zone`), a **dash** on the strokes
(`line`, `step`, `interval`, `path`, `rule`). Textures use no hue at all, so
they read in grayscale and for a color-blind reader alike. The first category is
always the plain one (solid), the rest textured:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * mean + x(species) + y(petal_length) +
  pattern(species) + y_label("Mean petal length (cm)")
```

*"Given the iris flowers: bars derived by mean, x is species, y is petal length,
pattern by species."*

A second use of the channel is redundant encoding: the same column on `color`
*and* `pattern`, so a reader separates the series by hue and by texture at once.
gog draws these as **one merged legend**, not two:

```{r}
#| echo: false
peek(channel_sales)
```

```{r}
data(channel_sales) + bar * sum * dodge + x(quarter) + y(revenue) + color(channel) +
  pattern(channel) + title("Grouped bars, separated by hue and hatch")
```

*"Given the channel sales table: bars derived by sum and dodge, x is quarter, y is
revenue, color by channel, pattern by channel."*

On a stroke the same channel dashes each series instead. The plot below carries
two teams, separated by hue and by dash:

```{r}
#| echo: false
peek(team_trend)
```

```{r}
data(team_trend) + line + x(month) + y(sales) + color(team) + pattern(team) +
  title("Two series, by hue and dash")
```

The texture palette is deliberately small: five hatches beside solid on a fill,
and four dashes beside solid on a stroke. A fill covers a large area, so it can
carry five different hatches. A dash needs enough uninterrupted line to be seen,
so a stroke carries four. That is why the two counts differ. So `pattern` suits
a few categories rather than dozens: past six on a fill, or five on a stroke,
the textures repeat and gog says so. To fix one texture for a whole layer
instead of mapping a column, use [`style(pattern = )`](style.qmd).

---

## `opacity`

Sometimes a number should decide how faint a mark is drawn, so that small values
fade and large ones stay solid.

`opacity(field)` maps a continuous column to transparency. Use it when marks
overlap, so the crowded regions look darker than the empty ones.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + opacity(population) +
  title("Opacity: larger populations drawn more solidly")
```

*"Given gapminder 2007: points, x is gdp, y is life, opacity by population."*

Mapped values run from 0.15 to 0.95 rather than 0 to 1: a fully transparent mark
is invisible, which is never what a reader wants. With no `opacity` binding, a
filled mark is drawn at 0.82 so that overlapping fills show through each other,
while a stroke and a `text` are drawn solid. A [`zone`](marks/zone.qmd) sets its
own, because a highlight has to let the plot behind it show.
[`style(opacity = )`](style.qmd) overrides any of them.

Like `size`, `opacity` produces a legend showing the minimum, midpoint, and
maximum of the column.

On `line`, `opacity` cannot be *mapped*: a stroke has *one* opacity, and one
value cannot carry a column. This is the same rule that refuses a per-row
`size` on a line, not a special case for `opacity`:

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + opacity(population)
```

That one stroke still *has* an opacity, so you can *set* it for the whole line:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(opacity = 0.4)
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, with
opacity 0.4."*

To distinguish lines *by data*, use `color` or [`group`](#group).

That refusal raises a fair question. `color` maps onto a line and `opacity` does
not, so the two look like one rule with an exception in it. They are not, and
the reason is worth stating exactly. A width, an
opacity and a dash are properties of the **whole stroke**: ask where along the
line the width is 3px and the question has no answer. A color is not like that.
You can point at one place on a line and ask what color it is, so a measure can
vary along it, and `color` on a stroke mark accepts either kind of column:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  color(gdp, scale = "log") + palette("viridis") +
  y_label("Life expectancy") + title("One line per country, colored by income")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, color
by gdp on a log scale, with the viridis palette."*

A category on `color` splits the mark into one stroke per group. A measure
varies the color *along* each stroke and earns the ramp legend instead of a
swatch per series. One channel and one mark give two behaviors, and the column's
type decides which behavior you get. `group` still splits the mark either way,
which is why the sentence above draws one line per country and runs the ramp
along each line. See [Setting vs mapping](style.qmd).

---

## `group`

The section above used `group` twice without saying what it does. It is the one
channel that splits a mark and encodes nothing.

`group(field)` draws a separate line for each value in the column, and gives
every line the same color. Use it when the lines must be kept apart but one
color is what you want.

```{r}
data(gapminder_asia) + line + x(year) + y(life) +
  group(country)
```

When you want color *and* grouping, use `color(field)` alone:
it implies grouping automatically.

---

## Multiple channels

Two positions and one refining channel carry three columns. A table usually has
a fourth you want on the page.

You have seen two at a time twice already. Channels compose freely, so you can
bind several of them on one mark:

```{r}
# Two positions and three refining channels
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + size(population) + shape(continent) +
  title("GDP and life expectancy, 2007")
```

---

## Palettes

Every plot so far used the colors gog chose by default. Those colors are yours
to choose instead, and `palette()` is where you choose them.

`palette()` chooses the colors the `color` channel assigns. There is no separate
atom for ramps, which ggplot2 calls gradients. One atom serves both, because the
column's kind decides which reading you get. For a categorical column, gog reads
a set of colors as *one color per category*. For a continuous column, it reads
the same set as **stops**, the fixed colors a ramp passes through.

There are three kinds of named palette, and they say three different things: a
**categorical** palette gives each category its own color, a **sequential** ramp
runs from low values to high, and a **diverging** ramp measures distance from a
center in either direction. The type of the column decides which kinds you may
name: a categorical column takes the first, a continuous column takes either of
the other two. The sections below follow that order.

gog's default palette holds 20 colors, which is more than most tables need. When
a column has more categories than that, gog builds the extra colors from the
HSL color wheel, and no two categories share a color.

### Categorical palettes

A categorical palette gives one color to each category, so a reader matches a
swatch in the legend to a name. gog ships three, and the last column says which
one to pick:

| Name | Colors | Best for |
|------|--------|----------|
| `"gog"` | 20 (default) | General use |
| `"okabe"` | 8 base + 8 light variants | Colorblind accessibility |
| `"soft"` | 8, muted | Filled marks: bars, areas, zones |

`"okabe"` is Okabe and Ito's set [@okabe2008], chosen so that its eight colors
stay distinct for readers with the common forms of color blindness. The eight
light variants beside them in the table are gog's, for a column with more than
eight categories:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("okabe")
```

The default is tuned for points and lines, where a color occupies a few pixels
and must stay visible from across the page. A fill is large, so it is already
easy to see, and the saturation that makes a small point findable is too strong
on a bar. A row of saturated bars draws more attention than the numbers the bars
report. `soft` is a different set of eight, chosen at lower saturation for the
same eight jobs. Compare these two. The only difference that changes the colors
is the palette:

```{r}
data(gapminder_2007) + bar * count + x(continent) + color(continent) +
  title("The default, over filled bars")
```

*"Given gapminder 2007: bars derived by count, x is continent, color by
continent."*

```{r}
data(gapminder_2007) + bar * count + x(continent) + color(continent) +
  palette("soft") + title("palette(\"soft\")")
```

The first one is correct, and on a scatter of the same data it is the better
choice. The difference is what the color is doing: a point has to be
findable, while a bar is already visible and only has to differ from its
neighbors.

### Ramps {#ramps}

For a **continuous** color column, `palette()` names a ramp instead. These run
one way, from low values to high. Lightness changes steadily along the ramp, and
that is what shows the order.

| Name | What it is | Use it when |
|------|-----------|-------------|
| `"blue"` | the default: one hue, light→dark | almost always |
| `"viridis"` | many hues, with equal color steps for equal value steps | dense data where fine levels must stay distinct |
| `"magma"` | viridis's family: black through purple to cream | a dark plot with strong contrast |
| `"inferno"` | the same family, hotter | as above, warmer |
| `"plasma"` | the same family, blue through magenta to yellow | a dark plot with strong contrast, cooler at the low end |
| `"cividis"` | the family's colorblind-optimized member | the ramp itself has to survive two-color vision (dichromacy) |
| `"gray"` | no hue at all: light gray to near-black | figures that will be printed in black and white |

Those seven names do not tell you what the ramps look like. Here are six of the
seven, all but the default `blue`, on the Maunga Whau crater. The crater is cut
into cells and colored by mean height. A dense continuous surface like this is
what a ramp is for. `palette()` is a property of the plot, so six ramps means
six plots, and `|` and `/` put them on one page:

```{r}
#| echo: false
peek(maunga_whau)
```

```{r}
#| fig-height: 7
((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
    color(elevation) + palette("viridis") + title("viridis")) |
   (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("magma") + title("magma"))) /
  ((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("inferno") + title("inferno")) |
     (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
        color(elevation) + palette("plasma") + title("plasma"))) /
  ((data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
      color(elevation) + palette("cividis") + title("cividis")) |
     (data(maunga_whau) + zone * bin(16) * mean + x(east) + y(north) +
        color(elevation) + palette("gray") + title("gray")))
```

*"Given Maunga Whau: zones derived by bin into 16 and mean, x is east, y is
north, color by elevation, with the viridis palette, beside the same with magma;
above the same pair with inferno and plasma; above the same pair with cividis
and gray."*

That is the same sentence six times with one word changed. It is written out in
full because a loop or a variable would be R syntax, and every sentence in this
book is drawn in four languages, each of which spells those differently.

Those six plots show four things, one to a paragraph.

`gray` runs the other way. Its crater is dark at the summit while the other
five are bright there. Both directions are correct. gog's own ramps (`gray`
and the default `blue`) run light to dark, because with one hue a darker color
reads as a larger value, and a photocopier keeps those lightness differences.
The viridis family runs dark to light, because that is what those ramps are, and
reversing them would give a ramp named `magma` that no longer looks like magma.
In any given plot the legend decides the question, and that is why a mapped
channel always gets a legend. Read the top label, and it names the value at that
end of the ramp.

Of the five imported ramps, `magma` and `inferno` end on a near-white. `magma`'s
last stop is nearly the same color as the panel behind it, so the very top of
the scale almost disappears. You can see this in the pale flat summits of the
`magma` and `inferno` craters. A ramp with equal color steps ends pale by
design, which is why `blue` is the default rather than one of these. `blue` and
`gray` stop short of white, so their pale end stays visible.

`cividis` travels through the fewest hues, and deliberately: it uses only the
two that survive the common forms of color blindness. The other ramps run
through many more, so they separate neighboring values by hue as well as by
lightness. Giving that up is the cost, and it is worth accepting whenever the
ramp itself has to stay readable.

`gray` is for the figure that has to survive photocopying. It is a ramp only,
and there is no categorical gray palette. Readers cannot distinguish more than about
four grays. For categories in a print figure, use [`pattern`](#pattern),
which separates them and leaves the `color` channel free. Asking for
`palette("gray")` on a categorical column says so:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("gray")
```

### Diverging ramps

A sequential ramp means *more*. A **diverging** ramp means *away from a center,
in two directions*, which is what a change, a residual, a correlation and an
anomaly all are. gog provides two:

| Name | What it is | Use it when |
|------|-----------|-------------|
| `"blue_red"` | blue → neutral → red | the usual choice |
| `"brown_teal"` | brown → neutral → teal | the sign of the value is what matters, and readers may be colorblind |

The name reads low to high, so `blue_red` paints the smallest values blue and
the largest red. Both pass through the same gray at the center of the scale.

The palette does not decide which value gets the neutral color. That value is
the middle of the scale's domain, so `limits` is what puts zero there. Without
`limits`, the middle falls on the data's own midpoint. Here `cashflow` runs from
-32 to 120, so the midpoint is 44. That value is painted gray, and zero is
painted blue.

```{r}
#| echo: false
peek(cashflow)
```

```{r}
data(cashflow) + point + x(step) + y(delta) +
  color(delta) + palette("blue_red") + style(size = 9) +
  title("The neutral lands at 44, not at 0")
```

State a symmetric domain and zero becomes the center of the scale, so zero is
the value painted gray. Read the "Closing" point, which is exactly zero:

```{r}
data(cashflow) + point + x(step) + y(delta) +
  color(delta, limits = c(-120, 120)) + palette("blue_red") + style(size = 9) +
  title("limits = c(-120, 120) puts zero on the neutral")
```

*"Given the cashflow table: points, x is step, y is delta, color by delta from
-120 to 120, with the blue red palette."*

Two things follow from doing it this way. The ramp stays **linear**: equal
distance in color is equal distance in the data on both sides of the center.
Some scales fix a midpoint and then rescale each side to the data's own ends,
and a scale built that way is no longer linear. And `limits` [removes rows
outside it](scales.qmd), so the domain has to be wide enough to hold the data:
`c(-50, 50)` here would drop the opening balance and say so.

Symmetric limits have one drawback, worth knowing before you set them. If a
signed column runs from -2 to 100, almost none of the blue half is used, because
the data reaches only a little way below zero.

### Stops of your own

A named ramp is not the only way to get one. The colors you pass become the
stops, and gog interpolates between them:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + palette(c("white", "navy")) + style(size = 6) +
  title("Custom stops")
```

*"Given gapminder 2007: points, x is gdp, y is life, color by life, with a
palette of white and navy."*

gog refuses a palette that does not match the column, rather than guessing what
you meant. A categorical palette cannot serve a continuous column:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(life) + palette("okabe")
```

### A ramp on a skewed column

No ramp reads well on a skewed column, and changing the palette does not fix it.
A linear ramp spreads its colors evenly across the domain, so when most rows sit
in a narrow part of it they all receive almost the same color. Color by
`population` and nearly every country falls at the light end, because two
countries are several times larger than any other. That is the data, not the
palette, so the fix is a [log scale](scales.qmd) on the channel:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log") +
  x_label("GDP per capita")
```

`color`, `size` and `opacity` all take `scale =`, because all three answer "how
far along?". `shape`, `pattern` and `group` do not: they answer "which one?".

### Custom colors

Often the colors are decided before you begin: a journal asks for them, or an
earlier figure already used them.

Pass the colors yourself: hex values, CSS color names such as `firebrick`, or a
mix of the two:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  palette(c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00"))
```

The same call takes CSS color names, which are easier to read back in six
months:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) +
  palette(c("firebrick", "steelblue", "seagreen", "rebeccapurple", "darkorange"))
```

gog refuses a color it cannot recognize, and offers the nearest match. It does
not hand the name to the renderer to be painted black:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette(c("firebrick", "stelblue"))
```

Colors are assigned in **first-appearance order** in the data (the first unique
value gets the first palette color, the second unique value gets the second, and
so on), so the same table always gets the same colors. (Unless the column
[declares its order](data.qmd); then that order sets the sequence.)

One mistake to know about: a single color is not a palette. `palette()`
takes a palette *name* or the *colors themselves*, and `palette("red")` is
neither; it is an attempt to set one constant color, which is
[`style(color = "red")`](style.qmd):

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("red")
```

---

## Legend placement

gog places every legend to the right of the plot area. The plot area shrinks by
itself to make room, so there is no margin to set, no position to choose, and no
way to hide a legend. A legend too tall for the plot is left out rather than
drawn past the edge, and gog says so and names the setting that gives it room.

## In short

A channel takes a column and never a value, so `color("red")` is refused and
the refusal points you to `style()`. The five refining channels earn a legend,
and `x`, `y` and `z` earn an axis. `group`, `label` and `play` earn neither.
Which channels a mark accepts follows from its geometry.

## What comes next

Every channel here reads a column, so every plot answers a question about the
data. [Setting vs mapping](style.qmd) covers the opposite case: a value you choose
yourself, which reads no column and earns no legend.
