# Cluster

```{r}
#| include: false
source("R/setup.R")
```

Which categories in a table are most alike? A table of numbers holds the answer,
but a reader cannot see it in the rows. The `cluster` transform computes it. It
treats each category of an axis as a **leaf** of a tree. It finds the two
closest leaves and joins them into one branch. Then it measures again, treating
that new branch as one item, and joins the closest two again. It repeats until
one tree holds every leaf. The finished tree is called a **dendrogram**. As with
a pie chart, no atom spells that name.

This chapter builds the tree, uses it to order a
[tile plot](marks/zone.qmd#cells-the-axes-already-made-the-tile-plot), and then
puts the two together as a clustered heatmap. `nutrients` holds one row per pair
of food and nutrient. There are eight everyday foods, and each one is measured
on five nutrients:

```{r}
#| echo: false
peek(nutrients)
```

## The tree

Which two of the eight foods are most alike, and which one is least like all the
rest? To answer that, the engine needs to know what to compare. Each leaf is
described by a **profile**. `cluster(amount, over = nutrient)` says that one
leaf's profile is its `amount` in every category of `nutrient`. So each food's
profile is five numbers. Two numbers place a point on a scatter plot, and five
numbers place a point the same way, along five axes that are never drawn. The
distance between two leaves is the straight-line distance between their points,
the **Euclidean distance**. The distance between two branches is the average of
the distances from every leaf on one branch to every leaf on the other. At each
step the two closest branches join, and the join is drawn at that distance. This
rule is called **average linkage**.

```{r}
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  title("Eight foods, joined by what they are made of")
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, x is
food."*

`path` draws the tree. A tree is a set of points that must be joined in a fixed
order, and `line` would sort them by `x` instead. `path` draws the rows in the
order they arrive, and `cluster` writes them in the order the tree needs. Here
you bind `food` to `x`, so `x` carries the leaves. This chapter calls that axis
the **leaf axis**. You bind nothing to the other position, and `cluster`
supplies it. The transform computes the distance at which each pair of branches
joins, and it names that axis Distance.

The place where two branches meet is a **join**. The further a join is drawn
from the leaf axis, the more different the two branches are. Beans and lentils join lowest,
and rice and oats join just above them. Almonds join at the greatest distance of
all, so almonds are the food least like the rest.

Bind `y` instead of `x`, and the same tree is drawn sideways, with the leaves
down the left edge:

```{r}
data(nutrients) + path * cluster(amount, over = nutrient) + y(food)
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, y is
food."*

Whichever axis carries the leaves, the tree is the same. Two things about how
`cluster` builds it cannot be changed. First, distance is measured on the values
as they are, with no rescaling. Fat runs from 0.3 to 49.9 and iron from 0.5 to
3.7, so fat has far more effect on the distance than iron does. If you want
every nutrient to count equally, rescale `amount` separately within each
nutrient, in the host language, before you draw. Second, branches join at their
average distance, so a join is never closer to the leaf axis than the joins
inside its two branches. And two leaves with identical profiles are at distance
zero, so their join is drawn at zero, where every leaf's line begins.

## The order

Reading the finished tree along the leaf axis also gives an order, with similar
foods next to each other. That order helps on a tile plot, because a pattern in
the tiles is hard to see when similar foods are far apart. A tile plot of the
same table draws its slots in the order the rows arrived. That order says
nothing about the foods:

```{r}
data(nutrients) + zone + x(food) + y(nutrient) + color(amount) +
  title("In the order the rows arrived")
```

*"Given nutrients: zones, x is food, y is nutrient, color by amount."*

`zone * cluster` draws the same tiles. The leaf axis now runs in the tree's
order. Chicken and salmon are next to each other on the left, and oats, rice,
beans and lentils are together between spinach and almonds:

```{r}
data(nutrients) + zone * cluster(over = nutrient) +
  x(food) + y(nutrient) + color(amount) +
  title("In the tree's order")
```

*"Given nutrients: zones derived by cluster over nutrient, x is food, y is
nutrient, color by amount."*

On a tile plot, `amount` is already bound to `color`, so `cluster` does not name
it again. `over` names the profile column, `nutrient`, which is on `y`. The
other axis, `food` on `x`, holds the leaves, and that is the axis the tree
reorders.

## The clustered heatmap

So far one tree has ordered one axis. Put a tree on two sides of the tiles,
above and to the right, and each axis takes its order from the tree along that
side. The result is usually called a **clustered heatmap**. It is three plots
joined by two operators: a tree above, the tile plot below it, and a second tree
to the right of the tiles. The plots align because they share axes, as the
marginal plot in [Composition](composition.qmd#the-marginal-plot) does. A second
rule from [that chapter](composition.qmd#a-clustered-panel-decides-the-order)
sets their order: a shared categorical axis takes the order a clustered panel
computed for it. The tile plot in the middle carries no `cluster` of its own. It
is the same plain tile plot drawn earlier in this chapter, without its title.
`theme(legend = "bottom")` puts its legend under the tiles, so the tree on the
right stands next to the tiles it orders. The tile plot takes its column order
from the tree above and its row order from the tree on the right.

```{r}
(data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
   theme(height = 150)) /
  ((data(nutrients) + zone + x(food) + y(nutrient) + color(amount) +
      theme(legend = "bottom")) |
     (data(nutrients) + path * cluster(amount, over = food) + y(nutrient) +
        theme(width = 150)))
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, x is food;
above zones, x is food, y is nutrient, color by amount, beside paths derived by
cluster on amount, over food, y is nutrient."*

Each column of tiles is one food. Follow the tree upward from a column to its
first join: that join shows which food, or which group of foods, the column's
food is most like. The tree on the right swaps the roles. Its leaves are the
five nutrients, and each nutrient's profile is its amount in the eight foods.
Iron and fiber join first. The trees are not decoration. They decide the order
of the foods across the page and of the nutrients down it.

## What cluster refuses

The tree decides the order of the leaf axis. You saw `order()` on a bar chart,
where it sorted the categories along an axis, and you may want to write it here
as well. An axis can hold only one order. `order()` sorts by a column, `cluster`
derives an order from the tree, and writing both is a contradiction:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  order(food)
```

A profile holds one value for every category of the profile column, so that
column must hold categories. In `gm_all`, `year` holds numbers, so the grammar
reads it as continuous, and a continuous column has no categories:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
#| error: true
data(gm_all) + path * cluster(life, over = year) + x(continent)
```

Each entry of a profile is one value. If more than one row holds the same leaf
and the same profile category, gog refuses. It will not choose a mean, a sum or
a median for you, and the message names the pair it found. `gm_eras` keeps two
of `gm_all`'s years, 1957 and 2007, as the categories of a column called `era`,
so `over = era` does name a categorical column. It still fails, because every
continent holds one row per country in each era:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
#| error: true
data(gm_eras) + path * cluster(life, over = era) + x(continent)
```

Aggregate to one row per continent and era in the host language first, for
example with each continent's mean `life` in each era.

## Where the limits are

Each refusal above names a fix for the sentence you wrote. The limits below are
different, because they are about what the engine draws and what the grammar
can say.

Trees are often drawn in a circle, with the leaves around the rim, so you may
want to draw one that way. [`polar()`](polar.qmd) is what bends a plot, and a
tree inside it is valid grammar that this engine does not draw. A circular tree
needs curved lines, and `path` draws straight ones. Write `polar()`, and the
engine says so:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) + polar()
```

You may want to color the tree by food. Every join connects two or more foods,
so no single column can give its lines one color, and the engine refuses
`color()` on the tree:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  color(food)
```

`style(color = )` paints the whole tree in one color. Coloring each group of
branches below a chosen distance is not part of the grammar.

And a tree uses its second position for the distance, so it cannot share a plot
with the tiles it orders. That is why the clustered heatmap composes three
plots.

## What comes next

`cluster` reorders an axis, and the plot still uses two positions.
[Space](space.qmd) adds the third position, `z`, and draws the marks inside a
cube you can turn.
