# Cluster

```{r}
#| include: false
source("R/setup.R")
```

Which categories in a table are most alike? A table of numbers holds the answer,
but a reader cannot see it in the rows. The `cluster` transform computes it. It
treats each category of an axis as a **leaf** of a tree. It finds the two
closest leaves and joins them into one branch. Then it looks again, treating
that new branch as a single thing, and joins the closest two again. It repeats
until one tree holds every leaf. The world calls the picture a dendrogram, and
as with the pie there is no atom that spells the name.

This chapter builds the tree, then uses it to order a tile plot. `nutrients`
holds one row per pair of food and nutrient. There are eight everyday foods, and
each one is measured on five nutrients:

```{r}
#| echo: false
peek(nutrients)
```

## The tree

Which two of the eight foods are most alike, and which one is least like all the
rest? To answer that, the engine needs to know what to compare. Each leaf is
described by a **profile**. `cluster(amount, over = nutrient)` says that one
leaf's profile is its `amount` in every category of `nutrient`. So each food's
profile is five numbers, and those five numbers give the position of one point.
The distance between two leaves is the straight-line distance between their two
points. Two branches join at an average distance. Measure from every leaf on one
branch to every leaf on the other, and average all of those distances.

```{r}
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  title("Eight foods, joined by what they are made of")
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, x is
food."*

`path` draws the tree. A tree is a set of points that only make the right shape
when they are joined in a fixed order. `path` is the mark that keeps that order,
because it strokes the rows as the table gives them. Here `x(food)` is the bound
position, so `x` carries the leaves, and this chapter calls it the leaf axis.
You bind nothing to the other position, and `cluster` supplies it. The transform
computes the distance at which each pair of branches joins, and it names that
axis Distance. The place where two branches meet is a **join**. The further a
join is drawn from the leaf axis, the more different the two branches are. Beans
and lentils join lowest, and rice and oats join just above them. Almonds join
highest of all, which makes almonds the food least like the rest.

Bind `y` instead of `x`, and the same tree is drawn sideways, with the leaves
down the left edge:

```{r}
data(nutrients) + path * cluster(amount, over = nutrient) + y(food)
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, y is
food."*

Three things about how `cluster` builds the tree cannot be changed. Distance is
measured on the values as they are, with no rescaling first. Fat runs from 0.3
to 49.9 and iron from 0.5 to 3.7, so fat has far more effect on the distance
than iron does. If you want every nutrient to count equally, rescale the columns
where your data lives. Branches join at their average distance, so every join is
further from the leaf axis than the joins it holds. And two leaves with
identical profiles join at zero, so their join is drawn on the leaf axis itself.

## The order

Reading the finished tree along the leaf axis also gives an order, with similar
foods next to each other. That order is worth having on a tile plot, because a
pattern in the tiles is hard to see when similar foods are far apart. A tile
plot of the same table draws its slots in the order the rows arrived. Nothing
chose that order:

```{r}
data(nutrients) + zone + x(food) + y(nutrient) + color(amount) +
  title("In the order the rows arrived")
```

*"Given nutrients: zones, x is food, y is nutrient, color by amount."*

`zone * cluster` draws the same tiles. The leaf axis now runs in the tree's
order. Chicken and salmon are next to each other on the left, and the four foods
that are mostly carbohydrate are together between spinach and almonds:

```{r}
data(nutrients) + zone * cluster(over = nutrient) +
  x(food) + y(nutrient) + color(amount) +
  title("In the tree's order")
```

*"Given nutrients: zones derived by cluster over nutrient, x is food, y is
nutrient, color by amount."*

On a tile plot the values are already bound to `color`, so `cluster` need not
name them again. `over` names the profile column. The other categorical axis
holds the leaves, and that is the axis the tree reorders.

## The clustered heatmap

So far one tree has ordered one axis. Put a tree on both sides of the tiles, and
both axes take an order, so you can see where each one came from. The world
calls the result a clustered heatmap. It is three plots joined by two operators:
a tree above, the tile plot below it, and a second tree to the right of the
legend. The alignment comes from one rule in
[Composition](composition.qmd#a-clustered-panel-decides-the-order): a shared
categorical axis takes the order a clustered panel computed for it. The tile
plot in the middle carries no `cluster` of its own. It is the same plain tile
plot drawn earlier in this chapter. Its column order arrives from the tree
above, and its row order from the tree on the right.

```{r}
(data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
   theme(height = 150)) /
  ((data(nutrients) + zone + x(food) + y(nutrient) + color(amount)) |
     (data(nutrients) + path * cluster(amount, over = food) + y(nutrient) +
        theme(width = 150)))
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, x is food,
above zones, x is food, y is nutrient, color by amount, beside paths derived by
cluster on amount, over food, y is nutrient."*

Each column of tiles is one food. Read the tree directly above a column, and it
shows which foods that one joined first. The trees are not decoration. They
decide which foods end up next to each other.

## What cluster refuses

You met `order()` on a bar chart, where it sorted the categories along an axis.
You may want to write it here as well. An axis holds one order. `order()` sorts
by a column, `cluster` derives an order from the tree, and writing both is a
contradiction:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  order(food)
```

A profile holds one value for every category of the profile column, so that
column must hold categories. A continuous column has no categories:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
#| error: true
data(gm_all) + path * cluster(life, over = year) + x(continent)
```

Each entry of a profile is one value. If one leaf and one profile category
together match more than one row, gog refuses rather than choosing between them,
and the message names the pair it found. `gm_eras` records the same two years as
categories, so `over = era` passes the previous rule. It breaks this one: every
continent holds one row per country in each era. Aggregate to one row per pair
first, where your data lives:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
#| error: true
data(gm_eras) + path * cluster(life, over = era) + x(continent)
```

## Where the limits are

Trees are often drawn in a circle, with the leaves around the rim, so asking for
a circular tree is a reasonable thing to do. That bend is
[`polar()`](polar.qmd), and a tree inside it is valid grammar that this engine
does not draw. Write `polar()`, and the engine says exactly that:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) + polar()
```

The grammar has no way to cut the tree into groups and color each group, and
`color()` on the tree says so:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  color(food)
```

And one plot clusters one axis. The clustered heatmap orders both, because it
composes two clustered plots, one for each axis.

## What comes next

A cluster reorders an axis, and the plot still uses two positions.
[Space](space.qmd) adds the third position, `z`, which lifts the marks off the
page into a cube you can turn.
