# Cluster

```{r}
#| include: false
source("R/setup.R")
```

Which things are most alike? A table of numbers holds the answer, but a
reader cannot see it in the rows. The `cluster` transform computes it. It
treats each category of an axis as a **leaf** of a tree. It joins the two
closest leaves, then the next two, and keeps joining until one tree holds
them all. The world calls the picture a dendrogram, and as with the pie
there is no atom that spells the name.

This chapter builds the tree, then uses it to order a tile plot. `nutrients`
holds one row per pair of food and nutrient: eight everyday foods, each
measured on five nutrients:

```{r}
#| echo: false
peek(nutrients)
```

## The tree

Each leaf is described by a **profile**. `cluster(amount, over = nutrient)`
says that one leaf's profile is its `amount` in every category of
`nutrient`. The distance between two leaves is the distance between their
profiles, and two branches join at the average distance between their
leaves.

```{r}
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  title("Eight foods, joined by what they are made of")
```

*"Given nutrients: paths derived by cluster on amount, over nutrient, x is food."*

`path` draws the tree, because the drawing is corner points in a stated
order, and `path` connects rows without sorting them. The bound position is
the leaf axis. The other position stays unbound: the tree computes the
distance at which two branches join, and labels that axis itself. Read the
height of a join as how different its two branches are. Rice and oats join
low, beans and lentils join about as low, and almonds join last of all.

Bind `y` instead of `x`, and the same tree lies on its side:

```{r}
data(nutrients) + path * cluster(amount, over = nutrient) + y(food)
```

Three things about the statistic cannot be changed. Distance is
straight-line distance on the values as given, so a nutrient measured in
large amounts counts for more than one measured in small amounts. Rescale
where your data lives if that is not what you mean. Branches join at their
average distance, so every join stands further from the leaves than the
joins inside it. And two leaves with identical profiles join at zero, drawn
as a join resting on the leaf axis.

## The order

The tree can also be read as an order. A tile plot of the same table draws
its slots in the order the rows arrived, and nothing chose that order:

```{r}
data(nutrients) + zone + x(food) + y(nutrient) + color(amount) +
  title("In the order the rows arrived")
```

`zone * cluster` draws the same tiles with the leaf axis in the tree's
order. Chicken and salmon now stand together on the left, and the four
foods that are mostly carbohydrate gather in the middle:

```{r}
data(nutrients) + zone * cluster(over = nutrient) +
  x(food) + y(nutrient) + color(amount) +
  title("In the tree's order")
```

On a tile plot the values are already bound to `color`, so the atom does not
name them again. `over` names the profile column, and the other categorical
axis is the leaf axis the tree reorders.

## The clustered heatmap

The full figure puts a tree beside the tiles on two sides, so a reader can
see where the order came from. The world calls it a clustered heatmap. It is
three plots and two operators: a tree above, the tile plot, and a second
tree on the right, beyond the color key. The alignment comes from one rule
in [Composition](composition.qmd#a-clustered-panel-decides-the-order): a
shared categorical axis takes the order a clustered panel computed for it.
The tile plot itself stays plain. Its column order arrives from the top
tree, and its row order from the side one.

```{r}
(data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
   theme(height = 150)) /
  ((data(nutrients) + zone + x(food) + y(nutrient) + color(amount)) |
     (data(nutrients) + path * cluster(amount, over = food) + y(nutrient) +
        theme(width = 150)))
```

Follow a column of tiles up to the tree above it, and the tree says which
columns it was joined with. The trees are not decoration: they decide which
column stands next to which.

## What cluster refuses

An axis holds one order. `order()` sorts by a column, `cluster` derives an
order from the tree, and saying both is a contradiction:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  order(food)
```

A profile holds one value for every category of the profile column, so that
column must hold categories. A numeric column has none:

```{r}
#| error: true
data(gm_all) + path * cluster(life, over = year) + x(continent)
```

A profile's entry is one value, so a pair that holds several rows is
refused, and the message names the pair. In `gm_eras` every continent holds
one row per country in each era. Aggregate to one row per pair first, where
the data lives:

```{r}
#| error: true
data(gm_eras) + path * cluster(life, over = era) + x(continent)
```

## Where the limits are

A circular tree is valid grammar that this engine does not draw yet. Adding
`polar()` says exactly that:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) + polar()
```

Cutting the tree into groups and coloring each group is not built, and
`color()` on the tree says what is missing:

```{r}
#| error: true
data(nutrients) + path * cluster(amount, over = nutrient) + x(food) +
  color(food)
```

And one plot clusters one axis. The heatmap above orders both, because it
composes two clustered plots, one for each axis.
