# Combinations

```{r}
#| include: false
source("R/setup.R")
```

Every mark down the side, every attribute across the top: the whole of what the
grammar lets you combine, on one page. Four crossings live here. The first is
what a mark lets you *map* (the channels); the second is what a mark lets you
*derive* (the transforms); the third is which two transforms can *stand together*
on one mark; the fourth is where a mark can be *drawn* (the coordinate spaces).
Not only what a mark takes, but what it does not, and the surprising-but-legal
pairings you would never think to try.

These grids are the orthogonality law made visible. [Law 1](design-laws.qmd) says
every compatible atom combines with every other; a table is that promise
enumerated, one cell at a time. It is the direct analog of the 자모 (*jamo*)
chart a Hangeul (한글) learner reads: consonants down the side, vowels across
the top, the whole syllable space in the crossing.

Legal is not the same as useful. [Law 8](design-laws.qmd) guarantees only the
first: where a cell says a combination renders, the engine draws it, but whether it
*should* be drawn is your judgment. The empty cells teach as much as the full
ones, marking the honest edge of what a geometry can express rather than an
oversight.

Nothing here is typed by hand. Every table is generated live from the engine's own
rule table each time the book builds, the same table the engine consults to accept
or refuse your plot. A grid that is *described* can fall out of step with the code;
a grid that is *generated* cannot. Add a mark, a channel, a transform or a space
and these pages fill in by themselves.

## Marks and channels

Down the side, the marks; across the top, the channels. A cell says whether you
may map a column to that channel on that mark.

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules`, which dumps `rule_for` (the one legality
# table) as JSON. The glyph for each cell is read straight off that verdict, so
# the grid can never claim something the engine would not enforce.
cli   <- find_gog_cli()
rules <- jsonlite::fromJSON(paste(system2(cli, "--rules", stdout = TRUE), collapse = "\n"))

marks <- rules$marks$name[rules$marks$drawable]  # the drawn marks, kernel order
chans <- rules$channels                          # every channel, kernel order
cells <- rules$cells

# Five states, each a straight reading of the rule:
#   ●  a required position (Must)
#   ✅ a column maps here and the engine draws it today
#   ○  not mappable per row, but style() can set it once for the layer
#   ◌  legal grammar the renderer refuses today — owed, blocked, or decided
#      against; the refusal message says which, and the legend below splits them
#   —  the mark has no such feature at all
glyph <- function(ob, renders, settable) {
  if (ob == "must")         "●"  # ●
  else if (!is.na(renders)) "✅"  # ✅
  else if (ob == "can")     "◌"  # ◌  grammar-legal, not drawn yet
  else if (settable)        "○"  # ○  set-only
  else                      "—"  # —
}
lookup <- setNames(
  mapply(glyph, cells$obligation, cells$renders, cells$settable),
  paste(cells$mark, cells$channel)
)

header  <- paste0("| Mark | ", paste0("`", chans, "`", collapse = " | "), " |")
divider <- paste0("|:--|", paste(rep(":--:", length(chans)), collapse = "|"), "|")
rows <- vapply(marks, function(m) {
  gs <- vapply(chans, function(c) lookup[[paste(m, c)]], character(1))
  paste0("| `", m, "` | ", paste(gs, collapse = " | "), " |")
}, character(1))
cat(header, divider, rows, sep = "\n")
```

Reading the five glyphs:

- **● must**: a required position. The mark does not render without it. A `point`
  needs an `x` and a `y`; a `text` needs its `label`.
- **✅ maps**: bind a column here and the engine draws it today. The ordinary
  case, a channel doing its job.
- **○ set**: the mark has this feature, but not row by row. A `line` is one
  stroke, so its `size` and `opacity` are a single value for the whole layer,
  fixed with [`style()`](style.qmd) rather than mapped. Those constants have their
  own [Mark × Setting grid](style.qmd).
- **◌ not drawn**: legal grammar the engine refuses today, with a direction rather
  than a wrong picture. One glyph, three verdicts, and the refusal message tells
  you which you have. **Owed**: a mapped font `size` on `text`.
  **Blocked** on infrastructure: `z` on `rule` and on `zone`, whose 3-D forms are a
  plane and a slab with no footprint to sort by, so they need an engine that can
  tell what hides what.
  **Decided**: `z` on `line`, `step`, `area` and `ribbon`, which read a domain left
  to right where a cube has none, and which are pointed at `path` instead. Only
  the first of the three will ever fill in. The Mark × Space grid below is the
  same three readings on its own blanks.
- **— none**: the mark has no such feature. A `bar` has no glyph to `shape`, a
  `point` nothing to `group` into a line. Binding it means nothing, and the engine
  says so.

Three things the crossing shows at a glance that prose keeps losing:

- **`group` splits a `line`, not a `bar`.** Run down the `group` column: ✅ on the
  path and distribution marks (`line`, `step`, `area`, `interval`, `box`,
  `ribbon`), — on `point`, `bar`, and `text`. A bar is split into groups by
  `color`, then piled or set side by side with `stack` / `dodge`; a bare `group`
  has no line to divide.
- **`label` rides `text` alone.** Its column is ● on `text` and — everywhere else.
  A labeled scatter is the superposition `point + text`, never a `label` hung on
  the point.
- **A category on `x` reaches every mark; on `y` it does not.** Read the two
  position rows against each other and the grid divides the marks in two. The
  glyph marks — `point`, `text` — and the marks whose slot width does the
  measuring — `bar`, `interval`, `box` — take either type on *both* positions,
  which is why a strip plot, a horizontal bar and a categorical box all fall out
  of one rule. So do `path`, which sorts nothing and therefore has no domain,
  `rule`, which measures nothing on either axis, and `zone`, which measures
  nothing by length at all — that last is what makes a heatmap cell a `zone`
  rather than a second kind of bar. The path and region marks —
  `line`, `step`, `area`, `ribbon` — take either type on `x` alone: their `x` is
  the domain the path is read along and their `y` is the quantity it traces, and
  those are not the same kind of axis. A glyph's two axes are.
- **Two rows have no ● at all.** Every other mark must be told where to put
  something; `rule` and `zone` show ✅ on both positions and ● on neither, because
  each needs *one* of the two and no per-channel cell can say "either, but not
  both". Those obligations are stated once in the engine instead, and a pair of
  ✅s with no ● is what a mark the panel finishes looks like from here
  ([Rule](marks/rule.qmd), [Zone](marks/zone.qmd)).
- **`zone`'s `color` is the only either-typed color on a fill.** Every other
  region takes a category; a zone takes a number too, and reads the ramp. A
  hairline has no area to decode a scale from, which is why `rule` and `line`
  refuse it — a rectangle has nothing but area.

## Marks and transforms

The second crossing: not what a mark maps, but what it *derives*. A transform
enters through `*` (`bar * bin`, `interval * range`, `point * jitter`), and which
transform a mark takes is decided by the mark's geometry against the transform's
kind. There are three kinds, and each has its own grid below.

Three glyphs read every cell:

- **✅ combines**: the transform applies to this mark and renders.
- **● required**: the transform's *kind* is the mark's minimum syllable. An
  `interval` and a `ribbon` float between a low and a high, so they will not draw
  until *one* transform of that kind supplies the pair. The `●` cells in a row are
  a choice, not a list: pick one.
- **— none**: the mark has no use for this kind of transform, and the engine
  refuses it with a direction toward the transform its geometry does want.

```{r}
#| include: false
# One dump, read once; each grid below is a slice of it by transform class, so
# adding a transform widens the right grid on its own.
tmarks <- rules$marks$name[rules$marks$drawable]
tf     <- rules$transforms          # name, class
tcells <- rules$transform_cells     # transform, mark, state
tglyph <- c(required = "●", combines = "✅", none = "—")
tlook  <- setNames(unname(tglyph[tcells$state]), paste(tcells$transform, tcells$mark))
tf_block <- function(cls) {
  cols    <- tf$name[tf$class == cls]
  header  <- paste0("| Mark | ", paste0("`", cols, "`", collapse = " | "), " |")
  divider <- paste0("|:--|", paste(rep(":--:", length(cols)), collapse = "|"), "|")
  rows <- vapply(tmarks, function(m) {
    gs <- vapply(cols, function(t) tlook[[paste(t, m)]], character(1))
    paste0("| `", m, "` | ", paste(gs, collapse = " | "), " |")
  }, character(1))
  cat(header, divider, rows, sep = "\n")
}
```

### Statistics — a value per group

The summarizing transforms each reduce the data to one value at each x: a bin's
count, a group's `mean`, a smoothed height. Any mark that draws a value at each x
can draw that summary, so the whole block moves together, which is why a
distribution has three spellings and no `histogram(element=)` family: `bar * bin`
fills it, `line * bin` is the frequency polygon, `step * bin` the staircase.

```{r}
#| echo: false
#| results: asis
tf_block("statistic")
```

The five drawing marks — `point`, `line`, `area`, `bar`, `step` — take the lot;
the other four take none, and each for a reason the next two grids make good. A
span mark (`interval`, `ribbon`) needs a *pair*, not a single value. A `box` takes
no statistic because it carries its own: the five-number summary is injected by the
mark, not composed, so it never appears as a column here. And a `text` draws its
`label`, which no statistic produces.

### Ranges and bands — a low and a high

Three transforms produce a *pair* of values, a low boundary and a high one, per
group. `range` and `confidence` compute the pair from your data; `bounds` reshapes
a pair you computed already. Four marks draw such a pair: `interval` whiskers it,
`ribbon` fills it, `line` and `step` trace its two boundaries as an unfilled band.

```{r}
#| echo: false
#| results: asis
tf_block("pair")
```

The `●` on `interval` and `ribbon` is the one place this crossing shows a
*requirement*, not just a permission. Neither mark has a baseline to fall back on,
so neither renders until a pair transform gives it a low and a high: `interval *
range` is the whole minimum syllable, a bare `interval` is refused. A `line` or a
`step` already draws fine on its own, so the same transform is optional there, a
second way to use the mark. Everywhere else the pair has nowhere to go: on a
`point` or a `bar` it is refused toward the whisker, on an `area` toward the
`ribbon` that is its filled twin.

### Collision modifiers — an offset, not a statistic

When marks collide at one position, three offsets pull them apart, and which one a
mark takes is decided by its geometry and by the *axis* the collision is on.
`dodge` subdivides a *width*, so it serves the marks that have one (`bar`, `box`,
`interval`). `stack` accumulates along the *measure* axis, so it serves the marks
that can spend a span: `bar` and `area` fill it with length, and a `point`, having
no length to stretch, spends it on how many dots there are, which is the dot plot.
`jitter` spreads a mark with *no* width along a *categorical* axis, which is
`point` alone.

```{r}
#| echo: false
#| results: asis
tf_block("collision")
```

No mark is left without one, and exactly one takes two: `point`, which is the only
mark that can meet two of the three preconditions, on different axes. `stack`
answers its overlap along the measure axis (a pile of dots, which can be counted);
`jitter` answers the same mark's overlap along a categorical one (a nudge, because a
category has no magnitude to spend). Everywhere else the division is exclusive,
which is why the engine can always name the right tool when you reach for the wrong
one. Ask for `point * dodge` and it points at `jitter`; ask for `area * dodge` and
it points at `stack`. A `box` shows the split cleanly: it takes no statistic (its own is
built in) yet still takes `dodge`, because setting grouped boxes side by side is a
question about position, orthogonal to the summary each box already carries.

## Transforms and transforms

The third crossing is the only one where a mark does not appear. Two transforms can
follow one mark — `bar * bin * mean` cuts the axis into bands, then averages inside
each. Which pairs can do that is decided by the transforms alone, and by one rule.

A transform has a **job**. There are four of them, and every transform in GOG does
at least one:

| job | the question it answers | the transforms that answer it |
|:--|:--|:--|
| **extent** | where are the cells? | `bin`, `density`, `smooth`, `partition`, and `bounds` on a `zone` |
| **measure** | what is in them? | `count`, `sum`, `mean`, `median`, `max`, `min`, `range`, `confidence`, `bounds` |
| **scale** | what scale is the answer read on? | `proportion`, and `stack(share = TRUE)` |
| **position** | where do the marks sit? | `dodge`, `stack`, `jitter` |

**Two transforms compose when they do different jobs, and contradict when they do
the same one.** A cell holds one extent, one measurement, one scale and one
arrangement. So a second transform doing a job the first already did leaves the
engine no reading that keeps both, and it refuses rather than drawing one and
discarding the other.

One transform is the exception, and it is the exception that makes chaining useful.
`bin` does two jobs: it cuts the axis, and it counts what lands in each band. The
count is a by-product of the cut, so `bin` can give it up — and it gives it up to a
transform you handed a *column*. That is why `bar * bin * mean` composes: `bin` supplies the
cells and `mean` supplies their value. And it is why `bar * bin * count` does not,
since a tally was never handed a column to reduce.

```{r}
#| echo: false
#| results: asis
# The chain grid, generated from the same predicate the engine refuses with, so it
# cannot claim a pair the engine would reject. Adding a transform widens it.
cc   <- rules$chain_cells
look <- setNames(ifelse(cc$legal, "✅", "—"), paste(cc$a, cc$b))
cols <- rules$transforms$name
cat(paste0("| | ", paste0("`", cols, "`", collapse = " | "), " |"),
    paste0("|:--|", paste(rep(":--:", length(cols)), collapse = "|"), "|"),
    vapply(cols, function(a) {
      gs <- vapply(cols, function(b) if (a == b) "·" else look[[paste(a, b)]], character(1))
      paste0("| `", a, "` | ", paste(gs, collapse = " | "), " |")
    }, character(1)),
    sep = "\n")
```

The grid is symmetric, and that is the rule showing rather than being claimed.
Whether two transforms contradict is a fact about their jobs. It has nothing to do
with which one you typed first. So order is yours to choose: write them in the order
they happen, and the sentence reads the way it runs.

Read the block of `—` in the middle. Nine transforms measure, so no two of them
combine, and that single square is the whole reason `bar * count * mean` is refused,
and `bar * mean * sum`, and `interval * range * confidence`. `bin` is the row with a
`✅` where the others have `—`, which is the exception above, drawn.

Two collisions the grid cannot show, because both depend on something outside the
pair. `stack(share = TRUE)` divides each pile by its own total, which is a *scale*,
so it contradicts `proportion` — two divisions, and the second reads as shares of
nothing. And `bounds` says where a `zone`'s sides are. On that one mark it does the *extent*
job, so it contradicts `bin`. On every other mark it names the low and the high on
the measure axis. The engine reads both cases correctly, and the grid above shows the
reading that applies everywhere else.

Chaining stops at four, and nobody chose the number. Four jobs, one transform each,
so a fifth transform has to repeat a job — and repeating one is the contradiction
this whole section is about.

### The shapes a long chain can take

Three and four transforms sound like a large space, and they are not. Once each job
is filled at most once, and the marks are asked which transforms they take, **every
chain of three or four has one of six shapes**. Two slots stand for a family rather
than a transform: `<aggregation>` is any of `sum`, `mean`, `median`, `max`, `min`,
and `<collision modifier>` is any of `dodge`, `stack`, `jitter`.

```{r}
#| echo: false
#| results: asis
# The shapes, derived rather than listed: every chain of three or four the pair grid
# above allows on some mark, collapsed to the family each slot stands for. Nothing is
# typed, so a new transform widens this by itself — and it cannot claim a chain the
# engine refuses, because it reads the same cells the grid does.
fam <- c(sum = "<aggregation>", mean = "<aggregation>", median = "<aggregation>",
         max = "<aggregation>", min = "<aggregation>",
         dodge = "<collision modifier>", stack = "<collision modifier>",
         jitter = "<collision modifier>")
jobof  <- setNames(rules$transforms$jobs, rules$transforms$name)
rankof <- c(extent = 1, measure = 2, scale = 3, position = 4)
slot   <- function(t) if (t %in% names(fam)) unname(fam[t]) else t
# A transform's plain job — the first the engine lists, which is the one it does
# whatever the mark and settings are. `bin` fills extent and measure and hands the
# second over; `stack` rescales only when given `share = TRUE`. Taking the first is
# what keeps `stack` a position transform here rather than a scale one.
jobt   <- function(t) jobof[[t]][1]
tlegal <- setNames(rules$transform_cells$state,
                   paste(rules$transform_cells$transform, rules$transform_cells$mark))
pairok <- setNames(rules$chain_cells$legal, paste(rules$chain_cells$a, rules$chain_cells$b))
shapes <- list()
for (m in rules$marks$name[rules$marks$drawable]) {
  take <- Filter(function(t) tlegal[[paste(t, m)]] != "none", rules$transforms$name)
  for (k in c(3, 4)) {
    if (length(take) < k) next
    combos <- combn(take, k, simplify = FALSE)
    for (C in combos) {
      if (!all(apply(combn(C, 2), 2, function(p) pairok[[paste(p[1], p[2])]]))) next
      # `base::order` — `order()` is a gog channel, and the grammar masks it.
      C <- C[base::order(vapply(C, function(t) rankof[[jobt(t)]], 1))]
      key <- paste(unique(vapply(C, slot, "")), collapse = " * ")
      if (is.null(shapes[[key]])) shapes[[key]] <- list(n = k, jobs = vapply(C, jobt, ""))
    }
  }
}
ord <- base::order(vapply(shapes, function(s) s$n, 1), names(shapes))
cat("| how many | the shape | the jobs it fills |",
    "|:--:|:--|:--|",
    vapply(ord, function(i) sprintf("| %d | `%s` | %s |",
      shapes[[i]]$n, names(shapes)[i],
      paste(unique(shapes[[i]]$jobs), collapse = ", ")), ""),
    sep = "\n")
```

Read the four-transform row first, because it is the whole ceiling in one line. It
fills each job exactly once, and only one of its slots has any real choice in it.
`bin` is the only extent transform that hands its measurement over. `proportion` is
the only one that rescales. The family in the last slot is picked by the mark. So the
aggregation is the only free decision, which is why four is reachable and not roomy.

The shapes say what the *transforms* allow. A mark can still refuse one of them for a
reason about its own geometry, and the grids earlier in this chapter are where that
is decided. `point * proportion * stack` is the clearest case. A dot plot piles one
whole dot per observation, and a share is less than one dot. So the chain is refused
on `point` and drawn on `bar`.

## Marks and spaces

The fourth crossing is not an attribute of the mark at all. It is the coordinate
space the whole plot sits in, and it asks a different question of every mark:
once the plane is bent, or stood up into a cube, can this geometry still be
drawn? A `polar()` bar is a wedge, a `polar()` line is a closed curve, and a `z`
scatter is a projected cloud, all from the same sentences that draw them flat.

Two glyphs here, not five. Either the engine draws this mark in this space today,
or it does not and says so:

- **✅ drawn**: the mark renders in this space.
- **— not drawn**: legal to write and refused with direction, never half-drawn.
  As on the channel grid, one glyph covers owed, blocked and decided alike; which
  one a blank is, is the subject of the three paragraphs under the table.

```{r}
#| echo: false
#| results: asis
spaces <- setdiff(rules$spaces, "globe")
scells <- rules$space_cells
slook  <- setNames(ifelse(scells$drawn, "✅", "—"), paste(scells$space, scells$mark))
header  <- paste0("| Mark | ", paste0("`", spaces, "`", collapse = " | "), " |")
divider <- paste0("|:--|", paste(rep(":--:", length(spaces)), collapse = "|"), "|")
rows <- vapply(marks, function(m) {
  gs <- vapply(spaces, function(s) slook[[paste(s, m)]], character(1))
  paste0("| `", m, "` | ", paste(gs, collapse = " | "), " |")
}, character(1))
cat(header, divider, rows, sep = "\n")
```

`flat` is full but for one cell, and that cell is worth stopping on: **`surface`
does not draw in the plane.** Every other mark was built there and some have since
learned to stand up, while a surface is a sheet through three positions and there
is no such thing without the third, so its minimum syllable includes the cube. It
is the only mark whose row reads the other way round from every other
([Surface](marks/surface.qmd)).

`space` holds six marks, and which six is a reading rather than a list of what got
built: a `point` is a place, a `path` is an order, a `surface` is a sheet, and
`bar`, `interval` and `box` are the three marks that stand in a slot and measure
along the other axis — a length, a span and a summary, which the grammar has
treated as one family since orientation was decided. Each means in a cube exactly
what it meant on the page.

The blanks in that column are **not one kind of thing**, and the engine now says
which of three each is. `line`, `step`, `area` and `ribbon` read a *domain* left to
right, and a cube has no left to right, so they are **refused** — a decision, not a
queue entry, with `path` as the direction. `rule` and `zone` are **blocked**: their
3-D forms are a plane and a slab, neither of which has a footprint to sort by, so
placing them among the data needs an engine that can tell what hides what. Until
2026-07-26 all six said
"not drawn yet", which promised a renderer for four marks the design had already
ruled out ([Space](space.qmd)).
`polar` is the one column with no empty half at all: every mark that draws on the
page draws in the circle. It reached that state last, and by losing an argument
rather than by winning one — five marks were held out on the ground that their
straight edges would have to become arcs, and three of them turned out never to
have needed an arc, a band's boundaries being the same chords a `line` draws. The
single blank is `surface`, and it is not a polar refusal but the same one its flat
cell carries: a sheet needs the cube ([Polar](polar.qmd)).

`nest` is the opposite shape of column: two cells filled and eleven blank, and
the blanks are not a queue. This space asks a different question from the other
four. Everywhere else a mark bends because its geometry survives a map of the
plane; a packing is not a map of the plane at all, so what it can draw is
whatever it can hand *a rectangle* to. `bar` can take one, because a bar is a
measure and this space is the third thing a measure can become, after a length
and an angle. `text` can too, and it is the only mark here that names a region
rather than filling one: the rectangle is already computed, so a label goes to
its center, which is what a published treemap has and what a legend cannot do
once the split runs to three figures ([Nest](nest.qmd)).

The eleven blanks have one reason between them, and it is not the one the
column's shape suggests. They are the marks placed by a *position*: `point` needs
somewhere to sit, `line`, `step`, `path`, `area` and `ribbon` need an order along
an axis (and two touching regions here are explicitly *not* near each other in
the data), `rule` and `zone` span an axis that is not there, and `box`,
`interval` and `surface` measure along one. `text` is not among them because it
is not placed by a position here either — it is placed by the region, which is
the same thing that places the bar.

`map` is empty all the way down, and it is in the grid for that reason. It is
named in the kernel and has no renderer behind it, so the column is the honest
edge of the engine rather than a gap in the table. A whole column of `—` is what
an unbuilt idea looks like when nothing is hidden.

**Every mark in the kernel is now in all three crossings.** `surface` was the last
one left out — a grammar-legal mark with no renderer, whose every cell would have
read the not-yet-drawn glyph — and it joined them in 2026. `path` was in that
sentence before it, and both arrivals are legible in the grids rather than merely
recorded: a path has a full row on Mark × Channel and an entirely **empty** one on
Mark × Transform, because a path *is* its rows in order and every statistic replaces
the rows. `rule` reads the same way on the transform grid for a reason one axis over,
having handed a whole axis to the panel, so there is no measure to compute. And
`surface` has a row of the transforms that **tile a floor** — `density`, which
estimates a value at every node for the sheet to interpolate between, and `bin`,
which cuts adjacent cells for it to lay a flat lid on, with the aggregation family
riding `bin` to name the column each cell holds. `path` keeps the shorter row beside
it, tracing the one field it derives where a surface raises it.

With these grids the orthogonality matrix is stated in full: what each mark maps,
what it sets ([the settings grid](style.qmd)), what it derives, and where it can be
drawn.
