# Combinations

```{r}
#| include: false
source("R/setup.R")
```

This page holds the whole of what the grammar lets you combine, in four
crossings. The first is what a mark lets you *map*, which is the channels. The
second is what a mark lets you *derive*, which is the transforms. The third is
which two transforms can be *chained* on one mark. The fourth is where a
mark can be *drawn*, which is the coordinate spaces. Three of them show what a
mark takes, what it does not, and the surprising pairings that are legal anyway.
The third asks the same question of two transforms.

These grids are the orthogonality law made visible. [Law 1](design-laws.qmd)
says every compatible atom combines with every other; a table is that promise
enumerated, one cell at a time. It is the direct analog of the letter chart a
Hangeul (한글) learner reads, with the consonants down one side and the vowels
across the top. A syllable block is what those letters assemble into. This book
has used that pairing from the start: a mark is a consonant and a position is a
vowel, and neither draws alone. Every syllable the grammar can form appears
somewhere in these grids.

Legal is not the same as useful. [Law 8](design-laws.qmd), Pronounceable ≠
Usable, guarantees only that the engine draws what a ✅ cell allows. Whether it
should be drawn is your judgment. A cell the engine refuses teaches as much as
one it draws. It records a real limit of the geometry, a ruling, or work the
engine still owes, and never an oversight.

Every crossing here is generated live from the engine's own rule table, each
time the book is built. That is the same table the engine consults to accept or
refuse your plot. A grid written by hand can stop matching the code; a grid
generated from it cannot. Add a mark, a channel, a transform or a space, and this
page gains the row or the column by itself.

## Marks and channels

Start with the channels. The marks run down the side and the channels across
the top, and each cell says whether you may map a column to that channel on that
mark.

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules`, which dumps `rule_for` (the one legality
# table) as JSON. The glyph for each cell is read straight off that verdict, so
# the grid can never claim something the engine would not enforce.
cli   <- find_gog_cli()
rules <- jsonlite::fromJSON(paste(system2(cli, "--rules", stdout = TRUE), collapse = "\n"))

marks <- rules$marks$name[rules$marks$drawable]  # the drawn marks, kernel order
chans <- rules$channels                          # every channel, kernel order
cells <- rules$cells

# Five states, each a straight reading of the rule:
#   ●  a required position (Must)
#   ✅ a column maps here and the engine draws it today
#   ○  not mappable per row, but style() can set it once for the layer
#   ◌  legal grammar the renderer refuses today: owed or blocked; the refusal
#      message says which, and the legend below splits them
#   —  the mark has no such feature, or the grammar does not allow the pairing
glyph <- function(ob, renders, settable) {
  if (ob == "must")         "●"  # ●
  else if (!is.na(renders)) "✅"  # ✅
  else if (ob == "can")     "◌"  # ◌  grammar-legal, not drawn yet
  else if (settable)        "○"  # ○  set-only
  else                      "—"  # no such feature, or decided
}
lookup <- setNames(
  mapply(glyph, cells$obligation, cells$renders, cells$settable),
  paste(cells$mark, cells$channel)
)

glyph_grid("Mark", marks, chans, function(m, c) lookup[[paste(m, c)]])
```

Reading the five glyphs:

- **● must**: a required channel. The mark does not render without it. A
  `point` needs an `x` and a `y`; a `text` needs its `label`.
- **✅ maps**: bind a column here and the engine draws it. This is the
  ordinary case: the channel shows the column.
- **○ set**: the mark has this feature, but not row by row. A `line` is one
  stroke, so its `size` and `opacity` are a single value for the whole layer,
  fixed with [`style()`](style.qmd) rather than mapped. Those constants have
  their own [settings grid](style.qmd#what-you-can-set). Seven of them never
  appear on the channel grid at all: `border_color`, `border_size`, `caps`,
  `center`, `nudge`, `arrow` and `reach` are settings and nothing else. [Two
  questions, not one](style.qmd#two-questions-not-one) shows in one table where
  each of five properties can be mapped and where it can only be set.
- **◌ not drawn**: the engine refuses this pairing and gives you a
  direction instead of a wrong picture. The glyph covers two verdicts, and the
  refusal message tells you which you have. **Owed**: a mapped `size` or
  `opacity` on `text`, and `z` on `text`. **Blocked** on
  infrastructure: `z` on `rule` and on `zone`. Their 3-D forms are a plane and a
  slab with no single depth to sort by, so they need an engine that can tell what
  hides what.
- **— none**: the mark has no such feature, or the grammar does not allow the
  pairing. A `bar` has no glyph for `shape` to change, and a `point` has no line
  for `group` to divide. Either binding means nothing, and the engine says so.
  **Decided**: `z` on `line`, `step`, `area` and `ribbon`. These marks sort by
  `x`, and in a cube `x` can point into the page, so the order cannot be seen.
  The refusal points to `path` instead, which for an `area` or a `ribbon` draws
  only the edge.

With the glyphs read, the grid itself shows five patterns:

- **`group` splits a `line`, not a `bar`.** Read down the `group` column: `—` on
  the five marks that draw one shape per row (`point`, `bar`, `text`, `rule`,
  `edge`), ✅ on every mark that builds one shape out of many rows. A bar is
  split into groups by `color`, then piled or set side by side with `stack` /
  `dodge`. A bare `group` has no line to divide.
- **`label` belongs to `text` alone.** Its column is ● on `text` and `—`
  everywhere else. A labeled scatter is two layers, `point + text`, never
  a `label` given to the `point`.
- **Categories on `x` and on `y`.** A category on `x` reaches every mark that
  has positions except `surface`, and on `y` it reaches fewer of them. Read the
  `x` and `y` columns against each other. Which types each accepts is in
  [Channels](channels.qmd), and the split divides the other marks with positions
  in two. The marks that sit at one place (`point`, `text`) and the marks that
  stand in a slot and measure along the other axis (`bar`, `interval`, `box`)
  take either type on *both* positions. A point's or a label's two axes are the
  same kind, which is why it takes either type on both. That one rule is where a
  strip plot, a horizontal bar and a categorical box all come from. Three more
  marks do the same. A `path` sorts nothing, so it has no domain. A `rule`
  measures nothing on either axis. A `zone` measures nothing by length, and that
  is what makes a heatmap cell a zone rather than a second kind of bar. The
  stroke and region marks (`line`, `step`, `area`, `ribbon`) take either type on
  `x` alone. Their `x` is the domain the mark is read along, and their `y` is
  the quantity it traces. Those are two different kinds of axis. `surface`
  stands outside the split, because both of its
  positions measure. A sheet is a height, and the two positions say where that
  height is read.
- **Three rows have no ● at all.** `rule` and `zone` show ✅ on both positions
  and ● on neither, because no single cell can state what they need. A `rule`
  needs exactly one of the two, and the panel supplies the other
  ([Rule](marks/rule.qmd)). A `zone` has no fixed need: a categorical position,
  `bounds`, `bin`, `density` or a boundary on a map can each give it its sides
  ([Zone](marks/zone.qmd)). `edge` is the third, and it takes no position at all.
  Those obligations are stated once in the engine instead.
- **`zone` and `surface` take either type on `color`.** Of the five filled
  regions, `zone` is the only one that takes a number: `bar`, `box`, `area` and
  `ribbon` take a category only. `surface`, a sheet rather than a region, takes a
  number as well, and both read the ramp.

## Marks and transforms

The second crossing: not what a mark maps, but what it *derives*. A transform
enters through `*` (`bar * bin`, `interval * range`, `point * jitter`), and
which transform a mark takes is decided by the mark's geometry against the
transform's kind. There are three kinds, and each has its own grid below.

Each cell holds one of three glyphs:

- **✅ combines**: the transform applies to this mark and renders.
- **● required**: on `interval`, `ribbon` and `edge`, the mark does not draw
  until one of its ● transforms is present, so that transform is part of its
  [minimum syllable](design-laws.qmd#law-7-minimum-syllable). The `●` cells in a
  row are a choice, not a list: pick one. A `zone` is the exception. Each ● is
  one way to give it its sides, and a categorical position is another, so a
  `zone` can draw with no transform at all.
- **— none**: the mark has no use for this kind of transform, and the engine
  refuses it with a direction toward the transform its geometry does want.

```{r}
#| include: false
# One dump, read once; each grid below is a slice of it by transform class, so
# adding a transform widens the right grid on its own.
tmarks <- rules$marks$name[rules$marks$drawable]
tf     <- rules$transforms          # name, class
tcells <- rules$transform_cells     # transform, mark, state
tglyph <- c(required = "●", combines = "✅", none = "—")
tlook  <- setNames(unname(tglyph[tcells$state]), paste(tcells$transform, tcells$mark))
tf_block <- function(cls) {
  cols <- tf$name[tf$class == cls]
  glyph_grid("Mark", tmarks, cols, function(m, t) tlook[[paste(t, m)]])
}
```

### Statistics: a value per group

The first of the three kinds is also the largest. Most summarizing transforms
reduce the data to one value at each x: a bin's count, a group's `mean`, a
smoothed height. Any mark that draws a value at each x can draw that summary, so
those marks are ✅ under every summary. That is why a distribution has several
spellings rather than one histogram with a shape option: `bar * bin` is the
histogram, `line * bin` is the frequency polygon, and `step * bin` is the
staircase.

```{r}
#| echo: false
#| results: asis
tf_block("statistic")
```

Five marks (`point`, `line`, `area`, `bar`, `step`) take the whole family of
summaries. `interval`, `box` and `rule` take none, and each has its own reason.
An `interval` needs a *pair*, not a single value. A `box` carries its own summary.
Its five numbers come from the mark, not from a transform you compose, so no
statistic applies to it. A `rule` takes its length from the panel, so there is
no measure to compute.

Four more transforms build a *shape* rather than a summary, and they are the
only statistics `text` and `edge` take. `partition` reads a hierarchy (`zone`, `text`), `flow`
reads a path through stages (`ribbon`, `zone`, `text`), `layout` places a
network (`edge`, `point`, `text`), and `cluster` orders the categories of an axis
by similarity (`path`, `zone`). `flow` and `cluster` each have a chapter, [Flow](flow.qmd)
and [Cluster](cluster.qmd). `layout` is taught in [Network](network.qmd), and
`partition` in [Transforms](transforms.qmd#partition-a-whole-among-nested-parts).

### Ranges and bands

The second kind answers with two numbers instead of one. Five transforms produce
a *pair* of values, a low boundary and a high one. Most give one pair per group;
`smooth_band` gives one at each x position along its curve.
[`range`](transforms.qmd#range-a-lowhigh-pair),
[`confidence`](transforms.qmd#confidence-the-interval-of-the-mean),
[`deviation`](transforms.qmd#deviation-the-spread-of-the-data) and
[`smooth_band`](transforms.qmd#smooth_band-the-band-around-the-curve) compute the
pair from your data; [`bounds`](transforms.qmd#bounds-a-pre-computed-lowhigh-pair)
reshapes a pair you computed already. Four marks draw any such pair: `interval`
shows it as a whisker, `ribbon` fills it, and `line` and `step` trace its two
boundaries. A fifth, `zone`, reads a `bounds` pair as two sides.

```{r}
#| echo: false
#| results: asis
tf_block("pair")
```

The `●` cells on `interval` and `ribbon` are a requirement, not a permission.
An `interval` has no baseline to start from, so it does not render until a pair
transform gives it a low and a high. A bare `interval` is refused, and
`interval * range` with an `x` and a `y` is the whole minimum syllable. A
`ribbon` needs a pair too, or one of the two statistics that draw it as a band,
`density` and `flow`. On a `zone`, the `●` under `bounds` is one of the ways it
gets its sides. A `line` or a `step` already draws on its own, so the same
transform is optional there, a second way to use the mark. Everywhere
else there is nothing for the pair to draw. On a `point` or a `bar`, gog refuses
it and names the two span marks instead, `interval` and `ribbon`. On an `area`
it names `ribbon`, which is the `area` with two boundaries instead of a
baseline.

### Collision modifiers: an offset, not a statistic

When marks collide at one position, four offsets separate them, and which one a
mark takes is decided by its geometry and by the *axis* the collision is on.
On the marks that have a *width* (`bar`, `box`, `interval`), `dodge` subdivides
it. A `point` has no width, so `dodge` moves each point sideways
within its category's slot, only as far as it must to touch no other point: the
[beeswarm](marks/point.qmd#a-categorical-x-the-strip-plot). `stack`
piles marks along the *measure* axis. A `bar` or an `area` piles as lengths. A
`point` has no length of its own, so it piles as dots, which is the dot plot.
`jitter` also moves a point along a *categorical* axis, by a small amount that
varies from point to point, and it serves `point` alone. Which marks take `repel` is decided by a
different question: not what geometry a mark has, but what it is made of. A label is as
wide as the word it draws, so two labels overlap where their points do not, and
`text` is the only mark made of words.

```{r}
#| echo: false
#| results: asis
tf_block("collision")
```

Six marks take an offset. `bar` takes two, `point` takes three, and the other
four take one. `bar` has both a width and a measure axis, so it takes `dodge` and
`stack`. For a `point`, `stack` answers an overlap along the measure axis, which
is a pile of dots you can count. `jitter` and `dodge` answer the same overlap by
moving points along a categorical axis, which carries no magnitude, so no value
changes. `jitter` moves each point a small amount that varies from point to
point; `dodge` moves each only as far as it must to touch no other point. Each of the six
takes a known offset, so when you ask one of them for an offset it does not take,
the engine can name one it does: `area * dodge` names `stack`. A `box` shows most clearly that an offset is not a statistic. It takes no statistic, because its own is built in, yet it
still takes `dodge`. Setting grouped boxes side by side is a question about
position, and that is orthogonal to the summary each box already carries.

## Transforms and transforms

The third crossing is the only one where a mark does not appear. Two transforms
can follow one mark: `bar * bin * mean` cuts the axis into bands, then averages
inside each. Which pairs can do that is decided by the transforms alone, and by
one rule.

A transform has a **job**. There are four of them, and every transform in gog
fills at least one, except `repel`:

| job | the question it answers | the transforms that answer it |
|:--|:--|:--|
| **extent** | where are the cells? | `bin`, `density`, `smooth`, `smooth_band`, `partition`, and `bounds` on a `zone` |
| **measure** | what is in them? | `count`, `sum`, `mean`, `median`, `max`, `min`, `quantile`, `range`, `confidence`, `deviation`, `bounds` |
| **scale** | what scale is the answer read on? | `proportion`, and `stack(share = TRUE)` |
| **position** | where do the marks sit? | `dodge`, `stack`, `jitter` |
| **all four at once** | what is the whole picture? | `flow`, `layout`, `cluster` |

`repel` fills no job at all. It moves a label after every job has run, so it
contradicts nothing. Its row in the grid below is the only one that is ✅ in
every column, and that includes the other three offsets. `flow`, `layout` and
`cluster` fill all four jobs at once. Each computes the whole picture, so
`repel` is the only transform that can follow one.

Two transforms compose when they do different jobs, and contradict when they do
the same one. Each cell a transform computes (a bin, a group, a point on a curve)
holds one extent, one measure, one scale and one position. So a second transform
doing a job the first already did leaves the engine no reading that keeps both, and it refuses rather than drawing one and
discarding the other.

One transform is the exception, and it is the exception that makes chaining
useful. `bin` does two jobs: it cuts the axis, and it counts what lands in each
band. The count is a by-product of the cut, so `bin` does not have to keep it.
It hands the measure job to a transform that reduces a column you named, such
as `mean`. The other extent transforms do two jobs as well, and none of them can
hand one over. A `density` or
a `smooth` cell is a point on an estimated curve, which sits *between* the rows
rather than holding any of them, so there is nothing inside it to reduce. That
is why `bar * bin * mean` composes: `bin` supplies the cells and `mean` supplies
their value. And it is why `bar * bin * count` does not, since `count` was never
handed a column to reduce.

```{r}
#| echo: false
#| results: asis
# The chain grid, generated from the same predicate the engine refuses with, so it
# cannot claim a pair the engine would reject. Adding a transform widens it.
cc   <- rules$chain_cells
look <- setNames(ifelse(cc$legal, "✅", "—"), paste(cc$a, cc$b))
cols <- rules$transforms$name
glyph_grid("", cols, cols, function(a, b) if (a == b) "·" else look[[paste(a, b)]])
```

The grid is symmetric, and the rule is the reason. Whether two transforms
contradict is a fact about their jobs, and it has nothing to do with which one
you typed first. So the order is yours to choose. Each job has its own step in the engine,
extent first and position last. Write the transforms in that order, and the
sentence reads the way it runs. The `·` on the diagonal pairs a transform with
itself. Written twice, a transform repeats its own job, and the engine refuses
it like any other repeat. `repel` has no job to repeat, so it is the one
transform the engine lets you write twice.

Eleven transforms measure, so no two of them combine: read their rows against
each other, and every cell between two of them is `—`. That is why chains such
as `bar * count * mean`, `bar * mean * sum` and `interval * range * confidence`
are refused. Now
compare `bin`'s row with `smooth`'s and `density`'s. Those two chain with no
measure transform at all. `bin` chains with ten of the eleven, every one but
`count`. That is the exception above, written out in the grid.

Two contradictions do not fit the grid, because each depends on something
outside the pair. `stack(share = TRUE)` divides each pile by its own
total, which is a *scale*. `proportion` divides again, and a number divided
twice is not a share of anything. And `bounds` says where a `zone`'s sides are. On that one mark
it does the *extent* job, so it contradicts `bin`. On every other mark it names
the low and the high on the measure axis. The engine refuses both, and the grid shows the
common case: `stack` without `share`, and `bounds` on any mark but `zone`.

Chaining stops at four, and the number was not chosen. There are four jobs and each
takes at most one transform, so a fifth transform has to repeat a job on the
same mark. Repeating one is the contradiction this whole section is about.

### The shapes a long chain can take

The rule above allows chains of three and four, so the next question is how many
there are. The answer is a short list. Once each job is filled at most once, and
the marks are asked which transforms they take, every chain of three or four
takes one of the shapes below. Two slots stand for a family rather than a
transform: `<aggregation>` is any of `sum`, `mean`, `median`, `max`, `min`,
`quantile`, and `<collision modifier>` is `dodge`, `stack` or `jitter`,
whichever the mark takes.

```{r}
#| echo: false
#| results: asis
# The shapes, derived rather than listed: every chain of three or four the pair grid
# above allows on some mark, collapsed to the family each slot stands for. Nothing is
# typed, so a new transform widens this by itself, and it cannot claim a chain the
# engine refuses, because it reads the same cells the grid does.
# `quantile` belongs to the aggregation family the prose names, and leaving it out
# printed five extra rows that said the same thing five times. The pair transforms
# are deliberately *not* folded: only `deviation` chains with `bin` and
# `proportion`, so a family slot there would claim three chains the engine refuses.
fam <- c(sum = "<aggregation>", mean = "<aggregation>", median = "<aggregation>",
         max = "<aggregation>", min = "<aggregation>", quantile = "<aggregation>",
         dodge = "<collision modifier>", stack = "<collision modifier>",
         jitter = "<collision modifier>", repel = "<collision modifier>")
jobof  <- setNames(rules$transforms$jobs, rules$transforms$name)
rankof <- c(extent = 1, measure = 2, scale = 3, position = 4)
slot   <- function(t) if (t %in% names(fam)) unname(fam[t]) else t
# A transform's plain job, the first the engine lists, which is the one it does
# whatever the mark and settings are. `bin` fills extent and measure and hands the
# second over; `stack` rescales only when given `share = TRUE`. Taking the first is
# what keeps `stack` a position transform here rather than a scale one.
jobt   <- function(t) jobof[[t]][1]
tlegal <- setNames(rules$transform_cells$state,
                   paste(rules$transform_cells$transform, rules$transform_cells$mark))
pairok <- setNames(rules$chain_cells$legal, paste(rules$chain_cells$a, rules$chain_cells$b))
shapes <- list()
for (m in rules$marks$name[rules$marks$drawable]) {
  take <- Filter(function(t) tlegal[[paste(t, m)]] != "none", rules$transforms$name)
  for (k in c(3, 4)) {
    if (length(take) < k) next
    combos <- combn(take, k, simplify = FALSE)
    for (C in combos) {
      if (!all(apply(combn(C, 2), 2, function(p) pairok[[paste(p[1], p[2])]]))) next
      # `base::order`, because gog exports its own `order()`, which masks base R's.
      C <- C[base::order(vapply(C, function(t) rankof[[jobt(t)]], 1))]
      key <- paste(unique(vapply(C, slot, "")), collapse = " * ")
      if (is.null(shapes[[key]])) shapes[[key]] <- list(n = k, jobs = vapply(C, jobt, ""))
    }
  }
}
ord <- base::order(vapply(shapes, function(s) s$n, 1), names(shapes))
cat("| how many | the shape | the jobs it fills |",
    "|:--:|:--|:--|",
    vapply(ord, function(i) sprintf("| %d | `%s` | %s |",
      shapes[[i]]$n, names(shapes)[i],
      paste(unique(shapes[[i]]$jobs), collapse = ", ")), ""),
    sep = "\n")
```

Read the four-transform row first, because it is the longest chain there is. It
fills each job exactly once, and two of its four slots are decided before you
choose anything. `bin` is the only extent transform that gives its measurement
to another, and `proportion` is the only transform whose whole job is to
rescale. That leaves two choices: the aggregation, which is free among six, and
the offset, which the mark narrows to one or two. So chains of four exist, but
with two slots fixed there are few of them.

The shapes say what the *transforms* allow. A mark can still refuse a chain
whose every pair is ✅, for a reason about its own geometry, and no grid shows
it. `point * proportion * stack` is the clearest case. A dot
plot piles one whole dot per observation, and a share is less than one dot. So
the chain is refused on `point` and drawn on `bar`.

## Marks and spaces

The fourth crossing is not about the mark itself. It is the coordinate
space the whole plot sits in, and it asks a different question of every mark.
Once the plane is bent, or extended into a cube, can this geometry still be
drawn? A `polar()` bar is a wedge, a `polar()` line runs around the center, and a `z`
scatter is a projected cloud, all from the same sentences that draw them flat.

Two glyphs here, not five. Either the engine draws this mark in this space,
or it does not and says so:

- **✅ drawn**: the mark renders in this space.
- **— not drawn**: refused with direction, never half-drawn. Here `—` has
  three verdicts: owed, blocked and decided. On the channel grid, owed and
  blocked show as ◌, and decided shows as `—`. A few blanks are simpler: the mark
  needs what the space lacks, as a `surface` needs the cube. The paragraphs under
  the table say why each column's blanks are blank.

```{r}
#| echo: false
#| results: asis
spaces <- rules$spaces
scells <- rules$space_cells
slook  <- setNames(ifelse(scells$drawn, "✅", "—"), paste(scells$space, scells$mark))
glyph_grid("Mark", marks, spaces, function(m, s) slook[[paste(s, m)]])
```

`flat` holds every mark except `surface` and `edge`. Read `surface` first: it
does not draw in the plane at all. Five of the marks that draw there also draw
in the cube. A surface is a sheet through three positions, and
there is no such thing without the third, so its minimum syllable includes the
cube. It is the only mark that draws in the cube and not in the plane
([Surface](marks/surface.qmd)). An `edge` is the other kind of blank, and it is
blank in every column but the last. It takes no position at all, so it draws
only in `network()`, the space that draws the places `layout` computes. In the columns that
follow, ignore the `edge` row: every blank in it has this one reason.

`space` holds six marks, and one rule explains all six: each means the same
thing in the cube as on the page. A `point` is a place, a `path` is its rows in
order, a `surface` is a sheet. `bar`, `interval` and `box` stand in a slot and
measure along the other axis. A bar gives a length, an interval a span, a box a
summary, and the grammar treats the three as one family.

The blanks in that column are not one kind of thing, and the refusal message
says which each is. `line`, `step`, `area` and `ribbon` read a *domain* left to
right, and a cube has no left to right. So they are **decided**, a ruling rather
than work still owed, and the refusal points to `path` instead. `rule` and
`zone` are **blocked**, the same verdict their `z` cells carry. A plane and a
slab have no single depth to sort by, so placing them among the data needs an
engine that can tell what hides what ([Space](space.qmd)). `text` is **owed**: the
engine does not draw a label in the cube, and the message says so rather than
drawing one flat ([Space](space.qmd)).

`polar` matches `flat`: the same twelve marks draw, and the same two do not.
Every mark that draws on the page also draws in the circle. Bending the plane
changes only how a mark's outline is drawn: a held value becomes an arc, and a
line between two points becomes a straight chord. A `ribbon`'s boundaries are
the same chords a `line` draws. The two blanks are `surface` and `edge`, and
neither is a polar refusal: a sheet needs the cube, and an edge needs
`network()` ([Polar](polar.qmd)).

`nest` is a column of the opposite shape: two cells filled and the rest blank.
Those blanks are not unfinished work. This space asks a different
question from `flat`, `space`, `polar`, `globe` and `map`. In those five, a mark
draws if its geometry survives what the space does to the plane. A nesting is not a transformation of the
plane at all, so it can draw any mark that accepts *a rectangle*. A `bar`
accepts one, because a bar is a measure. A measure has been a length on the page
and a radius or an angle in the circle, and here it becomes the size of a
rectangle. `text`
can take one too, and it is the only mark here that names a region rather than
filling one. The rectangle is already computed, so the label goes to its center.
That is what a published treemap shows. A legend cannot name a hundred regions,
so the label has to sit in the rectangle ([Nest](nest.qmd)).

Apart from `edge`, the blanks share one reason: they are the marks placed by a
*position*. `point` needs somewhere to sit. `line`, `step`, `area`, `ribbon` and
`path` join places in order, and two touching regions here are not near each
other in the data. `rule` and `zone` span an axis that is not there, and `box`,
`interval` and `surface` measure along one. `text` is missing from that list because no position places it here. The region places
it, and the region is what places the bar too.

`map`, the sixth column, fills five cells, and the shape of its column comes
from one fact: longitude and latitude use both positions for the place. The
marks that draw are the ones that place something. A `point` is a place, a
`path` is a route, and a `text` is a name at a place. A `rule` at one position
spans the other, which makes it a meridian or a parallel. A `zone` takes its
sides from a boundary, which is the choropleth ([Map](map.qmd)). The blanks
share one reason: each of them needs an axis of its own, either to measure along
or to read a domain along, and a map has neither to give. The quantity
moves to `size` or `color` instead, which is what cartography does with a
measure once both positions are gone.

`globe`, the column before it, fills six cells. Five of them are `map`'s five,
and the sixth is one a flat map cannot have. A globe and a map answer
one question: where a place is. So the marks that place something draw on both,
and only the reader's view changes. The far half of the earth is
hidden behind the sphere. A route bends along a great circle, and a rule draws
the whole meridian or parallel it sits on. The sixth cell is `bar`, because only
the sphere leaves one direction free: the radius, which runs straight out from
the surface at every place. A bar drawn along it is a spike, with `z` naming its
measure ([Globe](globe.qmd)). The blanks are `map`'s, minus that `bar`, and they
are blank for the same reason. A sphere has no axis of its own to give them
either.

`network` holds three cells, and one of its marks draws in no other space.
`edge` draws nowhere else, because an edge is placed by neither position:
`layout` computes where every node sits, `network()` draws those places, and the
edge joins two of them ([Network](network.qmd)). `point` and `text` join it there, as the nodes and
their names. The blanks are the marks a `layout` does not place.

That is the last column. Three of the four crossings have a mark on every row,
so one mark can also be read across all three at once. A path takes nearly
every channel on Mark × Channel and almost nothing on Mark × Transform, because a path *is* its rows in order
and every statistic replaces them. `rule` reads the same way on the transform
grid, and for the reason the statistics grid already gave: it takes its length
from the panel, so it has no measure to compute. And `surface`'s row on the
statistics grid holds the transforms that cover a plane. `density` estimates a
value at every point of a grid and `bin` cuts adjacent cells, while `sum`,
`mean`, `median`, `max` and `min` reduce the rows in each cell to one height. `path` derives two things. With `density` it draws
the contour of a field, the same field a `surface` lifts into a sheet. With
`cluster` it draws the dendrogram, the tree that joins similar categories.

With these four crossings and [the settings grid](style.qmd#what-you-can-set),
the orthogonality law is stated in full: what each mark maps, what it sets, what
it derives, which transforms can chain on it, and where it can be drawn.
