# Combinations

```{r}
#| include: false
source("R/setup.R")
```

Every mark down the side, every channel across the top: the whole of what the
grammar lets you combine, on one page. This page holds four crossings. The first
is what a mark lets you *map*, which is the channels. The second is what a mark
lets you *derive*, which is the transforms. The third is which two transforms
can *stand together* on one mark. The fourth is where a mark can be *drawn*,
which is the coordinate spaces. Each one shows what a mark takes, what it does
not, and the surprising pairings that are legal anyway.

These grids are the orthogonality law made visible. [Law 1](design-laws.qmd)
says every compatible atom combines with every other; a table is that promise
enumerated, one cell at a time. It is the direct analog of the letter chart a
Hangeul (한글) learner reads, with the consonants down one side and the vowels
across the top. A syllable block is what those letters assemble into. This book
has used that pairing from the start: a mark is a consonant and a position is a
vowel, and neither draws alone. Every syllable the grammar can form appears
somewhere in the crossing.

Legal is not the same as useful. [Law 8](design-laws.qmd), Pronounceable ≠
Usable, guarantees only that the engine draws what a cell allows. Whether it
should be drawn is your judgment. A cell the engine refuses teaches as much as
one it draws. It records a real limit of the geometry, not an oversight.

Every grid here is generated live from the engine's own rule table, each time
the book builds. That is the same table the engine consults to accept or refuse
your plot. A grid that is *described* can stop matching the code;
a grid that is *generated* cannot. Add a mark, a channel, a transform or a space
and these pages widen by themselves.

## Marks and channels

Take the channels first. Each cell says whether you may map a column to that
channel on that mark.

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules`, which dumps `rule_for` (the one legality
# table) as JSON. The glyph for each cell is read straight off that verdict, so
# the grid can never claim something the engine would not enforce.
cli   <- find_gog_cli()
rules <- jsonlite::fromJSON(paste(system2(cli, "--rules", stdout = TRUE), collapse = "\n"))

marks <- rules$marks$name[rules$marks$drawable]  # the drawn marks, kernel order
chans <- rules$channels                          # every channel, kernel order
cells <- rules$cells

# Five states, each a straight reading of the rule:
#   ●  a required position (Must)
#   ✅ a column maps here and the engine draws it today
#   ○  not mappable per row, but style() can set it once for the layer
#   ◌  legal grammar the renderer refuses today: owed, blocked, or decided
#      against; the refusal message says which, and the legend below splits them
#   —  the mark has no such feature at all
glyph <- function(ob, renders, settable) {
  if (ob == "must")         "●"  # ●
  else if (!is.na(renders)) "✅"  # ✅
  else if (ob == "can")     "◌"  # ◌  grammar-legal, not drawn yet
  else if (settable)        "○"  # ○  set-only
  else                      "—"  # no such feature
}
lookup <- setNames(
  mapply(glyph, cells$obligation, cells$renders, cells$settable),
  paste(cells$mark, cells$channel)
)

header  <- paste0("| Mark | ", paste0("`", chans, "`", collapse = " | "), " |")
divider <- paste0("|:--|", paste(rep(":--:", length(chans)), collapse = "|"), "|")
rows <- vapply(marks, function(m) {
  gs <- vapply(chans, function(c) lookup[[paste(m, c)]], character(1))
  paste0("| `", m, "` | ", paste(gs, collapse = " | "), " |")
}, character(1))
cat(header, divider, rows, sep = "\n")
```

Reading the five glyphs:

- **● must**: a required position. The mark does not render without it. A
  `point` needs an `x` and a `y`; a `text` needs its `label`.
- **✅ maps**: bind a column here and the engine draws it today. This is the
  ordinary case, a channel doing its job.
- **○ set**: the mark has this feature, but not row by row. A `line` is one
  stroke, so its `size` and `opacity` are a single value for the whole layer,
  fixed with [`style()`](style.qmd) rather than mapped. Those constants have
  their own [Mark × Setting grid](style.qmd). Seven of them never appear on this
  grid at all: `border_color`, `border_size`, `caps`, `center`, `nudge`,
  `arrow` and `reach` are settings and nothing else. [Two questions,
  not one](style.qmd#two-questions-not-one) puts both halves in one table.
- **◌ not drawn**: legal grammar the engine refuses today. It gives you a
  direction instead of a wrong picture. One glyph, three verdicts, and the
  refusal message tells you which you have. **Owed**: a mapped `size` or
  `opacity` on `text`, `z` on `text`, and `pattern` on `edge`. **Blocked** on
  infrastructure: `z` on `rule` and on `zone`. Their 3-D forms are a plane and a
  slab with no footprint to sort by, so they need an engine that can tell what
  hides what. **Decided**: `z` on `line`, `step`, `area` and `ribbon`. These
  marks read a domain from left to right, and a cube has no left to right, so
  the refusal points to `path` instead. The Mark × Space grid below is the same
  three readings on its own blanks.
- **— none**: the mark has no such feature. A `bar` has no glyph to `shape`, and
  a `point` has no line to `group` into. Binding it means nothing, and the
  engine says so.

Five things the grid shows that prose has to say one at a time:

- **`group` splits a `line`, not a `bar`.** Read down the `group` column: `—` on
  the five marks that draw one shape per row (`point`, `bar`, `text`, `rule`,
  `edge`), ✅ on every mark that builds one shape out of many rows. A bar is
  split into groups by `color`, then piled or set side by side with `stack` /
  `dodge`. A bare `group` has no line to divide.
- **`label` belongs to `text` alone.** Its column is ● on `text` and `—`
  everywhere else. A labeled scatter is the superposition `point + text`, never
  a `label` hung on the `point`.
- **Categories on `x` and on `y`.** A category on `x` reaches every mark that
  has positions except `surface`, and on `y` it reaches fewer of them. Read the
  `x` and `y` columns against each other. Which types each accepts is in
  [Channels](channels.qmd), and the split divides the marks with positions in
  two. The glyph marks (`point`, `text`) and the marks whose slot width does the
  measuring (`bar`, `interval`, `box`) take either type on *both* positions.
  That one rule is where a strip plot, a horizontal bar and a categorical box
  all come from. Three more marks do the same. A `path` sorts nothing, so it has
  no domain. A `rule` measures nothing on either axis. A `zone` measures nothing
  by length, and that is what makes a heatmap cell a zone rather than a second
  kind of bar. The path and region marks (`line`, `step`, `area`, `ribbon`) take
  either type on `x` alone. Their `x` is the domain the path is read along, and
  their `y` is the quantity it traces. Those are two different kinds of axis. A
  glyph's two axes are the same kind, which is why a `point` and a `text` take
  either type on both. `surface` stands outside the split, because both of its
  positions measure. A sheet is a height, and the two positions say where that
  height is read.
- **Three rows have no ● at all.** `rule` and `zone` show ✅ on both positions
  and ● on neither, because each needs *one* of the two and no per-channel cell
  can say "either, but not both". `edge` is the third, and it takes no position
  at all. Those obligations are stated once in the engine instead. Two ✅s on the
  positions and no ● means the mark needs one of them, and the panel supplies
  the other ([Rule](marks/rule.qmd), [Zone](marks/zone.qmd)).
- **`zone` and `surface` take either type on `color`.** They are the two fills
  that do: every other region's fill takes a category, and these two take a
  number as well, reading the ramp.

## Marks and transforms

The second crossing: not what a mark maps, but what it *derives*. A transform
enters through `*` (`bar * bin`, `interval * range`, `point * jitter`), and
which transform a mark takes is decided by the mark's geometry against the
transform's kind. There are three kinds, and each has its own grid below.

Three glyphs read every cell:

- **✅ combines**: the transform applies to this mark and renders.
- **● required**: the mark will not draw until a transform of this kind is
  present, so that kind is part of its minimum syllable. An `interval` and a
  `ribbon` span a low value and a high one, so neither draws until *one*
  transform of that kind supplies the pair. The `●` cells in a row are a choice,
  not a list: pick one. A `zone` shows ● wherever a transform is one of the ways
  it gets its sides.
- **— none**: the mark has no use for this kind of transform, and the engine
  refuses it with a direction toward the transform its geometry does want.

```{r}
#| include: false
# One dump, read once; each grid below is a slice of it by transform class, so
# adding a transform widens the right grid on its own.
tmarks <- rules$marks$name[rules$marks$drawable]
tf     <- rules$transforms          # name, class
tcells <- rules$transform_cells     # transform, mark, state
tglyph <- c(required = "●", combines = "✅", none = "—")
tlook  <- setNames(unname(tglyph[tcells$state]), paste(tcells$transform, tcells$mark))
tf_block <- function(cls) {
  cols    <- tf$name[tf$class == cls]
  header  <- paste0("| Mark | ", paste0("`", cols, "`", collapse = " | "), " |")
  divider <- paste0("|:--|", paste(rep(":--:", length(cols)), collapse = "|"), "|")
  rows <- vapply(tmarks, function(m) {
    gs <- vapply(cols, function(t) tlook[[paste(t, m)]], character(1))
    paste0("| `", m, "` | ", paste(gs, collapse = " | "), " |")
  }, character(1))
  cat(header, divider, rows, sep = "\n")
}
```

### Statistics: a value per group

The first of the three kinds is also the largest. Most summarizing transforms
reduce the data to one value at each x: a bin's count, a group's `mean`, a
smoothed height. Any mark that draws a value at each x can draw that summary, so
those marks are ✅ under every summary. That is why a distribution has three
spellings rather than one histogram with a shape option: `bar * bin` fills it,
`line * bin` is the frequency polygon, `step * bin` the staircase.

```{r}
#| echo: false
#| results: asis
tf_block("statistic")
```

Five marks (`point`, `line`, `area`, `bar`, `step`) take the whole family of
summaries. `interval`, `box` and `rule` take none, and each has its own reason.
A span mark needs a *pair*, not a single value. A `box` carries its own summary:
the five numbers come from the mark itself rather than from a transform you
compose, so no statistic column applies to it. A `rule` hands a whole axis to
the panel, so there is no measure to compute.

Four more transforms build a *shape* rather than a summary, and they fill most
of the rows in between. `partition` reads a hierarchy (`zone`, `text`), `flow`
reads a path through stages (`ribbon`, `zone`, `text`), `layout` places a
network (`edge`, `point`, `text`), and `cluster` orders leaves by similarity
(`path`, `zone`). Each has a chapter of its own.

### Ranges and bands

The second kind answers with two numbers instead of one. Four transforms produce
a *pair* of values, a low boundary and a high one, per group.
[`range`](transforms.qmd#range-a-lowhigh-pair),
[`confidence`](transforms.qmd#confidence-the-interval-of-the-mean) and
[`deviation`](transforms.qmd#deviation-the-spread-of-the-data) compute the pair
from your data; [`bounds`](transforms.qmd#bounds-a-pre-computed-lowhigh-pair)
reshapes a pair you computed already. Five marks draw such a pair: `interval`
shows it as a whisker, `ribbon` fills it, `line` and `step` trace its two
boundaries, and `zone` reads it as two sides.

```{r}
#| echo: false
#| results: asis
tf_block("pair")
```

The `●` cells on `interval`, `ribbon` and `zone` are a requirement, not a
permission. An `interval` and a `ribbon` have no baseline to start from, so
neither renders until a pair transform gives it a low and a high. `interval *
range` is the whole minimum syllable, and a bare `interval` is refused. A `line`
or a `step` already draws fine on its own, so the same transform is optional
there, a second way to use the mark. Everywhere else there is nothing for the
pair to draw. On a `point` or a `bar` it is refused toward the whisker, and on
an `area` toward the `ribbon`, which is the `area` with two boundaries instead
of a baseline.

### Collision modifiers: an offset, not a statistic

When marks collide at one position, four offsets separate them, and which one a
mark takes is decided by its geometry and by the *axis* the collision is on.
`dodge` subdivides a *width*, so it serves the marks that have one (`bar`,
`box`, `interval`). `stack` accumulates along the *measure* axis, so it serves
the marks that can spend a span. `bar` and `area` fill that span with length. A
`point` has no length of its own, so it spends the span on counting dots
instead, which is the dot plot. `jitter` spreads a mark with *no* width along a
*categorical* axis, which is `point` alone. `repel` divides on a different
question: not what geometry a mark has, but what it is made of. A label is as
wide as the word it draws, so two labels overlap where their points do not, and
`text` is the only mark made of words.

```{r}
#| echo: false
#| results: asis
tf_block("collision")
```

Six marks take an offset, and two of them take two. `bar` has both a width and a
measure axis, so it takes `dodge` and `stack`. A `point` has no width, so
`dodge` is out, and it takes `stack` and `jitter` instead. For a `point`,
`stack` answers an overlap along the measure axis, which is a pile of dots you
can count. `jitter` answers the same overlap along a categorical axis, which is
a nudge, because a categorical axis carries no magnitude. Everywhere else the
division is exclusive, which is why the engine can always name the right
transform when you ask for the wrong one. Ask for `point * dodge` and it names
`jitter`; ask for `area * dodge` and it names `stack`. A `box` shows the split
at its clearest. It takes no statistic, because its own is built in, yet it
still takes `dodge`. Setting grouped boxes side by side is a question about
position, and that is orthogonal to the summary each box already carries.

## Transforms and transforms

The third crossing is the only one where a mark does not appear. Two transforms
can follow one mark: `bar * bin * mean` cuts the axis into bands, then averages
inside each. Which pairs can do that is decided by the transforms alone, and by
one rule.

A transform has a **job**. There are four of them, and every transform in gog
fills at least one, except `repel`:

| job | the question it answers | the transforms that answer it |
|:--|:--|:--|
| **extent** | where are the cells? | `bin`, `density`, `smooth`, `partition`, and `bounds` on a `zone` |
| **measure** | what is in them? | `count`, `sum`, `mean`, `median`, `max`, `min`, `quantile`, `range`, `confidence`, `deviation`, `bounds` |
| **scale** | what scale is the answer read on? | `proportion`, and `stack(share = TRUE)` |
| **position** | where do the marks sit? | `dodge`, `stack`, `jitter` |

`repel` fills no job at all. It moves a label after every job has run, so it
contradicts nothing. Its row in the grid below is the only one that is ✅ from
end to end, and that includes the other three offsets. `flow`, `layout` and
`cluster` fill all four jobs at once. Each computes the whole picture, so
`repel` is the only transform that can follow one.

Two transforms compose when they do different jobs, and contradict when they do
the same one. Each of those cells holds one extent, one measure, one scale and
one position. So a second transform doing a job the first already did leaves the
engine no reading that keeps both, and it refuses rather than drawing one and
discarding the other.

One transform is the exception, and it is the exception that makes chaining
useful. `bin` does two jobs: it cuts the axis, and it counts what lands in each
band. The count is a by-product of the cut, so `bin` does not have to keep it.
It hands the measure job to any transform you gave a *column*. The other extent
transforms do two jobs as well, and none of them can do the same. A `density` or
a `smooth` cell is a point on an estimated curve, which sits *between* the rows
rather than holding any of them, so there is nothing inside it to reduce. That
is why `bar * bin * mean` composes: `bin` supplies the cells and `mean` supplies
their value. And it is why `bar * bin * count` does not, since `count` was never
handed a column to reduce.

```{r}
#| echo: false
#| results: asis
# The chain grid, generated from the same predicate the engine refuses with, so it
# cannot claim a pair the engine would reject. Adding a transform widens it.
cc   <- rules$chain_cells
look <- setNames(ifelse(cc$legal, "✅", "—"), paste(cc$a, cc$b))
cols <- rules$transforms$name
cat(paste0("| | ", paste0("`", cols, "`", collapse = " | "), " |"),
    paste0("|:--|", paste(rep(":--:", length(cols)), collapse = "|"), "|"),
    vapply(cols, function(a) {
      gs <- vapply(cols, function(b) if (a == b) "·" else look[[paste(a, b)]], character(1))
      paste0("| `", a, "` | ", paste(gs, collapse = " | "), " |")
    }, character(1)),
    sep = "\n")
```

The grid is symmetric, and the reason is in the rule rather than in the drawing.
Whether two transforms contradict is a fact about their jobs, and it has nothing
to do with which one you typed first. So the order is yours to choose. Write the
transforms in the order they run, and the sentence reads in that order too.

Read the block of `—` in the middle of the grid. Eleven transforms measure, so
no two of them combine. That block is the whole reason `bar * count * mean` is
refused, and `bar * mean * sum`, and `interval * range * confidence`. Now
compare `bin`'s row with `smooth`'s and `density`'s. Those two chain with no
measure transform at all, and `bin` chains with ten. That is the exception
above, written out in the grid.

Two collisions the grid cannot show, because both depend on something outside
the pair. `stack(share = TRUE)` divides each pile by its own total, which is a
*scale*. `proportion` divides again, and the second division has no whole left
to be a share of. And `bounds` says where a `zone`'s sides are. On that one mark
it does the *extent* job, so it contradicts `bin`. On every other mark it names
the low and the high on the measure axis. The engine reads both cases correctly,
and the grid above shows the reading that applies everywhere else.

Chaining stops at four, and nobody chose the number. There are four jobs and one
transform fills each of them, so a fifth transform has to repeat a job on the
same mark. Repeating one is the contradiction this whole section is about.

### The shapes a long chain can take

The rule above allows chains of three and four, so it is fair to ask how many
there are. The answer is a short list. Once each job is filled at most once, and
the marks are asked which transforms they take, every chain of three or four
takes one of the shapes below. Two slots stand for a family rather than a
transform: `<aggregation>` is any of `sum`, `mean`, `median`, `max`, `min`,
`quantile`, and `<collision modifier>` is `dodge`, `stack` or `jitter`,
whichever the mark takes.

```{r}
#| echo: false
#| results: asis
# The shapes, derived rather than listed: every chain of three or four the pair grid
# above allows on some mark, collapsed to the family each slot stands for. Nothing is
# typed, so a new transform widens this by itself, and it cannot claim a chain the
# engine refuses, because it reads the same cells the grid does.
# `quantile` belongs to the aggregation family the prose names, and leaving it out
# printed five extra rows that said the same thing five times. The pair transforms
# are deliberately *not* folded: only `deviation` chains with `bin` and
# `proportion`, so a family slot there would claim three chains the engine refuses.
fam <- c(sum = "<aggregation>", mean = "<aggregation>", median = "<aggregation>",
         max = "<aggregation>", min = "<aggregation>", quantile = "<aggregation>",
         dodge = "<collision modifier>", stack = "<collision modifier>",
         jitter = "<collision modifier>", repel = "<collision modifier>")
jobof  <- setNames(rules$transforms$jobs, rules$transforms$name)
rankof <- c(extent = 1, measure = 2, scale = 3, position = 4)
slot   <- function(t) if (t %in% names(fam)) unname(fam[t]) else t
# A transform's plain job, the first the engine lists, which is the one it does
# whatever the mark and settings are. `bin` fills extent and measure and hands the
# second over; `stack` rescales only when given `share = TRUE`. Taking the first is
# what keeps `stack` a position transform here rather than a scale one.
jobt   <- function(t) jobof[[t]][1]
tlegal <- setNames(rules$transform_cells$state,
                   paste(rules$transform_cells$transform, rules$transform_cells$mark))
pairok <- setNames(rules$chain_cells$legal, paste(rules$chain_cells$a, rules$chain_cells$b))
shapes <- list()
for (m in rules$marks$name[rules$marks$drawable]) {
  take <- Filter(function(t) tlegal[[paste(t, m)]] != "none", rules$transforms$name)
  for (k in c(3, 4)) {
    if (length(take) < k) next
    combos <- combn(take, k, simplify = FALSE)
    for (C in combos) {
      if (!all(apply(combn(C, 2), 2, function(p) pairok[[paste(p[1], p[2])]]))) next
      # `base::order`, because `order()` is a gog channel and the grammar masks it.
      C <- C[base::order(vapply(C, function(t) rankof[[jobt(t)]], 1))]
      key <- paste(unique(vapply(C, slot, "")), collapse = " * ")
      if (is.null(shapes[[key]])) shapes[[key]] <- list(n = k, jobs = vapply(C, jobt, ""))
    }
  }
}
ord <- base::order(vapply(shapes, function(s) s$n, 1), names(shapes))
cat("| how many | the shape | the jobs it fills |",
    "|:--:|:--|:--|",
    vapply(ord, function(i) sprintf("| %d | `%s` | %s |",
      shapes[[i]]$n, names(shapes)[i],
      paste(unique(shapes[[i]]$jobs), collapse = ", ")), ""),
    sep = "\n")
```

Read the four-transform row first, because it is the longest chain there is. It
fills each job exactly once, and two of its four slots are decided before you
choose anything. `bin` is the only extent transform that gives its measurement
to another, and `proportion` is the only one that rescales. That leaves two
choices: the aggregation, which is free among six, and the offset, which the
mark narrows to one or two. Four is reachable, and this is why it is rare.

The shapes say what the *transforms* allow. A mark can still refuse one of them
for a reason about its own geometry, and the grids earlier in this chapter are
where that is decided. `point * proportion * stack` is the clearest case. A dot
plot piles one whole dot per observation, and a share is less than one dot. So
the chain is refused on `point` and drawn on `bar`.

## Marks and spaces

The fourth crossing is not a channel of the mark at all. It is the coordinate
space the whole plot sits in, and it asks a different question of every mark.
Once the plane is bent, or turned into a cube, can this geometry still be
drawn? A `polar()` bar is a wedge, a `polar()` line is a closed curve, and a `z`
scatter is a projected cloud, all from the same sentences that draw them flat.

Two glyphs here, not five. Either the engine draws this mark in this space
today, or it does not and says so:

- **✅ drawn**: the mark renders in this space.
- **— not drawn**: legal to write and refused with direction, never half-drawn.
  As on the channel grid, one glyph stands for owed, blocked and decided alike.
  The paragraphs under the table say which one each blank is.

```{r}
#| echo: false
#| results: asis
spaces <- rules$spaces
scells <- rules$space_cells
slook  <- setNames(ifelse(scells$drawn, "✅", "—"), paste(scells$space, scells$mark))
header  <- paste0("| Mark | ", paste0("`", spaces, "`", collapse = " | "), " |")
divider <- paste0("|:--|", paste(rep(":--:", length(spaces)), collapse = "|"), "|")
rows <- vapply(marks, function(m) {
  gs <- vapply(spaces, function(s) slook[[paste(s, m)]], character(1))
  paste0("| `", m, "` | ", paste(gs, collapse = " | "), " |")
}, character(1))
cat(header, divider, rows, sep = "\n")
```

`flat` is full but for `surface` and `edge`, and `surface` is the one to read
first: it does not draw in the plane at all. Every other mark draws there, and
some also draw in the cube. A surface is a sheet through three positions, and
there is no such thing without the third, so its minimum syllable includes the
cube. It is the only mark that draws in the cube and not in the plane
([Surface](marks/surface.qmd)). An `edge` is the other kind of blank, and it is
blank in every column but the last. It takes no position at all, so the only
space that can draw it is the one that places its nodes. Read the columns that
follow with `edge` set aside.

`space` holds six marks, and the six follow from one reading rather than from a
list. A
`point` is a place, a `path` is an order, a `surface` is a sheet. `bar`,
`interval` and `box` are the three marks that stand in a slot and measure along
the other axis: a length, a span and a summary, which the grammar treats as one
family. Each means in a cube exactly what it meant on the page.

The blanks in that column are not one kind of thing, and the refusal message
says which each is. `line`, `step`, `area` and `ribbon` read a *domain* left to
right, and a cube has no left to right. So they are **decided**, a ruling rather
than work still owed, and the refusal points to `path` instead. `rule` and
`zone` are **blocked**, the same reading their `z` cells carry. A plane and a
slab have no footprint to sort by, so placing them among the data needs an
engine that can tell what hides what ([Space](space.qmd)). `text` is **owed**: a
label in the cube is not built yet, and the message says so rather than drawing
one flat ([Space](space.qmd)).

`polar` matches `flat`: the same twelve marks draw, and the same two do not.
Every mark that draws on the page also draws in the circle. Bending the plane
only curves a mark's edges, and every flat mark has edges that can curve. A
band's boundaries are the same chords a `line` draws. The two blanks are
`surface` and `edge`, and neither is a polar refusal: a sheet needs the cube,
and an edge needs the space that places its nodes ([Polar](polar.qmd)).

`nest` is a column of the opposite shape: two cells filled and the rest blank.
Those blanks are not work waiting to be done. This space asks a different
question from the other five. Everywhere else a mark bends because its geometry
survives a transformation of the plane. A nesting is not a transformation of the
plane at all, so it can draw any mark that accepts *a rectangle*. A `bar`
accepts one, because a bar is a measure. A measure has been a length on the page
and an angle in the circle, and here it becomes the size of a rectangle. `text`
can take one too, and it is the only mark here that names a region rather than
filling one. The rectangle is already computed, so the label goes to its center.
That is what a published treemap has, and what a legend cannot do once the split
runs past a hundred regions ([Nest](nest.qmd)).

The blanks share one reason, and it is not that the space is unfinished. They
are the marks placed by a *position*. `point` needs somewhere to sit. `line`,
`step`, `path`, `area` and `ribbon` need an order along an axis, and two
touching regions here are not near each other in the data. `rule` and `zone`
span an axis that is not there, and `box`, `interval` and `surface` measure
along one. `edge` is placed by a layout that this space does not compute. `text`
is missing from that list because no position places it here. The region places
it, and the region is what places the bar too.

`map`, the sixth column, fills five cells, and the shape of its column comes
from one fact: longitude and latitude spend both positions on the place. The
marks that draw are the ones that place something. A `point` is a place, a
`path` is a route, and a `text` is a name at a place. A `rule` at one position
spans the other, which makes it a meridian or a parallel. A `zone` takes its
sides from a boundary, which is the choropleth ([Map](map.qmd)). The blanks
share one reason: each of them needs an axis of its own, either to measure along
or to read a domain along, and a map has neither to give. The quantity
moves to `size` or `color` instead, which is what cartography does with a
measure once both positions are gone.

`globe`, the column before it, fills six cells. Five of them are `map`'s five,
and the sixth is one a flat map cannot have. The sphere and its flattening
answer one question: where does a place go. So the marks that place something
stand on both, and only the reader's view changes. The far half of the earth is
hidden behind the sphere, a route bends along a great circle, and a rule becomes
the whole meridian or parallel it holds. The sixth cell is `bar`, because the
sphere alone has a direction to spare: the radius, which runs straight out from
the surface at every place. A bar drawn along it is a spike, with `z` naming its
measure ([Globe](globe.qmd)). The blanks are `map`'s, minus that `bar`, and they
are blank for the same reason. A sphere has no axis of its own to give them
either.

`network` holds three cells, and one of its marks draws in no other space.
`edge` draws nowhere else, because an edge is placed by neither position: the
space computes where every node sits and the edge follows the two it joins
([Network](network.qmd)). `point` and `text` join it there, as the nodes and
their names.

Three of the four crossings have a mark on every row, so one mark can be read
across all three at once. A path takes nearly every channel on Mark × Channel
and almost nothing on Mark × Transform, because a path *is* its rows in order
and every statistic replaces them. `rule` reads the same way on the transform
grid, and for the reason the statistics grid already gave: it hands a whole axis
to the panel, so it has no measure to compute. And `surface` has a row of the
transforms that cover a plane. `density` estimates a value at every node and
`bin` cuts adjacent cells, while `sum`, `mean`, `median`, `max` and `min` name
the column each cell holds. `path` derives two things. With `density` it draws
the contour of a field, the same field a `surface` lifts into a sheet. With
`cluster` it draws the tree.

With these grids the orthogonality law is stated in full: what each mark maps,
what it sets ([the Mark × Setting grid](style.qmd)), what it derives, and where
it can be drawn.
