# Composition

```{r}
#| include: false
source("R/setup.R")
```

How do you put two different plots on one page? Faceting splits one plot into
panels. Composition does the opposite: it takes plots that were never one plot
and arranges them on a page. Wilkinson keeps the two apart, noting that "some
tabled graphics are really two or more graphics glued together". gog keeps them
apart in the operators you have already learned. `|` and `/` place things side
by side and one above the other, whatever those things are:

* `plot | facet(continent)`: one plot, split by a column, panels side by side
* `plot_a | plot_b`: two plots, arranged side by side
* `plot_a / plot_b`: two plots, one above the other

A facet splits one plot by a column of its data. A page arranges finished
plots, and says nothing about any plot's data. Which one you get is decided by
what you write on the right of the operator, and by nothing else.

## Two plots, one page

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent) + theme(tick_angle = 45))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

Two entirely different measurements: a country's income against its life
expectancy, and how many countries each continent holds. They share nothing,
not a scale, not an axis, not a legend, because there is nothing to share. The
page divides evenly between them.

`/` places the same two one above the other:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) /
  (data(gapminder_2007) + bar * count + x(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, above bars derived by
count, x is continent."*

Both operators take pages as well as plots, so a page can hold another page. `(a
| b) / c` is a row of two plots above a third. In R, Python and Julia, `/` is
applied before `|`, so `a | b / c` reads as `a | (b / c)`. Add parentheses when
the reading matters. JavaScript spells the operators as words, so there is no
order to remember. Its chapter shows the spelling.

## A shared column is a shared axis

Two composed plots are unrelated until they name the same column on the axis the
arrangement lines up: `x` for plots one above the other, and `y` for plots side
by side. When they do, that is not two axes that happen to look alike; it is one
axis, and gog treats it as one:

> The same column on the aligned axis of two composed plots is one axis: one
> scale, one panel extent, drawn once.

Here are two plots, both reading `gdp` on x. The `theme(height = 150)` on the
first asks for 150 pixels of the page's height:

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 150)) /
  (data(gapminder_2007) + point + x(gdp) + y(life))
```

*"Given gapminder 2007: bars derived by bin, x is gdp, above points, x is gdp, y
is life."*

Look at what the rule did. The histogram's panel is exactly as wide as the
scatter's, so a bar stands over the points it counts. Both plots measure `gdp`
over the same range, so equal distances across the two panels stand for equal
amounts of `gdp`. And the `gdp` axis is written once, along the bottom, because
there is one axis: the same decision faceting makes when it puts tick labels
only under the bottom row of panels.

That is why the histogram touches the scatter instead of standing a tick
label's height above it. A plot that draws no axis reserves no margin for one.

### One axis, but not one legend

Compose two plots that both map `color` and you get two legends, one for each:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)) |
  (data(gapminder_2007) + point + x(population) + y(life) + color(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, beside
points, x is population, y is life, color by continent."*

Look at what that figure draws once, and what it draws twice. Both plots read
`life` on `y`, so that axis is shared and written a single time, on the left.
Both color by `continent` as well, and there you get *two* legends, saying
exactly the same thing.

The asymmetry is deliberate. Axes merge because there is an exact test: the same
column on the same axis means one scale is provably right for both plots. A
legend has no such test. These two agree, but two composed plots are two plots,
and nothing in the grammar makes them agree. The next pair may split by
different columns, or map one column onto two scales. A single legend over that
is wrong for one of them. Nothing in the picture shows the error, because it
still looks like a legend.

gog merges no legend, because it cannot prove that one legend is right for both
plots. Two identical legends side by side show two plots that agree.

When you want one legend over several panels, use [faceting](faceting.qmd)
instead. A facet is *one* plot split by a column, so every panel is colored by
the same scale. It draws a single legend. The choice between `|` and `facet()`
is the choice between two plots and one.

## The marginal plot

The histogram above the scatter is a **marginal**: one column's distribution,
drawn along the edge of the plot that uses it. Put a second marginal on the
right and the picture is the marginal plot, which most tools offer as a named
chart type:

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 130)) /
  ((data(gapminder_2007) + point + x(gdp) + y(life)) |
     (data(gapminder_2007) + bar * bin + y(life) + theme(width = 130)))
```

*"Given gapminder 2007: bars derived by bin, x is gdp, above points, x is gdp, y
is life, beside bars derived by bin, y is life."*

Three ordinary plots and two operators. The top histogram shares `gdp` with the
scatter, so it is drawn exactly as wide as the scatter's panel. The right-hand
one shares `life`, so its panel is exactly as tall, and the `life` axis is drawn
once on the left. The empty top-right corner has no atom of its own. It is the
room left over once both shared axes have taken their width and height.

The marginals follow no special rule. `bar * bin + y(life)` is the histogram you
already know, with its bars running left to right, because that is what binding
the measure to `y` means. And the shared column need not be continuous: two
plots reading the same *category* on x share its slots, so a box plot's summary
stands over the points it summarizes, one category at a time:

```{r}
(data(gapminder_2007) + box + x(continent) + y(life) + theme(height = 150)) /
  (data(gapminder_2007) + point + x(continent) + y(gdp) + color(continent))
```

*"Given gapminder 2007: boxes, x is continent, y is life, above points, x is
continent, y is gdp, color by continent."*

## A clustered panel decides the order

A categorical axis needs an order, and most plots take the one the data arrived
in. A panel with [`cluster`](cluster.qmd) computes a better one. `cluster`
builds a tree of similar rows, and reading its leaves from left to right gives
an order that puts similar things next to each other. When such a panel shares
its axis, the shared axis takes that order.

> A shared categorical axis takes the order a clustered panel computed for it.

The panels beside the clustered one stay plain, and their slots arrive in the
tree's order. gog refuses two panels that derive different orders for one shared
axis, because one axis cannot hold two orders. Give both the same columns to
`cluster()`, or cluster in one panel alone. The
[Cluster](cluster.qmd#the-clustered-heatmap) chapter uses this rule: a tile plot
with a tree on two of its sides, each tree deciding one axis.

## How thin is the marginal histogram? `theme(width =, height =)`

A plot states how much room it wants in pixels, and the plots that ask for
nothing split what is left. A page that says nothing is 800 by 600, which is
what a plot drawn alone takes. That is the whole of the sizing above: the
marginal histogram, for example, asks for 130 pixels of height, the scatter
asks for nothing and takes the rest.

The same two words also size a plot drawn on its own, where the room being
divided is the whole image:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + theme(width = 420, height = 320)
```

One meaning in two places, which is Compositional Invariance: the sentence that
says a plot is 130 pixels tall says it whether or not anything is composed with
it. Do not confuse it with `theme(ratio = )`, which shapes the *panel* inside
whatever room the plot was given and never resizes the image.

Plots that ask for more height than the page has are refused, and the message
shows the arithmetic:

```{r}
#| error: true
render_svg(
  (data(gapminder_2007) + point + x(gdp) + y(life) + theme(height = 500)) /
    (data(gapminder_2007) + point + x(gdp) + y(life) + theme(height = 500))
)
```

## How big is the page itself?

Two plots set side by side divide the page's width, and each keeps the whole
of its height. So a height written into one of them changes nothing. Each plot
sits in a cell of the page, and a cell in a side-by-side page is already as
tall as the page. The height belongs to the page, so the page is
where it is written:

```{r}
((data(gapminder_2007) + point + x(gdp) + y(life)) |
   (data(gapminder_2007) + bar * count + x(continent))) + theme(height = 260)
```

The same two words, now written on the page instead of the plot. A plot states
how much room it wants, and so does a page. Composed, the first is a cell of the
second.

This matters most where a mark cannot be stretched to fill its panel. A
[cube](space.qmd) is drawn with one scale in both screen directions, so its
shape is never stretched. A tall, narrow cell therefore leaves it small, with
empty bands above and below. The figure is the wrong shape, and the page is
where you fix it.

Size is the only thing a page can say. Every other property `theme()` sets
describes a panel: its gridlines, its frame, its background, the angle its tick
labels are read at. A page has no panel, so each of those properties is refused,
and the refusal names the plot it belongs to:

```{r}
#| error: true
render_svg(((data(gapminder_2007) + point + x(gdp) + y(life)) |
              (data(gapminder_2007) + bar * count + x(continent))) + theme(grid = "none"))
```

## What a page is not

A page is plots arranged, so anything belonging to *one* plot is written into
that plot before it is composed. Apart from the size above, an atom added to a
page refuses and says so:

```{r}
#| error: true
render_svg(((data(gapminder_2007) + point + x(gdp) + y(life)) |
              (data(gapminder_2007) + bar * count + x(continent))) + title("Two views"))
```

And a page cannot be faceted, because a facet splits one plot by a column and a
page is not one plot:

```{r}
#| error: true
render_svg(((data(gapminder_2007) + point + x(gdp) + y(life)) |
              (data(gapminder_2007) + bar * count + x(continent))) | facet(continent))
```

Facet first, then compose: `(plot | facet(g)) | other_plot` is a faceted plot
beside another plot, and it is perfectly legal.

## Where these operators come from

Neither reading of `|` is this grammar's invention, and the two were borrowed
from different places.

The **facet** reading is the older. Trellis display wrote a conditioned plot as
`y ~ x | g` [@becker1996trellis], and lattice writes it that way still; the bar
means "given", which is the bar written in conditional probability. That is
exactly what `plot | facet(continent)` says.

The **composition** reading comes from patchwork [@pedersen2019patchwork]. That
package writes `p1 | p2` for side by side and `p1 / p2` for one above the other.
gog spells composition the same way on purpose: a reader who knows that package
already knows this chapter, and a third spelling of a settled idea would help
nobody.

Composing pictures with operators is older than either. Henderson's *Functional
Geometry* [@henderson1982] describes an algebra of pictures, and its two
combining operations are `beside` and `above`. *Structure and Interpretation of
Computer Programs* (SICP) teaches the same picture language, and Haskell's
diagrams library spells the two operations `|||` and `===`. The JavaScript
binding has no operators to overload, so it names its two functions `beside()`
and `below()` after that older algebra.

One divergence is worth naming, since Wilkinson settles most questions in this
project. His graph algebra also has a `/`, and it means **nest** (`a/b`, "a
within b"), not "below". gog's facet operators take his *crossing* and do not
use his nesting, so `/` here can mean the page.

## Layering, faceting, or composition?

The three answer different questions, so your data decides which one you want:

* **Layering** when the marks are *different tables on the same measurement*.
  One coordinate space, one pair of axes, and the second table annotates the
  first: a threshold, a band, a callout, a forecast. [Data](data.qmd) explains
  multi-table plots.
* **Faceting** when the panels are *the same measurement* on different subsets.
  One scale, one legend, and the comparison between panels is what you gain.
* **Composition** when the plots are *different measurements* that belong on one
  page. Each keeps its own space, and the only thing they may share is a column.

Layering is the one you have to verify, because it is the only one the engine
cannot check for you. A second table that measures something the axes do not
measure still draws when you layer it. The columns resolve, the marks are drawn,
and the picture is wrong. Sales figures on an income axis are a legal sentence.
So if the second table does not measure what the axes measure, compose two plots
instead of layering one onto the other.

A marginal plot is composition because the histogram and the scatter measure
different things: one counts countries, the other puts each country at its own
position. They share `gdp`, and that shared column is exactly what the rule
above matches.