# Composition

```{r}
#| include: false
source("R/setup.R")
```

How do you put two different plots on one page? Faceting splits one plot into
panels. Composition does the opposite: it takes plots that were never one plot
and arranges them on a page. Wilkinson separates the two, noting that "some
tabled graphics are really two or more graphics glued together". gog
separates them too, with the operators you have already learned. `|` places two
things side by side, and `/` places one above the other, whether those things
are panels or whole plots:

* `plot | facet(continent)`: one plot, split by a column, panels side by side
* `plot_a | plot_b`: two plots, arranged side by side
* `plot_a / plot_b`: two plots, one above the other

A facet splits one plot by a column of its data. A **page** arranges finished
plots, and says nothing about any plot's data. What you write on the right
of the operator decides which one you get, and nothing else does.

## Two plots, one page

Two views of one table often belong side by side, even when they share no
axis. Write each plot in full, then put `|` between them:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent) + theme(tick_angle = 45))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

The two plots measure entirely different things: a country's income against its
life expectancy, and how many countries each continent holds. They share
nothing, not a scale, not an axis, not a legend, because there is nothing to
share. The page divides evenly between them, so each plot keeps half the width.
That is why the bar plot turns its continent names with
`theme(tick_angle = 45)`.

`/` places the same two plots one above the other. Each one now has the full
width of the page, so the continent names fit without turning:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) /
  (data(gapminder_2007) + bar * count + x(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, above bars derived by
count, x is continent."*

Both operators take pages as well as plots, so a page can hold another page. `(a
| b) / c` is a row of two plots above a third, and the marginal plot later in
this chapter is built that way.

In R, Python and Julia, `/` is applied before `|`, so `a | b / c` reads as `a |
(b / c)`. Add parentheses whenever the order changes the picture. JavaScript
spells the operators as words, so there is no order to remember. The
[JavaScript](bindings/javascript.qmd) chapter shows those words.

## A shared column is a shared axis

The two plots in that figure share no column. One shared column changes what
the page draws. Two composed plots stay unrelated until both name the same
column on the same axis. The arrangement decides which axis: `x` when one plot
is above the other, and `y` when the two sit side by side. When both
name it, that is not two axes that happen to look alike; it is one axis, and gog
treats it as one:

> The same column on the aligned axis of two composed plots is one axis: one
> scale, one panel extent, drawn once.

Here are two plots, both reading `gdp` on x. The `theme(height = 150)` on the
first asks for 150 pixels of the page's height:

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 150)) /
  (data(gapminder_2007) + point + x(gdp) + y(life))
```

*"Given gapminder 2007: bars derived by bin, x is gdp, above points, x is gdp, y
is life."*

Look at what the rule did. The histogram's panel is exactly as wide as the
scatter's, so each bar is drawn directly above the points it counts. Both plots
measure `gdp` over the same range, so equal distances across the two panels
stand for equal amounts of `gdp`. And the `gdp` axis is written once, along the
bottom, because there is only one axis to write. Faceting makes the same
decision when it puts tick labels only under the bottom row of panels.

A plot that draws no axis reserves no margin for one. That is why the histogram
sits close to the scatter, with no strip of tick labels between the two panels.

### One axis, but not one legend

Having seen an axis merge, you might expect a shared legend to merge the same
way. Compose two plots that both map `color` and you get two legends, one for
each:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)) |
  (data(gapminder_2007) + point + x(population) + y(life) + color(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, beside
points, x is population, y is life, color by continent."*

Look at what that figure draws once, and what it draws twice. Both plots read
`life` on `y`, so that axis is shared and written a single time, on the left.
Both color by `continent` as well, and there you get *two* legends, saying
exactly the same thing.

Axes merge and legends do not, and that difference is deliberate. Axes merge
because there is an exact test: the same column on the same axis means one scale
must be right for both plots. A legend has no such test. These two plots do
agree, because both split by `continent`. But nothing in the grammar makes any
two composed plots agree. The next pair may split by different columns, or map
one column onto two scales. One legend over those two would be wrong for one of
them. Nothing in the picture shows the error, because it still looks like a
legend.

So gog merges no legend. Two identical legends side by side tell you that the
two plots split by the same column.

When you want one legend over several panels, use [faceting](faceting.qmd)
instead. A facet is *one* plot split by a column, so every panel is colored by
the same scale. It draws a single legend. The choice between `|` and `facet()`
is the choice between two plots and one.

## The marginal plot

The histogram sharing the scatter's `gdp` axis is a **marginal**: one column's
distribution,
drawn along the edge of the plot that uses it. Put a second marginal on the
right and the picture is the marginal plot, which most tools offer as a named
chart type:

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 130)) /
  ((data(gapminder_2007) + point + x(gdp) + y(life)) |
     (data(gapminder_2007) + bar * bin + y(life) + theme(width = 130)))
```

*"Given gapminder 2007: bars derived by bin, x is gdp, above points, x is gdp, y
is life, beside bars derived by bin, y is life."*

That figure is three ordinary plots and two operators. The top histogram shares
`gdp` with the scatter, so its panel is drawn exactly as wide as the scatter's.
The right-hand one shares `life`, so its panel is exactly as tall as the
scatter's, and the `life` axis is drawn once, on the left. The empty top-right
corner has no atom of its own. Both marginals stop where the scatter's panel
stops, and the corner is the room left over.

The marginals follow no special rule. `bar * bin + y(life)` is the histogram you
already know. Its bars run left to right, because `y` now carries the bins and
the count goes on `x`. The shared column need not be continuous either. Two
plots reading the same *category* on x share its slots, so each category keeps
the same place in both:

```{r}
(data(gapminder_2007) + box + x(continent) + y(life) + theme(height = 150)) /
  (data(gapminder_2007) + point + x(continent) + y(gdp) + color(continent))
```

*"Given gapminder 2007: boxes, x is continent, y is life, above points, x is
continent, y is gdp, color by continent."*

The five continents take the same slots in both panels, so each box is drawn
directly above the countries it summarizes.

## A clustered panel decides the order

A categorical axis needs an order, and most plots take the one the data arrived
in. A panel with [`cluster`](cluster.qmd) computes a better one. `cluster`
builds a tree of similar rows, and reading its leaves from left to right gives
an order that puts similar things next to each other. When such a panel shares
its axis, the shared axis takes that order.

> A shared categorical axis takes the order a clustered panel computed for it.

The panels beside the clustered one stay plain, and their slots arrive in the
tree's order. gog refuses two panels that derive different orders for one shared
axis, because one axis cannot hold two orders. Give both the same columns to
`cluster()`, or cluster in one panel alone. The
[Cluster](cluster.qmd#the-clustered-heatmap) chapter uses this rule: a tile plot
with a tree on two of its sides, each tree deciding one axis.

## How thin is the marginal histogram? `theme(width =, height =)`

A plot states how much room it wants in pixels, and the plots that ask for
nothing split what is left. A page that states no size is 800 by 600 pixels,
which is what a plot drawn alone takes. That is the whole of the sizing in the
marginal plot: the top histogram asks for 130 pixels of height, and the scatter
asks for nothing and takes the rest.

The same two arguments also size a plot drawn on its own, where the room being
divided is the whole image:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + theme(width = 420, height = 320)
```

That is Compositional Invariance, the sixth of the
[nine laws](design-laws.qmd): one meaning in two places. A sentence that says a
plot is 130 pixels tall says the same thing whether or not you compose it with
another plot. Do not confuse `theme(width =, height =)` with `theme(ratio = )`.
The second one shapes the *panel* inside the room the plot was given, and it
never resizes the image.

Plots that ask for more height than the page has are refused, and the message
gives both numbers:

```{r}
#| error: true
(data(gapminder_2007) + point + x(gdp) + y(life) + theme(height = 500)) /
  (data(gapminder_2007) + point + x(gdp) + y(life) + theme(height = 500))
```

## How big is the page itself?

That refusal compared both plots with the page's own height, so a page has a
size of its own. Two plots set side by side divide the page's width, and each
one keeps the full height of the page. Each plot sits in a cell, and a cell in a
side-by-side page is already as tall as the page. So a height written into one
of those plots changes nothing. The height belongs to the page, so write it on
the page:

```{r}
((data(gapminder_2007) + point + x(gdp) + y(life)) |
   (data(gapminder_2007) + bar * count + x(continent))) + theme(height = 260)
```

Here the same two arguments are written on the page instead of on a plot. A page
states how much room it wants, exactly as a plot does. When you compose them,
the plot becomes one cell of the page.

This matters most where a mark cannot be stretched to fill its panel. A
[cube](space.qmd) is drawn with one scale in both screen directions, so its
shape is never stretched. A tall, narrow cell therefore draws the cube small,
with empty bands above it and below it. The figure is then the wrong shape, and
you correct it by sizing the page.

Size is the only thing a page can say. Every other property `theme()` sets
describes a panel: its gridlines, its frame, its background, the angle its tick
labels are read at. A page has no panel, so each of those properties is refused,
and the refusal says to write it into the plot it describes:

```{r}
#| error: true
((data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))) + theme(grid = "none")
```

## What a page is not

A title over two plots seems to belong to the page rather than to either plot,
so you might try writing it there. A page is plots arranged, so anything
belonging to *one* plot is written into that plot before it is composed. Apart
from `width` and `height`, any atom added to a page is refused, and the message
names where it belongs:

```{r}
#| error: true
((data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))) + title("Two views")
```

And a page cannot be faceted, because a facet splits one plot by a column and a
page is not one plot:

```{r}
#| error: true
((data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))) | facet(continent)
```

Facet first, then compose: `(plot | facet(g)) | other_plot` is a faceted plot
beside another plot, and it is legal.

## Where these operators come from

Neither reading of `|` is this grammar's invention, and the two were borrowed
from different places.

The **facet** reading is the older of the two. Trellis display wrote a
conditioned plot as `y ~ x | g` [@becker1996trellis], and lattice writes it that
way still. There the bar means "given", the same meaning it has in conditional
probability. That is exactly what `plot | facet(continent)` says.

The **composition** reading comes from patchwork [@pedersen2019patchwork]. That
package writes `p1 | p2` for side by side and `p1 / p2` for one above the other.
gog spells composition the same way on purpose: a reader who already uses that
package has little new to learn here, and a third spelling of a settled idea
would only add work.

Composing pictures with operators is older than both of them. Henderson's
*Functional Geometry* [@henderson1982] describes an algebra of pictures, and its
two combining operations are `beside` and `above`. *Structure and
Interpretation of Computer Programs* (SICP) teaches the same picture language,
and Haskell's diagrams library spells the two operations `|||` and `===`. The
JavaScript binding has no operators to overload, so it spells the two
operations as words: `beside()`, and `below()` for what Henderson called
`above`.

One divergence is worth naming, since Wilkinson settles most questions in this
project. His graph algebra also has a `/`, and it means **nest** (`a/b`, "a
within b"), not "below". gog's facet operators use his *crossing* and not his
nesting, so `/` is free to mean the page here.

## Layering, faceting, or composition?

The three answer different questions, so your data decides which one you want:

* **Layering** when the marks are *different tables on the same measurement*.
  One coordinate space, one pair of axes, and the second table annotates the
  first: a threshold, a band, a label, a forecast. [Data](data.qmd) explains
  multi-table plots.
* **Faceting** when the panels are *the same measurement* on different subsets.
  One scale, one legend, and you can compare the panels directly.
* **Composition** when the plots are *different measurements* that belong on one
  page. Each keeps its own space, and the only thing they may share is a column.

Layering is the one you have to verify, because it is the only one the engine
cannot check for you. A second table that measures something the axes do not
measure still draws when you layer it. The columns resolve, the marks are drawn,
and the picture is wrong. Sales figures on an income axis are a legal sentence.
So if the second table does not measure what the axes measure, compose two plots
instead of layering one onto the other.

A marginal plot is composition because the histogram and the scatter measure
different things: one counts countries, the other puts each country at its own
position. They share `gdp`, and that one shared column is what the shared-axis
rule turns into a single axis.

## In short

`|` and `/` mean faceting when a column splits one plot into panels, and
composition when both sides are whole plots. A shared column becomes a shared
axis, and plots that share nothing keep their own scales. A page arranges the
plots it is given, and computes nothing of its own.

## What comes next

Composition arranges finished plots, and each one still draws the rows it was
given. [Flow](flow.qmd) computes a picture instead. It carries one amount
through a series of stages, and it reads those stages from the table's own
columns.
