# Composition

```{r}
#| include: false
source("R/setup.R")
```

How do you put two different plots on one page? Faceting splits one plot into
panels. Composition does the opposite: it takes plots that were never one plot
and arranges them on a page. Wilkinson keeps the two apart in the same words,
noting that "some tabled graphics are really two or more graphics glued
together", and gog keeps them apart in the operators you have already learned.
`|` and `/` place things side by side and one above the other, whatever those
things are:

* `plot | facet(continent)`: one plot, split by a column, panels side by side
* `plot_a | plot_b`: two plots, arranged side by side
* `plot_a / plot_b`: two plots, one above the other

A facet is *semantic*: one plot, one coordinate space, split by a variable and
sharing everything. A page is *presentational*: each plot keeps its own space,
its own marks, its own measurements. Which one you get is decided by what you
write on the right of the operator, and nothing else needs saying.

## Two plots, one page

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent) + theme(tick_angle = 45))
```

*"Points, x is gdp, y is life, beside bars derived by count, x is continent."*

Two entirely different measurements: a country's income against its life
expectancy, and how many countries each continent holds. They share nothing,
not a scale, not an axis, not a legend, because there is nothing to share. The
page divides evenly between them.

`/` stacks the same two:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) /
  (data(gapminder_2007) + bar * count + x(continent))
```

Both operators take pages as well as plots, so a page nests: `(a | b) / c` is two
plots in a row above one across the bottom. In R, and in Python and Julia,
**`/` binds tighter than `|`**, so `a | b / c` reads as `a | (b / c)`.
Parenthesize when the reading matters.

## A shared column is a shared axis

Two composed plots are unrelated until they name the same column on the same
axis. When they do, that is not two axes that happen to look alike; it is one
axis, and gog treats it as one:

> The same column on the same axis in two composed plots is one axis: one
> scale, one panel extent, drawn once.

Here are two plots, both reading `gdp` on x:

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 150)) /
  (data(gapminder_2007) + point + x(gdp) + y(life))
```

Look at what the rule did. The histogram's panel is exactly as wide as the
scatter's, so a bar stands over the points it counts. Both plots measure `gdp`
over the same range, so the bar edges and the point positions mean the same
distance. And the `gdp` axis is written once, along the bottom, because there is
one axis: the same decision faceting makes when it puts tick labels only under
the bottom row of panels.

That last part is why the histogram sits flush against the scatter rather than a
tick label's height above it. An axis a plot does not draw costs it no margin.

### One axis, but not one key

Compose two plots that both map `color` and you get two legends, one for each:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)) |
  (data(gapminder_2007) + point + x(population) + y(life) + color(continent))
```

Look at what that figure does twice and once. Both plots read `life` on `y`, so
that axis is shared and written a single time, on the left. Both color by
`continent` as well, and there you get **two** keys, saying exactly the same
thing.

The asymmetry is deliberate. An axis merges on a test that is exact: the same
column on the same channel means one scale is provably right for both plots. A
key has no such test. These two agree, but two composed plots are two plots, and
nothing in the grammar makes them agree. The next pair may split by different
columns, or cut one column on two scales, and a single key over that is wrong for
one of them. It is wrong quietly, because it still looks like a key.

gog will not merge a key it cannot promise, so it merges none. Identical keys
side by side are the honest picture of two plots that happen to agree.

When you want one key over several panels, use [faceting](faceting.qmd) instead.
A facet is *one* plot split by a column, so its panels share their colors by
construction. It draws a single legend. The choice between `|` and `facet()` is
the choice between two plots and one.

## The marginal plot

Put a second marginal on the right and the picture is the marginal plot, which
usually arrives as a chart type with a name of its own:

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 130)) /
  ((data(gapminder_2007) + point + x(gdp) + y(life)) |
     (data(gapminder_2007) + bar * bin + y(life) + theme(width = 130)))
```

Three ordinary plots and two operators. The top histogram shares `gdp` with the
scatter, so it is aligned and squeezed to the scatter's panel; the right-hand one
shares `life`, so it is aligned to the scatter's rows and the `life` axis is
drawn once, on the left. The empty top-right corner is not a spacer anyone asked
for and there is no atom for one: it is the room the shared extent left over.

Nothing about the marginals is special-cased. `bar * bin + y(life)` is the
histogram you already know, lying down because that is what binding the
measurement to `y` means. And the shared column need not be continuous: two
plots reading the same *category* on x share its slots, so a box plot's summary
stands over the cloud it summarizes, category for category:

```{r}
(data(gapminder_2007) + box + x(continent) + y(life) + theme(height = 150)) /
  (data(gapminder_2007) + point + x(continent) + y(gdp) + color(continent))
```

## How thin is the marginal? `theme(width =, height =)`

A plot states how much room it wants in pixels, and the plots that ask for
nothing split what is left. That is the whole of the sizing above: the marginal
histogram asks for 130 pixels of height, the scatter asks for nothing and takes
the rest.

The same words size a plot drawn on its own, where the room in question is the
image:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + theme(width = 420, height = 320)
```

One meaning in two places, which is Compositional Invariance: the sentence that
says a plot is 130 pixels tall says it whether or not anything is composed with
it. Do not confuse it with `theme(ratio = )`, which shapes the *panel* inside
whatever room the plot was given and never resizes the image.

Asking for more page than there is refuses, with the arithmetic in the message:

```{r}
#| error: true
render_svg(
  (data(gapminder_2007) + point + x(gdp) + y(life) + theme(height = 500)) /
    (data(gapminder_2007) + point + x(gdp) + y(life) + theme(height = 500))
)
```

## And how big is the page itself?

Two plots set side by side divide the page's width, and each keeps the whole of
its height. So a height written into one of them says nothing at all: the cell is
already as tall as the page. The page is the figure that has that height, and the
page is what states it:

```{r}
((data(gapminder_2007) + point + x(gdp) + y(life)) |
   (data(gapminder_2007) + bar * count + x(continent))) + theme(height = 260)
```

The same two words, one scope up. A plot states how much room it wants, and so
does a page. Composed, the first is a cell of the second. A page that says
nothing is 800 by 600, which is what a plot drawn alone takes.

This matters most where a mark cannot be stretched to fill its panel. A
[cube](space.qmd) is fitted with one scale on both axes, so it is never sheared.
A tall, narrow cell therefore leaves it small, with empty bands above and below.
The figure is the wrong shape, and the page is where that is said.

Size is the only thing a page can say. Every other property `theme()` sets
describes a panel: its gridlines, its frame, its background, the angle its tick
labels are read at. A page has no panel, so those refuse, and name the plot they
belong to:

```{r}
#| error: true
render_svg(((data(gapminder_2007) + point + x(gdp) + y(life)) |
              (data(gapminder_2007) + bar * count + x(continent))) + theme(grid = "none"))
```

## What a page is not

A page is plots arranged, so anything belonging to *one* plot is written into
that plot before it is composed. Apart from the size above, an atom added to a
page refuses and says so:

```{r}
#| error: true
render_svg(((data(gapminder_2007) + point + x(gdp) + y(life)) |
              (data(gapminder_2007) + bar * count + x(continent))) + title("Two views"))
```

And a page cannot be faceted, because a facet splits one plot by a column and a
page is not one plot:

```{r}
#| error: true
render_svg(((data(gapminder_2007) + point + x(gdp) + y(life)) |
              (data(gapminder_2007) + bar * count + x(continent))) | facet(continent))
```

Facet first, then compose: `(plot | facet(g)) | other_plot` is a faceted plot
beside another plot, and it is perfectly legal.

## Where these operators come from

Neither reading of `|` is this grammar's invention, and the two were borrowed
from different places.

The **facet** reading is the older. Trellis display wrote a conditioned plot as
`y ~ x | g` [@becker1996trellis], and lattice writes it that way still; the bar
reads "given", straight out of conditional-probability notation. That is exactly
what `plot | facet(continent)` says.

The **composition** reading is the one you will recognize if you use patchwork
[@pedersen2019patchwork], which pairs `p1 | p2` for side by side with `p1 / p2`
for one above the other. gog spells composition the same way on purpose: a reader
who knows that package already knows this page, and a third spelling of an
established idea would cost everyone something and buy nobody anything.

Composing pictures with operators is older than either. Henderson's *Functional
Geometry* [@henderson1982] set out an algebra of pictures whose combining
operations are `beside` and `above`, which is the picture language SICP went on
to teach and which Haskell's diagrams library spells `|||` and `===`. That
lineage is why the JavaScript binding, having no operators to overload, calls its
two words `beside()` and `below()`: the names were already there.

One divergence is worth naming, since Wilkinson is this project's usual
tiebreaker. His graph algebra also has a `/`, and it means **nest** (`a/b`, "a
within b"), not "below". gog's facet operators claim his *crossing* and
deliberately do not claim his nesting, so `/` here is free to mean the page.

## Layering, faceting, or composition?

The three answer different questions, and the choice is not a matter of taste:

* **Layering** when the marks are *different tables on the same measurement*.
  One coordinate space, one pair of axes, and the second table annotates the
  first: a threshold, a band, a callout, a forecast. [Data](data.qmd) owns
  multi-table plots.
* **Faceting** when the panels are *the same measurement* on different subsets.
  One scale, one legend, and the comparison between panels is the point.
* **Composition** when the plots are *different measurements* that belong on one
  page. Each keeps its own space, and what they have in common, if anything,
  is the variable they share.

Layering is the one you have to verify, because it is the only one the engine
cannot check for you. A table whose numbers mean something else still draws when
you layer it. The columns resolve, the marks land, and the picture is wrong.
Sales figures on an income axis are a legal sentence. So if the second table
does not measure what the axes measure, compose two plots instead of layering
one onto the other.

A marginal plot is composition because the histogram and the scatter measure
different things: one counts countries, the other places them. They share
`gdp`, and that shared column is exactly what the rule above picks up.
