# Composition

```{r}
#| include: false
source("R/setup.R")
```

How do you put two different plots on one page? Faceting cannot do it: every
panel it draws comes from one plot and one sentence. Composition takes plots
that were never one plot and arranges them on a page. Wilkinson separates the
two, noting that "some tabled graphics are really two or more graphics glued
together" [@wilkinson2005]. gog separates them too, with the operators you have
already learned. `|` places two things side by side, and `/` places one above
the other, whether those things are panels or whole plots:

* `plot | facet(continent)`: one plot, split by a column, panels side by side
* `plot_a | plot_b`: two plots, arranged side by side
* `plot_a / plot_b`: two plots, one above the other

A facet splits one plot by a column of its data. A **page** arranges finished
plots, and says nothing about any plot's data. Only one thing decides which of
the two you get: what you write on the right of the operator.

## Two plots, one page

Two views of one table often belong side by side, even when they share no
axis. Write each plot in full, then put `|` between them:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

The two plots measure entirely different things: a country's income against its
life expectancy, and how many countries each continent holds. They share
nothing, not a scale, not an axis, not a legend, because there is nothing to
share. Each plot sits in a **cell**, its own part of the page. The two cells are
equal, and a narrow gap between them leaves each a little under half the width.
Inside each cell, a panel holds the marks, and the axes and their labels take the
rest. The page does align the two panels at the top and the bottom.

`/` places the same two plots one above the other. Each one now has the full
width of the page:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) /
  (data(gapminder_2007) + bar * count + x(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, above bars derived by
count, x is continent."*

One above the other, the two panels align at the left and the right, even when
one plot's tick labels are wider.

Both operators take pages as well as plots, so a page can hold another page.
`(a | b) / c` is a row of two plots above a third. The marginal plot later in
this chapter is built the other way: one plot above a row of two.

In R, Python and Julia, `/` is applied before `|`, so `a | b / c` reads as `a |
(b / c)`. Write the parentheses whenever you mix the two operators. JavaScript
spells the operators as words, so there is no order to remember. The
[JavaScript](bindings/javascript.qmd) chapter shows those words.

## A shared column is a shared axis

The two plots in those figures share no column. One shared column changes what
the page draws. When two composed plots name the same column on the same axis,
both read it on one scale. The arrangement decides whether they also share one
drawn axis. Plots one above the other align their `x` axes, and plots side by
side align their `y` axes. When both plots name the column on the axis the
arrangement aligns, gog does not draw two matching axes. It draws one axis:

> The same column on the aligned axis of two composed plots is one axis: one
> scale, one panel extent, drawn once.

The shared column may sit on the other axis instead, such as `gdp` on x in two
plots side by side. Each plot then draws its own copy of that axis from the one
scale, so both copies mark the same values.

Here are two plots, both reading `gdp` on x. The `theme(height = 150)` on the
first asks for 150 pixels of the page's height:

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 150)) /
  (data(gapminder_2007) + point + x(gdp) + y(life))
```

*"Given gapminder 2007: bars derived by bin, x is gdp, above points, x is gdp, y
is life."*

Look at what the rule did. Both plots measure `gdp` over the same range, so each
bar is drawn directly above the points it counts. And the `gdp` axis is drawn once, along the
bottom, because there is only one axis to draw. Faceting makes the same
decision when it draws the tick labels only under the panels with nothing below
them.

A plot that does not draw an axis reserves no margin for it. That is why the
histogram sits close to the scatter, with no band of tick labels between the
two panels.

gog cannot know what a column measures, so on a page it compares column names.
Two plots built from different tables share a scale whenever both name a column
spelled the same way on the same axis. That is correct when the two columns
measure the same thing, such as `gdp` in two tables of countries. When they
measure different things, rename one of them before you compose. Otherwise the
plot with the smaller values is drawn in a small part of an axis sized for the
larger.

### One axis, but not one legend

You have just seen an axis merge, so you might expect two legends for one column
to merge the same way. Compose two plots that both map `color` and you get two
legends, one for each:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)) |
  (data(gapminder_2007) + point + x(population) + y(life) + color(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, beside
points, x is population, y is life, color by continent."*

Look at what that figure draws once, and what it draws twice. Both plots read
`life` on `y`, so that axis is shared and drawn once, on the left. Both color by
`continent` as well, and there you get *two* legends, saying exactly the same
thing.

Axes merge and legends do not, and that difference is deliberate. Axes merge
because there is an exact test: the same column on the same axis means one scale
is correct for both plots. A legend has no such test. These two plots do
agree, because both split by `continent`. But nothing in the grammar makes any
two composed plots agree. The next pair may split by different columns, or map
one column onto two scales. One legend over those two would be wrong for one of
them. Nothing in the picture shows the error, because it still looks like a
legend.

So gog merges no legend. Two identical legends side by side tell you that the
two plots split by the same column.

When you want one legend over several panels, use [faceting](faceting.qmd)
instead. A facet is *one* plot split by a column, so every panel is colored by
the same scale. It draws a single legend. The choice between composing and
faceting is the choice between two plots and one.

## The marginal plot

A scatter shows how two columns vary together, but not how the rows spread
along each one alone. Go back to the histogram that shared the scatter's `gdp`
axis. That histogram is a **marginal**: one column's distribution, drawn along
the edge of the plot that uses the column. Put a second marginal on the right
and the picture is the marginal plot, which most tools offer as a named chart
type:

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 130)) /
  ((data(gapminder_2007) + point + x(gdp) + y(life)) |
     (data(gapminder_2007) + bar * bin + y(life) + theme(width = 130)))
```

*"Given gapminder 2007: bars derived by bin, x is gdp, above points, x is gdp, y
is life, beside bars derived by bin, y is life."*

That figure is built from three ordinary plots and two operators. The top
histogram shares `gdp` with the scatter, so its panel is drawn exactly as wide
as the scatter's. The right-hand one shares `life`, so its panel is exactly as
tall as the scatter's, and the `life` axis is drawn once, on the left. Nothing
in the sentence draws the empty top-right corner. Both marginals stop where the
scatter's panel stops, and the corner is the room left over.

The marginals need no special rule either. `bar * bin + y(life)` is the
histogram you already know. Its bars run left to right, because `y` now carries
the bins and the count goes on `x`.

### A shared categorical column

You may want each group's summary drawn directly above the group's own rows.
The shared column need not be continuous for that. Two plots reading the same
*categorical* column on x share that column's slots, so each category keeps the
same place in both:

```{r}
(data(gapminder_2007) + box + x(continent) + y(life) + theme(height = 150)) /
  (data(gapminder_2007) + point + x(continent) + y(gdp) + color(continent))
```

*"Given gapminder 2007: boxes, x is continent, y is life, above points, x is
continent, y is gdp, color by continent."*

The five continents take the same slots in both panels, so each continent's box
sits directly above that continent's countries.

## A clustered panel decides the order

A categorical axis needs an order, and most plots take the one the data arrived
in. A panel with [`cluster`](cluster.qmd) computes a better one. `cluster`
builds a tree that joins the most similar categories first, and reading its
leaves along the axis gives an order that puts similar categories next to each
other. When a clustered panel is
composed with another, the axis they share takes the order the tree computed.

> A shared categorical axis takes the order a clustered panel computed for it.

The other panels need no `cluster` of their own, and their categories take the
tree's order. gog refuses two panels that derive different orders for one shared
axis, because one axis cannot hold two orders. Give both the same columns to
`cluster()`, or cluster in one panel alone. The
[Cluster](cluster.qmd#the-clustered-heatmap) chapter uses this rule: a tile plot
with a tree on two of its sides, each tree deciding one axis.

## How thin is the marginal histogram? `theme(width =, height =)`

In the marginal plot, both histograms are thin because each one asks for a small
size. The top histogram asks for 130 pixels of height, and the right-hand one
for 130 pixels of width. A plot asks for room in pixels with `theme(width = )`
and `theme(height = )`. The plots that ask for nothing divide what is left, so
the scatter takes the rest.

On a plot drawn alone, the same two arguments set the size of the whole image:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + theme(width = 420, height = 320)
```

Alone or on a page, the two arguments mean the same thing, which is
Compositional Invariance, the sixth of the [nine laws](design-laws.qmd). A plot
that asks for 130 pixels of height is 130 pixels tall, whether it is drawn alone
or above another plot. Beside another plot it is also 130 pixels tall, and the
plot beside it takes the same height.

Do not confuse `theme(width =, height =)` with
[`theme(ratio = )`](style.qmd#theme-settings-that-belong-to-no-mark). The
second one shapes the *panel* inside the room the plot was given, and it never
resizes the image. When a plot states a `ratio`, the page does not align its
panel with the other panels.

## How big is the page itself?

A page that states no size is 800 by 600 pixels, the same as a plot that states
none. Its plots can decide a different size. Plots one above the other that all
ask for a height make the page as tall as their heights plus the gaps. Plots
side by side that all ask for a width make it as wide as their widths plus the
gaps. Each gap is 20 pixels.

Plots side by side divide the page's width and share one height. A height
written into one of them is the height of all of them, and of the page. Plots
side by side that ask for different heights are refused. Plots one above the
other share one width, by the same rule. The marginal plot's page stays 800 by
600, because the scatter asks for nothing.

A plot that asks for nothing still needs room, so the other plots must leave
some of the page's height for it:

```{r}
#| error: true
(data(gapminder_2007) + point + x(gdp) + y(life) + theme(height = 700)) /
  (data(gapminder_2007) + bar * count + x(continent))
```

The bottom plot asks for nothing, so the page keeps its 600 pixels. gog added the
top plot's 700 pixels and the 20-pixel gap, and compared 720 with those 600.

When the plots do not decide the page's size, you can choose it: give the page
its own `theme()`. Put parentheses around the whole page before you add
`theme()` to it. In R and Python, `+` is applied before `|`, so without them the
`theme()` would go to the second plot instead of the page:

```{r}
((data(gapminder_2007) + point + x(gdp) + y(life)) |
   (data(gapminder_2007) + bar * count + x(continent))) + theme(height = 260)
```

The page is now 260 pixels tall instead of 600, and both plots take that full
height. A page takes the same two size arguments as a plot.

Sizing the page matters most when a plot cannot be stretched to fill its cell. A
[cube](space.qmd) is drawn with one scale in both screen directions, so its
shape is never stretched. In a tall, narrow cell the cube is therefore drawn
small, with an empty band above it and another below it. The figure is then the
wrong shape, and you correct it by sizing the page.

Size is the only thing a page can say. Every other property `theme()` sets
belongs to one plot: its gridlines, its frame, its background, the angle of its
tick labels. A page is not a plot, so each of those properties is refused, and
the refusal says to write it into the plot it describes:

```{r}
#| error: true
((data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))) + theme(grid = "none")
```

## What a page is not

A title over two plots seems to belong to the page rather than to either plot,
so you might try writing it there. A page is an arrangement of plots, so
anything belonging to *one* plot is written into that plot before it is
composed. A page accepts `theme(width = )` and `theme(height = )`. Every other
atom added to a page is refused, and the message names where it belongs:

```{r}
#| error: true
((data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))) + title("Two views")
```

gog draws no title over a whole page. On a page of plots one above the other,
the top plot's title sits above everything else. Otherwise, write the title as
the figure's caption in your document.

And a page cannot be faceted, because a facet splits one plot by a column and a
page is not one plot:

```{r}
#| error: true
((data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))) | facet(continent)
```

Facet first, then compose: `(plot | facet(g)) | other_plot` is a faceted plot
beside another plot, and it is legal.

## Where these operators come from

You have now seen `|` used both ways. Neither reading is this grammar's
invention, and the two were borrowed from different places.

The **facet** reading is the older of the two. Trellis display wrote a
conditioned plot as `y ~ x | g` [@becker1996trellis], and lattice writes it that
way still. There `|` means "given", the same meaning it has in conditional
probability. That is exactly what `plot | facet(continent)` says.

The **composition** reading comes from patchwork [@pedersen2019patchwork]. That
package writes `p1 | p2` for side by side and `p1 / p2` for one above the other.
gog spells composition the same way on purpose. A reader who already uses that
package already knows both operators. A different spelling for an idea that
already has one would only be one more thing to learn.

Composing pictures with operators is older than both of them. Henderson's
*Functional Geometry* [@henderson1982] describes an algebra of pictures, and it
places pictures with `beside` and `above`. The textbook *Structure and
Interpretation of Computer Programs* teaches the same picture language. Haskell's
diagrams library spells the two operations `|||` and `===`. The JavaScript
binding has no operators to overload, so it spells the two operations as words:
`beside()`, and `below()` for what Henderson called `above`.

gog follows Wilkinson in most of its decisions, and this is one place where it
does not. His graph algebra also has a `/`, and there it means *nesting*
(`a/b`, "a within b"), not one plot above another. That nesting is not gog's
`nest()` atom, which divides one panel into areas sized by a measure. gog's facet
operators always cross and never nest, as the [Faceting](faceting.qmd) chapter
explains, so gog can use `/` to place one plot above another.

## Layering, faceting, or composition?

Layering, faceting and composition answer different questions, so your data
decides which one you want:

* **Layering.** The layers draw *different tables* that measure the same thing.
  One coordinate space, one pair of axes, and the second table annotates the
  first: a threshold, a band, a note, a forecast.
  [Data](data.qmd#the-plots-table-and-a-layers-table) shows how to add a second
  table.
* **Faceting.** The panels show *the same measurement* for different subsets.
  One scale, one legend, and you can compare the panels directly.
* **Composition.** The plots show *different measurements* that belong on one
  page. Each keeps its own space, and the only thing they may share is a column.

In layering, the engine cannot check that a second table measures what the axes
measure, so you have to. A second table that measures something else still
draws when you layer it: gog finds every column, draws every mark, and the
picture is wrong. A sentence that puts sales figures on an income axis is legal,
and gog draws it. So if the second table does not measure what the axes
measure, compose two plots instead of layering one onto the other.

A marginal plot is composition because the histogram and the scatter measure
different things: one counts countries, the other puts each country at its own
position. They share `gdp`, and that one shared column is what the shared-axis
rule turns into a single axis.

## In short

`|` and `/` mean faceting when a column splits one plot into panels, and
composition when both sides are whole plots. Composed plots that name the same
column on the same axis read it on one scale. On the axis the arrangement
aligns, that column is one axis, drawn once. A page draws no mark of its own,
and its size is the only thing you can write on it.

## What comes next

Composition arranges finished plots on a page, and none of those plots shows
where each part of a total went. [Flow](flow.qmd) draws that: it carries each
amount through a series of stages, read from the table's own columns.
