# Coverage, both directions

```{r}
#| include: false
source("R/setup.R")
```

The [phrasebook](phrasebook.qmd) answers "how do I spell this in gog". This
appendix answers the question underneath it: *can* it be spelled. Every entry in
ggplot2's reference index was read against the kernel. The result is two lists
of very different lengths, and a table of differences between them.

Each table below has a ggplot2 column, and it is dated on purpose, because
nothing here can check it: reference index of ggplot2 4.0.3, read 2026-07-26
[@wickham2026ggplot2pkg]. ggplot2 is deliberately not a dependency of this book,
so building the book never runs it. The plots on this page, and the one refusal
shown further down, are different: the engine draws them each time the book is
built.

The first list is short: what gog does that ggplot2 does not. The second is
long: what ggplot2 draws that gog does not. Each gap on it is one of four kinds,
and three of the four were decided on purpose. A later section turns to
ggplot2's extension packages, which the reading of the index did not include,
and draws five of their plots from tables the host language computes.

## What gog does that ggplot2 does not

The first section below shows a picture ggplot2 cannot draw. The other two show
differences in vocabulary, not in what can be drawn.

### Anything in the cube

Binding `z` gives a plot a third position, and `surface` draws a height over a
plane:

```{r}
#| echo: false
peek(maunga_whau)
```

```{r}
data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  color(elevation) + palette("plasma") +
  title("Maunga Whau, one row per crossing")
```

*"Given Maunga Whau: a surface, x is east, y is north, z is elevation, color by
elevation."*

Six marks draw in the cube: `point`, `path`, `bar`, `interval`, `box` and
`surface`. `space(turn = , tilt = )` sets the viewing angle. `facet` and `play`
split a cube's rows as they split any other plot's: `| facet(g)` gives one
projected box per panel, `+ play(t)` runs the sequence inside a single box.
ggplot2 has no depth position, and no coordinate system that would read one. Its
own book says it does not support true 3-D surfaces [@wickham2016ggplot2]. In R
these plots come from plotly, rgl or lattice, each with a grammar of its own.
Here `z` is one more channel in the same sentence, and [Space](space.qmd) is the
chapter.

### Texture, a channel of its own

`pattern` maps a categorical column to a texture: a hatch on the filled regions
and a dash on the strokes. Put it on the same column as `color` and each series
is marked twice, by hue and by hatch, in one merged legend:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * mean + x(species) + y(petal_length) +
  color(species) + pattern(species) +
  title("Hue and hatch, one legend")
```

*"Given the iris flowers: bars derived by mean, x is species, y is petal length,
color by species, pattern by species."*

The plot above stays legible in grayscale and for readers with color blindness,
because hue and hatch both show the species. ggplot2 can fill a bar with a hatch
too. There, you build each hatch yourself with R's grid package, and it replaces
the fill color. In gog, `pattern` is a channel of its own, separate from `color`.
So hue and hatch can come from two different columns. On a line, the same word
draws dashes, where ggplot2 needs a second word, `linetype`.

### The same plot, without a word for it

This is also a picture ggplot2 can draw, and gog draws it without adding a word.
`geom_violin()` is a mark of its own there, with `stat_ydensity()` beside it.
Here it is [`ribbon *
density`](marks/ribbon.qmd#the-violin-a-distribution-per-category). A `ribbon`
fills the region between two boundaries, and `density` estimates one
distribution for each category.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + ribbon * density + x(continent) + y(life) +
  y_label("Life expectancy") + title("ribbon * density: no new word required")
```

*"Given gapminder 2007: ribbons derived by density, x is continent, y is life."*

An `area` fills from its curve to its category's center line, not to a second
boundary. So `area * density + x(continent) + y(life)` draws half of each
violin: the half violin. That is the purpose of orthogonality: gog draws the
plot without adding a word to its vocabulary.

The donut is the same case arriving from a *parameter* rather than a geom.
ggplot2 spells it `coord_radial(inner.radius = )`, a number given to the space;
here the hole is not a property of the circle at all. It is a stretch of the
radial axis with nothing standing on it, so a tree one level deep, drawn on an
axis that starts below the first ring, is a donut:

```{r}
#| echo: false
peek(spending)
```

```{r}
data(spending) + zone * partition(group) + x(amount) +
  y(depth, limits = c(0, 2)) + color(group) + polar() +
  title("A donut is a sunburst of a one-level tree")
```

*"Given the spending table: zones derived by partition through group, x is
amount, y is depth from 0 to 2, color by group, in polar."*

No atom in that sentence is about donuts. `limits` on `depth` also leaves the
middle of the three-level [sunburst](polar.qmd) empty, and leaves an empty band
under the flat icicle, which has no `polar()`. So the hole comes from the scale,
not from `polar()`. Only the plain pie cannot have a hole this way: nothing is
bound to its radius, so there is no domain to widen.

The marginal plot is the same again, this time between whole plots. ggExtra adds
`ggMarginal()` to ggplot2 and patchwork adds a layout algebra beside it; here a
page is `|` and `/` between two plots, and the marginals follow from one rule
about what a shared column means ([Composition](composition.qmd)):

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 130)) /
  ((data(gapminder_2007) + point + x(gdp) + y(life)) |
     (data(gapminder_2007) + bar * bin + y(life) + theme(width = 130)))
```

*"Given gapminder 2007: bars derived by bin, x is gdp; above points, x is gdp, y
is life, beside bars derived by bin, y is life."*

Those are all the differences in this direction. [What the extensions
add](#what-the-extensions-add) shows more plots that ggplot2 draws only with an
extension package. Everything else gog draws, ggplot2 draws too, usually with
more settings.

## What is different rather than missing

The rest is not about pictures at all. These rows are differences in how the two
packages work: each one follows from what its own package was built for.

| | ggplot2 | gog |
|---|---|---|
| Where it runs | R | R, Python, Julia and JavaScript, one engine, the same SVG out of each |
| What a plot *is* | R closures, alive in the session | an engine-neutral JSON specification you can store, compare, send or generate |
| A sentence the package cannot draw | a warning, and a plot with those rows removed (`Removed 12 rows containing missing values`) | no plot, and a message naming the sentence it would accept |
| What combines with what | documented per geom | one rule table, every pair decided, dumped from the engine into [Combinations](combinations.qmd) |
| Picking out a subset with the pointer | outside the package: the plot is a still image, and the answer is plotly or shiny | `brush` is a word in the sentence, so the selection appears in the PDF as well as on screen. It dims what you did not pick and removes no rows. [Selection](selection.qmd) is the chapter |
| Moving overlapping labels apart | ggrepel, a separate package, with a geom of its own to use instead | `text * repel`, a collision modifier beside `dodge`, `stack` and `jitter`, so it composes with the `text` mark that was already there. It draws a thin line back to the point when a label has moved far, and where no arrangement fits, it still draws every label and reports how many overlap |
| Spellings per plot | often two (`geom_bar`/`geom_col`, `geom_histogram`/`stat_bin`) | one |
| Saying "log" | `scale_x_log10()`, a scale function per transformation | `x(gdp, scale = "log")`, written on the binding |
| Non-positive values on a log axis | dropped, with a warning | refused, with the fix named |
| A mean under a log axis | the geometric mean | computed in the data's own units |
| Centering a diverging ramp | `scale_color_gradient2(midpoint = 0)`, which fixes the midpoint and stretches each side to the data's own ends | `limits = c(-k, k)`, the atom that already states a domain. The middle of a domain *is* a midpoint, so no parameter is added, and the ramp is linear on each side of the middle. |

## What ggplot2 draws that gog does not

The gaps below are sorted into four tables by reason, because the reason tells
you whether a gap is final. A gap with a written ruling behind it is a different
thing from a gap nobody has ruled on.

### Refused, with the ruling on record

gog will not draw these. Each has a written ruling, and the ruling names the
direction to go instead. [Design laws](design-laws.qmd) is where the laws behind
them live.

| In ggplot2 | The ruling |
|---|---|
| `sec_axis()`, `dup_axis()` | A second y-axis is two coordinate spaces drawn in one panel. Facet on the measure, or layer if the units really are shared. |
| `annotate()` at page coordinates, `annotation_custom()`, `annotation_raster()` | A shape at pixel (120, 340) is a drawing command, and the specification has to mean the same thing at another size, in a facet, or on a sphere. Annotation in *data* space needs no new word: layer a mark over a small table. |
| `ggproto()`, user-written geoms and stats | Every promise on the [Combinations](combinations.qmd) grid depends on the vocabulary being fixed. An atom missing from the rule table has no row on that grid, breaks No Exceptions, and cannot be written into the JSON. Compose instead. |
| `coord_flip()` | Nothing to write: orientation is read from which axis carries the categories, so swapping the two bindings turns the plot sideways. ggplot2 has since superseded it. |
| `geom_bar()` beside `geom_col()`; `geom_histogram()` beside `stat_bin(geom = "bar")` | gog spells each of these one way. `bar` draws your numbers, `bar * count` counts them, `bar * bin` bins them. |
| `geom_smooth(method = "lm")` | A family of fits behind one argument is exactly the enumeration this grammar refuses. A least-squares line is a different statistic from a local smoother, so it needs its own plain name. |
| `scale_x_log10()`, `scale_x_sqrt()` as *names* | One parameter derives the family: `base = 2`, `base = 10`, `base = exp(1)`. |
| `coord_polar(theta = "y")` | Which axis is the angle is read from the bindings, not chosen by an argument. |
| `qplot()`, and plotting bare vectors | A plot is a mapping from a table. Put the vectors in one. |
| `geom_vline()`, `geom_hline()` as separate marks | Three marks for one idea. They are one `rule`, which takes a *column*, so a table of thresholds draws all of them at once, and which is also the rug via `style(reach = "edge")`. |
| plotmath, `label_parsed()`, `label_bquote()` | A TeX engine belongs to one output format, not to the grammar. Unicode covers what readers need: superscript powers on log ticks, Greek, sub- and superscripts in a label. |
| `coord_cartesian(xlim = )` | This ggplot2 name covers two different operations. *Narrowing the domain and running the statistics again* is `limits`, which already reports how many rows it removed. *Magnifying the finished picture* changes how you look at it and not what the plot claims. So magnifying needs no word. Under every plot in the web edition, `−` and `+` magnify the picture, and [the frame button](utilities.qmd#the-buttons-under-a-plot) returns to the whole of it. |
| plotly-style interaction: click a point, and read what it is | This needs no word in the kernel, and the survey of what other tools do is what decided it. Plotly hands your application an *event*; Vega-Lite puts a predicate over named fields in the specification; D3 asks the author for a row identity, and uses it for animation rather than for selection. gog answers each of the three elsewhere. Reading a row belongs to the page a plot is shown on, not to the sentence. On a flat plot that carries `brush`, resting the pointer on a mark shows that row's values. A click keeps them on the plot as a [stamp](selection.qmd). A predicate over a column is `brush`. A row identity is what an animation needs, and the grammar has no word for it ([Animation](play.qmd)). |
| a free-shape selection, drawn around some points with the pointer | There is no `lasso` word in the kernel. Every other selection can be written down: `brush(gdp, at = c(1200, 45000))` names a range, and the PDF then shows the plot that range asks for. A free shape can only be written as a long list of corners, which nobody writes in a sentence. The *act* was not refused, only the word: in the web edition you draw the shape with the pointer and the sentence still says `brush`. |

The refusal a ggplot2 reader is most likely to meet is the first one, so
here it is, live. Two series in different units, the second asking for its own
scale:

```{r}
#| echo: false
peek(quarterly)
```

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) + line +
  point + y(sales, scale = "log")
```

Every refusal in gog has that shape: it says what the rule is, and what to write
instead.

### The host language's job

These are computations, not pictures. gog draws the table you give it. Compute
the fitted value, the theoretical quantile or a model turned into a table before
the plot, in R, Python, Julia or JavaScript. Those languages already have better
tools for them.

| In ggplot2 | Where it goes |
|---|---|
| `stat_ecdf()` | compute the cumulative shares, then `step` |
| `stat_qq()`, `stat_qq_line()` | compute the theoretical quantiles, then `point` and `line` |
| `stat_ellipse()` | compute the ellipse, then `path` |
| `stat_quantile()` | fit the quantile regression, then `line` |
| `stat_function()`, `geom_function()` | evaluate the function onto a grid, then `line` |
| `stat_unique()` | de-duplicate the table |
| `cut_interval()`, `cut_number()`, `cut_width()` | cut the column, or use `bin` |
| `mean_cl_boot()` | bootstrap a lower and an upper bound, then `ribbon * bounds(lower, upper)`. Two helpers documented beside it are gog atoms instead: `mean_sdl()` is `deviation`, and `median_hilow()` is `range(0.025, 0.975)` with `point * median` |
| `fortify()`, `autoplot()`, `map_data()` | get the model or the map boundaries in the host, then give gog a table |
| a correlation matrix | `cor()` where the data lives, reshape it long, then `zone + x(var1) + y(var2) + color(r)` |

The table makes the split between gog and the host language look simpler than it
is. gog *does* compute: `bin` cuts, `density` estimates, `smooth` fits a local
regression, `confidence` builds an interval, `deviation` measures spread, and
`quantile` cuts a distribution. The test is whether the picture can exist
without the computation. A histogram has no bars until something bins. A
quantile regression is a separate answer that is then drawn as a line. How the
test applies to a new case is not settled.

### Missing, with the direction already settled

gog does not draw these. Each one belongs to a family the grammar already has,
and nothing about them is disputed.

| In ggplot2 | Here |
|---|---|
| `labs(subtitle =, caption =)` | more of the label family; `title()` is the one that exists |
| `geom_sf()`, `geom_map()`, `coord_sf()`, `borders()` | the `map` space draws, so a mark at a place on a projected earth is an ordinary sentence, and `zone` with `group(country)` fills each region from a boundary the data carries, holes included ([Map](map.qmd)). gog reads no spatial file. A boundary arrives as an ordinary table of points, which is the shape `map_data()` returns |
| `geom_dotplot()` | gog would need a mesh that follows the data's clusters, and a second rule for choosing bin edges |
| `scale_x_binned()`, equal-count bins | the other cutpoint rule, the same parameter |

### Not decided either way

None of these has a ruling in either direction. [Many
groups](cookbook/groups.qmd#where-does-a-population-pyramid-come-from) gives
reasons against custom tick text, but that is an argument, not a ruling.

| In ggplot2 | What is missing here |
|---|---|
| `scale_*_continuous(breaks =, labels =)` | tick *positions* and tick *text*. `tick_count` asks for a number of ticks and the engine chooses them, so "tick at 0, 50 and 100" and relabeling a category cannot be written. |
| `scale_y_reverse()` | a reversed continuous axis. `x(continent) + order(continent, desc = TRUE)` sorts a categorical axis from Z to A, or a [declared order](data.qmd#declaring-the-order-of-categories) from last to first. |
| `scale_x_sqrt()`, `coord_trans()` | a power scale. The `log10`-style *naming* is refused; the capability never was. |
| `linewidth` as an aesthetic | a column mapped to stroke width. `size` can be set on a stroke for the whole layer, not mapped per row. |
| `labs(alt = )`, `get_alt_text()` | alt text on the rendered plot |
| `geom_label()` | text in a box, rather than bare `text` |
| `geom_curve()` | a curved connector between two points; `path` joins them straight |
| `after_stat()`, `after_scale()`, `stage()`, and `geom_count()` with them | sending a transform's *output* to a channel rather than to a position. `point * count + size(count)` is refused, because the count a transform invented is not a column the sentence can name, so sizing a point by how many rows landed on it has no spelling. In ggplot2, all four functions in this row do this one job, and it is the largest gap on this list. |
| `geom_contour()` over a supplied `z` | iso-lines of a field you computed. `path * density` traces a field gog estimated. |
| `aes(angle = )` on `geom_text()`, and ggplot2's polar text generally | a separate angle for each label. `style(angle = )` turns every label in a layer by the same angle, so a [sunburst](polar.qmd) cannot turn each label along its own arc, and the labels of small wedges overlap. Every plot in `polar` has this gap, not only the sunburst. |

## What the extensions add

Every table above compares gog with ggplot2 itself. Many plots drawn with
ggplot2 come instead from its extension packages: separate packages that add new
plots to ggplot2. gog already draws several of these plots: the node-link
diagram in [Network](network.qmd), the alluvial diagram in [Flow](flow.qmd), the
[beeswarm](marks/point.qmd#a-categorical-x-the-strip-plot),
[ridgelines](marks/ribbon.qmd#letting-the-ridges-overlap), a
[halo](marks/text.qmd#a-halo-around-each-label) behind a label, and animation
with [`play`](play.qmd). Other extension plots need a computation first, then an
ordinary drawing. The host language computes the tables, and gog draws them with
words it already has. Five of these plots follow.

### Stars from a significance test

Packages such as ggsignif and ggpubr mark a plot with the result of a
statistical test. The test is a computation, so it belongs to the host language,
which builds two small tables from its result. One holds the corners of each
bracket: a line that joins the two groups compared, with a short leg down at
each end. The other holds the stars that stand for each p-value. Here a t-test
compares the petal length of versicolor, and then of virginica, with that of
setosa, the group the other two are compared with:

```{r}
#| echo: false
peek(petal_brackets)
peek(petal_stars)
```

```{r}
data(iris_flowers) + box + x(species) + y(petal_length) +
  data(petal_brackets) + path + group(pair) +
  data(petal_stars) + text + label(stars) + style(nudge = "up") +
  y_label("Petal length") + title("Each species against setosa")
```

*"Given the iris flowers: boxes, x is species, y is petal length, and also paths
from the petal brackets, grouped by pair, and also text from the petal stars,
label by stars."*

Each bracket is four rows, which `path` joins in order: up from setosa, across,
and down to the species compared. `group(pair)` keeps the two brackets apart.
The stars sit above the species compared, and three stars mean a p-value below
0.001.

### A survival curve with the numbers at risk

A Kaplan–Meier curve estimates, for each day, the share of patients still
alive: the survival probability that [Step](marks/step.qmd) draws. Packages such
as survminer add a table beneath the curve. It gives the number at risk every
200 days: the patients still alive and still followed by the study. The host
language fits the curve and counts those at risk. These two tables were computed
with R's survival package, from its data on 228 patients with advanced lung
cancer:

```{r}
#| echo: false
peek(lung_survival)
peek(lung_at_risk)
```

gog draws the curve as a `step` and the counts as `text`, in a second plot below
the first. The two plots share the `days` axis, and both ask for 8 ticks, so a
tick stands above each count:

```{r}
(data(lung_survival) + step + x(days, tick_count = 8) + y(surviving) +
   color(sex) + palette(c(Female = "#e15759", Male = "#4e79a7")) +
   y_label("Share surviving") +
   title("Survival of 228 patients with advanced lung cancer, by sex")) /
  (data(lung_at_risk) + text + x(days, tick_count = 8) + y(sex) +
     label(at_risk) + color(sex, legend = FALSE) +
     palette(c(Female = "#e15759", Male = "#4e79a7")) +
     theme(height = 110) + y_label("At risk"))
```

*"Given the lung survival table: step outlines, x is days with 8 ticks, y is
surviving, color by sex; above text from the lung at risk table, x is days with
8 ticks, y is sex, label by at risk, color by sex, without a legend."*

Each plot chooses its own colors, so the palette is written in both plots. It
ties each sex to one color, whatever the order of the rows, and `legend = FALSE`
leaves the second legend out. Women lived longer in this study: at 400 days the
curve shows about half of them alive, against under a third of the men. The
counts fall faster than the curves, because 63 patients left the study alive,
and a patient counts as at risk only while the study still follows them.

### Names at the ends of the lines

Packages such as directlabels write each line's name at its end, so a reader
does not have to match colors against a legend. The lines come from
`gapminder_asia`. The names come from a second table, with one row for each
line, at its last year, 2007:

```{r}
#| echo: false
peek(gapminder_asia)
peek(asia_2007)
```

```{r}
data(gapminder_asia) + line + x(year, limits = c(1952, 2016)) + y(life) +
  color(country, legend = FALSE) +
  data(asia_2007) + text + label(country) + color(country) +
  style(nudge = "right") +
  y_label("Life expectancy") + title("Each line named at its end")
```

*"Given gapminder Asia: lines, x is year from 1952 to 2016, y is life, color by
country, without a legend, and also text from the Asia 2007 table, label by
country, color by country."*

The names replace the legend, so `legend = FALSE` leaves the legend out.
`color(country)` is written again after `text`, because a channel written after
one mark reaches that mark alone ([Encoding scope](encoding-scope.qmd)). The
`limits` on `year` extend the axis to 2016, past the last year of data, so the
names fit inside the panel.

### How sets overlap

When a person can give several answers, the question is how many gave each
combination. An UpSet plot shows this. Packages such as UpSetR and ComplexUpset
draw one bar for each combination of sets. Under the bars, a grid of dots shows
which sets belong to each combination. Suppose 85 people were asked where they
get their news: from a printed newspaper, the radio or the web. The host
language counts each combination of answers and writes three tables.
`news_counts` holds the count for each combination, largest first, which is the
order the bars stand in. `news_members` holds one row for each combination and
source, saying whether the source is in it. `news_links` keeps only the sources
that are, for the line that joins them:

```{r}
#| echo: false
peek(news_counts)
peek(news_members)
peek(news_links)
```

```{r}
(data(news_counts) + bar + x(combination) + y(people) + x_label("") +
   y_label("People") + title("Where 85 people get their news")) /
  (data(news_links) + path + x(combination) + y(source) + group(combination) +
     style(size = 2, color = "#333333") +
     data(news_members) + point + color(member, legend = FALSE) +
     palette(c(yes = "#333333", no = "#d6d6d6")) + style(size = 7) +
     x_label("") + y_label(""))
```

The two plots share the `combination` axis, so each bar stands above its column
of dots. A dark dot marks a source in the combination, and a line joins the dark
dots of one combination.

### A horizon chart

A horizon chart draws a long time series in a panel of small height. Each value
is measured from one chosen value, here the mean. Differences above it rise from
the baseline in shades of one color. Differences below it are turned upward too,
in shades of a second color. Each band holds a slice of fixed height, and the
part of a difference that does not fit is drawn again from the baseline, in a
darker shade, over the lighter band. Packages such as ggHoriPlot draw it.
Cutting each difference into slices is simple arithmetic, so the host language
does it. This table holds New York's daily high temperature from May to
September 1973, in degrees Fahrenheit, from R's airquality data. Each row holds
one day and one band: the part of that day's distance from the mean that falls
in the band, with three bands above the mean and three below:

```{r}
#| echo: false
peek(nyc_horizon)
```

The first rows are 0 because the first days of May were cooler than the mean,
so those days have nothing in the warm bands.

```{r}
data(nyc_horizon) + area + x(day) + y(degrees) + color(band) +
  palette(c(warm = "#fcbba1", warmer = "#fb6a4a", warmest = "#a50f15",
            cool = "#c6dbef", cooler = "#6baed6", coolest = "#08519c")) +
  style(opacity = 1) + y_label("") +
  title("New York's daily highs, May to September 1973") / facet(month)
```

*"Given the nyc horizon table: areas, x is day, y is degrees, color by band,
split into panel rows by month."*

Each month is one panel. Red shows the days warmer than the mean, blue the
cooler days, and a darker band a larger difference. `style(opacity = 1)` draws
each band solid, so a darker band hides the lighter one beneath it. The bands
are drawn in the order they first appear in the table, so the table lists each
lighter band before the darker one drawn over it.

## If your plot is not here

This chapter compares gog with ggplot2's reference index, not with its
extensions. So a plot from an extension may be absent from this chapter, unless
[What the extensions add](#what-the-extensions-add) draws it.

For a plot you can name, [Chart names](chart-names.qmd) is the dictionary and
says which ones are drawn. For a question rather than a name, start at the
[cookbook](parts/cookbook.qmd).
