# Coverage, both directions

```{r}
#| include: false
source("R/setup.R")
```

The [phrasebook](phrasebook.qmd) answers "how do I spell this in gog". This
appendix answers the question underneath it: *can* it be spelled. Every entry in
ggplot2's reference index was read against the kernel, and the result is two
lists of very different lengths.

The ggplot2 column is dated on purpose, because nothing here can check it:
**reference index of ggplot2 4.0.3, read 2026-07-26**
[@wickham2026ggplot2pkg]. Every gog claim below is
either a live plot on this page or a refusal the build re-runs, and ggplot2 is
deliberately not a dependency of this book, so its half is the half that will age.

The short list is the one people expect to be long, so it goes first and is not
padded. The long list is the interesting one, because every item on it is one of
four things, and three of the four were decided on purpose.

## What gog draws that ggplot2 does not

Three things, at the level of pictures.

### Anything in the cube

Binding `z` stands a plot up off the page, and `surface` draws a height over a
plane:

```{r}
data(maunga_whau) + surface + x(east) + y(north) + z(elevation) +
  color(elevation) + palette("plasma") +
  title("Maunga Whau, one row per crossing")
```

Six marks draw in the cube (`point`, `path`, `bar`, `interval`, `box` and
`surface`), the viewing angle is `space(turn = , tilt = )`, and the cube composes
with the two partitions like anything else: `| facet(g)` gives one projected box
per panel and `+ play(t)` runs the sequence inside one. ggplot2 has no
depth position and no coordinate system that would read one, and says so outright
in its own first chapter, so none of this is expressible in it: in R these plots
come from plotly, rgl or lattice, each with a grammar of its own. Here it is one
more vowel in the same sentence, and [Space](space.qmd) is the chapter.

### Texture, mapped from a column

`pattern` maps a categorical column to a hatch on fills and a dash on strokes.
Put it on the same column as `color` and the series are marked twice over,
in one merged key:

```{r}
data(iris_flowers) + bar * mean + x(species) + y(petal_length) +
  color(species) + pattern(species) +
  title("Hue and hatch, one legend")
```

ggplot2 has `linetype` for strokes and nothing for the inside of a filled shape;
hatching a bar there means the third-party ggpattern. So the print-safe and
color-blind-safe bar chart above is a gog sentence with no ggplot2 equivalent in
the box.

That is the whole list. Everything else gog draws, ggplot2 draws too, and
usually with more knobs on it.

### The same plot, without a word for it

Not a picture ggplot2 lacks, but one this grammar draws without being taught it.
`geom_violin()` is a mark of its own there, with `stat_ydensity()` beside it; here
it is [`ribbon * density`](marks/ribbon.qmd#the-violin-a-distribution-per-category),
the mark that closes on a second boundary, given the transform that estimates a
distribution, over a category:

```{r}
data(gapminder_2007) + ribbon * density + x(continent) + y(life) +
  y_label("Life expectancy") + title("ribbon * density: no new word required")
```

The half violin comes with it, for nothing, because `area` and `ribbon` already
differ by where a region closes: `area * density + x(continent) + y(life)`. That
is what the orthogonality is *for*: not that the plot is possible, but that the
vocabulary did not grow to make it so.

The donut is the same story arriving from a *parameter* rather than a geom.
ggplot2 spells it `coord_radial(inner.radius = )`, a number the space is told; here
the hole is not a property of the circle at all. It is a stretch of the radial axis
with nothing standing on it, so a tree one level deep, drawn on an axis that starts
below the first ring, is a donut:

```{r}
data(spending) + zone * partition(group) + x(amount) +
  y(depth, limits = c(0, 2)) + color(group) + polar() +
  title("a donut is a sunburst of a one-level tree")
```

Nothing there is about donuts, and the same `limits` that hollow this out hollow
the three-level [sunburst](polar.qmd) and leave an empty band under the flat
icicle, which is what tells you the hole belongs to the scale rather than to the
bend. The one reading that cannot state a hole is the plain pie, whose radius is a
constant with no domain to widen.

The marginal plot is the same story one level up. ggExtra adds `ggMarginal()` to
ggplot2 and patchwork adds a layout algebra beside it; here a page is `|` and `/`
between two plots, and the marginals fall out of one rule about what a shared
column means ([Composition](composition.qmd)):

```{r}
(data(gapminder_2007) + bar * bin + x(gdp) + theme(height = 130)) /
  ((data(gapminder_2007) + point + x(gdp) + y(life)) |
     (data(gapminder_2007) + bar * bin + y(life) + theme(width = 130)))
```

## What is different rather than missing

The rest is not about pictures at all. These rows are differences in how the two
packages work, not a scoreboard: each one follows from what its own package set
out to be.

| | ggplot2 | gog |
|---|---|---|
| Where it runs | R | R, Python, Julia and JavaScript, one engine, the same SVG out of each |
| What a plot *is* | R closures, alive in the session | an engine-neutral JSON description you can store, diff, send, or generate |
| A sentence the package dislikes | a warning, and a plot with those rows removed (`Removed 12 rows containing missing values`) | no plot, and a message naming the sentence it would accept |
| What combines with what | documented per geom | one rule table, every pair decided, dumped from the engine into [Combinations](combinations.qmd) |
| Picking out a subset with the mouse | outside the package: the plot is a still image, and the answer is plotly or shiny | `brush` is a word in the sentence, so the printed page shows the selection too. It dims what you did not pick and removes no rows. [Selection](selection.qmd) is the chapter |
| Spellings per plot | often two (`geom_bar`/`geom_col`, `geom_histogram`/`stat_bin`) | one |
| Saying "log" | `scale_x_log10()`, a scale function per transformation | `x(gdp, scale = "log")`, written on the binding |
| Non-positive values on a log axis | dropped, with a warning | refused, with the fix named |
| A mean under a log axis | the geometric mean | computed in the data's own units |
| Centering a diverging ramp | `scale_color_gradient2(midpoint = 0)`, which pins the midpoint and stretches each arm to the data's own ends | `limits = c(-k, k)`, the atom that already states a domain. The middle of a domain *is* a midpoint, so no parameter is added, and the ramp is linear on both arms. |

## What ggplot2 draws that gog does not

Sorted by *why*, because the sort is the useful part. A gap that was argued and
lost is a different thing from a gap nobody has looked at.

### Refused, and the reason is on record

These will not arrive. Each has a written ruling, and the ruling names the
direction to go instead. [Design laws](design-laws.qmd) is where the laws behind
them live.

| In ggplot2 | The ruling |
|---|---|
| `sec_axis()`, `dup_axis()` | A second y-axis is two coordinate spaces wearing one panel. Facet on the measure, or layer if the units really are shared. |
| `annotate()` at page coordinates, `annotation_custom()`, `annotation_raster()` | A shape at pixel (120, 340) is a drawing command, and the description has to mean the same thing at another size, in a facet, or on a sphere. Annotation in *data* space needs no new word: layer a mark over a small table. |
| `ggproto()`, user-written geoms and stats | The guarantees rest on a closed set. An atom the rule table never heard of is a hole in the completeness test, in No Exceptions, and in the wire format at once. Compose instead. |
| `coord_flip()` | There is no `flip` atom, because orientation is read off which axis carries the categories. ggplot2 has since superseded it. |
| `geom_bar()` beside `geom_col()`; `geom_histogram()` beside `stat_bin(geom = "bar")` | gog spells each of these one way. `bar` draws your numbers, `bar * count` counts them, `bar * bin` bins them. |
| `geom_smooth(method = "lm")` | A family of fits behind one buried argument is exactly the enumeration this grammar refuses. A least-squares line is a different statistic from a local smoother, so it earns its own plain name when it is built. |
| `scale_x_log10()`, `scale_x_sqrt()` as *names* | One parameter derives the family: `base = 2`, `base = 10`, `base = exp(1)`. |
| `coord_polar(theta = "y")` | Which axis is the angle is read off the bindings, not chosen by an argument. |
| `qplot()`, and plotting bare vectors | A plot is a mapping from a table. Put the vectors in one. |
| `geom_vline()`, `geom_hline()` as separate marks | Three marks for one idea. They are one `rule`, which takes a *column*, so a table of thresholds draws all of them at once, and which is also the rug via `style(reach = "edge")`. |
| plotmath, `label_parsed()`, `label_bquote()` | A TeX engine is one backend's typesetting. Unicode does the real work: superscript powers on log ticks, Greek, sub- and superscripts in a label. |
| `coord_cartesian(xlim = )` | One name here covers two different operations. *Narrowing the domain and running the statistics again* is `limits`, which already counts the rows it left out aloud. *Magnifying the finished picture* changes how you look and not what the plot claims, so it earns no word: it is a button under every plot in the web edition, `−`, `+` and *fit*. |
| plotly-style interaction: click a point, and read what it is | The kernel reserved the word `click` for this and gave it up in August 2026, and the survey of what other tools do is what decided it. Plotly hands your application an *event*; Vega-Lite puts a predicate over named fields in the specification; D3 asks the author for a key function, and uses it for animation rather than for selection. gog already holds all three. Reading a row belongs to the page, so the pointer does it, and in the web edition a click leaves that reading on the plot as a stamp. A predicate over a column is `brush`. A key is what an animation needs. |
| a free-shape selection, drawn around some points with the mouse | The kernel reserved the word `lasso` for this and gave it up. Every other selection can be written down: `brush(gdp, at = c(1200, 45000))` names a range, and the printed page then shows the plot that range asks for. A free shape can only be written as a long list of corners, which nobody writes in a sentence. The *act* was not refused, only the word: in the web edition you draw the shape with the mouse and the sentence still says `brush`. |

The refusal a ggplot2 reader is most likely to walk into is the first one, so
here it is, live. Two series in different units, the second asking for its own
scale:

```{r}
#| error: true
data(quarterly) + x(year) + y(sales) + line +
  point + y(sales, scale = "log")
```

The message is the pattern: what the rule is, and what to write instead.

### The host language's job

These are computations, not pictures. gog draws what you hand it, and the fitted
value, the theoretical quantile and the reshaped model belong upstream in R,
Python, Julia or JavaScript, where you already have better tools for them.

| In ggplot2 | Where it goes |
|---|---|
| `stat_ecdf()` | compute the cumulative shares, then `step` |
| `stat_qq()`, `stat_qq_line()` | compute the theoretical quantiles, then `point` and `line` |
| `stat_ellipse()` | compute the ellipse, then `path` |
| `stat_quantile()` | fit the quantile regression, then `line` |
| `stat_function()`, `geom_function()` | evaluate the function onto a grid, then `line` |
| `stat_unique()` | de-duplicate the table |
| `cut_interval()`, `cut_number()`, `cut_width()` | cut the column, or use `bin` |
| `mean_cl_boot()`, `mean_sdl()`, `median_hilow()` | compute the pair, then `ribbon * bounds(lower, upper)` |
| `fortify()`, `autoplot()`, `map_data()` | reach the model or the shapefile in the host, hand over a table |
| a correlation matrix | `cor()` where the data lives, reshape it long, then `zone + x(var1) + y(var2) + color(r)` |

The boundary is not as clean as that table makes it look, and pretending
otherwise would be dishonest. gog *does* compute: `bin` cuts, `density`
estimates, `smooth` fits a local regression, `confidence` builds an interval. The
working line is that a histogram has no bars until something bins, so the
computation is the picture, whereas a quantile regression is a separate answer
that happens to be drawn as a line. Where exactly that line falls is one of the
design's open hard parts, not a settled rule.

### Designed and owed

Named in the kernel or written down as wanted, with nothing decided against
them. These are scheduling, not argument.

| In ggplot2 | Here |
|---|---|
| `labs(subtitle =, caption =)` | more of the label family, and cheap |
| `theme(legend.position =)` | it has its home in `theme()` and is waiting for a case that needs it |
| `geom_smooth(se = TRUE)` | the smoother's band, its missing pair form, exactly as `confidence` is the mean's |
| ggrepel | `text * repel`, designed, waiting on a layout stage |
| `geom_sf()`, `geom_map()`, `coord_sf()`, `borders()` | the `map` space draws, so a mark at a place on a projected earth is an ordinary sentence ([Map](map.qmd)). What is still owed is the polygon geometry underneath a choropleth: a filled region whose boundary comes from the data, with holes in it |
| `geom_dotplot()` | a cluster-aware mesh, a second cutpoint rule for binning |
| `scale_x_binned()`, equal-count bins | the other cutpoint rule, the same parameter |

### Not decided either way

The useful output of the sweep. Nothing has been argued about these, in either
direction, and they are now written down where a future session will find them.

| In ggplot2 | What is missing here |
|---|---|
| `scale_*_continuous(breaks =, labels =)` | tick *positions* and tick *text*. `tick_count` asks for a number of ticks and the engine chooses them, so "tick at 0, 50 and 100" and relabeling a category are both unsayable. |
| `scale_*_manual(values = c(setosa = "red"))` | binding a color to a level *by name*, shown below |
| `scale_y_reverse()` | a reversed continuous axis. `order(desc = TRUE)` reverses a categorical one. |
| `scale_x_sqrt()`, `coord_trans()` | a power scale. The `log10`-style *naming* is refused; the capability never was. |
| `guides(color = "none")`, `show.legend = FALSE` | turning a key off. Every mapped channel earns one, always. |
| `linewidth` as an aesthetic | a column mapped to stroke width. `size` can be set on a stroke for the whole layer, not mapped per row. |
| `labs(alt = )`, `get_alt_text()` | alt text on the rendered plot |
| `geom_label()` | text in a box, as against bare `text` |
| `geom_curve()` | a curved connector between two points; `path` joins them straight |
| `after_stat()`, `after_scale()`, `stage()`, and `geom_count()` with them | sending a transform's *output* to a channel rather than to a position. `point * count + size(count)` is refused, because the count a transform invented is not a column the sentence can name, so sizing a point by how many rows landed on it has no spelling. One question, and it is the widest one on this list. |
| `geom_contour()` over a supplied `z` | iso-lines of a field you computed. `path * density` traces a field gog estimated. |
| `geom_text(angle = )`, and ggplot2's polar text generally | a rotated label. `text` places a name at a point in every space, upright, so a [sunburst](polar.qmd) cannot run its labels along their arcs and small wedges collide. general to `polar`, not owned by that plot. |
| `theme(axis.text = element_blank())` and friends | saying a plot has **no readable axes** as a request. `theme(grid = )` removes gridlines and `tick_count = 0` is refused by design, so the silence has to be *earned from what an axis carries* rather than asked for: a pie is recognized from its bindings, a packing has no axes at all, and a partition's radial axis carries a ring index the transform invented and so draws no ticks, no rings and no name. Three recognitions doing a general job, and the residue is the case where a reader wants a *real* quantity's axis silent: the measure round a sunburst, say, which today takes `x_label("")` and keeps its numbers. |

The color one is worth seeing, because it is not a missing knob so much as a
missing *kind* of statement. Three species, three colors:

```{r}
data(iris_flowers) + point + x(petal_length) + y(sepal_length) +
  color(species) + palette(c("tomato", "steelblue", "seagreen")) +
  title("Rows in the order the table has them")
```

Now the same sentence and the same palette, over the same rows sorted the other
way. Watch the legend:

```{r}
by_species <- iris_flowers[base::order(iris_flowers$species, decreasing = TRUE), ]
data(by_species) + point + x(petal_length) + y(sepal_length) +
  color(species) + palette(c("tomato", "steelblue", "seagreen")) +
  title("The same three colors, the other way up")
```

Nothing is wrong here: a palette hands out its colors in the categories' order,
that order is the axis's, and for a text column the axis order is the order the
values first appear (a factor's levels are honored when you supply one). It is
consistent and it is documented. What there is no way to say is *setosa is
tomato*, so a category cannot keep its color across two plots whose category
order differs, which is what a report of small multiples usually wants.

## If your plot is not here

The sweep was of ggplot2's reference index, so a plot that lives in one of its
extensions may be missing from both columns. Two families are absent from gog
with nothing written down about them at all: networks (nodes and edges, ggraph's
territory) and any shape with a hole in it, which is the same geometry a
choropleth and a ring-shaped filled contour both need.

For a plot you can name, [Chart names](chart-names.qmd) is the dictionary and
says which ones are drawn. For a question rather than a name, start at the
[cookbook](cookbook/one-variable.qmd).
