# Data

```{r}
#| include: false
source("R/setup.R")
```

Where does a plot get its data? Every sentence starts by naming a table,
and everything after that depends on the name. This chapter covers the two
ways to name one, the shape the table must have, and the declarations that
travel with its columns.

## Binding a table

Most of the time the table is already loaded and you know what you want to ask.
The only work left is the sentence. `data()` takes a table you already hold in
your session. Column names after that are bare words: no quotes, no table
prefix, no indexing.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

gog looks up `gdp` and `life` inside the `gapminder_2007` table automatically.
This is the **Bind-Once Law**: name the table once, then use columns freely.

`gapminder_2007` here is an ordinary table in your session, whatever your
language calls one: a data frame in R, a dict of columns or a pandas DataFrame
in Python, a named tuple of columns or a DataFrame in Julia, an object of
columns in JavaScript. No registration step, no special wrapping required. Each
chapter in the bindings part shows its own. The [R](bindings/r.qmd) chapter
also lists the ten names that loading gog masks in R, with the qualified form
of each.

Writing `data()` does more than supply the numbers. When you write
`data(gapminder_2007)`, gog keeps the name `gapminder_2007`, not only the
numbers inside it. The name matters as soon as a plot has more than one table,
because each layer looks up its bare column names in the nearest table by name.
Leave `data()` out and the table arrives with no name, so the plot refuses and
tells you what to write:

```{r}
#| error: true
gapminder_2007 + point + x(gdp) + y(life)
```

gog gives the same advice for a sentence with no data at all, `point + x(gdp)`
with nothing before it. Each binding words the advice for its own spelling of a
column, so this refusal is shown in R alone.

## The table has to be tidy

Every table in this book has the same shape. Each row is one observation, and
each column is one variable. That shape has a name, **tidy data**
[@wickham2014tidy], and gog assumes it in every sentence you write.

The reason is the rule that defines a channel in [Grammar](grammar.qmd): a
channel maps a column to a visual property, so anything you want to map has to
*be* a column. Suppose you record sales for two regions, and you give each
region a column of its own:

```{r}
wide <- data.frame(year  = 2019:2023,
                   north = c(120, 135, 128, 152, 168),
                   south = c( 90, 101, 140, 133, 155))
```

```{r}
#| echo: false
peek(wide)
```

Now ask for one line per region:

```{r}
#| error: true
data(wide) + line + x(year) + y(sales) + color(region)
```

The message names two problems. There is no column called `sales`, and no column
called `region`. Your two regions are column *names*, and a name is not a
column. Your sales numbers sit in two columns instead of one. The sentence asks
for columns the table does not have, so gog draws nothing.

The same numbers, one row per observation, look like this:

```{r}
long <- data.frame(year   = rep(2019:2023, times = 2),
                   region = rep(c("north", "south"), each = 5),
                   sales  = c(120, 135, 128, 152, 168,
                               90, 101, 140, 133, 155))
```

```{r}
#| echo: false
peek(long, n = 10)
```

Now `sales` is a column and `region` is a column, so the sentence draws the
plot and earns the legend:

```{r}
data(long) + line + x(year) + y(sales) + color(region)
```

*"Given the long table: lines, x is year, y is sales, color by region."*

gog never reshapes a table, and no transform turns columns into rows. You do
that work before you write the sentence, in the language you are already using.
R has `tidyr::pivot_longer()`. Python has `pandas.melt()`. Julia has
`DataFrames.stack()`. JavaScript has no standard table library, so you write a
short loop instead.

gog is stricter than ggplot2 here, but not about the shape. ggplot2 does not
reshape tables either, and it refuses the wide table above just as gog does. The
real difference is narrower. ggplot2 lets you calculate a value while you map
it, so `aes(x = log(gdp))` works there. gog does not allow that, because a
channel takes a column name and nothing else. Some of those calculations are
already part of the grammar. A log axis is one of them, written
`x(gdp, scale = "log")`. For anything else, add the column to your table first.
So tidy is not quite enough on its own. Every column your sentence names has to
be in the table already.

This book does not teach you how to tidy data, and it should not, because there
is a better book for it. Read *R for Data Science* by Hadley Wickham, Mine
Çetinkaya-Rundel and Garrett Grolemund [@wickham2023r4ds], which is free online
at <https://r4ds.hadley.nz/>. Its chapter on data tidying is the one that
matters here. If the shape of your data is not settled yet, read that chapter
before you read this one.

## Plot-level vs. layer-level data

A line and the points it passes through come from one table. You should not
have to name it twice. `data()` at the start of the expression is the
**plot-level default**: it applies to all layers that don't declare their own
table.

```{r}
# Both layers read from "gapminder_2007"
data(gapminder_2007) + x(gdp) + y(life) +
  line +
  point
```

*"Given gapminder 2007: x is gdp, y is life, a line and also points."*

A forecast continues a history of past values, and the two usually arrive as
separate tables. You want both on one pair of axes. `actuals` holds the sales
of five years that have passed, and `forecast` holds the three years still to
come:

```{r}
#| echo: false
peek(actuals)
```

```{r}
#| echo: false
peek(forecast)
```

A `data()` that appears **between two marks** overrides the table for
the layer that follows it:

```{r}
# line uses "actuals"; point uses "forecast"
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points."*

The rule: the first `data()` is the plot's table, and each later `data()`
applies to the mark written directly after it. A mark with no `data()` of
its own reads the plot's table. The refusal at the start of this chapter
holds for the second table too: write `+ data(forecast)`, not `+ forecast`.

Two things follow from that rule, and both are easy to miss.

A `data()` at the end of a sentence applies to nothing, because no mark comes
after it. The plot is the same as it would be without that line:

```{r}
# `forecast` is never drawn, because no mark comes after it
data(actuals) + x(year) + y(sales) +
  line +
  point +
  data(forecast)
```

A `data()` also stops after that one mark. The mark after it reads the
plot-level table again:

```{r}
# line reads actuals. point reads forecast. area reads actuals again.
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point +
  area
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points; then given the actuals again: an area."*

The two tables cover different years, which makes the change easy to see.
`actuals` runs from 2019 to 2023, and `forecast` from 2024 to 2026. The area
and the line both sit on the left. The points sit alone on the right.

That rule fits what a second table is usually for. The first table builds the
frame, and a later one annotates it: a threshold, a band, a note, a forecast. An
annotation is one mark, so a `data()` that reaches one mark is enough.
Every second table in this book is one of those.

To draw two marks from the second table, write `data()` before each of them.

The specification above is correct, and it still reads badly. `area` sits on the line
below `data(forecast)`, so it looks like part of the forecast. Only the rule
says otherwise. The same plot reads better with the marks grouped by the table
they read. Drawn again, it is the same picture:

```{r}
data(actuals) + x(year) + y(sales) +
  line + area +
  data(forecast) + point
```

*"Given the actuals: x is year, y is sales, a line and also an area; then given
the forecast: points."*

This is style, not grammar. Both spellings are legal, and both draw the same
plot. The order of `area` and `point` has changed, and here that changes
nothing. The two tables share no year, so neither mark is drawn over the other.
gog will never refuse either spelling, because legal is not the same as clear
([Law 8](design-laws.qmd)). Two sentences can mean the same thing and still be
very different to read. That difference is yours to control.

So group the marks by the table they read. Keep each extra `data()` on the
same line as the mark it applies to, and keep the marks that read the
plot-level table together. Every line is then one table and the marks that
read it, and no mark sits under a table it does not read. The rest of this
book follows that layout.

The first `data()` is the one exception. It applies to every mark below it,
so it takes a line of its own. A later `data()` applies to one mark only. On a
line of its own, it would look like it applies to every mark below it, as the
first one does. That is why `data(actuals)` stands alone above and
`data(forecast)` shares its line with `point`.

## When the second table names its columns differently

The plot above was the easy case: `actuals` and `forecast` both call their
columns `year` and `sales`, so one pair of positions written at the front
served both layers.

Plots often carry a note at a chosen point, and the note comes from a small
table of its own. That table's column names rarely match the first table's.
The `milestones` table marks two points on the income axis:

```{r}
#| echo: false
peek(milestones)
```

Its columns are `at`, `value` and `note`. None of those names appear in
`gapminder_2007`, whose columns are `gdp` and `life`. A layer may name its own
columns, and the axis does not change:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at) + y(value) + label(note) +
  style(size = 20, color = "red")
```

*"Given gapminder 2007: points, x is gdp, y is life; then given the milestones:
text, x is at, y is value, label by note."*

There is still one x axis and one y axis, one scale each, one set of ticks. Only
the *names* differ, because the two tables were written by different people for
different reasons.

What is shared is the **measurement**, and that is what an axis is. `at` holds
income and `value` holds years of life, the same two quantities `gdp` and `life`
hold. A table whose numbers mean something else does not belong on these axes,
however its columns are spelled: put sales figures on an income axis and the
plot is drawn correctly and reads as nonsense. Matching names are not the test.
Matching meanings are.

The same test says when a second table does not belong in the plot at all. A
layer is drawn on the axes the first table set, so its values must measure
what those axes measure. Values that measure something else need axes of their
own, and axes of their own means a plot of its own. Put that plot beside the
first one instead. [Composition](composition.qmd) shows how, with `|` and `/`.

A layer that names no position of its own reads the axis that is already
there. That is why the earlier plot worked with the positions written once at
the front.

## A table in a database

Not every table fits on one computer. Most working data sits in a database,
and a **warehouse** (a database built to hold a company's whole history) can
hold billions of rows. gog reaches those tables with one
word, **`query`**, and it stands exactly where `data` stands.

Nothing else in the sentence changes:

```
data(orders)                          + bar + x(status)
query(connection, "SELECT ...")       + bar + x(status)
```

The SQL stays inside that one word; nothing after it is SQL. After it, the columns are
still bare names. The marks, channels, transforms and operators are the ones you
have already learned. You are not writing a second language for databases. You
are naming the rows a different way.

Here is a real database. The code below builds a small one in memory, so it
needs no server and no password:

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
DBI::dbWriteTable(con, "countries", gapminder_2007)
```

Now the sentence. The connection is the only new part:

```{r}
query(con, "SELECT gdp, life, continent FROM countries WHERE gdp > 5000") +
  point + x(gdp) + y(life) + color(continent)
```

*"Given the query: points, x is gdp, y is life, color by continent."*

The plot is an ordinary scatter plot, because the grammar did an ordinary thing.
The database chose which rows to send. gog drew what arrived.

A query and a loaded table produce
*the same picture, byte for byte*, when they hold the same rows. Every binding
tests exactly that. So `query` is never a different way of drawing; it is a
different way of finding the rows to draw.

### Three things to know

The connection is yours. gog never opens one, and it never stores a password.
You connect with whatever your language already uses, and hand the open
connection to `query`. That also means gog adds no database driver to your
install. A reader who never writes SQL never installs one.

The query runs when the plot is drawn, not when the sentence is written.
Writing `query(con, "...")` asks the database nothing. The sentence is a
specification, exactly like every other sentence in this book, and the engine
draws it later. The delay is not only neatness; it makes something else
possible. A specification the engine can read *before* the database runs is
one where the engine could rewrite the question, and ask for a summary
instead of the rows. Counting is a job a database does well, and the design
allows for that later.

Every language reaches a different set of databases. The word is the same in
all four. What differs is the connection you hand it, because each language has
its own database convention. Your language's chapter says which:
[R](bindings/r.qmd), [Python](bindings/python.qmd), [Julia](bindings/julia.qmd),
[JavaScript](bindings/javascript.qmd).

### How much data comes back

`query()` brings *every row the query returns* to the computer running gog. The
database does the selecting. Your machine does the drawing. Know that before
you run this against a large table.

For most tables it does not matter. A few thousand rows arrive quickly, so
write the simple query and stop there. Selecting the columns you
need and letting gog do the rest is the normal case, and the example above is
exactly that.

It starts to matter somewhere around a hundred thousand rows. Past that, moving
the rows costs more than drawing them. A warehouse table can hold billions, and
`SELECT * FROM that_table` would try to bring all of them to you.

The way to avoid it is already familiar to anyone who works with a warehouse:
*summarize in SQL, then draw the summary*. This works because a plot is nearly
always a summary anyway. A bar chart of a billion orders shows only a few bars,
and a database can produce those bars far more cheaply than your laptop can.

```{r}
query(con, "SELECT continent, COUNT(*) AS countries
            FROM countries GROUP BY continent") +
  bar + x(continent) + y(countries)
```

*"Given the query: bars, x is continent, y is countries."*

Five rows crossed the connection, not 142. The plot is the same
one you would get by counting locally.

### Two directions, and which one suits you

There is another way to give SQL users a grammar of graphics. Posit's **ggsql**
[@posit2026ggsql] adds visual clauses to SQL itself: you write `VISUALIZE`,
`DRAW` and `SCALE` inside the query, and the database returns a picture.

It has two strengths. Its clauses read like SQL, so someone who
writes SQL all day stays in the language they already think in, and no host
language is installed at all. And it runs the whole pipeline in the database, one
query per layer, so a bar chart of ten billion rows fetches only the bar heights.
On that second point ggsql does something gog does not. gog brings the rows to
you, so on a large table you write the summary in SQL yourself, as the count
above did.

If SQL is where you work, ggsql is probably the better fit. gog would first
ask you to learn R, Python, Julia or JavaScript, which is a real cost for
someone who uses none of them.

gog is built in the opposite direction. The grammar stays in one place and
the data comes to it, so one sentence draws a loaded table, a table in
Postgres, and a table in a warehouse, in any of four languages. One syntax,
in as many languages and places as we can reach.

| You work | The lower cost is |
|---|---|
| Only in SQL | **ggsql**: nothing to install, and the syntax is the one you use |
| Across several languages | **gog**: one grammar covers all of them, and the database too |

```{r}
#| include: false
DBI::dbDisconnect(con)
```

## Declaring the order of categories

Low, Medium and High have an order every reader knows. A plot that draws them
in any other order makes the reader sort them. A column of plain text
has no order of its own, so gog draws its categories in the order the rows
happen to arrive:

```{r}
severity <- data.frame(
  level = c("High", "Low", "Medium"),
  count = c(30.0, 10.0, 20.0)
)

data(severity) + bar + x(level) + y(count) +
  title("Text column: the order the rows came in")
```

That is rarely what you want here. Declare the order you mean, where you build
the table. In R the declaration is a **factor**. Python, Julia and JavaScript
have no factor type, so gog supplies the declaration itself, under one name:
`ordered()`. It takes the column's values and the order its categories go in,
and everything after it reads the same in all four languages:

```{r}
severity <- data.frame(
  level = factor(c("High", "Low", "Medium"), levels = c("Low", "Medium", "High")),
  count = c(30.0, 10.0, 20.0)
)

data(severity) + bar + x(level) + y(count) +
  title("Declared: the order you asked for")
```

The order goes with the column, so *every part of the plot agrees*: the axis,
the color assignment, and the legend all read in the same order.

```{r}
data(severity) + point + x(count) + y(count) + color(level) +
  title("The legend follows the same order")
```

The order holds even when gog does the counting for you. These tickets arrive
`High` first, but `bar * count` still reads Low, Medium, High. A summary carries
the declared order onto its axis. It does not return to the order the rows
arrived in: 

```{r}
tickets <- data.frame(
  level = factor(
    c("High", "Low", "High", "Medium", "High", "Low", "Medium", "High"),
    levels = c("Low", "Medium", "High")
  )
)

data(tickets) + bar * count + x(level) +
  title("bar * count: the levels still order the axis")
```

*"Given the tickets: bars derived by count, x is level."*

In R, the list of levels can mean two things. `factor(x, levels = ...)` says
only which order the categories appear in. Adding `ordered = TRUE` says more:
that Low really is *less than* High. gog draws the categories in the listed
order either way. Most people list levels only to set the order on the page,
and a plot needs nothing more than that.

In Python, a pandas table can carry the same declaration as a `Categorical`
column, and gog reads its categories, in their order, as the levels.
`ordered()` is for a plain dictionary of lists, so that you can declare an
order without pandas installed.

Two rules worth knowing:

- **Levels set the order.** The data sets which categories appear. A level
  with no rows in the data gets no bar and no slot on the axis, because an
  empty labeled gap is harder to read than a shorter axis. A value in the data
  that the levels do not mention is not dropped. It is drawn after the listed
  ones.
- **`order()` in the plot wins.** The levels came with the table, so they apply
  to every plot drawn from it. `order()` is written in this one plot, so it
  applies here, and the more specific instruction wins. Name a different column
  in `order()`, and the categories follow that column's values instead. That is
  how you sort by value:

```{r}
data(severity) + bar + x(level) + y(count) + order(count, desc = TRUE) +
  title("order() overrides the levels")
```

*"Given the severity table: bars, x is level, y is count, ordered by count,
largest first."*

`order()` can also name the categorical column itself, as in `order(level)`.
That asks for the column's own order. For a column of plain text, that order is
alphabetical. For a column with declared levels, it is the declared order.
`desc = TRUE` reverses whichever applies. So a declared order cannot be turned
alphabetical from inside the plot. To change it, change the levels where the
table was built, in the language you are already using. The declaration
belongs to the table, and so does the change.

## Missing values

Real data has gaps, and survey tables are usually missing many values. A row
with no value where the plot needs one cannot be placed, so gog drops it and
*says how many*, rather than changing the count without telling you:

```{r}
#| message: true
patchy <- data.frame(
  weight = c(3.1, 4.2, NA, 5.8, 6.0, NA, 4.9, 5.1, 3.7, 5.5),
  batch  = c("A", "A", "B", "B", NA, "B", "A", "B", "A", "B")
)
data(patchy) + bar * bin + x(weight) +
  x_label("Weight") + title("Two rows have no weight")
```

*"Given the patchy table: bars derived by bin, x is weight."*

Only the columns the plot maps decide which rows are dropped. This plot maps
`weight`, so the message names `weight`, and the two rows with no weight are
left out. The table also has one row with no `batch`. The plot never reads
`batch`, so that row stays in the histogram. Bind `batch` to `color`, and the
same row is dropped as well, because it now has no color to take. The message
then names both columns:

```{r}
#| message: true
data(patchy) + bar * bin + x(weight) + color(batch) +
  x_label("Weight") + title("One more column mapped, one more row dropped")
```

*"Given the patchy table: bars derived by bin, x is weight, color by batch."*

So a missing value in a column the plot never reads changes nothing. A missing
value in a column the plot maps always drops the row, and the message says
which column.

## Special characters and other writing systems

The people who built a table chose its names, and a plot has to print them
unchanged. Category names, titles, and axis labels are drawn
exactly as you wrote them. Three characters have a special meaning inside an
SVG file: `&`, `<` and `>`.
gog writes them in the form the file needs, so they print as themselves, and a
name like `R&D` needs no special handling. The plot below puts all three into
its names and its title:

```{r}
special <- data.frame(
  firm  = c("R&D", "Sales & Ops", "<10% unit"),
  spend = c(10, 25, 15)
)
data(special) + bar + x(firm) + y(spend) + title("Q3 <spend> & forecast")
```

Many tables are not written in Latin letters at all. Their column names should
be drawn as they are, with no renaming first. A name in another writing system
is also drawn as written, and it can be a bare column name as well as a label.
The table below is in Korean. Its name, 지역별 (*jiyeokbyeol*), means by region.
Its two columns are 지역 (*jiyeok*), region, and 값 (*gap*), value. The three rows
are the cities 서울 (*Seoul*), 부산 (*Busan*) and 대구 (*Daegu*). The plot title 한국
지역별 값 (*Hanguk jiyeokbyeol gap*) reads *value by region, Korea*.

```{r}
지역별 <- data.frame(
  지역 = c("서울", "부산", "대구"),
  값   = c(10, 25, 15)
)
data(지역별) + bar + x(지역) + y(값) + title("한국 지역별 값")
```

A label's width is measured from its characters, and a Korean or Chinese
character is about twice as wide as a Latin one, so these labels get the margin
they need and no more.

That is everything a sentence needs from its table. The next chapter shows
where this book's own tables come from, and the mark chapters follow it.
