# Data

```{r}
#| include: false
source("R/setup.R")
```

## Binding a table

Where does a plot get its numbers? `data()` takes a data frame directly from
your R environment. Column names after that are bare words: no quotes, no `$`,
no indexing.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

GOG looks up `gdp` and `life` inside the `gapminder_2007` data frame automatically.
This is the **Bind-Once Law**: name the table once, then use columns freely.

`gapminder_2007` here is just a regular R data frame (or tibble) in your workspace:
no registration step, no special wrapping required.

`data()` is not ceremony, and skipping it refuses with direction. The call is
what *names* the table. The name travels with the plot, later layers resolve
their bare columns against the nearest one, and a bare frame has no name to
bind:

```{r}
#| error: true
gapminder_2007 + point + x(gdp) + y(life)
```

The same advice comes back if a bare frame appears mid-expression (write
`+ data(forecast)`, not `+ forecast`) or if the sentence has no data at all
(`point + x(gdp)` with nothing before it).

## Plot-level vs. layer-level data

`data()` at the start of the expression is the **plot-level default**:
it applies to all layers that don't declare their own table.

```{r}
# Both layers read from "gapminder_2007"
data(gapminder_2007) + x(gdp) + y(life) +
  line +
  point
```

A `data()` that appears **between two marks** overrides the table for
the layer that follows it:

```{r}
# line uses "actuals"; point uses "forecast"
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) +
  point
```

The rule: **nearest `data()` wins.**

## Multi-table layered plots

Each layer can draw from a completely different data frame as long as `x` and `y`
map to comparable columns:

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) +
  point
```

## Passing data in R

Pass any R data frame or tibble directly, no setup required:

```r
library(gog)

# gapminder_df is a regular data frame in your environment
data(gapminder_df) + point + x(gdp) + y(life) + color(continent)
```

That first line prints a long list of masked names, and it is worth knowing now
that most of the list cannot affect you. `mean(x)` and `sum(x)` keep working,
because R skips a name that is not a function when it resolves a call. Eight
names do take over, `data` among them, and the [R](bindings/r.qmd) chapter lists
all eight with the qualified form of each.

Under the hood the table currently crosses to the engine as an in-memory
`DataFrame`; Apache Arrow (zero-copy, no serialization penalty even for
millions of rows) is on the roadmap to replace it.

## Factors decide the order of categories

A plain text column has no order of its own, so GOG draws its categories in the
order the rows happen to arrive:

```{r}
severity <- data.frame(
  level = c("High", "Low", "Medium"),
  count = c(30.0, 10.0, 20.0)
)

data(severity) + bar + x(level) + y(count) +
  title("Text column — the order the rows came in")
```

That is rarely what you want here. Make it a **factor** and say what order you
mean:

```{r}
severity$level <- factor(severity$level, levels = c("Low", "Medium", "High"))

data(severity) + bar + x(level) + y(count) +
  title("Factor — the order you declared")
```

The `levels =` travels with the column, so **every part of the chart agrees**:
the axis, the color assignment, and the legend all read in the same order.

```{r}
data(severity) + point + x(count) + y(count) + color(level) +
  title("The legend follows the same order")
```

The order holds even when GOG does the counting for you. These tickets arrive
`High` first, but `bar * count` still reads Low, Medium, High. A summary carries
the declared order onto its axis, it does not fall back to first-appearance:

```{r}
tickets <- data.frame(
  level = factor(
    c("High", "Low", "High", "Medium", "High", "Low", "Medium", "High"),
    levels = c("Low", "Medium", "High")
  )
)

data(tickets) + bar * count + x(level) +
  title("bar * count — the levels still order the axis")
```

R has two kinds of factor: `factor(x, ordered = TRUE)`, meaning Low really is
*less than* High, and a plain factor that merely lists its levels. GOG honors
the order in both cases, because most people reach for a plain factor purely to
fix the display order and mean nothing mathematical by it.

### The same declaration in the other three languages

`factor()` is R's word for this. Python, Julia and JavaScript have no factor
type, so GOG supplies the declaration itself, under one name: `ordered()`. It
takes the column's values and the order its categories go in. Declare it where
you build the table, and everything after that reads the same in all four
languages.

**R**

```r
severity <- data.frame(
  level = factor(c("High", "Low", "Medium"), levels = c("Low", "Medium", "High")),
  count = c(30.0, 10.0, 20.0)
)
```

**Python**

```python
severity = {
    "level": ordered(["High", "Low", "Medium"], ["Low", "Medium", "High"]),
    "count": [30.0, 10.0, 20.0],
}
```

**Julia**

```julia
severity = Dict(
    "level" => ordered(["High", "Low", "Medium"], ["Low", "Medium", "High"]),
    "count" => [30.0, 10.0, 20.0],
)
```

**JavaScript**

```js
const severity = {
  level: ordered(["High", "Low", "Medium"], ["Low", "Medium", "High"]),
  count: [30.0, 10.0, 20.0],
};
```

Python reads a pandas `Categorical` as well, and takes its categories as the
levels. `ordered()` is what a plain dictionary of lists uses, because GOG does
not require pandas.

Two rules worth knowing:

- **Levels say what order, the data says what is there.** A level with no rows
  draws no bar and gets no slot on the axis: an empty labeled gap is harder to
  read than a shorter axis. A value in the data that is *not* in the levels is
  never dropped; it goes on the end.
- **An explicit `order()` wins.** You wrote it in the plot, which is nearer than
  the table. Name a *different* column and it overrides the levels, which is what
  lets you sort by value:

```{r}
data(severity) + bar + x(level) + y(count) + order(count, desc = TRUE) +
  title("order() overrides the levels")
```

Name the **category column itself** and you are asking for that column's own
order, so there is nothing to override. A plain text column carries nothing but
its spelling and sorts alphabetically; a factor carries its levels, so it keeps
them, and `desc = TRUE` reverses them. That leaves no way to alphabetize a
factor's labels from inside the plot, and it is the right thing to lose:
relevel it in the host, where the levels are.

## Missing values

Real data has gaps: `penguins` is missing two flipper measurements, and most
survey data far more. A row with no value where the plot needs one cannot be
placed, so GOG drops it and **says how many**, rather than quietly changing the
count under you:

```{r}
#| message: true
patchy <- data.frame(
  weight = c(3.1, 4.2, NA, 5.8, 6.0, NA, 4.9, 5.1, 3.7, 5.5),
  batch  = c("A", "A", "B", "B", "A", "B", "A", "B", "A", "B")
)
data(patchy) + bar * bin + x(weight) +
  x_label("Weight") + title("Two rows have no weight")
```

The drop is scoped to the columns the plot **maps**. The message names `weight`
because `weight` is bound to `x`, and a row with no `x` has nowhere to stand. A
missing value in a column the plot never reads costs nothing: leave `batch`
unbound and its gaps are ignored; bind it to `color` and a row missing its batch
drops too, because now it has no color to take. This is the rule other plotting
tools follow: only the channels actually in use decide which rows survive, so a
stray `NA` in an unused column never silently shrinks the picture.

## Labels are used verbatim

Category names, titles, and axis labels are drawn exactly as they appear in your
data. Characters with meaning in the output format (`&`, `<`, `>`) are escaped
for you, so a department called `R&D` needs no special handling:

```{r}
special <- data.frame(
  firm  = c("R&D", "Sales & Ops", "<10% unit"),
  spend = c(10, 25, 15)
)
data(special) + bar + x(firm) + y(spend) + title("Q3 <spend> & forecast")
```

Non-Latin scripts work the same way, including as bare column names:

```{r}
지역별 <- data.frame(
  지역 = c("서울", "부산", "대구"),
  값   = c(10, 25, 15)
)
data(지역별) + bar + x(지역) + y(값) + title("한국 지역별 값")
```

Label widths are measured per character, so full-width scripts reserve the right
amount of margin rather than three times too much.
