# Data

```{r}
#| include: false
source("R/setup.R")
```

## Binding a table

Where does a plot get its numbers? `data()` takes a data frame directly from
your R environment. Column names after that are bare words: no quotes, no `$`,
no indexing.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

GOG looks up `gdp` and `life` inside the `gapminder_2007` data frame automatically.
This is the **Bind-Once Law**: name the table once, then use columns freely.

`gapminder_2007` here is just a regular R data frame (or tibble) in your workspace:
no registration step, no special wrapping required.

Writing `data()` is not just a formality. When you write
`data(gapminder_2007)`, GOG keeps the name `gapminder_2007`, not only the
numbers inside it. The name matters as soon as a plot has more than one table,
because each layer looks up its bare column names in the nearest table by name.
Leave `data()` out and the table arrives with no name, so the plot refuses and
tells you what to write:

```{r}
#| error: true
gapminder_2007 + point + x(gdp) + y(life)
```

The same advice comes back if a bare frame appears mid-expression (write
`+ data(forecast)`, not `+ forecast`) or if the sentence has no data at all
(`point + x(gdp)` with nothing before it).

## The table has to be tidy

Every table in this book has the same shape. Each row is one observation, and
each column is one variable. That shape has a name, **tidy data**
[@wickham2014tidy], and GOG assumes it in every sentence you write.

The reason is a rule you have already read. A channel takes a column, so
anything you want to map has to *be* a column. Suppose you record sales for two
regions, and you give each region a column of its own:

```{r}
#| error: true
wide <- data.frame(year  = 2019:2023,
                   north = c(120, 135, 128, 152, 168),
                   south = c( 90, 101, 140, 133, 155))

data(wide) + line + x(year) + y(sales) + color(region)
```

The message names two problems. There is no column called `sales`, and no column
called `region`. Your two regions are column *names*, and a name is not a
column. Your sales numbers sit in two columns instead of one. The sentence asks
for columns the table does not have, so GOG draws nothing.

The same numbers, one row per observation, draw the plot and earn the legend:

```{r}
long <- data.frame(year   = rep(2019:2023, times = 2),
                   region = rep(c("north", "south"), each = 5),
                   sales  = c(120, 135, 128, 152, 168,
                               90, 101, 140, 133, 155))

data(long) + line + x(year) + y(sales) + color(region)
```

GOG never reshapes a table, and no transform turns columns into rows. You do
that work before you write the sentence, in the language you are already using.
R has `tidyr::pivot_longer()`. Python has `pandas.melt()`. Julia has
`DataFrames.stack()`. JavaScript has no standard table library, so you write a
short loop instead.

**GOG is stricter than ggplot2 here, but not about the shape.** ggplot2 does not
reshape tables either, and it refuses the wide table above just as GOG does. The
real difference is narrower. ggplot2 lets you calculate a value while you map
it, so `aes(x = log(gdp))` works there. GOG does not allow that, because a
channel takes a column name and nothing else. Some of those calculations are
already part of the grammar. A log axis is one of them, written
`x(gdp, scale = "log")`. For anything else, add the column to your table first.
So tidy is not quite enough on its own. Every column your sentence names has to
be in the table already.

This book does not teach you how to tidy data, and it should not, because there
is a better book for it. Read *R for Data Science* by Hadley Wickham, Mine
Çetinkaya-Rundel and Garrett Grolemund [@wickham2023r4ds], which is free online
at <https://r4ds.hadley.nz/>. Its chapter on data tidying is the one that
matters here. If the shape of your data is not settled yet, read that chapter
before you read this one.

## Plot-level vs. layer-level data

`data()` at the start of the expression is the **plot-level default**:
it applies to all layers that don't declare their own table.

```{r}
# Both layers read from "gapminder_2007"
data(gapminder_2007) + x(gdp) + y(life) +
  line +
  point
```

A `data()` that appears **between two marks** overrides the table for
the layer that follows it:

```{r}
# line uses "actuals"; point uses "forecast"
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point
```

The rule: **each `data()` applies to the mark written directly after it.** A
mark with no `data()` in front of it reads the plot-level table.

Two things follow from that rule, and both are easy to miss.

A `data()` at the end of a sentence applies to nothing, because no mark comes
after it. The plot is the same as it would be without that line:

```{r}
# `forecast` is never drawn, because no mark comes after it
data(actuals) + x(year) + y(sales) +
  line +
  point +
  data(forecast)
```

A `data()` also stops after that one mark. The mark after it reads the
plot-level table again:

```{r}
# line reads actuals. point reads forecast. area reads actuals again.
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point +
  area
```

The two tables cover different years, which makes the change easy to see.
`actuals` runs from 2019 to 2023, and `forecast` from 2024 to 2026. The area
and the line both sit on the left. The points sit alone on the right.

That rule fits what a second table is usually for. The first table builds the
frame, and a later one annotates it: a threshold, a band, a note, a forecast. An
annotation is one mark, so an extra `data()` that reaches one mark is a fit
rather than a limit. Every second table in this book is one of those.

To draw two marks from the second table, write `data()` before each of them.

That sentence is correct, and it still reads badly. `area` sits on the line
below `data(forecast)`, so it looks like part of the forecast. Only the rule
says otherwise.

This is style, not grammar. Both spellings are legal, and both build the same
plot. GOG will never refuse either one, because legal is not the same as clear
([Law 8](design-laws.qmd)). Two sentences can mean the same thing and still be
very different to read. That difference is yours to control.

So group the marks by the table they read. **Keep each extra `data()` on the
same line as the mark it applies to, and keep the marks that read the
plot-level table together.** The plot above is better written with
`line + area` on one line, and `data(forecast) + point` on the next. Every line
is then one table and the marks that read it, and no mark sits under a table it
does not read. The rest of this book follows that layout.

The first `data()` in a sentence does apply to every mark below it, so a line
of its own suits it. A later one never does. Giving a later one its own line
makes it look like it reaches further than it does.

## When the second table names its columns differently

The plot above was the easy case: `actuals` and `forecast` both call their
columns `year` and `sales`, so one pair of positions written at the front served
both layers.

A second table usually does not match. The `milestones` table marks two points on
the income axis, and its columns are `at`, `value` and `note`. None of those
names appear in `gapminder_2007`, whose columns are `gdp` and `life`. A layer may
name its own columns, and the axis does not change:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at) + y(value) + label(note) +
  style(size = 20, color = "red")
```

There is still one x axis and one y axis, one scale each, one set of ticks. Only
the *names* differ, because the two tables were written by different people for
different reasons.

What is shared is the **measurement**, and that is what an axis is. `at` holds
income and `value` holds years of life, the same two quantities `gdp` and `life`
hold. A table whose numbers mean something else does not belong on these axes,
however its columns are spelled: put sales figures on an income axis and the
plot is drawn correctly and reads as nonsense. Matching names are not the test.
Matching meanings are.

That test decides whether the second table belongs in this plot at all. A layer
joins a frame the first table built, so it has to land on the axes that frame
already has. When it does not, the two tables are measuring different things,
and two plots on one page say that honestly where one plot cannot.
[Composition](composition.qmd) is the operator for it.

A layer that names no position of its own reads the axis, whoever set it. That is
why the earlier plot worked with the positions written once at the front.

## A table in a database

Not every table fits in memory. Most working data sits in a database, and a
warehouse table can hold billions of rows. GOG reaches those tables with one
word, **`query`**, and it stands exactly where `data` stands.

Nothing else in the sentence changes:

```
data(orders)                          + bar + x(status)
query(connection, "SELECT ...")       + bar + x(status)
```

The SQL sits inside that one word and goes no further. After it, the columns are
still bare names. The marks, channels, transforms and operators are the ones you
have already learned. You are not writing a second language for databases. You
are naming the rows a different way.

Here is a real database. The chunk builds a small one in memory, so it needs no
server and no password:

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
DBI::dbWriteTable(con, "countries", gapminder_2007)
```

Now the sentence. The connection is the only new part:

```{r}
query(con, "SELECT gdp, life, continent FROM countries WHERE gdp > 5000") +
  point + x(gdp) + y(life) + color(continent)
```

The plot is an ordinary scatter plot, because the grammar did an ordinary thing.
The database chose which rows to send. GOG drew what arrived.

That last point is worth stating plainly. A query and a table in memory produce
**the same picture, byte for byte**, when they hold the same rows. Every binding
tests exactly that. So `query` is never a different way of drawing; it is a
different way of finding the rows to draw.

### Three things to know

**The connection is yours.** GOG never opens one, and it never stores a password.
You connect with whatever your language already uses, and hand the open
connection to `query`. That also means GOG adds no database driver to your
install. A reader who never writes SQL never installs one.

**The query runs when the plot is drawn, not when the sentence is written.**
Writing `query(con, "...")` asks the database nothing. The sentence is a
specification, exactly like every other sentence in this book, and the engine
draws it later. That is worth more than tidiness. A specification the engine
can read *before* the database runs is one where the engine could rewrite the
question, and ask for a summary instead of the rows. Counting is a job a database
does well, and the design leaves that door open.

**Every language reaches a different set of databases.** The word is the same in
all four. What differs is the connection you hand it, because each language has
its own database convention. Your language's chapter says which:
[R](bindings/r.qmd), [Python](bindings/python.qmd), [Julia](bindings/julia.qmd),
[JavaScript](bindings/javascript.qmd).

### How much data comes back

`query()` brings **every row the query returns** to the computer running GOG. The
database does the selecting. Your machine does the drawing. That is worth
knowing before you point this at a large table.

For most tables it does not matter. A few thousand rows arrive quickly, and you
should write the simple query and think no further. Selecting the columns you
need and letting GOG do the rest is the normal case, and the example above is
exactly that.

It starts to matter somewhere around a hundred thousand rows. Past that, moving
the rows costs more than drawing them. A warehouse table can hold billions, and
`SELECT * FROM that_table` would try to bring all of them to you.

The way to avoid it is already familiar to anyone who works with a warehouse:
**summarize in SQL, then plot the summary**. This works because a plot is nearly
always a summary anyway. A bar chart of a billion orders shows a handful of bars,
and a database can produce those bars far more cheaply than your laptop can.

```{r}
query(con, "SELECT continent, COUNT(*) AS countries
            FROM countries GROUP BY continent") +
  bar + x(continent) + y(countries)
```

Five rows crossed the connection, not a hundred and forty. The plot is the same
one you would get by counting locally.

The common summaries all have a SQL form:

| The plot you want | What to ask the database for |
|---|---|
| Bars of counts | `SELECT key, COUNT(*) AS n ... GROUP BY key` |
| A histogram | `SELECT FLOOR(v / 5000) * 5000 AS bucket, COUNT(*) AS n ... GROUP BY 1` |
| A time series | `SELECT date_trunc('month', t) AS month, SUM(v) AS total ... GROUP BY 1` |
| A scatter plot of a huge table | a sample: `TABLESAMPLE`, or `USING SAMPLE 10000` in DuckDB |

Two marks are harder. `smooth` and `density` fit a curve through the individual
rows, and a curve fitted through counts is not the same curve. A database *can*
compute them, and some tools do, but GOG does not ask it to yet. For now, take a
sample for those two and say so in the caption.

Writing the summary is your job today. The design is built so that it might not
always be: `query()` holds its SQL instead of running it, which is what would let
the engine send the `GROUP BY` itself and receive only the summary. The sentence
you write would not change. That is the direction GOG is built for, rather than a
promise with a date on it, and the summarizing is yours until it arrives.

### Two directions, and which one suits you

There is another way to give SQL users a grammar of graphics. Posit's **ggsql**
[@posit2026ggsql] adds visual clauses to SQL itself: you write `VISUALIZE`,
`DRAW` and `SCALE` inside the query, and the database returns a picture.

It has two strengths and both are real. Its clauses read like SQL, so someone who
writes SQL all day never leaves the language they are thinking in, and no host
language is installed at all. And it runs the whole pipeline in the database, one
query per layer, so a bar chart of ten billion rows fetches only the bar heights.
**On that second point ggsql is ahead of GOG**, which does not do this today.
The limit described above is ours, not the idea's.

**If SQL is where you work, ggsql is probably the better fit.** GOG would first
ask you to pick up R, Python, Julia or JavaScript, and that is a real cost for
someone with no other reason to.

GOG points the other way. The grammar stays in one place and the data comes to
it, so one sentence draws a table in memory, a table in Postgres, and a table in
a warehouse, in any of four languages. **One syntax, as many languages and
settings as we can reach.**

| You work | The lower cost is |
|---|---|
| Only in SQL | **ggsql** — nothing to install, and the syntax is the one you use |
| Across several languages | **GOG** — one grammar covers all of them, and the database too |

```{r}
#| include: false
DBI::dbDisconnect(con)
```

## Passing data in R

Pass any R data frame or tibble directly, no setup required:

```r
library(gog)

# gapminder_df is a regular data frame in your environment
data(gapminder_df) + point + x(gdp) + y(life) + color(continent)
```

That first line prints a long list of masked names, and it is worth knowing now
that most of the list cannot affect you. `mean(x)` and `sum(x)` keep working,
because R skips a name that is not a function when it resolves a call. Eight
names do take over, `data` among them, and the [R](bindings/r.qmd) chapter lists
all eight with the qualified form of each.

Under the hood the table currently crosses to the engine as an in-memory
`DataFrame`; Apache Arrow (zero-copy, no serialization penalty even for
millions of rows) is on the roadmap to replace it.

## Factors decide the order of categories

A plain text column has no order of its own, so GOG draws its categories in the
order the rows happen to arrive:

```{r}
severity <- data.frame(
  level = c("High", "Low", "Medium"),
  count = c(30.0, 10.0, 20.0)
)

data(severity) + bar + x(level) + y(count) +
  title("Text column — the order the rows came in")
```

That is rarely what you want here. Make it a **factor** and say what order you
mean:

```{r}
severity$level <- factor(severity$level, levels = c("Low", "Medium", "High"))

data(severity) + bar + x(level) + y(count) +
  title("Factor — the order you declared")
```

The `levels =` travels with the column, so **every part of the chart agrees**:
the axis, the color assignment, and the legend all read in the same order.

```{r}
data(severity) + point + x(count) + y(count) + color(level) +
  title("The legend follows the same order")
```

The order holds even when GOG does the counting for you. These tickets arrive
`High` first, but `bar * count` still reads Low, Medium, High. A summary carries
the declared order onto its axis, it does not fall back to first-appearance:

```{r}
tickets <- data.frame(
  level = factor(
    c("High", "Low", "High", "Medium", "High", "Low", "Medium", "High"),
    levels = c("Low", "Medium", "High")
  )
)

data(tickets) + bar * count + x(level) +
  title("bar * count — the levels still order the axis")
```

R has two kinds of factor: `factor(x, ordered = TRUE)`, meaning Low really is
*less than* High, and a plain factor that merely lists its levels. GOG honors
the order in both cases, because most people reach for a plain factor purely to
fix the display order and mean nothing mathematical by it.

### The same declaration in the other three languages

`factor()` is R's word for this. Python, Julia and JavaScript have no factor
type, so GOG supplies the declaration itself, under one name: `ordered()`. It
takes the column's values and the order its categories go in. Declare it where
you build the table, and everything after that reads the same in all four
languages.

**R**

```r
severity <- data.frame(
  level = factor(c("High", "Low", "Medium"), levels = c("Low", "Medium", "High")),
  count = c(30.0, 10.0, 20.0)
)
```

**Python**

```python
severity = {
    "level": ordered(["High", "Low", "Medium"], ["Low", "Medium", "High"]),
    "count": [30.0, 10.0, 20.0],
}
```

**Julia**

```julia
severity = Dict(
    "level" => ordered(["High", "Low", "Medium"], ["Low", "Medium", "High"]),
    "count" => [30.0, 10.0, 20.0],
)
```

**JavaScript**

```js
const severity = {
  level: ordered(["High", "Low", "Medium"], ["Low", "Medium", "High"]),
  count: [30.0, 10.0, 20.0],
};
```

Python reads a pandas `Categorical` as well, and takes its categories as the
levels. `ordered()` is what a plain dictionary of lists uses, because GOG does
not require pandas.

Two rules worth knowing:

- **Levels say what order, the data says what is there.** A level with no rows
  draws no bar and gets no slot on the axis: an empty labeled gap is harder to
  read than a shorter axis. A value in the data that is *not* in the levels is
  never dropped; it goes on the end.
- **An explicit `order()` wins.** You wrote it in the plot, which is nearer than
  the table. Name a *different* column and it overrides the levels, which is what
  lets you sort by value:

```{r}
data(severity) + bar + x(level) + y(count) + order(count, desc = TRUE) +
  title("order() overrides the levels")
```

Name the **category column itself** and you are asking for that column's own
order, so there is nothing to override. A plain text column carries nothing but
its spelling and sorts alphabetically; a factor carries its levels, so it keeps
them, and `desc = TRUE` reverses them. That leaves no way to alphabetize a
factor's labels from inside the plot, and it is the right thing to lose:
relevel it in the host, where the levels are.

## Missing values

Real data has gaps: `penguins` is missing two flipper measurements, and most
survey data far more. A row with no value where the plot needs one cannot be
placed, so GOG drops it and **says how many**, rather than quietly changing the
count under you:

```{r}
#| message: true
patchy <- data.frame(
  weight = c(3.1, 4.2, NA, 5.8, 6.0, NA, 4.9, 5.1, 3.7, 5.5),
  batch  = c("A", "A", "B", "B", "A", "B", "A", "B", "A", "B")
)
data(patchy) + bar * bin + x(weight) +
  x_label("Weight") + title("Two rows have no weight")
```

The drop is scoped to the columns the plot **maps**. The message names `weight`
because `weight` is bound to `x`, and a row with no `x` has nowhere to stand. A
missing value in a column the plot never reads costs nothing: leave `batch`
unbound and its gaps are ignored; bind it to `color` and a row missing its batch
drops too, because now it has no color to take. This is the rule other plotting
tools follow: only the channels actually in use decide which rows survive, so a
stray `NA` in an unused column never silently shrinks the picture.

## Labels are used verbatim

Category names, titles, and axis labels are drawn exactly as they appear in your
data. Characters with meaning in the output format (`&`, `<`, `>`) are escaped
for you, so a department called `R&D` needs no special handling:

```{r}
special <- data.frame(
  firm  = c("R&D", "Sales & Ops", "<10% unit"),
  spend = c(10, 25, 15)
)
data(special) + bar + x(firm) + y(spend) + title("Q3 <spend> & forecast")
```

Non-Latin scripts work the same way, including as bare column names:

```{r}
지역별 <- data.frame(
  지역 = c("서울", "부산", "대구"),
  값   = c(10, 25, 15)
)
data(지역별) + bar + x(지역) + y(값) + title("한국 지역별 값")
```

Label widths are measured per character, so full-width scripts reserve the right
amount of margin rather than three times too much.
