# Design laws

```{r}
#| include: false
source("R/setup.R")
```

gog is governed by nine laws.
They are not preferences or style guides; they are rules
the package *never* breaks.

Every new feature, every argument, every naming decision is checked against
these rules. If a proposed addition breaks one, it is rejected.

Laws 1 to 5 govern the daily grammar you write. Laws 6 to 8 govern the
grammar's structure: what composes, what renders, and what the engine may
refuse. Those three follow the Hangeul (한글) analysis most closely. Law 9 says
what a specification is, and what it must never contain.

## Law 1: Orthogonality

> Every compatible atom combines with every other atom.
> No atom is redundant.

If `color` works with `point`, it works with `line`, `bar`, `area`, and every
other mark. Where a channel makes sense for a mark, it is there. `color` is
never withheld from one mark and given to its neighbor.

One rule follows from this. A mark-specific version of a channel, such as
`point_color` beside `bar_color`, breaks the law. The channel must be
generalized, not specialized.

## Law 2: No Exceptions

> A transform behaves identically on every mark.
> No per-mark special cases.

Without this law you would have to learn each transform again for every mark
that can read it. `bin` cuts the same buckets and counts the same rows
whichever mark reads it. Change only the mark, and what changes is the
geometry:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
((data(gapminder_2007) + bar * bin + x(life)) |
   (data(gapminder_2007) + line * bin + x(life))) /
((data(gapminder_2007) + area * bin + x(life)) |
   (data(gapminder_2007) + step * bin + x(life)))
```

*"Given gapminder 2007: bars derived by bin, x is life, beside a line derived by
bin, x is life; above an area derived by bin, x is life, beside a step outline
derived by bin, x is life."*

One cut and one set of counts, four times over: rectangles, a line through the
bin centers, that same line filled, and the counts as a staircase. Read any
bar's height, then read the same bin in the panel beside it: the number is the
same. The transform never asks which mark will draw its result.

This is the hardest law to keep as the package grows. Every user request that
begins "but for maps I need…" or "histograms are special because…" is asking for
this law to be broken. The law holds because one table inside the engine holds
every permitted binding, for every mark × channel pair. A mark cannot leave a
channel out, because there is no place to leave it out.

## Law 3: Plain Names

> Every taught name is a common English word.
> No acronyms. No abbreviations. Two words maximum, joined by `_`.
> Only `x`, `y`, `z` are literal exceptions.

If a new user has to look up what a name means, it fails this law.

| The name you may know | The plain name here |
|---|---|
| `alpha` | `opacity` |
| `loess`, `stat_smooth` | `smooth` |
| `coord_polar` | `polar` |
| `geom_point`, `aes` | `point`, and channels called by their plain names: `x( )`, `color( )` |
| `color_by`, `size_by` | `color`, `size` |

The last row is the one where a word was *removed* rather than replaced, and it
is worth reading twice, because the word is not lost. A plotting library could
reasonably write `color_by(species)` and `size_by(population)`. The suffix marks
the argument as a column, which says nothing here, because every atom that takes
an argument takes a column. Now read `color(continent)` aloud. It is *"color by
continent"*. The word *by* is already in how you read the sentence, so writing
it into the name says it twice. That is a silent letter by Law 2's measure, and
every channel would have had to carry one.

### One spelling of English

The same reasoning settles how words are spelled. gog writes American English
everywhere: `color`, `center`, `gray`, and every word of the kernel. There is no British alternative. ggplot2 takes both
spellings, so `scale_colour_manual()` and `scale_color_manual()` are one
function there; here there is one spelling and no second one to learn.

Two spellings for one word is a silent letter, and Law 2 refuses silent letters
in the vocabulary as well as in the grammar. The cost is invisible to the person
who already knows both forms, and everyone else pays it. A reader of your code
must recognize two shapes for one idea, a search for `color(` finds only half
the code, and every document that lists the vocabulary lists it twice.

Most readers meet this law for the first time by typing `colour`, so the refusal
names the word to write instead:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + style(colour = "tomato")
```

That is the setting. The channel is refused the same way, and `colour` exists in
the package for no other purpose. It is not a channel. It is a word that exists
only so gog can name the channel you meant.

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + colour(continent)
```

The same holds for `border_colour` and `centre`. Everywhere else in the grammar,
the American spelling is simply the only one there has ever been.

## Law 4: Bind Once

> A table is named once.
> Columns are then bare names.
> The nearest `data()` wins.

Other plotting libraries often repeat the table in every call, or quote each
column as a string. Naming the table once is what lets the rest of the sentence
read as plain words.

```{r}
# Correct: table named once, columns bare
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

Quoting a column name is not a typo. It is a habit brought from other plotting
libraries, so gog names the fix instead of only reporting the error:

```{r}
#| error: true
data(gapminder_2007) + point + x("gdp") + y("life")
```

This law makes a plot specification readable as an English sentence. It also
lets your session offer column names as you type. And it lets gog refuse a
mistake while you are still writing, rather than when the plot is drawn.

## Law 5: Explicit Over Implicit

> Short is better than long, unless short is ambiguous.
> Then say it out loud.

Implicit behavior saves typing now and costs clarity later. gog chooses the
other way: when brevity would leave a reader guessing what the engine decided,
the engine makes you say it.

This law is also what the error messages are for. A refusal must tell you *what
gog assumed* or *what gog cannot do*, never merely that something went wrong.

### Setting is not mapping

The tempting short spelling for a constant color is `color("red")`, reusing the
channel and letting the quotes decide the meaning. gog does not reuse the
channel. So `color("continent")` would be genuinely ambiguous: a column, or a
color named "continent". gog could decide by asking *"is this a column in the
table?"*. Then the same expression would mean one thing against one table and
something else against the next.

So gog says it out loud. Mapping is a channel; setting is
[`style()`](style.qmd):

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "seagreen", opacity = 0.5)
```

*"Given gapminder 2007: points, x is gdp, y is life, colored seagreen, with
opacity 0.5."*

Separating them also lets each operation follow its own rules. A mapping gets a
legend and a setting does not. And `line` cannot *map* `opacity`, because one
stroke has one opacity, but it can still *set* it.

ggplot2 puts the same boundary in a different place. There the boundary is a
pair of parentheses: `color = "red"` inside `aes()` maps, and the same words
outside it set. gog moves that boundary from a punctuation mark to a word, so
the atom you write says which one you meant. The parentheses can fail in one way
a word cannot. `aes(color = "red")` is legal, and it maps every row to one
category whose only level is "red". You get a legend with one key labeled "red", and points drawn
in the palette's first color rather than in red.

### A missing name must not become a default

A palette name that matches nothing could be answered two ways: refuse it, or
quietly draw in the default colors. gog refuses, and the message lists every
name it would have accepted:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("red")
```

A default breaks this law without announcing it: the plot still appears, so
nothing looks wrong, and the person reading the plot is the one misled.

### Illegal is not the same as unsupported

Refusing is only half of the answer. gog separates two kinds of refusal,
because each one asks something different of you:

| Refusal | Meaning | What to do |
|---|---|---|
| **Illegal** | The grammar forbids it. It will never work. | Rewrite the expression. |
| **Unsupported** | The grammar allows it; this engine does not draw it. | Use another atom that renders. |

Calling a valid combination "illegal" would teach you a rule that does not
exist. So gog says which it is.

```{r}
#| echo: false
peek(medals)
```

```{r}
#| error: true
# Illegal: size encodes magnitude, and a category has none
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

```{r}
#| error: true
# Illegal: a bar has no size feature at all; its extent is y
data(medals) + bar + x(country) + y(gold) + size(gold)
```

```{r}
#| error: true
# Illegal: `z` stands a `point`, `path` or `bar` up into the cube (see Space),
# but a `line` is read along its x axis, and a cube has no left to right
data(medals) + line + x(country) + y(gold) + z(silver)
```

That third one is worth reading closely, because it is the kind that is easy to
mistake for a missing feature. A 3-D `line` is not unbuilt. It is ruled out, and
the refusal says why: a `line` sorts by `x` and draws one value for each. In a
cube, `x` is one of three equal positions, and at some viewing angles it runs
into the page and becomes depth. The line would be sorted by an axis the reader
cannot see. The direction the refusal gives is `path`, which is `line` with that
sort removed, and it does draw in space.

An Unsupported refusal looks different. Here is one:

```{r}
#| error: true
# Unsupported: `z` is valid grammar for `text`; this engine has not built it
data(medals) + text + x(country) + y(gold) + z(silver) + label(country)
```

In every case gog *refuses to render*. It does not quietly drop the channel and
hand you a plot that looks finished. A silently ignored binding is a silent
letter in software. You wrote it, and it changes nothing in the picture.

The four refusals in this section are *live*: the engine prints them as this
page builds.
The last one is a gap the engine can close; the first three are rulings that
never close. That is the whole distinction. A check re-runs every refusal in
this book and fails if one has started drawing instead.

### Drawing it anyway

Every refusal the engine makes ends with the same offer: fix it, *or set
`GOG_STRICT=0` to draw anyway.* That switch turns refusals into warnings. You
get the same message, in the same words, and you get the plot underneath it.

Set it on a line of its own, before the sentence you want drawn:

| Language | Set it with |
|---|---|
| R | `Sys.setenv(GOG_STRICT = 0)` |
| Python | `os.environ["GOG_STRICT"] = "0"` |
| Julia | `ENV["GOG_STRICT"] = "0"` |
| JavaScript | `process.env.GOG_STRICT = "0"` |
| Shell | `export GOG_STRICT=0` |

Only `0` works. `GOG_STRICT=false` and `GOG_STRICT=no` still refuse, because a
switch this simple is better with one spelling than with a list of words to
guess at.

What you get is the plot the engine can still draw. The refused binding is not
always dropped: `bar + size(gold)` draws the bars at one width, and beside them
a size key running from 17 to 46 that no bar obeys. That mismatch is what the
refusal was protecting you from.

It then stays on for the rest of your session. That is the part to remember:
every later refusal is a warning too, and readers stop reading warnings. Clear
the variable again once you have seen the plot you wanted.

Your binding makes some refusals before the engine is ever asked, and the switch
cannot change those, because nothing has been drawn to offer you. Your binding
catches a quoted
column name, a British spelling, a sentence that does not begin with `data()`,
and a malformed atom such as `box(whiskers = "middle")`.

## Law 6: Compositional Invariance

> A composed sub-expression means the same thing in every context.
> No enclosing expression may silently reinterpret an inner one.

The Korean syllable 하 (*ha*) is pronounced identically in 하늘 (*haneul*, sky), 하루
(*haru*, a day), and 하지만 (*hajiman*, but). 하 can be reused because its sound
never changes, not because it carries a meaning of its own. English composition is not stable: "ough" is a different sound
in *though*, *through*, *thought*, *tough*, *cough*, and *bough*. A gog
expression must be 하, never "ough": layering, faceting, or animating a plot may
add to it, but may not change what the existing part meant.

You can see the law by using one sub-expression in two sentences. `bar * bin`
is one sub-expression: a mark and its transform, composed once.

```{r}
data(gapminder_2007) + bar * bin + x(life)
```

Now put that same block inside a faceted sentence:

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

*"Given gapminder 2007: bars derived by bin, x is life, split into panel columns
by continent."*

The enclosing expression added panels. It did not reinterpret the
sub-expression: each panel bins and counts exactly as the single plot did. That
is the whole law, and it is what lets you read a long sentence one piece at a
time. A sub-expression can be held in a variable too, `histogram <- bar * bin`, and
spoken wherever you like.

## Law 7: Minimum Syllable

> A visible plot is a mark plus its required positions.
> A mark alone is silent; a position alone has no carrier.

In Hangeul the smallest pronounceable unit is a consonant plus a vowel: ㄱ (*g*)
alone is silent, and 가 (*ga*) is the minimum syllable. A bare vowel cannot stand
alone either: ㅏ (*a*) must be written 아 (*a*), on a silent carrier.

A mark is the silent consonant: `point` with no positions draws nothing, because
there is nowhere to draw it. A position is the vowel that needs its carrier:
`x(gdp)` without a mark answers no question. It does not say what sits at that
x. gog refuses the incomplete syllable, and the refusal names each position
that is missing:

```{r}
#| error: true
data(gapminder_2007) + point
```

## Law 8: Pronounceable ≠ Usable

> The grammar guarantees well-formedness, never good taste.
> Grammar is law; taste is advice.

꿲 (*kkwek*) obeys every Hangeul syllable rule and you can pronounce it, yet no
Korean word uses it. Plots work the same way. This renders, because it is
grammatically sound; whether it *says* anything is your judgment, not the
engine's:

```{r}
# Legal, and nearly invisible: the engine obeys the grammar and does not judge
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(opacity = 0.05)
```

It draws, and almost nothing is visible. No legal combination is ever refused
for being ugly: one that looks pointless today may be a real plot tomorrow, so
refusing it now because it looks bad would remove something nobody has thought
of yet.

The law has a second half. Two viewing angles of a cube show both halves at
once. `space(turn = 390)` is a full circle plus 30 degrees, so it is the same
view as `space(turn = 30)`, and gog draws it without comment. `space(tilt = 95)`
is refused. The difference is not that one is uglier. A bearing repeats every
360 degrees, so 390 is the same direction as 30, and you can stand there. A tilt
stops at 90, where you are looking straight down, so 95 is no direction at all.
gog gives advice about taste and never refuses on it. It enforces
well-formedness with no exceptions. [Space](space.qmd) draws both.

The engine will not judge, so here is the advice instead. gog exists so that
you can say a visualization in clean, minimal syntax. It will let you combine
anything the grammar allows, and that includes combinations that hide what you
wanted to show. Leave out anything that does not help the reader see the main
idea.

That advice is about the picture, and it applies to the sentence too. Two
sentences can build the identical plot while one is much harder to follow. A
second `data()` applies to one mark and stops there. The next mark written after
it reads the plot's first table again. A reader who has not learned that rule
will read such a sentence wrong. gog accepts either arrangement, so the fix is a habit rather
than a rule, and [Data](data.qmd) gives it.

## Law 9: Universal Transcription

> A specification describes the visual, never one renderer's drawing
> commands.

A specification is **declarative**: you say what you want to see, never how to
draw it. The form is far older than software. "And God said, Let there be light:
and there was light" [@kingjamesbible1769, Genesis 1:3] states an outcome and
nothing else. There is no procedure in it, no order of steps, and no mention of
how the light is made. Every declarative sentence leaves that part out, and
something else has to supply it. In gog the engine does, and every stroke on
these pages answers a sentence that never mentioned strokes.

What you write becomes a small, renderer-neutral specification: which marks,
which columns on which channels, which transforms. One renderer reads that
specification, and that renderer draws every SVG on these pages. The same
specification, unchanged, is what the Python, Julia and JavaScript bindings
build. It is also why a specification names no shape and no stroke: it says what
to see, and the renderer decides how to draw it.

### Why that renderer writes SVG

The law says the specification is neutral. It does not say which format the
renderer writes, and **SVG** is a choice with reasons.

The first reason is the one you can see. An SVG stores shapes and coordinates,
not a grid of colored dots. A plot therefore redraws at any size it is given,
and stays sharp at every one. Print one at poster size and its curves are still
curves.

The second is that text stays text. Every axis label, tick number and title is
written as characters in the file, not drawn as a picture of characters. You can
select them, a search can find them, and a screen reader can read them aloud. It
is also what lets a plot in this book carry Hangeul, or any other writing
system, without the engine knowing anything about the letters.

The third reason is the one this project depends on most, and it is easy to
miss. An SVG is a text file, so two of them can be compared character by
character. That is what makes the promise of four bindings checkable: every
sentence in this book is drawn from R, Python, Julia and JavaScript, and the
four files must match exactly. Comparing pictures instead would mean comparing
colored dots, and those dots change with the version of whatever drew them. An
exact
test would become an approximate one, and a small disagreement between two
languages could pass as a rounding difference.

The fourth is motion. A `play` animation is written into the same file as the
plot, using SVG's own timing elements. A time-lapse is one picture that moves,
with no video file and no JavaScript beside it.

### Why there is one renderer

Those four reasons settle the format. One question is left: why only one
renderer writes it. gog has exactly one renderer, and the reason is the set of
decisions a second renderer would have to copy.

A renderer decides five things the specification does not name: layout, tick
selection, palettes, axis ranges, and label formatting. Write a second renderer
beside the first and those five decisions exist twice. Two copies of a decision
do not stay equal. One would bin and count, the other would draw the raw rows,
and `bar * bin` would then mean two different pictures from one specification.

That would be Law 9 broken by the renderer rather than by the specification: a
renderer deciding things the specification is supposed to own, with nothing in
the design to make it visible. So there is exactly *one* renderer. A second one
is possible only once those five decisions live in a single shared layer that
both would read.

Raster output comes from converting the SVG rather than from a second renderer.
`render_svg()` hands you the plot as ordinary text, in every binding:

```{r}
svg <- render_svg(data(gapminder_2007) + point + x(gdp) + y(life))
cat(substr(svg, 1, 51))
```

Those are the first 51 characters; `svg` holds the whole plot, about 14,000 of
them. Writing it to a file saves it, and any tool that reads SVG turns it into
another format: `rsvg-convert -w 1600 plot.svg -o plot.png` gives you a raster
image, and `rsvg-convert -f pdf` is what places each plot in this book's PDF.
Converting is the deliberate answer rather than a missing feature. It keeps
every decision about what a plot looks like in the one place that is allowed to
make them, which is exactly what the law above requires.

The engine does that conversion itself in one case, and the case is worth seeing
because it shows where the boundary sits. A plot that binds `play()` moves in a
browser and nowhere else, so
[`save_gif()`](play.qmd#taking-the-sequence-with-you) writes the same sequence
as a picture that a slide or a message will play. That is still conversion in
the sense this section means. The frames come out of the one renderer, and the
GIF writer only turns them into pixels. It chooses no tick, no color and no
layout, so there is nothing for it to drift from.

## The settable rule

The nine laws are stated. One case is worth working through in full, because it
is where readers most often expect an exception to be hiding: a setting.

A **setting** (`style(…)`) is an atom like any other, so the first two laws
govern it as well. A setting is available on every mark whose geometry can carry
it, and it behaves the same way on all of them. It is absent only where
composing two marks already says the same thing better. The first of those is No
Exceptions: there is no mark-specific gap you have to memorize. The second is
the non-redundant half of Orthogonality: there is no setting for what two marks
already do by combining.

`style(border_color =, border_size =)` draws an outline distinct from the fill.
It is available on all five closed-glyph fills: `bar`, `box`, `point`, `zone`
and `surface`. Their perimeter is a shape you cannot trace by composition, so a
setting is the only way to draw it.

`style(pattern =)` is the paint's **texture**, and it is realized once per
geometry. On the six path strokes (`line`, `step`, `interval`, `path`, `rule`,
`edge`) it is a **dash**. On the five fills (`bar`, `box`, `area`, `ribbon`,
`zone`) it is a **hatch**. One setting, one form per geometry, the way `color`
is a stroke on a line and a fill on a bar.

A curve fill (`area`, `ribbon`) takes the hatch, because a region can be
textured. It takes no *border*, because its edge is a data curve, which already
*is* a `line` (`area + line`, `line * bounds`) and can do more than any border.
The classes, not the individual marks, decide.

This prevents one particular surprise: you learn a setting on one mark, then
find it missing on a sibling mark for no reason you can see. Where a class is
incomplete, gog says so with an **Unsupported** message rather than a silent
nothing, so the gap is visible.

## Why these laws matter

Each law refuses the same enemy, **the expert's shortcut**. It never arrives
under that name. It arrives as a request for one special case, and the request
is always a reasonable one:

- *"Can `bar` have a different color scale than `point`?"*
- *"Can `smooth` have a `span` parameter?"*
- *"Can I just write `alpha`? Everyone knows what it means."*

Each one is reasonable on its own, and each one costs more than the one before
it: the tenth exception makes the first nine harder to remember.

The laws are the answer: not "no, because we don't want to," but
"no, because that would break Law 2 / Law 3 / …"
The rule, not the maintainer, says no.

## What comes next

The laws say what the grammar will and will not do, in any language.
[R](bindings/r.qmd) opens the four chapters that say what belongs to your
language instead: its pipes, its masked names, and how it spells a column.
