# Design laws

```{r}
#| include: false
source("R/setup.R")
```

gog is governed by nine laws.
They are not preferences or style guides; they are rules
the package *never* breaks.

Every new feature, every argument, every naming decision is checked against
these rules. If a proposed addition breaks one, it is rejected.

Laws 1 to 5 govern the grammar you write every day. Laws 6 to 8 govern the
grammar's structure: what composes, what renders, and what the engine may
refuse. Those three follow Hangeul (한글), the Korean alphabet, most closely, and
each opens with an example from it. Law 9 says what a specification is, and what
it must never contain.

## Law 1: Orthogonality

> Every compatible atom combines with every other atom.
> No atom is redundant.

`color` works with `point`, `line`, `bar`, `area`, and every other mark.
Whenever a mark's geometry can show a channel, gog accepts that channel on that
mark. A mark can still limit the kind of column: a `bar` takes a categorical
`color` only.

One rule follows from this. A mark-specific version of a channel, such as
`point_color` beside `bar_color`, breaks the law. The channel must be
generalized, not specialized. This section shows no refusal, because gog has no
mark-specific channel to refuse.

## Law 2: No Exceptions

> A transform behaves identically on every mark.
> No per-mark special cases.
> No silent letters.

Without this law you would have to learn each transform again for every mark
that can read it. `bin` makes the same bins and counts the same rows, whichever
mark reads it. Change only the mark, and what changes is the geometry:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
((data(gapminder_2007) + bar * bin + x(life)) |
   (data(gapminder_2007) + line * bin + x(life))) /
((data(gapminder_2007) + area * bin + x(life)) |
   (data(gapminder_2007) + step * bin + x(life)))
```

*"Given gapminder 2007: bars derived by bin, x is life, beside a line derived by
bin, x is life; above an area derived by bin, x is life, beside a step outline
derived by bin, x is life."*

The same bins and the same counts are drawn four ways. You get rectangles, a
line through the bin centers, that same line filled, and the counts as a
staircase. Read any bar's height, then read the same bin in the panel beside
it: the number is the same. The transform never asks which mark will draw its
result. A transform that one mark ignored would be a **silent letter**: a
letter that is written but never pronounced, like the *k* in *knife*.

This is the hardest law to keep as the package grows. Every user request that
begins "but for maps I need…" or "histograms are special because…" is asking for
this law to be broken. The engine keeps the law in two places. The code that
computes a transform never looks at the mark, and one table records a decision
for every pair of a mark and a transform, even when the decision is a refusal.

## Law 3: Plain Names

> Every taught name is a common English word.
> No acronyms. No abbreviations. Two words maximum, joined by `_`.
> The only exceptions are `x`, `y` and `z`.

If a new user has to look up what a name means, it fails this law. `svg` and
`gif` are acronyms, but a new user does not have to look them up: they are the
usual names of two file formats. So `render_svg()`, `save_svg()` and
`save_gif()` keep them. The table pairs a
name you may know from another library with the plain name gog uses:

| The name you may know | The plain name here |
|---|---|
| `alpha` | `opacity` |
| `loess`, `stat_smooth` | `smooth` |
| `coord_polar` | `polar` |
| `geom_point`, `aes` | `point`, and channels called by their plain names: `x( )`, `color( )` |
| `color_by`, `size_by` | `color`, `size` |

In the last row a word was *removed* rather than replaced: the *by* of
`color_by`. A plotting library could reasonably write `color_by(species)` and
`size_by(population)`. The suffix marks the argument as a column, which adds
nothing in gog, because every channel takes a column. Now read
`color(continent)` aloud. It is *"color by continent"*. The word *by* is already
in how you read the sentence, so writing it into the name says it twice. In Law
2's terms, that is a silent letter, and every channel would have had to carry
one.

### One spelling of English

The same reasoning decides how words are spelled. gog writes American English
everywhere: `color`, `center`, the `gray` palette, and every word of the kernel.
There is no British alternative. ggplot2 takes both spellings, so
`scale_colour_manual()` and `scale_color_manual()` are one function in that
package. gog has one spelling and no second one to learn.

Two spellings for one word is a silent letter, and Law 2 refuses silent letters
in the vocabulary as well as in the grammar. The person who already knows both
spellings notices nothing. Everyone else has more to learn and more to check. If
gog took both spellings, a reader of your code would have to know both, a search
for `color(` would miss every `colour(`, and every list of the vocabulary would
list some words twice.

Many readers meet the spelling rule for the first time by typing `colour`, so
the refusal names the word to write instead:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + style(colour = "tomato")
```

That refusal was for a **setting**, a fixed value written in `style()`.
Writing `colour` as a channel is refused the same way. `colour()` is not a
channel: it exists only so that gog can name the channel you meant.

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + colour(continent)
```

`border_colour`, `centre` and `palette("grey")` are refused too, each with the
American word in its message. A color value is different: it is one of the color
names web pages use, not a word of the grammar, and that list holds both `gray`
and `grey`.

## Law 4: Bind Once

> A table is named once.
> Columns are then bare names.
> The nearest `data()` wins.

Other plotting libraries often repeat the table in every call, or quote each
column as a string. Naming the table once is what lets the rest of the sentence
read as plain words.

```{r}
# Correct: table named once, columns bare
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."*

Quoting a column name is not a typo. It is a habit brought from other plotting
libraries, so gog names the fix instead of only reporting the error:

```{r}
#| error: true
data(gapminder_2007) + point + x("gdp") + y("life")
```

This law makes a plot specification readable as an English sentence. It also
lets gog refuse a quoted column as soon as the sentence runs, before the engine
draws anything.

The law's third line is about a second table. A `data()` written directly
before a mark is the nearest one, and that mark reads it. Every other mark reads
the plot's table, the one named first. [Data](data.qmd) draws both cases.

## Law 5: Explicit Over Implicit

> Short is better than long, unless short is ambiguous.
> Then write the longer form.

Implicit behavior saves typing now and costs clarity later. gog chooses
clarity: when a short spelling would make a reader guess what the engine
decided, gog asks you for the longer form.

This law also decides what an error message must say. A refusal tells you what
gog cannot do and what to write instead, never only that something went wrong.

### Setting is not mapping

The clearest case of a short spelling that would make a reader guess is one
fixed color for every row. The spelling a reader might try first is
`color("red")`: the same channel, with quotes to say that `red` is a value and
not a column. The quotes cannot always decide. `color("continent")` is genuinely
ambiguous: a column, or a color named "continent". gog could decide by asking
*"is this a column in the table?"*. Then the same expression would mean one
thing with one table and something else with the next.

So gog does not reuse the channel. A fixed value gets a word of its own
instead: mapping is a channel, and setting is [`style()`](style.qmd):

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "seagreen", opacity = 0.5)
```

*"Given gapminder 2007: points, x is gdp, y is life, colored seagreen, with
opacity 0.5."*

Separating them also lets each operation follow its own rules. A mapping gets a
legend and a setting does not. And `line` cannot *map* `opacity`, because one
stroke has one opacity, but it can still *set* it.

ggplot2 puts the boundary between mapping and setting in a different place.
There the boundary is a pair of parentheses: `color = "red"` inside `aes()` is a
mapping, and the same words outside `aes()` are a setting. gog moves that
boundary from a punctuation mark to a word, so the atom you write says which one
you meant. The parentheses can fail in one way a word cannot.
`aes(color = "red")` is legal, and it maps every row to one category whose only
level is "red". You get a legend with one entry labeled "red", and points drawn in
the palette's first color rather than in red.

### An unknown name must not become a default

A palette name that matches nothing could be answered two ways: refuse it, or
quietly draw in the default colors. gog refuses, and the message lists every
name it would have accepted:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("red")
```

A default breaks this law without announcing it: the plot still appears, so
nothing looks wrong, and the person reading the plot is the one misled.

### Illegal is not the same as unsupported

Refusing is not enough by itself. gog separates two kinds of refusal,
because each one needs you to do something different:

| Refusal | Meaning | What to do |
|---|---|---|
| **Illegal** | The grammar forbids it. | Rewrite the sentence. |
| **Unsupported** | The grammar allows it; this engine does not draw it. | Remove the atom the message names, or use one this engine draws. |

Calling a valid combination "illegal" would teach you a rule that does not
exist. So an Unsupported refusal says that the grammar allows the sentence and
the engine does not draw it. An Illegal refusal gives the reason the grammar
forbids it. Here are three Illegal refusals, each for a different reason:

```{r}
#| echo: false
peek(medals)
```

```{r}
#| error: true
# Illegal: size encodes magnitude, and a category has none
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

```{r}
#| error: true
# Illegal: a bar has no size feature at all; its extent is y
data(medals) + bar + x(country) + y(gold) + size(gold)
```

```{r}
#| error: true
# Illegal: `z` gives a `point`, `path` or `bar` a third position (see Space),
# but a `line` is read along its x axis, and a cube has no left to right
data(medals) + line + x(country) + y(gold) + z(silver)
```

The third refusal is easy to mistake for an Unsupported one. It is not: the
grammar forbids a 3-D `line`, and the refusal above gives the reason. In a cube you
can look straight along `x`, and a line sorted along `x` would then have no
order you can see. The refusal suggests `path` instead, which is `line` without
that sort, and `path` draws in space.

An Unsupported refusal is worded differently. It says that what you wrote *is
valid grammar*, but *this engine does not draw it*:

```{r}
#| error: true
# Unsupported: `z` is valid grammar for `text`; this engine does not draw it
data(medals) + text + x(country) + y(gold) + z(silver) + label(country)
```

In every case gog *refuses to render*. It does not quietly drop the channel and
return a plot that looks finished. A silently ignored mapping is a silent
letter in software. You wrote it, and it changes nothing in the picture.

An Illegal refusal instead says why the grammar forbids what you wrote. That is
the whole distinction. All four refusals are *live*: the engine prints them when
this page is built. A check re-runs every refused sentence in the book, and fails
if one draws a plot.

### Drawing it anyway

Every refusal the engine makes ends with the same offer: fix it, *or set
`GOG_STRICT=0` to draw anyway.* That switch turns refusals into warnings. You
get the same message, in the same words, and the plot is drawn as well.

Set `GOG_STRICT` on a line of its own, before the sentence you want drawn. The
table gives that line in each language, and the line that clears the variable
again:

| Language | Set it with | Clear it with |
|---|---|---|
| R | `Sys.setenv(GOG_STRICT = 0)` | `Sys.unsetenv("GOG_STRICT")` |
| Python | `os.environ["GOG_STRICT"] = "0"` | `del os.environ["GOG_STRICT"]` |
| Julia | `ENV["GOG_STRICT"] = "0"` | `delete!(ENV, "GOG_STRICT")` |
| JavaScript | `process.env.GOG_STRICT = "0"` | `delete process.env.GOG_STRICT` |
| Shell | `export GOG_STRICT=0` | `unset GOG_STRICT` |

Only `0` works. `GOG_STRICT=false` and `GOG_STRICT=no` still refuse, because a
switch this simple is better with one spelling than with a list of words to
guess at.

What you get is the plot the engine can still draw. The refused mapping is not
always dropped: `bar + size(gold)` draws the bars at one width, and beside them
a size legend running from 17 to 46 that matches no bar. That mismatch is what
the refusal was protecting you from.

`GOG_STRICT` then stays set for the rest of your session. That is the part to
remember: every later refusal is a warning too, and after many warnings, people
stop reading them. Clear it once you have seen the plot you wanted.

Your binding makes some refusals before the engine is ever asked. The switch
cannot change those, because only the engine reads it, and the engine is never
reached. Your binding catches a quoted column name, a British spelling such as
`colour`, a sentence that does not begin with `data()`, and a malformed atom
such as `box(whiskers = "middle")`.

## Law 6: Compositional Invariance

> A composed sub-expression means the same thing in every context.
> No enclosing expression may silently reinterpret an inner one.

The Korean syllable 하 (*ha*) is pronounced identically in 하늘 (*haneul*, sky),
하루 (*haru*, a day), and 하지만 (*hajiman*, but). 하 can be reused because its
sound never changes, not because it carries a meaning of its own. English
letters do not keep their sound: "ough" is a different sound in *though*,
*through*, *thought*, *tough*, *cough*, and *bough*. A gog expression must be
하, never "ough": layering, faceting, or animating a plot may add to it, but may
not change what the existing part meant.

You can see the law by using one sub-expression in two sentences. `bar * bin`
is one sub-expression: a mark and its transform, composed once.

```{r}
data(gapminder_2007) + bar * bin + x(life)
```

Now put that same sub-expression inside a faceted sentence:

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

*"Given gapminder 2007: bars derived by bin, x is life, split into panel columns
by continent."*

The enclosing expression added panels. It did not reinterpret the
sub-expression: each panel bins and counts exactly as the single plot did. That
is the whole law, and it is what lets you read a long sentence one piece at a
time. A sub-expression can also be kept in a variable, and the variable then
stands for `bar * bin` wherever you write it.

## Law 7: Minimum Syllable

> A visible plot is a mark plus its required positions.
> A mark alone is silent; a position alone has no carrier.

In Hangeul the smallest pronounceable unit is a consonant plus a vowel: ㄱ (*g*)
alone is silent, and 가 (*ga*) is the minimum syllable. A bare vowel cannot stand
alone either: ㅏ (*a*) is written 아 (*a*), with the silent consonant ㅇ in front
of it as its carrier.

A mark is the silent consonant: `point` with no positions draws nothing, because
there is nowhere to draw it. gog refuses that sentence, and the refusal names
each position that is missing:

```{r}
#| error: true
data(gapminder_2007) + point
```

Add `x(gdp)` and `y(life)` and the same sentence draws. A position with no mark
is the vowel with no carrier: `x(gdp)` alone answers no question, because
nothing says what sits at that x.

## Law 8: Pronounceable ≠ Usable

> The grammar guarantees well-formedness, never good taste.
> Grammar is law; taste is advice.

꿲 (*kkwek*) obeys every Hangeul syllable rule and you can pronounce it, yet no
Korean word uses it. Plots work the same way. This renders, because it is
grammatically sound; whether it *says* anything is your judgment, not the
engine's:

```{r}
# Legal, and nearly invisible: the engine obeys the grammar and does not judge
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(opacity = 0.05)
```

It draws, and almost nothing is visible. No legal combination is ever refused
for being ugly. A combination that looks pointless to one person can be exactly
the plot another person needs.

The law has a second half, that well-formedness is enforced, and two `space()`
arguments show it. `space(turn = 390)` is a full circle plus 30 degrees, so it
is the same view as `space(turn = 30)`, and gog draws it without comment.
`space(tilt = 95)` is refused. The difference is not that one is uglier. A turn
repeats every 360 degrees, so 390 is the same direction as 30, and you can
stand there. A tilt stops at 90, where you are looking straight down, and going
past 90 would turn the scene upside down. Taste never makes gog refuse a plot,
and gog enforces well-formedness with no exceptions. [Space](space.qmd) shows
both sentences, and shows why one of them is refused.

The engine will not judge, so here is the advice instead. gog exists so that
you can describe a plot in clean, minimal syntax. It will let you combine
anything the grammar allows, and that includes combinations that hide what you
wanted to show. Leave out anything that does not help the reader see the main
idea.

That advice is about the picture. A sentence needs the same care, because
other people read it too. Two sentences can draw the identical plot while one is
much harder to follow. As Law 4 says, a second `data()` applies only to the mark
written directly after it, and the mark after that one reads the plot's table
again. A reader who does not know this reads such a sentence wrong. The grammar
accepts the marks in either order, so the fix is a writing habit: write each
mark next to the table it draws from. [Data](data.qmd) shows both orders.

## Law 9: Universal Transcription

> A specification describes the visual, never one renderer's drawing
> commands.

A specification is **declarative**: you say what you want to see, never how to
draw it. The form is far older than software. "And God said, Let there be light:
and there was light" [@kingjamesbible1769, Genesis 1:3] states an outcome and
nothing else. There is no procedure in it, no order of steps, and no mention of
how the light is made. Every declarative sentence leaves that part out, and
something else has to supply it. In gog the engine does, and every stroke on
these pages was drawn from a sentence that never mentioned strokes.

What you write becomes a small, renderer-neutral specification: which marks,
which columns on which channels, which transforms. One renderer reads that
specification, and that renderer draws every SVG on these pages. The same
specification, unchanged, is what all four bindings build. Being
renderer-neutral also means that a specification holds no drawing command, such
as "draw a line from here to there": it says what to see, and the renderer
decides how to draw it.

### Why that renderer writes SVG

The law says the specification is neutral. It does not say which format the
renderer writes, and SVG is a choice with reasons.

The first reason is visible in the picture itself. An SVG stores shapes and
coordinates, not a grid of colored dots. A plot therefore redraws at any size it
is given, and stays sharp at every one. Print one at poster size and its curves
are still curves.

The second is that text stays text. Every axis label, tick number and title is
written as characters in the file, not drawn as a picture of characters. You can
select them, a search can find them, and a screen reader can read them aloud. It
is also what lets a plot in this book carry Hangeul, or any other writing
system, without the engine drawing a single letter: it writes the characters,
and the font draws them.

The third reason is the one this project depends on most, and it is the least
obvious. An SVG is a text file, so two of them can be compared character by
character. That is what makes the promise of four bindings checkable: every
sentence in this book is drawn from R, Python, Julia and JavaScript, and the
four files must match exactly. Comparing pictures instead would mean comparing
colored dots, and those dots change with the version of whatever drew them. An
exact test would become an approximate one, and a small disagreement between
two languages could be mistaken for a rounding difference.

The fourth is motion. A `play` animation is written into the same file as the
plot, using SVG's own timing elements. A time-lapse is one picture that moves,
with no video file and no JavaScript beside it.

### Why there is one renderer

Those four reasons decide the format. One question is left: why only one
renderer writes it. gog has exactly one renderer, and the reason is the set of
decisions a second renderer would have to copy.

A renderer decides five things the specification does not name: layout, tick
selection, palettes, axis ranges, and label formatting. Write a second renderer
beside the first and those five decisions exist twice. Two copies of a decision
do not stay equal. One renderer would put ticks at 0, 20 and 40, the other at
0, 25 and 50, and one specification would then draw two different pictures.

That would break Law 9 from the renderer's side: one specification would no
longer describe one picture, and nothing in the design would show it. So there
is exactly *one* renderer. A second one is possible only once those five
decisions live in a single shared layer that both would read.

Raster output, a picture stored as a grid of colored dots such as a PNG file,
comes from converting the SVG rather than from a second renderer.
`render_svg()`, described in [Functions around the
grammar](utilities.qmd#render_svg), returns the plot as ordinary text. It has
the same name in all four languages, and the block below is R:

```{r}
svg <- render_svg(data(gapminder_2007) + point + x(gdp) + y(life))
cat(substr(svg, 1, 51))
```

Those are the first 51 characters; `svg` holds the whole plot, about 15,000 of
them. Write it to a file and you have the plot on disk. A tool that reads SVG
can convert that file to another format. `rsvg-convert`, a command-line SVG
converter, gives you a raster image with
`rsvg-convert -w 1600 plot.svg -o plot.png`, and its `-f pdf` option is what
places each plot in this book's PDF. Converting is the deliberate answer rather
than a missing feature. It keeps every decision about what a plot looks like in
the one place that is allowed to make them, which is exactly what Law 9
requires.

The engine does that conversion itself in one case, and that case shows where
drawing stops and converting begins. A plot that binds `play()` moves in a
browser and nowhere else, so
[`save_gif()`](play.qmd#taking-the-sequence-with-you) writes the same sequence
as a picture that a slide or a message will play. That is still conversion in
the sense this section means. The frames are drawn by the one renderer, and the
GIF writer only turns them into pixels. It chooses no tick and no layout. Its
one decision is how to fit each frame into the 256 colors a GIF file can hold.

## The settable rule

The nine laws are stated. One case needs a full explanation: a setting. It is
where readers most often expect to find an exception.

A setting (`style(…)`) is an atom like any other, so Orthogonality and No
Exceptions govern it as well. A setting is available on every mark whose
geometry can carry it, and it behaves the same way on all of them. That
uniformity is No Exceptions: there is no mark-specific gap you have to memorize.
A setting is absent only where composing two marks already says the same thing
better, and that absence follows Law 1's second line: no atom is redundant.

`style(border_color =, border_size =)` draws an outline distinct from the fill.
It is available on all five closed-shape fills: `bar`, `box`, `point`, `zone`
and `surface`. No second layer can draw their perimeter, so a setting is the
only way to draw it. An `area` or a `ribbon` is filled too, but a layered `line`
traces its edge, so neither takes a border. The letters of a label are filled
shapes too, so `text` takes both settings, and there they draw a **halo**, a
band of color under the letters. One setting takes one form in each class, as
`pattern` does below.

`style(pattern = )` gives a mark's paint a **texture**, and each class of
geometry shows that texture its own way. On the six path strokes (`line`,
`step`, `interval`, `path`, `rule`, `edge`) it is a **dash**. On the five filled
regions (`bar`, `box`, `area`, `ribbon`, `zone`) it is a **hatch**. One setting
takes one form in each class. That is what `color` already does: a stroke on a
line, a fill on a bar.

A curve fill (`area`, `ribbon`) takes the hatch, because a region can be
textured. It takes no *border*, because its edge is a data curve, which already
*is* a `line` (`area + line`, `line * bounds`) and can do more than any border.
The class decides which settings a mark takes, not the individual mark, and
[Setting vs mapping](style.qmd) defines each class.

This prevents one particular surprise: you learn a setting on one mark, then
find it missing on a sibling mark for no reason you can see. A setting written
outside its class is refused, and the refusal is Illegal: the grammar forbids
it.

## Why these laws matter

Each law refuses the same thing, **the expert's shortcut**: a special case
that saves an expert a few keystrokes and gives every beginner one more
exception to learn. Nobody asks for it by that name. It arrives as a request
like these:

- *"Can `bar` have a different color scale than `point`?"*
- *"Can `smooth` have a `span` parameter?"*
- *"Can I just write `alpha`? Everyone knows what it means."*

Each one is reasonable on its own, and each one costs more than the one before
it: the tenth exception makes the first nine harder to remember.

The laws are the answer: not "no, because we don't want to," but
"no, because that would break Law 2 / Law 3 / …"
The rule, not the maintainer, says no.

## What comes next

The laws say what the grammar will and will not do, in any language.
[R](bindings/r.qmd) opens the four chapters that say what belongs to your
language instead, such as how you install gog and how it reads from a
database.
