# Design laws

```{r}
#| include: false
source("R/setup.R")
```

gog is governed by nine laws.
They are not preferences or style guides; they are hard constraints
the package *never* violates.

Every new feature, every shortcut, every naming decision
is checked against these rules.
If a proposed addition breaks one, it is rejected.

Laws 1–5 govern the daily grammar you write. Laws 6–8 come from the Hangeul (한글)
analysis the project is founded on, and govern its structure: what composes,
what renders, and what the engine may refuse. Law 9 says what a specification
is, and what it must never contain.

## Law 1: Orthogonality

> Every compatible atom combines with every other atom.
> No atom is redundant.

If `color` works with `point`, it works with `line`, `bar`, `area`, and every other mark.
There are no atoms that "only work with certain marks."

The corollary: if you find yourself wanting a mark-specific version of a channel
(`point_color` vs `bar_color`), that is a violation.
The channel must be generalized, not specialized.

## Law 2: No Exceptions

> A transformation behaves identically on every mark.
> No per-mark special cases.

`bin` cuts the same buckets and counts the same rows whichever mark reads it.
Change only the mark, and what changes is the geometry:

```{r}
((data(gapminder_2007) + bar * bin + x(life)) |
   (data(gapminder_2007) + line * bin + x(life))) /
((data(gapminder_2007) + area * bin + x(life)) |
   (data(gapminder_2007) + step * bin + x(life)))
```

One cut and one set of counts, four times over: rectangles, a line through the
bin centers, that same line filled, and the counts as a stair. Read any bar's
height off the panel beside it and the number is the same. The transform never
asks which mark it is feeding.

This is the hardest law to keep as the package grows.
Every user request that begins "but for maps I need…" or "histograms are special
because…" is a pressure to break it. What keeps it is that the permitted
bindings live in **one table** in the engine (`gog-core/src/legality.rs`),
covering every mark × channel pair. A mark cannot quietly opt out of a channel,
because there is nowhere to opt out.

## Law 3: Plain Names

> Every taught name is a common English word.
> No acronyms. No abbreviations. Two words maximum, joined by `_`.
> Only `x`, `y`, `z` are literal exceptions.

| The name you may know | The plain name here |
|---|---|
| `alpha` | `opacity` |
| `loess`, `stat_smooth` | `smooth` |
| `coord_polar` | `polar` |
| `geom_point`, `aes` | `point`, and channels called by their plain names: `x( )`, `color( )` |
| `col`, `pos` | `column`, `position` |
| `color_by`, `size_by` | `color`, `size` |

If a new user has to look up what a name means, it fails this law.

The last row is the one where a word was **removed** rather than replaced, and
it is worth a moment because the word is not lost. Early drafts of this grammar
wrote `color_by(species)` and `size_by(population)`. The suffix marked the
argument as a column, which says nothing here, because every atom that takes an
argument takes a column. Now read `color(continent)` aloud. It is *"color by
continent"*. The word *by* lives in how the sentence is read, so writing it into
the name spells out what the reading already supplies. That is a silent letter
by Law 2's measure, and every channel would have had to carry one.

### One spelling of English, and it is American

gog writes American English everywhere: `color`, `center`, `gray`, and every
word of the kernel. There is no British alternative. ggplot2 takes both
spellings, so `scale_colour_manual()` and `scale_color_manual()` are one
function there; here there is one spelling and no second one to learn.

Two spellings for one word is a silent letter, which is Law 2's enemy wearing a
dictionary. The cost is invisible to the person who already knows both forms and
lands entirely on everyone else: a reader of your code must recognize two shapes
for one idea, a search for `color(` misses half the corpus, and every document
that lists the vocabulary has to list it twice.

Because ggplot2 is where most readers arrive from, typing `colour` is the
likeliest way to meet this law, so the refusal names the word to write:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + style(colour = "tomato")
```

That is the setting. The channel is refused the same way, and `colour` exists in
the package for no other purpose: it is not a channel, it is a word that says
where the channel went, exactly as `facet()` does in JavaScript.

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + colour(continent)
```

The same holds for `border_colour` and `centre`. Everywhere else in the grammar,
the American spelling is simply the only one there has ever been.

## Law 4: Bind Once

> A table is named once.
> Columns are then bare names.
> The nearest `data()` wins.

```{r}
# Correct: table named once, columns bare
data(gapminder_2007) + point + x(gdp) + y(life)
```

Quoting a column name is a habit carried in from other plotting libraries rather
than a typo, so gog says which fix it wants:

```{r}
#| error: true
data(gapminder_2007) + point + x("gdp") + y("life")
```

This law makes plot specifications readable as natural language.
It also enables autocomplete and early, friendly errors at spec-build time.

---

## Law 5: Explicit Over Implicit

> Short is better than long, unless short is ambiguous.
> Then say it out loud.

Implicit behavior saves typing and then charges interest in ambiguity. gog
takes the trade the other way: when brevity would leave a reader guessing what
the engine decided, the engine makes you say it.

This law is also what the error messages are for. A refusal must tell you
**what gog assumed** or **what gog cannot do**, never merely that something
went wrong.

### Setting is not mapping

The tempting short spelling for a constant color is `color("red")`, reusing
the channel and letting the quotes decide the meaning. gog does not, because
quoting a column name is a habit rather than a mistake, and `color("continent")`
would then be genuinely ambiguous: a column, or a color named "continent"?
Deciding by *"is it a column in this data?"* would make the same expression mean
different things against different tables.

So gog says it out loud. Mapping is a channel; setting is
[`style()`](style.qmd):

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "seagreen", opacity = 0.5)
```

This also gives the two operations the different rules they deserve. A mapping
earns a guide and a setting does not, and `line`, which cannot *map* `opacity`
because one stroke has one opacity, can perfectly well *set* it.

ggplot2 draws the same line in a different place. There the boundary is a
bracket: `color = "red"` inside `aes()` maps, and the same words outside it
set. gog moves that boundary from a bracket to a word, so the atom you write
says which one you meant. The bracket has one failure that a word cannot have.
Writing `aes(color = "red")` is legal, so it maps every row to a category with
a single level. You get a legend with one key labeled "red", and points drawn
in the palette's first color rather than in red.

### A missing name must not become a default

`palette()` once accepted any single string as a palette *name*, matched it
against the two real palettes, and fell through to the default when it matched
neither. `palette("red")` therefore drew a plot in the standard blue and
orange, silently:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + palette("red")
```

Falling back to a default is the most common way software breaks this law: the
plot still appears, so nothing looks wrong, and the reader is the one who ends
up misled.

### Illegal is not the same as unsupported

gog separates two refusals, because they ask different things of you:

| Refusal | Meaning | What to do |
|---|---|---|
| **Illegal** | The grammar forbids it. It will never work. | Rewrite the expression. |
| **Unsupported** | The grammar allows it; this engine cannot draw it yet. | Use another atom, or wait for the feature. |

Calling a valid combination "illegal" would teach you a rule that does not
exist. So gog says which it is.

```{r}
#| error: true
# Illegal: size encodes magnitude, and a category has none
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

```{r}
#| error: true
# Illegal: a bar has no size feature at all; its extent is y
data(medals) + bar + x(country) + y(gold) + size(gold)
```

```{r}
#| error: true
# Illegal: `z` stands a `point`, `path` or `bar` up into the cube (see Space),
# but a `line` is read along its x axis, and a cube has no left to right
data(medals) + line + x(country) + y(gold) + z(silver)
```

That third one is worth reading closely, because it is the kind that is easy to
mistake for a missing feature. A 3-D `line` is not unbuilt. It is ruled out, and
the refusal says why: a `line` sorts by `x` and draws one value for each, so in a
cube it would be sorted by an axis the reader cannot see. The direction it gives
is `path`, which is `line` with that sort removed, and it does draw in space.

Unsupported looks different, and here is one:

```{r}
#| error: true
# Unsupported: `z` is valid grammar for `text`; this engine has not built it
data(medals) + text + x(country) + y(gold) + z(silver) + label(country)
```

In every case gog **refuses to render**. It does not quietly drop the channel
and hand you a plot that looks finished. A silently ignored binding is the
software equivalent of a silent letter: you learn a rule that isn't real.

All four refusals are **live**: the engine prints them as this page builds. The
last one will stop being true when `z` on `text` is built, and the first three
never will, which is the whole distinction. A check re-runs every refusal in this
manual and fails if one has started drawing instead.

### Drawing it anyway

Every refusal above ends with the same offer: fix it, *or set `GOG_STRICT=0` to
draw anyway.* That switch turns refusals into warnings. You get the same
message, in the same words, and you get the plot underneath it.

Set it on a line of its own, before the sentence you want drawn:

| Language | Set it with |
|---|---|
| R | `Sys.setenv(GOG_STRICT = 0)` |
| Python | `os.environ["GOG_STRICT"] = "0"` |
| Julia | `ENV["GOG_STRICT"] = "0"` |
| JavaScript | `process.env.GOG_STRICT = "0"` |
| Shell | `export GOG_STRICT=0` |

Only `0` works. `GOG_STRICT=false` and `GOG_STRICT=no` still refuse, because a
switch this blunt is better with one spelling than with a list of words to guess
at.

It then stays on for the rest of your session, which is the part worth
remembering: every later refusal is a warning too, and a warning is easy to stop
reading. Turn it off again with `Sys.unsetenv("GOG_STRICT")` once you have seen
the plot you wanted.

You get the plot with the refused part left out, so `bar + size(gold)` draws the
bars and no size.

Three refusals it never reaches: a quoted column name, a British spelling, and a
sentence that does not begin with `data()`. Your package catches all three
before the engine is asked, so there is no half-drawn plot to offer.

---

## Law 6: Compositional Invariance

> A composed sub-expression means the same thing in every context.
> No enclosing expression may silently reinterpret an inner one.

The Korean syllable 하 (*ha*) is pronounced identically in 하늘 (*haneul*, sky),
하루 (*haru*, a day), and 하지만 (*hajiman*, but). Its reusability comes from
stability, not from meaning.
English composition betrays you instead: "ough" is a different sound in
*though*, *through*, *cough*, *rough*, and *bough*. A gog expression must be
하, never "ough": layering, faceting, or animating a plot may add to it, but
may not change what the existing part meant.

You can see the law by using one sub-expression in two sentences. `bar * bin`
is a block: a mark and its transform, composed once.

```{r}
data(gapminder_2007) + bar * bin + x(life)
```

Now put that same block inside a faceted sentence:

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

The enclosing expression added panels. It did not reinterpret the block: each
panel bins and counts exactly as the single plot did. That is the whole law,
and it is what lets you read a long sentence one piece at a time. A block can
be held in a variable too, `piled <- bar * bin`, and spoken wherever you like.

## Law 7: Minimum Syllable

> A visible plot is a mark plus its required positions.
> A mark alone is silent; a position alone has no carrier.

In Hangeul the smallest pronounceable unit is a consonant plus a vowel: ㄱ (*g*)
alone is silent, and 가 (*ga*) is the minimum syllable. A bare vowel cannot stand
alone either: ㅏ (*a*) must be written 아 (*a*), on a silent carrier.

A mark is the silent consonant: `point` with no positions draws nothing,
because there is nowhere to draw it. A position is the vowel that needs its
carrier: `x(gdp)` without a mark answers no question. *What* sits at that x?
gog refuses the incomplete syllable, and the refusal names the missing half:

```{r}
#| error: true
data(gapminder_2007) + point
```

## Law 8: Pronounceable ≠ Useable

> The grammar guarantees well-formedness, never good taste.
> Grammar is law; taste is advice.

꿲 (*kkwek*) obeys every Hangeul syllable rule, you can pronounce it, yet no
Korean word uses it. The same line runs through plots. This renders, because it is
grammatically sound; whether it *says* anything is your judgment, not the
engine's:

```{r}
# Legal, and nearly invisible: the engine's job is to obey, not to veto
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(opacity = 0.05)
```

That draws in silence. No legal combination is ever refused for being ugly: one
that looks pointless today may be a real chart tomorrow, so refusing it now on
taste grounds forecloses something you cannot see yet.

The engine will not judge, so here is the advice instead. gog exists so that
you can say a visualization in clean, minimal syntax. It will let you combine
anything the grammar allows, and that includes combinations that bury your
point. Be like a good writer, and do not let decoration distract from the main
idea.

That advice is about the picture, and it applies to the sentence too. Two
sentences can build the identical plot while one is much harder to follow. A
second `data()` applies to a single mark, so a mark written after it returns to
the plot's first table. A reader who has not learned that rule will read such a
sentence wrong. Both spellings are legal, and always will be. So the fix is a
habit rather than a rule, and [Data](data.qmd) gives it.

## Law 9: Universal Transcription

> A specification describes the visual, never one renderer's drawing
> commands.

A specification is **declarative**: you say what you want to see, never how to
draw it. The form is far older than software. "And God said, Let there be light:
and there was light" [@kingjamesbible1769, Genesis 1:3] states an outcome and
nothing else. There is no procedure in it, no order of steps, and no mention of
how the light is made. Every declarative sentence leaves that part out, and
something else has to supply it. In gog the engine does, and every stroke on
these pages answers a sentence that never mentioned strokes.

What you write in R becomes a small, engine-neutral description (which
marks, which columns on which channels, which transforms), and *that* is what
renders.
Today one renderer reads it, producing the SVGs on these pages; the same
description, unchanged, is what the Python, Julia and JavaScript front ends
already build, and what a GPU renderer would read later. This is also why the code in this book never mentions a
pixel: the specification has none to mention.

### Why that renderer writes SVG

The law says the description is neutral. It does not say which format comes out
of the other end, and **SVG** is a choice with reasons.

The first reason is the one you can see. An SVG stores shapes and coordinates,
not a grid of colored dots. A plot therefore redraws at whatever size it is
given, and stays sharp at all of them. Print one at poster size and its curves
are still curves.

The second is that text stays text. Every axis label, tick number and title is
characters in the file rather than a picture of characters. You can select them,
a search can find them, and a screen reader can read them aloud. It is also what
lets a plot in this book carry Hangeul, or any other writing system, without the
engine knowing anything about the letters.

The third reason is the one this project depends on most, and it is easy to miss.
An SVG is a text file, so two of them can be compared character by character.
That is what makes the promise of four bindings checkable: every sentence in this
book is drawn from R, Python, Julia and JavaScript, and the four files must match
exactly. Comparing pictures instead would mean comparing colored dots, which
changes with the version of whatever drew them. An exact test would become an
approximate one, and a small disagreement between two languages could hide inside
the tolerance.

The fourth is motion. A `play` animation is written into the same file as the
plot, using SVG's own timing elements. A time-lapse is one picture that moves,
with no video file and no JavaScript beside it.

### Why there is one renderer

There used to be two, and deleting the second one is the strongest evidence for
this law that the project has.

The second renderer wrote raster images directly. Because nothing sat between the
grammar and the output to hold the shared decisions, it had to make them again
for itself: layout, tick selection, palettes, axis ranges, label formatting. Five
decisions, written twice. The two copies then drifted, the way two copies always
do, and the drift ended somewhere specific. `bar * bin` drew binned counts in one
format and the raw, untransformed rows in the other. One specification, two
different plots, and only one of them correct.

That is Law 9 being broken from underneath. A renderer had quietly started
deciding things the description was supposed to own, and nothing in the design
made that visible. So the rule now is that there is exactly **one** renderer, and
a second one waits for a shared stage that holds those decisions where neither
can copy them.

Raster output comes from converting the SVG rather than from a second renderer.
`render_svg()` hands you the plot as ordinary text, in every binding:

```{r}
svg <- render_svg(data(gapminder_2007) + point + x(gdp) + y(life))
cat(substr(svg, 1, 51))
```

That string is the whole plot. `writeLines(svg, "plot.svg")` saves it, and any
tool that reads SVG turns it into a raster image: `rsvg-convert -w 1600 plot.svg
-o plot.png` is the one that builds this book's PDF. Converting is the deliberate
answer rather than a missing feature. It keeps every decision about what a plot
looks like in the one place that is allowed to make them, which is the whole of
the law above.

The engine does that conversion itself in one case, and the case is worth seeing
because it shows where the line falls. A plot that binds `play()` moves in a
browser and nowhere else, so [`save_gif()`](play.qmd#taking-the-sequence-with-you)
writes the same sequence as a picture that a slide or a message will play. That
is still conversion in the sense this section means. The frames come out of the
one renderer, and the GIF writer only turns them into pixels. It chooses no tick,
no color and no layout, so there is nothing for it to drift from.

---

## The settable rule: settings obey Laws 1–2 too

A **setting** (`style(…)`) is an atom like any other, so the first two laws govern
it as well: a setting is available on **every mark whose geometry can carry it**,
behaves the same across them, and is absent exactly where composition already
expresses the feature more capably. The first clause is No Exceptions: no
mark-specific gap you have to memorize; the second is Orthogonality's *non-redundant*
half: don't add a knob for what marks already do by combining.

That is why `style(border_color =, border_size =)`, an outline distinct from the
fill, is available on **all five closed-glyph fills** (`bar`, `box`, `point`,
`zone`, `surface`), not one of them: their perimeter is a shape you cannot trace
by composition, so a setting is the only way to draw it. And why `style(pattern =)`, the paint's
**texture**, is realized once per geometry: on the path strokes (`line`, `step`,
`path`, `interval`, `rule`) as a **dash**, and on the four fills (`bar`, `box`, `area`, `ribbon`)
as a **hatch**: one aesthetic, one form per geometry, the way `color` is a stroke
on a line and a fill on a bar. A curve fill (`area`, `ribbon`) takes the hatch (a
region can be textured) but no *border*: its edge is a data curve, which already
*is* a `line` (`area + line`, `line * bounds`), more capable than any rim. The
classes, not the individual marks, decide.

The failure this prevents is the quiet surprise: learning a setting on one mark and
finding it missing on its sibling for no reason you can see. When a class is still
being filled in, gog says so: an **unsupported** message ("designed but not drawn
yet"), never a silent nothing, so the gap is visible until it closes.

---

## Why these laws matter

Each law fights the same enemy: **the expert's shortcut.**

Real users constantly request special cases:
*"Can `bar` have a different color scale than `point`?"*
*"Can `smooth` have a `span` parameter?"*
*"Can I just write `alpha`? Everyone knows what it means."*

Each request sounds reasonable in isolation.
Each one erodes learnability non-linearly: the tenth exception makes
the first nine harder to remember.

The laws are the answer: not "no, because we don't want to," but
"no, because that would break Law 2 / Law 3 / …"
The rule, not the maintainer, says no.
