# Encoding scope

```{r}
#| include: false
source("R/setup.R")
```

If you write one `x()` and then two marks, which marks use it? Both do. Each
atom binds to a **scope**: the whole plot, or one layer. A layer is one mark
and the atoms bound to it. This chapter gives the rule that decides between the
two.

## The rule

The rule for every atom you write with `+` fits in one table. Each row names an
atom, the scope it binds to, and whether it matters where you write it.

| Atom | Scope | Depends on where you write it? |
|-------------------------------------------|-------------------------------------------|--------------|
| `x()`, `y()`, `z()` **before any mark** | **Plot**, every layer | Yes |
| the same positions **after a mark** | **Layer**, that mark's column for the shared axis; a layer that names no column of its own reads the first column written for that axis | Yes |
| the **first** `data()` or `query()` | **Plot**, default table | Yes |
| each **later** `data()` or `query()` | **Next layer** only | Yes |
| `color()`, `size()`, `shape()`, `pattern()`, `opacity()`, `group()`, `label()`, `play()` **before any mark** | **Plot**, every layer that can accept it | Yes |
| the same channels **after a mark** | **Layer**, that mark alone | Yes |
| `style()` | **Layer**, nearest preceding mark; refused when no mark precedes it | Yes |
| `title()`, `x_label()`, `y_label()`, `z_label()`, `palette()`, `theme()`, `order()`, `brush()`, and a space such as `polar()` | **Plot** | No |

In every row marked **Yes**, where you write the atom decides its scope. Write a
channel before any mark and it binds to the plot. Write a channel or `style()`
after a mark and it binds to that mark. `data()` and `query()` work
differently. The first `data()` or `query()` opens the sentence and is the
plot's table. Each later one binds to the mark written directly after it, not to
the mark before it. The atoms in the last row, marked **No**, belong to the plot
wherever you write them. They are the title and the axis labels, `palette()`,
`theme()`, `order()`, `brush()` and the spaces. The row for positions after a
mark says one thing more, and the next section shows it with an example.

Where two bindings could apply, the nearest one wins. That is the same rule
that lets a `data()` written between two marks give the next layer its own
table in place of the plot's.

## One axis, and which column each layer reads

The rule above never gives a layer axes of its own. Every layer of a plot
shares the same axes. That is what makes it one plot rather than two drawn on
top of each other. gog allows no second axis: one x scale, one set of ticks,
one label. So a second table's values must be in the same units as the first
table's, such as dollars in both. Values in different units need a plot of their
own. [Composition](composition.qmd) covers the choice between layering,
faceting, and a page of separate plots.

Sharing an axis is not the same as sharing a *column name*, and the two are
easy to confuse. Almost always you want to bind the positions once, before the
marks, so that every layer reads them:

```{r}
#| echo: false
peek(actuals)
```

```{r}
# x and y written once, before the marks: both marks use them
data(actuals) + x(year) + y(sales) +
  bar +
  line
```

*"Given the actuals: x is year, y is sales, bars and also a line."*

Write a position after a mark, and that mark reads the column you name. No
second axis appears. The layer only chooses which of *its* columns supplies the
values. That rule is why the `milestones` table from [Data](data.qmd) can keep
its own column names:

```{r}
#| echo: false
peek(gapminder_2007)
peek(milestones)
```

```{r}
# `milestones` calls its columns `at` and `value`, not `gdp` and `life`
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at) + y(value) + label(note)
```

*"Given gapminder 2007: points, x is gdp, y is life; then given the milestones:
text, x is at, y is value, label by note."*

Look at the axis labels: they are still `Gdp` and `Life`. Each axis is named
after the first column written for it, unless you name the axis with
`x_label()` or `y_label()`. Here those first columns are `gdp` and `life`,
written after `point`. The axis belongs to the plot, and only the choice of
column belongs to the layer. Without this rule, you would have to rename `at`
and `value` in `milestones` to `gdp` and `life`. Renaming is easy when you
build the table yourself. When the table comes from somewhere else, renaming
means changing data you did not write.

A layer that names no column of its own reads the first column written for each
axis, wherever the layer is written. So `point + x(gdp) + y(life) + line` draws
a line through the points. `point + line + x(gdp) + y(life)` draws the same
plot: the positions are written after `line`, and `point`, which names none,
reads them too. `color()` does not work this way. Written after one mark, it
never reaches another.

Naming a column per layer also lets you put two columns of one table on one
axis:

```{r}
#| echo: false
peek(sessions)
```

```{r}
data(sessions) + x(session) +
  line + y(open) + style(color = "steelblue") +
  line + y(close) + style(color = "tomato", pattern = "dashed") +
  y_label("Price")
```

*"Given the sessions: x is session, a line, y is open, colored steelblue, and
also a line, y is close, colored tomato, with pattern dashed."*

The solid steelblue line is `open` and the dashed tomato line is `close`, both
on one y axis. `y_label("Price")` names that axis. Without it, the axis would be
named `Open`, after the first column written for it.

That plot works, but there is usually a better way to draw two series. Reshape
the table to [tidy data](data.qmd#the-table-has-to-be-tidy): one `value` column
and a `type` column, mapped with `color(type)`. That form draws a legend, and it
works for any number of series. Name the columns per layer in two cases only:
when the two lines need different settings, and when the second set of values
sits in another table.

A layer may name its own column, as the `text` layer did with `milestones`. It
may not name its own scale:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at, scale = "log") + y(value) + label(note)
```

A second scale on one axis is a second axis. The points would be placed on a
linear x axis, and the notes on a log x axis. A panel draws only one x axis, so
nothing would show which axis places the notes. The `milestones` plot wrote
`x(at)`. This one writes `x(at, scale = "log")`, and that is the only
difference. A layer names its own *column*, never its own scale.

## Channels: plot-scoped or layer-scoped

Positions share one axis, which is why they needed a section of their own. The
other channels have no shared axis between layers. They follow the same rule:
where you write a channel decides its scope.

A channel is often meant for one layer alone. You may want the points colored
by continent, and one line in a single color drawn through them. Written *after
a mark*, a channel binds to that mark alone:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +   # color applies to point only
  line                         # line uses a single default color
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line."*

Each point is drawn in its continent's color. The line takes no `color`, so it
is one dark gray stroke, joining the 142 countries in order of gdp. Whenever a
plot has a color legend, a layer that neither maps nor sets `color` is drawn in
this gray. No continent in the legend is gray, so the line cannot be mistaken
for one.

Written *before any mark*, a channel has no mark to bind to. It binds to the
plot instead, so every layer that can accept it gets it:

```{r}
data(gapminder_2007) + x(gdp) + y(life) + color(continent) +
  point +
  line
```

*"Given gapminder 2007: x is gdp, y is life, color by continent, points and
also a line."*

Now the line takes `color` as well. A line colored by a column splits into one
line per category, so the plot draws five lines, one per continent.

That is the same rule that `x` and `y` follow. Written before the marks, they
apply to every mark. Written after one mark, they belong to that mark.
Positions add the one rule from the previous section: a layer that names no
column of its own reads the first column written for that axis.

### Why a channel binds to one mark only

Suppose you want the lines colored by country and only the points sized by
population. A channel written after a mark binds to that mark alone, never to
an earlier one:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
# `size` is written after `point`, so it belongs to the points;
# the line written above it does not change, and keeps its own color
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

*"Given gapminder Asia: x is year, y is life, lines colored by country, and
also points sized by population."*

This plot is possible only because a channel binds to one mark. If `size` also
bound to the earlier `line`, gog would refuse the whole plot: a line takes one
size for the whole layer, not one per row. There would then be no way to
describe the plot you wanted.

### One column per legend

Binding a channel to one mark has one limit. A plot draws one legend for
`color`, and that legend names one column. So when two layers both map `color`,
they must name the same column. The same is true of `size`, `shape`, `pattern`
and `opacity`, the other channels that draw a legend.

Suppose the lines above stay colored by country, and the points are colored by
population instead of sized by it. The one legend would name `country`, the
first column written, and nothing would explain the colors of the points. So gog
refuses the plot:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + color(population)
```

The message names two other channels for `population`, each with its own
legend: `size` or `opacity` on the points. The plot above uses `size`. Writing
[`legend = FALSE`](channels.qmd#leaving-a-legend-out) on either `color` lets the
plot draw, with no color legend at all.

### A plot-scoped channel and the marks left unchanged

A channel bound to one mark must be one that mark can accept, or gog refuses
the plot. A channel written for the whole plot works differently. What happens
when some marks can accept it and others cannot? `size` written for the whole
plot applies to the points, and not to the line. A line has one width for the
whole layer rather than one width per row, so it has no size to map. gog
applies `size` to every mark that has a size feature, and prints a message
naming the marks it left unchanged. That message is an **Assumption**: gog says
what it decided instead of refusing the sentence, so the plot still draws:

```{r}
#| message: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country) +
  point
```

*"Given gapminder Asia: x is year, y is life, size by population, lines grouped
by country, and also points."*

If *no* mark in the plot can accept the channel, the binding does nothing at
all, so gog refuses it:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country)
```

The message names the mark that could not take the binding, and it says what to
do: remove `size`, or add a mark that has a size feature. A channel that no mark
can take is never accepted in silence.

## `data()`: a second table in one sentence

A forecast, a threshold or a set of labels usually comes in a table of its own,
so one plot often reads two tables. The [Data](data.qmd) chapter already gave
the rule for two tables, and that rule is a scope rule. The first `data()` is
the plot's table. Each later `data()` binds to the mark written directly after
it.

```{r}
#| echo: false
peek(forecast)
```

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point     # forecast applies to point, not to line
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points."*

Where you write `data()` matters here. A later `data()` gives its table to the
next mark only. A mark written after that one reads the plot's table again.

## What comes next

Scope decides what an atom binds to inside one panel: the plot, or one layer.
[Faceting](faceting.qmd) makes more panels, one per category of a column. Every
panel is drawn by the same sentence, and by default every panel is read against
the same scales.
