# Encoding scope

```{r}
#| include: false
source("R/setup.R")
```

If you write one `x()` and then two marks, which marks use it? Both do. Each
atom binds to a **scope**: the whole plot, or one layer. This chapter gives the
rule that decides which.

## The rule

| Atom | Scope | Position-sensitive? |
|------|-------|-------------------|
| `x()`, `y()`, `z()` **before any mark** | **Plot**, every layer | Yes |
| the same positions **after a mark** | **Layer**, that mark's column for the shared axis | Yes |
| `data()` **before any mark** | **Plot**, default table | No |
| `data()` **after a mark** | **Next layer** only | Yes |
| `color()`, `size()`, `shape()`, `pattern()`, `opacity()`, `group()`, `label()`, `play()` **before any mark** | **Plot**, every layer that can accept it | Yes |
| the same channels **after a mark** | **Layer**, that mark alone | Yes |
| `style()` | **Layer**, nearest preceding mark; refused when no mark precedes it | Yes |
| `title()`, `x_label()`, `y_label()`, `z_label()`, `palette()`, `theme()`, `brush()` | **Plot** | No |

One rule covers the position-sensitive rows: a mark binds forward, so it takes
every atom written after it. The rows marked **No** (the labels, `palette()`,
and a `data()` before any mark) belong to the plot wherever they are written.
Write a channel before any mark and it binds to the plot; write it after a mark
and it binds to that mark.

Where two bindings could apply, the **nearest** one wins: the same rule that
lets a mid-expression `data()` give the next layer its own table.

## One axis, and which column each layer reads

Every layer of a plot shares the same axes. That is what makes it one plot
rather than two drawn on top of each other. gog allows no second axis: one x
scale, one set of ticks, one label. A table that does not belong on those axes
does not belong in this plot. [Composition](composition.qmd) covers the choice
between layering, faceting, and a page of separate plots.

Sharing an axis is not the same as sharing a *column name*, and the two are
easy to confuse. Bind the positions once before the marks and every layer reads
them, which is what you want almost always:

```{r}
#| echo: false
peek(actuals)
```

```{r}
# x and y defined once: both marks use them
data(actuals) + x(year) + y(sales) +
  bar +
  line
```

*"Given the actuals: x is year, y is sales, bars and also a line."*

Write a position after a mark and it belongs to that mark, exactly as `color()`
does. The axis is unchanged. The layer is only saying which of *its* columns
supplies the values. That rule is what lets `milestones` in [Data](data.qmd)
keep its own column names:

```{r}
#| echo: false
peek(gapminder_2007)
peek(milestones)
```

```{r}
# `milestones` calls its columns `at` and `value`, not `gdp` and `life`
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at) + y(value) + label(note)
```

*"Given gapminder 2007: points, x is gdp, y is life; then given the milestones:
text, x is at, y is value, label by note."*

Look at the axis labels: they still read `Gdp` and `Life`. The `milestones`
column names never reach the axis labels, because the axis belongs to the plot.
Only the column choice is local. Without this rule, `milestones` would have to
rename `at` and `value` to `gdp` and `life`. That is easy when you are inventing
a table, and tedious when the table already exists.

The same rule covers two columns of one table on one axis:

```{r}
#| echo: false
peek(sessions)
```

```{r}
data(sessions) + x(session) +
  line + y(open) + style(color = "steelblue") +
  line + y(close) + style(color = "tomato", pattern = "dashed")
```

*"Given the sessions: x is session, a line, y is open, colored steelblue, and
also a line, y is close, colored tomato."*

Reshaping to tidy data is still usually better: one `value` column plus a
`type` column, mapped with `color(type)`. That form earns a legend and handles
more than two series. Name the columns per layer when you want the two lines
styled differently. Do the same when the second set of values comes from
another table.

What a layer may *not* do is bring its own scale:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at, scale = "log") + y(value) + label(note)
```

A second scale on one axis is a second axis. Every mark would then be measured
against a scale you never wrote. This refusal and the plot before it differ by
one word: a layer names its own *column*, never its own scale.

## Channels: plot-scoped or layer-scoped, by position

Positions were the hard case, because every layer shares one axis. The other
channels share nothing, and they follow the same position rule.

A channel is often meant for one layer alone: colored points under a single
plain line, for example. Written *after a mark*, a channel binds to that mark
alone:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +   # color applies to point only
  line                         # line uses a single default color
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line."*

Written *before any mark*, a channel has no mark to bind to, so it binds to the
plot and covers every layer:

```{r}
data(gapminder_2007) + x(gdp) + y(life) + color(continent) +
  point +
  line
```

*"Given gapminder 2007: x is gdp, y is life, color by continent, points and also
a line."*

That is the same position rule `x` and `y` follow. Written before the marks,
they apply to every mark. Written after one mark, they belong to that mark. No
channel has a scoping rule of its own. For `x`, `y` and `z`, a layer-scoped
binding names a column, never a new axis.

### Why binding only goes forward

Suppose you want the lines colored by country and only the points sized by
population. Because a mark takes only the atoms written after it, the same
atoms always mean the same thing:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
# `size` is written after `point`, so it belongs to the points;
# the line beside them does not change, and keeps its own color
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

*"Given gapminder Asia: x is year, y is life, lines colored by country, and also
points sized by population."*

A channel that bound backward could not draw this plot. It would put `size` on
the line as well. A line has no size, so gog would refuse the whole plot, and
you could not say what you meant.

### A plot-scoped channel and the marks it skips

`size` written for the whole plot applies to the points, because a line has no
size. gog applies it to every mark that has the feature and prints a message
naming the marks it skipped. That message is an **assumption**: it reports a
choice rather than refusing one, so the plot still renders:

```{r}
#| message: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country) +
  point
```

*"Given gapminder Asia: x is year, y is life, size by population, lines grouped
by country, and also points."*

If *no* mark in the plot can accept the channel, the binding does nothing at
all, so gog refuses it:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country)
```

## `data()` mid-expression

A forecast, a threshold or a set of labels usually arrives in a table of its
own, so one plot often reads two tables. [Data](data.qmd) states this rule, and
it is a scope rule. The first `data()` is the plot's table. Each later `data()`
applies to the mark written directly after it.

```{r}
#| echo: false
peek(forecast)
```

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point     # ← forecast applies to point, not to line
```

Position matters here because each `data()` marks where one layer's table ends
and the next layer's begins.

## What comes next

Scope settles which layer reads an atom inside one panel.
[Faceting](faceting.qmd) makes more panels, one per category, every one drawn by
the same sentence and read against the same axes.
