# Encoding scope

```{r}
#| include: false
source("R/setup.R")
```

If you write one `x()` and then two marks, which marks use it? Both do. Each
atom binds to a **scope**: the whole plot, or one layer, and this chapter is
the rule for which.

## The rule

| Atom | Scope | Position-sensitive? |
|------|-------|-------------------|
| `x()`, `y()`, `z()` **before any mark** | **Plot**, every layer | Yes |
| the same positions **after a mark** | **Layer**, that mark's column for the shared axis | Yes |
| `data()` **before any mark** | **Plot**, default table | No |
| `data()` **after a mark** | **Next layer** only | Yes |
| `color()`, `size()`, `shape()`, `opacity()`, `group()` **before any mark** | **Plot**, every layer that can accept it | Yes |
| the same channels **after a mark** | **Layer**, that mark alone | Yes |
| `style()` | **Layer**, nearest preceding mark | Yes |
| `title()`, `x_label()`, `y_label()`, `z_label()`, `palette()` | **Plot** | No |

One rule covers the position-sensitive rows: an atom binds forward, never
backward, to whatever it is written after. The rows marked **No** (the labels,
`palette()`, and a `data()` before any mark) belong to the plot wherever they
are written. Write a channel before any mark and it binds to the plot; write it
after a mark and it binds to that mark.

Where two bindings could apply, the **nearest** one wins: the same rule that
lets a mid-expression `data()` replace the plot's table.

## One axis, and which column each layer reads

Every layer of a plot shares the same axes. That is what makes it one plot
rather than two drawn on top of each other. gog allows no second axis: one x
scale, one set of ticks, one label. A table that does not belong on those axes
does not belong in this plot. [Composition](composition.qmd) covers the choice
between layering, faceting, and a page of separate plots.

Sharing an axis is not the same as sharing a *column name*, and the two are
easy to confuse. Bind the positions once before the marks and every layer reads
them, which is what you want almost always:

```{r}
# x and y defined once: both marks use them
data(actuals) + x(year) + y(sales) +
  bar +
  line
```

*"Given the actuals: x is year, y is sales, bars and also a line."*

Write a position after a mark and it belongs to that mark, exactly as a color
does. The axis does not move; the layer is only saying which of *its* columns
supplies the values. That is the scope rule behind the second table in
[Data](data.qmd), which kept its own column names:

```{r}
# `milestones` calls its columns `at` and `value`, not `gdp` and `life`
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at) + y(value) + label(note)
```

Look at the axis labels: they still read `Gdp` and `Life`. The note's own column
names never reach the axis labels, because the axis is the plot's and only the
column choice is local. Without this, the note table would have to rename its
columns to match the base table, which is fine when you are inventing the table
and tedious when it already exists.

The same rule covers two columns of one table on one axis:

```{r}
data(actuals) + x(year) +
  line + y(sales) + style(color = "steelblue") +
  line + y(sales) + style(color = "tomato", pattern = "dashed")
```

Reshaping to tidy data (one `value` column plus a `type` column) and using
`color(type)` is still usually better: it earns a legend and handles more than
two series. Naming the columns per layer is for when you want the two lines
styled differently, or when the second set of values genuinely lives elsewhere.

What a layer may *not* do is bring its own scale:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at, scale = "log") + y(value) + label(note)
```

A second scale on one axis is a second axis, and every mark would then be read
against something the expression never named. That refusal and the feature above
differ by one word: a layer names its own **column**, never its own scale.

## Channels: plot-scoped or layer-scoped, by position

Written **after a mark**, a channel binds to that mark alone:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +   # color applies to point only
  line                         # line uses a single default color
```

Written **before any mark**, a channel has no mark to bind to, so it binds to
the plot and covers every layer:

```{r}
data(gapminder_2007) + x(gdp) + y(life) + color(continent) +
  point +
  line
```

That is the same position rule `x` and `y` follow: written before the marks they
apply to all of them, written after one they belong to it. No channel has a
scoping rule of its own. For `x`, `y` and `z`, a layer-scoped binding names a
column, never a new axis.

### Why binding only goes forward

Because a channel binds to what precedes it and never reaches back, the same
atoms always mean the same thing:

```{r}
# `size` is written after `point`, so it belongs to the points,
# the line beside them is untouched, and keeps its own color
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

A channel that bound backward could not draw this. It would put `size` on
the `line` too, and a line has no size, so the plot would be refused, with no
way to say what you meant.

### A plot-scoped channel skips marks without it

`size` written for the whole plot means the points; a line has no size.
gog applies it where it fits and says where it does not; this is an
**assumption**, not a refusal, so the plot still renders:

```{r}
#| message: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country) +
  point
```

If *no* mark in the plot can accept the channel, the binding would do nothing at
all, so it is refused instead:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country)
```

## `data()` mid-expression

[Data](data.qmd) gave the rule, and it is a scope rule: the first `data()` is
the plot's table, and each later `data()` applies to the mark written directly
after it.

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point     # ← forecast applies to point, not to line
```

Position matters here because `data()` is a boundary marker between layers.
