# Encoding scope

```{r}
#| include: false
source("R/setup.R")
```

If you write one x and then two marks, which marks does it apply to? Each atom
in a plot expression binds to a specific scope. Understanding this one rule
eliminates all ambiguity about which encoding belongs to which part of the
plot.

## The rule

| Atom | Scope | Position-sensitive? |
|------|-------|-------------------|
| `x()`, `y()`, `z()` **before any mark** | **Plot**, every layer | Yes |
| the same positions **after a mark** | **Layer**, that mark's column for the shared axis | Yes |
| `data()` at expression start | **Plot**, default table | No |
| `data()` between two marks | **Next layer** only | Yes |
| `color()`, `size()`, `shape()`, `opacity()`, `group()` **before any mark** | **Plot**, every layer that can accept it | Yes |
| the same channels **after a mark** | **Layer**, that mark alone | Yes |
| `style()` | **Layer**, nearest preceding mark | Yes |
| `title()`, `x_label()`, `y_label()`, `z_label()`, `palette()` | **Plot** | No |

One rule covers the whole table: an atom binds forward, never backward, and
what it binds to is whatever it is written after. Write a channel before any
mark and it has only the plot to attach to; write it after a mark and it attaches
to that mark.

Where two bindings could apply, the **nearer** one wins: the same rule that
makes a mid-expression `data()` beat the plot's table.

## One axis, and which column each layer reads

Every layer of a plot shares the same axes. That is what makes it one plot
rather than two drawn on top of each other, and it is not negotiable: there is
one x scale, one set of ticks, one label. A table that does not belong on those
axes does not belong in this plot. [Composition](composition.qmd) has the choice
between layering, faceting, and a page of separate plots.

Sharing an axis is not the same as sharing a *column name*, though, and those
two were run together for a long time. Bind the positions once before the marks
and every layer reads them, which is what you want almost always:

```{r}
# x and y defined once: both marks use them
data(actuals) + x(year) + y(sales) +
  bar +
  line
```

*"Given the actuals: x is year, y is sales, bars and also a line."*

Write a position after a mark and it belongs to that mark, exactly as a color
does. The axis does not move; the layer is only saying which of *its* columns
supplies the values. That is what lets a second table keep its own vocabulary:

```{r}
# `milestones` calls its columns `at` and `value`, not `gdp` and `life`
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at) + y(value) + label(note)
```

Look at the axis labels: they still read `Gdp` and `Life`. The note's own column
names never surface as chrome, because the axis is the plot's and only the
reading is local. Without this, the note table would have to rename its columns
to match the base data, which is fine when you are inventing the table and
tedious when it already exists.

The same rule answers two columns of one table on one axis:

```{r}
data(actuals) + x(year) +
  line + y(sales) + style(color = "steelblue") +
  line + y(sales) + style(color = "tomato", pattern = "dashed")
```

Reshaping to tidy form (one `value` column plus a `type` column) and using
`color(type)` is still usually better, because it earns you a legend and scales
past two series. Naming the columns per layer is for when you want the two lines
styled differently, or when the second set of values genuinely lives elsewhere.

What a layer may *not* do is bring its own scale:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at, scale = "log") + y(value) + label(note)
```

A second scale on one axis is a second axis, and every mark would then be read
against something the expression never named. That is the refusal in
[Design laws](design-laws.qmd), and the line between it and the feature above is
exactly one word wide: a layer names its own **column**, never its own scale.

## Channels: plot-scoped or layer-scoped, by position

Written **after a mark**, a channel binds to that mark alone:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +   # color applies to point only
  line                         # line uses a single default color
```

Written **before any mark**, there is no mark to attach to, so it belongs to the
plot and reaches every layer:

```{r}
data(gapminder_2007) + x(gdp) + y(life) + color(continent) +
  point +
  line
```

That is the same position rule `x` and `y` follow: written before the marks they
apply to all of them, written after one they belong to it. There is no channel
with a scoping rule of its own, and the positions differ only in what a layer-scoped
one *means*, which is a column for the shared axis rather than an axis of its
own.

### Why binding only goes forward

Because a channel binds to what precedes it and never reaches back, the same
atoms always mean the same thing:

```{r}
# `size` is written after `point`, so it belongs to the points,
# the line beside them is untouched, and keeps its own color
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

A channel that reached backwards could not express this. It would put `size` on
the `line` too, and a line has no size, so the plot would be refused, with no
way to say what you meant.

### A plot-scoped channel skips marks without the feature

`size` written for the whole plot means the points; a line has no size to give.
gog applies it where it fits and says where it did not; this is an
**assumption**, not a refusal, so the plot still renders:

```{r}
#| message: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country) +
  point
```

If *no* mark in the plot has the feature, the binding would do nothing at all,
so it is refused instead:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country)
```

## `data()` mid-expression

A `data()` that appears between two marks is layer-scoped:
it sets the table for the mark that immediately follows it.

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point     # ← forecast applies to point, not to line
```

Position matters here because `data()` is a boundary marker between layers.
