# Encoding scope

```{r}
#| include: false
source("R/setup.R")
```

If you write one `x()` and then two marks, which marks use it? Both do. Each
atom binds to a **scope**: the whole plot, or one layer. This chapter gives the
rule that decides between the two.

## The rule

The whole rule fits in one table. Each row names an atom, the scope it binds
to, and whether its position matters.

| Atom | Scope | Position-sensitive? |
|------|-------|-------------------|
| `x()`, `y()`, `z()` **before any mark** | **Plot**, every layer | Yes |
| the same positions **after a mark** | **Layer**, that mark's column for the shared axis | Yes |
| `data()` **before any mark** | **Plot**, default table | No |
| `data()` **after a mark** | **Next layer** only | Yes |
| `color()`, `size()`, `shape()`, `pattern()`, `opacity()`, `group()`, `label()`, `play()` **before any mark** | **Plot**, every layer that can accept it | Yes |
| the same channels **after a mark** | **Layer**, that mark alone | Yes |
| `style()` | **Layer**, nearest preceding mark; refused when no mark precedes it | Yes |
| `title()`, `x_label()`, `y_label()`, `z_label()`, `palette()`, `theme()`, `brush()` | **Plot** | No |

One rule covers the position-sensitive rows. A mark takes every atom written
after it, and none of the atoms written before it. Four kinds of atom are
marked **No**: the labels, `palette()`, `theme()` and `brush()`. Each belongs
to the plot wherever you write it. A `data()` before any mark is the plot's
table. Write a channel before any mark and it binds to the plot; write it after
a mark and it binds to that mark.

Where two bindings could apply, the **nearest** one wins. That is the same rule
that lets a `data()` written between two marks give the next layer its own
table.

## One axis, and which column each layer reads

Every layer of a plot shares the same axes. That is what makes it one plot
rather than two drawn on top of each other. gog allows no second axis: one x
scale, one set of ticks, one label. So a second table's values must fall on the
same scale as the first. Values measured on anything else need a plot of their
own. [Composition](composition.qmd) covers the choice between layering,
faceting, and a page of separate plots.

Sharing an axis is not the same as sharing a *column name*, and the two are
easy to confuse. Bind the positions once before the marks and every layer reads
them, which is what you want almost always:

```{r}
#| echo: false
peek(actuals)
```

```{r}
# x and y defined once: both marks use them
data(actuals) + x(year) + y(sales) +
  bar +
  line
```

*"Given the actuals: x is year, y is sales, bars and also a line."*

Write a position after a mark and it belongs to that mark, exactly as `color()`
does. The axis is unchanged. The layer is only saying which of *its* columns
supplies the values. That rule is why the `milestones` table from
[Data](data.qmd) can keep its own column names:

```{r}
#| echo: false
peek(gapminder_2007)
peek(milestones)
```

```{r}
# `milestones` calls its columns `at` and `value`, not `gdp` and `life`
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at) + y(value) + label(note)
```

*"Given gapminder 2007: points, x is gdp, y is life; then given the milestones:
text, x is at, y is value, label by note."*

Look at the axis labels: they still read `Gdp` and `Life`. The `milestones`
column names never appear in the axis labels, because the axis belongs to the
plot. Only the choice of column belongs to the layer. Without this rule,
`milestones` would have to rename `at` and `value` to `gdp` and `life`.
Renaming is easy when you build the table yourself. When the table comes from
somewhere else, renaming means changing data you did not write.

The same rule lets one table put two of its own columns on one axis:

```{r}
#| echo: false
peek(sessions)
```

```{r}
data(sessions) + x(session) +
  line + y(open) + style(color = "steelblue") +
  line + y(close) + style(color = "tomato", pattern = "dashed")
```

*"Given the sessions: x is session, a line, y is open, colored steelblue, and
also a line, y is close, colored tomato, with pattern dashed."*

That plot works, and it is not the first thing to try. Reshaping to tidy data
is usually better: one `value` column plus a `type` column, mapped with
`color(type)`. That form draws a legend, and it works for any number of series.
Name the columns per layer in two cases only: when the two lines need different
settings, and when the second set of values sits in another table.

A layer may not set its own scale:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at, scale = "log") + y(value) + label(note)
```

A second scale on one axis is a second axis. One panel would then hold two
coordinate spaces, and every mark would be placed by an axis that no sentence
in the plot names. This refusal and the `milestones` plot differ by one
argument: a layer names its own *column*, never its own scale.

## Channels: plot-scoped or layer-scoped, by position

Positions had to share one axis, which is why they needed a section of their
own. The other channels have no shared axis between layers. They follow the
same position rule.

A channel is often meant for one layer alone. You may want the points colored
by group and one plain line drawn through them. Written *after a mark*, a
channel binds to that mark alone:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +   # color applies to point only
  line                         # line uses a single default color
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line."*

Written *before any mark*, a channel has no mark to bind to. It binds to the
plot instead, so every layer gets it:

```{r}
data(gapminder_2007) + x(gdp) + y(life) + color(continent) +
  point +
  line
```

*"Given gapminder 2007: x is gdp, y is life, color by continent, points and
also a line."*

That is the same position rule that `x` and `y` follow. Written before the
marks, they apply to every mark. Written after one mark, they belong to that
mark. No channel has a scoping rule of its own. For `x`, `y` and `z`, a
layer-scoped binding names a column, never a new axis.

### Why binding works in one direction only

Suppose you want the lines colored by country and only the points sized by
population. A mark takes only the atoms written after it, so nothing you write
later can change what an earlier atom binds to:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
# `size` is written after `point`, so it belongs to the points;
# the line written above it does not change, and keeps its own color
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

*"Given gapminder Asia: x is year, y is life, lines colored by country, and
also points sized by population."*

This plot is possible only because binding goes one way. A channel that reached
backward as well would put `size` on the line too. A line has no per-row size,
so gog would refuse the whole plot, and there would be no way to describe the
plot you wanted.

### A plot-scoped channel and the marks left unchanged

`size` written for the whole plot applies to the points. A line has one width
for the whole layer rather than one width per row, so it has no size to map.
gog applies `size` to every mark that has a size feature, and prints a message
naming the marks it left unchanged. That message is an **assumption**: gog says
what it decided instead of refusing the sentence, so the plot still draws:

```{r}
#| message: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country) +
  point
```

*"Given gapminder Asia: x is year, y is life, size by population, lines grouped
by country, and also points."*

If *no* mark in the plot can accept the channel, the binding does nothing at
all, so gog refuses it:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country)
```

The message names the mark that could not take the binding, and it says what to
do: remove `size`, or add a mark that has a size feature. A binding that would
draw nothing is never accepted in silence.

## `data()`: a second table in one sentence

A forecast, a threshold or a set of labels usually arrives in a table of its
own, so one plot often reads two tables. The [Data](data.qmd) chapter already
gave the rule for two tables, and that rule is a scope rule. The first `data()`
is the plot's table. Each later `data()` applies to the mark written directly
after it.

```{r}
#| echo: false
peek(forecast)
```

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point     # forecast applies to point, not to line
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points."*

Position matters here. Each `data()` shows where one layer's table ends and the
next layer's table begins.

## What comes next

Scope decides which layer reads an atom inside one panel.
[Faceting](faceting.qmd) makes more panels, one per category of a column. Every
panel is drawn by the same sentence and read against the same axes.
