# Encoding scope

```{r}
#| include: false
source("R/setup.R")
```

If you write one `x()` and then two marks, which marks use it? Both do. Each
atom binds to a **scope**: the whole plot, or one layer. A layer is one mark
and the atoms bound to it. This chapter gives the rule that decides between the
two.

## The rule

The rule for every atom you write with `+` fits in one table. Each row names an
atom, the scope it binds to, and whether it matters where you write it.

| Atom | Scope | Depends on where you write it? |
|-------------------------------------------|-------------------------------------------|--------------|
| `x()`, `y()`, `z()` **before any mark** | **Plot**, every layer | Yes |
| `x()`, `y()`, `z()` **after a mark** | **Layer**, the column that mark reads. A layer with no `x()` of its own reads the first `x()` in the sentence, and the same holds for `y()` and `z()` | Yes |
| the **first** `data()` or `query()` | **Plot**, default table | Yes |
| each **later** `data()` or `query()` | **Next layer** only | Yes |
| `color()`, `size()`, `shape()`, `pattern()`, `opacity()`, `group()`, `label()`, `play()` **before any mark** | **Plot**, passed to every layer whose mark can take it. A layer that cannot take it is left unchanged, and gog says so ([below](#a-plot-scoped-channel-and-the-marks-left-unchanged)) | Yes |
| `color()`, `size()`, `shape()`, `pattern()`, `opacity()`, `group()`, `label()`, `play()` **after a mark** | **Layer**, that mark alone | Yes |
| `style()` | **Layer**, nearest preceding mark; refused when no mark precedes it | Yes |
| `title()`, `x_label()`, `y_label()`, `z_label()`, `palette()`, `theme()`, `order()`, `brush()`, and a space: `space()`, `polar()`, `nest()`, `map()`, `globe()` or `network()` | **Plot** | No |
| `facet()`, written after `|` or `/` | **Plot**, the whole plot | No |

In every row marked **Yes**, where you write the atom decides its scope. Write a
channel before any mark and it binds to the plot. Write a channel or `style()`
after a mark and it binds to that mark. `data()` and `query()` work
differently. The first `data()` or `query()` opens the sentence and is the
plot's table. Each later one binds to the mark written directly after it, not to
the mark before it. The atoms in the last two rows, marked **No**, belong to the
plot wherever you write them. They are the title and the axis labels,
`palette()`, `theme()`, `order()`, `brush()`, the spaces, and `facet()`, which
always splits the whole plot.

The positions, `x()`, `y()` and `z()`, need a closer look when they are written
after a mark, because this book writes them there in almost every sentence, as
in `data(gapminder_2007) + point + x(gdp) + y(life)`. These positions belong to
`point`. They still reach every other layer that names no column of its own for
that axis. So in practice they serve the whole plot, until a layer names a
column of its own. The next section shows that case with an example.

Where two bindings could apply, the nearest one wins. That is the same rule
that lets a `data()` written between two marks give the next layer its own
table in place of the plot's.

## One axis, and which column each layer reads

The rule above never gives a layer axes of its own. Every layer of a plot
shares the same axes. That is what makes it one plot rather than two drawn on
top of each other. gog allows no second axis. Each of `x`, `y` and `z` has one
scale, one set of ticks and one label. So a second table's values must be in the
same units as the first table's, such as dollars in both. Values in different
units need a plot of their own. [Composition](composition.qmd) covers the
choice between layering, faceting, and a page of separate plots.

Sharing an axis is not the same as sharing a *column name*, and the two are
easy to confuse. Almost always you want to bind the positions once, before the
marks, so that every layer reads them:

```{r}
#| echo: false
peek(actuals)
```

```{r}
# x and y written once, before the marks: all three marks use them
data(actuals) + x(year) + y(sales) +
  bar +
  line +
  point
```

*"Given the actuals: x is year, y is sales, bars, and also a line, and also
points."*

That sentence has three layers, one for each mark. The plot is not a fourth
layer. It holds the table and the positions that all three read, and it draws
the frame around them: the axes with their ticks and labels, and a title or a
legend when the sentence has one. Only the layers draw the data.

Write a position after a mark, and that mark reads the column you name. No
second axis appears. The layer only chooses which of *its* columns supplies the
values. That rule is why the `milestones` table from [Data](data.qmd) can keep
its own column names:

```{r}
#| echo: false
peek(gapminder_2007)
peek(milestones)
```

```{r}
# `milestones` calls its columns `at` and `value`, not `gdp` and `life`
data(gapminder_2007) + point + style(color = "#9e9e9e") + x(gdp) + y(life) +
  data(milestones) + text + x(at) + y(value) + label(note) +
  style(size = 15, border_size = 8)
```

*"Given gapminder 2007: points, x is gdp, y is life; then given the milestones:
text, x is at, y is value, label by note, with size 15 and border size 8."*

The two `style()` settings only make the notes easier to read: the points are
gray, and each note has a halo, a band of the background color under its letters
([A halo around each label](marks/text.qmd#a-halo-around-each-label)).

Look at the axis labels: they are still `Gdp` and `Life`. Each axis is named
after the first column written for it, unless you name the axis with
`x_label()` or `y_label()`. Here those first columns are `gdp` and `life`,
written after `point`. The axis belongs to the plot, and only the choice of
column belongs to the layer. Without this rule, you would have to rename `at`
and `value` in `milestones` to `gdp` and `life`. Renaming is easy when you
build the table yourself. When the table comes from somewhere else, renaming
means changing data you did not write.

A layer that names no column of its own reads the first column written for each
axis, wherever the layer is written. So `point + x(gdp) + y(life) + line` draws
a line through the points. `point + line + x(gdp) + y(life)` draws the same
plot: the positions are written after `line`, and `point`, which names none,
reads them too. `color()` does not work this way. Written after one mark, it
never reaches another. [Which orders matter](operators.qmd#which-orders-matter)
showed that difference with a trend line, and
[Channels: plot-scoped or layer-scoped](#channels-plot-scoped-or-layer-scoped)
below gives the rule.

Naming a column per layer also lets you put two columns of one table on one
axis:

```{r}
#| echo: false
peek(sessions)
```

```{r}
data(sessions) + x(session) +
  line + y(open) + style(color = "steelblue") +
  line + y(close) + style(color = "tomato", pattern = "dashed") +
  y_label("Price")
```

*"Given the sessions: x is session, a line, y is open, colored steelblue, and
also a line, y is close, colored tomato, with pattern dashed."*

The solid steelblue line is `open` and the dashed tomato line is `close`, both
on one y axis. `y_label("Price")` names that axis. Without it, the axis would be
named `Open`, after the first column written for it.

That plot works, but there is usually a better way to draw two series. Reshape
the table to [tidy data](data.qmd#the-table-has-to-be-tidy): one `value` column
and a `type` column, mapped with `color(type)`. That form draws a legend, and it
works for any number of series. A `y()` in each layer, as above, is the better
choice only in one of these situations:

- **A setting.** The lines need a setting that no channel can map. A line's
  width and its opacity can only be set, and a setting applies to a whole layer,
  so two widths need two layers. A color or a dash can be mapped instead, with
  `color(type)` or `pattern(type)`.
- **Another table.** The values sit in another table, as the notes in
  `milestones` did.

A layer may name its own column, as the `text` layer did with `milestones`. It
may not name its own scale:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(milestones) + text + x(at, scale = "log") + y(value) + label(note)
```

A second scale on one axis is a second axis. The points would be placed on a
linear x axis, and the notes on a log x axis. A panel draws only one x axis, so
nothing would show which axis places the notes. The `milestones` plot wrote
`x(at)`. This one writes `x(at, scale = "log")`, and that is the only
difference. A layer names its own *column*, never its own scale.

## Channels: plot-scoped or layer-scoped

The section above was about positions alone, because every layer shares the
same `x`, `y` and `z` axes, and that sharing gives positions their extra rules.
The other channels, such as `color()` and `size()`, have no axis. They follow
only the rule in the table at the start of this chapter: where you write a
channel decides its scope.

A channel is often meant for one layer alone. You may want the points colored
by country, and one trend line in a single color, fitted to all of them
together. Written *after a mark*, a channel binds to that mark alone. The table
`gapminder_asia` holds five Asian countries, one row per country and year:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + x(year) + y(life) +
  point + color(country) +   # color applies to point only
  line * smooth              # the trend line uses a single default color
```

*"Given gapminder Asia: x is year, y is life, points colored by country, and
also a line derived by smooth."*

Each point is drawn in its country's color. The line takes no `color`, so it is
one dark gray trend line, fitted to all five countries together. Whenever a plot
has a color legend, a layer that neither maps nor sets `color` is drawn in this
gray. No country in the legend is gray, so the line cannot be mistaken for one.

Written *before any mark*, a channel has no mark to bind to. It binds to the
plot instead, so every layer that can accept it gets it:

```{r}
data(gapminder_asia) + x(year) + y(life) + color(country) +
  point +
  line * smooth
```

*"Given gapminder Asia: x is year, y is life, color by country, points and
also lines derived by smooth."*

Now the line takes `color` as well. A line colored by a categorical column
splits into one line per category, so the plot draws five trends, one per
country.

That is the same rule that `x` and `y` follow. Written before the marks, they
apply to every mark. Written after one mark, they belong to that mark.
Positions add the one rule from the previous section: a layer that names no
column of its own reads the first column written for that axis.

### Why a channel binds to one mark only

Suppose you want the lines colored by country and only the points sized by
population. A channel written after a mark binds to that mark alone, never to
an earlier one:

```{r}
# `size` is written after `point`, so it belongs to the points;
# the line written above it does not change, and keeps its own color
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + size(population)
```

*"Given gapminder Asia: x is year, y is life, lines colored by country, and
also points sized by population."*

This plot is possible only because a channel binds to one mark. If `size` also
bound to the earlier `line`, gog would refuse the whole plot: a line takes one
size for the whole layer, not one per row. There would then be no way to
describe the plot you wanted.

### One column per legend

Binding a channel to one mark has one limit. A plot draws one legend for
`color`, and that legend names one column. So when two layers both map `color`,
they must name the same column. The same is true of `size`, `shape`, `pattern`
and `opacity`, the other channels that draw a legend.

Suppose the lines above stay colored by country, and the points are colored by
population instead of sized by it. The one legend would name `country`, the
first column written, and nothing would explain the colors of the points. So gog
refuses the plot:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) +
  line  + color(country) +
  point + color(population)
```

The message names two other channels for `population`, each with its own
legend: `size` or `opacity` on the points. The plot above uses `size`. Writing
[`legend = FALSE`](channels.qmd#leaving-a-legend-out) on either `color` lets the
plot draw, with no color legend at all.

### A plot-scoped channel and the marks left unchanged

A channel bound to one mark must be one that mark can accept, or gog refuses
the plot. A channel written for the whole plot works differently. What happens
when some marks can accept it and others cannot? `size` written for the whole
plot applies to the points, and not to the line. A line has one width for the
whole layer rather than one width per row, so it has no size to map. gog
applies `size` to every mark that has a size feature, and prints a message
naming the marks it left unchanged. That message is an **Assumption**: gog says
what it decided instead of refusing the sentence, so the plot still draws:

```{r}
#| message: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country) +
  point
```

*"Given gapminder Asia: x is year, y is life, size by population, lines grouped
by country, and also points."*

If *no* mark in the plot can accept the channel, the binding does nothing at
all, so gog refuses it:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + size(population) +
  line + group(country)
```

The message names the mark that could not take the binding, and it says what to
do: remove `size`, or add a mark that has a size feature. A channel that no mark
can take is never accepted in silence.

## `data()`: a second table in one sentence

A forecast, a threshold or a set of labels usually comes in a table of its own,
so one plot often reads two tables. The [Data](data.qmd) chapter already gave
the rule for two tables, and that rule is a scope rule. The first `data()` is
the plot's table. Each later `data()` binds to the mark written directly after
it.

```{r}
#| echo: false
peek(forecast)
```

```{r}
data(actuals) + x(year) + y(sales) +
  line +
  data(forecast) + point     # forecast applies to point, not to line
```

*"Given the actuals: x is year, y is sales, a line; then given the forecast:
points."*

Where you write `data()` matters here. A later `data()` gives its table to the
next mark only. A mark written after that one reads the plot's table again.

## What comes next

Scope decides what an atom binds to inside one panel: the plot, or one layer.
[Faceting](faceting.qmd) makes more panels, one per category of a column. Every
panel is drawn by the same sentence, and by default every panel is read against
the same scales.
