# Faceting

```{r}
#| include: false
source("R/setup.R")
```

How do you compare the same plot across groups? A facet splits one plot into
small multiples: one panel per category of a column. Wilkinson calls facets
*frames of frames*: the facet column's categories form an outer frame, and each
of its cells holds a copy of the inner x/y frame. Nothing else about the plot
changes. The marks, channels, and transforms you have already learned draw each
panel exactly as they would draw the whole.

The picture has been invented more than once, so you may know it under another
name. Becker and Cleveland's group at Bell Labs built it into S as **Trellis
display** [@becker1996trellis], the name borrowed from the garden trellis its
rows and columns resemble, and most people first saw it in S-PLUS. R draws it
that way still in the **lattice** package [@sarkar2026lattice], and its own
book is where the display is set out in full [@sarkar2008lattice]. Base R's
`coplot()` is the same idea
under a third name, the conditioning plot, because what a trellis conditions
*on* is the facet column. That is also why the operator in the next section is
`|`, the sign statistics already writes for conditioning.
[Composition](composition.qmd) finishes that history. Tufte's **small
multiples** [@tufte1990] names the result rather than the operation, which is the
division this book keeps: small multiples for the picture, `facet` for the thing
you write. And **facet** is Wilkinson's own word [@wilkinson2005], from the Latin
*facies*, a face: one side of a many-sided thing, like one face of a cut
diamond. gog takes his
because his is the widest of the four. A trellis, to him, is one arrangement a
facet can take, next to the scatterplot matrix and the row plot, and a grammar
needs the term that covers all three. The word every one of these systems shares
is **panel**, the cell a split produces, and this chapter uses it the same way.

## `|`: panels side by side

The `facet()` atom names the column that does the splitting, and `|` arranges
the panels in columns:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, split into panel columns by continent."*

Reading across, the panels share both axes, so a shift *between* continents
reads as plainly as a pattern within one. That shared frame is what faceting
is for: the alternative (one plot per continent, each with its own scale)
looks similar and stops the comparison working, with nothing on the page to
show it.

## `/`: panels stacked

`/` arranges the panels in rows. Stacking aligns the x axis, which is the
better direction when the comparison is *where things sit along x*:

```{r}
gm_ae <- gapminder_2007[gapminder_2007$continent %in% c("Africa", "Americas", "Europe"), ]
data(gm_ae) + bar * bin + x(life) / facet(continent)
```

The three distributions line up column for column: Africa's life expectancies
spread wide and low, Europe's cluster high and narrow.

That reading only works because the bars are the same width in all three
panels, and a histogram under a facet is where the count and the cut have to be
decided separately. Each panel counts its own rows, because a statistic always
sees the panel's subset as if it were the whole data, and that holds for every
transform on every mark: the No Exceptions law, met one panel at a time. But
`bin` is not only a statistic. It cuts the axis into cells *and*
tallies what lands in them, and the cut is a description of the axis rather
than a measurement of the rows, so it belongs with the shared scale. One set of
bins is cut from all three panels' rows at once, and each panel then counts
only its own into them.

It helps to see what the alternative would draw. Cut
per panel, Europe's narrow spread gets narrow bins and Africa's wide one gets
wide bins, so bars of different widths stand against a single axis. A bar four
years wide holding nine countries and a bar one year wide holding eight then
draw at nearly the same height, while the yearly rates they stand for differ by
more than a factor of three. The panels are still aligned and the axis still
looks correct, and yet the comparison this chapter began with has stopped
working.

## Crossing into a grid

Write both operators to cross two columns into a grid. Each operator applies
to the facet written after it, read left to right:

```{r}
gm_years <- gm_all[gm_all$year %in% c(1957, 2007), ]
gm_years$period <- as.character(gm_years$year)
data(gm_years) + point + x(gdp) + y(life) | facet(continent) / facet(period)
```

Fifty years, five continents, one frame: every panel column is a continent,
every panel row a period, and the shared scale lets a reader measure the
half-century's rise in life expectancy panel by panel.

The grid is a **crossing** in Wilkinson's sense: every row × column
combination gets a panel, and a combination with no rows still gets an empty
one. The frame says the combination is *possible*: an empty panel is
information, not an accident. (His *nesting*, draw only the combinations
that exist, is a different operation, and these operators deliberately do
not claim it.)

## Folding a long line of panels

One column with many categories makes a long line of panels, and a long line of
panels makes each one narrow. Here are the ten most populous countries of 2007,
each with its own half-century of life expectancy:

```{r}
ranked <- gapminder_2007[base::order(-gapminder_2007$population), ]
gm_top <- gm_all[gm_all$country %in% utils::head(ranked$country, 10), ]
data(gm_top) + line + group(country) + x(year) + y(life) | facet(country)
```

The shapes are still readable and nothing else on the panel is. Read the axis:
the tick labels have run into each other and print as `19502000`, and the last
panel's name is wider than the strip that prints it. Ten categories is a common
number, and one line of panels is already too narrow for it.

`wrap` folds that line into a rectangle. The number is how many panels to draw
before the line turns:

```{r}
data(gm_top) + line + group(country) + x(year) + y(life) | facet(country, wrap = 4)
```

Same ten panels, same shared scale, same sentence with one setting added. The
years are readable, `United States` fits its strip, and each country's rise is
now large enough to look at.

Notice what `wrap` does *not* say: which way the panels run. That was settled
before `wrap` was written, by the operator. `|` runs the line across, so four to
a row; `/` runs it down, so four to a column:

```{r}
data(gm_top) + line + group(country) + x(year) + y(life) / facet(country, wrap = 4)
```

The same number, the same ten countries, read down the columns instead of across
the rows. One number is enough because the direction is already in the sentence,
which is why there is no second setting for it. ggplot2 spells the same idea as
`nrow` and `ncol`.

Two details in the wrapped plots come from the fold itself. The first is the
bottom right, where the last row stops after two panels: those two places are
what is left over when ten panels are folded into a twelve-cell rectangle, so
nothing is drawn in them. They are not the empty panels of the previous section.
An empty panel in a crossing is a combination the frame says is possible and the
data has no example of, which is information; here there is no combination at
all, only room.

The second is the axis under `Mexico` and `Nigeria`. Tick labels belong to
panels that touch the margin the axis lives in, which in a full grid is the
bottom row. Once a row is ragged that rule stops describing the plot, so gog
uses the general rule it was always a special case of: a panel draws the x
axis when no panel sits below it. `Mexico` and `Nigeria` sit above the gap, so
they keep their year labels.

Wrapping a crossed grid is refused. The crossing has already made a rectangle
out of two columns, and a count beside it could only be a second answer about
the same shape:

```{r}
#| error: true
render_svg(data(gm_years) + point + x(gdp) + y(life) |
             facet(continent, wrap = 2) / facet(period))
```

## Giving a panel its own scale

Everything so far has used one scale across every panel, which is what makes
the panels comparable. It is also what makes some of them unreadable. Here is the
world's population by continent, one panel each:

```{r}
data(gm_all) + line * sum + x(year) + y(population) | facet(continent, wrap = 3)
```

Asia is nearly four billion people and Oceania about twenty-five million, so the
axis is sized for Asia and Oceania is a flat line along the bottom. Oceania's
line is drawn correctly, but the axis is too tall for its shape to show. The
plot is not wrong: it is telling you, correctly and
loudly, that these five quantities are not the same size. But if the question was
"how did each continent grow", it has answered a different one.

`free = TRUE` on the position fits that axis from each panel's own rows:

```{r}
data(gm_all) + line * sum + x(year) +
  y(population, free = TRUE) | facet(continent, wrap = 3)
```

Now every panel has a shape, and Oceania turns out to have more than doubled.
Notice the y axis: each panel carries its own numbers, because one set of labels
along the edge was enough only while the scale was shared.

Notice also what the second plot gives up, because it is the reason this is a
request and never a default. All five panels now look alike: five lines rising
left to right at about the same angle. Reading them side by side, nothing says
that one of those climbs is measured in billions and another in tens of millions,
and a reader who glances at the shapes without reading the axes will leave with
a false impression. The first plot cannot mislead that way and the second can, which
is why gog will not do this unless you ask.

Which axis is freed is decided by where you write it. `wrap` works the same way
with `|` and `/`: the sentence already names the axis, so `free`
does not name it again. There is no `free_x` or `free_y`, and freeing both is
writing it twice:

```{r}
data(gm_all) + point + x(gdp, free = TRUE) + y(life, free = TRUE) |
  facet(continent, wrap = 3)
```

There is a third position, and it takes `free` the same way. A cube sits in a
panel like any other plot ([Space](space.qmd)):

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  z(population, free = TRUE) + color(continent) | facet(continent)
```

Shared, the height axis has to reach China, so Oceania's two countries lie flat
on the floor with nothing to read. Freed, Oceania is measured in tens of millions
and has a shape again, with the loss this whole section is about: the five cubes
no longer compare. Nothing was added to the grammar to reach that plot. `free` is
written on a position and `z` is a position, so the third one accepts it like the
other two.

Three things follow a free scale, and none can be asked for on its own.
The first is the bin cut. The histogram earlier in this chapter shares one set of
bin edges across every panel, because a cut describes the axis and the axis was
shared. Free the axis and the edges follow it, for exactly the same reason:

```{r}
data(gm_all) + bar * bin + x(gdp, free = TRUE) | facet(continent, wrap = 3)
```

Each panel is cut from its own rows, so Africa gets bins the size of Africa's
spread rather than bins sized for the richest countries in the data. There is no
way to ask for a free scale
over a shared cut, which would be a picture whose bars are the panel's and whose
edges are the plot's.

The second is the panel's **guides**: its axis numbers and gridlines. A scale
says where the marks go, and where the axis prints its numbers. Freeing one has
to move both, or the numbers stop describing
the picture. Look again at the freed population plot above. Its gridlines sit at
different heights in every panel, and each panel's labels sit beside its own.
Shared, those same lines are drawn at one set of heights across all five. That is
the whole of what a freed guide means: the lines a reader measures against belong
to the scale that was freed.

The third is that a free scale belongs to a panel, never to an animation frame.
`play` ([Animation](play.qmd)) splits rows the same way a facet does, so asking for this
over frames is a sentence you can write, and it is refused:

```{r}
#| error: true
render_svg(data(gm_all) + point + x(gdp) + y(life, free = TRUE) + play(year))
```

The reason is in the message. Panels sit beside each other and a reader compares
them; a frame *replaces* the one before it, so an axis refitted every frame would
move under the data and the motion on screen would be the scale's rather than the
data's. Across panels the freedom is safe. Across frames it would misreport the
data.

Free scales are refused in two more places, both for the same kind of reason. On
a channel that is not a position there is no axis to free, only a legend, and
that one legend decodes the whole plot. And with no facet there are no panels:

```{r}
#| error: true
render_svg(data(gapminder_2007) + point + x(gdp) + y(life, free = TRUE))
```

## A layer without the facet column

A layer whose table does not have the facet column is drawn in full in every
panel. That is how one shared reference line is drawn behind small multiples:

```{r}
world_median <- data.frame(
  gdp  = c(min(gapminder_2007$gdp), max(gapminder_2007$gdp)),
  life = rep(median(gapminder_2007$life), 2)
)
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(world_median) + line + style(color = "gray", size = 1) |
  facet(continent)
```

The gray line, the world's median life expectancy, repeats in every panel,
so each continent is read against the same line. The engine reports this
as an assumption rather than doing it in silence: the table `world_median`
has no `continent` column, so its layer is drawn everywhere.

## The facet column must be a category

A facet column *names* the panels, so it must be a category column: a
number is a position along an axis, not a name for a frame. Faceting by a
numeric column refuses with direction:

```{r}
#| error: true
render_svg(data(gapminder_2007) + point + x(gdp) + y(life) | facet(population))
```

This is the judgment `shape` and `group` make about a mark's features, applied
to the plot's outer frame. In R, `factor()`, or any explicit cut into named
groups, is the way to say which panels you mean.

Trellis named that cut a **shingle**, and lattice still builds one with
`shingle()` or `equal.count()`. A shingle's intervals are allowed to overlap, so
one row can appear in two panels. That smooths the run of panels, and it counts
some rows twice. Cutting the column yourself with `cut()` gives up the
smoothing and puts the boundaries somewhere a reader can check them, in the strip
labels.

Faceting joins with the operators, never with `+`; a facet is a property of
the whole plot's frame, not one more encoding in the chain. Writing
`+ facet(continent)` refuses and points at `|` and `/`.
