# Faceting

```{r}
#| include: false
source("R/setup.R")
```

How do you compare the same plot across groups? A facet splits one plot into
small multiples: one panel per category of a column. Wilkinson calls a facet a
frame of frames. A **frame** is the space a plot is drawn in, with its own x and
y axes. The facet column's categories make the outer frame. Every cell of that
outer frame holds one copy of the inner frame, which is the plot itself. Nothing
else about the plot changes. The marks,
channels, and transforms you have already learned draw each panel exactly as
they would draw one plot with no facet.

The picture has been invented more than once, so you may know it under another
name. Becker and Cleveland's group at Bell Labs built it into S as **Trellis
display** [@becker1996trellis]. The name comes from the garden trellis, because
the panels are arranged in rows and columns like one. Most people first met the
display in S-PLUS. R draws
it that way still in the **lattice** package [@sarkar2026lattice], and its own
book is where the display is set out in full [@sarkar2008lattice]. Base R's
`coplot()` is the same idea under a third name, the conditioning plot, because
what a trellis conditions *on* is the facet column. That is also why the
operator in the next section is `|`, the sign that statistics already uses for
conditioning. [Composition](composition.qmd) finishes that history.

Tufte's **small multiples** [@tufte1990] names the result rather than the
operation, which is the division this book keeps: small multiples for the
picture, `facet` for the thing you write.

**Facet** is Wilkinson's own word [@wilkinson2005],
from the Latin *facies*, a face: one side of a many-sided thing, like one face
of a cut diamond. gog takes Wilkinson's word because it is the most general of
them. To him a trellis is one arrangement a facet can take, beside the
scatterplot matrix and the row plot. A grammar needs one term that covers all
three. The word every one of these systems shares is **panel**, the cell a split
produces, and this chapter uses it the same way.

## `|`: panels side by side

The scatter of income against life expectancy holds five continents in one
cloud. You want to see each on its own without losing the comparison between
them. The `facet()` atom names the column that does the splitting, and `|`
arranges the panels in columns:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, split into panel columns by
continent."*

Reading across, the panels share both axes, so a shift *between* continents
reads as plainly as a pattern within one. The shared scale is what faceting is
for. One plot per continent, each with its own scale, looks almost the same on
the page. It breaks the comparison, and nothing warns the reader.

## `/`: panels stacked

Are lives in Africa shorter than in Europe, and by how much? That is a question
about where each distribution sits along one axis. `/` arranges the panels in
rows. Stacking the panels aligns their x axes. That is the arrangement to choose
when the question is *where* the values sit along x:

```{r}
gm_ae <- gapminder_2007[gapminder_2007$continent %in% c("Africa", "Americas", "Europe"), ]
data(gm_ae) + bar * bin + x(life) / facet(continent)
```

*"Given the gm ae table: bars derived by bin, x is life, split into panel rows
by continent."*

Each panel's bars sit directly above the bars covering the same span of life
expectancy in the panel below. Africa's life expectancies spread wide toward
the low end, Europe's cluster in a narrow band at the high end, and the
Americas sit between the two.

That reading only works because the bars are the same width in all three
panels. A histogram under a facet needs two separate decisions: where the bins
are cut, and which rows are counted in them. Each panel counts its own rows,
because a statistic is computed on the panel's subset as if that subset were
the whole table. That is true of every transform on every mark, which is the [No
Exceptions](design-laws.qmd) law applied one panel at a time.

But `bin` is not only a statistic. It cuts the axis into cells *and* counts the
rows in each one. The cut describes the axis rather than measuring the rows, so
it belongs with the shared scale. One set of bins is cut from all three panels'
rows at once, and each panel then counts only its own rows into those bins.

It is worth seeing what the other choice would draw. Suppose each panel cut its
own bins. Europe's narrow spread would get narrow bins and Africa's wide spread
would get wide ones, so bars of different widths would be drawn against a
single axis. A bar four years wide holding nine countries and a bar one year wide
holding eight then draw at nearly the same height. The first holds about two
countries per year and the second holds eight, so the two rates differ by more
than three times. The panels are
still aligned and the axis still looks correct, and yet a reader cannot compare
the three distributions.

## Crossing into a grid

Every plot so far has split on one column. Two columns can matter at once. Here
the question is how each continent's countries moved between 1957 and 2007,
which is a split by continent and by period together. Write both operators to
cross two columns into a grid. Each operator applies to the facet written after
it, read left to right:

```{r}
gm_years <- gm_all[gm_all$year %in% c(1957, 2007), ]
gm_years$period <- as.character(gm_years$year)
data(gm_years) + point + x(gdp) + y(life) | facet(continent) / facet(period)
```

*"Given the gm years table: points, x is gdp, y is life, split into panel
columns by continent and panel rows by period."*

Fifty years and five continents in one grid: every panel column is a continent,
and every panel row a period. The shared scale lets you measure the rise in life
expectancy panel by panel.

The grid is a **crossing** in Wilkinson's sense: every combination of a panel
column and a panel row gets a panel, and a combination with no matching rows
still gets an empty one. The grid
shows that the combination is *possible*, so an empty panel is information
rather than an accident. Wilkinson's other operation, *nesting*, draws only the
combinations that exist. gog's facet operators always cross, so every
combination gets a panel. Neither is gog's `nest()` atom, which packs a whole
into areas.

## Folding a long line of panels

Back on one column, the opposite trouble appears. One column with many
categories makes a long line of panels, and a long line of panels makes each
one narrow. Here are the ten most populous countries of 2007.
Each panel shows one country's life expectancy from 1952 to 2007:

```{r}
ranked <- gapminder_2007[base::order(-gapminder_2007$population), ]
gm_top <- gm_all[gm_all$country %in% utils::head(ranked$country, 10), ]
data(gm_top) + line + group(country) + x(year) + y(life) | facet(country)
```

*"Given the gm top table: lines, x is year, y is life, grouped by country, split
into panel columns by country."*

The shapes are still readable. Nothing else on the panel can be read. Look at
the axis:
the tick labels touch and print as one run of digits, `196019802000`, and the
last panel's name is wider than the strip that prints it. Ten categories is a
common number, and ten panels in one line are already too narrow to read.

`wrap` folds that line into a rectangle. The number is how many panels to draw
before the line begins again:

```{r}
data(gm_top) + line + group(country) + x(year) + y(life) | facet(country, wrap = 4)
```

*"Given the gm top table: lines, x is year, y is life, grouped by country, split
into panel columns by country, wrapped at 4."*

The panels, the scale and the sentence are unchanged, and `wrap = 4` is the only
addition. The
years are readable, `United States` fits its strip, and each country's rise is
now large enough to read.

Notice what `wrap` does *not* say: which way the panels run. The operator
already decides it. `|` runs the line across, so four to a row; `/` runs it
down, so four to a column:

```{r}
data(gm_top) + line + group(country) + x(year) + y(life) / facet(country, wrap = 4)
```

The same number, the same ten countries, read down the columns instead of across
the rows. One number is enough because the direction is already in the sentence,
which is why `facet()` takes no second argument for it. ggplot2 spells the same
idea as `nrow` and `ncol`.

Two details in the wrapped plots come from the fold itself. The first is the
bottom right, where the last row stops after two panels. Ten panels folded four
to a row fill twelve cells, so two cells are left over and nothing is drawn in
them. They are not the empty panels of the previous section. In a crossing, an
empty panel stands for a combination the frame allows, and its emptiness reports
that the table holds no rows for it. Here there is no combination at all, only
leftover space.

The second is the extra axis beside the gap. Tick labels belong to the panels
that touch the edge where the axis is drawn. In a full rectangle that edge is
the bottom row. A folded line can leave the last row short, and then that rule
stops describing the plot. So gog uses a more general rule: a panel draws the x
axis when no panel sits below it. In the plot folded with `|`, `Mexico` and
`Nigeria` sit above the gap and keep their year labels. In the one folded with
`/`, the gap falls under `United States`.

Wrapping a crossed grid is refused. The crossing already makes a rectangle from
two columns, so a panel count beside it would answer the same question a second
time:

```{r}
#| error: true
data(gm_years) + point + x(gdp) + y(life) |
  facet(continent, wrap = 2) / facet(period)
```

A crossed grid already takes its shape from the two columns, so there is no
line left to fold.

## Giving a panel its own scale

Everything so far has used one scale across every panel, which is what makes the
panels comparable. It is also what makes some of them unreadable. Here is the
world's population by continent, one panel each:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + line * sum + x(year) + y(population) | facet(continent, wrap = 3)
```

*"Given all the gapminder years: a line derived by sum, x is year, y is
population, split into panel columns by continent, wrapped at 3."*

Asia is nearly four billion people and Oceania about twenty-five million, so the
axis is sized for Asia and Oceania is a flat line along the bottom. Oceania's
line is drawn correctly, but the axis is too tall for its shape to show. The
plot is correct. It shows that these five quantities are of very different
sizes. But the question was how each continent grew, and the plot answers a
different one.

Write `free = TRUE` inside a position, and that position's axis is fitted to
each panel's own rows:

```{r}
data(gm_all) + line * sum + x(year) +
  y(population, free = TRUE) | facet(continent, wrap = 3)
```

*"Given all the gapminder years: a line derived by sum, x is year, y is
population, freed, split into panel columns by continent, wrapped at 3."*

Now every panel has a shape, and Oceania has more than doubled.
Notice the y axis: each panel carries its own numbers, because one set of labels
along the edge is enough only when the scale is shared.

Notice also what the second plot loses. All five panels now look alike: five
lines rising left to right at about the same angle. Set the five panels side by
side, and nothing in them shows that one climb is measured in billions and
another in tens of millions. A
reader who looks at the shapes and not at the axes will draw the wrong
conclusion. The first plot cannot mislead a reader this way; the second one
can. So gog frees a scale only when you ask for it.

Which axis is freed is decided by where you write it. You write `free` inside
`x()` or inside `y()`, so the sentence has already named the axis and `free`
does not name it again. There is no `free_x` or `free_y`, and freeing both is
writing it twice:

```{r}
data(gm_all) + point + x(gdp, free = TRUE) + y(life, free = TRUE) |
  facet(continent, wrap = 3)
```

There is a third position, and it takes `free` the same way. A cube sits in a
panel like any other plot ([Space](space.qmd)):

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  z(population, free = TRUE) + color(continent) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, z is population, freed,
color by continent, split into panel columns by continent."*

If the height axis were shared, it would have to reach China, and Oceania's two
countries would sit at the very bottom, too close together to read. Freed,
Oceania is measured in tens of millions and has a shape again. The loss is the
one described above: the five cubes can no longer be compared with each other.
That plot needs no new word in the grammar. `free` is written on a position and
`z` is a position, so the third one accepts it like the other two.

Freeing any of the three positions has two further effects and one limit. gog
applies all three, and none is a setting you can write. The first is the bin
cut.
The histogram earlier in this chapter shares one set of bin edges across every
panel, because a cut describes the axis and the axis was shared. Free the axis
and the edges are cut per panel too, for exactly the same reason:

```{r}
data(gm_all) + bar * bin + x(gdp, free = TRUE) | facet(continent, wrap = 3)
```

Each panel is cut from its own rows, so Africa gets bins the size of Africa's
spread rather than bins sized for the richest countries in the data. You cannot
combine a free scale with a shared cut. Each panel would count only its own
rows, while the bin edges came from every panel's rows together, so the bars
would not fit the axis under them.

The second is the panel's **guides**: its axis numbers and gridlines. A scale
decides where the marks go and where the axis prints its numbers. Freeing the
scale has to move both, or the numbers stop describing the picture. Look again
at the plot where `y(population, free = TRUE)` was written. Its gridlines sit
at different heights in every panel, and each panel's labels sit beside its
own. When the scale is shared, those same lines are drawn at one set of heights
across all five. That is what a freed guide means: the lines a reader reads
values from belong to the scale that was freed.

The third is that a free scale belongs to a panel, never to an animation frame.
`play` ([Play](play.qmd)) splits rows the same way a facet does, so `free` on a
played plot looks like the same request. gog refuses it:

```{r}
#| error: true
data(gm_all) + point + x(gdp) + y(life, free = TRUE) + play(year)
```

The reason is in the message. Panels sit beside each other and a reader
compares them. A frame *replaces* the frame before it, so a refitted axis would
show different numbers in every frame. The reader would then see movement that
comes from the changing axis rather than from the data. Across panels a free
scale is safe. Across frames it would misreport the data.

Two more sentences are refused for the same kind of reason. Only `x`, `y` and
`z` take `free`, because a channel that is not a position has no axis to free,
only a legend, and that one legend decodes the whole plot. And with no facet
there are no panels:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life, free = TRUE)
```

Both messages name what is missing: an axis to free, or panels to free it in.

## A layer without the facet column

Five panels are easier to compare when every one of them carries the same
reference line, such as the world's median life expectancy. A layer whose table
does not have the facet column is drawn in full in every panel. That is how one
shared reference line is drawn behind small multiples:

```{r}
#| echo: false
peek(world_median)
```

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(world_median) + line + style(color = "gray", size = 1) |
  facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, and also a line from the
world median, split into panel columns by continent."*

The gray line, the world's median life expectancy, repeats in every panel,
so each continent is read against the same line. The engine reports this
as an assumption rather than doing it in silence: the table `world_median`
has no `continent` column, so its layer is drawn everywhere.

## The facet column's type

You may want one panel for small countries and one for large ones, which is a
split by `population`. A facet column *names* the panels, so it must be a
categorical column: a number is a position along an axis, not a name for a
frame. Faceting by a continuous column refuses with direction:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) | facet(population)
```

`shape` and `group` need a categorical column too, because each of them names
groups rather than measures them. A facet asks the same of the column that names
its panels. Turn the column into text, or cut it into named groups,
and the names you choose are the panels you get. In R that is `factor()` or
`cut()`.

Trellis named that cut a **shingle**, and lattice still builds one with
`shingle()` or `equal.count()`. A shingle's intervals are allowed to overlap, so
one row can appear in two panels. That makes the change from panel to panel
smoother, and it counts some rows twice. The cut you write yourself loses that
smoothing, and in exchange it prints its boundaries in the strip labels, where
a reader can check them.

Faceting joins with the operators, never with `+`; a facet is a property of
the whole plot's frame, not one more encoding in the chain. Write
`+ facet(continent)` and gog refuses, and the message points you at `|` and
`/`.

## In short

Every panel is drawn by the same sentence and read against the same scales. `|`
puts the panels in columns, `/` puts them in rows, and writing both crosses two
columns into a grid. `wrap` folds one long line into a rectangle, and `free`
releases the shared scale so each panel is fitted to its own rows. A layer
whose table does not carry the facet column is drawn in full in every panel.

## What comes next

Faceting cuts one plot into panels that share a sentence and both of their axes.
[Composition](composition.qmd) does the opposite: it arranges plots that were
never one plot on a single page.
