# Faceting

```{r}
#| include: false
source("R/setup.R")
```

How do you compare the same plot across groups? A **facet** splits one plot
into **small multiples**: one **panel** per category of a column. A **frame**
is the space a plot is drawn in, with its own x and y axes. Wilkinson calls a
facet a frame of frames. The facet column's categories make the outer frame,
and each of its cells is a panel. Each panel is an inner frame: the same plot,
drawn from one category's rows. The marks, channels, and transforms you have
already learned draw every panel, and by default all the panels share the same
scales.

The picture has been invented more than once, so you may know it under another
name. Becker and Cleveland's group at Bell Labs built it into S, the language R
was modeled on, as **Trellis display** [@becker1996trellis]. The name comes
from the garden trellis, because the panels are arranged in rows and columns
like one. R still draws it this way in the **lattice** package
[@sarkar2026lattice], and the package's own book explains the display in full
[@sarkar2008lattice]. Base R's `coplot()` is the same idea under a third name,
the conditioning plot, because what a trellis conditions *on* is the facet
column. That is also why the operator in the next section is `|`, the sign that
statistics already uses for conditioning.
[Composition](composition.qmd#where-these-operators-come-from) gives `|` a
second meaning, and says where both meanings come from.

Tufte's term, *small multiples* [@tufte1990], names the result rather than the
operation, and this book keeps the two apart: small multiples for the picture,
`facet` for the thing you write.

## `|`: panels side by side

The scatter of income against life expectancy holds five continents in one
cloud. You want to see each on its own without losing the comparison between
them. The `facet()` atom names the column that does the splitting, and `|`
arranges the panels in columns:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, split into panel columns by
continent."*

The panels share both axes, so a difference *between* continents is as easy to
see as a pattern within one. The shared scale is what faceting is for. Five
separate plots, each with its own scale, would look almost the same as these
five panels. Separate scales break the comparison, and nothing on the page
warns the reader.

A facet joins the plot with `|` or `/`, never with `+`, because the operator
decides where the panels go: `|` puts them in columns, `/` in rows. Write
`+ facet(continent)` and gog refuses, and the message names `|` and `/`.

## `/`: panels stacked

Are lives in Africa shorter than in Europe, and by how much? That is a question
about where each distribution sits along one axis. The plot also draws the
Americas as a third panel. `/` arranges the panels in rows. Stacking the panels
aligns their x axes. That is the arrangement to choose when the question is
*where* the values sit along x:

```{r}
gm_ae <- gapminder_2007[gapminder_2007$continent %in% c("Africa", "Americas", "Europe"), ]
data(gm_ae) + bar * bin + x(life) / facet(continent)
```

*"Given the gm ae table: bars derived by bin, x is life, split into panel rows
by continent."*

Each panel's bars sit directly above the bars covering the same span of life
expectancy in the panel below. Africa's life expectancies spread wide toward
the low end, Europe's cluster in a narrow band at the high end, and the
Americas' fall between the two, closer to Europe's.

That reading only works because the bars are the same width in all three
panels. A histogram under a facet needs two separate decisions: which rows are
counted, and where the bins are cut. Each panel counts its own rows, because a
statistic is computed on the panel's subset as if that subset were the whole
table. That is true of every transform on every mark, which is the [No
Exceptions](design-laws.qmd) law applied one panel at a time.

But `bin` is not only a statistic. It cuts the axis into bins *and* counts the
rows in each one. The cut is a set of positions on the x axis, so it belongs
with the shared scale. One set of bins is cut from all three panels' rows at
once, and each panel then counts only its own rows into those bins.

Suppose each panel cut its own bins. Europe's narrow spread would get narrow
bins and Africa's wide spread would get wide ones, so bars of different widths
would be drawn against a single axis. An African bar about five years wide and a
European bar under two years wide could each hold eight countries, and so be
drawn at the same height. But the African bar holds about one and a half
countries for each year of its width, and the European bar nearly five: more
than three times as many. The panels are still aligned and the axis still looks
correct, and yet a reader cannot compare the three distributions.

Like `bin`, [`density`](transforms.qmd#density-kernel-density-estimate) is not
only a statistic. Its bandwidth, which sets how smooth each curve is, belongs to
the whole plot, as the cut does: it is the average of the bandwidths each curve
would get from its own rows. Each panel estimates its curve from its own rows
with that bandwidth. A difference in shape between panels then comes from the
data and not from the smoothing.

## Crossing into a grid

Every plot so far has split on one column. Two columns can matter at once. Here
the question is how each continent's countries changed between 1957 and 2007,
which is a split by continent and by year together. The `gm_eras` table holds
those two years, with one row per country in each. It stores the year twice:
`year` is a number, and `era` is the same year as a label. A facet needs the
label, for the reason [the facet column's type](#the-facet-columns-type) gives.
Write both operators to cross the two columns into a grid: every category of one
is paired with every category of the other. Each operator applies to the facet
written after it, read left to right:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + point + x(gdp) + y(life) | facet(continent) / facet(era)
```

*"Given the gapminder eras: points, x is gdp, y is life, split into panel
columns by continent and panel rows by era."*

Five continents and two years, fifty years apart, in one grid: every panel
column is a continent, and every panel row a year. Because the scale is shared,
you can see each continent's rise in life expectancy by comparing its 1957
panel with the 2007 panel below it.

The grid is a **crossing** in Wilkinson's sense: every pair of a continent and a
year gets a panel, even a pair the table holds no rows for. Every pair in this
grid has rows, so no panel here is empty. When a pair has none, its panel is
still drawn, and the empty panel shows that the pair is *possible*: it is
information rather than an accident. Wilkinson's other operation, *nesting*,
draws only the pairs that exist. gog's facet operators always cross. Despite its
name, gog's `nest()` atom is not Wilkinson's nesting. It divides one panel into
areas sized by a measure ([Nest](nest.qmd)), and it makes no panels.

*Facet* is Wilkinson's word too [@wilkinson2005], from the Latin *facies*
(*face*): one side of a many-sided thing, like one face of a cut diamond. gog
takes his word because it is the most general of the names for this picture. To
him a trellis is one arrangement a facet can take among several, such as the
scatterplot matrix, a grid with one scatterplot for each pair of columns. A
grammar needs one term that covers them all.

## Folding a long line of panels

A single facet column can cause a problem of its own. One column with many
categories makes a long line of panels, and a long line of panels makes each
one narrow. Here are the ten most populous countries of 2007, taken from
`gm_all`, which holds every year of the gapminder data. Each panel shows one
country's life expectancy from 1952 to 2007:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
ranked <- gapminder_2007[base::order(-gapminder_2007$population), ]
gm_top <- gm_all[gm_all$country %in% utils::head(ranked$country, 10), ]
data(gm_top) + line + group(country) + x(year) + y(life) | facet(country)
```

*"Given the gm top table: lines, x is year, y is life, grouped by country, split
into panel columns by country."*

The lines are still readable, but the years under them are not: the year labels
overlap. The last panel's name is also wider than its strip, the band above the
panel that names it. Ten categories is a common number, and ten panels in one
line are already too narrow to read.

`wrap` folds that line into a rectangle. The number is how many panels to draw
before the line begins again:

```{r}
data(gm_top) + line + group(country) + x(year) + y(life) | facet(country, wrap = 4)
```

*"Given the gm top table: lines, x is year, y is life, grouped by country, split
into panel columns by country, wrapped at 4."*

The panels and the scale are unchanged, and `wrap = 4` is the only addition to
the sentence. The years are readable, `United States` fits its strip, and each
country's rise is now large enough to read.

Notice what `wrap` does *not* say: which way the panels run. The operator
already decides it. `|` runs the line across, so four to a row; `/` runs it
down, so four to a column:

```{r}
data(gm_top) + line + group(country) + x(year) + y(life) / facet(country, wrap = 4)
```

This is the same number and the same ten countries, now read down the columns
instead of across the rows. One number is enough because the direction is
already in the sentence, which is why `facet()` takes no second argument for it.
ggplot2 spells the same idea as `nrow` and `ncol`.

Two details in the wrapped plots come from the fold itself. The first is in the
bottom right corner, where two cells stay empty. Ten panels folded four at a
time fill twelve cells, so two cells are left over and nothing is drawn in them.
They are not the empty panels of a crossed grid. In a crossing, an empty panel
stands for a pair of categories that the table has no rows for. Here there is no
pair at all, only leftover space.

The second is the extra axis beside the gap. Tick labels belong to the panels
that touch the edge where the axis is drawn. In a full rectangle that edge is
the bottom row. A folded line can leave the last row short, and then the panels
above the gap would have no year labels. So gog uses a more general rule: a
panel draws the x axis when no panel sits below it. In the plot folded with
`|`, `Mexico` and `Nigeria` sit above the gap and keep their year labels. In the
plot folded with `/`, the gap falls under `United States`.

Wrapping a crossed grid is refused. The crossing already makes a rectangle from
two columns, so a `wrap` number would set the grid's shape a second time:

```{r}
#| error: true
data(gm_eras) + point + x(gdp) + y(life) |
  facet(continent, wrap = 2) / facet(era)
```

The message names the two columns that already make the rectangle, and asks you
to drop `wrap` or to facet by one column.

## Giving a panel its own scale

Everything so far has used one scale across every panel, which is what makes the
panels comparable. It is also what makes some of them unreadable. Here is the
world's population by continent, one panel each:

```{r}
data(gm_all) + line * sum + x(year) + y(population) | facet(continent, wrap = 3)
```

*"Given all the gapminder years: a line derived by sum, x is year, y is
population, split into panel columns by continent, wrapped at 3."*

Asia holds nearly four billion people and Oceania about twenty-five million. The
axis is sized for Asia, so Oceania's line lies flat along the bottom. The line
is drawn correctly, but the axis is too tall for its shape to show. What the
plot shows is that the five totals are of very different sizes. But the
question was how each continent grew, and the plot answers a different one.

Write `free = TRUE` inside a position, and that position's axis is fitted to
each panel's own rows:

```{r}
data(gm_all) + line * sum + x(year) +
  y(population, free = TRUE) | facet(continent, wrap = 3)
```

*"Given all the gapminder years: a line derived by sum, x is year, y is
population, freed, split into panel columns by continent, wrapped at 3."*

Now every line shows its rise, and you can see that Oceania's population more
than doubled. Notice the y axis: each panel carries its own numbers and its own
gridlines, because one set of labels along the edge is enough only when the
scale is shared.

Notice also what the second plot loses. Each axis is now fitted to its own
line, so every line rises from the bottom of its panel to the top, and the five
look much alike. Nothing in the five shapes shows that one rise is counted in
billions and another in tens of millions. A reader who looks at the shapes and
not at the axes will believe the five continents grew by similar amounts. The
first plot cannot mislead a reader this way; the second one can. So gog frees a
scale only when you ask for it.

Where you write `free` decides which axis is freed. You write `free` inside
`x()` or inside `y()`, so the sentence has already named the axis and `free`
does not name it again. There is no `free_x` or `free_y`. To free both axes,
write `free` in both:

```{r}
data(gm_all) + point + x(gdp, free = TRUE) + y(life, free = TRUE) |
  facet(continent, wrap = 3)
```

Each panel now has its own numbers on both axes: Asia's gdp axis reaches past
100K, and Oceania's starts near 10K.

There is a third position, `z`, and it takes `free` the same way, with no new
word in the grammar. A cube sits in a panel like any other plot
([Space](space.qmd)):

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  z(population, free = TRUE) + color(continent) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, z is population, freed,
color by continent, split into panel columns by continent."*

If the height axis were shared, it would have to reach China's population.
Oceania's two countries would then sit at the very bottom, almost on top of each
other. Freed, Oceania's height axis runs in tens of millions, and its two
countries sit far apart. The loss is the same one the freed population lines
had: the five cubes can no longer be compared with each other.

Freeing a position also changes the bin cut, and gog does this for you: it is
not a setting you can write. The histogram earlier in this chapter shares one
set of bin edges across every panel, because the cut is a set of positions on an
axis that was shared. Free the axis and the edges are cut per panel too, for
exactly the same reason:

```{r}
data(gm_all) + bar * bin + x(gdp, free = TRUE) | facet(continent, wrap = 3)
```

Each panel's bins are cut from its own rows, so Africa gets bins sized to
Africa's spread rather than bins sized for the richest countries in the data.
gog does not offer a free scale with a shared cut. The bin edges would span
every panel's range while each axis spans only its own, so some bars would
extend past the end of their axis.

Freeing an axis changes a `density` in the same way. On a freed axis, each
panel's curves are smoothed with a bandwidth chosen from that panel's rows alone.
A shared bandwidth is there so that panels can be compared, and freed panels
cannot be. Here it would only flatten the curve of a panel with a narrow spread.

Freeing also has limits. A free scale belongs to a panel, never to an animation
frame. `play`, which turns a plot into an animation ([Play](play.qmd)), splits
rows the same way a facet does, so `free` on a played plot looks like the same
request. gog refuses it:

```{r}
#| error: true
data(gm_all) + point + x(gdp) + y(life, free = TRUE) + play(year)
```

The reason is in the message. Panels sit beside each other and a reader
compares them. Each frame of an animation *replaces* the one before it, so a
refitted axis would show different numbers in every frame. The reader would then
see movement that comes from the changing axis rather than from the data. Across
panels a free scale is safe. Across the frames of an animation, it would
misreport the data.

Only `x`, `y` and `z` take `free`. A channel that is not a position has no axis
to free. It has a legend instead, and that one legend decodes the whole plot, so
`color()` has no `free` to write. One more sentence is refused: `free` with no
facet. A plot with no facet has one panel, and freeing its scale would change
nothing:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life, free = TRUE)
```

The message says the plot has one panel, and tells you to facet it or drop
`free`.

## A layer without the facet column

Return to the first plot in this chapter, with its five continents. Five panels
are easier to compare when every one of them carries the same reference line,
such as the world's median life expectancy. A layer whose table does not have
the facet column is drawn in full in every panel. That is how one shared
reference line is drawn across small multiples:

```{r}
#| echo: false
peek(world_median)
```

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  data(world_median) + line + style(color = "gray", size = 1) |
  facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, and also a line from the
world median, split into panel columns by continent."*

The gray line, the world's median life expectancy, repeats in every panel,
so each continent is read against the same line. The engine reports this
as an assumption rather than doing it in silence: the table `world_median`
has no `continent` column, so its layer is drawn everywhere.

## The facet column's type

You may want one panel for small countries and one for large ones, which is a
split by `population`. A facet column *names* the panels, so it must be a
categorical column: a number is a position along an axis, not a name for a
panel. gog refuses a facet on a continuous column, and the message says what to
do instead:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) | facet(population)
```

`shape` and `group` need a categorical column for the same reason: each of them
names groups rather than measures them. Cut the column into named groups, such
as small and large; in R that is `cut()`. Turning it into text, with
`factor()`, would give every country a panel of its own.

Trellis named that cut a **shingle**, and lattice still builds one with
`shingle()` or `equal.count()`. A shingle's intervals are allowed to overlap, so
one row can appear in two panels. That makes the change from panel to panel
smoother, and it counts some rows twice. The cut you write yourself loses that
smoothing, and it prints its boundaries in the strip labels, where a reader can
check them.

## In short

Every panel is drawn by the same sentence and, unless you free a scale, read
against the same scales. `|` puts the panels in columns, `/` puts them in rows,
and writing both crosses two columns of the table into a grid. `wrap` folds one
long line into a rectangle, and `free` gives each panel its own scale, fitted to
its own rows. A layer whose table does not carry the facet column is drawn in
full in every panel.

## What comes next

Faceting cuts one plot into panels that share a sentence and, by default, their
axes. [Composition](composition.qmd) does the opposite: it arranges plots that
were never one plot on a single page.
