# A first plot

```{r}
#| include: false
source("R/setup.R")
```

Do people in richer countries live longer?

That is a question about data: 142 countries, each with a GDP per capita and a
life expectancy.

```{r}
#| echo: false
peek(gapminder_2007)
```

In gog you answer a question like that by writing what the picture should
show, rather than by drawing anything:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

Read it aloud: *"Given gapminder 2007: points, x is gdp, y is life."*

The plot answers the opening question: yes, people in richer countries live
longer. It also shows a pattern: life expectancy climbs steeply across the
poorer countries and then flattens among the richest.

That one line of code is a complete gog sentence. You named a table, a mark,
and two positions; the engine chose the axes, the ticks, the margins, and drew
one point per row. You never call a drawing function, because a plot here is a
*specification*: a statement of what the picture should show. Rendering it is
the engine's job, not yours.

You do not define `data`, `point`, `x` and `y`. They are part of the package,
and loading gog makes them available. A word that takes no column and no value
is written without parentheses, like `point`.
Parentheses appear when there is something to put in them, like the column in
`x(gdp)`.

::: {.content-visible when-format="html"}

## The boxes under every plot

Before you learn the next word of the grammar, look at something on the page
that is not part of the grammar. A row of small boxes sits under the plot above, and under every
plot in this book. The two magnifiers zoom in and out. The frame returns
you to the whole picture. The camera saves what you are looking at as a picture
file, large enough for a slide or a journal figure. A box is dimmed when using it
would change nothing.

Once you have zoomed in, dragging the plot moves it. No box does this. When the
plot can be moved, the pointer changes to a hand.

Move the pointer over any box and a small label appears, saying what that box
does. A symbol means nothing to a reader who has not seen it before. That is
why the words are there.

Not one of these boxes has a word in the grammar, and that is deliberate. A
word belongs in the grammar only if it changes what the plot claims about the
data. Zooming in changes nothing: the points, the axes and the numbers stay
the ones the engine chose. To make the plot itself cover a narrower range,
write [`limits`](scales.qmd) in the sentence.
:::

::: {.content-visible when-format="pdf"}
The gog sentence at the top of this chapter has four spellings, and this
edition prints them under nearly every example: the R above, then Python, Julia
and JavaScript. None of the four depends on another. Each is a full way to write
the same specification, and one engine draws it. The plot is drawn once, and any
of the four would produce the same file. Read whichever is yours and skip the
others; the text around them is the same for all four.

They are in that order because the first three are spelled almost identically. The
operators are the same and only the way a column is named changes: a bare
`gdp` in R, `col.gdp` in Python, `:gdp` in Julia. JavaScript is the one that
reads differently, because it cannot give `+ * | /` new meanings and spells
them as words instead. It comes last, so the pattern the other three share is
seen first. Each language
also has a chapter of its own at the end, listing what differs:
[R](bindings/r.qmd), [Python](bindings/python.qmd), [Julia](bindings/julia.qmd),
[JavaScript](bindings/javascript.qmd).
:::

## Three steps to run any example

So far you have only looked at the first plot. Running its sentence yourself,
on your own computer, takes three steps:

1. **Install.** The command for your language is in
   [Installing gog](index.qmd#installing-gog).
2. **Load it.** `library(gog)` in R, `from gog import *` in Python,
   `using GrammarOfGraphics` in Julia, and an `import` from
   `grammar-of-graphics` in JavaScript.
3. **Read a table.** `gog_table()` comes with the package and downloads any
   of the book's tables over the internet. `gog_table("gapminder_2007")` reads
   the table shown above; store the result under the name `gapminder_2007`, and
   `data(gapminder_2007)` takes it. [Reading a table](book-data.qmd#reading-a-table)
   shows the line in all four languages.

Every example that reads one of the book's tables needs only those three
steps, and names its table inside `data()`. Where an example builds its own
table, the chapter shows the lines that build it.

## Add a word, ask a richer question

Which continent is which? Say so, one more word in the sentence:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)
```

*"…and color by continent."* The `color()` channel maps a column to a
visual property. A mapping raises a question (*which continent is this
dot?*), so the engine draws the legend that answers it. You never
asked for a legend. A mapped column earns one when there is something to
decode.

## An incomplete sentence is refused

What happens if you forget a position? An incomplete sentence is refused, just
as a syllable with no vowel cannot be spoken. The refusal names what is
missing:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp)
```

You meet a refusal this early because refusals and the small vocabulary of the
next chapter are the two things that make gog easy to learn. A **binding** is a
channel with a column in it, like `x(gdp)`. The engine never ignores a binding
without telling you, and every refusal also says what to do instead. In gog, the
error messages are part of the teaching; the [writing practice](writing.qmd)
chapter collects the ones you will meet first.

## Refine rather than rebuild

The `gdp` axis is crowded on the left: most countries have far lower incomes
than the richest few. A log scale gives every tenfold step,
such as 1,000 to 10,000, the same width on the axis. It changes how the axis is
spaced, not what is plotted, so it is written on the `x` binding. The only other
change is a title at the end:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."* After three changes, the sentence still reads left to right,
and the four words you learned first are all still there, in the same order. The
last word, `title()`, puts a title above the plot. It is a label: a word that
names the plot or an axis instead of binding a column. There are four labels,
and the [Grammar](grammar.qmd#labels-the-words-on-the-plot) chapter lists them.

## The shape of every sentence

The three plots in this chapter were each built by adding to the one before,
never by starting again. They share one shape, and so does every other plot in
this book, including the globes, networks and brushed plots at the end:

```
data(table) + mark + positions + refinements
```

The table comes first, named once; a database query can be used instead, as
[Data](data.qmd#a-table-in-a-database) shows. The mark is the shape drawn for
each row, and the positions are where it is drawn. Then come any number of
optional refinements: channels like `color()`, and labels like `title()`.

Putting a column into a channel has two names, and you will meet both. A
column is **bound** to a channel, and the binding is where a scale is written.
The same act is called **mapping** to stress that the channel received a
column, not a fixed value.
The difference between a column and a fixed value shows on the page: a mapped
column usually earns a **guide**, a legend or an axis that lets a reader decode
it. A fixed value you set yourself earns none. A table is bound too, by
`data()`. That is the Bind-Once law: name the table once, and no column needs
the table's name in front of it.

Two more parts join with operators of their own. A transform attaches to
its mark with [`*`](operators.qmd#derive), which is how the preface's
`bar * bin` made a histogram. A [facet](faceting.qmd) splits the finished plot
into panels with `|` or `/`. A scale is not
a separate part at all: it is written on the position it belongs to, the way
`x(gdp, scale = "log")` was written above. So the shape in full is:

```
data(table) + mark * transform + positions + refinements | facet(column)
```

The transform and the facet are optional, and so are the refinements. Delete
every one of them and a legal plot remains. What cannot be removed is the table,
the mark, and the positions that mark needs. That is the sentence you wrote at the
top of this chapter.

English grammar works the same way, which is why gog is called a grammar.
"Birds sing" is already a complete sentence in English. "The small birds
sing loudly in the morning" says much more, and every word added to it can be
taken out again. Remove the verb instead and there is no sentence left to
refine. A plot is built like an English sentence. Some of its parts are
required, and the rest refine it. That division is the same in every plot in
this book, so you learn it once.

## What comes next

Your first plot needed four words, and this chapter has not said how many
words there are in total. [The grammar](grammar.qmd) states gog's whole
vocabulary in one table, then gives each kind of word its own section.
