# A first plot

```{r}
#| include: false
source("R/setup.R")
```

Do people in richer countries live longer?

That is a question about data: 142 countries, each with a GDP per capita and a
life expectancy.

```{r}
#| echo: false
peek(gapminder_2007)
```

In gog you answer a question like that by writing what the picture should
show, rather than by drawing anything:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

Read it aloud: *"Given gapminder 2007: points, x is gdp, y is life."*

That sentence is the entire program. You named a table, a mark, and two
positions; the engine chose the axes, the ticks, the margins, and drew one
point per row. You never call a drawing function, because a plot here is a
*specification*: a sentence about the picture, not a claim about the data.
Rendering it is the engine's job, not yours.

None of those four words is yours to define. They arrive with the package, so
loading gog is what makes `point` and `x` available. A word is written bare
when it has nothing to take, which is why `point` has no parentheses.
Parentheses appear when there is something to put in them, like the column in
`x(gdp)`.

::: {.content-visible when-format="html"}

## The boxes under every plot

Look under the plot above. A row of small boxes sits there, and every plot in
this book has them. The two magnifiers zoom in and out. The frame returns
you to the whole picture. The camera saves what you are looking at as a picture
file, large enough for a slide or a journal figure. A box is dimmed when it can
do nothing.

Once you have zoomed in, dragging the plot moves it. There is no box for it;
the pointer shows it instead, becoming a hand over a plot you can move.

Move the pointer over any box and a small label appears, saying what that box
does. A symbol means nothing to a reader who has not seen it before. That is
why the words are there.

Not one of these boxes has a word in the grammar, and that is deliberate. A
word belongs in the grammar only if it changes what the plot claims about the
data. Zooming in changes nothing: the points, the axes and the numbers stay
the ones the engine chose. To make the plot itself cover a narrower range,
write [`limits`](scales.qmd) in the sentence.
:::

::: {.content-visible when-format="pdf"}
That sentence has four spellings, and this edition prints all of them under every
example: the R above, then Python, Julia and JavaScript. They are not translations
of each other. Each is its own language writing the same specification, which
one engine then draws, so the plot below is rendered once, and any of the four
would produce the same file. Read whichever is yours and ignore the rest;
nothing else in this book is language-specific.

They are in that order because the first three are very nearly one sentence. The
operators are the same and only the way a column is named changes: a bare
`gdp` in R, `col.gdp` in Python, `:gdp` in Julia. JavaScript is the one that
reads differently, because it cannot give `+ * | /` new meanings and spells
them as words instead. It comes last, so the pattern the other three share is
seen first. Each language
also has a chapter of its own at the end, a difference list about a page long:
[R](bindings/r.qmd), [Python](bindings/python.qmd), [Julia](bindings/julia.qmd),
[JavaScript](bindings/javascript.qmd).
:::

## Three steps to run any example

Everything in this book runs on your own computer, and it takes three steps:

1. **Install.** The command for your language is in
   [Installing gog](index.qmd#installing-gog).
2. **Load it.** `library(gog)` in R, `from gog import *` in Python,
   `using GrammarOfGraphics` in Julia, and an `import` from
   `grammar-of-graphics` in JavaScript.
3. **Read a table.** `gog_table("gapminder_2007")` fetches the one above;
   assign it that name, and `data()` takes it. It comes with the package, and
   [Reading a table](book-data.qmd#reading-a-table) shows the line in all
   four languages.

Every example after this one works from those three steps, and the table each
needs is named in the sentence itself.

## Add a word, ask a richer question

Which continent is which? Say so, one more word in the sentence:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)
```

*"…and color by continent."* The `color()` channel maps a column to a
visual property. A mapping raises a question (*which continent is this
dot?*), so the engine draws the legend that answers it. You never
asked for a legend. A mapped column earns one when there is something to
decode.

## An incomplete sentence is refused

What happens if you forget a position? An incomplete sentence is refused the
way a syllable missing its vowel cannot be pronounced, and the refusal tells
you what is missing:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp)
```

You meet this on your first page because refusals are half of what makes gog
learnable; the small vocabulary in the next chapter is the other half. A
**binding** is a channel with a column in it, like `x(gdp)`.
The engine never accepts a binding only to silently drop it,
and every *no* comes with a direction. The errors here do the teaching; the
[writing practice](writing.qmd) chapter collects the ones you will meet
first.

## Refine, don't rebuild

The `gdp` axis is crowded on the left: most countries have far lower incomes
than the richest few. A logarithmic view is a property of *how x is
read*, so it is written on the binding, and nothing else changes:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color
by continent, titled."* Three refinements later, the sentence still reads
left to right, and the four words you learned first are unchanged.

## The shape of every sentence

Every plot in this book, through the 3-D scenes at the end, is this same
shape:

```
data(table) + mark + positions + refinements
```

A table bound once, so columns are bare names. A mark, the visible form.
Positions, where the marks stand. Then any number of optional refinements:
channels like `color()`, and labels like `title()`.

Two words describe this, and the book uses both. A column is **bound** to a
channel, and the binding is where a scale gets written. The same act is called
**mapping** to emphasize that a channel received a column rather than a fixed
value.
The difference shows on the page: a mapped column earns a **guide** to decode
it, and a value you merely set earns none. A table is bound too, by `data()`,
which is the Bind-Once law: name it once, and its columns stay bare names.

Two more parts join with an operator of their own. A transform attaches to
its mark with `*`, which is how the preface's `bar * bin` made a histogram.
A facet splits the finished plot into panels with `|` or `/`. A scale is not
a separate part at all: it is written on the position it belongs to, the way
`x(gdp, scale = "log")` was written above. So the shape in full is:

```
data(table) + mark * transform + positions + refinements | facet(column)
```

The transform and the facet are optional, and so are the refinements. Delete
every one of them and a legal plot remains. What cannot go is the table, the
mark, and the positions that mark needs. That is the sentence you wrote at the
top of this chapter.

Written grammar works the same way, which is why this one is called a grammar at
all. "Birds sing" is already a complete sentence in English. "The small birds
sing loudly in the morning" says much more, and every word added to it can be
taken out again. Remove the verb instead and there is no sentence left to
refine. A plot is not a different kind of object. Some of its parts are
required, and the rest refine it. That division is the same in every plot in
this book, so you learn it one time.

The next chapter states the kernel on one page, and the page is short.
