# A first plot

```{r}
#| include: false
source("R/setup.R")
```

Do countries that earn more live longer?

That is a question about data (142 countries, each with a GDP per capita
and a life expectancy), and in GOG you answer it by saying what the data
means, not by drawing anything:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life)
```

Read it aloud: *"Given gapminder 2007: points, x is gdp, y is life."*

That sentence is the entire program. You named a table, a mark, and two
positions; the engine chose the axes, the ticks, the margins, and drew one
point per row. There is no draw call to make, because a plot here is a
*specification* (a sentence about the picture, not a claim about the data), and
rendering it is the engine's job, not yours.

::: {.content-visible when-format="pdf"}
That sentence has four spellings, and this edition prints all of them under every
example: the R above, then Python, Julia and JavaScript. They are not translations
of each other. Each is its own binding writing the same specification, which one
engine then draws, so the plot below is rendered once and would be
byte-identical from any of the four. Read whichever is yours and ignore the rest;
nothing else in this book is language-specific.

They are in that order because the first three are very nearly one sentence. The
operators are the same and only the way a column is named moves: a bare `gdp` in R,
`col.gdp` in Python, `:gdp` in Julia. JavaScript is the one that reads differently,
because it cannot give `+ * | /` new meanings and spells them as words instead, so
it comes last, after the pattern the other three share is established. Each binding
also has a chapter of its own at the end, a difference list about a page long:
[R](bindings/r.qmd), [Python](bindings/python.qmd), [Julia](bindings/julia.qmd),
[JavaScript](bindings/javascript.qmd).
:::

## Add a word, ask a richer question

Which continent is which? Say so, one more word in the sentence:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)
```

*"…and color by continent."* The `color()` channel maps a column to a
visual property, and because a mapping poses a question (*which continent
is this dot?*), the engine draws the legend that answers it. You never
asked for a legend. A mapped column earns one; that is the rule.

## An incomplete sentence is refused

What happens if you forget a position? An incomplete sentence is refused the
way a half-syllable is unpronounceable, and the refusal tells you what is
missing:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp)
```

This is worth meeting on your first page, because it is half of what makes
GOG learnable. A **binding** is a channel with a column in it, like `x(gdp)`.
The engine never accepts one and silently drops it,
and every *no* comes with a direction. The errors are teachers here; the
[writing practice](writing.qmd) chapter collects the ones you will meet
first.

## Refine, don't rebuild

The income axis is crowded on the left: most of the world earns far less
than the richest countries. A logarithmic view is a property of *how x is
read*, so it is written on the binding, and nothing else changes:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color
by continent, titled."* Four refinements in, the sentence still reads left
to right, and the first three words you learned are still doing the work.

## The shape of every sentence

Every plot in this book, through the 3-D scenes at the end, is this same
shape:

```
data(table) + mark + positions + refinements
```

A table bound once, so columns are bare names. A mark, the visible form.
Positions, where the marks stand. Then any number of optional refinements:
channels like `color()`, and labels like `title()`.

Two words describe this, and the book uses both. A column is **bound** to a
channel, and the result is a **binding**: `x(gdp)`, a channel with its column
in it. That is where a scale gets written. The same act is called **mapping**
when the point is that a channel received a column rather than a fixed value.
The difference shows on the page: a mapped column earns a legend, and a value
you merely set earns none. A table is bound too, by `data()`, which is the
Bind-Once law: name it once, and its columns stay bare names.

Two more parts join with an operator of their own. A transform attaches to
its mark with `*`, which is how the preface's `bar * bin` made a histogram.
A facet splits the finished plot into panels with `|` or `/`. A scale is not
a separate part at all: it is written on the position it belongs to, the way
`x(gdp, scale = "log")` was written above. So the shape in full is:

```
data(table) + mark * transform + positions + refinements | facet(column)
```

The transform and the facet are optional; everything before them is the
sentence you have already written twice. The next chapter lays out the whole
vocabulary on one page, and it is shorter than you expect.
