# Flow

```{r}
#| include: false
source("R/setup.R")
```

Where did each part of a total go? Some tables record movement: each row says
that some amount started in one place and finished in another. A flow diagram
draws that movement. The places are grouped into stages, and the stages are
drawn side by side. Bands carry each amount from stage to stage. The thicker the
band, the larger the amount it carries. This picture is an **alluvial diagram**,
one kind of **Sankey diagram**. [Where the limits are](#where-the-limits-are),
near the end of this chapter, says which Sankey diagrams gog does not draw. No
atom spells either name, just as there is no `pie()` atom.

gog builds that picture out of the table's own columns. A **stage** is one of
the table's categorical columns. Its distinct values are the possible places at
that stage, and each row sits in one of them. The `flow` transform computes the
whole diagram from one table of such columns. This chapter uses the passengers
and crew of the Titanic, the ship that sank in 1912. `titanic` holds one row per
group of people. Four
columns describe each group: `class`, `sex`, `age` (child or adult) and
`survived`. A fifth, `n`, counts the people in that group. The table has one row
for every combination of those four columns, so some rows count zero people:

```{r}
#| echo: false
peek(titanic)
```

## The finished diagram

Who survived the sinking, and did class and sex change their chance of
survival? The diagram below answers both:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("Who survived, by class and sex")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, and also zones derived by flow through class, sex
and survived, and text derived by flow through class, sex and survived, label by
name."*

Read it from left to right. Each white rectangle marks a **slot**: the part of a
stage that one place takes, such as a class, a sex or an outcome. Each band is a group of people who
share a class, a sex and an outcome, and its color is the outcome: blue for the
people who survived, orange for the people who did not. The blue bands that
leave the `Female` slot are much thicker than the orange ones, because 344 of
the 470 women and girls survived. At the `Male` slot the orange bands are
thicker, because only 367 of the 1,731 men and boys survived. Class mattered
too. 203 of the 325 people in first class survived, compared with 178 of the 706 in
third class.

## Three layers

The sentence above has three layers. In each one, `*` joins the same
transform, `flow(...)`, to a different mark, and each mark draws one part of the
picture:

| Layer | What it draws |
|---|---|
| `ribbon * flow(...)` | the bands, which carry the counts from stage to stage |
| `zone * flow(...)` | the slots, one rectangle for each place at each stage |
| `text * flow(...) + label(name)` | the names, one inside each slot |

The three layers compute the same positions from the same stages. So the bands
end at the edges of the slots, and each name sits inside its slot. The next
three sections add the layers one at a time, in the order of the table. They use
two stages, `class` and `survived`, and [Adding a stage](#adding-a-stage) then
adds `sex` between them.

## The bands

To build the diagram, start with the bands alone. Class is one stage and the
outcome is another. Name the stages inside `flow()`, from left to right, and
bind the count to `y`:

```{r}
data(titanic) + ribbon * flow(class, survived) + y(n) +
  title("Who survived, by class")
```

*"Given titanic: ribbons derived by flow through class and survived, y is n."*

Each band is one **path**: one place at each stage, here a class and then an
outcome. This plot does not outline the slots or name them, so find a path by
its position. Inside each stage the slots stack from the bottom, in the order
the column declares: `1st` to `Crew`, and `Yes` before `No`. So the lowest band leaves the
`1st` slot and arrives at the `Yes` slot. It carries the 203 first-class
passengers who survived, and it keeps that thickness at both ends. Rows that
share a path become one band: their counts are added, and the band carries the
total. The four `1st` rows that end in `Yes` differ only in `sex` and `age`, and
the atom names neither, so the four become one band.

`flow()` also sets both axes. With the count on `y`, the stages are drawn along
the horizontal axis, each named by its column, in the order you name them inside
`flow()`. The vertical axis is a continuous axis for the count. The bands at
each stage stack from zero with no gaps. So the top edge of any band is the
total of that band and every band below it.

The count is optional. Bind nothing and every row counts as 1, and the
diagram counts rows instead of summing a column. `partition` uses that default
too, and `count` tallies rows the same way. This table needs `y(n)`: it holds
one row per combination, so counting its rows would count combinations rather
than people.

## The slots

In the plot above, the bands that leave one place touch the bands that leave
the next, so you cannot see where one place ends and the next begins. A `zone`
layer draws the slots. Add `zone * flow(class, survived)`, with the same stages
as the bands. Both layers now use the count, so write `y(n)` before any mark,
where it belongs to the plot and every layer reads it ([Encoding
scope](encoding-scope.qmd#the-rule)):

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, survived) +
  zone * flow(class, survived) +
  title("The same flow, with its slots")
```

*"Given titanic: y is n, ribbons derived by flow through class and survived, and
also zones derived by flow through class and survived."*

Each rectangle marks one slot. It covers the same part of the count axis as the
bands that leave it or arrive at it, so each slot is as tall as the number of
people at that place. A [categorical
axis](scales.qmd#the-columns-type-picks-the-axis) also gives each category a
slot, but there every slot is the same size. The top slot of the class stage is
the crew, the largest group, with 885 people. At the outcome stage the larger
slot is `No`: 1,490 of the 2,201 people did not survive.

Each slot is white with a thin black line around it, so only the bands have
color and the edge of every slot is visible. Those are the defaults, and
[Choosing the colors](#choosing-the-colors) shows how to change them. The slots
still have no names, so you still count from the bottom to find each slot.

## The names

A `text` layer names the slots. The transform writes each slot's name into a
column called `name`, exactly as `partition` does. Add a third layer,
`text * flow(class, survived)`, with the same stages again, and bind
`label(name)` to draw that column:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, survived) +
  zone * flow(class, survived) +
  text * flow(class, survived) + label(name) +
  title("The same flow, with its slots named")
```

*"Given titanic: y is n, ribbons derived by flow through class and survived, and
also zones derived by flow through class and survived, and text derived by flow
through class and survived, label by name."*

Each name is drawn at the center of its slot. The layers are drawn in the order
they are written, so the `text` layer comes last. If the `text` layer came
before the `zone` layer, the white slots would hide the names.

The `1st` slot is now easy to find, and the thicker of its two bands arrives
at `Yes`. It carries the 203 of the 325 people in first class who survived, the
number given under the finished diagram. Every later plot in this chapter keeps
the three layers.

## Adding a stage

The finished diagram has a third stage, `sex`, between class and the outcome.
Add it by naming one more column inside `flow()`, in all three layers. Each path
now passes through three stages, and each band is as thick as its count at
every one:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("Class, then sex, then survival")
```

Each band is now a group of people with one class, one sex and one outcome.
`sex` declares no order, so its slots follow the order its values first appear
in the table: `Male` at the bottom, then `Female`. The band from `1st` through
`Female` to `Yes` carries 141 people, the women and girls in first class who
survived. Most people in the `Female` slot arrive at `Yes`, and most in the
`Male` slot arrive at `No`, as in the finished diagram.

## Color from a stage

With three stages the bands cross one another, and when every band has the same
color, one band is hard to follow through a crossing. A band runs through every
stage, so any stage can color it. The finished diagram at the start of this
chapter colors each band by `survived`, the slot it arrives at. Coloring by
`class` gives each band the color of the slot it starts from instead:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(class) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("The same paths, colored by class")
```

Now you can see which classes arrive in each later slot. Coloring by `survived`
and coloring by `class` answer two questions: who lived, and where each class
went.

`color()` is written after the `ribbon` layer, so it belongs to the bands alone.
Written before the marks, it would reach the slots and the names as well. gog
refuses that, because a slot and its name take their color from `style()`.

## Color from another column

Did the women in third class survive as often as the women in first class? A
color can show it inside each band, with no third stage. Any categorical column
can color the bands, not only a stage. A column that is not a stage splits each
band into one part for each of its values. A categorical color splits a line the
same way, into one line for each category. The parts of one band lie next to
each other across its thickness and follow its path, and the slots are the same
as without the color. With `survived` as a color instead of a stage, a flow
through class and sex shows who survived in every band:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex) + color(survived) +
  zone * flow(class, sex) +
  text * flow(class, sex) + label(name) +
  title("Class, then sex, colored by survival")
```

*"Given titanic: y is n, ribbons derived by flow through class and sex, colored
by survived, and also zones derived by flow through class and sex, and text
derived by flow through class and sex, label by name."*

Each band from a class to a sex now has a blue part for the people who survived
and an orange part for the people who did not. The band from `1st` to `Female`
is nearly all blue: 141 of its 145 people survived. The band from `3rd` to
`Female` is less than half blue: 90 of 196. The finished diagram shows the same
split with `survived` as a stage, where the outcome also gets slots of its own.
As a color, the outcome needs no slots of its own, and the legend names its
values.

Only a categorical column can split a band, because its values are groups. A
continuous column is refused, as it is on every `ribbon`.

## Choosing the colors

Every color in this chapter so far is a default. The bands take gog's default
palette and are translucent, the slots are white with a thin black line around
them, and the names are near-black. You choose each one in its own layer.

The bands take their colors from a column, and `palette()` chooses the color for
each value of that column. Write each value beside the color you want, as
[Colors by name](channels.qmd#colors-by-name) explains. Name every value, since
gog refuses a palette that leaves one out:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  palette(c(Yes = "steelblue", No = "darkgray")) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("The survivors in blue")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, with a palette naming Yes steelblue and No
darkgray, and also zones derived by flow through class, sex and survived, and
text derived by flow through class, sex and survived, label by name."*

The bands of the survivors are blue and the others are gray, so the survivors
are easy to find. A band is translucent, so it looks lighter than the color you
name, and two bands that cross both stay visible. To give every band the same
color, leave out `color()` and write `style(color = )` after the `ribbon` layer.
`style(opacity = )` there sets how translucent the bands are, with or without
`color()`.

A slot sits at one stage, so no single column can color every slot, and the
same is true of the names. The `zone` and `text` layers take their settings from
`style()` instead, one value for each layer. On the `zone` layer,
`color` fills the slots, and `border_color` and `border_size` draw the line
around each one. On the `text` layer, `color` and `size` set the color and the
size of the names:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  palette(c(Yes = "steelblue", No = "darkgray")) +
  zone * flow(class, sex, survived) +
  style(color = "midnightblue", border_color = "white", border_size = 2) +
  text * flow(class, sex, survived) + label(name) +
  style(color = "white", size = 14) +
  title("Dark slots with white names")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, with a palette naming Yes steelblue and No
darkgray, and also zones derived by flow through class, sex and survived,
colored midnightblue, with border color white and border size 2, and text
derived by flow through class, sex and survived, label by name, colored white,
with size 14."*

Each `style()` is written after the mark it changes, so it changes that layer
and no other. On the dark fill, the white line shows where each slot ends, and
`border_size = 0` would remove it. The `zone` layer also takes `opacity`, as
every mark does, and `pattern`, as the five filled regions do
([Style](style.qmd#what-you-can-set)).

## Room for long names

In a flow that runs left to right the slots are tall and narrow. The names in
`titanic` are short, so every one fits. A longer name, such as
`Member of the crew`, would be wider than its slot, and gog would draw it across
the slot's edges, over the bands. There are two ways to give it more space.

The first turns the names. `style(angle = 90)` on the `text` layer turns every
name upward, so it runs along its slot, as [Turning a
label](marks/text.qmd#turning-a-label) explains:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) + style(angle = 90) +
  title("Names turned 90 degrees")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, and also zones derived by flow through class, sex
and survived, and text derived by flow through class, sex and survived, label by
name, with angle 90."*

Each name is now written upward, so it must fit the slot's height, not the
slot's width.

The second places the stages from top to bottom. Bind the count to `x` instead
of `y`, and the first stage is at the top of the plot. Each slot is then wide
and short, so a long name fits across it unless the slot's count is small. The
sentence is the finished diagram's, with `x(n)` in place of `y(n)` and a new
title. gog takes the direction from the axis that holds the count, as it does
for a bar chart ([Horizontal bars](marks/bar.qmd#horizontal-bars)):

```{r}
data(titanic) + x(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("The same flow, from top to bottom")
```

*"Given titanic: x is n, ribbons derived by flow through class, sex and
survived, colored by survived, and also zones derived by flow through class, sex
and survived, and text derived by flow through class, sex and survived, label by
name."*

The stage names are now on the left: `class` at the top, then `sex`, then
`survived`. The count is on the bottom axis, and each name fits inside its slot.

## Leaving out the count axis

The count axis helped while the diagram was being built: it showed that the
bands at each stage add up to the same total, just over 2,000 people. A reader of the finished
diagram compares the thicknesses of the bands and the slots, and needs no
numbers for that. So the count axis, the gridlines and the frame (the two lines
along the bottom and the left of the panel) can be left out, and the gray panel
can be white. Write `axis = FALSE` on the count. In
`theme()`, `grid = "none"` and `frame = "none"` leave out the gridlines and the
frame, and `background = "white"` colors the panel ([Axes, legends, gridlines
and the frame](style.qmd#axes-legends-gridlines-and-the-frame)). The stage names
stay, because they belong to the other axis, which is still drawn:

```{r}
data(titanic) + y(n, axis = FALSE) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  theme(grid = "none", frame = "none", background = "white") +
  title("Who survived, by class and sex")
```

*"Given titanic: y is n without an axis, ribbons derived by flow through class,
sex and survived, colored by survived, and also zones derived by flow through
class, sex and survived, and text derived by flow through class, sex and
survived, label by name."*

Only the bands, the slots with their names, the legend and the stage names are
left. A reader can still see that most of the women survived and most of the
men did not. The exact counts are gone from the picture, so give them in the
paragraph beside the plot, as [The finished diagram](#the-finished-diagram)
does.

## What flow refuses

A flow needs at least two stages, because movement needs a place to start and
a place to finish:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class)
```

`flow()` puts the stages on the axis that does not hold the count. With `y(n)`
that is the x axis, so `x(class)` asks for the same axis as the stages, and gog
refuses it:

```{r}
#| error: true
data(titanic) + y(n) + x(class) + ribbon * flow(class, survived)
```

Only `ribbon`, `zone` and `text` can read what `flow` computes. Any other mark
is refused, and the message names the three:

```{r}
#| error: true
data(titanic) + y(n) + bar * flow(class, survived)
```

And a continuous column cannot color a band, just as it cannot color any
`ribbon`. `n` is a continuous column:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + color(n)
```

## Where the limits are

The two limits below are not mistakes in your sentence. The engine does not
draw the first, and it does not read the second.

Every mark that draws on the page also bends into a circle, so you would expect
a flow's bands to bend as well. [`polar()`](polar.qmd) is what bends a plot,
and a flow in `polar()` is valid grammar. The bands bent around a circle make a
**chord diagram**, which this engine does not draw:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + polar()
```

The input has a limit too. A flow reads its stages as columns of one table, so
each row has one place at every stage. Some tools accept a different shape. They
take a list of nodes, which are the slots, and a list of links between those
nodes. That shape can also describe flows that begin partway, skip a stage, or
return to a stage they already visited. A flow does not read it. In the
stage-column shape, every path passes through every stage, so every stage has
the same total. No path can return to a stage it already visited.

## What comes next

A flow draws its stages in the order you name them, and it never computes an
order itself. [Cluster](cluster.qmd) computes one from the data: it joins the
most alike categories into a tree, and puts an axis in the tree's order.
