# Flow

```{r}
#| include: false
source("R/setup.R")
```

Where did each part go? Some tables record movement: each row says that some
amount started in one place and finished in another. A flow diagram draws that
movement. The places stand side by side as stages, and bands carry each amount
from stage to stage. The wider the band, the larger the amount it carries. The
world calls this picture a Sankey diagram, or an alluvial diagram, and as with
the pie there is no atom that spells either name.

A **stage** is one of the table's categorical columns, and its distinct values
are the places a row can be at that point. The `flow` transform computes the
whole diagram from one table of such columns. This chapter builds the diagram
one layer at a time, using the passengers and crew of the Titanic. `titanic`
holds one row per group of people: the class they traveled in, their sex,
whether they were a child or an adult, whether they survived, and how many
people were in the group. The table carries every combination, so some of its
groups are empty:

```{r}
#| echo: false
peek(titanic)
```

## One row is one path

Name the stages in the atom, left to right, and bind the count to `y`:

```{r}
data(titanic) + ribbon * flow(class, survived) + y(n) +
  title("Who survived, by class")
```

*"Given titanic: ribbons derived by flow through class and survived, y is n."*

Each band is one path through the stages. The band that leaves `1st` and
arrives at `Yes` carries every first-class passenger who survived, and its
thickness is that count at both ends. Rows that share a path are added
together, so the four `1st` rows for the children, women and men who survived
become one band. That same addition merges the columns the atom does not name,
`sex` and `age` here.

The stages stand on the horizontal axis, named by their columns. `x()` has
nothing left to set: the atom's argument order is the order the stages are
drawn in. The vertical axis is the count, and it keeps its ticks. The bands at
each stage stack from zero with no gaps, so a height measured against the axis
is a true cumulative count.

## Adding a stage

Add a stage by adding a column. Each path now crosses three stages, and each
band still carries one path's count at both of its ends:

```{r}
data(titanic) + ribbon * flow(class, sex, survived) + y(n) +
  title("Class, then sex, then survival")
```

The picture answers a question the two-stage version could not. Follow any
band and you follow one group of people. You see their class, whether they
were men or women, and whether they survived.

## Color comes from a stage

A band runs through every stage, so any stage can color it. Write `color()`
after the `ribbon` layer and name the stage that should choose the color:

```{r}
data(titanic) + ribbon * flow(class, sex, survived) + y(n) + color(class) +
  title("The same paths, colored by class")
```

Coloring by `class` keeps each band the color of where it began, so the
mixing at `sex` and at `survived` becomes visible. Coloring by `survived`
would instead color every band by how its path ends. Only a stage can color
a band, because a band is a whole path, and only a stage holds a value every
path carries.

## The slots and their names

Two more layers finish the diagram. A `zone` layer draws each stage's slots,
the stacked blocks the bands leave and arrive at. A `text` layer names them.
The three layers read one layout, so they always agree about where everything
sits:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(class) +
  zone * flow(class, sex, survived) + style(color = "gray") +
  text * flow(class, sex, survived) + label(name) +
  title("The full diagram: bands, slots, names")
```

The slots take their color from `style()`, gray here so that only the bands
carry color. The names come from `label(name)`: the layout writes each slot's
name into a column called `name`, exactly as `partition` does.

The count on `y` is optional. Bind nothing and every row counts as 1, so the
diagram counts rows instead of summing a column. That is the same default
`partition` uses, and the same one `count` is built on.

## What flow refuses

A flow needs at least two stages, because movement needs a place to start and
a place to finish:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class)
```

The stage axis is drawn from the atom's own columns, so a bound `x` has
nothing left to set:

```{r}
#| error: true
data(titanic) + x(n) + ribbon * flow(class, survived)
```

Only three marks read the layout, and the refusal lists all three when
another mark is used:

```{r}
#| error: true
data(titanic) + y(n) + bar * flow(class, survived)
```

And a band's color must name a stage. A column outside the stages holds no
value a whole path shares:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + color(n)
```

## Where the limits are

Bending a flow into `polar()` is valid grammar. The bands bent around a
circle are the chord diagram, which this engine does not draw:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + polar()
```

The input has a limit too. A flow reads stage columns from one table, one row
per path. Some tools accept a different shape: a list of nodes, which are
these slots, and a list of links between them. That shape can also describe
flows that merge, skip a stage, or return to a stage they already visited.
gog does not read it. The stage-column shape settles three things by itself.
Every path passes through every stage, the totals match at each one, and no
cycle can be written.
