# Flow

```{r}
#| include: false
source("R/setup.R")
```

Where did each part end up? Some tables record movement: each row says that
some amount started in one place and finished in another. A flow diagram draws
that movement. The stages stand side by side, and bands carry each amount from
stage to stage. The wider the band, the larger the amount it carries.

The `flow` transform lays out the whole picture from one table. This chapter
builds the diagram one layer at a time, using the passengers and crew of the
Titanic. `titanic` holds one row per group of people, with the class they
traveled in, their sex, their age band, whether they survived, and how many
they were:

```{r}
#| echo: false
peek(titanic)
```

## One row is one path

Name the stages in the atom, in reading order, and bind the count to `y`:

```{r}
data(titanic) + ribbon * flow(class, survived) + y(n) +
  title("Who survived, by class")
```

*"Given titanic: ribbons derived by flow through class and survived, y is n."*

Each band is one path through the stages. The band that leaves `1st` and
arrives at `Yes` carries every first-class passenger who survived, and its
thickness is that count at both ends. Rows that share a path are added
together, so the two `1st` rows for men and women who survived become one
band. Columns the atom does not name, `sex` and `age` here, are set aside the
same way.

The stages stand on the horizontal axis, named by their columns. There is
nothing for `x()` to say: the atom's argument order is the order the stages
are drawn in. The vertical axis is the count, and it keeps its ticks. The
bands at each stage stack from zero with no gaps, so a height read off the
axis is a true cumulative count.

## More stages, more of the story

Add a stage by adding a column. The paths now run three stages long, and each
band still carries one path's count from end to end:

```{r}
data(titanic) + ribbon * flow(class, sex, survived) + y(n) +
  title("Class, then sex, then survival")
```

The picture answers a question the two-stage version could not. Follow any
band with your eye and you follow one group of people the whole way: which
class they traveled in, whether they were men or women, and how it ended.

## Color follows a stage

A band runs through every stage, so any stage can color it. Write `color()`
after the band layer and name the stage to read:

```{r}
data(titanic) + ribbon * flow(class, sex, survived) + y(n) + color(class) +
  title("The same paths, colored by class")
```

Coloring by `class` keeps each band the color of where it began, so the
mixing at `sex` and at `survived` becomes visible. Coloring by `survived`
would instead paint the whole picture by how it ends. Only a stage can color
a band, because a band is a whole path, and only a stage holds a value every
path carries.

## The slots and their names

Two more layers finish the diagram. A `zone` layer draws each stage's slots,
the stacked blocks the bands leave and arrive at. A `text` layer names them.
The three layers read one layout, so they always agree about where everything
sits:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(class) +
  zone * flow(class, sex, survived) + style(color = "gray") +
  text * flow(class, sex, survived) + label(name) +
  title("The full diagram: bands, slots, names")
```

The slots take their paint from `style()`, gray here so the bands keep the
color story to themselves. The names come from `label(name)`: the layout
publishes each slot's name under `name`, the same word a labeled treemap
reads.

The count on `y` is optional. Bind nothing and every row weighs 1, so the
diagram tallies rows instead of summing a column. That is the same default
`partition` uses, and the same one `count` is built on.

## What flow refuses

A flow needs at least two stages, because one column has no between:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class)
```

The stage axis is drawn from the atom's own columns, so a bound `x` has
nothing left to say:

```{r}
#| error: true
data(titanic) + x(n) + ribbon * flow(class, survived)
```

Three marks read the layout, and the refusal names them for any other mark:

```{r}
#| error: true
data(titanic) + y(n) + bar * flow(class, survived)
```

And a band's color must name a stage. A column outside the stages holds no
value a whole path shares:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + color(n)
```

## Where its edge is

Bending a flow into `polar()` is valid grammar. The bands bent around a rim
are the chord diagram. This engine does not draw it yet, and says so:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + polar()
```

The input has an edge too. A flow reads stage columns from one table, one row
per path. Some tools accept a different shape, a list of nodes and a list of
links between them, and that shape can also describe flows that merge, skip a
stage, or cycle back. gog does not read it. The stage-column shape is the one
where the guarantees hold by construction: every path passes through every
stage, so the totals agree at each one, and a cycle cannot be written at all.
