# Flow

```{r}
#| include: false
source("R/setup.R")
```

Where did each part go? Some tables record movement: each row says that some
amount started in one place and finished in another. A flow diagram draws that
movement. The places are sorted into stages that stand side by side, and bands
carry each amount from stage to stage. The wider the band, the larger the amount
it carries. The world calls this picture a Sankey diagram, or an alluvial
diagram, and as with the pie there is no atom that spells either name.

A **stage** is one of the table's categorical columns. Its distinct values are
the possible places at that step, and each row sits in one of them. The `flow`
transform computes the whole diagram from one table of such columns. This
chapter builds the diagram one layer at a time, using the passengers and crew of
the Titanic. `titanic` holds one row per group of people. The columns are
`class`, `sex`, `age` (child or adult) and `survived`. The last column, `n`, is
how many people the group held. The table has one row for every combination of
those four columns, so some rows count zero people:

```{r}
#| echo: false
peek(titanic)
```

## One row is one path

The first question of the Titanic table is who survived, and whether class
changed the chances. That is a movement from a class to an outcome, which is a
flow with two stages. Name the stages in the atom, left to right, and bind the
count to `y`:

```{r}
data(titanic) + ribbon * flow(class, survived) + y(n) +
  title("Who survived, by class")
```

*"Given titanic: ribbons derived by flow through class and survived, y is n."*

Each band is one path through the stages. This plot draws no names, so find a
path by its position: the lowest band leaves the `1st` slot and arrives at the
`Yes` slot. It carries the 203 first-class passengers who survived, and it holds
that thickness at both ends. Rows that share a path are added together, so the
four `1st` rows that end in `Yes`, one per sex and age, become one band. The
same addition combines rows that differ only in the columns the atom does not
name, `sex` and `age` here.

The stages stand on the horizontal axis, named by their columns. `x()` has
nothing left to set: the atom's argument order is the order the stages are drawn
in. The vertical axis is the count, with the same ticks any other plot would
get. The bands at each stage stack from zero with no gaps, so the top edge of
any band gives the total of that band and every band under it.

## Adding a stage

The next question is whether women survived more often than men, which the
two-stage diagram could not answer. Add a stage by adding a column. Each path
now crosses three stages, and each band still carries one path's count at both
of its ends:

```{r}
data(titanic) + ribbon * flow(class, sex, survived) + y(n) +
  title("Class, then sex, then survival")
```

*"Given titanic: ribbons derived by flow through class, sex and survived, y is
n."*

The three-stage picture answers a question the two-stage one could not: how many
female passengers in first class survived? Follow any band and you follow one
group of people. You see their class, whether they were male or female, and
whether they survived.

## Color comes from a stage

With three stages the bands cross one another, and a path is hard to follow
through a crossing unless something marks it. A band runs through every stage,
so any stage can color it. Write `color()` after the `ribbon` layer and name
the stage whose value gives each band its color:

```{r}
data(titanic) + ribbon * flow(class, sex, survived) + y(n) + color(class) +
  title("The same paths, colored by class")
```

*"Given titanic: ribbons derived by flow through class, sex and survived, y is
n, color by class."*

Coloring by `class` gives each band the color of the slot it starts from. You
can then see which classes meet in each later slot. Coloring by `survived` would
instead color every band by how its path ends. Only a stage can color a band,
because a band is a whole path, and only a stage holds a value every path
carries.

## The slots and their names

So far you have found a path by its position, because no slot carries a name.
Two more layers finish the diagram. A `zone` layer draws each stage's slots, the
stacked blocks the bands leave and arrive at. A `text` layer names them. The
three layers read one computation, so they always agree about where everything
sits:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(class) +
  zone * flow(class, sex, survived) + style(color = "gray") +
  text * flow(class, sex, survived) + label(name) +
  title("The full diagram: bands, slots, names")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, color by class, and also zones derived by flow through class, sex and
survived, and text derived by flow through class, sex and survived, label by
name."*

The slots take their color from `style()`, gray here so that only the bands
carry color. The names come from `label(name)`: the transform writes each slot's
name into a column called `name`, exactly as `partition` does.

The count on `y` is optional. Bind nothing and every row counts as 1, so the
diagram counts rows instead of summing a column. `partition` uses that default
too, and `count` tallies rows the same way.

## What flow refuses

A flow needs at least two stages, because movement needs a place to start and
a place to finish:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class)
```

The stage axis is drawn from the atom's own columns, so a bound `x` has
nothing left to set:

```{r}
#| error: true
data(titanic) + x(n) + ribbon * flow(class, survived)
```

Only three marks can read what `flow` computes, and the refusal lists all three
when another mark is used:

```{r}
#| error: true
data(titanic) + y(n) + bar * flow(class, survived)
```

And the column that colors a band must be one of the stages. A column outside
the stages holds no value a whole path shares:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + color(n)
```

## Where the limits are

Every mark that draws on the page also bends into a circle, so the flow is a
natural thing to try there. Bending a flow into `polar()` is valid grammar. The
bands bent around a circle are the chord diagram, which this engine does not
draw:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + polar()
```

The input has a limit too. A flow reads stage columns from one table, one row
per path. Some tools accept a different shape: a list of nodes, which are these
slots, and a list of links between them. That shape can also describe flows that
merge, skip a stage, or return to a stage they already visited. gog does not
read it. The stage-column shape settles three things by itself. Every path
passes through every stage, the totals match at each one, and no path can return
to a stage it already visited.