# Flow

```{r}
#| include: false
source("R/setup.R")
```

Where did each part of a total go? Some tables record movement: each row says
that some amount started in one place and finished in another. A flow diagram
draws that movement. The places are grouped into stages, and the stages are
drawn side by side. Bands carry each amount from stage to stage. The thicker the
band, the larger the amount it carries. This picture is an **alluvial diagram**,
one kind of **Sankey diagram**. The last section of this chapter says which
Sankey diagrams gog does not draw. No atom spells either name, just as there is
no `pie()` atom.

gog builds that picture out of the table's own columns. A **stage** is one of
the table's categorical columns. Its distinct values are the possible places at
that stage, and each row sits in one of them. The `flow` transform computes the
whole diagram from one table of such columns. This chapter builds the diagram
one layer at a time, using the passengers and crew of the Titanic. `titanic`
holds one row per group of people. Four columns describe each group: `class`,
`sex`, `age` (child or adult) and `survived`. A fifth, `n`, counts the people in
that group. The table has one row for every combination of those four columns,
so some rows count zero people:

```{r}
#| echo: false
peek(titanic)
```

## From rows to bands

The first question to ask of `titanic` is who survived, and whether a person's
class changed their chance of survival. Class is one stage and the outcome is
another, so this is a flow with two stages. Name the stages in the atom, left to
right, and bind the count to `y`:

```{r}
data(titanic) + ribbon * flow(class, survived) + y(n) +
  title("Who survived, by class")
```

*"Given titanic: ribbons derived by flow through class and survived, y is n."*

Each band is one **path**: one place at each stage, here a class and then an
outcome. Each place takes a **slot**, the part of its stage that its bands fill.
This plot neither draws nor names the slots, so find a path by its position.
Inside each stage the slots stack from the bottom, in the order the column
declares: `1st` to `Crew`, and `Yes` before `No`. So the lowest band leaves the
`1st` slot and arrives at the `Yes` slot. It carries the 203 first-class
passengers who survived, and it keeps that thickness at both ends. Rows that
share a path become one band: their counts are added, and the band carries the
total. The four `1st` rows that end in `Yes` differ only in `sex` and `age`, and
the atom names neither, so the four become one band.

The plot answers the question. The lowest band fills most of the `1st` slot: 203
of the 325 people in first class survived. In third class only 178 of 706 did,
so class did change the chance of survival.

`flow()` also sets both axes. The stages are drawn along the horizontal axis,
each named by its column, in the order you name them inside `flow()`. The
vertical axis is a continuous axis for the count. The bands at each stage stack
from zero with no gaps. So the top edge of any band is the total of that band and
every band below it.

The count on `y` is optional. Bind nothing and every row counts as 1, and the
diagram counts rows instead of summing a column. `partition` uses that default
too, and `count` tallies rows the same way. This table needs `y(n)`: it holds
one row per combination, so counting its rows would count combinations rather
than people.

## Adding a stage

The next question is whether women survived more often than men. The two-stage
diagram cannot answer it, because nothing in it records a person's sex. Add a
stage by naming one more column inside `flow()`. Each path now passes through
three stages, and each band is as thick as its count at every one:

```{r}
data(titanic) + ribbon * flow(class, sex, survived) + y(n) +
  title("Class, then sex, then survival")
```

*"Given titanic: ribbons derived by flow through class, sex and survived, y is
n."*

Each band is now a group of people with one class, one sex and one outcome.
`sex` declares no order, so its slots follow the order its values first appear
in the table: `Male` at the bottom, then `Female`. The band from `1st` through
`Female` to `Yes` carries 141 people, the women and girls in first class who
survived. Most people in the `Female` slot arrive at `Yes`, and most in the
`Male` slot arrive at `No`. Of 470 women and girls, 344 survived. Of 1,731 men
and boys, 367 did.

## Color from a stage

With three stages the bands cross one another, and when every band has the same
color, one band is hard to follow through a crossing. A band runs through every
stage, so any stage can color it. Write `color()` after the `ribbon` layer and
name the stage whose value gives each band its color:

```{r}
data(titanic) + ribbon * flow(class, sex, survived) + y(n) + color(class) +
  title("The same paths, colored by class")
```

*"Given titanic: ribbons derived by flow through class, sex and survived, y is
n, color by class."*

Coloring by `class` gives each band the color of its class, the slot it starts
from. You can then see which classes arrive in each later slot. Coloring by
`survived` would instead give each band the color of its outcome, the slot it
arrives at. Only a stage can color a band. The rows in one band can differ in
any other column, but they share one value at each stage.

## The slots and their names

So far you have found a path by its position, because no slot has a name. Two
more layers finish the diagram. A `zone` layer draws each stage's slots, which
are the stacked blocks where the bands meet each stage. A `text` layer names
them. All three layers repeat the same `flow(...)`, so they compute the same
positions for the bands, slots and names:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(class) +
  zone * flow(class, sex, survived) + style(color = "gray") +
  text * flow(class, sex, survived) + label(name) +
  title("The full diagram: bands, slots, names")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, color by class, and also zones derived by flow through class, sex and
survived, and text derived by flow through class, sex and survived, label by
name."*

The slots take their color from `style()`, which makes them gray so that only
the bands are colored. The names come from `label(name)`: the transform writes
each slot's name into a column called `name`, exactly as `partition` does.

## What flow refuses

A flow needs at least two stages, because movement needs a place to start and
a place to finish:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class)
```

The stage axis is drawn from the atom's own columns, so gog refuses an `x()`
you write:

```{r}
#| error: true
data(titanic) + x(class) + ribbon * flow(class, survived)
```

Only `ribbon`, `zone` and `text` can read what `flow` computes. Any other mark
is refused, and the message names the three:

```{r}
#| error: true
data(titanic) + y(n) + bar * flow(class, survived)
```

And the column that colors a band must be one of the stages. `age` is in the
table, but the rows inside one band can differ in `age`:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + color(age)
```

## Where the limits are

The two limits below are not mistakes in your sentence. The engine does not
draw the first, and it does not read the second.

Every mark that draws on the page also bends into a circle, so you would expect
a flow's ribbons to bend as well. [`polar()`](polar.qmd) is what bends a plot,
and a flow in `polar()` is valid grammar. The bands bent around a circle make a
**chord diagram**, which this engine does not draw:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + polar()
```

The input has a limit too. A flow reads its stages as columns of one table, so
each row has one place at every stage. Some tools accept a different shape. They
take a list of nodes, which are the slots, and a list of links between those
nodes. That shape can also describe flows that begin partway, skip a stage, or
return to a stage they already visited. A flow does not read it. In the
stage-column shape, every path passes through every stage, so every stage has
the same total. No path can return to a stage it already visited.

## What comes next

A flow draws its stages in the order you name them, and it never computes an
order itself. [Cluster](cluster.qmd) computes one from the data: it joins the
most alike categories into a tree, and puts an axis in the tree's order.
