# Flow

```{r}
#| include: false
source("R/setup.R")
```

Where did each part of a total go? Some tables record movement: each row says
that some amount started in one place and finished in another. A flow diagram
draws that movement. The places are grouped into stages, and the stages are
drawn side by side. Bands carry each amount from stage to stage. The thicker the
band, the larger the amount it carries. This picture is an **alluvial diagram**,
one kind of **Sankey diagram**. The section [Links in layers](#links-in-layers)
shows a Sankey diagram drawn from a table of links, and [Where the limits
are](#where-the-limits-are) says what gog does not draw. No atom spells either
name, just as there is no `pie()` atom.

gog builds that picture out of the table's own columns. A **stage** is one of
the table's categorical columns. Its distinct values are the possible places at
that stage, and each row sits in one of them. The `flow` transform computes the
whole diagram from one table of such columns. This chapter uses the passengers
and crew of the Titanic, the ship that sank in 1912. `titanic` holds one row per
group of people. Four
columns describe each group: `class`, `sex`, `age` (child or adult) and
`survived`. A fifth, `n`, counts the people in that group. The table has one row
for every combination of those four columns, so some rows count zero people:

```{r}
#| echo: false
peek(titanic)
```

## The finished diagram

Who survived the sinking, and did class and sex change their chance of
survival? The diagram below answers both:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("Who survived, by class and sex")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, and also zones derived by flow through class, sex
and survived, and text derived by flow through class, sex and survived, label by
name."*

Read it from left to right. Each white rectangle marks a **slot**: the part of a
stage that one place takes, such as a class, a sex or an outcome. Each band is a group of people who
share a class, a sex and an outcome, and its color is the outcome: blue for the
people who survived, orange for the people who did not. The blue bands that
leave the `Female` slot are much thicker than the orange ones, because 344 of
the 470 women and girls survived. At the `Male` slot the orange bands are
thicker, because only 367 of the 1,731 men and boys survived. Class mattered
too. 203 of the 325 people in first class survived, compared with 178 of the 706 in
third class.

## Three layers

The sentence above has three layers. In each one, `*` joins the same
transform, `flow(...)`, to a different mark, and each mark draws one part of the
picture:

| Layer | What it draws |
|---|---|
| `ribbon * flow(...)` | the bands, which carry the counts from stage to stage |
| `zone * flow(...)` | the slots, one rectangle for each place at each stage |
| `text * flow(...) + label(name)` | the names, one inside each slot |

The three layers compute the same positions from the same stages. So the bands
end at the edges of the slots, and each name sits inside its slot. The next
three sections add the layers one at a time, in the order of the table. They use
two stages, `class` and `survived`, and [Adding a stage](#adding-a-stage) then
adds `sex` between them.

## The bands

To build the diagram, start with the bands alone. Class is one stage and the
outcome is another. Name the stages inside `flow()`, from left to right, and
bind the count to `y`:

```{r}
data(titanic) + ribbon * flow(class, survived) + y(n) +
  title("Who survived, by class")
```

*"Given titanic: ribbons derived by flow through class and survived, y is n."*

Each band is one **path**: one place at each stage, here a class and then an
outcome. This plot does not outline the slots or name them, so find a path by
its position. Inside each stage the slots stack from the bottom, in the order
the column declares: `1st` to `Crew`, and `Yes` before `No`. So the lowest band leaves the
`1st` slot and arrives at the `Yes` slot. It carries the 203 first-class
passengers who survived, and it keeps that thickness at both ends. Rows that
share a path become one band: their counts are added, and the band carries the
total. The four `1st` rows that end in `Yes` differ only in `sex` and `age`, and
the atom names neither, so the four become one band.

`flow()` also sets both axes. With the count on `y`, the stages are drawn along
the horizontal axis, each named by its column, in the order you name them inside
`flow()`. The vertical axis is a continuous axis for the count. The bands at
each stage stack from zero with no gaps. So the top edge of any band is the
total of that band and every band below it.

The count is optional. Bind nothing and every row counts as 1, and the
diagram counts rows instead of summing a column. `partition` uses that default
too, and `count` tallies rows the same way. This table needs `y(n)`: it holds
one row per combination, so counting its rows would count combinations rather
than people.

## The slots

In the plot above, the bands that leave one place touch the bands that leave
the next, so you cannot see where one place ends and the next begins. A `zone`
layer draws the slots. Add `zone * flow(class, survived)`, with the same stages
as the bands. Both layers now use the count, so write `y(n)` before any mark,
where it belongs to the plot and every layer reads it ([Encoding
scope](encoding-scope.qmd#the-rule)):

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, survived) +
  zone * flow(class, survived) +
  title("The same flow, with its slots")
```

*"Given titanic: y is n, ribbons derived by flow through class and survived, and
also zones derived by flow through class and survived."*

Each rectangle marks one slot. It covers the same part of the count axis as the
bands that leave it or arrive at it, so each slot is as tall as the number of
people at that place. A [categorical
axis](scales.qmd#the-columns-type-picks-the-axis) also gives each category a
slot, but there every slot is the same size. The top slot of the class stage is
the crew, the largest group, with 885 people. At the outcome stage the larger
slot is `No`: 1,490 of the 2,201 people did not survive.

Each slot is white with a thin black line around it, so only the bands have
color and the edge of every slot is visible. Those are the defaults, and
[Choosing the colors](#choosing-the-colors) shows how to change them. The slots
still have no names, so you still count from the bottom to find each slot.

## The names

A `text` layer names the slots. The transform writes each slot's name into a
column called `name`, exactly as `partition` does. Add a third layer,
`text * flow(class, survived)`, with the same stages again, and bind
`label(name)` to draw that column:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, survived) +
  zone * flow(class, survived) +
  text * flow(class, survived) + label(name) +
  title("The same flow, with its slots named")
```

*"Given titanic: y is n, ribbons derived by flow through class and survived, and
also zones derived by flow through class and survived, and text derived by flow
through class and survived, label by name."*

Each name is drawn at the center of its slot. The layers are drawn in the order
they are written, so the `text` layer comes last. If the `text` layer came
before the `zone` layer, the white slots would hide the names.

The `1st` slot is now easy to find, and the thicker of its two bands arrives
at `Yes`. It carries the 203 of the 325 people in first class who survived, the
number given under the finished diagram. Every later plot in this chapter keeps
the three layers.

## Adding a stage

The finished diagram has a third stage, `sex`, between class and the outcome.
Add it by naming one more column inside `flow()`, in all three layers. Each path
now passes through three stages, and each band is as thick as its count at
every one:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("Class, then sex, then survival")
```

Each band is now a group of people with one class, one sex and one outcome.
`sex` declares no order, so its slots follow the order its values first appear
in the table: `Male` at the bottom, then `Female`. The band from `1st` through
`Female` to `Yes` carries 141 people, the women and girls in first class who
survived. Most people in the `Female` slot arrive at `Yes`, and most in the
`Male` slot arrive at `No`, as in the finished diagram.

## Color from a stage

With three stages the bands cross one another, and when every band has the same
color, one band is hard to follow through a crossing. A band runs through every
stage, so any stage can color it. The finished diagram at the start of this
chapter colors each band by `survived`, the slot it arrives at. Coloring by
`class` gives each band the color of the slot it leaves at the first stage
instead:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(class) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("The same paths, colored by class")
```

Now you can see which classes arrive in each later slot. Coloring by `survived`
and coloring by `class` answer two questions: who lived, and where each class
went.

`color()` is written after the `ribbon` layer, so it belongs to the bands alone.
Written before the marks, it would reach the slots and the names as well. gog
refuses that, because a slot and its name take their color from `style()`.

## Color from another column

Did the women in third class survive as often as the women in first class? A
color can show it inside each band, with no third stage. Any categorical column
can color the bands, not only a stage. A column that is not a stage splits each
band into one part for each of its values. A categorical color splits a line the
same way, into one line for each category. The parts of one band lie next to
each other across its thickness and follow its path, and the slots are the same
as without the color. With `survived` as a color instead of a stage, a flow
through class and sex shows who survived in every band:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex) + color(survived) +
  zone * flow(class, sex) +
  text * flow(class, sex) + label(name) +
  title("Class, then sex, colored by survival")
```

*"Given titanic: y is n, ribbons derived by flow through class and sex, colored
by survived, and also zones derived by flow through class and sex, and text
derived by flow through class and sex, label by name."*

Each band from a class to a sex now has a blue part for the people who survived
and an orange part for the people who did not. The band from `1st` to `Female`
is nearly all blue: 141 of its 145 people survived. The band from `3rd` to
`Female` is less than half blue: 90 of 196. The finished diagram shows the same
split with `survived` as a stage, where the outcome also gets slots of its own.
As a color, the outcome needs no slots of its own, and the legend names its
values.

Only a categorical column can split a band, because its values are groups. A
continuous column is refused, as it is on every `ribbon`.

## Choosing the colors

Every color in this chapter so far is a default. The bands take gog's default
palette and are translucent, the slots are white with a thin black line around
them, and the names are near-black. You choose each one in its own layer.

The bands take their colors from a column, and `palette()` chooses the color for
each value of that column. Write each value beside the color you want, as
[Colors by name](channels.qmd#colors-by-name) explains. Name every value, since
gog refuses a palette that leaves one out:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  palette(c(Yes = "steelblue", No = "darkgray")) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("The survivors in blue")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, with a palette naming Yes steelblue and No
darkgray, and also zones derived by flow through class, sex and survived, and
text derived by flow through class, sex and survived, label by name."*

The bands of the survivors are blue and the others are gray, so the survivors
are easy to find. A band is translucent, so it looks lighter than the color you
name, and two bands that cross both stay visible. To give every band the same
color, leave out `color()` and write `style(color = )` after the `ribbon` layer.
`style(opacity = )` there sets how translucent the bands are, with or without
`color()`.

A slot sits at one stage, so no single column can color every slot, and the
same is true of the names. The `zone` and `text` layers take their settings from
`style()` instead, one value for each layer. On the `zone` layer,
`color` fills the slots, and `border_color` and `border_size` draw the line
around each one. On the `text` layer, `color` and `size` set the color and the
size of the names:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  palette(c(Yes = "steelblue", No = "darkgray")) +
  zone * flow(class, sex, survived) +
  style(color = "midnightblue", border_color = "white", border_size = 2) +
  text * flow(class, sex, survived) + label(name) +
  style(color = "white", size = 14) +
  title("Dark slots with white names")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, with a palette naming Yes steelblue and No
darkgray, and also zones derived by flow through class, sex and survived,
colored midnightblue, with border color white and border size 2, and text
derived by flow through class, sex and survived, label by name, colored white,
with size 14."*

Each `style()` is written after the mark it changes, so it changes that layer
and no other. On the dark fill, the white line shows where each slot ends, and
`border_size = 0` would remove it. The `zone` layer also takes `opacity`, as
every mark does, and `pattern`, as the five filled regions do
([Style](style.qmd#what-you-can-set)).

## Room for long names

In a flow that runs left to right the slots are tall and narrow. The names in
`titanic` are short, so every one fits. A longer name, such as
`Member of the crew`, would be wider than its slot, and gog would draw it across
the slot's edges, over the bands. There are two ways to give it more space.

The first turns the names. `style(angle = 90)` on the `text` layer turns every
name upward, so it runs along its slot, as [Turning a
label](marks/text.qmd#turning-a-label) explains:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) + style(angle = 90) +
  title("Names turned 90 degrees")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, and also zones derived by flow through class, sex
and survived, and text derived by flow through class, sex and survived, label by
name, with angle 90."*

Each name is now written upward, so it must fit the slot's height, not the
slot's width.

The second places the stages from top to bottom. Bind the count to `x` instead
of `y`, and the first stage is at the top of the plot. Each slot is then wide
and short, so a long name fits across it unless the slot's count is small. The
sentence is the finished diagram's, with `x(n)` in place of `y(n)` and a new
title. gog takes the direction from the axis that holds the count, as it does
for a bar chart ([Horizontal bars](marks/bar.qmd#horizontal-bars)):

```{r}
data(titanic) + x(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  title("The same flow, from top to bottom")
```

*"Given titanic: x is n, ribbons derived by flow through class, sex and
survived, colored by survived, and also zones derived by flow through class, sex
and survived, and text derived by flow through class, sex and survived, label by
name."*

The stage names are now on the left: `class` at the top, then `sex`, then
`survived`. The count is on the bottom axis, and each name fits inside its slot.

## Leaving out the count axis

The count axis helped while the diagram was being built: it showed that the
bands at each stage add up to the same total, just over 2,000 people. A reader of the finished
diagram compares the thicknesses of the bands and the slots, and needs no
numbers for that. So the count axis, the gridlines and the frame (the two lines
along the bottom and the left of the panel) can be left out, and the gray panel
can be white. Write `axis = FALSE` on the count. In
`theme()`, `grid = "none"` and `frame = "none"` leave out the gridlines and the
frame, and `background = "white"` colors the panel ([Axes, legends, gridlines
and the frame](style.qmd#axes-legends-gridlines-and-the-frame)). The stage names
stay, because they belong to the other axis, which is still drawn:

```{r}
data(titanic) + y(n, axis = FALSE) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  theme(grid = "none", frame = "none", background = "white") +
  title("Who survived, by class and sex")
```

*"Given titanic: y is n without an axis, ribbons derived by flow through class,
sex and survived, colored by survived, and also zones derived by flow through
class, sex and survived, and text derived by flow through class, sex and
survived, label by name."*

Only the bands, the slots with their names, the legend and the stage names are
left. A reader can still see that most of the women survived and most of the
men did not. The exact counts are gone from the picture, so give them in the
paragraph beside the plot, as [The finished diagram](#the-finished-diagram)
does.

## Selecting one place

Color tells the bands apart, but it does not show where the people in one place
went. `brush` keeps the color of the bands that pass through one place and dims
the rest. Name a stage, and use `at` to name the place the selection starts
with, as a brush on a [categorical
column](selection.qmd#a-categorical-column-selects-slots) does:

```{r}
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  text * flow(class, sex, survived) + label(name) +
  brush(sex, at = "Female") +
  title("The women and girls, and where they went")
```

*"Given titanic: y is n, ribbons derived by flow through class, sex and
survived, colored by survived, and also zones derived by flow through class, sex
and survived, and text derived by flow through class, sex and survived, label by
name, and let the reader select sex, starting with Female."*

The bands that pass through `Female` keep their colors, and every other band is
dimmed. A brush on a stage never cuts a band in two. A band is a group of people
who share a place at every stage, so either all of them are in the selection or
none of them are. The `Male` slot is dimmed as well. The slots at the other two
stages hold both women and men, so they are not dimmed.

In the web edition, a click on a `sex` slot selects it. Click `Male` and the
selection moves there. Click the selected slot again, a band, or an empty part
of the panel, and nothing is dimmed. With a bare `brush`, a click selects a slot
at any stage. A drag selects nothing on a flow, and the pointer shows no values,
because each band and each slot stands for many rows. In the PDF, the plot keeps
the selection the sentence states.

A flow keeps or dims each band, or each part of a band, as a whole. So the
brushed column must be one that each band, or each part, holds a single value
of. A stage is one. So is a column given to the bands' `color`, because it
splits each band into parts, as [Color from another
column](#color-from-another-column) shows. Any other column can hold two values
in one band, and gog refuses it. `age` is a column of `titanic`, but not a stage
of this flow:

```{r}
#| error: true
data(titanic) + y(n) +
  ribbon * flow(class, sex, survived) + color(survived) +
  zone * flow(class, sex, survived) +
  brush(age, at = "Child")
```

## Two columns that name the same places

So far each column named in `flow()` has been a stage of its own. Some tables
record amounts between places of one kind instead. `trade_partners`, from the
[Network](network.qmd) chapter, records trade between countries: each row says
how much one country exported to another, in billions of dollars:

```{r}
#| echo: false
peek(trade_partners)
```

China appears in both columns. It exports to the United States, and it imports
from Korea. A flow through `exporter` and `importer` would put each column at a
stage of its own, so China would get two slots. `shared = TRUE` says that the two
columns name one set of places. Each country then gets one slot, and every slot
lies on one axis. The length of a country's slot is the total of its exports and
imports:

```{r}
data(trade_partners) + x(billions) +
  ribbon * flow(exporter, importer, shared = TRUE) + color(exporter) +
  zone * flow(exporter, importer, shared = TRUE) +
  text * flow(exporter, importer, shared = TRUE) + label(name) +
  title("Trade between countries, as arcs")
```

*"Given trade partners: x is billions, ribbons derived by flow through exporter
and importer, shared, colored by exporter, and also zones derived by flow through
exporter and importer, shared, and text derived by flow through exporter and
importer, shared, label by name."*

This picture is an **arc diagram**. Each band rises from the country that
exports and comes down at the country that imports. The farther apart the two
countries are on the axis, the higher the band rises. Each end of a band is as
wide as the amount it carries. `color(exporter)` gives each band the color of the
country it leaves.

A country's slot has up to two parts, with no line between them. The bands of
its exports leave the white part, and the bands of its imports arrive at the
gray part, to the right of the white part. Vietnam only imports, so its slot is
all gray. At China, the band to the United States leaves the white part, and the
band from Korea arrives at the gray part. The colors show this too: every band
of China's exports has China's color. The two parts show it even without the
legend. A `style(color = )` on the `zone` layer colors the white part, and the
gray part becomes a darker shade of that color.

The white part is for the first column in `flow()`, and the gray part is for the
second. To put the exports in the gray part instead, name `importer` first:

```{r}
data(trade_partners) + x(billions) +
  ribbon * flow(importer, exporter, shared = TRUE) + color(exporter) +
  zone * flow(importer, exporter, shared = TRUE) +
  text * flow(importer, exporter, shared = TRUE) + label(name) +
  title("Trade between countries, imports first")
```

Each slot now starts with the country's imports, in white, and its exports
follow in gray. The order of the countries changes too, but only at the start:
the United States now comes first and China second. gog puts the countries in
the order it first reads their names. In each row it reads the first column in
`flow()` before the second, so the first row now gives the United States before
China. If the columns have a declared order, gog uses that order instead.

The axis carries no numbers. Each amount takes space on the axis twice, once at
each end of its band, so an axis with numbers would end at twice the total
trade. The country names are written in place of the numbers, turned to read
upward when they do not fit side by side. With the amount on `y`, as in
`y(billions)`, the axis runs up the page, the names stand at its left, and the
bands reach to the right.

At the right end, four narrow slots stand side by side, and their names
overlap: Australia, Brazil, India and Spain. This plot is 800 pixels wide, the
default. At `theme(width = 1400)`, 1,400 pixels wide, the four names no longer
overlap. In the [chord diagram](#the-chord-diagram) below, each name is turned
to read outward from the ring. Neighboring names are close together at the ring
and farther apart at their outer ends. India and Spain still touch near the
ring.

The panel is white and has no frame, because the plot has no numbers to read
values from. `theme()` changes both, as it does for any plot: `background`
colors the panel, and `frame = "axes"` draws the two lines along the bottom and
the left ([Axes, legends, gridlines and the
frame](style.qmd#axes-legends-gridlines-and-the-frame)).

A shared flow, one with `shared = TRUE`, reads two columns: the place each
amount leaves, and the place it arrives at. A third column, for a route through
three places, would be valid grammar, but this engine does not draw it.

## The chord diagram

Add `polar()` to the arc diagram's sentence, the one with `exporter` first.
`polar()` puts the amounts on the angle, as it does for a pie, so the axis bends
into a ring. This picture is a **chord diagram**:

```{r}
data(trade_partners) + x(billions) +
  ribbon * flow(exporter, importer, shared = TRUE) + color(exporter) +
  zone * flow(exporter, importer, shared = TRUE) +
  text * flow(exporter, importer, shared = TRUE) + label(name) +
  polar() + title("Trade between countries, as chords")
```

*"Given trade partners: x is billions, ribbons derived by flow through exporter
and importer, shared, colored by exporter, and also zones derived by flow through
exporter and importer, shared, and text derived by flow through exporter and
importer, shared, label by name, in polar."*

Each country's slot is now a piece of the ring, and its length is still the
total of the country's exports and imports. It keeps its two parts, white for
exports and gray for imports. Each band curves toward the center
of the circle, from the country that exports to the country that imports. Like
the arc diagram's axis, the ring carries no numbers, and the names stand outside
it.

In a chord diagram, a band's area can mislead. A band's width at its two ends is
the amount it carries, but its area is not. A band between countries far apart
on the ring is long, and a band between two countries side by side on the ring
is short. The countries stand
around the ring in the order they first appear in the table, so the length of a
band says nothing about the amounts. Canada exported 420 billion dollars to the
United States, and China exported 500. Yet Canada's band covers more of the
picture, because China sits next to the United States. Read a band at its two
ends. The next section shows an easier way to read the diagram: one country at a
time.

## Selecting a place in either column

A brush on one column selects by that column alone. `brush(exporter, at =
"China")` keeps the bands that leave China and dims the bands that arrive there.
To keep both, brush the flow's own `name`, the column that `label(name)` reads.
That brush keeps every band that leaves or arrives at the place. Each band holds
two names, one at each end. [Selecting one place](#selecting-one-place) refuses
a column with two values in one band, and `name` is the one exception.

```{r}
data(trade_partners) + x(billions) +
  ribbon * flow(exporter, importer, shared = TRUE) + color(exporter) +
  zone * flow(exporter, importer, shared = TRUE) +
  text * flow(exporter, importer, shared = TRUE) + label(name) +
  brush(name, at = "China") +
  title("What China exports and imports")
```

*"Given trade partners: x is billions, ribbons derived by flow through exporter
and importer, shared, colored by exporter, and also zones derived by flow through
exporter and importer, shared, and text derived by flow through exporter and
importer, shared, label by name, and let the reader select name, starting with
China."*

Every band that leaves or arrives at China keeps its color. Every other band
is dimmed, and so is every other country's slot. In the web edition, a click on
a country's slot selects that country instead. A click on the selected slot, on
a band, or on an empty part of the panel clears the selection. `show rows` lists
the selected rows with both `exporter` and `importer`, so each row says whether
it is an export from China or an import into China.

The same brush and the same click work on the chord diagram:

```{r}
data(trade_partners) + x(billions) +
  ribbon * flow(exporter, importer, shared = TRUE) + color(exporter) +
  zone * flow(exporter, importer, shared = TRUE) +
  text * flow(exporter, importer, shared = TRUE) + label(name) +
  polar() + brush(name, at = "China") +
  title("What China exports and imports, as chords")
```

In the ring too, China's bands keep their colors, and every other band and
slot is dimmed. On a flow with stages, `brush(name, at = "Female")` keeps the
same bands as `brush(sex, at = "Female")`. The two differ only when one name
stands at two stages, and `name` then selects both.

## Links in layers

Some tables record movement whose paths do not all pass through the same
stages. `budget` holds one household's money for a month. Two incomes are added
into one, Income. From there the money goes to Taxes, Savings and four groups of
spending, and each group divides again. Each row is one **link**, an amount that
moves from one place to the next:

```{r}
#| echo: false
peek(budget)
```

Taxes and Savings are one step after Income, and no band leaves them. Rent is
one step later, and no band leaves it either. Utilities, at the same step as
Rent, divides again into Energy and Water. Stage columns cannot hold that,
because every row would need a value at every stage. `layered = TRUE` says that
each row is a link between two places. gog then puts the places in **layers**. A
layer is drawn like a stage, but gog computes it from the links instead of
reading a column. It is a different thing from a layer of the sentence, such as
the `ribbon` layer:

```{r}
data(budget) + y(amount) +
  ribbon * flow(source, target, layered = TRUE) + color(source) +
  zone * flow(source, target, layered = TRUE) +
  text * flow(source, target, layered = TRUE) + label(name) +
  title("A month of a household's money")
```

*"Given the budget: y is amount, ribbons derived by flow through source and
target, layered, colored by source, and also zones derived by flow through source
and target, layered, and text derived by flow through source and target,
layered, label by name."*

This picture is a Sankey diagram. Each place stands in the first layer that
comes after every place that sends to it. Salary and Side work send to Income,
so Income stands in the second layer. Taxes stands in the third layer, and no
band leaves it. Energy stands in the fifth layer, after Housing in the third and
Utilities in the fourth. Each slot is as tall as the larger of what arrives at
the place and what leaves it. A small gap separates the places of one layer.
`color(source)` gives each band the color of the place it leaves.

gog also computes the order of the places in each layer. It puts each place
near the places it links to, so that few bands cross. The slots are too narrow
to hold a name, so each name stands beside its slot. In every layer but the
last, it stands to the right, on top of any bands that leave the place. In the
last layer it stands to the left. Like the shared flow, a layered flow writes no numbers on its axes.
Its layers have no names to write along one axis. Along the other, the gaps
between places take space too, so the edge of a slot is not a running total
there.

## A Sankey diagram and an alluvial diagram

A layered flow adds up the amounts that arrive at a place. Salary and Side work
both arrive at Income. The band from Income to Housing holds money from each,
and the picture cannot show how much. The alluvial diagram at the start of this
chapter keeps each group of people on its own band, from the first stage to the
last. So when every row has a value at every stage, name those columns as
stages, and each band keeps its group. Use `layered = TRUE` for a table of
links, which stage columns cannot hold.

Adding up at each place also decides what a selection keeps.
`brush(name, at = "Housing")` keeps the links of Housing itself: the band that
arrives at Housing, and the bands that leave it for Rent, Utilities and
Repairs:

```{r}
data(budget) + y(amount) +
  ribbon * flow(source, target, layered = TRUE) + color(source) +
  zone * flow(source, target, layered = TRUE) +
  text * flow(source, target, layered = TRUE) + label(name) +
  brush(name, at = "Housing") +
  title("What arrives at Housing and what leaves it")
```

The bands from Utilities to Energy and Water stay dimmed, although here all of
their money passed through Housing. A place can receive money from several
places, as Income does. A band that leaves such a place carries some of each,
and the picture cannot show how much. So a selection keeps only the bands that
arrive at the place or leave it, whatever the table holds. In the web edition, a click on a slot
selects that place, as on the other flows.

## What flow refuses

A flow needs at least two stages, because movement needs a place to start and
a place to finish:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class)
```

`flow()` puts the stages on the axis that does not hold the count. With `y(n)`
that is the x axis, so `x(class)` asks for the same axis as the stages, and gog
refuses it:

```{r}
#| error: true
data(titanic) + y(n) + x(class) + ribbon * flow(class, survived)
```

Only `ribbon`, `zone` and `text` can read what `flow` computes. Any other mark
is refused, and the message names the three:

```{r}
#| error: true
data(titanic) + y(n) + bar * flow(class, survived)
```

And a continuous column cannot color a band, just as it cannot color any
`ribbon`. `n` is a continuous column:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + color(n)
```

## Where the limits are

The two limits below are not mistakes in your sentence. Each is valid grammar
that this engine does not draw.

A flow with stages is valid grammar in [`polar()`](polar.qmd) too. Each stage
would get a ring of its own, so the picture would be a circular alluvial
diagram, not the chord diagram above. This engine does not draw it, and the
message points to `shared = TRUE` for two columns that name one set of places:

```{r}
#| error: true
data(titanic) + y(n) + ribbon * flow(class, survived) + polar()
```

A layered flow has a limit too. Its bands may skip a layer, or stop before the
last layer, but its links may not form a cycle. Links that form a cycle have no
layers, because no place in the cycle comes first. A household that moves money
from its checking account to its savings account, and back again, makes a
cycle. gog refuses it, and the message names the cycle:

```{r}
#| error: true
accounts <- data.frame(source = c("Checking", "Savings"),
                       target = c("Savings", "Checking"), amount = c(200, 50))
data(accounts) + y(amount) + ribbon * flow(source, target, layered = TRUE)
```

Some tools draw a cycle as a band that runs back across the diagram. The
message points to `shared = TRUE` instead, which puts every place on one axis
and draws the same links as arcs.

## Choosing a flow diagram

This chapter has drawn four kinds of flow diagram: alluvial, Sankey, arc and
chord. To choose one, start from what one row of your table says. The last row
of the table is a fifth choice, the network diagram:

| What one row says | A question it answers | The diagram |
|---|---|---|
| a group and its place at each of several stages, as in `titanic` | How did each group move from stage to stage? | an alluvial diagram ([The finished diagram](#the-finished-diagram)) |
| a link, an amount that moves from one place to the next, with no cycles, as in `budget` | Where does the total go, step by step? | a Sankey diagram ([Links in layers](#links-in-layers)) |
| an amount between two places of one kind, often in both directions, as in `trade_partners` | Who sends to whom, and how much each way? | an arc diagram, or the same links on a ring, a chord diagram ([Two columns that name the same places](#two-columns-that-name-the-same-places)) |
| a connection between two places, when who is connected matters more than how much | What is connected to what? | a network diagram ([Network](network.qmd)) |

A flow diagram shows the pattern of the movement, but a band is hard to compare
with a band in another part of the picture. A [tile
plot](marks/zone.qmd#cells-the-axes-already-made-the-tile-plot) gives each pair
of places its own cell, so every pair has a fixed place in a grid:

```{r}
data(trade_partners) + zone + x(importer) + y(exporter) + color(billions) +
  title("Trade between countries, one cell per pair")
```

*"Given trade partners: zones, x is importer, y is exporter, color by
billions."*

Each row of cells is an exporter, and each column of cells is an importer. The
darkest cell is China's exports to the United States, the largest amount in the
table. The cells for Mexico and Canada are nearly as dark, so the shades alone
do not say which is larger. A reader compares lengths more accurately than
shades of a color, so draw the amounts as bars when the comparison must be
precise.

Four more suggestions help you draw a flow diagram. First, color the bands by
the column that answers your question. On a flow with stages, the first stage
shows where each group goes, and the last shows the outcomes. Second, add
`brush` to select one place: the bands that meet it keep their colors, and the
others are dimmed. With `brush` written, a click on a slot in the web edition
selects another place. Third, use few places. Before you draw, give the small
places one shared name in your table, such as Other. On a layered flow, give it
only to places that do not link to each other: a link from Other to Other is a
cycle. Fourth, in an arc or a chord diagram, the order of the places decides how
long the bands are and how often they cross. So declare an order that means
something, such as by continent.

## What comes next

A flow with stages stacks each stage's slots in the order you declare, or else
in the table's order. [Cluster](cluster.qmd) computes an order from the data
instead: it joins the categories that are most alike into a tree, and puts an
axis in the tree's order.
