# Globe

```{r}
#| include: false
source("R/setup.R")
```

What does the data look like on the earth itself? The map chapter flattened
the sphere to fit a page. This chapter keeps the sphere round, and turns it
toward you.

Here is the map chapter's first sentence, and beside it the same sentence with
`globe()` in place of `map()`:

```{r}
#| echo: false
peek(quakes_fiji)
```

```{r}
(data(quakes_fiji) + point + x(east) + y(north) + map() +
   title("map()")) |
  (data(quakes_fiji) + point + x(east) + y(north) +
     globe(turn = 178, tilt = -18) + title("globe(turn = 178, tilt = -18)"))
```

*"Given quakes Fiji: points, x is east, y is north, in the map space, beside the
same on the globe."*

`x` is longitude and `y` is latitude, exactly as on the map. What changed is the
meaning of a position on the page. Each point now stands at its place on the
sphere, and the sphere is drawn as the disk you see when you look at one.

## The view

A globe shows one half of the earth, so the choice of half matters. `turn` and
`tilt` name the place the view faces, and they are the same two arguments
`space()` takes. `turn` is the longitude at the center of the disk, and `tilt`
is the latitude. The chapter's first plot faces Fiji. Leave both out and the
view faces longitude zero and latitude zero, where the equator crosses the prime
meridian.

`turn` is a bearing, so it wraps: 190 and -170 name the same view. `tilt` is a
latitude, so it runs from -90 to 90 and no further.

In the web edition the angle you write is only the view the globe opens at. Drag
it and it turns, the way the cube in [Space](space.qmd) turns, and the far half
of the earth turns into view. The *reset* button returns to the view the
sentence asked for. On paper a plot keeps its one view, which is why this
chapter also draws some data at a second angle.

## What a view hides

Half the earth faces away, and data on that half is hidden behind the sphere.
gog counts what a view hides rather than dropping it in silence:

```{r}
#| echo: false
peek(world_borders)
```

```{r}
#| message: true
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) +
  style(color = "tomato", opacity = 0.35, size = 3) +
  globe(turn = 178, tilt = -18) + title("The map chapter's layers, kept round")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given quakes Fiji: points, x is east, y is north, on the globe."*

Read the message above the plot. The coastline table covers the whole earth, so
any one view hides some of it. From this view most of its rows face away, while
every earthquake faces you. A second angle brings the hidden rows into view:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  globe(turn = 20, tilt = 5) + title("The other side")
```

## The graticule

The gray lines on the sphere are the **graticule**: a meridian every thirty
degrees of longitude, and a parallel every thirty degrees of latitude. They
are this space's gridlines. A flat panel draws gridlines too, and here the same
lines are bent onto the sphere, the way the [polar](polar.qmd) space bends them
into rings and spokes. There are no axes beside the disk, because a sphere
has no edge to write one on, and the graticule is the reference instead.
`theme(grid = )` reaches it like any other grid:

```{r}
data(quakes_fiji) + point + x(east) + y(north) +
  globe(turn = 178, tilt = -18) + theme(grid = "none") +
  title("No graticule")
```

With the graticule off, nothing inside the disk says which part of the earth you
are looking at.

## A route bends with the sphere

The graticule bends with the sphere, and so does everything else drawn on it. A
route is the clearest case. On the map, `path` joins places with straight
strokes. On the globe a straight route does not exist. The shortest way between
two places bends with the surface, along a **great circle**: the line a tight
string would take if you pulled it across the sphere between the two places. A
`path` here follows it:

```{r}
#| echo: false
peek(flight)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(flight) + path + x(lon) + y(lat) +
  style(color = "#3d5a80", arrow = "end") +
  globe(turn = 178, tilt = -18) + title("Tokyo to Los Angeles")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the flight: a path, x is lon, y is lat, on the globe."*

Two rows describe that route, and the curve between them comes from the sphere.
The arc bends toward the pole because a route that swings north really is
shorter. A route that keeps to one latitude is the longer way.

## Many routes are a network

That table held one route. One table can hold every route at once, and the
sentence barely changes. Give each route a name, and give each of its two rows
a longitude and a latitude. `group()` then splits the table into one arc per
route, exactly as it split the coastline into pieces. A `point` layer marks the
cities and a `text` layer names them. The picture this draws is a route network,
and it needs no new word, because every position in it is a place the table
already holds:

```{r}
#| echo: false
peek(routes)
peek(cities)
```

```{r}
#| message: true
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(routes) + path + x(lon) + y(lat) + group(route) +
  style(color = "#3d5a80", opacity = 0.6) +
  data(cities) + point + x(lon) + y(lat) +
  data(cities) + text + x(lon) + y(lat) + label(city) + style(nudge = "up") +
  globe(turn = 95, tilt = 30) + title("Routes between cities")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the routes: paths, x is lon, y is lat, grouped by route; then given the
cities: points, x is lon, y is lat; then given the cities: text, x is lon, y is
lat, label by city, on the globe."*

Every arc bends with the surface. The count above the plot names one table, the
coastlines, because a table is counted only when the view hides some of its
rows. Every city and every arc faces this view. Dragging the globe in the web
edition turns those coastlines into view. A network between places needs
nothing new from the grammar. The places carry the positions, `group()` carries
the connections, and the sphere carries the shortest ways between them.

Looking up each city's longitude and latitude is the host language's job. The
spikes later in this chapter need the same kind of join, for each country's
center. What the sentence adds is only the layers: the routes, then the cities,
then their names.

## Meridians and parallels

A `rule` draws a line at one position and spans the axis it does not name. On
the map, a rule on latitude drew the equator as a straight line. Here the same
sentence bends it around the sphere:

```{r}
#| echo: false
peek(equator)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(equator) + rule + y(lat) + style(color = "#d4674f") +
  globe(turn = 20, tilt = 30) + title("The equator, bent around the sphere")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the equator: a rule, y is lat, on the globe."*

A rule on `y` is a parallel, a circle of constant latitude. A rule on `x` is a
meridian, and it runs pole to pole.

## The choropleth, kept round

A rule and a path draw lines on the sphere. A fill has to cover a whole region
instead, and the edge of the disk is where a fill is hardest to draw. The map
chapter filled each country by its continent, and the same sentence fills the
countries on the sphere:

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + globe(turn = 20, tilt = 5) +
  title("Every country, filled by continent")
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country, color
by continent, on the globe."*

Look at the countries that the edge of the disk cuts through. gog closes each
cut shape along that edge, so a half-visible country is still a filled region
rather than an open outline.

## A measure stands on the radius

On a flat map a quantity has to go on a channel, `size` or `color`, because both
positions carry the place. The sphere has a direction to spare: the radius,
which runs straight out from the surface. Every place on the sphere has its own
radius. A `bar` here stands at its place
and measures outward along that radius. It draws as a spike rising from the
surface, and `z` names the measure:

```{r}
#| echo: false
peek(population_spikes)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(population_spikes) + bar + x(lon) + y(lat) + z(reach) + color(continent) +
  globe(turn = 20, tilt = 12) +
  title("World population, standing on its places")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the spikes: bars, x is lon, y is lat, z is reach, color by continent, on
the globe."*

Look at the edge of the disk. The sphere itself clips a spike. A spike just
behind that edge has its base hidden, so only the part that rises clear of the
edge is drawn. One too short to clear it is not drawn at all. At the edge you
see a spike from the side, along its whole length. A spike facing you
foreshortens toward a point, which is why a plot like this is read by turning
it.

Finding a country's center and joining the population are the host language's
job, as the join was on the map. Shaping the measure belongs there too. The
`reach` column is the square root of the population, which keeps China's spike
from making every other spike too small to see. A scale parameter would only be
a second way of writing that one column.

## The depth goes to the cube

Each earthquake here carries a third number: `elevation`, its depth below the
surface. That number is not a length to raise from the surface. It is a third
coordinate of the place itself, and on the globe `z` belongs to the spike alone.
So gog refuses the sentence below rather than guessing at it. The refusal names
three ways forward: add a `bar` layer, drop `z()`, or draw the same table in the
cube.

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  globe(turn = 178, tilt = -18)
```

A place with a depth belongs in the cube, the space with a third position.
`space()` takes the same two arguments `globe()` does:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  color(slab) + space(turn = 130, tilt = 25) +
  title("The same earthquakes, read by depth")
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, color by
slab."*

`slab` names the depth band each quake falls in. The globe plots looked straight
down on these quakes, so a deep one and a shallow one landed in almost the same
place. Turned, they line up along the sinking sheet of ocean floor beneath them.
Drag this one too: the globe turns the earth, and the cube turns the cloud of
points, one gesture for both.

## What a globe refuses

The globe takes the same marks the map takes, and refuses the same ones for the
same reason. Suppose you try a `line` through the earthquakes, as you would on a
flat page. Both positions carry the place, exactly as on the map, so a mark that
measures along an axis has no axis left:

```{r}
#| error: true
data(quakes_fiji) + line + x(east) + y(north) + globe()
```

The quantity moves to a channel instead, `size()` or `color()`, or it stands on
the radius as a spike, the way the population spikes earlier in this chapter
already do.

A latitude past a pole names no place, so a view cannot face that latitude:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + globe(tilt = 100)
```

And a globe draws no axes at all. So gog refuses an axis label rather than
accepting it and drawing nothing:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + globe() + x_label("Longitude")
```

The map and the globe draw one meaning as two pictures. The map flattens the
sphere so a reader sees all of it at once. The globe keeps the sphere and hides
the far half. One word moves you between them.

## What comes next

A globe puts each row at the place its two numbers name. [Network](network.qmd)
draws the tables that hold no place at all, rows saying one thing is connected
to another, positioned by a layout rather than an axis.
