# Globe

```{r}
#| include: false
source("R/setup.R")
```

What does the data look like on the earth itself? The map chapter flattened
the sphere to fit a page. This chapter keeps it round, and turns it toward
you.

The sentence is the map chapter's first one, with the space swapped:

```{r}
(data(quakes_fiji) + point + x(east) + y(north) + map() +
   title("map()")) |
  (data(quakes_fiji) + point + x(east) + y(north) +
     globe(turn = 178, tilt = -18) + title("globe(turn = 178, tilt = -18)"))
```

*"Given quakes Fiji: points, x is east, y is north, on the globe."*

`x` is longitude and `y` is latitude, exactly as on the map. What changed is
what a position means on the page. Each point now stands at its place on the
sphere, and the sphere is drawn as the disk you would see looking at it.

## The view

A globe shows one half of the earth, so which half matters. `turn` and `tilt`
name the place the view faces, in the words `space()` already uses: `turn` is
the longitude at the center of the disk, and `tilt` is the latitude. The plot
above faces Fiji. Leave them out and the view faces zero and zero, where the
equator crosses the prime meridian.

`turn` is a bearing, so any number is a view: 190 and -170 face the same
place. `tilt` is a latitude, and it stops at the poles.

In the web edition the written angle is only where the globe starts. Drag it
and it turns, the way the cube in [Space](space.qmd) turns, and the far half
of the earth comes around; *reset* returns to the view the sentence asked
for. On paper a plot keeps its one view, which is why this chapter also
writes second views out.

Half the earth faces away, and data on that half is hidden behind the sphere.
gog counts what a view hides rather than dropping it in silence:

```{r}
#| message: true
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) +
  style(color = "tomato", opacity = 0.35, size = 3) +
  globe(turn = 178, tilt = -18) + title("The map chapter's layers, kept round")
```

Read the message above the plot. The coastline table covers the whole earth,
so most of its rows face away from any one view, while every earthquake faces
this one. Turning the globe brings the hidden rows around:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  globe(turn = 20, tilt = 5) + title("The other side")
```

## The graticule

The gray lines on the sphere are the **graticule**: a meridian every thirty
degrees of longitude, and a parallel every thirty degrees of latitude. They
are this space's gridlines, the same lines every panel draws, bent onto the
sphere the way polar bends them into rings and spokes. There are no axes
beside the disk, because a sphere has no edge to write one on; the graticule
is the reference instead. `theme(grid = )` reaches it like any other grid:

```{r}
data(quakes_fiji) + point + x(east) + y(north) +
  globe(turn = 178, tilt = -18) + theme(grid = "none") +
  title("No graticule")
```

## A route bends with the sphere

On the map, `path` joins places with straight strokes. On the globe a straight
route does not exist. The shortest way between two places bends with the
surface, along a great circle, and a `path` here follows it:

```{r}
flight <- data.frame(lon = c(139.69, -118.24), lat = c(35.69, 34.05))

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(flight) + path + x(lon) + y(lat) +
  style(color = "#3d5a80", arrow = "end") +
  globe(turn = 178, tilt = -18) + title("Tokyo to Los Angeles")
```

Two rows describe that route, and the curve between them is the sphere's own.
The arc leans toward the pole because the shortest way across an ocean does.

## Many routes are a network

One table can hold every route at once. Give each route a name, give each of
its two rows a place, and `group()` splits the table into one arc per route,
exactly as it split the coastline into pieces. A `point` layer marks the
cities and a `text` layer names them. The picture this draws is a route
network, and it needs no new word, because every position in it is a place
the data already knows:

```{r}
#| message: true
cities <- data.frame(
  city = c("Seoul", "Tokyo", "Singapore", "Delhi", "Cairo",
           "London", "Anchorage", "Sydney"),
  lon  = c(126.98, 139.69, 103.82, 77.21, 31.24, -0.13, -149.90, 151.21),
  lat  = c(37.57, 35.68, 1.35, 28.61, 30.04, 51.51, 61.22, -33.87)
)
legs <- data.frame(
  route = rep(c("Seoul-Tokyo", "Seoul-Singapore", "Seoul-Delhi",
                "Delhi-Cairo", "Cairo-London", "Tokyo-Anchorage",
                "Singapore-Sydney"), each = 2),
  city  = c("Seoul", "Tokyo", "Seoul", "Singapore", "Seoul", "Delhi",
            "Delhi", "Cairo", "Cairo", "London", "Tokyo", "Anchorage",
            "Singapore", "Sydney")
)
routes <- merge(legs, cities, by = "city", sort = FALSE)

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(routes) + path + x(lon) + y(lat) + group(route) +
  style(color = "#3d5a80", opacity = 0.6) +
  data(cities) + point + x(lon) + y(lat) +
  data(cities) + text + x(lon) + y(lat) + label(city) + style(nudge = "up") +
  globe(turn = 95, tilt = 30) + title("Routes between cities")
```

Looking up each city's longitude and latitude is the host language's job,
the same join the spike section below does for country centers. What the
sentence adds is only the layers: routes, cities, names, each reading its
own table.

Every arc bends with the surface, and the count above the plot is the same
honesty as before: the far side's coastlines are hidden, every city faces
this view, and dragging the globe in the web edition brings the rest around.
This is worth a moment, because a network between places asks nothing new of
the grammar. The places carry the positions, `group()` carries the
connections, and the sphere carries the shortest ways between them.

## Meridians and parallels

A `rule` holds one position and spans the axis it does not name. On the map
that drew the equator as a straight line. Here the same sentence draws it all
the way around:

```{r}
lines_of <- data.frame(lat = 0)

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(lines_of) + rule + y(lat) + style(color = "#d4674f") +
  globe(turn = 20, tilt = 30) + title("The equator, held whole")
```

A rule on `y` is a parallel, a circle of constant latitude. A rule on `x` is a
meridian, and it runs pole to pole.

## The choropleth, kept round

The map chapter filled each country by a value. The same sentence fills the
countries on the sphere:

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + globe(turn = 20, tilt = 5) +
  title("Every country, filled by continent")
```

Look at the countries the edge of the disk cuts through. A shape the horizon
cuts is closed again along it, so a half-visible country is still a filled
region rather than a shape left open.

## A measure stands on the radius

The flattened map has no axis to spare. The sphere has exactly one: the
radius, pointing away from every place. A `bar` here stands at its place and
measures outward along it, a spike from the surface, and `z` names its
measure:

```{r}
centers <- do.call(rbind, lapply(split(world_borders[, c("lon", "lat")],
                                       world_borders$country), colMeans))
spikes <- data.frame(country = rownames(centers), centers)
matched <- match(spikes$country, gapminder_2007$country)
spikes$people <- gapminder_2007$population[matched]
spikes$continent <- gapminder_2007$continent[matched]
spikes <- spikes[!is.na(spikes$people), ]
spikes$reach <- sqrt(spikes$people)

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(spikes) + bar + x(lon) + y(lat) + z(reach) + color(continent) +
  globe(turn = 20, tilt = 12) +
  title("World population, standing on its places")
```

Finding a country's center and joining the population are the host language's
job, as the join was on the map. So is shaping the measure: the square root
keeps China from dwarfing every other spike, and it is one line in R where a
scale parameter would be a second way of writing it.

Look at the rim. A spike is clipped by the sphere itself, so one standing just
behind the horizon still peeks over the limb when it is tall enough, and the
tallest spikes read in profile at the edge of the disk. A spike facing you
foreshortens toward a point, which is why a plot like this is read by turning
it.

## The depth goes to the cube

Each earthquake here carries a third number: `elevation`, its depth below the
surface. That is not a length to raise from the surface but a third place
coordinate, and on the globe `z` belongs to the spike alone. So gog refuses
the sentence below rather than guessing at it, and the red message is the
point of this chunk: it prints live as the page builds, and it names both
ways out.

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  globe(turn = 178, tilt = -18)
```

For a place with a depth, the cube is the space with a third position, and
the same table draws there, `space()` naming its view with the words `globe()`
uses:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  color(slab) + space(turn = 130, tilt = 25) +
  title("The same earthquakes, read by depth")
```

Drag this one too: the globe turns the earth, and the cube turns the cloud,
one gesture for both. The epicenters this chapter has drawn are the view from
above of the slab this plot shows sinking.

## What a globe refuses

Both positions are spent on the place, exactly as on the map, so a mark that
reads a domain along an axis has none to read:

```{r}
#| error: true
data(quakes_fiji) + line + x(east) + y(north) + globe()
```

The quantity moves to a channel instead, `size()` or `color()`, or it stands
on the radius as a spike, as the section above drew.

A latitude past a pole is not a place, so a view cannot face one:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + globe(tilt = 100)
```

And a globe draws no axes at all, so naming one is refused rather than
accepted and dropped:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + globe() + x_label("Longitude")
```

The map and the globe are one meaning with two pictures of it. The map trades
the sphere for a page a reader sees whole. The globe keeps the sphere and
trades away the far half. Moving between them costs one word.
