# Globe

```{r}
#| include: false
source("R/setup.R")
```

What does the data look like on the earth itself? The map chapter flattened
the sphere to fit a page. This chapter keeps the sphere round, and turns it
toward you.

`quakes_fiji` holds a thousand earthquakes recorded near Fiji, one row for each.
Here is the map chapter's first sentence, which draws them, and beside it the
same sentence with `globe()` in place of `map()`, turned to face Fiji:

```{r}
#| echo: false
peek(quakes_fiji)
```

```{r}
(data(quakes_fiji) + point + x(east) + y(north) + map() +
   title("map()")) |
  (data(quakes_fiji) + point + x(east) + y(north) +
     globe(turn = 178, tilt = -18) + title("globe(turn = 178, tilt = -18)"))
```

*"Given quakes Fiji: points, x is east, y is north, in the map space, beside the
same on the globe."*

`x` is longitude and `y` is latitude, exactly as on the map. What changed is how
the two numbers become a position on the page. Each point now stands at its
place on the sphere. A sphere seen from one side is a disk, and that disk is
what the plot draws.

## The view

A globe shows one half of the earth, so the choice of half matters. `turn` and
`tilt` name the place the view faces, and they are the same two arguments
`space()` takes. `turn` is the longitude at the center of the disk, and `tilt`
is the latitude. The chapter's first plot faces Fiji. Leave both out and the
view faces longitude zero and latitude zero, where the equator crosses the prime
meridian.

`turn` is a longitude, so 190 and -170 name the same view: 190 degrees east is
170 degrees west. `tilt` is a latitude, so it runs from -90 to 90 and no
further.

In the web edition the angle you write is only the view the globe opens at. Drag
it and it turns, the way the cube in [Space](space.qmd) turns, and the far half
of the earth comes into view. The *reset* button returns to the view the
sentence asked for. On paper a plot keeps its one view, which is why this
chapter also draws some data at a second angle.

## What a view hides

Half the earth faces away, and data on that half is hidden behind the sphere.
gog counts what a view hides rather than dropping it in silence. The next plot
draws the earthquakes over `world_borders`, the map chapter's table of the
points that trace every country's outline:

```{r}
#| echo: false
peek(world_borders)
```

```{r}
#| message: true
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) +
  style(color = "tomato", opacity = 0.35, size = 3) +
  globe(turn = 178, tilt = -18) + title("The map chapter's layers, kept round")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given quakes Fiji: points, x is east, y is north, on the globe."*

Read the message above the plot. Its count is out of 4150, the rows of
`world_borders`. That table covers the whole earth, so any one view hides some
of it. From this view most of its rows are on the far side of the earth, and
every earthquake is on the near side. A second angle
brings most of the hidden rows into view:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  globe(turn = 20, tilt = 5) + title("The other side")
```

## The graticule

Every globe so far has drawn a grid of thin gray lines over the sphere. It is
the **graticule**, this space's gridlines. Its **meridians** run from pole to
pole, one every thirty degrees of longitude, and its **parallels** circle the
earth, one every thirty degrees of latitude. A flat panel draws its gridlines
straight. A globe bends its gridlines onto the sphere, the way the
[polar](polar.qmd) space bends them into rings and spokes.

There are no axes beside the disk, because a sphere has no straight side to
write one along. The graticule is the reference instead. `theme(grid = )` works
on it as on any other grid: `"x"` keeps only the meridians, `"y"` keeps only the
parallels, and `"none"` removes it:

```{r}
data(quakes_fiji) + point + x(east) + y(north) +
  globe(turn = 178, tilt = -18) + theme(grid = "none") +
  title("No graticule")
```

With the graticule off, nothing inside the disk shows where the poles are, or
even that it is a sphere.

## A route bends with the sphere

The graticule bends with the sphere, and so does everything else drawn on it. A
route is the clearest case. On the map, `path` joins places with straight
strokes. On the globe a straight route does not exist. The shortest way between
two places bends with the surface, along a **great circle**: the line a tight
string would take if you pulled it across the sphere between the two places. A
`path` here follows it:

```{r}
#| echo: false
peek(flight)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(flight) + path + x(lon) + y(lat) +
  style(color = "#3d5a80", arrow = "end") +
  globe(turn = 178, tilt = -18) + title("Tokyo to Los Angeles")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the flight: a path, x is lon, y is lat, on the globe."*

Two rows describe that route, and the curve between them comes from the sphere.
The arc bends toward the pole because that way is shorter. A route that keeps to
one latitude is longer, unless that latitude is the equator.

## Many routes are a network

That table held one route. One table can hold every route at once, and the
sentence barely changes. Give each route a name, and give each of its two rows
a longitude and a latitude. `group()` then splits the table into one arc per
route, exactly as it split the world borders into pieces. A `point` layer marks
the cities and a `text` layer names them. The picture this draws is a route
network:

```{r}
#| echo: false
peek(routes)
peek(cities)
```

```{r}
#| message: true
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(routes) + path + x(lon) + y(lat) + group(route) +
  style(color = "#3d5a80", opacity = 0.6) +
  data(cities) + point + x(lon) + y(lat) +
  data(cities) + text + x(lon) + y(lat) + label(city) + style(nudge = "up") +
  globe(turn = 95, tilt = 30) + title("Routes between cities")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the routes: paths, x is lon, y is lat, grouped by route; then given the
cities: points, x is lon, y is lat; then given the cities: text, x is lon, y is
lat, label by city, on the globe."*

Every arc bends with the surface. The message above the plot counts the rows of
one table only, the world borders, because gog prints a count only for a layer
with rows the view hides. Every city and every arc faces this view. Dragging the
globe in the web edition brings the hidden borders into view. A network
between places needs nothing new from the grammar. The places give the
positions, `group()` says which places are joined, and the sphere gives the
shortest way between each pair.

Finding each city's longitude and latitude and joining them to the routes is
the host language's job, as joining two tables was in the map chapter. What the
sentence adds is only the layers: the routes, then the cities, then their names.

## Meridians and parallels

The graticule draws the equator in the same thin gray as every other parallel.
To give it a color of its own, draw it as a `rule`. A `rule` draws a line at one
position and spans the other axis. On the map, a rule on latitude drew the
equator as a straight line. The same
sentence here draws it bent around the sphere:

```{r}
#| echo: false
peek(equator)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(equator) + rule + y(lat) + style(color = "#d4674f") +
  globe(turn = 20, tilt = 30) + title("The equator, bent around the sphere")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the equator: a rule, y is lat, on the globe."*

A rule on `y` draws a parallel, as here, and a rule on `x` draws a meridian.

## The choropleth, kept round

A rule and a path draw lines on the sphere. A fill has to cover a whole region
instead, and the edge of the disk is where a fill is hardest to draw. A map that
fills each region by a column of the table is a **choropleth**. The map chapter
filled each country by its continent, and the same sentence fills the countries
on the sphere:

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + globe(turn = 20, tilt = 5) +
  title("Every country, filled by continent")
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country, color
by continent, on the globe."*

Look at the countries that the edge of the disk cuts through. gog closes each
cut shape along that edge, so a half-visible country is still a filled region
rather than an open outline.

## Thousands of storm tracks on a dark globe

The routes earlier in this chapter were a few arcs between cities. A table of
real records can hold thousands of lines, and then the lines are hard to tell
apart. Thin lines drawn over one another on a pale sphere merge into one gray
mass, and the lines inside it can no longer be followed.

`cyclones` is a table of that size. It holds all 2,174 storms from 1980 to 2025
that reached hurricane strength, a wind of at least 64 knots (119 kilometers per
hour). The US National Oceanic and Atmospheric Administration (NOAA) keeps the
record. Hurricanes, typhoons and cyclones are one kind of storm, with a
different name in each ocean. Scientists call all three tropical cyclones, which
gives the table its name.

Each row is one storm's position at one moment, and a storm's positions joined
in order are its **track**. The positions are twelve hours apart. They cover
only the time when the storm's wind was at least 34 knots, or 63 kilometers per
hour. At that speed a storm is given a name. `category` is the storm's highest
category on the Saffir-Simpson scale, which ranks a storm from 1 to 5 by its
strongest wind. It holds one value for each storm, in three bands: "1 or 2",
"3 or 4" and "5". The `month` column is not used in this chapter. The
[Play](play.qmd) chapter uses it to show the same storms one calendar month at a
time.

```{r}
#| echo: false
peek(cyclones)
```

Three settings give the globe its dark look. `theme(background = )` colors the
sphere itself, here almost black, and `grid = "none"` removes the graticule. The
land is the same `zone` layer as in the choropleth above. `style(color = )` sets
every country to one dark gray, so the continents stay darker than every track.
Bright lines on a dark sphere do not merge into gray, so the thin tracks stay
visible even where many of them cross:

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  style(color = "#1c2636") +
  data(cyclones) + path + x(lon) + y(lat) + group(storm) + color(category) +
  palette(c("mediumvioletred", "darkorange", "gold")) + style(opacity = 0.6) +
  globe(turn = 135, tilt = 10) + theme(background = "#050a14", grid = "none") +
  title("Every storm that reached hurricane strength, 1980 to 2025")
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country; then
given the cyclones: paths, x is lon, y is lat, grouped by storm, color by
category, with a palette of mediumvioletred, darkorange and gold, on the
globe."*

The busiest region is the western North Pacific, on both sides of the
Philippines. A third of the storms began there, 729 of the 2,174, and so did 125
of the 221 storms that reached category 5. In the north, most tracks run west,
and many then bend to the north and east. In the south, many bend to the south
and east. The band along the equator stays empty: no position lies closer to it
than 0.7 degrees, and only 33 of the 29,669 positions lie within 5 degrees of
it.

The gold tracks lie on top because of the table's order, not because of their
color. gog draws the storms in the order the table lists them, and a track drawn
later covers the tracks drawn before it. The table lists the category 5 storms
last. Reverse the rows, and the gold tracks go to the bottom, under all the
others.

## A measure stands on the radius

On a flat map a quantity has to go on a channel, `size` or `color`, because both
positions carry the place. The sphere leaves one direction free: the radius,
which runs straight out from the surface. Every place on the sphere has its own
radius. A `bar` here stands at its place and measures outward along that radius.
It draws as a spike rising from the surface, and `z` names the measure. Here the
measure is `reach`, the square root of each country's population:

```{r}
#| echo: false
peek(population_spikes)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(population_spikes) + bar + x(lon) + y(lat) + z(reach) + color(continent) +
  globe(turn = 20, tilt = 12) +
  title("World population, one spike per country")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the population spikes: bars, x is lon, y is lat, z is reach, color by
continent, on the globe."*

Look at the edge of the disk. The sphere itself can hide part of a spike. A
spike just behind that edge has its base hidden, so only the part that rises
past the edge is drawn. A spike too short to rise past that edge is not drawn at
all. At the edge you see a spike from the side, along its whole length. A spike
that points straight at you is seen from its tip, so it looks shorter than it
is. In the web edition, turn the globe to see such a spike from the side.

Finding each country's center and joining its population are the host
language's job, like the join for the cities above. Choosing how to scale the
measure belongs there too. The square root keeps China's spike from making every
other spike too small to see. A scale keeps an axis in the column's own units,
and a globe draws no axis, so the globe refuses `scale =` on `z()`.

## The depth goes to the cube

The Fiji earthquakes from the chapter's first plot carry a third number:
`elevation`, their depth below the surface, written as a negative number. That
number is not a height to raise above the surface. It is a third coordinate of
the place itself, and on the globe `z` belongs to `bar` alone.
So gog refuses the sentence below rather than guessing at it. The refusal names
three things you can do instead: add a `bar` layer, drop `z()`, or draw the same
table in the cube.

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  globe(turn = 178, tilt = -18)
```

A place with a depth belongs in the cube, the space with a third position.
`space()` takes the same two arguments `globe()` does:

```{r}
data(quakes_fiji) + point + x(east) + y(north) + z(elevation) +
  color(slab) + space(turn = 130, tilt = 25) +
  title("The same earthquakes, read by depth")
```

*"Given quakes Fiji: points, x is east, y is north, z is elevation, color by
slab."*

`slab` names the depth band each earthquake falls in. The globe plots looked
straight down on these earthquakes, so a deep one and a shallow one were drawn
at almost the same place. In the cube, depth is a third position, so they
separate. Most of them lie along one sloping sheet: the ocean floor, sinking
into the earth. Drag this one too. The drag that turns the earth on a globe
turns the cloud of points in the cube.

## What a globe refuses

The map and the globe draw the same places as two pictures. The map flattens the
sphere so a reader sees all of it at once. The globe keeps the sphere and hides
the far half. To go from one picture to the other, you change one word.

The globe takes the map's marks and one more: `bar`, which measures along the
radius. It refuses the rest for the same reason the map does. Suppose you try a
`line` through the earthquakes, as you would on a flat page. Both positions
carry the place, exactly as on the map, and only `bar` reads the radius. So a
`line` has no axis to measure along:

```{r}
#| error: true
data(quakes_fiji) + line + x(east) + y(north) + globe()
```

The refusal's advice is to put a quantity on a channel instead, `size` or
`color`. On a globe a quantity can also be a spike's length along the radius,
as in the population spikes earlier in this chapter.

A latitude past a pole names no place, so a view cannot face that latitude:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + globe(tilt = 100)
```

And a globe draws no axes at all. So gog refuses an axis label rather than
accepting it and drawing nothing:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + globe() + x_label("Longitude")
```

## What comes next

A globe can draw a row only at the place its two numbers name.
[Network](network.qmd) draws tables whose rows name no place, only a connection
between two nodes, and the engine computes a position for every node.
