# The grammar

```{r}
#| include: false
source("R/setup.R")
```

What do you have to say about your data before a plot can be drawn? gog's whole
vocabulary is a short list of ordinary English words. Learn them once and you
can read, and write, any plot.

## The kernel

| Kind | The words |
|---|---|
| **Tables** | `data` `query` |
| **Marks** | `point` `line` `area` `bar` `step` `interval` `box` `ribbon` `text` `path` `rule` `zone` `surface` |
| **Channels** | `x` `y` `z` `color` `size` `shape` `pattern` `opacity` `group` `label` `play` |
| **Transforms** | `bin` `smooth` `count` `density` `proportion` `sum` `mean` `median` `max` `min` `quantile` `range` `confidence` `deviation` `bounds` `partition` `flow`, and the four collision modifiers, `dodge` `stack` `jitter` `repel` |
| **Operators** | `+` layers, `*` derives, `|` and `/` arrange |
| **Scales** | `linear` `log` `time` `category` `order` |
| **Spaces** | `flat` `space` `polar` `nest` `globe` `map` |
| **Settings** | `style` `theme` `palette` |
| **Labels** | `title` `x_label` `y_label` `z_label` |
| **Facets** | `facet` |
| **Selections** | `brush` |

Every word in it is a common English word: no abbreviations, no acronyms, no
silent letters. A common word is often taken already. Several of these name
something in R, Python or Julia before gog is loaded, and
[the last section of this chapter](#names-the-host-language-already-uses) says
what happens then.

Read the kernel as two groups: what a plot needs, and what refines or arranges
it. Five kinds build a plot. A **table** supplies the rows and the columns. A
**mark** is the shape drawn for each row. A **channel** maps a column to
something you can see, such as a position or a color. A **transform** derives new
values from the columns you named. An **operator** joins the words into one
sentence.

The other six refine a plot or arrange it, and none of them is required. A
**scale** decides how a number becomes a position. A **space** is the surface the
plot is drawn on: a flat page, a cube, a circle, a map. A **setting** changes how
the plot looks without saying anything about the data, and a **label** names it.
A **facet** cuts one plot into panels. A **selection** picks rows out of a plot
that is already drawn.

So **the minimum sentence** is a table, a mark, and the positions that mark
needs:

```
data(table) + mark + x(column) + y(column)
```

Two promises hold across the whole kernel. The engine draws every word in it,
though not yet every *pairing* of words. And gog never accepts a word and then
quietly ignores it: a sentence it cannot draw is refused, and the refusal
names the rule. See
[Design laws](design-laws.qmd) for the difference between **illegal** (the
grammar forbids it) and **unsupported** (the grammar allows it; the engine
cannot draw it yet).

## The colors

Every specification in this book is colored by kind. The colors show the two
groups above; they are not decoration.

Four of the five kinds that build a plot are words, and each takes a color of its
own. The fifth is the operator, which is punctuation in most of these languages,
so it is set in heavier type instead of a color. The six kinds that refine or
arrange a plot share one muted color, because none of them is required.

Your own names keep the color of the text around them. That is the first thing
the coloring tells you: which words belong to gog, and which are yours.

```{r}
data(gapminder_2007) + bar * bin + x(life) + style(color = "tomato")
```

*"Given gapminder 2007: bars derived by bin, x is life, colored tomato."*

`data` names the table, `bar` is the mark, `bin` is the transform, and `x` is the
channel. Those four build the plot. `style` refines it, so it takes the quiet
color, and `gapminder_2007`, `life` and `"tomato"` are yours.

Inside `style`, the word `color` stays the color of the text. It names a property
here rather than a column, so it is not the `color` channel: a word you write as
an argument name is never one of gog's words.

One color says the same thing in every language. A highlighter cannot do this.
Each language sorts these words by its own rules, and the four disagree: R
calls `data` a function, Python marks it as nothing at all, and none of the
four can see that `point` is a mark. The four spellings of one specification
would be printed four different ways.

Each kind has a section below, in the order of the kernel above: the five that
build a plot first, then the six that refine or arrange one. Each section gives
the words and what they do, and names the chapter that teaches them in full.

## Tables: where the rows come from

Every sentence begins with a table, and there are two ways to name one.

| Table | Where the rows come from |
|---|---|
| `data` | A table already in memory |
| `query` | A table a database returns, named by SQL |

`query()` stands exactly where `data()` stands, and everything after it is
unchanged:

```
query(connection, "SELECT ...") + mark + x(column) + y(column)
```

The SQL stays inside `query()`. The columns after it are still bare names,
because a database returns a table, and every sentence begins with a table. See
[Data](data.qmd).

## Marks: the shape drawn for each row

A mark is the geometric form drawn for each row of data.

| Mark | What it draws |
|------|--------------|
| `point` | One glyph per row |
| `line` | Rows connected in *x order*: one value per x, read left to right |
| `bar` | Rectangle from baseline to value, vertical or horizontal |
| `area` | Filled region between a line and the baseline |
| `step` | A line that holds each value until it changes (cumulative curves, histogram outlines) |
| `interval` | A whisker spanning low→high at each x: error bars, ranges; needs a `range` transform |
| `box` | Distribution per group: box, median, whiskers to 1.5 times the interquartile range, outliers as dots; carries its own summary |
| `ribbon` | A filled band from a low boundary to a high one across x: confidence / spread bands; needs a `range` transform |
| `text` | A string at each position: its content comes from the `label` channel |
| `path` | Rows connected in *the table's order*: a route that may return the way it came; takes `style(arrow = )` |
| `rule` | A line at one position, spanning the other axis: reference lines, and rugs via `style(reach = "edge")` |
| `zone` | A shaded region: bounded where its sides are known, spanning the whole panel where they are not. Five things can fix the sides: a **categorical position** (each category gets its own slot: the tile plot), `bounds` (names them), `bin` (cuts them: the heatmap), `density` (cuts or traces them), and a **boundary** (the data draws them: the choropleth, in `map()`) |
| `surface` | A sheet above the x-y plane: one row per (x, y) crossing, with a face between each block of four, or one row per cut cell, drawn flat at its own height. The only mark that can be drawn in the cube by itself |

## Channels: a column made visible

A channel maps a data column to a visual property of the mark. Adding one word
to the sentence adds one thing the reader can see:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."* You
never asked for the legend. A mapped column earns one.

Channels do not all do the same work, and what separates them is what each
one *earns*:

| Channel | What it does | Earns |
|---------|-------------|-------|
| `x` | Horizontal position, from a numeric or categorical column | an axis |
| `y` | Vertical position, from a numeric or categorical column | an axis |
| `z` | Depth, the third position: it makes a plot 3-D. Drawn by `point`, `bar`, `interval`, `box`, `path` and `surface` | an axis |
| `color` | Fill or stroke color, from a categorical or continuous column | a legend |
| `size` | Radius, on `point` | a legend |
| `shape` | Glyph (circle, square, triangle) on `point` | a legend |
| `pattern` | Texture: a fill's hatch, a stroke's dash. Every mark but `point`, `text` and `surface` | a legend |
| `opacity` | Transparency, on `point` and `bar` | a legend |
| `group` | Splits a mark into one line per category, without encoding it. On `line`, `area`, `step`, `interval`, `box`, `ribbon`, `path`, `zone` and `surface` | nothing |
| `label` | The string a `text` mark draws, its content | nothing |
| `play` | Cuts the plot into frames and plays them in sequence, from a numeric or categorical column | a strip |

The **Earns** column sorts the channels into four groups, and a Korean
syllable shows how they fit together. It has three slots: an initial consonant,
then a vowel, then an optional final consonant. The mark is the initial
consonant, so the channels are the other two.

`x`, `y` and `z` are the **vowels**: a mark without one of them is silent, so
at least one is required, and each earns an axis to be read against.

The five that earn a legend are the **final consonants**, the optional ㅁ (*m*)
that turns 가 (*ga*) into 감 (*gam*). None is ever required, and each changes
what the plot says, which is exactly why each needs a key to decode it.

`group` and `label` are neither. `group` splits without encoding: on a `line`
it performs the very same split `color` does, one line per category, and
then stops. It is what is left of `color` once the decoding is taken away,
so it works like the space between words rather than like a letter. `label`
supplies a `text` mark's content rather than refining it.

`play` is the fourth kind, and it stands outside the syllable the way faceting
does. It does not refine one picture. It cuts the plot into frames and plays
them in order, so what it earns is a **strip** naming the frame on screen. See
[Play](play.qmd).

Not every channel applies to every mark: a `line` is drawn with a single
stroke, so it takes no per-row `size` or `opacity`. gog states the rule rather
than guessing; see [Design laws](design-laws.qmd), and
[Combinations](combinations.qmd) for the whole mark-by-channel grid on one page.

## Transforms: values the data does not hold

A transform derives new values from the columns you named, and attaches to a mark
with `*`. The mark stays what it is: `bar * bin` is still a bar, drawn over
counted bands.

Eleven of them replace values with a computed statistic.

| Transform | What it computes |
|---|---|
| `bin` | Cuts a continuous column into bands and counts the rows in each |
| `count` | Tallies the rows in each category |
| `proportion` | The same tally, as a share of the whole |
| `density` | A smooth curve of where the rows lie |
| `smooth` | A trend line fitted through the points |
| `sum` `mean` `median` `max` `min` | Reduce a column to one value per group |
| `quantile` | Reduce a column to the value that a given share of the rows fall below, `quantile(0.9)` |

Four produce a low and a high together, which is what a span needs:

| Transform | What it computes |
|---|---|
| `range` | A band between two quantiles, the whole group unless you name them |
| `confidence` | The interval around the mean, which shows how exactly the mean is known |
| `deviation` | A band of standard deviations around the mean, which is how spread the data is |
| `bounds` | A low/high pair the table already holds, in two named columns |

Four are the **collision modifiers**: they compute nothing, and they move
marks that would otherwise land on top of each other:

| Transform | What it does |
|---|---|
| `dodge` | Puts them side by side |
| `stack` | Piles them on top of one another |
| `jitter` | Spreads them sideways along a category |
| `repel` | Moves labels off one another when they overlap |

And two transforms lay out a whole picture. `partition` divides a whole among
nested parts, which is what a treemap and a mosaic are built from. `flow` lays
a magnitude through its stages, which is what a flow diagram is built from.

See [Transforms](transforms.qmd) for each one in detail, and
[Combinations](combinations.qmd) for which marks take which.

## The `+` operator: layering

`+` joins one word to the next. Every sentence in this book is built with it,
and the words can be read left to right in the order they were written.

```{r}
# One layer
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."* One mark, one layer.

```{r}
# Two layers: bars behind, points on top
data(actuals) + x(year) + y(sales) +
  bar + style(color = "lightsteelblue") +
  point + style(color = "steelblue", size = 6)
```

Each mark opens a new layer.
Layers are rendered in order: first mark drawn first (at the bottom).

Order appears more than once in this grammar, and it does not mean the same
thing every time. [Which orders matter](operators.qmd#which-orders-matter)
gathers all of it in one place: what you may rearrange freely, what the engine
refuses, and the two arrangements that change the picture with no warning.

## The `*` operator: combining

`*` binds a mark to a transform, creating a derived mark type.
The mark is always on the left; the transform is on the right. Where two
transforms follow one mark, write them in the order they happen:
`bar * bin * mean` cuts the axis into bands, then averages inside each one.

```r
bar * bin      # histogram
line * smooth  # smoothed trend line
zone * bin     # heatmap: the same `bin`, cutting both axes
```

The last line shows the pattern: one transform means one thing, and the mark
decides where it applies. A `bar` keeps one axis to measure the count along,
so `bin` cuts the other one. A `zone` measures nothing by length and needs an
extent on both axes, so the same word cuts both, and the count becomes the
color. When one of a zone's positions is categorical, `bin` cuts only the
other one, because a category is already cut into slots.

The summaries follow the same rule, read in the other direction. A summary
groups the rows by every position the mark does *not* measure with. It then
reduces the named column on the position the mark does measure with: `y` on a
bar, `color` on a zone, `z` in [`space()`](space.qmd). So
`zone * mean + x(a) + y(b) + color(v)` averages `v` inside every cell two
categories cross.

See [Operators](operators.qmd) for the full rules,
and [Transforms](transforms.qmd) for each transform in detail.

## The `|` and `/` operators: faceting, or composing

`|` puts things side by side and `/` puts one above the other. Whether those are
panels or whole plots is decided by what you write on the right.

With `facet()` on the right they split *one* plot into panels, one per category
of the column it names:

```{r}
# One panel per continent, side by side
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

Every panel shares one scale, that is what makes them comparable, and both
operators together cross into a grid. See [Faceting](faceting.qmd) for the
full rules.

With another *plot* on the right they arrange the two on one page, each keeping
its own coordinate space:

```{r}
# Two plots, side by side
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

Composed plots share nothing until they name the same column on the same axis,
and then they share that axis outright: one scale, one extent, drawn once. That
single rule is what makes a marginal plot (a scatter with a small distribution
drawn along each edge) two operators rather than a chart type.
See [Composition](composition.qmd).

## Scales: how a number becomes a position

A scale is written on the **binding**, the channel with its column in it. It is
not a word of its own, because it belongs to one channel and nothing else can
combine with it:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life."*

| Scale | What it does |
|-------|-------------|
| `linear` | Equal steps for equal differences. The default |
| `log` | Equal steps for equal ratios, on `x`, `y`, `color`, `size` and `opacity`, with any `base` |
| `time` | Positions dates and times, chosen from the column type: `Date` and `POSIXct` |
| `category` | One slot per distinct value, chosen from the column type: text and factor columns |
| `order` | Which way a categorical axis runs, or [declare it on the column](data.qmd): a factor in R, `ordered()` in the other three |

A scale is not a transform: it produces no new values, only different positions
for the ones you have. Where the two meet, what a log scale means for
`bar * bin` or `bar * sum`, is set out in [Scales](scales.qmd).

The same binding takes two more words for the same reason. `limits` says what
range the channel runs over when the data is not the authority on that, and
`tick_count` says how densely an axis should be labeled. Both describe the
scale, so both are written where the scale is written. [Scales](scales.qmd)
has all three.

## Spaces: the surface a plot is drawn on

A space decides what a position means. Every plot so far has been drawn on the
plane. That space is `flat`, and you never have to write it.

| Space | Where the marks stand |
|---|---|
| `flat` | The plane: `x` across, `y` up. The default |
| `space` | The cube: `x`, `y` and `z`, projected onto the page |
| `polar` | The plane bent into a circle: one position becomes an angle, the other a radius |
| `nest` | Rectangles inside rectangles, each sized by its share of the one holding it |
| `globe` | The earth itself, viewed: marks stand on the half of the sphere that faces you |
| `map` | The earth flattened: longitude across, latitude up |

One space holds the whole plot. See [Space](space.qmd), [Polar](polar.qmd),
[Nest](nest.qmd), [Map](map.qmd) and [Globe](globe.qmd).

## Settings: values you *set* rather than *map*

Every channel above answers a question about the data. `color(species)` asks
"which species?", and earns a legend so the reader can answer it.

Making every point one color asks nothing. It maps no column, needs no scale,
and has nothing to decode, so it is not a channel. That is `style`:

| Setting | What it fixes |
|---------|--------------|
| `style(color = )` | One color for the whole layer |
| `style(opacity = )` | One opacity, `0`–`1` |
| `style(size = )` | Point radius, or line stroke width |
| `style(shape = )` | One glyph for every point |
| `style(pattern = )` | The texture of a mark's paint: a stroke's dash (solid/dashed/dotted), a fill's hatch (hatch/crosshatch/grid/dots). Every mark but `point`, `text` and `surface` |
| `palette( )` | Which colors the `color` channel uses |
| `theme( )` | The page around the plot: grid, background, fonts, size |

```{r}
# map: color carries the continent, and a legend decodes it
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)
```

```{r}
# set: one color, no question asked, no legend drawn
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "tomato", opacity = 0.5)
```

See [Setting vs mapping](style.qmd) for the full rules.

## Labels: the words on the plot

An axis carries the name of the column bound to it. A label replaces that name
when the column name is not what a reader should see.

| Label | What it names |
|---|---|
| `title` | The plot |
| `x_label` `y_label` `z_label` | One axis each |

`title()` is the one that adds words rather than replacing them, because a plot
has no other way to say what it is about.

## Facets: one panel per category

| Facet | What it cuts |
|---|---|
| `facet` | Splits one plot into panels, one per category of the column it names |

A facet is written on the right of `|` or `/`, which is why it appeared in the
operator section above: `… | facet(continent)`. See [Faceting](faceting.qmd).

## Selections: choosing rows in a drawn plot

| Selection | What it does |
|---|---|
| `brush` | Dims the rows you did not pick, and removes none of them |

A selection is the only kind that acts on a plot after it is drawn. What it does
not change is the sentence, so the printed page still shows the selection that
sentence named. See [Selection](selection.qmd).

## Names the host language already uses

Every word in the kernel is an ordinary English word, and ordinary words are
already taken. `mean` names a statistic in every language that has one. `data`,
`order`, `range`, `text` and `box` all name something in R before gog arrives.
That is a consequence of choosing plain names rather than a sign that something
has gone wrong, and in ordinary code almost nothing changes.

Which name a bare word reaches is the host language's decision, and gog does not
override it. Each language decides differently, so the answer is different in
each one:

| Language | What the language does | To reach the other one |
|---|---|---|
| R | Nineteen kernel words collide with R names. Nine are objects, and a call skips them; ten are functions, and those do intercept the call | `base::order(x)`, `stats::density(x)` |
| Python | A star-import rebinds six builtins, because Python has no rule that protects them | `from builtins import sum` |
| Julia | Two modules exporting one name is an error, never a silent choice | `Base.sum([1, 2, 3])` |
| JavaScript | Nothing happens, because no name reaches your scope unless you import it | nothing to do |

R's case is the one to understand, because it looks the worst and matters the
least. A mark takes no argument, and neither does a statistic like `mean` or
`max`, so those words are plain objects rather than functions. When R evaluates
a call it looks for a *function* of that name, and skips any binding that
is not one. Base R therefore still answers `mean(c(1, 2, 3))`, and
`sapply(df, mean)` works for the same reason. The message R prints when the
package loads names every collision and cannot show you this distinction, which
is why the message looks worse than the problem is.

The words that take an argument are functions, and a function does intercept the
call. `range(0.25, 0.75)` names a band, so base R's `range()` needs its prefix.
Seven of the ten report the mistake instead of returning a wrong answer.
Handed a vector where a probability belongs, `range()` says which function you
reached and names the other one. Three cannot: `box()`, `data(mtcars)` and
`title("...")` are well formed under both readings, and a refusal needs a
mistake the engine can detect.

The collision has no good ending. Prefixing the vocabulary, `gog_mean` and
`gog_range`, would remove the reason for the vocabulary: a reader who has to
learn a prefix learns every word twice, and none of the words is English any
more. Renaming `order` to `sort_by` would take away a word every reader
already knows. So the words stay, and the few places that need a qualified
name write one.

Each binding chapter carries the full list for its language, the rule above in
code you can run, and the spellings that still fail. R's are five, and every one
of them gives a word to one function, which passes it to a second, as
`aggregate(v ~ g, df, mean)` does. See [R](bindings/r.qmd),
[Python](bindings/python.qmd), [Julia](bindings/julia.qmd) and
[JavaScript](bindings/javascript.qmd).

The next two chapters put this vocabulary to work: first you read ten plots,
then you write your own.
