# The grammar

```{r}
#| include: false
source("R/setup.R")
```

What do you have to say about your data before a plot can be drawn? gog's whole
vocabulary is a short list of ordinary English words. Learn them once and you
can read, and write, any plot.

## The kernel

| Kind | The words |
|---|---|
| **Tables** | `data` `query` |
| **Marks** | `point` `line` `area` `bar` `step` `interval` `box` `ribbon` `text` `path` `rule` `zone` `surface` |
| **Channels** | `x` `y` `z` `color` `size` `shape` `pattern` `opacity` `group` `label` `play` |
| **Transforms** | `bin` `smooth` `count` `density` `proportion` `sum` `mean` `median` `max` `min` `range` `confidence` `bounds` `partition`, and the three that resolve collisions, `dodge` `stack` `jitter` |
| **Operators** | `+` layers, `*` derives, `|` and `/` arrange |
| **Scales** | `linear` `log` `time` `category` `order` |
| **Spaces** | `flat` `space` `polar` `nest` `map` |
| **Settings** | `style` `theme` `palette` |
| **Labels** | `title` `x_label` `y_label` `z_label` |
| **Facets** | `facet` |
| **Selections** | `brush` |

Every word in it is a common English word: no abbreviations, no acronyms, no
silent letters.

Read the kernel as two groups: what a plot needs, and what refines or arranges
it. Five kinds build a plot. A **table** supplies the rows and the columns. A
**mark** is the shape drawn for each row. A **channel** maps a column to
something you can see, such as a position or a color. A **transform** derives new
values from the columns you named. An **operator** joins the words into one
sentence.

The other six refine a plot or arrange it, and none of them is required. A
**scale** decides how a number becomes a position. A **space** is the surface the
plot is drawn on: a flat page, a cube, a circle, a map. A **setting** changes how
the plot looks without saying anything about the data, and a **label** names it.
A **facet** cuts one plot into panels. A **selection** picks rows out of a plot
that is already drawn.

So **the minimum sentence** is a table, a mark, and the positions that mark
needs:

```
data(table) + mark + x(column) + y(column)
```

Two promises hold across the whole kernel. The engine draws every word in it.
And gog never accepts a word and then quietly ignores it: a sentence it cannot
draw is refused, and the refusal names the rule. See
[Design laws](design-laws.qmd) for the difference between **illegal** (the
grammar forbids it) and **unsupported** (the grammar allows it; the engine
cannot draw it yet).

Each kind has a section below, in the order of the kernel above: the five that
build a plot first, then the six that refine or arrange one. Each section gives
the words and what they do, and names the chapter that teaches them in full.

## Tables: where the rows come from

Every sentence begins with a table, and there are two ways to name one.

| Table | Where the rows come from |
|---|---|
| `data` | A table already in memory |
| `query` | A table a database returns, named by SQL |

`query()` stands exactly where `data()` stands, and everything after it is
unchanged:

```
query(connection, "SELECT ...") + mark + x(column) + y(column)
```

The SQL stays inside `query()`. The columns after it are still bare names,
because a database returns a table, and every sentence begins with a table. See
[Data](data.qmd).

## Marks: the shape drawn for each row

A mark is the geometric form drawn for each row of data.

| Mark | What it draws |
|------|--------------|
| `point` | One glyph per row |
| `line` | Rows connected in **x order**: a function read along a domain |
| `bar` | Rectangle from baseline to value, vertical or horizontal |
| `area` | Filled region between a line and the baseline |
| `step` | A line that holds each value until it changes (CDFs, histogram outlines) |
| `interval` | A whisker spanning low→high at each x: error bars, ranges; needs a `range` transform |
| `box` | Distribution per group: box, median, whiskers to 1.5×IQR, outliers as dots; carries its own summary |
| `ribbon` | A filled band from a low boundary to a high one across x: confidence / spread bands; needs a `range` transform |
| `text` | A string at each position: its content comes from the `label` channel |
| `path` | Rows connected in **the table's order**: a route that may double back; takes `style(arrow = )` |
| `rule` | A line at one position, spanning the other axis: reference lines, and rugs via `style(reach = "edge")` |
| `zone` | A shaded region: bounded where its sides are known, spanning the panel where they are not. Five things know them: a **categorical position** (a category owns a slot: the tile plot), `bounds` (names them), `bin` (cuts them: the heatmap), `density` (cuts or traces them), and a **boundary** (the data draws them: the choropleth, in `map()`) |
| `surface` | A sheet over a floor whose cells tile: one row per (x, y) crossing with a face between each block of four, or one row per cut cell with a flat plateau on each. The only mark that draws in the cube alone |

## Channels: a column made visible

A channel maps a data column to a visual property of the mark. Adding one word
to the sentence adds one thing the reader can see:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."* You
never asked for the legend. A mapped column earns one.

Channels do not all play the same part, and what tells them apart is what each
one *earns*:

| Channel | What it does | Earns |
|---------|-------------|-------|
| `x` | Horizontal position, from a numeric or categorical column | an axis |
| `y` | Vertical position, from a numeric or categorical column | an axis |
| `z` | Depth, the third position: it makes a plot 3-D. Drawn by `point`, `bar`, `interval`, `box`, `path` and `surface` | an axis |
| `color` | Fill or stroke color, from a categorical or continuous column | a legend |
| `size` | Radius, on `point` | a legend |
| `shape` | Glyph (circle, square, triangle) on `point` | a legend |
| `pattern` | Texture: a fill's hatch, a stroke's dash. Every mark but `point`, `text` and `surface` | a legend |
| `opacity` | Transparency, on `point` and `bar` | a legend |
| `group` | Splits a mark into one series per category, without encoding it. On `line`, `area`, `step`, `interval`, `box`, `ribbon`, `path`, `zone` and `surface` | nothing |
| `label` | The string a `text` mark draws, its content | nothing |
| `play` | Cuts the plot into frames and plays them in sequence, from a numeric or categorical column | a strip |

The **Earns** column sorts the channels into four kinds, and a Korean syllable
is the shape to hold them against. It has three slots: an initial consonant,
then a vowel, then an optional final consonant. The mark is the initial
consonant, so the channels are the other two.

`x`, `y` and `z` are the **vowels**: a mark without one of them is silent, so
at least one is required, and each earns an axis to be read against.

The five that earn a legend are the **final consonants**, the optional ㅁ (*m*)
that turns 가 (*ga*) into 감 (*gam*). None is ever required, and each changes
what the plot says, which is exactly why each needs a key to decode it.

`group` and `label` are neither. `group` splits without encoding: on a `line`
it performs the very same split `color` does, one polyline per category, and
then stops. It is what is left of `color` once the decoding is taken away,
which makes it closer to the space between words than to a letter. `label`
supplies a `text` mark's content rather than refining it.

`play` is the fourth kind, and it stands outside the syllable the way faceting
does. It does not refine one picture. It cuts the plot into frames and plays
them in order, so what it earns is a **strip** naming the frame on screen. See
[Play](play.qmd).

Not every channel applies to every mark: a `line` is drawn with a single
stroke, so it takes no per-row `size` or `opacity`. gog states the rule rather
than guessing; see [Design laws](design-laws.qmd), and
[Combinations](combinations.qmd) for the whole mark-by-channel grid on one page.

## Transforms: values the data does not hold

A transform derives new values from the columns you named, and attaches to a mark
with `*`. The mark stays what it is: `bar * bin` is still a bar, drawn over
counted bands.

Ten of them replace values with a computed statistic.

| Transform | What it computes |
|---|---|
| `bin` | Cuts a continuous column into bands and counts the rows in each |
| `count` | Tallies the rows in each category |
| `proportion` | The same tally, as a share of the whole |
| `density` | A smooth curve of where the rows lie |
| `smooth` | A trend line fitted through the points |
| `sum` `mean` `median` `max` `min` | Reduce a column to one value per group |

Three produce a low and a high together, which is what a span needs:

| Transform | What it computes |
|---|---|
| `range` | A low/high pair computed from the data |
| `confidence` | The interval around the mean |
| `bounds` | A low/high pair the table already holds, in two named columns |

Three do not compute anything. They move marks that would otherwise land on top
of each other:

| Transform | What it does |
|---|---|
| `dodge` | Puts them side by side |
| `stack` | Piles them on top of one another |
| `jitter` | Spreads them sideways along a category |

And `partition` divides a whole among nested parts, which is what a treemap and a
mosaic are built from.

See [Transforms](transforms.qmd) for each one in detail, and
[Combinations](combinations.qmd) for which marks take which.

## The `+` operator: layering

`+` joins one word to the next. Every sentence in this book is built with it,
and the words can be read left to right in the order they were written.

```{r}
# One layer
data(gapminder_2007) + point + x(gdp) + y(life)
```

*"Given gapminder 2007: points, x is gdp, y is life."* One mark, one layer.

```{r}
# Two layers: bars behind, points on top
data(actuals) + x(year) + y(sales) +
  bar + style(color = "lightsteelblue") +
  point + style(color = "steelblue", size = 6)
```

Each mark opens a new layer.
Layers are rendered in order: first mark drawn first (at the bottom).

Order comes up more than once in this grammar, and it does not mean the same
thing every time. [Which orders matter](operators.qmd#which-orders-matter)
gathers all of it in one place: what you may rearrange freely, what the engine
refuses, and the two arrangements that change the picture without saying so.

## The `*` operator: combining

`*` binds a mark to a transform, creating a derived mark type.
The mark is always on the left; the transform is on the right. Where two
transforms follow one mark, write them in the order they happen:
`bar * bin * mean` cuts the axis into bands, then averages inside each one.

```r
bar * bin      # histogram
line * smooth  # smoothed trend line
zone * bin     # heatmap: the same `bin`, cutting both axes
```

The last line is the pattern worth noticing: one transform means one thing, and
the mark decides how it lands. A `bar` leaves an axis free to measure the count
along, so `bin` cuts the other one; a `zone` measures nothing by length and
needs an extent on both, so the same word cuts every axis with a width to cut
and the count becomes the color. On a zone with one categorical position it
cuts the other one only, since a category arrived already cut.

The summaries divide by the same arithmetic, read the other way round. `zone *
mean + x(a) + y(b) + color(v)` averages `v` within every cell two categories
cross, because a summary groups by every position the mark does *not* measure
with and reduces the column named on the one it does, which on a bar is `y`,
on a zone is `color`, and in [`space()`](space.qmd) is `z`.

See [Operators](operators.qmd) for the full rules,
and [Transforms](transforms.qmd) for each transform in detail.

## The `|` and `/` operators: faceting, or composing

`|` puts things side by side and `/` puts one above the other. Whether those are
panels or whole plots is decided by what you write on the right.

With `facet()` on the right they split *one* plot into panels, one per category
of the column it names:

```{r}
# One panel per continent, side by side
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

Every panel shares one scale, that is what makes them comparable, and both
operators together cross into a grid. See [Faceting](faceting.qmd) for the
full rules.

With another *plot* on the right they arrange the two on one page, each keeping
its own coordinate space:

```{r}
# Two plots, side by side
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

Composed plots share nothing until they name the same column on the same axis,
and then they share that axis outright: one scale, one extent, drawn once. That
single rule is what makes a marginal plot two operators rather than a chart type.
See [Composition](composition.qmd).

## Scales: how a number becomes a position

A scale is written on the **binding**, the channel with its column in it. It is
not an atom of its own, because it belongs to one channel and nothing else can
combine with it:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life."*

| Scale | What it does |
|-------|-------------|
| `linear` | Equal steps for equal differences. The default |
| `log` | Equal steps for equal ratios, on `x`, `y`, `color`, `size` and `opacity`, with any `base` |
| `time` | Positions dates and times, chosen from the column type: `Date` and `POSIXct` |
| `category` | One slot per distinct value, chosen from the column type: text and factor columns |
| `order` | Which way a categorical axis runs, or [declare it on the column](data.qmd): a factor in R, `ordered()` in the other three |

A scale is not a transform: it produces no new values, only different positions
for the ones you have. Where the two meet, what a log scale means for
`bar * bin` or `bar * sum`, is set out in [Scales](scales.qmd).

The same binding takes two more words for the same reason. `limits` says what
range the channel runs over when the data is not the authority on that, and
`tick_count` says how densely an axis should be labeled. Both describe the
scale, so both are written where the scale is written. [Scales](scales.qmd)
has all three.

## Spaces: the surface a plot is drawn on

A space decides what a position means. Every plot so far has been drawn on the
plane, which is `flat` and needs no saying.

| Space | Where the marks stand |
|---|---|
| `flat` | The plane: `x` across, `y` up. The default |
| `space` | The cube: `x`, `y` and `z`, projected onto the page |
| `polar` | The plane bent into a circle: one position becomes an angle, the other a radius |
| `nest` | Rectangles inside rectangles, each sized by its share of the one holding it |
| `map` | The earth flattened: longitude across, latitude up |

One space holds the whole plot. See [Space](space.qmd), [Polar](polar.qmd),
[Nest](nest.qmd) and [Map](map.qmd).

## Settings: values you *set* rather than *map*

Every channel above answers a question about the data. `color(species)` asks
"which species?", and earns a legend so the reader can answer it.

Making every point one color asks nothing. It maps no column, needs no scale,
and has nothing to decode, so it is not a channel. That is `style`:

| Setting | What it fixes |
|---------|--------------|
| `style(color = )` | One color for the whole layer |
| `style(opacity = )` | One opacity, `0`–`1` |
| `style(size = )` | Point radius, or line stroke width |
| `style(shape = )` | One glyph for every point |
| `style(pattern = )` | The texture of a mark's paint: a stroke's dash (solid/dashed/dotted), a fill's hatch (hatch/crosshatch/grid/dots). Every mark but `point`, `text` and `surface` |
| `palette( )` | Which colors the `color` channel hands out |
| `theme( )` | The page around the plot: grid, background, fonts, size |

```{r}
# map: color carries the continent, and a legend decodes it
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)
```

```{r}
# set: one color, no question asked, no legend drawn
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "tomato", opacity = 0.5)
```

See [Setting vs mapping](style.qmd) for the full rules.

## Labels: the words on the plot

An axis carries the name of the column bound to it. A label replaces that name
when the column name is not what a reader should see.

| Label | What it names |
|---|---|
| `title` | The plot |
| `x_label` `y_label` `z_label` | One axis each |

`title()` is the one that adds words rather than replacing them, because a plot
has no other way to say what it is about.

## Facets: one panel per category

| Facet | What it cuts |
|---|---|
| `facet` | Splits one plot into panels, one per category of the column it names |

A facet is written on the right of `|` or `/`, which is why it appeared in the
operator section above: `… | facet(continent)`. See [Faceting](faceting.qmd).

## Selections: choosing rows in a drawn plot

| Selection | What it does |
|---|---|
| `brush` | Dims the rows you did not pick, and removes none of them |

A selection is the only kind that acts on a plot after it is drawn. What it does
not change is the sentence, so the printed page still shows the bound that
sentence named. See [Selection](selection.qmd).
