# The grammar

```{r}
#| include: false
source("R/setup.R")
```

What do you have to say about your data before a plot can be drawn? GOG's whole
vocabulary is a short list of ordinary English words. Learn them once and you
can read, and write, any plot.

## The kernel

| Kind | The words |
|---|---|
| **Tables** | `data` `query` |
| **Marks** | `point` `line` `area` `bar` `step` `interval` `box` `ribbon` `text` `path` `rule` `zone` `surface` |
| **Channels** | `x` `y` `z` `color` `size` `shape` `pattern` `opacity` `group` `label` `play` |
| **Selections** | `click`⬜ `brush` |
| **Transforms** | `bin` `smooth` `count` `density` `proportion` `sum` `mean` `median` `max` `min` `range` `confidence` `bounds` `partition`, and the three that resolve collisions, `dodge` `stack` `jitter` |
| **Scales** | `linear` `log` `time` `category` `order` |
| **Spaces** | `flat` `space` `polar` `nest` `map`⬜ |
| **Labels** | `title` `x_label` `y_label` `z_label` |
| **Settings** | `style` `theme` `palette` |
| **Facets** | `facet` |
| **Operators** | `+` layers, `*` derives, `|` and `/` arrange |

⬜ designed, not drawn yet.

Every sentence is a table, then words from this list:

```
data(table) + mark + x(column) + y(column)
```

No abbreviations, no acronyms, no silent letters.

A table does not have to be in memory. `query()` stands exactly where `data()`
stands, and everything after it is unchanged:

```
query(connection, "SELECT ...") + mark + x(column) + y(column)
```

The SQL sits inside that one word and goes no further. The columns after it are
still bare names, because the database returned a table and a table is what the
grammar has always begun with. Writing SQL is not a different way of drawing;
it is a different way of naming the rows.

The kernel above is the *designed* vocabulary. The three words marked ⬜ are not
drawn yet, and they are not callable either, so your language will report a
missing function rather than GOG explaining itself. What GOG never does is
accept one of these names and quietly ignore it. Every **mark** draws today,
`surface` being the last to arrive, and `play` draws too. What is left is the
two selections and the `map` space.
See [Design laws](design-laws.qmd) for the difference between
**illegal** (the grammar forbids it) and **unsupported** (the grammar allows it;
the engine cannot draw it yet). Each table below marks what renders today.

This list gets shorter in two ways, and the second one is worth knowing about. A
word leaves it when the engine learns to draw it. A word can also leave it
because GOG decides not to have it. There used to be a third selection, `lasso`,
for drawing a free shape around some points. It was removed in August 2026. Every
other word here can be written down: `x(gdp, limits = c(0, 5e4))` states a range,
and a reader holding the printed book sees the plot that range asks for. A free
shape can only be written as a long list of corners, which is not something a
person writes in a sentence. A word GOG cannot print is a word GOG will not keep.

## Marks: the "consonants"

A mark is the geometric form drawn for each row of data.

| Mark | What it draws | Today |
|------|--------------|-------|
| `point` | One glyph per row | ✅ |
| `line` | Rows connected in **x order**: a function read along a domain | ✅ |
| `bar` | Rectangle from baseline to value | ✅ vertical or horizontal |
| `area` | Filled region between a line and the baseline | ✅ |
| `step` | A line that holds each value until it changes (CDFs, histogram outlines) | ✅ |
| `interval` | A whisker spanning low→high at each x: error bars, ranges; needs a `range` transform | ✅ |
| `box` | Distribution per group: box, median, whiskers to 1.5×IQR, outliers as dots; carries its own summary | ✅ |
| `ribbon` | A filled band from a low boundary to a high one across x: confidence / spread bands; needs a `range` transform | ✅ |
| `text` | A string at each position: its content comes from the `label` channel | ✅ |
| `path` | Rows connected in **the table's order**: a route that may double back; takes `style(arrow = )` | ✅ |
| `rule` | A line at one position, spanning the other axis: reference lines, and rugs via `style(reach = "edge")` | ✅ |
| `zone` | A shaded rectangle: bounded where its sides are known, spanning the panel where they are not. Four things know them — a **categorical position** (a category owns a slot: the tile plot), `bounds` (names them), `bin` (cuts them: the heatmap), `density` (cuts or traces them) | ✅ |
| `surface` | A sheet over a floor whose cells tile: one row per (x, y) crossing with a face between each block of four, or one row per cut cell with a flat plateau on each. The only mark that draws in the cube alone | ✅ |

## Channels: the "vowels" and the "final consonants"

A channel maps a data column to a visual property of the mark. They do not all
play the same part, and what tells them apart is what each one *earns*:

| Channel | Visual property | Earns | Today |
|---------|----------------|-------|-------|
| `x` | Horizontal position | an axis | ✅ numeric or categorical |
| `y` | Vertical position | an axis | ✅ numeric or categorical |
| `z` | Depth: the third position, makes a plot 3-D | an axis | ✅ six marks: `point` cloud, `path` route, `bar` column, `interval` span, `box` summary, `surface` sheet |
| `color` | Fill / stroke color | a legend | ✅ categorical *or* continuous |
| `size` | Radius | a legend | ✅ |
| `shape` | Glyph (circle, square, triangle…) | a legend | ✅ `point` |
| `pattern` | Texture: a fill's hatch, a stroke's dash | a legend | ✅ all but `point`, `text` and `surface` |
| `opacity` | Transparency | a legend | ✅ `point`, `bar`; as a *setting* it reaches every mark |
| `group` | Splits a mark into one series per category, without encoding it | nothing | ✅ eight marks: `line`, `step`, `area`, `ribbon`, `interval`, `box`, `path`, `surface` |
| `label` | The string a `text` mark draws (its content) | nothing | ✅ `text` |
| `play` | Cuts the plot into frames and plays them in sequence | a strip | ✅ numeric *or* categorical |

The **Earns** column sorts the channels into four kinds. A Korean syllable has
three slots: an initial consonant, then a vowel, then an optional final
consonant. The mark is the initial consonant, which is why this section's title
names only the other two.

`x`, `y` and `z` are the **vowels**: a mark without one of them is silent, so
at least one is required, and each earns an axis to be read against.

The five that earn a legend are the **final consonants**, the optional ㅁ (*m*)
that turns 가 (*ga*) into 감 (*gam*). None is ever required, and each changes
what the plot says, which is exactly why each needs a key to decode it.

`group` and `label` are neither. `group` splits without encoding: on a `line`
it performs the very same split `color` does, one polyline per category, and
then stops. It is what is left of `color` once the decoding is taken away,
which makes it closer to the space between words than to a letter. `label`
supplies a `text` mark's content rather than refining it.

`play` is the fourth kind, and it stands outside the syllable the way faceting
does. It does not refine one picture. It cuts the plot into frames and plays
them in order, so what it earns is a **strip** naming the frame on screen. See
[Play](play.qmd).

Not every channel applies to every mark: a `line` is drawn with a single
stroke, so it takes no per-row `size` or `opacity`. GOG states the rule rather
than guessing; see [Design laws](design-laws.qmd), and
[Combinations](combinations.qmd) for the whole mark-by-channel grid on one page.

## Scales: how a number becomes a position

A scale is written on the **binding**, the channel with its column in it. It is
not an atom of its own, because it belongs to one channel and nothing else can
combine with it:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life."*

| Scale | What it does | Today |
|-------|-------------|-------|
| `linear` | Equal steps for equal differences | ✅ the default |
| `log` | Equal steps for equal ratios | ✅ on `x`, `y`, `color`, `size`, `opacity`; any `base` |
| `time` | Positions dates and times | ✅ chosen from the column type: `Date` / `POSIXct` |
| `category` | One slot per distinct value | ✅ chosen from the column type: text and factor columns |
| `order` | Which way a categorical axis runs | ✅, or [declare it on the column](data.qmd) — a factor in R, `ordered()` in the other three |

A scale is not a transform: it produces no new values, only different positions
for the ones you have. Where the two meet, what a log scale means for
`bar * bin` or `bar * sum`, is set out in [Scales](scales.qmd).

The same binding takes `limits`, which says what range the channel runs over
when the data is not the authority on that:

```{r}
data(tide) + line + x(hour, limits = c(0, 24)) + y(height) + polar()
```

The gauge reads every three hours, so the readings run from hour 1 to hour 22.
A circular axis joins its two ends. It cannot tell that the hours between 22 and
1 belong to the day as well. Left to fit the data, it would close the circle at
hour 22. The first and last readings would then land on the same spoke, though
21 hours separate them. `limits = c(0, 24)` says the day is 24 hours long, so
every reading sits where a clock would put it. The curve still does not close,
because no reading was taken at midnight and stating a domain does not invent
one. [Polar](polar.qmd) follows that through. `limits` works on every channel
that measures, `NA` leaves an end to the data, and the whole of what it does is
in [Scales](scales.qmd).

A position binding takes one more, `tick_count`, which says how densely the axis
should be labeled:

```{r}
#| layout-ncol: 2
data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  x_label("GDP per capita") + title("Three ticks")

data(gapminder_2007) + point + x(gdp, tick_count = 12) + y(life) +
  x_label("GDP per capita") + title("Twelve asked for, eleven drawn")
```

The left axis reads 0K, 20K, 40K. The right one steps by 5K, which gives eleven
ticks rather than the twelve asked for. `tick_count` is a target rather than a
promise. It chooses a step, and the step is then rounded to a number a person
would pick. What you can rely on is the direction: ask for more and you get more,
and every tick still lands on a round number.

It is written in the same place as `scale` and `limits`, and for the same reason.
How many ticks an axis gets is a property of the scale, so it belongs to the
sentence that describes the scale rather than to `theme()`, which is the page.
Unlike `limits` it reaches only `x`, `y` and `z`, because those are the three that
draw an axis at all. A legend names both ends and the middle, so there is no count
to choose. [Scales](scales.qmd) has the rest.

## Settings: values you *set* rather than *map*

Every channel above answers a question about the data. `color(species)` asks
"which species?", and earns a legend so the reader can answer it.

Making every point one color asks nothing. It maps no column, needs no scale,
and has nothing to decode, so it is not a channel. That is `style`:

| Setting | What it fixes | Today |
|---------|--------------|-------|
| `style(color = )` | One color for the whole layer | ✅ |
| `style(opacity = )` | One opacity, `0`–`1` | ✅ |
| `style(size = )` | Point radius, or line stroke width | ✅ |
| `style(shape = )` | One glyph for every point | ✅ |
| `style(pattern = )` | The texture of a mark's paint: a stroke's dash (solid/dashed/dotted), a fill's hatch (hatch/crosshatch/grid/dots) | ✅ all but `point`, `text` and `surface` |
| `palette( )` | Which colors the `color` channel hands out | ✅ |

```{r}
# map: color carries the continent, and a legend decodes it
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent)
```

```{r}
# set: one color, no question asked, no legend drawn
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "tomato", opacity = 0.5)
```

See [Setting vs mapping](style.qmd) for the full rules.

## The `+` operator: layering

`+` composes atoms into a specification.
The result is a `PlotSpec`, rendered to SVG. Raster formats come from
converting that SVG. Three dimensions already draw: `space()` projects the cube
into the same SVG, so there is one renderer rather than a second one for depth
(see [Space](space.qmd)). You can also turn that cube yourself: drag any 3-D plot
in the web edition and it rotates, with a readout of the two angles and a way
back to where it started. `space(turn =, tilt =)` names the angle a plot opens
at, and the mouse takes you anywhere else. What is still on the roadmap is
letting a nearer surface hide whatever sits behind it, which needs more than a
page can give.

```{r}
# One layer
data(gapminder_2007) + point + x(gdp) + y(life)
```

```{r}
# Two layers — bars behind, points on top
data(actuals) + x(year) + y(sales) +
  bar + style(color = "lightsteelblue") +
  point + style(color = "steelblue", size = 6)
```

Each mark opens a new layer.
Layers are rendered in order: first mark drawn first (at the bottom).

Order comes up more than once in this grammar, and it does not mean the same
thing every time. [Which orders matter](operators.qmd#which-orders-matter)
gathers all of it in one table: what you may rearrange freely, what the engine
refuses, and the two arrangements that change the picture without saying so.

## The `*` operator: combining a mark with a transform

`*` binds a mark to a transform, creating a derived mark type.
The mark is always on the left; the transform is on the right. Where two
transforms follow one mark, write them in the order they happen:
`bar * bin * mean` cuts the axis into bands, then averages inside each one.

```r
bar * bin      # histogram
line * smooth  # smoothed trend line
zone * bin     # heatmap — the same `bin`, cutting both axes
```

The last line is the pattern worth noticing: one transform means one thing, and
the mark decides how it lands. A `bar` leaves an axis free to measure the count
along, so `bin` cuts the other one; a `zone` measures nothing by length and
needs an extent on both, so the same word cuts every axis with a width to cut
and the count becomes the color. On a zone with one categorical position it
cuts the other one only, since a category arrived already cut.

The summaries divide by the same arithmetic, read the other way round. `zone *
mean + x(a) + y(b) + color(v)` averages `v` within every cell two categories
cross, because a summary groups by every position the mark does *not* measure
with and reduces the column named on the one it does — which on a bar is `y`,
on a zone is `color`, and in [`space()`](space.qmd) is `z`.

See [Operators](operators.qmd) for the full rules,
and [Transforms](transforms.qmd) for each transform in detail.

## The `|` and `/` operators: faceting, or composing

`|` puts things side by side and `/` puts one above the other. What things —
panels or whole plots — is decided by what you write on the right.

With `facet()` on the right they split *one* plot into panels, one per category
of the column it names:

```{r}
# One panel per continent, side by side
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

Every panel shares one scale, that is what makes them comparable, and both
operators together cross into a grid. See [Faceting](faceting.qmd) for the
full rules.

With another *plot* on the right they arrange the two on one page, each keeping
its own coordinate space:

```{r}
# Two plots, side by side
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

Composed plots share nothing until they name the same column on the same axis,
and then they share that axis outright — one scale, one extent, drawn once. That
single rule is what makes a marginal plot two operators rather than a chart type.
See [Composition](composition.qmd).
