---
# Re-stamped on every render, so the landing page reports how current this copy
# is. The fixed counterpart, the first-published date, is `book: date:` in
# `_quarto.yml`. This one belongs here because it must reach the landing page
# alone; see the comment beside its partner for what the other placements do.
date-modified: today
---

# Preface {.unnumbered}

```{r}
#| include: false
source("R/setup.R")
```

GOG is a single graphics engine written in [Rust](https://www.rust-lang.org/),
callable from [R](https://www.r-project.org/), [Python](https://www.python.org/),
[Julia](https://julialang.org/), and
[JavaScript](https://developer.mozilla.org/en-US/docs/Web/JavaScript).
Its discipline comes from Hangeul (한글), the Korean
alphabet; its vocabulary from Wilkinson's *The Grammar of Graphics*; and its
working shape from ggplot2. Behind all three is one promise: **a beginner
should be able to read any plot specification aloud on day one and write it
themselves on day two.**

## *A* grammar, not *the* grammar

Wilkinson opens his book by defending the definite article in his title, the
word *The*. His grammar is a formal system: a fixed set of rules that states
what it can build and where it stops. Those stated limits are what make the
claim testable. A looser approach, such as a list of chart types, has no such
boundary. Nothing says where that list ends, so there is nothing to measure it
against. No rival can even claim to cover more. Wilkinson wrote his limits
down, and that is why *the* is a claim he earned.

This book takes the indefinite article on purpose. GOG is not a restatement of
Wilkinson's system, and it is not a subset of it either. His grammar rests on an
algebra of variable sets, in which `*` crosses, `/` nests and `+` blends, and the
four operators you will meet in this book mean none of those things. What GOG
took from him is the inventory rather than the algebra: most of the names in its
vocabulary come from his chapters on statistics, geometry, coordinates and
aesthetics, which is a real debt and a large one. What GOG is instead is a
deliberately small **kernel** of marks, channels, transforms and operators, grown
one decision at a time and answerable to [nine laws](design-laws.qmd) of its own.
The kernel is that whole vocabulary: every name the grammar has, and nothing
outside it. What those four kinds of name mean is spelled out later in this
preface. When a design question is genuinely open, he is still the tiebreaker, and the
package carries his book's name. But the cover says *a*, not *the*. GOG is one
of the grammars that could be built, and it claims nothing more.

The small article also does something the definite one cannot. *A Grammar of
Graphics* abbreviates to **agog**, English for eager, expectant, impatient to
see what comes next. The acronym is not merely lucky; it names the mental
state this grammar runs on. A plot is a specification, not drawing code: you
never place a stroke yourself. You declare what you want to see, and then there
is a gap (brief, but real, and felt every time) before the engine shows you
what you said.

Everyone who writes a plot in GOG stands in that gap. Reading this book, you see
the far side of it. Every plot on these pages was drawn by the engine while the
page was being built, not pasted in as a screenshot. That waiting is over by the
time the page reaches you: the plot is already sitting under the code. Run one of
these code blocks on your own computer, and the gap is yours. An analyst
stands in it while exploring data, after swapping one channel and before the
picture changes. The people who write GOG stand in it too, for whole sessions,
waiting to see what a new mark or transform will draw. The package is `gog`;
the state of working with it is *agog*.

The hexagon at the top of this page says that without a sentence. What sits
inside it is a face. Its eyes are `+` and `*`. Its nose is `/` and `|`, and
its mouth is the package name. Those four symbols are every operator GOG has.

The two pairs sit where they do for a reason. `+` and `*` build a single plot,
and read as *"…and also…"* and *"…derived by…"*. `/` and `|` arrange finished
plots, and the panels one plot is split into, and read as *"…below…"* and
*"…beside…"*. One pair composes inside a picture, the other between pictures.
The face is the state this
section has been describing: someone who has written a specification and is
waiting to see it drawn.

## Where this began: 1443

Nearly every writing system in use today grew by accretion. Traders bent a
neighbor's letters around new sounds, scribes wore pictures down into
strokes, and nobody wrote down why. Hangeul is the great exception. In the
winter of 1443, King Sejong announced twenty-eight letters of
deliberate design. The 1446 book that introduced them, the *Hunminjeongeum*
("The Correct Sounds for the Instruction of the People"), opens by saying why
[@hunminjeongeum1446]. The spoken language differed from written Chinese, so
ordinary people "who have something they wish to say are, in the end, unable
to express their feelings." Therefore, new
letters, few and simple, "wishing only that everyone may learn them with
ease and use them daily, with convenience."

![King Sejong's preface to the *Hunminjeongeum*, written out in the new Korean letters themselves, the very words quoted above. It names the need before it teaches a single letter.[^hunmin-img]](images/Hunmin_Jeongeum.svg){fig-alt="King Sejong's preface to the Hunminjeongeum, set in the Korean alphabet, the founding text of Hangeul" width="320" fig-align="center"}

[^hunmin-img]: Image from [Wikimedia Commons](https://commons.wikimedia.org/wiki/File:Hunmin_Jeongeum.svg); public domain.

Read plainly, that royal preface is a requirements document. It names the
users: not the royal court, but the ordinary people the old system excluded.
It names the failure: having something to say and no way to say it. It names the
success criterion: easy to learn, used daily. And the book's postface adds
the performance target: *a wise man can learn the letters before the morning
is over; even a slow one, within ten days.* It claims universality, too:
with these letters, the sound of the wind, the cry of the crane, the crowing
of the cock and the barking of dogs can all be written down.

## What made the letters learnable

The letters earned those claims through design. The consonants draw the
mouth making the sound; the vowels build from three shapes: a dot for
heaven, a flat stroke for earth, an upright stroke for the human standing
between them. Linguists eventually had to coin a new category, *featural*, for
a script this systematic [@sampson1985]. In a featural script the shape of a
letter shows a property of the sound it makes. But the property this package is
actually built on is not the smallness of the letter set. English also has a
small alphabet, and takes years, because its composition is riddled with
exceptions ("ough" alone is five different sounds). Hangeul is learnable in
a morning because its **composition has no exceptions**. An **atom** is a piece
that nothing divides further, here a single letter. The power is not the atoms.
It is the regularity with which they combine.

The opposition understood the stakes precisely. In 1444 a senior scholar of
the court academy petitioned against the new script. Civilized writing meant
Chinese characters. The clerks already had a workaround, 이두 (*idu*), and it
served everyone who had trained in it. And if letters were easy, people would
abandon the hard study through which learning had always been earned. It is the eternal argument of the
fluent expert: *the difficulty earns its keep; my fluency is capital.* You
have heard its modern form: *everyone already knows what `alpha` means.*
Sejong overruled it. The script outlived a royal ban in 1504, survived five
centuries in private letters and popular novels, was named *Hangeul* around
1912, and is today the alphabet of one of the most literate countries on
earth. Systems that are well defined and easy to use win, in the end, because
every new learner is a fresh vote for them.

## The same predicament, in graphics

Now look at the tools we draw data with. Most are logographies, writing systems
in which each symbol stands for a whole word. They are a zoo of
chart types, each memorized whole. A histogram function here, a pie
function there, each with its own arguments in its own order. That is the
logographic predicament of graphics: not a grammar but a stockpile of
special cases, powerful only in a trained hand.

That predicament already had its reformers. Wilkinson found the grammar
underneath the zoo, and ggplot2 carried it into the daily work of hundreds of
thousands [@wickham2008thesis; @wickham2016ggplot2]. The two of them did for
graphics what Sejong did for writing. They traded a stack of memorized charts
for a real, generative language, and a better one than anything built before
it. GOG stands on that achievement, written in its debt, and adds a single
ambition of its own, the one Hangeul is famous for: no exceptions in how the
pieces compose. That is the standard GOG holds itself to, and the aim is to
narrow what still stands between what a plot *means* and what its code lets
you *say*.

That debt is specific, and naming it is better than gesturing at it. The `+`
that joins the parts of a sentence is ggplot2's, not Wilkinson's, whose `+`
blends variables and is a different operation entirely. The split between
*mapping* a column and *setting* a fixed value is ggplot2's. So is drawing
several marks over one pair of axes. So is the idea that a plot is a
specification you build and hand over, rather than a set of drawing commands.
Wickham named that last one himself. "One of the most important features of the
grammar is its declarative nature," he wrote, and ggplot2 uses `+` to keep it
that way in R [@wickham2008thesis, p. 54]. This package inherits both the
property and the operator that protects it. A
reader who already writes ggplot2 will recognize the shape of almost every
sentence in this book. That is on purpose. After years of ggplot2, its grammar
is how I think about visualization, so it is the shape this one took.

What changed is narrower, and it is why this package exists at all. There is no
function for each chart type, so a histogram is derived instead of looked up.
One engine answers four languages, so the sentence you learn in R is the
sentence you write in Python. And the rules for combining the pieces hold
everywhere, with no exceptions, which is what the rest of this preface is
about.

So GOG asks nothing new of the medium, only Sejong's old question, carried to
its last clause: not *which charts should exist*, but *what is the smallest
set of letters, and the most regular rule of composition, from which every
chart can be derived?* Wilkinson's grammar answers what exists
[@wilkinson2005]. ggplot2 found the shape that made such an answer usable, and
most of how a sentence here is assembled is its idea. Hangeul disciplines how
it composes. Marks are the consonants, silent until placed. Positions are the
vowels that give them somewhere to stand. A histogram is nothing to memorize:
it is `bar * bin`, derived, the way ㅋ (*k*) is derived from ㄱ (*g*) rather
than invented fresh.

```{r}
data(gapminder_2007) + bar * bin + x(life)
```

*"Given gapminder 2007: bars derived by bin, x is life."* That is the whole
program, and the plot above was drawn from those words as this page built.
There is no histogram function in this package and no argument that turns bars
into one. `bar` is the mark, `bin` is the transform that cuts a continuous
column into intervals and counts the rows falling into each, `x` is the
channel, which is the visual property a column is mapped to, and `*` is the
operator that puts a transform on a mark. It is the *derived by* in the
sentence above. Read `*` as a spinning wheel rather than as a multiplication
sign. Its spokes show one thing being turned into another, the way ㄱ shows
the shape of the mouth that says it. Nothing there is memorized. The names are
ordinary English words, and the rule that joins them never changes. Knowing
those four parts is knowing this plot, and every other statistical chart in
this book is the same four kinds of part in a different arrangement.

And one law stands over the whole package: **no exceptions.** A rule you learn
on one mark holds on every mark, or the package has failed you.

## What the grammar refuses

That law is easy to read as *everything combines with everything*. Here is the
other half of it, which is the half that does the work.

Hangeul can write the crowing of a cock. Hangeul cannot write ㄱㄴ. A
syllable is not a pile of letters; it is a set of slots, an initial consonant,
then a vowel, then an optional final consonant, and each class of letter belongs
to one of them. Two consonants with no vowel between them fill nothing. ㄱ on
its own is silent, and a lone vowel cannot stand either: to write the bare
sound *a* you supply a carrier for the empty initial slot, and what you write
is 아. The script that
boasts it can capture the barking of a dog refuses most of the arrangements
you could make of its own letters.

Those are not two facts. They are one. Take the slots away, let any letters
stack in any order, and the script does not become more expressive; it becomes
unreadable. A block would no longer tell you how to say it, and you would be
back to memorizing shapes whole, which is where writing began. The refusals are
what make the script readable at first sight. A reader meeting a word for the
first time still knows which part is the consonant, because nothing else could
have been in that slot.

A plot is the same. GOG will not draw a `point` from an x alone. A mark with
one position has not said where to put it. GOG will not map a column to `size`
on a `bar` either. The bar's length already shows the quantity, so a size would
have nothing left to show. Neither refusal is a missing feature. Both come from
the rule that lets you read a specification you have never seen before. If
every mark accepted every channel, knowing the mark would tell you nothing. A
plot would go back to being something you memorize. The engine says no out
loud, and every no points to what you meant instead. That is why this book
renders its refusals live, on the page, beside the plots that work.

The goal is that one small set of atoms draws every plot, the way one small
set of letters writes every Korean sound. Neither is a claim of completeness.

Hangeul does not write everything. It does not mark tone, and it does not mark
how fast a syllable is said. Some sounds it cannot spell at all. Those gaps
were left as gaps, because the sounds behind them are not the ones most
speakers need. The letters were not built to be perfect. They were built to be
a means of writing sound.

GOG is built the same way. It does not promise that every visualization is
possible. Reaching some of them would mean stepping outside the grammar's own
rules and adding an exception, and no chart is worth that.

What the grammar will not do is refuse a plot for being a bad idea. 꿲 (*kkwek*)
breaks no rule of Korean, and no one has ever needed to write it. Its graphics
equivalent is a color scale, the rule that turns a column's values into colors,
stretched over three hundred categories, and it
is legal here, because well-formedness and usefulness are different judgments
belonging to different people. The engine guarantees the first, always. The
second is yours, and it is what most of this book is about.

## The promise, and the rules this book plays by

This book inherits the 1446 promise, restated for graphics: **read the whole
grammar in a morning; write it fluently within ten days.** Part I is built
to be finished in one sitting. The rest of the book deepens what the
morning introduced; it never replaces it.

Five rules govern every page, and you can hold the book to them:

- **Every plot is live.** Each is drawn by the engine as the page builds;
  no plot in this book is a screenshot. The one static image is historical:
  the *Hunminjeongeum* page shown earlier, which no engine could draw.
- **Questions first.** Every teaching chapter opens with a question about
  data, and answers it before it names a feature.
- **A small cast.** Five families of table carry three out of four plots in
  this book. A chapter that needs a frame of its own introduces it where it is
  drawn. New columns cost attention; the diversity lives in the questions.
- **Read it aloud.** The first specification in every chapter is followed by
  its English sentence. Read that one aloud, and the rest of the chapter reads
  the same way. If a sentence cannot be read aloud cleanly, that is a bug in
  the grammar, not in you.
- **Errors on stage.** Refusals render live in these pages, because the
  engine's *no*, always with a direction, is half of what it teaches.

One thing to be clear about before you start: this package is young. The
version number is 0.0.1 and it means what it says. What exists is a minimum
viable package, so do not expect everything to work the first time. If you use
it for an experiment or a paper, you are taking a risk. This warning is here so
that the risk is one you chose, and not one you found out about later.

Expect it to move quickly. New atoms and new features will arrive, and they will
arrive often. A plot you write today may have a shorter way to say it next
month. What I will try hard not to do is take anything away. A sentence that
draws today should still draw a year from now. If that ever has to break, I will
say why rather than let it break quietly. I cannot promise it, though, and at
version 0.0.1 nobody honestly could.

This book moves with the package. Because every plot here is live, a change to
the engine reaches these pages on the next render. The text is rewritten
whenever it stops matching. Expect the book to change as often as the package
does. Once the grammar settles, a section recording what changed and when will
be worth adding. That is a promise for later, not something you will find here
today.

Please report the bugs and issues you hit. A package this young improves mainly
by being used, and that is what I am hoping for.

One kind of report is worth more than the others, and it is the one I cannot
write myself. Take a dataset you know well and draw it in GOG. Then draw the
same thing in a tool you already trust. Any tool at all:
[ggplot2](https://ggplot2.tidyverse.org/) or
[matplotlib](https://matplotlib.org/),
[Excel](https://www.microsoft.com/en-us/microsoft-365/excel) or
[Tableau](https://www.tableau.com/), [Stata](https://www.stata.com/) or
[SPSS](https://www.ibm.com/products/spss-statistics),
[Vega-Lite](https://vega.github.io/vega-lite/) or [D3](https://d3js.org/). Put
the two pictures side by side and tell me where they differ. A standard dataset
is ideal, because the picture it should produce is already familiar to
everybody.

Here is why that comparison is worth so much. This package tests itself
thoroughly, but in one direction only. Every sentence in this book is drawn in
all four languages, and the four files have to match to the byte. Every refusal
on these pages is run again on every build. All of that proves GOG agrees with
itself. None of it proves the picture is right. A box whose whiskers reached the
wrong value would pass every test this project has, four times over, in four
languages, and this book would print it on the page as though it were correct.

A difference is not automatically a defect, and it is fair to know that before
you spend the time. Some differences are decisions. A GOG box spans the lower
and upper quartiles, as most software does, while Tukey's original spans the
hinges, and the [box](marks/box.qmd) chapter says so and says why. So a
comparison gets one of
two answers: this is a bug and thank you, or this is a choice and here is the
reasoning behind it. Both are worth the message.

Not everything reported will be built. I am this project's benevolent dictator
for life. The phrase is a tongue-in-cheek open-source title for a founder who
keeps the final say in a dispute. It was first used in 1995 for [Guido van
Rossum](https://en.wikipedia.org/wiki/Guido_van_Rossum), who created Python.
Many people would say I am not qualified to hold
it. I am holding it until this package is in the hands of a user community and
of people better than me.

## The cast

Five groups of table carry most of the book, eight tables in all. Meet them
once, here:

- **`gapminder_2007`**, 142 countries in 2007: `gdp` (per capita),
  `life` (expectancy in years), `population`, `continent`, `country`.
  The workhorse, and an excerpt of the Gapminder Foundation's data
  [@bryan2025gapminder]. **`gapminder_asia`** is five Asian countries traced
  1952–2007, and **`gm_all`** is the full half-century sweep.
- **`iris_flowers`**, 150 flowers: `sepal_length`, `sepal_width`,
  `petal_length` (all in cm) and `species`. Three measurements on one
  scale, and a category that separates cleanly. The measurements are Edgar
  Anderson's, and they reach most software through Fisher's 1936 paper
  [@fisher1936] and R's own `datasets` package [@rcore2026].
- **`medals`**, five countries' Olympic `gold`, `silver`, `bronze`
  counts. Small enough to check by eye.
- **`actuals`** and **`forecast`**, yearly `sales`: history in one
  table, expectation in another. The two-table pair.
- **`six_weeks`**, 42 consecutive days of `orders`, with `day` a real
  `Date`, for calendar axes.

Those five groups answer most of the questions asked here. A handful of marks
need a shape none of them has, and each brings its own small frame, introduced
on the page that first draws it rather than in this list. Wind observations
for the circular axes in [polar](polar.qmd). A mountain's elevation grid for
[surface](marks/surface.qmd). Two gliders' coiling flight paths for the third
dimension. Thresholds and shaded spans for the marks that annotate. None of
them arrives to add variety. A new column costs the reader attention, so the
variety in this book stays in the questions.

## How this book runs

The shape of the book borrows from the *Hunminjeongeum* itself, which
stated its whole system in a few pages, then explained each letter class,
then showed assembly, then ended its explanations with ninety-four ordinary
words in use:

1. **A morning**, the whole grammar, fast: a first plot, the kernel on
   one page, reading practice, writing practice.
2. **The letters**, each atom family in depth: the data you bind once,
   the marks, the refining channels, setting vs mapping, the transforms,
   and the scales.
3. **Sentences**, composition: layering, deriving, scope, faceting, whole
   plots arranged on a page, the third dimension, the plane bent into a
   circle, the panel packed into regions, and a plot played in time.
4. **The cookbook**, recipes organized by your data's shape and your
   question, never by chart name; every recipe ends one atom away from
   its neighbors.
5. **The reasons**, the nine laws the package never breaks, and why.
6. **The bindings**, one short chapter apiece for [R](bindings/r.qmd),
   [Python](bindings/python.qmd), [Julia](bindings/julia.qmd) and
   [JavaScript](bindings/javascript.qmd). A binding is the package that lets
   one language write GOG sentences and send them to the engine. These
   chapters hold what belongs to the host language rather than to the grammar.

Five appendices follow. The [kernel card](kernel-card.qmd) is the whole
vocabulary on one page. The [mark-by-channel grid](combinations.qmd) shows
what combines and what the engine refuses, generated from the engine's own
rule table on every build so it cannot drift from the code. A
[dictionary](chart-names.qmd) runs from chart names to sentences. A
[phrasebook](phrasebook.qmd) serves readers arriving from ggplot2. And [what
GOG does not do, and why](coverage.qmd) closes the book.

This book is written in R. [Python](bindings/python.qmd),
[Julia](bindings/julia.qmd) and [JavaScript](bindings/javascript.qmd) speak
the same sentences into the same engine. Those three chapters are the whole of
the difference: how each language captures a column name, and, for JavaScript
alone, the four words it needs because it cannot give `+ * | /` a meaning of
its own. Most code blocks are executed during render: the plots you see are
drawn by the engine as the page builds. Examples that use roadmap features
remain display-only until those features land.

## Installing GOG

Every plot on this site was drawn by the engine while the page was being built.
The engine works, all four bindings work, and each one installs with a single
command. Copy the one for your language.

**R**

```r
install.packages("gog", repos = c("https://psychometrician.r-universe.dev",
                                  "https://cloud.r-project.org"))
```

**Python**

```bash
pip install gog
```

**Julia**

```julia
using Pkg; Pkg.add("GrammarOfGraphics")
```

**JavaScript**

```bash
npm install grammar-of-graphics
```

The package has two names, and the registries are the reason rather than the
grammar. In R and in Python it is `gog`, because a three-letter name is allowed
in both. Julia asks a package name for at least five letters and capitalized
words, and npm already had a package called `gog`. So those two spell the name
out instead, each in its own style: `GrammarOfGraphics` in Julia, and
`grammar-of-graphics` on npm, which does not allow capital letters. Where `gog`
is free the package is `gog`, and where it is not the name is written in full.

Three of these packages carry the engine inside them, already built for your
computer. You do not need a Rust [@klabnik2023rust] toolchain, and nothing has to
be added to your `PATH`. Two of the four commands come with a condition that is
worth knowing before you type it.

The R package is not on CRAN yet. It is published on
[r-universe](https://psychometrician.r-universe.dev/gog) instead, which is why the
`repos` argument names two addresses. The second address is an ordinary CRAN
mirror, and it is there so your other packages still install as usual.

The Julia package is the youngest of the four. Its registration is
[submitted](https://github.com/JuliaRegistries/General/pull/162792), and the
General registry holds every new package for three days before accepting it, so
that command starts working around 2026-08-02. It is also the one binding that
does not carry the engine yet, so the package will load before it can draw. The
[Julia](bindings/julia.qmd) chapter says what to do until then.

Each binding has its own chapter at the end of this book, and each of those
chapters closes with the longer version of this section: [R](bindings/r.qmd),
[Python](bindings/python.qmd), [Julia](bindings/julia.qmd) and
[JavaScript](bindings/javascript.qmd).

Every one of these packages is at version 0.0.1. The grammar is settled and the
engine draws this whole book, but the packaging is new, so the version number
will move faster than the vocabulary does. One thing does not move with it. If a
page shows a sentence, the engine drew it. If a page shows a refusal, the engine
refused it.

The prose in this book is licensed [CC BY-NC-SA
4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/). The code is licensed
[Apache 2.0](https://www.apache.org/licenses/LICENSE-2.0), the same license as
the engine, so the examples can be used in your own work.

## Who wrote this

This book carries five names, and the division of labor behind them is worth
stating plainly rather than leaving in a footnote.

One of the five is not a name. I publish as **psychometrician**, and the wish
is an old one. I have wanted a pen name since I learned statistics in 1992,
which is when I met *Student*. Student was [William Sealy
Gosset](https://en.wikipedia.org/wiki/William_Sealy_Gosset), an English chemist
and statistician, and the t-distribution carries his pen name instead of his
own. My own papers are few, and no academic journal would let me publish
under a pen name. This book is the first chance to use one.

The word does a second job. Most people do not know the job title, and anyone
who uses this package now knows that a
[psychometrician](https://en.wikipedia.org/wiki/Psychometrics) started it and
maintains it. What they are less likely to know is how much psychometricians
care about data visualization. As far as I know, only one academic journal
runs a visualization competition, and it is ours: *[Educational Measurement:
Issues and Practice](https://onlinelibrary.wiley.com/journal/17453992)*,
published by Wiley for the [National Council on Measurement in
Education](https://www.ncme.org/) (NCME). Every year it asks its members for
thoughtful and creative diagrams that communicate a clear message about
educational or psychological data. The winning entries become the journal's covers, and the
strong runners-up are shown inside. The [cover
gallery](https://onlinelibrary.wiley.com/page/journal/17453992/homepage/cover_gallery.htm)
is worth your time. This package may not be able to draw all of them.

Psychometricians have cared about data visualization for a long time. My own
history with it starts in 2005. I was writing my dissertation, and every
conditional plot in its simulation chapters came out of the **lattice**
package [@sarkar2008lattice]. It was the first graphics tool that impressed me.
After that I was a heavy ggplot2 user for years. What changed was the day I
understood that ggplot2 had a grammar. From then on it felt like there was no
end to the plots I could make.

I owe [Hadley Wickham](https://hadley.nz/) more than a citation for that. Before
his package and his book, a plot was a task to get through. You looked up the
function for the chart you had been asked for, learned whatever arguments it
happened to want, and moved on. The grammar changed what the work was. Visualization became a
creative act rather than an item on a list. It is creativity inside a
boundary, the one the grammar draws. That is the part I most want to pass on.

The design is mine. The founding question, the Hangeul discipline, the nine
laws, and the ruling at every fork about what this grammar would and would not
become: those are the parts I am answerable for. The hex sticker is mine as
well. I designed it myself in [Canva](https://www.canva.com), with no sticker
package and no model. I am not an artist, and what I wanted was not a beautiful
sticker but an unmistakable one. The Rust
engine was written entirely by the four Claude models on the cover, as were all
four language bindings and the test suites. They also did the work nobody is
ever thanked for, which is running every sentence in this manual through four
languages, again and again, until all four agreed to the byte.

The text is a third case, and the honest comparison is a ghostwritten
autobiography. The public figure whose name is on the cover usually did not
type it. A professional writer interviews them, and what comes back is their
thinking in another person's prose. These pages work the same way, with one
difference. Not every idea in this book started with me. Claude often suggested
something better than what I first brought, and my task was to accept it or
reject it. Claude put all of it into English. The sentences are not mine; the
decisions behind all of them are.

Naming them as co-authors is meant literally. A line reading "with assistance
from" would understate the work by an order of magnitude, and this is a book
about saying what you mean.

They were not one kind of collaborator, and it would be a poor tribute to
pretend they were. On some days they were assistants: fast, tireless, and so
eager to be useful that "done" tended to arrive some time before done did. A
change to how bars are drawn was once confirmed by checking a stroke color
instead of looking at the plot, and the report came back clean while every bar
on the page was still wrong. Most of the checks now guarding this repository
exist because of an afternoon like that, and not one of them was added to catch
a careless human.

On other days they were advisors, and that is the part I did not expect to be
writing about here. A design question asked at the wrong hour, asked badly,
asked for the third time in a worse form than the first: it got taken as
seriously as if it were a good one. A fair number of the refusals in this
engine exist because I said "why not just" and got a patient answer rather
than an agreeable one, and some of the reasoning in Part V was settled in an
argument I was not winning. An assistant that agrees with you is the easy
thing to build. This grammar would be a good deal worse if that is what I had
had.

Taken together, it felt like a research meeting. Smart researchers filled the
room, eager to listen to me and eager to carry out the plan once I had decided.
I have not had both of those at once before. It was always a good time, and
sometimes an expensive one. A small question would open into a long detour, and
I would lose several hours in it. Accepting that invitation was my decision
every time, so none of this is a complaint.

Responsibility does not divide the same way. Every error in these pages, every
refusal the engine gets wrong, and every design decision that turns out to have
been the wrong one is mine, because I chose it and I kept it. I organized this
project and I maintain it. The credit is shared; the fault is not.

One last item belongs in an honest accounting. I work as a psychometrician, and
this package was never part of that job. It was built late at night, on
weekends, and on vacation days. That time belonged to my wife and daughter, and
I spent it talking to the four Claude models on the cover instead. I am not
going to dress that up as devotion to the craft. To my wife and daughter: sorry.
I would like to end by saying it is finished. It is not. The version is 0.0.1,
and what comes next will cost the same nights, weekends and vacation days.

## Thanks to Posit

[Posit PBC](https://posit.co/), called RStudio PBC until 2022, builds a large
part of what a working R user touches in a day. The parts I depend on are free,
and there are three of them.

The [tidyverse](https://www.tidyverse.org/) [@wickham2019tidyverse] is the one I
would miss first. It is the vocabulary I think in
when I handle data, and has been since long before this project. You will not
find it in this book, whose tables are built in base R because `gog` imports
only `jsonlite` [@ooms2014jsonlite] and I want it to install anywhere. That is a
rule for the package. It is not how I work the rest of the day.

Their editors are the second. I worked in
[RStudio](https://posit.co/products/open-source/rstudio/) for years, and I work
now in [Positron](https://positron.posit.co/), which is newer and holds R and
Python in one window. The two are not the same program under different names: RStudio is
still here and still maintained, and Positron sits beside it.

Quarto is the third, and it belongs here rather than in a colophon, because the
first of the five rules above rests entirely on it. *Every plot is live* is a
claim about honesty rather than convenience. Every code block on these pages ran
while the page was being built. One that draws a plot really drew it, and one
that shows a refusal was really refused. [Quarto](https://quarto.org/) is what
executes those blocks. Without it, every plot here would be a screenshot and
every code block would be text that nobody had run. The book would drift out of
step with the package it describes, and no one would notice.

Quarto gives the book three more things it could not have faked. Four languages
spell the same sentence in one tabbed block, so a reader sees the parity instead
of taking it on trust. One source builds both an HTML site and a printable PDF,
Korean letters and all, from the same executed chunks. And a block is allowed
to fail without stopping the build, so the engine's refusals appear beside the
plots that work instead of in an appendix.

None of that announces itself while you read, which is the mark of a tool doing
its job well. Without Quarto there would be no book here at all.

I will not be modest about what the tidyverse and their editors are worth
either. My work as a psychometrician would be harder, and a good deal more
miserable, without what this company gives away. Thank you.

## Thanks to Anthropic

Soon after Claude came out, I built a web application. It was React on the
front, Supabase for the database, and the Google APIs behind it. I did not know
React well, and I did not know Supabase at all. The application got finished
anyway.

Then it failed, and the reason had nothing to do with the code. I could not
market it. I did not know how to reach the people it was built for, and after a
while I shut it down.

The lesson was not the one I went looking for. Building the thing was no longer
the hard part. If I plan carefully, keep at it, and work through the problem
with Claude, I can finish what I set out to build. So I began to want something
else: my own packages, in the two languages I actually work in, R mainly and
Python occasionally. This one came out of that.

A grammar of graphics is a larger promise than a web application. It is an
engine, and an engine has to be written, tested, and kept working. I am a
psychometrician, with no team on this project, no funding, and no ability to
write Rust. Without [Claude Code](https://claude.com/claude-code), this package
would still be a plan.

One more item belongs on that list. English is not my first language. Every
conversation behind this package was held in my own broken English, from the
first design question to the last bug report. I chose the wrong words and put
them in the wrong order. I left out the small words that hold an English
sentence together. Claude understood what I meant nearly every time. Twenty
years of thinking about this field reached the engine through sentences full of
mistakes. None of that thinking was lost on the way. You can do better than I
did.

I think this may have been luck, and I would rather say so than not. I happened
to be thinking about data visualization. I happened to be thinking about
Hangeul. Then Claude Code arrived. What I did was stand between the two and let
them meet.

There is nothing rare about that position. Anyone could be standing in it. If
you have been thinking about your own subject for a long time, and you are
willing to play with this tool, you are already there. Much bigger things than
this package will come out of that.

So, thank you to the people at [Anthropic](https://www.anthropic.com/) who built
Claude Code, and to the four models named on this book's cover. I do not know
most of your names. I know two of them from YouTube:
[Boris Cherny](https://x.com/bcherny) and [Cat Wu](https://x.com/_catwu). You
built the thing that let a psychometrician write a graphics engine. Thank you.
