# Map

```{r}
#| include: false
source("R/setup.R")
```

Where did it happen? Some data answers that with a place on the earth, and a
place is two numbers. **Longitude** says how far east or west, and **latitude**
how far north or south. Both are angles, measured in degrees.

Every plot in this book has used two numbers for position. So a map is not a new
kind of picture. It is the same sentence you already write, drawn in a space
that treats the two numbers as angles on a sphere.

`quakes_fiji` records earthquakes near Fiji, one row per quake:

```{r}
#| echo: false
peek(quakes_fiji)
```

```{r}
data(quakes_fiji) + point + x(east) + y(north) + map()
```

*"Given quakes Fiji: points, x is east, y is north, in the map space."*

That is a thousand earthquakes recorded near Fiji. `x` is longitude and `y` is
latitude, and `map()` is the whole of what makes it a map. The axes are labeled
in degrees, because that is what the columns hold.

In the web edition, `+` magnifies a crowded part, `−` makes the picture smaller,
and *fit* shows the whole map. A map needs that more than most plots do, because
places stay crowded however far you zoom in. Zooming is a **view**: it magnifies
the picture and changes nothing the plot claims.

## Flattening the sphere loses something

The earth is round and the page is flat. No projection can flatten a sphere
without distorting it, and the choice is not between a good projection and a bad
one. You choose which distortion you accept.

Two properties matter most: a region's area, and a small shape's form. A
projection can keep one of them, never both. `preserve` says which:

```{r}
#| echo: false
peek(world_borders)
```

```{r}
(data(world_borders) + path + x(lon) + y(lat) + group(piece) +
   map(preserve = "area") + title("preserve = \"area\"")) /
  (data(world_borders) + path + x(lon) + y(lat) + group(piece) +
     map(preserve = "angle") + title("preserve = \"angle\""))
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece, in the
map space preserving area, above the same preserving angle."*

Look at Greenland, at the top left of each map, and compare it with Africa. In
the first, Greenland is small, which is correct: Africa is about fourteen times
larger. In the second, Greenland has grown until it looks as large as Africa.
The second map keeps every small shape's true form and gets the sizes wrong.

Both plots name `lon` on `x`, so `/` gives them one axis and one width. The
projection sets each height, and the two heights differ. Keeping every small
shape's form is what stretches the second one north and south.

`preserve = "area"` is the default, because a map is usually read by comparing
how much space something takes. A projection that inflates the far north tells
the reader something false about every number placed there.

The world map above is a table like any other, and the next section reads it
column by column.

## Drawing the outlines

To draw a boundary of your own, you need to know what shape of table the engine
reads. The table has one row for every point along a coastline or a border:

```{r}
knitr::kable(head(world_borders, 4))
```

`path` connects rows in the order the table gives them, which is exactly what
tracing a coastline needs. `group(piece)` says where one outline ends and the
next begins. A ring is one closed loop of points, ending where it returns to its
starting point. A country is not always a single ring. An island is a ring of
its own, and a country that encloses another needs a second ring for the hole.
So `piece` counts rings rather than countries.

Without that split, `path` would join the last point of one country to the first
point of the next, and draw a line across the ocean between them.

## The second layer is the data

The chapter's first plot drew the earthquakes with no coastline, so nothing
said where on the earth they were. An outline on its own is only a background.
It becomes a plot when something is drawn on top of it, and that is ordinary
layering. The outline is one `data()`
and one mark, and the earthquakes are another.

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) +
  style(color = "tomato", opacity = 0.35, size = 3) +
  map() + title("A thousand earthquakes, and where Fiji is")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given quakes Fiji: points, x is east, y is north, in the map space."*

The earthquakes now have a coastline around them, so a reader can see where they
are. Nothing in the sentence is new: two tables, two marks, and the second
`data()` binds to the mark that follows it.

## A quantity goes on a channel

The earthquakes differ in magnitude, and the picture so far shows only where
they were. A flat plot, a cube and a polar plot all leave you an axis to
measure along. A bar stands on one axis and measures its length along the
other. A map has no axis left, because longitude and latitude use both.

So a quantity moves to a **channel**. Size each place by it:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) + size(magnitude) +
  style(color = "tomato", opacity = 0.4) +
  map() + title("Sized by magnitude")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given quakes Fiji: points, x is east, y is north, size by magnitude, in the map
space."*

That plot has a name, the **proportional-symbol map**, and no single atom draws
it. It is `point` with `size` bound, in this space.

Color works the same way, and takes a category rather than a quantity:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  color(continent) + map() +
  title("Each continent in its own color")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece, color by
continent, in the map space."*

## Meridians and parallels

A map often needs a reference line, and the most common one is the equator. A
`rule` draws a line at one position and spans the axis it does not name. In a
flat space that is a reference line. Here, a rule at one latitude is a
**parallel**, and a rule at one longitude is a **meridian**. The plot below
draws one parallel, the equator:

```{r}
#| echo: false
peek(equator)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(equator) + rule + y(lat) + style(color = "#d4674f") +
  map() + title("The equator")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the lines of table: a rule, y is lat, in the map space."*

A `rule` takes its position from a column, like every other mark, so the line
above comes from a one-row table. That is the same way a reference line is drawn
on any other plot.

Nothing was added to the grammar to draw that. The same sentence that puts a
reference line on a scatter puts the equator on a map. The space gives the mark
a new meaning, and the mark itself is unchanged.

## Naming places

A coastline alone does not say which part of the world it is. A few named
cities do. `text` writes a string at a position, and a position here is a
place:

```{r}
#| echo: false
peek(capitals)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(capitals) + point + x(lon) + y(lat) +
  style(color = "#3d5a80", size = 5) +
  data(capitals) + text + x(lon) + y(lat) + label(name) +
  style(nudge = "up", size = 11) +
  map() + title("Five cities")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the capitals: points, x is lon, y is lat; then given the capitals: text, x
is lon, y is lat, label by name, in the map space."*

## What a map refuses

You may want to count the earthquakes by longitude, which on a flat page is
`bar * count`. A mark that measures along an axis has no axis here. gog says so
rather than drawing something that looks like a map and is not one:

```{r}
#| error: true
data(quakes_fiji) + bar * count + x(east) + map()
```

The message names the marks that do draw, and it says what to do instead. Put
the quantity on `size`, as the magnitude map above does, or on `color`.

A position on a map is a place, so a category cannot be one:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| error: true
data(gapminder_2007) + point + x(continent) + y(life) + map()
```

And the panel's shape belongs to the projection:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + map() + theme(ratio = 1)
```

An equal-area map stretched to fit a box is no longer equal-area, though it
still looks like a map. That is the kind of error a reader cannot see, so the
engine refuses it instead of accepting the number and ignoring it.

## Where Mercator stops

`preserve = "angle"` is the **Mercator** projection, the one most web maps use.
It sends the poles infinitely far away, so it has to stop before it reaches
them. It stops at 85.05 degrees. A row beyond that is drawn at the edge, and gog
reports how many:

```{r}
#| echo: false
peek(far_north)
```

```{r}
data(far_north) + point + x(lon) + y(lat) + map(preserve = "angle")
```

Read the message above the plot. One row sits beyond the limit and was drawn at
the edge instead of where it belongs. Nothing was dropped, because dropping it
would misreport how many places there are. `preserve = "area"` reaches both
poles and has no such limit.

## Filling the regions: the choropleth

Every country so far has been drawn as an *outline*. Filling each country by a
value is the most familiar kind of map, and it has a name: the **choropleth**.

The mark is `zone`, which shades a region. Everywhere else in this book a zone
takes its sides from its own row: a category's slot, `bounds`, `bin` or
`density`. A coastline is a fifth way, and it is the first that gives many rows
for one shape. So the sentence needs a word for which rows belong to which
region, and `group` is that word already:

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + map() + title("Every country, filled by continent")
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country, color
by continent, in the map space."*

Compare that with the earlier map that colored each continent. Only two things
changed: `path` became `zone`, and `group(piece)` became `group(country)`. A
`path` traces one ring at a time, so it groups by ring. A `zone` fills a whole
country, so it groups by country.

## A value on the map

Where in the world do people live longest? `color` reads a number as readily as
a category, so the same sentence shades each country by a value. The value
comes from a table you already know: `gapminder_2007`, the same 142 countries
[A first plot](first-plot.qmd) draws.

Joining two tables is the host language's job, not the grammar's. In R the join
itself is one `match()` call. Every host language has its own one-call lookup.
The plot sentence under the join is the same in all four languages:

```{r}
filled <- world_borders
filled$life <- gapminder_2007$life[match(filled$country,
                                         gapminder_2007$country)]
filled <- filled[!is.na(filled$life), ]

data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  style(color = "#e6e6ec") +
  data(filled) + zone + x(lon) + y(lat) + group(country) + color(life) +
  map() + title("Life expectancy at birth, 2007")
```

Two layers again, and the first one is not decoration. Of the 176 countries in
the outline table, 120 have a number in `gapminder_2007`. The pale gray layer
underneath is what the reader sees for the other 56. Without it those countries
would simply be absent, and a reader could not tell "no data" from "no country".

Read the map and the scatter together. The chapter [A first
plot](first-plot.qmd) shows life expectancy against income as a cloud of points.
Here is the same column, and the map answers a question the scatter cannot:
*where*.

## Rings, and the country inside another country

A region is made of rings, and the rings are found by a rule rather than by a
column: a ring ends where it returns to the point it started from. Boundary
data is written that way, so nothing extra has to be said.

That rule is what lets a country contain a hole. Lesotho lies entirely inside
South Africa, and South Africa's boundary has a second ring around Lesotho:

```{r}
enclave <- world_borders[world_borders$country %in%
                           c("South Africa", "Lesotho"), ]
enclave$life <- gapminder_2007$life[match(enclave$country,
                                          gapminder_2007$country)]

data(enclave) + zone + x(lon) + y(lat) + group(country) + color(life) +
  map() + title("Lesotho is not painted on top of South Africa")
```

Lesotho is a hole in South Africa, not a patch laid over it. The difference
matters because the two countries have different values. A patch is correct only
when it is drawn second, and nothing fixes the drawing order.

If a ring in your data does not close, gog cannot tell where one ring ends and
the next begins. It draws the region, warns you that the ring ends had to be
guessed, and names what to fix.

## What comes next

A map flattens the sphere so a page can hold it, and every sentence in this
chapter also draws on the sphere itself. That is [Globe](globe.qmd): the same
marks at the same places, seen from one side.
