# Map

```{r}
#| include: false
source("R/setup.R")
```

Where did it happen? Some data answers that with a place on the earth, and a
place is two numbers: **longitude**, how far east or west, and **latitude**, how
far north or south. Both are angles, measured in degrees.

Two numbers is what every plot in this book has taken. So a map is not a new
kind of picture. It is the same sentence you already write, drawn in a space
that knows the two numbers are angles on a sphere.

```{r}
data(quakes_fiji) + point + x(east) + y(north) + map()
```

*"Given quakes Fiji: points, x is east, y is north, in the map space."*

That is a thousand earthquakes recorded near Fiji. `x` is longitude and `y` is
latitude, and `map()` is the whole of what makes it a map. The axes read in
degrees, because that is what the columns hold.

## The sphere has to be flattened, and something is lost

The earth is round and the page is flat. No projection can flatten a sphere
without distorting it, and the choice is not between a good projection and a bad
one. It is between which mistake you prefer.

Two properties matter most, and a projection can keep one of them, never both.
`preserve` says which:

```{r}
#| layout-ncol: 2
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  map(preserve = "area") + title("preserve = \"area\"")

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  map(preserve = "angle") + title("preserve = \"angle\"")
```

Look at Greenland, at the top left of each map, and compare it with Africa. On
the left, Greenland is small, which is correct: Africa is about fourteen times
larger. On the right, Greenland has grown until it rivals Africa. The right-hand
map has kept every small shape's true form and paid for it in size.

`preserve = "area"` is the default, and the reason is what a reader does with a
map. A map is usually read by comparing how much space something takes. A
projection that inflates the far north tells the reader something false about
every number placed there.

## Drawing the outlines

The world map above is a table like any other. It has one row for every point
along a coastline or a border:

```{r}
knitr::kable(head(world_borders, 4))
```

`path` connects rows in the order the table gives them, which is exactly what
tracing a coastline needs. `group(piece)` says where one outline ends and the
next begins. A country is not always one closed shape: islands are separate
rings, and a country lying wholly inside another is a ring of its own. So
`piece` counts rings rather than countries.

Without that split, `path` would join the last point of one country to the first
point of the next, and draw a line across the ocean between them.

## Two layers, and the second one is the data

An outline on its own is a backdrop. The plot begins when something is drawn on
top of it, and that is ordinary layering: the outline is one `data()` and one
mark, the observations are another.

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) +
  style(color = "tomato", opacity = 0.35, size = 3) +
  map() + title("A thousand earthquakes, and where Fiji is")
```

The earthquakes now sit somewhere a reader can find. Nothing in the sentence is
new: two tables, two marks, and the second `data()` binds to the mark that
follows it.

## Both positions are used up, so a quantity goes on a channel

Every other space leaves you an axis to measure along. A bar stands on one axis
and measures up the other. A map has no axis to spare, because longitude and
latitude have taken both.

So a quantity moves to a **channel**. Size each place by it:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) + size(magnitude) +
  style(color = "tomato", opacity = 0.4) +
  map() + title("Sized by magnitude")
```

That plot has a name, the **proportional-symbol map**, and there is no atom that
spells it. It is `point` with `size` bound, in this space.

Color works the same way, and reads a category rather than a quantity:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  color(continent) + map() +
  title("Each continent in its own color")
```

## Meridians and parallels are a mark you already have

A `rule` draws a line at one position and spans the axis it does not name. Flat,
that is a reference line. Here, a rule at one longitude is a **meridian**, and a
rule at one latitude is a **parallel**:

```{r}
lines_of <- data.frame(lat = 0)

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(lines_of) + rule + y(lat) + style(color = "#d4674f") +
  map() + title("The equator")
```

A `rule` takes its position from a column, like every other mark, so the line
above comes from a one-row table. That is the same way a reference line is drawn
on any other plot.

Nothing was added to the grammar to draw that. The same sentence that puts a
reference line on a scatter puts the equator on a map, because the space bends
what the mark means without changing what the mark is.

## Naming places

`text` writes a string at a position, and a position here is a place:

```{r}
capitals <- data.frame(
  lon  = c(-0.13, 139.69, -74.01, 151.21, 18.42),
  lat  = c(51.51,  35.69,  40.71, -33.87, -33.92),
  name = c("London", "Tokyo", "New York", "Sydney", "Cape Town"))

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(capitals) + point + x(lon) + y(lat) +
  style(color = "#3d5a80", size = 5) +
  data(capitals) + text + x(lon) + y(lat) + label(name) +
  style(nudge = "up", size = 11) +
  map() + title("Five cities")
```

## What a map refuses

A mark that measures along an axis has no axis here. GOG says so rather than
drawing something that looks like a map and is not one:

```{r}
#| error: true
data(quakes_fiji) + bar * count + x(east) + map()
```

The message names the four marks that do draw, and it names the way out: put the
quantity on `size` or `color`, which is what the two plots above did.

A position on a map is a place, so a category cannot be one:

```{r}
#| error: true
data(gapminder_2007) + point + x(continent) + y(life) + map()
```

And the panel's shape belongs to the projection:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + map() + theme(ratio = 1)
```

An equal-area map stretched to fit a box is no longer equal-area, though it
still looks like a map. That is the kind of error a reader cannot see, so the
engine refuses it instead of accepting the number and ignoring it.

## Where Mercator stops

`preserve = "angle"` sends the poles infinitely far away, so it has to stop
short of them. It stops at 85.05 degrees, and it says how many rows it could not
reach rather than moving them in silence:

```{r}
far_north <- data.frame(
  lon = c(0, 15, -45, 10),
  lat = c(60, 78, 83, 89))

data(far_north) + point + x(lon) + y(lat) + map(preserve = "angle")
```

Read the message above the plot. One row sits beyond the limit and was drawn at
the edge instead of where it belongs. Nothing was dropped, because dropping it
would misreport how many places there are. `preserve = "area"` reaches both
poles and has no such limit.

## What is not drawn yet

The **choropleth** shades each country by a value: a map of life expectancy, or
of income. It is the plot most people picture when they hear "map", and it is
not drawn yet.

What is missing is not the space. It is the geometry. A choropleth needs a
filled region whose boundary comes from the data, and `zone`, which is the mark
that shades a region, knows four ways to find its sides today. None of them is a
coastline. GOG says exactly that:

```{r}
#| error: true
data(world_borders) + zone + x(lon) + y(lat) + group(piece) +
  color(continent) + map()
```

Until it lands, a quantity on a map goes on `size` or `color` at a point, which
is what this chapter has been doing.
