# Map

```{r}
#| include: false
source("R/setup.R")
```

Where did it happen? Some data answers that with a place on the earth, and a
place is two numbers: **longitude**, how far east or west, and **latitude**, how
far north or south. Both are angles, measured in degrees.

Two numbers is what every plot in this book has taken. So a map is not a new
kind of picture. It is the same sentence you already write, drawn in a space
that knows the two numbers are angles on a sphere.

```{r}
data(quakes_fiji) + point + x(east) + y(north) + map()
```

*"Given quakes Fiji: points, x is east, y is north, in the map space."*

That is a thousand earthquakes recorded near Fiji. `x` is longitude and `y` is
latitude, and `map()` is the whole of what makes it a map. The axes read in
degrees, because that is what the columns hold.

In the web edition, `−`, `+` and *fit* under the plot look closer at a crowded
part of it and return to the whole. A map wants that more than most plots,
because places are crowded at every scale. Zooming is a **view**: it magnifies
the picture and changes nothing the plot claims.

## Flattening the sphere loses something

The earth is round and the page is flat. No projection can flatten a sphere
without distorting it, and the choice is not between a good projection and a bad
one. It is between which mistake you prefer.

Two properties matter most, and a projection can keep one of them, never both.
`preserve` says which:

```{r}
#| layout-ncol: 2
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  map(preserve = "area") + title("preserve = \"area\"")

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  map(preserve = "angle") + title("preserve = \"angle\"")
```

Look at Greenland, at the top left of each map, and compare it with Africa. On
the left, Greenland is small, which is correct: Africa is about fourteen times
larger. On the right, Greenland has grown until it rivals Africa. The right-hand
map has kept every small shape's true form and paid for it in size.

`preserve = "area"` is the default, and the reason is what a reader does with a
map. A map is usually read by comparing how much space something takes. A
projection that inflates the far north tells the reader something false about
every number placed there.

## Drawing the outlines

The world map above is a table like any other. It has one row for every point
along a coastline or a border:

```{r}
knitr::kable(head(world_borders, 4))
```

`path` connects rows in the order the table gives them, which is exactly what
tracing a coastline needs. `group(piece)` says where one outline ends and the
next begins. A country is not always one closed shape: islands are separate
rings, and a country lying wholly inside another is a ring of its own. So
`piece` counts rings rather than countries.

Without that split, `path` would join the last point of one country to the first
point of the next, and draw a line across the ocean between them.

## The second layer is the data

An outline on its own is a backdrop. The plot begins when something is drawn on
top of it, and that is ordinary layering: the outline is one `data()` and one
mark, the observations are another.

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) +
  style(color = "tomato", opacity = 0.35, size = 3) +
  map() + title("A thousand earthquakes, and where Fiji is")
```

The earthquakes now sit somewhere a reader can find. Nothing in the sentence is
new: two tables, two marks, and the second `data()` binds to the mark that
follows it.

## A quantity goes on a channel

Every other space leaves you an axis to measure along. A bar stands on one axis
and measures up the other. A map has no axis to spare, because longitude and
latitude have taken both.

So a quantity moves to a **channel**. Size each place by it:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) + size(magnitude) +
  style(color = "tomato", opacity = 0.4) +
  map() + title("Sized by magnitude")
```

That plot has a name, the **proportional-symbol map**, and there is no atom that
spells it. It is `point` with `size` bound, in this space.

Color works the same way, and reads a category rather than a quantity:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  color(continent) + map() +
  title("Each continent in its own color")
```

## Meridians and parallels

A `rule` draws a line at one position and spans the axis it does not name. Flat,
that is a reference line. Here, a rule at one longitude is a **meridian**, and a
rule at one latitude is a **parallel**:

```{r}
lines_of <- data.frame(lat = 0)

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(lines_of) + rule + y(lat) + style(color = "#d4674f") +
  map() + title("The equator")
```

A `rule` takes its position from a column, like every other mark, so the line
above comes from a one-row table. That is the same way a reference line is drawn
on any other plot.

Nothing was added to the grammar to draw that. The same sentence that puts a
reference line on a scatter puts the equator on a map, because the space bends
what the mark means without changing what the mark is.

## Naming places

`text` writes a string at a position, and a position here is a place:

```{r}
capitals <- data.frame(
  lon  = c(-0.13, 139.69, -74.01, 151.21, 18.42),
  lat  = c(51.51,  35.69,  40.71, -33.87, -33.92),
  name = c("London", "Tokyo", "New York", "Sydney", "Cape Town"))

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(capitals) + point + x(lon) + y(lat) +
  style(color = "#3d5a80", size = 5) +
  data(capitals) + text + x(lon) + y(lat) + label(name) +
  style(nudge = "up", size = 11) +
  map() + title("Five cities")
```

## What a map refuses

A mark that measures along an axis has no axis here. GOG says so rather than
drawing something that looks like a map and is not one:

```{r}
#| error: true
data(quakes_fiji) + bar * count + x(east) + map()
```

The message names the four marks that do draw, and it names the way out: put the
quantity on `size` or `color`, which is what the two plots above did.

A position on a map is a place, so a category cannot be one:

```{r}
#| error: true
data(gapminder_2007) + point + x(continent) + y(life) + map()
```

And the panel's shape belongs to the projection:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + map() + theme(ratio = 1)
```

An equal-area map stretched to fit a box is no longer equal-area, though it
still looks like a map. That is the kind of error a reader cannot see, so the
engine refuses it instead of accepting the number and ignoring it.

## Where Mercator stops

`preserve = "angle"` sends the poles infinitely far away, so it has to stop
short of them. It stops at 85.05 degrees, and it says how many rows it could not
reach rather than moving them in silence:

```{r}
far_north <- data.frame(
  lon = c(0, 15, -45, 10),
  lat = c(60, 78, 83, 89))

data(far_north) + point + x(lon) + y(lat) + map(preserve = "angle")
```

Read the message above the plot. One row sits beyond the limit and was drawn at
the edge instead of where it belongs. Nothing was dropped, because dropping it
would misreport how many places there are. `preserve = "area"` reaches both
poles and has no such limit.

## Filling the regions: the choropleth

Everything so far has drawn *outlines*. Filling each country by a value is the
plot most people picture when they hear "map", and it has a name: the
**choropleth**.

The mark is `zone`, which shades a region. Everywhere else in this book a zone
finds its sides from its own row: a category's slot, `bounds`, `bin` or
`density`. A coastline is a fifth way, and it is the first that arrives as
**many rows for one shape**. So the sentence needs a word for which rows belong
to which region, and `group` is that word already:

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + map() + title("Every country, filled by continent")
```

Compare that with the outline map earlier. Only two things changed: `path`
became `zone`, and `group(piece)` became `group(country)`. A `path` traces one
ring at a time, so it groups by ring. A `zone` fills a whole country, so it
groups by country.

## A value on the map

`color` reads a number as readily as a category, and that is the choropleth
proper. The value comes from a table you already know: `gapminder_2007`, the
same 142 countries the scatter chapters draw.

Joining two tables is the host language's job, not the grammar's. In R that is
one line with `match()`:

```{r}
filled <- world_borders
filled$life <- gapminder_2007$life[match(filled$country,
                                         gapminder_2007$country)]
filled <- filled[!is.na(filled$life), ]

data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  style(color = "#e6e6ec") +
  data(filled) + zone + x(lon) + y(lat) + group(country) + color(life) +
  map() + title("Life expectancy at birth, 2007")
```

Two layers again, and the first one is doing real work. `gapminder_2007` covers
120 of the 176 countries in the outline table, so the pale grey layer underneath
is what the reader sees where there is no number. Without it those countries
would simply be absent, and a reader could not tell "no data" from "no country".

Read the map and the scatter together. The chapter on
[first plots](first-plot.qmd) shows life expectancy against income as a cloud of
points. Here is the same column, and the map answers a question the scatter
cannot: *where*.

## Rings, and the country inside another country

A region is made of rings, and the rings are found by a rule rather than by a
column: a ring ends where it returns to the point it started from. Boundary
data is written that way, so nothing extra has to be said.

That rule is what lets a country contain a hole. Lesotho lies entirely inside
South Africa, and South Africa's boundary has a second ring around it:

```{r}
enclave <- world_borders[world_borders$country %in%
                           c("South Africa", "Lesotho"), ]
enclave$life <- gapminder_2007$life[match(enclave$country,
                                          gapminder_2007$country)]

data(enclave) + zone + x(lon) + y(lat) + group(country) + color(life) +
  map() + title("Lesotho is not painted on top of South Africa")
```

Lesotho is a hole in South Africa, not a patch laid over it. The difference
matters because the two countries have different values, and a patch would be
correct only if it happened to be drawn second.

If a ring in your data does not close, GOG cannot tell where one ring ends and
the next begins. It says so rather than guessing quietly, and names what to fix.
