# Map

```{r}
#| include: false
source("R/setup.R")
```

Where did it happen? Some data answers that with a place on the earth, and a
place is given by two numbers. **Longitude** says how far east or west of
Greenwich a place is, and **latitude** how far north or south of the equator.
Both are angles in degrees, and south and west are negative.

Most plots in this book have placed their marks with two numbers, one on `x` and
one on `y`. So a map is not a new kind of picture. It is the same sentence you
already write, drawn in a space that treats the two numbers as angles on a
sphere.

`quakes_fiji` records earthquakes near Fiji, one row per earthquake:

```{r}
#| echo: false
peek(quakes_fiji)
```

```{r}
data(quakes_fiji) + point + x(east) + y(north) + map()
```

*"Given quakes Fiji: points, x is east, y is north, in the map space."*

That is a thousand earthquakes recorded near Fiji. `x` is longitude and `y` is
latitude, and `map()` is the only word that makes it a map. The axes are labeled
in degrees, because that is what the columns hold. Every `north` value is
negative, because Fiji lies south of the equator. `east` keeps counting past 180
instead of turning negative, so 185 means 175 degrees west.

In the web edition, the buttons under the plot work here too: the two
magnifiers zoom in and out, and the frame button shows the whole map again. Once you
have zoomed in, dragging moves the map around inside the panel, and the frame
button undoes both. A map needs those controls more than most plots do, because a map
is full of detail at every zoom level, from a whole ocean down to one street.
Zooming changes only the **view**: it magnifies the picture and changes nothing
the plot claims.

Zooming cannot change the **projection**, the method that flattens the round
earth onto the page. You choose the projection, and the next section shows what
each choice keeps and what it loses.

## Choosing a projection

The earth is round and the page is flat. No projection can flatten a sphere
without distorting it, and the choice is not between a good projection and a bad
one. You choose which distortion you accept.

Two properties matter most: the area of a region, and the shape of a small
region. A projection can keep one of them, never both. A projection that keeps
every angle keeps every small shape, so `preserve` takes `"area"` or `"angle"`:

```{r}
#| echo: false
peek(world_borders)
```

```{r}
(data(world_borders) + path + x(lon) + y(lat) + group(piece) +
   map(preserve = "area") + title("preserve = \"area\"")) /
  (data(world_borders) + path + x(lon) + y(lat) + group(piece) +
     map(preserve = "angle") + title("preserve = \"angle\""))
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece, in the
map space preserving area, above the same preserving angle."*

Look at Greenland, at the top left of each map, and compare it with Africa.
Under `preserve = "area"`, Greenland is small, which is correct: Africa is about
fourteen times larger. Under `preserve = "angle"`, the **Mercator** projection
most web maps use, Greenland is drawn as large as Africa. The `"angle"` map
keeps the true shape of every small region, and the areas it draws are wrong.

Both plots name `lon` on `x`, so `/` treats that as one axis: one scale, and one
panel width. The projection sets each height, and the two heights differ.
Keeping every small shape is what stretches the second one north and south.

`preserve = "area"` is the default, because a map is usually read by comparing
how much area each region covers. A projection that draws the far north too
large makes every region there look bigger than it is.

The world map above comes from a table like any other, and the next section says
how that table is laid out.

## Drawing the outlines

To draw a boundary of your own, you need a table laid out the way the engine
reads it. `world_borders` holds one row for every point along a coastline or a
border, and it repeats the country and the continent on each of those rows.

`path` connects rows in the order the table gives them, which is exactly what
tracing a coastline needs. A **ring** is one closed loop of points, ending where
it returns to its starting point. `group(piece)` says where one ring ends and
the next begins. A country is not always a single ring. An island is a ring of
its own, and a country that encloses another needs a second ring for the hole.
So `piece` has one value for each ring, not one for each country.

Without `group(piece)`, `path` would join the last point of one country to the
first point of the next. That would draw a line straight across the ocean
between them. Grouping by `country` is not enough either, because a `path`
joins every row of its group in order: South Africa's outline would gain a line
to the ring around Lesotho.

## The second layer is the data

The chapter's first plot drew the earthquakes with no coastline, so nothing said
where on the earth they were. The outline supplies that, and here it is the
background rather than the subject. Your own data goes on top of it, which is
ordinary layering. The outline is one `data()` and one mark, and the earthquakes
are another.

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) +
  style(color = "tomato", opacity = 0.35, size = 3) +
  map() + title("A thousand earthquakes near Fiji")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given quakes Fiji: points, x is east, y is north, in the map space."*

The earthquakes now sit on a world map, north of New Zealand and east of
Australia, so a reader can see where they are. Nothing in the sentence is new:
two tables, two marks, and the second `data()` binds to the mark that follows
it.

## A quantity goes on a channel

The earthquakes differ in magnitude, and the picture so far shows only where
they were. In a flat plot, a cube or a polar plot, an axis can carry a quantity.
A bar has its base on one axis and its length along the other. A map has no axis
left for a quantity, because longitude and latitude use both.

So a quantity moves to a **channel**. Here `size` carries each earthquake's
magnitude:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) + size(magnitude) +
  style(color = "tomato", opacity = 0.4) +
  map() + title("Sized by magnitude")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given quakes Fiji: points, x is east, y is north, size by magnitude, in the map
space."*

That plot has a name, the **proportional-symbol map**, and no single atom draws
it. It is `point` with `size` bound, in this space.

Color works the same way. Here the column bound to it is a category, so each
continent is given its own hue:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  color(continent) + map() +
  title("Each continent in its own color")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece, color by
continent, in the map space."*

`color` reads the category exactly as it does on a flat plot. The space changed;
the channel did not. The legend's seventh entry, *Seven seas (open ocean)*, is
not a continent. The outline comes from Natural Earth, which puts the French
Southern and Antarctic Lands under that name. They are islands in the southern
Indian Ocean.

## Meridians and parallels

A map often needs a reference line, and the most common one is the equator. A
`rule` draws a line at one position and spans the other axis. On a flat plot
that is a reference line. Here, a `rule` at one latitude is a
**parallel**, and a `rule` at one longitude is a **meridian**. The plot below
draws one parallel, the equator:

```{r}
#| echo: false
peek(equator)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(equator) + rule + y(lat) + style(color = "#d4674f") +
  map() + title("The equator")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the equator: a rule, y is lat, in the map space."*

A `rule` takes its position from a column, like every other mark, so the
equator comes from a one-row table. Nothing was added to the grammar to draw it:
the sentence that puts a reference line on a scatter puts the equator on a map.
The space gives the mark a new meaning, and the mark itself is unchanged.

## Naming places

A coastline alone does not say which part of the world it is. A few city names
tell the reader where to look. `text` writes a string at a position, and a
position here is a place:

```{r}
#| echo: false
peek(capitals)
```

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(capitals) + point + x(lon) + y(lat) +
  style(color = "#3d5a80", size = 5) +
  data(capitals) + text + x(lon) + y(lat) + label(name) +
  style(nudge = "up", size = 11) +
  map() + title("Five capitals")
```

*"Given the world borders: paths, x is lon, y is lat, grouped by piece; then
given the capitals: points, x is lon, y is lat; then given the capitals: text, x
is lon, y is lat, label by name, in the map space."*

Each city is drawn by two layers: a point for the place, and text above it for
the name. `style(nudge = "up")` moves each name up so that it does not cover its
dot.

## What a map refuses

You may want a histogram of the earthquakes by longitude, which on a flat plot
is `bar * bin`. A mark that measures along an axis needs one of its own, and on
a map longitude and latitude use both. gog says so rather than drawing something
that looks like a map and is not one:

```{r}
#| error: true
data(quakes_fiji) + bar * bin + x(east) + map()
```

The message names the marks that do draw, and it says what to do instead. Put
the quantity on `size`, as the magnitude map above does, or on `color`.

A category cannot be a position on a map either, because a position there is a
place:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
#| error: true
data(gapminder_2007) + point + x(continent) + y(life) + map()
```

The projection also sets the panel's width divided by its height, which
`theme(ratio = )` would change:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + map() + theme(ratio = 1)
```

A map drawn with `preserve = "area"` and stretched into a square panel no longer
keeps area, though it still looks like a map. A reader cannot see that error, so
the engine refuses `ratio` on a map instead of accepting the number and ignoring
it.

## Where Mercator stops

Some sentences are drawn with a message instead of refused. Mercator has a limit
that `preserve = "area"` does not. It would draw the poles infinitely far from
the equator, so it has to stop before it reaches them. It stops at 85.05 degrees
north and south. A row beyond that is drawn at the edge, and gog reports how
many rows were moved that way:

```{r}
#| echo: false
peek(far_north)
```

```{r}
#| message: true
data(far_north) + point + x(lon) + y(lat) + map(preserve = "angle")
```

Read the message above the plot. One row sits beyond the limit and was drawn at
the edge instead of where it belongs. Nothing was dropped, because dropping it
would misreport how many places there are. `preserve = "area"` reaches both
poles and has no such limit.

## Filling the regions: the choropleth

Every country so far has been drawn as an *outline*. Filling each country by a
quantity is the most familiar kind of map, and it has a name: the
**choropleth**. The map below fills each country by its continent first; the
next section fills it by a quantity.

The mark is `zone`, which shades a region. Everywhere else in this book a zone
takes its sides from its own row: a category's slot, a pair of `bounds`, a bin,
a density cell, or a region that `partition` or `flow` computes. A country's
outline is different: it is the only case where many rows make one shape. So
the sentence needs a word for which rows belong to which region, and `group` is
that word already:

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + map() + title("Every country, filled by continent")
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country, color
by continent, in the map space."*

Compare that with the earlier map that colored each continent. Apart from the
title, only two things changed: `path` became `zone`, and `group(piece)` became
`group(country)`. A `path` traces one ring at a time, so it groups by ring. A
`zone` fills a whole country, so it groups by country.

## Life expectancy on the map

Where in the world do people live longest? `color` accepts a quantity as well
as a category, so the same sentence can shade each country by life expectancy.
That column comes from a table you already know: `gapminder_2007`, the same 142
countries that the chapter [A first plot](first-plot.qmd) draws.

Joining two tables is the host language's job, not the grammar's, so this chunk
is shown in R only. In R the join is one call to `match()`, and each of the
other languages has its own way to look up one table's values in another. The
plot sentence under the join is the same sentence in every language, spelled
the way the tabs above spell it:

```{r}
filled <- world_borders
filled$life <- gapminder_2007$life[match(filled$country,
                                         gapminder_2007$country)]
filled <- filled[!is.na(filled$life), ]

data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  style(color = "#e6e6ec") +
  data(filled) + zone + x(lon) + y(lat) + group(country) + color(life) +
  map() + title("Life expectancy at birth, 2007")
```

Two layers again. The first draws all 176 countries in pale gray, and the second
colors the 120 whose names `match()` found in `gapminder_2007`. A name matches
only when both tables spell it the same way. The outline says *United States of
America* and `gapminder_2007` says *United States*, so the United States stays
gray, and so do 14 other countries whose names the two tables spell differently.
Without the gray layer, the 56 unmatched countries would simply be absent, and a
reader could not tell "no match" from "no country".

Compare the map with the scatter plot in [A first plot](first-plot.qmd), which
shows life expectancy against GDP per capita, one point per country. The map
shows that same `life` column. It answers a question the scatter cannot:
*where*.

## Two years on one color scale

The map above shows a single year. Did life expectancy rise everywhere between
1957 and 2007, or only in some places? To compare two years, draw the same map
once for each year, as two panels of one plot. `gm_eras` holds life expectancy
for both years, one row per country and year:

```{r}
#| echo: false
peek(gm_eras)
```

It stores the year twice: `year` is a number, and `era` is the same year as a
label. A facet needs the label, because a facet column names its panels
([Faceting](faceting.qmd#the-facet-columns-type)).

The join works as in the section above, and is again shown in R only. Each year
is joined to its own copy of the outline, and the table `eras` holds both
copies. Its `era` column says which year each copy shows:

```{r}
#| message: true
e57 <- gm_eras[gm_eras$era == "1957", ]
e07 <- gm_eras[gm_eras$era == "2007", ]
then <- world_borders
then$era <- "1957"
then$life <- e57$life[match(then$country, e57$country)]
now <- world_borders
now$era <- "2007"
now$life <- e07$life[match(now$country, e07$country)]
eras <- rbind(then, now)
eras <- eras[!is.na(eras$life), ]

data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  style(color = "#e6e6ec") +
  data(eras) + zone + x(lon) + y(lat) + group(country) + color(life) +
  palette("viridis") + map() +
  title("Life expectancy at birth, 1957 and 2007") / facet(era)
```

*"Given the world borders: zones, x is lon, y is lat, grouped by country; then
given the eras table: zones, x is lon, y is lat, grouped by country, color by
life, with the viridis palette, in the map space, split into panel rows by
era."*

`/ facet(era)` stacks the two panels, one above the other. In the default
projection a world map is about twice as wide as it is tall, so side by side
each map would be about half as wide. Both panels share one color scale, and
that is what makes them comparable: the same shade means the same life
expectancy in 1957 and in 2007.

`palette("viridis")` colors the shortest life expectancy dark purple and the
longest yellow. In the map above, dark meant the longest. Dark purple is clearly
different from the pale gray of the unmatched countries, so a short life
expectancy never looks like a missing value.

Most countries are greener or yellower in 2007 than in 1957. Counted from the
table, life expectancy rose in 117 of the 120 colored countries, and the median
rose from 48 years to 72. Life expectancy fell only in Lesotho, Zambia and
Zimbabwe, all three in the south of Africa.

The gray layer appears in both panels. Its table, `world_borders`, has no `era`
column, so gog draws that layer in every panel and says so in a message. [A
layer without the facet column](faceting.qmd#a-layer-without-the-facet-column)
explains the rule.

## Rings, and the country inside another country

A `zone` groups by `country`, not by `piece`, so nothing in its sentence says
where one ring ends. gog finds each ring by closure: a ring ends where it
returns to the point it started from. Boundary data is written that way, so
nothing extra has to be said.

Finding rings by closure is what lets a country contain a hole. Lesotho, one of
the three countries above, lies entirely inside South Africa, and South Africa's boundary has a second ring
around Lesotho:

```{r}
enclave <- world_borders[world_borders$country %in%
                           c("South Africa", "Lesotho"), ]
enclave$life <- gapminder_2007$life[match(enclave$country,
                                          gapminder_2007$country)]

data(enclave) + zone + x(lon) + y(lat) + group(country) + color(life) +
  map() + title("Lesotho is not painted on top of South Africa")
```

Lesotho is a hole in South Africa, not a patch laid over it. The difference
matters because the two countries have different values. A patch is right only
when it is drawn after the region beneath it, and nothing in the sentence says
which region is drawn first.

Real boundary data is not always written this carefully. If a ring in your data
does not close, gog cannot tell where one ring ends and the next begins. It
draws the region, warns you that the ring ends had to be guessed, and names what
to fix.

## What comes next

A map flattens the sphere so a page can hold it. The next chapter,
[Globe](globe.qmd), draws every mark from this chapter on the sphere itself: the
same marks at the same places, seen from one side.
