# Map

```{r}
#| include: false
source("R/setup.R")
```

Where did it happen? Some data answers that with a place on the earth, and a
place is two numbers: **longitude**, how far east or west, and **latitude**, how
far north or south. Both are angles, measured in degrees.

Every plot in this book has used two numbers for position. So a map is not a new
kind of picture. It is the same sentence you already write, drawn in a space
that treats the two numbers as angles on a sphere.

```{r}
data(quakes_fiji) + point + x(east) + y(north) + map()
```

*"Given quakes Fiji: points, x is east, y is north, in the map space."*

That is a thousand earthquakes recorded near Fiji. `x` is longitude and `y` is
latitude, and `map()` is the whole of what makes it a map. The axes are labeled
in degrees, because that is what the columns hold.

In the web edition, `−`, `+` and *fit* under the plot magnify a crowded part and
then show the whole map again. A map needs that more than most plots do,
because places are crowded at every scale. Zooming is a **view**: it magnifies
the picture and changes nothing the plot claims.

## Flattening the sphere loses something

The earth is round and the page is flat. No projection can flatten a sphere
without distorting it, and the choice is not between a good projection and a bad
one. You choose which distortion you accept.

Two properties matter most, and a projection can keep one of them, never both.
`preserve` says which:

```{r}
(data(world_borders) + path + x(lon) + y(lat) + group(piece) +
   map(preserve = "area") + title("preserve = \"area\"")) /
  (data(world_borders) + path + x(lon) + y(lat) + group(piece) +
     map(preserve = "angle") + title("preserve = \"angle\""))
```

The world map above is a table like any other, and the next section reads it
column by column.

Look at Greenland, at the top left of each map, and compare it with Africa. In
the first, Greenland is small, which is correct: Africa is about fourteen times
larger. In the second, Greenland has grown until it looks as large as Africa.
The second map keeps every small shape's true form and gets the sizes wrong.

The two are the same width and not the same height, and the projection sets that
height. Both cover every degree of longitude, so `/`
gives them one `lon` axis and one width. Keeping every small shape's form is
what stretches the second one north and south.

`preserve = "area"` is the default, and the reason is how a reader uses a
map. A map is usually read by comparing how much space something takes. A
projection that inflates the far north tells the reader something false about
every number placed there.

## Drawing the outlines

The table has one row for every point along a coastline or a border:

```{r}
knitr::kable(head(world_borders, 4))
```

`path` connects rows in the order the table gives them, which is exactly what
tracing a coastline needs. `group(piece)` says where one outline ends and the
next begins. A country is not always one closed shape: islands are separate
rings, and a country that encloses another needs a second ring for the hole (a
ring ends where it returns to its starting point). So `piece` counts rings
rather than countries.

Without that split, `path` would join the last point of one country to the first
point of the next, and draw a line across the ocean between them.

## The second layer is the data

An outline on its own is only a background. It becomes a plot when something is
drawn on top of it, and that is ordinary layering: the outline is one `data()`
and one mark, the observations are another.

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) +
  style(color = "tomato", opacity = 0.35, size = 3) +
  map() + title("A thousand earthquakes, and where Fiji is")
```

The earthquakes now sit somewhere a reader can find. Nothing in the sentence is
new: two tables, two marks, and the second `data()` binds to the mark that
follows it.

## A quantity goes on a channel

Every other space leaves you an axis to measure along. A bar stands on one axis
and measures its length along the other. A map has no axis left, because
longitude and latitude use both.

So a quantity moves to a **channel**. Size each place by it:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(quakes_fiji) + point + x(east) + y(north) + size(magnitude) +
  style(color = "tomato", opacity = 0.4) +
  map() + title("Sized by magnitude")
```

That plot has a name, the **proportional-symbol map**, and no single atom draws
it. It is `point` with `size` bound, in this space.

Color works the same way, and takes a category rather than a quantity:

```{r}
data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  color(continent) + map() +
  title("Each continent in its own color")
```

## Meridians and parallels

A `rule` draws a line at one position and spans the axis it does not name. In a
flat space that is a reference line. Here, a rule at one longitude is a
**meridian**, and a
rule at one latitude is a **parallel**:

```{r}
lines_of <- data.frame(lat = 0)

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(lines_of) + rule + y(lat) + style(color = "#d4674f") +
  map() + title("The equator")
```

A `rule` takes its position from a column, like every other mark, so the line
above comes from a one-row table. That is the same way a reference line is drawn
on any other plot.

Nothing was added to the grammar to draw that. The same sentence that puts a
reference line on a scatter puts the equator on a map. The space gives the mark
a new meaning, and the mark itself is unchanged.

## Naming places

`text` writes a string at a position, and a position here is a place:

```{r}
capitals <- data.frame(
  lon  = c(-0.13, 139.69, -74.01, 151.21, 18.42),
  lat  = c(51.51,  35.69,  40.71, -33.87, -33.92),
  name = c("London", "Tokyo", "New York", "Sydney", "Cape Town"))

data(world_borders) + path + x(lon) + y(lat) + group(piece) +
  style(color = "#c9ccd6") +
  data(capitals) + point + x(lon) + y(lat) +
  style(color = "#3d5a80", size = 5) +
  data(capitals) + text + x(lon) + y(lat) + label(name) +
  style(nudge = "up", size = 11) +
  map() + title("Five cities")
```

## What a map refuses

A mark that measures along an axis has no axis here. gog says so rather than
drawing something that looks like a map and is not one:

```{r}
#| error: true
data(quakes_fiji) + bar * count + x(east) + map()
```

The message names the marks that do draw, and it says what to do instead: put
the quantity on `size` or `color`, which is what the two plots above did.

A position on a map is a place, so a category cannot be one:

```{r}
#| error: true
data(gapminder_2007) + point + x(continent) + y(life) + map()
```

And the panel's shape belongs to the projection:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + map() + theme(ratio = 1)
```

An equal-area map stretched to fit a box is no longer equal-area, though it
still looks like a map. That is the kind of error a reader cannot see, so the
engine refuses it instead of accepting the number and ignoring it.

## Where Mercator stops

`preserve = "angle"` is the **Mercator** projection, the one behind most web
maps. It sends the poles infinitely far away, so it has to stop short of them.
It stops at 85.05 degrees, and it says how many rows it could not reach rather
than moving them and saying nothing:

```{r}
far_north <- data.frame(
  lon = c(0, 15, -45, 10),
  lat = c(60, 78, 83, 89))

data(far_north) + point + x(lon) + y(lat) + map(preserve = "angle")
```

Read the message above the plot. One row sits beyond the limit and was drawn at
the edge instead of where it belongs. Nothing was dropped, because dropping it
would misreport how many places there are. `preserve = "area"` reaches both
poles and has no such limit.

## Filling the regions: the choropleth

Everything so far has drawn *outlines*. Filling each country by a value is the
plot most people picture when they hear "map", and it has a name: the
**choropleth**.

The mark is `zone`, which shades a region. Everywhere else in this book a zone
takes its sides from its own row: a category's slot, `bounds`, `bin` or
`density`. A coastline is a fifth way, and it is the first that arrives as
**many rows for one shape**. So the sentence needs a word for which rows belong
to which region, and `group` is that word already:

```{r}
data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  color(continent) + map() + title("Every country, filled by continent")
```

Compare that with the outline map earlier. Only two things changed: `path`
became `zone`, and `group(piece)` became `group(country)`. A `path` traces one
ring at a time, so it groups by ring. A `zone` fills a whole country, so it
groups by country.

## A value on the map

`color` reads a number as readily as a category, and that is a true choropleth.
The value comes from a table you already know: `gapminder_2007`, the
same 142 countries the scatter chapters draw.

Joining two tables is the host language's job, not the grammar's. In R the join
itself is one `match()` call. Every host language has its own one-call lookup,
and the plot sentence below it is the same in all four:

```{r}
filled <- world_borders
filled$life <- gapminder_2007$life[match(filled$country,
                                         gapminder_2007$country)]
filled <- filled[!is.na(filled$life), ]

data(world_borders) + zone + x(lon) + y(lat) + group(country) +
  style(color = "#e6e6ec") +
  data(filled) + zone + x(lon) + y(lat) + group(country) + color(life) +
  map() + title("Life expectancy at birth, 2007")
```

Two layers again, and the first one is not decoration. `gapminder_2007` covers
120 of the 176 countries in the outline table, so the pale gray layer underneath
is what the reader sees where there is no number. Without it those countries
would simply be absent, and a reader could not tell "no data" from "no country".

Read the map and the scatter together. The chapter on
[first plots](first-plot.qmd) shows life expectancy against income as a cloud of
points. Here is the same column, and the map answers a question the scatter
cannot: *where*.

## Rings, and the country inside another country

A region is made of rings, and the rings are found by a rule rather than by a
column: a ring ends where it returns to the point it started from. Boundary
data is written that way, so nothing extra has to be said.

That rule is what lets a country contain a hole. Lesotho lies entirely inside
South Africa, and South Africa's boundary has a second ring around it:

```{r}
enclave <- world_borders[world_borders$country %in%
                           c("South Africa", "Lesotho"), ]
enclave$life <- gapminder_2007$life[match(enclave$country,
                                          gapminder_2007$country)]

data(enclave) + zone + x(lon) + y(lat) + group(country) + color(life) +
  map() + title("Lesotho is not painted on top of South Africa")
```

Lesotho is a hole in South Africa, not a patch laid over it. The difference
matters because the two countries have different values, and a patch would be
correct only when it is drawn second, and nothing guarantees that.

If a ring in your data does not close, gog cannot tell where one ring ends and
the next begins. It says so rather than guessing without saying so, and names
what to fix.

Every sentence in this chapter also draws with the sphere kept round: that is
[Globe](globe.qmd), the same marks at the same places, viewed.
