# Nest

```{r}
#| include: false
source("R/setup.R")
```

What share of the whole is each part? A plot has to show that share with
something you can see. On a flat panel the share is a **length**: a segment's
height as a fraction of the column it sits in. On a panel bent into a circle it
is an **angle**, and the column becomes a pie. There is a third way, and it fits
the most categories into the least space: **area**.

That is what `nest()` does. Each row's measure becomes the size of a region,
and the regions fill the panel with no gaps. Dividing a panel that way is a
**packing**. The plot has a name, **treemap**, and as with the pie there is no
atom that spells it.

The treemap began with a real and ordinary problem. In 1990 the 80-megabyte hard
disk in Ben Shneiderman's lab, shared by 14 people, was full, and a directory
listing could not show what was filling it. So he worked out a way to draw the
disk as a rectangle packed with nested rectangles, each sized by what it holds
[@shneiderman1992]. The treemap is now the usual way to draw a share of a whole
when there are more categories than a pie has room for. This chapter teaches
that plot.

## One sentence, three spaces

How is the world's population divided among the continents? You already know
one plot that shows a share: the stacked bar. Start there, because the same
sentence will answer the question three times. Here is the world's population as
one stacked bar, a single column divided into one part for each continent:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * sum * stack + y(population) + color(continent) +
  title("A column divided")
```

*"Given gapminder 2007: bars derived by sum and stack, y is population, color by
continent."*

Now the same sentence in the circle. The column's height becomes the whole turn,
and each segment's share of the height becomes its slice's share of the circle:

```{r}
data(gapminder_2007) + bar * sum * stack + y(population) + color(continent) +
  polar() + title("The same shares, as angles")
```

*"Given gapminder 2007: bars derived by sum and stack, y is population, color by
continent, in polar."*

And the same sentence again, packed into areas. One word has gone from it, and
the end of this section says which word and why:

```{r}
data(gapminder_2007) + bar * sum + y(population) + color(continent) +
  nest() + title("The same shares, as areas")
```

*"Given gapminder 2007: bars derived by sum, y is population, color by
continent, in the nest."*

Three pictures, one grammar. No sentence names a chart. Each says a bar, summed
and colored by continent, and then its space: `polar()`, `nest()`, or no word at
all for the flat page. Asia is three fifths of the world's population in all
three, and you can check that against the column, the pie and the treemap in
turn.

The word missing from the third sentence is `stack`: the flat and polar forms
carry it, and the packed form does not. `stack` decides what happens when two
marks are drawn at the same place, so it assumes two marks can share a place. A
packing gives every region a place of its own, so two marks never collide.
`stack` would have nothing to resolve, and gog refuses it here.

## Why area, when a column already works

Read the column again and look for Oceania. It holds 0.4% of the world's
population, so its segment is about two pixels tall. The border drawn around it
takes about a quarter of that height. In the packing the same share is a region
about six pixels wide, because a column has only its length to divide, and a
packing divides both width and height.

Oceania's region is three times as wide as its segment was tall, and it still
reports the same 0.4%. With only five parts, that extra width is all a packing
adds. The advantage grows with the number of parts. A packing is for the case
where there are *many* shares and the question is which of them are large. In a
column, every new category makes the segments thinner, and small ones become
too thin to see long before they would in a packing.

Area has one weakness, the same one the pie has: the eye compares two lengths
more accurately than two areas. If the comparison has to be precise, draw the
flat column.

## Two levels, from a position

How much did each continent grow between 1957 and 2007? A packing can show that
by nesting one level of regions inside another. Binding a position is what adds
the outer level. Each category gets a region sized by its own total, and the
rows inside it are packed into that region. Here each continent's region is
divided between its two years:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * sum + x(continent) + y(population) + color(era) +
  nest() + title("Two eras, inside each continent")
```

*"Given the gapminder eras: bars derived by sum, x is continent, y is
population, color by era, in the nest."*

Each continent's region is sized by its two years added together. No continent
ever had that population at one time, so the region's size is only a frame, and
its two parts are what the plot compares. The thicker light lines mark the
continents, and the thinner ones divide each continent between the two eras. The
legend decodes only the eras, and no continent is named in the plot. Asia fills
the left side. Europe, Africa and the Americas fill the right side, from top to
bottom, and Oceania is a thin strip, about seven pixels wide, at the bottom
right. Africa's region is the most uneven: more than three quarters of it is the
2007 part, because its population more than tripled in those fifty years.

The nesting needs no new word. Levels inside levels are called a **dimension
hierarchy**, and Wilkinson shows that a dimension hierarchy *is* a nested facet
[@wilkinson2005]. gog's facets always cross, so here the nesting happens inside
one panel. `x` divides that panel as an outer facet would, but by share, not
into panels of equal size.

The inner level comes from `color`, and the legend is what decodes it. So the
inner level needs few enough values to fit in the legend beside the plot. A
common treemap puts countries inside continents, and its legend would need 142
entries. The section [Naming the regions](#naming-the-regions), later in this
chapter, shows how to write each name inside its own region instead.

## What a packed panel does not have

Look for the axes. There are none, and no setting adds them.

A flat panel's horizontal direction *means* something: it stands for a
variable, and a tick is a place on it. In a packing neither direction is a
variable. Wilkinson
states it exactly: the two dimensions of a treemap "have no intrinsic meaning
related to the data because they can be reordered without changing the metric
properties of the tree" [@wilkinson2005]. That two regions touch says nothing
about how close they are in the data. Moving one to the other side of the panel
would not change any share the plot reports.

So there is no tick to place and no axis label to print, and gog refuses
`x_label()`:

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population) + color(continent) + nest() +
  x_label("Population")
```

The color legend decodes a packed panel, exactly as it decodes a pie.

## Naming the regions

A legend works while it has few enough entries to read. With many entries, the
reader has to search the legend for each color, and 142 countries are far too
many to search. So put the name where the region is. This sentence has no `sum`:
without it, each row is its own region, so every country gets one inside its
continent.

```{r}
data(gapminder_2007) + bar + x(continent) + y(population) + color(continent) +
  text + label(country) + style(color = "white") + nest() +
  title("Countries inside continents")
```

*"Given gapminder 2007: bars, x is continent, y is population, colored by
continent, and also text, label by country, in the nest."*

The sentence has two layers, and the second is the ordinary `text` mark: `label`
says which column supplies the string, exactly as it does on a scatter. `label`
does not say where to put the name. In a nest, the packing decides. A label goes
to the center of the region its own row was packed into. That is the same region
the bar drew. Both marks read one packing, so no name is drawn inside another
country's region.

Notice what is missing from that sentence: the text layer carries no `x` and no
`y` of its own. The sentence's `x(continent)` and `y(population)` place the names
instead. `x` chooses the continent's region, and `y` sizes each country's region
inside it. Once a region has a size and a place, the name inside it has a place
too. On a flat panel a `text` mark needs two
positions before it can be drawn at all. Here it needs a measure and a string,
and the packing places it.

All 142 countries are named. A name that does not fit its region is drawn
smaller, at the size that fits. Most of the small names can be read when you
zoom in on the plot, and the few smallest stay too small even then. The largest countries come first inside each continent, so
their regions are close to square and their names keep their full size. A
sunburst names its sectors by the same rule ([Naming the
nodes](polar.qmd#naming-the-nodes)).

A share too small to have a region at all has nowhere to put its name. Only then
is a name left out, and a message above the plot says how many names were drawn.
Without that message, a treemap with a missing name would read as a complete
list.

Two things make the small names larger: fewer categories, and a larger plot with
`theme(width =, height =)`. Each one gives every region more room.
`style(size = )` sets the size a name is drawn at when it fits, and a name never
grows past its region. gog has no separate setting for fitting names into
regions.

## The other refusals

`x_label()` is one of several refusals in a nest, and the three below come from
one fact: a region has an area, and nothing else. A `point` is drawn at the
coordinates it is given, and a packing has no coordinates to give:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + nest()
```

An area cannot be negative. A measure with values below zero cannot be packed
into regions. On the flat page, a bar below the baseline points down, and the
reader can see the sign. A region has no direction to point in, so nothing in
the picture could show the sign. The parts would also no longer fit the whole:
in the `cashflow` table below, the positive parts add up to 165, and the whole
is only 101.

```{r}
#| echo: false
peek(cashflow)
```

```{r}
#| error: true
data(cashflow) + bar * sum + y(delta) + color(step) + nest()
```

A log scale is refused too. A region's size is its share of the total, and that
share is computed from the values as they are. A log scale places the values so
that equal distances stand for equal ratios. A packing draws no distances, so a
log scale has nothing to change, and gog refuses it.

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population, scale = "log") +
  color(continent) + nest()
```

## Sorting the regions

Which region goes where? Two things decide: the packing method, which is fixed,
and the order of the categories, which you can change. The method is the
*squarified* one of Bruls, Huizing and van Wijk [@bruls2000]: regions are placed
in rows, and each row is chosen so its rectangles are as close to square as they
can be. A simpler method divides one direction at a time, which makes long thin
regions, and a long thin area is the hardest kind to judge. gog uses the
squarified method for every packing.

Squarified packing gives the squarest regions when the rows arrive sorted by
size, which is why published treemaps sort largest first. Inside a region, the
rows stand on no position, so nothing else gives them an order, and gog packs
them largest first. The regions themselves are different. They are the
categories of a position, and they follow the order in which the continents
first appear in the table, or a declared order if the column has one. To sort
them, you add `order()`:

```{r}
data(gapminder_2007) + bar * sum + x(continent) + y(population) +
  color(continent) + order(population, desc = TRUE) + nest() +
  title("Largest continent first")
```

*"Given gapminder 2007: bars derived by sum, x is continent, y is
population, color by continent, ordered by population, largest first, in the
nest."*

That is the same `order()` that reorders a bar chart, and it does the same work
here. If the packing sorted on its own, `order()` would change nothing in this
space. The same word would then mean one thing on a bar chart and nothing in a
nest.

`order()` sorts the categories on a position, `x` or `y`, so a plot needs a
categorical position before it can be sorted. In a nest that position draws no
axis, but its categories still have an order. A one-level treemap binds no
position at all, so there the atom has nothing to sort:

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population) + color(continent) +
  order(population, desc = TRUE) + nest()
```

That refusal does not come from the packed space. It comes from `order()`, and
it applies in every space.

## What comes next

Regions in a nest are sized by the measure, and the packing decides where each
one goes. [Map](map.qmd) places a mark at the point its own row names instead:
longitude and latitude, projected onto the page.
