# Nest

```{r}
#| include: false
source("R/setup.R")
```

How much of the whole does each part take? A share of a whole has to be carried
by something you can see. Flat, it is a **length**: a segment's height as a
fraction of the column it sits in. Bent into a circle, it is an **angle**, and
the column becomes a pie. There is a third answer, and it is the one that fits
the most categories into the least page: **area**.

That is what `nest()` draws. Each row's measure becomes the size of a region,
and the regions partition the panel. The plot has a name, treemap, and as with
the pie there is no atom that spells it.

That name has a precise origin and a mundane one. Ben Shneiderman's 80-megabyte
hard disk was full in 1990, and a directory listing could not tell him what was
filling it, so he wrote a display that packs a rectangle with nested rectangles
sized by what they hold [@shneiderman1992]. Thirty years on it is how a share of a
whole gets drawn whenever there are more categories than a pie can carry, which is
the argument this chapter opened with, arrived at from the other end.

## One sentence, three spaces

Here is the population of each continent as a single column, divided into its
parts:

```{r}
data(gapminder_2007) + bar * sum * stack + y(population) + color(continent) +
  title("A column divided")
```

*"Given gapminder 2007: bars derived by sum and stacked, y is population, color by continent."*

Now the same sentence in the circle. The column's height becomes the whole turn,
and each segment's share of the height becomes its slice's share of the circle:

```{r}
data(gapminder_2007) + bar * sum * stack + y(population) + color(continent) +
  polar() + title("The same shares, as angles")
```

And the same sentence again, packed:

```{r}
data(gapminder_2007) + bar * sum + y(population) + color(continent) +
  nest() + title("The same shares, as areas")
```

Three pictures, one grammar. Nothing in any of them names a chart: a bar, summed,
colored by continent, and then a word for the space. Asia is a little over half
the world's population in all three, and you can check that against the column,
the pie and the packing in turn.

One word did change, and it is worth the sentence: the flat and polar forms carry
`stack` and the packed form does not. `stack` decides what happens when two marks
land in the same place, which presupposes that they can. A packing gives every
piece a region of its own, so there is no collision left to resolve, and GOG
refuses `stack` here rather than accepting a word with nothing to do.

## Why area, when a column already worked

Read the column again and look for Oceania. At 0.3% of the world it is about
**two pixels** tall, which is thinner than the line drawn around it. In the
packing the same share is a strip about **six pixels** wide, because a column
spends its whole length on one dimension and a packing has two to spend.

Three times better is honest, and it is not a transformation. With five parts a
packing buys you a little; the gain grows with the number of parts, because a
column can only ever get longer while a panel gets subdivided in both directions
at once. A packing is for the case where there are *many* shares and the question
is which of them are large, and where a column would have run out of pixels long
before it ran out of categories.

The cost is on the other side, and it is the objection that has always been made
to the pie: two lengths can be compared exactly, and two areas cannot. If the
comparison has to be precise, the flat column is still the plot, and this space
is not trying to replace it.

## Two levels, from a position

Bind a position and the packing gains a level. Each category gets a region sized
by its own total, and the rows inside it are packed into that region:

```{r}
data(gm_eras) + bar * sum + x(continent) + y(population) + color(era) +
  nest() + title("Two eras, inside each continent")
```

Each continent is a block sized by its total across both eras, and inside it the
two eras are packed against each other. The heavier lines are the continents; the
thinner ones divide what is inside them. Asia's block is not only the largest, it
is the most lopsided: half a century put far more people into its 2007 half than
into any other continent's.

The nesting is not new vocabulary. Wilkinson makes the point directly
[@wilkinson2005]: a
dimension hierarchy *is* a nested facet, so `x` here is doing what the outer
level of a facet does, with the panel divided by share instead of into equal
boxes.

The inner level comes from `color`, which sets what the second level can
usefully be when the legend is what decodes it: a split with few enough values
to read down the side. Countries inside continents, the treemap most often
printed, would be a legend of 142 entries, and the section after next names the
regions in place instead.

## What a packed panel does not have

Look for the axes. There are none, and that is the space's defining property
rather than a setting you could switch back on.

A flat panel's horizontal direction *means* something: it is a variable, and a
tick is a place on it. In a packing neither direction is a variable. Wilkinson
puts it as sharply as it can be put: the two dimensions of a treemap "have no
intrinsic meaning related to the data because they can be reordered without
changing the metric properties of the tree" [@wilkinson2005]. Two regions that touch are not near
each other in the data, and moving one to the other side of the panel would not
change a single share the plot reports.

So there is nothing to tick and nothing to label, and GOG refuses to pretend
otherwise:

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population) + color(continent) + nest() +
  x_label("Population")
```

What decodes a packed panel is the color legend, exactly as it decodes a pie.

## Naming the regions

A legend works while there are few enough entries to scan. Past that it stops
being a key and becomes a second plot to read, and the countries of the world
are well past it. So put the name where the region is:

```{r}
#| message: true
data(gapminder_2007) + bar + x(continent) + y(population) + color(continent) +
  text + label(country) + style(color = "white") + nest() +
  title("Countries inside continents")
```

Two layers, and the second is the ordinary `text` mark: `label` says which
column supplies the string, exactly as it does on a scatter. What it does *not*
say is where to put it, because in this space nothing does. A label goes to the
center of the region its own row was packed into, which is the same region the
bar drew — one packing, read by both marks, so a name cannot end up in a
rectangle that belongs to someone else.

Notice what is missing from that sentence: there is no `x` on the text layer and
no `y` either, and the plot's own `y(population)` is doing both jobs at once. It
sizes the regions, and sizing them is what places the labels. Everywhere else in
this book a `text` mark needs two positions before it can be drawn at all; here
it needs a measure and a string, and a packing does the rest.

Count the names and there are far fewer than 142, which is what the line above
the plot is for. Most of the world's countries have a share narrower than
their own name, so a label that does not fit its rectangle is not drawn. The
engine says how many it left out, rather than letting you read the named
countries as though they were all of them. Nigeria and Ethiopia are named and
most of Africa is not, and every one of those unnamed countries still has a
region exactly as wide as its share of the world. The picture is complete and
the labeling is not, and the plot is the thing that tells you which.

That message is why this section is not the trap it looks like. A treemap that
printed the two dozen names that happened to fit, and said nothing about the
rest, would read as a complete labeling of two dozen countries.

Three things move the number, and each is an atom you have already met: fewer
categories, a bigger canvas with `theme(width =, height =)`, or smaller type
with `style(size = )`. None of them is a setting for *labels* in particular, and
none of them will make a name fit a region narrower than the word.

## The other refusals

A packing cannot hold a mark that is placed by a position, because it has no
positions to place one at:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + nest()
```

An area cannot be negative, so a measure that runs below zero has no packing.
Flat, a bar below the baseline reads as a loss; a region has no direction to run
in, and a treemap whose parts no longer sum to the whole is a treemap that lies.

A log scale is refused too, and the reason is the one worth remembering: a
region's size is its **share of the total**, which is arithmetic on the values
themselves. Re-spacing them changes what a distance means, and a packing has no
distances. Accepting the scale would have left the picture byte-for-byte
identical, which is the one thing the engine will not do quietly.

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population, scale = "log") +
  color(continent) + nest()
```

## The order is yours

Published treemaps usually sort by size, largest first, because the packing
algorithm produces squarer rectangles that way. GOG does not do it for you: the
regions follow the order the categorical axis already has, so the sort is
something you say rather than something the layout decides behind you.

```{r}
data(gm_eras) + bar * sum + x(continent) + y(population) + color(era) +
  order(population, desc = TRUE) + nest() + title("Largest continent first")
```

That is the same `order()` that turns a bar chart round, doing the same job. Had
the packing sorted on its own, `order()` would have been a word with no effect
here, and the atom would have quietly meant less than it says.

It sorts a categorical *position* axis, so it needs one. The one-level treemap
above binds no position at all, and there the atom has nothing to sort:

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population) + color(continent) +
  order(population, desc = TRUE) + nest()
```

That refusal is not the packed space's. It is `order()`'s, and it was always
true: the same sentence flat, a single divided column, had been accepting the
word and dropping it in silence. Writing this chapter is what found it, which is
the ordinary way a new space earns its keep — it asks an old atom a question
nobody had asked.

## Where the rectangles come from

The packing is *squarified*, after Bruls, Huizing and van Wijk [@bruls2000]: each
run of regions is laid down in the strip that keeps its rectangles closest to square.
The naive alternative cuts one direction at a time, which turns any long tail
into slivers, and slivers are the shape an area is hardest to judge. There is no
setting for this. One layout that fixes a real defect is a feature; a second one
beside it, for completeness, would be the kind of vocabulary this grammar exists
to refuse.
