# Nest

```{r}
#| include: false
source("R/setup.R")
```

What share of the whole is each part? A share of a whole has to be carried
by something you can see. Flat, it is a **length**: a segment's height as a
fraction of the column it sits in. Bent into a circle, it is an **angle**, and
the column becomes a pie. There is a third answer, and it is the one that fits
the most categories into the least space: **area**.

That is what `nest()` draws. Each row's measure becomes the size of a region,
and the regions fill the panel with no gaps. Dividing a panel that way is a
**packing**. The plot has a name, **treemap**, and as with the pie there is no
atom that spells it.

That name comes from a real and ordinary problem. Ben Shneiderman's 80-megabyte hard
disk was full in 1990, and a directory listing could not tell him what was
filling it. So he wrote a display that packs a rectangle with nested rectangles,
each sized by what it holds [@shneiderman1992]. It is now the usual way to draw
a share of a whole when there are more categories than a pie has room for. That
display is the one this chapter teaches.

## One sentence, three spaces

How is the world's population divided among the continents? You already know
one plot that shows a share: the divided column. Start there, because the same
sentence will answer the question three times. Here is the population of each
continent as a single column, divided into its parts:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + bar * sum * stack + y(population) + color(continent) +
  title("A column divided")
```

*"Given gapminder 2007: bars derived by sum and stack, y is population, color by
continent."*

Now the same sentence in the circle. The column's height becomes the whole turn,
and each segment's share of the height becomes its slice's share of the circle:

```{r}
data(gapminder_2007) + bar * sum * stack + y(population) + color(continent) +
  polar() + title("The same shares, as angles")
```

*"Given gapminder 2007: bars derived by sum and stack, y is population, color by
continent, in polar."*

And the same sentence again, packed (one word is gone, and the end of this
section says why):

```{r}
data(gapminder_2007) + bar * sum + y(population) + color(continent) +
  nest() + title("The same shares, as areas")
```

*"Given gapminder 2007: bars derived by sum, y is population, color by
continent, in the nest."*

Three pictures, one grammar. Nothing in any of them names a chart: a bar,
summed, colored by continent, and then a word for the space. Asia is three
fifths of the world's population in all three, and you can check that against
the column, the pie and the treemap in turn.

That missing word is `stack`: the flat and polar forms carry it and the packed
form does not. `stack` decides what happens when two marks land in the same
place, so it assumes two marks can share a place. A packing gives every region a
place of its own, so two marks never collide. `stack` would have nothing to
resolve, and gog refuses it here.

## Why area, when a column already works

Read the column again and look for Oceania. At 0.4% of the world it is about
two pixels tall. The border drawn around it takes about a quarter of that
height. In
the packing the same share is a region about six pixels wide, because a column
has only its length to divide, and a packing divides both width and height.

Oceania's region is three times as wide as the bar was tall, and the share it
reports is unchanged. With only five parts, that is all a packing gains. The
gain grows with the number of parts, because a column divides in one direction
and a panel divides in two. A packing is for the case where there are *many* shares and
the question is which of them are large. A column has one fixed length to
divide, so it stops showing new categories long before a packing does.

Area has one weakness, the same one the pie has: two lengths can be compared
exactly, and two areas cannot. If the comparison has to be precise, draw the
flat column.

## Two levels, from a position

A share often has a share inside it: each continent's population, and within it
how much belongs to 1957 and how much to 2007. A packing can nest one split
inside another. Binding a position is what adds the second level. Each
category gets a region sized by its own total, and the rows inside it are
packed into that region:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * sum + x(continent) + y(population) + color(era) +
  nest() + title("Two eras, inside each continent")
```

*"Given the gapminder eras: bars derived by sum, x is continent, y is
population, color by era, in the nest."*

Each continent is a region sized by its total across both eras, and inside it
the two eras are packed against each other. The heavier lines are the
continents; the thinner ones divide what is inside them. Africa's region is the
most uneven: more than three quarters of it is the 2007 part.

The nesting is not new vocabulary. Wilkinson makes the point directly
[@wilkinson2005]: a **dimension hierarchy** (his name for levels inside levels)
*is* a nested facet. So `x` here does what the outer level of a facet does. The one
difference is that the panel is divided by share, not into equal boxes.

The inner level comes from `color`, and the legend is what decodes it. So the
second level needs few enough values to fit in the legend beside the plot.
Countries inside continents is the treemap printed most often, and its legend
would need 142 entries. [Naming the regions](#naming-the-regions) writes each
name inside its own region instead.

## What a packed panel does not have

Look for the axes. There are none, and no setting adds them.

A flat panel's horizontal direction *means* something: it is a variable, and a
tick is a place on it. In a packing neither direction is a variable. Wilkinson
states it exactly: the two dimensions of a treemap "have no intrinsic meaning
related to the data because they can be reordered without changing the metric
properties of the tree" [@wilkinson2005]. Two regions that touch are not near
each other in the data. Moving one to the other side of the panel would not
change a single share the plot reports.

So there is no tick to place and no axis name to print, and gog refuses
`x_label()`:

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population) + color(continent) + nest() +
  x_label("Population")
```

What decodes a packed panel is the color legend, exactly as it decodes a pie.

## Naming the regions

A legend works while it has few enough entries to read. With more entries it
becomes a second thing to read, and 142 countries are far too many. So put the
name where the region is. The sentence drops `sum` as well: without it each row
is its own region, so every country gets one inside its continent.

```{r}
#| message: true
data(gapminder_2007) + bar + x(continent) + y(population) + color(continent) +
  text + label(country) + style(color = "white") + nest() +
  title("Countries inside continents")
```

*"Given gapminder 2007: bars, x is continent, y is population, color by
continent, and also text, label by country, in the nest."*

Two layers, and the second is the ordinary `text` mark: `label` says which
column supplies the string, exactly as it does on a scatter. What it does *not*
say is where to put it, because in this space nothing says where a mark goes. A
label goes to the center of the region its own row was packed into. That is the same
region the bar drew. Both marks read one packing, so no name is drawn inside
another country's region.

Notice what is missing from that sentence: the text layer carries no `x` and no
`y` of its own. The plot's own `x(continent)` and `y(population)` do that work
for it. They size the regions. Once a region has a size and a place, the
name inside it has a place too. On a flat panel a `text` mark needs two
positions before it can be drawn at all. Here it needs a measure and a string,
and the packing places it.

Count the names and there are far fewer than 142. Most of the world's countries
get a region narrower than their own name, so a label that does not fit its
rectangle is not drawn. The engine says how many names it drew and why the rest
are missing, rather than letting you read the named countries as though they
were all of them.

A few African countries are named, and most of Africa is not. Almost every
unnamed country still has a region whose area is exactly its share of the
world, and one share is too small even for that. What is missing from this plot
is names, not shares. The count above the plot is the number you can check
against the names on the page.

That message is why the missing names do not mislead you: a treemap that printed
only the names that fit, and said nothing, would read as a complete list.

Three things change how many names are drawn, and you have used each of them
already: fewer categories, a larger plot with `theme(width =, height =)`, or
smaller text with `style(size = )`. None of them is a setting for *labels* in
particular, and none of them will make a name fit a region narrower than the
word.

## The other refusals

The refusals in this section come from one fact: a region has an area, and
nothing else. A `point` is drawn at the coordinates it is given, and a packing
has no coordinates to give:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + nest()
```

An area cannot be negative. A measure with values below zero cannot be packed
into regions. On the flat page, a bar below the baseline points down, and the
reader can see the sign. A region has no direction to point in, so nothing in
the picture could show it. The parts would also stop summing to the whole.

```{r}
#| echo: false
peek(cashflow)
```

```{r}
#| error: true
data(cashflow) + bar * sum + y(delta) + color(step) + nest()
```

A log scale is refused too. A region's size is its share of the total, and that
share is computed from the values as they are. A log scale re-spaces the values
so that equal distances stand for equal ratios. A packing draws no distances,
so there is nothing for the scale to change, and gog refuses a word that would
draw the same picture.

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population, scale = "log") +
  color(continent) + nest()
```

The packing method is not a choice you make. It is *squarified*, after Bruls,
Huizing and van Wijk [@bruls2000]: each group of regions is placed so its
rectangles are as close to square as they can be. The simpler method divides one
direction at a time, which makes long thin regions, and a long thin area is the
hardest kind to judge. gog offers no choice here, because a second layout would
add a word that draws nothing new.

## Sorting the regions

Squarified packing does its best work on a sorted list, which is why published
treemaps usually sort by size, largest first. gog does not sort them for you. The
regions follow the order the categorical axis already has, so the sort is
something you write.

```{r}
data(gm_eras) + bar * sum + x(continent) + y(population) + color(era) +
  order(population, desc = TRUE) + nest() + title("Largest continent first")
```

*"Given the gapminder eras: bars derived by sum, x is continent, y is
population, color by era, ordered by population, largest first, in the nest."*

That is the same `order()` that reorders a bar chart, and it does the same work
here.
If the packing sorted on its own, `order()` would change nothing in this space,
and its name would describe something it does not do.

`order()` sorts a categorical *position* axis, so a plot needs one before it
can be sorted. A one-level treemap binds no position at all, so there the atom
has nothing to sort:

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population) + color(continent) +
  order(population, desc = TRUE) + nest()
```

That refusal is not the packed space's. It is `order()`'s, and it applies in
every space.

## What comes next

Regions in a nest are sized by the measure, and the packing puts them wherever
they fit. [Map](map.qmd) places a mark at the point its own row names instead:
longitude and latitude, projected onto the page.
