# Nest

```{r}
#| include: false
source("R/setup.R")
```

How much of the whole does each part take? A share of a whole has to be carried
by something you can see. Flat, it is a **length**: a segment's height as a
fraction of the column it sits in. Bent into a circle, it is an **angle**, and
the column becomes a pie. There is a third answer, and it is the one that fits
the most categories into the least space: **area**.

That is what `nest()` draws. Each row's measure becomes the size of a region,
and the regions partition the panel. The plot has a name, treemap, and as with
the pie there is no atom that spells it.

That name has an exact and ordinary origin. Ben Shneiderman's 80-megabyte
hard disk was full in 1990, and a directory listing could not tell him what was
filling it, so he wrote a display that packs a rectangle with nested rectangles
sized by what they hold [@shneiderman1992]. It is now the usual way to draw a
share of a whole when there are more categories than a pie has room for. That is
where this chapter started.

## One sentence, three spaces

Here is the population of each continent as a single column, divided into its
parts:

```{r}
data(gapminder_2007) + bar * sum * stack + y(population) + color(continent) +
  title("A column divided")
```

*"Given gapminder 2007: bars derived by sum and stack, y is population, color by continent."*

Now the same sentence in the circle. The column's height becomes the whole turn,
and each segment's share of the height becomes its slice's share of the circle:

```{r}
data(gapminder_2007) + bar * sum * stack + y(population) + color(continent) +
  polar() + title("The same shares, as angles")
```

And the same sentence again, packed (one word is gone, and the paragraph below
says why):

```{r}
data(gapminder_2007) + bar * sum + y(population) + color(continent) +
  nest() + title("The same shares, as areas")
```

Three pictures, one grammar. Nothing in any of them names a chart: a bar, summed,
colored by continent, and then a word for the space. Asia is three fifths of
the world's population in all three, and you can check that against the column,
the pie and the treemap in turn.

One word did change, and it is worth explaining: the flat and polar forms carry
`stack` and the packed form does not. `stack` decides what happens when two marks
land in the same place, and that assumes they can. A packing gives every
region a place of its own, so there is no collision left to resolve, and gog
refuses `stack` here rather than accepting a word with nothing to do.

## Why area, when a column already worked

Read the column again and look for Oceania. At 0.4% of the world it is about
**two pixels** tall, so nearly half of it is the border drawn around it. In the
packing the same share is a region about **six pixels** wide, because a column
has only its length to divide, and a packing divides both width and height.

The packing gives Oceania three times more room, and the share it reports is
unchanged. With five parts a packing gains little; the gain grows with the
number of parts, because a column can only ever get longer while a panel gets
subdivided in both directions at once. A packing is for the case where there are
*many* shares and the question is which of them are large. A column would have
run out of pixels long before it ran out of categories.

Area has one weakness, the same one the pie has: two lengths can be compared
exactly, and two areas cannot. If the comparison has to be precise, draw the
flat column.

## Two levels, from a position

Bind a position and the packing gains a level. Each category gets a region sized
by its own total, and the rows inside it are packed into that region:

```{r}
data(gm_eras) + bar * sum + x(continent) + y(population) + color(era) +
  nest() + title("Two eras, inside each continent")
```

Each continent is a region sized by its total across both eras, and inside it the
two eras are packed against each other. The heavier lines are the continents; the
thinner ones divide what is inside them. Africa's region is the most lopsided:
more than three quarters of it is the 2007 half.

The nesting is not new vocabulary. Wilkinson makes the point directly
[@wilkinson2005]: a **dimension hierarchy** (his name for levels inside levels)
*is* a nested facet. So `x` here is doing what the outer level of a facet does,
with the panel divided by share instead of into equal boxes.

The inner level comes from `color`, and the legend is what decodes it. So the
second level needs few enough values to read down the side of the plot.
Countries inside continents, the treemap most often printed, would be a legend
of 142 entries, and the section after next names the regions in place instead.

## What a packed panel does not have

Look for the axes. There are none, and that is the space's defining property
rather than a setting you could switch back on.

A flat panel's horizontal direction *means* something: it is a variable, and a
tick is a place on it. In a packing neither direction is a variable. Wilkinson
puts it plainly: the two dimensions of a treemap "have no
intrinsic meaning related to the data because they can be reordered without
changing the metric properties of the tree" [@wilkinson2005]. Two regions that
touch are not near each other in the data. Moving one to the other side of the
panel would not change a single share the plot reports.

So there is nothing to tick and nothing to label, and gog refuses to pretend
otherwise:

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population) + color(continent) + nest() +
  x_label("Population")
```

What decodes a packed panel is the color legend, exactly as it decodes a pie.

## Naming the regions

A legend works while there are few enough entries to scan. With more entries it
becomes a second thing to read, and 142 countries are far too many. So put the
name where the region is:

```{r}
#| message: true
data(gapminder_2007) + bar + x(continent) + y(population) + color(continent) +
  text + label(country) + style(color = "white") + nest() +
  title("Countries inside continents")
```

Two layers, and the second is the ordinary `text` mark: `label` says which
column supplies the string, exactly as it does on a scatter. What it does *not*
say is where to put it, because in this space nothing does. A label goes to the
center of the region its own row was packed into. That is the region the bar
drew, so both marks read one packing and no name is drawn in another country's
region.

Notice what is missing from that sentence: there is no `x` on the text layer and
no `y` either, and the plot's own `y(population)` is doing both jobs at once. It
sizes the regions, and sizing them is what places the labels. Everywhere else in
this book a `text` mark needs two positions before it can be drawn at all. Here
it needs a measure and a string, and the packing places it.

Count the names and there are far fewer than 142, which is what the message above
the plot is for. Most of the world's countries have a share narrower than
their own name, so a label that does not fit its rectangle is not drawn. The
engine says how many it left out, rather than letting you read the named
countries as though they were all of them. Egypt and Sudan are named and most of
Africa is not. Every one of those unnamed countries still has a region exactly as
wide as its share of the world. The picture is complete and the labeling is not,
and the message above the plot says how many names are missing.

That message is why the missing names are safe: a treemap that printed only the
names that fit, and said nothing, would read as a complete list.

Three things move the number, and each is a change you have already met: fewer
categories, a larger plot with `theme(width =, height =)`, or smaller type
with `style(size = )`. None of them is a setting for *labels* in particular, and
none of them will make a name fit a region narrower than the word.

## The other refusals

A packing cannot hold a mark that is placed by a position, because it has no
positions to place one at:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + nest()
```

An area cannot be negative, so a measure that runs below zero has no packing.
On the flat page, a bar below the baseline points down, and the reader can see
the sign. A region has no direction to point in, and its parts would no longer
sum to the whole.

```{r}
#| error: true
data(cashflow) + bar * sum + y(delta) + color(step) + nest()
```

A log scale is refused too, and the reason is the one worth remembering: a
region's size is its **share of the total**, which is arithmetic on the values
themselves. Re-spacing them changes what a distance means, and a packing has no
distances. Accepting the scale would draw exactly the same picture, and gog
refuses a word that changes nothing.

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population, scale = "log") +
  color(continent) + nest()
```

One more thing is not offered. The packing is *squarified*, after Bruls, Huizing
and van Wijk [@bruls2000]: each group of regions is placed so its rectangles come
out as close to square as they can. The simpler method divides one direction at a
time, which makes long thin regions, and a long thin area is the hardest kind to
judge. gog offers no choice here, because a second layout would add a word that
draws nothing new.

## The order is yours

Published treemaps usually sort by size, largest first, because the packing
produces squarer rectangles that way. gog does not do it for you: the
regions follow the order the categorical axis already has, so the sort is
something you write, not something the layout decides for you.

```{r}
data(gm_eras) + bar * sum + x(continent) + y(population) + color(era) +
  order(population, desc = TRUE) + nest() + title("Largest continent first")
```

That is the same `order()` that reorders a bar chart, doing the same job here. Had
the packing sorted on its own, `order()` would have been a word with no effect
here, and the atom would have quietly meant less than it says.

It sorts a categorical *position* axis, so it needs one. The one-level treemap
above binds no position at all, and there the atom has nothing to sort:

```{r}
#| error: true
data(gapminder_2007) + bar * sum + y(population) + color(continent) +
  order(population, desc = TRUE) + nest()
```

That refusal is not the packed space's. It is `order()`'s, and it holds in every
space.
