# Network

```{r}
#| include: false
source("R/setup.R")
```

What is connected to what? Some tables record relations rather than
measurements: each row says that one thing is related to another. A network
diagram draws those relations. Each thing becomes a **node**, one marked place
in the picture, and each relation becomes a stroke between two nodes. Nodes with
many relations are drawn near the middle.

Nothing in such a table says where any node should sit. That is the whole
difficulty, and this chapter shows how gog answers it. The engine computes a
position for every node. A space named `network()` draws the result, with no
axes at all.

`trade_partners` holds one row per trade relation: who exports, who imports,
and the value of the goods, in billions of dollars, rounded:

```{r}
#| echo: false
peek(trade_partners)
```

## One row is one edge

The first question of the trade table is who trades with whom, and one stroke
per row draws exactly that. Write the `edge` mark derived by `layout()`, name
the two endpoint columns in order, and say the space:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in
the network."*

Each stroke is one row of the table. The `layout` reads both columns and
finds every distinct name. All nodes push apart, and nodes that share a row
also pull together, so the connected ones settle near each other. The engine
computes every position itself, and computes it the same way everywhere, so
this sentence draws the same picture in R, Python, Julia and JavaScript, to
the byte.

The panel has no axes, no ticks and no grid, because `network()` draws none. A
layout's positions mean nothing as quantities. Flip the picture left to right
and it means the same, so a number on an axis would state nothing about the
data. A `layout` therefore draws in `network()` and in no other space.

## The nodes and their names

The strokes alone do not say which node is which country, or which countries
trade the most. Two more layers finish the diagram. A `point` layer draws the
nodes, and a `text` layer names them. Each one writes the same `layout`, and
the engine computes it once per layer. Every computation lands on the same
positions, because the placement is deterministic. The layout publishes two
columns about each node: its `name`, and its `degree`, the count of its
relations. `size(degree)` draws the nodes with the most relations largest. The
`text` layer also composes `repel`, the modifier that moves labels off each
other, so a name rests beside its dot instead of covering it:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) * repel + label(name) +
  network() +
  title("Who trades with whom")
```

*"Given trade partners: edges derived by layout from exporter to importer, and
also points derived by layout from exporter to importer, size by degree, and
also text derived by layout from exporter to importer and repel, label by name,
in the network."*

The nodes with the most relations sit near the middle. No row of the table says
that the United States, China and Germany are centers of trade; saying it is
what the layout is for.

## An edge carries its own columns

Every stroke so far is drawn alike, whether the trade is 500 billion dollars or
50 billion. An edge is one row, so its color and its opacity can follow any
column of the table. `opacity(billions)` maps each stroke's opacity to its
value, so the heavier relations draw darker:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) * repel + label(name) +
  network() +
  title("Heavier trade, darker stroke")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, size by degree, and also text derived by layout from exporter to
importer and repel, label by name, in the network."*

A node is different. A node stands for many rows, and those rows carry different
values. So a node layer maps only `name` and `degree`. Every other column
belongs on the `edge` layer, where one row carries one value.

## An angle asks for the cube

A plane gives the layout two directions to spread its nodes, and a cube gives
it three, in a picture you can turn. Bare `network()` computes the layout in
the plane. State a viewing angle and the same edges are laid out in a cube
instead. Every node takes a new position, and
the picture is drawn from the angle you asked for:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  network(turn = 35, tilt = 20) +
  title("The same network, given depth")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, size by degree, in the network, turned 35, tilted 20."*

Nothing in the table gives a node a third position, so there is nothing to bind
to `z`. The third dimension belongs to the space, exactly as the viewing angle
does, so asking for the angle is asking for the cube. The box is drawn for depth
and carries no numbers, for the same reason the flat form has no axes.

In the web edition the cube turns with a drag, the way the cube in
[Space](space.qmd) turns, and *reset* returns to the angle the sentence asked
for. On paper it keeps the view it was given.

Naming a node in the cube is valid grammar. This engine does not draw those
names:

```{r}
#| error: true
data(trade_partners) +
  edge * layout(exporter, importer) +
  text * layout(exporter, importer) * repel + label(name) +
  network(turn = 35, tilt = 20)
```

## What the network refuses

The likeliest mistake is to leave `network()` off, since a flat plot needs no
word for its space. A `layout` outside its space would draw axes, and the
numbers along them would mean nothing:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

Only three marks read a layout. Write any other mark and the refusal names all
three:

```{r}
#| error: true
data(trade_partners) + bar * layout(exporter, importer) + network()
```

The layout has already placed every node, so `x()` has nothing left to set:

```{r}
#| error: true
data(trade_partners) + x(billions) +
  edge * layout(exporter, importer) + network()
```

And `network()` with no layout has nothing to place, so it refuses:

```{r}
#| error: true
data(trade_partners) + point + network()
```

## Where the limits are

A row from a node to itself is refused. Such a stroke would have no length,
and a fact about one node belongs on the node. A row missing an endpoint is
left out, and the engine says how many it left out.

gog does not read a second table of node facts. A node shows only what the edges
give it: its name and its degree. Joining more columns onto the nodes is a table
operation, so do it in the host language, before the plot. A `brush` is refused
too, and its answer is also upstream: a brush names a range on the data's axes,
and a network has none. Select the rows in your table first, then color the
edges they make.