# Network

```{r}
#| include: false
source("R/setup.R")
```

What is connected to what? Some tables record relations rather than
measurements: each row says that one thing is related to another. A network
diagram draws those relations. Each thing becomes a **node**, one marked place
in the picture, and each relation becomes an **edge**, a stroke between two
nodes.

Nothing in such a table says where any node should sit. Choosing those positions
is the problem, and this chapter shows how gog solves it. The engine computes a
position for every node. A space named `network()` draws the result, with no
axes at all.

`trade_partners` holds one row per trade relation: who exports, who imports,
and the value of the goods, in billions of dollars, rounded:

```{r}
#| echo: false
peek(trade_partners)
```

## One row is one edge

The first question this table raises is who trades with whom. One stroke per row
answers exactly that. Write the `edge` mark derived by `layout()`, give it the
*from* column first and the *to* column second, and add the space:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in
the network."*

Each row of the table becomes one stroke. Five pairs of countries trade in both
directions, and each pair's two strokes lie on one line. So you can see only 17
separate strokes for the 22 rows. The `layout` reads both columns and makes one
node for each distinct name. The engine then pushes every node away from every
other node, and pulls the two nodes of each row toward each other. Two countries
that trade therefore sit near each other. The positions are computed by the
engine that all four languages share, never by the language you write in. So
this sentence draws the same picture in R, Python, Julia and JavaScript, and the
four image files are identical.

The panel has no axes, no ticks and no grid, because `network()` draws none. A
layout's positions mean nothing as quantities. Flip the picture left to right
and it means the same, so a number on an axis would state nothing about the
data. gog therefore draws a `layout` only inside `network()`, and refuses it in
every other space.

## The nodes and their names

The strokes alone do not say which node is which country, or which countries
have the most trade relations. Two more layers finish the diagram. A `point`
layer draws the nodes, and a `text` layer names them. Each one carries the same
`layout`, and the engine computes it once per layer. The layout uses no random
numbers, so all three layers get the same positions. That is why every dot sits
exactly where its strokes end.

The layout also computes two columns for the nodes: `name`, and `degree`, the
number of rows the node appears in. A pair that trades both ways counts twice,
so China, which trades with 7 countries, has a degree of 9. `size(degree)` then
draws a node larger when it appears in more rows. The `text` layer also
composes [`repel`](marks/text.qmd#repel-moving-labels-off-each-other), the
collision modifier that moves each label off the dots and off the other labels.
So a name sits just outside its dot instead of covering it:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) * repel + label(name) +
  network() +
  title("Who trades with whom")
```

*"Given trade partners: edges derived by layout from exporter to importer, and
also points derived by layout from exporter to importer, sized by degree, and
also text derived by layout from exporter to importer and repel, label by name,
in the network."*

China, the United States and Germany have the largest dots, and they sit near
the middle. No row of the table says that these three have the most trade
relations. The layout found them by counting each country's rows.

## An edge carries its own columns

Every stroke so far is drawn alike, whether the trade is 500 billion dollars or
50 billion. Each edge is drawn from one row, so its color, its dash and its
opacity can come from that row's own columns. `color` and `pattern` need a
categorical column, and `opacity` needs a continuous one. `opacity(billions)`
maps each stroke's opacity to that row's `billions`, so a larger trade draws a
darker stroke. A pair that trades both ways is two strokes on one line, so it
looks darker than either row alone:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) * repel + label(name) +
  network() +
  title("More trade, darker stroke")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, sized by degree, and also text derived by layout from exporter to
importer and repel, label by name, in the network."*

A node is different. It stands for every row it appears in, and those rows can
hold different values. So the `point` and `text` layers map only `name` and
`degree`, and gog refuses any other column there. Every other column belongs on
the `edge` layer, where one row carries one value.

## An angle asks for the cube

Every picture so far is flat. A plane gives the layout two directions for
placing its nodes. A cube gives it three, and a cube can be turned. A crowded
network has more room in three directions, and turning the cube shows which
strokes pass in front of the others. Bare `network()` computes the layout in the
plane. State a viewing angle and the same edges are laid out in a cube instead.
Every node takes a new position, and the picture is drawn from the angle you
asked for. This engine does not draw names in the cube, so the sentence below
has no `text` layer:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  network(turn = 35, tilt = 20) +
  title("The same network, given depth")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, sized by degree, in the network, turned 35, tilted 20."*

Nothing in the table gives a node a third position, so there is nothing to bind
to `z`. The third dimension belongs to the space, exactly as the viewing angle
does. The box outlines the cube, so you can see how it is turned. It has no
numbers along its edges, because a layout's positions are not measurements. The
flat form has no axes for the same reason.

In the web edition the cube turns with a drag, the way the cube in
[Space](space.qmd) turns. The *reset* button returns it to the angle the
sentence asked for. In the PDF it keeps the view it was given.

A `text` layer in the cube is valid grammar that this engine does not draw. The
refusal says so, and it suggests the flat form for the labels:

```{r}
#| error: true
data(trade_partners) +
  edge * layout(exporter, importer) +
  text * layout(exporter, importer) * repel + label(name) +
  network(turn = 35, tilt = 20)
```

## What the network refuses

That refusal is about something this engine does not draw. The four refusals in
this section are different: each one is a mistake in the sentence. The most
likely mistake is to leave `network()` off, because most plots need no space at
all. Without it the plot would draw axes under the layout, and the numbers along
them would mean nothing:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

Only three marks read a layout. Write any other mark and the refusal names all
three:

```{r}
#| error: true
data(trade_partners) + bar * layout(exporter, importer) + network()
```

The layout has already placed every node, so `x()` has nothing left to set:

```{r}
#| error: true
data(trade_partners) + x(billions) +
  edge * layout(exporter, importer) + network()
```

And `network()` with no layout has nothing to place, so it refuses:

```{r}
#| error: true
data(trade_partners) + point + network()
```

## Where the limits are

A sentence can be right, and the table can still hold rows a layout cannot use.
A row from a node to itself is refused, because such a stroke would have no
length. Remove such rows from your table before you draw the plot. A row with a
missing value in the *from* or the *to* column draws no stroke, and a message
says how many rows were left out.

You may know more about each country than the edges say, such as its
continent. gog does not read a second table describing the nodes, so a node
shows only what the edges give it: its name and its degree. You can still show
such a fact, but only on the edges. Before you draw the plot, join it onto each
row in the host language, as a column such as the exporter's continent. Then map
that column on the `edge` layer.

You may want to drag a rectangle over part of the picture and select what it
covers. A `brush` is refused. A brush selects a range of values on the plot's
axes, and a network has no axes. The fix is again in your table: add a column
that marks the rows you want, and map it with `color()` on the `edge` layer.

If your table also has a year column, you may want to show how trade changes.
For most plots, `play` does that: it draws one frame per year and shows the
frames in order. In a network, `play` is valid grammar that this engine does not
draw, like names in the cube.

Instead, split the plot by year with `| facet(year)`, which draws one panel per
year. All panels share one layout, so a country sits in the same place in every
panel where it appears. Each panel draws only the edges from its own rows. An
edge missing from a later panel is a trade relation that ended.

## What comes next

A network draws one table at one moment. The section above suggested panels for
comparing years, because this engine does not play a network. Most other plots
can be played. [Play](play.qmd) adds time to a plot: it cuts the plot into
frames and runs them in order, so the reader watches the picture change.
