# Network

```{r}
#| include: false
source("R/setup.R")
```

What is connected to what? Some tables record relations rather than
measurements: each row says that one thing is related to another. A network
diagram draws such a table. Each thing becomes a **node**, one marked place in
the picture, and each relation becomes a stroke between two nodes. Nodes with
many relations end up near the middle.

Nothing in such a table says where any node should sit. That is the whole
difficulty, and this chapter shows how gog answers it. The engine computes a
position for every node, the same way in every language. A space named
`network()` draws the result, with no axes at all.

`trade_partners` holds one row per trade relation: who exports, who imports,
and the value of the goods, in billions of dollars, rounded:

```{r}
#| echo: false
peek(trade_partners)
```

## One row is one edge

Name the two endpoint columns in `layout()`, give the result to the `edge`
mark, and say the space:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in the network."*

Each stroke is one row of the table. The `layout` reads both columns and
finds every distinct name. All nodes push apart, and nodes that share a row
also pull together, so the connected ones settle near each other. The engine
computes every position itself, and computes it the same way everywhere, so
this sentence draws the same picture in R, Python, Julia and JavaScript, to
the byte.

The panel has no axes, no ticks and no grid, because `network()` draws none.
A layout's positions mean nothing as quantities. Flip the picture left to
right and it means the same, so no axis number would be true. `network()` is
the one space that makes that statement, and a `layout` is read nowhere
else.

## The nodes and their names

Two more layers finish the diagram, and both read the same layout: written on
each layer, computed once each, and always landing on the same positions,
because the computation is exact. A `point` layer draws the nodes, and a
`text` layer names them. The layout publishes two columns about each node:
its `name`, and its `degree`, the count of its relations. `size(degree)`
draws the nodes with the most relations largest. The `text` layer also
composes `repel`, the same modifier every labeled scatter uses, so a name
rests beside its dot instead of striking through it:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) * repel + label(name) +
  network() +
  title("Who trades with whom")
```

The nodes with the most relations end up near the middle. No row of the
table says that the United States, China and Germany are centers of trade;
saying it is what the layout is for.

## An edge carries its own columns

An edge is one row, so its color and its opacity can follow any column of
the table. `opacity(billions)` fades each relation by its value, so the
heavier relations stand out:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) * repel + label(name) +
  network() +
  title("Heavier trade, darker stroke")
```

A node is different. A node stands for many rows, and those rows carry
different values. So a node layer maps only `name` and `degree`, and any
other column is refused toward the `edge` layer, where it belongs.

## An angle asks for the cube

Bare `network()` computes the layout in the plane. State a viewing angle and
the same layout is computed in a cube instead, and drawn from that angle:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  network(turn = 35, tilt = 20) +
  title("The same network, given depth")
```

There is no third column behind this and nothing to bind to `z`. The third
dimension belongs to the space, exactly as the viewing angle does, so asking
for the angle is asking for the cube. The box is drawn for depth and carries
no numbers, for the same reason the flat form has no axes.

In the web edition the cube turns with a drag, the way the cube in
[Space](space.qmd) turns, and *reset* returns to the angle the sentence asked
for. On paper it keeps the view it was given.

## What the network refuses

A `layout` outside its space would put meaningless numbers under real axes:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

Only three marks read a layout. Write any other one and the refusal lists
them:

```{r}
#| error: true
data(trade_partners) + bar * layout(exporter, importer) + network()
```

The layout has already placed every node, so `x()` has nothing left to set:

```{r}
#| error: true
data(trade_partners) + x(billions) +
  edge * layout(exporter, importer) + network()
```

And `network()` with no layout has nothing to place, so it refuses:

```{r}
#| error: true
data(trade_partners) + point + network()
```

## Where the limits are

A row from a node to itself is refused. Such a stroke would have no length,
and a fact about one node belongs on the node. A row missing an endpoint is
left out, and the engine says how many it left out.

gog does not read a second table of node facts. A node shows only what the
edges give it: its name and its degree. Joining more columns onto the nodes
is a table operation, so do it in the host language, before the plot. A
`brush` is refused for the same underlying reason: a brush names a range on
the data's axes, and a network has none. Select the rows in your table
first, then color the edges they make.
