# Network

```{r}
#| include: false
source("R/setup.R")
```

What is connected to what? Some tables record relations rather than
measurements: each row says that one thing is linked to another. A network
diagram draws those relations. The things become points, the relations become
strokes between them, and things with many relations sit near the middle of
the picture.

Nothing in such a table says where anything should sit. That is the whole
difficulty, and this chapter is how gog answers it. The engine computes a
position for every node, the same way in every language, and a space of its
own draws the result with no axes at all.

`trade_partners` holds one row per trade relation: who exports, who imports,
and the yearly value, rounded to billions:

```{r}
#| echo: false
peek(trade_partners)
```

## One row is one edge

Name the two endpoint columns in `layout()`, give the result to the `edge`
mark, and say the space:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in the network."*

Each stroke is one row of the table. The `layout` reads both columns, finds
every distinct name, and computes a position for each: names that share many
rows pull together, and names that share none push apart. The engine computes
these positions itself, so the same sentence draws the same picture in R,
Python, Julia and JavaScript, to the byte.

The panel has no axes, no ticks and no grid, and that is the space speaking.
A layout's positions mean nothing as quantities. The picture could be
mirrored without changing anything it says, so there is no number an axis
could honestly put under it. `network()` is the one space that makes that
statement, and a `layout` draws nowhere else.

## The nodes and their names

Two more layers finish the diagram, reading the same layout. A `point` layer
draws the nodes, and a `text` layer names them. The layout publishes two
columns about each node: its `name`, and its `degree`, the count of its
relations. `size(degree)` makes the busy nodes large:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) + label(name) +
  network() +
  title("Who trades with whom")
```

The hubs find the middle by themselves. No row of the table says that the
United States, China and Germany are centers of trade; saying it is what the
layout is for.

## An edge carries its own columns

An edge is one row, so the edge layer can map any column the table holds.
`opacity(billions)` fades each relation by its value, which keeps a dense
diagram legible and earns a key:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) + label(name) +
  network() +
  title("The heavier the trade, the darker the stroke")
```

A node is different. It stands for every row that names it, and those rows
disagree about everything except the name, so a node layer maps only what a
node has: its `name` and its `degree`. A column of the edge table belongs on
the `edge` layer.

## Stating an angle states the cube

Bare `network()` computes the layout in the plane. State a viewing angle and
the same layout is computed in a cube instead, and drawn from that angle:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  network(turn = 35, tilt = 20) +
  title("The same network, given depth")
```

There is no third column behind this and nothing to bind to `z`. The third
dimension belongs to the space, exactly as the viewing angle does, so asking
for the angle is asking for the cube. The glass box is drawn for depth and
carries no numbers, for the same reason the flat form has no axes.

In the web edition the cube turns with a drag, the way the cube in
[Space](space.qmd) turns, and *reset* returns to the angle the sentence asked
for. On paper it keeps the view it was given.

## What the network refuses

A `layout` outside its space would put meaningless numbers under real axes:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

Three marks read a layout, and the refusal names them for any other mark:

```{r}
#| error: true
data(trade_partners) + bar * layout(exporter, importer) + network()
```

The nodes place themselves, so a bound position has nothing left to say:

```{r}
#| error: true
data(trade_partners) + x(billions) +
  edge * layout(exporter, importer) + network()
```

And the space without a layout is a room with nothing in it:

```{r}
#| error: true
data(trade_partners) + point + network()
```

## Where the limits are

A row from a node to itself is refused: a stroke from a point to the same
point has no extent, and a fact about one node belongs on the node. A row
missing an endpoint is left out, and the plot says how many.

A table of node facts is not read. What a node shows here derives from the
edges, its name and its degree, and joining a second table of attributes onto
the nodes is a table operation that belongs in your data language, before the
plot. The same holds for a `brush`: a brush names a range of the data's own
axes, and a network's positions are not data, so selection here means
selecting rows upstream and coloring the edges they make.
