# Network

```{r}
#| include: false
source("R/setup.R")
```

What is connected to what? Some tables record relations rather than
measurements: each row says that one thing is related to another. A network
diagram draws those relations. Each thing becomes a **node**, one marked place
in the picture, and each relation becomes an **edge**, a stroke between two
nodes.

Nothing in such a table says where any node should sit. Choosing those positions
is the problem, and this chapter shows how gog solves it. The engine computes a
position for every node. A space named `network()` draws the result, with no
axes at all.

`trade_partners` holds one row per trade relation: who exports, who imports,
and the value of the goods, in billions of dollars, rounded:

```{r}
#| echo: false
peek(trade_partners)
```

## One row is one edge

The first question this table raises is who trades with whom. One stroke per row
answers exactly that. Write the `edge` mark derived by `layout()`, name the two
endpoint columns in order, and add the space:

```{r}
data(trade_partners) + edge * layout(exporter, importer) + network()
```

*"Given trade partners: edges derived by layout from exporter to importer, in
the network."*

Each stroke is one row of the table. The `layout` reads both columns and
finds every distinct name. The engine then pushes every node away from every
other one, and pulls back together each pair that shares a row, so two
countries that trade sit near each other. The engine computes every position
itself, and computes it the same way everywhere, so this sentence draws the
same picture in R, Python, Julia and JavaScript, to the byte.

The panel has no axes, no ticks and no grid, because `network()` draws none. A
layout's positions mean nothing as quantities. Flip the picture left to right
and it means the same, so a number on an axis would state nothing about the
data. A `layout` is therefore read inside `network()` and in no other space.

## The nodes and their names

The strokes alone do not say which node is which country, or which countries
trade the most. Two more layers finish the diagram. A `point` layer draws the
nodes, and a `text` layer names them. Each one carries the same `layout`, and
the engine computes it once per layer. Every computation gives the same
positions, because the placement is deterministic.

The layout computes two columns about each node: its `name`, and its `degree`,
the count of its relations. `size(degree)` then draws a node larger when it has
more relations. The `text` layer also composes `repel`, the modifier that moves
labels off each other, so a name rests beside its dot instead of covering it:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) * repel + label(name) +
  network() +
  title("Who trades with whom")
```

*"Given trade partners: edges derived by layout from exporter to importer, and
also points derived by layout from exporter to importer, size by degree, and
also text derived by layout from exporter to importer and repel, label by name,
in the network."*

The nodes with the most relations sit near the middle. No row of the table says
that China, the United States and Germany have the most trade relations. The
layout is what shows it.

## An edge carries its own columns

Every stroke so far is drawn alike, whether the trade is 500 billion dollars or
50 billion. An edge is one row, so its color and its opacity can come from that
row's own columns. `color` needs a categorical column, and `opacity` needs a
continuous one. `opacity(billions)` maps each stroke's opacity to that row's
`billions`, so the heavier trades draw darker:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  text * layout(exporter, importer) * repel + label(name) +
  network() +
  title("Heavier trade, darker stroke")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, size by degree, and also text derived by layout from exporter to
importer and repel, label by name, in the network."*

A node is different. A node stands for many rows, and those rows carry different
values. So a node layer maps only `name` and `degree`. Every other column
belongs on the `edge` layer, where one row carries one value.

## An angle asks for the cube

Every picture so far is flat. A plane gives the layout two directions to place
its nodes in. A cube gives it three, and a cube can be turned. Bare `network()`
computes the layout in the plane. State a viewing angle and the same edges are
laid out in a cube instead. Every node takes a new position, and the picture is
drawn from the angle you asked for:

```{r}
data(trade_partners) +
  edge * layout(exporter, importer) + opacity(billions) +
  point * layout(exporter, importer) + size(degree) +
  network(turn = 35, tilt = 20) +
  title("The same network, given depth")
```

*"Given trade partners: edges derived by layout from exporter to importer,
opacity by billions, and also points derived by layout from exporter to
importer, size by degree, in the network, turned 35, tilted 20."*

Nothing in the table gives a node a third position, so there is nothing to bind
to `z`. The third dimension belongs to the space, exactly as the viewing angle
does. State the angle, and you have stated the cube. The box shows where the
depth is. It carries no numbers, because a layout's positions are not
measurements, which is the same reason the flat form has no axes.

In the web edition the cube turns with a drag, the way the cube in
[Space](space.qmd) turns, and *reset* returns to the angle the sentence asked
for. On paper it keeps the view it was given.

Naming a node in the cube is valid grammar that this engine does not draw. The
refusal says so, and it names the flat form as the place for the labels:

```{r}
#| error: true
data(trade_partners) +
  edge * layout(exporter, importer) +
  text * layout(exporter, importer) * repel + label(name) +
  network(turn = 35, tilt = 20)
```

## What the network refuses

That refusal is a gap in this engine. The rest are mistakes in the sentence, and
the likeliest is to leave `network()` off, because a flat plot names no space at
all. Without it the plot would draw axes under the layout, and the numbers along
them would mean nothing:

```{r}
#| error: true
data(trade_partners) + edge * layout(exporter, importer)
```

Only three marks read a layout. Write any other mark and the refusal names all
three:

```{r}
#| error: true
data(trade_partners) + bar * layout(exporter, importer) + network()
```

The layout has already placed every node, so `x()` has nothing left to set:

```{r}
#| error: true
data(trade_partners) + x(billions) +
  edge * layout(exporter, importer) + network()
```

And `network()` with no layout has nothing to place, so it refuses:

```{r}
#| error: true
data(trade_partners) + point + network()
```

## Where the limits are

A sentence can be right and the table still hold rows a layout cannot use. A row
from a node to itself is refused. Such a stroke would have no length, and a fact
about one node belongs on the node. A row missing an endpoint is left out, and
the engine says how many it left out.

gog does not read a second table of node facts. A node shows only what the edges
give it: its name and its degree. Joining more columns onto the nodes is a table
operation, so do it in the host language, before the plot.

A `brush` is refused, and its answer is in the table too. A brush names a range
on the data's axes, and a network has none. Select the rows in your table first,
then color the edges they make.

A network cannot be played either. Every frame would lay the graph out again, so
the movement would come from the layout rather than from a change in the data.
Draw the frames as separate plots beside each other with `|`.

## What comes next

A network draws one table at one moment, and nothing in the picture moves.
[Play](play.qmd) adds time to a plot: the same sentence cut into frames and run
in order, so the reader watches the picture change.
