# Operators

```{r}
#| include: false
source("R/setup.R")
```

How do you put two things in one picture? GOG has exactly **four operators**.
They are the assembly rules, not atoms themselves, but the rules that say how
atoms combine.

Think of them the way Hangeul (한글) works:
consonants and vowels are the atoms;
the rule "initial + vowel + optional final = syllable block" is the operator.
The rule is fixed; the atoms are infinite.

```
+   compose   (add a layer, a channel, or an annotation to a spec)
*   derive    (combine a mark with a transform into a compound mark)
|   facet ↔   (small multiples, arranged in columns)
/   facet ↕   (small multiples, arranged in rows)
```

---

## `+`: Compose

`+` is the general assembly operator.
Every atom enters a specification via `+`:

```r
data(df) + point + x(col_a) + y(col_b) + color(group) + title("My plot")
```

Adding a **mark** (`point`, `line`, `bar`, …) opens a new layer.
Adding a **channel** (`x()`, `color()`, …) or **annotation** (`title()`) attaches
to the appropriate scope (plot-level or the nearest preceding mark, see
[Encoding scope](encoding-scope.qmd)).

### Layering with `+`

If you have edited video, you already know this idea. A video is a stack of
tracks: the picture, the sound, the captions, the logo, the effects. Each one
is separate, they play at the same time, and you watch one thing. A plot layers
the same way. It is not a new idea to learn.

Multiple marks in the same expression produce multiple layers, rendered in order:

```{r}
# Scatter plot: points, and a horizontal reference line at each band
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  data(life_bands) + rule + style(color = "darkgrey", pattern = "dashed") +
  title("GDP vs Life Expectancy — 2007")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and also rules from the life bands."*

Layers draw in the order they are written: the first mark sits at the
bottom. Add a `line` after `point` to draw the line on top of the points.

### The order of `+`: *data, mark, positions, refinements*

`*` is rigid: the mark is always on the left, and no transform may stand there
(`bin * bar` is a type error). `+` is
the opposite. Inside a single layer the channels **commute**: write them in
any order and the plot is the same. The two lines below draw the identical
histogram; the only difference is that `color` moved ahead of `x`:

```{r}
#| layout-ncol: 2
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
data(iris_flowers) + bar * bin + color(species) + x(petal_length)
```

So if the order does not change the picture, why does every plot in this book
follow the same one: *data, mark, positions, refinements*? Because a fixed
order lets a plot **read like a sentence**, and the sentence has a natural,
subject-first shape:

- **`data(...)` first**, the table the columns come from. A bare name like
  `petal_length` means nothing until a table is bound (Bind-Once), so the
  binding leads. In R the sentence *must* start here.
- **the mark and its `*`-chain second**, the subject. A visual is a mark
  before it is anything else (Minimum Syllable): `bar * mean * dodge` says
  *what you are drawing* and *what was done to it*, as one unbroken word. It is
  the initial consonant of the block; everything after is said about it.
- **the positions third**, the frame. `x` and `y` say *where* the subject
  stands, and a mark cannot render without them (the vowel to the mark's
  consonant), so they follow immediately.
- **the refinements last**, `color`, `size`, `order`, labels, `title`:
  everything optional. Delete every refinement and a legal plot remains; delete
  the mark or a position and nothing renders. Writing the removable parts last
  is also how you *build* a plot: the smallest legal sentence first, then
  refine ([First plot](first-plot.qmd)).

The order is a reading convention, not a grammar rule: the engine accepts any
permutation that scopes the same way. Where position genuinely *does* matter
(a channel written before or after a mark across two or more layers) is the
subject of [Encoding scope](encoding-scope.qmd). The payoff of
keeping one order everywhere is that once your eye learns it, every plot in the
book, and every plot you write, parses the same way.

#### Which parts of the order are yours to choose

The order above is what this book prefers. It is not all the engine enforces,
and the difference is worth knowing, because it tells you where to be careful.
There are three kinds of ordering, and only one of them needs your attention.

| Kind | What it means | Where |
|---|---|---|
| **Free** | Either way draws the same picture, to the byte. | The mark and its positions, in any arrangement. `x` against `y`. Refinements among themselves. |
| **Refused** | The engine stops and names what to move. | A sentence that does not begin with `data()`. A `style()` written before any mark. |
| **Meaningful** | No error, and a different picture. | A channel before the marks or after one of them, once a plot has two. |

**Free** is most of it. Every one of these draws the identical plot, and the
book writes the first only because a fixed order is easier to read:

```{r}
#| layout-ncol: 2
data(gapminder_2007) + point + x(gdp) + y(life)
data(gapminder_2007) + x(gdp) + point + y(life)
```

**Refused** is the engine telling you the sentence has no subject yet. A
setting needs a mark to settle on, and in R a plot needs its table first,
because `+` has to have a plot on its left before it can add anything:

```{r}
#| error: true
data(gapminder_2007) + style(size = 3) + point + x(gdp) + y(life)
```

**Meaningful** is the only one to watch, because it is the only one that stays
silent. With two marks in a sentence, a channel written before them belongs to
the plot and reaches both, and a channel written after one belongs to that
mark. Both are legal, neither warns, and they draw different pictures. That is
the subject of [Encoding scope](encoding-scope.qmd), and it is the reason this
book keeps one order: when every sentence is arranged the same way, the one
that is arranged differently is saying so on purpose.

#### When a refinement sits far from what it belongs to

The convention has one spot that catches people out. A grouped bar chart pairs
`dodge` with `color`, yet they land at opposite ends of the sentence, with the
positions in between:

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  title("Mean life expectancy by continent — 1957 vs 2007")
```

`dodge` and `color(era)` are plainly related: `dodge` has nothing to set side
by side until a `color` (or `group`) names the groups. So why not write them
together? Because they live on opposite sides of the grammar *by design*.
`dodge` rides `*`: it modifies the mark, and it is the same modifier that works
on `box * dodge` and `interval * range * dodge`, neither of which involves
color. `color` rides `+`: it binds a column to a channel. `color * dodge` is
not a thing: `*` wants a mark on its left. Welding the two together would cost
`dodge` its independence from *which* channel does the splitting, and that
independence is the whole point of a modifier.

And the order is not a compromise; it is exactly right, once you see what
`dodge` does. `dodge` subdivides the **x** slot: each continent's space is cut
into one sub-bar per era. So it needs *both* neighbors: `x(continent)` says
*which slot* to divide, `color(era)` says *into how many pieces*. The sentence
lists them in the order `dodge` consumes them: the axis it splits along, then
the groups it splits into. The positions are not standing between `dodge` and
`color`; one of them *is* `dodge`'s other operand.

You also never have to scan the line to check the split is there.
`bar * mean * dodge` with no `color` or `group` does not draw a lone bar; it is
refused, pointing you at the missing channel. Legality holds the pair together,
so writing `color` far from `dodge` costs nothing: the two are bound whether or
not they sit side by side on the page.

---

## `*`: Derive

`*` modifies a mark **in place**, producing a compound form.

```r
bar * bin     # histogram
line * smooth # smoothed trend line
line * density # kernel density curve
```

The rule is fixed: **mark on the left, transform on the right.**

### Why the order is fixed

In Hangeul, a syllable is always *initial consonant + vowel*, never the reverse.
GOG's `*` works the same way: the mark is the subject; the transform is
what is done to it.

`bin * bar` is a type error, just as ㅏ + ㄱ is not a valid Hangeul syllable.

```r
bar * bin   # ✅  "a bar mark, derived by binning"
bin * bar   # ❌  Error — a transform cannot derive a mark
```

### What `*` is rigid about

The rigidity that does the work is the one above: the mark is on the left, always,
and no transform may stand there. That is what makes `bar * bin` a compound mark
rather than a function call, and it is why the atoms can be nouns.

Among the transforms themselves, the engine appends each `*` in turn and applies them
left to right. What that order can decide, though, turns out to be nothing, and the
reason is worth reading because it is the grammar being tight rather than loose. Two
transforms may only stand together when they answer *different* questions: where the
cells are, what is in them, or what scale the answer is read on. A pair that answers
the same question twice is refused rather than sequenced, so `bin` and `count` — both
of them tallies with cells of their own — cannot both be in one layer:

```{r}
#| error: true
data(gapminder_2007) + bar * bin(12) * count + x(life)
```

So the pairs that survive are exactly the ones whose halves cannot be swapped. A cell
has to exist before anything can be measured in it, which makes the cut *prior* rather
than merely earlier, and `bar * bin * mean` and `bar * mean * bin` draw the identical
plot. Order is not carrying meaning here because the two transforms already carry it
between them. See [Chaining transforms](transforms.qmd#chaining-transforms).

### Live examples

```{r}
# Histogram: life expectancy across all countries in 2007
data(gapminder_2007) + bar * bin + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Distribution of Life Expectancy — 2007")
```

```{r}
# Scatter + LOESS smooth layer on top
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  title("GDP vs Life Expectancy with LOESS Smooth")
```

```{r}
# Kernel density estimate
data(gapminder_2007) + line * density + x(life) + y(density) +
  x_label("Life expectancy (years)") + y_label("Density") +
  title("Kernel Density — Life Expectancy 2007")
```

### Chaining transforms

Each `*` appends one more transform to the layer, and they are applied left to right.
Which pairs are legal, and what each one means, is [its own
section](transforms.qmd#chaining-transforms). Two transforms have to divide
the work between saying **where the cells are** and saying **what is in
them**. A pair that answers the same half twice is refused rather than
resolved by picking one.

---

## `|` and `/`: Facet, or compose

`|` puts things side by side and `/` puts one above the other. Which things is
decided by what is written on the right: a `facet()` splits one plot into
panels, another *plot* arranges the two on a page.

With `facet()`, `|` produces small multiples arranged in columns, one panel per
category of the column it names.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

`/` produces small multiples stacked in rows:

```{r}
gm_two <- gapminder_2007[gapminder_2007$continent %in% c("Africa", "Europe"), ]
data(gm_two) + point + x(gdp) + y(life) / facet(continent)
```

Write both to cross them into a grid: each operator applies to the facet
written after it, read left to right. See [Faceting](faceting.qmd) for the
grid, what the panels share, and why the facet column must be a category.

With a plot on the right, the same two operators arrange separate plots on one
page — each keeping its own coordinate space, sharing an axis only where the two
name the same column on it:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

One pair of operators, two meanings, told apart by the operand rather than by a
second pair of words. See [Composition](composition.qmd).

---

## Why `+`, and not a pipe

Readers arriving from the tidyverse ask this, and it deserves a real answer,
because ggplot2's own author has said publicly that he would spell it
differently today [@wickham2018pipe]. The regret is on the record: ggplot,
ggplot2's predecessor, was written as function composition; the pipe did not
exist yet; `+` was what he reached for instead. Had `%>%` been available, he
says, the grammar would have piped, and there would have been no need for
ggplot2 at all. Three reasons are given for preferring the pipe. Two of them do
not reach GOG, and the third is the one that settles the question the other way.

**The first is about one language.** A pipe is a single idea that works
everywhere in R, so a grammar spelling composition differently is one more thing
to learn, and switching between the two spellings is a frequent source of errors
(his own included). That is true, and it is an argument about R's ecosystem. GOG
is one syntax spoken by four languages, and the arithmetic there comes out
differently. `+` can be given a meaning in R, Python and Julia. The pipe cannot:
Julia has `|>` in its base library and R has two of them, but **Python has no
pipe operator, and no way to add one**, since it lets you give new meanings only
to the operators it already has. Piping would have meant piping in two languages
and spelling it some other way in the third, which is the fork this project
exists to avoid. JavaScript decides nothing here, because it cannot overload any
operator at all, which is why it spells all four as words
([JavaScript](bindings/javascript.qmd)).

**The second is about arithmetic**, and it is the sharp one: you expect `x + y`
to equal `y + x`, and ggplot2's `+` does not always oblige. That objection is
answered a few sections up. Here the channels commute, and so does the mark
among them: move `color` ahead of `x`, or write the positions before the mark,
and the engine emits the same bytes. One thing does depend on order, two marks
swapped, and there the order is the z-order you asked for: the second draws on
top.

The other half of that objection is grouping, whether `x + (y + z)` means what
`(x + y) + z` means. The question cannot arise here, because there is no way to
write the left-hand side. `+` always takes a plot on its left, so two atoms with
no data beneath them are not a smaller expression waiting to be grouped. They
are refused:

```{r}
#| error: true
point + x(gdp)
```

No pile of atoms accumulates, so there is no grouping to get wrong.

**The third reason is the one that decides it**, and it is less an argument
against the pipe than a bill the pipe would present. A pipe applies functions:
`x |> f` means `f(x)`, so every atom would have to become a function taking a
plot and returning one. Make `bar` a function, then ask what `bar * bin` could
mean. Nothing at all: you cannot multiply two functions and get a histogram. The
derive operator exists only because marks and transforms are **things** rather
than actions, values that combine with each other before either one has met a
plot.

That is the same decision this book opens with, arriving a second time. ㄱ is a
letter, not an instruction. You put it in a slot next to other letters, and the
block is assembled from its pieces rather than performed in sequence. Hangeul's
atoms are nouns, which is exactly why they compose, and GOG's are nouns for the
same reason. A grammar built on the pipe is a grammar of verbs: it can layer,
because layering really is a sequence of additions, but it cannot derive.
Wickham's own pre-ggplot2 experiment was function composition, and it had no `*`
either.

So `+` here is a choice rather than something inherited. It is the one operator
available in every language this grammar speaks, it commutes wherever nothing
should turn on order, and it leaves the atoms as nouns so that `*` can exist.

---

## Summary

| Operator | Reads as | Example |
|----------|----------|---------|
| `+` | "…and also…" | `point + color(continent)` |
| `*` | "…derived by…" | `bar * bin` |
| <code>&#124;</code> | "…one panel per…" (columns) | <code>plot &#124; facet(continent)</code> |
| `/` | "…one panel per…" (rows) | `plot / facet(continent)` |
| <code>&#124;</code> | "…beside…" | <code>plot_a &#124; plot_b</code> |
| `/` | "…below…" | `plot_a / plot_b` |

A finite set of rules.
An infinite space of plots.
