# Operators

```{r}
#| include: false
source("R/setup.R")
```

How do you put two things in one picture? You join them with an **operator**,
and gog has exactly four. They are not atoms. They are the assembly rules: how
atoms combine.

Hangeul (한글) works the same way. Its consonants and vowels are the atoms. Its
operator is one rule: an initial consonant, a vowel and an optional final
consonant make a syllable block. The rule never changes, and a small set of
letters builds every syllable the language needs.

```
+   layer     (add a layer, a channel, or a label to a specification)
*   derive    (combine a mark with a transform into a compound mark)
|   facet ↔   (small multiples, arranged in columns)
/   facet ↕   (small multiples, arranged in rows)
```

`|` and `/` have a second use: they join whole plots into a **page**. A page is
one picture that holds two or more finished plots, each in its own part of it.
It is not a page of a document. That use has a chapter of its own
([Composition](composition.qmd)).

## `+`: layer

Every sentence you have written so far holds several atoms, and `+` joins most
of them. Each mark, channel, label and setting enters a specification through
it:

```r
data(my_table) + point + x(height) + y(weight) + color(species) + title("My plot")
```

Adding a **mark** (`point`, `line`, `bar`, …) starts a new layer. A
**channel** (`x()`, `color()`, …) written before any mark belongs to the plot,
and one written after a mark belongs to that mark. A **label atom** (`title()`,
`x_label()`) belongs to the plot wherever you write it ([Encoding
scope](encoding-scope.qmd)).

### Layering with `+`

A second mark draws a second layer over the first, and video editing works the
same way. A video is a stack of tracks: the picture, a logo in the corner, the
captions. Each track is separate, they play at the same time, and you see one
finished video.

Here the points are the first layer and the reference lines the second:

```{r}
#| echo: false
peek(gapminder_2007)
peek(life_bands)
```

```{r}
# Scatterplot: points, and a horizontal reference line at each band
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  data(life_bands) + rule + style(color = "darkgray", pattern = "dashed") +
  title("GDP vs life expectancy, 2007")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also rules from the life bands."*

The rules come after `point` in the sentence, so they are drawn over the
points.

### The order of `+`: data, mark, positions, refinements

This book writes its sentences in one order, and you may wonder whether the
engine requires it. Inside a single layer the mark and the channels
**commute**, which means that any order of them draws the same plot. The mark,
`x` and `color` can be arranged six ways, so here are all six.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
((data(iris_flowers) + bar * bin + x(petal_length) + color(species)) |
   (data(iris_flowers) + bar * bin + color(species) + x(petal_length))) /
((data(iris_flowers) + x(petal_length) + bar * bin + color(species)) |
   (data(iris_flowers) + x(petal_length) + color(species) + bar * bin)) /
((data(iris_flowers) + color(species) + bar * bin + x(petal_length)) |
   (data(iris_flowers) + color(species) + x(petal_length) + bar * bin))
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."*

Six sentences, and the same histogram six times. The six are not plots that
merely look alike. The engine writes six files, identical byte for byte.

So if the order does not change the picture, why does this book use one order:
*data, mark, positions, refinements*? Because a fixed order lets the code read
like an English sentence, with a natural, subject-first shape:

- **`data(...)` first.** It names the table the columns come from. A bare name
  like `petal_length` means nothing until a table is bound (Bind-Once), so the
  table is named first. The sentence must start here.
- **The mark and its `*`-chain second.** This is the subject. Nothing is drawn
  without a mark (Minimum Syllable): `bar * mean * dodge` says *what you are
  drawing* and *what was done to it*, as one compound word. It is the initial
  consonant of the block; everything after is said about it.
- **The positions third.** `x` and `y` say *where* the subject stands, and a
  mark cannot be drawn without them (the vowel to the mark's consonant), so they
  follow immediately.
- **The refinements last.** `color`, `size`, `order`, labels, `title`:
  everything optional. Delete every refinement and a legal plot remains; delete
  the mark or a position and nothing is drawn. Writing the removable parts last
  is also how you *build* a plot: the minimum sentence first, then refine
  ([First plot](first-plot.qmd)).

Only two parts of the order are rules: the sentence starts with a source
(`data()` or `query()`), and `style()` follows its mark. The rest is a reading
convention: inside one layer, any arrangement draws the same plot. Where you
write a channel *does* matter once a plot has two layers, because a channel
written before a mark and one written after it bind differently. [Encoding
scope](encoding-scope.qmd) gives the rule.
Learn one order and use it everywhere. Then your plots and the plots in this
book read the same way.

[Which orders matter](#which-orders-matter) lists what the engine enforces under
`+` and `*`, and what you are free to arrange.

#### A refinement far from what it refines

One part of the convention is easy to misread. A grouped bar chart pairs
`dodge` with `color`, yet the sentence puts them at opposite ends, with the
positions in between:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

*"Given the gapminder eras: bars derived by mean and dodge, x is continent, y is
life, color by era."*

`dodge` and `color(era)` are plainly related: on a bar, `dodge` has nothing to set
side by side until a channel such as `color` names the groups. So why not write them
together? Because the grammar deliberately gives them different operators.
`dodge` belongs to `*`, because it modifies the mark. The same modifier works on
`box * dodge` and `interval * range * dodge`. `color` belongs to `+`: it binds a
column to a channel.
`color * dodge` cannot be written: `*` needs a mark on its left. Joining the two
would make `dodge` depend on which channel splits the groups, and a modifier
must not depend on that.

Where `color(era)` stands is the book's convention, not something `dodge` asks
for. `color(era)` is optional, so it goes with the refinements, at the end.
`dodge` is optional too, but it needs a channel that splits the bars, here
`color(era)`. So you delete the two together: delete both and a legal plot
remains, with one bar per continent. You do not have to remember that `dodge`
needs a split, because the engine checks it. Without one, `bar * mean * dodge`
is refused rather than drawn as one bar per continent. The message tells you to
add a channel that splits the bars, such as `color`. So `color` can stand far
from `dodge` in the sentence and still work with it.

Three channels can make this split: `color`, `pattern` and `group`. Each cuts
the mark into one piece per era, and `dodge` sets the pieces side by side.
Without `dodge`, the pieces in one slot are drawn over one another. `color`
gives each piece its own color, `pattern` gives each its own texture, and both
draw a legend. `group` splits the mark and shows nothing more: no color, no
texture, no legend. A box and an interval take all three, and a bar takes only
`color` and `pattern`. The three sentences below differ only in that channel and
in their titles: the top one splits the boxes by `color(era)`, the middle one by
`pattern(era)`, and the bottom one by `group(era)`. The `theme(height = 900)`
after the outer parentheses makes the whole page 900 pixels tall, so each plot
has room ([How big is the page itself?](composition.qmd#how-big-is-the-page-itself)).

```{r}
((data(gm_eras) + box * dodge + x(continent) + y(life) + color(era) +
    title("color(era)")) /
   (data(gm_eras) + box * dodge + x(continent) + y(life) + pattern(era) +
      title("pattern(era)")) /
   (data(gm_eras) + box * dodge + x(continent) + y(life) + group(era) +
      title("group(era)"))) +
  theme(height = 900)
```

*"Given the gapminder eras: boxes derived by dodge, x is continent, y is life,
color by era, above the same with pattern by era, above the same grouped by
era."*

The boxes stand in the same places in all three plots. In the bottom one, every
box has the same color, and nothing says which box is 1957 and which is 2007. Use
`group` when another layer already shows the groups. The last plot in
[Interval](marks/interval.qmd#a-bar-with-its-error-bar) has two layers, and both
use `dodge`. The bars are split by `color(era)` and the error bars by
`group(era)`. So each error bar is drawn in black on top of its own bar, and the
bars' legend explains both.

## `*`: derive

A mark on its own draws the rows as they are. A histogram counts them first,
and a trend line fits a curve to them first. That computing step needs a word
of its own. `*` joins a transform to a mark, and the result is one
**compound mark**.

```r
bar * bin     # histogram
line * smooth # smoothed trend line
line * density # kernel density curve
```

`*` has one rule and one freedom. The mark goes on the left and the transforms
go on the right. Among the transforms themselves, any order draws the same plot.
The rest of this section explains both.

### Why the order is fixed

Start with the rule. In Hangeul, a syllable is always *initial consonant +
vowel*, never the reverse. gog's `*` works the same way: the mark is the
subject; the transform is what is done to it.

`bin * bar` is refused, just as ㅏ (*a*) + ㄱ (*g*) is not a valid Hangeul
syllable. In the right order, ㄱ (*g*) + ㅏ (*a*) makes 가 (*ga*). In the same
way, the mark and then the transform make `bar * bin`:

```r
bar * bin   # a bar mark, derived by binning
```

The refusal names the fix:

```{r}
#| error: true
bin * bar
```

### One fixed place, and one free order

The mark stands on the left, always, and no transform may stand there. That is
what makes `bar * bin` a compound mark rather than a function call. [Why `+`,
and not a pipe](#why-and-not-a-pipe) explains why it lets the atoms stay nouns.

After the mark, the transforms **commute**, as the mark and channels do under
`+`: any order of them draws the same plot. Here `bin` and `mean` are written
both ways:

```{r}
(data(gapminder_2007) + bar * bin * mean + x(gdp) + y(life) +
   title("bar * bin * mean")) |
  (data(gapminder_2007) + bar * mean * bin + x(gdp) + y(life) +
     title("bar * mean * bin"))
```

*"Given gapminder 2007: bars derived by bin and mean, x is gdp, y is life,
beside bars derived by mean and bin, x is gdp, y is life."*

The two plots are identical. The engine does not run the transforms in the order
you type them. It finds the order the work needs, and here it runs `bin` first
in both.

The order you write is therefore a matter of style, not a rule. This book
writes the transforms in the order they run, so the sentence reads in the same
order as the work: `bar * bin * mean` cuts the axis into bands, then averages inside each
one.

`bin` must run first because of what `mean` needs. A statistic reduces a
**group** to one number, and a continuous column has no groups: every value
stands alone, so its mean is itself. Ask for the mean of `life` at each value of `gdp`, and each group holds
one country:

```{r}
#| message: true
data(gapminder_2007) + point * mean + x(gdp) + y(life)
```

*"Given gapminder 2007: points derived by mean, x is gdp, y is life."*

The picture is the plain scatterplot, and the note printed with it says why:
each mean is one row's own value. `bin` is what makes the groups, by turning
many distinct numbers into one band. So the cut comes first: without it, the
mean has nothing to average.

So two transforms work together only when they do different jobs. There are four
jobs: saying where the cells are, what is in them, what scale the answer is read
on, and where the marks sit ([Chaining
transforms](transforms.qmd#chaining-transforms) explains each one). A pair that
does the same job twice is refused rather than run in sequence. `bin` and
`count` both count the rows, so each cell would be measured twice:

```{r}
#| error: true
data(gapminder_2007) + bar * bin(12) * count + x(life)
```

The order does not decide a refusal either: `bar * count * bin(12)` is refused
just like the pair above. [What cannot be chained, and
why](transforms.qmd#what-cannot-be-chained-and-why) shows why a pair is
refused, one refusal at a time, and
[Combinations](combinations.qmd#transforms-and-transforms) has the whole grid of
pairs.

### Live examples

The start of this section listed three compound marks: `bar * bin`,
`line * smooth` and `line * density`. You will use them often. Here each one is
drawn, so you can see what its transform made from the rows.

```{r}
# Histogram: life expectancy across all countries in 2007
data(gapminder_2007) + bar * bin + x(life) +
  x_label("Life expectancy (years)") +
  title("Distribution of life expectancy, 2007")
```

```{r}
# Scatterplot, with a LOESS smooth on top
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  title("GDP vs life expectancy, with a LOESS smooth")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

```{r}
# Kernel density estimate
data(gapminder_2007) + line * density + x(life) +
  x_label("Life expectancy (years)") +
  title("Kernel density: life expectancy, 2007")
```

*"Given gapminder 2007: a line derived by density, x is life."*

In each plot, the compound mark draws what its transform computed rather than
the rows. The bars are counts of countries in bands of life expectancy. The line
through the points is a fitted curve, and the last line is an estimated density.
`gapminder_2007` has no `count` or `density` column. `bin` and `density` compute
those numbers and put them on y themselves, so neither sentence needs a `y()`,
and the axes are labeled Count and Density. You can still give the column a
name of your own with `y()` ([Columns a transform
writes](transforms.qmd#columns-a-transform-writes)).

## `|` and `/`: facet, or compose

Two questions lead to the last two operators: how the same plot looks for each
group, and how two different plots fit on one page. `|` puts things side by
side and `/` puts one above the other. What you write on the right decides the
job: a `facet()` splits one plot into panels, and a second plot is placed beside
or below the first.

With `facet()`, `|` produces small multiples arranged in columns, one panel per
category of the column it names.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, split into panel columns by
continent."*

`/` stacks the panels in rows instead. The sentence is the same, with `/` in
place of `|`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) / facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, split into panel rows by
continent."*

Stacked, the panels line up their x axes, so a value sits directly above the
same value in the panel below. Side by side, they line up their y axes instead.

Write `|` and `/` in one sentence to combine them into a grid. Each operator
applies to the facet written after it, read from left to right. See
[Faceting](faceting.qmd) for when to choose rows or columns, the grid, what the
panels share, and why the facet column must be categorical.

With a plot on the right, the same two operators arrange separate plots on one
page, and each plot keeps its own axes:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

The plot on the left shows life expectancy on `y`, and the plot on the right
shows a count, so each draws its own vertical axis. Two plots side by side that
name the same column on `y` share that axis, and it is drawn once. In the same
way, two plots stacked with `/` that name the same column on `x` share that
axis.

Neither reading is new. The facet reading comes from Trellis display
[@becker1996trellis], and the composition reading from patchwork
[@pedersen2019patchwork], which writes `p1 | p2` and `p1 / p2` the same way
([Where these operators come
from](composition.qmd#where-these-operators-come-from)).

One pair of operators, two jobs. [Composition](composition.qmd) covers the
second.

## Which orders matter

This book writes every sentence in the same order: data, mark, positions,
refinements. Most of that order is the book's habit, and the engine draws the
same plot if you change it. Only a few parts are rules that the engine enforces.
The table below sorts the orderings under `+` and under `*` into three kinds,
and only one kind needs your attention.

| Kind | What it means | Under `+` | Under `*` |
|---|---|---|---|
| **Free** | Either way draws exactly the same picture. | In a plot with one mark: the mark, its positions and its other channels, in any arrangement. Labels go anywhere. Only `style()` must follow its mark. | The transforms after one mark, in any order. The book writes them in the order they run, as a style; the engine finds that order itself. |
| **Refused** | The engine stops and names what to move. | Atoms with no table named before them: `point + x(gdp)`. A `style()` written before any mark. | A transform on the left: `bin * bar`. |
| **Meaningful** | No error, and a different picture. | A channel before the marks, or after one of them, once a plot has two. Two marks swapped: the second draws on top. | Nothing. No order under `*` changes the picture. |

Start with the cell where `*` meets **Meaningful**. `*` fixes one thing only,
which is that the mark stands on its left. Once the mark is there, no
arrangement of what follows changes what you get.

**Free** covers most arrangements. The two sentences below draw the identical
plot, and the book writes the first because a fixed order is easier to read:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + x(gdp) + point + y(life))
```

Both name `life` on `y`, so the two plots share that axis and it is drawn once,
on the left. Each plot still draws its own `gdp` axis underneath it.

**Refused** means an atom comes before the thing it needs. A plot needs its
table first, so `+` must have a plot on its left. A setting needs a mark, so a
`style()` before any mark is refused:

```{r}
#| error: true
data(gapminder_2007) + style(size = 3) + point + x(gdp) + y(life)
```

**Meaningful** is the only kind that needs care, because it is the only kind
the engine says nothing about. With two marks in a sentence, a channel written
before them belongs to the plot and reaches both marks, and a channel written
after one belongs to that mark. Both are legal, neither warns, and they draw
different pictures. Here only `color(era)` moves, from before the marks to after
`point`:

```{r}
(data(gm_eras) + x(gdp, scale = "log") + y(life) + color(era) +
   point + line * smooth + title("color(era) before the marks")) |
  (data(gm_eras) + x(gdp, scale = "log") + y(life) + point + color(era) +
     line * smooth + title("color(era) after point"))
```

*"Given the gapminder eras: x is gdp on a log scale, y is life, color by era,
points and also lines derived by smooth, beside x is gdp on a log scale, y is
life, points colored by era, and also a line derived by smooth."*

On the left, `color(era)` reaches both marks, so the trend splits into one line
per era. On the right, it belongs to the points alone, and one gray line is
fitted to both eras together. [Encoding scope](encoding-scope.qmd) gives the
whole rule. It is also why this book keeps one order: when every sentence looks
the same, a different arrangement is a deliberate signal.

The order of the marks is the second case, and here too the engine gives no
message. The marks are drawn in the order you wrote them, so the second mark
sits on top of the first. Here the bars are light gray, so you can see which
mark is on top:

```{r}
(data(gapminder_2007) + x(continent) + y(life) +
   bar * mean + style(color = "lightgray") + point +
   title("bar * mean, then point")) |
  (data(gapminder_2007) + x(continent) + y(life) +
     point + bar * mean + style(color = "lightgray") +
     title("point, then bar * mean"))
```

*"Given gapminder 2007: x is continent, y is life, bars derived by mean, colored
lightgray, and also points, beside points and also bars derived by mean, colored
lightgray."*

On the left, every point is drawn over the bars. On the right, the bars are
drawn last, so they cover the points below each mean, which show only faintly
through the gray.

### What groups first

The three kinds above, free, refused and meaningful, are about the order in
which you write the atoms. **Precedence** answers a different question: when a
sentence holds more than one operator, which operator groups its atoms first. Arithmetic has the same idea. `2 + 3 * 4` is 14 and not 20,
because `*` groups before `+`. It is read as `2 + (3 * 4)`, as if the
parentheses were written. A sentence in gog is read the same way, so precedence
decides which atoms each operator joins.

This order of grouping belongs to the language you write in, not to gog, and
gog cannot change it. R and Python share one order. Julia differs in one place:
it groups `|` together with `+`. When the two meet, the one written first groups
first ([Julia](bindings/julia.qmd#the-operators-and-their-precedence) shows
where this changes a sentence). The table gives the order in each of the three
languages:

| Groups | R | Python | Julia |
|---|---|---|---|
| First | `*` and `/` | `*` and `/` | `*` and `/` |
| Next | `+` | `+` | `+` and `|`, from left to right |
| Last | `|` | `|` | |

JavaScript has no column in the table, because it spells each operator as a
word, such as `beside()`. The things the word joins are written inside its
parentheses, so you see the grouping instead of remembering it
([JavaScript](bindings/javascript.qmd)).

For one plot, this is the order you want in all three languages. So a plot needs
no parentheses of its own, even with a facet:

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

R reads that sentence as
`(data(gapminder_2007) + (bar * bin) + x(life)) | facet(continent)`. The
transform joins its mark first, the atoms join the plot next, and the facet
splits the finished plot last. Python and Julia read it the same way.

A row facet needs no parentheses either, because a facet always splits the
whole plot.

An atom can also be written after a facet, like the title in the sentence below:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) |
  facet(continent) + title("Each continent in its own panel")
```

The title belongs to the whole plot, so it appears once, above all five panels,
in every language.

### Parentheses on a page

That order works for one plot, with or without a facet. It can go wrong when
`|` or `/` has no `facet()` on its right, and joins two separate plots into one
page instead ([Composition](composition.qmd)). Then each side of the operator
must be a whole plot, from its `data()` to its last atom. Without parentheses,
precedence can join part of one plot to part of the other before either plot is
complete. Here are the two plots from earlier in this chapter, set one above
the other with no parentheses:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) /
  data(gapminder_2007) + bar * count + x(continent)
```

`/` groups before `+`, so it takes only the atom on each side of it: `y(life)`
on the left and the second `data()` on the right. R reads the sentence as
`… + x(gdp) + (y(life) / data(gapminder_2007)) + …`. For a facet, that grouping
does no harm, because a facet splits the whole plot anyway. For a page it does:
`/` must place a whole plot below another, and `y(life)` is only one atom of a
plot. So gog refuses the sentence, and its message names `y()`, the atom that
`/` took. Python and Julia give the same refusal.

With `|` in place of `/`, the result depends on the language. R and Python
group `|` last, so both plots are complete before `|` joins them, and the page
draws. Julia groups `|` together with `+`, from left to right. So `|` joins the
first plot to the second plot's `data()` alone, and gog refuses the sentence.

So whenever this book composes plots in R, Python or Julia, it writes each plot
inside its own parentheses. Then every plot is complete before `|` or `/` joins
it, and the page means the same thing in all three languages. In JavaScript each
plot is written as `plot(…)`, so it is complete already.

Two more places need parentheses. The first is a page inside a page. `/` groups
before `|` in R, Python and Julia, so `a | b / c` is read as `a | (b / c)`: one
plot beside two plots, one above the other. The examples below use three plots
from this chapter, titled a, b and c:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) + title("a")) |
  (data(gapminder_2007) + bar * count + x(continent) + title("b")) /
  (data(gapminder_2007) + bar * bin + x(life) + title("c"))
```

*"Given gapminder 2007: points, x is gdp, y is life; beside bars derived by
count, x is continent, above bars derived by bin, x is life."*

To put a row of two above a third plot, put parentheses around the row,
`(a | b) / c`:

```{r}
((data(gapminder_2007) + point + x(gdp) + y(life) + title("a")) |
   (data(gapminder_2007) + bar * count + x(continent) + title("b"))) /
  (data(gapminder_2007) + bar * bin + x(life) + title("c"))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent; above bars derived by bin, x is life."*

Both pages draw the same in all three languages, because each one writes its
parentheses out.

The second place is a `theme()` for the whole page. Put parentheses around the
page before you add it, and the `theme()` sizes the page. This page is 500
pixels wide, and the two plots share that width:

```{r}
((data(gapminder_2007) + point + x(gdp) + y(life) + title("a")) |
   (data(gapminder_2007) + bar * count + x(continent) + title("b"))) +
  theme(width = 500)
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

Without the outer parentheses, the languages disagree. Julia still gives the
`theme()` to the page. R and Python give it to the second plot, because `+`
groups before `|`, so only b is 500 pixels wide and the page keeps its 800. To
draw that page in every language, Julia included, put the `theme()` inside the
second plot's parentheses:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) + title("a")) |
  (data(gapminder_2007) + bar * count + x(continent) + title("b") +
     theme(width = 500))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

Written this way, with every parenthesis in place, each of the four pages draws
the same in R, Python and Julia ([How big is the page
itself?](composition.qmd#how-big-is-the-page-itself)).

The table collects the rules for a page. Each `(…)` holds one whole plot, from
its `data()` to its last atom, and the last column gives the same page in
JavaScript's words:

| Page | R, Python and Julia | JavaScript |
|---|---|---|
| Two plots side by side | `(…) | (…)` | `beside(plot(…), plot(…))` |
| One plot above another | `(…) / (…)` | `below(plot(…), plot(…))` |
| A row of two above a third | `((…) | (…)) / (…)` | `below(beside(plot(…), plot(…)), plot(…))` |
| A height for the whole page | `((…) | (…)) + theme(height = 260)` | `beside(plot(…), plot(…), theme({ height: 260 }))` |

Each language's chapter shows these rules in its own spelling:
[R](bindings/r.qmd), [Python](bindings/python.qmd), [Julia](bindings/julia.qmd)
and [JavaScript](bindings/javascript.qmd).

### What parentheses may group

The table above shows what parentheses group: whole plots and whole pages. You
might also try them around several marks, to give a second table to all of
them. That does not work, because parentheses do not group *marks*. A second
`data()` applies only to the mark written directly after it ([Data](data.qmd)):

```{r}
#| echo: false
peek(actuals)
peek(forecast)
```

```{r}
#| error: true
data(actuals) + x(year) + y(sales) +
  line +
  (data(forecast) + point + area)
```

gog refuses that instead of drawing it. The refusal is better than the
alternative. The engine cannot give the second table to both marks inside the
parentheses. Its only other choice is to discard both marks and draw a plot with
two layers missing, without saying so. To put two marks on the second table,
write `data()` before each of them.

The same refusal covers a position or a title inside the parentheses, for the
same reason. In JavaScript, where the operators are words, the same mistake is
a `plot()` handed to another `plot()`, and it is refused there too
([JavaScript](bindings/javascript.qmd#a-second-table) shows both spellings).

## Why `+`, and not a pipe

Readers who know R's tidyverse packages ask why gog does not use a pipe. A pipe
passes a value into a function: `a |> f` means `f(a)`. Hadley Wickham,
ggplot2's author, has said publicly that the pipe would have been the better
interface [@wickham2018pipe]. ggplot, ggplot2's predecessor, was written as
function composition, and it could have used a pipe if one had existed. ggplot2
used `+` instead.

He also answers the question a reader asks next: why not change ggplot2 now?
It cannot change: its functions take the mapping and the data first, while a
pipe needs the plot as the first argument. And he judges that a new package
built only to make this change would cost more work than the change would be
worth. So ggplot2's `+` comes from its history, not a fresh choice, and a
grammar designed today has to make that choice for itself. He gives two reasons
for preferring the pipe, and neither applies to gog. A third consideration is
gog's rather than his, and it is the one that decides the question.

His first reason is about one language. A pipe is a single idea that works
everywhere in R, so a grammar that spells function composition differently is
one more thing to learn. Switching between the two spellings is a frequent
source of errors, his own included. That is true, and it is an argument about R
and its packages. gog is one syntax written in four languages, and that changes
the answer. `+` can be given a meaning in R, Python and Julia. The pipe cannot
be given a gog meaning. Julia's `|>` and R's two pipes already mean something
gog must not change. Python has no pipe operator at all, and no way to add one:
you can give new meanings only to the operators it already has. In R and Julia
a pipe would work with the meaning it already has. Python would need another
spelling, and the languages would then have two different grammars. This
project exists to avoid that. JavaScript decides nothing here, because it cannot
overload any operator at all, which is why it spells all four as words
([JavaScript](bindings/javascript.qmd)).

His second reason is about arithmetic, and it is the stronger of his two
reasons. A `+` looks like the plus sign of arithmetic, so you expect it to
behave like one, in two ways. The order should not matter: `a + b` should equal
`b + a`. And the grouping should not matter: `a + (b + c)` should equal
`(a + b) + c`. ggplot2's `+` does not always behave either way.

The six permutations drawn earlier in this chapter answer the first half, the
order. In a plot with one mark, the mark and its channels commute. Move `color`
ahead of `x`, or write the positions before the mark, and the engine produces
the identical file. Once a plot has two marks, two orders change the picture:
where a channel is written, and the order of the marks. In both, the order says
something you chose: which marks a channel reaches, and which mark is drawn on
top ([Which orders matter](#which-orders-matter)).

The second half, the grouping, never comes up in gog. To group `b + c` first,
`b + c` would have to mean something without `a`. In gog it cannot, because a
sentence must start with a table. `+` joins an atom to a plot, and a plot starts
with `data()`. So two atoms joined with no table before them, such as `point`
and `x(gdp)`, are refused:

```{r}
#| error: true
point + x(gdp)
```

The third point is gog's own, and it is what decides the question. It is not an
argument against the pipe. It is a statement of what a pipe would remove. A
pipe applies functions, so every atom would have to become a function taking a
plot and returning one. Make `bar` a function, then ask what `bar * bin` could
mean. There is no answer: you cannot multiply two functions and get a histogram.
The derive operator exists only because marks and transforms are *things*
rather than actions. They are values, and they
combine with each other before either one is joined to a plot.

This is the decision the book opens with: a plot is a specification, not drawing
code. Hangeul shows the same idea. ㄱ is a letter, not an instruction: it fills
one slot of a syllable block, and the block is built from its letters, not
performed one step after another. gog's atoms fill slots in the same way. The
mark is the initial consonant, `x`, `y` and `z` are the vowel, and a channel
with a legend is the final consonant ([The grammar](grammar.qmd#channels-a-column-made-visible)).
A layer is built from them as a block is built from its letters. Hangeul's
letters are nouns, which is exactly why they combine, and gog's atoms are nouns
for the same reason. A grammar built on the pipe is a grammar of
verbs: it can layer, because layering really is a sequence of additions, but it
cannot derive. ggplot, ggplot2's predecessor, was written as function
composition, and it had no `*` either.

So `+` here is a choice rather than something inherited. Unlike the pipe, it can
be given a meaning in R, Python and Julia. It commutes wherever order should not
matter, and it leaves the atoms as nouns so that `*` can exist.

## Summary

Each operator reads as a phrase when a plot is read aloud, and `|` and `/` have
two readings each:

| Operator | Reads as | Example |
|----------|----------|---------|
| `+` | "…and also…" | `point + line` |
| `*` | "…derived by…" | `bar * bin` |
| <code>&#124;</code> | "…split into panel columns by…" | <code>data(gm) + point + x(gdp) + y(life) &#124; facet(continent)</code> |
| `/` | "…split into panel rows by…" | `data(gm) + point + x(gdp) + y(life) / facet(era)` |
| <code>&#124;</code> | "…beside…" | <code>(one specification) &#124; (another)</code> |
| `/` | "…above…" | `(one specification) / (another)` |

On a page, put each plot inside its own parentheses, and the page inside
parentheses before a `theme()`. Then the page draws the same in R, Python and
Julia.

A finite set of rules, and more plots than you will ever need to draw.

## What comes next

The operators say how atoms combine into a sentence. Where you write each atom
decides what it binds to, and [Encoding scope](encoding-scope.qmd) gives that
rule in full.
