# Operators

```{r}
#| include: false
source("R/setup.R")
```

How do you put two things in one picture? You join them with an **operator**,
and gog has exactly four. They are not atoms. They are the assembly rules: how
atoms combine.

Hangeul (한글) works the same way. Its consonants and vowels are the atoms. Its
operator is one rule: an initial consonant, a vowel and an optional final
consonant make a syllable block. The rule never changes, and a small set of
letters builds every syllable the language needs.

```
+   layer     (add a layer, a channel, or a label to a specification)
*   derive    (combine a mark with a transform into a compound mark)
|   facet ↔   (small multiples, arranged in columns)
/   facet ↕   (small multiples, arranged in rows)
```

`|` and `/` also arrange whole plots on a page. That second use has a chapter
of its own ([Composition](composition.qmd)).

---

## `+`: layer

Every sentence you have written so far holds several atoms, and `+` joins most
of them. Each mark, channel, label and setting enters a specification through
it:

```r
data(my_table) + point + x(height) + y(weight) + color(species) + title("My plot")
```

Adding a **mark** (`point`, `line`, `bar`, …) starts a new layer. A
**channel** (`x()`, `color()`, …) written before any mark belongs to the plot,
and one written after a mark belongs to that mark. A **label atom** (`title()`,
`x_label()`) belongs to the plot wherever you write it ([Encoding
scope](encoding-scope.qmd)).

### Layering with `+`

A second mark draws a second layer over the first, and video editing works the
same way. A video is a stack of tracks: the picture, the sound, the captions.
Each track is separate, they play at the same time, and you see one finished
video.

Here the points are the first layer and the reference lines the second:

```{r}
#| echo: false
peek(gapminder_2007)
peek(life_bands)
```

```{r}
# Scatterplot: points, and a horizontal reference line at each band
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  data(life_bands) + rule + style(color = "darkgray", pattern = "dashed") +
  title("GDP vs life expectancy, 2007")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also rules from the life bands."*

The rules come after `point` in the sentence, so they are drawn over the
points.

### The order of `+`: data, mark, positions, refinements

This book writes its sentences in one order, and you may wonder whether the
engine requires it. Inside a single layer the mark and the channels
**commute**, which means that any order of them draws the same plot. The mark,
`x` and `color` can be arranged six ways, so here are all six.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
((data(iris_flowers) + bar * bin + x(petal_length) + color(species)) |
   (data(iris_flowers) + bar * bin + color(species) + x(petal_length))) /
((data(iris_flowers) + x(petal_length) + bar * bin + color(species)) |
   (data(iris_flowers) + x(petal_length) + color(species) + bar * bin)) /
((data(iris_flowers) + color(species) + bar * bin + x(petal_length)) |
   (data(iris_flowers) + color(species) + x(petal_length) + bar * bin))
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."*

Six sentences, and the same histogram six times. The six are not plots that
merely look alike. The engine writes six files, identical byte for byte.

So if the order does not change the picture, why does this book use one order:
*data, mark, positions, refinements*? Because a fixed order lets the code read
like an English sentence, with a natural, subject-first shape:

- **`data(...)` first.** It names the table the columns come from. A bare name
  like `petal_length` means nothing until a table is bound (Bind-Once), so the
  table is named first. The sentence must start here.
- **The mark and its `*`-chain second.** This is the subject. Nothing is drawn
  without a mark (Minimum Syllable): `bar * mean * dodge` says *what you are
  drawing* and *what was done to it*, as one compound word. It is the initial
  consonant of the block; everything after is said about it.
- **The positions third.** `x` and `y` say *where* the subject stands, and a
  mark cannot be drawn without them (the vowel to the mark's consonant), so they
  follow immediately.
- **The refinements last.** `color`, `size`, `order`, labels, `title`:
  everything optional. Delete every refinement and a legal plot remains; delete
  the mark or a position and nothing is drawn. Writing the removable parts last
  is also how you *build* a plot: the minimum sentence first, then refine
  ([First plot](first-plot.qmd)).

Only two parts of the order are rules: the sentence starts with `data()`, and
`style()` follows its mark. The rest is a reading convention: inside one layer,
any arrangement draws the same plot. Where you write a channel *does* matter once
a plot has two layers, because a channel written before a mark and one written
after it bind differently. [Encoding scope](encoding-scope.qmd) gives the rule.
Learn one order and use it everywhere. Then your plots and the plots in this
book read the same way.

[Which orders matter](#which-orders-matter) lists what the engine enforces under
`+` and `*`, and what you are free to arrange.

#### A refinement far from what it refines

One part of the convention is easy to misread. A grouped bar chart pairs
`dodge` with `color`, yet the sentence puts them at opposite ends, with the
positions in between:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

*"Given the gapminder eras: bars derived by mean and dodge, x is continent, y is
life, color by era."*

`dodge` and `color(era)` are plainly related: `dodge` has nothing to set side by
side until a channel such as `color` names the groups. So why not write them
together? Because the grammar deliberately gives them different operators.
`dodge` belongs to `*`, because it modifies the mark. The same modifier works on
`box * dodge` and `interval * range * dodge`, where `group` can make the split
instead of `color`. `color` belongs to `+`: it binds a column to a channel.
`color * dodge` cannot be written: `*` needs a mark on its left. Joining the two
would make `dodge` depend on which channel splits the groups, and a modifier
must not depend on that.

The order is not a compromise. It follows from what `dodge` does. `dodge`
subdivides the `x` slot: each continent's slot is cut into one bar per era. So
`dodge` uses both atoms: `x(continent)` says *which slot* to divide,
`color(era)` says *into how many pieces*. The sentence lists them in the order
`dodge` uses them: the axis it splits along, then the groups it splits into. The
positions do not separate `dodge` from what it needs; one of them is the slot
`dodge` divides.

You also never have to read back through the sentence to check that the split is
there. If no channel splits the mark, `bar * mean * dodge` is refused rather
than drawn as one bar per slot. The message tells you to add a channel that
splits it, such as `color`. So `color` can stand far from `dodge` in the
sentence and still work with it.

---

## `*`: derive

A mark on its own draws the rows as they are. A histogram counts them first,
and a trend line fits a curve to them first. That computing step needs a word
of its own. `*` joins a transform to a mark, and the result is one
**compound mark**.

```r
bar * bin     # histogram
line * smooth # smoothed trend line
line * density # kernel density curve
```

`*` has one rule and one freedom. The mark goes on the left and the transforms
go on the right. Among the transforms themselves, any order draws the same plot.
The rest of this section explains both.

### Why the order is fixed

Start with the rule. In Hangeul, a syllable is always *initial consonant +
vowel*, never the reverse. gog's `*` works the same way: the mark is the
subject; the transform is what is done to it.

`bin * bar` is refused, just as ㅏ (*a*) + ㄱ (*g*) is not a valid Hangeul
syllable.

```r
bar * bin   # a bar mark, derived by binning
```

The refusal names the fix:

```{r}
#| error: true
bin * bar
```

### One fixed place, and one free order

The mark stands on the left, always, and no transform may stand there. That is
what makes `bar * bin` a compound mark rather than a function call. [Why `+`,
and not a pipe](#why-and-not-a-pipe) explains why it lets the atoms stay nouns.

Write the transforms in the order they happen. This is a reading convention: any
order draws the same plot, and this one lets you follow the steps as you read.
`bar * bin * mean` cuts the axis into bands, then averages inside each one. That
is the order it runs in, and there is no other order it could run in.

The reason is what `mean` needs. A statistic reduces a **group** to one number,
and a continuous column has no groups: every value stands alone, so its mean is
itself. `bin` is what makes the groups, by turning many distinct numbers into
one band. So the cut comes first: without it, the mean has nothing to average.

So two transforms work together only when they do different jobs. There are four
jobs: making the groups, measuring inside them, rescaling the result, and
arranging the marks. A pair that does the same job twice is refused rather than
run in sequence. `bin` and `count` both count the rows, so each cell would be
measured twice:

```{r}
#| error: true
data(gapminder_2007) + bar * bin(12) * count + x(life)
```

The engine decides the running order from the transforms themselves, not from
where you typed them. The order you type can never be the wrong order, although
the pair itself can still be refused, as the refusal above shows.
[Chaining transforms](transforms.qmd#chaining-transforms) shows why a pair is
refused, one refusal at a time, and
[Combinations](combinations.qmd#transforms-and-transforms) has the whole grid of
pairs.

### Live examples

The three compound marks named when `*` was introduced appear in the sentences
you will write most often. Here each one is drawn, so you can see what the
transform did to the rows before the mark drew them.

```{r}
# Histogram: life expectancy across all countries in 2007
data(gapminder_2007) + bar * bin + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Distribution of life expectancy, 2007")
```

```{r}
# Scatterplot, with a LOESS smooth on top
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  title("GDP vs life expectancy, with a LOESS smooth")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

```{r}
# Kernel density estimate
data(gapminder_2007) + line * density + x(life) + y(density) +
  x_label("Life expectancy (years)") + y_label("Density") +
  title("Kernel density: life expectancy, 2007")
```

*"Given gapminder 2007: a line derived by density, x is life, y is density."*

In each plot, the compound mark draws what its transform computed rather than
the rows. The bars are counts of countries in bands of life expectancy. The line
through the points is a fitted curve, and the last line is an estimated density.

---

## `|` and `/`: facet, or compose

Two questions lead to the last two operators: how the same plot looks for each
group, and how two different plots fit on one page. `|` puts things side by
side and `/` puts one above the other. What you write on the right decides the
job: a `facet()` splits one plot into panels, and a second plot is placed beside
or below the first.

With `facet()`, `|` produces small multiples arranged in columns, one panel per
category of the column it names.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, split into panel columns by
continent."*

`/` stacks the panels in rows. Five rows would make each panel too short, so
this example uses two continents:

```{r}
gm_two <- gapminder_2007[gapminder_2007$continent %in% c("Africa", "Europe"), ]
data(gm_two) + point + x(gdp) + y(life) / facet(continent)
```

*"Given the gm two table: points, x is gdp, y is life, split into panel rows by
continent."*

Write `|` and `/` in one sentence to combine them into a grid. Each operator
applies to the facet written after it, read from left to right. See
[Faceting](faceting.qmd) for the grid, what the panels share, and why the facet
column must be categorical.

With a plot on the right, the same two operators arrange separate plots on one
page, and each plot keeps its own axes:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

These two measure different things on `y`, life expectancy and a count, so each
draws its own vertical axis. When two plots side by side name the same column
on `y`, that axis is shared and drawn once. Two plots one above the other share
`x` the same way.

One pair of operators, two jobs. [Composition](composition.qmd) covers the
second.

---

## Which orders matter

Most of the order *data, mark, positions, refinements* is a reading convention,
and the engine enforces the rest. The difference tells you where to be careful.
There are three kinds of ordering, and only one of them needs your attention.

| Kind | What it means | Under `+` | Under `*` |
|---|---|---|---|
| **Free** | Either way draws exactly the same picture. | Inside one layer: the mark, its positions and its other channels, in any arrangement. Labels go anywhere. Only `style()` must follow its mark. | Two transforms after one mark. Write them in the order they happen; the engine runs them in that order either way. |
| **Refused** | The engine stops and names what to move. | Atoms with no table named before them: `point + x(gdp)`. A `style()` written before any mark. | A transform on the left: `bin * bar`. |
| **Meaningful** | No error, and a different picture. | A channel before the marks, or after one of them, once a plot has two. Two marks swapped: the second draws on top. | Nothing. No order under `*` changes the picture. |

Start with the cell where `*` meets **Meaningful**. `*` fixes one thing only,
which is that the mark stands on its left. Once the mark is there, no
arrangement of what follows changes what you get.

**Free** covers most arrangements. The two sentences below draw the identical
plot, and the book writes the first because a fixed order is easier to read:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + x(gdp) + point + y(life))
```

Both name `life` on `y`, so the two plots share that axis and it is drawn once,
on the left. Each plot still draws its own `gdp` axis underneath it.

**Refused** means an atom comes before the thing it needs. A plot needs its
table first, so `+` must have a plot on its left. A setting needs a mark, so a
`style()` before any mark is refused:

```{r}
#| error: true
data(gapminder_2007) + style(size = 3) + point + x(gdp) + y(life)
```

**Meaningful** is the only kind that needs care, because it is the only kind
the engine says nothing about. With two marks in a sentence, a channel written
before them belongs to the plot and reaches both, and a channel written after
one belongs to that mark. Both are legal, neither warns, and they draw
different pictures. That is the subject of [Encoding
scope](encoding-scope.qmd). It is also why this book keeps one order: when
every sentence looks the same, a different arrangement is a deliberate signal.

The marks themselves are the other case the engine does not report. Two marks
swapped draw in the order you wrote them, so the second sits on top of the
first.

### What groups first

Those three kinds are about where you may put a word. Precedence is a
different question: how the words you did write get grouped. It is what lets
the sentence below carry no parentheses at all:

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

`*` and `/` group first, then `+`, then `|`. So the transform joins its mark
first, the atoms join the plot next, and the facet splits the finished plot
last. `/` has the same precedence as `*`, so in the row facet drawn earlier,
`y(life) / facet(continent)` groups first. The engine still splits the whole
plot: that sentence draws the same file as the one with the plot in parentheses.
[What parentheses may group](#what-parentheses-may-group) has the rule.

Those are R's own precedences, and Python's. Julia puts `|` in the addition
tier, beside `+` rather than below it. A sentence that writes `+` after a `|`
therefore groups differently there, which is why a composed page is
parenthesized. [Julia](bindings/julia.qmd) compares four expressions in both
languages, and only one of them groups differently. JavaScript can give no
operator a meaning of its own, so it spells the four as words, which makes the
grouping something you see instead of something you remember
([JavaScript](bindings/javascript.qmd)).

### What parentheses may group

Precedence decides the grouping when you write no parentheses, and parentheses
are how you change it. That is the whole of what they do. They group **plots**,
which is why a composed page uses them. `/` is applied before `|`, so
`a | b / c` reads as `a | (b / c)` unless you say otherwise
([Composition](composition.qmd)).

They do not group **marks**. A second `data()` applies to the mark written
directly after it ([Data](data.qmd)). Parentheses around several marks do not
extend that table to all of them:

```{r}
#| echo: false
peek(actuals)
peek(forecast)
```

```{r}
#| error: true
data(actuals) + x(year) + y(sales) +
  line +
  (data(forecast) + point + area)
```

gog refuses that instead of drawing it. The refusal is better than the
alternative. The engine cannot give the second table to both marks inside the
parentheses. Its only other choice is to discard both marks and draw a plot with
two layers missing, without saying so. To put two marks on the second table,
write `data()` before each of them.

The same refusal covers a position or a title inside the parentheses, for the
same reason. In JavaScript, where the operators are words, the same mistake is
a `plot()` handed to another `plot()`, and it is refused there too.

---

## Why `+`, and not a pipe

Readers who know R's tidyverse packages ask why gog does not use a pipe. A pipe
passes a value into a function: `a |> f` means `f(a)`. Hadley Wickham,
ggplot2's author, has said publicly that the pipe would have been the better
interface [@wickham2018pipe]. ggplot, ggplot2's predecessor, was written as
function composition, and it could have used a pipe if one had existed. ggplot2
used `+` instead.

He also answers the question a reader asks next: why not change ggplot2 now?
It cannot change: its functions take the mapping and the data first, while a
pipe needs the plot as the first argument. And he judges that a new package
built only to make this change would cost more work than the change would be
worth. So ggplot2's `+` comes from its history, not a fresh choice, and a
grammar designed today has to make that choice for itself. He gives two reasons
for preferring the pipe, and neither applies to gog. A third consideration is
gog's rather than his, and it is the one that decides the question.

His first reason is about one language. A pipe is a single idea that works
everywhere in R, so a grammar that spells function composition differently is
one more thing to learn. Switching between the two spellings is a frequent
source of errors, his own included. That is true, and it is an argument about R
and its packages. gog is one syntax written in four languages, and that changes
the answer. `+` can be given a meaning in R, Python and Julia. The pipe cannot
be given a gog meaning. Julia's `|>` and R's two pipes already mean something
gog must not change. Python has no pipe operator at all, and no way to add one:
you can give new meanings only to the operators it already has. In R and Julia
a pipe would work with the meaning it already has. Python would need another
spelling, and the languages would then have two different grammars. This
project exists to avoid that. JavaScript decides nothing here, because it cannot
overload any operator at all, which is why it spells all four as words
([JavaScript](bindings/javascript.qmd)).

The second is about arithmetic, and it is the stronger of the two. You expect
`a + b` to equal `b + a`, and ggplot2's `+` does not always do that. That
objection is answered by the six permutations drawn earlier in this chapter.
Here the mark and the channels commute. Move `color` ahead of `x`, or write the
positions before the mark, and the engine produces the identical file. One
thing does depend on order: two marks swapped. There the order is the drawing
order you asked for, and the second mark draws on top.

The other half of that objection is grouping, whether `a + (b + c)` means what
`(a + b) + c` means. The question cannot arise here, because `a + (b + c)`
cannot be written at all. `+` always takes a plot on its left, so two atoms with
no table named before them never form a smaller expression to group. They are
refused:

```{r}
#| error: true
point + x(gdp)
```

The third point is gog's own, and it is what decides the question. It is not an
argument against the pipe. It is a statement of what a pipe would remove. A
pipe applies functions, so every atom would have to become a function taking a
plot and returning one. Make `bar` a function, then ask what `bar * bin` could
mean. There is no answer: you cannot multiply two functions and get a histogram.
The derive operator exists only because marks and transforms are *things*
rather than actions. They are values, and they
combine with each other before either one is joined to a plot.

This is the decision the book opens with: a plot is a specification, not drawing
code. ㄱ is a letter, not an instruction. You put it in a slot next to other
letters, and the block is assembled from its pieces rather than performed in
sequence. Hangeul's atoms are nouns, which is exactly why they combine, and
gog's are nouns for the same reason. A grammar built on the pipe is a grammar of
verbs: it can layer, because layering really is a sequence of additions, but it
cannot derive. ggplot, ggplot2's predecessor, was written as function
composition, and it had no `*` either.

So `+` here is a choice rather than something inherited. Unlike the pipe, it can
be given a meaning in R, Python and Julia. It commutes wherever order should not
matter, and it leaves the atoms as nouns so that `*` can exist.

---

## Summary

Each operator reads as a phrase when a plot is read aloud, and `|` and `/` have
two readings each:

| Operator | Reads as | Example |
|----------|----------|---------|
| `+` | "…and also…" | `point + line` |
| `*` | "…derived by…" | `bar * bin` |
| <code>&#124;</code> | "…split into panel columns by…" | <code>(data(gm) + point + x(gdp) + y(life)) &#124; facet(continent)</code> |
| `/` | "…split into panel rows by…" | `(data(gm) + point + x(gdp) + y(life)) / facet(era)` |
| <code>&#124;</code> | "…beside…" | <code>(one specification) &#124; (another)</code> |
| `/` | "…above…" | `(one specification) / (another)` |

A finite set of rules, and more plots than you will ever need to draw.

## What comes next

The operators say how atoms combine into a sentence. Where you write each atom
decides what it binds to, and [Encoding scope](encoding-scope.qmd) gives that
rule in full.
