# Operators

```{r}
#| include: false
source("R/setup.R")
```

How do you put two things in one picture? gog has exactly **four operators**.
They are not atoms. They are the assembly rules: how atoms combine.

Think of them the way Hangeul (한글) works: consonants and vowels are the atoms;
the rule "initial + vowel + optional final = syllable block" is the operator.
The rule never changes, and a small set of letters builds every syllable the
language needs.

```
+   compose   (add a layer, a channel, or a label to a specification)
*   derive    (combine a mark with a transform into a compound mark)
|   facet ↔   (small multiples, arranged in columns)
/   facet ↕   (small multiples, arranged in rows)
```

`|` and `/` also arrange whole plots on a page. That second use has a section
of its own.

---

## `+`: compose

Every sentence you have written so far holds several atoms, and one operator
joins them. `+` is the general assembly operator, and every atom enters a
specification through it:

```r
data(my_table) + point + x(height) + y(weight) + color(species) + title("My plot")
```

Adding a **mark** (`point`, `line`, `bar`, …) starts a new layer.
Adding a **channel** (`x()`, `color()`, …) or a **label atom** (`title()`,
`x_label()`) attaches to the appropriate scope, which is the plot or the nearest
preceding mark (see [Encoding scope](encoding-scope.qmd)).

### Layering with `+`

Video editing works the same way, and the comparison is worth making. A video is
a stack of tracks: the picture, the sound, the captions. Each track is separate,
they play at the same time, and you see one finished video. A plot layers the
same way.

Multiple marks in the same expression produce multiple layers, rendered in
order:

```{r}
#| echo: false
peek(gapminder_2007)
peek(life_bands)
```

```{r}
# Scatter plot: points, and a horizontal reference line at each band
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  data(life_bands) + rule + style(color = "darkgray", pattern = "dashed") +
  title("GDP vs Life Expectancy, 2007")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also rules from the life bands."*

Layers draw in the order they are written: the first mark sits at the
bottom. Add a `line` after `point` to draw the line on top of the points.

### The order of `+`: data, mark, positions, refinements

Every sentence in this book is written in one order, and you may wonder
whether the engine requires it. Inside a single layer the channels **commute**,
and the mark commutes among them. Write the mark, `x` and `color` in any order
and the plot is the same. Three things can be arranged six ways, so here are
all six.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
((data(iris_flowers) + bar * bin + x(petal_length) + color(species)) |
   (data(iris_flowers) + bar * bin + color(species) + x(petal_length))) /
((data(iris_flowers) + x(petal_length) + bar * bin + color(species)) |
   (data(iris_flowers) + x(petal_length) + color(species) + bar * bin)) /
((data(iris_flowers) + color(species) + bar * bin + x(petal_length)) |
   (data(iris_flowers) + color(species) + x(petal_length) + bar * bin))
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."*

Six sentences, and the same histogram six times. These are not six plots that
merely look alike. The engine produces one identical file for each of the six.

So if the order does not change the picture, why does every plot in this book
follow the same one: *data, mark, positions, refinements*? Because a fixed order
lets a plot read like a sentence, and the sentence has a natural, subject-first
shape:

- **`data(...)` first.** It names the table the columns come from. A bare name
  like `petal_length` means nothing until a table is bound (Bind-Once), so the
  table is named first. The sentence must start here.
- **The mark and its `*`-chain second.** This is the subject. A visual is a mark
  before it is anything else (Minimum Syllable): `bar * mean * dodge` says *what
  you are drawing* and *what was done to it*, as one unbroken word. It is the
  initial consonant of the block; everything after is said about it.
- **The positions third.** `x` and `y` say *where* the subject stands, and a
  mark cannot render without them (the vowel to the mark's consonant), so they
  follow immediately.
- **The refinements last.** `color`, `size`, `order`, labels, `title`:
  everything optional. Delete every refinement and a legal plot remains; delete
  the mark or a position and nothing renders. Writing the removable parts last
  is also how you *build* a plot: the minimum sentence first, then refine
  ([First plot](first-plot.qmd)).

The order is a reading convention, not a grammar rule: the engine accepts any
permutation that scopes the same way. Where position genuinely *does* matter (a
channel written before or after a mark across two or more layers) is the subject
of [Encoding scope](encoding-scope.qmd). One order everywhere is worth learning:
every plot in the book, and every plot you write, then reads the same way.

[Which orders matter](#which-orders-matter) sets out what the engine enforces
under `+` and `*`, and what you are free to arrange.

#### A refinement far from what it refines

One part of the convention is easy to misread. A grouped bar chart pairs
`dodge` with `color`, yet the sentence puts them at opposite ends, with the
positions in between:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

*"Given the gapminder eras: bars derived by mean and dodge, x is continent, y is
life, color by era."*

`dodge` and `color(era)` are plainly related: `dodge` has nothing to set side by
side until a `color` (or `group`) names the groups. So why not write them
together? Because the grammar deliberately gives them different operators.
`dodge` belongs to `*`, because it modifies the mark. The same modifier works on
`box * dodge` and `interval * range * dodge`, and neither of those involves
color. `color` belongs to `+`: it binds a column to a channel. `color * dodge`
cannot be written: `*` needs a mark on its left. Joining the two would make
`dodge` depend on which channel splits the groups, and a modifier must not
depend on that.

The order is not a compromise. It follows from what `dodge` does. `dodge`
subdivides the **x** slot: each continent's space is cut into one sub-bar per
era. So it needs both of them: `x(continent)` says
*which slot* to divide, `color(era)` says *into how many pieces*. The sentence
lists them in the order `dodge` uses them: the axis it splits along, then the
groups it splits into. The positions do not separate `dodge` from what it needs;
one of them is the slot `dodge` divides.

You also never have to read back through the sentence to check that the split is
there. `bar * mean * dodge` with no `color` or `group` does not draw a lone bar;
it is refused, and the message points at the missing channel. The legality check
requires both, so writing `color` far from `dodge` is safe. The two are bound
whether or not they sit side by side on the page.

---

## `*`: derive

A mark on its own draws the rows as they are. A histogram counts them first,
and a trend line fits a curve to them first. That computing step needs a word
of its own. `*` joins a transform to a mark, and the result is one
**compound mark**.

```r
bar * bin     # histogram
line * smooth # smoothed trend line
line * density # kernel density curve
```

`*` has one rule and one freedom. The mark goes on the left and the transforms
go on the right. Among the transforms themselves the order you write does not
matter, and any order draws the same plot. The rest of this section explains
both.

### Why the order is fixed

In Hangeul, a syllable is always *initial consonant + vowel*, never the reverse.
gog's `*` works the same way: the mark is the subject; the transform is
what is done to it.

`bin * bar` is refused, just as ㅏ (*a*) + ㄱ (*g*) is not a valid Hangeul
syllable.

```r
bar * bin   # a bar mark, derived by binning
```

Written the other way round, the engine refuses the sentence and names the fix:

```{r}
#| error: true
bin * bar
```

### The one fixed position under `*`

The rule that matters is the one just stated: the mark is on the left, always,
and no transform may stand there. That is what makes `bar * bin` a compound mark
rather than a function call, and it is why the atoms can be nouns.

Write the transforms in the order they happen. This is a reading convention: any
order draws the same plot, and this one lets you follow the steps as you read.
`bar * bin * mean` cuts the axis into bands, then averages inside each one. That
is the order it runs in, and there is no other order it could run in.

The reason is what `mean` needs. A statistic reduces a **group** to one number,
and a continuous column has no groups: every value stands alone, so its mean is
itself. `bin` is what makes the groups, by turning many distinct numbers into
one band. So the cut comes first because otherwise the mean has nothing to
average.

So two transforms work together only when they do different jobs. One makes the
groups, and the other measures inside them or rescales the result. A pair that
answers the same question twice is refused rather than sequenced. `bin` and
`count` are both tallies with groups of their own, so they cannot both be in one
layer:

```{r}
#| error: true
data(gapminder_2007) + bar * bin(12) * count + x(life)
```

The engine decides the running order from the transforms themselves, not from
where you typed them. The order you type can never be the wrong order, although
the pair itself can still be refused, as the chunk above shows. Write the
transforms in the order that reads best to you, and the sentence will read in
the order it runs. [Chaining transforms](transforms.qmd#chaining-transforms)
lists the legal transform pairs, one pair at a time.

### Live examples

The three compound marks named at the top of this section appear in the
sentences you will write most often. Here each one is drawn, so you can see
what the transform did to the rows before the mark drew them.

```{r}
# Histogram: life expectancy across all countries in 2007
data(gapminder_2007) + bar * bin + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Distribution of Life Expectancy, 2007")
```

```{r}
# Scatter + LOESS smooth layer on top
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  title("GDP vs Life Expectancy with LOESS Smooth")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

```{r}
# Kernel density estimate
data(gapminder_2007) + line * density + x(life) + y(density) +
  x_label("Life expectancy (years)") + y_label("Density") +
  title("Kernel Density: Life Expectancy 2007")
```

*"Given gapminder 2007: a line derived by density, x is life, y is density."*

---

## `|` and `/`: facet, or compose

Two questions lead to the last two operators: how the same plot looks for each
group, and how two different plots fit on one page. `|` puts things side by
side and `/` puts one above the other. What is written on the right decides
which. A `facet()` splits one plot into panels, and a second plot is placed
beside or below the first.

With `facet()`, `|` produces small multiples arranged in columns, one panel per
category of the column it names.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, split into panel columns by
continent."*

`/` produces small multiples stacked in rows:

```{r}
gm_two <- gapminder_2007[gapminder_2007$continent %in% c("Africa", "Europe"), ]
data(gm_two) + point + x(gdp) + y(life) / facet(continent)
```

*"Given the gm two table: points, x is gdp, y is life, split into panel rows by
continent."*

Write both to combine them into a grid. Each operator applies to the facet
written after it, read from left to right. See [Faceting](faceting.qmd) for the
grid, what the panels share, and why the facet column must be a category.

With a plot on the right, the same two operators arrange separate plots on one
page. Each plot keeps its own coordinate space. Two plots side by side share
the vertical axis when they name the same column on it. Two plots one above the
other share the horizontal axis the same way:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

One pair of operators, two meanings, and what you write on the right decides
which one you get. See [Composition](composition.qmd).

---

## Which orders matter

This book writes every sentence the same way: data, mark, positions,
refinements. That is a preference rather than a rule, and the engine enforces
only part of it. The difference is worth knowing, because it tells you where to
be careful. There are three kinds of ordering, and only one of them needs your
attention.

| Kind | What it means | Under `+` | Under `*` |
|---|---|---|---|
| **Free** | Either way draws exactly the same picture. | Inside one layer, everything: the mark, its positions and its refinements, in any arrangement. | Two transforms after one mark. Write them in the order they happen; the engine works that order out either way. |
| **Refused** | The engine stops and names what to move. | Atoms with no table behind them: `point + x(gdp)`. A `style()` written before any mark. | A transform on the left: `bin * bar`. |
| **Meaningful** | No error, and a different picture. | A channel before the marks, or after one of them, once a plot has two. Two marks swapped: the second draws on top. | Nothing. No order under `*` changes the picture. |

Start with the last cell, `*` under Meaningful. `*` fixes one thing only, which
is that the mark stands on its left. Once the mark is there, no arrangement of
what follows changes what you get.

**Free** is most of it. Every one of these draws the identical plot, and the
book writes the first only because a fixed order is easier to read:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + x(gdp) + point + y(life))
```

Both name `life` on `y`, so the two panels share that axis and it is drawn once,
on the left. Each panel still draws its own `gdp` axis underneath it.

**Refused** is the engine telling you the sentence has no subject yet. A setting
needs a mark to attach to, and a plot needs its table first. `+` has to have a
plot on its left before it can add anything:

```{r}
#| error: true
data(gapminder_2007) + style(size = 3) + point + x(gdp) + y(life)
```

**Meaningful** is the only kind that needs care, because it is the only kind
the engine says nothing about. With two marks in a sentence, a channel written
before them belongs to the plot and reaches both, and a channel written after
one belongs to that mark. Both are legal, neither warns, and they draw
different pictures. That is the subject of [Encoding
scope](encoding-scope.qmd). It is also why this book keeps one order: when
every sentence looks the same, a different arrangement is a deliberate signal.

The marks themselves are the other case the engine does not report. Two marks
swapped draw in the order you wrote them, so the second sits on top of the
first.

### What groups first

Those three kinds are about where you may put a word. Precedence is a
different question: how the words you did write get grouped. It is what lets a
sentence like this one carry no parentheses at all.

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

`*` and `/` group first, then `+`, then `|`. So the transform joins its mark
first, the atoms join the plot next, and the facet splits the finished plot
last. `/` sits in the tighter tier with `*`, which is why a row facet reads as
one piece with the atom beside it. [What parentheses may
group](#what-parentheses-may-group) has the rule.

Those are R's own precedences, and Python's. Julia puts `|` in the addition
tier, beside `+` rather than below it. A sentence that writes `+` after a `|`
therefore groups differently there, which is why a composed page is
parenthesized. [Julia](bindings/julia.qmd) measures the four shapes side by
side. JavaScript can give no operator a meaning of its own, so it spells the
four as words, which makes the grouping something you see instead of something
you remember ([JavaScript](bindings/javascript.qmd)).

### What parentheses may group

Parentheses change which operator runs first, and that is the whole of what they
do. They group **plots**, which is why a composed page uses them. `/` is
applied before `|`, so `a | b / c` reads as `a | (b / c)` unless you say
otherwise ([Composition](composition.qmd)).

They do not group **marks**. A second `data()` applies to the mark written
directly after it ([Data](data.qmd)). Parentheses around several marks do not
extend that table to all of them:

```{r}
#| echo: false
peek(actuals)
peek(forecast)
```

```{r}
#| error: true
data(actuals) + x(year) + y(sales) +
  line +
  (data(forecast) + point + area)
```

gog refuses that instead of drawing it. The refusal is better than the
alternative. The engine cannot give both marks inside the parentheses the second
table. Its only other choice is to discard both marks and draw a plot with two
layers missing, without saying so. To put two marks on the
second table, write `data()` before each of them.

The same refusal covers a position or a title inside the parentheses, for the
same reason. In JavaScript, where the operators are words, the same mistake is
a `plot()` handed to another `plot()`, and it is refused there too.

---

## Why `+`, and not a pipe

Readers who know R's tidyverse packages ask why gog does not use a pipe. Hadley
Wickham, ggplot2's author, has said publicly that the pipe would have been the
better interface [@wickham2018pipe]. ggplot, ggplot2's predecessor, was written
as function composition, and it would have used a pipe if one had existed.
ggplot2 used `+` instead.

He also answers the question a reader asks next. ggplot2 cannot change now: its
functions take the mapping and the data first, while a pipe needs the plot. And
he judges that a new package built only to make this change would cost more than
it returns. For ggplot2 the answer therefore comes from its history rather than
from a fresh choice, and a grammar designed today has to answer it from the
beginning. He gives two reasons for preferring the pipe, and neither applies to
gog. A third consideration settles it the other way, and that one is gog's
rather than his.

The first is about one language. A pipe is a single idea that works everywhere
in R, so a grammar spelling composition differently is one more thing to learn.
Switching between the two spellings is a frequent source of errors, his own
included. That is true, and it is an argument about R and its packages. gog is
one syntax written in four languages, and that changes the answer. `+` can be
given a meaning in R, Python and Julia. The pipe cannot be given a gog meaning.
Julia's `|>` and R's two pipes already mean something gog must not change.
Python has no pipe operator at all, and no way to add one: you can give new
meanings only to the operators it already has. Piping would work in R and Julia
and need some other spelling in Python. The languages would then have two
different grammars, and this project exists to avoid that. JavaScript decides
nothing here, because it cannot overload any operator at all, which is why it
spells all four as words ([JavaScript](bindings/javascript.qmd)).

The second is about arithmetic, and it is the strongest. You expect `x + y` to
equal `y + x`, and ggplot2's `+` does not always do that. That objection is
answered by the six permutations drawn earlier in this chapter. Here the
channels commute, and so does the mark among them. Move `color` ahead of `x`,
or write the positions before the mark, and the engine produces the identical
file. One thing does depend on order: two marks swapped. There the order is the
drawing order you asked for, and the second mark draws on top.

The other half of that objection is grouping, whether `x + (y + z)` means what
`(x + y) + z` means. The question cannot arise here, because `x + (y + z)`
cannot be written at all. `+` always takes a plot on its left, so two atoms with
no table behind them never form a smaller expression to group. They are refused:

```{r}
#| error: true
point + x(gdp)
```

Atoms never collect on their own, so there is no grouping to get wrong.

The third point is gog's own, and it is what decides the question. It is not an
argument against the pipe. It is a statement of what a pipe would remove. A
pipe applies functions: `x |> f` means `f(x)`, so every atom would have to
become a function taking a plot and returning one. Make `bar` a function, then
ask what `bar * bin` could mean. It could mean nothing: you cannot multiply two
functions and get a histogram. The derive operator exists only because marks
and transforms are *things* rather than actions. They are values, and they
combine with each other before either one is joined to a plot.

That is the same decision this book opens with, arriving a second time. ㄱ is a
letter, not an instruction. You put it in a slot next to other letters, and the
block is assembled from its pieces rather than performed in sequence. Hangeul's
atoms are nouns, which is exactly why they compose, and gog's are nouns for the
same reason. A grammar built on the pipe is a grammar of verbs: it can layer,
because layering really is a sequence of additions, but it cannot derive.
Wickham's own pre-ggplot2 experiment was function composition, and it had no `*`
either.

So `+` here is a choice rather than something inherited. It is the one operator
that can be given a meaning in R, Python and Julia. It commutes wherever order
should not matter, and it leaves the atoms as nouns so that `*` can exist.

---

## Summary

| Operator | Reads as | Example |
|----------|----------|---------|
| `+` | "…and also…" | `point + color(continent)` |
| `*` | "…derived by…" | `bar * bin` |
| <code>&#124;</code> | "…split into panel columns by…" | <code>(data(gm) + point + x(gdp) + y(life)) &#124; facet(continent)</code> |
| `/` | "…split into panel rows by…" | `(data(gm) + point + x(gdp) + y(life)) / facet(era)` |
| <code>&#124;</code> | "…beside…" | <code>(one specification) &#124; (another)</code> |
| `/` | "…above…" | `(one specification) / (another)` |

A finite set of rules, and more plots than you will ever need to draw.

## What comes next

The operators say how sentences combine, and not which layer a channel belongs
to once there are two. [Encoding scope](encoding-scope.qmd) gives the rule:
where you write an atom decides what it binds to.
