# Operators

```{r}
#| include: false
source("R/setup.R")
```

How do you put two things in one picture? gog has exactly **four operators**.
They are not atoms. They are the assembly rules: how atoms combine.

Think of them the way Hangeul (한글) works: consonants and vowels are the atoms;
the rule "initial + vowel + optional final = syllable block" is the operator.
The rule never changes, and the number of things you can build with it has no
limit.

```
+   compose   (add a layer, a channel, or a label to a specification)
*   derive    (combine a mark with a transform into a compound mark)
|   facet ↔   (small multiples, arranged in columns)
/   facet ↕   (small multiples, arranged in rows)
```

`|` and `/` also arrange whole plots on a page, and that second use has its own
section below.

---

## `+`: compose

Every sentence you have written so far holds several atoms, and one operator
joins them. `+` is the general assembly operator.
Every atom enters a specification via `+`:

```r
data(df) + point + x(col_a) + y(col_b) + color(group) + title("My plot")
```

Adding a **mark** (`point`, `line`, `bar`, …) starts a new layer.
Adding a **channel** (`x()`, `color()`, …) or **label** (`title()`) attaches
to the appropriate scope (plot-level or the nearest preceding mark, see
[Encoding scope](encoding-scope.qmd)).

### Layering with `+`

If you have edited video, you already know this idea. A video is a stack of
tracks: the picture, the sound, the captions, the logo, the effects. Each track
is separate, they play at the same time, and you see one finished video. A plot
layers the same way.

Multiple marks in the same expression produce multiple layers, rendered in
order:

```{r}
#| echo: false
peek(gapminder_2007)
peek(life_bands)
```

```{r}
# Scatter plot: points, and a horizontal reference line at each band
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  data(life_bands) + rule + style(color = "darkgray", pattern = "dashed") +
  title("GDP vs Life Expectancy, 2007")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also rules from the life bands."*

Layers draw in the order they are written: the first mark sits at the
bottom. Add a `line` after `point` to draw the line on top of the points.

### The order of `+`: data, mark, positions, refinements

Every sentence in this book is written in one order, and you may wonder
whether the engine requires it. Inside a single layer the channels **commute**,
and the mark commutes among them. Write the mark, `x` and `color` in any order
and the plot is the same. Three things can be arranged six ways, so here are
all six.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
((data(iris_flowers) + bar * bin + x(petal_length) + color(species)) |
   (data(iris_flowers) + bar * bin + color(species) + x(petal_length))) /
((data(iris_flowers) + x(petal_length) + bar * bin + color(species)) |
   (data(iris_flowers) + x(petal_length) + color(species) + bar * bin)) /
((data(iris_flowers) + color(species) + bar * bin + x(petal_length)) |
   (data(iris_flowers) + color(species) + x(petal_length) + bar * bin))
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."*

Six sentences, and the same histogram six times. Not six plots that look alike:
the engine emits the same bytes for every one of them.

So if the order does not change the picture, why does every plot in this book
follow the same one: *data, mark, positions, refinements*? Because a fixed order
lets a plot read like a sentence, and the sentence has a natural, subject-first
shape:

- **`data(...)` first.** It names the table the columns come from. A bare name
  like `petal_length` means nothing until a table is bound (Bind-Once), so the
  table is named first. The sentence must start here.
- **The mark and its `*`-chain second.** This is the subject. A visual is a mark
  before it is anything else (Minimum Syllable): `bar * mean * dodge` says *what
  you are drawing* and *what was done to it*, as one unbroken word. It is the
  initial consonant of the block; everything after is said about it.
- **The positions third.** `x` and `y` say *where* the subject stands, and a
  mark cannot render without them (the vowel to the mark's consonant), so they
  follow immediately.
- **The refinements last.** `color`, `size`, `order`, labels, `title`:
  everything optional. Delete every refinement and a legal plot remains; delete
  the mark or a position and nothing renders. Writing the removable parts last
  is also how you *build* a plot: the minimum sentence first, then refine
  ([First plot](first-plot.qmd)).

The order is a reading convention, not a grammar rule: the engine accepts any
permutation that scopes the same way. Where position genuinely *does* matter (a
channel written before or after a mark across two or more layers) is the subject
of [Encoding scope](encoding-scope.qmd). One order everywhere is worth learning:
every plot in the book, and every plot you write, then reads the same way.

[Which orders matter](#which-orders-matter) below sets out what the engine
enforces under `+` and `*`, and what you are free to arrange.

#### A refinement far from what it refines

One part of the convention is easy to misread. A grouped bar chart pairs
`dodge` with `color`, yet the sentence puts them at opposite ends, with the
positions in between:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

*"Given the gapminder eras: bars derived by mean and dodge, x is continent, y is
life, color by era."*

`dodge` and `color(era)` are plainly related: `dodge` has nothing to set side by
side until a `color` (or `group`) names the groups. So why not write them
together? Because the grammar deliberately gives them different operators.
`dodge` belongs to `*`, because it modifies the mark. The same modifier works on
`box * dodge` and `interval * range * dodge`, and neither of those involves
color. `color` belongs to `+`: it binds a column to a channel. `color * dodge`
cannot be written: `*` needs a mark on its left. Joining the two would make
`dodge` depend on which channel splits the groups, and a modifier must not
depend on that.

And the order is not a compromise; it is exactly right, once you see what
`dodge` does. `dodge` subdivides the **x** slot: each continent's space is cut
into one sub-bar per era. So it needs *both* neighbors: `x(continent)` says
*which slot* to divide, `color(era)` says *into how many pieces*. The sentence
lists them in the order `dodge` uses them: the axis it splits along, then the
groups it splits into. The positions do not separate `dodge` from what it needs;
one of them is the slot `dodge` divides.

You also never have to scan the line to check the split is there.
`bar * mean * dodge` with no `color` or `group` does not draw a lone bar; it is
refused, pointing you at the missing channel. The legality check requires both,
so writing `color` far from `dodge` is safe: the two are bound whether or not
they sit side by side on the page.

---

## `*`: derive

A mark on its own draws the rows as they are. A histogram counts them first,
and a trend line fits a curve to them first. That computing step needs a word
of its own. `*` joins a transform to a mark, and the result is one
**compound mark**.

```r
bar * bin     # histogram
line * smooth # smoothed trend line
line * density # kernel density curve
```

Two rules govern `*`, and only two. The mark goes on the left, and the
transforms go on the right. Then, among the transforms themselves, the order
you write them in does not matter: any order draws the same plot. The rest of
this section explains each rule.

### Why the order is fixed

In Hangeul, a syllable is always *initial consonant + vowel*, never the reverse.
gog's `*` works the same way: the mark is the subject; the transform is
what is done to it.

`bin * bar` is refused, just as ㅏ (*a*) + ㄱ (*g*) is not a valid Hangeul
syllable.

```r
bar * bin   # ✅  "a bar mark, derived by binning"
bin * bar   # ❌  Refused: `*` is not defined for transform * mark
```

### What `*` is rigid about

The rule that matters is the one above: the mark is on the left, always,
and no transform may stand there. That is what makes `bar * bin` a compound mark
rather than a function call, and it is why the atoms can be nouns.

Write the transforms in the order they happen. This is a reading convention: any
order draws the same plot, and this one lets you follow the steps as you read.
`bar * bin * mean` cuts the axis into bands, then averages inside each one. That
is the order it runs in, and there is no other order it could run in.

The reason is what `mean` needs. A statistic reduces a **group** to one number,
and a continuous column has no groups: every value stands alone, so its mean is
itself. `bin` is what makes the groups, by turning many distinct numbers into
one band. So the cut comes first because otherwise the mean has nothing to
average.

So two transforms work together only when they do different jobs. One makes the
groups, and the other measures inside them or rescales the result. A pair that
answers the same question twice is refused rather than sequenced, so `bin` and
`count` (both of them tallies with groups of their own) cannot both be in one
layer:

```{r}
#| error: true
data(gapminder_2007) + bar * bin(12) * count + x(life)
```

The engine decides the running order from the transforms themselves, not from
where you typed them, so you cannot get it wrong. Write it the way you mean it,
and the sentence will read in the order it runs. [Chaining
transforms](transforms.qmd#chaining-transforms) goes through the legal transform
pairs one by one.

### Live examples

The three compound marks named above appear in the sentences you will write
most often. Here each one is drawn, so you can see what the transform did to
the rows before the mark drew them.

```{r}
# Histogram: life expectancy across all countries in 2007
data(gapminder_2007) + bar * bin + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Distribution of Life Expectancy, 2007")
```

```{r}
# Scatter + LOESS smooth layer on top
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  title("GDP vs Life Expectancy with LOESS Smooth")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and
also a line derived by smooth."*

```{r}
# Kernel density estimate
data(gapminder_2007) + line * density + x(life) + y(density) +
  x_label("Life expectancy (years)") + y_label("Density") +
  title("Kernel Density: Life Expectancy 2007")
```

*"Given gapminder 2007: a line derived by density, x is life, y is density."*

---

## `|` and `/`: facet, or compose

Two questions lead to the last two operators: how the same plot looks for each
group, and how two different plots fit on one page. `|` puts things side by
side and `/` puts one above the other. What is written on the right decides
which. A `facet()` splits one plot into panels; a second
plot is placed beside or below the first.

With `facet()`, `|` produces small multiples arranged in columns, one panel per
category of the column it names.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, split into panel columns by
continent."*

`/` produces small multiples stacked in rows:

```{r}
gm_two <- gapminder_2007[gapminder_2007$continent %in% c("Africa", "Europe"), ]
data(gm_two) + point + x(gdp) + y(life) / facet(continent)
```

*"Given the gm two table: points, x is gdp, y is life, split into panel rows by
continent."*

Write both to cross them into a grid: each operator applies to the facet
written after it, read left to right. See [Faceting](faceting.qmd) for the
grid, what the panels share, and why the facet column must be a category.

With a plot on the right, the same two operators arrange separate plots on one
page. Each plot keeps its own coordinate space. Two plots side by side share the
vertical axis when they name the same column on it, and two plots one above the
other share the horizontal axis the same way:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

*"Given gapminder 2007: points, x is gdp, y is life, beside bars derived by
count, x is continent."*

One pair of operators, two meanings, and what you write on the right decides
which one you get. See [Composition](composition.qmd).

---

## Which orders matter

This book writes every sentence the same way: data, mark, positions,
refinements. That is a preference rather than a rule, and the engine enforces
only part of it. The difference is worth knowing, because it tells you where to
be careful. There are three kinds of ordering, and only one of them needs your
attention.

| Kind | What it means | Under `+` | Under `*` |
|---|---|---|---|
| **Free** | Either way draws exactly the same picture. | The mark and its positions, in any arrangement. `x` before `y`, or `y` before `x`. Refinements among themselves. | Two transforms after one mark. Write them in the order they happen; the engine works that order out either way. |
| **Refused** | The engine stops and names what to move. | A sentence that does not begin with `data()`. A `style()` written before any mark. | A transform on the left: `bin * bar`. |
| **Meaningful** | No error, and a different picture. | A channel before the marks, or after one of them, once a plot has two. Two marks swapped: the second draws on top. | Nothing. No order under `*` changes the picture. |

Read the bottom right cell first. `*` is rigid about one thing only, which is
that the mark stands on its left. Once the mark is there, no arrangement of what
follows changes what you get.

**Free** is most of it. Every one of these draws the identical plot, and the
book writes the first only because a fixed order is easier to read:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + x(gdp) + point + y(life))
```

Both name `life` on `y`, so the two panels share that axis and it is drawn once,
on the left. Each panel still draws its own `gdp` axis underneath it.

**Refused** is the engine telling you the sentence has no subject yet. A setting
needs a mark to attach to, and a plot needs its table first. `+` has to have a
plot on its left before it can add anything:

```{r}
#| error: true
data(gapminder_2007) + style(size = 3) + point + x(gdp) + y(life)
```

**Meaningful** is the only one to watch, because it is the only one that stays
silent. With two marks in a sentence, a channel written before them belongs to
the plot and reaches both, and a channel written after one belongs to that mark.
Both are legal, neither warns, and they draw different pictures. That is the
subject of [Encoding scope](encoding-scope.qmd). It is also why this book keeps
one order: when every sentence looks the same, a different arrangement is a
deliberate signal.

The other silent one is the marks themselves. Two marks swapped draw in the
order you wrote them, so the second sits on top of the first.

### What groups first

The three kinds above are about where you may put a word. Precedence is a
different question: how the words you did write get grouped. It is what lets a
sentence like this one carry no parentheses at all.

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

`*` groups first, then `+`, then `|`. So the transform joins its mark first, the
atoms join the plot next, and the facet splits the finished plot last. `/`
groups before `|`; the next section has the rule.

Those are R's own precedences, and Python's and Julia's, so one sentence groups
the same way in three of the four languages. JavaScript can give no operator a
meaning of its own, so it spells the four as words, which makes the grouping
something you see instead of something you remember
([JavaScript](bindings/javascript.qmd)).

### What parentheses may group

Parentheses change which operator runs first, and that is the whole of what they
do. They group **plots**, which is why a composed page uses them: `/` is
applied before `|`, so `a | b / c` reads as `a | (b / c)` unless you say
otherwise ([Composition](composition.qmd)).

They do not group **marks**. A second `data()` applies to the mark written
directly after it ([Data](data.qmd)). Parentheses around several marks do not
extend that table to all of them:

```{r}
#| echo: false
peek(actuals)
peek(forecast)
```

```{r}
#| error: true
data(actuals) + x(year) + y(sales) +
  line +
  (data(forecast) + point + area)
```

gog refuses that instead of drawing it. The refusal is worth having, because the
alternative is worse than an error. The engine cannot give both marks inside the
parentheses the second table. Its only other choice would be to discard both
marks and draw a plot missing two layers, in silence. To put two marks on the
second table, write `data()` before each of them.

The same refusal covers a position or a title inside the parentheses, for the
same reason. In JavaScript, where the operators are words, the same mistake is
a `plot()` handed to another `plot()`, and it is refused there too.

---

## Why `+`, and not a pipe

Readers who know R's tidyverse packages ask why gog does not use a pipe. Hadley
Wickham, ggplot2's author, has said publicly that the pipe would have been the
better interface [@wickham2018pipe]. ggplot, ggplot2's predecessor, was written
as function composition, and it would have used a pipe if one had existed.
ggplot2 used `+` instead.

He also answers the question a reader asks next. ggplot2 cannot change now: its
functions take the mapping and the data first, while a pipe needs the plot. And
he judges that a new package built only to change this would not be worth what
it would cost. So for ggplot2 the matter is settled by history rather than by
taste, and a grammar starting fresh has to answer it on its own. Three reasons
are given for preferring the pipe. Two of them do not apply to gog, and the
third is the one that settles the question the other way.

The first is about one language. A pipe is a single idea that works everywhere
in R, so a grammar spelling composition differently is one more thing to learn,
and switching between the two spellings is a frequent source of errors (his own
included). That is true, and it is an argument about R and its packages. gog is
one syntax spoken by four languages, and that changes the answer. `+` can be
given a meaning in R, Python and Julia. The pipe cannot be given a gog meaning.
Julia's `|>` and R's two pipes already mean something gog must not change.
Python has no pipe operator at all, and no way to add one: you can give new
meanings only to the operators it already has. Piping would work in R and Julia
and need some other spelling in Python. The languages would then have two
different grammars, and this project exists to avoid that. JavaScript decides
nothing here, because it cannot overload any operator at all, which is why it
spells all four as words ([JavaScript](bindings/javascript.qmd)).

The second is about arithmetic, and it is the strongest. You expect `x + y` to
equal `y + x`, and ggplot2's `+` does not always do that. That objection is
answered a few sections up. Here the channels commute, and so does the mark
among them: move `color` ahead of `x`, or write the positions before the mark,
and the engine emits the same bytes. One thing does depend on order: two marks
swapped. There the order is the drawing order you asked for, and the second mark
draws on top.

The other half of that objection is grouping, whether `x + (y + z)` means what
`(x + y) + z` means. The question cannot arise here, because `x + (y + z)`
cannot be written at all. `+` always takes a plot on its left, so two atoms with
no table behind them never form a smaller expression to group. They are refused:

```{r}
#| error: true
point + x(gdp)
```

Atoms never collect on their own, so there is no grouping to get wrong.

The third reason is the one that decides it, and it is not an argument against
the pipe. It is what a pipe would take away. A pipe applies functions: `x |> f`
means `f(x)`, so every atom would have to become a function taking a plot and
returning one. Make `bar` a function, then ask what `bar * bin` could mean.
Nothing at all: you cannot multiply two functions and get a histogram. The
derive operator exists only because marks and transforms are *things* rather
than actions. They are values, and they combine with each other before either
one is joined to a plot.

That is the same decision this book opens with, arriving a second time. ㄱ is a
letter, not an instruction. You put it in a slot next to other letters, and the
block is assembled from its pieces rather than performed in sequence. Hangeul's
atoms are nouns, which is exactly why they compose, and gog's are nouns for the
same reason. A grammar built on the pipe is a grammar of verbs: it can layer,
because layering really is a sequence of additions, but it cannot derive.
Wickham's own pre-ggplot2 experiment was function composition, and it had no `*`
either.

So `+` here is a choice rather than something inherited. It is the one operator
that can be given a meaning in R, Python and Julia. It commutes wherever order
should not matter, and it leaves the atoms as nouns so that `*` can exist.

---

## Summary

| Operator | Reads as | Example |
|----------|----------|---------|
| `+` | "…and also…" | `point + color(continent)` |
| `*` | "…derived by…" | `bar * bin` |
| <code>&#124;</code> | "…split into panel columns by…" | <code>plot &#124; facet(continent)</code> |
| `/` | "…split into panel rows by…" | `plot / facet(continent)` |
| <code>&#124;</code> | "…beside…" | <code>plot_a &#124; plot_b</code> |
| `/` | "…above…" | `plot_a / plot_b` |

A finite set of rules. An unlimited number of plots.