# Operators

```{r}
#| include: false
source("R/setup.R")
```

How do you put two things in one picture? gog has exactly **four operators**.
They are not atoms. They are the assembly rules: how atoms combine.

Think of them the way Hangeul (한글) works:
consonants and vowels are the atoms;
the rule "initial + vowel + optional final = syllable block" is the operator.
The rule is fixed; what you can build with it is not.

```
+   compose   (add a layer, a channel, or a label to a specification)
*   derive    (combine a mark with a transform into a compound mark)
|   facet ↔   (small multiples, arranged in columns)
/   facet ↕   (small multiples, arranged in rows)
```

`|` and `/` also arrange whole plots on a page, and that second use has its own
section below.

---

## `+`: compose

`+` is the general assembly operator.
Every atom enters a specification via `+`:

```r
data(df) + point + x(col_a) + y(col_b) + color(group) + title("My plot")
```

Adding a **mark** (`point`, `line`, `bar`, …) starts a new layer.
Adding a **channel** (`x()`, `color()`, …) or **label** (`title()`) attaches
to the appropriate scope (plot-level or the nearest preceding mark, see
[Encoding scope](encoding-scope.qmd)).

### Layering with `+`

If you have edited video, you already know this idea. A video is a stack of
tracks: the picture, the sound, the captions, the logo, the effects. Each one
is separate, they play at the same time, and you watch one thing. A plot layers
the same way. It is not a new idea to learn.

Multiple marks in the same expression produce multiple layers, rendered in order:

```{r}
# Scatter plot: points, and a horizontal reference line at each band
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  data(life_bands) + rule + style(color = "darkgray", pattern = "dashed") +
  title("GDP vs Life Expectancy, 2007")
```

*"Given gapminder 2007: x is gdp, y is life, points colored by continent, and also rules from the life bands."*

Layers draw in the order they are written: the first mark sits at the
bottom. Add a `line` after `point` to draw the line on top of the points.

### The order of `+`: *data, mark, positions, refinements*

Inside a single layer the channels **commute**, and the mark commutes among them:
write the three in any order and the plot is the same. Three things can be
arranged six ways, so here are all six.

```{r}
((data(iris_flowers) + bar * bin + x(petal_length) + color(species)) |
   (data(iris_flowers) + bar * bin + color(species) + x(petal_length))) /
((data(iris_flowers) + x(petal_length) + bar * bin + color(species)) |
   (data(iris_flowers) + x(petal_length) + color(species) + bar * bin)) /
((data(iris_flowers) + color(species) + bar * bin + x(petal_length)) |
   (data(iris_flowers) + color(species) + x(petal_length) + bar * bin))
```

Six sentences, and the same histogram six times. Not six plots that look alike:
the engine emits the same bytes for every one of them.

So if the order does not change the picture, why does every plot in this book
follow the same one: *data, mark, positions, refinements*? Because a fixed
order lets a plot **read like a sentence**, and the sentence has a natural,
subject-first shape:

- **`data(...)` first**, the table the columns come from. A bare name like
  `petal_length` means nothing until a table is bound (Bind-Once), so the
  table is named first. In R the sentence *must* start here.
- **the mark and its `*`-chain second**, the subject. A visual is a mark
  before it is anything else (Minimum Syllable): `bar * mean * dodge` says
  *what you are drawing* and *what was done to it*, as one unbroken word. It is
  the initial consonant of the block; everything after is said about it.
- **the positions third**, the frame. `x` and `y` say *where* the subject
  stands, and a mark cannot render without them (the vowel to the mark's
  consonant), so they follow immediately.
- **the refinements last**, `color`, `size`, `order`, labels, `title`:
  everything optional. Delete every refinement and a legal plot remains; delete
  the mark or a position and nothing renders. Writing the removable parts last
  is also how you *build* a plot: the minimum sentence first, then
  refine ([First plot](first-plot.qmd)).

The order is a reading convention, not a grammar rule: the engine accepts any
permutation that scopes the same way. Where position genuinely *does* matter
(a channel written before or after a mark across two or more layers) is the
subject of [Encoding scope](encoding-scope.qmd). Keeping one order everywhere
is worth the trouble: once you learn it, every plot in the book, and every plot
you write, reads the same way.

Which parts of that order the engine actually enforces, and which are yours to
arrange, is gathered for all four operators in [Which orders
matter](#which-orders-matter) below.

#### A refinement far from what it refines

One part of the convention is easy to misread. A grouped bar chart pairs
`dodge` with `color`, yet the sentence puts them at opposite ends, with the
positions in between:

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

`dodge` and `color(era)` are plainly related: `dodge` has nothing to set side
by side until a `color` (or `group`) names the groups. So why not write them
together? Because the grammar puts them on opposite sides, deliberately.
`dodge` belongs to `*`: it modifies the mark, and it is the same modifier that
works on `box * dodge` and `interval * range * dodge`, neither of which
involves color. `color` belongs to `+`: it binds a column to a channel.
`color * dodge` cannot be written: `*` needs a mark on its left. Joining the
two would make `dodge` depend on which channel splits the groups, and a
modifier must not depend on that.

And the order is not a compromise; it is exactly right, once you see what
`dodge` does. `dodge` subdivides the **x** slot: each continent's space is cut
into one sub-bar per era. So it needs *both* neighbors: `x(continent)` says
*which slot* to divide, `color(era)` says *into how many pieces*. The sentence
lists them in the order `dodge` uses them: the axis it splits along, then the
groups it splits into. The positions are not in the way; one of them *is*
`dodge`'s other operand.

You also never have to scan the line to check the split is there.
`bar * mean * dodge` with no `color` or `group` does not draw a lone bar; it is
refused, pointing you at the missing channel. The legality check requires both,
so writing `color` far from `dodge` is safe: the two are bound whether or not
they sit side by side on the page.

---

## `*`: derive

`*` joins a transform to a mark, and the result is one **compound mark**.

```r
bar * bin     # histogram
line * smooth # smoothed trend line
line * density # kernel density curve
```

Two rules govern `*`, and only two. The mark goes on the left, and the
transforms go on the right. Then, among the transforms themselves, the order
you write them in does not matter: any order draws the same plot. The rest of
this section explains each rule.

### Why the order is fixed

In Hangeul, a syllable is always *initial consonant + vowel*, never the reverse.
gog's `*` works the same way: the mark is the subject; the transform is
what is done to it.

`bin * bar` is a type error, just as ㅏ (*a*) + ㄱ (*g*) is not a valid Hangeul
syllable.

```r
bar * bin   # ✅  "a bar mark, derived by binning"
bin * bar   # ❌  Error: a transform cannot derive a mark
```

### What `*` is rigid about

The rule that matters is the one above: the mark is on the left, always,
and no transform may stand there. That is what makes `bar * bin` a compound mark
rather than a function call, and it is why the atoms can be nouns.

**Write the transforms in the order they happen**, as a reading convention: any
order draws the same plot, and this order lets a reader run the sentence in
their head.
`bar * bin * mean` cuts the axis into bands, then averages inside each one. That is
the order it runs in, and there is no other order it could run in.

The reason is what `mean` needs. A statistic reduces a **group** to one number, and
a continuous column has no groups: every value stands alone, so its mean is itself.
`bin` is what makes the groups, by turning many distinct numbers into one band. So
the cut comes first because otherwise the mean has nothing to average.

That is also why the pairs that can stand together are the ones that do *different*
jobs: saying where the groups' cells are and what is in them, or what scale the
answer is read on. A pair that answers the same question twice is refused rather
than sequenced, so `bin` and `count` (both of them tallies with groups of their
own) cannot both be in one layer:

```{r}
#| error: true
data(gapminder_2007) + bar * bin(12) * count + x(life)
```

The engine decides the running order from the transforms themselves, not from
where you typed them, so you cannot get it wrong. Write it the way you mean it
and it will read the way it runs. [Chaining
transforms](transforms.qmd#chaining-transforms) takes the pairs one at a time.

### Live examples

```{r}
# Histogram: life expectancy across all countries in 2007
data(gapminder_2007) + bar * bin + x(life) + y(count) +
  x_label("Life expectancy (years)") + y_label("Count") +
  title("Distribution of Life Expectancy, 2007")
```

```{r}
# Scatter + LOESS smooth layer on top
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  title("GDP vs Life Expectancy with LOESS Smooth")
```

```{r}
# Kernel density estimate
data(gapminder_2007) + line * density + x(life) + y(density) +
  x_label("Life expectancy (years)") + y_label("Density") +
  title("Kernel Density: Life Expectancy 2007")
```

---

## `|` and `/`: facet, or compose

`|` puts things side by side and `/` puts one above the other. Which things is
decided by what is written on the right: a `facet()` splits one plot into
panels, another *plot* arranges the two on a page.

With `facet()`, `|` produces small multiples arranged in columns, one panel per
category of the column it names.

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) | facet(continent)
```

`/` produces small multiples stacked in rows:

```{r}
gm_two <- gapminder_2007[gapminder_2007$continent %in% c("Africa", "Europe"), ]
data(gm_two) + point + x(gdp) + y(life) / facet(continent)
```

Write both to cross them into a grid: each operator applies to the facet
written after it, read left to right. See [Faceting](faceting.qmd) for the
grid, what the panels share, and why the facet column must be a category.

With a plot on the right, the same two operators arrange separate plots on one
page, each keeping its own coordinate space, sharing an axis only where the two
name the same column on it:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + bar * count + x(continent))
```

One pair of operators, two meanings, and the operand decides which, rather than
a second pair of words. See [Composition](composition.qmd).

---

## Which orders matter

This book writes every sentence the same way: data, mark, positions,
refinements. That is a preference rather than a rule, and it is not all the
engine enforces. The difference is worth knowing, because it tells you where to
be careful. There are three kinds of ordering, and only one of them needs your
attention.

| Kind | What it means | Under `+` | Under `*` |
|---|---|---|---|
| **Free** | Either way draws exactly the same picture. | The mark and its positions, in any arrangement. `x` before `y`, or `y` before `x`. Refinements among themselves. | Two transforms after one mark. Write them in the order they happen; the engine works that order out either way. |
| **Refused** | The engine stops and names what to move. | A sentence that does not begin with `data()`. A `style()` written before any mark. | A transform on the left: `bin * bar`. |
| **Meaningful** | No error, and a different picture. | A channel before the marks, or after one of them, once a plot has two. Two marks swapped: the second draws on top. | Nothing. No order under `*` changes the picture. |

Read the last cell first. `*` is rigid about one thing only, which is that the
mark stands on its left. Once the mark is there, no arrangement of what follows
changes what you get.

**Free** is most of it. Every one of these draws the identical plot, and the
book writes the first only because a fixed order is easier to read:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life)) |
  (data(gapminder_2007) + x(gdp) + point + y(life))
```

Both name `life` on `y`, so the two panels share one axis and it is drawn once.
That is composition doing what the section above said it does, and it is why the
right panel carries no numbers of its own.

**Refused** is the engine telling you the sentence has no subject yet. A
setting needs a mark to attach to, and in R a plot needs its table first,
because `+` has to have a plot on its left before it can add anything:

```{r}
#| error: true
data(gapminder_2007) + style(size = 3) + point + x(gdp) + y(life)
```

**Meaningful** is the only one to watch, because it is the only one that stays
silent. With two marks in a sentence, a channel written before them belongs to
the plot and reaches both, and a channel written after one belongs to that
mark. Both are legal, neither warns, and they draw different pictures. That is
the subject of [Encoding scope](encoding-scope.qmd), and it is the reason this
book keeps one order: when every sentence is arranged the same way, the one
that is arranged differently is saying so on purpose.

The other silent one is the marks themselves. Two marks swapped draw in the
order you wrote them, so the second sits on top of the first.

### What groups first

The three kinds above are about where you may put a word. Precedence is a
different question: how the words you did write get grouped. It is what lets a
sentence like this one carry no parentheses at all.

```{r}
data(gapminder_2007) + bar * bin + x(life) | facet(continent)
```

`*` binds tightest, then `+`, then `|`. So the transform joins its mark first,
the atoms join the plot next, and the facet splits the finished plot last. `/`
is applied before `|`; the next section has the rule.

Those are R's own precedences, and Python's and Julia's, so one sentence groups
the same way in three of the four languages. JavaScript can give no operator a
meaning of its own, so it spells the four as words, which makes the grouping
something you see instead of something you remember
([JavaScript](bindings/javascript.qmd)).

### What parentheses may group

Parentheses change which operator runs first, and that is the whole of what they
do. They group **plots**, which is why a composed page uses them: `/` is
applied before `|`, so `a | b / c` reads as `a | (b / c)` unless you say
otherwise ([Composition](composition.qmd)).

They do not group **marks**. A second `data()` applies to the mark written
directly after it ([Data](data.qmd)), and putting parentheses around a run of
marks does not extend it to all of them:

```{r}
#| error: true
data(actuals) + x(year) + y(sales) +
  line +
  (data(forecast) + point + area)
```

gog refuses that instead of drawing it. The refusal is worth having, because
the alternative is worse than an error. There is no way to keep the marks
inside those parentheses. The only other choice is to drop them, and draw a
plot that is missing two layers without saying so. To put two marks on the
second table, write `data()` before each of them.

The same refusal covers a position or a title inside the parentheses, for the
same reason. In JavaScript, where the operators are words, the same mistake is
a `plot()` handed to another `plot()`, and it is refused there too.

---

## Why `+`, and not a pipe

Readers who know R's tidyverse packages ask this, and it deserves a real
answer, because Hadley Wickham, ggplot2's author, has said publicly that the
pipe would have been the better interface [@wickham2018pipe]. The history is
written down. ggplot, ggplot2's predecessor, was written as function
composition, and it could have piped had the pipe existed; ggplot2 used `+`
instead.

He also answers the question a reader asks next. ggplot2 cannot change now: its
functions take the mapping and the data first, where a pipe needs the plot. And
he judges that a new package built only to change this would not be worth
what it would cost. So for ggplot2 the matter is settled by history rather than
by taste, and a grammar starting fresh has to answer it on its own. Three
reasons are given for preferring the pipe. Two of them do not apply to gog, and
the third is the one that settles the question the other way.

The first is about one language. A pipe is a single idea that works
everywhere in R, so a grammar spelling composition differently is one more thing
to learn, and switching between the two spellings is a frequent source of errors
(his own included). That is true, and it is an argument about R and its
packages. gog is one syntax spoken by four languages, and that changes the
answer. `+` can be given a meaning in R, Python and Julia. The pipe cannot be
given a gog meaning. Julia's `|>` and R's two pipes already mean something gog
must not change, and Python has no pipe operator at all, and no way to add one,
since it lets you give new meanings only to the operators it already has.
Piping would have meant piping in two languages and spelling it some other way
in the third, which would leave the languages with two different grammars, and
this project exists to avoid that. JavaScript decides nothing here, because it
cannot overload any operator at all, which is why it spells all four as words
([JavaScript](bindings/javascript.qmd)).

**The second is about arithmetic**, and it is the strongest: you expect `x + y`
to equal `y + x`, and ggplot2's `+` does not always oblige. That objection is
answered a few sections up. Here the channels commute, and so does the mark
among them: move `color` ahead of `x`, or write the positions before the mark,
and the engine emits the same bytes. One thing does depend on order, two marks
swapped, and there the order is the drawing order you asked for: the second draws on
top.

The other half of that objection is grouping, whether `x + (y + z)` means what
`(x + y) + z` means. The question cannot arise here, because there is no way to
write the left-hand side. `+` always takes a plot on its left, so two atoms with
no data beneath them are not a smaller expression waiting to be grouped. They
are refused:

```{r}
#| error: true
point + x(gdp)
```

No pile of atoms accumulates, so there is no grouping to get wrong.

The third reason is the one that decides it, and it is not an argument against
the pipe. It is what a pipe would take away. A pipe applies functions:
`x |> f` means `f(x)`, so every atom would have to become a function taking a
plot and returning one. Make `bar` a function, then ask what `bar * bin` could
mean. Nothing at all: you cannot multiply two functions and get a histogram. The
derive operator exists only because marks and transforms are **things** rather
than actions, values that combine with each other before either one has met a
plot.

That is the same decision this book opens with, arriving a second time. ㄱ is a
letter, not an instruction. You put it in a slot next to other letters, and the
block is assembled from its pieces rather than performed in sequence. Hangeul's
atoms are nouns, which is exactly why they compose, and gog's are nouns for the
same reason. A grammar built on the pipe is a grammar of verbs: it can layer,
because layering really is a sequence of additions, but it cannot derive.
Wickham's own pre-ggplot2 experiment was function composition, and it had no `*`
either.

So `+` here is a choice rather than something inherited. It is the one operator
available in every language this grammar speaks, it commutes wherever order
should not matter, and it leaves the atoms as nouns so that `*` can exist.

---

## Summary

| Operator | Reads as | Example |
|----------|----------|---------|
| `+` | "…and also…" | `point + color(continent)` |
| `*` | "…derived by…" | `bar * bin` |
| <code>&#124;</code> | "…one panel per…" (columns) | <code>plot &#124; facet(continent)</code> |
| `/` | "…one panel per…" (rows) | `plot / facet(continent)` |
| <code>&#124;</code> | "…beside…" | <code>plot_a &#124; plot_b</code> |
| `/` | "…below…" | `plot_a / plot_b` |

A finite set of rules.
An infinite space of plots.
