# Play

```{r}
#| include: false
source("R/setup.R")
```

How did the whole picture change from year to year? `play` cuts the plot into
frames and shows them in sequence. It is the last channel in the kernel to
draw, and the shortest way to say what it does is that it is
[faceting](faceting.qmd) read in time. A facet splits the rows by a column's
distinct values and lays the pieces out across the page; `play` splits them
exactly the same way and lays them out along the clock.

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + color(continent) +
  play(year)
```

*"Given all the gapminder years: points, x is gdp on a log scale, y is life, color by continent, played by year."*

Twelve frames, one per census year in the data, running from 1952 to 2007. The
strip above the panel names the one you are looking at.

## The same sentence, in space

Everything `play` does, `facet` does too, and swapping one operator for the
other is the whole difference:

```{r}
gm_four <- gm_all[gm_all$year %in% c(1952, 1972, 1992, 2007), ]
gm_four$period <- as.character(gm_four$year)
data(gm_four) + point + x(gdp, scale = "log") + y(life) + color(continent) |
  facet(period)
```

```{r}
data(gm_four) + point + x(gdp, scale = "log") + y(life) + color(continent) +
  play(period)
```

Four panels, then four frames, over identical rows. The grid is better for
comparing two moments carefully, because both are in front of you and your eye
can go back and forth. The sequence is better for seeing *movement*, because a
point that shifts left is a point you watch move rather than one you have to
find twice. Neither is a different plot; they are one plot read two ways, which
is why they are one grammar with one column bound differently.

Print gets the first frame. This page in the PDF shows 1952 with its strip, and
that is why the faceted version is here beside it: the grid is what a reader
holding the printed book sees instead of the motion.

## The column needs an order

`play` and `facet` split the rows the same way, so you might expect the same
columns to suit both. They do not. The test is not the column's type. It is
whether the column has an order.

A facet lays its panels out in space, and you read space in any order you like.
Your eye goes to the panel it wants, then comes back. A sequence lays its pieces
out in time, and time arrives in one order only. So a sequence claims something
a grid never claims: this frame comes after that one.

Bind a column with no order, and the plot shows movement the data does not have:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + play(continent)
```

Watch the strip: Asia, Europe, Africa, Americas, Oceania. That order is not
alphabetical, and it is not geographic. It is the order the continents first
appear in the table, because the rows are sorted by country and Afghanistan
comes before Albania. The animation is showing you how a file was sorted. The
points jump between frames, and none of the jumps mean anything.

This is the same mistake as a rainbow ramp on a numeric column
([Channels](channels.qmd)). The encoding carries an order the thing itself does
not have, so a reader looks for a pattern that was never there. Written as
`| facet(continent)`, the five groups sit side by side and claim nothing about
sequence.

Ordered does not mean numeric. A year is ordered, and so is a date. So is a text
column whose levels you declared: months in calendar order, or a rating from low
to high. gog runs the frames in the order the levels declare, the same way it
runs an axis ([Data](data.qmd)). A text column with no declared levels falls
back to the order its values first appear, which is the trap above.

How many values there are is a separate question, and the next section takes it
up. Twelve
frames and a hundred and forty frames are both worth animating, as long as each
one means something beside the one before it. A long sequence costs time rather
than sense, and [`speed`](#how-fast) is there for that.

## Too many panels for one page

A sequence can hold more pieces than a facet can, and this section is about that
difference. It splits the rows the way a facet does, and nothing limits how many
pieces it shows, because it shows them one at a time.

Whether a column has an order decides whether it can be played. How many values
it has decides something else: whether a facet of it is still readable. A facet
has to fit every piece on one page. A sequence does not.

Having an order is not on its own a reason to play a column. A short sequence
earns its motion only when its frames hold the same things in different states.
Then a reader can follow one of them across the frames. The four frames above are
that case: the same countries at four dates, and the cloud climbs. Four frames
holding four different sets of things are not that case. Nothing moves between
them. The reader waits, and then has to remember what a facet would have shown at
once. This is worth saying plainly, because animation is the choice most people
want to make.

The other end is where the panels stop. They share one page, so every panel added
makes every panel smaller. `wrap` folds a long line into a rectangle, which makes
each panel wider and shorter ([Faceting](faceting.qmd)). The rectangle is still
one page.

Here are the two gliders from [the cube](space.qmd), cut into one panel per
second instead:

```{r}
thermal_moments <- thermals
thermal_moments$moment <- factor(thermal_moments$second)
data(thermal_moments) + point + x(east) + y(north) + color(glider) |
  facet(moment, wrap = 13)
```

170 panels, already folded into a rectangle. The shapes are still there: two dots
in every cell, one for each glider. Their positions are not. The tick labels have
printed over one another. A dot cannot be read against them, and its position is
all this plot measures. A reader can see that something moves, and cannot see
where.

The same rows along the clock instead:

```{r}
#| message: true
data(thermals) + point + x(east) + y(north) + color(glider) +
  play(second, speed = 3)
```

One panel at full size, one pair of axes, and the two gliders circling through
all 170 frames.

gog drew both of those sentences. It refuses neither, and it never will. A plot
is refused for not being well formed, never for being a poor idea ([Design
laws](design-laws.qmd)). Which of the two says more about gliders is your
judgment, not the engine's. What the engine does instead is name the one cost it
can measure. The note above counts the frames and reports how long one loop
takes. It offers two answers: [`speed`](#how-fast), or a column with fewer values.

That judgment is worth making before anyone else sees the plot. A facet that was
the wrong choice costs a reader one glance. A sequence asks an audience to watch,
and holds them for as long as the loop runs. If it then says nothing, that time
was spent for nothing, and an audience can usually tell.

The choice is easiest to see with a large number. Imagine a year of hourly
temperatures for New York, drawn as one curve for each day. As a facet that is
365 panels, and no fold leaves a curve large enough to read. As a sequence it is
365 frames, and a reader watches a year of weather pass. Below about ten pieces
the panels usually read better, and [Faceting](faceting.qmd) shows ten of them
wrapped and readable. Above that the page cannot hold them, and the sequence is
what is left.

Holding more does not make a sequence the better picture. What it gives up is
what a facet does best: two pieces side by side, where a reader compares them
instead of remembering them.

## Nothing moves that the data did not move

Every scale, the color map and every legend are fitted across the whole
sequence at once, never per frame. That is the property that makes an
animation worth trusting, and it is worth seeing what its absence would look
like. If each frame fitted its own axis, 1952's poorest countries and 2007's
would both sit at the left edge. A world that grew six-fold would draw as a
world that stood still. The only thing actually moving would be the axis,
silently, with nothing on the page to say so.

Because the frames share their scales, the opposite reads correctly: in the plot
above, the cloud genuinely walks right and up, and the countries that do not
move are countries that did not.

The color map is shared for a second reason. A reader tracks a bubble by its
color, so a palette reassigned partway through would break the one thread they
are holding on to. Every continent keeps its color in every frame, including
frames where it happens to have no rows.

What is *not* shared is the statistics. A transform sees each frame's rows as if
they were the whole data, exactly as it sees a facet panel's: the No Exceptions
law applied to frames:

```{r}
gm_past60 <- gm_all[gm_all$life > 60, ]
data(gm_past60) + bar * count + x(continent) + play(year)
```

Each bar is that year's tally of the countries that had passed sixty years of
life expectancy. Asia climbs from four to thirty, Africa opens with none and ends
with twelve, and Europe starts at twenty-three and then runs out of countries to
add.

A tally can also have nothing to report, and it is worth seeing that once,
because a still animation is sometimes an answer rather than a fault:

```{r}
data(gm_all) + bar * count + x(continent) + play(year)
```

Twelve frames run and not a bar moves. The statistic is being recomputed in every
one of them; it keeps arriving at the same number, because gapminder holds the
same 142 countries in every year, so counting rows per continent counts the
*table* and not the year. Only the strip changes, which is the plot saying so.
Filtering first, as the sentence above it does, is what turns a count of countries
into a count of something that happened.

What the frames do share is the *cut*, when there is one. A `bin`'s cells
describe the axis rather than measuring the rows. One set of bins is cut from
every frame's rows at once, and each frame then counts only its own into them.
[Faceting](faceting.qmd) sets out why, and the rule is the same word for word
with "panel" read as "frame".

```{r}
data(gm_all) + bar * bin + x(life) + play(year)
```

The distribution walks right across half a century, and it does so against bars
of one width, which is the only way the walk is legible.

## A layer that does not play stands still

`play` is a channel, so it obeys the ordinary [scope](encoding-scope.qmd) rule:
written before any mark it reaches every layer, and written after one it binds
that layer alone. A layer that never names it is drawn whole in every frame,
which is how a fixed reference sits behind something that moves.

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + play(year) +
  data(life_bands) + rule + y(life) + style(color = "#c0392b")
```

The two lines hold still while the cloud climbs past them. Nothing was added to
the grammar to allow that: the reference layer simply never mentioned `year`,
and a layer with no `play` binding has no frames to be cut into.

The same scope rule, in the cube and across panels at once. The route is the
whole flight drawn at once, and the marker is where the glider actually is:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  data(thermal_marks) + point + play(instant, speed = 3) | facet(glider)
```

`play` is bound to the second layer because `x`, `y` and `z` were bound to the
first, and a channel written after a mark reaches that mark alone. So the coil
stands still in both panels while a dot climbs it, and you can see the tighter
glider on the right outclimbing the wider one on the left, which is the whole
reason the two tables are drawn together.

One detail of the second table is worth knowing, because it decides the size of
the plot rather than its meaning. `thermal_marks` calls its time column `instant`
where `thermals` calls the same measurement `second`. The frames are gathered
from every table that carries the played column, not only from the layer that
bound `play`, on the same rule that gives the plot one scale rather than one per
layer. Had the marker also called its column `second`, the backdrop's 170 samples
would have set the frame count and the sequence would have run 170 frames instead
of 17.

## How fast

`speed` is a multiple of the normal pace, written on the binding beside
`scale` and `limits`, because it belongs to `play` and to nothing else:

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + color(continent) +
  play(year, speed = 3)
```

The same twelve frames this chapter opened with, three times as fast; `speed =
0.5` would run them at half pace. Long sequences are the reason it exists, and
gog says so before you ask: past about thirty frames it reports how many seconds
the loop will take, because the length of a loop is a default you did not choose
and cannot see.

## The column that cannot do both jobs

There is one shape worth knowing because it looks so reasonable. Bind the same
column to `x` and to `play`, on a mark that reads a function along `x`, and every
frame is left holding a single position:

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + color(country) +
  play(year, speed = 3)
```

A `line` needs at least two positions to read between. If each frame is one year,
there is no domain left in it, and what you get is axes and a legend around an
empty panel. That refusal is here because the first draft of this chapter
contained exactly that sentence, and it rendered, silently, exit 0, a frame
strip counting up over nothing. A reader spotted it.

The same is true of the same column split across the page rather than the clock,
`| facet(year)`, and for the same reason; the engine refuses both, because
[panels and frames are one partition asked twice](faceting.qmd). Marks that place
each row on its own (`point`, `bar`, `box`) have no domain to lose, so they
sweep through happily:

```{r}
data(gapminder_asia) + point + x(year) + y(life) + color(country) + play(year)
```

## Crossed with a facet

Panels split the page and frames split the clock, so the two compose and every
panel advances in step:

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + play(year) |
  facet(continent)
```

Five continents, twelve frames, one shared pair of axes. Each panel is its own
story and the strip above them all says which year every story is at.

## The cube also moves

Nothing about `play` belongs to the plane. A frame is a subset of the rows, and
the [third position](space.qmd) is a way of looking at whatever rows you have, so
the two compose, and neither needed anything added for the other:

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + z(population) +
  color(continent) + play(year)
```

That one is worth reading slowly, because it is the sentence this kernel was
designed around. The vocabulary was drawn up with four words held in reserve for
things it could not yet draw: the third position, the animation, the per-layer
table, and the selection. The sentence above spends three of them at once. The
fourth, `brush`, draws now too. [Selection](selection.qmd) closes by adding it to
this very sentence, so all four are spent in one line.

The cube is fitted once across the whole sequence, exactly as a flat plot's axes
are, and it earns more here than it does on the page. A box that resized itself
per frame would swing its own corners around a cloud that had not moved, and
every apparent motion would belong to the guide rather than to the countries.
Fitted once, the height axis has room for 2007's China from the first frame
onward, which is why 1952 opens with the cloud pressed low and only China and
India standing well clear of the floor.

A transform inside the cube is the same transform. The floor is cut once from
every frame's rows and each frame then counts only its own into the cells, which
is this chapter's shared-cut rule read with a second axis present:

```{r}
data(gm_all) + bar * bin(12) + x(life) + y(continent) + color(continent) +
  palette("okabe") + space() + play(year)
```

Follow one continent's row of columns rather than the whole floor. Each pile
slides toward longer lives as the sequence runs, and it slides against cells that
never move, so what you are watching is the countries changing which cell they
fall in. `color(continent)` names the same column the `y` axis does, and it is there so
that following one row is possible at all. A floor axis in a cube runs
diagonally into the picture. Picking a row off it by eye means tracing a
receding edge while five rows of columns stand in the way. The legend does in one
glance what the axis makes you work for.

The palette is a choice the cube makes for you, and it is the same argument the
[`okabe` set](channels.qmd) was built on arriving from an unexpected direction. A
solid in space is shaded per face, so one category is already three tones before
any two categories are compared, and a palette whose hues sit close together
loses the distinction that shading eats. Okabe-Ito's eight are spaced for readers
who cannot separate some hues at all, which turns out to be the same requirement
as separating them after a third of the light has been taken out.

What the frames do not touch is the viewing angle. `space(turn = , tilt = )` is a
property of the space rather than of the data, so one angle serves the whole
sequence and every frame is seen from the same place. That is not a gap waiting to
be filled: a camera that swung while the sequence played would be motion carrying
no column, and a reader would then have to subtract it from the motion that does
carry one. `play` moves the data past a fixed eye.

You can still turn the plot yourself while it runs, and that is a different
thing. The refusal above is about a camera that moves on its own, with no column
behind it. A reader who drags knows which motion is theirs, and can stop it.
This is worth more here than anywhere else in the book. Something is moving
inside a solid shape. The only way to see what the near edge hides is to look
around it.

Panels and frames do not come apart here, which is the point. They are one
split asked twice, so crossing them costs nothing: the panels divide the page,
the frames divide the clock, and each panel projects the cube its own rectangle
has room for.

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + z(population) +
  color(continent) + play(year) | facet(continent, wrap = 3)
```

Five cubes, twelve moments, and one strip above them all naming the year, because
a frame is the whole plot where a panel is one piece of it. That is four of the
kernel's ideas in one sentence: the third position, the panels, the frames, and a
scale shared across every one of them so the cubes stay comparable while they run.

And the sheet moves too. `ripples` is a wave tank tapped in one corner, measured
on a grid at twelve instants, in a shallow tank and a deep one.

```{r}
#| echo: false
peek(ripples)
```

One row is one grid node at one instant in one tank, so `instant` counts frames
from 1 to 12 rather than measuring a time, and `height` is a signed height in
millimeters. Depth sets how fast water carries a wave, so the same tapping makes
crests 140 mm apart in the shallow tank and 220 mm apart in the deep one:

```{r}
data(ripples) + surface + x(across) + y(along) + z(height) +
  color(height) + palette("plasma") + play(instant, speed = 4) | facet(tank)
```

Watch one arc leave the corner and cross the floor. It is the same wave in both
panels and it is drawn against one shared height scale, so the difference you see
is a wavelength and not a scaling. `speed = 4` is doing real work here: at the
normal pace twelve frames take nine and a half seconds, which is a slideshow of a
wave rather than a wave.

The height is said twice, once as the third position and once as the color, and
that is worth doing rather than a redundancy to apologize for. A sheet seen from
one angle hides part of itself: the far corner is foreshortened and the near
crests stand in front of the troughs behind them, so a shape that is plain at the
front is guesswork at the back. Bright crest bands survive the projection where
the height alone does not, and the wavelength this plot exists to compare is the
spacing of those bands. `palette("plasma")` runs dark magenta through orange to
yellow; [Scales](scales.qmd) has the rest, including two diverging ramps whose
neutral would sit at the still water line if the sign were the point instead.

## The circle also moves

One space over, the same holds and for the same reason. [`polar()`](polar.qmd)
bends the panel, the frames choose the rows, and neither has to know about the
other:

```{r}
data(gm_all) + bar * mean + x(continent) + y(life) + polar() + play(year)
```

A rose of mean life expectancy, one petal per continent, opening across half a
century. The radial axis is fitted over the whole sequence exactly as a flat one
is, which is why the ring at 80 is already drawn in 1952 with nothing reaching it:
the petals grow inside a circle that does not resize under them. Asia's is the one
to watch, from 46 years to 71. Africa's climbs steadily until the late 1980s and
then stops for fifteen years, which is a decade the average is reporting rather
than smoothing away.

The pie is where that shared axis says something surprising, and it is worth
following, because it is the flat rule bent rather than a rule of its own:

```{r}
data(gm_all) + bar * sum * stack + y(population) + color(continent) + polar() +
  play(year)
```

The circle does not close until the last frame. A pie's turn *is* its measure
axis, that axis is fitted across the sequence like every other, and 1952's world
holds a little under two fifths of 2007's, so 1952 sweeps a little under two
fifths of the turn. The identical sentence without `polar()` draws a stacked
column that stops well short of the top of its axis, and nobody finds that
surprising. A pie that does not close is that column bent.

If the question is each year's *shares* rather than each year's total, the
sentence can say so, and then every frame closes:

```{r}
data(gm_all) + bar * sum * stack(share = TRUE) + y(population) + color(continent) +
  polar() + play(year)
```

The turn is one whole in every frame now, and what moves is how it divides: Asia
widens a little from 58% to 61%, Europe falls from 17% to 9%, and Africa grows by
half again, from 10% to 15%. Two sentences, one atom apart, answering two
different questions about the same column. Choosing between them is the reader's
job and the grammar's business is to keep both sayable.

## Taking the sequence with you

Every plot on this page moves because the SVG carries its own timing. A browser
reads that timing. Most other places do not. A message to a friend, a slide in a
talk, a post on a social site: each of them shows the first frame and stops.

`save_gif()` writes the same sequence as a file those places do play:

```{r}
wave <- file.path(tempdir(), "wave.gif")
save_gif(data(ripples) + surface + x(across) + y(along) + z(height) +
           color(height) + palette("plasma") + play(instant, speed = 4) |
           facet(tank),
         wave, scale = 2)
cat(basename(wave), round(file.size(wave) / 1024), "KB\n")
```

Nothing has to be installed first. The engine converts and encodes the file
itself, so the same line works on every machine that can draw the plot at all.

`scale` multiplies the canvas. A plot is 800 by 600 pixels unless its theme says
otherwise, which is small for a post, so `scale = 2` doubles both numbers. The
pace carries over without being restated: this sequence was written `speed = 4`,
and each frame holds for a fifth of a second in the file as it does on the page.

The file cannot disagree with the plot beside it, and the reason is worth one
paragraph, because it is the same reason the last section gives for having one
renderer. The frames are not drawn again one at a time. The plot is drawn once,
with every scale, the color map and each legend fitted across the whole sequence,
and each frame is that single drawing with one moment left showing. So there is
no second set of decisions that could come out different.

Two sentences are refused. A plot with no `play()` has no moments to write, and
saying so is more useful than writing a file of one frame:

```{r}
#| error: true
still <- data(gapminder_2007) + point + x(gdp) + y(life)
save_gif(still, file.path(tempdir(), "still.gif"))
```

And the path has to end in `.gif`, because the name of a file is a claim about
what is inside it:

```{r}
#| error: true
moving <- data(gm_all) + point + x(gdp) + y(life) + play(year)
save_gif(moving, file.path(tempdir(), "wave.png"))
```

One thing changes on the way out, and it is worth knowing before you send the
file rather than after. A GIF is pixels. The text stops being text, so it no
longer sharpens when a reader zooms in, and the colors are fitted to a palette of
256. The SVG remains the plot itself. The GIF is a copy of it made for somewhere
that cannot read the original.

### Why a GIF and not a video

A video file would be smaller and sharper than a GIF, and a few places accept
video and nothing else. gog writes no video, and the reason is about what a
package can carry rather than about what a plot is.

Making a video needs a video encoder. The common one is more than ten times the
size of this whole engine, once for every platform a package supports, and its
license does not allow it to travel inside a package licensed the way this one
is. The other route is to require the reader to install it separately. That
sounds harmless until you count who cannot. A reader who installs a prebuilt
package rather than building from source gets whatever the package carries, so a
tool that cannot fit inside it never reaches them at all. The feature would work
for some readers and refuse for others, which is a worse answer than not having
it.

A GIF asks for nothing. The engine writes it on every machine that can draw the
plot, in every one of the four languages. When the file looks soft, raise
`scale`. When you genuinely need a video, converting a GIF into one is a solved
problem, and there are free converters online that do it in a step.

## What it refuses

One column names the panels or names the frames, and not both. Drawn, it would
put one panel's rows in one frame and leave every other panel empty:

```{r}
#| error: true
data(gm_four) + point + x(gdp) + y(life) + play(period) | facet(period)
```

A `speed` that is not a pace is refused where you wrote it:

```{r}
#| error: true
data(gm_all) + point + x(gdp) + y(life) + play(year, speed = 0)
```

`speed` belongs to `play` and to nothing else, because nothing else in the
grammar spends time: a color is drawn all at once, so there is no duration for
a multiplier to scale. That one is not shown here as a gog refusal, and the
reason is worth saying. No binding gives any other channel a `speed` argument at
all, so `color(continent, speed = 2)` fails in R (or in Python, or Julia, or
JavaScript) before a specification is ever built. The engine still checks it,
because a specification can also be written by hand, but you will never see that
message from a sentence: the language you are writing in refuses first, which is
the earlier and better place for it.

## What is not here yet

Two things people expect from an animation are missing. Neither is an oversight,
and the two are missing for different reasons. One waits on a word the grammar
does not have. The other waits on nothing but the work.

There is **no tweening**: frames cut, they do not glide. A bubble in 1957 and a
bubble in 1962 are two complete pictures shown in turn, not one bubble eased
between two positions. What easing needs first is a **key**, something that says
Afghanistan in one frame is the same country as Afghanistan in the next, and the
grammar has no word for that yet. `group` looks like the word until you notice
that a `point` does not take it, and the bubble everyone wants to see glide is a
point. So the missing piece is a vocabulary decision rather than a drawing
technique, which is why it is owed rather than half-built.

There is **no transport**: no play button, no pause, no scrub bar. The sequence
loops.

For a while this looked like the harder of the two. A scrub bar seems to pick out
a stretch of frames, and picking things out is what the `brush` word is for. That
reading turned out to be wrong, and seeing why takes one paragraph. It is worth
the paragraph, because it decides where the buttons belong.

Every frame is already in the file. The whole sequence is drawn once, and a clock
decides which frame you see at each moment. So a scrub bar moves the clock. It
chooses no rows, hides no rows, and changes nothing the sentence said. Play and
pause do the same thing, one step smaller.

That puts all three on the same side of a line this book has already drawn. You
can turn a plot in the cube with the mouse ([Space](space.qmd)). Turning it
changes how you look, never what is drawn, so no word was added to the grammar
for it. The transport is the same kind of control, for time instead of angle. It
belongs to the page a plot is shown on, not to the sentence that describes the
plot. Nothing in the grammar has to grow to get these buttons. They are simply
not built yet.
