# Polar

```{r}
#| include: false
source("R/setup.R")
```

Which way does the wind blow most often? Some questions do not fit a flat page.
Every plot so far has been drawn on one: an `x` running across, a `y` running
up. That is one coordinate
space, and the kernel names six. It is also the wrong shape for a whole class
of variables. Compass direction, the hour of the day, the month of the year,
the phase of anything that repeats: for all of them the last value is the
neighbor of the first, and a straight axis has nowhere to say so.

Two of the oldest statistical graphics already live in this space. Playfair's
*Statistical Breviary* of 1801 cut a circle into sectors to show the Turkish
empire's territory and gave us the pie chart [@playfair1801]. Florence
Nightingale's 1858 diagram of what killed British soldiers in the Crimea
[@nightingale1858] is a bar chart in polar
coordinates, and it answers to three names: a **coxcomb**, a **rose**, or, most
precisely, a **polar area diagram**. That third name is the one carrying
information. Her wedges put each month's death count into the sector's *area*, so
their radius grows as the square root of the number, while a polar bar here maps
the value straight to the radius. Both are honest drawings and they are not the
same drawing, which is worth knowing before you set one beside the other.

The line version has collected as many names: **radar** plot, **spider** plot,
**star** plot, and **Kiviat** diagram in engineering. As with the rose and the
pie, none of them is an atom. Each is a mark and a channel or two with `polar()`
at the end, which is what the rest of this chapter is about.

## The axis that breaks a circle

Here is a season of wind observations as an ordinary bar chart, one bar per
compass point:

```{r}
data(winds) + bar * count + x(direction)
```

*"Given the winds: bars derived by count, x is direction."*

Read the two ends. North is at the far left and north-west at the far right,
with the whole compass laid out between them, when on the ground they are
45 degrees apart. The chart has cut the circle open and flattened it, and the
cut is invisible: nothing in the picture says that the two ends belong together.
A reader who did not already know the compass could not recover it from this.

Now add one atom to the same sentence:

```{r}
data(winds) + bar * count + x(direction) + polar()
```

Same mark, same transform, same binding, same data. The only thing that changed
is the space the plot is drawn in, and north-west is beside north because on the
compass it is beside north. The prevailing westerly is now a shape rather than a
tall bar you have to look up in the labels.

This is the point of a coordinate space, and it is why a pie chart, a rose and a
radar plot are not chart types in this grammar. They are marks you already know,
read in a circle.

## What `polar()` changes, and what it leaves alone

`x` becomes the **angle** and `y` becomes the **radius**. That order is
Wilkinson's [@wilkinson2005], and his reason is that the first dimension of a plot is its domain:
the thing you look values up by, which in a circle is the bearing you look along.
The second is the range, the amount, which becomes how far out from the center
the mark reaches. (That is the rule when both positions are bound. Bind only one
and there is nothing to choose, so the one you have becomes the angle: that is
the pie, and it has its own section below.)

So `x(direction)` still says what it always said, that direction is the position
each bar stands at. `polar()` only decides that positions are bearings rather
than places along a line. A bar still runs from its baseline to its value, which
is why the count still sets how far each wedge reaches: the center is zero.

Nothing else in the grammar knows the space changed, so nothing else has to be
rewritten for it:

```{r}
data(winds) + bar * count * dodge + x(direction) + color(season) + polar()
```

`color` splits the count by season, `dodge` sets the two halves of the split side
by side inside each slot instead of on top of each other, and the legend decodes
the hues. Every one of those works in a circle exactly as it works on a page,
because none of them was ever written in terms of pixels.

[`play`](play.qmd) is in that list too, and it is the one worth naming out
loud, because a rose that advances through time asks for nothing new. Frames
choose the rows and the space bends whatever they chose, so the two never
meet. That chapter's "The circle also moves" runs a rose and two pies through
half a century, and it is also where the pie's turn gets read as what it is, a
measure axis shared across the sequence like any other.

Change the one word and the seasons pile outward from the center instead, one
ring on top of another:

```{r}
data(winds) + bar * count * stack + x(direction) + color(season) + polar()
```

Summer lies against the center and winter rides on top of it, so each wedge now
reaches the season total: west runs out to 63, where the dodged rose above
stopped at the taller of its two halves, 38. The radial axis grew to fit, and
that is why `stack` is a transform rather than a drawing option. What it
accumulates is what the scale has to read.

The pair is worth reading together, because each buries what the other shows.
Stacked, the compass reads by total wind, and west is plainly the prevailing
direction. Dodged, the seasons stand side by side, and you can see that the
south-west blows in summer (30 against 17) while the west blows in winter (38
against 25). Neither is the correct rose. The rose is not a chart with its own
options at all. It is `bar`, with its axis bent.

## Where the circle starts

The angular axis begins at twelve o'clock and runs clockwise, which is how a clock
and a compass are both read, and the first category sits *on* that mark rather
than beside it. That is why the rose above needed nothing said to it: north points
straight up, because north is the first of the eight directions and the axis
starts at the top.

It is worth knowing why that is not free. A categorical scale runs from half a
slot before the first category to half a slot after the last, so the scale's own
origin is padding rather than any place in the data. Aim the start angle at the
padding and north lands at 22.5 degrees on an eight-point compass, leaving you to
work out `-180/8` to bring it back. A flat categorical axis already puts its tick
at the category's center, so the circle agrees with it: `start` points at the
category, not at the gap before it.

From there, `polar(start = )` turns the whole space by a number of degrees:

```{r}
data(day_cycle) + line + x(hour) + y(trips) + polar(start = -180)
```

The same day as before, half a turn round: noon is at the top now and midnight at
the bottom, which is the right way up if what you are reading is the working day
rather than the calendar one. `start` is a property of the space rather than of
the data, the polar counterpart of the `space(turn =, tilt =)` that sets the angle
a 3-D plot is viewed from (see [Space](space.qmd)). It changes how you look, never
what is drawn.

## One turn, whatever the data spans

The angular axis is **periodic**: one full turn covers exactly the range the axis
covers, so its two ends land on the same spoke. That is what closes the circle
with no seam and no dead wedge, and it has a consequence worth saying out loud.
The circle needs the whole cycle, or the two ends will meet in the wrong place.

Here is a day of trips, hour by hour. The table runs from hour 0 to hour 24 with
*both* endpoints present, because midnight is where the day ends and also where
it begins:

```{r}
data(day_cycle) + line + x(hour) + y(trips) + polar()
```

The curve closes at the top because the reading at 24 is the reading at 0, and
the two sit on the same spoke. Most periodic data is not so obliging. A tide
gauge reads every three hours and reaches neither midnight, so the axis it gets
spans 1 to 22 — and the failure that causes is not subtle:

```{r}
data(tide) + line + x(hour) + y(height) + polar()
```

One turn covers 1 to 22, so hour 1 and hour 22 land on the **same spoke**: the
first and last readings are drawn on top of each other, three hours apart in the
world and zero degrees apart on the page. Every reading between them is off true
as well, spread over 21 hours of angle instead of 24.

A periodic axis cannot tell that your variable is periodic; nothing in a column
of hours says the day wraps. So say it:

```{r}
data(tide) + line + x(hour, limits = c(0, 24)) + y(height) + polar()
```

Now each reading sits where a clock would put it — hour 1 a sixteenth of a turn
past midnight, hour 22 a sixteenth short of it — and the two are 45 degrees
apart, which is the three hours they really are.

**The curve still does not close, and that is the honest answer.** The gauge took
no reading at midnight, and `limits` states a domain; it does not invent an
observation. The gap you can see between hour 22 and hour 1 *is* the three-hour
gap in the data, drawn at the size it is. Contrast the trips above, which close
because a reading at 24 genuinely exists. A categorical angle is the third case
and closes for a third reason, below: there every slot is filled, so the last
category is adjacent to the first by construction rather than by luck.

`limits` states the domain the axis runs over rather than deriving it from the
data, which is the general fix and not a polar one — the same words hold a color
ramp still across facets. [Scales](scales.qmd) has the rest of it. Note what it
did *not* do here either: no reading was dropped, because 1 through 22 all lie
inside 0 through 24. Stating a domain widens as readily as it narrows.

The same cycle filled is `area`, which closes on the baseline ring rather than on
a straight line under the curve:

```{r}
data(day_cycle) + area + x(hour) + y(trips) + polar()
```

## The radar: a categorical angle

A named angle closes differently, and the difference is worth following because
it is the data's doing rather than the renderer's. An hour axis closes because
*you* gave it both ends of the cycle, 0 and 24, and the scale put them on one
spoke. Categories cannot do that: each appears exactly once, by definition, so
there is no repeated endpoint to land on. What they have instead is that they
exhaust the turn between them, which makes the last one genuinely adjacent to the
first. So a path across categories closes:

```{r}
data(gapminder_2007) + line * mean + x(continent) + y(life) + polar() +
  title("line * mean + x(continent) + polar()")
```

That is a radar plot, and there is no `radar` atom in this grammar — the same
sentence flat is the profile from the [Line](marks/line.qmd) chapter, with a
straight axis and therefore two ends. Bending the space is what joins them.
Filled, it is the shape people usually mean by the name:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) + polar() +
  title("area * mean — the filled radar")
```

Split it and each group is its own outline, the split reading exactly as it does
on a flat profile:

```{r}
data(gm_eras) + line * mean + x(continent) + y(life) + color(era) + polar() +
  title("half a century apart, two rings")
```

The radar collects a criticism the rose does not escape either: a category's
distance from the center is a radius, so its area grows as the square, and the
enclosed shape invites you to compare regions that no number in the table
corresponds to. Reach for it when the categories really are a closed set with no
first or last member, which is the case a straight axis misreports by having to
pick an order. When the question is which continent is highest, the flat bar
chart of the same sentence answers it precisely, and is two atoms away.

## An angle that is measured, not named

An angular position does not have to be a category. `bearing` is the same wind
observation recorded in degrees, a continuous position that happens to run round
a circle:

```{r}
data(winds) + point + x(bearing) + y(speed) + color(season) + polar()
```

Each dot sits at its own bearing, at the radius its speed earns, and the eight
clusters the observations were recorded in are visible as clusters rather than as
eight separate columns. The radial gridlines are read outward from the center,
so a dot near the rim is a fast wind whichever way it was blowing.

`bin` cuts a measured angle into wedges the same way it cuts a flat axis into
bars:

```{r}
data(winds) + bar * bin + x(bearing) + polar()
```

That is the *rosa ventorum* that cartographers drew by hand from the 13th
century, and it is one atom different from a histogram: `bar * bin` in both
cases, read on a line or read in a circle. The bins touch here as they touch
there, because a histogram cuts a continuous axis into adjacent intervals, and
that is true whichever shape the axis has been bent into.

A compass is a cycle, though, and the periodic axis has the same blind spot it
had with the hours. The range is fitted to what was observed, so the turn is
the span of the *bins* rather than the 360 degrees a bearing actually runs
over. The wrap lands a little off north. Saying the period fixes it here exactly as it
did there:

```{r}
data(winds) + bar * bin + x(bearing, limits = c(0, 360)) + polar()
```

The wedges now tile the compass rather than the data, so north on the plot is
north.

## The pie: where you put the category

Everything so far put the category on a *position*: `x(direction)` gave each
compass point its own slot, and the count decided how far out the wedge reached.
A pie does the opposite. Its categories are not positions at all; they are a
split, and what goes round the circle is the measure itself.

That is the whole difference, and it is sayable. Take the category off `x` and
give it to `color`, then say that the pieces lie end to end:

```{r}
data(winds) + bar * count * stack + color(direction) + polar()
```

Read the sentence: a bar, counted, stacked, colored by direction, in polar. Not
one word of it is about pies. The slices are the segments of a stacked bar, the
angles are their shares, and the legend is the key.

The flat form is the same sentence without the last atom, and it is a real chart
in its own right, the one column that shows how a total divides:

```{r}
data(winds) + bar * count * stack + color(direction)
```

Those two pictures carry identical numbers. Wilkinson's second chapter is called
*How To Make a Pie*, and his answer is exactly this: a pie chart is a stacked bar
in polar coordinates. The column's height is the pie's full turn, and each
segment's share of the height is its slice's share of the circle.

Nothing new was added to the grammar to reach it. There is no `pie()` atom, no
`type = "pie"`, and no argument anywhere that says which chart to draw. Two
things happen on their own. First, a bar whose split *is* its segmentation has
only one slot to stand in, so it needs no position axis, and dropping `x` is what
says so. Second, a plot with one position bound has nothing to choose between:
that position becomes the angle, and the radius is left as the constant that sets
the pie's size. Both were in Wilkinson's book, where the one-argument polar
function is a separate tool from the two-argument one.

The measure can be anything a group can be reduced to, not just a count. Swap
`count` for `sum` and name the column to total, and the same eight directions are
weighted by how hard the wind blew rather than by how often:

```{r}
data(winds) + bar * sum * stack + y(speed) + color(direction) + polar()
```

Compare it with the pie above: the same categories, a different verdict. West is
both the commonest direction and the strongest, so its slice widens from about a
quarter of the circle to a third. The northerlies and southerlies are the gentlest
winds, so theirs shrink by roughly two fifths, even though exactly the same
observations went into them. One atom decided which question the circle answers.

And `proportion` puts the shares on the axis rather than leaving them implied:

```{r}
data(winds) + bar * proportion * stack + color(season) + polar()
```

The split has to be there. A stack with nothing to pile, or a bar with neither a
position nor a split, is refused rather than guessed at, and so is a statistic
that only means something *along* an axis:

```{r}
#| error: true
render_svg(data(winds) + bar * bin * stack + color(season) + polar())
```

## What a circle costs

The grammar will draw this, and drawing it is not the same as recommending it.
A wedge's *area* grows with the square of its radius, so a count that doubles
does not draw twice the ink, it draws four times. Wilkinson says it plainly of
the rose: the polar bar chart confounds area with radius, and taking the square
root of the radius only partly repairs it. A reader who judges these by area, and
most readers judge filled shapes by area, will read a stronger west wind than the
numbers support.

That is the eighth law working as intended. The engine guarantees a plot is well
formed, never that it is well chosen. Reach for a circle when the variable really
is a circle, which is the case the flat axis genuinely misreports, and reach back
for the flat bar chart when what you need is to compare two counts precisely.

The pie escapes that particular criticism and collects its own. Its slices are
angles rather than radii, so a share of the total really is a share of the ink,
which is why a child can judge one. What angles are bad at is *ranking*: two
slices of 22% and 19% are hard to order in a circle and trivial to order in a
column. Since both come from the same sentence, the honest move when the ranking
is the question is to drop the last atom and read the bar.

## Every mark bends

There is no mark that draws on the page and refuses the circle. The summary
marks were the last to arrive, and they read here exactly as they read flat: a
box is still a five-number summary, a whisker is still a reach, a band is still a
spread. Only the geometry changed.

```{r}
data(winds) + box + x(direction) + y(speed) + polar()
```

Each box is a wedge of the ring between its quartiles, its median an arc across
it, its whiskers running straight out from the center. Compare it with the same
sentence flat and nothing has been added or taken away.

The whisker alone is the sparser reading, and the one to reach for when the
quartiles are more detail than the question needs:

```{r}
data(winds) + interval * range + x(direction) + y(speed) + polar()
```

A band works the same way, and gives the shape that has a name of its own: bind
a low and a high boundary across the compass and `polar()` turns it into the
**radar band**, the filled counterpart of the radar `line` already drew.

```{r}
data(winds) + ribbon * range + x(direction) + y(speed) + polar()
```

A staircase holds each value across its own slot rather than sloping between
them, so bent it is a ring of arcs joined by radial jumps. And because the
categories exhaust the turn, the last one's tread carries round to the first
and the shape closes:

```{r}
data(winds) + step * mean + x(direction) + y(speed) + polar()
```

Look at where the value changes. Every jump is a straight line out from the
center, and every held value is a genuine arc rather than a chord cutting across
it. That distinction is the whole of what these marks needed. A segment that
*holds* a value across a span has to follow the ring. A straight line between
the same two points falls inside the circle, and every point along it would be
at a radius the data never took.

A region is a wedge, so a `zone` shades one. Two categorical axes give the tile
plot bent into a disc, a cell per pair:

```{r}
data(winds) + zone * count + x(direction) + y(season) + polar()
```

The one thing that does not bend is `surface`, and it is not a polar refusal: a
sheet through three positions has no reading in a space with two, which is the
same reason it does not draw on a flat page either. A plot is also drawn in one
space and not two, so asking for `polar()` and `z()` together is refused rather
than one of the two quietly winning.

One mesh is refused, and the reason is worth reading because it is the third
time the same sentence has been the answer:

```{r}
#| error: true
render_svg(data(gm_all) + zone * bin(tiling = "hex") + x(gdp) + y(life) + polar())
```

`bin(tiling = )` says how to partition a **plane**. It is refused on a
one-dimensional bin because an interval is not a plane, and on a categorical axis
because slots are an order rather than a distance. Here it is refused for the
third form of the same problem: a bent plane has no distance for a hexagon to be
regular against, since a step of angle at the rim is a longer step than the same
one near the center. That is precisely the perceptual defect `hex` exists to fix,
arriving from the other side. A rectangle survives the bend, because a bent
rectangle is a sector and that is a shape the engine draws.

Which marks draw in which coordinate space is stated in full, on one page, by the
Mark × Space grid in [Combinations](combinations.qmd). It is generated from the
engine's own table, so it cannot promise a plot the engine will not draw.

## The sunburst: depth is the radius

A **sunburst** draws a tree: one ring per level, each arc as wide as its share of
the whole and sitting inside its parent's. It has no mark of its own and needs
none. An arc of a ring is a rectangle in angle and radius, `zone` is the mark
bounded by a pair on each axis, and this space is what bends a rectangle into a
sector. What was missing until now was the arithmetic between a hierarchy and
those four numbers, and that is `partition`.

The hierarchy arrives as **columns**. `spending` is a household budget with one
row per leaf, and the first three columns spell the path down to it:

```{r}
knitr::kable(head(spending, 6))
```

`NA` is not missing data there. It says the branch stops: rent has no sub-parts
worth naming, so it reaches the second level and no further.

```{r}
data(spending) + zone * partition(group, item, detail) + x(amount) +
  color(group) + polar()
```

`partition(group, item, detail)` names the levels outermost first, so `group` is
the innermost ring and `detail` the rim. `x(amount)` is what each branch is
weighed by. Nothing in the sentence is about drawing, and nothing in it is about
circles.

Bind nothing to `x` and every leaf weighs 1, which is the tally `count` already
does when nothing else measured, so the rings divide by how many leaves each
branch has rather than by what they cost:

```{r}
data(spending) + zone * partition(group, item, detail) + color(group) + polar()
```

Housing fills half the circle above and a third of it here, because it is 1,275 of
the 2,490 spent and 4 of the 12 lines that spend it. That is the same swap the pie
made between `count` and `sum`, arriving at a tree.

Read the outer rim. It is blank across Rent and Groceries and solid across
Utilities and Car, because those two branches are the only ones with a third
level. A ragged rim is what a real hierarchy looks like, and it comes from the
`NA`s rather than from anything the plot was told.

The radial axis carries **depth**, which is the level a node sits at: 1 for the
groups on the inner ring, 2 for the items outside them, 3 for the details on the
rim. The transform computes it and gives it that name, so it is a column you can
bind like any other even though it is not in your table, and `partition` is what
put it there. Naming it is how you say what the radius runs over, and stating a
domain that starts below the first ring is what hollows out the middle:

```{r}
data(spending) + zone * partition(group, item, detail) + x(amount) +
  y(depth, limits = c(0, 4)) + color(group) + polar()
```

That is worth a sentence, because it is not a setting. A hole is a stretch of the
radial axis with nothing standing on it: the innermost ring starts at 1, and
saying the axis runs from 0 leaves the first unit of radius empty. Widen the
stretch and the hole widens with it, since `limits` states a domain and the rings
keep the places they always had — `c(-1, 4)` gives up two fifths of the radius
rather than a quarter.

## The donut: one level is still a tree

A hierarchy one level deep is a hierarchy, and everything above holds for it. One
level is one ring, that ring is at depth 1, and an axis running from 0 to 2 leaves
the first of its two units empty — which is a **donut**:

```{r}
data(spending) + zone * partition(group) + x(amount) +
  y(depth, limits = c(0, 2)) + color(group) + polar()
```

There is no `donut` atom, no `hole =` setting, and nothing in that sentence about
donuts. It is the sunburst's sentence with one level named instead of three, drawn
on a radial axis with one empty unit at the bottom of it.

The published version of this chart writes its names in the ring, and that is the
`text` layer from [Naming the nodes](#naming-the-nodes) with one level instead of
two:

```{r}
data(spending) + x(amount) + y(depth, limits = c(0, 2)) +
  zone * partition(group) + color(group) +
  text * partition(group) + label(name) + style(color = "white") +
  polar() + x_label("")
```

The legend is now saying twice what the ring says once, which is the moment to drop
one of them; the chapter keeps both so that the two ways of decoding a wedge are
side by side.

The pie is the one plot in this space that cannot say it, and the reason is worth
keeping rather than fixing. A [pie](#the-pie-where-you-put-the-category) spends its
measurement on the angle and leaves the radius a constant, so it has no radial
domain to widen; a hole would have to be a number handed to the space, which is the
one thing a hole is not. The sentence that wants a hole is the sentence whose radius
carries something, and that sentence is one atom away.

## The same sentence, unbent

Take `polar()` off the three-level sunburst and change nothing else:

```{r}
data(spending) + zone * partition(group, item, detail) + x(amount) +
  y(depth, limits = c(0, 4)) + color(group)
```

That is an **icicle chart**: depth going up instead of outward, share going
across instead of round. The two are one sentence and one coordinate space apart,
which is this chapter's argument arriving at a chart famous enough to have two
names. A rose and a bar chart are the same pair, and so are a pie and a stacked
bar.

The empty band along the bottom is the hole, and seeing it flat is the clearest
statement of what a hole is. It is not a round thing. It is a stretch of the
depth axis with nothing on it, and bending the picture wraps that stretch into a
disc at the center.

## Naming the nodes

`partition` publishes each node's **center** as well as its four edges, which is
why a second mark can read the same computation. `text` takes the center and
labels the node:

```{r}
data(spending) + x(amount) + y(depth, limits = c(0, 4)) +
  zone * partition(group, item, detail) + color(group) +
  text * partition(group, item) + label(name) + style(color = "white") +
  polar() + x_label("")
```

`label(name)` reads a column the transform wrote, the way `color(count)` reads
a binned tally. And the text layer names *two* levels where the zone names
three, which is how the rim is left unlabeled. A shallower partition of the
same table puts its nodes in exactly the same places, because the arcs are
decided by the leaves underneath them, and those did not change.

One thing about that plot is an honest limit rather than a choice. The names
sit upright where a published sunburst runs them along their arcs, because
`text` places a label at a point in this space and has no rotation to follow
one with.

There is an `x_label("")` in that sentence and no `y_label("")` beside it, and
the asymmetry is the whole of how these plots are labeled. The radial axis
carries a **ring index** the transform invented, and nobody reads a level off
a scale, so it draws no ticks, no rings and no name of its own. That is
decided from what the axis carries rather than asked for, the way a pie is
recognized from its bindings. The measure keeps its guide, because an amount
round a circle is a quantity in the data's own units, and silencing *that* is
what the empty string is for.

## What the grammar can now check

A hierarchy has an invariant that rectangles do not: every child lies inside its
parent, and siblings do not overlap. Handed the four numbers directly the engine
cannot know that, and will draw a confident, wrong sunburst without a word.
Handed the **tree**, it can, and one case is worth meeting because it is a real
ambiguity rather than a mistake:

```{r}
#| error: true
mixed <- data.frame(group = c("Housing", "Housing"),
                    item  = c(NA, "Rent"),
                    amount = c(50, 980))
render_svg(data(mixed) + x(amount) + zone * partition(group, item) + polar())
```

Housing has a number of its own *and* a child with a number, and those two
readings draw different pictures: either its arc is the child's total, or its own
50 sits beside the 980 and widens it. Plotly answers it with a setting called
`branchvalues`. GOG does not choose for you, because the arithmetic belongs to
the accounts rather than to the grammar. Put every number on a leaf and let the
parents be the sums, which is what a partition computes.

The other refusals are the same rule in smaller print. A level must be a **name**
that a branch shares, so a numeric column is refused toward `x()`, where a
measurement belongs. A branch may stop early but may not have a hole in the
middle, since nothing below a stop can be reached. And a mark with no reading for
a region is sent to the two that have one.

What is still outside the grammar is the shape of the input. A hierarchy
written as **columns**, two or three deep, is what this atom reads and what
faceting already understood. A tree of arbitrary depth arriving as a parent
column and a child column is the shape a *network* has. Networks are declined
here deliberately rather than by omission, so that a famous chart does not
decide that question by walking in through the side.