# Polar

```{r}
#| include: false
source("R/setup.R")
```

Which way does the wind blow most often? Some questions do not fit a flat page.
Every plot so far has been drawn on one: an `x` running across, a `y` running
up. That is one coordinate space, `flat`, and the kernel names six more. The
flat space is also the wrong shape for a whole class of variables. Compass
direction, the hour of the day, the month of the year: in each of them the last
value is the neighbor of the first. A straight axis has no way to show that.

Two of the oldest statistical graphics were drawn in a circle instead, in the
space gog calls `polar`.
Playfair's *Statistical Breviary* of 1801 cut a circle into sectors to show the
Turkish empire's territory, and that is the first pie chart [@playfair1801].
Florence Nightingale's 1858 diagram of what killed British soldiers in the
Crimea [@nightingale1858] is a bar chart in polar coordinates. It has three
names: a **coxcomb**, a **rose**, or, most precisely, a **polar area diagram**.
That third name is the only one that says what the wedges mean. Her wedges put
each month's death rate into the wedge's *area*, so their radius grows as the
square root of the rate. In gog, a polar bar maps the value directly to the
radius. Both are correct drawings, but they are not the same drawing, so check
which one you have before you compare them. The section [Nightingale's
rose](#nightingales-rose), later in this chapter, draws a rose from her own
numbers.

The line version has even more names: **radar** plot, **spider** plot, **star**
plot, and **Kiviat** diagram in engineering. As with the rose and the pie, none
of them is an atom. Each is a mark and a channel or two with `polar()` at the
end, which is what the rest of this chapter is about.

## A circular variable on a straight axis

How often the wind blew from each direction is a count per compass point, so
start with the plot you know. It shows what a straight axis loses. Here are wind
observations from two seasons, summer and winter, drawn as an ordinary bar chart
with one bar per compass point:

```{r}
#| echo: false
peek(winds)
```

```{r}
data(winds) + bar * count + x(direction)
```

*"Given the winds: bars derived by count, x is direction."*

Read the two ends. North is at the far left and north-west at the far right,
with the whole compass between them, although the two directions are 45 degrees
apart in the world. The chart has cut the circle open and flattened it, and the
cut is invisible: nothing in the picture says that the two ends belong together.
A reader who did not already know the compass could not tell from this chart
that north-west is next to north.

Now add one atom to the same sentence:

```{r}
data(winds) + bar * count + x(direction) + polar()
```

*"Given the winds: bars derived by count, x is direction, in polar."*

Same mark, same transform, same binding, same data. The only thing that changed
is the space the plot is drawn in, and north-west is beside north because on the
compass it is beside north. The west wind, which blew most often, is now a long
wedge pointing west, rather than a tall bar you find by reading the labels.

This is what a coordinate space is for, and it is why a pie chart, a rose and a
radar plot are not chart types in this grammar. They are marks you already know,
read in a circle.

## What `polar()` changes and what it keeps

One word changed the whole picture, so the next thing to settle is which of the
two positions became the angle. `x` becomes the **angle** and `y` becomes the
**radius**. That order is Wilkinson's [@wilkinson2005]. His reason is that the
first position says where a mark stands, and in a circle that is an angle. The
second says how much, and that becomes how far out from the center the mark
reaches. (That is the rule when both positions are
bound. Bind only one and there is nothing to choose, so the one you have becomes
the angle: that is the pie, and it has its own section below.)

So `x(direction)` still says what it always said, that direction is the position
each bar stands at. `polar()` changes one thing only: how positions are read.
`x` now gives a direction around the circle, and `y` a distance from the center.
A bar still runs from its baseline to its value. The baseline is the center of
the circle, so the count sets how far out each wedge reaches.

The next question is whether the rest of the grammar still works in a circle: a
split by season, for example, with the two seasons side by side. Nothing else in
the grammar knows the space changed, so nothing else has to be rewritten for it:

```{r}
data(winds) + bar * count * dodge + x(direction) + color(season) + polar()
```

*"Given the winds: bars derived by count and dodge, x is direction, color by
season, in polar."*

`color` splits the count by season, and `dodge` sets the two seasons' bars side
by side inside each slot instead of on top of each other. The legend decodes the
colors. Every one of those works in a circle exactly as it works on a flat page,
because none of them was ever written in terms of pixels.

Change `dodge` to `stack` and the seasons pile outward from the center instead,
one season outside the other:

```{r}
data(winds) + bar * count * stack + x(direction) + color(season) + polar()
```

*"Given the winds: bars derived by count and stack, x is direction, color by
season, in polar."*

Summer sits against the center and winter is stacked outside it, so each wedge
now reaches the total of both seasons. West runs out to 63. In the dodged rose
above, west stopped at 38, the longer of its two bars. The radial axis now
reaches 63 rather than 38. That is why `stack` is a transform and not a drawing
option: a transform runs before the scale is set, so the scale can read the
totals `stack` adds up.

Read the pair together, because each hides what the other shows. Stacked, each
wedge shows the total wind from its direction, and west is plainly the most
common direction. Dodged, the seasons stand side by side, and you can see that
the south-west wind comes mostly in summer (30 against 17), and the west wind
mostly in winter (38 against 25). Neither is the correct rose. The rose is not a
chart with its own options at all. It is `bar`, with its axis bent.

[`play`](play.qmd) also works in a circle, with no new atom. The animation
chooses the rows for each frame, and `polar()` bends those rows into a circle,
so neither step changes the other. The Play chapter's "Frames in polar"
animates a rose and two pies across half a century.

## Where the circle starts

The angular axis begins at the top and runs clockwise, which is how a clock and
a compass are both read. The middle of the first category's slot sits at the
top, not the edge of the slot. That is why north points straight up in the rose
above: north is the first of the eight directions.

The `day_cycle` table holds one day of trips, hour by hour:

```{r}
#| echo: false
peek(day_cycle)
```

The axis starts at the top, so midnight sits there. A working day reads better
with noon at the top. `polar(start = )` turns the whole space by a number of
degrees:

```{r}
data(day_cycle) + line + x(hour) + y(trips) + polar(start = -180)
```

*"Given the day cycle: a line, x is hour, y is trips, in polar, starting at
-180."*

Those are the same trips, turned half a circle. Noon is at the top now and
midnight at the bottom, which is the arrangement you want if you are reading the
working day rather than the day from midnight to midnight. `start` is a property
of the space rather than of the data, the polar counterpart of the
`space(turn =, tilt =)` that sets the angle a 3-D plot is viewed from (see
[Space](space.qmd)). It changes how you look at the plot, never what is drawn.

On a categorical angle, `start` points at the first category rather than at the
scale's origin, and the reason is in how a categorical scale is built. A
categorical scale runs from half a slot before the first category to half a slot
after the last, so the scale's own origin is padding rather than any place in
the data. A start angle aimed at the padding would put north at 22.5 degrees, so
you would have to write `-180/8` to correct it. A flat categorical axis already
puts its tick at the category's center, so the circle agrees with it: `start`
points at the category, not at the gap before it.

## One turn, whatever the data spans

The angular axis is **periodic**: one full turn is the whole of the axis, so its
first value and its last value land on the same spoke, the same line out from
the center. That is what closes the circle without a gap or an empty wedge, and
it has one consequence: the circle needs the whole cycle, or the two ends will
meet in the wrong place.

The day of trips, turned half a circle above, closes, and the reason is in the
table's first and last rows. It runs from hour 0 to hour 24 with *both*
endpoints present, because midnight is where the day ends and also where it
begins:

```{r}
data(day_cycle) + line + x(hour) + y(trips) + polar()
```

The curve closes at the top because the reading at 24 is the reading at 0, and
the two sit on the same spoke. That is the first of three cases, and in one of
them the circle does not close. Most periodic data does not include both ends. A
tide gauge reads every three hours and misses both ends of the day, hour 0 and
hour 24. Its axis spans 1 to 22, and the failure is easy to see:

```{r}
#| echo: false
peek(tide)
```

```{r}
data(tide) + line + x(hour) + y(height) + polar()
```

One turn covers 1 to 22, so hour 1 and hour 22 land on the *same spoke*. The
first and last readings are drawn on top of each other, three hours apart in the
world and zero degrees apart on the page. Every reading between them sits at the
wrong angle too, spread over 21 hours instead of 24.

The axis cannot know how long your cycle is: nothing in a column of hours says
that a day has 24 of them. So say it:

```{r}
data(tide) + line + x(hour, limits = c(0, 24)) + y(height) + polar()
```

*"Given the tide: a line, x is hour from 0 to 24, y is height, in polar."*

Now each reading sits where a 24-hour clock would put it. Hour 1 is a
twenty-fourth of a turn past midnight, and hour 22 is a twelfth of a turn before
it. The two are 45 degrees apart, which matches the three hours between them.

The curve still does not close, which is the second case, and that is the
correct picture. The gauge took no reading at midnight, and `limits` states a
domain; it does not invent an observation. The gap you can see between hour 22
and hour 1 *is* the three-hour gap in the data, drawn at the size it is.
Contrast the trips above, which close because a reading at 24 genuinely exists.
A categorical angle is the third case, and it closes for a different reason.
Every slot is filled there, so the last category is next to the first whatever
the data holds.

`limits` states the domain the axis runs over rather than deriving it from the
data. That is the general fix and not a polar one. The same argument, `limits`,
keeps one color ramp fixed across plots built from different tables, which
[Scales](scales.qmd) explains in full. Note what `limits` did not do. No reading
was dropped, because 1 through 22 all lie inside 0 through 24. Stating a domain
widens an axis as easily as it narrows one.

Fill the day of trips with `area`. Its baseline is zero, which in a circle is
the center, so the area fills everything between the curve and the center:

```{r}
data(day_cycle) + area + x(hour) + y(trips) + polar()
```

*"Given the day cycle: an area, x is hour, y is trips, in polar."*

## The radar: a categorical angle

A categorical angle closes differently. An hour axis closes only when the table
holds both ends of the cycle, 0 and 24, which the scale puts on one spoke.
Categories cannot do that: each appears exactly once, by definition, so there is
no repeated endpoint to land on. Instead the categories together fill the whole
turn, which makes the last one genuinely next to the first. So the engine joins
the last category back to the first, and a line across categories closes:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + line * mean + x(continent) + y(life) + polar() +
  title("line * mean + x(continent) + polar()")
```

*"Given gapminder 2007: a line derived by mean, x is continent, y is life, in
polar."*

That is a radar plot, and there is no `radar` atom in this grammar. The same
sentence without `polar()` is the profile from the [Line](marks/line.qmd)
chapter, minus its point layer. There the axis is straight, so the line has two
ends. Bending the space is what joins them. Filled, it is the shape people
usually mean by the name:

```{r}
data(gapminder_2007) + area * mean + x(continent) + y(life) + polar() +
  title("area * mean: the filled radar")
```

Split it with `color` and each group is its own outline, and the split reads
exactly as it does on a flat profile:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + line * mean + x(continent) + y(life) + color(era) + polar() +
  title("half a century apart, two rings")
```

The 2007 outline lies outside the 1957 one for every continent, and the gap is
widest at Asia, about 21 years.

The radar shares a problem with the rose, explained under
[What a circle makes harder to read](#what-a-circle-makes-harder-to-read): the
eye reads the area the outline encloses, and no number in the table states that
area. That area also changes when the categories are put in a different order.
Use it when the categories really are a closed set with no first or last member,
which is the case a straight axis misreports by having to pick an order. When
the question is which continent is highest, the flat bar chart of the same
sentence answers it precisely, and is two atoms away.

## An angle that is measured, not named

An angular position does not have to be a category. The eight compass points are
names, but an instrument records the same directions as degrees, with a speed
beside each reading. `bearing` is the same wind observation recorded in degrees,
a continuous position that happens to run around a circle:

```{r}
data(winds) + point + x(bearing) + y(speed) + color(season) + polar()
```

*"Given the winds: points, x is bearing, y is speed, color by season, in
polar."*

Each dot sits at its own bearing. Its distance from the center is its speed. No
two readings share a bearing, so the dots fill the circle rather than standing
in eight slots. They crowd on the western side, where most of the wind came
from. The gridline circles mark speed, read outward from the center. A dot near
the rim is a fast wind, whichever way it was blowing.

`bin` cuts a measured angle into wedges the same way it cuts a flat axis into
bars:

```{r}
data(winds) + bar * bin + x(bearing) + polar()
```

*"Given the winds: bars derived by bin, x is bearing, in polar."*

That is a wind rose, one atom different from a histogram: `bar * bin` in both
cases, read on a line or read in a circle. The name comes from the *rosa
ventorum*, Latin for wind rose, which cartographers drew on maps from the 13th
century to mark the directions of the winds. The bins touch here as they touch
there, because a histogram cuts a continuous axis into adjacent intervals, and
that is true whichever shape the axis has been bent into.

A compass turn is 360 degrees, but the axis cannot know that, just as it could
not know the length of a day. The axis is fitted to what was observed, so the
turn is the span of the *bins*, from 2.6 to 359.7 degrees. The readings here
cover nearly the whole compass, so north lands only 2.6 degrees counterclockwise
of the top.

Stating the domain with `limits`, as with the tide readings, puts north at the
top:

```{r}
data(winds) + bar * bin + x(bearing, limits = c(0, 360)) + polar()
```

*"Given the winds: bars derived by bin, x is bearing from 0 to 360, in polar."*

The turn now runs from 0 to 360 degrees, whatever the readings span. `bin` cuts
its wedges on the stated domain as well. It makes ten wedges, as above, so each
covers 36 degrees and the first starts at north. The label 0 at the top, missing
from the rose above, marks where that wedge begins.

## The pie: where you put the category

The roses so far put the direction on a *position*: `x(direction)` gave each
compass point its own slot, and the count decided how far out the wedge reached.
A pie does the opposite. Its categories are not positions at all; they are a
split, and what goes around the circle is the measure itself.

That is the whole difference, and the grammar can say it. Take the category off
`x` and give it to `color`, then say that the pieces lie end to end:

```{r}
data(winds) + bar * count * stack + color(direction) + polar()
```

Read the sentence: a bar, counted, stacked, colored by direction, in polar. Not
one word of it is about pies. The slices are the segments of a stacked bar, and
the angles are their shares. The legend says which direction each slice is.

The flat form is the same sentence without the last atom, and it is a real chart
of its own, the one column that shows how a total divides:

```{r}
data(winds) + bar * count * stack + color(direction)
```

Those two pictures carry identical numbers. Wilkinson's second chapter is called
*How To Make a Pie*, and his answer is exactly this: a pie chart is a stacked
bar in polar coordinates. The column's height is the pie's full turn, and each
segment's share of the height is its slice's share of the circle.

Nothing new was added to the grammar to reach it. There is no `pie()` atom, no
`type = "pie"`, and no argument anywhere that says which chart to draw. Two
things happen on their own. First, a bar split only by color stands in one
slot, so it needs no position axis. Dropping `x` is how you say that. Second, a
plot with one position bound has nothing to choose between: that position
becomes the angle, and the radius is left as the constant that sets the pie's
size. Wilkinson's book has both rules: its one-argument polar function is a
separate tool from the two-argument one.

The measure can be any single number that summarizes a group, not just a count.
Swap `count` for `sum`, name the column it should total, and the same eight
directions are weighted by wind speed rather than by how often the wind blew:

```{r}
data(winds) + bar * sum * stack + y(speed) + color(direction) + polar()
```

*"Given the winds: bars derived by sum and stack, y is speed, color by
direction, in polar."*

Compare it with the pie above: the same categories, a different answer. West is
both the most common direction and the strongest, so its slice widens from about
a quarter of the circle to a third. Winds from the north and south are the
gentlest, so theirs shrink by roughly two fifths, even though exactly the same
observations went into them. Swapping one atom changed which question the circle
answers.

Swap `count` for `proportion` instead, and the angles stay the same, because in
a pie each slice's angle is already its share of the whole. In the pie, only the
axis name changes. The flat stacked bar is where the change shows, because its
axis now runs from 0 to 1:

```{r}
data(winds) + bar * proportion * stack + color(direction)
```

*"Given the winds: bars derived by proportion and stack, color by direction."*

Each segment's height is now its share. West runs from 0.66 to 0.90, a share of
0.24, which is the quarter of the circle its slice takes in the count pie.

The split has to be there. A stack with nothing to pile, or a bar with neither a
position nor a split, is refused rather than guessed at. So is a statistic that
only means something *along* an axis:

```{r}
#| error: true
data(winds) + bar * bin * stack + color(season) + polar()
```

## Nightingale's rose

The start of this chapter named Florence Nightingale's diagram of the deaths in
the British army during the Crimean War. The numbers behind it are in the
`nightingale` table: the deaths in each month from April 1854 to March 1856,
from three causes:

```{r}
#| echo: false
peek(nightingale)
```

Each row is one month and one cause, and `deaths` counts the soldiers who died
of it. The cause `Disease` means infectious diseases, which she called
preventable or mitigable. `Wounds` means wounds, and `Other` every other cause.
`period` says which year of the war a month belongs to. Each of her years runs
from April to March, so the second year begins in April 1855. A sanitary
commission arrived in March 1855 to improve the hygiene of the camps and the
hospitals. Her two years compare the deaths before it and after it.

```{r}
data(nightingale) + bar * stack + x(month) + y(deaths) + color(cause) +
  palette(c(Disease = "lightsteelblue", Wounds = "lightcoral",
            Other = "dimgray")) +
  polar(start = -90) +
  title("Deaths in the British army in the Crimean War, by cause") |
  facet(period)
```

*"Given the nightingale table: bars derived by stack, x is month, y is deaths,
color by cause, with a palette naming Disease lightsteelblue, Wounds lightcoral
and Other dimgray, in polar, starting at -90, split into panel columns by
period."*

Each wedge is one month, and its length from the center is the number of
soldiers who died that month. The blue segment shows the deaths from disease,
and the pink and gray segments outside it show the deaths from wounds and from
other causes. `polar(start = -90)` turns the circle 90 degrees counterclockwise,
so April, the first month, sits on the left, where her diagrams start. Two atoms
are new to this chapter. The
[palette](style.qmd#palette-choosing-the-colors-a-mapping-uses) gives each cause
a lighter form of her blue, red and black, and `| facet(period)` draws one panel
for each year, earliest first ([Faceting](faceting.qmd)).

The first winter has the longest wedges. In January 1855, 3,168 soldiers died:
2,761 of disease, 83 of wounds and 324 of other causes. A year later, January
1856 is a small wedge near the center: 92 deaths from all three causes together.

Her diagram differs from this one in four ways. She put her first year on the
right. She drew the three causes on top of each other, each measured from the
center, while `stack` piles them, so each wedge here reaches the month's total.
Her wedges show the yearly death rate per 1,000 soldiers, which allows for an
army that grew from about 8,600 men to over 40,000. And her radius grows as the
square root of that rate, so a wedge's area is in proportion to it. Here the
radius is the number of deaths itself, as for every polar bar in gog. So the
first winter looks larger, beside the other months, than it does in her
diagram. The next section explains why that matters.

## What a circle makes harder to read

The grammar draws every rose in this chapter, and drawing one is not the same as
recommending it. A wedge's *area* grows with the square of its radius, so a
count that doubles gets four times the area, not twice. Wilkinson says it
plainly of the rose: the polar bar chart confounds area with
radius, and growing the radius as the square root only partly repairs it. Most
readers judge filled shapes by area. A reader who does that will read a stronger
west wind than the numbers support.

[Law 8](design-laws.qmd#law-8-pronounceable-usable) states the limit: the engine
guarantees a plot is well formed, never that it is well chosen. Use a circle
when the variable really is a cycle, which is the case a flat axis misreports.
Use the flat bar chart when the comparison must be precise.

The pie avoids that criticism and has one of its own. Its slices are angles
rather than radii, so a share of the total really is the same share of the
circle, and readers judge that share well. Angles are poor for *ranking*: the
east and south-west slices of the count pie, 16% and 18%, are hard to order in
the circle and easier to order as lengths in the stacked bar. When ranking is
the question, drop the last atom and read the bar. Both pictures come from the
same sentence.

## The marks in a circle

The rose counted observations per direction. How fast the wind blew from each
direction, and how much that varied, is a question for the marks that
summarize: `box`, `interval` and `ribbon`. Every mark that draws on a flat page
also draws in a circle. These three read here exactly as they read flat: a box
still shows the quartiles and the median, a whisker still shows how far the
values reach, a band is still a spread. Only the geometry changed.

```{r}
data(winds) + box + x(direction) + y(speed) + polar()
```

*"Given the winds: boxes, x is direction, y is speed, in polar."*

Each box is a piece of a ring, from the lower quartile out to the upper. Its
median is an arc across it, and its whiskers lie along lines through the center.
Drawn flat, the same sentence gives boxes with the same parts.

An `interval` shows less: one line from the lowest to the highest value, with no
box. It is the one to use when the quartiles are more detail than the question
needs:

```{r}
data(winds) + interval * range + x(direction) + y(speed) + polar()
```

*"Given the winds: intervals derived by range, x is direction, y is speed, in
polar."*

A ribbon works the same way. `range` gives each compass point a low and a high
value, and `polar()` turns the ribbon into a **radar band**, the filled
counterpart of the radar that `line` drew.

```{r}
data(winds) + ribbon * range + x(direction) + y(speed) + polar()
```

*"Given the winds: a ribbon derived by range, x is direction, y is speed, in
polar."*

A staircase holds each value from its own category around to the next rather
than sloping between them, so in a circle it is a series of arcs joined by
straight jumps. And because the categories fill the whole turn, the last arc
carries all the way around to the spoke the first one started from:

```{r}
data(winds) + step * mean + x(direction) + y(speed) + polar()
```

*"Given the winds: a step outline derived by mean, x is direction, y is speed,
in polar."*

Look at where the value changes. Every jump is a straight line out from the
center, and every held value is a genuine arc rather than a chord cutting across
it. That distinction is the only thing the marks in this section needed in order
to bend. A segment that *holds* a value across a span has to follow the circle
at its radius. A straight line between the same two points falls inside the
circle, and every point along it would be at a radius no reading has.

A `zone` shades a region, and in this space a region is a wedge. Put a category
on each axis and a tile plot bends into a disc, with one cell for every pair:

```{r}
data(winds) + zone * count + x(direction) + y(season) + polar()
```

*"Given the winds: zones derived by count, x is direction, y is season, in
polar."*

Two marks do not draw in a circle, and in neither case is the circle the reason.
A `surface` needs three positions, so it needs the cube, the 3-D space that
[Space](space.qmd) describes; it does not draw on a flat page either. An `edge`
needs the [network](network.qmd) space, which places its nodes. A plot is also
drawn in one space and not two, so asking for `polar()` and `z()`
together is refused rather than one of the two being ignored.

One sentence is refused here, because a hexagon needs a plane where the same
step is the same distance everywhere:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
#| error: true
data(gm_all) + zone * bin(tiling = "hex") + x(gdp) + y(life) + polar()
```

`bin(tiling = )` says how to divide a *plane* into cells. It is refused on a
one-dimensional bin because an interval is not a plane, and on a categorical
axis because slots are an order rather than a distance. Here it is refused
because a bent plane is not such a plane either. A hexagon exists to keep every
cell the same size and shape. A bent plane has no fixed distance to measure that
against, because a step of angle at the rim is longer than the same step near
the center. A rectangle still works after the bend, because a bent rectangle is
a sector and that is a shape the engine draws.

Which marks draw in which coordinate space is stated in full, on one page, by
the Mark × Space grid in [Combinations](combinations.qmd). It is generated from
the engine's own table, so it cannot promise a plot the engine will not draw.

## The sunburst: depth is the radius

A pie divides one whole into parts. The rest of this chapter is about dividing
each part again. Where does a household's money go, and inside each group, which
items cost the most? That is a tree of shares, one level inside another. A
**sunburst** draws a tree: one ring per level, and each piece is as wide as its
share of the whole. A child sits one ring further out, within its parent's
angle. The sunburst has no mark of its own and needs none. A piece of a ring is
a rectangle in angle and radius. `zone` is the mark bounded by a pair of values
on each axis. This space is what bends that rectangle into a sector.
[`partition`](transforms.qmd#partition-a-whole-among-nested-parts) is what turns
a hierarchy into those two pairs of values, one rectangle per node, meaning
every group, item and detail in the tree.

The hierarchy is stored as **columns**. `spending` is a household budget with
one row per leaf, meaning the last level of a branch, and the first three
columns hold the path down to it:

```{r}
#| echo: false
peek(spending, n = 6)
```

In this table, a missing value does not mean an unknown value. It means the
branch stops: Rent has no sub-parts, so it reaches the second level and no
further.

```{r}
data(spending) + zone * partition(group, item, detail) + x(amount) +
  color(group) + polar()
```

*"Given the spending table: zones derived by partition through group, item and
detail, x is amount, color by group, in polar."*

`partition(group, item, detail)` names the levels from the top of the tree down,
so `group` is the innermost ring and `detail` the rim. `x(amount)` is the column
that weights each branch: a branch's share of the circle is its share of the
total amount. Apart from `polar()`, nothing in the sentence is about circles,
and nothing in it names a sunburst.

Bind nothing to `x` and every leaf weighs 1, the same weight `count` gives every
row. The rings are then split by how many leaves each branch has rather than by
what they cost:

```{r}
data(spending) + zone * partition(group, item, detail) + color(group) + polar()
```

Housing fills half the circle above and a third of it here, because it holds
1,275 of the 2,490 spent, and 4 of the 12 leaves. That is the same swap the pie
made between `count` and `sum`, this time on a tree.

Read the outer rim. It is solid outside Utilities and Car and blank outside
every other item, because those two are the only items with a third level. A
ragged rim is what a real hierarchy looks like, and it comes from the missing
values rather than from anything in the sentence.

The radial axis carries **depth**, which says which ring a node sits on. The
groups fill the ring from 1 to 2, the items the ring from 2 to 3, and the
details the ring from 3 to 4. `partition` computes depth and writes it under
that name, so it is a column you can bind like any other, even though your
table never held it. Binding it to `y` is how you say what the radius runs over,
and stating a domain that starts below the first ring is what hollows out the
middle:

```{r}
data(spending) + zone * partition(group, item, detail) + x(amount) +
  y(depth, limits = c(0, 4)) + color(group) + polar()
```

The hole is not a setting, and the reason matters. A hole is a stretch of the
radial axis with nothing standing on it: the innermost ring starts at 1, and
saying the axis runs from 0 leaves the first unit of radius empty. Widen the
stretch and the hole widens with it, because `limits` states a domain and the
rings keep the places they always had. With `c(-1, 4)` the hole takes two fifths
of the radius rather than a quarter.

## The donut: one level is still a tree

A hierarchy one level deep is still a hierarchy, and everything above is true of
it as well. A **donut** is the hollow sunburst with one level. One level is one
ring, that ring runs from 1 to 2, and an axis running from 0 to 2 leaves the
first of its two units empty:

```{r}
data(spending) + zone * partition(group) + x(amount) +
  y(depth, limits = c(0, 2)) + color(group) + polar()
```

There is no `donut` atom, no `hole =` setting, and nothing in that sentence
about donuts. It is the sunburst's sentence with one level named instead of
three, drawn on a radial axis whose first unit is empty.

The donut you usually see published writes its names in the ring, and that is
the `text` layer explained under [Naming the nodes](#naming-the-nodes), two
sections later, with one level instead of two. In the sentence below,
`label(name)` reads a column that `partition` writes, one of the [columns a
transform writes](transforms.qmd#a-name-the-transform-sets):

```{r}
data(spending) + x(amount) + y(depth, limits = c(0, 2)) +
  zone * partition(group) + color(group) +
  text * partition(group) + label(name) + style(color = "white") +
  polar() + x_label("")
```

*"Given the spending table: x is amount, y is depth from 0 to 2, zones derived
by partition through group, colored by group, and also text derived by partition
through group, label by name, in polar."*

The group names are now written in two places, in the legend and in the ring, so
a published plot would keep only one. Both are kept here, so that the two ways
of decoding a wedge stand side by side.

The pie is the one plot in this space that cannot have a hole, and that follows
from what a pie binds. A [pie](#the-pie-where-you-put-the-category) puts its
measurement on the angle and leaves the radius constant, so it has no radial
domain to widen. No setting adds a hole to a pie. A hole appears only
where the radial axis runs below the innermost ring, so only a plot that binds
the radius can have one. The donut binds it with `y(depth, limits = c(0, 2))`.

## The same sentence on a flat page

This chapter has said that the circle is one word, `polar()`. If that is true,
taking the word away should leave a readable plot. Take `polar()` off the
three-level sunburst and change nothing else:

```{r}
data(spending) + zone * partition(group, item, detail) + x(amount) +
  y(depth, limits = c(0, 4)) + color(group)
```

That is an **icicle chart**: depth going up instead of outward, share going
across instead of around. The two pictures are one coordinate space apart, and
that space is the only difference between two charts with two names. A rose and
a bar chart differ in the same way, and so do a pie and a stacked bar.

The empty band along the bottom is the hole, and seeing it flat shows most
clearly what a hole is. A hole is not a shape. It is a stretch of the depth
axis with nothing on it, and bending the picture wraps that stretch into a disc
at the center.

## Naming the nodes

Go back to the three-level sunburst. The legend decodes the groups and nothing
inside them, so the items on the middle ring are unnamed. `partition` writes the
**center** of each node, the middle of its own region, as well as its four
edges. That is why a second mark can read the same computation. `text` takes
that center and labels the node:

```{r}
data(spending) + x(amount) + y(depth, limits = c(0, 4)) +
  zone * partition(group, item, detail) + color(group) +
  text * partition(group, item) + label(name) + style(color = "white") +
  polar() + x_label("")
```

`label(name)` reads a column the transform wrote. The text layer names *two*
levels where the zone draws three. The third level gets no label, which is why
the rim is bare. A shallower partition of the same table puts its nodes in
exactly the same places, because each node's width is decided by the leaves
underneath it, and those did not change.

One thing about that plot is a limit of the engine rather than a choice. A
published sunburst runs its names along their arcs. Here they sit upright,
because `text` places a label at a point and cannot turn it to follow an arc.

That sentence carries an `x_label("")` and no `y_label("")` beside it, because
only one of the two axes has a label to remove. The radial axis carries the
depth the transform computed. Nobody needs a scale to tell the first ring from
the second, so that axis draws no ticks, no gridlines and no name of its own,
and nothing in the sentence has to ask for that. The engine reads what the axis
carries. It never knows that the plot is a sunburst, in the same way that it
never knows a pie is a pie: in both, what is drawn follows from what the
sentence binds. The amount around the circle keeps its ticks and its name,
because it is a quantity in the data's own units. `x_label("")` removes the name
and leaves the ticks. To leave out both, drop `x_label("")` and write
`x(amount, axis = FALSE)`, as [Scales](scales.qmd#axis-leaving-an-axis-out)
explains. gog refuses the two together, because the name is part of the axis.

## What a tree lets the grammar check

A hierarchy has two rules that rectangles do not: every child lies inside its
parent, and children of one parent do not overlap. Give the engine four edge
numbers per node and it cannot check either rule. It will draw a wrong sunburst
and say nothing. Give it the *tree* and it can check both. One case is shown
here because it is a real ambiguity rather than a mistake:

```{r}
#| error: true
mixed <- data.frame(group = c("Housing", "Housing"),
                    item  = c(NA, "Rent"),
                    amount = c(50, 980))
data(mixed) + x(amount) + zone * partition(group, item) + polar()
```

Housing has a number of its own *and* a child with a number. That can be read
two ways, and the two draw different pictures: either Housing is as wide as its
child's 980, or Housing's own 50 is added to the 980 and Housing is wider.
Plotly answers it with a setting called `branchvalues`. gog does not choose for
you, because that sum is a decision about your data, not about the plot. Put
every number on a leaf and let the parents be the sums, which is what a
partition computes.

Three other refusals are smaller. A level must be a **name** that a branch
shares, so a continuous column is refused and sent to `x()`, where a
measurement belongs. A branch may stop early but may not skip a level: a row
with a detail and no item is refused, because the detail has no item to sit
inside. And a mark that cannot read a region, such as `point` or `bar`, is
refused. The message sends you to `zone`, which draws the region, or to `text`,
which labels its center.

`partition` deliberately does not read a tree stored as a parent column and a
child column. It reads a hierarchy written as columns, one per level, which is
what faceting reads too. A parent column and a child column can describe a tree
of any depth, and that is the shape a *network* has. [Network](network.qmd) is
where that shape is read: `layout(parent, child)` places it, in a space with its
own rules. Each transform reads the shape it is built for.

## In short

`x` becomes the angle and `y` the radius, because the first position says where
and the second says how much. Bind only one position and it becomes the angle: a
stacked bar then draws a pie. A categorical angle closes, so the last category
sits beside the first. A cycle whose table lacks its ends needs `limits` to fill
the turn. `partition` draws a hierarchy as rings of depth, and a domain below
the first ring makes the hole.

## What comes next

Polar bends the axes, and it still has axes. [Nest](nest.qmd) has none: it packs
a panel into regions whose areas are the measure, which is how a treemap shows a
share of a whole.
