# Reading practice

```{r}
#| include: false
source("R/setup.R")
```

Can you say what a plot will look like from its specification alone? You have
the vocabulary; now read with it. Ten plots follow, ending in a cube. For each
one: read the specification aloud before looking at the picture, then check the
picture against what you said. The English sentence is given under each plot.
Try to say it yourself first.

## Gold medals by country

Which country won the most gold, and by how much? `medals` holds one row per
country, with its counts of gold, silver and bronze:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold)
```

*"Given the medals: bars, x is country, y is gold."* A bar measures an
amount from zero, so the amount needs a continuous axis and the category takes
the other one. The bindings put countries on `x` and gold on `y`. There is
no transform here. `bar` draws the numbers exactly as the table gives them.

## The same bars, horizontal

A medal table is usually printed as a list, with the leader at the top. The
same five counts can be read that way.

```{r}
data(medals) + bar + x(gold) + y(country)
```

*"Given the medals: bars, x is gold, y is country."* Swap which axis carries
the category and the bars become horizontal. There is no "horizontal bar" atom,
and no argument that turns the bars. The sentence already says which axis holds
the category, so the orientation follows from the bindings.

## One measurement's shape

The first two plots drew five counts that were already in the table. Often a
table holds only measurements, and no counts.

Do petal lengths gather around one typical value, or do they separate into
groups? One measurement answers that, and it comes from a new table.
`iris_flowers` holds one row per flower: three measurements, and the species
the flower belongs to.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

*"Given the iris flowers: bars derived by bin, x is petal length."* The `*`
derives the counts: `bin` cuts the axis into bins and counts the rows in each,
and `bar` draws them. You did not bind `y`. `bin` invents the count itself, so
the `y` axis is labeled `Count`. Note the two
peaks. Plot eight explains them.

## The same question, smoothly

A histogram draws the shape as flat steps, one per bin, and sometimes you want
one smooth curve instead. The measurement this time is life
expectancy, across the 142 countries.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"Given gapminder 2007: a line derived by density, x is life."* The question is
the same as the histogram's: how is this measurement distributed? The answer is
drawn differently. `density` replaces `bin`, and `line` replaces `bar`, because
a smooth estimate is one curve, not many bars.

## One strip of dots per continent

The density curve in plot four reduced 142 countries to one shape. Do the
continents differ in life expectancy, and how widely do the countries
inside each one vary? Both answers are visible when every country is shown.

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Given gapminder 2007: points, x is continent, y is life."* A categorical `x`
gives each category a slot on the axis, and that category's points sit in a
strip above the slot. The same rule placed bars on country slots in plot one.
Here it applies to a different mark. A plot shaped like this one has a name: the
**strip plot**.

## The typical value per continent

You often want one number per continent, its average, instead of 142 separate
values.

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*
Each bar is one continent's average. Here `y(life)` names
the column that `mean` averages, because a transform that reads a column must
be told which column.

## Five countries across the decades

Plots four to six show only 2007. A question about change needs many years,
so the next plot uses another table.

How has life expectancy changed since 1952, and did five Asian countries
change at the same rate? `gapminder_asia` holds one row per country *and
year*, so each country appears many times.

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*
On a `line`, a categorical `color` first *splits* the rows into one line
per category, then colors them. One channel does two things, and you write it
once.

## Three species on one axis

Plot three showed two peaks in petal length, and the table has three species.
How do three species make two peaks?

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."* The left peak is setosa alone. The right peak is versicolor and
virginica together, because their petal lengths overlap. All three histograms
use the same bin edges, and each one is drawn where its own species' values
fall. Each fill is translucent under a solid outline, so every histogram stays
visible. `color` splits the rows here into one histogram per species,
by the same rule as in plot seven.

## One panel per continent

Richer countries live longer across the world as a whole, as the first chapter
showed. Is that also true inside each continent?

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split into
panel columns by continent."* Inside each continent, life expectancy also rises
with income. The rise is clearest in Europe and Asia and least clear in Africa,
where the points are most scattered. The `|` operator splits the plot into panel
columns, one panel per category. All the panels share one pair of scales, so
they can be compared. As in the first chapter, `scale = "log"` spreads out the
countries that would otherwise crowd against the left edge.

## Three measurements on three axes

Petal length alone left two species overlapping in plot eight. Two more
measurements might separate them, if all three measurements can be seen at once.

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is petal
length, color by species."* The third dimension is not a different package, and not a
different mark. It is one more position channel, and the same specification
without `z(petal_length)` would draw a flat scatter. In the web edition the cube
turns with a drag: drag right and the side facing you moves right, drag down and
it tilts. Every angle hides some points and shows others, so turn the cube more
than once.

## What you just did

Those ten plots used twelve distinct words, and every plot reads left to right:
*data, mark, positions, refinements*. A transform joined the mark and a facet
came last, exactly where the first chapter put them. The same `color` split the bars
and lines, and colored the points. The same `*` derived counts, means and
densities. Each word meant the same thing in every plot that used it.

## What comes next

Every sentence here was already written, so nothing asked you to choose a mark.
[Writing practice](writing.qmd) reverses that: it starts from a question and asks
you to write the sentence.
