# Reading practice

```{r}
#| include: false
source("R/setup.R")
```

Can you tell what a plot shows before you see it? You have the vocabulary; now
read with it. Ten plots follow, ending in a cube. For each one: read the
specification aloud before looking at the picture, then check the picture
against what you said. The English sentence is given under each plot. Try to
say it yourself first.

## Gold medals by country

`medals` holds one row per country, with its counts of gold, silver and
bronze:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold)
```

*"Given the medals: bars, x is country, y is gold."* A bar measures an
amount from zero, so the categories sit on x and the number on y. No
transform anywhere: `bar` draws the numbers exactly as they arrive in the
table.

## The same bars, lying down

```{r}
data(medals) + bar + x(gold) + y(country)
```

*"Given the medals: bars, x is gold, y is country."* Swap which axis carries
the category and the bars lie down. There is no "horizontal bar" atom, and no argument that
flips one: orientation is read off the bindings, because the sentence
already says which axis holds the category.

## One measurement's shape

`iris_flowers` holds one row per flower: four measurements, and the species
the flower belongs to.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

*"Given iris flowers: bars derived by bin, x is petal length."* The `*`
derives: `bin` cuts the axis into intervals and counts rows into them, and
`bar` draws the counts. You did not bind `y`; `bin` invents the count, and
the axis says so. Note the two peaks. Plot eight explains them.

## The same question, smoothly

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"Given gapminder 2007: a line derived by density, x is life."* Same question
as the histogram (*how is this measurement distributed?*), different
derivation. Swapping one atom gives the smooth estimate instead of the binned
one.

## One dot per country, grouped by continent

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Given gapminder 2007: points, x is continent, y is life."* A categorical x
gives each category a slot, and the points for that category sit in a strip
above it. That is the rule that placed bars on country slots in plot one,
working here on a different mark. A plot shaped like this one has a name: the
**strip plot**.

## The typical value per continent

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*
The strip plot above
showed every country; `* mean` collapses each continent to its average.
Here `y(life)` names the column the mean *reads*: a transform that reads
a column makes you say which one.

## Five countries across the decades

`gapminder_asia` holds one row per country *and year*, so each country
appears many times.

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*
On a `line`, a categorical `color` first *splits* the rows into one line
per category, then colors them. One channel does two things, and you write it
once.

## Three species on one axis

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Given iris flowers: bars derived by bin, x is petal length, color by
species."* The two peaks from plot three now sit in separate histograms, one
per species. All three use the same bin edges, and each is drawn where its own
values fall. Each fill is translucent under a solid outline, so every
histogram stays visible. The channel that split the lines in plot seven splits
the rows here: same word, same rule.

## One panel per continent

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split into
panel columns by continent."* The `|` operator splits the plot into panel
columns, one panel per category. All the panels share one pair of scales, so
they can be compared.

## Three measurements on three axes

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Given iris flowers: points, x is sepal length, y is sepal width, z is petal
length, color by species."* The third dimension is not a different package, and not a
different mark. It is one more position channel, and the sentence grew by one
word.

## What you just did

Ten plots, and twelve distinct words. Every one read left to right
as *data, mark, positions, refinements*. Nothing you learned for one plot had
to be unlearned for the next. The same `color` split bars, lines and points.
The same `*` derived counts, means and densities. Now you write.
