# Reading practice

```{r}
#| include: false
source("R/setup.R")
```

Can you tell what a plot shows before you see it? You have the vocabulary; now
read with it. Ten plots follow, ending in a cube. For each one: read the
specification aloud before looking at the picture, then check the picture
against what you said. The English sentence is given under each plot. Try to
say it yourself first.

## Gold medals by country

Which country won the most gold, and by how much? `medals` holds one row per
country, with its counts of gold, silver and bronze:

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold)
```

*"Given the medals: bars, x is country, y is gold."* A bar measures an
amount from zero, so the categories sit on x and the number on y. No
transform anywhere: `bar` draws the numbers exactly as they arrive in the
table.

## The same bars, lying down

A medal table is usually printed as a list, with the leader at the top. The
same five counts can be read that way.

```{r}
data(medals) + bar + x(gold) + y(country)
```

*"Given the medals: bars, x is gold, y is country."* Swap which axis carries
the category and the bars lie down. There is no "horizontal bar" atom, and no argument that
flips one: orientation is read off the bindings, because the sentence
already says which axis holds the category.

## One measurement's shape

Do petal lengths gather around one typical value, or do they fall into groups?
One measurement answers that, and it comes from a new table. `iris_flowers`
holds one row per flower: four measurements, and the species the flower
belongs to.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

*"Given the iris flowers: bars derived by bin, x is petal length."* The `*`
derives: `bin` cuts the axis into intervals and counts rows into them, and
`bar` draws the counts. You did not bind `y`; `bin` invents the count, and
the axis says `Count`. Note the two peaks. Plot eight explains them.

## The same question, smoothly

Bars jump from one bin to the next, and sometimes you want a distribution as
one smooth curve instead. The measurement this time is life expectancy, across
the 142 countries.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"Given gapminder 2007: a line derived by density, x is life."* Same question
as the histogram (*how is this measurement distributed?*), different
derivation. Swapping one atom gives the smooth estimate instead of the binned
one.

## One dot per country, grouped by continent

Do the continents differ in life expectancy, and how widely do the countries
inside each one vary? Both answers are visible when every country is shown.

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Given gapminder 2007: points, x is continent, y is life."* A categorical x
gives each category a slot, and the points for that category sit in a strip
above it. That is the rule that placed bars on country slots in plot one,
working here on a different mark. A plot shaped like this one has a name: the
**strip plot**.

## The typical value per continent

You often want one number per continent, its average, rather than the spread
of every country.

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."*
The strip plot above showed every country; `* mean` reduces each continent to
one value, its average. Here `y(life)` names the column the mean *reads*: a
transform that reads a column makes you say which one.

## Five countries across the decades

How has life expectancy changed since 1952, and did five Asian countries
change at the same rate? `gapminder_asia` holds one row per country *and
year*, so each country appears many times.

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*
On a `line`, a categorical `color` first *splits* the rows into one line
per category, then colors them. One channel does two things, and you write it
once.

## Three species on one axis

Plot three showed two peaks in petal length, and the table has three species.
How do three species make two peaks?

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."* The two peaks from plot three now sit in separate histograms, one
per species. All three use the same bin edges, and each is drawn where its own
values fall. Each fill is translucent under a solid outline, so every
histogram stays visible. `color` does the splitting here, the way
it split the lines in plot seven: same word, same rule.

## One panel per continent

Richer countries live longer across the world as a whole. Does the same
pattern hold inside each continent?

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split into
panel columns by continent."* The `|` operator splits the plot into panel
columns, one panel per category. All the panels share one pair of scales, so
they can be compared.

## Three measurements on three axes

Petal length alone left two species overlapping in plot eight. Two more
measurements might separate them, if all three can be seen at once.

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is petal
length, color by species."* The third dimension is not a different package, and not a
different mark. It is one more position channel, and the sentence grew by one
word. In the web edition the cube turns with a drag: drag right and the side
facing you moves right, drag down and it tips. Each angle shows what another
hides, so try a few.

## What you just did

Ten plots, and twelve distinct words. Every one read left to right
as *data, mark, positions, refinements*. Nothing you learned for one plot had
to be unlearned for the next. The same `color` split the bars and lines, and
colored the points. The same `*` derived counts, means and densities. Now
you write.
