# Reading practice

```{r}
#| include: false
source("R/setup.R")
```

Can you tell what a plot shows before you see it? You have the vocabulary; now
read with it. Ten plots follow, each a little more composed than the last. For
each one: read the code aloud before looking at the picture, then check the
picture against what you said. The reading is given under each plot, but it
works better if you get there first.

## Golds by country

```{r}
data(medals) + bar + x(country) + y(gold)
```

*"Given the medals: bars, x is country, y is gold."* A bar measures an
amount from zero, so the categories sit on x and the number rides on y. No
transform anywhere: `bar` draws the numbers exactly as they arrive in the
table.

## The same bars, lying down

```{r}
data(medals) + bar + x(gold) + y(country)
```

*"Bars, x is gold, y is country."* Swap which axis carries the category and
the bars lie down. There is no "horizontal bar" atom and no flip switch:
orientation is read off the bindings, because the sentence already says
everything.

## One measurement's shape

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

*"Bars derived by bin: x is petal length."* The `*` derives: `bin` cuts
the axis into intervals and counts rows into them, and `bar` draws the
counts. You did not bind `y`; the count is synthesized, and the axis says
so. Note the two humps; hold that thought for plot eight.

## The same question, smoothly

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"A line derived by density: x is life."* Same question as the histogram
(*how is this measurement distributed?*), different derivation. Swap one
atom, get the smooth estimate instead of the binned one.

## Every country, one dot per continent

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Points, x is continent, y is life."* A categorical x gives each category a
slot and stacks the points above it. That is the rule that placed bars on
country slots in plot one, working here on a different mark. A plot shaped
like this one has a name: the **strip plot**.

## The typical value per continent

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Bars derived by mean: x is continent, y is life."* The strip plot above
showed every country; `* mean` collapses each continent to its average.
Here `y(life)` names the column the mean *reads*: a transform that reads
a column makes you say which one.

## Five countries across the decades

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*
On a `line`, a categorical `color` first *splits* the rows into one line
per category, then colors them. One channel, two duties, stated once.

## Three species on one axis

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Bars derived by bin: x is petal length, color by species."* The two humps
from plot three come apart. There is now one histogram per species. All three
share the same bin edges and are drawn in place. Translucent fills sit under
solid outlines, so nothing hides behind anything. The channel that split the
lines in plot seven splits the bins here: same word, same rule.

## One panel per continent

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) | facet(continent)
```

*"Points, x is gdp on a log scale, y is life, split into panel columns by
continent."* The `|` operator multiplies the plot across panels, one panel
per category. All the panels share one pair of scales, so they can be
compared.

## Into the page

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Points: x is sepal length, y is sepal width, z is petal length, color
by species."* The third dimension is not a different package, and not a
different mark. It is one more position channel, and the sentence grew by one
word.

## What you just did

Ten plots, and roughly a dozen distinct words. Every one read left to right
as *data, mark, positions, refinements*. Nothing you learned for one plot had
to be unlearned for the next. The same `color` split bars, lines and points.
The same `*` derived counts, means and densities. Now you write.
