# Scales

```{r}
#| include: false
source("R/setup.R")
```

What do you do when most of your data is crowded into one corner? You change
the **scale**, which decides how a number becomes a position. It is written on
the binding, not as an atom of its own:

```r
x(gdp, scale = "log")
```

A scale belongs to one channel, so it is written where that channel is bound.
There is no `scale()` atom to write separately. A separate atom would make you
name the channel twice, once to bind it and once to scale it, and the two could
then disagree.

Writing the scale on the binding also makes a sentence easier to read. To see
how `gdp` is placed, you look at `x(gdp, scale = "log")` and nowhere else.
Nothing else in the sentence can change it, and there is no second function to
look up.

## A scale is not a transform

A log scale is not the only way to see a spread of incomes. You could take the
logarithm of the column yourself and plot the result.
That is a different operation, and the difference decides what the axis says.

`bin` and `mean` are **transforms**: they read your table and compute new
values from it. A scale produces no new values at all. It changes how far along
the axis a number lands and what the ticks say, and nothing else.

The difference is visible in the finished plot. Here is `gdp` plotted with a
log *scale*:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) +
  x_label("GDP per capita") +
  title("Income and life expectancy: log scale")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

The axis reads 1K and 10K. Those are dollars, the units you supplied.

The word *log* appears nowhere on the axis, and that is deliberate. The units
did not change, so the axis label does not claim they did. The ticks are the
only sign: each is a whole power of ten, so equal steps along the axis mean
equal ratios. Write `x_label("GDP per capita (log scale)")` when a reader needs
the scale named in words. The axis label is yours, and gog never writes the
scale into it.

Now the same picture, drawn after taking the logarithm of the column yourself:

```{r}
gm_logged <- gapminder_2007
gm_logged$log_gdp <- log10(gm_logged$gdp)

data(gm_logged) + point + x(log_gdp) + y(life) +
  color(continent) +
  title("The same shape, with the axis in log units")
```

The shape is identical, and the axis now reads 3 and 4. To know a country's
income, the reader has to raise ten to that power. That arithmetic is what a
scale does for the reader.

So the choice between the two is about what the quantity *is*. Take the
logarithm of the column when that logarithm is itself the quantity of interest:
a log-odds, a decibel, a pH. Use a log scale when the quantity is still dollars
or people, and you only want to see it across orders of magnitude.

## The order of a scale and a transform

Once a transform is in the plot, a scale is no longer only about display,
because a transform has to choose which numbers to work on. gog follows one
rule, and it has no per-transform exceptions:

> A transform groups by one axis and measures on the other. The scale is
> applied before the transform on the grouping axis, and after it on the
> measured axis.

### Scaling the grouping axis

A histogram of income is the first plot where that rule changes the result. Its
bins have to be cut somewhere, either in dollars or in the logarithms of those
dollars.

`bin` groups by x, so a log x is applied first. The bins are cut on the
logarithms, not on the dollars. Each bin then spans one constant *ratio*. Every
bar is therefore the same width, and the bars fill the axis from end to end:

```{r}
data(gapminder_2007) + bar * bin + x(gdp, scale = "log") +
  x_label("GDP per capita") +
  title("Bins cut in log space are evenly spaced")
```

*"Given gapminder 2007: bars derived by bin, x is gdp on a log scale."*

Suppose the bins had been cut in dollars and only the *drawing* had been
logarithmic. The first bin alone would cover most of the axis, and the rest
would crowd into the right edge. That reads as a mistake rather than as a
modeling choice, which is why the grouping axis is scaled first.

### The measured value keeps its own units

Put the log on the measured axis instead and a different question follows. Does
a bar on a log scale stand for a sum of dollars, or for a sum of logarithms?

`sum` writes to y, so a log y is applied *after* the sum. The two bars in this
plot stand for 100 and 10:

```{r}
#| echo: false
peek(receipts)
```

```{r}
data(receipts) + bar * sum + x(store) + y(sales, scale = "log") +
  title("A sum is still a sum")
```

*"Given the receipts: bars derived by sum, x is store, y is sales on a log
scale."*

If the scale had been applied first, gog would have added the *logarithms*,
giving the log of a product, which is not a quantity anyone asked for. gog adds
the dollars first and scales the total afterwards, so `sum` reports a sum of
dollars.

If you have used ggplot2, note the order, because ggplot2 does the opposite.
Under `scale_y_log10()`, a `stat_summary(fun = mean)` averages the logarithms of
the values. The result is a geometric mean, the average computed on the
logarithms. gog computes the summary in your data's units and scales the result,
so the number the bar stands for is the number you would get by taking the sum
or the mean yourself.

### `smooth`: a curve on a log axis

The grouping axis raises the same question for `smooth`: is the curve fitted to
dollars, or to their logarithms? On incomes that run from a few hundred dollars
to nearly fifty thousand, the answer changes the shape of the curve.

`smooth` groups by x, so with a log x the curve is fitted against log income.
Life expectancy rises by about the same amount for each tenfold rise in income,
so the curve is close to a straight line:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  style(color = "lightgray") +
  line * smooth + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") +
  title("A trend fitted against log income")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, and also a
line derived by smooth, x is gdp on a log scale, y is life."*

## What a log scale refuses

Some columns cannot take a log scale, and they are not rare: a depth that goes
below zero, a balance that went negative, a count of zero.

A logarithm is undefined at zero and below, so rows with such values have no
position on the axis. gog says so rather than dropping them quietly:

```{r}
#| echo: false
peek(depth_readings)
```

```{r}
#| error: true
data(depth_readings) + point + x(depth, scale = "log") + y(temp)
```

The message names how many rows are affected and the smallest value in the
column, because "some of your data cannot be drawn" is not enough to act on.

Text has no logarithm either, and a column of text already gets a categorical
axis without being asked:

```{r}
#| error: true
data(gapminder_2007) + point + x(continent, scale = "log") + y(life)
```

## Choosing the base

Not every quantity is best read in powers of ten. Ten is the usual base for
money and for population, but a pitch doubles with every octave, and a decay is
counted in factors of *e*.

`scale = "log"` means base 10 unless you say otherwise:

```r
x(freq, scale = "log", base = 2)
```

The base is a *number*, not a name. There is no `"log2"` or `"ln"` scale. One
`base` argument already covers every base, so a separate name for each would
add words without adding meaning. The same objection rules out ggplot2's
`scale_x_log10` family.

### The base changes the labels, not the picture

You might expect a new base to move the points. It does not move a single one.
Any two logarithms of the same column differ by a constant factor, and
gog normalizes every axis by its own range. The factor cancels, so the points
land in *the same places in every base*. The transforms are unaffected too:
bins of equal width in base 2 are equal width in base 10, so `bar * bin` cuts
at the same values whichever base you name.

What the base actually chooses is where the gridlines fall and how they read.
Here are six frequencies, each one an octave above the one before it, on base
10 and then base 2:

```{r}
#| echo: false
peek(octaves)
```

```{r}
data(octaves) + point + x(freq, scale = "log") + y(level) +
  x_label("Frequency (Hz)") + title("Base 10: gridlines at 100, 1K")
```

```{r}
data(octaves) + point + x(freq, scale = "log", base = 2) + y(level) +
  x_label("Frequency (Hz)") + title("Base 2: gridlines on every octave")
```

*"Given the octaves: points, x is freq on a log scale in base 2, y is level."*

The points have not moved. The second plot puts a gridline on each doubling,
which is what "an octave" means. So for pitch, bits, or doubling times, base 2
is the one whose labels name what you are counting.

### Natural log

The natural base is for a quantity that decays at a constant rate. The amounts
in the `decay` table shrink by a factor of *e* every hour, so there is one
gridline between each reading and the next. A change by a factor of *e* is
called an **e-folding**.

R has no constant named *e*, so you write `base = exp(1)`. The other three
bindings have one: Python's `math.e`, Julia's `ℯ`, and JavaScript's `Math.E`.

```{r}
#| echo: false
peek(decay)
```

```{r}
data(decay) + point + x(amount, scale = "log", base = exp(1)) + y(hours) +
  x_label("Amount remaining") +
  title("Gridlines at each e-folding: one step is a factor of e")
```

Notice the labels: *e*, *e²*, *e³*, *e⁴*, not 2.718, 7.389, 20.086, 54.598.
That follows from one rule gog applies to every base:

> Label the quantity when it is a short number; otherwise label the power.

Base 10 gives 1, 10, 100, 1K, 10K and so on, up to 100B. K means a thousand, M
a million, and B a billion (10⁹). The letters stop at B, so 10¹² would read
1000B, with four digits before the letter. From 10¹² upward, a tick is labeled as
a power instead, such as `10¹²`. Base 2 gives 1, 2, 4 and so on, up to 8388608. A number with more
than seven digits is labeled as a power instead, such as `2²⁴`. Base *e* has no
short quantity above 1, so every tick past the first reads as a power, which is
what a reader counting e-foldings wants. If instead you want the axis to read
0, 1, 2, 3 in ln units, you do not want a log scale at all. You want `log()`
applied to the column, and an axis labeled in log units, like the `log_gdp` plot
earlier in this chapter. A scale keeps the reader's units; a column of
logarithms changes them.

A base only means something as the base *of* a logarithm, so gog refuses one
that has no log scale to belong to:

```{r}
#| error: true
data(octaves) + point + x(freq, base = 2) + y(level)
```

## Time: a date column gets a calendar axis

A column of dates raises a question the other columns do not. Where do the
ticks go, and what do they say, when the units are years and months rather than
tens and hundreds?

The time scale is one you never have to write. A column of dates or date-times
(`Date` or `POSIXct` in R) *is* temporal, because its type already says so. gog
reads the calendar from the type, the same way a column of text already gets a
categorical axis:

```{r}
#| echo: false
peek(revenue)
```

```{r}
data(revenue) + line + x(day) + y(sales) +
  title("Twenty-six years, ticked in years")
```

The sentence names no scale at all. Writing `scale = "time"` is allowed and adds
nothing, exactly like writing `scale = "linear"` on a number.

The ticks land on **calendar boundaries**, because the calendar is not decimal.
An axis of years steps 1, 2 or 5 like any number line. For units smaller than a
year those steps mean nothing: nobody reads an axis cut into fifths of a year.
Months step by 1, 2, 3 or 6, and each of those divides twelve. A gridline
therefore always falls on January, and a 3-month step marks each quarter. Days
step by 1 or 2, and weeks by 1 or 2, always on Mondays. Hours and minutes step in
the units a clock uses.

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
# six_weeks is the cast's daily orders table, see the Preface
data(six_weeks) + line + x(day) + y(orders) +
  title("Six weeks, ticked on Mondays")
```

A tick is labeled at its own resolution: `2010` for a year step, `Jan 2024` for
a month step, `Mar 4` for a day step. A column of dates never gets clock ticks,
however narrow its range. A column of date-times (`POSIXct` in R) can, because
it stores a time of day:

```{r}
#| echo: false
peek(monitoring)
```

```{r}
data(monitoring) + line + x(at) + y(load) +
  title("A day and a half, ticked by the clock")
```

`00:00` happens every day, so that label names no particular day. The midnight
tick shows the date instead. Every other label on a time axis follows one rule:
*a tick is labeled at its own resolution, and it leaves out what it shares with
its neighbors.*

### Bars on a date axis

Daily orders make a natural bar chart, one bar per day, and forty-two bars
raise the question of where the ticks go.

Bars positioned at dates are an ordinary time series. The ticks still come from
the calendar, and there is not a tick for every bar. Wilkinson's stock-price
example puts its ticks on Sundays rather than on every trading day
[@wilkinson2005], and gog ticks these six weeks at the start of each week:

```{r}
data(six_weeks) + bar + x(day) + y(orders) +
  title("Daily bars, weekly ticks")
```

A bar that measures a date is refused. A bar's *length* is an amount, and a
moment in time is not an amount. A bar "reaching" 2007 would be measured from
1970, the year a stored date counts from, which nobody chose as a baseline:

```{r}
#| error: true
data(revenue) + bar + x(sales) + y(day)
```

### What a time axis refuses

A date is stored as a number underneath. You might therefore expect a date to
take a log scale, and a number to take a time scale. Neither is allowed.

A moment has no logarithm. The calendar's zero is an arbitrary convention, and
a logarithm measured from an arbitrary zero measures nothing:

```{r}
#| error: true
data(revenue) + line + x(day, scale = "log") + y(sales)
```

gog can read the calendar off a date column, but you cannot declare a plain
number temporal at the binding. The engine cannot know whether a number like
20656 counts days, seconds, or years:

```{r}
#| error: true
data(revenue) + point + x(sales, scale = "time") + y(sales)
```

The fix belongs in the data, where the answer is known: convert the column to
your language's date or timestamp type, which in R is `as.Date()` or
`as.POSIXct()`.

Timezones are never converted. The engine draws the clock time you gave it. In
R, a `POSIXct` is formatted in its own timezone when it is drawn, so the axis
shows the times your session showed you.
If two series must share an axis, put them in one timezone before plotting;
that decision belongs to the analysis, not the engine.

### Time on the other channels

A scale answers *how far along?* on every channel that measures, and a date
column is no exception. Coloring rows by how recent they are is common. An axis
tick can read `Mar 4`, because the ticks beside it place it in a sequence. A
legend row stands alone, so each row carries the full date:

```{r}
data(six_weeks) + point + x(day) + y(orders) + color(day) +
  title("Recent days, darker")
```

## The scales

The chapter has used `linear`, `log` and `time` so far, and `category` comes
later, in [The column's type picks the axis](#the-columns-type-picks-the-axis).
The table names all four and what each one does.

| Scale | What it does |
|---|---|
| `linear` | Equal steps for equal differences. The default |
| `log` | Equal steps for equal ratios, on `x`, `y`, `color`, `size` and `opacity`, with any `base` above 1 |
| `time` | Positions dates and times, chosen from the column type: columns of dates and date-times, on every channel that measures |
| `category` | One slot per distinct value, chosen from the column type: columns of text, and columns with a declared order |

## The axis fits your data

Where should an axis end, on the data or on the next round number past it?
Padding to a round number leaves part of the panel blank, and implies that the
data reaches further than it does.

An axis shows round numbers on its ticks, but the *range* follows your data, not
the round numbers. Each tick is a round number that falls *inside* the data, or
at most a little past it into the margin. The axis does not extend to the next
multiple of ten only to end on a round number. A span of 1952–2007 is drawn as
1952–2007, with ticks at 1960, 1980 and 2000, not padded to 1940–2020 with a
third of the panel left blank:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + area + x(year) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("The year axis fits 1952–2007; the ticks fall inside it")
```

*"Given gapminder Asia: areas, x is year, y is life, grouped by country."*

Two more rules refine that fit. When the data sets an end, that end is a **free
end**, and a free end leaves a small margin, so a glyph never touches the panel
edge. A point at the exact maximum would otherwise be cut in half by that edge:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  x_label("GDP per capita") + title("Every point sits clear of the panel edge")
```

A baseline gets no margin. The area of Asian life expectancies above shows it:
the shape starts on the panel's bottom edge, not a little way above it. When a
bar or an area measures from zero, the zero is part of what it measures. A gap
beneath it would show a stretch of axis below zero that no bar can reach, so the
baseline sits on the panel edge. An `area` has no glyph for the edge to cut, so
it runs from the left edge of the panel to the right. That is why the area plot
above spans the full width while the scatterplot of incomes keeps its margins.
An area ends exactly at its first and last values. A point's circle reaches past
its value by its radius, so it needs the margin.

## A scale is not only for axes

Most of this chapter has put a scale on an axis. A color ramp can crowd its data
into one end, just as a linear axis can crowd it into one corner of the panel.
`x` and `y` turn a number into a position, but so do `color`, `size` and
`opacity`: a ramp position, a radius, a transparency. All five answer *how far
along?*, so all five take a scale. In three dimensions `z` takes a scale too, but
it refuses `log`: see [Space](space.qmd).

This matters most for a column spread across orders of magnitude. `population`
runs from about 200 thousand to 1.3 billion. On a linear ramp the median
country sits in the bottom one percent of that range, so nearly every point
gets the same color:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population) +
  x_label("GDP per capita") + title("Linear color: nearly one flat color")
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log") +
  x_label("GDP per capita") + title("Log color: the whole ramp is used")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
population on a log scale."*

The same is true of `size`, where a linear scale leaves almost every point at
the minimum radius:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  size(population, scale = "log") + style(color = "steelblue") +
  x_label("GDP per capita") + title("Log size")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, size by
population on a log scale."*

Look again at the middle row of each of the two color legends. On a linear ramp
it is the arithmetic midpoint; on a log ramp it is the *geometric* one: √(min
· max), about 16M rather than 660M. The label has to be the geometric midpoint,
because it names the color painted at the middle of the ramp.

`shape` and `group` take no `scale` argument. They answer *which one?*, and
there is no distance between circle and square for a scale to run along.

## `limits`: stating the range yourself

Every rule so far takes the range from the data. That is the right default, but
it is wrong when the data does not cover every value the axis should show. The
range an axis should cover is its **domain**, and it is not always the range the
data reaches. The `tide` table holds a gauge's readings, one height every three
hours from hour 1 to hour 22, so none falls at midnight. `polar()` draws an axis
as a circle, and [Polar](polar.qmd) teaches that space in full. Nothing in a
column of hours tells that circle that hour 24 is the same place as hour 0.
`limits` states the domain:

```{r}
#| echo: false
peek(tide)
```

```{r}
data(tide) + line + x(hour, limits = c(0, 24)) + y(height) + polar() +
  title("The whole day, whatever hours the gauge recorded")
```

*"Given the tide: a line, x is hour from 0 to 24, y is height, in polar."*

Without `limits`, one turn would cover only 1 to 22, and the first and last
readings would be drawn on the same spoke. With `limits`, each hour sits at its
share of the day: hour 6 a quarter of the way around, hour 12 halfway. The curve
still does not close, and should not. The gauge took no reading at midnight, and
a domain says what range the axis covers rather than inventing an observation.
The gap is in the readings, and the plot shows it at its true size, the three
hours from 22 to 1. [Polar](polar.qmd) describes three cases: two in which the
curve closes, and this one, in which it does not.

`limits` is written on the binding, beside `scale` and `base`, for the same
reason they are: a domain belongs to one channel, and an atom of its own would
name that channel twice. A stated end is drawn exactly where you state it. The
margin exists to keep a glyph off the panel edge, and an end you chose is not a
data extreme that needs the room.

A domain can widen the axis as easily as it narrows it. Nothing was dropped in
that plot: the readings run 1 through 22 and all of them lie inside 0 through
24. Narrowing is the direction that removes rows, and gog reports how many rows
it removes:

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp, limits = c(0, 30000)) + y(life) +
  color(continent) + x_label("GDP per capita") +
  title("A domain of 0 to $30,000")
```

The note above the plot is that report: it names how many rows fell outside and
what the domain was. The plot still draws, because removing those rows is what
you asked for, but gog does not remove data without saying so. Ask for a domain
no row can satisfy and gog refuses instead, because the alternative is an empty
panel with axes drawn over nothing.

Keeping every country on the page and marking one group of them is a different
operation, and it has a different word. `limits` runs before the transforms and
removes rows. `brush` runs after the picture is composed and removes nothing.
[Selection](selection.qmd) shows the two side by side.

You can state one end alone. The end you leave out stays a free end, which is
how you start an axis at zero and let the top follow the data:

```{r}
data(gapminder_2007) + point * mean + x(continent) + y(life, limits = c(0, NA)) +
  title("y from zero, top still fitted to the data")
```

*"Given gapminder 2007: points derived by mean, x is continent, y is life from
0."*

`limits` reaches more than the axes. A domain is a property of the scale, so it
reaches every channel that measures. That is how you keep one color ramp the
same across plots built from different data:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log", limits = c(1e5, 2e9)) +
  x_label("GDP per capita") + title("A ramp fixed from 100K to 2B")
```

On a categorical axis `limits` is refused, and the refusal names the two ways to
get what you meant. A category has no range to lie inside. Choosing *which*
categories appear is filtering your table, and choosing what order they appear
in is `order()`.

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent, limits = c(0, 3)) + y(life)
```

## The column's type picks the axis

The time section showed a column of dates choosing its own axis without being
asked. A column of text does the same, and this section states the rule behind
both. Other tools make you name a scale whenever you want to change one, so you
may expect to write `scale = "category"`.

The four scales are not all chosen the same way. You ask for `log`. You do not
ask for `time` or `category`. gog reads those off the column: a column of
dates gets a calendar axis, and a column of text gets one slot per distinct value.

You may still write them. As with `scale = "time"` on a date column, writing
`scale = "category"` on a categorical column is allowed and changes nothing.
The plot below is identical to the same sentence with the scale left off:

```{r}
data(gapminder_2007) + bar * mean + x(continent, scale = "category") + y(life) +
  title("Writing it changes nothing")
```

*"Given gapminder 2007: bars derived by mean, x is continent as a category, y
is life."*

What you cannot do is use a scale to contradict the column. Ask for a
categorical axis on a number and gog refuses:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp, scale = "category") + y(life)
```

The reason behind that refusal also explains the others in this chapter. A
scale says *how* a measured column is placed. The column's type decides whether
the axis measures at all. So `log` on a categorical column is refused, and
`time` on a plain number is refused. A date column cannot be turned back into
raw numbers by asking for `linear` either. None of those refusals offers another
scale. Each tells you to change the column or remove the scale, because the
column's type is what the axis reads.

There are two things you might mean by "draw this number as categories", and
there is a different answer for each. For one slot per distinct value, declare
the column's categories: `factor()` in R and `ordered()` in the other three
bindings, both shown in [Data](data.qmd). For ranges, use `bin`, which cuts the
numbers into intervals and is covered in [Transforms](transforms.qmd).

## `tick_count`: how many ticks the axis aims for

An axis picks about five ticks by default. After `limits`, the count is the
second thing about an axis you can choose. A crowded axis is easier to read with
fewer; an axis someone will take numbers from is easier with more. `tick_count`
is written on the binding, beside `scale`, `base` and `limits`, and for the
same reason. How many ticks an axis gets is a property of the scale, so it
belongs to the sentence that describes the scale rather than to `theme()`,
which controls the page.

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
   x_label("GDP per capita") + title("A low count: the shape, not the numbers")) |
  (data(gapminder_2007) + point + x(gdp, tick_count = 12) + y(life) +
     x_label("GDP per capita") + title("A high count: for reading values off"))
```

*"Given gapminder 2007: points, x is gdp with 3 ticks, y is life, beside the
same with 12 ticks."*

The count is a target, not an exact number. The count sets the *step* between
ticks, and the step is then rounded to 1, 2 or 5 times a power of ten. So asking
for eight ticks on an axis from 0 to 100 gives a step of 10, and eleven ticks.
What you can rely on is the direction: a higher count never gives fewer ticks
than a lower one. Every tick is a round number. The count also has an upper
limit: an axis draws at most 26 ticks. If the rounded step would give more than
26 ticks, the engine uses a larger round step. The axis is still labeled from
one end to the other, and a message names both steps.

Both panels in the plot above cover the same range. That is the rule the count obeys: it
changes how densely an axis is *labeled*, never where the axis stops. Ask for
two ticks and the axis does not shrink to the two values it draws.

```{r}
data(gapminder_asia) + line + x(year, tick_count = 2) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("Two ticks, and still 1952 to 2007")
```

A log axis and a date axis take a count too, but their steps are rounded to
different values. On a log axis the ticks can be at every power of the base, or
skip one or more powers between them. In base 10 they can also be at 1, 2 and 5
times each power. A date axis uses calendar steps, such as a day, a week, a
month or a year. [The time section](#time-a-date-column-gets-a-calendar-axis)
describes these steps. On these two axes, the engine picks the step whose number
of ticks is closest to your count.

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log", tick_count = 12) + y(life) +
  x_label("GDP per capita") + title("Seven ticks: 1, 2 and 5 times each power")
```

*"Given gapminder 2007: points, x is gdp on a log scale with 12 ticks, y is
life."*

The sentence asks for twelve ticks, and the axis draws seven. In base 10 no log
step is finer than 1, 2 and 5 times each power, so seven is the closest count to
twelve. Without a count, this axis reads only 1K and 10K, as at the start of the
chapter.

```{r}
data(six_weeks) + line + x(day, tick_count = 3) + y(orders) +
  title("Three ticks: one every two weeks")
```

*"Given the six weeks table: a line, x is day with 3 ticks, y is orders."*

By default these six weeks get a tick every Monday, six in all. Asked for three,
the axis steps two weeks at a time, still on Mondays.

A categorical axis has one place per category, so the data sets the count and
you do not. The refusal names the same two fixes the `limits` refusal named:

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent, tick_count = 3) + y(life)
```

A legend is not an axis, and it takes no tick count. `limits` reaches every
channel that measures, because every one of them has a domain; `tick_count`
reaches only the three position channels that draw an axis: `x`, `y` and `z`. A
color legend names three rows, both ends and the middle, and the scale, linear
or log, sets their values, not a count you choose. The middle row is the
midpoint explained above: the label for the color painted halfway along the
ramp. So `color()`, `size()` and `opacity()` have no tick count to set, the same
way `shape()` has no `scale` and no `limits`: there is nothing for the count to
describe.

### Crowding in the plane and in the cube

A flat panel and a cube follow one rule for crowded numbers: no number may be
printed over another. A flat continuous axis draws every tick it chose, as long
as no two numbers overlap. Where panels are side by side, an end number that
would reach past its panel is moved inward. In narrow panels, such as the five
panels of a plot faceted by continent ([Faceting](faceting.qmd)), the numbers
can still overlap. The axis then draws fewer numbers, and the ticks and
gridlines of the others are left out too. The numbers it keeps are the ones a
larger round step would choose, so 0, 10K, 20K, 30K and 40K become 0, 20K and
40K. A number left out can be read from the numbers on either side, so no
message is printed. If you stated a `tick_count`, a message says how many ticks
were drawn. A polar plot, a map, and an axis freed with `free = TRUE` still draw
every number they chose.

The sentence below asks for twelve ticks, but each panel has room for three,
and a message says so:

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp, tick_count = 12) + y(life) |
  facet(continent)
```

*"Given gapminder 2007: points, x is gdp with 12 ticks, y is life, split into
panel columns by continent."*

A cube can also have too little room for its numbers. In a 3-D plot
([Space](space.qmd)), each axis is longer or shorter depending on the viewing
angle. Turning the cube lengthens one axis and shortens another. The cube then
draws as many numbers as fit and leaves the rest out. When a count you stated
cannot fit, the engine says so rather than quietly drawing fewer.

Names on a categorical axis follow the same rule: no name may be printed over
another. The axis tries to name every category, and a name is usually much wider
than a number. When names do not fit side by side, a flat axis first turns them
90 degrees, to read upward. A turned name needs room along the axis only for its
height, not its width. When even turned names are too close, the flat axis
leaves some out, as the cube does. It draws one name in every two or three, each
with its tick, and prints a message. The 142 countries of 2007 do not fit, even
turned:

```{r}
#| message: true
data(gapminder_2007) + bar + x(country) + y(life)
```

Every bar is still drawn, and a bar without a name keeps its place between two
named ones. The message gives two ways to name every country: put `country` on
the y axis in a taller plot, or draw fewer rows.

## In short

A scale changes where a number lands on the axis. It never changes the number
itself. `limits` states the domain: narrowing it removes rows and reports how
many, while a brush removes none. `tick_count` is a target, not an exact
number. A crowded continuous axis, flat or in a cube, draws fewer numbers. A crowded
categorical axis turns its names, then leaves some out and says so.

## What comes next

A scale decides how far along a channel a number lands. It does not decide
which visual property shows the column.
[Channels](channels.qmd) covers that: which property carries a column, and what
guide it earns.
