# Scales

```{r}
#| include: false
source("R/setup.R")
```

What do you do when most of your data is crowded into one corner? You change
the **scale**, which decides how a number becomes a position. It is written on
the binding, not as an atom of its own:

```r
x(gdp, scale = "log")
```

Scales are a property of the axis, so they live where the axis is bound. There
is no `scale()` atom to write separately. A separate atom would make you name
the channel twice, once to bind it and once to scale it, and the two could then
disagree.

Writing the scale on the binding helps a second way. This one is about
remembering rather than about being correct.
The scale is written inside the binding it belongs to, so there is only one
place to look. A reader who wants to know how `gdp` is being read finds the
answer in `x(gdp, scale = "log")` itself. Nothing else in the sentence can
change it, and there is no second function to look up.

## A scale is not a transform

A log scale is not the only way to see a spread of incomes. You could take the
logarithm of the column yourself and plot the result.
That is a different operation, and the difference decides what the axis says.

`bin` and `mean` are **transforms**: they read your table and compute new
values from it. A scale produces no new values at all. It changes how far along
the axis a number lands and what the ticks say, and nothing else.

The difference is visible in the finished plot. Here is `gdp` plotted with a
log **scale**:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) +
  x_label("GDP per capita") +
  title("Income and life expectancy: log scale")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

The axis reads 1K and 10K. Those are dollars, the units you supplied.

The word *log* appears nowhere on the axis, and that is deliberate. The units
did not change, so the axis name does not claim they did. The ticks are the
only sign: each is a whole power of ten, so equal steps along the axis mean
equal ratios. Write `x_label("GDP per capita (log scale)")` when a reader needs
it in words. The axis name is yours, and gog never writes the scale into it.

Now the same picture drawn by logging the column yourself:

```{r}
gm_logged <- gapminder_2007
gm_logged$log_gdp <- log10(gm_logged$gdp)

data(gm_logged) + point + x(log_gdp) + y(life) +
  color(continent) +
  title("The same shape, with the axis in log units")
```

The shape is identical, and the axis now reads 3 and 4. To know a country's
income, the reader has to raise ten to that power. That arithmetic is what a
scale does for the reader.

So the choice between the two is about what the quantity *is*. Log the column
when the logarithm is itself the quantity of interest: a log-odds, a decibel, a
pH. Use a log scale when the quantity is still dollars or people, and you only
want to see it across orders of magnitude.

## The order of a scale and a transform

Once a transform is in the plot, a scale is no longer only about display,
because a transform has to choose which numbers to work on. gog follows one
rule, and it has no per-transform exceptions:

> A transform groups by one axis and measures on the other. The scale is
> applied before the transform on the grouping axis, and after it on the
> measured axis.

### Scaling the grouping axis

A histogram of income is the first plot where that rule changes the result. Its
bins have to be cut somewhere, either in dollars or in the logarithms of those
dollars.

`bin` groups by x, so a log x is applied first. The bins are cut on the
logarithms, not on the dollars. Each bin then spans one constant *ratio*. Every
bar is therefore the same width, and the bars fill the axis from end to end:

```{r}
data(gapminder_2007) + bar * bin + x(gdp, scale = "log") +
  x_label("GDP per capita") +
  title("Bins cut in log space are evenly spaced")
```

*"Given gapminder 2007: bars derived by bin, x is gdp on a log scale."*

Suppose the bins had been cut in dollars and only the *drawing* had been
logarithmic. The first bin alone would cover most of the axis, and the rest
would crowd into the right edge. That reads as a mistake rather than as a
modeling choice, which is why the grouping axis is scaled first.

### The measured value keeps its own units

Put the log on the measured axis instead and a different question follows. Does
a bar on a log scale stand for a sum of dollars, or for a sum of logarithms?

`sum` writes to y, so a log y is applied *after* the sum. The two bars in this
plot stand for 100 and 10:

```{r}
#| echo: false
peek(receipts)
```

```{r}
data(receipts) + bar * sum + x(store) + y(sales, scale = "log") +
  title("A sum is still a sum")
```

*"Given the receipts: bars derived by sum, x is store, y is sales on a log
scale."*

If the scale had been applied first, gog would have added the *logarithms*,
giving the log of a product, which is not a quantity anyone asked for. gog adds
the dollars first and scales the total afterwards, so `sum` reports a sum of
dollars.

The order is worth knowing if you have used ggplot2, which does the opposite.
Under `scale_y_log10()`, a `stat_summary(fun = mean)` averages the logged
values. The result is a geometric mean, the average computed on the logs. gog
computes the summary in your data's units and scales the result, so the number
the bar stands for is the number you would get by taking the sum or the mean
yourself.

### `smooth` fits inside its groups

The same question comes back from the grouping side: is the curve fitted to
dollars, or to their logarithms? On incomes that run from a few hundred dollars
to nearly fifty thousand, the answer changes the shape of the curve.

`smooth` groups by x, so a log x fits the curve against log income. Fitting
against log income spreads the points across the panel, so the curve comes out
close to a straight line:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  style(color = "lightgray") +
  line * smooth + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") +
  title("A trend fitted against log income")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, and also a
line derived by smooth, x is gdp on a log scale, y is life."*

## What a log scale refuses

Some columns cannot take a log scale, and they are not rare: a depth that goes
below zero, a balance that went negative, a count of zero.

A logarithm is undefined at zero and below, so those rows have no position on
the axis. gog says so rather than dropping them quietly:

```{r}
#| echo: false
peek(depth_readings)
```

```{r}
#| error: true
data(depth_readings) + point + x(depth, scale = "log") + y(temp)
```

The message names how many rows are affected and the smallest value in the
column, because "some of your data cannot be drawn" is not enough to act on.

Text has no logarithm either, and a column of text already gets a categorical
axis without being asked:

```{r}
#| error: true
data(gapminder_2007) + point + x(continent, scale = "log") + y(life)
```

## Choosing the base

Ten is the usual base for money and for population. It is not the right base
for everything you measure. A pitch doubles with every octave, and a decay is
counted in factors of *e*.

`scale = "log"` means base 10 unless you say otherwise:

```r
x(freq, scale = "log", base = 2)
```

The base is a **number**, not a name. There is no `"log2"` or `"ln"` scale. One
`base` argument already covers every base, so a separate name for each would
add words without adding meaning. The same objection rules out ggplot2's
`scale_x_log10` family.

### The base changes the labels, not the picture

One result here surprises most people, and it is worth knowing before you set a
base. Any two logarithms of the same column differ by a constant factor, and
gog normalizes every axis by its own range. The factor cancels, so the points
land in *the same places in every base*. The transforms are unaffected too:
bins of equal width in base 2 are equal width in base 10, so `bar * bin` cuts
at the same values whichever base you name.

What the base actually chooses is where the gridlines fall and how they read.
Here are six frequencies, each one an octave above the one before it, on base
10 and then base 2:

```{r}
#| echo: false
peek(octaves)
```

```{r}
data(octaves) + point + x(freq, scale = "log") + y(level) +
  x_label("Frequency (Hz)") + title("Base 10: gridlines at 100, 1K")
```

```{r}
data(octaves) + point + x(freq, scale = "log", base = 2) + y(level) +
  x_label("Frequency (Hz)") + title("Base 2: gridlines on every octave")
```

*"Given the octaves: points, x is freq on a log scale in base 2, y is level."*

The points have not moved. The second plot puts a gridline on each doubling,
which is what "an octave" means, so for pitch, bits, or doubling times, base 2
is the one whose labels name what you are counting.

### Natural log

The natural base is for a quantity that decays at a constant rate. The amounts
in the table fall by a factor of *e* every hour, so one gridline falls between
every two consecutive readings.

R has no bare *e* to write, so it is `base = exp(1)`. The other three bindings
have one: Python's `math.e`, Julia's `ℯ`, and JavaScript's `Math.E`.

```{r}
#| echo: false
peek(decay)
```

```{r}
data(decay) + point + x(amount, scale = "log", base = exp(1)) + y(hours) +
  x_label("Amount remaining") +
  title("Gridlines at each e-folding: one step is a factor of e")
```

Notice the labels: *e*, *e²*, *e³*, *e⁴*, not 2.718, 7.389, 20.09, 54.598. That
follows from one rule gog applies to every base:

> Label the quantity when it is a short number; otherwise label the power.

Base 10 gives 1, 10, 100, 1K, always short. Base 2 gives 1, 2, 4 … 1048576,
short until the number grows too wide to sit under a tick, and then `2²⁴`. Base
*e* has no short quantity above 1, so every tick past the first reads as a
power, which is what a reader counting e-foldings wants. If instead you want
the axis to read 0, 1, 2, 3 in ln units, you do not want a log scale at all.
You want `log()` applied to the column, and an axis labeled in log units, as at
the top of this chapter. A scale keeps the reader's units; a logged column
changes them.

A base only means something as the base *of* a logarithm, so gog refuses one
that has no log scale to belong to:

```{r}
#| error: true
data(octaves) + point + x(freq, base = 2) + y(level)
```

## Time: a date column gets a calendar axis

A column of dates raises a question the other columns do not. Where do the
ticks go, and what do they say, when the units are years and months rather than
tens and hundreds?

The time scale is one you never have to write. A `Date` or `POSIXct` column
*is* temporal, the type already says so, and gog reads the calendar from the
type, the same way a column of text already gets a categorical axis:

```{r}
#| echo: false
peek(revenue)
```

```{r}
data(revenue) + line + x(day) + y(sales) +
  title("Twenty-six years, ticked in years")
```

No `scale = "time"` anywhere: writing it is allowed and means nothing extra,
exactly like writing `scale = "linear"` on a number.

The ticks land on **calendar boundaries**, because the calendar is not decimal.
An axis of years steps 1, 2 or 5 like any number line. For units smaller than a
year those steps mean nothing: nobody reads an axis cut into fifths of a year.
Months step by 1, 2, 3 or 6. A gridline therefore always falls on January, and
a 3-month step marks each quarter. Weeks land on **Mondays**, and days, hours
and minutes step in the units a clock uses.

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
# six_weeks is the cast's daily orders table, see the Preface
data(six_weeks) + line + x(day) + y(orders) +
  title("Six weeks, ticked on Mondays")
```

A tick is labeled at its own resolution: `2010` for a year step, `Jan 2024` for
a month step, `Mar 4` for a day step. A `Date` column never gets clock ticks,
however narrow its range; a `POSIXct` may:

```{r}
#| echo: false
peek(monitoring)
```

```{r}
data(monitoring) + line + x(at) + y(load) +
  title("A day and a half, ticked by the clock")
```

`00:00` happens every day, so it names no particular one. The midnight tick
shows the date instead. One rule covers every label on a time axis: *a tick
names the finest unit that distinguishes it from its neighbors.*

### Bars sit on dates, and never measure one

Daily orders make a natural bar chart, one bar per day, and forty-two bars
raise the question of where the ticks go.

Bars positioned at dates are an ordinary time series. The ticks still come from
the calendar, and not one tick per bar. Wilkinson's stock-price example ticks
Sundays, not trades [@wilkinson2005], and the same rule here ticks the six
weeks of orders at week boundaries:

```{r}
data(six_weeks) + bar + x(day) + y(orders) +
  title("Daily bars, weekly ticks")
```

The other direction is refused. A bar's *length* is an amount, and a moment in
time is not an amount. A bar "reaching" 2007 would be measured from 1970, the
year a stored date counts from, which nobody chose as a baseline:

```{r}
#| error: true
data(revenue) + bar + x(sales) + y(day)
```

### What a time axis refuses

A date is stored as a number underneath. You might therefore expect a date to
take a log scale, and a number to take a time scale. Neither is allowed.

A moment has no logarithm. The calendar's zero is an arbitrary convention, and
a logarithm measured from an arbitrary zero measures nothing:

```{r}
#| error: true
data(revenue) + line + x(day, scale = "log") + y(sales)
```

Reading a type works in one direction only. gog can read the calendar off a
date column, but you cannot declare a plain number temporal at the binding. The
engine cannot know whether a number like 20656 counts days, seconds, or years:

```{r}
#| error: true
data(revenue) + point + x(sales, scale = "time") + y(sales)
```

The fix belongs in the data, where the answer is known: convert the column to
your language's date or timestamp type, which in R is `as.Date()` or
`as.POSIXct()`.

Timezones are never converted. The engine draws the clock time you gave it. A
`POSIXct` is formatted in its own timezone when it is drawn, so the axis can
never disagree with what your session showed you.
If two series must share an axis, put them in one timezone before plotting;
that decision belongs to the analysis, not the engine.

### Time on the other channels

A scale answers *how far along?* on every channel that measures, and a date
column is no exception. How recent a row is, shown as a color ramp, is a normal
question. A legend row has no ticks beside it, so a reader cannot work out the
year from a short label. Each row therefore carries the full date:

```{r}
data(six_weeks) + point + x(day) + y(orders) + color(day) +
  title("Recent days, darker")
```

## The scales

You have now met all four scales. The table names what each one does, and which
channels accept it.

| Scale | What it does |
|---|---|
| `linear` | Equal steps for equal differences. The default |
| `log` | Equal steps for equal ratios, on `x`, `y`, `color`, `size` and `opacity`, with any `base` above 1 |
| `time` | Positions dates and times, chosen from the column type: `Date` and `POSIXct` columns, on every channel that measures |
| `category` | One slot per distinct value, chosen from the column type: columns of text, and columns with a declared order |

## The axis fits your data

Where should an axis end, on the data or on the next round number past it?
Padding to a round number leaves part of the panel blank, and implies that the
data reaches further than it does.

An axis shows round numbers on its ticks, but the **range** follows your data,
not the round numbers. Each tick is a round number that falls *inside* the
data, or at most a little past it into the margin. The axis does not stretch
out to the next ten just to end on one. A span of 1952–2007 is drawn as
1952–2007, with ticks at 1960, 1980 and 2000, not padded to 1940–2020 with a
third of the panel left blank:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + area + x(year) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("The year axis fits 1952–2007; the ticks fall inside it")
```

*"Given gapminder Asia: areas, x is year, y is life, grouped by country."*

Two more rules refine that fit, and the next two plots show the reason for
each. An end the data sets is a **free end**, and a free end leaves a small
margin, so a mark never touches the panel edge. A point at the exact maximum
would otherwise be cut in half by that edge:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  x_label("GDP per capita") + title("Every point sits clear of the panel edge")
```

A baseline gets no margin. When a bar or an area measures from zero, that zero
is a real coordinate. A gap beneath it would draw the baseline in the wrong
place, so the baseline sits on the panel edge. An `area` has no glyph to clip,
so it runs to both panel edges along the axis it fills. That is why the year
area spans the full width while the income scatterplot keeps its margins. An
area's edges are part of the shape it draws. A point's edges are not.

## A scale is not only for axes

`x` and `y` turn a number into a position, but so do `color`, `size` and
`opacity`: a ramp position, a radius, a transparency. All five answer *how far
along?*, so all five take a scale. In three dimensions `z` takes one too, but
the `z` axis is always linear, so `z` refuses a log scale: see
[space](space.qmd).

This matters most for a column spread across orders of magnitude. `population`
runs from about 200 thousand to 1.3 billion. On a linear ramp the median
country sits in the bottom one percent of that range, so nearly every point
gets the same color:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population) +
  x_label("GDP per capita") + title("Linear color: nearly one flat color")
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log") +
  x_label("GDP per capita") + title("Log color: the ramp is actually used")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
population on a log scale."*

The same is true of `size`, where a linear scale leaves almost every point at
the minimum radius:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  size(population, scale = "log") + style(color = "steelblue") +
  x_label("GDP per capita") + title("Log size")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, size by
population on a log scale."*

Look again at the middle row of each of the two color legends. On a linear ramp
it is the arithmetic midpoint; on a log ramp it is the **geometric** one: √(min
· max), about 16M rather than 660M. It has to be, because that label names the
color painted at the middle of the color ramp.

`shape` and `group` take no `scale` argument. They answer *which one?*, and
there is no distance between circle and square for a scale to run along.

## `limits`: when your data is not the authority

Every rule so far derives the range from the data, which is right by default
and wrong whenever the data does not cover the whole range. The range an axis
should cover is its **domain**, and it is not always the range the data
reaches. The `tide` table holds a gauge's readings, one height every three
hours, so it never reaches midnight. An axis drawn as a circle cannot know the
missing hours are still part of the day, because nothing in a column of hours
says the day wraps. `polar()` draws the axis as a circle, and
[Polar](polar.qmd) teaches that space in full. `limits` states the domain:

```{r}
#| echo: false
peek(tide)
```

```{r}
data(tide) + line + x(hour, limits = c(0, 24)) + y(height) + polar() +
  title("The whole day, whatever hours the gauge recorded")
```

*"Given the tide: a line, x is hour from 0 to 24, y is height, in polar."*

Without that, one turn would cover 1 to 22 and the first and last readings
would be drawn on the same spoke. With it, each sits where a clock would put
it. The curve still does not close, and should not. The gauge took no reading
at midnight, and a domain says what range the axis covers rather than inventing
an observation. The gap is in the readings, and the plot draws it at the width
they left. [Polar](polar.qmd) sets out the three ways a circle does and does
not close.

`limits` is written on the binding, beside `scale` and `base`, for the same
reason they are: a domain belongs to one channel, and an atom of its own would
name that channel twice. A stated end is drawn exactly where you state it. The
margin exists to keep a glyph off the panel edge, and an end you chose is not a
data extreme that needs the room.

A domain can widen the axis as easily as it narrows it. Nothing was dropped in
that plot: the readings run 1 through 22 and all of them lie inside 0 through
24. Narrowing is the direction that removes rows, and gog reports every row it
removes:

```{r}
data(gapminder_2007) + point + x(gdp, limits = c(0, 30000)) + y(life) +
  color(continent) + x_label("GDP per capita") +
  title("Under $30,000, and the countries above it are reported")
```

It also prints a line on your console, naming how many rows fell outside and
what the domain was. This page cannot show you that line, because only refusals
appear in a rendered book. That is deliberate: removing them is what you asked
for, so the plot draws, but gog will not remove data without saying so. Ask for
a domain no row can satisfy and it refuses instead, because the alternative is
an empty panel with axes drawn over nothing.

Keeping every country on the page and marking one group of them is a different
operation, and it has a different word. `limits` runs before the transforms and
removes rows. `brush` runs after the picture is composed and removes nothing.
[Selection](selection.qmd) sets the two side by side.

You can state one end alone. The end you leave out stays a free end, which is
how you fix a baseline and let the top follow the data:

```{r}
data(gapminder_2007) + point * mean + x(continent) + y(life, limits = c(0, NA)) +
  title("y from zero, top still fitted to the data")
```

*"Given gapminder 2007: points derived by mean, x is continent, y is life from
0."*

`limits` reaches more than the axes. A domain is a property of the scale, so it
reaches every channel that measures. That is how you fix one color ramp across
plots built from different data:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log", limits = c(1e5, 2e9)) +
  x_label("GDP per capita") + title("A ramp fixed from 100K to 2B")
```

On a categorical axis `limits` is refused, and the refusal points you at the
two fixes that do the job you probably meant. A category has no range to lie
inside. Choosing *which* categories appear is filtering your table, and
choosing what order they appear in is `order()`.

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent, limits = c(0, 3)) + y(life)
```

## The column's type picks the axis

Other tools make you name a scale whenever you want to change one, so a reader
may arrive expecting to write `scale = "category"`.

The four scales are not all chosen the same way. You ask for `log`. You do not
ask for `time` or `category`. gog reads those off the column: a `Date` column
gets a calendar axis, and a column of text gets one slot per distinct value.

You may still write them. As with `scale = "time"` on a date column, writing
`scale = "category"` on a categorical column is allowed and changes nothing.
The plot below is identical to the same sentence with the scale left off:

```{r}
data(gapminder_2007) + bar * mean + x(continent, scale = "category") + y(life) +
  title("Writing it changes nothing")
```

*"Given gapminder 2007: bars derived by mean, x is continent as a category, y
is life."*

What you cannot do is use a scale to contradict the column. Ask for a
categorical axis on a number and gog refuses:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp, scale = "category") + y(life)
```

The reason behind that refusal also explains the others in this chapter. A
scale says *how* a measured column is placed. Whether the axis measures at all
is the column's type decides whether the axis measures at all. So `log` on a
categorical column is refused, and `time` on a plain number is refused. A date
column cannot be turned back into raw numbers by asking for `linear` either.
Every one of those refusals tells you to change the column, because the
column's type is what decides.

There are two things you might mean by "draw this number as categories", and
there is a different answer for each. For one slot per distinct value, declare
the column's categories: `factor()` in R and `ordered()` in the other three
bindings, both shown in [Data](data.qmd). For ranges, use `bin`, which cuts the
numbers into intervals and is covered in [Transforms](transforms.qmd).

## `tick_count`: how many ticks the axis aims for

An axis picks five ticks or so by default, and the count is the second part of
the fit you can choose, after `limits`. A crowded axis is easier to read with
fewer; an axis someone will take numbers from is easier with more. `tick_count`
is written on the binding, beside `scale`, `base` and `limits`, and for the
same reason. How many ticks an axis gets is a property of the scale, so it
belongs to the sentence that describes the scale rather than to `theme()`,
which controls the page.

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
   x_label("GDP per capita") + title("A low count: the shape, not the numbers")) |
  (data(gapminder_2007) + point + x(gdp, tick_count = 12) + y(life) +
     x_label("GDP per capita") + title("A high count: for reading values off"))
```

*"Given gapminder 2007: points, x is gdp with 3 ticks, y is life, beside the
same with 12 ticks."*

The count is a target rather than a promise. The count chooses a *step*, and
the step is then rounded to a number a person would pick. So asking for eight
on an axis running 0 to 100 gets a step of 10 and eleven ticks. What you can
rely on is the direction: ask for more and you never get fewer, and every tick
still falls on a round number.

Both of those panels cover the same range. That is the rule the count obeys: it
changes how densely an axis is *labeled*, never where the axis stops. Ask for
two ticks and the axis does not shrink to the two values it draws.

```{r}
data(gapminder_asia) + line + x(year, tick_count = 2) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("Two ticks, and still 1952 to 2007")
```

A categorical axis has one tick per category, so the data sets the count and
you do not. The refusal names the same two fixes the `limits` refusal named:

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent, tick_count = 3) + y(life)
```

And a legend is not a short axis. `limits` reaches every channel that measures,
because every one of them has a domain; `tick_count` reaches only the three
position channels that draw an axis: `x`, `y` and `z`. A color legend names
three rows, both ends and the middle, and those come from the scale's own shape
rather than from a count you could choose. The middle row is the one the color
ramps above explained: the label for the color painted halfway along the ramp.
So `color()`, `size()` and `opacity()` have no tick count to set, the same way
`shape()` has no `scale` and no `limits`: there is nothing for the count to
describe.

### Crowding in the plane and in the cube

A flat panel and a cube count their ticks differently, and neither answer is
wrong. A flat axis draws *every* tick it chose, however close together the
ticks get. The panel's width is already fixed, and the labels sit in a margin
reserved for them. An edge label that would overhang its panel is anchored
inward instead. Nothing is dropped, so every tick the axis chose is one you can
count.

A cube cannot make that promise. An axis in [space](space.qmd) is drawn at
whatever length the viewing angle leaves it. Turning the cube lengthens one
axis and shortens another, so an axis can be left with no room for its numbers.
The cube then draws as many as fit and leaves the rest out. Both are the same
rule underneath, *do not print one number through another*. They differ only in
how much room each has, because the plane can reserve room in advance and the
cube cannot. When a count you stated cannot fit, the engine says so rather than
quietly drawing fewer.

## In short

A scale changes where a number lands on the axis. It never changes the number
itself. `limits` states the domain: narrowing it removes rows and reports how
many, while a brush removes none. In a flat panel `tick_count` is a target
rather than a promise, and a cube drops the ticks that will not fit.

## What comes next

A scale decides how far along a channel a number lands. It does not decide
which visual property carries the column in the first place.
[Channels](channels.qmd) covers that: which property carries a column, and what
guide it earns.
