# Scales

```{r}
#| include: false
source("R/setup.R")
```

What do you do when most of your data is crowded into one corner? A **scale**
decides how a number becomes a position. It is written on the binding, not as
an atom of its own:

```r
x(gdp, scale = "log")
```

Scales are a property of the axis, so they live where the axis is bound. There
is no `scale()` atom to write separately: that would make you name the channel
twice, once to bind it and once to scale it, and the two could then disagree.

There is a second gain, and it is about remembering rather than being correct.
The scale sits inside the thing it measures, so there is only one place to
look. A reader who wants to know how `gdp` is being read finds the answer in
`x(gdp, scale = "log")` itself. Nothing else in the sentence can change it, and
there is no second function to go and find.

## A scale is not a transform

`bin` and `mean` are **transforms**: they read your data and produce different
data. A scale produces no new values at all. It changes how far along the axis a
number lands and what the ticks say, and nothing else.

The difference shows up in what the reader sees. Here is `gdp` plotted with a
log **scale**:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) +
  x_label("GDP per capita") +
  title("Income and life expectancy: log scale")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by continent."*

The axis reads 100, 1K, 10K, 100K. Those are dollars, the units you supplied.

The word *log* appears nowhere on the plot, and that is deliberate. The units
did not change, so the axis name does not claim they did. The ticks are the
only sign, and they carry it by their spacing: whole powers of ten, evenly
placed. Write `x_label("GDP per capita (log scale)")` when a reader needs it
in words. The axis name is yours, and gog never writes the scale into it.

Now the same picture drawn by logging the column yourself:

```{r}
gm_logged <- gapminder_2007
gm_logged$log_gdp <- log10(gm_logged$gdp)

data(gm_logged) + point + x(log_gdp) + y(life) +
  color(continent) +
  title("The same shape, in units nobody asked for")
```

Identical shape, and the axis now reads 2, 3, 4, 5. To know what a point is
worth, the reader has to raise ten to a power in their head. That arithmetic is
the work a scale exists to do for them.

So the choice between the two is about what the quantity *is*. Log the column
when the logarithm is itself the quantity of interest: a log-odds, a decibel,
a pH. Use a log scale when the quantity is still dollars or people, and you
only want to see it across orders of magnitude.

## Where a scale sits in the pipeline

Once a transform is in the plot, "the scale is only about display" stops being
quite true, because a transform has to decide *which space* to work in. gog
follows one rule, and it has no per-transform exceptions:

> A scale applies *before* the transform on the axis it groups by, and
> *after* it on the axis the transform writes.

### Grouping happens in the space you will see

`bin` groups by x, so a log x is applied first and the bins are cut in log
space. Each bin then spans one constant *ratio*, so the bars land at a constant
spacing and cover the axis:

```{r}
data(gapminder_2007) + bar * bin + x(gdp, scale = "log") +
  x_label("GDP per capita") +
  title("Bins cut in log space are evenly spaced")
```

Had the bins been cut in dollars and only the *drawing* been logarithmic, they
would have bunched into the right-hand end. On this data the gaps between bars
run 176, 82, 54, 40, 32, 27 pixels, and the left half of the plot is empty. That
reads as a broken picture rather than as a modeling choice, which is why the
grouping axis is scaled first.

### The measured value keeps its own units

`sum` writes to y, so a log y is applied *after* the sum. The bars below total
100 and 10:

```{r}
receipts <- data.frame(
  store = c("North", "North", "South", "South"),
  sales = c(10.0, 90.0, 1.0, 9.0)
)

data(receipts) + bar * sum + x(store) + y(sales, scale = "log") +
  title("A sum is still a sum")
```

If the scale had been applied first, gog would have added the *logarithms*,
giving the log of a product, which is not a quantity anyone asked for. Summing
first and displaying second keeps `sum` meaning what its name says.

The order is worth knowing if you arrive from ggplot2, where it runs the other
way: under `scale_y_log10()`, a `stat_summary(fun = mean)` averages the logged
values and returns a geometric mean. gog computes the summary in your data's
units and scales the result, so the number the bar stands for is the number you
would get from `sum()` or `mean()` yourself.

### A smoother fits within its groups

`smooth` groups by x, so a log x means the curve is fitted against log income,
which is what makes a straight line out of a relationship that is otherwise all
bunched into the left margin:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  style(color = "lightgrey") +
  line * smooth + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") +
  title("A LOESS fitted against log income")
```

## What a log scale refuses

A logarithm is undefined at zero and below, so those rows have no position on
the axis. gog says so rather than dropping them quietly:

```{r}
#| error: true
readings <- data.frame(
  depth = c(1.0, 0.0, -4.0, 100.0),
  temp  = c(12.0, 14.0, 15.0, 9.0)
)

data(readings) + point + x(depth, scale = "log") + y(temp)
```

The message names how many rows are affected and how far down the column goes,
because "some of your data cannot be drawn" is not enough to act on.

Text has no logarithm either, and a text column already gets a categorical axis
without being asked:

```{r}
#| error: true
data(gapminder_2007) + point + x(continent, scale = "log") + y(life)
```

## Choosing the base

`scale = "log"` means base 10 unless you say otherwise:

```r
x(freq, scale = "log", base = 2)
```

The base is a **number**, not a name. There is no `"log2"` or `"ln"` scale,
because that would enumerate what one parameter already derives, the same
objection that rules out ggplot2's `scale_x_log10` / `scale_x_sqrt` family.

### The base is almost entirely cosmetic

This is the surprising part, and it is worth knowing before you use it.
log_b(x) differs from log₁₀(x) by a constant factor, and gog normalizes every
axis by its own range, so the factor cancels and **every base draws the same
picture**. It even survives the transforms: bins of equal width in log₂ space
are equal width in log₁₀ space, so `bar * bin` cuts at the same values whichever
base you name.

What the base actually chooses is **where the gridlines fall and how they read**.
Here are the same six octaves on base 10 and base 2:

```{r}
octaves <- data.frame(
  freq  = c(55.0, 110.0, 220.0, 440.0, 880.0, 1760.0),
  level = c(3.0, 6.0, 9.0, 7.0, 4.0, 2.0)
)

data(octaves) + point + x(freq, scale = "log") + y(level) +
  x_label("Frequency (Hz)") + title("Base 10: gridlines at 100, 1K")
```

```{r}
data(octaves) + point + x(freq, scale = "log", base = 2) + y(level) +
  x_label("Frequency (Hz)") + title("Base 2: gridlines on every octave")
```

The points have not moved. The second chart puts a gridline on each doubling,
which is what "an octave" means, so for pitch, bits, or doubling times, base 2
is the one that makes the axis say something.

### Natural log

R has no `e` constant, so it is `base = exp(1)`:

```{r}
decay <- data.frame(
  hours  = 0:5 + 0.0,
  amount = exp(-(0:5)) * 100
)

data(decay) + point + x(amount, scale = "log", base = exp(1)) + y(hours) +
  x_label("Amount remaining") +
  title("Gridlines at each e-folding")
```

Notice the labels: **e, e², e³**, not 2.718, 7.389, 20.09. That follows from one
rule gog applies to every base:

> Label the quantity when it reads cleanly; otherwise label the power.

Base 10 gives 1, 10, 100, 1K, always clean. Base 2 gives 1, 2, 4 … 1048576,
clean until it grows too wide to sit under a tick, and then `2²⁴`. Base *e*
has no clean quantities at all, so it always reads as a power, which is
precisely what somebody counting e-foldings wanted. If instead you want the
axis to read 0, 1, 2, 3 in ln units, you do not want a log scale at all. You
want `log()` applied to the column, and an axis honestly labeled in log units,
as at the top of this chapter. A scale keeps the reader's units; a logged
column changes them.

A base only means something as the base *of* a logarithm, so gog refuses one
that has no log scale to belong to:

```{r}
#| error: true
data(octaves) + point + x(freq, base = 2) + y(level)
```

## Time: a date reads as a calendar

The time scale is the one scale you never write. A `Date` or `POSIXct` column
*is* temporal, the type already says so, and gog reads the calendar from the
type, the same way a text column already gets a categorical axis:

```{r}
revenue <- data.frame(
  day   = as.Date(paste0(1994:2019, "-06-30")),
  sales = c(52, 57, 55, 61, 64, 60, 68, 66, 71, 75, 74, 80, 78, 85,
            72, 72, 79, 84, 88, 87, 93, 96, 94, 101, 105, 103)
)

data(revenue) + line + x(day) + y(sales) +
  title("Twenty-six years, ticked in years")
```

No `scale = "time"` anywhere: writing it is allowed and means nothing extra,
exactly like writing `scale = "linear"` on a number.

The ticks land on **calendar boundaries**, because the calendar is not decimal.
An axis of years walks 1-2-5 like any other number line, but below the year the
`nice_number` habit has no meaning: nobody reads an axis cut in fifths of a
year. Months step by 1, 2, 3 or 6 so that January stays a gridline and
3-month steps read as quarters; weeks land on **Mondays**; days, hours and
minutes take the steps a clock face suggests.

```{r}
# six_weeks is the cast's daily orders table, see the Preface
data(six_weeks) + line + x(day) + y(orders) +
  title("Six weeks, ticked on Mondays")
```

A tick is labeled at its own resolution: `2010` for a year step, `Jan 2024`
for a month step, `Mar 4` for a day step. A `Date` column never grows clock
ticks, however narrow its range; a `POSIXct` may:

```{r}
monitoring <- data.frame(
  at   = as.POSIXct("2024-03-04 06:00", tz = "UTC") + 3600 * 0:36,
  load = round(40 + 25 * sin(0:36 / 4) + (0:36 %% 5))
)

data(monitoring) + line + x(at) + y(load) +
  title("A day and a half, ticked by the clock")
```

Midnight's clock face says nothing, `00:00` happens every day, so the
midnight tick borrows the day's name instead. That is the label rule stated
once: *a tick names the finest unit that distinguishes it from its
neighbors.*

### Bars sit on dates; they never measure one

Bars positioned at dates are an ordinary time series, and the ticks still come
from the calendar rather than one per bar. Wilkinson's stock-price example
[@wilkinson2005] ticks Sundays, not trades:

```{r}
data(six_weeks) + bar + x(day) + y(orders) +
  title("Daily bars, weekly ticks")
```

The other direction is refused. A bar's *length* is an amount, and a moment in
time is not an amount. A bar "reaching" 2007 would be measured from 1970,
an origin nobody chose:

```{r}
#| error: true
data(revenue) + bar + x(sales) + y(day)
```

### What a time axis refuses

A moment has no logarithm. The calendar's zero is an arbitrary convention,
and a logarithm measured from an arbitrary zero measures nothing:

```{r}
#| error: true
data(revenue) + line + x(day, scale = "log") + y(sales)
```

And the request only runs one way. A date column is temporal by type, so a
plain number cannot be *declared* temporal at the binding. The engine cannot
know whether `20656` means a day, a second, or a year:

```{r}
#| error: true
data(revenue) + point + x(sales, scale = "time") + y(sales)
```

The fix belongs in the data, where the answer is known: convert the column
with `as.Date()` or `as.POSIXct()`.

A word on timezones: the engine is naive on purpose. It draws the clock time
you see in R: a `POSIXct` is formatted in its own timezone on the way out and
never converted again, so the axis can never disagree with what `print()`
showed you. If two series must share an axis, put them in one timezone before
plotting; that decision belongs to the analysis, not the renderer.

### Time on the other channels

A scale answers *how far along?* on every channel that measures, and a date
column is no exception: recency as a color ramp is a perfectly good
question. A legend row has no neighboring ticks to borrow context from, so
it carries the full self-contained date:

```{r}
data(six_weeks) + point + x(day) + y(orders) + color(day) +
  title("Recent days, darker")
```

## The scales

| Scale | What it does |
|---|---|
| `linear` | Equal steps for equal differences. The default |
| `log` | Equal steps for equal ratios, on `x`, `y`, `color`, `size` and `opacity`, with any `base` |
| `time` | Positions dates and times, chosen from the column type: `Date` and `POSIXct` columns, on every channel that measures |
| `category` | One slot per distinct value, chosen from the column type: text and factor columns |

## The axis fits your data

An axis shows round numbers on its ticks, but the **range** follows your data,
not the round numbers. The ticks are the nice values that fall *inside* the
data; the axis does not stretch out to the next ten just to end on one. A span
of 1952–2007 is drawn as 1952–2007, with ticks at 1960, 1980 and 2000, not
padded to 1940–2020 with a third of the panel left blank:

```{r}
data(gapminder_asia) + area + x(year) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("The year axis fits 1952–2007; the ticks fall inside it")
```

Two things are calibrated on top of that fit, each for a reason you can see in
the plot. A **free end breathes** a little, so a mark is not welded to the
frame. A point at the exact maximum would otherwise be sliced in half by the
panel edge:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  x_label("GDP per capita") + title("Points breathe off every edge")
```

A baseline does not breathe. When a bar or an area measures from zero, that
zero is a real coordinate, and a gap beneath it would draw the baseline where
the baseline is not, so it sits flush. An `area` also has no glyph to clip, so
on the axis it fills along it reaches the panel edges instead of leaving an
empty strip each side. That is why the area above spans the full width while the
scatter here keeps its margins: a fill's edges *are* the shape, a point's are
not.

## A scale is not only for axes

`x` and `y` turn a number into a position, but so do `color`, `size` and
`opacity`: a ramp position, a radius, a transparency. All four answer *how far
along?*, so all four take a scale.

This matters most for a column spread across orders of magnitude. `population`
runs from about 200 thousand to 1.3 billion, and on a linear ramp the median
country sits in the bottom 2% of the range, so nearly every point comes out
the same color:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population) +
  x_label("GDP per capita") + title("Linear color: nearly one flat color")
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log") +
  x_label("GDP per capita") + title("Log color: the ramp is actually used")
```

The same is true of `size`, where a linear scale leaves almost every point at
the minimum radius:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  size(population, scale = "log") + style(color = "steelblue") +
  x_label("GDP per capita") + title("Log size")
```

Look at the middle of the color legend above. On a linear ramp it is the
arithmetic midpoint; on a log ramp it is the **geometric** one: √(min · max),
about 16M rather than 660M. It has to be, because that label names the color
painted at the middle of the strip.

`shape` and `group` take no `scale` argument. They answer *which one?*, and
there is no distance between circle and square for a scale to run along.

## `limits`: when your data is not the authority

Everything above derives the range from the data, which is right by default and
wrong whenever the data is not the whole story. A tide gauge reading every three
hours never reaches midnight, and a periodic axis has no way to know the missing
hours are still part of the day, and nothing in a column of hours says the day
wraps. `limits` says the domain out loud:

```{r}
data(tide) + line + x(hour, limits = c(0, 24)) + y(height) + polar() +
  title("The whole day, whatever the gauge happened to catch")
```

Without that, one turn would cover 1 to 22 and the first and last readings would
be drawn on the same spoke. With it, each sits where a clock would put it. The
curve still does not close, and should not: the gauge took no reading at
midnight, and a domain says what range the axis covers rather than inventing an
observation to fill it. That gap is the data's, drawn at its true width.
[Polar](polar.qmd) sets out the three ways a circle does and does not close.

`limits` is written on the binding, beside `scale` and `base`, for the same
reason they are: a domain belongs to one channel, and an atom of its own would
name that channel twice. A stated end is drawn exactly where you state it, and the
breathing margin exists to keep a glyph off the frame, and an end you chose is
not a data extreme that needs the room.

Stating a domain widens as readily as it narrows. Nothing was dropped above:
the readings run 1 through 22 and all of them lie inside 0 through 24. Narrowing
is the direction that removes rows, and it says so:

```{r}
data(gapminder_2007) + point + x(gdp, limits = c(0, 30000)) + y(life) +
  color(continent) + x_label("GDP per capita") +
  title("Under $30,000, and the countries above it are counted aloud")
```

Rendering that prints a line to the console naming how many rows fell outside
and what the domain was. That is deliberate: removing them is what you asked
for, so the plot draws, but gog will not remove data without saying so. Ask for
a domain no row can satisfy and it refuses instead, because the alternative is
an empty panel with axes drawn over nothing.

Keeping every country on the page and making one group of them stand out is a
different operation, and it has a different word. `limits` runs before the
statistics and removes rows. `brush` runs after the picture is composed and
removes nothing. [Selection](selection.qmd) sets the two side by side.

**Either end alone.** `NA` leaves that end to the data, which is how you pin a
baseline without capping the top:

```{r}
data(gapminder_2007) + point * mean + x(continent) + y(life, limits = c(0, NA)) +
  title("y from zero, top still fitted to the data")
```

It is not only for axes. A domain is a property of the scale, so it reaches
every channel that measures, which is how you hold one color ramp still across
plots that have different data in them:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log", limits = c(1e5, 2e9)) +
  x_label("GDP per capita") + title("A ramp fixed from 100K to 2B")
```

On a categorical axis `limits` is refused, and the refusal points you at the
atom that does the job you probably meant. A category has no range to lie
inside. Choosing *which* categories appear is filtering your table, and
choosing what order they appear in is `order()`.

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent, limits = c(0, 3)) + y(life)
```

## The column's type picks the axis

The scales in the table above are not all chosen the same way. You ask for
`log`. You do not ask for `time` or `category`, because gog reads those off the
column: a `Date` column gets a calendar axis, and a text column gets one slot
per distinct value.

You may still say them out loud. Writing `scale = "category"` on a text column
is allowed and changes nothing, in the same way `scale = "linear"` on a number
changes nothing. The plot below is identical to the one without it:

```{r}
data(gapminder_2007) + bar * mean + x(continent, scale = "category") + y(life) +
  title("Saying it out loud draws the same plot")
```

What you cannot do is use a scale to contradict the column. Ask for a
categorical axis on a number and gog refuses:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp, scale = "category") + y(life)
```

The reason behind that refusal also explains three others in this chapter. A
scale says **how** a measured column is placed. Whether the axis measures at
all is the column's type. So `log` on a text column is refused, `time` on a
plain number is refused, and a date column cannot be turned back into raw
numbers by asking for `linear`. Every one of those refusals tells you to change
the column, because the column is where the answer lives.

There are two things you might mean by "draw this number as categories", and
gog has a word for each. For one slot per distinct value, make the column text:
a factor in R, `ordered()` in the other three bindings, both shown in
[Data](data.qmd). For ranges, use `bin`, which cuts the numbers into intervals
and is covered in [Transforms](transforms.qmd).

## `tick_count`: how many ticks the axis aims for

An axis picks five ticks or so by default, and the count is the one thing
about the fit you may want a say in. A crowded axis reads better with fewer;
an axis someone will read values off reads better with more. `tick_count` is
written on the binding, beside `scale`, `base` and `limits`, and for the same
reason. How many ticks an axis gets is a property of the **scale**, so it
belongs to the sentence that describes the scale rather than to `theme()`,
which is the page.

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
   x_label("GDP per capita") + title("Three ticks: the shape, not the numbers")) |
  (data(gapminder_2007) + point + x(gdp, tick_count = 12) + y(life) +
     x_label("GDP per capita") + title("Twelve: for reading values off"))
```

It is a **target rather than a promise**, and the difference is worth
understanding because it is not a defect. The count chooses a *step*, and the
step is then rounded to a number a person would pick, so asking for eight on an
axis running 0 to 100 gets a step of 10 and eleven ticks. What you can rely on is
the direction: ask for more and you get more, and every tick still falls on a
round number.

The two ends of the range are the same range in both panels above. That is the
rule the parameter is careful about: it changes how densely an axis is
*labeled*, never where the axis stops. Ask for two ticks and the axis does not
shrink to the two values it draws.

```{r}
data(gapminder_asia) + line + x(year, tick_count = 2) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("Two ticks, and still 1952 to 2007")
```

A **categorical axis has one tick per category**, so the count there is the
data's rather than yours, and the refusal points at the two atoms that do the
jobs you might have meant:

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent, tick_count = 3) + y(life)
```

And a legend is not a short axis. `limits` reaches every channel that
measures, because every one of them has a domain; `tick_count` reaches only
the three that draw an axis. A color bar names three rows, both ends and the
middle, and those come from the scale's own shape rather than from a count you
could choose. On a log ramp the middle row is the geometric mean, the label
for the color actually painted halfway along the strip. So `color()`, `size()`
and `opacity()` take no `tick_count` at all, the same way `shape()` takes no
`scale` and no `limits`: the parameter is absent rather than refused, which
is the strongest form the refusal can take.

### Crowding in the plane and in the cube

Worth knowing if you read the same plot both ways, because the two answers
look inconsistent and are not. A flat axis draws *every* tick it chose,
however tight they get. The panel is as wide as it is going to be, and the
labels sit in a margin reserved for them. An edge label that would overhang
its panel is anchored inward instead. Nothing is dropped, so what you asked
for is what you count.

A cube cannot make that promise. An axis in [space](space.qmd) is drawn at
whatever length the viewing angle leaves it, and tilted far enough it has no room
for its numbers at all, so the frame draws as many as fit and leaves the rest
out. Both are the same rule underneath, *do not print one number through
another*, and they differ only in what is available to spend: the plane can
reserve room in advance, and the cube cannot. When a count you stated has to be
thinned that way, the engine says so rather than quietly drawing fewer.
