# Scales

```{r}
#| include: false
source("R/setup.R")
```

What do you do when most of your data is crowded into one corner? You change
the **scale**, which decides how a number becomes a position. It is written
on the binding, not as an atom of its own:

```r
x(gdp, scale = "log")
```

Scales are a property of the axis, so they live where the axis is bound. There
is no `scale()` atom to write separately. That would make you name the channel
twice, once to bind it and once to scale it, and the two could then disagree.

Writing the scale on the binding helps a second way, and that way is about
remembering rather than being correct.
The scale sits inside the thing it measures, so there is only one place to
look. A reader who wants to know how `gdp` is being read finds the answer in
`x(gdp, scale = "log")` itself. Nothing else in the sentence can change it, and
there is no second function to look up.

## A scale is not a transform

You could also take the logarithm of the column yourself and plot the result.
That is a different act, and the difference decides what the axis says.

`bin` and `mean` are **transforms**: they read your data and produce different
data. A scale produces no new values at all. It changes how far along the axis a
number lands and what the ticks say, and nothing else.

The difference is visible in the finished plot. Here is `gdp` plotted with a
log **scale**:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) +
  x_label("GDP per capita") +
  title("Income and life expectancy: log scale")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

The axis reads 1K and 10K. Those are dollars, the units you supplied.

The word *log* appears nowhere on the plot, and that is deliberate. The units
did not change, so the axis name does not claim they did. The ticks are the only
sign: each is a whole power of ten, so equal steps along the axis mean equal
ratios. Write `x_label("GDP per capita (log scale)")` when a reader needs it in
words. The axis name is yours, and gog never writes the scale into it.

Now the same picture drawn by logging the column yourself:

```{r}
gm_logged <- gapminder_2007
gm_logged$log_gdp <- log10(gm_logged$gdp)

data(gm_logged) + point + x(log_gdp) + y(life) +
  color(continent) +
  title("The same shape, with the axis in log units")
```

Identical shape, and the axis now reads 3 and 4. To know a country's income, the
reader has to raise ten to that power. That arithmetic is the work a scale
exists to do for them.

So the choice between the two is about what the quantity *is*. Log the column
when the logarithm is itself the quantity of interest: a log-odds, a decibel,
a pH. Use a log scale when the quantity is still dollars or people, and you
only want to see it across orders of magnitude.

## The order of a scale and a transform

Once a transform is in the plot, a scale is no longer only about display,
because a transform has to choose which numbers to work on. gog
follows one rule, and it has no per-transform exceptions:

> A transform groups by one axis and measures on the other. The scale is
> applied before the transform on the grouping axis, and after it on the
> measured axis.

### Scaling the grouping axis

A histogram of income is the first place the rule matters. Its bins have to be
cut somewhere, in dollars or in their logarithms.

`bin` groups by x, so a log x is applied first. The bins are cut on the
logarithms, not on the dollars. Each bin then spans one constant *ratio*, so the
bars land at a constant spacing and cover the axis:

```{r}
data(gapminder_2007) + bar * bin + x(gdp, scale = "log") +
  x_label("GDP per capita") +
  title("Bins cut in log space are evenly spaced")
```

*"Given gapminder 2007: bars derived by bin, x is gdp on a log scale."*

Suppose the bins had been cut in dollars and only the *drawing* had been
logarithmic. The first bin alone would cover most of the axis, and the rest
would crowd into the right edge. That reads as a broken picture rather than as a
modeling choice, which is why the grouping axis is scaled first.

### The measured value keeps its own units

Put the log on the measured axis instead and a different question arises. Does a
bar on a log scale stand for a sum of dollars, or for a sum of logarithms?

`sum` writes to y, so a log y is applied *after* the sum. The two bars below
stand for 100 and 10:

```{r}
#| echo: false
peek(receipts)
```

```{r}
data(receipts) + bar * sum + x(store) + y(sales, scale = "log") +
  title("A sum is still a sum")
```

*"Given the receipts: bars derived by sum, x is store, y is sales on a log
scale."*

If the scale had been applied first, gog would have added the *logarithms*,
giving the log of a product, which is not a quantity anyone asked for. Summing
first and displaying second keeps `sum` meaning what its name says.

The order is worth knowing if you have used ggplot2, which does the opposite.
Under `scale_y_log10()`, a `stat_summary(fun = mean)` averages the logged
values. What comes back is a geometric mean, the average computed on the logs.
gog computes the summary in your data's units and scales the result, so the
number the bar stands for is the number you would get from `sum()` or `mean()`
yourself.

### `smooth` fits inside its groups

The same question arises from the grouping side: is the curve fitted to dollars,
or to their logarithms? On incomes that run from a few hundred dollars to nearly
fifty thousand, the answer changes the shape of the curve.

`smooth` groups by x, so a log x fits the curve against log income. That is what
straightens a relationship otherwise crowded into the left of the panel:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  style(color = "lightgray") +
  line * smooth + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") +
  title("A trend fitted against log income")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, and also a
line derived by smooth, x is gdp on a log scale, y is life."*

## What a log scale refuses

Some columns cannot take a log scale, and you will meet them: a depth that goes
below zero, a balance that went negative, a count of zero.

A logarithm is undefined at zero and below, so those rows have no position on
the axis. gog says so rather than dropping them quietly:

```{r}
#| echo: false
peek(depth_readings)
```

```{r}
#| error: true
data(depth_readings) + point + x(depth, scale = "log") + y(temp)
```

The message names how many rows are affected and the smallest value in the
column, because "some of your data cannot be drawn" is not enough to act on.

Text has no logarithm either, and a column of text already gets a categorical axis
without being asked:

```{r}
#| error: true
data(gapminder_2007) + point + x(continent, scale = "log") + y(life)
```

## Choosing the base

Ten is the usual base for money and for population, and not for everything you
measure. A pitch doubles with every octave, and a decay is counted in factors of
*e*.

`scale = "log"` means base 10 unless you say otherwise:

```r
x(freq, scale = "log", base = 2)
```

The base is a **number**, not a name. There is no `"log2"` or `"ln"` scale,
because that would enumerate what one parameter already derives, the same
objection that rules out ggplot2's `scale_x_log10` family.

### The base changes the labels, not the picture

One result here surprises most people, and it is worth knowing before you set a
base. Any two logarithms of the same column differ by a constant factor, and gog
normalizes every axis by its own range. The factor cancels, so the points land
in *the same places in every base*. The transforms are unaffected too: bins of
equal width in log₂ space are equal width in log₁₀ space, so `bar * bin` cuts at
the same values whichever base you name.

What the base actually chooses is where the gridlines fall and how they read.
Here are six frequencies, each one an octave above the last, on base 10 and
then base 2:

```{r}
#| echo: false
peek(octaves)
```

```{r}
data(octaves) + point + x(freq, scale = "log") + y(level) +
  x_label("Frequency (Hz)") + title("Base 10: gridlines at 100, 1K")
```

```{r}
data(octaves) + point + x(freq, scale = "log", base = 2) + y(level) +
  x_label("Frequency (Hz)") + title("Base 2: gridlines on every octave")
```

*"Given the octaves: points, x is freq on a log scale in base 2, y is level."*

The points have not moved. The second plot puts a gridline on each doubling,
which is what "an octave" means, so for pitch, bits, or doubling times, base 2
is the one whose labels name what you are counting.

### Natural log

The natural base is for a quantity that decays at a constant rate. The amount
below falls by a factor of *e* every hour, so one gridline falls between every
two consecutive readings.

R has no `e` constant, so it is `base = exp(1)`:

```{r}
#| echo: false
peek(decay)
```

```{r}
data(decay) + point + x(amount, scale = "log", base = exp(1)) + y(hours) +
  x_label("Amount remaining") +
  title("Gridlines at each e-folding: one step is a factor of e")
```

Notice the labels: **e, e², e³**, not 2.718, 7.389, 20.09. That follows from one
rule gog applies to every base:

> Label the quantity when it is a short number; otherwise label the power.

Base 10 gives 1, 10, 100, 1K, always short. Base 2 gives 1, 2, 4 … 1048576,
short until the number grows too wide to sit under a tick, and then `2²⁴`.
Base *e* has no short quantity above 1, so every tick past the first reads as a
power, which is what a reader counting e-foldings wants. If instead you want the
axis to read 0, 1, 2, 3 in ln units, you do not want a log scale at all. You
want `log()` applied to the column, and an axis honestly labeled in log units,
as at the top of this chapter. A scale keeps the reader's units; a logged column
changes them.

A base only means something as the base *of* a logarithm, so gog refuses one
that has no log scale to belong to:

```{r}
#| error: true
data(octaves) + point + x(freq, base = 2) + y(level)
```

## Time: a date column gets a calendar axis

A column of dates raises a question the other columns do not. Where do the ticks
go, and what do they say, when the units are years and months rather than tens
and hundreds?

The time scale is one you never have to write. A `Date` or `POSIXct` column
*is* temporal, the type already says so, and gog reads the calendar from the
type, the same way a column of text already gets a categorical axis:

```{r}
#| echo: false
peek(revenue)
```

```{r}
data(revenue) + line + x(day) + y(sales) +
  title("Twenty-six years, ticked in years")
```

No `scale = "time"` anywhere: writing it is allowed and means nothing extra,
exactly like writing `scale = "linear"` on a number.

The ticks land on **calendar boundaries**, because the calendar is not decimal.
An axis of years steps 1, 2 or 5 like any number line. Below the year those
steps mean nothing: nobody reads an axis cut in fifths of a year. Months step by
1, 2, 3 or 6, so January stays a gridline and 3-month steps read as quarters.
Weeks land on **Mondays**, and days, hours and minutes take the steps a clock
face suggests.

```{r}
#| echo: false
peek(six_weeks)
```

```{r}
# six_weeks is the cast's daily orders table, see the Preface
data(six_weeks) + line + x(day) + y(orders) +
  title("Six weeks, ticked on Mondays")
```

A tick is labeled at its own resolution: `2010` for a year step, `Jan 2024` for
a month step, `Mar 4` for a day step. A `Date` column never gets clock ticks,
however narrow its range; a `POSIXct` may:

```{r}
#| echo: false
peek(monitoring)
```

```{r}
data(monitoring) + line + x(at) + y(load) +
  title("A day and a half, ticked by the clock")
```

`00:00` happens every day, so it names no particular one. The midnight tick
shows the date instead. One rule covers every label on a time axis: *a
tick names the finest unit that distinguishes it from its neighbors.*

### Bars sit on dates; they never measure one

Daily orders make a natural bar chart, one bar per day, and forty-two bars raise
the question of where the ticks go.

Bars positioned at dates are an ordinary time series, and the ticks still
come from the calendar rather than one per bar. Wilkinson's stock-price
example ticks Sundays, not trades [@wilkinson2005], and the same rule here
ticks the six weeks of orders at week boundaries:

```{r}
data(six_weeks) + bar + x(day) + y(orders) +
  title("Daily bars, weekly ticks")
```

The other direction is refused. A bar's *length* is an amount, and a moment in
time is not an amount. A bar "reaching" 2007 would be measured from 1970,
an origin nobody chose:

```{r}
#| error: true
data(revenue) + bar + x(sales) + y(day)
```

### What a time axis refuses

A date is stored as a number underneath. You might therefore expect a date to
take a log scale, and a number to take a time scale. Neither is allowed.

A moment has no logarithm. The calendar's zero is an arbitrary convention,
and a logarithm measured from an arbitrary zero measures nothing:

```{r}
#| error: true
data(revenue) + line + x(day, scale = "log") + y(sales)
```

The reading works in one direction only. A date column is temporal by type, so a
plain number cannot be *declared* temporal at the binding. The engine cannot
know whether a plain number like 20656 counts days, seconds, or years:

```{r}
#| error: true
data(revenue) + point + x(sales, scale = "time") + y(sales)
```

The fix belongs in the data, where the answer is known: convert the column
with `as.Date()` or `as.POSIXct()`.

Timezones are never converted. The engine draws the clock time you gave it. A
`POSIXct` is formatted in its own timezone when it is drawn, so the axis can
never disagree with what `print()` showed you.
If two series must share an axis, put them in one timezone before plotting; that
decision belongs to the analysis, not the engine.

### Time on the other channels

A scale answers *how far along?* on every channel that measures, and a date
column is no exception: how recent a row is, shown as a color ramp, is a normal
question. A legend row stands alone, with no neighboring ticks to give it
context, so it carries the full date:

```{r}
data(six_weeks) + point + x(day) + y(orders) + color(day) +
  title("Recent days, darker")
```

## The scales

| Scale | What it does |
|---|---|
| `linear` | Equal steps for equal differences. The default |
| `log` | Equal steps for equal ratios, on `x`, `y`, `color`, `size` and `opacity`, with any `base` above 1 |
| `time` | Positions dates and times, chosen from the column type: `Date` and `POSIXct` columns, on every channel that measures |
| `category` | One slot per distinct value, chosen from the column type: columns of text, and columns with a declared order |

## The axis fits your data

Where should an axis end, on the data or on the next round number past it?
Padding to a round number leaves part of the panel blank, and implies that the
data reaches further than it does.

An axis shows round numbers on its ticks, but the **range** follows your data,
not the round numbers. Each tick is a round number that falls *inside* the
data, or at most a little past it into the margin; the axis does not stretch out
to the next ten just to end on one. A span of 1952–2007 is drawn as 1952–2007,
with ticks at 1960, 1980 and 2000, not padded to 1940–2020 with a third of the
panel left blank:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + area + x(year) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("The year axis fits 1952–2007; the ticks fall inside it")
```

*"Given gapminder Asia: areas, x is year, y is life, grouped by country."*

Two more rules refine that fit, and the plot shows the reason for each. A *free
end* leaves a small margin, so a mark never touches the panel edge. A point at
the exact maximum would otherwise be clipped by the panel edge:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  x_label("GDP per capita") + title("Every point sits clear of the panel edge")
```

A baseline gets no margin. When a bar or an area measures from zero, that zero
is a real coordinate. A gap beneath it would draw the baseline in the wrong
place, so the baseline sits on the panel edge. An `area` has no glyph to clip,
so along the axis it fills it runs to both panel edges. That is why the year
area spans the full width while the income scatterplot keeps its margins: a
fill's edges *are* the shape, a point's are not.

## A scale is not only for axes

`x` and `y` turn a number into a position, but so do `color`, `size` and
`opacity`: a ramp position, a radius, a transparency. All five answer *how far
along?*, so all five take a scale. In three dimensions `z` takes one too, but
its axis is linear, so a log depth is refused: see [space](space.qmd).

This matters most for a column spread across orders of magnitude.
`population` runs from about 200 thousand to 1.3 billion. On a linear ramp the
median country sits in the bottom 2% of that range, so nearly every point gets
the same color:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population) +
  x_label("GDP per capita") + title("Linear color: nearly one flat color")
```

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log") +
  x_label("GDP per capita") + title("Log color: the ramp is actually used")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
population on a log scale."*

The same is true of `size`, where a linear scale leaves almost every point at
the minimum radius:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  size(population, scale = "log") + style(color = "steelblue") +
  x_label("GDP per capita") + title("Log size")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, size by
population on a log scale."*

Look again at the middle row of the two color legends further up. On a linear
ramp it is the arithmetic midpoint; on a log ramp it is the **geometric** one:
√(min · max), about 16M rather than 660M. It has to be, because that label names
the color painted at the middle of the color ramp.

`shape` and `group` take no `scale` argument. They answer *which one?*, and
there is no distance between circle and square for a scale to run along.

## `limits`: when your data is not the authority

Everything above derives the range from the data, which is right by default and
wrong whenever the data does not cover the domain. A tide gauge reading every
three hours never reaches midnight. An axis drawn as a circle cannot know the
missing hours are still part of the day, because nothing in a column of hours
says the day wraps. The `tide` table holds a gauge's readings, one height every
three hours. `polar()` draws the axis as a circle, and [Polar](polar.qmd)
teaches that space in full. `limits` states the domain:

```{r}
#| echo: false
peek(tide)
```

```{r}
data(tide) + line + x(hour, limits = c(0, 24)) + y(height) + polar() +
  title("The whole day, whatever hours the gauge recorded")
```

*"Given the tide: a line, x is hour from 0 to 24, y is height, in polar."*

Without that, one turn would cover 1 to 22 and the first and last readings would
be drawn on the same spoke. With it, each sits where a clock would put it. The
curve still does not close, and should not. The gauge took no reading at
midnight, and a domain says what range the axis covers rather than inventing an
observation. That gap is the data's, drawn at its true width. [Polar](polar.qmd)
sets out the three ways a circle does and does not close.

`limits` is written on the binding, beside `scale` and `base`, for the same
reason they are: a domain belongs to one channel, and an atom of its own would
name that channel twice. A stated end is drawn exactly where you state it. The
margin exists to keep a glyph off the panel edge, and an end you chose is not a
data extreme that needs the room.

A domain can widen the axis as easily as it narrows it. Nothing was dropped
above: the readings run 1 through 22 and all of them lie inside 0 through 24.
Narrowing is the direction that removes rows, and it says so:

```{r}
data(gapminder_2007) + point + x(gdp, limits = c(0, 30000)) + y(life) +
  color(continent) + x_label("GDP per capita") +
  title("Under $30,000, and the countries above it are reported")
```

It also prints a line on your console, naming how many rows fell outside and
what the domain was. This page cannot show you that line, because only refusals
appear in a rendered book. That is deliberate: removing them is what you asked
for, so the plot draws, but gog will not remove data without saying so. Ask for
a domain no row can satisfy and it refuses instead, because the alternative is
an empty panel with axes drawn over nothing.

Keeping every country on the page and marking one group of them is a
different operation, and it has a different word. `limits` runs before the
transforms and removes rows. `brush` runs after the picture is composed and
removes nothing. [Selection](selection.qmd) sets the two side by side.

You can state one end alone. An empty end leaves that end to the data, which is
how you fix a baseline and let the top follow the data:

```{r}
data(gapminder_2007) + point * mean + x(continent) + y(life, limits = c(0, NA)) +
  title("y from zero, top still fitted to the data")
```

*"Given gapminder 2007: points derived by mean, x is continent, y is life from
0."*

It is not only for axes. A domain is a property of the scale, so it reaches
every channel that measures. That is how you fix one color ramp across plots
built from different data:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(population, scale = "log", limits = c(1e5, 2e9)) +
  x_label("GDP per capita") + title("A ramp fixed from 100K to 2B")
```

On a categorical axis `limits` is refused, and the refusal points you at the
two fixes that do the job you probably meant. A category has no range to lie
inside. Choosing *which* categories appear is filtering your table, and
choosing what order they appear in is `order()`.

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent, limits = c(0, 3)) + y(life)
```

## The column's type picks the axis

Other tools make you name a scale whenever you want to change one, so `scale =
"category"` is a habit a reader may arrive with.

The scales in the table above are not all chosen the same way. You ask for
`log`. You do not ask for `time` or `category`. gog reads those off the column:
a `Date` column gets a calendar axis, and a column of text gets one slot per
distinct value.

You may still write them. Writing `scale = "category"` on a categorical column is
allowed and changes nothing, in the same way `scale = "linear"` on a number
changes nothing. The plot below is identical to the same sentence with the
scale left off:

```{r}
data(gapminder_2007) + bar * mean + x(continent, scale = "category") + y(life) +
  title("Writing it changes nothing")
```

*"Given gapminder 2007: bars derived by mean, x is continent as a category, y is
life."*

What you cannot do is use a scale to contradict the column. Ask for a
categorical axis on a number and gog refuses:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp, scale = "category") + y(life)
```

The reason behind that refusal also explains the others in this chapter. A scale
says *how* a measured column is placed. Whether the axis measures at all is the
column's type. So `log` on a categorical column is refused, `time` on a plain number is
refused, and a date column cannot be turned back into raw numbers by asking for
`linear`. Every one of those refusals tells you to change the column, because
the column is where the answer lives.

There are two things you might mean by "draw this number as categories", and gog
has a word for each. For one slot per distinct value, declare the column's
categories: `factor()` in R and `ordered()` in the other three bindings, both
shown in [Data](data.qmd). For ranges, use `bin`, which cuts the numbers into
intervals and is covered in [Transforms](transforms.qmd).

## `tick_count`: how many ticks the axis aims for

An axis picks five ticks or so by default, and the count is the one part of the
fit you may want to choose. A crowded axis is easier to read with fewer; an axis
someone will take numbers from is easier with more. `tick_count` is written on
the binding, beside `scale`, `base` and `limits`, and for the same reason. How
many ticks an axis gets is a property of the scale, so it belongs to the
sentence that describes the scale rather than to `theme()`, which controls the
page.

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
   x_label("GDP per capita") + title("A low count: the shape, not the numbers")) |
  (data(gapminder_2007) + point + x(gdp, tick_count = 12) + y(life) +
     x_label("GDP per capita") + title("A high count: for reading values off"))
```

*"Given gapminder 2007: points, x is gdp with 3 ticks, y is life, beside the
same with 12 ticks."*

The count is a target rather than a promise, and the difference is worth
understanding. The count chooses a *step*, and the step is then rounded to a
number a person would pick, so asking for eight on an axis running 0 to 100 gets
a step of 10 and eleven ticks. What you can rely on is the direction: ask for
more and you never get fewer, and every tick still falls on a round number.

Both panels above cover the same range. That is the rule the count obeys: it
changes how densely an axis is *labeled*, never where the axis stops. Ask for
two ticks and the axis does not shrink to the two values it draws.

```{r}
data(gapminder_asia) + line + x(year, tick_count = 2) + y(life) +
  group(country) + style(opacity = 0.4) +
  title("Two ticks, and still 1952 to 2007")
```

A categorical axis has one tick per category, so the count there is the data's
rather than yours, and the refusal points at the two fixes that do the jobs you
might have meant:

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent, tick_count = 3) + y(life)
```

And a legend is not a short axis. `limits` reaches every channel that measures,
because every one of them has a domain; `tick_count` reaches only the three
position channels that draw an axis: `x`, `y` and `z`. A color legend names
three rows, both ends and the middle, and those come from the scale's own shape
rather than from a count you could choose. On a log ramp the middle row is the
geometric mean, the label for the color actually painted halfway along the ramp.
So `color()`, `size()` and `opacity()` have no tick count to set, the same way
`shape()` has no `scale` and no `limits`: there is nothing for the count to
describe.

### Crowding in the plane and in the cube

A flat panel and a cube count their ticks differently, and neither answer is
wrong. A flat axis draws *every* tick it chose, however tight they get. The
panel's width is already fixed, and the labels sit in a margin reserved for
them. An edge label that would overhang its panel is anchored inward instead.
Nothing is dropped, so what you asked for is what you count.

A cube cannot make that promise. An axis in [space](space.qmd) is drawn at
whatever length the viewing angle leaves it. Tilted far enough, it has no room
for its numbers, so the cube draws as many as fit and leaves the rest out. Both
are the same rule underneath, *do not print one number through another*. They
differ only in how much room each has, because the plane can reserve room in
advance and the cube cannot. When a count you stated has to be thinned that way,
the engine says so rather than quietly drawing fewer.

## In short

A scale moves where a number lands, and never what it is. `limits` states the
domain: narrowing it removes rows and prints how many, where a brush removes
none. In a flat panel `tick_count` is a target rather than a promise, and a cube
drops the ticks that will not fit.

## What comes next

A scale decides where a number lands on an axis, and says nothing about color,
size or shape. [Channels](channels.qmd) covers the rest of them: which visual
property carries a column, and what guide it earns.
