# Selection

```{r}
#| include: false
source("R/setup.R")
```

Which of these points are the rich countries, and where do they sit? A scatter
with 142 dots answers slowly. Add one word and the reader can draw a box around
any group and see the rest step back:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, and let the reader select."*

On a page that can run the engine, drag across that plot. On paper it is the
scatter itself, because nothing is selected until someone selects it.

Bare `brush` means *the positions this plot binds*, so here it is a region over
`gdp` and `life` together. It says the plot is explorable without deciding in
advance what is worth exploring.

Every country is still drawn while you drag. The ones inside your box keep their
color and the rest are pushed back far enough to read as a background, so the
group is read *against* what it was taken from. Remove the other rows and the
plot no longer shows you that the rich countries are the top right corner of
something.

## Saying where the selection opens

Name a column and the brush is a bound on that column alone, which is how you
point at something rather than let a reader look for it:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(gdp, at = c(20000, 60000))
```

The thirty or so countries above 20,000 dollars are lit, and you can see at once
that they sit almost entirely above 75 years. That is closer to an annotation
than to a tool: the sentence has done the selecting, and a reader can still drag
to move the bound.

The two forms answer different questions, and the difference is who chooses.
`brush` hands the choice to the reader. `brush(gdp, at = ...)` makes it in the
sentence, and prints.

## It highlights. It never removes rows.

This is the one rule to learn, and it separates `brush` from an atom you already
know. [`limits`](scales.qmd) also takes a range in a column's own units. It does
something different with it.

`limits` runs **before** the statistics. It removes rows, and the transform then
sees the survivors as if they were the whole data. `brush` runs **after** the
picture is composed. It removes nothing and recomputes nothing.

One range, one mark, two outcomes:

```{r}
#| layout-ncol: 2
data(gapminder_2007) + point + x(gdp, limits = c(20000, 60000)) + y(life) +
  theme(width = 380, height = 300) + title("limits: 114 countries gone")

data(gapminder_2007) + point + x(gdp) + y(life) +
  theme(width = 380, height = 300) + brush(gdp, at = c(20000, 60000)) +
  title("brush: all 142 still here")
```

The left plot is a different plot. Its axis runs from 20,000 to 60,000, only the
countries inside that range are drawn, and both axes were fitted to the ones
that survived. Look at the y axis: it starts near 70, because no short-lived
country is left to stretch it downward. GOG counts the countries it removed
aloud rather than dropping them quietly.

The right plot is the whole world, with a group of it standing out. The axes are
the axes of all 142 countries, so the selection sits where it always sat.

Which of the two you want is a real question and the grammar keeps both sayable.
Ask *how does this subgroup look on its own* and you want `limits`. Ask *where
does this subgroup sit inside the whole* and you want `brush`.

## Two bounds make a rectangle

One `brush` is one column. Write two, and a row has to satisfy both:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(gdp, at = c(20000, 60000)) + brush(life, at = c(78, 85))
```

The rich countries that also live longest. Two of the countries that survived
the first bound do not survive this one, and you can see exactly which two,
because they are still on the page.

Writing it as two atoms rather than one with two arguments follows the rest of
the grammar. A position is one column at a time, and so is a bound.

## A column of categories selects slots

`at` is where the selection opens. What you give it says which kind of column
you meant, in the same way a column decides whether `color` hands out a ramp or
a palette. Numbers are a range. Names are a set of slots:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) + color(continent) +
  theme(width = 620) + brush(continent, at = c("Europe", "Oceania"))
```

One word, two readings, decided by the column rather than by a second atom. Drag
across the slots and the selection follows them, whole slots at a time, because
a category has no half.

The column is on the `x` axis here and that is not decoration. **A drag can only
reach a column the panel places.** Written over the scatter this chapter opened
with, where `continent` is a color and nothing else, `brush(continent, at = ...)`
is still a legal sentence and still dims the right countries — but there is
nowhere to drag, because no axis is measuring continents. The sentence has said
everything, and the mouse has nothing left to say.

## Two views, one selection

This is the use a brush was invented for, and everything above it has been the
setup. Selecting inside one plot tells you little you could not already see. The
question worth asking is *where does this group sit in a different picture*, and
that needs a second picture:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) +
   brush(gdp, at = c(20000, 60000)) + theme(height = 230)) /
  (data(gapminder_2007) + point + x(gdp) + y(population, scale = "log") +
     brush(gdp, at = c(20000, 60000)) + theme(height = 200))
```

Two plots, one page, and one bound lighting the same countries in both. Drag in
either and the other follows. The rich countries live long, which the top plot
says; they are also mostly small, which only the bottom one says, and neither
plot could have told you both.

**Nothing was added to the grammar to link them.** Look at what the two sentences
share: the column `gdp`, named on `x` in each. A bound is a fact about a column's
values rather than about a panel, so two plots that read that column are already
answering the same question. There is no name to declare, no identifier to match
up, and no third atom joining them.

That falls out of [composition](composition.qmd)'s one rule, arriving somewhere
new. Two composed plots naming the same column on the same axis share that axis,
which is why these two are the same width and the `gdp` numbers are written once.
The selection travels the same road for the same reason.

## Which one is that?

The other thing people want from a selection is a name. Layer `text` over the
points and the brush decides which labels are legible:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  text + label(country) + style(size = 7) +
  brush(gdp, at = c(30000, 60000)) + theme(width = 620)
```

A hundred and forty-two labels is unreadable, and that is the point: the ones
you did not select step back with their points, and about twenty names are left
standing. Drag the bound and a different twenty are readable.

This works because `text` is one of the marks a brush reaches, on the same test
as `point`: one row is one label. Nothing here is about selection *and* labels
being designed together, because they were not.

## What can be brushed

A selection picks out rows, so a mark can be brushed when **one row is one
shape**: `point`, `text`, `rule` and `zone`.

A `line` is the other case. It draws one shape *through* many rows, so a row is
a corner of the line rather than a thing of its own. Selecting half of a line
would mean cutting it in two, and cutting a line in two already has a word:

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  brush(life, at = c(60, 80))
```

The refusal names both ways forward. Split the line with `group()`, or brush a
mark that draws one shape per row.

A summarized layer refuses for a different reason. A bar of `bar * count` stands
for forty countries at once, so "twelve of these are selected" has no honest
picture: dimming the whole bar would be wrong, and dimming part of it would
invent a mark nobody asked for. GOG says so instead of guessing:

```{r}
#| error: true
data(gapminder_2007) + bar * count + x(continent) +
  brush(gdp, at = c(20000, 60000))
```

Beside a layer that *can* answer, neither of these is a refusal. The plot draws,
and GOG reports which layer it left whole:

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  line * smooth +
  brush(gdp, at = c(20000, 60000))
```

The points answer the selection and the trend line does not, so the line is
drawn once, at full strength, through all 142 countries. That is the right
answer: a trend fitted to the selection alone would be a different statistic,
and the sentence did not ask for one.

A range that does not run upward selects nothing, so it is a mistake rather than
an empty selection:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  brush(gdp, at = c(60000, 20000))
```

## Selecting while the plot plays

A frame is a subset of the rows, and a bound is a test on a column's values, so
the two compose and neither needed anything added for the other. The bound is
tested again inside every frame:

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + color(continent) +
  theme(width = 620) + play(year) + brush(gdp, at = c(3000, 20000))
```

Watch countries cross into the selection and light up. The bound holds still at
3,000 to 20,000 dollars and the world moves through it: in 1952 the lit group is
the middle-income countries of the day, and by 2007 many of them have left it
through the top while others have arrived from below. Nothing pauses, and no
frame is skipped. The same test simply gets a different answer in each one, which
is what makes a brush a threshold you watch the world cross.

Choosing a bound the data actually crosses is worth a moment, and it is the one
piece of judgment this atom asks of you. The two ends of a sequence are where an
animation misleads. Count what a bound catches in each frame before you write it
down: 3,000 to 20,000 dollars lights between 52 and 74 countries in every frame
of this sequence, while 20,000 and up lights none at all in 1952 and 1957, one in
1962, and four by 1972. The second is a true picture and it reads as a broken
plot, because a reader cannot tell *nothing is selected here* from *this is not
working*. Drag the bound upward and you will watch it happen.

One column cannot do both jobs. Brushing the column the frames are cut on would
select frames rather than rows inside one, and that is a scrub bar rather than a
selection:

```{r}
#| error: true
data(gm_all) + point + x(gdp) + y(life) + play(year) + brush(year, at = c(1980, 2007))
```

A scrub bar belongs to the page a plot is shown on, not to the sentence that
describes the plot. [Play](play.qmd) says why.

## All three at once

The [third position](space.qmd), the frames and the selection are three separate
ideas, and none of them was taught about the others. So they compose:

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + z(population) +
  color(continent) + play(year) + brush(gdp, at = c(3000, 20000))
```

A cube of 142 countries, twelve moments, and one bound tested inside every one
of them. Drag the plot to turn it and the selection turns with the cloud, because
the bound picks rows and the angle picks a viewpoint, and neither asks the other
anything.

There is one difference here, and it is worth knowing rather than working out.
In the cube the drag already belongs to the camera, so the sentence has to name
the bound: `brush` on its own would declare a region with no way to draw one.
That is the same rule as everywhere else in this chapter, arriving from the other
side. A drag means one thing at a time, and in the cube it means turn.

## Moving the bound yourself

Every plot in this chapter after the first names where its selection *opens*. On a page that can
run the engine, you can drag the panel to move the bound, and the plot redraws
under your hand.

This changes nothing about the sentence, and it is the same arrangement the cube
uses. `space(turn =, tilt =)` chooses the angle a plot opens at and the mouse
takes you anywhere else; `brush(gdp, at = )` chooses the bound a plot opens with
and the mouse does the same. Drag to find a selection, then write it into the
sentence.

On paper, and in the PDF edition, each plot shows the bound its sentence asked
for. That is why naming it is worth doing even though a reader can move it. The
printed figure shows the selection you chose, and a reader who can drag starts
from it.

Written with no `at` at all, nothing is selected. The plot then draws exactly as
it would with no `brush` in the sentence, which is the resting state a page shows
before the first drag.

## What the page adds, and what it does not

A brushed plot on a web page carries a line of controls under it. None of them
are in the sentence, and that is the rule rather than an omission. They report a
selection or change how you look at one. Neither changes what the plot claims
about the data, so neither earns a word in the grammar. It is the same reason a
plot in the cube has a line of controls and `space()` has no word for the mouse.

Reading left to right, the line looks like this:

| Control | What it does |
|---|---|
| `drag: select` / `drag: pan` | Says what dragging the plot will do, and switches it |
| `17 of 142 selected` | How many rows the current bound catches |
| `show rows` | Opens a table of those rows, and closes it again |
| `clear` | Puts the selection back to what the sentence asked for |
| `−` `+` | Makes the picture smaller or larger |
| `fit` | Returns to the whole picture |

### Selecting

Drag across the panel. A dotted box follows the pointer, and it shows **exactly
what is bound**. On a plot with one brush the box spans the whole panel in the
other direction, because that direction was not selected. On a plot with two, it
is a rectangle. A box that narrowed an axis nobody bound would be claiming a
range you did not choose.

Dragging in either direction gives the same bound, so you can start from any
corner. A **single click clears** the selection, which is what a click on empty
space does everywhere else.

### Reading what you caught

The count is always there. `show rows` opens a table underneath, listing the
columns the sentence maps rather than every column in the table, because twelve
columns of a spreadsheet are not what you are looking at. A long selection says
how many rows the table left out instead of quietly showing fewer.

That table answers the question people usually reach for a click to ask. Draw a
bound tight around one point and it names that point.

For a quick look without selecting anything, **move the pointer over a mark** and
that row's values appear beside it. Nothing is selected and nothing changes.

### Looking closer

`+` and `−` magnify the picture and `fit` returns to the whole of it. Zooming in
switches the drag to panning, because a reader who has just looked closer usually
wants to move around inside it; `fit` hands the drag back to selecting. You can
switch by hand at any time with the button on the left, and the pointer tells you
which mode you are in: a crosshair to draw a box, an open hand to move the
picture.

This is a *view*, so it **recomputes nothing**. The bars are the same bars and the
ticks are the same ticks, drawn larger. Zoom far enough into a histogram and you
will see the same bins, not finer ones. If you want the statistics re-run over a
narrower range, that is [`limits`](scales.qmd), and it is the other operation this
chapter opened with.

### Two words, two things

`clear` acts on the **selection** and `fit` acts on the **view**. They are next to
each other and they undo different things, so they are named for what they act on
rather than both being called reset.

None of this exists on paper, and none of it is missing there. The PDF shows the
bound the sentence named, which is why naming one is worth doing even where a
reader can move it.
