# Selection

```{r}
#| include: false
source("R/setup.R")
```

Which countries in a scatter are the rich ones, and where do they sit? Finding
them among 142 dots takes a long time. Add one word, `brush`, to the sentence
and the reader can draw a box around any group. The countries inside the box
keep their color, and the rest are dimmed:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, and let
the reader select."*

In the web edition, drag across that plot. In the PDF it is the
plain scatter, because a bare `brush` selects nothing until a reader drags.

Bare `brush` means *the positions this plot binds*, so here it is a region over
`gdp` and `life` together. The sentence offers the reader a selection without
choosing in advance which part of the plot to select.

Every country is still drawn while you drag, so the group is read *against* what
it was taken from. Remove the other rows and the plot no longer shows that the
rich countries sit at the top right of the whole world.

## Saying where the selection starts

In the PDF nothing can be dragged, so a reader of the first plot
must still find the rich countries without help. Name a column and the brush
states a **bound** on it: the low and high values the selection starts at. That
is how you point at something rather than let a reader look for it:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(gdp, at = c(20000, 60000))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, and let
the reader select gdp, starting from 20000 to 60000."*

The thirty or so countries above 20,000 dollars keep their color, and almost all
of them sit above 75 years. A bound written into the sentence points the reader
at a group, as a note on the plot would. The sentence has done the selecting,
and a reader can still drag across the panel to replace the bound.

The two forms differ in who chooses the group. `brush` leaves the choice to the
reader. `brush(gdp, at = ...)` makes the choice in the sentence, so the
selection also appears in the PDF.

## Highlighting, never removing rows

A brush highlights rows and never removes them. That rule separates `brush`
from an atom you already know. [`limits`](scales.qmd) also takes a range in a
column's own units, and it puts that range to a different use.

`limits` on a column of the table runs *before* the statistics. It removes rows,
and the transform then reads the rows that are left as if they were the whole
table. `brush` runs *after* the axes and statistics are computed. It removes
nothing and recomputes nothing.

One range, one mark, two outcomes:

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp, limits = c(20000, 60000)) + y(life) +
  theme(width = 620, height = 300) + title("limits: 109 countries gone")
```

*"Given gapminder 2007: points, x is gdp from 20000 to 60000, y is life."*

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  theme(width = 620, height = 300) + brush(gdp, at = c(20000, 60000)) +
  title("brush: all 142 still here")
```

The `limits` plot is drawn from fewer rows. Its axis runs from 20,000 to 60,000,
only the countries inside that range are drawn, and both axes were fitted to the
rows that were left. Look at its y axis: the lowest number on it is 74, because
the poorer countries were removed, and they included every country with a low
life expectancy. gog prints how many it removed rather than removing them in
silence.

The `brush` plot is the whole world, with one group keeping its color. Its axes
are the axes of all 142 countries, so the selected points are exactly where an
unbrushed plot would have put them.

The two plots do not share an axis, even though one is drawn above the other.
A country at 30,000 dollars sits a quarter of the way along the top plot's axis
and more than halfway along the bottom one's. Read each plot against its own
axes.

Which of the two you want depends on your question, and the grammar lets you
write either one. Ask *how does this subgroup look on its own* and you want
`limits`. Ask *where does this subgroup sit inside the whole* and you want
`brush`.

## Two bounds make a rectangle

Which of the rich countries also live longest? That is two conditions on one
row. One `brush` tests one column. Write two, and a row has to satisfy both:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(gdp, at = c(20000, 60000)) + brush(life, at = c(78, 85))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, and let
the reader select gdp, starting from 20000 to 60000, and life, starting from 78
to 85."*

These are the rich countries that also live longest. Six countries inside the
`gdp` bound fall outside the `life` bound, and you can see which six, because
they are still drawn, dimmed.

Two bounds needed `brush` written twice, and that follows the rest of the
grammar. A
position is one column at a time, and so is a bound.

## A categorical column selects slots

Both bounds so far have been ranges of numbers. What happens when the column
holds names instead? `at` is still where the selection starts, and the column
decides which reading you get, in the same way it decides whether `color` uses a
ramp or a palette. Numbers are a range. Names are a set of slots:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) + color(continent) +
  theme(width = 620) + brush(continent, at = c("Europe", "Oceania"))
```

*"Given gapminder 2007: points, x is continent, y is life, color by continent,
and let the reader select continent, starting with Europe and Oceania."*

Drag across the slots and the selection follows them, whole slots at a time,
because a slot cannot be split.

The column is on the `x` axis here, and that matters. A drag can only change a
bound on a column that one of the panel's axes shows. Add the same
`brush(continent, at = ...)` to the first scatter in this chapter, where
`continent` is only a color. The sentence is still legal and still dims the
right countries. But no axis shows continents, so a drag cannot change the
selection the sentence chose.

It does not matter *which* axis places the column. Here is the same sentence
with `x` and `y` exchanged, in a shorter panel:

```{r}
data(gapminder_2007) + point + x(life) + y(continent) + color(continent) +
  theme(width = 620, height = 320) + brush(continent, at = c("Europe", "Oceania"))
```

The slots now run up the left side, and the same two continents keep their
color. Drag on the panel and the selected band is horizontal instead of
vertical, still one whole slot at a time. The bound never mentions an axis. It
names a column, so moving the column to the other axis turns the drag from
vertical to horizontal, and the `brush` stays exactly as it was written.

## Two views, one selection

A brush is most useful across two plots. A selection inside one plot marks a
group you can already see there. The useful
question is *where does this group sit in a different picture*, and that needs
a second picture:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) +
   brush(gdp, at = c(20000, 60000)) + theme(height = 230)) /
  (data(gapminder_2007) + point + x(gdp) + y(population, scale = "log") +
     brush(gdp, at = c(20000, 60000)) + theme(height = 200))
```

*"Given gapminder 2007: points, x is gdp, y is life, and let the reader select
gdp, starting from 20000 to 60000, above points, x is gdp, y is population on a
log scale, and let the reader select gdp, starting from 20000 to 60000."*

Two plots sit on one page. In both plots, the same countries keep their color.
Drag in either and the other follows. The top plot says the rich countries live
long. The bottom plot says their populations run from under a million to over
three hundred million. Neither plot could have told you both.

Nothing was added to the grammar to link them. Look at what the two sentences
share: each one says `brush(gdp, at = c(20000, 60000))`. A bound is a fact about
a column's values rather than about a panel, so two plots that brush the same
column answer the same question. There is no name to declare, no identifier to
match, and no third atom joining them.

[Composition](composition.qmd) has a matching rule for axes. Two composed plots
naming the same column on the same axis share that axis, which is why these two
are the same width and the `gdp` numbers are written once. The selection is
shared for a related reason: both sentences brush the same column.

## One bound reaches every panel

[Faceting](faceting.qmd) is the other way to get more than one panel, and a
brush treats it the same way. The bound is a fact about a column, and every
panel reads that column, so every panel answers:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  brush(gdp, at = c(20000, 60000)) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, and let the reader select
gdp, starting from 20000 to 60000, split into panel columns by continent."*

Five panels, one bound, and the rich countries keeping their color. Europe has
many and Africa has none, which is the comparison the split was made for.
`facet` divides the rows into panels, `brush` dims the rows outside the bound,
and neither atom refers to the other.

In the web edition, resting the pointer on a mark shows that row's values. A
bound is shared across panels, because it names a column. The pointer is not:
its values come from the panel you are pointing at, never from a row another
panel drew.

## Which one is that?

A selection shows you where a group sits. It does not say which country each dot
is, and that is usually the next question. Layer `text` over the points, and the
bound decides which labels stay legible:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  text + label(country) + style(size = 7) +
  brush(gdp, at = c(30000, 60000)) + theme(width = 620)
```

*"Given gapminder 2007: points, x is gdp, y is life, and also text, label by
country, and let the reader select gdp, starting from 30000 to 60000."*

A hundred and forty-two labels on one scatter cannot be read, and a selection
solves that here. The labels you did not select are dimmed with their points,
and twenty names keep their color. Drag across the panel for a new bound, and a
different group of names becomes readable.

This works because `text`, like `point`, draws one shape for each row, so a
brush can dim each label on its own. No special rule joins `brush` to `text`:
each atom follows its own rule, and this plot uses both.

## What can be brushed

A selection marks rows, so a mark can be brushed only when *one row is one
shape*. Four marks meet that test and draw a selection: `point`, `text`, `rule`
and `zone`. An `interval` with the
[`bounds`](transforms.qmd#bounds-a-pre-computed-lowhigh-pair) transform meets
the test too, because each whisker's two ends come from one row.

A plain `bar` meets the test as well and still refuses. One row is one bar, but
a bar's thickness comes from the smallest gap between neighboring bars. The
engine draws the selected rows and the other rows as two groups, and each group
would measure its own gap, so the two groups would get different thicknesses. A
jittered `point` and a repelled `text` refuse for a similar reason. A jittered
point's offset is derived from its place in the table. A repelled label is
placed by what the other labels around it are doing. In both cases every mark
would move when the selection changed. An `edge` meets the test too, and it
still refuses. The [Network](network.qmd) chapter gives the reason: a network
has no axes, so there is nothing for a range of values to be measured on.

A `line` does not meet the test at all. It draws one shape *through* many rows,
so a row is a corner of the line rather than a thing of its own. Selecting half
of a line would mean cutting it in two. `group()` cuts a line into pieces, but
each piece still runs through many rows, so the engine refuses even this
sentence:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  brush(life, at = c(60, 80))
```

The refusal lists the layers that draw a selection. A brushed plot needs at
least one of them.

A summarized layer refuses for a different reason. A bar of `bar * count` stands
for as many as fifty-two countries at once. There is no way to draw "twelve of
these are selected": dimming the whole bar would be wrong, and dimming part of
it would invent a mark nobody asked for. gog says so instead of guessing:

```{r}
#| error: true
data(gapminder_2007) + bar * count + x(continent) +
  brush(gdp, at = c(20000, 60000))
```

Apart from the `edge`'s, none of these refusals applies when the plot also
carries a layer that *can* answer. The plot draws, and gog reports which layer
it left whole:

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  line * smooth +
  brush(gdp, at = c(20000, 60000))
```

*"Given gapminder 2007: points, x is gdp, y is life, and also a line derived by
smooth, and let the reader select gdp, starting from 20000 to 60000."*

The points answer the selection and the trend line does not, so the line is
drawn once, undimmed, through all 142 countries. That is the right answer: a
trend fitted to the selection alone would be a different statistic, and the
sentence did not ask for one.

A range written from high to low selects nothing, so it is a mistake rather than
an empty selection:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  brush(gdp, at = c(60000, 20000))
```

A map cannot be brushed at all. It projects longitude and latitude to new
positions before it draws anything, and a brush on it would count one set of
places while it dims another. So gog refuses any sentence with both `brush` and
`map()`. `quakes_fiji` records earthquakes near Fiji, with longitude in `east`
and latitude in `north`:

```{r}
#| echo: false
peek(quakes_fiji)
```

The sentence below adds `brush` to a map of them:

```{r}
#| error: true
data(quakes_fiji) + point + x(east) + y(north) + map() +
  brush(east, at = c(178, 182))
```

To select places by longitude and latitude, brush the same columns without
`map()`, as the message says. Without a projection, the axes are in degrees, so
the band from 178 to 182 has straight sides:

```{r}
data(quakes_fiji) + point + x(east) + y(north) +
  brush(east, at = c(178, 182))
```

The earthquakes inside the band keep their color.

Apart from the backward range, the `edge` and the map, every refusal in this
section comes from two questions. Can the engine say which rows your selection holds?
And can it dim them while every mark keeps its place and its size? A played
plot raises the first question again, because it draws only some of its rows at
a time. The next section answers it.

## Selecting while the plot plays

A sequence draws one year at a time, and you still want one group marked in
every year. A frame is a subset of the rows, and a bound is a test on a column's
values, so the two combine without either one changing. The bound is tested
again inside every frame:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + color(continent) +
  theme(width = 620) + play(year) + brush(gdp, at = c(3000, 20000))
```

*"Given all the gapminder years: points, x is gdp on a log scale, y is life,
color by continent, played by year, and let the reader select gdp, starting from
3000 to 20000."*

In the web edition, watch countries cross into the selection, frame by frame.
In the PDF, the plot shows the first frame, 1952. The bound stays
at 3,000 to 20,000 dollars while the countries move. In 1952 the selected group
is that year's middle-income countries. By 2007 many have grown past 20,000 and
poorer ones have grown into the range. A brush does not pause the sequence or
skip a frame.

### What the count includes

Stop the clock, draw a small box, and the count under the plot (the line reading
`17 of 1704 selected`) looks far too large. The number is right. A bound is a
test on a column's values, so it is true or false for every row in the table. It
does not stop at the frame you can see. The count and `show rows`, a button
under the plot, report the whole selection across every frame. The picture shows
you one frame of it.

The played plot in this section makes that concrete. `gm_all` is 1,704 rows,
which is 142 countries in each of twelve years, and one frame draws 142 of them.
Stop on 1982 and drag across the four countries furthest to the left, and the
count reads far more than four. This brush tests only `gdp`, so the box covers
the whole height of the panel, and it tests every year's rows, not only 1982's.
In other years, other countries had a `gdp` in that range too. The four you can
see are the 1982 rows of a selection that spans all twelve years.

The table that `show rows` opens is where you check it. It lists the `year` of
every selected row, beside the other columns the sentence maps. You can see
which frame each row belongs to, and pick out the four that belong to the frame
on screen.

A faceted plot never raises this question. There every selected row is drawn at
once, so a count of the whole selection is a count of what you see. A sequence
shows one frame of twelve at a time, so most of the selected rows are not on the
screen. They are in other frames. The rule is the same one; only the picture
differs.

### Choosing a bound the data crosses

A bound chosen for the late frames can mislead a reader in the early ones. Count
what a bound selects in each frame before you write it down. From 3,000 to
20,000 dollars selects between 52 and 74 countries in every frame here. From
20,000 to 60,000 dollars selects none at all in 1952 and 1957, one in 1962, and
four by 1972. The second bound draws a true picture, and it still reads as a
broken plot, because a reader cannot tell *nothing is selected here* from *this
is not working*. To see it, drag across the panel from 20,000 to 60,000 dollars
and step back to 1952: every country in that frame is dimmed.

One column cannot both cut the frames and carry a bound. Brushing the column the
frames are cut on would select frames rather than rows inside one, and that is a
control for moving through time rather than a selection:

```{r}
#| error: true
data(gm_all) + point + x(gdp) + y(life) + play(year) + brush(year, at = c(1980, 2007))
```

Moving the clock belongs to the page a plot is shown on, not to the sentence
that describes the plot. That is what the buttons under a played plot do, and
[Play](play.qmd#what-the-page-adds) says why they have no word here.

## All three at once

A bound and a sequence combined without a new rule, and a
[third position](space.qmd) joins them the same way. [Play](play.qmd) drew the
cube of countries through the years, and nothing in it marks the middle-income
countries. The third position, the frames and the selection are three separate
ideas, and none of them was designed around the others. So all three can be
written in one sentence, and none of them changes what the others do:

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + z(population) +
  color(continent) + play(year) + brush(gdp, at = c(3000, 20000))
```

*"Given all the gapminder years: points, x is gdp on a log scale, y is life, z
is population, color by continent, played by year, and let the reader select
gdp, starting from 3000 to 20000."*

A cube of 142 countries, twelve frames, and one bound tested inside every one of
them. Drag the plot to turn it and the selection turns with the cloud. The
bound picks rows and the drag picks the viewing angle, and neither one depends
on the other.

One thing is different in the cube. There the drag turns the plot, so it cannot
also draw a selection, and you name the bound in the sentence. `brush` on its
own would declare a region that no drag could draw. The rule is the one behind
every plot in this chapter: a drag does one thing at a time, and in the cube
that one thing is turning.

So in the cube you cannot change the selection, but you can still read it. The
count and `show rows` sit on their own line under the plot, and you can read the
selected rows while you turn it.

## Changing the bound yourself

After the first one, every brushed plot in this chapter states where its
selection starts. In the web edition, drag across a flat panel and your range
replaces the sentence's bound. The plot redraws as you drag.

This changes nothing about the sentence, and it is the same arrangement the cube
uses. `space(turn =, tilt =)` chooses the angle a plot starts at, and the
pointer moves it from there. `brush(gdp, at = )` chooses the bound a plot starts
with, and the pointer replaces it from there. Drag to find a selection, read its
two ends off the axis, and write them into the sentence.

In the PDF, each plot shows the bound its sentence asked for. That
is why naming it matters even though a reader of the web edition can replace
it: that reader starts from it too.

A `brush` that names a column but no `at` also selects nothing until a reader
drags.

## What the page adds

A flat brushed plot on a web page carries two lines of controls under it. None
of them are in the sentence, and that is the rule rather than an omission. They
report a selection, or they change how you look at one. Not one of them changes
what the plot claims about the data, so none of them needs a word in the
grammar. A plot in the cube follows the same rule: `space()` names the angle
the plot starts at, and turning it needs no word. It carries two lines of
controls, or three when it is brushed.

The first line is [the one every plot
carries](first-plot.qmd#the-boxes-under-every-plot):

| Control | What it does |
|---|---|
| the magnifiers, `−` and `+` | Makes the picture smaller or larger |
| the frame button (four corners around a small square) | Returns to the whole picture |
| the camera | Saves what you are looking at as a picture file |

It sits first, directly under the picture, on a plain plot and a brushed one
alike. This matters. A reader who has found the magnifiers once finds them in
the same place under every plot in this book, rather than in a place that moves
with whatever else the plot needed.

A plot that also plays carries three more buttons at the end of that same line,
which pause the sequence and step through its frames.
[Play](play.qmd#what-the-page-adds) sets them out.

The second line is what a brush adds:

| Control | What it does |
|---|---|
| `drag:` and its icons | Says what dragging the plot will do, and switches it: a rectangle, a free shape, or moving the picture |
| `40 of 142 selected` | How many rows the current bound selects. On a played plot this counts every frame, not the one on screen. Plots composed with `|` or `/` share this count. When they read the same table, each row is counted once, and only if all their bounds select it |
| `show rows` | Opens a table of those rows; while the table is open the button reads `hide rows`. A long selection arrives ten at a time, with `‹` and `›` under the table |
| `clear` | Puts the selection back to what the sentence asked for |
| `unstamp` | Removes every stamp at once ([Stamping a point](#stamping-a-point) explains stamps). With two or more, the button shows how many. Absent until you have made one |

In the cube, this line holds only the count and `show rows`. It comes third,
under the line that shows the angle.

A control drawn as an icon names itself when the pointer rests on it. So do
`clear` and `unstamp`, which say what they undo.

### Dragging a box

A bound written in the sentence is a starting point, and a drag replaces it.
Drag across the panel. A dotted box follows the pointer, and it shows *exactly*
what is bound. On a plot with one brush the box spans the whole panel in the
other direction, because that direction was not selected. On a plot with two, it
is a rectangle. A box that narrowed an axis you never bound would claim a range
you did not choose. The box disappears when you release the pointer, and from
then on the dimming shows the bound.

Dragging in either direction gives the same bound, so you can start from any
corner. A click on empty space empties the selection, so nothing is dimmed. A
click that lands on a
mark does something else, and [Stamping a point](#stamping-a-point) is about
that.

### Drawing around a group

A rectangle is what a sentence can say. Some groups are not rectangles. A cloud
of points often has a shape: a narrow tail to one side, a curve, a group sitting
apart from the rest. A box drawn around such a group includes points you did not
mean.

The second icon beside `drag:` is the **free shape**. Choose it, then draw
around the group with the pointer held down. The outline closes itself when you
release the pointer, and everything inside it is selected. The count and the row
table read exactly as they do for a box, because a shape and a bound ask the
same question: which rows do you mean.

Nothing about the sentence changes. You do not type a shape, and the PDF
still shows the bound the sentence named. Drawing one is a way of looking,
like turning a cube, rather than a claim about the data.

That is also why the grammar has no word for it. Other tools call the gesture a
*lasso*, and gog has no such word. Written down, a drawn shape is a long list of
corner points, too long to type or read as part of a sentence. So you can draw
the shape, but you cannot write it.

Two restrictions apply. A free shape needs two continuous axes, so where one
axis is categorical the drag stays a rectangle. And `clear` removes a free shape
and returns to the bound the sentence named. It cannot restore the shape,
because no sentence can state a drawn shape.

### Reading what you selected

The dimming shows where a group sits, but not which rows are in it. The count
says how many rows are selected, and `show rows` opens the row table.
It lists the columns the sentence maps, not every column of your data, because
the columns you did not plot are not what you are looking at.

A long selection arrives ten rows at a time. The line under the table says which
rows you are reading, as `11–20 of 40`, and the two arrows beside it move you
through the rest. Every row you selected is reachable, and none of them pushes
the plot off the screen. A selection short enough to fit shows no arrows at all.

The row table lists every row you selected, with the columns the sentence maps,
in one place, away from the picture. It shows country names only when the
sentence maps `country`, as the labeled plot in
[Which one is that?](#which-one-is-that) does. Draw a bound tight around one
point and the table lists that point alone.

### Stamping a point

The pointer shows one row's values at a time, and they disappear as soon as you
move away from the mark. That is right for a quick look and wrong for a
comparison. Suppose three points on a scatter interest you. Reading them one
after another means remembering the first two while you look at the third.

Click a mark and its values stay. That is a **stamp**: a dot on the mark, a
short line, and a small card saying what the pointer said. Click as many as you
want to compare. The card sits on the picture beside the mark it belongs to,
which is the difference between a stamp and the row table. The row table names
rows. A stamp labels the plot.

A card needs space on the picture. On a crowded scatter a card can cover the
points you want to read. So drag a card and it goes where you put it. The
line stretches to follow, and a long line ends in an arrowhead at the stamped
point. Move four cards out to the edges and the picture underneath stays whole.

A card keeps the distance you gave it from its point. Zoom in, pan across, or
drag a new bound, and the card moves with its point. Zoom far enough that the
point leaves the picture on screen, and the whole stamp goes out of sight with
it. It has nothing left to point at, and a card left over an
unrelated corner would claim a point that is not there. Bring the point back and
the card is where you left it.

A played plot works the same way, for the same reason. A row there is one
country in one year, so a stamp made in 1972 names something the 1987 frame does
not draw. The stamp is hidden while the other frames run, and it returns to its
point when the sequence comes back to 1972. The card names the year it holds.
Stamp the same country in three frames and each card appears in its own frame.

There are three ways to take a stamp off. The `×` in the corner of a card
removes that card. Clicking the card itself does the same, as long as you did
not move it: if the pointer moved between press and release, the click counted
as a drag instead. `unstamp` removes every card at once.

`clear` does not remove the cards, because a stamp is not a selection. A click
on empty space does not remove them either. It empties the selection, so every
point returns to full strength, and every card is still there. `clear`,
`unstamp` and the frame button each undo one thing, which [Three controls,
three things](#three-controls-three-things) sets out.

The camera saves the cards too. Arrange the cards, press the camera, and the
picture file carries the plot with your annotations on it, each card where you
put it and each line pointing at its row. The camera saves what you are looking
at, the same way it saves how far you have zoomed in and the angle a cube is
turned to.

You do not type a stamp, and the PDF carries none.

A stamp is for deciding which points are worth naming. A plot with a name on
every point is unreadable, and a plot with a name on the few that matter is the
one you wanted. Stamping lets you try a few before you choose. Then put those
rows in a small table of their own, give it to a `text` layer, and the picture
carries the names itself, in the web edition and the PDF alike.
[A note is a layer](marks/text.qmd#a-note-is-a-layer) shows how.

Arranging the cards by hand has a written form too. `text * repel` does the same
job inside the sentence. Every label is placed outside its own dot, and one that
moved far keeps a thin line back to the point it names. That is the stamp's line
again, written down rather than dragged. Stamp to decide what is worth naming,
then write the labels down and let
[`repel`](marks/text.qmd#repel-moving-labels-off-each-other) place them.

A stamp is for looking at the data. `text` is for telling a reader what it
says.

Both can go into a presentation, and they are not the same there. A saved
picture is fixed: it keeps one moment of one session and cannot be drawn again.
A sentence with
`text` in it draws again for anyone who has the data, in the PDF and on the page
and in all four languages. Stamp and save to keep this one picture. Write the
labels down to keep the plot itself.

### Where pointing cannot answer

On a brushed polar plot, the pointer shows no values. The reason is how the
page finds a row. It computes where each row was drawn, from the row's own
value and the two ends of the axis the panel records. That is exact wherever a
mark stands at its value. A polar plot puts every mark somewhere else on
purpose: it turns `x` into an angle and `y` into a distance from the center.

So the pointer says nothing rather than naming a row that is not there, and a
line under the plot gives the reason. Nothing can be stamped either. The page
computes a position rather than reading one, so it stays silent wherever it
cannot compute one.

In the cube, the pointer shows no values either, and nothing can be stamped.
There the viewing angle moves every mark, so the page cannot compute where each
row was drawn.

A layer whose marks do not stand at their rows' values does not silence the
pointer. `jitter` sets each point beside its value, `repel` moves each label
away from the others, and a summary such as `bin` or `mean` draws one shape for
many rows. As [What can be brushed](#what-can-be-brushed) showed, a layer like
that either refuses the brush or is drawn whole beside a layer that answers it.
The pointer then reads that other layer, whose marks stand at their values.

### Looking closer

A crowded corner is easier to read when it is larger. `+` and `−` magnify the
picture, and the frame button returns to the whole of it. Zooming in switches
the drag to panning, because a reader who has just looked closer usually wants
to move around inside the picture. The frame button sets the drag back to
selecting. You can switch by hand at any time with the icons beside `drag:`. The
pointer also tells you which mode you are in: a crosshair to draw a box, an open
hand to move the picture.

This is a *view*, so it *recomputes nothing*. The bars are the same bars and the
ticks are the same ticks, drawn larger. Zoom far enough into a histogram and you
will see the same bins, not finer ones. If you want the statistics re-run over a
narrower range, that is [`limits`](scales.qmd), the atom this chapter compared
with `brush`.

### Three controls, three things

Three of these controls undo something, and pressing the wrong one removes work
you meant to keep. `clear` acts on the **selection**, `unstamp` acts on the
**stamps**, and the frame button acts on the **view**. Each one undoes only its
own thing, so pressing one never costs you the other two. The frame button sits
a line above the other two, beside the magnifiers. It puts back the view they
changed.

The `×` on a card is a fourth undo, and it is not under the plot at all. It
belongs to the card rather than to the plot. A control under the picture acts on
the whole plot: all the cards, the whole selection or the whole view. A control
on a card acts on that card alone.

The PDF has none of these controls, and it shows the bound the sentence named.

### Your data on the page

Before you publish a brushed plot, know what data it carries. The browser
redraws the plot as the reader drags, so the table has to be in the page. The
page carries every row of the table, but only the columns the sentence names.

The rows are all there, though, each with its exact values in the columns the
page carries. So check those values before you publish, the way you would check
any file before sharing it.
[Plots that carry the table](utilities.qmd#plots-that-carry-the-table) lists
every kind of plot that carries its table, and what a saved SVG file carries
instead.

## In short

A selection runs after the axes and statistics are computed: it dims rows,
removes none, and leaves the axes where they were. The column decides the
reading, a range for numbers and a set for categories. A bound tests a column's
values rather than a panel, so plots that brush the same column share the
selection. Everything a brush needs is already inside the web page: every row
of the table, but only the columns the sentence names.

## What comes next

Selection is the last atom this book teaches. The chapters so far started from
the grammar; [the cookbook](cookbook/one-variable.qmd) starts from your data
instead: one variable first, then two, then more.
