# Selection

```{r}
#| include: false
source("R/setup.R")
```

Which of these points are the rich countries, and where do they sit? A scatter
with 142 dots answers slowly. Add one word and the reader can draw a box around
any group and see the rest step back:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, and let the reader select."*

On a page that can run the engine, drag across that plot. On paper it is the
scatter itself, because nothing is selected until someone selects it.

Bare `brush` means *the positions this plot binds*, so here it is a region over
`gdp` and `life` together. It says the plot is explorable without deciding in
advance what is worth exploring.

Every country is still drawn while you drag. The ones inside your box keep their
color and the rest are pushed back far enough to read as a background, so the
group is read *against* what it was taken from. Remove the other rows and the
plot no longer shows you that the rich countries are the top right corner of
something.

## Saying where the selection opens

Name a column and the brush is a bound on that column alone, which is how you
point at something rather than let a reader look for it:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(gdp, at = c(20000, 60000))
```

The thirty or so countries above 20,000 dollars are lit, and you can see at once
that they sit almost entirely above 75 years. That is closer to an annotation
than to a tool: the sentence has done the selecting, and a reader can still drag
to move the bound.

The two forms answer different questions, and the difference is who chooses.
`brush` hands the choice to the reader. `brush(gdp, at = ...)` makes it in the
sentence, and prints.

## It highlights. It never removes rows.

This is the one rule to learn, and it separates `brush` from an atom you already
know. [`limits`](scales.qmd) also takes a range in a column's own units. It does
something different with it.

`limits` runs **before** the statistics. It removes rows, and the transform then
sees the survivors as if they were the whole data. `brush` runs **after** the
picture is composed. It removes nothing and recomputes nothing.

One range, one mark, two outcomes:

```{r}
#| layout-ncol: 2
data(gapminder_2007) + point + x(gdp, limits = c(20000, 60000)) + y(life) +
  title("limits: 114 countries gone")

data(gapminder_2007) + point + x(gdp) + y(life) +
  brush(gdp, at = c(20000, 60000)) + title("brush: all 142 still here")
```

The left plot is a different plot. Its axis runs from 20,000 to 60,000, only the
countries inside that range are drawn, and both axes were fitted to the ones
that survived. Look at the y axis: it starts near 70, because no short-lived
country is left to stretch it downward. GOG counts the countries it removed
aloud rather than dropping them quietly.

The right plot is the whole world, with a group of it standing out. The axes are
the axes of all 142 countries, so the selection sits where it always sat.

Which of the two you want is a real question and the grammar keeps both sayable.
Ask *how does this subgroup look on its own* and you want `limits`. Ask *where
does this subgroup sit inside the whole* and you want `brush`.

## Two bounds make a rectangle

One `brush` is one column. Write two, and a row has to satisfy both:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(gdp, at = c(20000, 60000)) + brush(life, at = c(78, 85))
```

The rich countries that also live longest. Two of the countries that survived
the first bound do not survive this one, and you can see exactly which two,
because they are still on the page.

Writing it as two atoms rather than one with two arguments follows the rest of
the grammar. A position is one column at a time, and so is a bound.

## A column of categories selects slots

`at` is where the selection opens. What you give it says which kind of column
you meant, in the same way a column decides whether `color` hands out a ramp or
a palette. Numbers are a range. Names are a set of slots:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(continent, at = c("Europe", "Oceania"))
```

One word, two readings, decided by the column rather than by a second atom.

## What can be brushed

A selection picks out rows, so a mark can be brushed when **one row is one
shape**: `point`, `text`, `rule` and `zone`.

A `line` is the other case. It draws one shape *through* many rows, so a row is
a corner of the line rather than a thing of its own. Selecting half of a line
would mean cutting it in two, and cutting a line in two already has a word:

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  brush(life, at = c(60, 80))
```

The refusal names both ways forward. Split the line with `group()`, or brush a
mark that draws one shape per row.

A summarized layer refuses for a different reason. A bar of `bar * count` stands
for forty countries at once, so "twelve of these are selected" has no honest
picture: dimming the whole bar would be wrong, and dimming part of it would
invent a mark nobody asked for. GOG says so instead of guessing:

```{r}
#| error: true
data(gapminder_2007) + bar * count + x(continent) +
  brush(gdp, at = c(20000, 60000))
```

Beside a layer that *can* answer, neither of these is a refusal. The plot draws,
and GOG reports which layer it left whole:

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  line * smooth +
  brush(gdp, at = c(20000, 60000))
```

The points answer the selection and the trend line does not, so the line is
drawn once, at full strength, through all 142 countries. That is the right
answer: a trend fitted to the selection alone would be a different statistic,
and the sentence did not ask for one.

A range that does not run upward selects nothing, so it is a mistake rather than
an empty selection:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  brush(gdp, at = c(60000, 20000))
```

## Selecting while the plot plays

A frame is a subset of the rows, and a bound is a test on a column's values, so
the two compose and neither needed anything added for the other. The bound is
tested again inside every frame:

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + color(continent) +
  play(year) + brush(gdp, at = c(20000, 60000))
```

Watch a country cross into the selection and light up. In 1952 almost nothing is
selected, because almost nothing was that rich. By 2007 the bright group is
large. Nothing pauses, and no frame is skipped: the same test simply gets a
different answer in each one, which is what makes the brush a threshold you
watch the world cross.

One column cannot do both jobs. Brushing the column the frames are cut on would
select frames rather than rows inside one, and that is a scrub bar rather than a
selection:

```{r}
#| error: true
data(gm_all) + point + x(gdp) + y(life) + play(year) + brush(year, at = c(1980, 2007))
```

A scrub bar belongs to the page a plot is shown on, not to the sentence that
describes the plot. [Play](play.qmd) says why.

## Moving the bound yourself

Every plot in this chapter after the first names where its selection *opens*. On a page that can
run the engine, you can drag the panel to move the bound, and the plot redraws
under your hand.

This changes nothing about the sentence, and it is the same arrangement the cube
uses. `space(turn =, tilt =)` chooses the angle a plot opens at and the mouse
takes you anywhere else; `brush(gdp, at = )` chooses the bound a plot opens with
and the mouse does the same. Drag to find a selection, then write it into the
sentence.

On paper, and in the PDF edition, each plot shows the bound its sentence asked
for. That is why naming it is worth doing even though a reader can move it. The
printed figure shows the selection you chose, and a reader who can drag starts
from it.

Written with no `at` at all, nothing is selected. The plot then draws exactly as
it would with no `brush` in the sentence, which is the resting state a page shows
before the first drag.
