# Selection

```{r}
#| include: false
source("R/setup.R")
```

Which countries in a scatter are the rich ones, and where do they sit? Finding
them among 142 dots takes a long time. Add one word to the sentence and the
reader can draw a box around any group. The countries inside the box keep their
color, and the rest are dimmed:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, and let
the reader select."*

On a page that can run the engine, drag across that plot. On paper it is the
scatter itself, because a bare `brush` selects nothing until a reader drags.

Bare `brush` means *the positions this plot binds*, so here it is a region over
`gdp` and `life` together. The sentence offers the reader a selection without
choosing in advance which part of the plot is worth selecting.

Every country is still drawn while you drag. The ones inside your box keep their
color and the rest are dimmed, so the group is read *against* what it was taken
from. Remove the other rows and the plot no longer shows that the rich countries
sit at the top right of the whole world.

## Saying where the selection starts

On paper the first plot is a plain scatter, and the reader still has to find
the rich countries. Name a column and the brush states a **bound** on it: the
low and high values the selection starts at. That is how you point at something
rather than let a reader look for it:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(gdp, at = c(20000, 60000))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, and let
the reader select gdp, starting from 20000 to 60000."*

The thirty or so countries above 20,000 dollars keep their color, and almost all
of them sit above 75 years. A bound written into the sentence works more like an
annotation than like a tool. The sentence has done the selecting, and a reader
can still drag to move the bound.

The two forms answer different questions, and the difference is who chooses.
`brush` leaves the choice to the reader. `brush(gdp, at = ...)` makes the choice
in the sentence, so the selection appears on paper too.

## Highlighting, never removing rows

This is the one rule to learn, and it separates `brush` from an atom you already
know. [`limits`](scales.qmd) also takes a range in a column's own units, and it
puts that range to a different use.

`limits` runs *before* the statistics. It removes rows, and the transform then
reads the rows that are left as if they were the whole table. `brush` runs
*after* the
picture is composed. It removes nothing and recomputes nothing.

One range, one mark, two outcomes:

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp, limits = c(20000, 60000)) + y(life) +
  theme(width = 620, height = 300) + title("limits: 109 countries gone")
```

*"Given gapminder 2007: points, x is gdp from 20000 to 60000, y is life."*

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  theme(width = 620, height = 300) + brush(gdp, at = c(20000, 60000)) +
  title("brush: all 142 still here")
```

The `limits` plot is a different plot. Its axis runs from 20,000 to 60,000, only
the countries inside that range are drawn, and both axes were fitted to the rows
that were left. Look at its y axis: the lowest number on it is 74, because every
country with a low life expectancy was removed. gog prints how many it removed
rather than removing them in silence.

The `brush` plot is the whole world, with one group keeping its color. Its axes
are the axes of all 142 countries, so the selected points are exactly where an
unbrushed plot would have put them.
The two are stacked rather than set side by side, and that is on purpose. They
share no fitted axis. Set next to each other, they would invite a reader to
compare positions that were fitted to different data.

Which of the two you want depends on your question, and the grammar lets you
write either one. Ask *how does this subgroup look on its own* and you want
`limits`. Ask *where does this subgroup sit inside the whole* and you want
`brush`.

## Two bounds make a rectangle

Which of the rich countries also live longest? That is two conditions on one
row. One `brush` is one column. Write two, and a row has to satisfy both:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  theme(width = 620) + brush(gdp, at = c(20000, 60000)) + brush(life, at = c(78, 85))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, and let
the reader select gdp, starting from 20000 to 60000, and life, starting from 78
to 85."*

These are the rich countries that also live longest. Six of the countries the
first bound selected are not selected by this one, and you can see exactly which
six, because they are still on the page.

Two bounds needed two atoms, and that follows the rest of the grammar. A
position is one column at a time, and so is a bound.

## A categorical column selects slots

Both bounds so far have been ranges of numbers. What happens when the column
holds names instead? `at` is still where the selection starts, and the column
decides which reading you get, in the same way it decides whether `color` uses a
ramp or a palette. Numbers are a range. Names are a set of slots:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) + color(continent) +
  theme(width = 620) + brush(continent, at = c("Europe", "Oceania"))
```

*"Given gapminder 2007: points, x is continent, y is life, color by continent,
and let the reader select continent, starting with Europe and Oceania."*

One word gives two readings, and the column decides which one you get, not a
second atom. Drag across the slots and the selection follows them, whole slots
at a time, because a slot cannot be split part way.

The column is on the `x` axis here, and that matters. A drag can only reach a
column the panel places. Add the same `brush(continent, at = ...)` to the first
scatter in this chapter, where `continent` is only a color. The sentence is
still legal and still dims the right countries. But there is nowhere to drag,
because no axis is measuring continents. The sentence has chosen the selection,
and dragging cannot change it.

*Which* axis places it does not matter. Here is the same sentence with `x` and
`y` exchanged, in a shorter panel:

```{r}
data(gapminder_2007) + point + x(life) + y(continent) + color(continent) +
  theme(width = 620, height = 320) + brush(continent, at = c("Europe", "Oceania"))
```

The slots now run up the left side, and the same two continents keep their
color. Drag on the panel and the selected band is horizontal instead of
vertical, still one whole slot at a time. The bound never mentions an axis. It
names a column,
so turning the plot moves the drag and leaves the sentence alone.

## Two views, one selection

This is what a brush is most useful for, and the chapter has prepared for it.
Selecting inside one plot shows you what the plot already showed. The question
worth asking is *where does this group sit in a different picture*, and that
needs a second picture:

```{r}
(data(gapminder_2007) + point + x(gdp) + y(life) +
   brush(gdp, at = c(20000, 60000)) + theme(height = 230)) /
  (data(gapminder_2007) + point + x(gdp) + y(population, scale = "log") +
     brush(gdp, at = c(20000, 60000)) + theme(height = 200))
```

*"Given gapminder 2007: points, x is gdp, y is life, and let the reader select
gdp, starting from 20000 to 60000, above points, x is gdp, y is population on a
log scale, and let the reader select gdp, starting from 20000 to 60000."*

Two plots, one page, and one bound keeping the same countries in color in both.
Drag in either and the other follows. The top plot says the rich countries live
long. The bottom plot says their populations run from under a million to over
three hundred million. Neither plot could have told you both.

Nothing was added to the grammar to link them. Look at what the two sentences
share: the column `gdp`, named on `x` in each. A bound is a fact about a
column's values rather than about a panel, so two plots that read that column
are already answering the same question. There is no name to declare, no
identifier to match, and no third atom joining them.

That is [composition](composition.qmd)'s one rule again, in a new place. Two
composed plots naming the same column on the same axis share that axis, which is
why these two are the same width and the `gdp` numbers are written once. The
selection is shared for the same reason.

## One bound reaches every panel

[Faceting](faceting.qmd) is the other way to get more than one panel, and a
brush treats it the same way. The bound is a fact about a column, and every
panel reads that column, so every panel answers:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  brush(gdp, at = c(20000, 60000)) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, and let the reader select
gdp, starting from 20000 to 60000, split into panel columns by continent."*

Five panels, one bound, and the rich countries keeping their color. Europe has
many and Africa has none, which is the comparison the split was made for.
`facet` divides the rows and `brush` dims the rest, and neither atom refers to
the other.

The panels differ in one way worth knowing. A bound is shared, because it names
a column. Pointing is not shared. In the web edition, resting the pointer on a
mark shows that row's values. They come from the panel you are pointing at,
never from a row another panel drew.

## Which one is that?

A selection shows you where a group sits. It does not say which country each dot
is, and that is usually the next question. Layer `text` over the points, and the
bound decides which labels stay legible:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  text + label(country) + style(size = 7) +
  brush(gdp, at = c(30000, 60000)) + theme(width = 620)
```

*"Given gapminder 2007: points, x is gdp, y is life, and also text, label by
country, and let the reader select gdp, starting from 30000 to 60000."*

A hundred and forty-two labels on one scatter cannot be read, and that is what a
selection answers here. The labels you did not select are dimmed with their
points, and twenty names keep their color. Drag the bound and a different twenty
become readable.

This works because `text` is one of the marks a brush reaches, on the same test
as `point`: one row is one label. No rule joins selection to labels. Each
follows its own rule, and this plot uses both at once.

## What can be brushed

A selection picks out rows, so a mark can be brushed only when *one row is one
shape*. Four marks meet that test and draw a selection: `point`, `text`, `rule`
and `zone`.

A plain `bar` meets the test as well and still refuses. One row is one bar, but
a bar's thickness is measured from the gap between its neighbors, so selected
and unselected bars would be drawn at different widths. A jittered `point` and a
repelled `text` refuse for the same kind of reason. A jittered point's offset is
derived from its place in the table. A repelled label is placed by what the
other labels around it are doing. In both cases every mark would move when the
selection changed.

A `line` is the other case. It draws one shape *through* many rows, so a row is
a corner of the line rather than a thing of its own. Selecting half of a line
would mean cutting it in two, and the grammar already has a word for that:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
#| error: true
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  brush(life, at = c(60, 80))
```

The refusal names two fixes. This sentence has already used the first:
`group()` splits the line, and each piece still runs through twelve years. So
use the second, and brush a mark that draws one shape per row.

A summarized layer refuses for a different reason. A bar of `bar * count` stands
for as many as fifty-two countries at once. There is no way to draw "twelve of
these are selected": dimming the whole bar would be wrong, and dimming part of
it would invent a mark nobody asked for. gog says so instead of guessing:

```{r}
#| error: true
data(gapminder_2007) + bar * count + x(continent) +
  brush(gdp, at = c(20000, 60000))
```

When a layer that *can* answer is present, none of these is a refusal. The
plot draws, and gog reports which layer it left whole:

```{r}
#| message: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  line * smooth +
  brush(gdp, at = c(20000, 60000))
```

*"Given gapminder 2007: points, x is gdp, y is life, and also a line derived by
smooth, and let the reader select gdp, starting from 20000 to 60000."*

The points answer the selection and the trend line does not, so the line is
drawn once, undimmed, through all 142 countries. That is the right answer: a
trend fitted to the selection alone would be a different statistic, and the
sentence did not ask for one.

A range that does not run upward selects nothing, so it is a mistake rather than
an empty selection:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  brush(gdp, at = c(60000, 20000))
```

Every refusal in this section comes from one question: can the engine say which
rows your selection holds? A played plot asks it again, because it draws only
some of its rows at a time.

## Selecting while the plot plays

A frame is a subset of the rows, and a bound is a test on a column's values, so
the two combine without either one changing. The bound is tested again inside
every frame:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + color(continent) +
  theme(width = 620) + play(year) + brush(gdp, at = c(3000, 20000))
```

*"Given all the gapminder years: points, x is gdp on a log scale, y is life,
color by continent, played by year, and let the reader select gdp, starting from
3000 to 20000."*

Watch countries cross into the selection, frame by frame. The bound stays at
3,000 to 20,000 dollars while the countries move. In 1952 the selected group is
that year's middle-income countries. By 2007 many have grown past 20,000 and
poorer ones have grown into the range. Nothing pauses, and no frame is skipped.
The bound stays still, and the countries move across it.

### What the count includes

Stop the clock, draw a small box, and the count under the plot (the line reading
`17 of 1704 selected`) looks far too large. The number is right. A bound is a
test on a column's values, so it holds or fails for every row in the table. It
does not stop at the frame you can see. The count and `show rows`, a button
under the plot, report the whole selection across every frame. The picture shows
you one frame of it.

The played plot in this section makes that concrete. `gm_all` is 1,704 rows,
which is 142 countries in each of twelve years, and one frame draws 142 of them.
Stop on 1982 and draw a box around four countries in the low corner, and the
count reads far more than four. The same box tests every year's rows, not only
1982's, and in earlier years other countries sat in that corner too. The four
you can see are the 1982 rows of a selection that spans all twelve years.

`show rows` is where you check it. The table lists the `year` of every row it
found, beside the columns the sentence maps. You can see which frame each row
belongs to, and pick out the four that belong to the frame on screen.

A faceted plot never raises this question. There every selected row is on the
page at once, so a count of the whole selection is a count of what you see. A
sequence shows one frame of twelve at a time, so most of the selected rows are
not on the screen. They are hidden in time rather than past the edge of the
page. The rule is the same one; only the picture differs.

### Choosing a bound the data crosses

The first frames of a sequence are where an animation misleads a reader. Count
what a bound selects in each frame before you write it down. From 3,000 to
20,000 dollars selects between 52 and 74 countries in every frame here. From
20,000 to 60,000 dollars selects none at all in 1952 and 1957, one in 1962, and
four by 1972. The second bound draws a true picture, and it still reads as a
broken plot, because a reader cannot tell *nothing is selected here* from *this
is not working*. Drag the bound upward and you will see it.

One column cannot both cut the frames and carry a bound. Brushing the column the
frames are cut on would select frames rather than rows inside one, and that is a
control for moving through time rather than a selection:

```{r}
#| error: true
data(gm_all) + point + x(gdp) + y(life) + play(year) + brush(year, at = c(1980, 2007))
```

Moving the clock belongs to the page a plot is shown on, not to the sentence
that describes the plot. That is what the buttons under a played plot do, and
[Play](play.qmd#what-the-page-adds) says why they have no word here.

## All three at once

[Play](play.qmd) drew the cube of countries through the years, and nothing in
it marks the middle-income countries. The [third position](space.qmd), the
frames and the selection are three separate ideas, and none of them was designed
around the others. So all three can be written in one sentence, and none of them
changes what the others do:

```{r}
data(gm_all) + point + x(gdp, scale = "log") + y(life) + z(population) +
  color(continent) + play(year) + brush(gdp, at = c(3000, 20000))
```

*"Given all the gapminder years: points, x is gdp on a log scale, y is life, z
is population, color by continent, played by year, and let the reader select
gdp, starting from 3000 to 20000."*

A cube of 142 countries, twelve frames, and one bound tested inside every one of
them. Drag the plot to turn it and the selection turns with the cloud, because
the bound picks rows and the angle picks a viewpoint, and neither asks the other
anything.

There is one difference here, and it is better stated than left for you to work
out. In the cube the drag already turns the plot, so name the bound in the
sentence. `brush` on its own would declare a region with no way to draw one.
That is the same rule as everywhere else in this chapter, reached from the
opposite direction. A drag means one thing at a time, and in the cube it means
turn.

## Moving the bound yourself

Every brushed plot in this chapter after the first states where its selection
starts. On a page that can run the engine, you can drag the panel to move the
bound, and the plot redraws as you drag.

This changes nothing about the sentence, and it is the same arrangement the cube
uses. `space(turn =, tilt =)` chooses the angle a plot starts at, and the
pointer moves it from there. `brush(gdp, at = )` chooses the bound a plot starts
with, and the pointer moves that from there too. Drag to find a selection, then
write it into the sentence.

On paper, and in the PDF edition, each plot shows the bound its sentence asked
for. That is why naming it is worth doing even though a reader can move it. The
printed figure shows the selection you chose, and a reader who can drag starts
from it.

Written with no `at` at all, `brush` selects nothing. The picture is then the
same as it would be with no `brush` in the sentence, and that is the resting
state a page shows before the first drag.

## What the page adds

A brushed plot on a web page carries a line of controls under it. None of them
are in the sentence, and that is the rule rather than an omission. They report a
selection, or they change how you look at one. Not one of them changes what the
plot claims about the data, so none of them needs a word in the grammar. It is
the same reason a plot in the cube has a line of controls and `space()` has no
word for the pointer.

They come in two lines, and the first is [the one every plot
carries](first-plot.qmd#the-boxes-under-every-plot):

| Control | What it does |
|---|---|
| `−` `+` | Makes the picture smaller or larger |
| the frame (the box that shows the whole plot) | Returns to the whole picture |
| the camera | Saves what you are looking at as a picture file |

It sits first, directly under the picture, on a plain plot and a brushed one
alike. This matters. A reader who has found the magnifiers once finds them in
the same place under every plot in this book, rather than in a place that moves
with whatever else the plot needed.

A plot that also plays carries three more buttons at the end of that same line,
which stop the clock and step it. [Play](play.qmd#what-the-page-adds) sets them
out.

The second line is what a brush adds:

| Control | What it does |
|---|---|
| `drag:` and its icons | Says what dragging the plot will do, and switches it: a rectangle, a free shape, or moving the picture |
| `40 of 142 selected` | How many rows the current bound selects. On a played plot this counts every frame, not the one on screen |
| `show rows` | Opens a table of those rows; while the table is open the button reads `hide rows`. A long selection arrives ten at a time, with `‹` and `›` under the table |
| `clear` | Puts the selection back to what the sentence asked for |
| `unstamp` | Removes every card at once. Once there are several, the button shows how many. Absent until you have made one |

A control drawn as an icon names itself when the pointer rests on it. The ones
with a word on them need no name.

### Dragging a box

A bound written in the sentence is a starting point, and dragging moves it. Drag
across the panel. A dotted box follows the pointer, and it shows *exactly*
what is bound. On a plot with one brush the box spans the whole panel in the
other direction, because that direction was not selected. On a plot with two, it
is a rectangle. A box that narrowed an axis nobody bound would be claiming a
range you did not choose.

Dragging in either direction gives the same bound, so you can start from any
corner. A click on empty space clears the selection. A click that lands on a
mark does something else, and [Stamping a point](#stamping-a-point) is about
that.

### Drawing around a group

A rectangle is what a sentence can say. Some groups are not rectangles. A cloud
of points often has a shape: a narrow tail to one side, a curve, a group sitting
apart from the rest. A box drawn around such a group includes points you did not
mean.

The second icon beside `drag:` is the **free shape**. Choose it, then draw
around the group with the pointer held down. The outline closes itself when you
let go, and everything inside it is selected. The count and the row table read
exactly as they do for a box, because a shape and a bound ask the same question:
which rows do you mean.

Nothing about the sentence changes. You do not type a shape, and the printed
page still shows the bound the sentence named. Drawing one is a way of looking,
like turning a cube, rather than a claim about the data.

That is also why the grammar has no word for it. Other tools call the gesture a
*lasso*, and gog has no such word. A shape can only be written down as a long
list of corners, and a specification is something you can write down and print.
You can still draw the shape; you cannot write it.

Two restrictions are worth knowing. A free shape needs two axes that measure, so
where one axis carries categories the drag stays a rectangle. A slot cannot be
split part way, and whole slots are what a reader needs there. And `clear`
removes the shape instead of restoring it, because a shape has no resting form
for a sentence to state.

### Reading what you selected

The count is always there. `show rows` opens a table underneath, listing the
columns the sentence maps rather than every column in the table, because the
columns you did not plot are not what you are looking at.

A long selection arrives ten rows at a time. The line under the table says which
rows you are reading, as `11–20 of 40`, and the two arrows beside it move you
through the rest. Every row you selected is reachable, and none of them pushes
the plot off the screen. A selection short enough to fit shows no arrows at all.

The table names every row you selected, in one place, away from the picture.
Draw a bound tight around one point and it names that point alone.

For a quick look without selecting anything, move the pointer over a mark and
that row's values appear beside it. Nothing is selected and nothing changes.

### Stamping a point

The pointer names one row at a time, and the name disappears as soon as you
move. That is right for a quick look and wrong for a comparison. Suppose three
points on a scatter interest you. Reading them one after another means
remembering the first two while you look at the third.

Click a mark and its values stay. That is a **stamp**: a dot on the mark, a
short line, and a small card saying what the pointer said. Click as many as you
want to compare. The card sits on the picture beside the mark it belongs to,
which is the difference between a stamp and the row table. One names rows, the
other labels the plot.

A card needs space on the picture. On a crowded scatter it covers the very
points you stamped it to read. So drag a card and it goes where you put it. The
line stretches to follow, and a long line ends in an arrow head pointing at the
row. Move four cards out to the edges and the picture underneath stays whole.

Where you put a card is where it stays. Zoom in, pan across, or drag a new
bound: the card holds its place beside its point while the picture moves under
it. Zoom far enough that the point leaves the panel and the whole stamp goes out
of sight with it. It has nothing left to point at, and a card left over an
unrelated corner would claim a point that is not there. Bring the point back and
the card is where you left it.

A played plot works the same way, for the same reason. A row there is one
country in one year, so a stamp made in 1972 names something the 1987 frame does
not draw. The stamp is hidden while the other frames run, and it returns to its
point when the sequence comes back to 1972. The card names the year it holds.
Stamp the same country in three frames and each card appears in its own frame.

There are three ways to take a stamp off. The `×` in the corner of a card
removes that card. Clicking the card itself does the same, as long as you did
not move it: if the pointer moved between press and release, the click counted
as a drag instead. `unstamp` removes every card at once.

`clear` leaves them alone, because a stamp is not a selection: clear a bound to
read your stamped rows against the whole table and they are all still there.
`clear`, `unstamp` and the frame each undo one thing, which [Three words, three
things](#three-words-three-things) sets out.

The camera saves them. Arrange the cards, press the camera, and the picture file
carries the plot with your annotations on it, each card where you put it and
each line pointing at its row. That is the camera's rule rather than a rule
about stamps. It saves what you are looking at, the same way it saves how far
you have zoomed in and the angle a cube is turned to.

You do not type a stamp, and the printed page carries none.

A stamp is for deciding which points are worth naming. A plot with a name on
every point is unreadable, and a plot with a name on the few that matter is the
one you wanted. Stamping lets you try a few before you commit to them. Write
those names into the sentence and the picture carries them itself, on the page
and in the PDF alike. That is [`text`](marks/text.qmd), and it is a layer like
any other.

Arranging the cards by hand has a written form too. `text * repel` does the same
job inside the sentence. Every label is placed outside its own dot, and one that
moved far keeps a thin line back to the point it names. That is the stamp's line
again, written down rather than dragged. Stamp to decide what is worth naming,
then write the labels down and let
[`repel`](marks/text.qmd#repel-moving-labels-off-each-other) place them.

A stamp is for looking at the data. `text` is for telling a reader what it
says.

Both can go into a presentation, and they are not the same there. A saved
picture is one session at one angle and cannot be redrawn. A sentence with
`text` in it draws again for anyone who has the data, in the PDF and on the page
and in all four languages. Stamp and save to keep this one picture. Write the
labels down to keep the plot itself.

### Where pointing cannot answer

The page computes where each row was drawn, from the row's own value and the two
ends of the axis the panel records. That is exact wherever a mark stands at its
value, and some plots put a mark somewhere else on purpose.

`jitter` sets each point beside its value so that a crowded group can be read.
`dodge` moves marks aside to clear their neighbors, and `stack` sets each on top
of the one below. A summary such as `bin` or `mean` draws one shape for many
rows, so there is no single row under the pointer to name. A polar plot bends
both axes around a circle, and a map turns two columns into places on the page
before it draws anything.

On any plot like those the pointer says nothing rather than naming a row that is
not there, and a line under the plot gives the reason. Nothing can be stamped
either. The page computes a position rather than reading one, so it stays silent
wherever it cannot compute one.

### Looking closer

A crowded corner is easier to read when it is larger. `+` and `−` magnify the
picture, and the frame returns to the whole of it. Zooming in switches the drag
to panning, because a reader who has just looked closer usually wants to move
around inside it, and the frame sets the drag back to selecting. You can switch
by hand at any time with the button on the left. The pointer also tells you
which mode you are in: a crosshair to draw a box, an open hand to move the
picture.

This is a *view*, so it *recomputes nothing*. The bars are the same bars and the
ticks are the same ticks, drawn larger. Zoom far enough into a histogram and you
will see the same bins, not finer ones. If you want the statistics re-run over a
narrower range, that is [`limits`](scales.qmd), the atom this chapter compared
with `brush`.

### Three words, three things

`clear` acts on the **selection**, `unstamp` acts on the **stamps**, and the
frame acts on the **view**. The three controls undo three different things, so
each is named for what it acts on. One word for all three would hide the
difference. The
frame sits a line above the other two, beside the zoom buttons. It puts back the
view they changed.

The `×` on a card is a fourth undo, and it is not in this line. It belongs to
the card rather than to the plot. A control under the picture acts on all of
something. A control on a card acts on that card.

None of this exists on paper, and a printed plot needs none of it. The PDF shows
the bound the sentence named.

### Your data on the page

The browser answers a brush, so the table has to be in the page. It is the whole
table, including columns the sentence never maps. The table that `show rows`
opens lists only the mapped columns, and the page carries every column.

This matters when a table has a column you would not publish. Drop the columns
you do not plot before you draw, the way you would check any file before sharing
it.
[What the file carries](utilities.qmd#what-the-file-carries) states the rule for
both kinds of plot, with the one line that applies it.

## In short

A selection runs after the picture is composed: it dims rows, removes none, and
leaves the axes where they were. The column decides the reading, a range for
numbers and a set for categories. A bound tests a column's values rather than a
panel, so plots that share a column share the selection. Everything a brush
needs is already inside the web page, the whole table included, so check what
that table carries before you publish one.

## What comes next

Selection is the last of the atoms, and every part so far has taught them one
family at a time. [The cookbook](cookbook/one-variable.qmd) starts from your
data instead: one variable first, then two, then more.
