# Space

```{r}
#| include: false
source("R/setup.R")
```

What do you do when a third measurement matters as much as the first two? Most
plots so far have been drawn on a plane: a mark, an `x`, a `y`. The grammar has a
third position channel, `z`. Binding it is enough to draw the mark in three
dimensions.

## `z`: one more vowel

Here are three flower measurements. Two of them make an ordinary scatter, and
the species already separate, although *versicolor* and *virginica* overlap a
little where they meet along the diagonal:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + color(species)
```

*"Given the iris flowers: points, x is petal length, y is sepal length, color by
species."*

Now add the third measurement as `z`. Nothing else about the sentence changes:
the same `point`, the same `x` and `y`, and one more position, which the
[Grammar](grammar.qmd) chapter calls a *vowel*. The flat plot becomes a cube:

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + z(sepal_width) + color(species)
```

*"Given the iris flowers: points, x is petal length, y is sepal length, z is
sepal width, color by species."*

That is the whole idea. A 3-D scatter is not a different chart with a different
name; it is `point` with its third position bound, the same way a horizontal bar
is just `bar` with its categories on `y`. A flat plot is a 3-D plot with `z`
held constant. So every flat plot you have read so far was a special case of
the cube, and the same rules place a mark in both.

Drawing a 3-D plot on a flat page takes two steps, and the second is the one
every flat plot already uses. First the engine turns the three coordinates of
every point to match the angle you are looking from. Then it draws the turned
points on the page, the way it draws any flat plot. Points nearer to you are
drawn last, so they cover the ones behind them. The fill is translucent, so a
point behind still shows through the one in front. The drawing order is what
gives a flat picture its depth.

The faint box behind the cloud of points is a guide rather than part of the
data. A bare cloud of points gives you no way to tell which direction is `x`,
which is `y` and which is `z`. The box draws the edges of the space the points
stand in, and each of the three axes puts its numbers along one of those edges.
Those three edges are drawn darker than the other nine.

Now look at the flowers. At this opening angle *setosa* sits apart from the
other two, as it did on the flat page, but *versicolor* and *virginica* overlap
much more than they did there. The data has not changed, so the extra overlap
comes from the angle you are looking from.

In the web edition you can turn this cube yourself. Drag it in any direction and
watch *versicolor* and *virginica*. Some angles separate the two colors, and
others leave them mixed together. When you find an angle that separates them,
read the two numbers under the plot. They are called `turn` and `tilt`, and they
describe the view you are looking at. The next section shows how to write those
two numbers into the sentence, so the plot opens at the angle you found.

## The viewing angle

A 3-D plot has to be seen from *somewhere*, and that angle is not part of the
data. It is a property of the space the data sits in, the way a polar plot needs
to know where its circle starts. [Polar](polar.qmd) is the next chapter, and
`polar(start = )` is the same kind of parameter. So the angle is not a channel
(it maps no column) and not a mark. It is `space()`:

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + z(sepal_width) + color(species) +
  space(turn = -50, tilt = 15)
```

*"Given the iris flowers: points, x is petal length, y is sepal length, z is
sepal width, color by species, turned -50, tilted 15."*

Same data and same bindings, turned 80 degrees from the default and seen from a
little lower down. The choice matters. From this side both lengths grow toward
the right of the page, and *virginica* is the longer flower on both, so it sits
to the right of *versicolor*. At the default angle the two lengths ran in
opposite directions across the page, so most of the difference between the two
flowers was lost. They still overlap in the middle, because this side spreads
the flowers by sepal length more than by petal length, and the two overlap far
more on sepal length. A low tilt points the `z` axis up the page and flattens
the floor, so in this view you can read sepal width easily and the two lengths
less easily. A higher tilt does the reverse: the floor takes up more of the page
and the heights grow shorter.

Turning the plot changes *how you look* at the data, never *what the plot
claims* about it. That is why the angle is set in `space()`, and not by a
channel or a mark. Omit `space()` and the plot opens at a default angle; use
`space()` when that default hides what you want to see.

Both angles are given in degrees, and both move your eye rather than the data.

`turn` says which side of the cube you look at. It moves your eye around the
plot, the way you walk around an object to see another side of it. At
`turn = 0` you look straight along the `x` axis, so `x` runs into the page. As
`turn` grows from 0, the direction you look moves from the `x` axis toward the
`y` axis. At `turn = 90` you look straight along `y`, so `y` runs into the page.
A negative turn moves your eye around the other way, and a full circle is 360
degrees.

`tilt` says how high your eye is. At `tilt = 0` you are level with the floor and
see the cube from the side. Raise it and you look down from higher up. At
`tilt = 90` you are directly above the cube, looking straight down. The floor
fills the picture. The `z` axis now points straight at you, so it has no length
left on the page and even its label disappears. A negative tilt puts your eye
below the floor, looking up, as far as `tilt = -90` from directly underneath.
Dragging stops at `tilt = 90` and at `tilt = -90`. A written tilt must also stay
between -90 and 90, so dragging reaches every tilt you can write.

A plot with no `space()` opens at `turn = 30, tilt = 25`. That is far enough
around and high enough up to show all three axes at once, which is why it is the
default. Name one angle and the other keeps its default, so `space(tilt = 85)`
looks down from high up without changing which side you stand on. Writing
`space()` with nothing inside it takes both defaults, which is how the histogram
sentences later in this chapter ask for a cube. (Astronomy calls these two
angles *azimuth* and *elevation*, and the plain words say the same thing.)

## Turning the plot yourself

In the web edition you turn every cube in this book the same way, by dragging
it. The cube follows your pointer. Drag right and the side facing you moves
right; drag down and it tips down to show you the top. A row under each cube
shows the two angles of the current view, and the numbers change as you drag.
So you can find an angle by hand, then write it into the sentence as
`space(turn = , tilt = )`.

A *reset* button sits beside those numbers. It returns the plot to the view it
opened at, which is the angle the sentence asked for, or the default angle if
the sentence named none. It is dimmed until you turn the plot, because until
then there is nothing to undo.

The same [buttons under every plot](first-plot.qmd#the-boxes-under-every-plot)
sit in a row of their own, between the plot and the angles, and each button does
the same work here as on a flat plot. The camera saves the picture as you are
looking at it, so the angle you turned the cube to is saved with it.

The cube differs from a flat plot in two ways, and both follow from one fact:
dragging turns the cube and everything in it, rather than moving the picture
across the page. So a zoomed cube cannot be dragged to one side, and the cube
has a second control that undoes something: *reset*. The two undo different
things: the frame button undoes the zoom, and *reset* undoes the angle. So zooming in
never loses the angle you found, and going back to the opening angle keeps the
zoom you set. A flat plot has no angle to restore, so it has no *reset*.

Turning a cube with the pointer changes nothing in its sentence. `space()` sets
the angle the plot opens at, and dragging moves the view to any other angle from
there. The data, the marks and the scales are the same at every angle.

Dragging needs the web edition. In the PDF, each plot is the still image its
sentence asks for. That is the reason to name the angle even for readers who
can drag: the PDF shows the view you chose, and the web edition opens at it.

## `path`: the route through the cube

So far the cube has held one mark. A scatter is a cloud: every dot stands alone
and the table's order means nothing. When the order of the rows is what matters,
the mark that says so is `path`, which joins rows in the order the table gives
them. Give it three positions and the route runs through the cube. Here are two
gliders circling one thermal (a rising column of warm air), sampled as they
climb:

```{r}
#| echo: false
peek(thermals)
```

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider)
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider."*

Nothing in that sentence asks for a spiral. `path` was told to join the rows in
order, and the gliders circled while they climbed. The spiral is what that
looks like when it is drawn. The same three bindings would draw a straight line
for an aircraft flying level, and a tangle for a tumbling one, because the route
is only the order of the rows.

Look at where the two routes cross. The one in front covers the one behind, and
which is which changes along the climb, so neither glider is simply "on top". A
stroke running through the cube has no single distance from the eye. Its far end
and its near end are different distances. So the engine sorts the *segments*
rather than the strokes, and sorts the segments of both routes together. Sorting
whole strokes would put one glider entirely in front of the other: a picture of
two coils that never meet, instead of two that pass through each other along the
climb.

`path` is the only mark that draws a route through the cube. The reason is the
one that already separates it from `line` on the flat page: the order of the
rows. An order belongs to no axis, so it means the same thing when the plane
becomes a cube. A `line` sorts by `x` and draws one `y` for each, read left to
right along a domain. The cube has no left to right. `x` is one of three equal
positions, and at some angles it runs straight into the page and becomes depth.
A line in the cube would be sorted by an axis you cannot see, so the engine
refuses it instead:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude)
```

## The 3-D histogram

Every cube so far bound a column to each of the three axes. The third axis can
also carry something the data does not hold. The [heatmap](marks/zone.qmd) cuts
a plane into cells and colors each one by how many rows fell in it. Those cells
already cover a plane, and in the cube that plane is the floor. The color on
each cell shows its count. Two shades of one color are hard to compare exactly,
and two heights are easy. Here is the heatmap first, and then the same counts
drawn as heights:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + zone * bin(20) + x(gdp, scale = "log") + y(life)
```

*"Given all the gapminder years: zones derived by bin into 20, x is gdp on a log
scale, y is life."*

```{r}
data(gm_all) + bar * bin(20) + x(gdp, scale = "log") + y(life) + space() +
  style(color = "#2a9d8f")
```

*"Given all the gapminder years: bars derived by bin into 20, x is gdp on a log
scale, y is life, in the cube."*

The mark changed from `zone` to `bar`, and `space()` asks for the cube. The
transform is the same `bin`, cutting the same two axes into the same cells, and
the count it invents is the same count. What differs is the mark that receives
the count. The mark decides which visual attribute carries it. A `zone` measures
nothing by length, so its count has to become a color. A `bar` *is* a length
from a baseline, so its count becomes a height. The bar's color then carries
nothing, so every bar has the same color, and `style()` only chooses which one.

Nothing was added to the grammar to do that, and the rule that decided it was
written for the flat page. How many axes a transform cuts is decided by the
mark. `bar * bin` cuts one: a bar uses one axis for the domain and keeps the
other to measure along. `zone * bin` cuts two, because a zone keeps none. A cube
gives the bar a third axis. It can use two for the floor and still have one to
measure along, so the same `bar * bin` that cuts one axis on the page cuts two
here. Read that way, the 3-D histogram needs no feature of its own. It is what
the flat rule already said, applied to three axes.

Note also what you did *not* have to write. There is no `z()` in that sentence,
for the same reason a flat histogram has no `y()`: the count is invented rather
than bound, so naming it would be naming an output.

## Transforms in the cube

That floor was cut from two continuous axes. A floor is made of cells, so the
question is where the cells come from. The [zone](marks/zone.qmd) chapter
answered it for a plane, and the same two answers hold here. A continuous axis
has to be **cut** into cells before anything can be counted in them; a
categorical axis arrives already cut, because a category *is* a cell. So which
transform you want depends on which kind of axis you have. Two categorical axes
arrive already cut:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * count + x(continent) + y(era) + space() + style(color = "#2a9d8f")
```

*"Given the gapminder eras: bars derived by count, x is continent, y is era, in
the cube."*

That sentence counts the rows for every pair of categories, and each count is
drawn as a bar's height. A floor with one axis of each kind is a distribution
per category, drawn as bars:

```{r}
data(gm_all) + bar * bin(12) + x(life) + y(continent) + space() + style(color = "#2a9d8f")
```

Compare the two floors. The `count` floor holds a category on both axes, so its
bars leave a gap in both directions. The `bin` floor holds one axis of each
kind, so its bars *touch* along the binned axis and leave a gap along the
categorical one. That is not a decision about how a 3-D plot should look; it is
the flat rule, per axis. A histogram's bins are adjacent intervals, so its bars
fill them and meet. A category is one slot, and its neighbors are separate slots
rather than adjacent intervals. A bar fills four fifths of its slot and leaves
one fifth empty, and that gap is what says the categories are separate. So a
floor cut on one axis and slotted on the other has no gaps one way and gaps the
other. Each axis follows the rule for its own kind of column, and neither
depends on the other.

Ask for a cut where there is nothing to cut and the engine names the other
transform, exactly as it does on the plane:

```{r}
#| error: true
data(gm_eras) + bar * bin + x(continent) + y(era) + space()
```

`count` and `bin` both *invent* what they measure: you give `count` no column to
reduce, because it counts rows. The other kind of statistic is the kind you give
a column to reduce: a mean, a sum, a median. Those work on a floor too, and the
sentence names the column on the axis that measures:

```{r}
data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) + space() +
  style(color = "#2a9d8f")
```

*"Given the gapminder eras: bars derived by mean, x is continent, y is era, z is
life."*

Each bar sits on the cell of one continent and one era, and its height is that
group's mean life expectancy. Read that beside the flat
`bar * mean + x(continent) + y(life)`. The rule is the
same in both. A summary reduces the column on the axis the mark measures along,
and groups by the other positions. On the flat page, a bar measures with `y`, so
it groups by `x` and averages `y`. In the cube it measures with `z`, so it
groups by the *pair* and averages `z`. Nothing about `mean` changed. The mark
simply has one more position to group by.

That is the same rule the histogram used, applied a second time: take the
positions the space offers, and set aside the one the mark measures along.
Applied to `bin`, it says how many axes are cut. Applied to `mean`, it says how
many columns the rows are grouped by. One rule answers both questions, and in
both the answer depends on the mark.

The two kinds of statistic differ in what you give them, not in what they can
draw. `count` and `bin` invent a measurement, so they need no column, and a
color or a height shows the count. `mean`, `sum`, `median`, `max`, `min` and
`quantile` reduce a column you name, so the sentence has to name it. Leave `z()`
off and the engine says so rather than guessing.

One statistic does stay on the plane. It is `smooth`, which fits a curve
instead of reducing a column:

```{r}
#| error: true
data(gm_eras) + bar * smooth + x(continent) + y(era) + z(life) + space()
```

The engine gives two reasons, and only the second belongs to the cube. The first
would refuse the same sentence on the flat page, because both floor columns are
categories. The second holds on any floor: a smoother traces a curve of one
column against another, and a curve needs a domain to run along. The cube's
floor is a pair of positions with no left to right. That is why `line` and
`area` cannot stand here either. It is the same reason, stated by a transform
this time instead of by a mark.

## `interval` and `box` in the cube

A mean per cell hides how far the values in that cell spread, and the spread is
often the question. A bar is not the only mark that sits in a slot and
measures along the other axis. An `interval` does the same with a whisker, which
spans a range from end to end, and a `box` does it with its quartiles and its
median. So all three can be drawn in the cube, each on one *cell* of the floor,
measuring along `z`:

```{r}
data(gm_eras) + interval * range + x(continent) + y(era) + z(life) + space() +
  style(color = "#2a9d8f")
```

*"Given the gapminder eras: intervals derived by range, x is continent, y is
era, z is life."*

Every cell's whisker spans that cell's own low and high. The caps (the short
lines across each end of a whisker) are a **cross** rather than a single tick,
and that is not decoration. On the flat page a cap runs across the slot, in the
one direction the whisker does not use. A cube has two such directions. Picking
one of them would be arbitrary, and a reader would look for a meaning that is
not there. So the cap runs in both.

Ask for `confidence`, which carries a center as well as two ends, and gog draws
the **pointrange**: a dot at the center with a whisker on each side.

```{r}
data(gm_eras) + interval * confidence + x(continent) + y(era) + z(life) +
  space() + style(color = "#2a9d8f")
```

*"Given the gapminder eras: intervals derived by confidence, x is continent, y
is era, z is life."*

The box plot stands in the cube the same way. Its box is a solid between the
two quartiles, and its whiskers reach beyond the box:

```{r}
data(gm_eras) + box + x(continent) + y(era) + z(life) + space() + style(color = "#2a9d8f")
```

*"Given the gapminder eras: boxes, x is continent, y is era, z is life."*

Look at the median on each box. It is drawn only where the box's own sides face
you. The median is a plane *inside* an opaque solid, so the far half of it is
behind the box. That is the back-face rule (a face turned away is not drawn),
the same rule the solid's own six faces follow. One case is handled differently.
When every value in a group is the same, its two quartiles are equal, so the box
has no height and no sides to turn toward you. The whole median is drawn
instead, as a flat rectangle. On the flat page, a box of no height becomes a
single line in the same way.

The small dots beyond the whiskers are outliers, each drawn at the
center of its cell. [Interval](marks/interval.qmd) and [Box](marks/box.qmd) are
the chapters for these two marks on the flat page.

## `z_label()`: naming the third axis

Those three marks all stood on the same two axes and measured along the third.
Whatever stands in the cube, the third axis still needs a label. Each axis takes
its label from the name of the column bound to it, and you can replace any of
the three. Each position has its own label function, and `z_label()` is
the third. Units are the usual reason to set all three at once:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  x_label("Meters east") + y_label("Meters north") + z_label("Altitude (m)") +
  title("Two gliders, one thermal")
```

Ask for a `z_label()` on a plot with no `z` and the label has no axis to name,
so gog says so and draws the plot anyway. It is guidance rather than a refusal,
because nothing else about the plot is wrong.

[Path](marks/path.qmd) is the chapter for the mark itself; the same two
gliders appear in the [third-variable cookbook](cookbook/three-or-more.qmd).

## What the cube refuses

Three sentences in this chapter have been refused already, each with a message
that says why. What follows is what the cube itself will not do, and why. The
first two are not refusals but fixed drawing orders, and the engine draws the
plot without a message. The first concerns the guide box: the engine does not
sort it by depth with the marks. The box is drawn *behind* the marks and its
labels are drawn over them, whatever the angle. So an edge of the box never
passes in front of a point, even where the edge is nearer to you. And no mark
can cover the numbers that say what it measures, because a guide describes the
scene rather than being part of it.

Turning the whole cube is cheap, and the engine does it every time you drag.
Working out, piece by piece, which part of the box hides which point costs far
more computing, and the renderer does not do it. That is a limit of the
renderer, not a rule of the grammar, so the box stays behind the marks however
far you turn the scene.

Layers keep a fixed order as well. The depth sort runs *within* a layer, and
layers are painted in the order you wrote them, exactly as they are on the flat
page. So a `point` cloud layered over a `path` sits in front of the whole route
rather than passing through it. That is the ordinary layering rule applied in
the cube, not a new exception, so read a layered 3-D plot as layers, not as one
depth-sorted scene.

A tilt beyond 90 or -90 is refused, and a turn beyond a full circle is not. One
of the two angles has an end and the other does not, and that is the whole
reason:

```{r}
#| error: true
data(thermals) + point + x(east) + y(north) + z(altitude) + space(tilt = 95)
```

`tilt` is how high your eye is, and there is a highest point. At 90 you are
directly above the cube, and at -90 directly underneath. Going past either end
would turn the scene upside down, and an upside-down cube shows nothing the
sentence asked for. So dragging stops at both ends, and a written tilt must stay
between them.

`turn` is which side you stand on, and it has no end. A full circle brings you
back, so 390 is the same view as 30 and gog draws it without comment:

```{r}
data(thermals) + point + x(east) + y(north) + z(altitude) + space(turn = 390)
```

That is `space(turn = 30)` drawn a second time: 390 is 30 plus a full circle, so
the two sentences produce exactly the same file. `tilt` has two ends and `turn`
has none, so the two rules come from the angles themselves, not from a choice
gog made.

When the cube does not draw a mark, there are three possible reasons, and the
engine's message says which one you have. It was **decided** against, or it is
**blocked**, or it is simply **not built**. A blocked mark would need the
renderer to work out, piece by piece, what hides what; a mark that is not built
only lacks its own drawing code.

Some are **decided**. `line`, `step`, `area` and `ribbon` read a *domain* left
to right. They sort by `x` and draw one value for each, which is what makes a
line a function rather than a route. A cube has no left to right, so all four
are refused there. The `path` section showed that refusal for `line`. It is a
rule rather than a missing feature, and the refusal says so: the route through
three dimensions is [`path`](marks/path.qmd), which is `line` with exactly that
sort removed. On the page `line` and `path` differ by one rule: whether the rows
are sorted by `x`. In the cube that one rule decides which of the two can be
drawn at all.

## What the cube blocks

The second reason is blocked, and it is a different kind of refusal. A `rule`
in a cube is a *plane*. It takes one position, as it does on a flat plot, and
spans the other two axes. Here `z(altitude)` places the rule, so each row would
be a horizontal plane at its own altitude, and the engine refuses the sentence:

```{r}
#| error: true
data(thermals) + rule + z(altitude) + space()
```

Marks in the cube are painted in order, from far to near. The engine places each
solid, such as a bar, in that order by its footprint, the part of the floor it
stands on. A horizontal plane spans the whole floor, so part of it is nearer to
you than a given point and part is farther away. Drawn in one piece, it could
only be entirely in front of the data or entirely behind it, but its purpose is
to cut *through* the data. A `zone` in a cube is a slab, a plane with thickness,
and it is refused for the same reason. Neither is drawn, because drawing one
means working out, piece by piece, what hides what, and the engine does not do
that.

The third reason is the simplest, and `text` is the example. A label at a third
position is valid grammar, and no part of the grammar forbids it. The engine
does not draw it, and says exactly that:

```{r}
#| error: true
data(iris_flowers) + text + x(petal_length) + y(sepal_length) + z(sepal_width) +
  label(species)
```

The third reason covers a scale as well. The `z` axis is linear, so a log `z` is
refused, and the message says to use a log scale on `x` or `y` instead.

Every refusal above comes with a message that says why. Silence would be worse
than a refusal, because a plot that quietly dropped a dimension would still look
finished and would answer a question you never asked.

## Faceting: one cube per panel

A cube of 142 countries mixes five continents together. The question is how one
continent's cloud differs from another's. A facet splits the rows into panels,
and a panel is a rectangle. The cube is drawn inside a rectangle. So a faceted
3-D plot is one cube per panel, each projected into its own panel:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + z(population) +
  color(continent) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, z is population, color by
continent, split into panel columns by continent."*

The scales are shared, exactly as they are on the flat page, so the five cubes
share one set of axes, and a given height means the same population in every
panel. Oceania
holds two countries, so its cube is nearly empty, and that is the correct
reading. The shared scale is what makes the difference visible.

The same split, on a route rather than a cloud:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) |
  facet(glider)
```

The guides are the one thing that needs explaining. Each cube draws its own tick
numbers and its own three axis labels. A faceted flat plot writes each axis once,
along an edge of the page. That is a rule rather than an oversight. A guide
drawn on the panel's boundary is drawn once, because the panels share that
boundary. A guide drawn inside the panel appears in every panel, because there
is no boundary to share it on. A faceted polar plot works the same way: each
panel draws the category names around its own circle. A cube's three axes are
edges of the cube,
and the cube sits inside the panel. A shared column of numbers beside the page
would have nothing to point at. The tick reading -200 lands in a different place
in every panel, because each panel projects from its own rectangle.

A cube's three axes are its three darker edges. Which edges they are depends on
the viewing angle, and the engine picks them by a rule. The cube has four edges
parallel to each axis. `x` and `y` take the two floor edges nearest you. `z`
takes the vertical edge on the left side of the cube's outline. Seen from above,
the two floor edges form a V shape at the bottom of the picture. In that view
all three lie on the outline, so their numbers sit outside the cube, away from
the marks. [Surface](marks/surface.qmd#one-sheet-four-angles) shows the same
rule: the surface fills most of the picture, and the height numbers still sit
outside it. Before any text is drawn, the engine checks the numbers and axis
labels against each other for overlap. Where two would overlap, it moves one of
them farther out, or leaves it out. So numbers from two axes never overlap where
their edges meet, and an axis label never covers its own last number.

One case is left: an axis that is drawn very short. Tilt the eye up and the
measure axis is drawn shorter, until its numbers no longer fit. Here is the same
summary at two tilts; the second is seen from almost straight above:

```{r}
((data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) + space() +
    style(color = "#2a9d8f") +
    z_label("Mean life expectancy") + title("The three-quarter view")) |
   (data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) +
      space(tilt = 85) + style(color = "#2a9d8f") +
      z_label("Mean life expectancy") + title("Tilted up, with fewer numbers"))) +
  theme(height = 430)
```

*"Given the gapminder eras: bars derived by mean, x is continent, y is era, z is
life, beside the same tilted 85."*

The right-hand plot draws fewer numbers up its measure axis, and that is all it
does about the crowding: it drops *numbers* and never changes the scale. A
coarser tick step would also clear the crowding, and gog does not use it. The
step decides where the axis stops, so a bigger step pushes the top of the axis
above the tallest bar and leaves empty space above the bars. The numbers that
remain are always a subset of the ones the left-hand plot shows. Each one sits
at the same height in both, so a value read from either panel is the same value.

A second kind of split lays the panels out in time instead of across the page.
[`play`](play.qmd) is that chapter, and it also shows the two kinds of split
used together.

## In short

Binding `z` places a mark in the cube, and nothing else in the sentence
changes. A transform cuts or groups by each position the mark does not measure
along, so `bin` cuts two axes in the cube and one on the page. Only `path` draws
a route through the cube: it follows the order of the rows, and an order belongs
to no axis. Turning the plot changes how you look at it, never what it claims
about the data.

## What comes next

A cube keeps the axes straight and adds one. [Polar](polar.qmd) bends them
instead: the same marks on the same data, with `x` running around a circle and
`y` out from its center.
