# Space

```{r}
#| include: false
source("R/setup.R")
```

What do you do when a third measurement matters as much as the first two? Every
plot so far has been drawn on a plane: a mark, an `x`, a `y`. The grammar has a
third position channel, `z`. Binding it is enough to lift the mark off the
page.

## `z`: one more vowel

Here are three flower measurements. Two of them make an ordinary scatter, and
the species already begin to separate, but *versicolor* and *virginica* still
blur together along the diagonal:

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + color(species)
```

*"Given the iris flowers: points, x is petal length, y is sepal length, color by
species."*

Now add the third measurement as `z`. Nothing else about the sentence changes
(the same `point`, the same `x` and `y`, one more vowel), and the plot lifts
into space:

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + z(sepal_width) + color(species)
```

*"Given the iris flowers: points, x is petal length, y is sepal length, z is
sepal width, color by species."*

That is the whole idea. A 3-D scatter is not a different chart with a different
name; it is `point` with its third position bound, the same way a horizontal bar
is just `bar` with its categories on `y`. A flat plot is a spatial one with `z`
held constant, so the same rules that place a mark on the page place it in the
cube. Every flat plot you have read so far was the special case.

Drawing a 3-D plot on a flat page takes two steps, and neither one is new.
First the engine turns the three coordinates of every point to match the angle
you are looking from. Then it draws the turned points on the page, the way it
draws any flat plot. Points nearer to you are drawn last, so they sit on top of
the ones behind them. The fill is translucent, so a point behind still shows
through the one in front. The drawing order is what gives a flat picture its
depth.

The faint box behind the cloud of points is a guide rather than part of the
data. A bare cloud of points gives you no way to tell which direction is `x`,
which is `y` and which is `z`. The box draws the edges of the space the points
stand in, and each of the three axes puts its numbers along one of those edges.

On a screen you can turn this cube yourself. Drag it in any direction and watch
*versicolor* and *virginica*. Some angles separate the two colors, and others
leave them mixed together. When you find an angle that separates them, read the
two numbers under the plot. They are called `turn` and `tilt`, and they describe
the view you are looking at. The next section shows how to write those two
numbers into the sentence, so the plot opens at the angle you found.

## The viewing angle

A 3-D plot has to be seen from *somewhere*, and that angle is not part of the
data. It is a property of the space the data sits in, the way a polar plot needs
to know where its circle starts. [Polar](polar.qmd) is the next chapter, and
`polar(start = )` is the same kind of parameter. So the angle is not a channel
(it maps no column) and not a mark. It is `space()`:

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + z(sepal_width) + color(species) +
  space(turn = -50, tilt = 15)
```

*"Given the iris flowers: points, x is petal length, y is sepal length, z is
sepal width, color by species, turned -50, tilted 15."*

Same data, same bindings: turned around and seen from lower down. The choice
matters. A low tilt stands the `z` axis up the page and flattens the floor, so
this view reads sepal width well and the two lengths poorly. *Setosa* stands
clear on sepal width, but *versicolor* and *virginica* overlap on it almost
completely, so this is the wrong view for telling those two apart. A higher tilt
does the reverse: the floor opens out and the heights compress. Turning the plot
changes *how you look*, never *what is drawn*, which
is exactly why it belongs to the coordinate space and not to the grammar of the
plot itself. Omit `space()` and the plot opens at a default angle; use `space()`
when that default hides what you came to see.

Both angles are given in degrees, and both move your eye rather than the data.

`turn` says which side of the cube you look at. It moves your eye around the
plot, the way you walk around an object to see another side of it. A positive
turn moves you one way around, a negative turn moves you the other way, and a
full circle is 360 degrees.

`tilt` says how high your eye is. At `tilt = 0` you are level with the floor and
see the cube from the side. Raise it and you look down from further up. At
`tilt = 90` you are directly above the cube, looking straight down. The floor
fills the picture. The `z` axis now points straight at you, so it has no length
left on the page and even its name disappears. A negative tilt puts your eye
below the floor, looking up, as far as `tilt = -90` from directly underneath.
Dragging stops at `tilt = 90` and at `tilt = -90`. Those are the same two ends a
written angle stops at, so the mouse reaches every angle you can name.

A plot with no `space()` opens at `turn = 30, tilt = 25`. That is far enough
around and high enough up to show all three axes at once, which is why it is the
default. Name one angle and the other keeps its default, so `space(tilt = 85)`
looks down from high up without changing which side you stand on. Writing
`space()` with nothing inside it takes both defaults, which is how the histogram
sentences later in this chapter ask for a cube. (Astronomy calls these two
angles *azimuth* and *elevation*, and the plain words say the same thing.)

## Turning the plot yourself

You turned the first cube in this chapter with the mouse, and every cube in the
book turns the same way. The cube follows your pointer. Drag right and the side
facing you moves right; drag down and it tips down to show you the top. A line
under each plot names the two angles of the view it is showing, and the numbers
change as you drag. So you can find an angle by hand, then write it into the
sentence as `space(turn = , tilt = )`.

A *reset* button sits beside those numbers. It returns the plot to the view it
opened at, which is the angle the sentence asked for, or the default angle if
the sentence named none. It is dimmed until you turn the plot, because until
then there is nothing to undo.

Beside it sit the same boxes that
[every plot in this book](first-plot.qmd#the-boxes-under-every-plot) has.
They mean here what they mean everywhere. The camera saves the picture as you
are looking at it, so the angle you turned the cube to is saved with it.

The cube differs from a flat plot in two ways, and both follow from one fact:
dragging turns the scene rather than moving the picture. So a zoomed cube cannot
be dragged to one side, and the magnifiers are the only way to look closer.

And the cube is the one plot with *two* boxes that undo something. There are
two because they undo different things: the frame undoes the zoom, and *reset*
undoes the angle. So zooming in never loses the angle you found, and
straightening the angle keeps the zoom you set. A flat plot has no
angle to return, so it has the frame alone.

This changes nothing about the sentence. `space()` chooses where a plot opens,
and the mouse takes you anywhere else from there. The data, the marks and the
scales are the same at every angle.

Dragging needs a web page to run in. On paper, and in the PDF edition, each plot
is the still image its sentence asks for. That is the reason to name the angle
even for readers who can drag: the printed figure shows the view you chose, and
a reader with a mouse starts from it.

## `path`: the route through the cube

So far the cube has held one mark. A scatter is a cloud: every dot stands alone
and the table's order means nothing. When the order of the rows is what matters,
the mark that says so is `path`, which joins rows in the order the table gives
them. Give it three positions and the route runs through the cube. Here are two
gliders circling one thermal (a rising column of warm air), sampled as they
climb:

```{r}
#| echo: false
peek(thermals)
```

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider)
```

*"Given the thermals: paths, x is east, y is north, z is altitude, color by
glider."*

Nothing in that sentence asks for a spiral. `path` was told to join the rows in
order, and the gliders went round while they went up. The coil is what that
looks like when it is drawn. The same three bindings would draw a straight line
for an aircraft flying level, and a tangle for a tumbling one, because the route
is only the order of the rows.

Look at where the two routes cross. The one in front covers the one behind, and
which is which changes along the climb, so neither glider is simply "on top". A
stroke running through the cube has no single distance from the eye. Its far end
and its near end are different distances. So the engine sorts the *segments*
rather than the strokes, and sorts the segments of both routes together. Sorting
whole strokes would put one glider entirely in front of the other: a picture of
two coils that never meet, instead of two that pass through each other along the
climb.

`path` is the only mark that draws a route through the cube. The reason is the
one that already separates it from `line` on the flat page: the order of the
rows. An order belongs to no axis, so it means the same thing when the plane
becomes a cube. A `line` sorts
by `x` and
draws one `y` for each, read left to right along a domain. The cube has no left
to right. `x` is one of three equal positions, and at some angles it runs
straight into the page and becomes depth. A line in space would be sorted by an
axis you cannot see, so the engine refuses it instead:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude)
```

## The 3-D histogram

Every cube so far bound a column to each of the three axes. The third axis can
also carry something the data does not hold. The [heatmap](marks/zone.qmd) cuts
a plane into cells and colors each one by how many rows fell in it. Those cells
already cover a plane, which is the cube's floor once the plot lifts, and the
color represents a quantity. Two shades of one
color are hard to compare exactly, and two heights are easy. Bind that quantity
to a height instead and the same counts are drawn as standing bars:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + zone * bin(20) + x(gdp, scale = "log") + y(life)
```

*"Given all the gapminder years: zones derived by bin into 20, x is gdp on a log
scale, y is life."*

```{r}
data(gm_all) + bar * bin(20) + x(gdp, scale = "log") + y(life) + space() +
  style(color = "#2a9d8f")
```

*"Given all the gapminder years: bars derived by bin into 20, x is gdp on a log
scale, y is life, in the cube."*

The mark changed from `zone` to `bar`, and `space()` asks for the cube. The
transform is the same `bin`, cutting the same two axes into the same cells, and
the count it invents is the same count. What differs is the mark the count was
handed to. The mark decides which visual attribute carries it. A `zone` measures
nothing by length, so its count has to become a color. A `bar` *is* a length
from a baseline, so its count becomes a height.

Nothing was added to the grammar to do that, and the rule that decided it was
written for the flat page. How many axes a transform cuts is decided by the
mark. `bar * bin` cuts one: a bar spends an axis on the domain and keeps the
other to measure along. `zone * bin` cuts two, because a zone keeps none. A cube
hands the bar a third axis. It can spend two on the floor and still have one to
measure along, so the same sentence that cuts one axis on the page cuts two
here. Read that way the 3-D histogram is not a feature. It is what the flat rule
already said, applied to three axes.

Note also what you did *not* have to write. There is no `z()` in that sentence,
for the same reason a flat histogram has no `y()`: the count is invented rather
than bound, so naming it would be naming an output.

## Transforms in the cube

That floor was cut from two continuous axes. A floor is made of cells, so the
question is where the cells come from. The [zone](marks/zone.qmd) chapter asked
it of a plane, and the answers here are the same two. A continuous axis has to
be **cut** into cells before anything can be counted in them; a categorical axis
arrives already cut, because a category *is* a cell. So which transform you want
depends on which kind of axis you have. Two categorical axes arrive already cut:

```{r}
#| echo: false
peek(gm_eras)
```

```{r}
data(gm_eras) + bar * count + x(continent) + y(era) + space() + style(color = "#2a9d8f")
```

*"Given the gapminder eras: bars derived by count, x is continent, y is era, in
the cube."*

That one counts into slots on both axes: a table of counts, standing up. And a
floor with one axis of each kind is a distribution per category, drawn as bars:

```{r}
data(gm_all) + bar * bin(12) + x(life) + y(continent) + space() + style(color = "#2a9d8f")
```

Compare the two floors. The `count` floor holds a category on both axes, so its
bars leave a gap in both directions. The `bin` floor holds one axis of each
kind, so its bars *touch* along the binned axis and leave a gap along the
categorical one. That is not a decision about how a 3-D plot should look; it is
the flat rule, per axis. A histogram's bins are adjacent intervals, so its bars
fill them and meet. A category is one slot, and its neighbors are separate slots
rather than adjacent intervals. A bar fills four fifths of its slot and leaves
one fifth empty, and that gap is what says the categories are separate. So a
floor cut on one axis and slotted on the other has no gaps one way and gaps the
other. Each axis follows its own kind, and neither looks at the other.

Ask for a cut where there is nothing to cut and the engine names the other
transform, exactly as it does on the plane:

```{r}
#| error: true
data(gm_eras) + bar * bin + x(continent) + y(era) + space()
```

`count` and `bin` both *invent* what they measure: you give `count` no column to
reduce, because it counts rows. The other kind of statistic is the kind you give
a column to reduce: a mean, a sum, a median. Those work on a floor too, and the
sentence names the column on the axis that measures:

```{r}
data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) + space() +
  style(color = "#2a9d8f")
```

*"Given the gapminder eras: bars derived by mean, x is continent, y is era, z is
life."*

Read that beside the flat `bar * mean + x(continent) + y(life)`. The rule is the
same in both. A summary reduces the column on the axis the mark measures along,
and groups by the other positions. On the flat page, a bar measures with `y`, so
it groups by `x` and averages `y`. In the cube it measures with `z`, so it
groups by the *pair* and averages `z`. Nothing about `mean` changed. The mark
simply has one more position to group by.

That is the same rule the histogram used, read a second time: take the positions
the space offers, and set aside the one the mark measures along. The first
reading says how many axes `bin` cuts. This one says how many columns `mean`
groups by. One rule, two questions, and the answer to both is read off the mark.

The two families differ in what you give them, not in what they can draw.
`count` and `bin` invent a measurement, so they need no column, and color or
height carries the answer. `mean`, `sum`, `median`, `max`, `min` and `quantile`
reduce a column you name, so the sentence has to name it. Leave `z()` off and
the engine says so rather than guessing.

One statistic really does stay on the plane. It is `smooth`, which fits a curve
instead of reducing a column:

```{r}
#| error: true
data(gm_eras) + bar * smooth + x(continent) + y(era) + z(life) + space()
```

A smoother traces a curve of one column against another, and a curve needs a
domain to run along. The cube's floor is a pair of positions with no left to
right. That is why `line` and `area` cannot stand here either. It is the same
reason, stated by a transform this time instead of by a mark.

## `interval` and `box` in the cube

A mean per cell hides how far the values in that cell spread, and the spread is
often the question. A bar is not the only mark that stands in a slot and
measures along the other axis. A whisker spans a range from end to end. A box
summarizes a range with its quartiles and its median. `bar`, `interval` and
`box` ask the same question of the same pair of axes, so all three take the
third dimension. Give the pair a cube and each stands on a *cell* of the floor,
measuring along `z`:

```{r}
data(gm_eras) + interval * range + x(continent) + y(era) + z(life) + space() +
  style(color = "#2a9d8f")
```

*"Given the gapminder eras: intervals derived by range, x is continent, y is
era, z is life."*

Every cell's whisker spans that cell's own low and high. The caps are a
**cross** rather than a single tick, and that is not decoration. On the flat
page a cap runs across the slot, in the one direction the plot has left over. A
cube leaves two directions over. Picking one of them would be arbitrary, and a
reader would look for a meaning that is not there. So the cap runs in both.

Ask for the interval that carries a center and gog draws the **pointrange**: a
dot at the center with a whisker on each side.

```{r}
data(gm_eras) + interval * confidence + x(continent) + y(era) + z(life) +
  space() + style(color = "#2a9d8f")
```

*"Given the gapminder eras: intervals derived by confidence, x is continent, y
is era, z is life."*

And the five-number summary stands the same way, its box a solid between the
quartiles with the whiskers reaching past it:

```{r}
data(gm_eras) + box + x(continent) + y(era) + z(life) + space() + style(color = "#2a9d8f")
```

*"Given the gapminder eras: boxes, x is continent, y is era, z is life."*

Look at the median on each box. It is drawn only where the box's own sides face
you. The median is a plane *inside* an opaque solid, so the far half of it is
behind the box. That is the back-face rule (a face turned away is not drawn),
the same rule the solid's own six faces follow. One case escapes the rule. When
every value in a group is the same, its two quartiles are equal, so the box has
no height and no sides to turn toward you. The whole median is drawn instead,
and you see the plane face-on. A box of no height collapses to the same flat
line on the page.

The small dots standing clear of the whiskers are outliers, each drawn at the
center of its cell. [Interval](marks/interval.qmd) and [Box](marks/box.qmd) are
the chapters for these two marks on the flat page.

## `z_label()`: naming the third axis

Those three marks all stood on the same two axes and measured along the third.
Whatever stands in the cube, that third axis still needs its name. The cube's
three axes take their names from the columns bound to them, and you can replace
any of the three. Each position has its own label function, and `z_label()` is
the third. Units are the usual reason to set all three at once:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  x_label("Meters east") + y_label("Meters north") + z_label("Altitude (m)") +
  title("Two gliders, one thermal")
```

Ask for a `z_label()` on a plot with no `z` and the label has no axis to name,
so gog says so and draws the plot anyway. It is guidance rather than a refusal,
because nothing else about the plot is wrong.

[Path](marks/path.qmd) is the chapter for the mark itself; the same two
gliders appear in the [third-variable cookbook](cookbook/three-or-more.qmd).

## What the cube refuses

Every sentence so far in this chapter draws. What follows is the other half:
what the cube will not do, and why. Two of those limits come from the guide box,
so start with how the box is drawn. The box is drawn *behind* the marks, so the
points are always in front of it. That is a fixed order rather than a
calculation, and it
means no point disappears behind the far face of the box. Its *labels* go on
afterwards, over the marks. A guide is an annotation about the scene rather than
a member of it, so no mark can cover the numbers that say what it measures.

Turning the plot is what makes a 3-D plot readable at all, and the mouse does
it. Deciding what hides what, piece by piece, is a separate and harder job. It
takes far more computing than turning does, and the renderer decides it rather
than the grammar. So the box stays behind the marks however far you turn the
scene.

A tilt past straight up or straight down is refused, and a turn past a full
circle is not. One of the two angles has an end and the other does not, and that
is the whole reason:

```{r}
#| error: true
data(thermals) + point + x(east) + y(north) + z(altitude) + space(tilt = 95)
```

`tilt` is how high your eye is, and there is a highest point. At 90 you are
directly
above the cube and at -90 directly underneath, so there is nothing past either
end to look from. Past either end the scene turns over, and an upside-down cube
shows nothing the sentence asked for. Dragging stops at both ends, and writing
an angle stops there too.

`turn` is which side you stand on, and you can keep walking. A full circle
brings you back, so 390 is the same view as 30 and gog draws it without
comment:

```{r}
data(thermals) + point + x(east) + y(north) + z(altitude) + space(turn = 390)
```

The same picture as `space(turn = 30)`, to the byte. Two angles and two rules.
One of them has ends to stop at and the other comes back around, so the
difference is in the angles and not in a preference.

A mark the cube does not draw has three possible reasons, and the engine says
which one you have. It was **decided** against, it is **blocked** on work the
renderer cannot do yet, or it is simply **not built**.

Some are **decided**. `line`, `step`, `area` and `ribbon` read a *domain* left
to right. They sort by `x` and draw one value for each, which is what makes a
line a function rather than a route. A cube has no left to right, so all four
are refused in space. The `path` section showed that refusal for `line`. It is
a rule rather than a gap, and the refusal says so: the route through three
dimensions is [`path`](marks/path.qmd), which is `line` with exactly that sort
removed. Two marks that differ by one rule on the page turn out to differ by a
whole dimension in the cube.

## What the cube blocks

The second reason is **blocked**, and it is a different kind of refusal. A
`rule` marks a value on one axis and spans the ones it does not name: in a cube
that is *two* axes, so a rule here is a **plane**:

```{r}
#| error: true
data(thermals) + rule + x(east) + z(altitude) + space()
```

Marks in space are painted far-to-near and sorted by the area they cover, and a
plane covers the whole floor. It could only be drawn wholly in front of or
wholly behind the data, but cutting *through* the data is the whole job of a
reference plane. A `zone` is the same problem one dimension wider. Neither is
drawn, because the engine places each mark by its whole footprint rather than
piece by piece.

The third reason is the shortest, and `text` is where you meet it. A label at a
third position is legal grammar, and no rule stands against it. The engine does
not draw it yet, and says exactly that:

```{r}
#| error: true
data(iris_flowers) + text + x(petal_length) + y(sepal_length) + z(sepal_width) +
  label(species)
```

The third reason covers a scale as well. The `z` axis is linear, so a log `z` is
refused, and the message says to put the log channel on `x` or `y` instead.

In every case the engine tells you what it will not do. Silence would be worse
than a refusal, because a plot that quietly dropped a dimension would still look
finished and would answer a question you never asked.

## Faceting: one cube per panel

A cube of 142 countries mixes five continents together. The question is how one
continent's cloud differs from another's. A facet splits the rows into panels,
and a panel is a rectangle. The cube is drawn inside a rectangle. So a faceted
3-D plot is one cube per panel, each projected into its own panel:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + z(population) +
  color(continent) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp, y is life, z is population, color by
continent, split into panel columns by continent."*

The scales are shared, exactly as they are on the flat page, so the five cubes
are the same cube five times and a height in one is a height in another. Oceania
holds two countries, so its cube is nearly empty, and that is the correct
reading. The shared scale is what makes the difference visible.

The same split, on a route rather than a cloud:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) |
  facet(glider)
```

The guides are the one thing that needs explaining. Each cube draws its own tick
numbers and its own three axis names. A faceted flat plot writes each axis once,
along an edge of the page. That is a rule rather than an oversight. A guide
drawn on the panel's boundary is drawn once, because the panels share that
boundary. A guide drawn inside the panel appears in every panel, because there
is no boundary to share it on. A faceted polar plot already works this way, with
a category ring inside every panel. A cube's three axes are edges of the cube,
and the cube sits inside the panel. A shared column of numbers beside the page
would have nothing to point at. The tick reading -200 lands in a different place
in every panel, because each panel projects from its own rectangle.

Which edges carry the numbers is chosen per viewing angle, and the choice
follows a rule. The cube offers four parallel edges per axis. The two domains
take the floor edges nearest you, the pair that makes the V at the bottom of the
picture. The measure axis takes whichever vertical edge sits on the cube's
outline. All three lie outside the data, so nothing drawn inside the cube can be
in front of them. The same choice is visible in
[Surface](marks/surface.qmd#one-sheet-four-angles): the vertical edge on the
outline still carries its heights while the surface fills the rest of the
picture. The labels are then measured against each other before any of them is
drawn. So two axes cannot print their numbers in the same
place where their edges meet, and an axis name cannot land on its own last tick.

One case is left: an axis that is drawn very short. Tilt the eye up and the
measure axis is drawn shorter, until its numbers have nowhere to sit. The same
summary at two tilts, and the second one is seen from almost straight above:

```{r}
((data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) + space() +
    style(color = "#2a9d8f") +
    z_label("Mean life expectancy") + title("The three-quarter view")) |
   (data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) +
      space(tilt = 85) + style(color = "#2a9d8f") +
      z_label("Mean life expectancy") + title("Tilted up, with fewer numbers"))) +
  theme(height = 430)
```

*"Given the gapminder eras: bars derived by mean, x is continent, y is era, z is
life, beside the same tilted 85."*

The right-hand plot draws fewer numbers up its measure axis, and that is all it
does about the crowding: it drops *labels* and never changes the scale. The
A coarser tick step would also clear the crowding, and gog does not use it. The
step decides where the axis stops, so a bigger step pushes the top of the axis
above the tallest bar and wastes the space above the bars. The numbers that
remain are always a subset of the
ones the left-hand plot shows. Each one sits at the same height in both, so a
height read off either panel is the same height.

One more case, and it is about layers rather than marks. The depth sort runs
*within* a layer, and layers are painted in the order you wrote them, exactly as
they are on the flat page. So a `point` cloud layered over a `path` sits in
front of the whole route rather than passing through it. That is the ordinary
layering rule applied in space, not a new exception, so read a layered 3-D plot
as layers, not as one depth-sorted scene.

A second kind of split lays the panels out in time instead of across the page.
[`play`](play.qmd) is that chapter, and it is where the two are combined.

## In short

Binding `z` is the whole of it. The mark lifts, and nothing else in the sentence
is rewritten. Turning the plot changes how you look and never what is drawn,
which is why the angle belongs to the space.

## What comes next

A cube keeps the axes straight and adds one. [Polar](polar.qmd) bends them
instead: the same marks on the same data, with `x` running around a circle and
`y` out from its center.
