# Space

```{r}
#| include: false
source("R/setup.R")
```

What do you do when a third measurement matters as much as the first two? Every
plot so far has lived in a plane: a mark, an `x`, a `y`. The grammar has always
held a third position channel, `z`, waiting for the mark to stand up off the
page. Binding it is all it takes.

## One more vowel

Here are three flower measurements. Two of them make an ordinary scatter, and
the species already begin to sort themselves out, but *versicolor* and
*virginica* still blur together along the diagonal:

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + color(species)
```

*"Given the iris flowers: points, x is petal length, y is sepal length, color by species."*

Now add the third measurement as `z`. Nothing else about the sentence changes
(the same `point`, the same `x` and `y`, one more vowel), and the plot lifts into
space:

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + z(sepal_width) + color(species)
```

That is the whole idea. A 3-D scatter is not a different chart with a different
name; it is `point` with its third position bound, the same way a horizontal bar
is just `bar` with its categories on `y`. Two dimensions are the degenerate case
of three (a flat plot is a spatial one with `z` held constant), so there are not
two engines here, one for planes and one for space. There is one, and the flat
plot you have been reading all along was the quiet special case.

The marks come cheaply: each point is rotated to the viewing angle and dropped
onto the page, near points painted last so they sit in front of far ones. The
faint box behind the cloud is the guide: the frame that tells you which way is
which, since a bare cloud of dots carries no orientation on its own.

## The viewing angle

A 3-D plot has to be seen from *somewhere*, and that angle is not part of the
data. It is a property of the space the data sits in, the way a polar plot
needs to know where its circle starts. [Polar](polar.qmd) is the sibling
chapter, and `polar(start = )` is the same kind of parameter. So it is not a
channel (it maps no column) and not a mark. It is `space()`:

```{r}
data(iris_flowers) + point +
  x(petal_length) + y(sepal_length) + z(sepal_width) + color(species) +
  space(turn = -50, tilt = 15)
```

Same data, same bindings: swung around and viewed from lower down. The choice
matters: from this angle the three species peel apart along the depth axis more
cleanly than they did from the default three-quarter view, and a different angle
again would bury them. Rotating the plot changes *how you look*, never *what is
drawn*, which is exactly why it belongs to the coordinate space and not to the
grammar of the plot itself. Omit `space()` and you get the default view that
shows all three axes at once; reach for it when the default hides what you came
to see.

`turn` swings the scene around its upright axis (which side you look at it from),
and `tilt` lifts your eye above the floor, from a flat side-on view up to
looking straight down. Both are in degrees. (Astronomy calls these two angles
*azimuth* and *elevation*, and the plain words say the same thing.)

## Turning the plot yourself

Every plot in the cube can be turned with the mouse. Drag left or right to swing
the scene around. Drag up or down to raise or lower your eye. A line under each
plot reads out the two angles while you move it. So you can find an angle by
hand, then write it into the sentence as `space(turn = , tilt = )`. A *reset*
button returns the plot to the view it opened with.

Beside it are the two magnifiers and the frame that sit under
[every plot in this book](first-plot.qmd#the-four-boxes-under-the-plot). They
mean here what they mean everywhere: look closer, look wider, return to the whole
picture, and change nothing about what the plot says.

The cube differs from a flat plot in two ways, and both follow from the drag
being spoken for. There is **no hand**, because dragging turns the scene rather
than moving the picture, so the zoom is on the buttons alone.

And the cube is the one plot with *two* buttons that undo something. They are two
because they undo different things: the frame returns the zoom, and *reset*
returns the angle. So looking closer never costs you an angle you worked to
find, and straightening the angle never costs you your zoom. A flat plot has no angle, so it carries *fit* alone.

This changes nothing about the sentence. `space()` chooses where a plot opens,
and the mouse takes you anywhere else from there. The data, the marks and the
scales are the same at every angle. You are changing *how you look*, never
*what is drawn*.

Dragging needs a web page to run in. On paper, and in the PDF edition, each plot
is the still image its sentence asks for. So the angle is still worth naming,
even though a reader can move it. The printed figure shows the view you chose,
and a reader who can drag starts from it.

## A route, not a cloud

A scatter is a cloud: every dot stands alone and the table's order means nothing.
When the order *is* the story, the mark that says so is `path`, which joins rows
in the order the table gives them. Give it three positions and the route runs
through the cube. Here are two gliders circling one thermal, sampled as they
climb:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider)
```

Nothing in that sentence asks for a spiral. `path` was told to join the rows in
order; the gliders went round while they went up; the coil is what that is. The
same three bindings would draw a straight line for an aircraft on a heading, and
a knot for a tumbling one, because the route is whatever the rows did.

Look at where the two threads cross. The one in front covers the one behind, and
which is which changes along the climb, so neither glider is simply "on top". A
stroke running through the cube has no single distance from the eye (its far end
and its near end are different distances), so the engine sorts the *segments*
rather than the strokes, and sorts those of both routes together. Sorting whole
strokes would put one glider entirely in front of the other: a picture of two
coils that never meet, instead of two that thread through each other.

`path` is the only stroke that takes `z`, and the reason is the one that
separates it from `line` down on the plane. An order belongs to no axis, so it
means the same thing when the plane becomes a cube. A `line` sorts by `x` and
draws one `y` for each, read left to right along a domain, and the cube has no
left to right: `x` is one of three equal positions, and at some viewing angles it
runs straight into the page and becomes depth. A line in space would be sorted by
an axis you cannot see, so it says so instead:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude)
```

## A field read as a height

The [heatmap](marks/zone.qmd) cuts a plane into cells and colors each one by how
many rows fell in it. The cells are already a floor plan; what a color is standing
in for is a quantity. Bind that quantity to a height instead and the same sentence
draws the same field as columns:

```{r}
data(gm_all) + zone * bin(20) + x(gdp, scale = "log") + y(life)
```

```{r}
data(gm_all) + bar * bin(20) + x(gdp, scale = "log") + y(life) + space() +
  style(color = "#2a9d8f")
```

One word changed. The transform is the same `bin`, cutting the same two axes into
the same cells, and the count it invents is the same count; what differs is the
mark it was handed to, and therefore where the answer goes. A `zone` measures
nothing by length, so its tally has to become a color. A `bar` *is* a length from
a baseline, so its tally stands up.

Nothing was added to the grammar to do that, and the rule that decided it was
written for the flat page. How many axes a transform cuts is read off the mark:
`bar * bin` cuts one, because a bar spends an axis on the domain and keeps the
other to measure along, and `zone * bin` cuts two, because a zone keeps none. A
cube hands the bar a third axis, so it can spend two on the floor and still have
one to measure along — and the same sentence that cuts one axis on the page cuts
two here. Read that way the 3-D histogram is not a feature; it is what the rule
already said, once there were three axes for it to say it about.

Note also what you did *not* have to write. There is no `z()` in that sentence,
for the same reason a flat histogram has no `y()`: the count is invented rather
than bound, so naming it would be naming an output.

## What the floor is made of

A floor is cells, and cells come from the axes — which is the [zone](marks/zone.qmd)
chapter's question arriving one dimension up, with the same four answers. A
continuous axis has to be **cut** into cells before anything can be counted in
them; a categorical axis arrives already cut, because a category *is* a cell. So
which transform you want depends on which kind of axis you have, and a floor can
be one of each:

```{r}
data(gm_eras) + bar * count + x(continent) + y(era) + space() + style(color = "#2a9d8f")
```

That one counts into slots on both axes — a table of counts, standing up. And a
floor with one axis of each kind is a distribution per category, drawn as columns:

```{r}
data(gm_all) + bar * bin(12) + x(life) + y(continent) + space() + style(color = "#2a9d8f")
```

Look at the two floors side by side. The binned axis's columns **touch**, and the
categorical one's leave air between them. That is not a decision about how a 3-D
plot should look; it is the flat rule, per axis. A histogram's bins are adjacent
intervals, so its bars fill them and meet; a category is a slot with its neighbors
somewhere else entirely, and the fifth of the slot left empty is what says so. A
floor cut on one axis and slotted on the other is contiguous one way and gapped the
other, because each axis is answering only for itself.

Ask for a cut where there is nothing to cut and the engine names the other
transform, exactly as it does on the plane:

```{r}
#| error: true
data(gm_eras) + bar * bin + x(continent) + y(era) + space()
```

Both of those transforms *invent* what they measure — nobody hands `count` a
column to reduce, it tallies rows. The other kind of statistic is the kind you
point at a column: a mean, a sum, a median. Those read a floor too, and the
sentence says which column out loud, on the axis the answer will stand along:

```{r}
data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) + space() +
  style(color = "#2a9d8f")
```

Read that beside the flat `bar * mean + x(continent) + y(life)` and the rule is
one sentence in both: **a summary groups by every position the mark does not
measure with, and reduces the column named on the one it does.** Flat, a bar
measures with `y`, so it groups by `x` and averages `y`. In the cube it measures
with `z`, so it groups by the *pair* and averages `z`. Nothing about `mean`
changed; the mark simply has one more position to subtract from.

That is the same subtraction this chapter opened with — the positions the space
offers, less the one the mark measures along — read a second time. The first
reading says how many axes `bin` cuts. This one says how many keys `mean` groups
by. One rule, two questions, and the answer to both is read off the mark.

The five summaries divide from the four measuring transforms on what they are
handed, not on what they can reach: `count` and `bin` invent a measurement, so
they need no column and color or height carries the answer; `mean`, `sum`,
`median`, `max` and `min` reduce a column you name, so the sentence has to name
it. Leave `z()` off and the engine says so rather than guessing.

One statistic really does stay on the plane, and it is the one that *fits*
rather than reduces:

```{r}
#| error: true
data(gm_eras) + bar * smooth + x(continent) + y(era) + z(life) + space()
```

A smoother traces a curve of one column against another, and a curve needs a
domain to run along. The cube's floor is a pair of positions with no left to
right, which is exactly why `line` and `area` cannot stand here either — the
same reason, arriving through a transform instead of through a mark.

## The other two things that stand on a floor

A bar is not the only mark that stands in a slot and measures along the other
axis. A whisker spans, and a box summarizes, and those are the same question
asked of the same pair of axes — which is why all three get the third dimension
together rather than one at a time. Give the pair a cube and each stands on a
*cell* of the floor, measuring along `z`:

```{r}
data(gm_eras) + interval * range + x(continent) + y(era) + z(life) + space() +
  style(color = "#2a9d8f")
```

Every cell's whisker spans that cell's own low and high. The caps are a **cross**
rather than a single tick, and that is not decoration: flat, a cap runs across the
slot in the one direction left over, and a cube leaves two. Picking one of them
would be an arbitrary choice the reader then has to un-read as meaning something,
so the cap says what it says in both.

Ask for the interval that carries a center and the pointrange comes with it:

```{r}
data(gm_eras) + interval * confidence + x(continent) + y(era) + z(life) +
  space() + style(color = "#2a9d8f")
```

And the five-number summary stands the same way, its box a solid between the
quartiles with the whiskers reaching past it:

```{r}
data(gm_eras) + box + x(continent) + y(era) + z(life) + space() + style(color = "#2a9d8f")
```

Look at the median on each box. It is drawn only where the box's own sides face
you, because it is a plane *inside* an opaque solid and the far half of it is
behind the box — the same back-face rule the solid's own six faces follow. A group
with a single observation is the exception, and it goes the other way: its
quartiles coincide, so there is no solid at all, and what you see is the whole
median quad seen face-on. Flat, that same group draws as a line, which is what a
zero-height box has always collapsed to.

## Naming the third axis

The cube's three edges take their names from the columns bound to them, and each
can be overridden. `z_label()` is the third member of a family you have already
met, and it exists for the reason the family does rather than as a convenience:
`x_label()` and `y_label()` shipped without it for several sessions, which made
the positions a family of three where two had a label and one did not. Units are
the usual reason to reach for all three at once:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) +
  x_label("Meters east") + y_label("Meters north") + z_label("Altitude (m)") +
  title("Two gliders, one thermal")
```

Ask for a `z_label()` on a plot with no `z` and the label has no axis to land on,
so GOG says so and draws the plot anyway. It is guidance rather than a refusal:
nothing about the plot is wrong, and one inert label is not worth refusing a
whole figure over.

[Path](marks/path.qmd) owns the mark; the same two gliders appear in the
[third-variable cookbook](cookbook/three-or-more.qmd).

## What the box does not draw yet

The frame is honest about being a still photograph. The box is drawn *behind* the
marks so the cloud always reads on top, which is a cheap stand-in for working out
what hides what, and a point never disappears behind the far face of the box. Its
*labels* go on afterwards, over the marks, because a guide is an annotation about
the scene rather than a member of it: no mark can take away the numbers that say
what it measures.

These were once described here as one job, and they are two. Turning the plot is
built, and it is what makes a 3-D plot legible in the first place. Working out
what hides what, piece by piece, is the harder one and is still to come: it needs
more than turning does, and that is where the grammar stops and the renderer's
real work begins. So the box stays behind the marks however far you turn the
scene, and that is the same honest photograph at every angle rather than
something dragging is about to fix.

Some things stay outside this cut, and the useful part is that they are **not all
the same kind of thing**. A blank in the cube's column can mean three different
sentences, and the engine now says which.

Some are **decided**. `line`, `step`, `area` and `ribbon` read a *domain* left to
right: they sort by `x` and draw one value for each, which is what makes a line a
function rather than a route. A cube has no left to right. `x` is one of three
equal positions, and at some viewing angles it runs straight into the page and
becomes depth, so a line in space would be sorted by an axis the reader cannot
see:

```{r}
#| error: true
data(thermals) + line + x(east) + y(north) + z(altitude)
```

That is a ruling rather than a queue entry, and the direction says so: the route
through three dimensions is [`path`](marks/path.qmd), which is `line` with exactly
that sort removed. Two marks that differ by one rule on the page turn out to
differ by a whole dimension in the cube.

Some are **blocked**, which is a different sentence. A `rule` marks a value on one
axis and spans the ones it does not name — in a cube that is *two* axes, so a rule
here is a **plane**:

```{r}
#| error: true
data(thermals) + rule + x(east) + z(altitude) + space()
```

Marks in space are painted far-to-near and sorted by their footprint, and a
plane's footprint is the whole floor. It could only be drawn wholly in front of or
wholly behind the data, when cutting *through* the data is the entire job of a
reference plane. A `zone` is the same problem one dimension wider. Both need an
engine that can tell, piece by piece, what hides what, which is the next step
rather than a refusal.

And one thing is simply not built: the `z` axis is linear for now, so a log depth
is declined with direction rather than plotted on the wrong scale. In every case
the engine tells you what it will not do, because a plot that silently drops a
dimension is the one failure this project refuses above all others.

## One cube per panel

A facet splits the rows into panels, and a panel is a rectangle. The cube is
drawn inside a rectangle. So a faceted 3-D plot is one cube per panel, each
projected into the room its own panel has, and nothing else needs saying:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + z(population) +
  color(continent) | facet(continent)
```

The scales are shared, exactly as they are on the flat page, so the five cubes
are the same cube five times and a height in one is a height in another. Oceania
holds two countries and almost no ink, which is the honest reading: it is the
comparison that makes it look empty.

The same split, on a route rather than a cloud:

```{r}
data(thermals) + path + x(east) + y(north) + z(altitude) + color(glider) |
  facet(glider)
```

The one thing worth explaining is the guides, because each cube draws its own
tick numbers and its own three axis names, where a faceted flat plot writes
each axis once along an edge of the page. That is a rule rather than an
oversight. **A guide that lives on the panel's boundary is drawn once, because
the panels share that boundary. A guide that lives inside the panel is drawn
in every panel, because there is no boundary to share it on.** The circle
already works this way, a category ring per panel. A cube's three axes are
edges of the cube, and the cube is inside the panel, so a shared column of
numbers beside the page would have nothing to point at: `-200` lands at a
different place in every panel, because every panel projects from its own
rectangle.

The other partition splits the same rows and lays the pieces along the clock
instead of across the page. [`play`](play.qmd) is that chapter, and it is where
the two are crossed.

Which edges carry the numbers is chosen per viewing angle, and it is worth
knowing why. The cube offers four parallel edges per axis. The two domains
take the floor edges nearest you, the pair that makes the V at the bottom of
the picture, and the measurement takes whichever vertical strut sits on the
cube's silhouette. All three are outside the data by construction, so nothing
drawn inside the cube can be in front of them. Look at the left-hand strut of
the tilted-up sheet in [Surface](marks/surface.qmd): you can still read
heights off it while the mesh fills the rest of the frame. The labels are then
measured against each other before any of them is drawn, so two axes cannot
print their numbers in the same place where their edges meet, and an axis name
cannot land on top of its own last tick.

The honest edge that remains is what happens when an axis projects very short. Tilt
the eye up and the measure axis foreshortens, until its numbers have nowhere to sit.
The same summary at two tilts, and the second one is nearly a plan view:

```{r}
#| layout-ncol: 2
data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) + space() +
  style(color = "#2a9d8f") +
  z_label("Mean life expectancy") + title("The three-quarter view")

data(gm_eras) + bar * mean + x(continent) + y(era) + z(life) +
  space(tilt = 85) + style(color = "#2a9d8f") +
  z_label("Mean life expectancy") + title("Tilted up, and thinned")
```

The right-hand frame draws fewer numbers up its measure axis, and that is the whole
of what it does about the squeeze: it thins the *labels* and never the scale. A
coarser step would have been the other way to spend it, and it is the wrong way,
because the step sets where the axis stops, so a bigger one raises the ceiling and
leaves the cube half empty. The numbers that survive are always a subset of the ones
the left-hand frame shows, at exactly the spacing it uses, so a height read off
either panel is the same height.

One more honest edge, and it is about layers rather than marks. The depth sort
runs *within* a layer, and layers paint in the order you wrote them, exactly as
they do on the flat page. So a `point` cloud layered over a `path` sits in front
of the whole route rather than threading through it. That is the ordinary
layering rule holding in space rather than a new exception, but it is worth
knowing before you read a layered 3-D plot as if it were depth-sorted end to end.
