# Setting vs mapping

```{r}
#| include: false
source("R/setup.R")
```

Should this color come from your data, or from you? There are two ways to make
something red, and gog keeps them apart on purpose.

**Mapping** lets the data decide a visual property. `color(continent)` says
*let the continent decide the color*, so gog builds a scale, assigns a color
per continent, and draws a legend to decode it.

**Setting** gives the property a value yourself. `style(color = "tomato")`
says *make them all tomato*. No column is read, no scale is built, and no
legend appears, because there is nothing to decode.

```{r}
# MAP: color answers "which continent?"
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  title("Mapped: a legend decodes the color")
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."*

```{r}
# SET: color answers nothing; note the missing legend
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "tomato") +
  title("Set: no legend, because there is nothing to look up")
```

## Why a constant is not a channel

Every channel in gog maps a data column to a visual feature. A constant maps
nothing, so it fails the definition: the same reason rotating a 3-D plot is a
property of the coordinate space rather than a channel.

A practical test agrees: a mapping draws a legend, a setting draws none. If
the plot needs a legend to be read, it was a mapping.

## What you can set

Every mark is a row and every setting is a column. The grid shows what each
mark takes and what it refuses, including combinations you would not have
guessed were legal. **Legal is not the same as useful** (that is
[Law 8](design-laws.qmd): the grammar guarantees the first, you judge the second);
the table shows the limit of what can be expressed. Its companion grid,
what a mark lets you *map* rather than set, is [Combinations](combinations.qmd).

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules` (its `settings` block), so this grid cannot
# drift from what `style()` actually accepts: the five channel-backed settings
# come from each channel's `settable` flag, the seven style-only ones from the
# engine's own checks. `mark_takes_setting` is the single source behind both the
# grid and the refusal you get for an illegal one.
cli    <- find_gog_cli()
rules  <- jsonlite::fromJSON(paste(system2(cli, "--rules", stdout = TRUE), collapse = "\n"))
marks  <- rules$marks$name[rules$marks$drawable]
setts  <- rules$settings
lookup <- setNames(rules$setting_cells$settable,
                   paste(rules$setting_cells$setting, rules$setting_cells$mark))
header  <- paste0("| Setting | ", paste0("`", marks, "`", collapse = " | "), " |")
divider <- paste0("|:--|", paste(rep(":--:", length(marks)), collapse = "|"), "|")
rows <- vapply(setts, function(s) {
  gs <- vapply(marks, function(m) if (isTRUE(lookup[[paste(s, m)]])) "✅" else "—", character(1))
  paste0("| `", s, "` | ", paste(gs, collapse = " | "), " |")
}, character(1))
cat(header, divider, rows, sep = "\n")
```

(✅ the setting is available · `—` the mark's geometry cannot carry it. Generated live from the
engine, like the [Combinations](combinations.qmd) grid, so it cannot drift from
what `style()` accepts.)

`color` (CSS name or hex) and `opacity` (`0`–`1`) are set on every mark. The
other rows are not arbitrary permissions: each follows from the mark's geometry,
and a setting is available on *every* mark of a kind, never one by accident. That
is the [settable rule](design-laws.qmd): a setting spans its *geometry class*.

`size` is a measurement in pixels: a radius, a stroke width, or a font size.
It belongs to every glyph and stroke, and not to `bar`, `area` or `ribbon`,
whose size their positions already fix. `shape` (one of five glyphs) is
`point`'s alone.

`pattern` is the **texture** of a mark's paint: a general aesthetic, realized by
geometry the way `color` is a fill on a bar and a stroke on a line. On the
five path strokes (`line`, `step`, `path`, `interval`, `rule`) it is the
**dash**: `"solid"`/`"dashed"`/`"dotted"`, a dashed error bar included. On
the five fills (`bar`, `box`, `area`, `ribbon`, `zone`) it is a **fill
texture**: `"solid"`/`"hatch"`/`"crosshatch"`/`"grid"`/`"dots"`, the
hatchings a plot can be read by in grayscale or in print, where hue is gone.
It is paint, never geometry: a dashed line and a solid one trace the same
path. A `step` changes the path itself.

```{r}
data(gapminder_asia) + line * mean + x(year) + y(life) +
  style(pattern = "dashed", size = 1.5) + y_label("Mean life expectancy") +
  title("A dashed trend line")
```

On a fill the same setting hatches the region, a print-ready look that survives
a black-and-white photocopy:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  style(pattern = "crosshatch") +
  y_label("Mean life expectancy") +
  title("A crosshatched bar")
```

The texture is one per layer: `style()` gives every bar in the layer the *same*
hatch, so it cannot give each category of a color split its own. Doing that is a
*mapping* (a category to a texture), which is the
[`pattern()` channel](channels.qmd#pattern), the texture twin of `shape()`.

`border_color` and `border_size` (an outline distinct from the fill, `0` = no
border) belong to the five **closed-glyph fills** (`bar`, `box`, `point`, `zone`
and `surface`) where the border is a shape no composition can trace, so a
setting is the only way to draw it. `color` fills the shape; `border_color` outlines it; the two are
independent, so a box can be pale inside and dark-edged:

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  style(color = "lightsteelblue", border_color = "#26456e", border_size = 1.6) +
  y_label("Life expectancy") + title("A box: light fill, dark outline")
```

The same setting outlines a `point`, so overlapping points stay separate,
exactly as on `bar` and `box`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "steelblue", opacity = 0.7, size = 5, border_color = "white", border_size = 1) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("A point with a white rim")
```

A **curve fill** (`area`, `ribbon`) takes no border: its edge is a *data curve*,
which already is a `line` (`area + line`, `line * bounds`), more capable than
a border setting could be. And a `cross` point has no fill to outline, so a border on a
`cross`-shaped glyph simply has nothing to draw on.

The last five rows are **display toggles**, each tied to one mark's own
geometry. `caps` and `center` show or hide an
[`interval`](marks/interval.qmd)'s end caps and center dot. `nudge` moves a
[`text`](marks/text.qmd) label off its point. `arrow` puts a head on a
[`path`](marks/path.qmd)'s end. And `reach` says how far a
[`rule`](marks/rule.qmd) crosses the axis it does not name. They are single-mark on
purpose (a cap is a thing only a whisker has), which is why their rows are nearly
empty, and that emptiness is information: it says the setting belongs to that mark
and no other.

`arrow` shows this best: its row has one cell, and that one cell is the
reason only `path` takes it. A head marks a direction, and only a path has
one: a `line` sorts its vertices by `x`, so its last point is wherever the domain
ends rather than where the data stopped.

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Each path ends at 2007")
```

It takes `"end"`, `"start"`, or `"both"`: a value rather than a `TRUE`/`FALSE`,
because a double-headed arrow is a common request, and `TRUE`/`FALSE` would
have needed a second setting.

### One color, any mark

`style(color = )` is the same instruction whichever mark it lands on. It fills
points, strokes lines, and fills bars:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "red") +
  title("point")
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(color = "red") +
  title("line")
```

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(color = "red") +
  title("bar")
```

Note what is *not* the way to do this:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + color("red")
```

`color()` names a **column**, so `color("red")` asks for a column called `red`.
Setting a value and mapping a column are different requests, and gog makes you
say which one you mean.

### Several at once

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "#2a6f97", opacity = 0.45, size = 7, shape = "diamond")
```

### Colors are CSS colors

Both forms work anywhere a color is accepted, in `style()` and in `palette()`:

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(color = "darkslateblue") +
  title("A CSS color name")
```

A name gog does not recognize is refused, with the nearest real one offered,
rather than being passed to the renderer and painted black:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(color = "stelblue")
```

R users predictably assume R color names are CSS color names. They are not.
R's `colors()` includes numbered shades (`gray80`, `gray50`,
`steelblue3`) that CSS does not have. gog uses CSS colors because that is
what SVG accepts and what the other three bindings use too, so the
vocabulary is the same in every language:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(color = "gray80")
```

For an exact shade, give the hex value.

## The case setting exists for

Two jobs come up constantly, and neither is a mapping.

**Overplotting.** With enough points, the marks cover one another and the
shape disappears. Lowering opacity
is a statement about the *rendering*, not about the data:

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(opacity = 0.2, size = 3) +
  title("1704 points: transparency reveals the density")
```

**A reference layer.** When one layer carries the message, the others should
be drawn faintly:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "lightgray") +
  line * smooth + style(color = "firebrick", size = 3) +
  title("Five countries in gray; the trend in red")
```

## `line` takes settings it refuses as channels

A polyline is drawn with *one* stroke. There is no per-row anything to vary
along it, so `size` and `opacity` are not channels on `line`:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + line + group(country) + opacity(life)
```

But one stroke still *has* a width and an opacity. Those are settings:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(size = 4, opacity = 0.5) +
  title("Set on a line: one width, one opacity")
```

This is where the distinction matters most. "Can a column be mapped here?" and
"does this mark have this property?" are different questions, and `line` is
where they diverge.

## You cannot map and set the same property

Doing both would mean one of the two instructions is silently discarded, which
gog never does:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + style(color = "tomato")
```

Pick the one you meant.

## `style` applies to one layer only

`color()` also applies to earlier layers that have none. `style()` attaches
to the nearest mark before it and goes no further, because a decoration
written for one layer should not change another:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  line * smooth +
  point + style(color = "tomato", opacity = 0.4) +
  title("Only the points are tomato")
```

To style both layers, say so on both.

## `palette`: choosing the colors a mapping uses

`style(color = )` removes the mapping. `palette()` keeps it and chooses which
colors the categories receive:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  palette(c("tomato", "steelblue", "seagreen", "goldenrod", "orchid"))
```

A palette needs a mapping to give its colors to. With nothing on `color` it
has nothing to color, so it is refused rather than accepted and then ignored:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + palette("okabe")
```

The message names two ways forward, because there are two things a reader may
have meant. To give each category its own color, bind the column with
`color(continent)`. To paint every mark one color, that is a setting rather than
a palette, and `style(color = "steelblue")` is where it belongs.

A mapping need not be written to be there. A `zone * density` measures by color
with nothing bound ([Zone](marks/zone.qmd)), so a palette works on one.

One color, though, is not a palette. `palette()` takes either a **palette
name** (`"gog"`, `"okabe"`) or a **vector of colors**; a single color name
is neither, so it is refused rather than quietly using the default:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  palette("red")
```

To make everything one color you are *setting*, not choosing a palette; use
`style(color = "red")`.

## `theme`: the page rather than the ink

`style()` sets a mark's own appearance. Some settings belong to no mark at all:
the gridlines, the shape of the panel, the angle the tick labels are read at.
Those belong to the whole plot rather than to any mark, and they have their
own word.

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme(grid = "none")
```

`grid` names its sets by the **axis whose ticks they mark**, not by the direction
they run: `"x"` keeps the lines standing up from the x axis and drops the rest.
Naming them that way is what lets the setting survive a bent plane, where the y
axis's gridlines are rings and nothing runs vertically at all:

```{r}
data(winds) + bar * proportion * stack + color(season) + polar() +
  theme(grid = "x")
```

`ratio` sets the panel's width divided by its height, which is how a circle
comes out round. The image keeps the size it was given and the panel shrinks
inside it, so the plot still takes the same room on a page:

```{r}
data(winds) + bar * count + x(direction) + polar() + theme(ratio = 1)
```

And `tick_angle` turns the x labels, which is the answer to category names
that overlap. The bottom margin grows to hold them:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  theme(tick_angle = 45)
```

`font_size` sets how big the plot's text is. It is **one number, not three**.
The number itself is the size of a tick label, in pixels. The axis names and
the title are a fixed proportion above it, so the default of `11` draws ticks
at 11, axis names at 13 and the title at 16. Ask for 16 and those three become 16, 19 and 23:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  title("Bigger type, one number") + theme(font_size = 16)
```

Look at the margins. They grew with the text, because the room an axis needs is
measured from the labels that go in it. Text and the space it sits in are one
decision, so larger text never covers the plot.

A smaller number keeps a plot readable when it has to be printed small:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  title("Smaller type, same plot") + theme(font_size = 8)
```

One number covers all the text because the three sizes were never
independent. If the three sizes were set separately, one plot could carry
three unrelated type sizes. Asking for a big title alone is not possible
here, and that is the design rather than a gap in it.

The number is a **measurement in pixels, not a multiplier**. `font_size = 1.5`
is refused, because pixels are easy to mistake for a multiplier:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme(font_size = 1.5)
```

`font_size` does not name a typeface. The engine measures text with its own table
of character widths, which is what lets it lay out a plot without a font
installed, so there is no font for it to choose. The reader's own system supplies
the letterforms.

A theme can also be asked for by name, and then adjusted. A preset you cannot
adjust would leave you setting every property one at a time, which is the
thing this atom exists to avoid:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  theme("minimal", ratio = 1)
```

## The journal look

`theme("bw")` is the one most often wanted for publication: a white panel inside
a full rectangle.

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme("bw")
```

The name is the misleading part, and it is worth being exact about it. What
goes black and white is the page around the data, never the data: the
continents above keep every color the palette handed them. In a "bw" plot
only the data carries color; everything around it is black, white or gray.

Three settings carry the whole preset. This sentence draws the same picture
as `theme("bw")` above:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme(background = "white", frame = "full", strip = "white")
```

That is the rule every preset in gog follows: a preset only names settings
you could have written yourself. One that went further would be a second
vocabulary inside the first.

## The strip, and why print is the test

The third property is `strip`. It colors the band above each panel that names
the level the panel holds. You only see it on a faceted plot, and it shows
itself in print rather than on screen.

By default the band is a light gray. On a screen that is fine, and it ties
each name to its own panel:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life)) |
  facet(continent)
```

On paper it is not fine. A light gray is printed as a pattern of tiny dots, and
that pattern sits directly under small text. The label gets harder to read, and
a photocopy makes it worse. This is the same reason `pattern` exists for fills:
what survives on a screen and what survives on a page are different questions.

So `theme("bw")` makes the band white as well. The framed panel already separates
one panel from the next, so the tint has nothing left to do:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) + theme("bw")) |
  facet(continent)
```

You can also name the color yourself, for a poster or a slide where a tint helps
rather than hurts:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "#dce8f0", frame = "full")) |
  facet(continent)
```

An animated plot ([Play](play.qmd), later in the book) gets the same band,
because it names its frame the same way. `theme(strip = )` colors that one
too. A property that moved one and left the
other would be exactly the kind of exception [Law 2](design-laws.qmd) forbids.

### The label color follows the band

A dark strip with light type is a common look, and it takes one property:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "#22223b", frame = "full")) |
  facet(continent)
```

You did not name a text color, and the type came out white. gog reads the band
you asked for and picks whichever of its two default inks contrasts more against
it. A pale band keeps the dark ink, so nothing above this point changed.

gog works the ink color out; it does not guess to save you typing. Without
it, `theme(strip = "black")` would paint the near-black default label onto a
near-black band, and every panel would lose its name while the plot still
looked finished. gog will not draw a strip whose name cannot be read, so the
second half of the instruction is supplied rather than required.

When the ink is a real choice rather than a legibility question, name it. A navy
band with gold type is legible, and nothing should argue with it:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "navy", strip_text = "gold", frame = "full")) |
  facet(continent)
```

That is the shape of every derivation in gog: it decides what you did not
say, and stops deciding as soon as you say it. The engine guarantees the plot is
well-formed and leaves taste to you, which is [Law 8](design-laws.qmd).

One band has no color to read. `"transparent"` shows whatever is behind it,
and gog cannot know what that is, so the ink stays dark, which is
right for the page a transparent band is usually sitting on.

`frame` says how the panel is bounded, and it has three answers rather than two.
`"full"` closes the axis lines into a rectangle, `"axes"` is the default pair
along the bottom and left, and `"none"` removes them:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  theme(frame = "none", grid = "y")
```

It is called `frame` and not `border` because `style(border_color = )`
already uses that word, and it means something else: the border of a *mark*,
a bar's outline or a point's border. A frame surrounds the picture; a border is what a shape is drawn
with. One word, one meaning.

`background` takes any color the rest of the grammar takes, which includes the
one a journal usually asks for, since the figure is going onto a page whose
color it does not know:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  theme(background = "transparent", frame = "full")
```

Everything `theme()` accepts is validated, and a name it does not know is refused
rather than quietly ignored:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme("dark")
```

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme(grid = "vertical")
```

The line between the two settings atoms is worth stating once, because it is not
about which is more important. `style()` is the ink: it belongs to a mark, and a
plot with three layers can carry three different ones. `theme()` is the page:
there is one of it, and no single layer has gridlines or a background of its
own.
That the two never overlap is why they are two words rather than one word read
differently depending on where you put it.

Type is where that line is easiest to see, because both atoms can set a size.
`style(size = 17)` on a [`text`](marks/text.qmd) layer sets how big *those labels*
are, and they are labels the data asked for. `theme(font_size = 16)` sets how big
the tick labels, the axis names and the title are, and those belong to the page.
One is a property of a mark and the other is a property of the plot, so neither
can do the other's job.
