# Setting vs mapping

```{r}
#| include: false
source("R/setup.R")
```

Should a mark's color come from your data, or from you? There are two ways to
make something red, and gog keeps them apart on purpose.

**Mapping** lets the data decide a visual property. `color(continent)` says
*let the continent decide the color*, so gog builds a scale, assigns a color
per continent, and draws a legend to decode it.

**Setting** means you give the property a value yourself. `style(color =
"tomato")` says *make them all tomato*. No column is read, no scale is built,
and no legend appears, because there is nothing to decode.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
# MAP: color answers "which continent?"
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  title("Mapped: a legend decodes the color")
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."*

```{r}
# SET: color answers nothing; note the missing legend
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "tomato") +
  title("Set: no legend, because there is nothing to decode")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored tomato."*

## Why a constant is not a channel

Every channel in gog maps a data column to a visual property. A constant maps
nothing, so it fails the definition. Rotating a 3-D plot is excluded for the
same reason: it is a
property of the coordinate space rather than a channel.

A practical test agrees: a mapping to color, size, shape, pattern or opacity
draws a legend, and a setting draws none. If the plot needs a legend to be read,
it was a mapping.

## What you can set

Before you write a setting, you need to know whether the mark can carry it, and
the answer differs between a bar and a point.

Every setting is a row and every mark is a column. The grid shows what each mark
takes and what it refuses, including combinations you would not have guessed
were legal. Legal is not the same as useful. [Law 8](design-laws.qmd) says the
grammar guarantees a plot is legal, and you judge whether it is useful. The
table shows the limit of what can be expressed. Its companion grid, what a mark
lets you *map* rather than set, is [Combinations](combinations.qmd).

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules` (its `settings` block), so this grid cannot
# drift from what `style()` actually accepts: the five channel-backed settings
# come from each channel's `settable` flag, the seven style-only ones from the
# engine's own checks. `mark_takes_setting` is the single source behind both the
# grid and the refusal you get for an illegal one.
cli    <- find_gog_cli()
rules  <- jsonlite::fromJSON(paste(system2(cli, "--rules", stdout = TRUE), collapse = "\n"))
marks  <- rules$marks$name[rules$marks$drawable]
setts  <- rules$settings
lookup <- setNames(rules$setting_cells$settable,
                   paste(rules$setting_cells$setting, rules$setting_cells$mark))
header  <- paste0("| Setting | ", paste0("`", marks, "`", collapse = " | "), " |")
divider <- paste0("|:--|", paste(rep(":--:", length(marks)), collapse = "|"), "|")
rows <- vapply(setts, function(s) {
  gs <- vapply(marks, function(m) if (isTRUE(lookup[[paste(s, m)]])) "✅" else "—", character(1))
  paste0("| `", s, "` | ", paste(gs, collapse = " | "), " |")
}, character(1))
cat(header, divider, rows, sep = "\n")
```

(✅ the setting is available · `—` the mark's geometry cannot carry it. Generated
live from the engine, like the [Combinations](combinations.qmd) grid, so it
cannot drift from what `style()` accepts.)

The twelve rows are every setting there is. `x`, `y`, `z`, `group`, `label` and
`play` are absent because they are channels and nothing else, which [Two
questions, not one](#two-questions-not-one) works through.

`color` (CSS name or hex) and `opacity` (`0`–`1`) are set on every mark. The
other rows are not arbitrary permissions: each follows from the mark's geometry,
and a setting is available on *every* mark of a kind, never on one mark by
accident.
That is the [settable rule](design-laws.qmd): a setting spans its *geometry
class*, the group of marks built the same way.

`size` is a measurement in pixels: a radius, a stroke width, or a font size. It
belongs to every glyph and stroke, and to `box`, whose width it sets. The grid
leaves it blank for `bar`, `area`, `ribbon`, `zone` and `surface`, whose
positions already decide their size. `shape`
(one of seven glyphs) is `point`'s alone.

`pattern` is the **texture** of a mark's paint: one visual property, realized by
geometry the way `color` is a fill on a bar and a stroke on a line. On the six
path strokes (`line`, `step`, `path`, `interval`, `rule`, `edge`) it is the
**dash**: `"solid"`/`"dashed"`/`"dotted"`/`"dotdash"`/`"longdash"`, a dashed
error bar included. On the five fills (`bar`, `box`, `area`, `ribbon`, `zone`)
it is a **fill texture**:
`"solid"`/`"hatch"`/`"crosshatch"`/`"stripes"`/`"grid"`/`"dots"`, the hatchings
that keep a plot readable in grayscale or in print, where there is no color. It is paint, never
geometry: a dashed line and a solid one trace the same path. A `step`, by
contrast, is a different mark because it traces a different path.

A trend drawn through the data is often dashed, so that it reads as a summary
rather than as a measurement:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line * mean + x(year) + y(life) +
  style(pattern = "dashed", size = 1.5) + y_label("Mean life expectancy") +
  title("A dashed trend line")
```

*"Given gapminder Asia: a line derived by mean, x is year, y is life, with
pattern dashed and size 1.5."*

On a fill the same setting hatches the region, a print-ready look that survives
a black-and-white photocopy:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  style(pattern = "crosshatch") +
  y_label("Mean life expectancy") +
  title("A crosshatched bar")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life, with
pattern crosshatch."*

The texture is one per layer: `style()` gives every bar in the layer the *same*
hatch, so it cannot give each category of a color split its own. Doing that is a
*mapping* (a category to a texture), which is the [`pattern()`
channel](channels.qmd#pattern), the texture channel that matches `shape()`.

`border_color` and `border_size` draw an outline separate from the fill, and `0`
means no border. They belong to the five **closed-glyph fills** (`bar`, `box`,
`point`, `zone` and `surface`). On those marks no composition can trace the
border, so a setting is the only way to draw it. `color` fills the shape;
`border_color` outlines it; the two are independent, so a box can be pale inside
and dark-edged:

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  style(color = "lightsteelblue", border_color = "#26456e", border_size = 1.6) +
  y_label("Life expectancy") + title("A box: light fill, dark outline")
```

*"Given gapminder 2007: boxes, x is continent, y is life, colored
lightsteelblue, with border color #26456e and border size 1.6."*

The same two settings outline a `point`, so overlapping points stay separate,
exactly as on `bar` and `box`:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "steelblue", opacity = 0.7, size = 5, border_color = "white", border_size = 1) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("A point with a white border")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored steelblue, with
opacity 0.7, size 5, border color white and border size 1."*

A **curve fill** (`area`, `ribbon`) takes no border. Its edge is a data curve,
so you draw it as a `line` layer (`area + line`, `line * bounds`), which can do
more than a border setting. And a `cross` point has no fill to outline, so a
border on a `cross`-shaped glyph simply has nothing to draw on.

The last five rows turn one part of a mark's drawing on or off, each belonging
to one mark's own
geometry. `caps` and `center` show or hide an [`interval`](marks/interval.qmd)'s
end caps and center dot. `nudge` moves a [`text`](marks/text.qmd) label off its
point. `arrow` puts a head on a [`path`](marks/path.qmd)'s end. And `reach` says
how far a [`rule`](marks/rule.qmd) crosses the axis it does not name. They are
single-mark on purpose, because only a whisker has a cap. That is why their rows
are nearly empty, and an empty row tells you the setting belongs to one mark.

`arrow` shows this best: only `path` is checked in its row, and the reason is
geometry. A head marks a direction, and only a path has one: a `line` sorts its
vertices by `x`, so its last point is the largest `x`, not the point the data
ended on.

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Each path ends at 2007")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
end."*

It takes `"end"`, `"start"`, or `"both"`: a value rather than a `TRUE`/`FALSE`,
because a double-headed arrow is a common request, and `TRUE`/`FALSE` would
have needed a second setting.

### One color, any mark

Having set a point's color once, you should not need a second spelling for a
line or a bar.

`style(color = )` is the same instruction whichever mark it lands on. It fills
points, strokes lines, and fills bars:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "red") +
  title("point")
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(color = "red") +
  title("line")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, colored
red."*

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(color = "red") +
  title("bar")
```

This next sentence looks like the same request, and it is not:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + color("red")
```

`color()` names a **column**, so `color("red")` asks for a column called `red`.
Setting a value and mapping a column are different requests, and gog makes you
say which one you mean.

### Several at once

A figure for publication rarely needs only one setting. Color, opacity, size and shape
are decided together, so they are written together:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "#2a6f97", opacity = 0.45, size = 7, shape = "diamond")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored #2a6f97, with
opacity 0.45, size 7 and shape diamond."*

One `style()` call carries as many settings as the mark can take, and every
row in the layer is drawn with all of them. The four here are independent:
remove any one and the other three are unchanged. Each mark's chapter lists
what you can set on it, under [What you can set](marks/point.qmd#what-you-can-set).

### Colors are CSS colors

A color arrives as a hex value from a style guide, or as a name remembered from
another tool. Not every name you remember is one gog knows.

A CSS name and a hex value both work anywhere a color is accepted, in `style()`
and in `palette()`:

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(color = "darkslateblue") +
  title("A CSS color name")
```

A name gog does not recognize is refused, and the closest valid name is offered,
rather than being passed to the renderer and painted black:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(color = "stelblue")
```

R users often assume R color names are CSS color names. They are not. R's
`colors()` includes numbered shades (`gray80`, `gray50`, `steelblue3`) that CSS
does not have. gog uses CSS colors because that is what SVG accepts and what the
other three bindings use too, so the vocabulary is the same in every language:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(color = "gray80")
```

For an exact shade, give the hex value.

## What settings are for

Two jobs are constant in practice, and neither is a mapping.

**Overplotting.** With enough points, the marks cover one another and the shape
of the cloud disappears. Lowering opacity is a statement about the *rendering*,
not about the data:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(opacity = 0.2, size = 3) +
  title("1704 points: transparency reveals the density")
```

**A reference layer.** When one layer is what the plot is about, the others
should be drawn faintly:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "lightgray") +
  line * smooth + style(color = "firebrick", size = 3) +
  title("Five countries in gray; the trend in red")
```

*"Given gapminder Asia: x is year, y is life, lines, grouped by country, colored
lightgray, and also a line derived by smooth, colored firebrick, with size 3."*

## Two questions, not one

A mapped `opacity` works on a point. Try the same binding on a line and gog
refuses it.

A line is drawn with *one* stroke from its first point to its last. Nothing
varies from row to row along that stroke, so `size` and `opacity` are not
channels on `line`:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + line + group(country) + opacity(life)
```

But one stroke still *has* a width and an opacity. Those are settings:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(size = 4, opacity = 0.5) +
  title("Set on a line: one width, one opacity")
```

This is where the distinction matters most. "Can a column be mapped here?" and
"Does this mark have this property?" are different questions, and `line` is
where the answers differ.

Every mark answers both. Five properties are a channel and a setting under the
same name, and this is what each mark says about them:

```{r}
#| echo: false
#| results: asis
# Generated from the same `gog-cli --rules` dump behind the grid above and the
# one in Combinations, so this table cannot claim a state the engine would not
# enforce. Three answers rather than four: the engine tests that whatever a mark
# can map it can also set, so "map only" cannot occur in these five columns.
#
# `renders` and not `obligation` alone decides the mapping half. Three cells are
# legal grammar the engine does not draw yet (`size` and `opacity` on `text`,
# `pattern` on `edge`), and a table reading them as mappable would promise a
# plot nobody gets. Combinations marks those with a separate glyph; here they
# read as what you can do today, and the sentence under the table says which.
r     <- .gog_rules()
props <- c("color", "size", "shape", "pattern", "opacity")
cells <- r$cells
state <- function(m, p) {
  row  <- cells[cells$mark == m & cells$channel == p, ]
  maps <- nrow(row) == 1 && row$obligation %in% c("must", "can") && !is.na(row$renders)
  sets <- nrow(row) == 1 && row$settable
  if (maps && sets) "map or set" else if (sets) "set only" else "—"
}
cat(paste0("| Mark | ", paste0("`", props, "`", collapse = " | "), " |"),
    paste0("|:--|", paste(rep(":--:", length(props)), collapse = "|"), "|"),
    vapply(r$marks$name[r$marks$drawable], function(m) {
      paste0("| `", m, "` | ",
             paste(vapply(props, function(p) state(m, p), character(1)),
                   collapse = " | "), " |")
    }, character(1)),
    sep = "\n")
```

Read a `set only` cell as a mark with one of something: one stroke, one region,
one band. Bind a column there and the refusal says as much, and names the
setting to write instead.

Three of those cells are temporary rather than geometric. `size` and `opacity`
on a `text`, and `pattern` on an `edge`, are legal grammar the engine has not
built yet, so today they behave as set only and the refusal calls itself
unsupported rather than illegal. [Combinations](combinations.qmd) gives them a
glyph of their own.

Six channels are missing from that table because they are only ever mapped.
`x`, `y`, `z`, `group`, `label` and `play` have no row on the grid above either,
and the reason is the same: a position, a grouping or a name comes from data or
it does not exist. There is nothing to set. Seven settings run the other way and
never appear on the [mapping grid](combinations.qmd): `border_color`,
`border_size`, `caps`, `center`, `nudge`, `arrow` and `reach` decorate a mark
and say nothing about a column.

## You cannot map and set the same property

It is easy to write both by accident: a `color(continent)` left from an earlier
draft, and a `style(color = )` added to try one shade.

Doing both would mean one of the two instructions is silently discarded, which
gog never does:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + style(color = "tomato")
```

Pick the one you meant.

## `style` applies to one layer only

A trend over a cloud of points usually wants the points faint and the line
plain, so the two layers cannot look the same.

`color()` written before any mark applies to every layer that can take it.
`style()` attaches to the nearest mark before it and goes no further, because a
decoration written for one layer should not change another:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  line * smooth +
  point + style(color = "tomato", opacity = 0.4) +
  title("Only the points are tomato")
```

To style both layers, say so on both.

## `palette`: choosing the colors a mapping uses

You may want both at once: the continents separated by color, and those five
colors chosen by you rather than by gog.

`style(color = )` is written instead of a mapping. `palette()` keeps the mapping
and chooses which colors the categories receive:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  palette(c("tomato", "steelblue", "seagreen", "goldenrod", "orchid"))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, with a
palette of tomato, steelblue, seagreen, goldenrod and orchid."*

A palette needs a mapping to give its colors to. With no column bound to `color`
it has nothing to color, so it is refused rather than accepted and then ignored:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + palette("okabe")
```

The message names two fixes, because there are two things you may have meant.
To give each category its own color, bind the column with `color(continent)`. To
paint every mark one color, that is a setting rather than a palette, and
`style(color = "steelblue")` is where it belongs.

A mapping need not be written to be there. A `zone * density` colors its cells
by a measure it computes itself ([Zone](marks/zone.qmd)), so `palette()` works
on that plot.

One color, though, is not a palette. `palette()` takes either a **palette
name** (`"gog"`, `"okabe"`) or the **colors themselves**; a single color name
is neither, so it is refused rather than quietly using the default:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  palette("red")
```

Making everything one color is *setting*, not choosing a palette, so use
`style(color = "red")`.

## `theme`: settings that belong to no mark

The gridlines distract on a slide. The category names overlap along the bottom.
Neither problem belongs to any mark.

`style()` sets a mark's own appearance. Some settings belong to no mark at all:
the gridlines, the shape of the panel, the angle the tick labels are read at.
Those belong to the whole plot rather than to any mark, and they have their
own word.

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme(grid = "none")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

`grid` names its sets by the axis whose ticks they mark, not by the direction
they run: `"x"` keeps the lines standing up from the x axis and removes the
others.
Naming them that way is what lets the setting survive a bent plane, where the y
axis's gridlines are rings and nothing runs vertically at all:

```{r}
#| echo: false
peek(winds)
```

```{r}
data(winds) + bar * proportion * stack + color(season) + polar() +
  theme(grid = "x")
```

*"Given the winds: bars derived by proportion and stack, color by season, in
polar."*

`ratio` sets the panel's width divided by its height, which is how a circle
comes out round. The image keeps the size it was given and the panel shrinks
inside it, so the plot still takes the same room on a page:

```{r}
data(winds) + bar * count + x(direction) + polar() + theme(ratio = 1)
```

*"Given the winds: bars derived by count, x is direction, in polar."*

And `tick_angle` turns the x labels, which is the answer to category names
that overlap. The bottom margin grows to hold them:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  theme(tick_angle = 45)
```

`axis_label` says where an axis's *name* sits. By default each name is centered
along its own axis, which turns the y name through 90 degrees. Ask for `"end"`
and both names move to their axis's far end and stay horizontal, so nothing on
the plot has to be read sideways. It is one property rather than one per axis
because it states a convention, and a plot that followed two of them would have
no rule to explain:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  x_label("GDP per person") + y_label("Life expectancy (years)") +
  theme(axis_label = "end")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, with the
axis names at their ends."*

`font_size` sets how big the plot's text is. It is **one number, not three**.
The number itself is the size of a tick label, in pixels. The axis names and the
title are a fixed proportion larger. The default of `11` draws ticks at 11, axis names
at 13 and the title at 16. Ask for 16 and those three become 16, 19 and 23:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  title("Bigger text, one number") + theme(font_size = 16)
```

Look at the margins. They grew with the text, because the room an axis needs is
measured from the labels that go in it. Text and the space it sits in are one
decision, so larger text never covers the plot.

A smaller number keeps a plot readable when it has to be printed small:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  title("Smaller text, same plot") + theme(font_size = 8)
```

One number covers all the text because the three sizes were never independent.
If the three sizes were set separately, one plot could carry three unrelated
text sizes. Asking for a big title alone is not possible here, and that is the
design rather than a gap in it.

The number is a **measurement in pixels, not a multiplier**. `font_size = 1.5`
is refused, because a number that small is almost always meant as a multiplier:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme(font_size = 1.5)
```

`font_size` does not name a typeface. The engine measures text with its own
table of character widths, which is what lets it lay out a plot without a font
installed, so there is no font for it to choose. The computer that opens the
plot supplies the letters.

A theme can also be asked for by name. A named bundle of these properties is a
**preset**, and you can still adjust it. A preset you could not adjust would
leave you setting every property one at a time:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  theme("minimal", ratio = 1)
```

## The journal look

A journal often asks for a figure on a white background inside a framed panel.

`theme("bw")` is the one most often wanted for publication: a white panel inside
a full rectangle.

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme("bw")
```

The name is the misleading part, and it is worth being exact about it. What becomes
black and white is everything around the data, never the data itself: the
continents
above keep every color the palette gave them. In a "bw" plot only the data
carries color; everything around it is black, white or gray.

Three settings carry the whole preset. This sentence draws the same picture
as `theme("bw")` above:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme(background = "white", frame = "full", strip = "white")
```

That is the rule every preset in gog follows: a preset only names settings
you could have written yourself. A preset that did more than that would be a
second
vocabulary inside the first.

## The strip, and how it prints

The third property is `strip`. It colors the band above each panel, the band
that names the category the panel holds. You only see it on a faceted plot,
and its color matters more in print than on screen.

By default the band is a light gray. On a screen that is fine, and it ties
each name to its own panel:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life)) |
  facet(continent)
```

*"Given gapminder 2007: points, x is gdp with 3 ticks, y is life, split into
panel columns by continent."*

On paper it is not fine. A light gray is printed as a pattern of tiny dots, and
that pattern sits directly under small text. The label gets harder to read, and
a photocopy makes it worse. This is the same reason `pattern` exists for fills:
what survives on a screen and what survives on a page are different questions.

So `theme("bw")` makes the band white as well. The framed panel already
separates one panel from the next, so the tint has nothing left to do:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) + theme("bw")) |
  facet(continent)
```

You can also name the color yourself, for a poster or a slide where a tint makes the name
easier to read rather than harder:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "#dce8f0", frame = "full")) |
  facet(continent)
```

An animated plot ([Play](play.qmd), later in the book) gets the same band,
because it names the value it is showing the same way. `theme(strip = )` colors
that one too. A property that colored one band and not the other would be exactly
the kind of exception [Law 2](design-laws.qmd) forbids.

### The label color follows the band

A dark band is a real choice, and it creates one problem: the default label is
dark
too.

A dark strip with light text is a common look, and it takes one property:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "#22223b", frame = "full")) |
  facet(continent)
```

You did not name a text color, and the text came out white. gog reads the band
you asked for and picks whichever of its two default label colors contrasts more
against it. A pale band keeps the dark label color, so the plots earlier in this
section are unchanged.

gog chooses the label color itself; it does not guess to save you typing. Without
that rule, `theme(strip = "black")` would paint the near-black default label
onto a near-black band. Every panel name would be unreadable while the plot
still looked finished. gog will not draw a panel name that cannot be read, so it
supplies the label color you did not write.

When the label color is a real choice rather than a legibility question, name
it. A navy band with gold text is legible, so gog uses the color you name:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "navy", strip_text = "gold", frame = "full")) |
  facet(continent)
```

Every derivation in gog works this way: it decides what you did not
say, and stops deciding as soon as you say it. The engine guarantees the plot is
well-formed and leaves taste to you, which is [Law 8](design-laws.qmd).

One band gives gog no color to read. `"transparent"` shows whatever is behind
it, and gog cannot know what that is. So the label stays dark, which suits the
white page a transparent band usually sits on.

`frame` says how the panel is bounded, and it takes three values. `"full"`
closes the axis lines into a rectangle, `"axes"` is the default pair along the
bottom and left, and `"none"` removes them:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  theme(frame = "none", grid = "y")
```

It is called `frame` and not `border` because `style(border_color = )`
already uses that word, and it means something else: the border of a *mark*,
a bar's or a point's. A frame surrounds the picture; a border is what a shape is
drawn with. One word, one meaning.

`background` takes any color the rest of the grammar takes, `transparent` among
them. A journal usually asks for that one, because the figure goes onto a page
whose color gog does not know:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  theme(background = "transparent", frame = "full")
```

Everything `theme()` accepts is validated, and a name it does not know is
refused rather than quietly ignored:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme("dark")
```

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme(grid = "vertical")
```

The difference between the two settings atoms is worth stating once, because it
is not
about which is more important. `style()` is the ink: it belongs to a mark, and a
plot with three layers can carry three different ones. `theme()` is the page:
there is one of it, and no single layer has gridlines or a background of its
own. The two never overlap, and that is why they are two words rather than one.

Text size is where that split is easiest to see, because both atoms can set a
size. `style(size = 17)` on a [`text`](marks/text.qmd) layer sets how big *those
labels* are, and they are labels the data asked for. `theme(font_size = 16)`
sets how big the tick labels, the axis names and the title are, and those belong
to the page. One is a property of a mark and the other is a property of the
plot, so neither can do the other's job.

## In short

A mapping earns a guide and a setting earns none. That is the quickest test for
which one you wrote. Five names are both: `color`, `size`, `shape`, `pattern`
and `opacity`. Which call you write decides which you get.

## What comes next

A setting changes how a mark looks and computes nothing from the rows.
[Transforms](transforms.qmd) derives values the table does not hold: counts,
means, bins, densities, and the spreads a span mark needs.
