# Setting vs mapping

```{r}
#| include: false
source("R/setup.R")
```

Should a mark's color come from your data, or from you? There are two ways to
make something red, and gog keeps them apart on purpose.

**Mapping** lets the data decide a visual property. `color(continent)` says
*let the continent decide the color*, so gog builds a scale, assigns a color
per continent, and draws a legend to decode it.

**Setting** means you give the property a value yourself. `style(color =
"tomato")` says *make them all tomato*. No column is read, no scale is built,
and no legend appears, because there is nothing to decode.

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
# MAP: color answers "which continent?"
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  title("Mapped: a legend decodes the color")
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent."*

```{r}
# SET: color answers nothing; note the missing legend
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "tomato") +
  title("Set: no legend, because there is nothing to decode")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored tomato."*

## Why a constant is not a channel

Every channel in gog maps a data column to a visual property. A constant maps
nothing, so it does not meet the definition. The same test settles a question
from later in the book. Rotating a 3-D plot is not a channel either, because it
is a property of the coordinate space.

A practical test gives the same answer: a mapping to color, size, shape, pattern
or opacity draws a legend, and a setting draws none. If the plot needs a legend
to be read, it was a mapping. A mapping written with `legend = FALSE` [leaves
its legend out](channels.qmd#leaving-a-legend-out), so a plot with no legend can
still use a mapping.

## What you can set

You have just seen one setting on a point. Before you write another, you need to
know whether the mark can carry it, and the answer differs between a bar and a
point.

In the grid below, every setting is a row and every mark is a column. The grid
shows what each mark takes and what it refuses, including combinations you would
not have guessed were legal. Legal is not the same as useful. [Law
8](design-laws.qmd#law-8-pronounceable-usable) says the grammar guarantees a
plot is legal, and you judge whether it is useful. The grid shows the limit of
what can be expressed. [Combinations](combinations.qmd) is the companion grid:
what a mark lets you *map* rather than set.

```{r}
#| echo: false
#| results: asis
# Generated from `gog-cli --rules` (its `settings` block), so this grid cannot
# drift from what `style()` actually accepts: the five channel-backed settings
# come from each channel's `settable` flag, the seven style-only ones from the
# engine's own checks. `mark_takes_setting` is the single source behind both the
# grid and the refusal you get for an illegal one.
cli    <- find_gog_cli()
rules  <- jsonlite::fromJSON(paste(system2(cli, "--rules", stdout = TRUE), collapse = "\n"))
marks  <- rules$marks$name[rules$marks$drawable]
setts  <- rules$settings
lookup <- setNames(rules$setting_cells$settable,
                   paste(rules$setting_cells$setting, rules$setting_cells$mark))
header  <- paste0("| Setting | ", paste0("`", marks, "`", collapse = " | "), " |")
divider <- paste0("|:--|", paste(rep(":--:", length(marks)), collapse = "|"), "|")
rows <- vapply(setts, function(s) {
  gs <- vapply(marks, function(m) if (isTRUE(lookup[[paste(s, m)]])) "✅" else "—", character(1))
  paste0("| `", s, "` | ", paste(gs, collapse = " | "), " |")
}, character(1))
cat(header, divider, rows, sep = "\n")
```

(✅ the setting is available · `—` the mark's geometry cannot carry it. Generated
live from the engine, like the [Combinations](combinations.qmd) grid, so it
always matches what `style()` accepts.)

Those rows are every setting there is. `x`, `y`, `z`, `group`, `label` and
`play` are absent because they are channels and nothing else, which [Two
questions, not one](#two-questions-not-one) explains.

`color` (CSS name or hex) and `opacity` (`0`–`1`) can be set on every mark. The
other rows follow from each mark's geometry rather than from a list someone
chose. A setting is available on *every* mark whose geometry can carry it, and
it behaves the same way on all of them. That is the [settable
rule](design-laws.qmd#the-settable-rule): a setting covers its whole *geometry
class*, the group of marks drawn with the same kind of shape.

`size` is a measurement in pixels: a radius, a stroke width, or a font size. It
belongs to `point`, `text`, every stroke mark, and the lines of a `box`. The
grid marks it `—` for `bar`, `area`, `ribbon`, `zone` and `surface`. Their
positions already decide their size, so there is nothing to set. `shape` chooses
one of seven glyphs. Only `point` takes it.

`pattern` is the **texture** of a mark's paint. It is one property drawn
differently by each geometry, the way `color` is a fill on a bar and a stroke on
a line. On the six stroke marks (`line`, `step`, `path`, `interval`, `rule`,
`edge`) it is the **dash**:
`"solid"`/`"dashed"`/`"dotted"`/`"dotdash"`/`"longdash"`, so an error bar can be
dashed too. On the five filled regions (`bar`, `box`, `area`, `ribbon`, `zone`)
it is a **fill texture**: `"solid"`/`"hatch"`/`"crosshatch"`/`"stripes"`/
`"grid"`/`"dots"`. A `surface` takes no hatch: the hatch is drawn flat on the
screen, so on a tilted sheet it would look denser on some faces than on others.
Those hatchings keep a plot readable in grayscale or in black-and-white print.
It changes the paint, never the shape: a dashed line and a solid one pass
through the same points. A `step` is a different mark because it joins those
points by a different route.

A trend drawn through the data is often dashed, so that it reads as a summary
rather than as a measurement:

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line * mean + x(year) + y(life) +
  style(pattern = "dashed", size = 1.5) + y_label("Mean life expectancy") +
  title("A dashed trend line")
```

*"Given gapminder Asia: a line derived by mean, x is year, y is life, with
pattern dashed and size 1.5."*

On a fill the same setting hatches the region, and the hatch stays visible on a
black-and-white photocopy:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  style(pattern = "crosshatch") +
  y_label("Mean life expectancy") +
  title("A crosshatched bar")
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life, with
pattern crosshatch."*

The texture is one per layer: `style()` gives every bar in the layer the *same*
hatch. It cannot give each category its own hatch. That is a *mapping* (a
category to a texture), and the [`pattern()` channel](channels.qmd#pattern) does
it for fills and strokes the way `shape()` does it for points.

`border_color` and `border_size` draw an outline separate from the fill. A
`border_size` of `0` means no border. Both belong to the five **closed-glyph
fills** (`bar`, `box`, `point`, `zone` and `surface`). On those marks you cannot
draw the outline as a second layer. A setting is the only way to get it. `color`
fills the shape; `border_color` outlines it; the two are independent, so a box
can be pale inside and dark-edged:

```{r}
data(gapminder_2007) + box + x(continent) + y(life) +
  style(color = "lightsteelblue", border_color = "#26456e", border_size = 1.6) +
  y_label("Life expectancy") + title("A box: light fill, dark outline")
```

*"Given gapminder 2007: boxes, x is continent, y is life, colored
lightsteelblue, with border color #26456e and border size 1.6."*

The same two settings outline a `point`, exactly as on `bar` and `box`, so each
point stays visible where points overlap:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "steelblue", opacity = 0.7, size = 5, border_color = "white", border_size = 1) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("A point with a white border")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored steelblue, with
opacity 0.7, size 5, border color white and border size 1."*

`opacity` fades what `color` paints, which here is the fill. The white outline
stays fully opaque. To make an outline partly transparent, put the transparency
in `border_color`: in `"#4682b44d"`, the last two hex digits make steelblue 30%
opaque.

On `text`, the same two settings draw a **halo**, a band of color under the
letters. The [Text](marks/text.qmd#a-halo-around-each-label) chapter shows one.

A **curve fill** (`area`, `ribbon`) takes no border. Its edge is a data curve,
so you draw it as a `line` layer (`area + line`, `line * bounds`), and a line
can be dashed, which a border cannot. And a `cross` point has no fill to
outline, so a border on a `cross`-shaped glyph has nothing to draw on, and gog
refuses it. When `shape` is mapped to a column, the categories drawn as crosses
go without a border, and gog prints a message that names them.

The last five rows of the grid above each control one piece of a mark's
drawing. `caps` shows or hides the caps at the ends of a whisker, on an
[`interval`](marks/interval.qmd) and a [`box`](marks/box.qmd). `center` shows or
hides an interval's center dot. `nudge` moves a [`text`](marks/text.qmd) label
off its point. `arrow` draws an arrowhead at one end, or both ends, of a
[`path`](marks/path.qmd) or an [`edge`](marks/edge.qmd). And `reach` says how
far a [`rule`](marks/rule.qmd) reaches across the panel. A row has a ✅ under
every mark that has the part its setting controls. For `caps` that part is a
whisker; for `arrow`, a start and an end that come from the data.

A `line` shows why `arrow` does not belong to it. An arrowhead marks the
direction the data runs in. A `line` sorts its vertices by `x`, so its last
point is simply the largest `x` in the table, not the last row, and an arrowhead
there would mark the largest `x` rather than the end of the data.

```{r}
data(gapminder_asia) + path + x(gdp) + y(life) + color(country) +
  style(arrow = "end") +
  x_label("GDP per person") + y_label("Life expectancy") +
  title("Each path ends at 2007")
```

*"Given gapminder Asia: paths, x is gdp, y is life, color by country, with arrow
end."*

It takes `"end"`, `"start"` or `"both"`, not true or false. Arrowheads at both
ends are often wanted, and a true-or-false value would need a second setting for
them.

### One color, any mark

Having set a point's color once, you should not need a second spelling for a
line or a bar.

`style(color = )` is the same instruction whichever mark it attaches to. It
fills points and bars, and it colors the stroke of a line:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "red") +
  title("point")
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(color = "red") +
  title("line")
```

*"Given gapminder Asia: lines, x is year, y is life, grouped by country, colored
red."*

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(color = "red") +
  title("bar")
```

This next sentence looks like the same request, and it is not:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + color("red")
```

`color()` takes a *column*. `"red"` in quotes is a value, so gog refuses it;
`color(red)` without quotes would map a column called `red`. Setting a value and
mapping a column are different requests, and gog makes you say which one you
mean.

### Several at once

A figure for publication rarely needs only one setting. Color, opacity, size and shape
are decided together, so they are written together:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) +
  style(color = "#2a6f97", opacity = 0.45, size = 7, shape = "diamond")
```

*"Given gapminder 2007: points, x is gdp, y is life, colored #2a6f97, with
opacity 0.45, size 7 and shape diamond."*

One `style()` call carries as many settings as the mark can take, and every row
in the layer is drawn with all of them. The four here are independent: remove
any one and the other three are unchanged. Each mark's chapter lists its
settings in a section called *What you can set*, starting with
[Point](marks/point.qmd#what-you-can-set).

### CSS colors

A color usually comes as a hex value from a style guide, or as a name remembered
from another tool. Not every name you remember is one gog knows.

A name from CSS, the language web pages are styled with, and a hex value both
work anywhere a color is accepted, in `style()` and in `palette()`:

```{r}
data(medals) + bar + x(country) + y(gold) +
  style(color = "darkslateblue") +
  title("A CSS color name")
```

gog refuses a name it does not recognize and offers the closest valid one,
instead of passing it on to be painted black:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(color = "stelblue")
```

R users often assume R color names are CSS color names. They are not. R's
`colors()` includes numbered shades (`gray80`, `gray50`, `steelblue3`) that CSS
does not have. gog uses CSS colors because that is what SVG accepts and what the
other three bindings use too, so the vocabulary is the same in every language:

```{r}
#| error: true
data(medals) + bar + x(country) + y(gold) + style(color = "gray80")
```

For an exact shade, give the hex value.

## What settings are for

The grid above says what you *can* set. Two uses cover most of what you will
actually set.

**Overplotting.** With enough points, the marks cover one another and the shape
of the cloud disappears. Lowering opacity changes how the points are *drawn*,
not the data:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
data(gm_all) + point + x(gdp) + y(life) +
  style(opacity = 0.2, size = 3) +
  title("1704 points: low opacity reveals the density")
```

Where many points overlap, the cloud is darker, so its dense regions show.

**A reference layer.** When one layer is what the plot is about, the others
should be drawn faintly:

```{r}
data(gapminder_asia) + x(year) + y(life) +
  line + group(country) + style(color = "lightgray") +
  line * smooth + style(color = "firebrick", size = 3) +
  title("Five countries in gray; the trend in red")
```

*"Given gapminder Asia: x is year, y is life, lines, grouped by country, colored
lightgray, and also a line derived by smooth, colored firebrick, with size 3."*

## Two questions, not one

A mapped `opacity` works on a point. Try the same binding on a line and gog
refuses it.

A line is drawn with *one* stroke from its first point to its last. The stroke
has one width and one opacity, so `size` and `opacity` are not channels on
`line`:

```{r}
#| error: true
data(gapminder_asia) + x(year) + y(life) + line + group(country) + opacity(life)
```

But one stroke still *has* a width and an opacity. Those are settings:

```{r}
data(gapminder_asia) + line + x(year) + y(life) + group(country) +
  style(size = 4, opacity = 0.5) +
  title("Set on a line: one width, one opacity")
```

This is where the distinction matters most. "Can a column be mapped here?" and
"Does this mark have this property?" are different questions, and `line` is
where the answers differ.

Every mark answers both. Five properties are a channel and a setting under the
same name, and this is what each mark says about them:

```{r}
#| echo: false
#| results: asis
# Generated from the same `gog-cli --rules` dump behind the grid above and the
# one in Combinations, so this table cannot claim a state the engine would not
# enforce. Three answers rather than four: the engine tests that whatever a mark
# can map it can also set, so "map only" cannot occur in these five columns.
#
# `renders` and not `obligation` alone decides the mapping half. Two cells are
# legal grammar the engine does not draw yet (`size` and `opacity` on `text`),
# and a table reading them as mappable would promise a plot nobody gets. Combinations marks those with a separate glyph; here they
# read as what you can do today, and the sentence under the table says which.
r     <- .gog_rules()
props <- c("color", "size", "shape", "pattern", "opacity")
cells <- r$cells
state <- function(m, p) {
  row  <- cells[cells$mark == m & cells$channel == p, ]
  maps <- nrow(row) == 1 && row$obligation %in% c("must", "can") && !is.na(row$renders)
  sets <- nrow(row) == 1 && row$settable
  if (maps && sets) "map or set" else if (sets) "set only" else "—"
}
cat(paste0("| Mark | ", paste0("`", props, "`", collapse = " | "), " |"),
    paste0("|:--|", paste(rep(":--:", length(props)), collapse = "|"), "|"),
    vapply(r$marks$name[r$marks$drawable], function(m) {
      paste0("| `", m, "` | ",
             paste(vapply(props, function(p) state(m, p), character(1)),
                   collapse = " | "), " |")
    }, character(1)),
    sep = "\n")
```

A `set only` cell means the mark has the property, but one value for the whole
layer, not one per row: a line has one width, and a rule layer one opacity. Bind
a column there and the refusal explains that, and names the setting to write
instead.

Two `set only` cells are exceptions. The grammar allows `size` and `opacity` on
a `text`, but the engine does not draw them. gog refuses the mapping as
**Unsupported** rather than **Illegal**, and the message names no setting. Set
either one with `style()` instead. [Combinations](combinations.qmd) marks these
two cells with a symbol of their own, ◌.

The table above has a column for five of the kernel's eleven channels. The other
six, `x`, `y`, `z`, `group`, `label` and `play`, are only ever mapped, and they
have no row on the settings grid either. Positions, groups, labels and the steps
of an animation come from data, or they do not exist at all. There is nothing to
set. Seven settings are only ever set, and never appear on the [mapping
grid](combinations.qmd). `border_color`, `border_size`, `caps`, `center`,
`nudge`, `arrow` and `reach` decorate a mark and read no column.

## Mapping and setting the same property

It is easy to write both by accident: a `color(continent)` left from an earlier
draft, and a `style(color = )` added to try one shade.

Doing both would mean one of the two instructions is silently discarded, which
gog never does:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) +
  color(continent) + style(color = "tomato")
```

Write the one you meant.

## `style` and the layer it follows

A trend line is easiest to see when the points under it are faint and the line
is not, so the two layers need different settings.

`color()` written before any mark applies to every layer that can take it.
`style()` attaches to the nearest mark before it, and to no other mark, because
a setting written for one layer should not change another:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  line * smooth +
  point + style(color = "tomato", opacity = 0.4) +
  title("Only the points are tomato")
```

To style both layers, write a `style()` after each mark.

## `palette`: choosing the colors a mapping uses

Setting a color and mapping a column are the two choices this chapter has kept
apart. Sometimes you want part of each: the continents separated by color, and
those five colors chosen by you rather than by gog.

`style(color = )` is written instead of a mapping. `palette()` keeps the mapping
and chooses which colors the categories receive:

```{r}
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  palette(c("tomato", "steelblue", "seagreen", "goldenrod", "orchid"))
```

*"Given gapminder 2007: points, x is gdp, y is life, color by continent, with a
palette of tomato, steelblue, seagreen, goldenrod and orchid."*

A palette needs a mapping to give its colors to. With no column bound to `color`
it has nothing to color, so it is refused rather than accepted and then ignored:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + palette("okabe")
```

The message names two fixes, because there are two things you may have meant. To
give each category its own color, bind the column with `color(continent)`. To
paint every mark one color, write `style(color = "steelblue")`, which is a
setting rather than a palette.

A plot can have a color mapping that you did not write. A `zone * density`
colors its cells by a measure it computes itself ([Zone](marks/zone.qmd)), so
`palette()` works on that plot.

One color, though, is not a palette. `palette()` takes either a **palette name**
(`"gog"`, `"okabe"`) or the **colors themselves**, and it reads a single word as
a palette name. `"red"` is not a palette name, so gog refuses it rather than
quietly using the default palette:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + color(continent) +
  palette("red")
```

Making everything one color is *setting*, not choosing a palette, so use
`style(color = "red")`.

## `theme`: settings that belong to no mark

The gridlines distract on a slide. You want the category names along the
bottom at 45 degrees.

`style()` sets a mark's own appearance. The gridlines, the shape of the panel
and the angle of the tick labels belong to no mark, so they have a word of their
own: `theme()`.

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme(grid = "none")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent."*

The gridlines are gone, and nothing about the points changed, because `theme()`
never changes a mark.

`grid` names each set of gridlines by the axis whose ticks they mark, not by the
direction they run. So `"x"` keeps the lines belonging to the x axis and removes
the rest. That naming is the same when the plane is bent into a circle. The next
plot keeps the x axis gridlines in [polar](polar.qmd), a space later in the
book. They radiate from the center instead of running from bottom to top:

```{r}
#| echo: false
peek(winds)
```

```{r}
data(winds) + bar * count + x(direction) + polar() + theme(grid = "x")
```

*"Given the winds: bars derived by count, x is direction, in polar."*

`ratio` sets the panel's width divided by its height. The image keeps the size
it was given and the panel shrinks inside it, so the plot still takes the same
room on a page:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) + theme(ratio = 1)
```

`tick_angle` turns the x tick labels to the angle you give, in degrees. You do
not need it to keep names apart. When category names do not fit side by side,
gog turns them 90 degrees by itself, to read upward.
[Scales](scales.qmd#crowding-in-the-plane-and-in-the-cube) shows what gog does
when even turned names do not fit. `tick_angle` is for choosing the angle
yourself. gog keeps the angle you choose, even when fewer names fit at it, and
says so.
The bottom margin grows to hold the turned labels, and the axis name sits below
them:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  theme(tick_angle = 45)
```

`axis_label` says where an axis's *name* sits. By default each name is centered
along its own axis, so the y name is turned through 90 degrees. Ask for `"end"`
and both names move to their axis's far end and stay horizontal, so nothing on
the plot has to be read sideways. It is one property, not one per axis, so the
two names never follow different rules:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  x_label("GDP per person") + y_label("Life expectancy (years)") +
  theme(axis_label = "end")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, with the
axis names at their ends."*

`font_size` sets how big the plot's text is. It is *one* number, not three. The
number itself is the size of a tick label, in pixels. The axis names and the
title are a fixed proportion larger. The default of `11` draws tick labels at
11, axis names at 13 and the title at 16. Ask for 16 and those three become 16,
19 and 23:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  title("Bigger text, one number") + theme(font_size = 16)
```

Look at the margins. They grew with the text, because the room an axis needs is
measured from the labels that go in it. Larger text therefore never overlaps the
panel.

A smaller number leaves more room for the data in a small figure:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  title("Smaller text, same plot") + theme(font_size = 8)
```

One number sets all the text, so one plot never carries three unrelated text
sizes. You cannot ask for a bigger title on its own: the title grows only when
the tick labels grow.

The number is a measurement in *pixels*, not a multiplier. `font_size = 1.5`
is refused, because a number that small is almost always meant as a multiplier:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme(font_size = 1.5)
```

`font_size` does not name a typeface. The engine measures text with its own
table of character widths, so it can lay out a plot with no font installed.
It asks for the reader's system font, and the computer that opens the plot
supplies the letters.

A **preset** is a named bundle of these properties, and you ask for one by its
name. You can still change any property in it, so one small change never means
writing every property yourself:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  theme("minimal", ratio = 1)
```

gog has three presets: `"gog"`, the default, `"minimal"` and `"bw"`. `"minimal"`
removes the gridlines, the same as `theme(grid = "none")`, and `ratio = 1` then
makes the panel square.

## A figure for a journal

A journal often asks for a figure on a white background inside a framed panel.

`theme("bw")` is the one most often wanted for publication: a white panel inside
a full rectangle.

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme("bw")
```

`"bw"` stands for black and white, and the name misleads. What becomes black and
white is everything around the data, never the data itself. The continents in
that plot keep every color the palette gave them. In a "bw" plot only the data
carries color, and everything around it is black, white or gray.

The whole preset is three settings, written out below. This sentence draws the
same picture `theme("bw")` just drew:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  color(continent) + theme(background = "white", frame = "full", strip = "white")
```

Here `background` makes the panel white and `frame` draws a rectangle around it.
`strip` is explained in the next section, and the other two at the end of this
chapter.

Every preset in gog follows the same rule: a preset only names settings you
could have written yourself. A preset that did more than that would be a second
set of words to learn.

## The strip, and how it prints

The third setting inside `theme("bw")` is `strip`. It colors the band above each
panel, the band that names the category the panel holds. You only see it on a
faceted plot, and its color matters more in print than on screen.

By default the band is a light gray. On a screen that is fine, and it shows
which panel each name belongs to:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life)) |
  facet(continent)
```

*"Given gapminder 2007: points, x is gdp with 3 ticks, y is life, split into
panel columns by continent."*

On paper it is not fine. A light gray is printed as many tiny black dots, and
those dots sit directly under small text. The label gets harder to read, and a
photocopy makes it worse. This is the same reason `pattern` exists for fills:
what is easy to see on a screen can be hard to see on paper.

So `theme("bw")` makes the band white as well. The framed panel already
separates one panel from the next, so the tint is no longer needed:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) + theme("bw")) |
  facet(continent)
```

You can also name the color yourself, for a poster or a slide where a tint makes the name
easier to read rather than harder:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "#dce8f0", frame = "full")) |
  facet(continent)
```

An animated plot has the same band ([Play](play.qmd), later in the book). It
names the moment the animation shows, such as the year, the way a panel band
names its category. `theme(strip = )` colors that one too. A property that
colored one band and not the other would be exactly the kind of exception [Law
2](design-laws.qmd#law-2-no-exceptions) forbids.

### The label color on a dark band

Some plots use a dark band on purpose, and that creates one problem: the default
label is dark too.

A dark band with light text is often wanted, and you need only one property for
it:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "#22223b", frame = "full")) |
  facet(continent)
```

You did not name a text color, and the text is drawn white. gog reads the band
color you asked for. It has two default label colors, and it picks the one that
contrasts more with that band. On a pale band the label stays dark, which is why
the faceted plots earlier in this chapter have dark labels.

gog chooses the label color because an unreadable name is a defect, not to save
you a line of code. Without that rule, `theme(strip = "black")` would paint the
near-black default label onto a black band. Every panel name would be unreadable
while the rest of the plot looked correct. So when you do not name a label
color, gog chooses a readable one.

When you want a particular label color, and not only a readable one, name it.
gog then uses the color you name as written, without checking it against the
band:

```{r}
(data(gapminder_2007) + point + x(gdp, tick_count = 3) + y(life) +
  theme(strip = "navy", strip_text = "gold", frame = "full")) |
  facet(continent)
```

Every default in gog works this way: it decides what you did not say, and stops
deciding as soon as you say it. The engine guarantees the plot is well-formed
and leaves taste to you, which is [Law
8](design-laws.qmd#law-8-pronounceable-usable).

A `"transparent"` band gives gog no color to read, because it shows whatever is
behind it. So the label stays dark, which suits the white page a band usually
sits on.

`frame` says how the panel is bounded, and it takes three values. `"full"`
closes the axis lines into a rectangle, `"axes"` is the default pair along the
bottom and left, and `"none"` removes them:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  theme(frame = "none", grid = "y")
```

It is called `frame` and not `border` because `style(border_color = )` already
uses that word for something else. A border belongs to a *mark*, such as a bar
or a point. A frame bounds the panel. One word, one meaning.

`background` colors the panel, and it takes any color the rest of the grammar
takes, a CSS name such as `linen` among them:

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) +
  theme(background = "linen", frame = "full")
```

`"transparent"` is the one value that changes more than the panel. If only the
panel were transparent, the margin around it would still be painted. So
`"transparent"` also leaves the margin and the legend's box unpainted, and the
page under the figure shows through. A journal often asks for this, because it
prints the figure on a page whose color gog does not know.

Everything `theme()` accepts is validated, and a name it does not know is
refused rather than quietly ignored:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme("dark")
```

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + theme(grid = "vertical")
```

`style()` and `theme()` differ in what they belong to. `style()` belongs to a
mark, and a plot with three layers can carry three different ones. `theme()`
belongs to the whole plot, and there is only one of it: no single layer has
gridlines or a background of its own. The two never overlap, and that is why
they are two words rather than one.

Text size is where that split is easiest to see, because both atoms can set a
size. `style(size = 17)` on a [`text`](marks/text.qmd) layer sets how big *those
labels* are, and they are labels that come from a column.
`theme(font_size = 16)` sets how big the tick labels, the axis names and the
title are, and those belong to the whole plot. One is a property of a mark and
the other is a property of the plot, so neither can replace the other.

## In short

A mapping to color, size, shape, pattern or opacity earns a legend, and a
setting earns none. That is the quickest test for which one you wrote; a
mapping with `legend = FALSE` also has none. Five
names are a channel and a setting at once: `color`, `size`, `shape`, `pattern`
and `opacity`. Which call you write decides which you get.

## What comes next

A setting changes how a mark looks and computes nothing from the rows.
[Transforms](transforms.qmd) shows how gog derives values the table does not
hold: counts, means, bins, densities, and the low and high values an `interval`
needs.
