# Transforms

```{r}
#| include: false
source("R/setup.R")
```

How is a measurement distributed across your rows? A transform is the
"add-a-stroke" derivation in the grammar. It modifies how data reaches the mark,
preprocessing it statistically before any pixel is drawn.

The family has other names. ggplot2 says **stat**, and the chapter Wilkinson
devotes to it says **statistics** [@wilkinson2005]; SQL says **aggregate**, dplyr says
**summarize**, a spreadsheet says **pivot**. Each of those covers about half of
what GOG groups here. `mean`, `count` and `density` really are statistics, and
they replace rows with a summary. But `bin` cuts an axis into cells, `bounds`
reshapes two columns you already have and computes nothing at all, and `stack`,
`dodge` and `jitter` are about collisions between marks rather than about the
values behind them. Calling all of that a statistic would misname most of it, so
the word here is the widest one that stays true: a transform is anything that
changes what reaches the mark.

Transforms are applied with the `*` operator:

```r
mark * transform
```

The result is a derived mark type, not a new atom to memorize.

| Transform | What it does |
|---|---|
| `bin` | cuts a continuous axis into equal-width cells, and counts the rows in each |
| `smooth` | fits a LOESS curve through the `(x, y)` pairs |
| `count` | counts the rows at each distinct `x` |
| `density` | estimates the distribution of `x` with a Gaussian kernel |
| `proportion` | divides the measurement by its total, so the shares sum to 1 |
| `sum` | totals `y` within each `x` |
| `mean` | averages `y` within each `x` |
| `median` | the middle `y` within each `x` |
| `max` | the largest `y` within each `x` |
| `min` | the smallest `y` within each `x` |
| `range` | the lowest and highest `y` per group, as a pair |
| `confidence` | the mean's confidence interval per group: low, high and center |

The middle five (`sum`, `mean`, `median`, `max`, `min`) form the **aggregation
family**. All share the same pattern: group by x, reduce y to one value. `range`
groups the same way but reduces to *two* values, a low and a high, the extents
an [`interval`](marks/interval.qmd) spans.

## One variable, or two?

Twelve of the transforms fall into two families by how many columns they
*read*, and that number is exactly what decides whether you write `y()`. The
rest of the kernel's seventeen come later in this chapter: `bounds`, which
reshapes rather than computes, `partition`, which cuts a panel into regions,
and the three collision modifiers.

**Inventing** transforms (`bin`, `count`, `density`) read a
*single* variable and make the measured axis from it. `proportion` joins them
whenever it stands alone, having then no measurement to rescale but the tally it
makes itself. You name only the
variable whose shape you are asking about, and the transform invents the y column
and labels it ("Count", "Proportion", "Density"). `bar * bin + x(life)` and
`line * density + x(life)` are the same grammatical shape: one name in, a new
axis out. Leaving `y()` off is not a shortcut the engine tolerates:
there is exactly one sensible y here (the count, the density), so the omission
cannot be misread and the engine fills it in silently.

**Reading** transforms (`smooth`, `sum`, `mean`, `median`, `max`, `min`,
`range`, `confidence`) read *two* columns and reshape the second against the
first. They fit or aggregate an existing `y`, so a `y()` must be in scope:
`bar * mean + x(country) + y(gold)` has no way to guess which column to average.
Leave `y()` off and the engine invents nothing: only you know which column was
meant, so it stops and names what is missing (`add y(<column>)`). `range` and
`confidence` are the two that read a `y` like the others but *write more than
one*: `range` a low and a high, `confidence` a low, high and center, the
extents an [`interval`](marks/interval.qmd) spans.

So `y()`'s presence is not a quirk to memorize per transform; it reads straight
off the family. This is [Explicit Over Implicit](design-laws.qmd) in miniature:
the short form is allowed exactly when it cannot be misread, and refused the
moment it could.

A statistic runs **within each group** when `color` (or `group`) is bound.
`bar * bin + color(species)` is one histogram *per species*, see [overlaid
histograms](marks/bar.qmd#overlaid-histograms), not
one combined bar with the split thrown away; and the same rule gives a density
curve, a count, or a mean per group. Each output keeps its group, so the mark
colors it and the legend agrees.

---

## `bin`: histogram

`bin` groups `x` values into equal-width buckets and counts the observations
in each bucket.
Combine it with `bar` to draw a histogram:

```{r}
data(gapminder_2007) + bar * bin + x(life) +
  x_label("Life expectancy (years)") +
  title("Histogram of Life Expectancy, 2007")
```

*"Given gapminder 2007: bars derived by bin, x is life."*

**Bin count** defaults to Sturges' rule [@sturges1926]: `k = ⌈log₂(n)⌉ + 1`. Notice there is no
`y()` in the expression: `bin` invents the count column internally, and the
y-axis is labeled "Count" automatically.

### Choosing the bin count

Sturges' rule is a sensible default, not a mandate. Pass a number to set the bin
count, or name a `width` to fix the bin width in the data's own units:

```{r}
data(gapminder_2007) + bar * bin(30) + x(life) +
  x_label("Life expectancy (years)") + title("Thirty bins: a finer histogram")
```

```{r}
data(gapminder_2007) + bar * bin(width = 5) + x(life) +
  x_label("Life expectancy (years)") + title("Bins five years wide")
```

The positional number is the *count*: `bin(30)` is thirty bins. A width is the
less common request, so it is said out loud, `bin(width = 5)`; the two are
mutually exclusive. Writing `bin(30, width = 5)` refuses with direction rather
than guessing which you meant: the same reason `bin(30)` alone can't mean a
width, since on this data five bins and five-wide bins are opposite plots.

### Why `bar * bin`, not a `histogram()` shortcut?

Because `bin` does the same computation regardless of the mark, and the *mark*
chooses how the bins are drawn. `bar * bin` fills bars. `line * bin` connects the
bin centers into a [frequency polygon](marks/line.qmd), and
[`step * bin`](marks/step.qmd) traces the histogram as a staircase outline. Add
`stack` and a fourth reading appears, the dot plot further down this chapter: one
dot per observation rather than one mark per bin. Adding a `histogram()` shortcut
would violate Law 2 (No Exceptions) by special-casing a particular mark+transform
pair.

### The same `bin`, cutting two axes

The mark decides something else too: **how many axes get cut**. Each mark above
leaves one axis free to measure the count along, so `bin` cuts the other one. A
[`zone`](marks/zone.qmd) measures nothing by length and needs an extent on both
axes before it is a rectangle at all, so the same word cuts both, and the count
goes to `color` because that is the only channel left holding a number:

```{r}
data(gm_all) + zone * bin(24) + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("One bin, two axes: the heatmap")
```

That is a heatmap, with no atom added to the kernel.

### One of each: the mixed mesh

"Cuts both" is a shade too strong. What `bin` cuts is every axis that has a
width to cut, and a categorical axis has none, because it arrived already cut:
a category *is* a cell. Give a zone one of each and the transform does the only
coherent thing, which is to cut the continuous axis and leave the other alone:

```{r}
data(gm_all) + zone * bin(20) + x(life) + y(continent) +
  palette("viridis") + title("One axis cut, one already cut")
```

A row of cells per continent, every row cut on the same edges, so a column
compares straight down. It is a distribution per category shown as shade rather
than as height, which is why five of them fit where one histogram would go. The
[Zone chapter](marks/zone.qmd) reads all three cases as one rule.

### Which mesh: `tiling`

`bin` also decides the *shape* of the cells it cuts, because a different mesh
puts different rows in different cells and so changes the counts. Rectangles are
the default; hexagons stagger alternate rows, which stops the eye reading the
mesh's own alignment as structure in the data [@carr1987]:

```{r}
data(gm_all) + zone * bin(20, tiling = "hex") + x(gdp, scale = "log") + y(life) +
  x_label("GDP per capita") + y_label("Life expectancy") +
  title("Carr's staggered mesh: tiling = hex")
```

A tiling means nothing to a bin that cuts one axis, since there the cells are
intervals and an interval has no shape, so `bar * bin(tiling = )` is refused
pointing at `zone`. The [Zone chapter](marks/zone.qmd) works through both meshes,
including why the empty cells are left as panel.

---

## `smooth`: LOESS trend line

`smooth` fits a locally weighted linear regression (LOESS) through the
`(x, y)` points and evaluates it at 100 evenly spaced x positions
[@cleveland1979].

Combine it with `line` to overlay a trend line on a scatter plot:

```{r}
data(gapminder_2007) + x(gdp) + y(life) +
  point + color(continent) +
  line * smooth +
  title("GDP vs Life Expectancy with LOESS Smooth")
```

**Span** defaults to 0.75: 75 % of the data is used for each local regression.
A larger span gives a smoother curve; a smaller span follows the data more
closely. `smooth` requires both `x` and `y` to be numeric columns, and it
preserves the field names, so no extra `y()` binding is needed.

---

## `count`: frequency aggregation

`count` counts the number of rows for each unique value of `x`.
Works on both string and numeric columns.

```{r}
# How many countries are in each continent?
data(gapminder_2007) + bar * count + x(continent) +
  title("Countries per Continent, 2007")
```

This is the categorical counterpart to `bin`:
`bin` groups a continuous variable; `count` groups a discrete one. The bars show
it: a histogram's touch (adjacent slices of one continuum), a bar chart's stand
apart (distinct categories).

`count`, `bin`, and `density` are the three **inventing** transforms, and
`proportion` behaves as one when it stands alone, so `y()` is optional for all
four and the y-axis labels itself automatically ("Count", "Proportion",
"Density"). Bind `y(my_name)` only when you want a custom axis label.

### The same `count`, tallying two axes

`bin` cuts one axis on a bar and two on a zone, and `count` divides the same way for
the same reason. A bar has a length to spend the tally on, so it groups by one axis;
a [`zone`](marks/zone.qmd) measures nothing by length, so it tallies rows into the
cell where *two* categories cross and the count goes to `color`:

```{r}
data(winds) + zone * count + x(direction) + y(season) +
  title("One count, two axes: the tile plot")
```

That is the tile plot (a confusion matrix is the same sentence with different
columns), and like the heatmap, no atom was added to the kernel.

The pair of them states the whole rule: `bin` cuts, `count` tallies. A
continuous axis has to be cut into cells before there is anything to count in; a
categorical axis arrives already cut, because a category *is* a cell. So each is
refused where the other belongs, and each refusal names the other.

Read one axis at a time and that is three plots rather than two. Cut both axes and
it is the heatmap; cut neither and it is the tile plot; cut one of two and it is
the mixed mesh above. A `bin` is refused only where there is nothing left for it
to cut at all.

---

## `proportion`: relative frequency

`proportion` divides a measurement by its total, producing shares in [0, 1] that
sum to 1. Left to itself it has nothing to divide yet, so it tallies the rows first
exactly as `count` does:

```{r}
# What share of countries belongs to each continent?
data(gapminder_2007) + bar * proportion + x(continent) +
  title("Continent share, 2007")
```

Use `proportion` when you care about relative composition rather than raw
counts, e.g. "what fraction of respondents chose each option?" Because
`count` and `proportion` differ only by a constant factor (total n), `order`
works identically for both: sorting by the y column produces the same order.

It follows `count` into two dimensions as well, so `zone * proportion` reads a tile
plot as shares (see [Zone](marks/zone.qmd)).

### The total is the whole plot, always

The denominator is every row the plot draws, in every context. That is the one thing the
word is allowed to mean. Split the bars by a second column and they still sum to 1
between them, so each bar reads as *this* pair's share of everything:

```{r}
data(winds) + bar * proportion + x(direction) + color(season) +
  title("Every bar is a share of all 264 observations")
```

Add up the sixteen segments there and you get 1, not 2. What you cannot get that way
is the *conditional* reading, "of the summer winds, what share came from the west",
and that is the facet, which cuts the data into panels before any of this runs, so
each panel normalizes inside itself:

```{r}
data(winds) + bar * proportion + x(direction) | facet(season)
```

Read the two plots against each other and the difference is the whole distinction: in
the first, summer's bars and winter's bars are competing for one total; in the second
each season gets its own.

### Normalizing something other than a tally

Because `proportion` divides whatever measurement it finds, it composes with any
transform that leaves one number per cell. Put it after a `bin` and the histogram's
own counts come back as fractions, the **relative-frequency histogram**, on exactly
the mesh a plain `bar * bin` cuts:

```{r}
data(gm_all) + bar * bin(20) * proportion + x(life) +
  title("Relative frequency, not counts")
```

Put it after a summary and the column you named is the thing rescaled, so each slot
reads as its share of the grand total:

```{r}
data(gapminder_2007) + bar * sum * proportion + x(continent) + y(population) +
  title("Each continent's share of world population, 2007")
```

What it refuses is what leaves it nothing to divide by. A `density` already
integrates to 1, and its cells are sample points rather than parts of a whole. A
`smooth` gives a value *at* each x, so adding those values up answers nothing. A
pair transform leaves two numbers per cell, and a span has no total it is a part
of.

```{r}
#| error: true
data(gm_all) + line * density * proportion + x(life)
```

---

## `density`: kernel density estimate

`density` estimates the probability density function of `x` using a Gaussian
kernel, evaluated at 256 evenly spaced points. By default the bandwidth (how
wide each kernel is, and so how smooth the curve) is chosen automatically by
Silverman's rule-of-thumb [@silverman1986], IQR-adjusted for robustness.

Combine it with `line` to draw a smooth density curve:

```{r}
data(gapminder_2007) + line * density + x(life) +
  x_label("Life expectancy (years)") +
  title("Kernel Density: Life Expectancy 2007")
```

As with the other inventing transforms, there is no `y()` to write: the
density column and its axis label are both made for you.

### Choosing the bandwidth

The automatic bandwidth is a sensible default, not a mandate, and it is
`density`'s one knob, just as the bin count is `bin`'s. It is set the same two
ways. A positional number is an *adjust* multiplier on the automatic bandwidth:
`density(2)` is twice as smooth, `density(0.5)` half as smooth and so more
jagged.

```{r}
data(gapminder_2007) + line * density(2) + x(life) +
  x_label("Life expectancy (years)") + title("density(2): twice as smooth")
```

To fix the bandwidth in the data's own units instead, name it. Here
`bandwidth = 1` sets each kernel one year wide, narrower than the default, so
the curve keeps more of the distribution's fine structure:

```{r}
data(gapminder_2007) + line * density(bandwidth = 1) + x(life) +
  x_label("Life expectancy (years)") + title("bandwidth = 1 year: an absolute width")
```

This mirrors `bin` exactly. The positional number is the *adjust* multiplier:
the common request, since you rarely know the bandwidth you want up front and
only need the automatic one a little smoother or rougher. An absolute width is
the less common request, so it is said out loud, `density(bandwidth = 1)`; the
two are mutually exclusive, and `density(2, bandwidth = 1)` refuses with
direction rather than guessing which you meant.

### One transform, four readings

`density` estimates a distribution; what it is *drawn as* depends on how much
room the bindings leave it, and the answer is different in each case rather than
being four transforms wearing one name.

Bind one number and it is a **curve** along the free axis, which is everything
above. Bind two numbers to a mark with no measure axis and it is a **field** over
the plane: [`path * density`](marks/path.qmd) traces its contours and
[`zone * density`](marks/zone.qmd) paints it cell by cell, with
[`surface * density`](marks/surface.qmd) raising it into the cube. Bind a
**category and a number** and there is no free axis left, so the estimate is
drawn across the category's slot, one distribution per group, which is
[the violin](marks/ribbon.qmd#the-violin-a-distribution-per-category):

```{r}
data(gapminder_2007) + ribbon * density + x(continent) + y(life) +
  y_label("Life expectancy") + title("ribbon * density: the slot reading")
```

Each of `density`'s knobs belongs to one of those readings, and asking for one in
a reading it cannot mean is refused rather than ignored. `bandwidth` is a length
in one column's units, so it means nothing to a field, whose two axes measure
different quantities. `levels` counts iso-lines, and a curve has none to count.
And the slot reading has two of its own: `compare` says what a violin's width
*means* from one slot to the next, and `reach` says how far it goes, in slots, so
that past half a slot the shapes run into their neighbors and the plot is a
ridgeline. Neither means anything to a plot with no slots:

```{r}
#| error: true
data(gapminder_2007) + line * density(compare = "count") + x(life)
```

---

## What `bin`, `density` and `smooth` refuse

The two families above sort the transforms by how many columns they read. A
second property cuts across that sort: these three all describe how values are
spread *along* an axis, so the axis they read has to carry a number. A category
is one slot rather than a stretch of line, so there is nothing to cut into
intervals, nothing for a curve to spread over, and nothing to fit through.

The violin above looks like an exception, and it is worth being precise
about why it is not one. There the category is not the axis being estimated
along: `y(life)` is, and the category only says *which rows* each estimate is
made from. Take the number away and there is nothing left to spread over, so the
refusal returns, which is why every chunk below binds a category and nothing
else.

Each refusal names the atom that asks the same question of categories. Binning
one is really counting it:

```{r}
#| error: true
data(gapminder_2007) + bar * bin + x(continent)
```

A density over categories is really a share of the total:

```{r}
#| error: true
data(gapminder_2007) + bar * density + x(continent)
```

And a curve fitted across categories is really a summary within each one:

```{r}
#| error: true
data(gapminder_2007) + point * smooth + x(continent) + y(life)
```

Which axis must be numeric is read off the bindings rather than fixed to `x`.
The [horizontal histogram](marks/bar.qmd) is `bar * bin + y(life)`: it cuts `y`,
because `y` is the axis it was handed and `x` is the one it invents. `smooth` is
the strict one of the three, since it fits `y` against `x` and so needs a number
on both.

---

## Chaining transforms

Two transforms on one mark have to divide the work, and there are only four jobs to
divide. A transform can say **where the cells are**, it can say **what is in them**,
it can say **what scale the answer is read on**, and it can say **where the marks
sit**. `bounds` names the sides of a rectangle. `count` tallies into the cells your
categories already own, and says nothing about where those cells are. `mean` and its
four siblings reduce a column you name, and say nothing about where the cells are
either. `proportion` does the third job and only the third: it divides by a total,
which is why it can stand after almost anything. `dodge`, `stack` and `jitter` do the
fourth, moving marks that would otherwise land on top of one another.

Two transforms compose when they do different jobs, and contradict when they do the
same one. A cell holds one extent, one measurement, one scale and one arrangement.
A second transform doing a job the first already did therefore leaves no reading that
keeps both. The engine refuses, rather than drawing one and quietly throwing the other
away. [Combinations](combinations.qmd) has the whole grid, generated from the same
rule.

Because each job is filled once, the order you write them in does not change the
picture. Write them in the order they happen and the sentence reads the way it
runs.

`bin` is the one that does both, and it is the only one that can give half of it up.
Cutting an axis is what makes it a `bin`; the tally it hands back is a by-product of
the cut. So when you compose it with a statistic, `bin` keeps the cut and the
statistic keeps the measurement:

```{r}
data(gm_all) + bar * bin(20) * mean + x(gdp, scale = "log") + y(life) +
  title("Mean life expectancy by income band")
```

Read the sentence as three separate claims. `bin(20)` cuts income into twenty bands.
`mean` reduces `life` within each one. `bar` draws the result as a length from zero.
The bars touch because the bands tile the axis, exactly as a histogram's do, and the
y axis is now labeled for the column that was actually reduced.

Any mark that takes both transforms takes the pair, so the same three claims read as
points instead:

```{r}
data(gm_all) + point * bin(20) * median + x(gdp, scale = "log") + y(life) +
  title("Median life expectancy by income band")
```

### What cannot be chained, and why

The other two inventing transforms cannot give their measurement up, because
measuring is all they do. Take it away from `count` and nothing is left of it:

```{r}
#| error: true
data(gm_all) + bar * count * mean + x(continent) + y(life)
```

`density` fails a different test. Its cells are the points where the estimate was
sampled, and the estimate exists *between* your observations rather than in buckets
holding them, so there are no rows inside one for a statistic to reduce:

```{r}
#| error: true
data(gm_all) + bar * density * mean + x(gdp) + y(life)
```

`smooth` is refused against all three, including `bin`, for a reason none of them
share. It fits a curve through the rows and already averages locally as it goes, so
cutting them into cells first changes nothing it was not doing:

```{r}
#| error: true
data(gm_all) + bar * bin * smooth + x(gdp) + y(life)
```

And two transforms that each invent their own measurement are the same contradiction
with neither able to give way, so a cell would be measured twice:

```{r}
#| error: true
data(gm_all) + bar * bin * count + x(life)
```

Each refusal names both ways out, and the first two also name the way *through*:
`bin` is the transform that cuts without keeping the measurement.

The same rule reaches the other three jobs, and it reads the same way each time. Two
transforms that both produce a low and a high measure the cell twice. So `range` and
`confidence` cannot stand together. A minimum and a maximum are not a 95% confidence
interval, and one of the two would have to be discarded:

```{r}
#| error: true
data(gm_all) + interval * range * confidence + x(continent) + y(life)
```

The same holds when one of the pair is a single value and the other is two of them:

```{r}
#| error: true
data(gm_all) + line * sum * range + x(year) + y(population)
```

And two arrangements are the same contradiction on the fourth job. A bar cannot be
set beside its neighbor and piled on top of it at once, because it sits in one place:

```{r}
#| error: true
data(gm_all) + bar * count * dodge * stack + x(continent) + color(continent)
```

A transform can also turn out to be unnecessary rather than contradictory, and that
is a different answer. `proportion` divides a tally by its total, so it already
counts the rows, which leaves `count` beside it with nothing to add. The plot is
still exactly the plot you asked for, so the engine draws it:

```{r}
data(gm_all) + bar * count * proportion + x(continent) +
  title("Share of countries by continent")
```

It also prints a warning on your console. This page cannot show you that warning,
because only refusals appear in a rendered book. It reads: *"`bar * count *
proportion` draws the same plot as `bar * proportion` — a share is a share of a
tally, so `proportion` already counts the rows and `count` adds nothing."*

That is the line between the two answers. When keeping both transforms would mean
throwing one away, the engine refuses. When both can stay and one of them simply
changes nothing, it draws and says so. Neither answer is silence, which is the rule
underneath both: a word you wrote is never read and then ignored.

---

## Output columns

| Transform | Reads from | Writes to |
|-----------|-----------|-----------|
| `bin` | `x(field)` | `x(field)` (bin centers), `y(your_name)` (counts) |
| `count` | `x(field)` | `x(field)` (unique values), `y(your_name)` (counts) |
| `proportion` | whatever measured, or `x(field)` if nothing did | the same column, divided by its total (sums to 1) |
| `smooth` | `x(field)`, `y(field)` | same `x`, same `y` (smoothed values) |
| `density` | `x(field)` | `x(field)` (eval points), `y(your_name)` (density) |
| `sum` / `mean` / `median` / `max` / `min` | `x(field)`, `y(field)` | `x(field)` (unique), `y(field)` (aggregated) |
| `range` | `x(field)`, `y(field)` | `x(field)` (each group twice), `y(field)` (low, then high) |
| `confidence` | `x(field)`, `y(field)` | `x(field)` (each group twice), `y(field)` (low, then high), `center` (the mean) |

For `bin`, `count`, `proportion`, and `density`, the output y column name is
whatever you bind with `y()` in the spec, or an empty string when `y()` is
omitted (the renderer finds it either way).  The y-axis label defaults to the
transform name ("Count", "Proportion", "Density") when `y()` is absent.

One row of that table changes when `bin` is composed with a statistic. Cutting is
then all it does: it writes the bin centers to `x(field)` as before and leaves
`y(field)` alone for the statistic to reduce, so the y column holds your column's
summary rather than a count.

---

## Aggregation family: `sum`, `mean`, `median`, `max`, `min`

When a dataset has multiple rows for the same x value (e.g., medals from
different years, all labeled "USA"), `bar` alone draws one bar per row:
the last row silently wins.

The aggregation family groups rows by x and reduces y to a single value
per group. All five follow exactly the same grammar:

```
bar * sum     + x(country) + y(gold)
bar * mean    + x(country) + y(gold)
bar * median  + x(country) + y(gold)
bar * max     + x(country) + y(gold)
bar * min     + x(country) + y(gold)
```

**`sum`**, total y per group, the most common aggregation for counts and amounts:

```{r}
medals_dup <- data.frame(
  country = c("USA", "USA", "GBR", "GBR", "JPN"),
  gold    = c(10, 5, 8, 4, 6)
)
data(medals_dup) + bar * sum + x(country) + y(gold) +
  order(gold, desc = TRUE) +
  y_label("Total gold medals") + title("Sum per country: bar * sum")
```

**`mean`**, average y per group, useful for comparing typical values:

```{r}
data(medals_dup) + bar * mean + x(country) + y(gold) +
  order(gold, desc = TRUE) +
  y_label("Mean gold medals") + title("Mean per country: bar * mean")
```

**`median`**, the middle value, robust to outliers:

```{r}
data(medals_dup) + bar * median + x(country) + y(gold) +
  order(gold, desc = TRUE) +
  y_label("Median gold") + title("Median per country: bar * median")
```

**`max`**, the largest value within each group:

```{r}
data(medals_dup) + bar * max + x(country) + y(gold) +
  order(gold, desc = TRUE) +
  y_label("Max gold") + title("Max per country: bar * max")
```

**`min`**, the smallest value within each group:

```{r}
data(medals_dup) + bar * min + x(country) + y(gold) +
  order(gold, desc = TRUE) +
  y_label("Min gold") + title("Min per country: bar * min")
```

All five aggregations work with both string and numeric x columns. For string
x, groups preserve first-appearance order (before any `order`); for numeric x,
groups are sorted ascending.

| Transform | What it computes per group |
|-----------|---------------------------|
| `sum`     | total of y values          |
| `mean`    | arithmetic average of y    |
| `median`  | middle value of y          |
| `max`     | largest y value            |
| `min`     | smallest y value           |

### The same five, grouping by a pair

`count` follows a bar onto a zone and tallies two axes instead of one, and these
five follow it, for the same reason and by the same rule. A bar has a length to
spend, so it groups by one position and measures along the other; a
[zone](marks/zone.qmd) measures nothing by length, so both its positions are free
to be keys and the answer goes to `color`:

```{r}
data(winds) + zone * mean + x(direction) + y(season) + color(speed) +
  title("The same mean, grouped by a pair")
```

What tells the transform which column to reduce is the same channel the answer
comes back on. That is not a special case for the zone: a summary reduces **in
place** everywhere, and on `bar * mean + x(country) + y(gold)` it is `y` that both
names `gold` and receives the average. A zone measures by color, so color does
both jobs there.

So the rule is one sentence, and it is the dimensional rule read a second time:
a summary groups by every position the mark does not measure with, and reduces
the column named on the one it does.

| mark | positions | measures with | groups by | reduces |
|---|---|---|---|---|
| `bar` | `x`, `y` | `y` | `x` | `y` |
| `bar` in [`space()`](space.qmd) | `x`, `y`, `z` | `z` | `x`, `y` | `z` |
| [`surface`](marks/surface.qmd) | `x`, `y`, `z` | `z` | `x`, `y` | `z` |
| `zone` | `x`, `y` | `color` | `x`, `y` | `color` |

The middle two rows are identical, and that is the rule working rather than a
coincidence: a 3-D bar and a surface have the same three positions and measure
along the same one, so they group and reduce alike. What differs is only what each
lays on the cell the pair names, a column or a plateau.

The division from `count`, `bin` and `density` is not about how many
axes anything can reach: all of them read a mesh. It is about what each was handed.
Those three invent their measurement, so they need no column named and publish
their answer under a name of their own; these five are given a column, so the
sentence has to say which one, and the answer replaces it. `proportion` sits outside
the division because it is handed neither: it takes whatever measurement the sentence
already made and divides it.

### Why verbs?

`count`, `sum`, `mean`, `median`, `max`, `min` are all **verbs**: they describe
*what to do* to y, not a data type or a chart name. The `*` operator makes
this explicit: `bar * sum` reads as "a bar that sums". Every aggregation is
a plain English word; none are abbreviations (Law 3: Plain Names).

---

## `range`: a low/high pair

Every transform above writes a single measured value per group. `range` is the
first to write *two*: the minimum and the maximum of `y` within each group,
emitted as a low then a high. That pair is exactly what an
[`interval`](marks/interval.qmd) spans:

```{r}
data(iris_flowers) + interval * range + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("Petal length range per species")
```

`range` reads a `y` and groups by `x` exactly like the aggregation family: it
needs a `y()` in scope and works on string or numeric x alike. What differs is
the *output*: two rows per group instead of one. This low/high-rows shape is
reused elsewhere: the [`box`](marks/box.qmd) mark encodes its own five-number
summary the same way (min and max as the pair of rows, the quartiles alongside),
and the [`ribbon`](marks/ribbon.qmd) mark fills between the same pair as a
continuous band. See [Interval](marks/interval.qmd) for the whisker, and for why
its extents are invented here rather than bound to two extra channels the
kernel would never otherwise have.

---

## `confidence`: the interval of the mean

`range` shows the *full* spread; often you want the **uncertainty of the mean**
instead: an error bar. `confidence` computes the mean's confidence interval per
group (mean ± t·se, the t-interval), emitting a low, a high, *and* a center.
That center is what turns the whisker into a pointrange:

```{r}
data(iris_flowers) + interval * confidence(0.95) + x(species) + y(petal_length) +
  x_label("Species") + y_label("Petal length (cm)") +
  title("95% CI of mean petal length")
```

The level is a parameter, exactly like `bin(30)` and `density(2)`: `confidence`
alone is 95%, `confidence(0.99)` is wider. It uses the **t**-distribution (not a
fixed 1.96), so a small group's interval is honestly wider, and a group of one,
no spread, collapses to a point.

The center dot appears because the statistic *has* a center; `range` has none, so
it draws a bare whisker. That is the rule for the whole family: the transform
decides whether there is a center, and the mark draws what it is given. So the
difference between an error bar and a pointrange is which statistic you ask for,
not another mark to remember.

---

## `bounds`: a *pre-computed* low/high pair

`range` and `confidence` **compute** the pair from raw values. `bounds` is their
counterpart that computes nothing. You already have the low and the high in two
columns, worked out upstream: a model's standard error, a psychometric SEM, a
bootstrap interval. `bounds(lower, upper)` reshapes those two columns into the
very same low/high pair, so a [`ribbon`](marks/ribbon.qmd) or
[`interval`](marks/interval.qmd) draws them with no new machinery:

```{r}
data(score_band) + ribbon * bounds(lower, upper) + x(score) +
  y_label("Expected true score") + x_label("Raw score") +
  title("A pre-computed band: ribbon with bounds")
```

It differs from every transform above in two ways that both follow from *reshaping,
not computing*. Its arguments are **column names**, not a number: it has to be told
which columns hold the bounds. And it needs **no `y()`**: those two columns *are* the
extents, so (like `count`, which invents its own y) the mark takes its low and
high from `bounds` and the y-axis fits them. This is GOG's line held exactly: the
statistics it *does* compute (`bin`, `smooth`, `density`, `range`, `confidence`) are
a deliberate few; everything else you compute yourself and GOG draws, and `bounds`
is the door for a band you already have.

It is legal on the five marks that can draw a low/high pair, each a different
geometry over the *same* statistic, the [No Exceptions](design-laws.qmd) rule that
gives one `bin` three histograms:

- [`ribbon`](marks/ribbon.qmd) **fills** the pair (a filled band),
- [`interval`](marks/interval.qmd) draws a **whisker** across it at each x,
- [`line`](marks/line.qmd) traces its **two boundaries** (the *unfilled* band),
- [`step`](marks/step.qmd) traces them as **staircases** (stepped control limits), and
- [`zone`](marks/zone.qmd) **shades the region between them**, the band that names
  its own sides rather than tracking a curve.

So `line * bounds(lower, upper) + style(pattern = "dashed")` is the unfilled, dashed
error band (the two ± curves with no fill between them) and it is the same pair,
drawn by a different mark, that `ribbon * bounds` fills. On any mark that draws
*neither* a pair nor a single value, `bounds` is refused with direction.

---

## `partition`: a whole among nested parts

Every transform so far reads a column and writes a summary of it. `partition`
reads a **hierarchy** (two or three columns spelling the path down to each row)
and writes the region each node of that hierarchy occupies: four edges and a
center, which is exactly what [`bin`](#bin-histogram) publishes for a cell. That
is what makes it a transform rather than a coordinate space, and it is why
[`zone`](marks/zone.qmd) needs no new machinery to draw the result.

The levels can run two ways, and the difference is one parameter. **Nested** is
the default: every level divides the same axis, and the other axis steps out one
ring per level. Flat that is the icicle, and in [`polar()`](polar.qmd) it is the
sunburst, which is where that reading is worked through in full.

**Crossed** is the other, and it is the one this chapter draws, because it stays
flat. The levels alternate: the first divides the width, the second divides the
height *within* each of those columns, a third would divide the width again. The
plot it makes is the **mosaic**.

### The mosaic

`commutes` is a survey of how people get to work in four cities, one row per
city-and-mode with a count:

```{r}
knitr::kable(head(commutes, 6))
```

```{r}
data(commutes) + zone * partition(city, mode, cross = TRUE) + x(people) +
  color(mode) + style(border_color = "white", border_size = 2) +
  x_label("People surveyed") + title("How four cities get to work")
```

Two variables, and the plot reads them in two directions at once. A column's
**width** is that city's share of everyone surveyed, so Ashford's column is three
times Millbrook's; a cell's **height** is that mode's share *within its own city*,
so every column fills the panel whatever it weighs. The consequence is the one a
mosaic is drawn for: a cell's **area** is the number of people in it, and the eye
compares areas across the whole panel without being told to.

That is also what makes Millbrook readable rather than misleading. Its cycling
share is the largest of the four and its column is the narrowest, so the plot
reports the finding and the weight of evidence behind it in the same gesture,
which four equal-width stacked bars cannot do at all.

The two axes therefore carry different quantities, which is worth looking at
before reading anything off them. The horizontal runs `0` to the total in the
data's own units, 3,790 people, exactly as the icicle's measure axis does, so
composing with [`proportion`](#proportion-relative-frequency) turns it into
shares. The vertical runs `0` to `1` and can do nothing else, because what a
height means is a share of its own column and the grand total is not the thing
being divided.

Naming the columns is the [`text`](marks/text.qmd) layer the sunburst already
uses, and for the same reason: a shallower partition of the same table lands its
nodes in exactly the same places, so a second reader needs no filtering and no
arithmetic of its own.

```{r}
data(commutes) + x(people) +
  zone * partition(city, mode, cross = TRUE) + color(mode) +
  style(border_color = "white", border_size = 2) +
  text * partition(city, cross = TRUE) + label(name) +
  x_label("People surveyed") + title("The same plot, with its columns named")
```

The labels sit at the middle of each column because that is where the center of a
column is, and `style(nudge = )` is what moves them if you want them along the
top. That is an honest limit rather than a choice: a mosaic's column names belong
on an axis, and GOG's category axis gives every slot the same width, which is the
one thing a mosaic's columns are not. Until that axis learns to take its positions
from the data, the names arrive as a layer.

The cell edges are a `style()` setting like any other, and they matter more here
than anywhere else in the book. Without them, two neighboring cells that share a
color merge into one shape and the column boundary disappears:

```{r}
data(commutes) + zone * partition(city, mode, cross = TRUE) + x(people) +
  color(mode) + x_label("People surveyed") +
  title("The same plot with no edges")
```

Both plots draw the same numbers. The second is the reason a
[`zone`](marks/zone.qmd) carries `style(border_color = , border_size = )` at all:
the mark began as a highlight behind a line, where a frame would compete with the
data, and it now also draws the cells that *are* the data.

One level is legal and means something: the **spine plot**, columns of varying
width and nothing dividing them, which is the marginal distribution on its own.

```{r}
data(commutes) + zone * partition(city, cross = TRUE) + x(people) +
  color(city) + style(border_color = "white", border_size = 2) +
  x_label("People surveyed") + title("The margin alone: a spine plot")
```

Everything `partition` refuses nested it refuses crossed, because the refusals
are about the **tree** and not about the layout. A level must be a name rather
than a number. A branch may stop early, but it may not have a hole in the
middle. And an interior node carrying a value of its own is the one genuine
ambiguity, so it is refused rather than guessed. The [sunburst
chapter](polar.qmd) meets each of them where they are easiest to see.

---

## `dodge`: side by side instead of on top

Every transform above is a *statistic*: it reads columns and writes new ones.
`dodge` is a different kind of atom: a **collision modifier**. It writes no new
values; it *moves* the marks that a color split would otherwise stack at one
position, setting them side by side instead.

When `color` splits a `bar`, a `box` or an `interval` into groups, the groups
share their category's slot and pile up. For a histogram that reads well as
translucent overlays, but for a grouped bar chart you want them separated.
`dodge` does exactly that:

```{r}
data(gm_eras) + bar * mean * dodge + x(continent) + y(life) + color(era) +
  y_label("Mean life expectancy") +
  title("Mean life expectancy by continent, 1957 vs 2007")
```

The two eras sit next to each other in every continent, each bar half as wide,
together filling exactly the slot one bar would; the gap between continents is
untouched. There is no width to set: `dodge` derives it from the number of groups
(`G` groups, each `1/G` of the slot). And because the offset resolves the overlap
outright, dodged bars draw **solid**: the translucent fill exists only to see
through a pile-up, and there is none left.

It rides the `*` slot like any transform, so it composes cleanly:
`bar * mean * dodge` aggregates first and dodges the result; `box * dodge` dodges
a box's own summary; `interval * range * dodge` dodges grouped whiskers. Here it
sets each continent's two distributions side by side:

```{r}
data(gm_eras) + box * dodge + x(continent) + y(life) + color(era) +
  y_label("Life expectancy") +
  title("Life expectancy distribution, 1957 vs 2007")
```

Which marks `dodge` serves is decided by what a mark *is*. It subdivides a
*width* (a bar's thickness, a box's body, a whisker's slot) so it is defined
for `bar`, `box` and `interval`. A `point` has no width to subdivide (its
overlap answer is `jitter`, below); a `line` or `area` is offset by
*accumulating* rather than subdividing (that is `stack`, next). Ask for
`dodge` on one of those and GOG refuses, naming the offset that fits. That is
the discipline the rest of the kernel follows: an atom combines with the marks
it was defined for, and the refusal points at the right tool rather than
drawing something wrong.

`dodge` also needs a split to separate: a `color` (or `group`) binding. Without
one there is a single mark per slot and nothing to set beside anything, so
`bar * dodge` on its own is refused toward adding the color, never accepted as a
silent no-op.

## `stack`: piled up, not side by side

`stack` is `dodge`'s sibling: the other way to resolve a color split's overlap.
Where `dodge` sets the groups *beside* each other, `stack` piles them *on top* of
each other along the measure axis, each group starting where the one below it
ended. It is the offset for the marks that **sum**: `bar` and `area`.

Take the dodge example and change one atom:

```{r}
data(gm_eras) + bar * sum * stack + x(continent) + y(population) + color(era) +
  y_label("Population") +
  title("Population by continent, 1957 and 2007 stacked")
```

Notice the measure also changed, from `mean life` to `sum population`. You stack
quantities that *add up*: two populations sum to a total that means something, so
the full bar reads as "everyone, both eras". Two mean life expectancies do not add
(a bar reaching 130 years is nonsense), so that comparison wants `dodge`, side by
side. The grammar writes either; the data decides which is honest. Because the pile
resolves the overlap, the bars draw **solid**, and the axis grows to the tallest
*stack*: Asia clears five billion, the sum of its two eras.

### Stacked areas: the case `stack` was built for

A split `area` is where overlap hurts most: filled regions bury one another and no
opacity really fixes it (the [area chapter](marks/area.qmd) shows the warning GOG
raises). `stack` is the answer. Each region sits on the cumulative height of the
ones below, so they abut into one band instead of hiding each other, the classic
"parts of a whole, over time":

```{r}
data(gm_all) + area * sum * stack + x(year) + y(population) + color(continent) +
  y_label("Population") +
  title("World population by continent, stacked")
```

Asia fills the floor; Europe, Africa, the Americas and Oceania pile on top, and the
band's upper edge is the world total climbing from two billion to six and a half.
Read any one color's *thickness* for that continent's share.

### `stack(share = TRUE)`: every pile filled to one

Reading a thickness is the hard part of the plot above, because each band's baseline
moves under it. When the composition is the question and the total is not,
`stack(share = TRUE)` divides every pile by its own slot's total, so all of them
reach 1 and only the proportions are left:

```{r}
data(gm_all) + area * sum * stack(share = TRUE) + x(year) + y(population) +
  color(continent) + title("Every year's composition, filled to one")
```

Now Asia's band is a constant-ish two thirds and Europe's visibly narrows across the
century, both of which the plot above hides inside a rising total. The axis says
`Share`, because whatever the numbers were before, they are fractions of one now.

This is not `proportion` wearing a second face. The two divide by different
totals. `proportion` divides by the **whole plot's**, so its bars still say how
big each slot is. This one divides by the **slot's own**, which deliberately
throws that away. And only this one composes with a measurement it was not
given: `sum` of a population column is exactly the case `proportion` has no way
to say. Filling is a *position* adjustment:
it changes where the marks sit and what the scale reads, never what was counted,
which is why it belongs to `stack` and not to a statistic.

A `bar` fills the same way, and that reading is the chart the name comes from:
the **100% stacked bar**, worked through in the [bar
chapter](marks/bar.qmd#the-100-stacked-bar). One parameter, both geometries,
exactly as plain `stack` hands both marks the same span and lets each draw it:
[No Exceptions](design-laws.qmd).

Which marks `stack` serves is decided by what a mark *is*: the same rule
`dodge` follows. What `stack` computes is one thing for every mark, a span from a
foot to a top along the measure axis, and what differs is how a geometry draws that
span. A `bar` fills it, an `area` fills it across x, and a `point`, having no length
to stretch across it, spends it on *how many dots there are* (the dot plot, just
below). A `box` or `interval` is set beside its neighbors rather than piled (that
is `dodge`); an unfilled `line` has nothing to fill and stack (use `area`); a
`ribbon` already spans a low to a high, so it measures no height from a baseline.
Ask for `stack` on one of those and GOG refuses, naming the offset that fits. For
`bar` and `area` it also needs a `color` (or `group`) split to pile: `bar * stack`
alone is refused toward adding the color, never accepted as a silent no-op.

### `stack(baseline = )`: where the pile hangs

A pile has two free choices. Its heights are one, and `share` is how you change
them. Where the pile *stands* is the other, and `baseline` is how you change that.

The default is `"zero"`. Every pile stands on the axis, which is both plots above.
That layout has a known weakness. Only the bottom band has a shape you can read.
Every band above it rests on the sum of the bands below, so it inherits all of
their movement as well as its own.

The weakness is easiest to see when the series take turns. Here are four music
genres over half a year, each popular at a different time:

```{r}
data(listening) + area * stack + x(week) + y(plays) + color(genre) +
  y_label("Plays") + title("Stacked on zero")
```

Techno is the red band. It looks like it grows to week 19 and then collapses. Part
of that shape is Techno, and part of it is Folk and Jazz shrinking underneath it.
You cannot tell which part is which.

`"center"` hangs each pile so that its middle is at zero:

```{r}
data(listening) + area * stack(baseline = "center") + x(week) + y(plays) +
  color(genre) + title("Centered: the ThemeRiver")
```

That symmetric layout is the **ThemeRiver** [@havre2000], named for the plot it
was built to draw: how much each theme is written about, over time.

`"wiggle"` instead picks the foot that makes the bands as flat as it can. Thick
bands are held steadiest, because a reader looks at them first. Thin bands absorb
the movement instead:

```{r}
data(listening) + area * stack(baseline = "wiggle") + x(week) + y(plays) +
  color(genre) + title("Flattest: the streamgraph")
```

Now Techno is a clean lens shape that swells and fades, which is what the numbers
actually say. Folk, the blue band, is a steady decline rather than a wedge. No band
changed thickness between these three plots. Only the place each pile hangs changed.

The last plot is a streamgraph [@byron2008]. The middle one is a ThemeRiver. The two are often
confused and they are not the same layout: one is symmetric, the other is chosen to
reduce movement. Readers compare values more accurately on a streamgraph than on
either of the other two, and that result is the reason this option exists.

Notice also that these two plots have no numbers on the measure axis. Once a pile
has moved off zero, no value on that axis means anything. A band drawn from 12 to
20 says "this group is 8 here", and the 12 and the 20 only record where the pile
was hung. So GOG draws no ticks and no axis name, for the same reason it leaves an
empty heatmap cell unpainted. The *scale* is unchanged, so thicknesses stay
comparable everywhere in the plot, and thickness is what a streamgraph asks you to
read.

`baseline` is not an area-only parameter. A `bar` has a pile too, so it hangs the
same way:

```{r}
data(listening) + bar * stack(baseline = "center") + x(week) + y(plays) +
  color(genre) + title("The same baseline, drawn as bars")
```

One parameter, both geometries. `share` follows the same rule, and so does plain
`stack`, which hands both marks one span and lets each draw it:
[No Exceptions](design-laws.qmd).

Moving a pile costs the origin of the measure axis. Only a flat plot has an origin
to spare, so the other spaces refuse it:

```{r}
#| error: true
data(listening) + area * stack(baseline = "wiggle") + x(week) + y(plays) +
  color(genre) + polar()
```

In `polar()` the measure is an angle or a radius. A radius of zero is the center of
the circle, so moving a pile off it asks for a negative radius, which is not a
place on the plot.

### A pile has one direction

`stack` has a third condition. The first two are about the sentence. This one is
about the numbers.

A pile puts its groups end to end. Every member must therefore point the same way.

Negative values are allowed. A pile of them grows *downward* from zero, exactly as
a positive pile grows up:

```{r}
drawdown <- data.frame(
  quarter = rep(c("Q1", "Q2", "Q3"), 2),
  cost    = c(-5, -4, -6, -3, -2, -3),
  kind    = rep(c("refunds", "chargebacks"), each = 3))

data(drawdown) + bar * stack + x(quarter) + y(cost) + color(kind) +
  y_label("Cost") + title("A pile that grows down")
```

A pile whose members disagree has no correct drawing. Suppose one group is 5 and
another is -3 at the same position. The negative band would begin where the first
band ended, at 5, and reach back down to 2. That region already belongs to the first
group. A reader would see a solid block of length 3 where the number is negative. So
GOG refuses:

```{r}
#| error: true
mixed <- data.frame(
  quarter = rep(c("Q1", "Q2", "Q3"), 2),
  amount  = c(5, 5, 5, 2, -3, 2),
  kind    = rep(c("sales", "returns"), each = 3))

data(mixed) + bar * stack + x(quarter) + y(amount) + color(kind)
```

The message names the position, both groups, and both numbers. A message that said
only "some values are negative" would make you search the whole table for a pile you
cannot see on the page.

There are two ways out, and you choose by what you want the plot to show. If the
sign matters, use `dodge`. Each bar then keeps its own baseline, so a negative bar
hangs below the axis where a reader can see that it is negative. If the total
matters, take the absolute value in the host first. You are then stacking
magnitudes, and the sentence says so.

`stack(share = TRUE)` fails for the same reason, and the failure is easier to see. A
share is a fraction of the pile's total. Here the total is 5 plus -3, which is 2, so
the two shares come out as 2.5 and -1.5. No rescaling fixes that. The problem is in
the pile itself, not in how the pile is drawn.

Piles at *different* positions may still point different ways. Each pile is read on
its own. One quarter can be positive and the next negative in the same plot.

### The dot plot

A point has no length. Give it the same span every other mark gets and the only
thing it can spend it on is *itself*: one dot per unit, piled up from the foot. The
unit of a count is one row, so a pile of dots is a count you can count:

```{r}
data(gapminder_2007) + point * bin * stack + x(life) +
  x_label("Life expectancy (years)") + title("One dot per country")
```

That is a dot plot [@wilkinson1999dotplots], and it is the same sentence as the histogram at the top of this
chapter with one atom changed and one added. Compare the two: `bar * bin` draws each
tally as a length, and you read the shape off the outline; `point * bin * stack`
draws each tally as that many dots, and every country in the table is on the page.
Count the tallest pile and you get its bin's count exactly, which no bar can offer.

Nothing here is a color split, and none is needed. `bar` and `area` pile *groups*,
so they ask for one; a point piles *rows*, which every table already has. What it
asks for instead is a transform that counts them, because a pile of 3.7 dots means
nothing:

```{r}
#| error: true
data(gapminder_2007) + point * mean * stack + x(continent) + y(life)
```

The refusal names the two sentences that do count: `bin` for a continuous axis, and
`count` for a categorical one. The second draws the bar chart's abstraction with the
unit left visible, one dot per row:

```{r}
data(gapminder_2007) + point * count * stack + x(continent) +
  title("One dot per country, piled by continent")
```

The dots pack tightly here and loosely on a small table, and that is not a
setting: the gap between two rungs is one count unit on the y axis, so a
taller pile subdivides the panel more finely. A pile of ten dots in a tall
panel is airy. `style(size = )` will fatten the dots, but making them *touch*
means fixing the panel's height to the tallest pile. That is the aspect-ratio
control GOG does not have yet, the same one [polar](polar.qmd)'s circles and
the hexagonal [mesh](marks/zone.qmd) want. The count axis is real either way,
which is the trade taken deliberately: the alternative is to draw the dots
touching and leave the axis with no meaning.

### How many rows is too many

Push the same sentence far enough and the dots stop being dots. Every year of
gapminder rather than one is 1704 rows, and the tallest pile is 314 of them:

```{r}
#| message: true
data(gm_all) + point * bin * stack + x(life) +
  x_label("Life expectancy (years)") + title("Every year, every country: 1704 dots")
```

GOG draws it, and says what happened. The piles have become solid sticks: at 314
rungs the gap between two of them is under a pixel, so the dots overlap and nothing
can be counted, which was this plot's entire claim. What you are looking at is a
histogram drawn dot by dot, and an expensive one, since it ships 1704 circles where
`bar * bin` ships fourteen rectangles.

The warning is not a row limit, and that matters. It compares one rung's gap
against one dot's diameter, both of which GOG knows once the panel is laid
out. So it fires exactly when the picture degrades and stays quiet when it
does not. Shrink the dots with `style(size = )` and the same table goes
silent; widen the piles and a smaller one speaks up. A number like "more than
500 rows" would be wrong at some panel size or dot size, and there is no need
to guess one when the real condition is available.

Which also means the warning is a prompt, not a verdict. Two atoms usually answer it
without giving up a single dot, and the first is the one people forget: **more bins**
spread the same rows over more piles, so every pile gets shorter without anything
getting smaller. Then the dots themselves:

```{r}
data(gm_all) + point * bin(40) * stack + x(life) + style(size = 1.5) +
  x_label("Life expectancy (years)") + title("The same 1704 rows, cut finer and drawn smaller")
```

The columns are made of visible dots again, and every one of the 1704 is still on the
page. [Faceting](faceting.qmd) is the third lever, since fewer rows per panel means
shorter piles in each.

What none of them can beat is the arithmetic. Counting needs at least one
dot-diameter of axis per row, so the tallest honest pile is about the panel's
height divided by a dot: a few hundred with dots you can see, a few dozen if
you mean to count them. Past that a summary is not second best but the
better answer, and the reason is the one that makes the dot plot worth having
in the first place. A bin width or a bandwidth is a guess about structure when
there are thirty rows. It is a reliable description when there are ten
thousand. The plot runs out exactly where the summaries stop needing an
apology.

### One column, split into its parts

Every stacked bar so far has stood at a position, one pile per category on the
x-axis. Take the position away and the split still has something to divide: a
single column, showing how one total breaks up.

```{r}
data(winds) + bar * count * stack + color(direction)
```

There is no `x()` in that sentence, and none is needed. A bar normally requires a
position because it has to stand somewhere, but here the split *is* the
segmentation and there is one slot for it to divide, so the requirement lifts.
That is the only place in the grammar where a required position may be dropped,
and it is why it is worth knowing: bend this same plot into a circle and it is a
pie, which is the subject of [Polar](polar.qmd).

Without a split there is nothing to divide, so the position is required again and
the refusal says so. The statistic has to be one that means something with
nothing to group by, too: `count` and `sum` reduce a set of rows to one number,
while `bin` and `density` describe how values spread *along* an axis, and there is
no axis here.

## `jitter`: spread sideways, along the category

`dodge` and `stack` resolve the overlap of *grouped* marks. The third and last
collision modifier resolves a different overlap: not groups colliding, but a
mark's *own* points landing on top of each other. A [strip
plot](marks/point.qmd), every country as a dot above its continent, stacks its
points into a single vertical line, and heavy overlap hides how many there are:

```{r}
data(gapminder_2007) + point + x(continent) + y(life) + style(opacity = 0.5) +
  y_label("Life expectancy") + title("Strip plot: the points pile onto one line")
```

`jitter` nudges each point a little to the side so the pile fans out and the shape
of each group shows:

```{r}
data(gapminder_2007) + point * jitter + x(continent) + y(life) + style(opacity = 0.5) +
  y_label("Life expectancy") + title("point * jitter: the same points, spread to show density")
```

Africa's long tail of low values and Europe's tight cluster near the top were both
there before; the spread is what lets you read them.

`jitter` moves the *category*, never the *value*. Look again: every dot sits at
exactly the same height as before; the spread is entirely sideways. This is the
rule, not a coincidence. The horizontal axis is a category, and a category has no
magnitude, so nudging a point along it changes nothing you could misread. The
vertical axis is a *measurement*, and moving a point off its measured value would be
a lie about the data, so `jitter` never touches it. (Spreading in both directions
is the other way to do this. It moves a point off its measured value, which is
why GOG spreads along one axis only.) On a horizontal strip, the category on `y`,
the spread is vertical instead; `jitter` always spreads the categorical axis and
leaves the measured one exact.

The spread is **deterministic**: it is computed from the data itself, not from a
random-number generator or the clock, so the same plot drawn twice is pixel-for-pixel
the same picture. A spec always means one plot.

### How much to spread: `jitter(amount)`

The default spread fills a fraction of each category's slot: wide enough to show
the pile, narrow enough to leave a gap between categories so no point strays into a
neighbor. When you want more or less, `jitter(amount)` scales it: a plain number
multiplying the default, so `jitter(0.5)` is half the spread and `jitter(2)` twice
it (a bare `jitter` is `jitter(1)`).

```{r}
data(gapminder_2007) + point * jitter(0.5) + x(continent) + y(life) +
  style(opacity = 0.5) + y_label("Life expectancy") +
  title("point * jitter(0.5): a tighter spread")
```

The knob has the same shape as `density`'s `adjust`, a dimensionless multiple
of an automatic value. It is there for the same reason `density` and `bin`
take a parameter while `dodge` does not. The jitter amount is a *legibility
choice* with no single right value, whereas `dodge`'s width is *determined* by
the group count. A
free parameter earns a knob; a settled one does not. (`jitter(0)` is legal and draws
the plain strip plot, un-spread: GOG never forbids the harmless.)

Two things set `jitter` apart from its siblings. It needs **no `color` split**:
`dodge` and `stack` separate groups, but `jitter` separates a mark's own coincident
points, so a plain `point * jitter` with nothing but `x` and `y` is the whole
expression. And it needs a **categorical position axis** to spread within: ask for
`jitter` when both `x` and `y` are continuous and GOG refuses, because there is no
band to spread across and no axis it could move without misplacing a value. The
right tool for overplotting on two continuous axes is `style(opacity = )`, which
reveals density without moving any point.

Which marks `jitter` serves is decided by what a mark *is*: the rule that
divides the whole trio. `jitter` spreads a scatter of individual points, so it is
defined for `point` alone. A `bar`, `box` or `interval` has a width, and its groups
are set apart by `dodge`; a `line` or `area` is one connected shape, not a cloud of
points. Ask for `jitter` on any of those and GOG refuses, naming the offset that
fits.

The three offsets divide the marks by geometry **and by axis**: **`dodge`**
subdivides a width, **`stack`** accumulates along the *measure* axis, **`jitter`**
spreads along the *categorical* one. Every mark has an honest answer to overlap, and
`point` has two, because it is the one mark that can collide on either axis: piled
into a countable stack of dots along a measure, nudged apart within a slot along a
category. The offset follows the axis, which is why neither reading needs a third
name.
