# Functions around the grammar

```{r}
#| include: false
source("R/setup.R")
```

Every word in the kernel belongs to a sentence. `point` is a mark, `x` is a
channel, `bin` is a transform, and none of them means anything on its own. They
are words, and a word needs a sentence.

The functions in this appendix are not words. They take a plot you have already
written and act on it, or they hand you a table to write a sentence about. None of
them joins a sentence with `+` or `*`, and none of them is part of the grammar.

The name alone will not tell you which is which. `x_label` is a word of the
grammar and joins two words with an underscore, exactly as `render_svg` does.
What separates them is where each one can stand. A word of the grammar stands
inside a sentence, and these three never do. That is why none of them appears
in [The kernel card](kernel-card.qmd).

| | What it gives you |
|---|---|
| `render_svg(plot)` | The plot as text, in one string |
| `save_gif(plot, path)` | A played plot as a file that moves |
| `gog_table(name)` | One of this book's example tables |

The list is short. Everything on it follows the same rule: it acts on a finished
plot, or it hands you a table, and it has no place in a sentence.

## `render_svg()`

A plot is a specification until something draws it. `render_svg()` draws it and
gives you the result as ordinary text:

```{r}
picture <- render_svg(data(gapminder_2007) + point + x(gdp) + y(life))
cat(substr(picture, 1, 51), "\n")
cat(nchar(picture), "characters\n")
```

The first line of output shows where that text begins. The second line gives its
whole length. So the picture you have been looking at is also this: a run of
characters that starts `<svg` and states a width. After that it names every
point, every tick and every label.

That text is the whole plot. It is what every notebook, every web page and this
book itself receives. Writing it to a file with a `.svg` name gives you a
picture that any browser opens and a vector drawing program edits.

The function has the same name in all four languages, and takes the plot as its
one argument.

### What the file carries

An SVG file is text, so anyone who opens it can read it. If your table holds
something private, open the file and read it yourself before you publish. The
file carries the drawing and nothing else.

A plot draws the columns you map. Columns you never map are absent from the
file, and so are their names:

```{r}
private <- data.frame(
  patient = c("A. Kim", "B. Lee", "C. Park", "D. Choi"),
  email   = c("ak@ex.com", "bl@ex.com", "cp@ex.com", "dc@ex.com"),
  weight  = c(61.2, 74.8, 55.4, 68.0)
)

picture <- render_svg(data(private) + bar * bin + x(weight))
sapply(c("A. Kim", "email", "ex.com", "61.2"), grepl, x = picture, fixed = TRUE)
```

Every search returns `FALSE`. The sentence maps one column, so the drawing used
one column. Even the weights are gone, because `bin` counts rows into intervals
and the bars show the counts.

The missing weights come from the transform rather than from a general rule. A
mark that summarizes draws no individual row. A mark that does not summarize
draws one glyph per row, and each position is computed from one value:

```{r}
picture <- render_svg(data(private) + point + x(weight) + y(patient))
lengths(gregexpr("<circle", picture))
```

Four rows, four circles, and the four names printed along the axis. Every
plotting tool works this way, and it is what you asked for. A scatter plot draws
each value you gave it.

So the rule is not that the file is safe. It is narrower and more useful. What
you map is what you publish. A column you never map cannot be recovered from the
file, and a private column that stays out of the sentence stays private.

### Plots that carry the table

One case works differently, and you should know it before you publish.

Four kinds of plot are drawn twice: one that binds [`brush()`](selection.qmd),
one in the [cube](space.qmd), one on the [globe](globe.qmd), and a
[network](network.qmd) given an angle. Once when you write it, and once again in
the reader's browser, because the reader can move a selection or turn the view.
The second drawing needs the numbers, so the page carries the table beside the
picture.

It carries the whole table, including columns the sentence never maps. The
table that `show rows` (a button under the plot, explained in
[Selection](selection.qmd)) opens lists only the mapped columns, and the page
itself carries every column.

An SVG file never carries the table, even when its sentence names a brush.
`render_svg()` gives you the picture and stops.

One line of code keeps a private column out of the page. Bind the columns you
plot and drop the rest:

```{r}
shown <- private[, c("patient", "weight")]
data(shown) + point + x(weight) + y(patient) + brush(weight)
```

*"Given the shown table: points, x is weight, y is patient, and let the reader
select weight."*

That sentence names a brush, so this page carries a table. The table it carries
is `shown`, which has two columns. Bind `private` instead and `email` would be
in this page too.

### What "byte for byte" means

This book says in several places that the four languages draw the same picture
byte for byte. `render_svg()` is the function that claim is about, and the idea
behind it is simpler than the phrase sounds.

A plot looks like a picture on your screen. Underneath, it is the text you just
saw. Two plots are the same when that text is the same, byte for byte. Not
similar. Not close enough to look the same. The same.

That is how the four languages are kept in step. Nearly every sentence in this
book is written in R, in Python, in Julia and in JavaScript. Each one that draws
is turned into its text, and the four texts are compared byte by byte. One wrong
byte counts as a disagreement and has to be explained.

The reason for such a strict test is that a loose one would miss things. Nobody
looks at four pictures and decides whether they match. A person cannot see a bar
that is half a pixel too wide, or a color one step away from the right one.
This check sees both, and it compares the four texts the same way every time.

## `save_gif()`

A plot that binds `play()` moves in a browser, because the SVG carries its own
timing. A message, a slide in a talk and a post on a social site cannot play
that timing. [`save_gif()`](play.qmd#taking-the-sequence-with-you) writes the
same sequence as a file that they do play:

```{r}
ring <- file.path(tempdir(), "ring.gif")
save_gif(data(gm_all) + bar * mean + x(continent) + y(life) + polar() +
           play(year),
         ring, scale = 2)
cat(basename(ring), round(file.size(ring) / 1024), "KB\n")
```

Nothing has to be installed. `scale` multiplies the canvas, which is 800 by 600
pixels unless the plot's theme says otherwise.

Three of the four languages take `scale` as a named argument. JavaScript has no
named arguments, so it takes the same setting in the options object it uses
everywhere else:

| | |
|---|---|
| R | `save_gif(p, "ring.gif", scale = 2)` |
| Python | `save_gif(p, "ring.gif", scale=2)` |
| Julia | `save_gif(p, "ring.gif", scale = 2)` |
| JavaScript | `save_gif(p, "ring.gif", { scale: 2 })` |

A plot with no `play()` is refused rather than written as a file of one frame.
The [Play](play.qmd) chapter has that refusal, and the reason there is no video
version.

## `gog_table()`

Every shared table this book plots is available to you by name, so you can run
any example here without writing a data reader first:

```r
gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life)
```

That block does not run on this page, and the reason is worth one line. The call
reads the table over the network. A page that fetched on every build would leave
the book unable to render without a working connection.

[The book's data](book-data.qmd) is the chapter for this function. It lists
every table in the book and shows the call in each of the four languages.

`gog_table()` exists for reading this book rather than for your own work. When
the data is yours, bind it with `data()` and skip this entirely.
