# Functions around the grammar

```{r}
#| include: false
source("R/setup.R")
```

Every word in the kernel belongs to a sentence. `point` is a mark, `x` is a
channel, `bin` is a transform, and none of them means anything on its own. They
are words, and a word needs a sentence.

The functions in this appendix are not words. Two of them take a finished plot
and act on it. The third gives you a table that your sentence can name. None of
them can be written into a sentence with `+` or `*`, and none of them is part of
the grammar.

The name alone will not tell you whether a function is a word of the grammar.
`x_label` is a word of the grammar, and its name joins two parts with an
underscore, exactly as `render_svg` does. What separates them is where each one
is written. A word of the grammar is written inside a sentence. The three
functions below are always written outside one, which is why none of them
appears in [The kernel card](kernel-card.qmd).

| | What it gives you |
|---|---|
| `render_svg(plot)` | The plot as text, in one string |
| `save_gif(plot, path)` | A played plot as a GIF file |
| `gog_table(name)` | One of this book's shared tables |

Each function has a section below, in the same order as the rows above. The
first turns a finished plot into text.

## `render_svg()`

A plot is a specification until the engine draws it. Sometimes you want the
drawing itself: to save it as a file, or to put it in a web page of your own.
`render_svg()` draws the plot and gives you the drawing as ordinary text. It
takes the plot as its only argument, and it has the same name in all four
languages:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
picture <- render_svg(data(gapminder_2007) + point + x(gdp) + y(life))
cat(substr(picture, 1, 51), "\n")
cat(nchar(picture), "characters\n")
```

The first line of output shows where that text begins. The second line gives its
whole length. So every plot in this book is also a piece of text. It starts with
`<svg` and gives the width of the picture. After that it gives the position of
every point, every tick and every label.

That text is the whole picture. A notebook displays it, and so does this book.
Write it to a file whose name ends in `.svg`, and any web browser can open that
file as a picture. A vector drawing program, such as Inkscape, can edit it too,
because the file describes shapes rather than a grid of colored dots.

For a plot `p`, Python, Julia and JavaScript write that file with `save()`. R
already has a `save()` of its own, which stores R objects, so in R you write
the text yourself:

| | |
|---|---|
| R | `cat(render_svg(p), file = "life.svg")` |
| Python | `p.save("life.svg")` |
| Julia | `save(p, "life.svg")` |
| JavaScript | `save(p, "life.svg")` |

Each call writes the text that `render_svg()` returns, so the four files show
the same picture.

### What the file carries

An SVG file is text, so anyone who opens it in a text editor can read everything
in it. The file carries the drawing and nothing else, but the drawing itself can
show private values. If your table holds something private, open the file and
read it yourself before you publish.

A plot uses only the columns its sentence names. Columns the sentence never
names are absent from the file, and so are their names:

```{r}
patients <- data.frame(
  patient = c("A. Kim", "B. Lee", "C. Park", "D. Choi"),
  email   = c("ak@ex.com", "bl@ex.com", "cp@ex.com", "dc@ex.com"),
  weight  = c(61.2, 74.8, 55.4, 68.0)
)

picture <- render_svg(data(patients) + bar * bin + x(weight))
sapply(c("A. Kim", "email", "ex.com", "61.2"), grepl, x = picture, fixed = TRUE)
```

None of the four searches finds anything. The sentence names one column, so the
drawing uses one column. Even the weights are gone: `bin` cuts the weights into
intervals and counts the rows in each, so the bars show only the counts.

The weights are absent because of the transform, not because a plot never
carries values. A `point` with no transform draws one glyph per row, and each
glyph is placed from one value:

```{r}
picture <- render_svg(data(patients) + point + x(weight) + y(patient))
lengths(gregexpr("<circle", picture))
```

The four rows are drawn as four circles, and the four patient names are printed
along the axis. This is not special to gog: a scatterplot in any plotting tool
draws a point for every row you give it.

The weights are not written in the file as numbers. But each circle is placed by
its weight, so a reader who measures the drawing can recover every weight.

So the rule is not that the file is safe. It is narrower: what your sentence
names is what you publish. That includes a column named only in `facet()`,
whose values are printed as panel labels. A column the sentence never names
cannot be recovered from the file, so it stays private.

### Plots that carry the table

A saved SVG file carries only the drawing. A web page can carry more, and that
includes a page that a notebook or Quarto makes for you. You should know what it
carries before you publish it.

Some plots are drawn twice. The engine makes the first drawing. The reader's
browser makes the second, because the reader can move a selection or turn the
view. Four kinds of plot do this: one that binds [`brush()`](selection.qmd), one
in the [cube](space.qmd), one on the [globe](globe.qmd), and a
[network](network.qmd) given an angle. The second drawing needs the numbers, so
the web page also holds a copy of the table. That copy is in the page's code:
the screen does not show it, but anyone who opens the page's code can read it.

The copy holds every row of the table, but only the columns the sentence names.
As in a saved SVG file, a column the sentence never names is not in the page.
The difference is in the values: the page holds each row's exact numbers, and
the file holds only a drawing of them. Here is `patients` again, with a brush:

```{r}
data(patients) + point + x(weight) + y(patient) + brush(weight)
```

*"Given the patients: points, x is weight, y is patient, and let the reader select
weight."*

That sentence names a brush, so the web version of this page carries a copy of
the table. The copy has two columns, `weight` and `patient`, because the
sentence names those two. The `email` column is not in that copy at all. In the
web version, the plot has a `show rows` button under it, which
[Selection](selection.qmd) explains. The table that button opens lists the same
two columns.

An SVG file never carries the table, even when its sentence names a brush.
`render_svg()` returns only the picture.

### What "byte for byte" means

This book says in several places that the four languages draw the same picture
byte for byte. That claim is about the text that `render_svg()` returns.

A plot looks like a picture on your screen, and that picture is drawn from this
text. A byte is the unit a computer stores text in: one English letter, digit or
space takes one byte. Two plots are the same when their text is the same, byte
for byte. Text that is only similar does not count, even when the two pictures
look the same.

That is how gog checks that the four languages draw the same picture. Nearly
every sentence in this book is written in R, in Python, in Julia and in
JavaScript. Each one that draws is turned into its text, and the text from
Python, Julia and JavaScript is compared with R's, byte for byte. One wrong byte
counts as a disagreement, and one disagreement makes the whole check fail.

A test that allowed small differences would accept pictures that are not the
same. The pictures are not compared by eye, because a person cannot see a bar
that is half a pixel too wide, or a color that is slightly wrong. Comparing the
text finds both differences, and it works the same way every time.

## `save_gif()`

A plot that binds `play()` is a sequence of pictures rather than one. It moves
in a browser, because the SVG carries its own timing. A chat message, a slide in
a talk and a post on a social site cannot show that movement.
[`save_gif()`](play.qmd#taking-the-sequence-with-you) writes the same sequence
as a file that those places do play:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
ring <- file.path(tempdir(), "ring.gif")
save_gif(data(gm_all) + bar * mean + x(continent) + y(life) + polar() +
           play(year),
         ring, scale = 2)
cat(basename(ring), round(file.size(ring) / 1024), "KB\n")
```

The output names the file that was written and its size. The engine writes the
GIF itself, so no other program has to be installed.

`scale` multiplies the width and the height of the picture. A plot is 800 by 600
pixels unless its theme says otherwise, so `scale = 2` writes a GIF of 1600 by
1200 pixels.

Three of the four languages take `scale` as a named argument. JavaScript has no
named arguments, so it takes `scale` in an object at the end of the call:

| | |
|---|---|
| R | `save_gif(p, "ring.gif", scale = 2)` |
| Python | `save_gif(p, "ring.gif", scale=2)` |
| Julia | `save_gif(p, "ring.gif", scale = 2)` |
| JavaScript | `save_gif(p, "ring.gif", { scale: 2 })` |

A plot with no `play()` is refused rather than written as a file of one frame.
The [Play](play.qmd) chapter shows that refusal. It also explains why gog writes
no video.

## `gog_table()`

Every example in this book starts from a table. The examples draw from a set of
shared tables, such as `gapminder_2007`, and each one is published with the book
as a CSV file. `gog_table()` reads one by its name, so you can run any example
without finding and saving the file yourself. Unlike `render_svg()` and
`save_gif()`, this function comes before the plot rather than after it:

```r
gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life)
```

No plot appears under that block, because the book does not run it. The call
reads the table over the network, and running it here would mean the book could
be built only with a network connection.

`gog_table()` is there so that you can run this book's examples, not for your
own work. When the data is yours, bind it with `data()` and skip `gog_table()`
entirely.

`gog_table()` takes a second argument, `text`. A CSV file records each value but
never what kind of value it is. A column of `01`, `02`, `03` looks like numbers,
so it is read as 1, 2, 3, and the leading zeros are lost. To keep such a column
as text, list it in `text`.
[Labels that look like numbers](book-data.qmd#labels-that-look-like-numbers)
names the three columns in the book's tables that need it.

[The book's data](book-data.qmd) explains this function in full. It lists every
shared table in the book and shows the call in each of the four languages.
