# Functions around the grammar

```{r}
#| include: false
source("R/setup.R")
```

Every word in the kernel belongs to a sentence. `point` is a mark, `x` is a
channel, `bin` is a transform, and none of them means anything on its own. They
are words, and a word needs a sentence.

The functions in this appendix are not words. They take a plot you have already
written and do something with it, or they give you a table to write one about.
None of them can appear inside a specification, and none of them is part of the
grammar.

You can tell the two apart by looking at the name. A word of the grammar is one
ordinary English word: `bar`, `color`, `smooth`, `play`. A function around the
grammar is two words joined by an underscore: `render_svg`, `save_gif`,
`gog_table`. A separate rule exempts `x`, `y` and `z` as the only single-letter
names, which is why these three are not listed in
[The kernel card](kernel-card.qmd).

| | What it gives you |
|---|---|
| `render_svg(plot)` | The plot as text, in one string |
| `save_gif(plot, path)` | A played plot as a file that moves |
| `gog_table(name)` | One of this book's example tables |

The list is short today and it will grow. Anything added to it follows the same
rule: two words, an underscore between them, and no place in a sentence.

## `render_svg()`

A plot is a specification until something draws it. `render_svg()` draws it and
gives you the result as ordinary text:

```{r}
picture <- render_svg(data(gapminder_2007) + point + x(gdp) + y(life))
cat(substr(picture, 1, 51), "\n")
cat(nchar(picture), "characters\n")
```

The first line is where that text begins, and the second says how long the whole
of it is. So a plot you have been looking at as a picture is also this: a run of
characters that starts `<svg`, states a width, and goes on to name every point,
every tick and every label in turn.

That text is the whole plot. It is what every notebook, every web page and this
book itself receives. Writing it to a file with a `.svg` name gives you a picture
that any browser opens and any drawing program edits.

The call is spelled the same way in all four languages.

### What the file carries

An SVG file is text, so anyone who opens it can read it. Check that before you
publish, if your table holds something private. The file carries the drawing and
nothing else.

A plot draws the columns you map. Columns you never map are absent from the file,
and so are their names:

```{r}
private <- data.frame(
  patient = c("A. Kim", "B. Lee", "C. Park", "D. Choi"),
  email   = c("ak@ex.com", "bl@ex.com", "cp@ex.com", "dc@ex.com"),
  weight  = c(61.2, 74.8, 55.4, 68.0)
)

picture <- render_svg(data(private) + bar * bin + x(weight))
sapply(c("A. Kim", "email", "ex.com", "61.2"), grepl, x = picture, fixed = TRUE)
```

Every search returns `FALSE`. The sentence maps one column, so the drawing used
one column. Even the weights are gone, because `bin` counts rows into intervals
and the bars show the counts.

The missing weights come from the transform rather than from a general rule. A
mark that summarizes draws no individual row. A mark that does not summarize
draws one glyph per row, and each position is computed from one value:

```{r}
picture <- render_svg(data(private) + point + x(weight) + y(patient))
lengths(gregexpr("<circle", picture))
```

Four rows, four circles, and the four names printed along the axis. Every
plotting tool works this way, and it is what you asked for. A scatter plot draws
each value you gave it.

So the rule is not that the file is safe. It is narrower and more useful. What
you map is what you publish. A column you never map cannot be recovered from the
file, and a private column that stays out of the sentence stays private.

### Plots that carry the table

One case works differently, and you should know it before you publish.

A plot that binds [`brush()`](selection.qmd) or draws in the [cube](space.qmd) is
drawn twice. Once here, and once again in the reader's browser, because they can
move a selection or turn the view. The second drawing needs the numbers, so the
file carries the table as well as the picture.

It carries the whole table, including columns the sentence never maps. The table
that `show rows` (a button under the plot, explained in
[Selection](selection.qmd)) opens lists only the mapped columns, and the file
itself carries every column.

A static plot never does this, even when its sentence names a brush.
`render_svg()` gives you the picture and stops.

One line of code prevents it. Bind the columns you plot and drop the rest:

```{r}
shown <- private[, c("patient", "weight")]
data(shown) + point + x(weight) + y(patient) + brush(weight)
```

That sentence names a brush, so this page carries a table. The table it carries
is `shown`, which has two columns. Bind `private` instead and `email` would be in
this page too.

### What "byte for byte" means

This book says in several places that the four languages draw the same picture
byte for byte. `render_svg()` is the function that claim is about, and the idea
behind it is simpler than the phrase sounds.

A plot looks like a picture on your screen. Underneath, it is the text you just
saw. Two plots are the same when that text is the same, character for character.
Not similar. Not close enough to look the same. The same.

That is how the four languages are held together. Every sentence in this book is
written in R, in Python, in Julia and in JavaScript. Each one is turned into its
text, and the four are compared character by character. One character out of
place counts as a disagreement and has to be explained.

The reason for such a strict test is that a loose one would miss things. Nobody
looks at four pictures and decides whether they match. A person cannot see a bar
that is half a pixel too wide, or a color that is one shade off. This check can,
and it compares them the same way every time.

## `save_gif()`

A plot that binds `play()` moves in a browser, because the SVG carries its own
timing. A message, a slide and a post cannot play that timing.
[`save_gif()`](play.qmd#taking-the-sequence-with-you) writes the same sequence as
a file that they do play:

```{r}
ring <- file.path(tempdir(), "ring.gif")
save_gif(data(gm_all) + bar * mean + x(continent) + y(life) + polar() +
           play(year),
         ring, scale = 2)
cat(basename(ring), round(file.size(ring) / 1024), "KB\n")
```

Nothing has to be installed. `scale` multiplies the canvas, which is 800 by 600
pixels unless the plot's theme says otherwise.

Three of the four languages take `scale` as a named argument. JavaScript has no
named arguments, so it takes the same setting in the options object it uses
everywhere else:

| | |
|---|---|
| R | `save_gif(p, "ring.gif", scale = 2)` |
| Python | `save_gif(p, "ring.gif", scale=2)` |
| Julia | `save_gif(p, "ring.gif", scale = 2)` |
| JavaScript | `save_gif(p, "ring.gif", { scale: 2 })` |

A plot with no `play()` is refused rather than written as a file of one frame.
The [Play](play.qmd) chapter has that refusal, and the reason there is no video
version.

## `gog_table()`

Every table this book plots is available to you by name, so you can run any
example here without writing a data reader first:

```r
gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life)
```

That block does not run on this page, and the reason is worth one line. The call
reads the table over the network, and a page that fetched on every build would
make the book depend on a working connection to describe itself.

[The book's data](book-data.qmd) is the chapter for this one. It lists every
table in the book and shows the call in each of the four languages.

`gog_table()` exists for reading this book rather than for your own work. When
the data is yours, bind it with `data()` and skip this entirely.
