# Functions around the grammar

```{r}
#| include: false
source("R/setup.R")
```

Every word in the kernel belongs to a sentence. `point` is a mark, `x` is a
channel, `bin` is a transform, and none of them means anything on its own. They
are words, and a word needs a sentence.

The functions in this appendix are not words. They take a plot you have already
written and act on it, or they hand you a table to write a sentence about. None
of them can be written into a sentence with `+` or `*`, and none of them is part
of the grammar.

The name alone will not tell you whether a function is a word of the grammar.
`x_label` is a word and joins two words with an underscore, exactly as
`render_svg` does. What separates them is where each one can stand. A word of
the grammar stands inside a sentence. The three functions below never do, which
is why none of them appears in [The kernel card](kernel-card.qmd).

| | What it gives you |
|---|---|
| `render_svg(plot)` | The plot as text, in one string |
| `save_gif(plot, path)` | A played plot as a GIF file |
| `gog_table(name)` | One of this book's example tables |

The list is short, and each function has a section below, in the order of the
table. The first turns a finished plot into text.

## `render_svg()`

A plot is a specification until something draws it. `render_svg()` takes the
plot as its one argument, and it has the same name in all four languages. It
draws the plot and gives you the result as ordinary text:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
picture <- render_svg(data(gapminder_2007) + point + x(gdp) + y(life))
cat(substr(picture, 1, 51), "\n")
cat(nchar(picture), "characters\n")
```

The first line of output shows where that text begins. The second line gives its
whole length. So the picture you have been looking at is also this: a run of
characters that starts `<svg` and states a width. After that it names every
point, every tick and every label.

That text is the whole plot. It is what every notebook, every web page and this
book itself shows you. Writing it to a file with a `.svg` name gives you a
picture that any browser opens and a vector drawing program edits.

### What the file carries

An SVG file is text, so anyone who opens it can read it. If your table holds
something private, open the file and read it yourself before you publish. The
file carries the drawing and nothing else.

A plot draws the columns you map. Columns you never map are absent from the
file, and so are their names:

```{r}
private <- data.frame(
  patient = c("A. Kim", "B. Lee", "C. Park", "D. Choi"),
  email   = c("ak@ex.com", "bl@ex.com", "cp@ex.com", "dc@ex.com"),
  weight  = c(61.2, 74.8, 55.4, 68.0)
)

picture <- render_svg(data(private) + bar * bin + x(weight))
sapply(c("A. Kim", "email", "ex.com", "61.2"), grepl, x = picture, fixed = TRUE)
```

Every search returns `FALSE`. The sentence maps one column, so the drawing used
one column. Even the weights are gone, because `bin` counts rows into intervals
and the bars show the counts.

The weights are absent because of the transform, not because a plot never
carries values. A mark that summarizes draws no individual row. A mark that does
not summarize draws one glyph per row, and each glyph is placed from one value:

```{r}
picture <- render_svg(data(private) + point + x(weight) + y(patient))
lengths(gregexpr("<circle", picture))
```

The four rows drew four circles, and the four patient names are printed along
the axis. Every plotting tool works this way, and it is what you asked for. A
scatter plot draws each value you gave it.

The number itself is not written in the file. Each circle sits where that number
puts it, so a reader who measures the drawing can recover the value.

So the rule is not that the file is safe. The real rule is narrower and more
useful. What you map is what you publish. A column you never map cannot be
recovered from the file, and a private column the sentence never names stays
private.

### Plots that carry the table

One case works differently, and you should know it before you publish.

Some plots are drawn twice. You make the first drawing. The reader's browser
makes the second, because the reader can move a selection or turn the view. Four
kinds of plot do this: one that binds [`brush()`](selection.qmd), one in the
[cube](space.qmd), one on the [globe](globe.qmd), and a
[network](network.qmd) given an angle. The second drawing needs the numbers, so
the page carries the table beside the picture.

It carries the whole table, including columns the sentence never maps. `show
rows` is a button under the plot, explained in [Selection](selection.qmd). The
table it opens lists only the mapped columns. The page itself carries every
column.

The remedy is to bind a table that holds only the columns you plot. Here is
`private` again with the email column left out:

```{r}
shown <- data.frame(
  patient = c("A. Kim", "B. Lee", "C. Park", "D. Choi"),
  weight  = c(61.2, 74.8, 55.4, 68.0)
)

data(shown) + point + x(weight) + y(patient) + brush(weight)
```

*"Given the shown table: points, x is weight, y is patient, and let the reader
select weight."*

That sentence names a brush, so this page carries a table. The table it carries
is `shown`, which has two columns. If you bound `private` instead, `email` would
be on this page too. Each language has its own way to drop a column from a table
before you bind it, and that work belongs to the language rather than to the
grammar.

An SVG file never carries the table, even when its sentence names a brush.
`render_svg()` gives you the picture and stops.

### What "byte for byte" means

This book says in several places that the four languages draw the same picture
byte for byte. That claim is about `render_svg()`, and the idea is simple.

A plot looks like a picture on your screen. Underneath, it is the text that
`render_svg()` returns. Two plots are the same when that text is the same, byte
for byte. Not similar. Not close enough to look the same. The same.

That is how the four languages are kept the same. Nearly every sentence in this
book is written in R, in Python, in Julia and in JavaScript. Each one that draws
is turned into its text, and the four texts are compared one byte at a time. One
wrong byte counts as a disagreement and has to be explained.

A test that allowed small differences would accept pictures that are not the
same. Nobody looks at four pictures and decides whether they match. A person
cannot see a bar that is half a pixel too wide, or a color that is slightly
wrong. Comparing the text finds both differences, and it works the same way
every time.

## `save_gif()`

A plot that binds `play()` is a sequence of pictures rather than one. It moves
in a browser, because the SVG carries its own timing. A message, a slide in a
talk and a post on a social site cannot play that timing.
[`save_gif()`](play.qmd#taking-the-sequence-with-you) writes the same sequence
as a file that those places do play:

```{r}
#| echo: false
peek(gm_all)
```

```{r}
ring <- file.path(tempdir(), "ring.gif")
save_gif(data(gm_all) + bar * mean + x(continent) + y(life) + polar() +
           play(year),
         ring, scale = 2)
cat(basename(ring), round(file.size(ring) / 1024), "KB\n")
```

The output names the file that was written and its size. Nothing has to be
installed to write it.

`scale` multiplies the canvas. A plot is 800 by 600 pixels unless its theme says
otherwise, so `scale = 2` doubles both numbers.

Three of the four languages take `scale` as a named argument. JavaScript has no
named arguments, so it takes the same setting in the options object it uses
everywhere else:

| | |
|---|---|
| R | `save_gif(p, "ring.gif", scale = 2)` |
| Python | `save_gif(p, "ring.gif", scale=2)` |
| Julia | `save_gif(p, "ring.gif", scale = 2)` |
| JavaScript | `save_gif(p, "ring.gif", { scale: 2 })` |

A plot with no `play()` is refused rather than written as a file of one frame.
The [Play](play.qmd) chapter shows that refusal. It also explains why gog writes
no video.

## `gog_table()`

`render_svg()` and `save_gif()` act on a finished plot. This one comes before
the plot. Every shared table this book plots is available to you by name, so you
can run any example here without downloading a file first:

```r
gapminder_2007 <- gog_table("gapminder_2007")
data(gapminder_2007) + point + x(gdp) + y(life)
```

That block does not run on this page, and the reason is worth one line. The call
reads the table over the network. If the book did that on every build, it could
not render without a connection.

`gog_table()` exists for reading this book rather than for your own work. When
the data is yours, bind it with `data()` and skip `gog_table()` entirely.

[The book's data](book-data.qmd) is the chapter for this function. It lists
every table in the book and shows the call in each of the four languages.
