# Writing practice

```{r}
#| include: false
source("R/setup.R")
```

How does a question about your data turn into a specification? Reading was
recognition; writing is recall. Psychologists use those two words for two
kinds of remembering. Recognition starts from something in front of you, and
asks only whether you have seen it before. That is what the last chapter did:
it gave you finished sentences and asked what they say. Recall starts from
nothing in front of you, and asks you to produce the thing itself. It is the
harder of the two, and it is what writing a specification asks of you.

This chapter builds three sentences from three questions, the way you will
actually work: start with the smallest legal sentence, look, refine. Then it
introduces the other half of writing: what happens when a sentence is wrong,
and why gog's refusals are the fastest teacher in the book.

## Which continents live longest?

Start by looking at everything, the smallest sentence that shows the data:

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Points, x is continent, y is life."* Informative, but the question asked
for a comparison of *typical* values, not a cloud. Derive the summary,
swap the mark's derivation, keep everything else:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

Comparison wants order. `order()` sorts a categorical axis by a value,
something no bare column can say:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  order(life, desc = TRUE)
```

Three steps, and each step *added* a word or swapped one: no rewriting,
no second syntax. That is what compositionality buys you in practice.

## How do petal lengths differ by species?

The one-variable question first, then the split:

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

And if the overlap bothers you, the outline form is one mark away; the
staircase silhouette holds each bin's height as an unfilled stroke, so the
three species never paint over each other:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species)
```

The transform never changed. `bin` cut the same intervals all three times;
only the geometry drawing them moved. A derivation you learn once works
under every mark that can carry it.

## Where is the world's wealth, and its health?

The full sentence from the first chapter, written in one breath this time:

```{r}
data(gapminder_2007) + point +
  x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color
by continent, size by population, titled."* Six atoms past the minimum,
and still one sentence with no new grammar in it.

## When the engine says no

You will write wrong sentences. Everyone does. gog's design promise is
that a wrong sentence is refused *with a direction*, never absorbed into a
silently wrong picture. Meet the refusals you are most likely to see this
week.

Forgetting to bind the table. The sentence must start with the data:

```{r}
#| error: true
gapminder_2007 + point + x(gdp) + y(life)
```

Quoting a column name, a habit from other packages. The refusal says
which fix it wants:

```{r}
#| error: true
data(gapminder_2007) + point + x("gdp") + y("life")
```

Mapping a channel to a column that cannot answer its question. `size`
encodes magnitude, and a category has none:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

Asking a transform to read a column you never named:

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent)
```

Binning a category. `bin` cuts a continuous axis, and the refusal points
at the atom that does count categories:

```{r}
#| error: true
data(gapminder_2007) + bar * bin + x(continent)
```

Each message names the rule and the way forward. Nothing rendered behind
your back; nothing was dropped to make the picture come out. When you hit
a refusal outside this list, read it the same way: it is the grammar
telling you which law your sentence broke, and the fix is in
the message.

Every refusal closes with a second offer, to set `GOG_STRICT=0` and draw
the plot anyway. That is an environment variable, and what it does to each
kind of refusal (along with the one line that sets it in R, Python, Julia
and JavaScript) is [Drawing it anyway](design-laws.qmd#drawing-it-anyway).
For now, take the first offer: at this stage the refusals are the lesson.

## The morning checkpoint

If you started this part in one sitting, check yourself against its promise:

- Read any sentence in this book aloud before looking at its plot.
- Given a question (*distribution? comparison? relationship? change over
  time?*), name the mark and transform that answer it.
- Predict whether a sentence will render or refuse, and roughly what the
  refusal will say.

If that mostly works, the morning did its job. Part II gives every letter
its full chapter; or jump straight to [the cookbook](cookbook/one-variable.qmd)
and start answering your own questions, following the links back whenever
a recipe uses a word you want more of.
