# Writing practice

```{r}
#| include: false
source("R/setup.R")
```

How does a question about your data turn into a specification? Reading was
recognition; writing is recall. Psychologists use those two words for two kinds
of remembering. Recognition starts from something in front of you, and asks only
whether you have seen it before. That is what [Reading practice](reading.qmd)
did: it gave you finished sentences and asked what they say. Recall starts from
nothing in front of you: you produce the answer yourself. It is the harder of
the two, and it is what writing a specification asks of you.

From here on, this chapter asks you to recall. Three worked examples come first.
The first two show the method you will use: start with the minimum sentence,
look, refine. The third asks you to write the whole sentence at once. Five
questions follow with nothing in front of you, each answered by a sentence you
met in Reading practice: write yours before you read on, then check it against
the printed one. Last comes the other half of writing: what happens when a
sentence is wrong, and what a refusal tells you to do next.

## Which continents live longest?

This is the first of the three worked examples. Before reading on, try to
write the sentence that answers the question above: which mark, which
transform, which positions? Then build it one step at a time. Start with the
minimum sentence, the smallest one that shows the data:

```{r}
#| echo: false
peek(gapminder_2007)
```

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Given gapminder 2007: points, x is continent, y is life."* The plot is
informative, but the question asked for a comparison of *typical* values, and
this one shows every country. Swap the mark and give it a derivation; everything
else stays:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life."* Now
there is one bar per continent, as tall as the mean life expectancy of its
countries.

Bars are easiest to compare when they are sorted, with the largest first. A
categorical column can carry a fixed order, declared when the table is built.
The order you want here depends on the means the plot has just computed, so the
table cannot carry it. The sentence has to ask for it: `order()` sorts the axis
by the column you name, and `desc = TRUE` puts the largest first:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  order(life, desc = TRUE)
```

*"Given gapminder 2007: bars derived by mean, x is continent, y is life, ordered
by life, largest first."*

That took two changes, and neither rewrote the sentence. Each added or swapped
words and left the rest in place, which is what composition gives you:
every word you keep still means the same thing.

## How do petal lengths differ by species?

Reading practice drew the first two plots below. Write the two sentences from
memory: the one-variable question first, then the same question split by
species.

```{r}
#| echo: false
peek(iris_flowers)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

*"Given the iris flowers: bars derived by bin, x is petal length."* Now split
the same bars by species, with one word.

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: bars derived by bin, x is petal length, color by
species."*

If the overlapping bars are hard to compare, one mark changes that: `step`
draws each bin's height as an outline with no fill, so the three species never cover
each other:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species)
```

*"Given the iris flowers: step outlines derived by bin, x is petal length, color by
species."*

The transform never changed. `bin` cut the same intervals all three times; what
changed was the color split, and then the mark drawing them. A transform you
learn once works with every mark that can take it.

## Where is the world's wealth, and its health?

The first two questions were each about one measurement. This one is about two,
income and life expectancy, and they fill both positions. A reader also wants to
know how many people each dot stands for. [A first plot](first-plot.qmd) built
this sentence step by step. Write the whole sentence now, adding `size` for
population, then check:

```{r}
data(gapminder_2007) + point +
  x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color by
continent, size by population."* Three atoms were added to the minimum
sentence: `color`, `size` and `title`. The log scale is written inside `x`, so it
adds no word, and the sentence still uses no new grammar.

## Five questions, nothing in front of you

Each question below matches one plot from Reading practice. Write your
sentence first; the printed one is there to check against, not to read first.

Which countries won the most gold? Every country name has to stay readable,
so the bars should be horizontal.

```{r}
#| echo: false
peek(medals)
```

```{r}
data(medals) + bar + x(gold) + y(country)
```

*"Given the medals: bars, x is gold, y is country."* The two bindings
were swapped: `gold` on `x`, `country` on `y`. That makes the bars horizontal,
and no new word was needed.

How is life expectancy distributed, drawn as a smooth curve rather than
counted into bins?

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"Given gapminder 2007: a line derived by density, x is life."* A histogram
answers the same question; here `line` replaces `bar`, and `density` replaces
`bin`.

How did life expectancy move across the decades in the five Asian countries,
one line per country?

```{r}
#| echo: false
peek(gapminder_asia)
```

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*
`color` does two things here. It splits the rows into one line per country,
then gives each line its own color, exactly as it did in Reading practice.

Do richer countries also live longer inside each continent, one panel per
continent?

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split
into panel columns by continent."* The `|` operator and `facet` split the
plot into panels. All the panels share one pair of scales, so they can be
compared.

Do the three iris measurements separate the species when seen together?

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Given the iris flowers: points, x is sepal length, y is sepal width, z is
petal length, color by species."* The cube again, and you wrote it yourself
this time.

## When gog says no

Every sentence so far was correct. The other half of writing is what happens
when one is not.

You will write wrong sentences. Everyone does. gog promises that a wrong
sentence is refused *with a direction*, never drawn with the wrong part quietly
removed. Here are the refusals you are most likely to see
first.

Forgetting to bind the table. The sentence must start with the data:

```{r}
#| error: true
gapminder_2007 + point + x(gdp) + y(life)
```

The message above is R's. Each binding writes this advice with its own way of
naming a column.

Quoting a column name, a habit from other packages. The refusal names the
fix:

```{r}
#| error: true
data(gapminder_2007) + point + x("gdp") + y("life")
```

That message is R's too. Python and JavaScript name the fix `x(col.gdp)`, and
Julia names it `x(:gdp)`.

Mapping a channel to a column that cannot answer its question. `size`
encodes magnitude, and a category has none:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

Forgetting the column a summary reads. The refusal names the missing `y`,
because `mean` must be told which column to average:

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent)
```

Binning a category. `bin` cuts a continuous axis, and the refusal names
`count`, the transform that tallies rows per category:

```{r}
#| error: true
data(gapminder_2007) + bar * bin + x(continent)
```

Each message names the rule and the fix. No wrong plot was drawn, and no binding
was dropped to complete one. Read any refusal outside this list the same way.

The last three refusals end with a second option: set `GOG_STRICT=0` and draw
the plot anyway. The first two never reach the engine, so there is nothing to
draw. `GOG_STRICT` is an environment variable, and [Drawing it
anyway](design-laws.qmd#drawing-it-anyway) shows which refusals `GOG_STRICT`
affects, with the one line that sets it in each language. For now, make the
fix the message asks for: at this stage the refusals are the lesson.

## The morning checkpoint

That is the whole method: start with the minimum sentence, look, refine, and read
the refusal when one comes. If you have read this part from the start, check
yourself against what it promised:

- Read any sentence in this book aloud before looking at its plot.
- Given a question (*distribution? comparison? relationship? change over
  time?*), name the mark and transform that answer it.
- Predict whether a sentence will render or refuse, and roughly what the
  refusal will say.

If you can do most of that, the morning has reached its goal. You can read the
next part in order, or go directly to [the cookbook](cookbook/one-variable.qmd)
and answer your own questions, following a link back whenever a recipe uses a
word you want explained.

## What comes next

A morning teaches the shape of a sentence. It does not show everything one word
can do.
[Data](data.qmd) opens the letters with the table every sentence starts from:
the shape gog needs, and what gog does with the columns it finds.
