# Writing practice

```{r}
#| include: false
source("R/setup.R")
```

How does a question about your data turn into a specification? Reading was
recognition; writing is recall. Psychologists use those two words for two
kinds of remembering. Recognition starts from something in front of you, and
asks only whether you have seen it before. That is what the last chapter did:
it gave you finished sentences and asked what they say. Recall starts from
nothing in front of you: you produce the answer yourself. It is the
harder of the two, and it is what writing a specification asks of you.

This chapter works in the recall direction from here on. Three builds come
first, showing the method you will actually use: start with the minimum
sentence, look, refine. Five questions follow with nothing in front of you,
each answered by a sentence you met in the last chapter: write yours before
you scroll, then check it against the printed one. Last comes the other half
of writing: what happens when a sentence is wrong, and how quickly a refusal
teaches the rule.

## Which continents live longest?

Before reading on, try to write the sentence that will answer it: which mark,
which transform, which positions? Then build it the way this chapter always
will. Start with the
minimum sentence, the smallest one that shows the data:

```{r}
data(gapminder_2007) + point + x(continent) + y(life)
```

*"Given gapminder 2007: points, x is continent, y is life."* Informative, but
the question asked for a comparison of *typical* values, and this shows every
country. Swap the mark and give it a derivation; everything else stays:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life)
```

A comparison reads best in order. A categorical column can carry a fixed
order, declared where the data is made. The order wanted here depends on the
means this plot just computed, so only the sentence can ask for it: `order()`
sorts the axis by the column you name:

```{r}
data(gapminder_2007) + bar * mean + x(continent) + y(life) +
  order(life, desc = TRUE)
```

Three steps, and no step rewrote the sentence: words were added or swapped,
never restructured. That is what composition gives you: each step keeps every
word you already wrote.

## How do petal lengths differ by species?

The last chapter drew both of these. Write the two sentences from memory: the
one-variable question first, then the same question split by species.

```{r}
data(iris_flowers) + bar * bin + x(petal_length)
```

```{r}
data(iris_flowers) + bar * bin + x(petal_length) + color(species)
```

And if the overlap bothers you, one mark changes it: `step` draws each bin's
height as an unfilled line, so the three species never cover each other:

```{r}
data(iris_flowers) + step * bin + x(petal_length) + color(species)
```

The transform never changed. `bin` cut the same intervals all three times;
only the mark drawing them changed. A derivation you learn once works with
every mark that can take it.

## Where is the world's wealth, and its health?

Chapter one built this sentence step by step. Write it in one pass now,
adding `size` for population, then check:

```{r}
data(gapminder_2007) + point +
  x(gdp, scale = "log") + y(life) +
  color(continent) + size(population) +
  title("The health and wealth of nations, 2007")
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, color
by continent, size by population, titled."* Three atoms added to the minimum
sentence (`color`, `size` and `title`; the log scale rides inside `x` rather
than counting as a word), and still no new grammar in it.

## Five questions, nothing in front of you

Each question below matches one plot from the last chapter. Write your
sentence first; the printed one is there to check against, not to read first.

Which countries won the most gold? Every country name has to stay readable,
so lay the bars down.

```{r}
data(medals) + bar + x(gold) + y(country)
```

*"Given the medals: bars, x is gold, y is country."* One swap of the two
bindings lies the bars down, and no new word was needed.

How is life expectancy distributed, drawn as a smooth curve rather than
counted into bins?

```{r}
data(gapminder_2007) + line * density + x(life)
```

*"Given gapminder 2007: a line derived by density, x is life."* The
histogram's sibling: one atom swapped, same question.

How did life expectancy move across the decades in the five Asian countries,
one line per country?

```{r}
data(gapminder_asia) + line + x(year) + y(life) + color(country)
```

*"Given gapminder Asia: lines, x is year, y is life, color by country."*
`color` splits first and colors second, exactly as it did in the last
chapter.

Does the wealth and health relationship hold inside each continent, one
panel per continent?

```{r}
data(gapminder_2007) + point + x(gdp, scale = "log") + y(life) | facet(continent)
```

*"Given gapminder 2007: points, x is gdp on a log scale, y is life, split
into panel columns by continent."* The `|` and its facet, shared scales and
all.

Do the three iris measurements separate the species when seen together?

```{r}
data(iris_flowers) + point +
  x(sepal_length) + y(sepal_width) + z(petal_length) +
  color(species)
```

*"Given iris flowers: points, x is sepal length, y is sepal width, z is
petal length, color by species."* The cube again, and you wrote it yourself
this time.

## When the engine says no

You will write wrong sentences. Everyone does. gog's design promise is
that a wrong sentence is refused *with a direction*, never drawn with the
wrong part quietly left out. Here are the refusals you are most likely to see
first.

Forgetting to bind the table. The sentence must start with the data:

```{r}
#| error: true
gapminder_2007 + point + x(gdp) + y(life)
```

Quoting a column name, a habit from other packages. The refusal names the
fix:

```{r}
#| error: true
data(gapminder_2007) + point + x("gdp") + y("life")
```

Mapping a channel to a column that cannot answer its question. `size`
encodes magnitude, and a category has none:

```{r}
#| error: true
data(gapminder_2007) + point + x(gdp) + y(life) + size(continent)
```

Forgetting the column a summary reads. The refusal names the missing `y`
first, because the bar cannot stand without it:

```{r}
#| error: true
data(gapminder_2007) + bar * mean + x(continent)
```

Binning a category. `bin` cuts a continuous axis, and the refusal points
at the atom that does count categories:

```{r}
#| error: true
data(gapminder_2007) + bar * bin + x(continent)
```

Each message names the rule and the fix. Nothing was drawn without telling
you, and no binding was dropped to complete the plot. Read any refusal outside
this list the same way: it names the law the sentence broke, and it names the
fix.

The last three refusals close with a second offer: set `GOG_STRICT=0` and draw
the plot anyway. The first two never reach the engine, so there is nothing to
draw. `GOG_STRICT` is an environment variable, and [Drawing it
anyway](design-laws.qmd#drawing-it-anyway) shows what it does to each kind of
refusal, with the one line that sets it in each language. For now, make the
fix the message asks for: at this stage the refusals are the lesson.

## The morning checkpoint

If you started this part in one sitting, check yourself against its promise:

- Read any sentence in this book aloud before looking at its plot.
- Given a question (*distribution? comparison? relationship? change over
  time?*), name the mark and transform that answer it.
- Predict whether a sentence will render or refuse, and roughly what the
  refusal will say.

If that mostly works, the morning worked. Part II gives every letter its full
chapter. Or go straight to [the cookbook](cookbook/one-variable.qmd) and
answer your own questions, following a link back whenever a recipe uses a word
you want explained.
